### 2014-05-24 paint-contour-tree.tex
adding files

\documentclass[11pt]{article}
\usepackage[margin=2.54cm]{geometry}
\usepackage[dvipsnames,usenames]{color}
\usepackage{amsfonts,amsmath,amssymb,amsthm,mathtools}
\usepackage{paralist}
\usepackage{algorithmic,algorithm}
\usepackage{bm}
\usepackage{xspace}
\usepackage{centernot}
\usepackage{fancybox}
\usepackage{framed}

%\usepackage{mathabx}
\usepackage[pagebackref,letterpaper=true,colorlinks=true,pdfpagemode=none,urlcolor=blue,linkcolor=blue,citecolor=BrickRed,pdfstartview=FitH]{hyperref}
\usepackage{xspace,prettyref}
\usepackage{color}
\usepackage{graphics}
%\usepackage{MnSymbol}
%% For hyperlinks. Should always be the last package.
%\usepackage[colorlinks,urlcolor=blue,citecolor=blue,linkcolor=blue]{hyperref}
\let\pref=\prettyref
\newcommand{\mathcalavehyperref}[2]{\texorpdfstring{\hyperref[#1]{#2}}{#2}}
\newcommand{\comment}[1]{ {\color{BrickRed} \footnotesize[#1]}\marginpar{\footnotesize\textbf{\color{red} To Do!}}}
\newcommand{\ignore}[1]{}

\newtheorem{theorem}{Theorem}
\newtheorem{lemmata}[theorem]{Lemmata}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{claim}[theorem]{Claim}
\newtheorem{subclaim}{Subclaim}
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{corollary}[theorem]{Corollary}
\newtheorem{fact}[theorem]{Fact}
\newtheorem{conjecture}[theorem]{Conjecture}
\newtheorem{question}[theorem]{Question}
\newtheorem{example}[theorem]{Example}
\newtheorem{definition}[theorem]{Definition}
\theoremstyle{definition}
\newtheorem{remark}{Remark}
\newtheorem{observation}{Observation}

\newcommand{\EPSfigure}[5]{

        % #1 = File name and other arguments to \psfig macro
        % #2 = Caption Text
        % #3 = Positioning Letters: h, t|b|p
        % #4 = {*} causes double column figure, {} for single
        % #5 = Label to apply to figure

        \begin{figure#4}[#3]
                \centering
                \ \psfig{file=#1}
%               \label{#5}
%
                \ \caption{{\em #2}\label{#5}}\hfill\break

        \end{figure#4}
}

\newcommand{\innbd}{\Gamma^-}
\newcommand{\nbd}{\Gamma}
\newcommand{\outnbd}{\Gamma^+}

\newcommand{\bT}{{\bf T}}
\newcommand{\cS}{{\cal S}}
\newcommand{\cB}{{\cal B}}
\newcommand{\cC}{{\cal C}}
\newcommand{\cD}{{\cal D}}
\newcommand{\cE}{{\cal E}}
\newcommand{\cF}{{\cal F}}
\newcommand{\cG}{{\cal G}}
\newcommand{\cH}{{\cal H}}
\newcommand{\cL}{{\cal L}}
\newcommand{\cP}{{\cal P}}
\newcommand{\cV}{{\cal V}}

\newcommand{\LL}{\mathbb{L}}
\newcommand{\MM}{\mathbb{M}}
\newcommand{\NN}{\mathbb{N}}
\newcommand{\RR}{\mathbb{R}}
\newcommand{\SSS}{\mathbb{S}}

\newcommand{\bmu}{\overline{\mu}}
\newcommand{\poly}{\textrm{poly}}
\newcommand{\mymod}{\textrm{mod} \ }
\newcommand{\otilde}{\widetilde{O}}
%\newcommand{\qed}{\hfill $\Box$}
\newcommand{\eps}{\varepsilon}
\newcommand{\fu}{\varphi}
\newcommand{\restr}[1]{{|_{{\textstyle #1}}}}
\newcommand{\proj}{\mbox{\rm proj}}
\newcommand{\vol}{\mbox{\tt vol}\,}
\newcommand{\area}{\mbox{\tt area}\,}
\newcommand{\conv}{\mbox{\tt conv}\,}
\newcommand{\diam}{\hbox{\tt diam}\,}
\newcommand{\hdisc}{\mbox{\rm herdisc}}
\newcommand{\ldisc}{\mbox{\rm lindisc}}
\newcommand{\EX}{\hbox{\bf E}}
\newcommand{\prob}{{\rm Prob}}
\newcommand{\proofend}{{\medskip\medskip}}
%\newcommand{\proof}{{\noindent\bf Proof: }}
%\newcommand{\reals}{{\rm I\!\hspace{-0.025em} R}}
%\newcommand{\dist}{\hbox{dist}}
\newcommand{\defeq}{\mbox{\,$\stackrel{\rm def}{=}$\,}}
\newcommand{\boxalg}[1]
{\begin{center}\fbox{\parbox{\columnwidth}{\tt
\begin{tabbing}
\=mm\=mm\=mm\=mm\=mm\=mm\=mm\=mm\=mm\kill
#1
\end{tabbing} } } \end{center} }

%% HYPER-LINKED REFERENCES
\newcommand{\Sec}[1]{\hyperref[sec:#1]{\S\ref*{sec:#1}}} %section
\newcommand{\Eqn}[1]{\hyperref[eqn:#1]{(\ref*{eqn:#1})}} %equation
\newcommand{\Clm}[1]{\hyperref[clm:#1]{Claim~\ref*{clm:#1}}} %claim
\newcommand{\Fig}[1]{\hyperref[fig:#1]{Figure~\ref*{fig:#1}}} %figure
\newcommand{\Tab}[1]{\hyperref[tab:#1]{Table~\ref*{tab:#1}}} %table
\newcommand{\Thm}[1]{\hyperref[thm:#1]{Theorem~\ref*{thm:#1}}} %theorem
\newcommand{\Lem}[1]{\hyperref[lem:#1]{Lemma~\ref*{lem:#1}}} %lemma
\newcommand{\Prop}[1]{\hyperref[prop:#1]{Proposition~\ref*{prop:#1}}} %property
\newcommand{\Cor}[1]{\hyperref[cor:#1]{Corollary~\ref*{cor:#1}}} %corollary
\newcommand{\Def}[1]{\hyperref[def:#1]{Definition~\ref*{def:#1}}} %definition
\newcommand{\Alg}[1]{\hyperref[alg:#1]{Algorithm~\ref*{alg:#1}}} %algorithm
\newcommand{\Ex}[1]{\hyperref[ex:#1]{Example~\ref*{ex:#1}}} %example

%% Comments to ourselves
\newcommand{\Reminder}[1]{{\color{red}#1}}
\newcommand{\Sesh}[1]{\Reminder{Sesh interjects: #1}}

%%Ben Added
\DefineNamedColor{named}{RedViolet} {cmyk}{0.07,0.90,0,0.34}
\providecommand{\AlgorithmI}[1]{{\textcolor[named]{RedViolet}{\texttt{\bf{#1}}}}}
\providecommand{\Algorithm}[1]{{\AlgorithmI{#1}\index{algorithm!#1@{\AlgorithmI{#1}}}}}
\newcommand{\zip}{\Algorithm{ZIP}}
\newcommand{\link}{\Algorithm{LINK}}
\newcommand{\sizes}{\Algorithm{SIZES}}
\newcommand{\AMT}{\ensuremath{\mathsf{AMT}}\xspace}
\newcommand{\MT}{\ensuremath{\mathsf{MT}}\xspace}
\newcommand{\BCS}{\ensuremath{\mathsf{BCS}}\xspace}
\newcommand{\sub}{\mathsf{sub}}
\newcommand{\Merge}{\mathsf{Merge}}
\newcommand{\stack}{\mathsf{S}}
\newcommand{\AQ}{\mathsf{AQ}}

\newcommand{\cut}{{\tt cut}}
\newcommand{\cont}{\psi}
\newcommand{\lift}{{\tt lift}}
\newcommand{\paint}{{\tt paint}}
\newcommand{\rain}{{\tt rain}}
\newcommand{\reeb}{R}
\newcommand{\surgery}{{\tt surgery}}
\newcommand{\wet}{{\tt wet}}

\newcommand{\MTAlg}{\Algorithm{MTAlg}\xspace}
\newcommand{\Init}{\Algorithm{Init}\xspace}

\newcommand{\CodeComment}[1]{\textcolor{blue}{\texttt{#1}}}

\author{}

\title{Follow the Flow of Rain and Paint: An Instance Optimal Contour Tree Algorithm}
\date{}

\begin{document}

\maketitle

\section{Preliminaries}
Consider a piecewise-linear (PL) $2$-manifold $\MM$ with boundaries.
This is represented by a triangulation (given as a DCEL) where edges
can be oriented and the boundary of a face is always counterclockwise. Some faces will \emph{not}
be triangular, and each of these corresponds to a \emph{boundary face}.
Let $G(\MM) = (V,E)$ denote this triangulation. 
We will assume that $\MM$ is embedding in $\RR^d$.
The manifold within each face is just linear.
We define a \emph{height function} on $\MM$, $f:\MM \mapsto \RR$. For each point in $x \in \MM$, let $f(x)$ simply be the value of $x_1$.
We distinguish between vertices and points in $\MM$. A point simply denotes any $x \in \MM$. A vertex is a point
corresponding to some vertex of the triangulation $G(\MM)$.

The ``status" of an internal point is decided by the gradient at that point. Points of zero gradient are deemed \emph{critical};
all other points are regular.
For PL manifolds, we need a discrete version of this definition.
For vertex $v$, we use $\nbd(v)$ to denote the neighborhood of $v$ in $M$. Let $\outnbd(v) := \{ w | f(v) > f(w), w \in \Gamma(v)\}$
and $\innbd(v) := \{w | f(v) < f(w), w \in \Gamma(v)\}$. 
%It is convenient to define $D(\MM)$, a directed variant of $G(\MM)$,
%where edges are directed from higher to lower $f$-value. So $\outnbd(v)$ and $\innbd(v)$ are exactly the out- and in-neighborhoods
%of $v$. A vertex in $v \in V$ can be classified in four types.
\smallskip
\begin{asparaenum}
	\item Regular: Both $\outnbd(v)$ and $\innbd(v)$ are non-empty and are contiguous in $\nbd(v)$.
	\item Maxima: $\nbd(v) = \outnbd(v)$.
	\item Minima: $\nbd(v) = \innbd(v)$.
	\item Saddle: Both $\outnbd(v)$ and $\innbd(v)$ are not contiguous in $\nbd(v)$.
\end{asparaenum}
\smallskip
Because $\MM$ is PL, the gradient at all non-vertex points of $\MM$ is zero.
So all non-vertex points are regular. We will assume that $f$ is Morse, so the function value on all critical points is distinct. We will assume the stronger condition that all internal
vertices have distinct function values.
A value $t$ is \emph{non-critical} if there exists no critical point $x$ such that $f(x) = t$.

Furthermore, the boundary vertices have special properties, encapsulated in the following definition.

\begin{definition} \label{def:bound} A manifold is \emph{boundary critical} if the following holds.
Consider a boundary face $F$. All vertices in $F$ have the same function value. Furthermore, all
internal neighbors of $F$ either have function values strictly greater (or strictly less)
than this value.
\end{definition}

In other words, the vertices in a boundary face collectively behave like a maximum or minimum. 
Abusing notation, the term ``critical point", ``maxima", and ``minima" will also refer to boundaries.
Critical values include these boundary values.

\begin{definition} \label{def:desc} A \emph{descending path} in $\MM$ is a continuous function $\lambda:[0,1] \mapsto \MM$
such that for all $a < b$, $f(\lambda(a)) \geq f(\lambda(b))$. A \emph{PL descending path} from $x$ to $y$ is a sequence
of points $x = s_0, s_1, s_2, \ldots, s_k = y$ with the following properties: for all $s_i$ but the first and last,
$s_i$ lies on an edge of $\MM$. $s_i$ and $s_{i+1}$ lie in the same face of $\MM$. 
For $i < j$, $f(s_i) \geq f(s_j)$.
\end{definition}

Since $\MM$ is PL, if there is a descending path from $x$ to $y$, it must be PL. So we will always refer to
PL descending paths.

\begin{definition} \label{def:cont} A \emph{contour} in $\MM$ is a continuous function $\phi:\SSS^1 \mapsto \MM$
such that $\forall a, b \in \SSS^1$, $f(\phi(a)) = f(\phi(b))$. This is called a \emph{$t$-contour} if $f(\lambda(a)) = t$.
A contour is \emph{simple} if $\forall a, b \in \SSS^1$, $\lambda(a) \neq \lambda(b)$. 
\end{definition}

Just as in the case of paths, contours are also PL and have a corresponding discrete representation.

\begin{claim} \label{clm:cont} If two distinct contours intersect, then the intersection point
is critical.
\end{claim}

\begin{proof} Consider the intersection point $x$. It cannot be internal to a face, otherwise $\MM$ self-intersects.
By the distinctness of the contours, there must be $4$ faces incident to $x$, where each face
has vertex with a larger function value.
It cannot be internal to an edge, otherwise this edge would participate in $4$ faces. So it must be a vertex. 
These $4$ faces lead to at least $2$ neighbors of $x$ with higher function value. One can also show the existence of $2$
interleaved neighbors with lower function value. So $x$ is critical.
\end{proof}

Note that there is a unique contour containing a regular point, motivating the following definition.
\begin{definition} \label{def:p-cont}
For regular point $x$, $\cont(x)$ denotes the unique contour containing $x$.
\end{definition}

By the Morse property of distinct critical values, we get the following important lemma.

\begin{lemma} \label{lem:cont} Let $t$ be a non-critical value. The set of $t$-contours is a set
of disjoint simple PL contours.
\end{lemma}

We can use contours to classify saddle points. For a saddle point, we call two increasing (or decreasing) neighbors \emph{opposite}
if they lie in disjoint part of $\Gamma^+(w)$ (or $\Gamma^-(w)$).

\begin{definition} \label{def:merge} Consider two increasing (resp. decreasing) opposite edges incident to saddle point $x$,
and let $y$ and $z$ be points on these edges such that $f(y) = f(z) = f(x) + \eps$ (resp. $=f(x) - \eps$).
If $\cont(y)$ and $\cont(z)$ are disjoint, then $x$ is called a \emph{merge} (resp. \emph{split}) vertex.
\end{definition}

\begin{claim} \label{clm:saddle} Every saddle is either a merge or a split, but not both.
\end{claim}

\subsection{The L-homotopy} \label{sec:l-hom}
We need a specific notion of homotopy between contours. The L below stands for ``level".

\begin{definition} \label{def:cont-hom} Two simple contours $\phi_0, \phi_1$ are \emph{L-homotopic}
if: there exists a continuous function $H:\SSS^1 \times [0,1] \mapsto \MM$
such that $H(\SSS^1,0) = \phi_0$, $H(\SSS^1,1) = \phi_1$, and $\forall a \in [0,1]$,
$H(\SSS^1,a)$ is a contour.
\end{definition}

\begin{claim} \label{clm:non-hom} Consider a critical point $x$. Take an increasing edge incident to $x$
and let $y$ be a point on this edge within an $\eps$-ball of $x$. Similarly, take a decreasing edge
and point $z$ on this edge. The contours through $y$ and $z$ are not L-homotopic.
\end{claim}

\begin{proof} If the contours are L-homotopic, the homotopy must ``pass through" a contour $\phi$ at $x$.
There are two distinct simple contours through $x$, and there are three possible choices for $\phi$. None of these
choices can work.
\end{proof}

\begin{definition} \label{def:hom-mono} Two simple contours are $\phi_0, \phi_1$ are \emph{monotonically L-homotopic}
if the $L$-homotopy $H$ has the following property. Consider function $\alpha:[0,1] \mapsto \RR$
where $\alpha(a) = f^{-1}(H(\SSS^1,a))$. The function $\alpha$ is strictly increasing or strictly decreasing.
\end{definition}

\begin{claim} \label{clm:mono} If two simple contours are L-homotopic, then they are monotonically L-homotopic.
\end{claim}

\begin{proof} Consider some L-homotopy $H$. We can assume wlog that for all $a \neq b$, $H(\SSS^1,a) \neq H(\SSS^1,b)$.
(Otherwise, one could simplify $H$ and ``shortcut" between $a$ and $b$.)
By continuity of $H$, the function $\alpha$ is continuous. Hence, if it is not strictly
increasing or decreasing, $\alpha$ attains its maximum at some $a \in (0,1)$. There exist $a_1 = a - \eps_1$
and $a_2 = a + \eps_2$ (for small appropriate small values $\eps_1, \eps_2$) such that $\alpha(a_1) = \alpha(a_2)$.
Furthermore, $H(\SSS^1,a_1)$ and $H(\SSS^1,a_2)$ intersect (and are distinct). By \Clm{cont}, $\alpha(a_1)$
is a critical value. 
\end{proof}

\begin{claim} \label{clm:reg} Consider a contour $\phi$ containing only regular points. 
Take the edge cut by $\phi$ whose upper endpoint $x$ has minimum value, and let $e$
denote the portion about $\phi$. Then $\phi$ is L-homotopic to the contour
through any point in this interior of $e$. If $x$ is regular, then $\phi$ is L-homotopic
to the contour through $x$.
\end{claim}

\begin{proof} One should able to explicit construct the L-homotopy as a piecewise linear map.
\end{proof}

\subsection{Reeb graphs} \label{sec:reeb}

Now we get to the real definitions. Call a contour regular if it only contains regular points.

\begin{definition} \label{def:rel} Define a relation $\sim$ between regular contours, so $\phi \sim \phi'$ if $\phi$
is L-homotopic to $\phi$'.
\end{definition}

It is easy to check that $\sim$ is an equivalence relation, so the set of regular contours
is partitioned into equivalence classes. All contours in a class are L-homotopic to each other.

\begin{claim} \label{clm:equiv} Consider such an equivalence class $\Phi$. Then $f(\Phi)$
is an open interval $(a,b)$, where $a$ and $b$ are critical values.
\end{claim}

\begin{proof} \Clm{mono} should help us argue that for distinct $\phi, \phi' \in \Phi$,
for any $t \in [f(\phi),f(\phi')]$, there exists $\phi'' \in \Phi$ such that $f(\phi'') = t$.
So $f(\Phi)$ is an interval. \Clm{reg} should tell us that this interval must be open
with critical value endpoints.
\end{proof}

We can now define the Reeb Graph $\reeb(\MM)$. We use the fact that critical points 
have distinct values. (Technically, the Reeb Graph depends on the function $f$.
Because it is implicit by the embedding of $\MM$, we will remove it from the notation.)

\begin{definition} \label{def:reeb} The \emph{Reeb Graph} $\reeb(\MM)$ has as vertex set
the set of critical points. The edge $(x,y)$ is present if there exists equivalence class
$\Phi$ such that $f(\Phi) = [f(x),f(y)]$.
\end{definition}

We define a ``cut" operation on manifolds that cuts along a contour to create
a manifold with a new boundary. Intuitively, any cut creates two disjoint boundaries, one
above and one below.

\medskip
\fbox{
\begin{minipage}{0.9\textwidth}
{\bf $\cut(\MM,\phi)$}

\smallskip
\begin{asparaenum}
	\item Insert the vertices of $\phi$ in $\MM$.
	\item Take an arbitrary edge $e$ intersecting $\phi$.
	\item Take a point $x$ (resp. $y$) on $e$ at distance $\eps$ above (resp. below) $\phi$.
	\item Let $\phi_x$ be the contour through $x$, and analogously define $\phi_y$. Insert the vertices
	of these contours in the manifold. Triangulate all new faces.
	\item Delete the vertices (and incident edges, faces, etc.) of $\phi$ from the manifold.
\end{asparaenum}
\end{minipage}}

\medskip
This results in a new manifold that is also boundary critical. There are two new disjoint
boundary faces added. The manifold may be disconnected. Abusing notation, we use $\cut(\MM,\Phi)$
for a set $\Phi$ of contours to mean the manifold obtained by cutting along all contours in $\Phi$.

We now describe a procedure that converts the Reeb Graph of $\cut(\MM,\phi)$ to the Reeb Graph of $\MM$.

\medskip
\fbox{
\begin{minipage}{0.9\textwidth}
{\bf $\surgery(\MM,\phi)$}

\smallskip
\begin{asparaenum}
	\item Let $\MM' = \cut(\MM,\phi)$.
	\item Construct $\reeb(\MM')$ and let $B, C$ be the vertices corresponding to the new boundaries created
	in $\MM'$. (One is a minima and the other is maxima.)
	\item Since $B, C$ are leaves, they each have unique neighbors $B'$ and $C'$, respectively. Insert
	edge $(B',C')$ and delete $B,C$ to obtain a new graph.
\end{asparaenum}
\end{minipage}}

\medskip
\begin{lemma} \label{lem:surgery} For any regular contour $\phi$, the output of $\surgery(\MM,\phi)$ is $\reeb(\MM)$.
\end{lemma}

\begin{proof} The only difference between $\reeb(\MM)$ and $\reeb(\cut(\MM,\phi))$ is in the equivalence
class corresponding to $\phi$. This class is split into two classes in $\cut(\MM,\phi)$, which are joined
back by $\surgery(\MM,\phi)$.
\end{proof}

\section{Raining to partition $\MM$} \label{sec:rain}

In this section, we assume that $\MM$ is homeomorphic to $\SSS^2$ with boundaries.

\begin{definition} \label{def:dom} A manifold is \emph{extremum dominant} if there exists
a maximum (resp. minimum) $x$ such that every non-maximal (resp. non-minimal) \emph{vertex} in the manifold has an increasing (resp. decreasing) path to $x$.
\end{definition}

We will describe a linear time algorithm that partitions $\MM$ into a set of extremum dominant manifolds.
The first part of this requires a ``raining" procedure. Start at some point $x \in \MM$ and imagine rain at $x$.
The water will flow downwards along descending paths and ``wet" all the points encountered.

\begin{definition} \label{def:wet} The set of points $y$ such that there is a descending path from $x$ to $y$
is denoted by $\wet(x,\MM)$. A point $z$ is on the \emph{interface} of $\wet(x,\MM)$ if every neighbor
has non-trivial intersection with $\wet(x,\MM)$.
\end{definition}

\begin{claim} \label{clm:inter} For any $x$, the interface of $\wet(x,\MM)$ is a set of disjoint, simple, non-regular contours.
Furthermore, each such contour contains a merge vertex.
\end{claim}

\begin{proof} If $y \in \wet(x,\MM)$, then any contour containing $y$ is also in $\wet(x,\MM)$.
Consider a $y$ in the interface. We argue that all points in a contour $\phi$ containing $y$
are also in the interface. If such a contour is regular, we can find a contour just above it that
is also wet.

If two simple contours intersect, the intersection point $y$ is critical. We can argue that the only
way to reach $y$ from a descending path is reach one of the contours and then follow it to get to $y$.
But that would imply the existence of another critical point on this contour, violating the Morse condition.

Let us argue that $y$ is a merge vertex. Consider the descending path from $x$ to $y$. The final edge
that reaches $y$ is obviously all wet, and is an increasing edge from $y$. Consider an opposite increasing
edge. Take two point in these edges that are infinitesimally higher than $y$ and show that the contours
through them as disjoint. So $y$ is a merge.
\end{proof}

We will define a \lift{} operation on the interface contours. Consider such a contour $\phi$ containing
merge vertex $y$. Take any dry increasing edge incident to $y$, and pick the point $z$ on this edge at height
$f(y) + \eps$. Let $\lift(y) = \cont(z)$.

\begin{claim} \label{clm:cut-int} Let $\phi$ be an interface contour. Then $\cut(\MM,\lift(\phi))$
results in two disjoint manifolds, one consisting entirely of dry points, and the other containing
all wet points.
\end{claim}

\begin{proof} Note that $\lift(\phi)$ is a closed, dry curve. But the Jordan Curve Theorem,
$\cut(\MM,\lift(\phi))$ results in two disjoint manifolds. Let $\NN$ be the manifold containing
the point $x$, and $\NN'$ be the other manifold. If there exists a wet point $z$ in $\NN'$,
then there exists a descending path from $x$ to $z$ intersecting $\phi$ (Jordan Curve Theorem again).
That contradicts the fact that $\phi$ is dry.
\end{proof}

We now define the main partitioning procedure that cuts up a manifold $\NN$ homeomorphic to $\SSS^2$ with boundaries into a series of extremum
dominant manifolds. It takes as input the manifold and a maximum $x$. To initialize,
we begin with $\NN$ set to $\MM$ and $x$ as an arbitrary maximum.

\medskip
\fbox{
\begin{minipage}{0.9\textwidth}
{\bf $\rain(x,\NN)$}

\smallskip
\begin{compactenum}
	\item Compute the interface of $\wet(x,\NN)$. 
	\item If the interface is empty, simply output $\NN$. Otherwise, denote the contours by $\phi_1, \phi_2, \ldots, \phi_k$, and set $\phi'_i = \lift(\phi_i)$.
	\item Initialize $\NN_1 = \NN$.
	\item For $i$ from $1$ to $k$: 
	\begin{compactenum}
		\item Construct $\cut(\phi'_i,\NN_1)$ to get the dry manifold $\LL_i$ and wet manifold $\NN_{i+1}$.
		\item Let the new boundary of $\LL_i$ be $B_i$. Invert $\LL_i$ so that $B_i$ is a maximum. Recursively
		call $\rain(B_i,\LL_i)$.
	\end{compactenum}
	\item Output $\NN_{k+1}$.
\end{compactenum}
\end{minipage}}

\medskip
For convenience, denote the total output of $\rain(x,\MM)$ by $\MM_1, \MM_2, \ldots, \MM_r$.

\begin{lemma} \label{lem:rain-1} Each output manifold $\MM_i$ is extremum dominant.
\end{lemma}

\begin{proof} Consider a call to $\rain(x,\NN)$. If the interface is empty, then all of $\NN$
is in $\wet(x,\NN)$, so $\NN$ is trivially extremum dominant. So suppose the interface
is non-empty and consists of $\phi_1, \phi_2, \ldots, \phi_k$ (as denoted in the procedure).
By repeated applications of \Clm{cut-int}, $\NN_{k+1}$ consists of wet points. 
Consider $\wet(x,\NN_{k+1})$. The interface must exactly be $\phi_1, \phi_2, \ldots, \phi_k$.
So the only dry vertices are those in the boundaries $B_1, B_2, \ldots, B_k$. But these
boundaries are maxima.
\end{proof}

\begin{lemma} \label{lem:rain-time} The total running time of $\rain(x,\MM)$ is $O(n)$.
\end{lemma}

\begin{proof} Consider some invocation to $\rain(x,\NN)$. The total running time to determine
$\wet(x,\NN)$ (and the interface) is at most the number of vertices in $\wet(x,\NN)$. Note that
this includes original vertices in $\NN$ as well as new vertices created by constructing
$\phi_1, \phi_2, \ldots, \phi_k$. Essentially, we can bound the total time by $O(n)$ plus
the total sizes of all the interface boundaries ever constructed (during recursive calls).

To bound the latter quantity, we will simply argue that any edge contains at most
one boundary vertex. Consider some edge $e$ and take the first time a boundary vertex is created in $e$.
This happens when some contour $\phi$ intersects $e$. On applying $\cut(\NN,\lift(\phi))$,
consider the dry manifold $\LL$ and wet manifold $\MM$ constructed. The edge $e$ is split into two parts, the dry and wet part.
The wet part is in $\MM$ and will never be involved in any recursive call.

The dry part of $e$ (call it $e'$) needs to be considered.
A boundary vertex $v$ (an new endpoint of $e$) is created, which is part of the boundary $B$ in $\LL$,
such that $\rain(B,\LL)$ is called to the inverted $\LL$. In this call, since the water starts from $v$,
all of $e'$ becomes wet. So there is no further interface contour that can intersect this portion of $e'$.
\end{proof}

\begin{claim} \label{clm:rain-reeb} Given $\reeb(\MM_1), \reeb(\MM_2), \ldots, \reeb(\MM_r)$,
$\reeb(\MM)$ can be computed in $O(n)$ time.
\end{claim}

\begin{proof} Basically apply the $\surgery$ procedure, and invoke \Lem{surgery}.
\end{proof}

\section{Painting to compute contour trees} \label{sec:paint}

What the previous section allows us to restrict attention to extremum dominant manifolds.
We will orient so that the extremum in question is a \emph{minimum}.
We will fix such a manifold $\MM$, with the dominant minimum $x$.

We begin by associating a color with each maximum. If the set of maxima is $S$,
this is simply an arbitrary bijection $c:S \mapsto [|S|]$. Imagine there being a large
can of paint of color $c(y)$ at maxima $y$. We will spill different paint from each maximum and watch it flow down.
This is analogous to the raining in the previous section, but paint is a much more viscous liquid.
\emph{So paint only flows down edges, and it does not color the interior of faces.} Furthermore, paints
do not mix, so eventually, every vertex and edge of $\MM$ gets a unique color.

We begin with the procedure \paint. We can think of this procedure as extending the map $c:S \mapsto [|S|]$
to $c:G(\MM) \mapsto [|S|]$.

\end{document}

### 2014-05-24 paint.tex
adding files

\documentclass[A4paper,11pt]{article}

\usepackage{sariel}
\usepackage[noend]{algorithmic}

\usepackage{latexsym}
\usepackage{epsfig}
\usepackage{cite}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{psfrag}
\usepackage{verbatim}
\usepackage{paralist}
%\usepackage[usenames]{color}

\usepackage{graphicx}
\usepackage{euscript}%
\usepackage{xspace}%
\usepackage{color}%
\usepackage{picins}%
\usepackage{enumerate}
\usepackage{multirow}
\usepackage[noend]{algorithmic}
%\usepackage[breaklinks=true]{hyperref}

%% For hyperlinks. Should always be the last package.
\usepackage[colorlinks,urlcolor=blue,citecolor=blue,linkcolor=blue]{hyperref}

%%% \documentstyle[11pt,psfig]{article}
%%\addtolength{\textwidth}{3cm}
\newcommand{\ignore}[1]{}
\addtolength{\textwidth}{4cm}
\addtolength{\textheight}{4cm}
\addtolength{\oddsidemargin}{-2cm}
\addtolength{\topmargin}{-2.5cm}

\newcommand{\EPSfigure}[5]{

        % #1 = File name and other arguments to \psfig macro
        % #2 = Caption Text
        % #3 = Positioning Letters: h, t|b|p
        % #4 = {*} causes double column figure, {} for single
        % #5 = Label to apply to figure

        \begin{figure#4}[#3]
                \centering
                \ \psfig{file=#1}
%               \label{#5}
%
                \ \caption{{\em #2}\label{#5}}\hfill\break

        \end{figure#4}
}

% \newtheorem{theorem}{Theorem}[section]
% \newtheorem{thm}{Theorem}[section]
% \newtheorem{lemma}[theorem]{Lemma}
% \newtheorem{lem}[theorem]{Lemma}
% \newtheorem{claim}[theorem]{Claim}
% \newtheorem{fact}[theorem]{Fact}
% \newtheorem{definition}[theorem]{Definition}
% \newtheorem{prop}[theorem]{Proposition}
% \newtheorem{corol}[theorem]{Corollary}
% \newtheorem{obs}[theorem]{Observation}
% \newtheorem{defn}[theorem]{Definition}
% \newtheorem{example}[theorem]{Example}
% \newtheorem{observation}[theorem]{Observation}
% \newtheorem{corollary}[theorem]{Corollary}
% \newtheorem{property}[theorem]{Property}
% \newtheorem{remark}[theorem]{Remark}

\newcommand{\bT}{{\bf T}}
\newcommand{\cS}{{\cal S}}
\newcommand{\cB}{{\cal B}}
\newcommand{\cC}{{\cal C}}
\newcommand{\cD}{{\cal D}}
\newcommand{\cE}{{\cal E}}
\newcommand{\cF}{{\cal F}}
\newcommand{\cG}{{\cal G}}
\newcommand{\cH}{{\cal H}}
\newcommand{\cL}{{\cal L}}
\newcommand{\cP}{{\cal P}}
\newcommand{\cV}{{\cal V}}

\newcommand{\RR}{\mathbb{R}}
\newcommand{\bmu}{\overline{\mu}}
\newcommand{\poly}{\textrm{poly}}
\newcommand{\mymod}{\textrm{mod} \ }
\newcommand{\otilde}{\widetilde{O}}
%\newcommand{\qed}{\hfill $\Box$}
%\newcommand{\eps}{\varepsilon}
\newcommand{\fu}{\varphi}
\newcommand{\restr}[1]{{|_{{\textstyle #1}}}}
\newcommand{\proj}{\mbox{\rm proj}}
\newcommand{\vol}{\mbox{\tt vol}\,}
\newcommand{\area}{\mbox{\tt area}\,}
\newcommand{\conv}{\mbox{\tt conv}\,}
\newcommand{\diam}{\hbox{\tt diam}\,}
\newcommand{\hdisc}{\mbox{\rm herdisc}}
\newcommand{\ldisc}{\mbox{\rm lindisc}}
\newcommand{\EX}{\hbox{\bf E}}
\newcommand{\prob}{{\rm Prob}}
\newcommand{\proofend}{{\medskip\medskip}}
%\newcommand{\proof}{{\noindent\bf Proof: }}
%\newcommand{\reals}{{\rm I\!\hspace{-0.025em} R}}
%\newcommand{\dist}{\hbox{dist}}
\newcommand{\defeq}{\mbox{\,$\stackrel{\rm def}{=}$\,}}
\newcommand{\boxalg}[1]
{\begin{center}\fbox{\parbox{\columnwidth}{\tt
\begin{tabbing}
\=mm\=mm\=mm\=mm\=mm\=mm\=mm\=mm\=mm\kill
#1
\end{tabbing} } } \end{center} }

%% Hyper-linked References
\newcommand{\Sec}[1]{\hyperref[sec:#1]{\S\ref*{sec:#1}}} %section
\newcommand{\Eqn}[1]{\hyperref[eqn:#1]{(\ref*{eqn:#1})}} %equation
\newcommand{\Fig}[1]{\hyperref[fig:#1]{Figure~\ref*{fig:#1}}} %figure
\newcommand{\Tab}[1]{\hyperref[tab:#1]{Table~\ref*{tab:#1}}} %table
\newcommand{\Thm}[1]{\hyperref[thm:#1]{Theorem~\ref*{thm:#1}}} %theorem
\newcommand{\Lem}[1]{\hyperref[lem:#1]{Lemma~\ref*{lem:#1}}} %lemma
\newcommand{\Prop}[1]{\hyperref[prop:#1]{Property~\ref*{prop:#1}}} %property
\newcommand{\Cor}[1]{\hyperref[cor:#1]{Corollary~\ref*{cor:#1}}} %corollary
\newcommand{\Def}[1]{\hyperref[def:#1]{Definition~\ref*{def:#1}}} %definition
\newcommand{\Alg}[1]{\hyperref[alg:#1]{Algorithm~\ref*{alg:#1}}} %algorithm
%\newcommand{\Ex}[1]{\hyperref[ex:#1]{Example~\ref*{ex:#1}}} %example

%% Comments to ourselves
\newcommand{\Reminder}[1]{{\color{red}#1}}
\newcommand{\Sesh}[1]{\Reminder{Sesh interjects: #1}}

%%Ben Added
\DefineNamedColor{named}{RedViolet} {cmyk}{0.07,0.90,0,0.34}
\providecommand{\AlgorithmI}[1]{{\textcolor[named]{RedViolet}{\texttt{\bf{#1}}}}}
\providecommand{\Algorithm}[1]{{\AlgorithmI{#1}\index{algorithm!#1@{\AlgorithmI{#1}}}}}
\newcommand{\zip}{\Algorithm{ZIP}}
\newcommand{\link}{\Algorithm{LINK}}
\newcommand{\sizes}{\Algorithm{SIZES}}
\newcommand{\AMT}{\ensuremath{\mathsf{AMT}}\xspace}
\newcommand{\MT}{\ensuremath{\mathsf{MT}}\xspace}
\newcommand{\BCS}{\ensuremath{\mathsf{BCS}}\xspace}
\newcommand{\sub}{\mathsf{sub}}
\newcommand{\Merge}{\mathsf{Merge}}
\newcommand{\stack}{\mathsf{S}}
\newcommand{\AQ}{\mathsf{AQ}}

\newcommand{\MTAlg}{\Algorithm{MTAlg}\xspace}
\newcommand{\Init}{\Algorithm{Init}\xspace}

\newcommand{\CodeComment}[1]{\textcolor{blue}{\texttt{#1}}}

\author{}

\title{Painted Mountaintops: \break An Optimal Output Sensitive Merge Tree Algorithm}
\date{}

\begin{document}

\maketitle

\section{Preliminaries}
In the following let $G=(V,E)$ be a connected undirected graph such that $V\subseteq \Re^d$ and $E$ is the set of edges of some simplicial complex defined over $V$.  Let $f:V \mapsto \RR^+$ be a height function defined over the vertices of $V$.  For a vertex $v\in V$, we refer to the value $f(v)$ as the height of $v$, and we use $G^{\geq h}$ (resp. $G^{>h}$) to denote the induced subgraph of $G$ on the vertices with height $\geq h$ (resp. $>h$).  For a given height, $h$, we sometimes refer to a connected component of the induced subgraph $G^{\geq h}$ as a \emphi{superlevel set} of $G$ at height $h$.

We will assume no two vertices share the same height value.  Also for simplicity for now we assume $d=2$.

\subsection{Definitions}

\begin{definition}
\deflab{maxima}
We classify each vertex $v\in V$ as one of four types.  $v$ is a \emphi{maxima} (resp. \emphi{minima}) if for all $w$ such that $\{w,v\}\in E$, $f(v)>f(w)$ (resp. $f(v)<f(w)$).  If the neighbors of $v$ with greater function value all appear in a contiguous chain in either the clockwise or counterclockwise ordering of the neighbors around $v$ (determined by the given embedding of $G$), then we call $v$ a non-critical point.  Otherwise, if $v$ has neighbors of greater and smaller function value that do not appear in contiguous chains, then we call $v$ a \emphi{saddle point}.  We refer to maxima, minima, and saddle points as \emphi{critical points}.
\end{definition}

\begin{definition}
\deflab{aug}
The \emphi{augmented merge tree}, denoted $\AMT(G)$, of a graph $G=(V,E)$ with an associated height function $f:V \mapsto \RR^+$ and a vertex set labeled $\{v_1,\dots,v_n\}$ in increasing order of $f$ value, is the graph on the same vertex set with the same associated height function such that for $i<j$, $v_i$ and $v_j$ are adjacent in $\AMT(G)$ iff:
\begin{enumerate}
 \item $v_j$ is the vertex of smallest height in some connected component $C$ of $G^{>f(v_i)}$.
 \item In the graph $G$, $v_i$ is adjacent to some vertex in $C$.
\end{enumerate}
\end{definition}

We call vertices with $\geq 2$ neighbors with higher function value in $\AMT(G)$ \emphi{merge vertices}.  Note that by the definition of $\AMT(G)$ such a vertex $v$ must be connected to vertices from two distinct connected components in $G^{>f(v)}$ (and hence the appropriate name).  Only saddle points can be merge vertices (for all other vertex types, since the graph is triangulated, all neighbors of higher function value will be in the same connected component).  However, not all saddle points are merge vertices, and we refer to such vertices as \emphi{false merges}.  For simplicity we assume all merge vertices merge exactly two connected components, i.e. the number of neighbors of greater function value for of every vertex in $\AMT(G)$ is $\leq 2$.

\begin{definition}
\deflab{merge}
The \emphi{merge tree} of a graph $G=(V,E)$ with an associated height function $f:V \mapsto \RR^+$, denoted $\MT(G)$, is the tree obtained from $\AMT(G)$ by contracting all maximal paths of degree two nodes down to single edges.
\end{definition}

One can now see where the term ``augmented'' comes from.  Namely, it is the structure obtained by taking the merge tree and augmenting it with all non-critical vertices, that is the edges are subdivided with these vertices.  However, one can instead consider augmenting with respect to only a subset of these vetices.  Moreover, this structure can be obtained from $\AMT(G)$ in the same way as $\MT(G)$ was constructed. Specifically, for a subset $S\subseteq V(G)$, let $\AMT(S, G)$ denote the structure obtained from $\AMT(G)$ by contracting all maximal paths of degree two nodes, which do not contain vertices from $S$, down to single edges. Note that $\AMT(G) = \AMT(V,G)$ and $\MT=\AMT(\emptyset, G)$\footnote{Technically, if the global minimum is not also a merge vertex, then the merge contains this additional node, hanging from the lowest merge vertex.  We will largely ignore this vertex}.

\remove{
Given $\AMT(G)$, $\MT(G)$ can be computed in $O(\cardin{\AMT(G)}) = O(n)$ time (where $n$ is the number of vertices of $G$).
By how $\MT(G)$ is constructed from $\AMT(G)$, and \lemref{corr} and \lemref{tree}, it is not hard to see that $\MT(G)$ is a tree and that if we cut $\MT(G)$ at height $h$ then each subtree we get is a corresponding subset of some connected component of $G^{\geq h}$.  Specifically, it is the subset corresponding to all maxima and merging vertices from this component.

One can now see where the term ''augmented" comes from in augmented merge tree.  Namely, it is the structure obtained by taking the merge tree and augmenting it with all non-critical vertices, that is the edges are subdivided with these vertices.  However, one can instead consider augmenting with respect to only a subset of these vetices.  Moreover, this structure can be obtained from $\AMT(G)$ in the same way $\MT(G)$ was constructed. Specifically, for a subset $S\subseteq V(G)$, let $\AMT(S, G)$ denote the structure obtained from $\AMT(G)$ by contracting all maximal paths of degree two nodes, which do not contain vertices from $S$, down to single edges. Note that $\AMT(G) = \AMT(V,G)$ and $\MT=\AMT(\emptyset, G)$.
}

\subsection{Properties}
An \emphi{increasing rooted tree} is a tree on a vertex set with an associated function value for each vertex, such that the function values increase along any root to leaf path (naturally the root must be the vertex of smallest function value).
The following lemma with allow us to prove that $\AMT(G)$ is the unique increasing tree on the same vertex set as G with the same superlevel sets.

\begin{lemma}
\lemlab{unique}
Given an undirected graph $G=(V,E)$ with an associated height function $f:V \mapsto \RR^+$, $\AMT(G)$ is the unique increasing forest on the same vertex set and associated height function such that for any height $h$, the connected components of $G^{\geq h}$ are identical to the connected components of $\AMT(G)^{\geq h}$.
\end{lemma}
\begin{proof}
 The full proof is in the other writeup.
\end{proof}

Due to the manner in which $\MT(G)$ is constructed from the $\AMT(G)$, it is not hard to see that a similar lemma holds for $\MT(G)$.

\begin{lemma}
\lemlab{unique2}
Given an undirected graph $G=(V,E)$ with an associated height function $f:V \mapsto \RR^+$, let $W$ denote the set of maxima and merge vertices from $V$.  Then $\MT(G)$ is the unique increasing forest on $W$ with the same associated height function such that for any height $h$, the subset of vertices in each connected component of $\MT(G)^{\geq h}$ is a subset of vertices of a unique connected component in $G^{\geq h}$ (i.e. the components of $\MT(G)^{\geq h}$ and $G^{\geq h}$ are in one to one correspondence).
\end{lemma}

\section{Constructing the Merge Tree}
%For now we assume the input is a triangulated terrain defined as a height function above a triangulation in the plane, that all vertices have distinct heights, and that at most two components merge at a saddle point.
%
Let $M$ be the set of maxima and $S$ the set of saddle points in $G$.  The goal is to produce an algorithm for computing the merge tree that runs in $O\pth{n+\sum_{x\in M\cup S} \log depth(x)}$ time, where $n$ is the number of vertices in the triangulation and $depth(x)$ is the tree distance of $x$ from the root in the merge tree augmented with all saddle points.  
Specifically, if $F$ is the set of false merges, then depth is defined by the tree $\AMT(F, G)$.  
%$\MT(G)$ can then be computed by contracting all maximal paths of degree two vertices (which can be done in $O(n)$ time).  Therefore, for simplicity of exposition below we refer to $\AMT(F, G)$ as the merge tree (that is for now define $\MT(G) = \AMT(F, G)$)

First we describe the paint spilling subroutine which will be used in the initialization phase of our algorithm.  Intuitively this subroutine partitions $G$ into large connected regions, specifically one for each maxima.

\paragraph{Paint Spilling:}
First do a linear time search to determine all maxima.  Let $M$ denote this set.  We give each vertex in M a different color (i.e. a label). Arbitrarily pick a maxima in $M$, and do an ``any''-first search to explore all monotonically decreasing paths from this maxima.  That is, we do a search where if we run into an edge that requires us to walk up hill, we do not explore that edge.  In the process of this search we will mark all encountered vertices with the color of the maxima we started at, hence the name paint spilling.  Now repeat this process from other maxima in $M$, except now if an explored path from a maxima runs into an already colored vertex, we do not explore that vertex in the search.  If a vertex gets hit by multiple colors, we will say it gets colored only by the first to hit it, i.e. the vertex that explores it.

Note that saddle points are hit by colors from different sides.  Specifically, as defined above saddle points have two contiguous sets of neighboring high vertices (by assumption it is exactly two, but in general it could be more).  Let these sets be called $S_1$ and $S_2$.  
For saddle points we will record both the first color that hits it from $S_1$ and the first color that hits it from $S_2$.  For simplicity of exposition we will say the saddle point gets colored by both these colors (though it still only gets explored by the first color to hit it).

\begin{lemma}
All vertices get colored during Paint Spill.
\end{lemma}
\begin{proof}
Every vertex on the terrain either has a neighbor above it or is a mamixa, therefore by induction every vertex has a monotonically increasing path to a maxima.
\end{proof}
\begin{lemma}
Any two vertices in the augmented merge tree which received the same color during Paint Spill must be on the same root to leaf path in the AMT. 
\end{lemma}
\begin{proof}Let $a_1$ and $a_2$ be two vertices of the same color.  Let h be the min height of $a_1$ and $a_2$.  Let $\AMT^{\geq h}$ be the induced subgraph of \AMT on all vertices of height $\geq h$.  If $a_1$ and $a_2$ are not on the same root to leaf path then they are in different connected components in this induced subgraph.  By \lemref{unique}, they are then in different superlevel sets in the terrain at height $h$.  Therefore they do not have monotonically increasing paths to any common maxima.
\end{proof}
\begin{corollary}
\corlab{samePath}
Let $M\cup S$ be the set of all maxima and saddle points in $G$.  Then any two vertices in $M\cup S$ which received the same color during Paint Spill must be on the same root to leaf path in $\MT(G)$. 
\end{corollary}

\paragraph{Intuition.}  Before describing the algorithm in detail, lets build some intuition.  First, consider the paint spilling subroutine.  After paint spilling each color class in $G$ will be a large connected region containing exactly one maxima.  The hope is that the corresponding vertices in $\MT(G)$ were colored in a similar fashion, that is suppose the vertex colors when viewed in $\MT(G)$ look as if they were the result of paint spilling in the graph $\MT(G)$.  More formally, suppose that we had a stronger version of the above corollary in which not only are all vertices of the same color in $\MT(G)$ in the same root to leaf path, but also each color class is a connected path in $\MT(G)$.  In this case our paint spilling subroutine has luckily decomposed the entire merge tree into branches (i.e. connected decreasing paths), each containing one leaf.  Therefore, in order to construct the merge tree one simply needs to determine how to glue the branches together at merge vertices.

We have the following algorithm for this special case. \Sesh{Show counterexample to special case. This exists in 1D.} 
First sort the maxima and saddles from each color class independently into a decreasing path (in other words construct the branches for the color classes).  Now pick the branch containing the root of the merge tree (i.e. the branch with the lowest height vertex).  Starting at the leaf of this color class, walk down this branch one vertex at a time.  Each time you come to a new saddle vertex, we recursively build the subtree that needs to be glued in there.  Specifically, each vertex which is not a maxima is a saddle, and so we recorded a second color for that vertex.  Now recursively go down the branch of that other color class and build the subtree (i.e. go to its leaf and walk down , recursively gluing on subtrees as you go).  Once you reach the bottom vertex of that branch, you will have built the appropriate subtree to be glued on, at which point you attach it and move on.

In short, we constructed the merge tree by doing a post order traversal of the vertices in the merge tree\footnote{Yes, we agree that this does sound bizarrely circular.  We construct the tree by doing a post order traversal of the tree we are trying to construct.}.  Now in reality however we will not be so lucky with the coloring of vertices in $\MT(G)$, i.e. the color classes might not be contiguous.  However, it turns out that one can still construct the merge with a post order traversal.  That is, the basic informal outline of the algorithm above still works for this more difficult case, however, the details are naturally a little trickier.

\subsection{Initialization}
\seclab{init}
The first step is to identify the set of maxima, $M$, and the set of saddle points, $S$, with a linear time scan of $G$.  Next, we run the paint spilling algorithm described above which gives a unique color to each maxima, and gives two colors to each point in $S$.  Our merge tree algorithm will process vertices in $M\cup S$ by color class.  In order to facilitate this processing, we setup three objects for each color class, each of which we describe individually below.

The first thing we construct for each color class is a priority queue, which will be used to process the vertices of that color in decreasing height order.  This queue will contain exactly one entry for each vertex in $M\cup S$ that was labeled by the queue's color.  
Specifically an element in the queue is a pair of the form $(h,c)$ where $h$ is the height value of a saddle (or maxima) and $c$ is (a pointer to) the other color class that hit it (if it is a maxima then $c$ is just the color of the current queue).  The priority queue is a max queue on the $h$ values.

The second thing we construct for each color class is an entry in a union-find data structure, which is used to keep track of which colors have merged at any given time during the merge tree construction.  Specifically, we will maintain cross pointers between the union find components and their respective priority queues, so that each color query can be made efficiently by each queue.

The last thing we construct for each color class is simply an empty tree, which will be filled in (leaf first) as we process the associated color class.  Specifically, we associate a pointer with the queue of the corresponding color class which always points to the root of this tree (which is defined as the lowest height vertex in the tree).

We will also construct one global data structure, namely a stack which we denote by $\stack$.  Each entry in $\stack$ has the following form $[c_1, (h, c_2)]$, where $c_1$ and $c_2$ are color classes, and $h$ is a height value.  Specifically, during the execution of our merge tree algorithm, priority queues may push their top entry onto $\stack$, and when they do they record their own color (i.e. $c_1$) and then the queue entry (i.e. $(h, c_2)$).

\subsection{Handling a Merge}  
During the execution of our merge tree algorithm, periodically we will merge color classes.  Let $c_1$ and $c_2$ be two given color classes.  We now describe the subroutine $\Merge(c_1, c_2)$, which merges all the associated objects of these colors.  Specifically, we need to merge the associated trees, priority queues, and union find elements corresponding to $c_1$ and $c_2$.  First we look up $c_1$ and $c_2$ in the union find data structure (note that we need to do a look up first as $c_1$ and $c_2$ may be old labels, i.e. they may have already merged with other colors).  This gives us a pointer to the associated queues of these colors. 
Now each of these queues has a pointer to the root of its associated tree.  First we merge these trees.  Specifically, create a new vertex, label it the root, and add edges from it to the roots of the two trees\footnote{As described below, we only perform a merge when the entry from each queue is the same saddle point, which is the saddle point at which the corresponding merge is happening.  Therefore, each new vertex created can be labeled with an associated saddle point from the graph.}.  Now perform a standard priority queue merge of the two priority queues, and set its associated tree pointer to point to the new root we created.  Finally, we perform a union of $c_1$ and $c_2$ in the union find data structure.

%Note that it is possible that during the course of our merge tree construction that we try to merge a color class with itself, i.e. we call $\Merge(c,c)$ (specifically this happens at false merges), however this does not invalidate the merging procedure outlined above.

\remove{
\paragraph{Understanding how the merge tree is built: }
At a high level, the merge tree is constructed as follows.  Initially create a set of isolated nodes, one for each vertex in $M$.  As time goes on these nodes will merge, such that at any point in time we will have a forest of trees on these nodes.  In the end all of these tree will merge together producing the global merge tree.

At all times in this process, for each tree, we will keep three things, a pointer to the root, a color (i.e. label) for the tree, and an associated priority queue.  Merging two trees will always be done as follows.  A new vertex is added, and connected to the roots of the two merging subtrees.  This new vertex is labeled as the root of the merged tree.  Both the colors and priority queues of the merging trees are merged together.  Changing the root takes constant time, as we have pointers to the old roots.  
%Merging colors with be done using a union-find data structure, and the total time for all union find ops will be $\cardin{M}\alpha(\cardin{M})$ (i.e. it will be dominated by other terms in the runtime analysis as so we can ignore it).  
The time for merging priority queues will depend on their size and what data structure we use to implement them.

The elements in each queue will correspond to saddle points in the graph.  Specifically an element in the queue is a pair of the form $(h,c)$ where $h$ is the height value of a saddle and $c$ is the color of the other queue whose color hit it (the current queue color is the other color that hit it).  The priority queue is a max queue on the the $h$ values.

Merging colors will be done using a union-find data structure.  Specifically, we maintain cross pointers between the union find components and their respective queues.  Each time we merge two queues we merge their respective components in the union find data structure, and set a cross pointer between its root and the new queue.  Now when we look at a queue element $(h,c)$, we can find the queue corresponding to $c$, regardless of whether it has merge with other queue, by finding the component of that color in the union find structure and then following the cross pointer to the respective queue (i.e. c is really a pointer to an element in the union find data structure).  By the magic of union-find, all operations take amortized $\alpha(\cardin{M})$, and as we will show below we only perform $O(\cardin{M})$ such operations so the total time to manage the union find, i.e. $O(\cardin{M}\alpha(\cardin{M}))$, is negligible as it will dominated by other terms in the run time analysis.

\paragraph{Merge Forest Initialization: }
Before initializing the merge forest we run the Paint Spilling algorithm described above, in the process identifying the set $M$, and for each maxima in M the set of saddle points its color hits.
Now to initialize the merge forest we create a bunch of isolated nodes, one for each vertex in $M$.  We also create a set of $\cardin{M}$ singleton color elements to be managed by a union find data structure, where the colors used are the same from the Paint Spilling algorithm.  Each isolated node in the merge forest will contain a pointer to the receptive color element in the union find data structure sharing the same color that we ``spilled'' from the respective maxima. For each isolated node in the merge forest we also create a pointer an associated priority queue.  Initially each priority queue contains exactly one element for each saddle point hit by the respective color.  Specifically for each such saddle point we insert the element $(h,c)$ where $h$ is the height of the saddle point and $c$ is the other color that hit it (specifically $c$ is a pointer to the respective element in the union-find data structure).  
}

\subsection{The Algorithm}
\seclab{algorithm}

We now describe our algorithm for computing the merge tree $\MT(G)$ of a given graph $G$.  The first step is to perform the initialization procedures outlined in \secref{init}.  Let $\stack$ denote the global stack described in that section (which initially is empty).  Consider the queue of the color class with the lowest height saddle point (there are actually two, but just pick one arbitrarily).  Throughout the entire execution of the algorithm, this queue will be referred to as the active queue, and denoted $\AQ$.    
The algorithm for computing the merge tree is shown in \figref{algorithm}.  In this algorithm we use $\Init(G)$ to denote the initializations phase of \secref{init}.

\begin{figure}[h]%
    \begin{center}
    \fbox{\begin{minipage}{0.5\linewidth}%{\FigWidth}%
           $\MTAlg(G)$
               \begin{algorithmic}[1]
               \STATE $\Init(G)$
               \WHILE{$\AQ$ or $\stack$ not empty}

               \IF{$\stack$ empty}
               \STATE Pop $\AQ$ and push entry onto $\stack$
               \ENDIF

               \STATE %\linelab{gen:Random}%
               $[c_1, (h,c_2)] \leftarrow$ Peek$(\stack)$

               \IF{$c_1 = c_2$} 
               \STATE Pop$(\stack)$
               \STATE Continue           
               \ENDIF

               \STATE
               $(h',c')\leftarrow$  Pop($c_2$ queue).

               \IF{$h'=h$}
               \STATE $\Merge(c_1,c_2)$
               \STATE Pop$(\stack)$
               \ELSE
               \STATE Push$([c_2, (h',c')])$
               \ENDIF

%               \LINEIF{blahblah}
%               {%\linelab{bounded1x} %
%                  ``something''.} %
%
%               \CodeComment{// blah}%

               \ENDWHILE

           \end{algorithmic}
       \end{minipage}}%
    \caption{Algorithm to compute the merge tree.  Note that whenever a color is used it is first looked up in the union-find structure.}
    \figlab{algorithm}%
    \end{center}
\end{figure}%

Observe that each queue entry refers to a specific saddle (or maxima) in $G$ (namely the vertex in $M\cup S$ with corresponding height value).  Actually, in practice one would keep a cross pointer between this vertex and the queue entry (which we left out for simplicity of exposition).  Therefore in the analysis below, we will often refer to the queue entries as the vertices in the graph that they correspond to.

Now each time an entry is removed from the stack in the algorithm, the $\Merge$ subroutine is called and a new vertex is created in the forest of trees that we are maintaining.  As justified by the lemma below, we will also refer to stack entries and the vertices we create in the maintained forest as the vertices in $G$ that they correspond to.

\begin{lemma}
 Each entry popped from the stack (and corresponding vertex created in the maintained forest), 
 corresponds to the unique vertex in $M\cup S$, of the same height.
\end{lemma}
\begin{proof}
THIS PROOF MAY BE OVERKILL.
 Clearly by construction every queue entry corresponds to a specific vertex in $M\cup S$.  Specifically, since each saddle gets colored by exactly two colors, there are exactly two queue entries created for each saddle.  Now in the above algorithm description, an entry gets removed from the stack only after the corresponding queue entry from the queue that put it on the stack and the queue entry of the queue that it pointed to have been removed from the corresponding queue.  These two entries have the same height and hence corresponds to the same vertex in $G$.  Therefore each vertex popped from the stack corresponds to a unique vertex in $M\cup S$.
 (Note technically the above was the argument only for merge vertices, but the argument for maxima and false saddles in analogous.) 
\end{proof}

\begin{lemma}
 \lemlab{eventually}
 \begin{compactenum}
  \item
 Every vertex that gets pushed onto the stack eventually gets popped off. 
\item
 At all times the vertices on the stack are sorted in increasing order by height (with the highest at the top.)
 \end{compactenum}
\end{lemma}
\begin{proof}
 NEEDS TO BE FILLED IN
\end{proof}

As a consequence of the fact that the stack is always sorted we have the following corollary.

\begin{corollary}
\corlab{queueStatus}
 Each priority queue can have at most one entry on the stack at any given time.  In particular, a queue cannot be modified (i.e. popped or merged) until all its entries on the stack have been removed.
\end{corollary}

%First we initialize the merge forest (which in turn calls the Paint Spilling subroutine).  Now we go through the priority queues one at a time, handling all events in a given queue before moving on to the next one.\footnote{Currently the order in which we visit the queues does not seem relevant for our analysis, however likely in practice there is a benefit to moving through them in the same order we spilled paint (or the reverse order).}

%We now describe how to handle a queue event $(h,c)$.  First, find the priority queue that $c$ references, i.e. follow the pointer to some component in the union data structure and then follow the pointer to the queue associated with the root of that component.  Check the element at the top of the queue $(h',c')$.  If $h'=h$ then both queues highest height event is the same saddle point (since we assumed distinct point heights), and we are ready to handle this event. First we check if $c'$ and $c$ have merged in the union find data structure.  If so then this is a "false" merge point, i.e. only the topology of the current component is changing rather than two components merging.  
%In this case we simply clear these events from the tops of the queues and proceed to go down the first queue.  Otherwise, if $c'\neq c$ then this is a true merge in which case we merge the trees in the merge forest, the queues, and the colors in the union-find, all as described above.  Then we proceed to handle the next event in the merged queue.

%Now suppose that $h\neq h'$.  First observe that we always have $h\leq h'$ (we leave this to the reader to verify), and so in this case $h'>h$.  In this case we will build a stack of queue events (of increasing heights) which need to be recursively processed.%
%\footnote{Intuitively, the process of building up and clearing this stack will correspond to exploring one of the subtrees in the merge tree of the saddle point at height $h$.}
%Specifically, the stack will contain pointers to queues whose top queue event is waiting for its corresponding queue to clear off higher events.  Initially the stack only contains two pointers, one to the queue whose top event is $(h,c)$ (pushed onto the stack first) and a pointer to the queue whose top event is $(h',c')$.  We then proceed to handle the stack events in the same manner we handled queue events, i.e. until the stack is cleared it is treated as the the current priority queue which we are trying to clear (whose events happen to already be sorted by height in the stack ordering).

\section{Analysis}
\subsection{Running Time}
Let $M$ be the set of maxima and $S$ the set of saddle points in $G$, and for a vertex $v\in M\cup S$, let $depth(v)$ be the distance of $v$ from the root in $\MT(G)$

\begin{lemma}
 $\MT(G)$ runs in $O\pth{n+\sum_{v\in M\cup S} \log depth(v)}$ time.
\end{lemma}
\begin{proof}
To get the stated running time bound we will assume each priority queue is implemented using a Fibonacci heap, although there are other heap types that also work (such as binomial heaps).

First, we show that $\Init(G)$ easily runs in the given time bound.  The paint spilling subroutine runs in $O(n)$.  After that we must build the associated structures for each of the $\cardin{M}$ color classes.  Setting up the empty trees and union find data structure takes constant time per color.  Since each vertex in $M\cup S$ appears in at most two color classes and insertions in Fibonacci heaps take constant time, the total time spent to setup the queues is $O(\cardin{M\cup S})$, and so the bound follows.

Now we show the while loop terminates in the given time bound.  
First, we prove that for any priority queue used during the execution of $\MTAlg(G)$, at all times the set of vertices contained in the queue are on the same root to leaf path.  Clearly \corref{samePath} implies that this is true initially.  Now we never insert or modify elements in a queue, so the only time this might break is after two queues merge.  So suppose $\Merge(c_1, c_2)$ was called on two color classes $c_1$ and $c_2$.  As argued above, in order for $\Merge$ to be called on these color classes, the vertex we popped from the stack must be a saddle at which these colors meet, and all elements left in the queues of $c_1$ and $c_2$ must have height less than this saddle.  Specifically, below this saddle point these colors have merged, and so all vertices left in the merging queues are in the same root to leaf path in the merge tree.

Observe that in each iteration of the while loop, the only non-trivial work done is in popping elements from the queues and the $\Merge$ operation, and the only non-trivial thing in a $\Merge$ is the merge of priority queues.  We charge these queue operations to the respective vertices in $M\cup S$ that caused them.  Specifically, for a given vertex $v\in M\cup S$, $v$ is extracted from a queue at most twice and was the vertex at which a merge occurred at most once.  Merges and extracting the max in the queues take $O(\log m)$ time\footnote{Actually merges are constant time in Fibonacci heaps, but this argument shows that binomial heaps should work as well.}, where $m$ is the size of the queue.  However, the fact that all vertices contained in the queue are always on the same root to leaf path implies that when we perform these queue operations for $v$, the size of its queue is bounded by its depth in the merge tree (since $v$ is the higher than any other element in the queue when these operations are 
performed).  Therefore the total amount charged to $v$ is $\log depth(v)$, and since at least one element gets removed from a queue in each iteration of the while loop, this completes the proof.  (Note that since there are only a constant amount of union find operations per iteration of the while loop, these operations can be ignored.)
\end{proof}

\subsection{Correctness}

\begin{lemma}
 At any point during the execution of the algorithm, the forest of trees being maintained corresponds to a forest of disjoint rooted subtrees of vertices in $\MT(G)$.
\end{lemma}
\begin{proof}
 The statement will be proven by induction on the $i$th vertex to be created (correspondingly the $i$th vertex popped from the stack).
 Initially the forest is an empty forest and so the base case trivially holds.

 Consider the $i$th vertex, $v_i$, to be created and removed from the stack.  Let $c_1$ and $c_2$ be the two colors that labeled this vertex.  It is clear from the algorithm description that when vertex $v_i$ is popped from the stack, the entry corresponding to it in the queue for $c_1$ and $c_2$ must be removed from the corresponding queues.  In particular, by \corref{queueStatus}, this implies that at the time $v_i$ is created, the set of vertices from the $c_1$ and $c_2$ queues that have been removed and added to the current forest is exactly the set of vertices with height greater than $v_i$.

 REMAINING ARGUMENT SKETCH:  The above then implies the subtrees rooted at vertices of colors $c_1$ and $c_2$ have been correctly reconstructed.  This issue is that some of the other vertices of other colors that need to be in the subtrees of $v_i$, might not be have been made yet.  However, one should be able to argue that if some subset of the colors have been correctly constructed then always at least one of the not constructed colors should be adjacent at a saddle and hence also should have been constructed.
\end{proof}

Observe that the above lemma implies that the algorithm correctly constructs the entire merge tree since all vertices in the active queue clearly get put onto the stack.  Specifically, by \lemref{eventually} all vertices that get put onto the stack must get popped off, and the active queue contains the vertex of lowest height in the $\MT(G)$, and so by the above lemma, the algorithm must construct the subtree of the root of $\MT(G)$.

\remove{
\begin{lemma}
 The order in which vertices get popped of the stack in the algorithm of \secref{algorithm} is a valid post order traversal ordering of a DFS of $\MT(G)$.
\end{lemma}
\begin{proof}
We prove the above lemma by arguing inductively that the $i$th vertex to be popped from the stack is the correct next vertex in a post order traversal. First observe that the first vertex popped from the stack is the highest vertex in the active queue (i.e. it is the first to be put on the stack and is then immediately popped off), which is maxima and hence a leaf in $\MT(G)$, and so is the first vertex in some post order traversal.

Consider the $i$th vertex, $v_i$, to be popped off the stack.
\end{proof}

Given the above lemma it is not hard to prove that the algorithm of \secref{algorithm} constructs $\MT(G)$. 
}

\section{Lower Bound}
This section should be ignored now, it will just be good to remember for later.

In \figref{reduction} we see a reduction from sorting $n$ values to computing the merge tree on a planar triangulation on $O(n)$ vertices.  In the figure '$+$' and '$-$' mean $+\infty$ and $-\infty$, respectively.  This specific example was constructed such that the values we are sorting are merge vertices, i.e. they show up in sorted in order in not only the augmented merge tree, but also the regular merge tree.  If one removed the top row of alternating $+\infty$ and $-\infty$ then one would get a simpler example but it would only be sorted in the augmented merge tree.

\begin{figure}[h]
\begin{center}
\includegraphics[width=4.0in]{lowerBound}
\end{center}
\vspace{-.2in}
\caption{Reduction from sorting to computing merge trees.}
\figlab{reduction}
\end{figure}

%\bibliographystyle{alpha}
%\bibliography{connected-components}

\end{document}

### 2014-06-10 paint-contour-tree.tex
writing paint section

@@
 \newcommand{\surgery}{{\tt surgery}}
+\newcommand{\touch}{T}
 \newcommand{\wet}{{\tt wet}}
@@
 We begin by associating a color with each maximum. If the set of maxima is $S$,
-this is simply an arbitrary bijection $c:S \mapsto [|S|]$. Imagine there being a large
+this is simply an arbitrary bijection $\chi:S \mapsto [|S|]$. Imagine there being a large
 can of paint of color $c(y)$ at maxima $y$. We will spill different paint from each maximum and watch it flow down.
@@
 do not mix, so eventually, every vertex and edge of $\MM$ gets a unique color.
-
-We begin with the procedure \paint. We can think of this procedure as extending the map $c:S \mapsto [|S|]$
-to $c:G(\MM) \mapsto [|S|]$.
+%
+\begin{definition} \label{def:paint} A  \emph{painting} of $\MM$ is a map $\chi:V \cup E \mapsto [|S|]$
+with the following properties. (Remember that $G(\MM) = (V,E)$.)
+\begin{asparaitem}
+	\item The restriction $c:S \mapsto [|S|]$ is a bijection.
+	\item Consider an edge $e$. There exists a descending path from some maximum $x$ to $e$
+	consisting of edges in $E$, such that all edges and vertices along this path have the same color. (The maximum $x$
+	is well-defined to be the unique maximum with the same color as $e$.)
+\end{asparaitem}
+\end{definition}
+%
+Note that paintings are not unique. It is straightforward to construct a painting of $\MM$ in $O(n)$ time, using a descending BFS from
+each maximum. 
+%
+\begin{definition} \label{def:color-set} Fix a painting $\chi$, and a color $c$.
+A critical point $x$ is \emph{$c$-touched} if $x$ is incident to a $c$-colored
+increasing edge. The set of $c$-touched critical points is denoted $\touch(c)$.
+\end{definition}
+%
+We will store each set $\touch(c)$ in a binomial heap. Abusing notation, $\touch(c)$ refers
+both to the set and the data structure used to store it.
 

### 2014-06-10 paint-contour-tree.tex
more updates to painting section

@@
 
+\newcommand{\col}{col}
+\newcommand{\mcol}{mcol}
 \newcommand{\cut}{{\tt cut}}
@@
 \newcommand{\reeb}{R}
+\newcommand{\rep}{rep}
+\newcommand{\st}{K}
 \newcommand{\surgery}{{\tt surgery}}
 \newcommand{\touch}{T}
+\newcommand{\update}{{\tt update}}
 \newcommand{\wet}{{\tt wet}}
@@
 \section{Painting to compute contour trees} \label{sec:paint}
-
-What the previous section allows us to restrict attention to extremum dominant manifolds.
+The previous section allows us to restrict attention to extremum dominant manifolds.
 We will orient so that the extremum in question is a \emph{minimum}.
-We will fix such a manifold $\MM$, with the dominant minimum $x$.
+We will fix such a manifold $\MM$, with the dominant minimum $u$.
 
@@
 %
-Note that paintings are not unique. It is straightforward to construct a painting of $\MM$ in $O(n)$ time, using a descending BFS from
-each maximum. 
-%
+Note that paintings are not unique. It is straightforward to construct a painting of $\MM$ in $O(n)$ time, using a descending BFS from each maximum.
+
+\subsection{The data structures} \label{sec:struct}
+
+{\bf The binomial heaps $\touch(c)$:} For each color $c$, we store a binomial
+heap $T(c)$ of non-extremal critical points.
+
 \begin{definition} \label{def:color-set} Fix a painting $\chi$, and a color $c$.
-A critical point $x$ is \emph{$c$-touched} if $x$ is incident to a $c$-colored
-increasing edge. The set of $c$-touched critical points is denoted $\touch(c)$.
+A critical point $x$ is \emph{touched by $c$} if $x$ is incident to a $c$-colored
+increasing edge. The set of non-extremal critical points touched by $c$ is denoted $\touch(c)$.
+For $x$, $\col(x)$ is the set of colors that touch $x$.
 \end{definition}
 %
-We will store each set $\touch(c)$ in a binomial heap. Abusing notation, $\touch(c)$ refers
+We will store each set $\touch(c)$ in a binomial heap, keyed by the heights. 
+Abusing notation, $\touch(c)$ refers
 both to the set and the data structure used to store it.
 
+\medskip
+\noindent
+{\bf The union-find data structure on colors:} We will repeatedly perform unions
+of classes of colors, and this will be maintained as a standard union-find data structure.
+For any color $c$, $\rep(c)$ denotes the representative of its class. We initialize
+by simply setting $\rep(c) = c$ for all colors.
+
+\medskip
+\noindent
+{\bf The stack $\st$:} This consists of non-extremal critical points with the following structure.
+Each point $x \in \st$ has an associated subset of $\col(x)$, denoted $\mcol(x)$.
+Both $\mcol(x)$ and its complement are stored as hash table. So lookups, inserts, and deletes
+are in these sets are all constant time operations. The stack is guaranteed to satisfy 
+the following invariants.
+\begin{asparaitem}
+	\item For every $x \in \st$: For every $c \in \mcol(x)$, $x$ is the highest element
+	in $T(c)$. Furthermore, $c = \rep(c)$.
+	\item Consider $x, y \in \st$ such that $y$ was pushed on $x$. There exists $c \in \col(x) \setminus
+	\mcol(x)$ such that $c \in \col(y)$.
+\end{asparaitem}
+
+\subsection{Updating the stack} \label{sec:stack}
+
+We have a simple, but critical, procedure that updates the stack $\st$ by pushing
+points on to it.
+
+\medskip
+\fbox{
+\begin{minipage}{0.9\textwidth}
+{\bf $\update(\st)$}
+
+\smallskip
+\begin{compactenum}
+	\item Let $h$ be head of $\st$.
+	\item While $\col(h) \setminus \mcol(h)$ non-empty:
+	\begin{compactenum}
+		\item Pick some $c \in \col(h) \setminus \mcol(h)$. Replace it by $c' = \rep(c)$.
+		\item Obtain highest point $x$ in $T(c')$. 
+		\item If $x = h$, put $c'$ in $\mcol(h)$. Otherwise ($x \neq h$), push $x$ onto stack
+		and set $\mcol(x) = \{c'\}$.
+		\item Let $h$ be new head of $\st$.
+	\end{compactenum}
+\end{compactenum}
+\end{minipage}}
+
+\medskip
+When $\update(\st)$ terminates, the head $h$ has the following property. For every $c \in \col(h)$,
+$h$ is the highest point in $\touch(c)$.
+
+\begin{claim} \label{clm:cont} Consider an increasing path from $h$ to some maxima $x$.
+For any $y$ in this path, $\cont(y)$ only contains points of color $c(x)$.
+Furthermore, for all vertices on $\cont(y)$, their increasing edges are colored $c(x)$.
+\end{claim}
+
+\begin{proof} Consider this increasing path, and think of moving $y$ continuously downwards from $x$
+to $h$. Initially (when $y = x$), the lemma is trivially true. Consider the first $y$
+when the condition fails. We have $\cont(y)$ only containing points of color $c(x)$,
+but there is a vertex $z \in \cont(y)$ that has increasing edges of a different color $c'$.
+All the increasing edges of $z$ that are colored $c(x)$ are contiguous (since we can construct
+a contour intersecting all of them). Thus, there are two disjoint increasing neighborhoods for $z$,
+and $z$ is a saddle point. But that contradicts the fact that $h$ is the highest saddle point
+of color $c(x)$.
+
+So, as $y$ decreases, for all vertices on $\cont(y)$, their increasing edges are colored $c(x)$.
+This also means that $\cont(y)$ only contains points of color $c(x)$.
+\end{proof}
+
+We have important lemmas characterizing $|\col(h)|$.
+
+\begin{lemma} \label{lem:mono} If $h$ is a merge vertex, then $|\col(h)| > 1$.
+\end{lemma}
+
+\begin{proof} Suppose not. Consider the edges incident to $h$ in circular order. There exist (at least) $2$
+disjoint contiguous stretches of edges that are increasing. 
+All these edges have the same color $c$, associated with some maximum $x$. Consider two neighbors $y,z$
+that lie in different stretches. By the definition of a painting, there exist descending paths (along
+edges) from $x$ to $y$ and $x$ to $z$. We can construct a Jordan curve $C$ from $x \rightarrow y
+\rightarrow h \rightarrow z \rightarrow x$. All points in this curve except $h$ have 
+heights larger than $f(h)$. By the Jordan Curve Theorem, this partitions $\MM$ into
+an ``inside" and an ``outside".
+
+One of the lower neighbors of $h$ (say $p$) is inside, and another (say $q$) is outside.
+Because $h$ is a merge, neither of these can a minimum other than the dominant minimum $u$.
+Hence, both $p$ and $q$ have descending paths to $u$. So there is a path from $p$
+to $q$ where all heights are strictly less than $f(h)$. But this path must intersect $C$.
+Contradiction.
+\end{proof}
+
+\begin{lemma} \label{lem:col} If $h$ is a merge vertex, $|\col(h)| = 2$.
+If $h$ is a split vertex, $|\col(h)| = 1$.
+\end{lemma}
+
+\begin{proof} Consider one contiguous sequence of increasing edges from $h$.
+There exists a contour intersecting all of these edges. By \Clm{cont}, they all
+must have the same color. There are two such contiguous sequences.
+
+If $h$ is a merge vertex, then $|\col(h)| = 2$ by \Lem{mono}. If $h$ is a split vertex,
+there is a single contour passing through all increasing edges. By \Clm{cont}, they
+all have the same color.
+\end{proof}
+
 \end{document}

### 2014-06-11 paint-contour-tree.tex
done with first very preliminary pass

@@
 
+\newcommand{\build}{{\tt build}}
 \newcommand{\col}{col}
-\newcommand{\mcol}{mcol}
 \newcommand{\cut}{{\tt cut}}
 \newcommand{\cont}{\psi}
+\newcommand{\h}{high}
 \newcommand{\lift}{{\tt lift}}
+\newcommand{\mcol}{mcol}
+\newcommand{\onestep}{{\tt piece}}
 \newcommand{\paint}{{\tt paint}}
@@
 be triangular, and each of these corresponds to a \emph{boundary face}.
-Let $G(\MM) = (V,E)$ denote this triangulation. 
+Let $G(\MM) = (V,E)$ denote this triangulation. We assume this triangulation
+has bounded degree.
 We will assume that $\MM$ is embedding in $\RR^d$.
@@
 
+\medskip
+\noindent
+{\bf Highest points $\h(c)$:} For each color $c$, we maintain a special critical point $\h(c)$ of this color.
+It is supposed to denote the ``highest" point of color $c$, but the notion of highest evolves
+with the algorithm.
+
 \subsection{Updating the stack} \label{sec:stack}
@@
 \begin{compactenum}
+	\item If $K$ is empty, pick arbitrary $c$. Let $x$ be the highest point in $T(c)$.
+	Push $x$ onto stack and set $\mcol(x) = \{c\}$.
 	\item Let $h$ be head of $\st$.
@@
 
+\subsection{Constructing $\reeb(\MM)$} \label{sec:const}
+
+The main construction procedure is $\onestep(\MM)$. It pieces together $\reeb(\MM)$
+incrementally.
+
+\medskip
+\fbox{
+\begin{minipage}{0.9\textwidth}
+{\bf $\onestep(\MM)$}
+
+\smallskip
+\begin{compactenum}
+	\item Run $\update(\st)$. Pop $\st$ to get $h$.
+	\item Suppose $\mcol(h) = \{c\}$ (so $h$ is split). 
+	\begin{compactenum}
+		\item Connect $h$ (in $\reeb(\MM)$) to $\h(C)$ and the unique minimum corresponding to split $h$. 
+		\item Delete $h$ from $T(c)$, set $\h(C) = h$.
+		\item End procedure.
+	\end{compactenum}
+	\item Let $\mcol(h) = \{c_1, c_2\}$ (so $h$ is merge).
+	\item Connect $h$ in $\reeb(\MM)$ to $\h(c_1)$ and $\h(c_2)$.
+	\item Delete all copies of $h$ from $T(c_1)$ and $T(c_2)$.
+	\item Perform union of colors $c_1$ and $c_2$ (denote merged color as $c$). Merge heaps
+	to get $T(c) = T(c_1) \cup T(c_2)$.
+	\item Set $\h(c) = h$.	
+\end{compactenum}
+\end{minipage}}
+
+\medskip
+\begin{theorem} \label{thm:piece} Consider the very first call to $\onestep(\MM)$. Let $y$ be a point
+infinitesimally below $h$ on a decreasing edge. Consider $\cut(\MM,\cont(y)) = \{\MM_u,\MM_\ell\}$ (the upper and lower manifolds). Then $\onestep(\MM)$
+outputs $\reeb(\MM_u)$ and implicitly gives a valid painting of $\MM_\ell$.
+\end{theorem}
+
+We have an important lemma that helps to bound the running time.
+
+\begin{lemma} \label{lem:tree} For every color $c$, all points in $T(c)$ lie along a monotone path in 
+$\reeb(\MM)$.
+\end{lemma}
+
+\begin{proof} Prove by induction of the number of edges in $\reeb(\MM)$. Consider the head $h$ of $\st$ in the first call to $\onestep(\MM)$,
+and the contour $\cont(y)$ as defined in \Thm{piece}. Take $\cut(\MM,\cont(y)) = \{\MM_u,\MM_\ell\}$. Apply induction
+hypothesis on $\MM_\ell$, and use that to argue about $T(c_1)$ and $T(c_2)$ in the original manifold.
+\end{proof}
+
 \end{document}

### 2014-06-27 paint-contour-tree.tex
Many changes, mostly before section 1.2

@@
 
-\title{Follow the Flow of Rain and Paint: An Instance Optimal Contour Tree Algorithm}
+\title{Follow the Flow of Rain and Paint: \break An Output Sensitive Contour Tree Algorithm}
 \date{}
@@
 
-The ``status" of an internal point is decided by the gradient at that point. Points of zero gradient are deemed \emph{critical};
+We define the size of the manifold $\MM$, denoted $|\MM|$, to be the number of faces, edges, and vertices.  
+Let $n$ denote the number of vertices.  
+Since we are working with $2$-manifolds, the size of the manifold is $\Theta(n)$.
+
+For smooth manifolds the ``status" of an internal point is decided by the gradient at that point. 
+Points of zero gradient are deemed \emph{critical};
 all other points are regular.
 For PL manifolds, we need a discrete version of this definition.
-For vertex $v$, we use $\nbd(v)$ to denote the neighborhood of $v$ in $M$. Let $\outnbd(v) := \{ w | f(v) > f(w), w \in \Gamma(v)\}$
+First, all non-vertex points are considered regular.
+For vertex $v$, we use $\nbd(v)$ to denote the neighborhood of $v$ in $G(\MM)$. Let $\outnbd(v) := \{ w | f(v) > f(w), w \in \Gamma(v)\}$
 and $\innbd(v) := \{w | f(v) < f(w), w \in \Gamma(v)\}$. 
@@
 \smallskip
-Because $\MM$ is PL, the gradient at all non-vertex points of $\MM$ is zero.
-So all non-vertex points are regular. We will assume that $f$ is Morse, so the function value on all critical points is distinct. We will assume the stronger condition that all internal
-vertices have distinct function values.
+%Because $\MM$ is PL, the gradient at all non-vertex points of $\MM$ is zero.
+%So all non-vertex points are regular.
+
+We will assume that $f$ is Morse, so the function value on all critical points is distinct. 
+We will also assume the stronger condition that all internal vertices have distinct function values.
+Note that assuming $f$ is Morse implies that $\MM$ has no multi-saddles 
+(i.e. $\outnbd(v)$ and $\innbd(v)$ can each have at most two non-contiguous components).
 A value $t$ is \emph{non-critical} if there exists no critical point $x$ such that $f(x) = t$.
@@
 
-Since $\MM$ is PL, if there is a descending path from $x$ to $y$, it must be PL. So we will always refer to
+Since $\MM$ is PL, if there is a descending path from $x$ to $y$ then it has PL representation. So we will always refer to
 PL descending paths.
 
-\begin{definition} \label{def:cont} A \emph{contour} in $\MM$ is a continuous function $\phi:\SSS^1 \mapsto \MM$
-such that $\forall a, b \in \SSS^1$, $f(\phi(a)) = f(\phi(b))$. This is called a \emph{$t$-contour} if $f(\lambda(a)) = t$.
-A contour is \emph{simple} if $\forall a, b \in \SSS^1$, $\lambda(a) \neq \lambda(b)$. 
+\begin{definition} \label{def:cont} A \emph{contour} in $\MM$ is the image of a continuous and injective function $\phi:\SSS^1 \mapsto \MM$ such that $\forall a, b \in \SSS^1$, $f(\phi(a)) = f(\phi(b))$.
+This is called a \emph{$t$-contour} if $f(\phi(a)) = t$.
+%A contour is \emph{simple} if $\forall a, b \in \SSS^1$, $\lambda(a) \neq \lambda(b)$. 
 \end{definition}
 
-Just as in the case of paths, contours are also PL and have a corresponding discrete representation.
+Note that just as in the case of paths, contours are also PL and have a corresponding discrete representation.
 
-\begin{claim} \label{clm:cont} If two distinct contours intersect, then the intersection point
-is critical.
-\end{claim}
+\begin{remark}
+For contours which pass through critical points, our definition of a contour differs from the ``standard" definition (i.e. contours which are the result of a horizontal cutting).
 
-\begin{proof} Consider the intersection point $x$. It cannot be internal to a face, otherwise $\MM$ self-intersects.
-By the distinctness of the contours, there must be $4$ faces incident to $x$, where each face
-has vertex with a larger function value.
-It cannot be internal to an edge, otherwise this edge would participate in $4$ faces. So it must be a vertex. 
-These $4$ faces lead to at least $2$ neighbors of $x$ with higher function value. One can also show the existence of $2$
-interleaved neighbors with lower function value. So $x$ is critical.
+In particular, the injective requirement does not allow contours through maxima or minima.
+We can add these single point contours at the maxima and minima to our set of contours (though this is not really necessary).
+\end{remark}
+
+\begin{lemma} \label{lem:cont} If two distinct contours intersect, then their intersection is a saddle point.
+\end{lemma}
+
+\begin{proof}
+Consider any (closed) triangular face $f$, interior to the $2$-manifold.
+The set of points on this face at a given height $h$ forms a closed line segment, call it $l_f$.
+Let $\phi$ be any contour that intersects some point in the interior of this face at height $h$.
+Then $\phi$ must contain all of $l_f$, since otherwise continuity or simplicity would be violated.
+
+So consider two distinct intersecting contours $\phi_1$ and $\phi_2$.
+Since they are distinct, there must be a point on the boundary of their intersection, i.e.
+any (arbitrarily small) ball centered at this point contains contour points both inside and outside the intersection of the contours.
+Call this point $p$.
+
+If $p$ lies in the interior of a face $f$ then both $\phi_1$ and $\phi_2$ must contain the entire above mentioned closed lines segment $l_f$, and hence $p$ was not on the intersection boundary. 
+Similarly suppose that $p$ is in the interior of an edge (clearly such an edge would need to be interior to the manifold for $\phi_1$ and $\phi_2$ to be contours).
+Since this is a $2$-manifold each edge is adjacent to exactly two faces.
+Let $l_a$ and $l_b$ be the above described closed line segments on each face.
+Then again $\phi_1$ and $\phi_2$ contain all of $l_a$ and $l_b$, and $p$ was not on the intersection boundary.
+
+So $p$ must be a vertex $v$ of the manifold.
+Let $\phi$ be any contour through $v$.
+There must be two distinct faces $f_a$ and $f_b$ adjacent to $v$ which $\phi$ intersects.
+Any face adjacent to $v$ has exactly two bounding edges adjacent to $v$.
+Since faces are linear interpolates of the vertices, and since $\phi$ intersects $f_a$ and $f_b$, for either face one of these two edges must be increasing from $v$ and the other decreasing from $v$.
+Call such a face a middle face.
+
+Now if $v$ is on the boundary of the intersection of $\phi_1$ and $\phi_2$ then we need at least 3 middle faces adjacent to $v$ (since otherwise it again is like the case we $p$ was interior to an edge).
+However, the only vertices the have at least 3 such faces are by definition saddle points.
+As no two vertices are at the same height, this saddle point can be the only point of intersection.
 \end{proof}
 
-Note that there is a unique contour containing a regular point, motivating the following definition.
+%\begin{proof} Consider the intersection point $x$. It cannot be internal to a face, otherwise $\MM$ self-intersects.
+%By the distinctness of the contours, there must be $4$ faces incident to $x$, where each face
+%has vertex with a larger function value.
+%It cannot be internal to an edge, otherwise this edge would participate in $4$ faces. So it must be a vertex. 
+%These $4$ faces lead to at least $2$ neighbors of $x$ with higher function value. One can also show the existence of $2$
+%interleaved neighbors with lower function value. So $x$ is critical.
+%\end{proof}
+
+\begin{corollary}
+There is a unique contour containing a regular point.
+\end{corollary}
+
 \begin{definition} \label{def:p-cont}
-For regular point $x$, $\cont(x)$ denotes the unique contour containing $x$.
+For a regular point $x$, $\cont(x)$ denotes the unique contour containing $x$.
 \end{definition}
 
-By the Morse property of distinct critical values, we get the following important lemma.
+%By the Morse property of distinct critical values, we get the following important lemma.
 
-\begin{lemma} \label{lem:cont} Let $t$ be a non-critical value. The set of $t$-contours is a set
-of disjoint simple PL contours.
-\end{lemma}
+\begin{corollary} \label{cor:cont} Let $t$ be a non-critical value. The set of $t$-contours is a set
+of disjoint PL contours.
+\end{corollary}
 
 We can use contours to classify saddle points. For a saddle point, we call two increasing (or decreasing) neighbors \emph{opposite}
-if they lie in disjoint part of $\Gamma^+(w)$ (or $\Gamma^-(w)$).
+if they lie in disjoint parts of $\Gamma^+(w)$ (or $\Gamma^-(w)$).
 
-\begin{definition} \label{def:merge} Consider two increasing (resp. decreasing) opposite edges incident to saddle point $x$,
+In the following definition and, unless otherwise stated, throughout the rest of this writeup $\eps$ is some infinitesimally small amount
+(in particular smaller than the minimum height difference between vertices in the manifold).
+
+\begin{definition} \label{def:merge} Consider two increasing (resp. decreasing) opposite edges incident to a saddle point $x$,
 and let $y$ and $z$ be points on these edges such that $f(y) = f(z) = f(x) + \eps$ (resp. $=f(x) - \eps$).
@@
 
-\begin{claim} \label{clm:saddle} Every saddle is either a merge or a split, but not both.
+As $f$ is a Morse function over a $2$-manifold, we have the following.
+
+\begin{claim} \label{clm:saddle} 
+  \begin{compactenum}[(1)]
+    \item Every saddle is either a merge or a split, but not both.
+    \item Every saddle has exactly two distinct contours which contain it.
+  \end{compactenum}
 \end{claim}
@@
 
-\begin{definition} \label{def:cont-hom} Two simple contours $\phi_0, \phi_1$ are \emph{L-homotopic}
+\begin{definition} \label{def:cont-hom} Two contours $\phi_0, \phi_1$ are \emph{L-homotopic}
 if: there exists a continuous function $H:\SSS^1 \times [0,1] \mapsto \MM$
@@
 
-\begin{claim} \label{clm:non-hom} Consider a critical point $x$. Take an increasing edge incident to $x$
+\begin{claim} \label{clm:non-hom} Consider a saddle point $x$. Take an increasing edge incident to $x$
 and let $y$ be a point on this edge within an $\eps$-ball of $x$. Similarly, take a decreasing edge
-and point $z$ on this edge. The contours through $y$ and $z$ are not L-homotopic.
+and a point $z$ on this edge. $\psi(y)$ and $\psi(z)$ are not L-homotopic.
 \end{claim}
 
-\begin{proof} If the contours are L-homotopic, the homotopy must ``pass through" a contour $\phi$ at $x$.
-There are two distinct simple contours through $x$, and there are three possible choices for $\phi$. None of these
-choices can work.
+\begin{proof}
+Without loss of generality assume that $x$ is a merge vertex.  By \Clm{saddle} there are exactly two distinct contours, 
+$\phi_1$ and $\phi_2$ through $x$.  Let $E_1$ and $E_2$ be the sets of manifold edges that are not adjacent to $x$ and that $\phi_1$ and $\phi_2$ intersect, respectively.
+Observe that as contours are by definition non-trivial, both $E_1$ and $E_2$ must non-empty.  Moreover since $\phi_1$ and $\phi_2$ only intersect at $x$, $E_1\cap E_2 = \emptyset$
+
+As $x$ is a merge vertex, $\psi(y)$ cannot intersect all edges in $E_1 \cup E_2$.
+On the other hand since $x$ cannot be both a merge and a split by \Clm{saddle}, $\psi(z)$ must intersect all of $E_1 \cup E_2$.  As $x$ is the only vertex in between $z$ and $y$ 
+and no edge in $E_1\cup E_2$ is adjacent to $x$, it is not hard to see that by the continuity requirement $\psi(z)$ cannot be L-homotopic to $\psi(y)$.
+%
+%
+% Without loss of generality assume that $x$ is a merge vertex.  By \ref{clm:saddle} there are exactly two distinct contours, 
+% $\phi_1$ and $\phi_2$ through $x$.  Observe that by continuity of L-homotopy, if $\psi(z)$ is $L$-homotopic to $\psi(y)$, 
+% then $\psi(z)$ must L-homotopic to either $\phi_1$ or $\phi_2$, as these are the only two possible contours at the height of $x$ 
+% (in the connected component of the manifold in between $y$ and $z$ and containing $x$).
+% We now show that $\psi(z)$ is not L-homotopic to either $\phi_1$ or $\phi_2$.
+% 
+% First observe that since contours are non-trivial (except at extrema), by continuity $\psi(z)$ cannot be 
+% simultaneously L-homotopic to both contours through $x$.  So suppose $\psi(z)$ is L-homotopic to one of the contours, say $\phi_1$.
+% 
+% As $\eps$ is smaller than the minimum height difference of two vertices in the manifold, 
+% and $z$ is $\eps$ lower than $x$, there are no vertices in between $z$ and $x$.  In particular, whatever edges $\phi_1$ and $\phi_2$ intersect (in their interier), 
+% $\psi(z)$ must also intersect.
 \end{proof}
 
-\begin{definition} \label{def:hom-mono} Two simple contours are $\phi_0, \phi_1$ are \emph{monotonically L-homotopic}
+\begin{definition} \label{def:hom-mono} Two contours $\phi_0, \phi_1$ are \emph{monotonically L-homotopic}
 if the $L$-homotopy $H$ has the following property. Consider function $\alpha:[0,1] \mapsto \RR$
@@
 
-\begin{claim} \label{clm:mono} If two simple contours are L-homotopic, then they are monotonically L-homotopic.
+\begin{claim} \label{clm:mono} If two contours are L-homotopic, then they are monotonically L-homotopic.
 \end{claim}
@@
 By continuity of $H$, the function $\alpha$ is continuous. Hence, if it is not strictly
-increasing or decreasing, $\alpha$ attains its maximum at some $a \in (0,1)$. There exist $a_1 = a - \eps_1$
-and $a_2 = a + \eps_2$ (for small appropriate small values $\eps_1, \eps_2$) such that $\alpha(a_1) = \alpha(a_2)$.
-Furthermore, $H(\SSS^1,a_1)$ and $H(\SSS^1,a_2)$ intersect (and are distinct). By \Clm{cont}, $\alpha(a_1)$
-is a critical value. 
+increasing or decreasing, $\alpha$ attains a local maximum at some $a \in (0,1)$. So there exist $a_1 = a - \eps_1$
+and $a_2 = a + \eps_2$ such that $\alpha(a_1) = \alpha(a_2)$.
+
+By continuity, $\eps_1$ and $\eps_2$ can be chosen small enough such that there are no vertices 
+in between $f^{-1}(H(\SSS^1,a_1))$ and the local maximum. 
+Now the local maximum at $a$ may contain a vertex, but by \Clm{non-hom} this vertex cannot be a saddle.
+Therefore $H(\SSS^1,a_1)$ and $H(\SSS^1,a_2)$ must intersect the same set of edges and therefore cannot be distinct.
 \end{proof}
@@
 Take the edge cut by $\phi$ whose upper endpoint $x$ has minimum value, and let $e$
-denote the portion about $\phi$. Then $\phi$ is L-homotopic to the contour
-through any point in this interior of $e$. If $x$ is regular, then $\phi$ is L-homotopic
+denote the portion above $\phi$. Then $\phi$ is L-homotopic to the contour
+through any point in the interior of $e$. If $x$ is regular, then $\phi$ is L-homotopic
 to the contour through $x$.
@@
 
-\begin{proof} One should able to explicit construct the L-homotopy as a piecewise linear map.
+\begin{proof} 
+Let $a$ be the height of $\phi$ and $b$ the height of $x$. Consider the (potentially disconnected) 
+manifold obtained by removing all manifold points above $b$ and below $a$.  
+Specifically consider connected component that contains $\phi$.  
+By continuity of L-homotopy the claim holds iff it holds on this manifold. 
+However, this manifold contains no vectices except at the ends, and therefore
+an L-homotopy can be constructed using the natural piecewise linear map.
 \end{proof}
@@
 \begin{asparaenum}
-	\item Insert the vertices of $\phi$ in $\MM$.
+%	\item Insert the vertices of $\phi$ in $\MM$.
 	\item Take an arbitrary edge $e$ intersecting $\phi$.
 	\item Take a point $x$ (resp. $y$) on $e$ at distance $\eps$ above (resp. below) $\phi$.
-	\item Let $\phi_x$ be the contour through $x$, and analogously define $\phi_y$. Insert the vertices
-	of these contours in the manifold. Triangulate all new faces.
+	\item Let $\phi_x$ be the contour through $x$, and analogously define $\phi_y$. Insert a vertex
+	at each intersection point of these contours with edges of the manifold. Triangulate all new faces.
 	\item Delete the vertices (and incident edges, faces, etc.) of $\phi$ from the manifold.
@@
 
-\begin{claim} \label{clm:inter} For any $x$, the interface of $\wet(x,\MM)$ is a set of disjoint, simple, non-regular contours.
+Note that $\wet(x,\MM)$ (and its interface) can easily be computed in linear time in the size of the wet sub-manifold, using a descending BFS from $x$.
+
+\begin{claim} \label{clm:inter} For any $x$, the interface of $\wet(x,\MM)$ is a set of disjoint, non-regular contours.
 Furthermore, each such contour contains a merge vertex.
@@
 
-If two simple contours intersect, the intersection point $y$ is critical. We can argue that the only
+If two contours intersect, the intersection point $y$ is critical. We can argue that the only
 way to reach $y$ from a descending path is reach one of the contours and then follow it to get to $y$.
@@
 
-\begin{lemma} \label{lem:rain-time} The total running time of $\rain(x,\MM)$ is $O(n)$.
+The procedure $\rain$ creates new boundary vertices on the edges it cuts each time it call the 
+subroutine $\cut$.  The key to understanding the running time of this procedure is to bound the 
+number of newly created vertices, for which we have the following lemma.
+
+\begin{lemma}\label{lem:new-verts}
+The total number of boundary vertices created by a call to $\rain(x, \NN)$ is $O(n)$, where $n=|\NN|$
 \end{lemma}
-
-\begin{proof} Consider some invocation to $\rain(x,\NN)$. The total running time to determine
-$\wet(x,\NN)$ (and the interface) is at most the number of vertices in $\wet(x,\NN)$. Note that
-this includes original vertices in $\NN$ as well as new vertices created by constructing
-$\phi_1, \phi_2, \ldots, \phi_k$. Essentially, we can bound the total time by $O(n)$ plus
-the total sizes of all the interface boundaries ever constructed (during recursive calls).
-
-To bound the latter quantity, we will simply argue that any edge contains at most
-one boundary vertex. Consider some edge $e$ and take the first time a boundary vertex is created in $e$.
+\begin{proof}
+As the number of edges is $O(n)$, arguing that there is at most one boundary vertex created along any given edge will imply the claim.
+Consider some edge $e$ and take the first time a boundary vertex is created in $e$.
 This happens when some contour $\phi$ intersects $e$. On applying $\cut(\NN,\lift(\phi))$,
@@
 
-\begin{claim} \label{clm:rain-reeb} Given $\reeb(\MM_1), \reeb(\MM_2), \ldots, \reeb(\MM_r)$,
-$\reeb(\MM)$ can be computed in $O(n)$ time.
+As each original vertex in $\NN$ appears in only one of the sub-manifolds output by $\rain(x,\NN)$, we have the following corollary.
+
+\begin{corollary}
+The total size of a sub-manifolds output by $\rain(x,\MM)$ is $O(n)$.
+\end{corollary}
+
+\begin{corollary} \label{cor:rain-time} The total running time of $\rain(x,\MM)$ is $O(n)$.
+\end{corollary}
+
+\begin{proof} Consider some invocation to $\rain(x,\MM)$. The only non-trivial work done in $\rain$ is by the calls to $\wet$.
+We know that each invocation to $\wet$ runs in time linear 
+to the portion of the manifold which is wet (including the new boundary vertices created).  
+As the output of $\rain$ is precisely the distinct sub-manifolds wet by each distinct call to $\wet$, the claim follows.
+\end{proof}
+
+Consider the set of manifolds output by $\rain(x,\MM)$, there is a natural tree ordering defined over these manifolds where the 
+root corresponds the first manifold that was wet when we rained from $x$, and its children correspond to the recursive calls 
+(i.e. where we rained from the interface contours of the first raining).  Call this the \emph{Manifold Ordering Tree}.
+$\rain$ can easily be modified to also output this tree (without affecting the asymptotic running time).
+
+%Consider one of the manifolds $\MM_i$ output by $\rain$.  This manifold has a set of boundary maxima/minima that are the result 
+%of cutting along interface contours.  For each such contour there is exactly one other output manifold $\MM_j$ for which 
+%the contour appears as a boundary maxima/minima.
+
+\begin{claim} \label{clm:rain-reeb} Let $\reeb(\MM_1), \reeb(\MM_2), \ldots, \reeb(\MM_r)$, 
+be the output of $\rain(x,\MM)$. Suppose that for each reeb graph we know its corresponding vertex in the Manifold Ordering Tree. 
+Then $\reeb(\MM)$ can be computed in $O(n)$ time.
 \end{claim}
 
-\begin{proof} Basically apply the $\surgery$ procedure, and invoke \Lem{surgery}.
+\begin{proof}
+We will walk through the Manifold Ordering Tree in a leaf first ordering.  Each time we visit a node we connect its reeb graph to the 
+reeb graph of its children in the tree using the $\surgery$ procedure.
+This takes $O(n)$ time overall by \Lem{new-verts} and since $\surgery$ takes constant time (given the appropriate vertices in each reeb graph).
+The correctness follows by induction and  \Lem{surgery}.
 \end{proof}

### 2014-06-28 paint-contour-tree.tex
small changes to section 3

@@
 this is simply an arbitrary bijection $\chi:S \mapsto [|S|]$. Imagine there being a large
-can of paint of color $c(y)$ at maxima $y$. We will spill different paint from each maximum and watch it flow down.
+can of paint of color $\chi(y)$ at maxima $y$. We will spill different paint from each maximum and watch it flow down.
 This is analogous to the raining in the previous section, but paint is a much more viscous liquid.
@@
 %
-\begin{definition} \label{def:paint} A  \emph{painting} of $\MM$ is a map $\chi:V \cup E \mapsto [|S|]$
+\begin{definition} \label{def:paint} A  \emph{painting} of $\MM$ is a map $\chi:S \cup E \mapsto [|S|]$
 with the following properties. (Remember that $G(\MM) = (V,E)$.)
 \begin{asparaitem}
-	\item The restriction $c:S \mapsto [|S|]$ is a bijection.
+	\item The restriction $\chi:S \mapsto [|S|]$ is a bijection.
 	\item Consider an edge $e$. There exists a descending path from some maximum $x$ to $e$
-	consisting of edges in $E$, such that all edges and vertices along this path have the same color. (The maximum $x$
+	consisting of edges in $E$, such that all edges along this path have the same color as $x$. (The maximum $x$
 	is well-defined to be the unique maximum with the same color as $e$.)
@@
 %
-Note that paintings are not unique. It is straightforward to construct a painting of $\MM$ in $O(n)$ time, using a descending BFS from each maximum.
+Note that paintings are not unique. It is straightforward to construct a painting of $\MM$ in $O(n)$ time, using a descending BFS from each maximum that does not explore previously colored edges.
 
 \subsection{The data structures} \label{sec:struct}
-
-{\bf The binomial heaps $\touch(c)$:} For each color $c$, we store a binomial
-heap $T(c)$ of non-extremal critical points.
 
@@
 %
-We will store each set $\touch(c)$ in a binomial heap, keyed by the heights. 
+%We will store each set $\touch(c)$ in a binomial heap, keyed by the heights. 
 Abusing notation, $\touch(c)$ refers
 both to the set and the data structure used to store it.
+
+\medskip
+\noindent
+{\bf The binomial heaps $\touch(c)$:} For each color $c$, we store a binomial
+heap $T(c)$ of non-extremal critical points, keyed by vertex heights.
 
@@
 	\item Consider $x, y \in \st$ such that $y$ was pushed on $x$. There exists $c \in \col(x) \setminus
-	\mcol(x)$ such that $c \in \col(y)$.
+	\mcol(x)$ such that $x$ is not highest in $T(c)$ but $y$ is highest in $T(c)$.
 \end{asparaitem}
@@
 		\item If $x = h$, put $c'$ in $\mcol(h)$. Otherwise ($x \neq h$), push $x$ onto stack
-		and set $\mcol(x) = \{c'\}$.
+		and add $c'$ to $\mcol(x)$.
 		\item Let $h$ be new head of $\st$.
@@
 	\begin{compactenum}
-		\item Connect $h$ (in $\reeb(\MM)$) to $\h(C)$ and the unique minimum corresponding to split $h$. 
-		\item Delete $h$ from $T(c)$, set $\h(C) = h$.
+		\item Connect $h$ (in $\reeb(\MM)$) to $\h(c)$ and the unique minimum corresponding to split $h$. 
+		\item Delete $h$ from $T(c)$, set $\h(c) = h$.
 		\item End procedure.

### 2014-06-29 paint-contour-tree.tex
Minor section 2 changes

@@
 
-\begin{proof} \Clm{mono} should help us argue that for distinct $\phi, \phi' \in \Phi$,
+\begin{proof} \Clm{mono} and continuity of L-homotopy immediately imply that for $\phi, \phi' \in \Phi$,
 for any $t \in [f(\phi),f(\phi')]$, there exists $\phi'' \in \Phi$ such that $f(\phi'') = t$.
-So $f(\Phi)$ is an interval. \Clm{reg} should tell us that this interval must be open
-with critical value endpoints.
+So $f(\Phi)$ is an interval. As equivalence classes are defined over only regular contours, \Clm{reg} 
+tells us that this interval must be open with critical value endpoints.
 \end{proof}
@@
 
-\begin{definition} \label{def:reeb} The \emph{Reeb Graph} $\reeb(\MM)$ has as vertex set
+\begin{definition} \label{def:reeb} The \emph{Reeb Graph} $\reeb(\MM)$ has as a vertex set
 the set of critical points. The edge $(x,y)$ is present if there exists equivalence class
-$\Phi$ such that $f(\Phi) = [f(x),f(y)]$.
+$\Phi$ such that $f(\Phi) = (f(x),f(y))$.
 \end{definition}
@@
 %	\item Insert the vertices of $\phi$ in $\MM$.
-	\item Take an arbitrary edge $e$ intersecting $\phi$.
+	\item Take an arbitrary edge $e$ whose interior intersect $\phi$.
 	\item Take a point $x$ (resp. $y$) on $e$ at distance $\eps$ above (resp. below) $\phi$.
@@
 \medskip
-This results in a new manifold that is also boundary critical. There are two new disjoint
-boundary faces added. The manifold may be disconnected. Abusing notation, we use $\cut(\MM,\Phi)$
-for a set $\Phi$ of contours to mean the manifold obtained by cutting along all contours in $\Phi$.
+This results in a new manifold which is boundary critical if the input manifold was. There are two new disjoint
+boundary faces added. The manifold may be disconnected. Abusing notation, we use $\cut(\MM,\Psi)$
+for a set $\Psi$ of contours to mean the manifold obtained by cutting along all contours in $\Psi$.
 
@@
 \begin{definition} \label{def:wet} The set of points $y$ such that there is a descending path from $x$ to $y$
-is denoted by $\wet(x,\MM)$. A point $z$ is on the \emph{interface} of $\wet(x,\MM)$ if every neighbor
+is denoted by $\wet(x,\MM)$. A point $z$ is on the \emph{interface} of $\wet(x,\MM)$ if every neighborhood of $z$
 has non-trivial intersection with $\wet(x,\MM)$.
@@
 
-\begin{proof} If $y \in \wet(x,\MM)$, then any contour containing $y$ is also in $\wet(x,\MM)$.
-Consider a $y$ in the interface. We argue that all points in a contour $\phi$ containing $y$
-are also in the interface. If such a contour is regular, we can find a contour just above it that
-is also wet.
+\begin{proof}
+ First observe that any point which in the interface must be wet.  Second, for $y \in \wet(x,\MM)$, all the points in any contour containing $y$ are also in $\wet(x,\MM)$, 
+ as one can follow the descending path to $y$ and then walk along the contour.
 
-If two contours intersect, the intersection point $y$ is critical. We can argue that the only
-way to reach $y$ from a descending path is reach one of the contours and then follow it to get to $y$.
-But that would imply the existence of another critical point on this contour, violating the Morse condition.
+ So let $y$ be a point in the interface.
+ For $\eps$ sufficiently small, the points in any $\eps$ neighborhood of $y$ that lie below $y$ must have a descending path from $y$ and hence must be wet.
+ Therefore in order for $y$ to have non-trivial intersection with $\wet(x,\MM)$ for any arbitrarily small $\eps$ there must be a dry point which lies 
+ strictly above $y$.  In particular, this implies there must be an entire dry contour which lies just above $y$.
 
-Let us argue that $y$ is a merge vertex. Consider the descending path from $x$ to $y$. The final edge
-that reaches $y$ is obviously all wet, and is an increasing edge from $y$. Consider an opposite increasing
-edge. Take two point in these edges that are infinitesimally higher than $y$ and show that the contours
-through them as disjoint. So $y$ is a merge.
+ However, since $y$ is in the interface, it is also wet.  Therefore for any contour $\phi$ which contains $y$, for any arbitrarily small $\eps$ there is 
+ a wet point which lies above $\phi$ and has a descending path of length at most $\eps$ to $\phi$.
+ This implies there is some contour which lies just 
+ above $y$ which is wet.  So any contour $\phi$ containing $y$ has two contours which lie just above it, one which is dry, and one which is wet.  
+ This can only happen if $\phi$ contains a merge vertex, because otherwise all contours which lie just above would be in the same L-homotopy equivalence class and so by 
+ \Clm{mono}, if one is wet, all below it are wet.
+
+ By similar logic one can argue that of the two contours through a given merge vertex at most one can contain interface points other than the merge vertex, 
+ and that if one of the contours does contain such interface points, then the entire contour consists of interface points.
 \end{proof}
 
+% \begin{proof} If $y \in \wet(x,\MM)$, but not in the interface, then any contour containing $y$ is also in $\wet(x,\MM)$.
+% So consider a point $y$ in the interface. We argue that all points in a contour $\phi$ containing $y$
+% are also in the interface. If such a contour is regular, we can find a contour just above it that
+% is also wet. 
+% 
+% If two contours intersect, the intersection point $y$ is critical. We can argue that the only
+% way to reach $y$ from a descending path is reach one of the contours and then follow it to get to $y$.
+% But that would imply the existence of another critical point on this contour, violating the Morse condition.
+% 
+% Let us argue that $y$ is a merge vertex. Consider the descending path from $x$ to $y$. The final edge
+% that reaches $y$ is obviously all wet, and is an increasing edge from $y$. Consider an opposite increasing
+% edge. Take two point in these edges that are infinitesimally higher than $y$ and show that the contours
+% through them as disjoint. So $y$ is a merge.
+% \end{proof}
+
 We will define a \lift{} operation on the interface contours. Consider such a contour $\phi$ containing
-merge vertex $y$. Take any dry increasing edge incident to $y$, and pick the point $z$ on this edge at height
+a merge vertex $y$. Take any dry increasing edge incident to $y$, and pick the point $z$ on this edge at height
 $f(y) + \eps$. Let $\lift(y) = \cont(z)$.
@@
 \begin{claim} \label{clm:cut-int} Let $\phi$ be an interface contour. Then $\cut(\MM,\lift(\phi))$
-results in two disjoint manifolds, one consisting entirely of dry points, and the other containing
-all wet points.
+results in two disjoint manifolds, one consisting entirely of dry points.
 \end{claim}
 
-\begin{proof} Note that $\lift(\phi)$ is a closed, dry curve. But the Jordan Curve Theorem,
+\begin{proof} Note that $\lift(\phi)$ is a closed, dry curve. By the Jordan Curve Theorem,
 $\cut(\MM,\lift(\phi))$ results in two disjoint manifolds. Let $\NN$ be the manifold containing
-the point $x$, and $\NN'$ be the other manifold. If there exists a wet point $z$ in $\NN'$,
-then there exists a descending path from $x$ to $z$ intersecting $\phi$ (Jordan Curve Theorem again).
-That contradicts the fact that $\phi$ is dry.
+the point $x$ that we called $\wet(x,\MM)$ from, and let $\NN'$ be the other manifold. 
+If there exists a wet point $z$ in $\NN'$, then there exists a descending path from $x$ to $z$. 
+As $x$ is in $\NN$ and $z$ is in $\NN'$, again by the Jordan Curve Theorem this path must intersect $\phi$, 
+contradicting the fact that $\phi$ is dry.
 \end{proof}
@@
 	\begin{compactenum}
-		\item Construct $\cut(\phi'_i,\NN_1)$ to get the dry manifold $\LL_i$ and wet manifold $\NN_{i+1}$.
+		\item Call $\cut(\phi'_i,\NN_1)$ to get the dry manifold $\LL_i$ and wet manifold $\NN_{i+1}$.
 		\item Let the new boundary of $\LL_i$ be $B_i$. Invert $\LL_i$ so that $B_i$ is a maximum. Recursively

### 2014-07-30 paint-contour-tree.tex
forgot to stage file for commit

@@
 \newcommand{\Ex}[1]{\hyperref[ex:#1]{Example~\ref*{ex:#1}}} %example
+\newcommand{\Obs}[1]{\hyperref[obs:#1]{Observation~\ref*{obs:#1}}} %observation
 
@@
 \newcommand{\AQ}{\mathsf{AQ}}
+
+\newcommand{\etal}{\textit{et~al.}\xspace}
 
@@
 
-\author{}
+\author{
+  Seshadhri Comandur
+  \and
+  Kashyap Dixit
+  \and
+  Benjamin Raichel
+}
 
 \title{Follow the Flow of Rain and Paint: \break An Output Sensitive Contour Tree Algorithm}
-\date{}
+%\date{}
 
@@
 \maketitle
+
+\begin{abstract}
+Reeb graphs are a fundamental topological structure which tracks the evolution of levels sets of a real valued function over a manifold.
+In the case when this graph is loop free it is called a contour tree.
+Due to the use of the contour trees in practice on large data sets, there has been significant previous work on algorithms for their efficient computation.
+
+Some of these previous results are describe as optimal, however they are only worst case optimal.
+Here we present an output sensitive approach to computing contour trees.
+Our approach leads to a faster algorithm when the contour tree is nicely behaved (i.e.
+more balanced), yet is still optimal in the worst case.
+Our approach is novel and we believe will lead to faster algorithms for other applications as well.
+\end{abstract}
+
+\section{Introduction}
+Describe what reeb graphs and contour trees are.
+
+\medskip
+\noindent
+Explain why useful:  Lets you know the ``shape of your date''.
+Map of the critical points where contours change.
+Fast extraction of seed sets in a small sized structure, which can let you do threshold queries.
+
+\medskip
+\noindent
+Paragraph on specific applications.  Scientific visualization, in particular combustion research.  GIS.
+
+\subsection{Previous Work}
+
+\paragraph{Contour Trees.}
+As the application areas of contour trees usually involve very large data sets, highly efficient algorithms are not only desirable but often necessary.
+As such there have been a long sequence of papers with improved running times for computing contour trees.
+Kreveld \etal \cite{kobps-ctsssit-97} presented an $O(N\log N)$ time algorithm for functions over 2D meshes and an $O(N^2)$ algorithm for higher dimensions, where $N$ is the number of simplices in the mesh.
+Tarasov and Vyalya \cite{tv-cct-98} improved the running time to $O(N\log N)$ for the 3D case by making several passes over the input.
+
+Carr \etal \cite{csa-cctad-00} improved the running time for all dimensions to $O(n\log n + N)$, where $n$ is the number of mesh vertices.
+Their algorithm works in 3 simple stages.
+The first stage consist of a top down sweep to compute what is known as the merge (or join) tree, which tracks the evolution of super-level sets (whereas the contour tree tracks level sets).
+Similarly, the second stage does a bottom up sweep to compute the split tree, which tracks sub-level sets.
+Their main contribution is the third and final step where
+they present an $O(n)$ time algorithm to merge together the merge and split trees to get the contour tree.  Chiang \etal \cite{cllr-sooscctmp-05} 
+have since
+
+%The most expensive part of the Carr \etal algorithm comes from sorting.  They sort all vertices of the mesh to facilitate an easing sweeping of the mesh.
+%asdfas \cite{} improve the running to $O()$ by sorting only critical points of the mesh.  However, without looking at the non-critical points in the mesh 
+%one does not know how the critical points connect to each other in the contour tree.  To overcome this \cite{} use ``mountain climbers'' from each saddle 
+%to determine which critical points they belong to above.
+
+\paragraph{Output Sensitivity.}
+Let $t$ be the number of critical points in the input mesh.
+There is room to improve the results of Carr \etal, as $t$ could potentially be much smaller than $n$.
+In particular, Cole-McLaughlin and Pascucci \cite{pc-ectls-02} provide an $O(n+t\log n)$ time algorithm for the case of 3 dimensional structured meshes.
+Chiang \etal \cite{cllr-sooscctmp-05} improve upon this result, and provide an $O(N+t\log t)$ which works for structured or unstructured meshes in any dimension.
+
+As the the authors point out in \cite{cllr-sooscctmp-05}, by reduction from sorting, one can show that any algorithm to compute the contour tree in the worst case takes $\Omega(N+t\log t)$ time.
+However, this lower bound only holds when the contour tree is very unbalanced (i.e.
+long and skinny) and therefore contains a sorted path with a large fraction of the vertices.
+Therefore the bound of $\Omega(N+t\log t)$ is only worst case optimal.
+In this paper we will consider an output sensitive approach which takes into account the shape of the contour tree.
+
+\paragraph{Reeb Graphs.} UNDER CONSTRUCTION.
+There have also been a number of result for the more general Reeb graph, i.e.
+when the loopless assumption is removed.
+Shinagawa and Kunii \cite{sk-crgacs-91} provided the first provably correct algorithm for the case of triangulated $2$-manifolds.
+Cole-McLaughlin \etal \cite{cehnp-lrbm-03} later improved the running time to $O(n\log n)$.
+
+\subsection{Our Approach and Contribution.}
+
+\paragraph{To Sort or not to Sort.}
+As discussed above, the algorithm of Carr \etal \cite{csa-cctad-00} computes the contour tree by first constructing the merge and split trees. 
+As vertices appear in sorted order in any root to leaf path in either tree, sorting the input vertices is a natural step.
+However, not all vertices must be sorted together.
+In particular, Chiang \etal \cite{cllr-sooscctmp-05} improved the running time by observing that as the merge and split tree only consists of critical points, it suffices to only sort critical points.
+%
+%However, without looking at the non-critical points in the mesh 
+%one does not know how the critical points connect to each other in the contour tree.
+\cite{} easily overcome this by sending out increasing monotone paths (i.e.
+``mountain climbers'') from each critical point to determine which critical points they connect to above.
+%
+If a large fraction of the vertices appear in a single root to leaf path in the merge or split tree, paying to sort all the critical 
+points is unavoidable in order to compute the merge and split trees.  However, if the merge and split trees are balanced then sorting all the critical points is potentially very costly.
+
+There is also a more fundamental problem with the now standard approach of first computing the merge and split trees in order to compute the contour tree.
+In particular, this approach can never lead to an algorithm that is optimal with respect to the shape of the contour tree.
+This is because it is possible to have merge and split trees which are long and skinny (i.e.
+costly to compute) whereas the contour tree is balanced.
+For example consider the case when the mesh looks like two (thickened) balanced binary trees, one upside down, and joined at their roots.
+In this case the contour tree is potentially cheap to compute as there is no long sorted path.
+However, the merge and split trees will have long sorted tails (consisting of the vertices in the bottom and top balanced trees, respectively), and therefore require $\Omega(n\log n)$ time to compute.
+
+\paragraph{Our Contribution.} UNDER CONSTRUCTION.
+Give a high level overview.
+Emphasize the spilling partitioning scheme used to avoid sorting, and say it may be useful else where.
+State our running time.
+Say it may be optimal (we still need to investigate).
+It matches previous on worst case (i.e.
+optimal in worst case).
 
@@
 
+\pagebreak
+\section{Lower Bound}
+First we setup some basic terminology.
+Any tree $T$ considered in this section will be a rooted full binary tree where the height of a vertex is its distance 
+from the root $r$.  As such, the children of a vertex $v\in T$ are the adjacent vertices of larger height
+(and $v$ is the parent of such vertices).  Then the subtree rooted at $v$, denoted $T_v$ consists of the graph induced on all 
+vertices which are descendants of $v$ (including $v$ itself).  For two vertices $v$ and $w$ in $T$ let $d(v,w)$ denote the 
+length of the path between $v$ and $w$.
+
+\begin{definition}
+ A coloring of a tree $T$ is any function $\chi:L(T)\times E \rightarrow [n]$, where $L(T)$ is the set of leaves of $T$ and $n = |L(T)|$.  
+ We say a coloring is valid if
+ \begin{itemize}
+  \item The restriction $\chi:L(T) \rightarrow [n]$ is a bijection.
+  \item Let $e$ be any edge in $E$ and let $u$ be its higher endpoint, then $\chi(e) \in \chi(L(T_u))$. 
+ \end{itemize}
+ We consider each internal vertex as having two colors, namely the colors of the two edges for which it is a lower endpoint.
+\end{definition}
+
+For a vertex $v\in T$, let $H_v$ be the subset of 
+vertices along the path from the root to $v$ such that one of the two colors of each vertex is in $\chi(L(T_v))$.
+We will refer to $H_v$ as the \emph{heap} at $v$.
+Then define the \emph{algorithm} cost function as $f_{alg}(\chi(T)) = \sum_{v\in T} \log |H_v|$.
+
+\begin{lemma}
+\label{lem:algCost}
+NEEDS TO BE WRITTEN:
+Here we need a lemma showing that the running time of our algorithm is $O(f_{alg}(\chi(T)))$, where $T$ is the merge tree of the given input, 
+and $\chi$ is any valid coloring of $T$.  (maybe add one in the definition of $H_v$ so that $\log |H_v|$ is always $>1$).
+\end{lemma}
+
+\subsection{Path Decomposition Background}
+We now introduce the concept of path decompositions and their cost.  In the next subsection we then relate this to the
+running time of our algorithm.
+
+\begin{definition}
+ Let $T$ be a rooted binary tree.  We call a path $p$ in $T$ a \emph{leaf path} if $p$ contains a leaf of $T$ and 
+ is a contiguous sub-path of a root to leaf path in $T$.  A \emph{path decomposition}, $P(T)$, is a partition of $T$ into 
+ a set of vertex disjoint leaf paths.  Note that the number of paths in $P(T)$ must equal the number of leaves in $T$.
+ For $p\in P(T)$, we use $r_p$ to denote the lowest vertex in $p$ (i.e. its root) and 
+ $|p|$ to denote its length.
+
+ We define the cost of a particular path decomposition, $P(T)$, as $f_{path}(P(T)) = \sum_{p\in P(T)} |p| \log |p|$.
+\end{definition}
+
+% \begin{observation}
+% DELETE IF NOT NEEDED:
+% Note that $f_{path}(P(T)) = \Theta( \sum_{v\in V} l_v )$, where $l_v$ is the distance from $v$ to the leaf of its containing path in $P(T)$.  
+% That is, conceptually we can think of each vertex as charging the $\log$ of the length to its path's leaf.
+% \end{observation}
+
+\begin{definition}
+ Let T be a binary tree and let $\chi(T)$ and $P(T)$ be a valid coloring and path decomposition of $T$, respectively.  
+ For $p\in P(T)$, we say that $p$ is \emph{light} if $4 |p| < |H_{r_p}|$, and we say it is \emph{heavy} otherwise.
+\end{definition}
+
+\begin{observation}
+\label{obs:decrease}
+ Let $v$ be a vertex in $T$ and let $w$ be its parent.  Then $|H_w| \geq |H_v| -1$, as $L(T_v) \subseteq L(T_w)$ and the path from 
+ $w$ to the root has one less vertex than the path from $v$ to the root.
+
+ In particular, we have the following more general property.
+ Let $v$ and $u$ be any two vertices in the same root to leaf path of $T$, such that $u$ is lower than $v$.  
+ Then $|H_v| \leq |H_u| + d(u,v)$. 
+\end{observation}
+
+\begin{lemma}
+\label{lem:pathBounds}
+ Let T be a binary tree and let $\chi(T)$ and $P(T)$ be a valid coloring and path decomposition of $T$, respectively.
+ Then, 
+ \begin{compactenum}
+  \item for any heavy path $p\in P(T)$, $\sum_{v\in p} \log |H_v| = O(|p| \log |p|)$.
+  \item for any light path $p\in P(T)$, $\sum_{v\in P} |H_v| = O(|p| \log |H_{r_p}|)$.
+ \end{compactenum}
+\end{lemma}
+\begin{proof}
+ By $\Obs{decrease}$ we know that for any vertex $v\in p$, $|H_v|\leq |H_{r_p}| + |p|$ (as $r_p$ lies below $v$ on $p$).
+ As such, 
+ \[
+ \sum_{v\in p} \log |H_v| \leq \sum_{v\in p} \log(|H_{r_p}| + |p|)  = |p| \log (|H_{r_p}| + |p|).
+ \]
+ If $p$ is a heavy path then $|p| \log (|H_{r_p}| + |p|) = O(|p| \log |p|)$, and if $p$ is light then 
+ $|p| \log (|H_{r_p}| + |p|) = O(|p| \log |H_{r_p}| )$.
+\end{proof}
+
+\begin{definition}
+ We say two paths $p,q \in P(T)$ are adjacent if in $T$ there is an edge between a vertex in $p$ and a vertex in $q$.  
+Note that as $p$ and $q$ are vertex disjoint leaf paths, it follows that this edge must be between the root of either 
+$p$ or $q$ and its parent vertex in $T$.  wlog suppose this edge is between the root of $p$ and its parent in $q$. 
+Then we call this parent vertex the \emph{connection} of $p$ and refer to it as $c_p$.
+\end{definition}
+
+The following lemma can be thought of as a generalization of $\Obs{decrease}$.
+
+\begin{lemma}
+\label{lem:adjacent}
+ Let $T$ be a binary tree and let $\chi(T)$ and $P(T)$ be a valid coloring and path decomposition of $T$, respectively.
+ Let $p$ be any path in $P(T)$ and let $\{q_1, \dots q_k\}$ be the set of all light paths adjacent to $p$.
+
+ Let $H_i$ be the heap at $r_{q_i}$ and let $H_0$ be the heap at $r_p$.  Then $\sum_{i=1}^k |H_i| \leq |H_0|+|p|$.
+ %Then $\sum_{i=1}^k \sum_{v\in q_i} \log |H_v| = O(|p|\log |p|)$
+\end{lemma}
+\begin{proof}
+ %Let $c_1, \dots, c_k$ be the connections of the paths $q_1, \dots, q_k$ to the path $p$, and let $c_0=r_p$.  
+ %Let $d_i = d(c_i, c_{i+1})$, and let $d_k$ be the distance from $c_k$ to the leaf of $p$.  
+ %Observe that $\sum_{i=0}^k d_i = l_p$.
+ %
+ % 
+ %Let $H_i$ be the heap at $r_{q_i}$, and let $H_0$ be the heat to $r_p$.  
+ First observe that for all $i$, $r_{q_i}$ lies above $r_p$ on some root to leaf path. 
+ Let $H_i'$ be the subset of $H_i$ that lies below $r_p$.  Observe that $H_i'\subseteq H_0$, as $L(T_{r_{q_i}})\subseteq L(T_{r_p})$.
+
+ Now for any $i\neq j$, $H_i \cap H_j = \emptyset$, as $r_{q_i}$ and $r_{q_j}$ are not on the same root to leaf path.
+ In particular, $(H_i\setminus H_i') \cap (H_j \setminus H_j') = \emptyset$, and so $\sum_{i=1}^k |H_i\setminus H_i'| \leq |p|$ 
+ as $H_i\setminus H_i'$ is precisely the subset of $H_i$ that lies on $p$.
+ Moreover, for any $i\neq j$, $ H_i' \cap H_j' = \emptyset$, and so since $H_i'\subseteq H_0$ we have $\sum_{i=1}^k |H_i'|\leq |H_0|$.
+ Putting these together gives
+ \[
+  \sum_{i=1}^k |H_i| = \sum_{i=1}^k |H_i'| + \sum_{i=1}^k |H_i\setminus H_i'| \leq |H_0|+|p|.
+ \]
+\end{proof}
+
+\begin{corollary}
+ Let $p$ be a heavy path in $P(T)$ and let $\{q_1, \dots q_k\}$ be the set of all light paths adjacent to $p$.
+ Then $\sum_{i=1}^k \sum_{v\in q_i} \log |H_v| = O(|p|\log |p|)$.
+\end{corollary}
+\begin{proof}
+ Let $H_i$ be as defined in \Lem{adjacent}.  By \Lem{pathBounds}, for all $1\leq i\leq k$, 
+ $\sum_{v\in q_i} \log |H_v| = O(|q_i|\log |H_i|) = O(|H_i|\log |H_i|)$, as $q_i$ is a light path.
+ Now by \Lem{adjacent}, $\sum_{i=1}^k |H_i| \leq |H_0|+|p| = O(|p|)$, as $p$ is a heavy path. 
+ Therefore, as $x\log x$ is a convex function, Jensen's inequality implies the claim.
+
+\end{proof}
+
+\subsection{Lower Bound by Path Decomposition}
+
+\begin{lemma}
+\label{lem:pathDecomp}
+NEEDS TO BE WRITTEN:
+Have a lemma here showing that our algorithm must pay at least the cost of any valid path decomposition.
+\end{lemma}
+
+For a given coloring $\chi$ of a merge tree $T$, we now describe a particular path decomposition such that 
+$f_{alg}(\chi(T)) = \Theta(f_{path}(P(T)))$.  By \Lem{algCost} and \Lem{pathDecomp} this will imply the running time of our algorithm 
+is optimal with respect to the shape of the merge tree.
+
+%We define two cost functions over validly colored trees.  To do so, we first define two quantities for each vertex $v\in T$.  
+%First let $P_v$ be the longest path from $v$ to a descendant leaf of $v$.  Second, let $H_v$ be the subset of 
+%vertices along the path from the root to $v$ such that one of the two colors  is in $\chi(L(T_v))$.
+
+\bibliographystyle{alpha}
+\bibliography{contour}
+
 \end{document}

### 2014-08-01 paint-contour-tree.tex
Small changes, section 5.2 not finished

@@
 
+Now that we have a simple expression upper bounding the running time of our algorithm, we spend the remainder 
+of this section proving that this running time is optimal.
+
 \subsection{Path Decomposition Background}
@@
   \item for any heavy path $p\in P(T)$, $\sum_{v\in p} \log |H_v| = O(|p| \log |p|)$.
-  \item for any light path $p\in P(T)$, $\sum_{v\in P} |H_v| = O(|p| \log |H_{r_p}|)$.
+  \item for any light path $p\in P(T)$, $\sum_{v\in p} \log |H_v| = O(|p| \log |H_{r_p}|)$.
  \end{compactenum}
@@
 
- Let $H_i$ be the heap at $r_{q_i}$ and let $H_0$ be the heap at $r_p$.  Then $\sum_{i=1}^k |H_i| \leq |H_0|+|p|$.
+ For $1\leq i\leq k$ let $r_i = r_{q_i}$ and let $H_i = H_{r_i}$.
+ Then $\sum_{i=1}^k |H_i| \leq |H_p|+|p|$, where $H_p = H_{r_p}$.
  %Then $\sum_{i=1}^k \sum_{v\in q_i} \log |H_v| = O(|p|\log |p|)$
@@
  %Let $H_i$ be the heap at $r_{q_i}$, and let $H_0$ be the heat to $r_p$.  
- First observe that for all $i$, $r_{q_i}$ lies above $r_p$ on some root to leaf path. 
- Let $H_i'$ be the subset of $H_i$ that lies below $r_p$.  Observe that $H_i'\subseteq H_0$, as $L(T_{r_{q_i}})\subseteq L(T_{r_p})$.
-
- Now for any $i\neq j$, $H_i \cap H_j = \emptyset$, as $r_{q_i}$ and $r_{q_j}$ are not on the same root to leaf path.
+ First observe that for all $i$, $r_{i}$ lies above $r_p$ on some root to leaf path. 
+ Let $H_i'$ be the subset of $H_i$ that lies below $r_p$.  Observe that $H_i'\subseteq H_p$, as $L(T_{r_i})\subseteq L(T_{r_p})$.
+
+ Now for any $i\neq j$, $H_i \cap H_j = \emptyset$, as $r_{i}$ and $r_{j}$ are not on the same root to leaf path (i.e. $L(T_{r_i}) \cap L(T_{r_j}) = \emptyset$).
  In particular, $(H_i\setminus H_i') \cap (H_j \setminus H_j') = \emptyset$, and so $\sum_{i=1}^k |H_i\setminus H_i'| \leq |p|$ 
  as $H_i\setminus H_i'$ is precisely the subset of $H_i$ that lies on $p$.
- Moreover, for any $i\neq j$, $ H_i' \cap H_j' = \emptyset$, and so since $H_i'\subseteq H_0$ we have $\sum_{i=1}^k |H_i'|\leq |H_0|$.
+ Moreover, for any $i\neq j$, $ H_i' \cap H_j' = \emptyset$, and so since $H_i'\subseteq H_p$ we have $\sum_{i=1}^k |H_i'|\leq |H_p|$.
  Putting these together gives
  \[
-  \sum_{i=1}^k |H_i| = \sum_{i=1}^k |H_i'| + \sum_{i=1}^k |H_i\setminus H_i'| \leq |H_0|+|p|.
+  \sum_{i=1}^k |H_i| = \sum_{i=1}^k |H_i'| + \sum_{i=1}^k |H_i\setminus H_i'| \leq |H_p|+|p|.
  \]
@@
  $\sum_{v\in q_i} \log |H_v| = O(|q_i|\log |H_i|) = O(|H_i|\log |H_i|)$, as $q_i$ is a light path.
- Now by \Lem{adjacent}, $\sum_{i=1}^k |H_i| \leq |H_0|+|p| = O(|p|)$, as $p$ is a heavy path. 
+ Now by \Lem{adjacent}, $\sum_{i=1}^k |H_i| \leq |H_p|+|p| = O(|p|)$, as $p$ is a heavy path. 
  Therefore, as $x\log x$ is a convex function, Jensen's inequality implies the claim.
@@
 
-For a given coloring $\chi$ of a merge tree $T$, we now describe a particular path decomposition such that 
-$f_{alg}(\chi(T)) = \Theta(f_{path}(P(T)))$.  By \Lem{algCost} and \Lem{pathDecomp} this will imply the running time of our algorithm 
-is optimal with respect to the shape of the merge tree.
-
-%We define two cost functions over validly colored trees.  To do so, we first define two quantities for each vertex $v\in T$.  
-%First let $P_v$ be the longest path from $v$ to a descendant leaf of $v$.  Second, let $H_v$ be the subset of 
-%vertices along the path from the root to $v$ such that one of the two colors  is in $\chi(L(T_v))$.
+For a given valid coloring $\chi$ of a tree $T$, we now describe a particular path decomposition for which 
+we can prove $f_{alg}(\chi(T)) = O(f_{path}(P(T)))$ using the observations about decompositions from the 
+previous subsection.  By the above lemma this will imply $f_{alg}(\chi(T)) = \Theta(f_{path}(P(T)))$.
+
+\begin{definition}
+ Let $\chi(T)$ be a valid coloring of a binary tree $T$.  We define a set of disjoint paths over the vertices of 
+ $T$ as follows.  For each internal vertex $v\in T$ add the edge from $v$ to $\arg \max_{v_l, v_r} \{|H_{v_l}|, |H_{v_r}|\}$ 
+ where $v_l$ and $v_r$ are the children of $v$ (if $|H_{v_l}|=|H_{v_r}|$ then pick one arbitrarily).
+ The corresponding collection of paths that all such edges correspond to is a path decomposition of $T$, which 
+ we refer to as a \emph{maximum} path decomposition, and is denoted by $\mathcal{P}(T)$.  
+\end{definition}
+
+\begin{definition}
+ Let $p$ be a path in a path decomposition $P(T)$.  Call a light path in $P(T)$ reachable from $p$ if there is a path in the tree
+ $T$ from some vertex in $p$ which only moves to vertices of larger height and only vertices which lie in light paths.
+ Call this set of reachable paths the \emph{light path tree} of $p$.  Naturally this set can be viewed as a tree 
+ (i.e. $p$ is the root of this tree and each other path has a well defined depth from $p$).
+\end{definition}
+
+\begin{lemma}
+ Let $\chi(T)$ be a valid coloring of a tree $T$, and let $\mathcal{P}(T)$ be the corresponding maximum path decomposition.
+ Let $p$ be a light path in $\mathcal{P}(T)$ and let $LPT(p)$ be the light path tree of $p$.
+
+ Then $\sum_{q\in L(p)} \sum_{v\in q} \log |H_v| = |H_p|\log |H_p|$.
+\end{lemma}
+\begin{proof}UNDER CONSTRUCTION / NOTES:
+
+ Claim: Let $q$ be some path of depth $i$ in $LPT(p)$, and let $H_q$ be the heap at the root of this path.
+ Then $|H_q|< |H_p|/c^i$ for some constant $c>1$.  This should follow from using a maximum path decomposition.
+
+ This claim would imply the depth of the tree is $O(\log |H_p|)$ and the total number of paths in the tree is $O(|H_p|)$, 
+ thought I am not sure if either of these statements are useful.
+\end{proof}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\subsection{Junk, don't read}
+The following can be viewed as stronger version of \Lem{adjacent}, which is possible as the path is now assumed to be maximum.
+
+\begin{lemma}
+ Let $\chi(T)$ be a valid coloring of a tree $T$, and let $\mathcal{P}(T)$ be the corresponding maximum path decomposition.
+ Let $p$ be a light path in $\mathcal{P}(T)$ and let $\{q_1, \dots q_k\}$ be the set of all light paths adjacent to $p$.
+
+ For $1\leq i\leq k$ let $r_i = r_{q_i}$ and let $H_i = H_{r_i}$.
+  Then $\sum_{i=1}^k |H_i| \leq |H_p|/c$ for some constant $c>1$, where $H_p = H_{r_p}$.
+\end{lemma}
+
+% \begin{proof}UNDER CONSTRUCTION: 
+% For simplicity, let $H_{k+1}$ denote the heap at the leaf vertex of $p$.  As the $H_k+1$ is disjoing from all other $H_{i}$, \Lem{adjacent} implies that 
+% $H_p > \sum_{j=1}^{k+1} H_i - |p|$.  
+% 
+% Let $x$ be the largest index such that $H_x > H_p / 3$.
+% By the above observation and since $p$ is a light path we have 
+% \[
+%  H_p > \sum_{j=1}^{k} H_i + H_{k+1} - |p|
+% \] 
+% \end{proof}
 

### 2014-08-07 paint-contour-tree.tex
Various small changes

@@
  a set of vertex disjoint leaf paths.  Note that the number of paths in $P(T)$ must equal the number of leaves in $T$.
- For $p\in P(T)$, we use $r_p$ to denote the lowest vertex in $p$ (i.e. its root) and 
- $|p|$ to denote its length.
+ For $p\in P(T)$, we use $|p|$ to denote its length and $r_p$ to denote the lowest vertex in $p$, which we call the \emph{root} of $p$.
 
@@
  Let T be a binary tree and let $\chi(T)$ and $P(T)$ be a valid coloring and path decomposition of $T$, respectively.  
- For $p\in P(T)$, we say that $p$ is \emph{light} if $4 |p| < |H_{r_p}|$, and we say it is \emph{heavy} otherwise.
+ For $p\in P(T)$, we say that $p$ is \emph{light} if $|p| < \sqrt{|H_{r_p}|}/100$, and we say it is \emph{heavy} otherwise.
 \end{definition}
@@
 \begin{definition}
- We say two paths $p,q \in P(T)$ are adjacent if in $T$ there is an edge between a vertex in $p$ and a vertex in $q$.  
-Note that as $p$ and $q$ are vertex disjoint leaf paths, it follows that this edge must be between the root of either 
-$p$ or $q$ and its parent vertex in $T$.  wlog suppose this edge is between the root of $p$ and its parent in $q$. 
-Then we call this parent vertex the \emph{connection} of $p$ and refer to it as $c_p$.
-\end{definition}
+Let $P(T)$ be a path decomposition of $T$.  Then the paths in $P(T)$ naturally define a tree\footnote{Please excuse the 
+overloading of the term 'tree', it is the most natural term to use here.} of there own.  Specifically, in the original 
+tree $T$ contract each path down to its root.  Call the resulting tree the \emph{path tree} of $T$, corresponding to 
+the path decomposition $P(T)$.
+
+We say two path in $P(T)$ are adjacent if their corresponding vertices are adjacent in the path tree.  Similarly, for 
+two adjacent paths in $P(T)$ we can define a parent/child relationship based on this relationship in the path tree. 
+\end{definition}
+
+%\begin{definition}
+%We say two paths $p,q \in P(T)$ are adjacent if in $T$ there is an edge between a vertex in $p$ and a vertex in $q$.  
+%Note that as $p$ and $q$ are vertex disjoint leaf paths, it follows that this edge must be between the root of either 
+%$p$ or $q$ and its parent vertex in $T$.  wlog suppose this edge is between the root of $p$ and its parent in $q$. 
+%Then we call this parent vertex the \emph{connection} of $p$ and refer to it as $c_p$.
+%\end{definition}
 
@@
  Let $T$ be a binary tree and let $\chi(T)$ and $P(T)$ be a valid coloring and path decomposition of $T$, respectively.
- Let $p$ be any path in $P(T)$ and let $\{q_1, \dots q_k\}$ be the set of all light paths adjacent to $p$.
-
+ Let $p$ be any path in $P(T)$ and let $\{q_1, \dots q_k\}$ be the set of all adjacent child paths of $p$.
  For $1\leq i\leq k$ let $r_i = r_{q_i}$ and let $H_i = H_{r_i}$.
- Then $\sum_{i=1}^k |H_i| \leq |H_p|+|p|$, where $H_p = H_{r_p}$.
- %Then $\sum_{i=1}^k \sum_{v\in q_i} \log |H_v| = O(|p|\log |p|)$
+ Also, let $H_0$ denote the heap at the leaf of $p$.
+
+ Then $\sum_{i=0}^k |H_i| \leq |H_p|+|p|$, where $H_p = H_{r_p}$.
 \end{lemma}
@@
  Now for any $i\neq j$, $H_i \cap H_j = \emptyset$, as $r_{i}$ and $r_{j}$ are not on the same root to leaf path (i.e. $L(T_{r_i}) \cap L(T_{r_j}) = \emptyset$).
- In particular, $(H_i\setminus H_i') \cap (H_j \setminus H_j') = \emptyset$, and so $\sum_{i=1}^k |H_i\setminus H_i'| \leq |p|$ 
+ In particular, $(H_i\setminus H_i') \cap (H_j \setminus H_j') = \emptyset$, and so $\sum_{i=0}^k |H_i\setminus H_i'| \leq |p|$ 
  as $H_i\setminus H_i'$ is precisely the subset of $H_i$ that lies on $p$.
- Moreover, for any $i\neq j$, $ H_i' \cap H_j' = \emptyset$, and so since $H_i'\subseteq H_p$ we have $\sum_{i=1}^k |H_i'|\leq |H_p|$.
+ Moreover, for any $i\neq j$, $ H_i' \cap H_j' = \emptyset$, and so since $H_i'\subseteq H_p$ we have $\sum_{i=0}^k |H_i'|\leq |H_p|$.
  Putting these together gives
  \[
-  \sum_{i=1}^k |H_i| = \sum_{i=1}^k |H_i'| + \sum_{i=1}^k |H_i\setminus H_i'| \leq |H_p|+|p|.
+  \sum_{i=0}^k |H_i| = \sum_{i=0}^k |H_i'| + \sum_{i=0}^k |H_i\setminus H_i'| \leq |H_p|+|p|.
  \]
@@
 
+Using the notation defined in the above lemma, we have the following crucial lemma, which we will use to 
+argue we can charge that cost of all paths in $P(T)$ to the cost of the heavy paths.
+
+THERE IS A PROBLEM in the following lemma.
+
+\begin{lemma}
+Let $p$ be a heavy path in $P(T)$ and let $\{q_1, \dots q_k\}$ be any set of light paths in $P(T)$ such that 
+$\sum_{i=1}^k |H_i| = O(H_p)$.  Then $\sum_{i=1}^k \sum_{v\in q_i} \log |H_v| = O(|p|\log |p|)$
+\end{lemma}
+\begin{proof}
+By \Lem{pathBounds}, for all $1\leq i\leq k$, 
+ $\sum_{v\in q_i} \log |H_v| = O(|q_i|\log |H_i|) = O(|H_i|\log |H_i|)$, as $q_i$ is a light path.
+ By assumption $\sum_{i=1}^k |H_i| = O(|H_p|)$, 
+ As $x\log x$ is a convex function, Jensen's inequality then implies 
+ $\sum_{i=1}^{k} |H_i|\log |H_i| = O(|H_p| \log |H_p|) = O(|p|\log |p|)$, since $p$ is a heavy path.
+\end{proof}
+
 \begin{corollary}
+\label{cor:heavyLight}
  Let $p$ be a heavy path in $P(T)$ and let $\{q_1, \dots q_k\}$ be the set of all light paths adjacent to $p$.
@@
  $T$ from some vertex in $p$ which only moves to vertices of larger height and only vertices which lie in light paths.
- Call this set of reachable paths the \emph{light path tree} of $p$.  Naturally this set can be viewed as a tree 
- (i.e. $p$ is the root of this tree and each other path has a well defined depth from $p$).
-\end{definition}
+ Call this set of reachable paths the \emph{light path set} of $p$.  
+ Consider the natural tree over the paths in the light path set of $p$, where each vertex in the tree is a light path 
+ $p$ is the root.  We call this tree the \emph{light path tree} of $p$
+\end{definition}
+
+Let $\mathcal{L}$ be a light path tree of some light path $p$ in a maximum path decomposition $\mathcal{P}(T)$.
+For a path $v\in \mathcal{L}$, let $H_v$ denote the heap at the root of $v$, and let $H_{v_f}$ denote the heap 
+at the leaf of $v$.  Let $S_l(v)$ (resp. $S_h(v)$) be the subset of light (resp. heavy) paths adjacent to $v$ in $T$
+(i.e. $S_l(v)$ are the children of $v$ in $\mathcal{L}$).  Then define the cost of $v$ as 
+$c(v) = H_{v_f}+\sum_{i\in S_h(v)} H_i$.
+
+We now define a weighted version of $\mathcal{L}$ based on these cost.  First we augment $\mathcal{L}$ by attaching one 
+leaf to each vertex in $\mathcal{L}$.  Now for each newly created leaf (which are the only leaves now) assign the cost 
+$c(v)$ defined above, where $v$ is that leaf's corresponding parent.  Now for any vertex $u$ in this tree, define 
+its weight to be the sum of the weights of its children (where at the leaves the weights are just the costs).  
+To simplify notation, we will also refer to this new weighted tree as $\mathcal{L}$.
+
+As the weight $w(v)$ includes the sizes of the heaps of all leaves in the subtree rooted at $v$, we have the following.
+
+\begin{claim}
+For any $v\in \mathcal{L}$, $w(v)\geq |H_v|$
+\end{claim}
+
+Let $W= \sum_{v\in \mathcal{L}} w(v)$.  Suppose one could prove $W=O(H_r)$, where $r$ is the root of $\mathcal{L}$.  
+Then by the above claim, 
+$\sum_{v\in \mathcal{L}} H_v = O(H_r)$, and so by convexity and \Cor{heavyLight}, we could argue the entire cost 
+of all light paths in the light path tree can be charge to a heavy path.  So proving $W=O(H_r)$ is our new goal.
+
+\pagebreak
+\subsection{Junk, don't read}
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\subsection{Junk, don't read}
+\subsection{More Junk}
+
 The following can be viewed as stronger version of \Lem{adjacent}, which is possible as the path is now assumed to be maximum.

### 2014-08-17 paint-contour-tree.tex
subsection5.2 added

@@
 
+\newcommand{\XSays}[2]{{
+      {$\rule[-0.12cm]{0.2in}{0.5cm}$\fbox{\tt
+            #1:} }
+      \textcolor{red}{#2}
+      \marginpar{\textcolor{blue}{#1}}
+      {$\rule[0.1cm]{0.3in}{0.1cm}$\fbox{\tt
+            end}$\rule[0.1cm]{0.3in}{0.1cm}$}
+      }
+   }
+\newcommand{\Ben}[1]{{\XSays{Ben}{#1}}}
+
 \author{
@@
 \section{Lower Bound}
+
+In this section we prove a lower bound on the time it takes to compute a merge tree.
+
+\begin{definition}
+ We say that a contour tree of a $2$-manifold is a \emph{merge tree} if it has a well defined root vertex such that height values 
+ strictly increase along any root to leaf path.
+\end{definition}
+
+\begin{observation}
+ The contour tree of an minimum dominant manifold (see \Def{dom}) is close to a merge tree.  Specifically, there is a root such that on any root 
+ to leaf path height values strictly increase, with the only exception being potentially at the leaves themselves.  As such, up to constant 
+ factors, any lower bound for computing the merge tree is a lower bound for computing the contour tree of a minimum dominant
+ manifold (i.e. intuitively just clip off the leaves or shift them to lie above their parent).
+\end{observation}
+
 First we setup some basic terminology.
@@
 
+\Ben{The following lemma should be immediate once previous section is written properly.}
 \begin{lemma}
@@
 \subsection{Path Decomposition Background}
+\label{sec:background}
 We now introduce the concept of path decompositions and their cost.  In the next subsection we then relate this to the
@@
 
+\newcommand{\pathTree}{P_{\mathcal{T}}}
 \begin{definition}
-Let $P(T)$ be a path decomposition of $T$.  Then the paths in $P(T)$ naturally define a tree\footnote{Please excuse the 
+Let $P(T)$ be a path decomposition of $T$.  Then the paths in $P(T)$ naturally defines a tree\footnote{Please excuse the 
 overloading of the term 'tree', it is the most natural term to use here.} of there own.  Specifically, in the original 
-tree $T$ contract each path down to its root.  Call the resulting tree the \emph{path tree} of $T$, corresponding to 
-the path decomposition $P(T)$.
+tree $T$ contract each path down to its root.  Call the resulting tree the \emph{path tree} of $T$ corresponding to 
+the path decomposition $P(T)$, and denote it by $\pathTree(T)$.
 
@@
 
-THERE IS A PROBLEM in the following lemma.
+\Ben{There is a fundamental flaw in the following lemma and corollary.}
 
@@
 \subsection{Lower Bound by Path Decomposition}
+We say two path trees $T_1$ and $T_2$ (of merge trees of potentially different surfaces) are equivalent if there is an isomorphism, $f$,
+between the two trees such that for any node $v\in T_1$ corresponding to a path of size $|p_v|$, $f(v)$ also corresponds to a path of size $|p_v|$.
+For the purposes of the next lemma, we say two 2-manifolds are different, if their corresponding contour trees are the different.
+
+\begin{figure}[h]\centering
+    \includegraphics[width=.15\linewidth]{figs/diamond}\hspace{2cm}
+    \includegraphics[width=.125\linewidth]{figs/topView}\hspace{2cm}
+    \includegraphics[width=.095\linewidth]{figs/sideView}
+    \caption{From left to right: the cone portion of a diamond; the top down view and then the side view of a cone face with a platform added.}
+    \label{fig:diamond}
+\end{figure}
 
@@
 \label{lem:pathDecomp}
-NEEDS TO BE WRITTEN:
-Have a lemma here showing that our algorithm must pay at least the cost of any valid path decomposition.
+Let $T$ be a merge tree, $P = P(T)$ any path decomposition of $T$, and $\pathTree$ the corresponding path tree.  
+Then there are at least $\Pi_{p_i\in P} |p_i|!$ different 2-manifolds with an equivalent path tree.
 \end{lemma}
-
-For a given valid coloring $\chi$ of a tree $T$, we now describe a particular path decomposition for which 
-we can prove $f_{alg}(\chi(T)) = O(f_{path}(P(T)))$ using the observations about decompositions from the 
-previous subsection.  By the above lemma this will imply $f_{alg}(\chi(T)) = \Theta(f_{path}(P(T)))$.
+\begin{proof}
+ We describe a simple input mesh, with heights specified for all vertices except for the saddles.  The main idea is that 
+ this mesh will be flexible enough to allow almost any assignment of heights to the saddles, thus defining a class of merge trees 
+ whose size is that described in the lemma statement.
+
+ First, we describe the basic building block of this input mesh, which roughly speaking corresponds to a fixed path $p\in P$.  
+ Specifically, we say a \emph{diamond} of size $m$ is the following mesh.  To start, construct a cycle 
+ of length $m$, where the height of the $i$th vertex in the cycle is say $i/2m$.  Next, create an apex vertex at height 
+ $\infty$ (or any appropriately large finite height), and add an edge between this vertex and each vertex in the cycle.  At this point 
+ we have built an upside down cone with $m$ triangular faces (see \Fig{diamond}). To make this cone a closed surface we add a triangular cap at the bottom. 
+ Specifically, add a $3$ cycle of vertices at heights $0$, $1/20m$, and $1/10m$, and connect each vertex of this $3$-cycle to an arbitrary distinct 
+ vertex on the $m$ cycle (with matching cyclic ordering), and triangulate all remaining faces.  For consistency, if the size $m$ is $\leq 3$, the diamond 
+ will be just a tetrahedron.
+
+ Now above each face of the cone of the diamond, we will create a triangular ``platform'', which will act as the location to glue on other (shrunken) diamonds, 
+ and whose highest point of connection to the face below will be a saddle vertex.  Specifically the heights of the vertices of the triangle will be $h$, $h-\eps$, and $h-2\eps$, where 
+ $h$ is some unspecified height value.  This near horizontal platform is then connected to its respective face with vertical edges down to the face. See \Fig{diamond}.
+ %
+ %The vertex at height $h$ will lie in the plane of the respective face of the cone, and the other two will lie above the face, 
+ %with edges connecting down to the face, hence creating an actual ``platform''.
+
+ At this point we can construct the entire input mesh, by gluing together diamonds.  Specifically, for each path $p_i\in P$ we create a diamond of size $p_i$.
+ Now we glue together diamonds of different paths in the same way the paths are connected in $\pathTree$.  Specially, for two paths $p,q\in P$ where 
+ $p$ is the parent of $q$ in $\pathTree$, we glue the bottom triangular cap of the diamond for $q$ onto the triangular ``platform'' of some face of the diamond for $p$.
+ (Naturally for this construction to work, diamonds for a given path will be scaled down relative to the size of the diamond of their parent).
+
+ Observe now that the only saddle points in this construction are the highest connecting point on a face of an edge from the platform above.  Moreover, the only maxima are the apexes of the 
+ diamonds, and there is exactly one minima corresponding to the bottom cap of the diamond representing the root of $\pathTree$.  Therefore, the saddles on a given 
+ diamond will appear contiguously on a root to leaf path in the merge tree of the mesh, where the leaf corresponds to the maxima of the diamond 
+ (since all these saddles have a direct line of sight to this apex).  In particular, 
+ this implies that, regardless of the heights assigned to the platforms, the merge tree of this mesh has a path decomposition whose corresponding path 
+ tree is equivalent to $\pathTree$.
+
+ Now there is a valid instantiation of this described mesh for any relative ordering of the heights of the saddles on a given diamond.  In particular, 
+ there are $|p|!$ possible orderings of the heights of the saddles on the diamond for $p$, and hence $\Pi_{p_i\in P} |p_i|!$ possible instantiations of 
+ the mesh overall.  Also, each one of these instantiations will result in a different merge.  Specifically, a permutation of the heights on a given diamond 
+ corresponds to a permutation of the vertices along a path in $P$, as by construction all saddles on a given diamond will appear in sorted order in the merge tree.
+\end{proof}
+
+\begin{corollary}
+ Let $A(S)$ be any algorithm to compute the merge tree of any input surface $S$ (whose contour tree is a merge tree).  
+ Then the running time of $A(S)$ is $\Omega(f_{path}(P))=\Omega(\sum_{p_i\in P} |p_i|\log|p_i|)$ where $P$ is any path decomposition of the 
+ merge tree of $S$.
+\end{corollary}
+\begin{proof}
+ Given an input surface $S$ whose merge tree has a particular path decomposition $P$, by the above lemma, there are $\Pi_{p_i\in P} |p_i|!$ other surfaces 
+ with distinct merge trees, all of which have a path tree equivalent to $\pathTree$.  Therefore, any algorithm which computes the merge tree must distinguish between 
+ these possibilities, and therefore its corresponding decision tree must have $\Pi_{p_i\in P} |p_i|!$ leaves.  Hence by Stirling's approximation, 
+ this decision tree must have depth $\Omega(\sum_{p_i\in P} |p_i|\log|p_i|)$.
+\end{proof}
+
+\subsection{Upper Bound by Path Decomposition}
+\Ben{This section needs revision.}
+
+From the previous section we now have a lower bound of $\Omega(f_{path}(P))$ for computing the merge tree in terms of any path decomposition $P$. 
+We now show that the running time of our algorithm for computing the merge tree is upper bounded by this expression for a particular path decomposition.
+
+Specifically, for a given valid coloring $\chi$ of a tree $T$, we describe a path decomposition for which 
+we can prove $f_{alg}(\chi(T)) = O(f_{path}(P(T)))$ using the observations about path decompositions from 
+\Sec{background}, which will then imply $f_{alg}(\chi(T)) = \Theta(f_{path}(P(T)))$.
 

### 2014-08-19 paint-contour-tree.tex
Changes to 5.3

@@
 \newcommand{\Ben}[1]{{\XSays{Ben}{#1}}}
+\newcommand{\pth}[2][\!]{#1\left({#2}\right)}
 
@@
 \newcommand{\pathTree}{P_{\mathcal{T}}}
+\newcommand{\pathTreeA}{\mathcal{P}_{\mathcal{T}}}
 \begin{definition}
@@
 
+We now prove that in a maximum path decomposition, when going between adjacent light paths, the heap size decreases by a constant fraction.
+
+\begin{lemma}
+\label{lem:geometric}
+ Let $p$ and $q$ be two light paths such that $p$ is the parent of $q$ in the path tree $\pathTreeA$.  
+ Then $|H_q|\leq \frac{3}{5} |H_p|$, where $H_p$ and $H_q$ are the heaps at the roots of $p$ and $q$, respectively. 
+\end{lemma}
+\begin{proof}
+ Let $w$ be the vertex in the path $p$ to which the root of the path $q$ is adjacent to in $T$, 
+ i.e. $w$ is the parent of $r_q$ in $T$.  Let $c$ be the other child of $w$, which lies in $p$.
+ As $\mathcal{P}(T)$ is a maximum path decomposition of $T$ and $c$ rather than $r_q$ is in 
+ the same path as $w$, we have $|H_q|\leq |H_c|$.  Therefore, $|H_w|\geq |H_q|+|H_c|-1\geq 2|H_q|-1$. 
+ Therefore, by \Obs{decrease}, $|H_p|+ d(r_p, w) \geq |H_w|\geq 2|H_q|-1$.  
+ Since $p$ is a light path, $d(r_p, w)\leq \sqrt{|H_p|}/100$ and so we have $|H_p|+\sqrt{|H_p|}/100\geq 2|H_q|-1$.
+ Assuming $|H_p|>11$ (and we can safely assume $|H_p|$ is bigger than any constant), this yields $\frac{3}{5} |H_p| \geq |H_q|$
+\end{proof}
+
+Naturally, this geometric series behavior can be extended to more than two adjacent light paths, thus motivating 
+the following definition.
+
 \begin{definition}
- Let $p$ be a path in a path decomposition $P(T)$.  Call a light path in $P(T)$ reachable from $p$ if there is a path in the tree
- $T$ from some vertex in $p$ which only moves to vertices of larger height and only vertices which lie in light paths.
- Call this set of reachable paths the \emph{light path set} of $p$.  
- Consider the natural tree over the paths in the light path set of $p$, where each vertex in the tree is a light path 
- $p$ is the root.  We call this tree the \emph{light path tree} of $p$
-\end{definition}
-
-Let $\mathcal{L}$ be a light path tree of some light path $p$ in a maximum path decomposition $\mathcal{P}(T)$.
+ Let $p$ be a path in a path decomposition $P(T)$, whose corresponding path tree is $\pathTree$.
+ Call a light path $q\in P(T)$ light reachable from $p$ if the path from $q$ to $p$ in $\pathTree$ 
+ contains only vertices which correspond to light paths.  Call the subtree of $\pathTree$ consisting 
+ of all light reachable paths from $p$ the \emph{light path tree} of $p$.
+\end{definition}
+
+Let $\mathcal{L}$ be a light path tree of some light path $r$ in a maximum path decomposition $\mathcal{P}(T)$.
+Our goal is to use \Lem{geometric}  to show that $\sum_{v\in \mathcal{L}} |H_v| = O(|H_r|)$, namely the total sum of the sizes of the heaps 
+at the roots of all paths in the light path tree is on the order of the size of the heap at the root of $r$.  
+ADD THE ARGUMENT saying how this fact will be used.
+
 For a path $v\in \mathcal{L}$, let $H_v$ denote the heap at the root of $v$, and let $H_{v_f}$ denote the heap 
-at the leaf of $v$.  Let $S_l(v)$ (resp. $S_h(v)$) be the subset of light (resp. heavy) paths adjacent to $v$ in $T$
-(i.e. $S_l(v)$ are the children of $v$ in $\mathcal{L}$).  Then define the cost of $v$ as 
-$c(v) = H_{v_f}+\sum_{i\in S_h(v)} H_i$.
-
-We now define a weighted version of $\mathcal{L}$ based on these cost.  First we augment $\mathcal{L}$ by attaching one 
+at the leaf of $v$.  Let $S_l(v)$ (resp. $S_h(v)$) be the subset of light (resp. heavy) paths adjacent to $v$ in $\mathcal{L}$.
+
+We want to understand how the heap size changes as we move around the paths in $\mathcal{L}$.  Specifically, there are 
+three places the vertices in the heap along a light path can come from; adjacent light paths, adjacent heavy paths, and the initial heap at the leaf of the path.  
+Given this observation we now define a weight function which upper bounds the size of the heap at the root of any path in $\mathcal{L}$.
+
+First, for all $v\in \mathcal{L}$ we define the local cost function $c(v) = H_{v_f}+\sum_{i\in S_h(v)} H_i$, this is the addition to the heap at $v$ which 
+was not already charged to descendant light paths.  We now define a weighted version of $\mathcal{L}$ based on these costs.  
+To do so, we first we augment $\mathcal{L}$ by attaching one 
 leaf to each vertex in $\mathcal{L}$.  Now for each newly created leaf (which are the only leaves now) assign the cost 
-$c(v)$ defined above, where $v$ is that leaf's corresponding parent.  Now for any vertex $u$ in this tree, define 
-its weight to be the sum of the weights of its children (where at the leaves the weights are just the costs).  
-To simplify notation, we will also refer to this new weighted tree as $\mathcal{L}$.
-
-As the weight $w(v)$ includes the sizes of the heaps of all leaves in the subtree rooted at $v$, we have the following.
+$c(v)$ defined above, where $v$ is that leaf's corresponding parent.  
+Now we can define a weight function $w(u)$ over any node $u$ in the tree, which is defined to be the total cost of all leaves in its subtree.   
+To simplify notation, we now use $\mathcal{L}$ to refer to this new weighted tree (rather than its unweighted counterpart).
+
+As the weight $w(v)$ includes the sizes of the heaps of all contributing heavy paths and leaves in the subtree rooted at $v$, we have the following.
 
 \begin{claim}
-For any $v\in \mathcal{L}$, $w(v)\geq |H_v|$
+For any $v\in \mathcal{L}$, $w(v)\geq |H_v|$.
 \end{claim}
 
-Let $W= \sum_{v\in \mathcal{L}} w(v)$.  Suppose one could prove $W=O(H_r)$, where $r$ is the root of $\mathcal{L}$.  
-Then by the above claim, 
-$\sum_{v\in \mathcal{L}} H_v = O(H_r)$, and so by convexity and \Cor{heavyLight}, we could argue the entire cost 
-of all light paths in the light path tree can be charge to a heavy path.  So proving $W=O(H_r)$ is our new goal.
+Let $W= \sum_{v\in \mathcal{L}} w(v) \geq \sum_{v\in \mathcal{L}} |H_v|$.  Suppose we could show that conversely $W=O(H_r)$.  
+Then we would have $\sum_{v\in \mathcal{L}} H_v = O(H_r)$, which was our original goal.  So proving $W=O(H_r)$ is now our new goal.
+
+To this end, first observe that $|H_r|\geq w(r)-\sum_{v\in \mathcal{L}} |v|$.  This holds by the same argument as in \Lem{adjacent}, namely, 
+$w(r)$ by definition counts all the contributing heavy path heaps and initial leaf heaps that contribute to the heap at $r$ and $\sum_{v\in \mathcal{L}} |v|$
+is an upper bound on the number of vertices that could have been removed from one of these heaps before reach the root of $r$ (since a vertex in a heap can only 
+get removed if it appears on one of the paths).
+
+As all paths in $\mathcal{L}$ are light, we thus have the following.
+
+\begin{claim}
+$|H_r|\geq w(r)-\sum_{v\in \mathcal{L}} \sqrt{H_v}/100 \geq w(r)-\sum_{v\in \mathcal{L}} \sqrt{w(v)}/100$
+\end{claim}
+
+We will need the following lemma.
+
+\begin{lemma}
+\label{lem:distribution}
+Let $C\subseteq \mathcal{P}$ be the subset of paths that are leaves in $\mathcal{L}$.
+
+Then $\sum_{v\in \mathcal{P}} \sqrt{w(v)}/100 \leq \sum_{c\in C} \sqrt{w(c)}/20$
+\end{lemma}
+
+\begin{proof}
+The idea is to recursively redistribute the contribution to the sum of each internal vertex to its children, until all the weight sits on the children.
+Specifically, let $v$ be some path in $\mathcal{L}$.  Its contribution to the sum is $\sqrt{w(v)}/100$, which we will call the value at $v$.  
+We redistribute this value to its children, 
+in proportion by the weight of each child.  For example, if $v$ has a child $x$ then the fraction it gets assigned is $w(x)/w(v)$ 
+(recall that by definition $w(v)$ is the sum of the weights of its children and so the total value will get reassigned to the children).
+
+Suppose we keep redistributing the value of $v$ until it reaches the leaves of $\mathcal{L}$.  Let's see how much gets assigned to a given leaf.
+Specifically, let $u_0, u_1, \ldots, u_k$ be a path in $\mathcal{L}$ (of paths from the set $\mathcal{P}$), such that $v=u_0$,
+$u_{i+1}$ is child of $u_i$ for all $i$, and $u_k$ is a leaf of $\mathcal{L}$.
+
+Then the total value from $u_0$ that ends up at $u_k$ is 
+\[
+\frac{\sqrt{w(u_0)}}{100} \pth{ \frac{w(u_1)}{w(u_0)} \frac{w(u_2)}{w(u_1)} \cdots \frac{w(u_k)}{w(u_{k-1})} }
+= \frac{\sqrt{w(u_0)}}{100} \pth{ \frac{w(u_k)}{w(u_0)} } = \frac{w(u_k)}{100\sqrt{w(u_0)}}
+\]
+
+For a given leaf in $\mathcal{L}$, we can calculate the total value reassigned to it from it its ancestors.
+Specifically, let $u_0, u_1, u_2, \dots, u_m$ be a root to leaf path in $\mathcal{L}$ (i.e. $u_0$ is the root $r$ of $\mathcal{L}$).
+
+Then by the above calculation and \Lem{geometric}, the total reassigned to $u_m$ is
+
+\Ben{There is an issue here.  \Lem{geometric} is in terms of the heaps not the weights.}
+
+\[
+ \sum_{i=0}^m \frac{w(u_m)}{100\sqrt{w(u_i)}} \leq \frac{w(u_m)}{100} \sum_{i=0}^m 1 \Big/ \sqrt{\pth{\frac{5}{3}}^iw(u_m)} 
+ \leq \frac{\sqrt{w(u_m)}}{100} \sum_{i=0}^{\infty} \pth{\frac{3}{5}}^{i/2} \leq \frac{\sqrt{w(u_m)}}{20}
+\]
+
+The claim now follows by summing this quantity over the children of $\mathcal{L}$.
+\end{proof}
+
+By definition, $w(r) = \sum_{c\in C} w(c)$, where $C$ is the subset of leaves of $\mathcal{L}$.  
+Therefore combining the above lemma and claim gives the following.
+
+\begin{claim}
+ $|H_r|\geq w(r)/2$
+\end{claim}
+
+\begin{lemma}
+ $W=O(H_r)$
+\end{lemma}
+\begin{proof}
+MISSING.  Is this what we are claiming?  I don't see the proof.
+\end{proof}
+%
+%
+% So $w(r)$ upper bound $H_r$, but by how much?  Consider a path $v\in \mathcal{L}$.  
+% As we move from the leaf to the root of $v$ we now want to understand 
+% how the heap can decreases.  Now the only way a vertex in the heap can leave that 
+% heap is if that vertex is one of the vertices we traverse in $v$, i.e. the total 
+% loss in vertices from the heap along $v$ is bounded by $|v|$.  
+% Therefore, $|H_r|\geq W-\sum_{v\in \mathcal{L}} |v|$.  
+% (This is an extension of \Lem{adjacent}, and probably should be merged with it.)
+% As $\mathcal{L}$ consists of only light paths we have $|v|\leq \sqrt{|H_v|}/100 \leq \sqrt{w(v)}/100$, 
+% and therefore $|H_r|\geq W-\sum_{v\in \mathcal{L}} \sqrt{w(v)}/100$.
 

### 2014-08-20 paint-contour-tree.tex
small changes

@@
 Let $\mathcal{L}$ be a light path tree of some light path $r$ in a maximum path decomposition $\mathcal{P}(T)$.
-Our goal is to use \Lem{geometric}  to show that $\sum_{v\in \mathcal{L}} |H_v| = O(|H_r|)$, namely the total sum of the sizes of the heaps 
-at the roots of all paths in the light path tree is on the order of the size of the heap at the root of $r$.  
-ADD THE ARGUMENT saying how this fact will be used.
+We now define a weight function $w(v)$ for all $v\in \mathcal{L}$.  Such that $w(v)=\Theta(|H_v|)$.  
+For this alternative quantity we will be able to show BLAH BLAH BLAH.
+
+%Our goal is to use \Lem{geometric}  to show that $\sum_{v\in \mathcal{L}} |H_v| = O(|H_r|)$, namely 
+%the total sum of the sizes of the heaps at the roots of all paths in the light path tree is on the 
+%order of the size of the heap at the root of $r$.  ADD THE ARGUMENT saying how this fact will be used.
 
@@
 
-Let $W= \sum_{v\in \mathcal{L}} w(v) \geq \sum_{v\in \mathcal{L}} |H_v|$.  Suppose we could show that conversely $W=O(H_r)$.  
-Then we would have $\sum_{v\in \mathcal{L}} H_v = O(H_r)$, which was our original goal.  So proving $W=O(H_r)$ is now our new goal.
-
-To this end, first observe that $|H_r|\geq w(r)-\sum_{v\in \mathcal{L}} |v|$.  This holds by the same argument as in \Lem{adjacent}, namely, 
-$w(r)$ by definition counts all the contributing heavy path heaps and initial leaf heaps that contribute to the heap at $r$ and $\sum_{v\in \mathcal{L}} |v|$
-is an upper bound on the number of vertices that could have been removed from one of these heaps before reach the root of $r$ (since a vertex in a heap can only 
-get removed if it appears on one of the paths).
+%Let $W= \sum_{v\in \mathcal{L}} w(v) \geq \sum_{v\in \mathcal{L}} |H_v|$.  
+%Suppose we could show that conversely $W=O(H_r)$.  Then we would have 
+%$\sum_{v\in \mathcal{L}} H_v = O(H_r)$, which was our original goal.  
+%So proving $W=O(H_r)$ is now our new goal.
+
+We now would like to show conversely that $w(v)= O(|H_v|)$.  
+To do so we first introduce some notation.  
+Let $C\subseteq \mathcal{P}$ be the subset of paths that are leaves in $\mathcal{L}$.
+For any path $v$, let $\mathcal{L}_v$ (resp. $C_v$) be the subset of paths in $\mathcal{L}$ (resp $C$) 
+that are contained in the subtree of $\mathcal{L}$ which is rooted at $v$.
+
+First, observe that $|H_v|\geq w(v)-\sum_{u\in \mathcal{L}_v} |u|$.  
+This holds by the same argument as in \Lem{adjacent}, namely, $w(u)$ by definition 
+counts all the contributing heavy path heaps and initial leaf heaps that contribute 
+to the heap at $v$ and $\sum_{u\in \mathcal{L}_v} |u|$ is an upper bound on the number 
+of vertices that could have been removed from one of these heaps before reaching the 
+root of $v$ (since a vertex in a heap can only get removed if it appears on one of 
+the paths).
 
@@
 \begin{claim}
-$|H_r|\geq w(r)-\sum_{v\in \mathcal{L}} \sqrt{H_v}/100 \geq w(r)-\sum_{v\in \mathcal{L}} \sqrt{w(v)}/100$
+$|H_v|\geq w(v)-\sum_{u\in \mathcal{L}_v} \sqrt{|H_u|}/100 \geq w(v)-\sum_{u\in \mathcal{L}_v} \sqrt{w(u)}/100$
 \end{claim}
 
-We will need the following lemma.
+By the above claim, if we can prove $\sum_{u\in \mathcal{L}_v} \sqrt{w(u)}/100 \leq w(v)/c$ for some constant $c>1$, 
+then we will have proven that $|H_v| = \Theta(w(v))$.
+
+Recall that \Lem{geometric} told us that that $|H_{u}|\leq (3/5) |H_{u'}|$, if $u'$ is the parent $u$ in $\mathcal{L}$. 
+The following lemma shows that if we had a similar statement in terms of the weights rather than the heap sizes, 
+then we get the desired bound on the sum.
 
@@
 \label{lem:distribution}
-Let $C\subseteq \mathcal{P}$ be the subset of paths that are leaves in $\mathcal{L}$.
-
-Then $\sum_{v\in \mathcal{P}} \sqrt{w(v)}/100 \leq \sum_{c\in C} \sqrt{w(c)}/20$
+Let $v$ be a path in $\mathcal{L}$.  Suppose that for any two paths $x,y\in \mathcal{L}_v$ 
+such that $y$ is the parent of $x$, $w(x)\leq (4/5) w(y)$.
+Then $\sum_{u\in \mathcal{P}_v} \sqrt{w(u)}/100 \leq \sum_{c\in C_v} \sqrt{w(c)}/20$
 \end{lemma}
@@
 \begin{proof}
-The idea is to recursively redistribute the contribution to the sum of each internal vertex to its children, until all the weight sits on the children.
-Specifically, let $v$ be some path in $\mathcal{L}$.  Its contribution to the sum is $\sqrt{w(v)}/100$, which we will call the value at $v$.  
-We redistribute this value to its children, 
-in proportion by the weight of each child.  For example, if $v$ has a child $x$ then the fraction it gets assigned is $w(x)/w(v)$ 
-(recall that by definition $w(v)$ is the sum of the weights of its children and so the total value will get reassigned to the children).
-
-Suppose we keep redistributing the value of $v$ until it reaches the leaves of $\mathcal{L}$.  Let's see how much gets assigned to a given leaf.
-Specifically, let $u_0, u_1, \ldots, u_k$ be a path in $\mathcal{L}$ (of paths from the set $\mathcal{P}$), such that $v=u_0$,
-$u_{i+1}$ is child of $u_i$ for all $i$, and $u_k$ is a leaf of $\mathcal{L}$.
+The idea is to recursively redistribute the contribution to the sum of each internal vertex to its children, 
+until all the weight sits on the leaves.  Specifically, let $v$ be some path in $\mathcal{L}$.  Its 
+contribution to the sum is $\sqrt{w(v)}/100$, which we will call the value at $v$.  We redistribute this 
+value to its children, in proportion by the weight of each child.  For example, if $v$ has a child $x$ then 
+the fraction it gets assigned is $w(x)/w(v)$ (recall that by definition $w(v)$ is the sum of the weights of 
+its children and so the total value will get reassigned to the children).
+
+Suppose we keep redistributing the value of $v$ until it reaches the leaves of its subtree in $\mathcal{L}$.  
+Let's see how much gets assigned to a given leaf.  Specifically, let $u_0, u_1, \ldots, u_k$ 
+be a path in $\mathcal{L}$ (of paths from the set $\mathcal{P}_v$), such that $v=u_0$, $u_{i+1}$ 
+is child of $u_i$ for all $i$, and $u_k$ is a leaf of $\mathcal{L}$.
 
@@
 For a given leaf in $\mathcal{L}$, we can calculate the total value reassigned to it from it its ancestors.
-Specifically, let $u_0, u_1, u_2, \dots, u_m$ be a root to leaf path in $\mathcal{L}$ (i.e. $u_0$ is the root $r$ of $\mathcal{L}$).
-
-Then by the above calculation and \Lem{geometric}, the total reassigned to $u_m$ is
-
-\Ben{There is an issue here.  \Lem{geometric} is in terms of the heaps not the weights.}
+Specifically, let $u_0, u_1, u_2, \dots, u_m$ be a root to leaf path in $\mathcal{L}$ 
+(i.e. $u_0$ is the root $r$ of $\mathcal{L}$).
+
+By assumption from the lemma statement we have $w(u_{i+1})\leq (3/5) w(u_i)$.  Therefore by the above calculation, 
+the total reassigned to $u_m$ is
 
@@
 
-The claim now follows by summing this quantity over the children of $\mathcal{L}$.
-\end{proof}
-
-By definition, $w(r) = \sum_{c\in C} w(c)$, where $C$ is the subset of leaves of $\mathcal{L}$.  
-Therefore combining the above lemma and claim gives the following.
+The claim then follows by summing this quantity over the children of $\mathcal{L}$.
+\end{proof}
+
+By combining the above lemma and claim we have the following.
 
 \begin{claim}
- $|H_r|\geq w(r)/2$
+For any path $v\in \mathcal{L}$ if it holds that 
+$w(x)\leq (4/5) w(y)$ for any two paths $x,y\in \mathcal{L}_v$ such that $y$ is the parent of $x$, then
+
+\[
+|H_v|\geq w(v)-\sum_{u\in \mathcal{L}_v} \sqrt{w(u)}/100 \geq w(v)- \sum_{c\in C_v} \sqrt{w(c)}/20 \geq \frac{4}{5} w(v)
+\]
 \end{claim}
@@
 \begin{lemma}
- $W=O(H_r)$
+Let $\mathcal{L}$ be the light path tree of some path $r$.  
+Then for any $v\in \mathcal{L}$, we have $|H_v|\geq \frac{4}{5} w(v)$.
 \end{lemma}
 \begin{proof}
-MISSING.  Is this what we are claiming?  I don't see the proof.
-\end{proof}
+We will actually prove the stronger statement that for any $v\in \mathcal{L}$, 
+we have $|H_v|\geq \frac{4}{5} w(v)$ and $w(v)\leq (4/5) w(p)$ for any two paths 
+$x,y\in \mathcal{L}_v$ such that $y$ is the parent of $x$.
+
+Suppose otherwise, and let $v$ be the lowest path in $\mathcal{L}$ such that this fails 
+(i.e. the statement holds for all descendants of $v$ in $\mathcal{L}$).  Then for any 
+two paths $x,y\in \mathcal{L}_v$ such that $y$ is the parent of $x$, 
+we have that $w(x)\leq (4/5) w(y)$ and so by \Lem{geometric},
+
+\end{proof}
+
 %

### 2014-08-20 paint-contour-tree.tex
Structural changes to section 5

@@
 \newcommand{\Ben}[1]{{\XSays{Ben}{#1}}}
+\newcommand{\remove}[1]{}
 \newcommand{\pth}[2][\!]{#1\left({#2}\right)}
@@
 \pagebreak
+\newcommand{\pathTree}{P_{\mathcal{T}}}
+\newcommand{\pathTreeA}{\mathcal{P}_{\mathcal{T}}}
 \section{Lower Bound}
@@
 
-\subsection{Path Decomposition Background}
-\label{sec:background}
-We now introduce the concept of path decompositions and their cost.  In the next subsection we then relate this to the
-running time of our algorithm.
+To do so, we now introduce the concept of path decompositions and their cost.  Then in the following two subsections 
+we then relate path decompositions costs (defined below) to the running time of our algorithm using $\Lem{algCost}$.
 
 \begin{definition}
+\label{def:path}
  Let $T$ be a rooted binary tree.  We call a path $p$ in $T$ a \emph{leaf path} if $p$ contains a leaf of $T$ and 
@@
 
-% \begin{observation}
-% DELETE IF NOT NEEDED:
-% Note that $f_{path}(P(T)) = \Theta( \sum_{v\in V} l_v )$, where $l_v$ is the distance from $v$ to the leaf of its containing path in $P(T)$.  
-% That is, conceptually we can think of each vertex as charging the $\log$ of the length to its path's leaf.
-% \end{observation}
+Let $P(T)$ be a path decomposition of $T$.  Then the paths in $P(T)$ naturally define a tree\footnote{Please excuse the 
+overloading of the term 'tree', it is the most natural term to use here.} of there own.  Specifically, in the original 
+tree $T$ contract each path down to its root.  Call the resulting tree the \emph{path tree} of $T$ corresponding to 
+the path decomposition $P(T)$, and denote it by $\pathTree(T)$.
+
+We say two paths in $P(T)$ are adjacent if their corresponding vertices are adjacent in the path tree.  Similarly, for 
+two adjacent paths in $P(T)$ we can define a parent/child relationship based on this relationship in the path tree.
+
+\remove{
+\Ben{Maybe move the remainder of this subsection to 5.3 since heavy and light paths and not used until then}
 
@@
 
-\newcommand{\pathTree}{P_{\mathcal{T}}}
-\newcommand{\pathTreeA}{\mathcal{P}_{\mathcal{T}}}
-\begin{definition}
-Let $P(T)$ be a path decomposition of $T$.  Then the paths in $P(T)$ naturally defines a tree\footnote{Please excuse the 
-overloading of the term 'tree', it is the most natural term to use here.} of there own.  Specifically, in the original 
-tree $T$ contract each path down to its root.  Call the resulting tree the \emph{path tree} of $T$ corresponding to 
-the path decomposition $P(T)$, and denote it by $\pathTree(T)$.
-
-We say two path in $P(T)$ are adjacent if their corresponding vertices are adjacent in the path tree.  Similarly, for 
-two adjacent paths in $P(T)$ we can define a parent/child relationship based on this relationship in the path tree. 
-\end{definition}
-
-%\begin{definition}
-%We say two paths $p,q \in P(T)$ are adjacent if in $T$ there is an edge between a vertex in $p$ and a vertex in $q$.  
-%Note that as $p$ and $q$ are vertex disjoint leaf paths, it follows that this edge must be between the root of either 
-%$p$ or $q$ and its parent vertex in $T$.  wlog suppose this edge is between the root of $p$ and its parent in $q$. 
-%Then we call this parent vertex the \emph{connection} of $p$ and refer to it as $c_p$.
-%\end{definition}
-
 The following lemma can be thought of as a generalization of $\Obs{decrease}$.
@@
 \begin{proof} 
- %Let $c_1, \dots, c_k$ be the connections of the paths $q_1, \dots, q_k$ to the path $p$, and let $c_0=r_p$.  
- %Let $d_i = d(c_i, c_{i+1})$, and let $d_k$ be the distance from $c_k$ to the leaf of $p$.  
- %Observe that $\sum_{i=0}^k d_i = l_p$.
- %
- % 
- %Let $H_i$ be the heap at $r_{q_i}$, and let $H_0$ be the heat to $r_p$.  
  First observe that for all $i$, $r_{i}$ lies above $r_p$ on some root to leaf path. 
@@
 
-Using the notation defined in the above lemma, we have the following crucial lemma, which we will use to 
-argue we can charge that cost of all paths in $P(T)$ to the cost of the heavy paths.
-
-\Ben{There is a fundamental flaw in the following lemma and corollary.}
-
-\begin{lemma}
-Let $p$ be a heavy path in $P(T)$ and let $\{q_1, \dots q_k\}$ be any set of light paths in $P(T)$ such that 
-$\sum_{i=1}^k |H_i| = O(H_p)$.  Then $\sum_{i=1}^k \sum_{v\in q_i} \log |H_v| = O(|p|\log |p|)$
-\end{lemma}
-\begin{proof}
-By \Lem{pathBounds}, for all $1\leq i\leq k$, 
- $\sum_{v\in q_i} \log |H_v| = O(|q_i|\log |H_i|) = O(|H_i|\log |H_i|)$, as $q_i$ is a light path.
- By assumption $\sum_{i=1}^k |H_i| = O(|H_p|)$, 
- As $x\log x$ is a convex function, Jensen's inequality then implies 
- $\sum_{i=1}^{k} |H_i|\log |H_i| = O(|H_p| \log |H_p|) = O(|p|\log |p|)$, since $p$ is a heavy path.
-\end{proof}
-
-\begin{corollary}
-\label{cor:heavyLight}
- Let $p$ be a heavy path in $P(T)$ and let $\{q_1, \dots q_k\}$ be the set of all light paths adjacent to $p$.
- Then $\sum_{i=1}^k \sum_{v\in q_i} \log |H_v| = O(|p|\log |p|)$.
-\end{corollary}
-\begin{proof}
- Let $H_i$ be as defined in \Lem{adjacent}.  By \Lem{pathBounds}, for all $1\leq i\leq k$, 
- $\sum_{v\in q_i} \log |H_v| = O(|q_i|\log |H_i|) = O(|H_i|\log |H_i|)$, as $q_i$ is a light path.
- Now by \Lem{adjacent}, $\sum_{i=1}^k |H_i| \leq |H_p|+|p| = O(|p|)$, as $p$ is a heavy path. 
- Therefore, as $x\log x$ is a convex function, Jensen's inequality implies the claim.
-
-\end{proof}
+% \Ben{There is a fundamental flaw in the following lemma and corollary.  
+% As such, the remainder of this subsection will likely be deleted.}
+% 
+% Using the notation defined in the above lemma, we have the following crucial lemma, which we will use to 
+% argue we can charge that cost of all paths in $P(T)$ to the cost of the heavy paths.
+% 
+% \begin{lemma}
+% Let $p$ be a heavy path in $P(T)$ and let $\{q_1, \dots q_k\}$ be any set of light paths in $P(T)$ such that 
+% $\sum_{i=1}^k |H_i| = O(H_p)$.  Then $\sum_{i=1}^k \sum_{v\in q_i} \log |H_v| = O(|p|\log |p|)$
+% \end{lemma}
+% \begin{proof}
+% By \Lem{pathBounds}, for all $1\leq i\leq k$, 
+%  $\sum_{v\in q_i} \log |H_v| = O(|q_i|\log |H_i|) = O(|H_i|\log |H_i|)$, as $q_i$ is a light path.
+%  By assumption $\sum_{i=1}^k |H_i| = O(|H_p|)$, 
+%  As $x\log x$ is a convex function, Jensen's inequality then implies 
+%  $\sum_{i=1}^{k} |H_i|\log |H_i| = O(|H_p| \log |H_p|) = O(|p|\log |p|)$, since $p$ is a heavy path.
+% \end{proof}
+% 
+% 
+% \begin{corollary}
+% \label{cor:heavyLight}
+%  Let $p$ be a heavy path in $P(T)$ and let $\{q_1, \dots q_k\}$ be the set of all light paths adjacent to $p$.
+%  Then $\sum_{i=1}^k \sum_{v\in q_i} \log |H_v| = O(|p|\log |p|)$.
+% \end{corollary}
+% \begin{proof}
+%  Let $H_i$ be as defined in \Lem{adjacent}.  By \Lem{pathBounds}, for all $1\leq i\leq k$, 
+%  $\sum_{v\in q_i} \log |H_v| = O(|q_i|\log |H_i|) = O(|H_i|\log |H_i|)$, as $q_i$ is a light path.
+%  Now by \Lem{adjacent}, $\sum_{i=1}^k |H_i| \leq |H_p|+|p| = O(|p|)$, as $p$ is a heavy path. 
+%  Therefore, as $x\log x$ is a convex function, Jensen's inequality implies the claim.
+%  
+% \end{proof}
+
+}
 
 \subsection{Lower Bound by Path Decomposition}
+
+In this section we show that any algorithm for computing the merge tree $T$ must pay at least $f_{path}(P(T))$ (see \Def{path}) for any path decomposition $P(T)$.
+
 We say two path trees $T_1$ and $T_2$ (of merge trees of potentially different surfaces) are equivalent if there is an isomorphism, $f$,
@@
 \subsection{Upper Bound by Path Decomposition}
-\Ben{This section needs revision.}
+\Ben{This section is a work in progress.}
 
 From the previous section we now have a lower bound of $\Omega(f_{path}(P))$ for computing the merge tree in terms of any path decomposition $P$. 
-We now show that the running time of our algorithm for computing the merge tree is upper bounded by this expression for a particular path decomposition.
-
-Specifically, for a given valid coloring $\chi$ of a tree $T$, we describe a path decomposition for which 
-we can prove $f_{alg}(\chi(T)) = O(f_{path}(P(T)))$ using the observations about path decompositions from 
-\Sec{background}, which will then imply $f_{alg}(\chi(T)) = \Theta(f_{path}(P(T)))$.
+We now show that the running time of our algorithm for computing the merge tree is upper bounded by this expression for a specific path decomposition, 
+which we will call the maximum path decomposition (defined below).
+Specifically, for a given valid coloring $\chi$ of a tree $T$, we will prove that for the maximum path decomposition $f_{alg}(\chi(T)) = O(f_{path}(P(T)))$. 
+%using the observations about path decompositions from 
+%\Sec{background}, which will then imply $f_{alg}(\chi(T)) = \Theta(f_{path}(P(T)))$.
+
+\subsubsection{Heavy and light paths}
+
+Before describing the maximum path decomposition, we introduce the notion of heavy and light paths 
+(which will help motivate what the decomposition should be).
+
+Recall we are ultimately trying to show $f_{alg}(\chi(T)) = O(f_{path}(P(T)))$ for some path decomposition $P=P(T)$, 
+or more specifically $\sum_{v\in T} \log |H_v| = O(\sum_{p\in P} |p| \log |p|)$.  In particular, if we could 
+show for each path $p\in P$ that $\sum_{v\in p} \log |H_v|= |p|\log|p|$ then we would be done.
+In other words we can think of each path as having a budget of $|p|\log|p|$, and ideally the budget 
+of each path will pay for the algorithm's cost due to this path.  This is too good to hope for, so instead 
+we will divide paths into two categories, called heavy and light, which we show below in \Lem{pathBounds} roughly corresponds 
+to whether a path can pay for itself or not.
+
+\begin{definition}
+ Let T be a binary tree and let $\chi(T)$ and $P(T)$ be a valid coloring and path decomposition of $T$, respectively.  
+ For $p\in P(T)$, we say that $p$ is \emph{light} if $|p| < \sqrt{|H_{r_p}|}/100$, and we say it is \emph{heavy} otherwise.
+\end{definition}
+
+\begin{observation}
+\label{obs:decrease}
+ Let $v$ be a vertex in $T$ and let $w$ be its parent.  Then $|H_w| \geq |H_v| -1$, as $L(T_v) \subseteq L(T_w)$ and the path from 
+ $w$ to the root has one less vertex than the path from $v$ to the root.
+
+ In particular, we have the following more general property.
+ Let $v$ and $u$ be any two vertices in the same root to leaf path of $T$, such that $u$ is lower than $v$.  
+ Then $|H_v| \leq |H_u| + d(u,v)$. 
+\end{observation}
+
+The following lemma demonstrates that heavy paths can ``pay'' for themselves.
+
+\begin{lemma}
+\label{lem:pathBounds}
+ Let T be a binary tree and let $\chi(T)$ and $P(T)$ be a valid coloring and path decomposition of $T$, respectively.
+ Then, 
+ \begin{compactenum}
+  \item for any heavy path $p\in P(T)$, $\sum_{v\in p} \log |H_v| = O(|p| \log |p|)$.
+  \item for any light path $p\in P(T)$, $\sum_{v\in p} \log |H_v| = O(|p| \log |H_{r_p}|)$.
+ \end{compactenum}
+\end{lemma}
+\begin{proof}
+ By $\Obs{decrease}$ we know that for any vertex $v\in p$, $|H_v|\leq |H_{r_p}| + |p|$ (as $r_p$ lies below $v$ on $p$).
+ As such, 
+ \[
+ \sum_{v\in p} \log |H_v| \leq \sum_{v\in p} \log(|H_{r_p}| + |p|)  = |p| \log (|H_{r_p}| + |p|).
+ \]
+ If $p$ is a heavy path then $|p| \log (|H_{r_p}| + |p|) = O(|p| \log |p|)$, and if $p$ is light then 
+ $|p| \log (|H_{r_p}| + |p|) = O(|p| \log |H_{r_p}| )$.
+\end{proof}
+
+We are now ready to define the maximum path decomposition, which will allow us to recharge the light paths to heavy paths.
+
+\subsubsection{The decomposition and its main property}
 
@@
  contains only vertices which correspond to light paths.  Call the subtree of $\pathTree$ consisting 
- of all light reachable paths from $p$ the \emph{light path tree} of $p$.
-\end{definition}
-
-Let $\mathcal{L}$ be a light path tree of some light path $r$ in a maximum path decomposition $\mathcal{P}(T)$.
-We now define a weight function $w(v)$ for all $v\in \mathcal{L}$.  Such that $w(v)=\Theta(|H_v|)$.  
-For this alternative quantity we will be able to show BLAH BLAH BLAH.
-
-%Our goal is to use \Lem{geometric}  to show that $\sum_{v\in \mathcal{L}} |H_v| = O(|H_r|)$, namely 
-%the total sum of the sizes of the heaps at the roots of all paths in the light path tree is on the 
-%order of the size of the heap at the root of $r$.  ADD THE ARGUMENT saying how this fact will be used.
+ of all light reachable paths from $p$ the \emph{light path tree} of $p$, and denote is $\mathcal{L}(p)$.
+\end{definition}
+
+In the following subsection we show how to exploit this geometric series behavior in light path trees.
+
+\subsubsection{An asymptotically equivalent weight function}
+Let $\mathcal{L} = \mathcal{L}(r)$ be a light path tree of some light path $r$ in a maximum path decomposition $\mathcal{P}(T)$.
+The intuitive idea is that due to the geometric decreasing behavior of light path trees described in the previous subsection, 
+one can show that for the entire light path tree our algorithm pays roughly the cost due to the root path $r$.  This argument 
+is hard to make directly, so instead we define an alternative quantity for which we can make the argument.
+Specifically, we now define a weight function $w(v)$ for all $v\in \mathcal{L}$, and show that $w(v)=\Theta(|H_v|)$.
 
@@
 such that $y$ is the parent of $x$, $w(x)\leq (4/5) w(y)$.
-Then $\sum_{u\in \mathcal{P}_v} \sqrt{w(u)}/100 \leq \sum_{c\in C_v} \sqrt{w(c)}/20$
+Then $\sum_{u\in \mathcal{P}_v} \sqrt{w(u)}/100 \leq \sum_{c\in C_v} \sqrt{w(c)}/10$
 \end{lemma}
@@
 
-By assumption from the lemma statement we have $w(u_{i+1})\leq (3/5) w(u_i)$.  Therefore by the above calculation, 
+By assumption from the lemma statement we have $w(u_{i+1})\leq (4/5) w(u_i)$.  Therefore by the above calculation, 
 the total reassigned to $u_m$ is
@@
 \[
- \sum_{i=0}^m \frac{w(u_m)}{100\sqrt{w(u_i)}} \leq \frac{w(u_m)}{100} \sum_{i=0}^m 1 \Big/ \sqrt{\pth{\frac{5}{3}}^iw(u_m)} 
- \leq \frac{\sqrt{w(u_m)}}{100} \sum_{i=0}^{\infty} \pth{\frac{3}{5}}^{i/2} \leq \frac{\sqrt{w(u_m)}}{20}
+ \sum_{i=0}^m \frac{w(u_m)}{100\sqrt{w(u_i)}} \leq \frac{w(u_m)}{100} \sum_{i=0}^m 1 \Big/ \sqrt{\pth{\frac{5}{4}}^iw(u_m)} 
+ \leq \frac{\sqrt{w(u_m)}}{100} \sum_{i=0}^{\infty} \pth{\frac{4}{5}}^{i/2} \leq \frac{\sqrt{w(u_m)}}{10}
 \]
@@
 \[
-|H_v|\geq w(v)-\sum_{u\in \mathcal{L}_v} \sqrt{w(u)}/100 \geq w(v)- \sum_{c\in C_v} \sqrt{w(c)}/20 \geq \frac{4}{5} w(v)
+|H_v|\geq w(v)-\sum_{u\in \mathcal{L}_v} \sqrt{w(u)}/100 \geq w(v)- \sum_{c\in C_v} \sqrt{w(c)}/10 \geq \frac{4}{5} w(v)
 \]
@@
 \end{lemma}
+% \begin{proof}
+% We will actually prove the stronger statement that for any $v\in \mathcal{L}$, 
+% we have $|H_v|\geq \frac{4}{5} w(v)$ and $w(v)\leq (4/5) w(p)$ where $p$ is the parent 
+% of $v$ in $\mathcal{L}$.
+% 
+% Suppose otherwise, and let $v$ be the lowest path in $\mathcal{L}$ such that this fails 
+% (i.e. the statement holds for all descendants of $v$ in $\mathcal{L}$).  Then for any 
+% two paths $x,y\in \mathcal{L}_v$ such that $y$ is the parent of $x$, 
+% we have that $w(x)\leq (4/5) w(y)$ and so by \Lem{distribution}, $|H_v|\geq \frac{4}{5} w(v)$.
+% Combining this fact with \Lem{geometric} gives,
+% \[
+% w(v) \leq \frac{5}{4}|H_v| \leq \frac{5}{4} \frac{3}{5} |H_p| \leq \frac{3}{4} w(p).
+% \]
+% 
+% \end{proof}
+%
 \begin{proof}
-We will actually prove the stronger statement that for any $v\in \mathcal{L}$, 
-we have $|H_v|\geq \frac{4}{5} w(v)$ and $w(v)\leq (4/5) w(p)$ for any two paths 
-$x,y\in \mathcal{L}_v$ such that $y$ is the parent of $x$.
-
-Suppose otherwise, and let $v$ be the lowest path in $\mathcal{L}$ such that this fails 
-(i.e. the statement holds for all descendants of $v$ in $\mathcal{L}$).  Then for any 
-two paths $x,y\in \mathcal{L}_v$ such that $y$ is the parent of $x$, 
-we have that $w(x)\leq (4/5) w(y)$ and so by \Lem{geometric},
-
-\end{proof}
+By the above claim, we know that $|H_v|\geq \frac{4}{5} w(v)$ if it holds that 
+$w(x)\leq (4/5) w(y)$ for any two paths $x,y\in \mathcal{L}_v$ such that $y$ is the parent of $x$.
+So let $v$ be the lowest path in $\mathcal{L}$ such that this fractional decrease fails to hold, i.e. $w(c)> (4/5) w(v)$ 
+where $c$ is a child of $v$ in $\mathcal{L}$, and conversely all vertices in $\mathcal{L}_c$ 
+the fractional decrease holds.  As the fractional decrease holds for a paths in $\mathcal{L}_c$, 
+we have $|H_c|\geq \frac{4}{5} w(c)$.  Therefore,
+\[
+w(c) \leq \frac{5}{4}|H_c| \leq \frac{5}{4} \frac{3}{5} |H_v| \leq \frac{3}{4} w(v).
+\]
+However, this is a contradiction with the assumption $w(c)> (4/5) w(v)$ , and hence the 
+fractional decrease always holds implying the claim always holds.
+\end{proof}
+
+\begin{corollary}
+ Let $\mathcal{L}$ be a light path tree of a light path $r$.  Then for any $v\in \mathcal{L}$, $w(v) = O(|H_v|)$.
+\end{corollary}
 

### 2014-08-31 paint-contour-tree.tex
Many changes in Section 5.2

@@
 \pagebreak
-\newcommand{\pathTree}{P_{\mathcal{T}}}
-\newcommand{\pathTreeA}{\mathcal{P}_{\mathcal{T}}}
+\newcommand{\pathTree}{P_{\mathcal{S}}}
+\newcommand{\pathTreeA}{\mathcal{P}_{\mathcal{S}}}
 \section{Lower Bound}
@@
 \begin{observation}
- The contour tree of an minimum dominant manifold (see \Def{dom}) is close to a merge tree.  Specifically, there is a root such that on any root 
+ The contour tree of a minimum dominant manifold (see \Def{dom}) is close to a merge tree.  Specifically, there is a root such that on any root 
  to leaf path height values strictly increase, with the only exception being potentially at the leaves themselves.  As such, up to constant 
@@
 To do so, we now introduce the concept of path decompositions and their cost.  Then in the following two subsections 
-we then relate path decompositions costs (defined below) to the running time of our algorithm using $\Lem{algCost}$.
+we relate path decomposition costs (defined below) to the running time of our algorithm using $\Lem{algCost}$.
 
@@
 Let $P(T)$ be a path decomposition of $T$.  Then the paths in $P(T)$ naturally define a tree\footnote{Please excuse the 
-overloading of the term 'tree', it is the most natural term to use here.} of there own.  Specifically, in the original 
-tree $T$ contract each path down to its root.  Call the resulting tree the \emph{path tree} of $T$ corresponding to 
+overloading of the term 'tree', it is the most natural term to use here.} of their own.  Specifically, in the original 
+tree $T$ contract each path down to its root.  Call the resulting tree the \emph{shrub} of $T$ corresponding to 
 the path decomposition $P(T)$, and denote it by $\pathTree(T)$.
 
-We say two paths in $P(T)$ are adjacent if their corresponding vertices are adjacent in the path tree.  Similarly, for 
-two adjacent paths in $P(T)$ we can define a parent/child relationship based on this relationship in the path tree.
+We say two paths in $P(T)$ are adjacent if their corresponding vertices are adjacent in the shrub.  Similarly, for 
+two adjacent paths in $P(T)$ we can define a parent/child relationship based on this relationship in the shrub.
 
@@
 
-We say two path trees $T_1$ and $T_2$ (of merge trees of potentially different surfaces) are equivalent if there is an isomorphism, $f$,
-between the two trees such that for any node $v\in T_1$ corresponding to a path of size $|p_v|$, $f(v)$ also corresponds to a path of size $|p_v|$.
-For the purposes of the next lemma, we say two 2-manifolds are different, if their corresponding contour trees are the different.
+We say two shrubs $S_1$ and $S_2$ (of merge trees of potentially different surfaces) are equivalent if there is an isomorphism, $f$,
+between the two shrubs such that for any node $v\in S_1$ corresponding to a path of size $|v|$, $f(v)$ also corresponds to a path of size $|v|$.
+For the purposes of the next lemma, we say two $2$-manifolds are different, if their corresponding contour trees are the different.
 
@@
 \label{lem:pathDecomp}
-Let $T$ be a merge tree, $P = P(T)$ any path decomposition of $T$, and $\pathTree$ the corresponding path tree.  
-Then there are at least $\Pi_{p_i\in P} |p_i|!$ different 2-manifolds with an equivalent path tree.
+Let $T$ be a merge tree, $P = P(T)$ any path decomposition of $T$, and $\pathTree$ the corresponding shrub.  
+Then there are at least $\Pi_{p_i\in P} |p_i|!$ different 2-manifolds with an equivalent shrub.
 \end{lemma}
@@
  (since all these saddles have a direct line of sight to this apex).  In particular, 
- this implies that, regardless of the heights assigned to the platforms, the merge tree of this mesh has a path decomposition whose corresponding path 
- tree is equivalent to $\pathTree$.
+ this implies that, regardless of the heights assigned to the platforms, the merge tree of this mesh has a path decomposition whose corresponding shrub is equivalent to $\pathTree$.
 
@@
  Given an input surface $S$ whose merge tree has a particular path decomposition $P$, by the above lemma, there are $\Pi_{p_i\in P} |p_i|!$ other surfaces 
- with distinct merge trees, all of which have a path tree equivalent to $\pathTree$.  Therefore, any algorithm which computes the merge tree must distinguish between 
+ with distinct merge trees, all of which have a shrub equivalent to $\pathTree$.  Therefore, any algorithm which computes the merge tree must distinguish between 
  these possibilities, and therefore its corresponding decision tree must have $\Pi_{p_i\in P} |p_i|!$ leaves.  Hence by Stirling's approximation, 
@@
 
+Throughout this section, for a path $p\in P(T)$, we use the notation $H_p$ to refer to the heap at the root of $p$, i.e. $H_p = H_{r_p}$.
+
 \subsubsection{Heavy and light paths}
@@
 of each path will pay for the algorithm's cost due to this path.  This is too good to hope for, so instead 
-we will divide paths into two categories, called heavy and light, which we show below in \Lem{pathBounds} roughly corresponds 
+we will divide paths into two categories, called heavy and light, which as we show below in \Lem{pathBounds} roughly corresponds 
 to whether a path can pay for itself or not.
@@
  Let T be a binary tree and let $\chi(T)$ and $P(T)$ be a valid coloring and path decomposition of $T$, respectively.  
- For $p\in P(T)$, we say that $p$ is \emph{light} if $|p| < \sqrt{|H_{r_p}|}/100$, and we say it is \emph{heavy} otherwise.
+ For $p\in P(T)$, we say that $p$ is \emph{light} if $|p| < \sqrt{|H_{p}|}/100$, and we say it is \emph{heavy} otherwise.
 \end{definition}
@@
  $w$ to the root has one less vertex than the path from $v$ to the root.
-
  In particular, we have the following more general property.
- Let $v$ and $u$ be any two vertices in the same root to leaf path of $T$, such that $u$ is lower than $v$.  
+ Let $v$ and $u$ be any two vertices in the same root to leaf path of $T$, such that $v$ is a descendant of $u$.  
  Then $|H_v| \leq |H_u| + d(u,v)$. 
 \end{observation}
+
+The following can be thought of as a generalization of the above observation, and will be useful in later sections.
+
+\begin{lemma}
+\label{lem:adjacent}
+ Let $T$ be a binary tree and let $\chi(T)$ and $P(T)$ be a valid coloring and path decomposition of $T$, respectively.
+ Let $p$ be any path in $P(T)$ and let $\{q_1, \dots q_k\}$ be the set of all adjacent child paths of $p$ in $\pathTree(T)$.
+ For $1\leq i\leq k$ let $r_i = r_{q_i}$ and let $H_i = H_{r_i}$.
+ Also, let $H_0$ denote the heap at the leaf of $p$.
+
+ Then $\sum_{i=0}^k |H_i| \leq |H_p|+|p|$, where $H_p = H_{r_p}$.
+\end{lemma}
+\begin{proof} 
+ First observe that for all $i$, $r_{i}$ lies above $r_p$ on some root to leaf path. 
+ Let $H_i'$ be the subset of $H_i$ that lies below $r_p$.  Observe that $H_i'\subseteq H_p$, as $L(T_{r_i})\subseteq L(T_{r_p})$.
+
+ Now for any $i\neq j$, $H_i \cap H_j = \emptyset$, as $r_{i}$ and $r_{j}$ are not on the same root to leaf path (i.e. $L(T_{r_i}) \cap L(T_{r_j}) = \emptyset$).
+ In particular, $(H_i\setminus H_i') \cap (H_j \setminus H_j') = \emptyset$, and so $\sum_{i=0}^k |H_i\setminus H_i'| \leq |p|$ 
+ as $H_i\setminus H_i'$ is precisely the subset of $H_i$ that lies on $p$.
+ Moreover, for any $i\neq j$, $ H_i' \cap H_j' = \emptyset$, and so since $H_i'\subseteq H_p$ we have $\sum_{i=0}^k |H_i'|\leq |H_p|$.
+ Putting these together gives
+ \[
+  \sum_{i=0}^k |H_i| = \sum_{i=0}^k |H_i'| + \sum_{i=0}^k |H_i\setminus H_i'| \leq |H_p|+|p|.
+ \]
+\end{proof}
 
@@
   \item for any heavy path $p\in P(T)$, $\sum_{v\in p} \log |H_v| = O(|p| \log |p|)$.
-  \item for any light path $p\in P(T)$, $\sum_{v\in p} \log |H_v| = O(|p| \log |H_{r_p}|)$.
+  \item for any light path $p\in P(T)$, $\sum_{v\in p} \log |H_v| = O(|p| \log |H_{p}|)$.
  \end{compactenum}
@@
 \begin{proof}
- By $\Obs{decrease}$ we know that for any vertex $v\in p$, $|H_v|\leq |H_{r_p}| + |p|$ (as $r_p$ lies below $v$ on $p$).
+ By $\Obs{decrease}$ we know that for any vertex $v\in p$, $|H_v|\leq |H_{p}| + |p|$ (as $v$ is a descendant of $r_p$ along $p$).
  As such, 
  \[
- \sum_{v\in p} \log |H_v| \leq \sum_{v\in p} \log(|H_{r_p}| + |p|)  = |p| \log (|H_{r_p}| + |p|).
+ \sum_{v\in p} \log |H_v| \leq \sum_{v\in p} \log(|H_{p}| + |p|)  = |p| \log (|H_{p}| + |p|).
  \]
- If $p$ is a heavy path then $|p| \log (|H_{r_p}| + |p|) = O(|p| \log |p|)$, and if $p$ is light then 
- $|p| \log (|H_{r_p}| + |p|) = O(|p| \log |H_{r_p}| )$.
-\end{proof}
-
-We are now ready to define the maximum path decomposition, which will allow us to recharge the light paths to heavy paths.
+ If $p$ is a heavy path then $|p| \log (|H_{p}| + |p|) = O(|p| \log |p|)$, and if $p$ is light then 
+ $|p| \log (|H_{p}| + |p|) = O(|p| \log |H_{p}| )$.
+\end{proof}
+
+For a light path $p$ we can think of the quantity $|p|(\log |H_p| - \log |p|)$ as the excess of $p$, i.e. what was not ``paid for'' locally by $p$.
+Our eventual goal will be to show that by choosing the right path decomposition, this excess 
+can be recharged to heavy paths. In the next couple sections we introduce the appropriate path decomposition and prove 
+some useful facts about it.  Then given these facts, in \Sec{} we are able to make the recharging argument.
 
@@
 \label{lem:geometric}
- Let $p$ and $q$ be two light paths such that $p$ is the parent of $q$ in the path tree $\pathTreeA$.  
+ Let $p$ and $q$ be two light paths such that $p$ is the parent of $q$ in the shrub $\pathTreeA$.  
  Then $|H_q|\leq \frac{3}{5} |H_p|$, where $H_p$ and $H_q$ are the heaps at the roots of $p$ and $q$, respectively. 
@@
 \begin{definition}
- Let $p$ be a path in a path decomposition $P(T)$, whose corresponding path tree is $\pathTree$.
- Call a light path $q\in P(T)$ light reachable from $p$ if the path from $q$ to $p$ in $\pathTree$ 
+ Let $p$ be a path in a path decomposition $P(T)$, with corresponding shrub $\pathTree$.
+ Call a light path $q\in P(T)$ light reachable from $p$ if the path from $p$ to $q$ in $\pathTree$ 
  contains only vertices which correspond to light paths.  Call the subtree of $\pathTree$ consisting 
- of all light reachable paths from $p$ the \emph{light path tree} of $p$, and denote is $\mathcal{L}(p)$.
-\end{definition}
-
-In the following subsection we show how to exploit this geometric series behavior in light path trees.
+ of all light reachable paths from $p$ the \emph{light shrub} of $p$, and denote it $\mathcal{L}(p)$.
+\end{definition}
+
+In the following subsection we show how to exploit this geometric series behavior of light shrubs.
 
 \subsubsection{An asymptotically equivalent weight function}
-Let $\mathcal{L} = \mathcal{L}(r)$ be a light path tree of some light path $r$ in a maximum path decomposition $\mathcal{P}(T)$.
-The intuitive idea is that due to the geometric decreasing behavior of light path trees described in the previous subsection, 
-one can show that for the entire light path tree our algorithm pays roughly the cost due to the root path $r$.  This argument 
+\label{sec:weight}
+Let $\mathcal{L} = \mathcal{L}(r)$ be a light shrub of some light path $r$ in a maximum path decomposition $\mathcal{P}(T)$.
+The intuitive idea is that due to the geometric decreasing behavior of light shrubs described in the previous subsection, 
+one can show that for the entire light shrub our algorithm pays roughly the cost due to the root path $r$.  This argument 
 is hard to make directly, so instead we define an alternative quantity for which we can make the argument.
@@
 For a path $v\in \mathcal{L}$, let $H_v$ denote the heap at the root of $v$, and let $H_{v_f}$ denote the heap 
-at the leaf of $v$.  Let $S_l(v)$ (resp. $S_h(v)$) be the subset of light (resp. heavy) paths adjacent to $v$ in $\mathcal{L}$.
+at the leaf of $v$.  Let $S_l(v)$ (resp. $S_h(v)$) be the subset of light (resp. heavy) paths adjacent to $v$ in $\pathTreeA(T)$.
 
@@
 was not already charged to descendant light paths.  We now define a weighted version of $\mathcal{L}$ based on these costs.  
-To do so, we first we augment $\mathcal{L}$ by attaching one 
+To do so, we first augment $\mathcal{L}$ by attaching one 
 leaf to each vertex in $\mathcal{L}$.  Now for each newly created leaf (which are the only leaves now) assign the cost 
@@
 
-As the weight $w(v)$ includes the sizes of the heaps of all contributing heavy paths and leaves in the subtree rooted at $v$, we have the following.
+As the weight $w(p$ includes the sizes of the heaps of all contributing heavy paths and leaves in the subtree rooted at $p$, we have the following.
 
 \begin{claim}
-For any $v\in \mathcal{L}$, $w(v)\geq |H_v|$.
+For any $p\in \mathcal{L}$, $w(p)\geq |H_p|$.  Moreover, for any $v\in p$, $w(p)\geq |H_v|$.
 \end{claim}
@@
 For any path $v$, let $\mathcal{L}_v$ (resp. $C_v$) be the subset of paths in $\mathcal{L}$ (resp $C$) 
-that are contained in the subtree of $\mathcal{L}$ which is rooted at $v$.
+that are in the subtree of $\mathcal{L}$ which is rooted at $v$.
 
 First, observe that $|H_v|\geq w(v)-\sum_{u\in \mathcal{L}_v} |u|$.  
-This holds by the same argument as in \Lem{adjacent}, namely, $w(u)$ by definition 
+This holds by the same argument as in \Lem{adjacent}, namely, $w(v)$ by definition 
 counts all the contributing heavy path heaps and initial leaf heaps that contribute 
@@
 \begin{lemma}
-Let $\mathcal{L}$ be the light path tree of some path $r$.  
+Let $\mathcal{L}$ be the light shrub of some path $r$.  
 Then for any $v\in \mathcal{L}$, we have $|H_v|\geq \frac{4}{5} w(v)$.
@@
 \begin{corollary}
- Let $\mathcal{L}$ be a light path tree of a light path $r$.  Then for any $v\in \mathcal{L}$, $w(v) = O(|H_v|)$.
+\label{cor:equivalent}
+ Let $\mathcal{L}$ be a light shrub of a light path $r$.  Then for any $v\in \mathcal{L}$, $|H_v| \leq w(v) \leq \frac{5}{4} |H_v|$.
 \end{corollary}
 
-%
-%
-% So $w(r)$ upper bound $H_r$, but by how much?  Consider a path $v\in \mathcal{L}$.  
-% As we move from the leaf to the root of $v$ we now want to understand 
-% how the heap can decreases.  Now the only way a vertex in the heap can leave that 
-% heap is if that vertex is one of the vertices we traverse in $v$, i.e. the total 
-% loss in vertices from the heap along $v$ is bounded by $|v|$.  
-% Therefore, $|H_r|\geq W-\sum_{v\in \mathcal{L}} |v|$.  
-% (This is an extension of \Lem{adjacent}, and probably should be merged with it.)
-% As $\mathcal{L}$ consists of only light paths we have $|v|\leq \sqrt{|H_v|}/100 \leq \sqrt{w(v)}/100$, 
-% and therefore $|H_r|\geq W-\sum_{v\in \mathcal{L}} \sqrt{w(v)}/100$.
-
-\pagebreak
-\subsection{Junk, don't read}
+Now we are finally ready to make use of the properties we have proven about the weight function we introduced.
+
+\Ben{The constant in the lemmas below, but it is no more than twice that.}
 
 \begin{lemma}
- Let $\chi(T)$ be a valid coloring of a tree $T$, and let $\mathcal{P}(T)$ be the corresponding maximum path decomposition.
- Let $p$ be a light path in $\mathcal{P}(T)$ and let $LPT(p)$ be the light path tree of $p$.
-
- Then $\sum_{q\in L(p)} \sum_{v\in q} \log |H_v| = |H_p|\log |H_p|$.
-\end{lemma}
-\begin{proof}UNDER CONSTRUCTION / NOTES:
-
- Claim: Let $q$ be some path of depth $i$ in $LPT(p)$, and let $H_q$ be the heap at the root of this path.
- Then $|H_q|< |H_p|/c^i$ for some constant $c>1$.  This should follow from using a maximum path decomposition.
-
- This claim would imply the depth of the tree is $O(\log |H_p|)$ and the total number of paths in the tree is $O(|H_p|)$, 
- thought I am not sure if either of these statements are useful.
-\end{proof}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{More Junk}
-
-The following can be viewed as stronger version of \Lem{adjacent}, which is possible as the path is now assumed to be maximum.
+\label{lem:lightCost}
+ Let $\mathcal{L}$ be a light shrub of a light path $r$.  Let $V$ be the set of heavy paths from the full shrub which are adjacent to paths in $\mathcal{L}$.  
+  Then $\sum_{p\in \mathcal{L}}\sum_{v\in p} \log |H_v| + \sum_{q\in V} |H_q|\log |H_q| \leq \frac{4}{3} |H_r|\log |H_r|$
+\end{lemma}
+\begin{proof}
+\Ben{
+The following is a proof for $\frac{3}{2}+\eps$, rather than $\frac{4}{3}$.  The discrepancy is due to a corrected calculation error.  
+4/3 can be achieved by tightening constants above, and is left in the lemma statement as it is used later.
+}
+
+Recall that in a proper coloring of a binary tree $T$, the color of any vertex must be the color of one of its descendants.  
+Therefore, any vertex in the light path shrub must appear in the heap of either the root of an adjacent 
+path from the full shrub or from the heap of a leaf of a path in $\mathcal{L}$.  
+Also note $|H_r|$ precisely counts sum of the number of vertices from these heaps that do not appear in $\mathcal{L}$.
+As there are no adjacent light paths to a light shrub, the total number of vertices over all paths in $\mathcal{L}$ is at most 
+$m = \sum_{p\in \mathcal{L}} |H_{p_f}| +\sum_{q\in V} |H_q| - |H_r| = w(r)-|H_r|$, where $H_{p_f}$ is the heap at the leaf of $p$.
+
+Using this observation and \Cor{equivalent} we then have, 
+\begin{align*}
+ \sum_{p\in \mathcal{L}}\sum_{v\in p} \log |H_v| + \sum_{q\in V} |H_q|\log |H_q|
+ &\leq \sum_{p\in \mathcal{L}}\sum_{v\in p} \log(w(r)) + \sum_{q\in V} |H_q|\log(w(r))\\
+ &\leq \log(w(r))\pth{ w(r)-|H_r| } + \log(w(r))w(r)\\ 
+ &= (2w(r)-|H_r|)\log(w(r)) \leq \pth{\frac{5}{2}|H_r|-|H_r|}\log\pth{\frac{5}{4}|H_r|}\\
+ &= \frac{3}{2}|H_r|\log\pth{\frac{5}{4}|H_r|}
+\end{align*}
+\end{proof}
+
+\subsubsection{Paying it down the line}
+
+% \begin{claim}
+% \label{clm:sizes}
+%  Let $p$ be a path in a maximum path decomposition, such that $|H_p|>c|p|$ for some constant $c> 6$.
+%  Let $H_f$ denote the heap at the leaf of $p$.  We then have the following
+%  \begin{compactenum}
+%   \item $|H_f|\leq (1+1/c)|H_p|$
+%   \item $|H_f|\geq |H_p|/3$
+%   \item For any path $q$ adjacent to $p$ in the shrub, $|H_q|<\frac{2}{3}|H_p|$
+%  \end{compactenum} 
+% \end{claim}
+% \begin{proof}
+%   For any vertex $v\in p$ \Obs{decrease} implies that $|H_v|\leq |H_p| + |p| < (1+1/c) |H_p|$.
+%   First observe that by setting $v=f$ this immediately implies the first part of the claim.
+%  % By a similar argument, $|H_f|\geq |H_p| - |p| > (1-1/c)|H_p|$.
+%   
+%   To prove the last part of the claim, suppose otherwise that $|H_q|\geq \frac{2}{3}|H_p|$ for some 
+%   adjacent path $q$.  As this is a maximum path decomposition, this implies that there is some 
+%   $v\in p$ such that $|H_v|\geq \frac{4}{3}|H_p|-1$.
+%   By \Obs{decrease} we then have, 
+%    $|H_p| \geq |H_v| - |p|\geq \frac{4}{3}|H_p| - 1 - |H_p|/c \geq (4/3-1/c)|H_p|-1$.
+%   When rearranged this gives $1\geq (1/3-1/c)|H_p|$.  As $c>6$ we have $1\geq |H_p|/6$, 
+%   which is a contradiction as $|H_p|>c|p|>6$.
+% \end{proof}
+
+\Ben{I had originally thought the following claim would be useful in proving the lemma below.}
+\begin{claim}
+\label{clm:sizes}
+ Let $p$ be a path in a maximum path decomposition, such that $|H_p|>c|p|$ for some constant $c> 6$.
+ We then have the following
+% \begin{compactenum}
+%  \item 
+  For any path $q$ adjacent to $p$ in the shrub, $|H_q|<\frac{2}{3}|H_p|$
+%  \item If there does not exist and adjacent path $q$ such $H_q\geq H_p/3$, then $|H_f|\geq |H_p|/3$
+% \end{compactenum} 
+\end{claim}
+\begin{proof}
+  To prove the claim, suppose otherwise that $|H_q|\geq \frac{2}{3}|H_p|$ for some 
+  adjacent path $q$.  As this is a maximum path decomposition, this implies that there is some 
+  $v\in p$ such that $|H_v|\geq \frac{4}{3}|H_p|-1$.
+  By \Obs{decrease} we then have, 
+   $|H_p| \geq |H_v| - |p|\geq \frac{4}{3}|H_p| - 1 - |H_p|/c \geq (4/3-1/c)|H_p|-1$.
+  When rearranged this gives $1\geq (1/3-1/c)|H_p|$.  As $c>6$ we have $1\geq |H_p|/6$, 
+  which is a contradiction as $|H_p|>c|p|>6$.  
+\end{proof}
 
 \begin{lemma}
- Let $\chi(T)$ be a valid coloring of a tree $T$, and let $\mathcal{P}(T)$ be the corresponding maximum path decomposition.
- Let $p$ be a light path in $\mathcal{P}(T)$ and let $\{q_1, \dots q_k\}$ be the set of all light paths adjacent to $p$.
-
- For $1\leq i\leq k$ let $r_i = r_{q_i}$ and let $H_i = H_{r_i}$.
-  Then $\sum_{i=1}^k |H_i| \leq |H_p|/c$ for some constant $c>1$, where $H_p = H_{r_p}$.
-\end{lemma}
-
-% \begin{proof}UNDER CONSTRUCTION: 
-% For simplicity, let $H_{k+1}$ denote the heap at the leaf vertex of $p$.  As the $H_k+1$ is disjoing from all other $H_{i}$, \Lem{adjacent} implies that 
-% $H_p > \sum_{j=1}^{k+1} H_i - |p|$.  
-% 
-% Let $x$ be the largest index such that $H_x > H_p / 3$.
-% By the above observation and since $p$ is a light path we have 
+\label{lem:excess}
+ Let $p$ be a heavy path, and let $S_1, \dots S_k$ and $R_1, \dots, R_l$ be the sizes of the heaps at the roots of the light and heavy children of $p$, respectively, in the shrub.  
+ Let $H_f$ be the heap at the leaf of the path $p$.
+ Then,  $|H_p| \log |H_p|   \geq  \sum_{v\in p} \log|H_v| + \frac{4}{3}\sum_{i=1}^k S_i \log S_i + \sum_{i=1}^l R_i \log R_i  -  c|p|\log |p|$, for some sufficiently large constant $c$
+ \end{lemma}
+\begin{proof}
+ First observe that the lemma holds for $|H_p|$ less than any fixed constant, since otherwise \Lem{adjacent} implies we can choose $c$ 
+ large enough such that $c|p|\log|p|$ will dominate all other terms.
+ Next, observe that $\sum_{v\in p} \log|H_v| \leq \sum_{v\in p} \log(|H_v|+|p|) = |p|\log(|H_v|+|p|) = O(|p|\log|p|)$, as $p$ is a heavy path.  
+ Therefore, in the remainder of the proof  we can safely ignore the $\sum_{v\in p} \log|H_v|$ term as we can choose $c$ large enough to cover it.  
+ That is we now instead prove the inequality,
+ \[
+  |H_p| \log |H_p|   \geq \frac{4}{3}\sum_{i=1}^k S_i \log S_i + \sum_{i=1}^l R_i \log R_i  -  c|p|\log |p|
+ \]
+ To do so we  break the analysis into two case based on the relative size of $|p|$ and $|H_p|$. 
+ In the following, $\alpha>10$ is some yet to be determined constant.
+
+\paragraph{Case 1, $|p|\geq |H_p|/\alpha$:\\}
+First observe that by \Lem{adjacent}, $|H_p|+|p| \geq \sum_{i=1}^k S_i+ \sum_{i=1}^l R_i$.
+By choosing $c$ large enough we therefore have,
+\begin{align*}
+ c|p|\log|p| &\geq 2(1+\alpha)^2 |p|\log |p| \geq 2(1+\alpha)|p| \log((1+\alpha)|p|)\\
+             &\geq 2(|H_p|+|p|)\log(|H_p|+|p|) \geq 2\log(|H_p|+|p|) \pth{\sum_{i=1}^k S_i + \sum_{i=1}^l R_i}\\
+             &\geq \frac{4}{3}\sum_{i=1}^k S_i \log S_i + \sum_{i=1}^l R_i \log R_i 
+\end{align*}
+
+In this case $\log(|H_p|+|p|)\leq \log(2|p|) \leq 1+\log(|p|)$ (where $\log$ is base $2$).
+Therefore, assuming $|p|>1$, we have 
+\[
+(|H_p|+|p|)\log(|H_p|+|p|)\leq (1+c)|p|\log((1+c)|p|) %(|H_p|+|p|)(1+\log |p|) \leq 2(1+c)|p|\log |p|
+\]
+which when combined with the inequality from \Lem{adjacent} gives, 
+\[
+ 4(1+c)|p|\log |p| \geq 2(|H_p|+|p|)\log(|H_p|+|p|)\geq \pth{|H_f| + \frac{4}{3}\sum_{i=1}^k S_i + \sum_{i=1}^l R_i}\log(|H_p|+|p|).
+\]
+This easily implies the desired inequality as $\log(H_p+p)$ is larger than any other $\log$ term in the inequality.
+
+\paragraph{Case 2, $|p|< |H_p|/\alpha$:\\}
+
+\Ben{Was not able to handle this case.}
+
+%  First, lets assume we can show $\sum_{i=1}^k S_i + \sum_{i=1}^l R_i \leq \frac{3}{4}|H_p|$.  
+%  In this case, using the last part of \Clm{sizes}, we can prove the desired inequality as follows.
+%  
+%  \begin{align*}
+%   \frac{4}{3}\sum_{i=1}^k S_i \log S_i + \sum_{i=1}^l R_i \log R_i  &\leq  \frac{4}{3} \pth{\sum_{i=1}^k S_i \log S_i + \sum_{i=1}^l R_i \log R_i} \\
+%   &\leq  \frac{4}{3} \log |H_p| \pth{\sum_{i=1}^k S_i + \sum_{i=1}^l R_i}  \leq  |H_p| \log |H_p|
+%  \end{align*}
+%  
+%  Finally, to complete the proof we must prove our assumption that $\sum_{i=1}^k S_i + \sum_{i=1}^l R_i \leq \frac{3}{4}|H_p|$.
+%  By \Lem{adjacent} and \Clm{sizes} we have, 
 %  \[
-%  H_p > \sum_{j=1}^{k} H_i + H_{k+1} - |p|
+%  \sum_{i=1}^k S_i + \sum_{i=1}^l R_i 
+%  \leq |H_p|+|p|-|H_f| 
+%  \leq (1+1/\alpha-1/3)|H_p| = (2/3+1/\alpha)|H_p|
 %  \]
+%  
+%  Therefore $\sum_{i=1}^k S_i + \sum_{i=1}^l R_i \leq \frac{3}{4}|H_p|$ if $(2/3 + 1/\alpha)\leq 3/4$, which rearranged gives $\alpha \geq 12$
+\end{proof}
+
+\remove{
+% \subsubsection{Paying it down the line}
+% 
+% \begin{lemma}
+% \label{lem:excess}
+%  Let $p$ be a heavy path, and let $S_1, \dots S_k$ and $R_1, \dots, R_l$ be the sizes of the heaps at the roots of the light and heavy children of $p$, respectively, in the shrub.  
+%  Let $H_f$ be the size of the heap at the leaf of the path $p$, and let $X = \sum_{i=0}^k H_i \log H_i$.  
+%  Then,  $|H_p| \log |H_p|   \geq  H_f\log H_f + \sum_{i=0}^k S_i \log S_i + \sum_{i=0}^l R_i \log R_i  -  c|p|\log |p|$, for some sufficiently large constant $c$
+%  \end{lemma}
+% \begin{proof}
+% By \Lem{adjacent} we know that $|H_p|+|p| \geq  H_f + \sum_{i=0}^k S_i + \sum_{i=0}^l R_i$.  
+% %Now as $p$ is a heavy path we have $|p|\geq \sqrt{|H_p|}/100$, and so, $(|H_p|+|p|)\log(|H_p|+|p|)$
+% 
+% We  break the analysis into two case based on the relative size of $|p|$ and $|H_p|$.  In the following $c$ is some yet to be determined constant greater than $1$.
+% 
+% \paragraph{Case 1, $|p|\geq |H_p|/c$:}  
+% In this case $\log(|H_p|+|p|)\leq \log(2|p|) \leq 1+\log(|p|)$ (where $\log$ is base $2$).
+% Therefore, assuming $|p|>1$, we have 
+% \[
+% (|H_p|+|p|)\log(H_p+p)\leq (|H_p|+|p|)(1+\log |p|) \leq (2|p|)(2\log |p|)\leq 4|p|\log |p|
+% \]
+% which when combined with the inequality from \Lem{adjacent} gives, 
+% \[
+%  4|p|\log |p| \geq (|H_p|+|p|)\log(|H_p|+|p|)\geq \pth{H_f + \sum_{i=0}^k S_i + \sum_{i=0}^l R_i}\log(|H_p|+|p|).
+% \]
+% This easily implies the claim as $\log(H_p+p)$ is larger than any other $\log$ term in the inequality.
+% 
+% \paragraph{Case 2, $|p|< |H_p|/c$:}
 % \end{proof}
+}
+
+Define the crust of the shrub to be the first (i.e. closest to the root) layer of light paths in the shrub.  Apply \Lem{lightCost} to each of the light paths in the crust and then 
+say by \Lem{excess} and since the root heap goes to zero we are done.
 

### 2014-09-17 paint-contour-tree.tex
modified authors
updating upper bounds

@@
 \newcommand{\cont}{\psi}
+\newcommand{\cost}{\mathop{cost}}
 \newcommand{\h}{high}
@@
 \newcommand{\paint}{{\tt paint}}
+\newcommand{\pmax}{P_{\max}}
 \newcommand{\rain}{{\tt rain}}
@@
 \author{
-  Seshadhri Comandur
+  Benjamin Raichel
   \and
-  Kashyap Dixit
-  \and
-  Benjamin Raichel
+  C. Seshadhri
 }
@@
 \title{Follow the Flow of Rain and Paint: \break An Output Sensitive Contour Tree Algorithm}
-%\date{}
+\date{}
 
@@
 \newcommand{\pathTreeA}{\mathcal{P}_{\mathcal{S}}}
-\section{Lower Bound}
-
-In this section we prove a lower bound on the time it takes to compute a merge tree.
+\section{Leaf assignments and path decompositions}
+
+In this section, we set up a framework to analyze the time taken to compute a merge tree.
 
@@
 
-First we setup some basic terminology.
 Any tree $T$ considered in this section will be a rooted full binary tree where the height of a vertex is its distance 
@@
 length of the path between $v$ and $w$.
+We use $A(v)$ to denote the set of ancestors of $v$.
+For set of nodes $U$, $A(U) = \bigcup_{u \in U} A(u)$.
 
 \begin{definition}
- A coloring of a tree $T$ is any function $\chi:L(T)\times E \rightarrow [n]$, where $L(T)$ is the set of leaves of $T$ and $n = |L(T)|$.  
- We say a coloring is valid if
- \begin{itemize}
-  \item The restriction $\chi:L(T) \rightarrow [n]$ is a bijection.
-  \item Let $e$ be any edge in $E$ and let $u$ be its higher endpoint, then $\chi(e) \in \chi(L(T_u))$. 
- \end{itemize}
- We consider each internal vertex as having two colors, namely the colors of the two edges for which it is a lower endpoint.
-\end{definition}
-
-For a vertex $v\in T$, let $H_v$ be the subset of 
-vertices along the path from the root to $v$ such that one of the two colors of each vertex is in $\chi(L(T_v))$.
-We will refer to $H_v$ as the \emph{heap} at $v$.
-Then define the \emph{algorithm} cost function as $f_{alg}(\chi(T)) = \sum_{v\in T} \log |H_v|$.
-
-\Ben{The following lemma should be immediate once previous section is written properly.}
+ A \emph{leaf assignment} $\chi$ of a tree $T$ assigns \emph{two} distinct leaves in each internal vertex $v$,
+ one from the left and one from the right subtree of $v$.
+\end{definition}
+
+For a vertex $v\in T$, we use $H_v$ to denote the \emph{heap} at $v$.
+Formally, $H_v = \{u | u \in A(v), \chi(u) \cap L(T_v) \neq \emptyset\}$.
+In words, $H_v$ is the set of ancestors of $v$ which are colored by some leaf in $T_v$.
+The following lemma should justify these technical definitions.
+
 \begin{lemma}
-\label{lem:algCost}
-NEEDS TO BE WRITTEN:
-Here we need a lemma showing that the running time of our algorithm is $O(f_{alg}(\chi(T)))$, where $T$ is the merge tree of the given input, 
-and $\chi$ is any valid coloring of $T$.  (maybe add one in the definition of $H_v$ so that $\log |H_v|$ is always $>1$).
-\end{lemma}
-
-Now that we have a simple expression upper bounding the running time of our algorithm, we spend the remainder 
-of this section proving that this running time is optimal.
-
-To do so, we now introduce the concept of path decompositions and their cost.  Then in the following two subsections 
-we relate path decomposition costs (defined below) to the running time of our algorithm using $\Lem{algCost}$.
+The running time of our algorithm is $O(\sum_{v \in T} \log |H_v|)$.
+% where $T$ is the merge tree of the given input, 
+%and $\chi$ is any valid coloring of $T$.  (maybe add one in the definition of $H_v$ so that $\log |H_v|$ is always $>1$).
+\end{lemma}
+
+\begin{proof} \Sesh{add}
+
+\end{proof}
+
+%vertices along the path from the root to $v$ such that one of the two colors of each vertex is in $L(T_v)$.
+%We will refer to $H_v$ as the \emph{heap} at $v$.
+%Then define the \emph{algorithm} cost function as $f_{alg}(\chi(T)) = \sum_{v\in T} \log |H_v|$.
+%
+%\begin{lemma}
+%\label{lem:algCost}
+%NEEDS TO BE WRITTEN:
+%Here we need a lemma showing that the running time of our algorithm is $O(f_{alg}(\chi(T)))$, where $T$ is the merge tree of the given input, 
+%and $\chi$ is any valid coloring of $T$.  (maybe add one in the definition of $H_v$ so that $\log |H_v|$ is always $>1$).
+%\end{lemma}
+%
+%Now that we have a simple expression upper bounding the running time of our algorithm, we spend the remainder 
+%of this section proving that this running time is optimal.
+
+We introduce the concept of path decompositions. A \emph{leaf path} is simply a contiguous path in $T$
+containing a leaf.
+% Then in the following two subsections 
+%we relate path decomposition costs (defined below) to the running time of our algorithm using $\Lem{algCost}$.
 
 \begin{definition}
-\label{def:path}
- Let $T$ be a rooted binary tree.  We call a path $p$ in $T$ a \emph{leaf path} if $p$ contains a leaf of $T$ and 
- is a contiguous sub-path of a root to leaf path in $T$.  A \emph{path decomposition}, $P(T)$, is a partition of $T$ into 
+\label{def:path} A \emph{path decomposition}, $P(T)$, is a partition of $T$ into 
  a set of vertex disjoint leaf paths.  Note that the number of paths in $P(T)$ must equal the number of leaves in $T$.
- For $p\in P(T)$, we use $|p|$ to denote its length and $r_p$ to denote the lowest vertex in $p$, which we call the \emph{root} of $p$.
-
- We define the cost of a particular path decomposition, $P(T)$, as $f_{path}(P(T)) = \sum_{p\in P(T)} |p| \log |p|$.
-\end{definition}
-
-Let $P(T)$ be a path decomposition of $T$.  Then the paths in $P(T)$ naturally define a tree\footnote{Please excuse the 
+
+ The \emph{cost} of $P(T)$ is $\cost(P(T)) = \sum_{p\in P(T)} |p| \log |p|$. (We use $|p|$ to denote the number of vertices in $p$.)
+\end{definition}
+
+Fix $\chi(T)$, which fixes all the heaps $H_v$. We choose a special path decomposition, which can be related
+to the running time of the algorithm. This is defined best as a subset of edges in $T$ such that each internal
+vertex has degree exactly $1$. This naturally gives as path decomposition.
+For each internal vertex $v\in T$, add the edge from $v$ to $\arg \max_{v_l, v_r} \{|H_{v_l}|, |H_{v_r}|\}$ 
+ where $v_l$ and $v_r$ are the children of $v$ (if $|H_{v_l}|=|H_{v_r}|$ then pick one arbitrarily).
+This is called the \emph{maximum} path decomposition, denoted by $\pmax(T)$.
+
+Our main theorem is the following.
+
+\begin{theorem} \label{thm:runtime} $\sum_{v \in T} \log |H_v| = O(\cost(\pmax(T)))$.
+\end{theorem}
+
+\subsection{Shrubs, heavy paths, and light paths}
+
+The paths in $P(T)$ naturally define a tree\footnote{Please excuse the 
 overloading of the term 'tree', it is the most natural term to use here.} of their own.  Specifically, in the original 
 tree $T$ contract each path down to its root.  Call the resulting tree the \emph{shrub} of $T$ corresponding to 
-the path decomposition $P(T)$, and denote it by $\pathTree(T)$.
+the path decomposition $P(T)$. Abusing notation, we simply use $P(T)$ to denote the shrub.
+As a result, we use terms like `parent', `child', `sibling', etc. for paths as well.
+The shrub gives a handle on the heaps of a path.
+We use $b(p)$ to denote the \emph{base} of the path, which is vertex in $p$ closest
+to root of $T$. We use $\ell(p)$ to denote the leaf in $p$.
+We use $H_p$ to denote the $H_{b(p)}$.
+% and denote it by $\pathTree(T)$.
+
+\begin{lemma}
+\label{lem:adjacent}
+ Let $p$ be any path in $P(T)$ and let $\{q_1, \dots q_k\}$ be the children on $p$.
+Then $H_{\ell(p)} + \sum_{i=1}^k |H_{q_i}| \leq |H_p|+2|p|$. 
+\end{lemma}
+
+\begin{proof} For convenience, denote $H_i = H_{q_i}$ and $H_0 = H_{\ell(p)}$.
+Consider $v \in \bigcup_{i} H_i$ that lies below
+$b(p)$ in $T$. Note that $v \in H_p$ and has only one color in $\bigcup_{i} L(T_(b(q_i))) \cup \ell(p)$.
+Since the latter sets are all disjoint, any $u \in H_i \cap H_j$ must lie in $p$.
+Any $u \in p$ can be in at most two of the $H_i$'s. Hence, $\sum_i |H_i| \leq |H_p| + 2|p|$.
+\ignore{
+ First observe that for all $i$, $r_{i}$ lies above $r_p$ on some root to leaf path. 
+ Let $H_i'$ be the subset of $H_i$ that lies below $r_p$.  Observe that $H_i'\subseteq H_p$, as $L(T_{r_i})\subseteq L(T_{r_p})$.
+
+ Now for any $i\neq j$, $H_i \cap H_j = \emptyset$, as $r_{i}$ and $r_{j}$ are not on the same root to leaf path (i.e. $L(T_{r_i}) \cap L(T_{r_j}) = \emptyset$).
+ In particular, $(H_i\setminus H_i') \cap (H_j \setminus H_j') = \emptyset$, and so $\sum_{i=0}^k |H_i\setminus H_i'| \leq |p|$ 
+ as $H_i\setminus H_i'$ is precisely the subset of $H_i$ that lies on $p$.
+ Moreover, for any $i\neq j$, $ H_i' \cap H_j' = \emptyset$, and so since $H_i'\subseteq H_p$ we have $\sum_{i=0}^k |H_i'|\leq |H_p|$.
+ Putting these together gives
+ \[
+  \sum_{i=0}^k |H_i| = \sum_{i=0}^k |H_i'| + \sum_{i=0}^k |H_i\setminus H_i'| \leq |H_p|+|p|.
+ \]
+}
+\end{proof}
+
+We wish to prove $\sum_{v\in T} \log |H_v| = O(\sum_{p\in P} |p| \log |p|)$. 
+The simplest approach is to prove $\forall p\in P$, $\sum_{v\in p} \log |H_v|= O(|p|\log|p|)$.
+This is unfortunately not true, which is why we divide paths into two categories.
+
+\begin{definition}
+ For $p\in P(T)$, $p$ is \emph{light} if $|p| < \sqrt{|H_{p}|}/100$, and \emph{heavy} otherwise.
+\end{definition}
+
+\ignore{
+\begin{observation}
+\label{obs:decrease}
+ Let $v$ be a vertex in $T$ and let $w$ be its parent.  Then $|H_w| \geq |H_v| -1$, as $L(T_v) \subseteq L(T_w)$ and the path from 
+ $w$ to the root has one less vertex than the path from $v$ to the root.
+ In particular, we have the following more general property.
+ Let $v$ and $u$ be any two vertices in the same root to leaf path of $T$, such that $v$ is a descendant of $u$.  
+ Then $|H_v| \leq |H_u| + d(u,v)$. 
+\end{observation}
+
+The following can be thought of as a generalization of the above observation, and will be useful in later sections.
+}
+
+The following lemma demonstrates that heavy paths can ``pay'' for themselves.
+
+\begin{lemma}
+\label{lem:pathBounds} If $p$ is heavy, $\sum_{v\in p} \log |H_v| = O(|p| \log |p|)$.
+If $p$ is light, $\sum_{v\in p} \log |H_v| = O(|p| \log |H_p|)$.
+% Let T be a binary tree and let $\chi(T)$ and $P(T)$ be a valid coloring and path decomposition of $T$, respectively.
+% Then, 
+% \begin{compactenum}
+%  \item for any heavy path $p\in P(T)$, $\sum_{v\in p} \log |H_v| = O(|p| \log |p|)$.
+%  \item for any light path $p\in P(T)$, $\sum_{v\in p} \log |H_v| = O(|p| \log |H_{p}|)$.
+% \end{compactenum}
+\end{lemma}
+\begin{proof} \ignore{
+ By $\Obs{decrease}$ we know that for any vertex $v\in p$, $|H_v|\leq |H_{p}| + |p|$ (as $v$ is a descendant of $r_p$ along $p$).
+}
+For $v \in p$, $|H_v|\leq |H_{p}| + |p|$ (as $v$ is a descendant of $b(p)$ along $p$).
+Hence, $\sum_{v\in p} \log |H_v| \leq \sum_{v\in p} \log(|H_{p}| + |p|)  = |p| \log (|H_{p}| + |p|)$.
+ If $p$ is a heavy path, then $|p| \log (|H_{p}| + |p|) = O(|p| \log |p|)$. If $p$ is light, then 
+ $|p| \log (|H_{p}| + |p|) = O(|p| \log |H_{p}| )$.
+\end{proof}
+
+\ignore{
+For a light path $p$ we can think of the quantity $|p|(\log |H_p| - \log |p|)$ as the excess of $p$, i.e. what was not ``paid for'' locally by $p$.
+Our eventual goal will be to show that by choosing the right path decomposition, this excess 
+can be recharged to heavy paths. In the next couple sections we introduce the appropriate path decomposition and prove 
+some useful facts about it.  Then given these facts, in \Sec{} we are able to make the recharging argument.
+}
+
+We have a claim that also bounds heap sizes for special set of paths.
+
+\begin{claim} \label{clm:above} Let $Q$ be a set of incomparable nodes in the shrub $P(T)$ (which are paths in $T$).
+Suppose all paths in $A(Q)$ are heavy.
+Then $\sum_{q \in Q} |H_q| \log |H_q| = O(\sum_{p \in A(Q)} |p|\log |p|)$.
+\end{claim}
+
+\begin{proof} Pick $q \in Q$ and some $v \in H_q$. As we traverse up the tree passing
+through $A(q)$, we must encounter some $p \in A(q)$ such that $v \in p$. Furthermore,
+there is an order in which these vertices (of $H_q$) occur. Let $H'$ be the first half
+of vertices that occur. For all $v \in H'$, $|H_v| \geq |H_q|/2$. Hence,
+$|H_q| \log |H_q| \leq 4 \sum_{v \in H'} \log |H_v| \leq 4 \sum_{v \in H_q} |H_v|$.
+So $\sum_{q \in Q} |H_q| \log |H_q| \leq 4 \sum_{q \in Q} \sum_{v \in H_q} |H_v|$.
+
+Since nodes in $Q$ are incomparable, all the $H_q$'s (for $q \in Q$) are disjoint.
+Each such $v$ must occur in some path $p \in A(Q)$. Thus,
+$\sum_{q \in Q} |H_q| \log |H_q| \leq 4 \sum_{p \in A(Q)} \sum_{v \in p} |H_v|$.
+We apply the bound of \Lem{pathBounds} to complete the proof.
+\end{proof}
+
+Now, for the critical piece of the approach. Focus on $\pmax(T)$. Let $L$ be the set of light paths, 
+and $\hat{L} \subseteq L$ be the set of paths such that no ancestor of $p \in L'$
+is light. Obviously, $\hat{L}$ is an anti-chain in $P(T)$.
+
+\begin{lemma} \label{lem:light} $\sum_{p \in L} \sum_{v \in p} \log |H_v| = O(\sum_{p \in L'} |H_p| \log |H_p|)$.
+\end{lemma}
+
+With this lemma, the proof of \Thm{runtime} is straightforward.
+
+\begin{proof} (of \Thm{runtime}) We split into heavy and light paths.
+We have $\sum_{v \in T} \log |H_v| = \sum_{p \in L} \sum_{v \in p} \log |H_v| + \sum_{p \notin L} \sum_{v \in p} \log |H_v|$.
+For the first term, we apply \Lem{light} to bound by $O(\sum_{p \in L'} |H_p| \log |H_p|)$.
+Now, we apply \Clm{above} to the set $L'$, noting that it satifies the claim conditions and that $A(Q) \cap L = \emptyset$.
+Hence, the first term is $O(\sum_{p \notin L} |p|\log |p|)$.
+The latter is also bounded by the same expression, by \Lem{pathBounds}. By definition,
+$\sum_{p \notin L} |p|\log |p| \leq \cost(\pmax(T))$.
+\end{proof}
+
+The main challenge is proving \Lem{light}, which we defer to the next section.
+
+\ignore{
+\begin{definition}
+ Let $\chi(T)$ be a valid coloring of a binary tree $T$.  We define a set of disjoint paths over the vertices of 
+ $T$ as follows.  
+ The corresponding collection of paths that all such edges correspond to is a path decomposition of $T$, which 
+ we refer to as a \emph{maximum} path decomposition, and is denoted by $\mathcal{P}(T)$.  
+\end{definition}
+}
+
+\subsection{Upper Bound by Path Decomposition}
+
+\ignore{
+From the previous section we now have a lower bound of $\Omega(f_{path}(P))$ for computing the merge tree in terms of any path decomposition $P$. 
+We now show that the running time of our algorithm for computing the merge tree is upper bounded by this expression for a specific path decomposition, 
+which we will call the maximum path decomposition (defined below).
+Specifically, for a given valid coloring $\chi$ of a tree $T$, we will prove that for the maximum path decomposition $f_{alg}(\chi(T)) = O(f_{path}(P(T)))$. 
+%using the observations about path decompositions from 
+%\Sec{background}, which will then imply $f_{alg}(\chi(T)) = \Theta(f_{path}(P(T)))$.
+
+Throughout this section, for a path $p\in P(T)$, we use the notation $H_p$ to refer to the heap at the root of $p$, i.e. $H_p = H_{r_p}$.
+}
+
+We focus on $\pmax(T)$, and prove the key property of this path decomposition.
+
+\begin{lemma}
+\label{lem:geometric}
+ Let $p$ be the parent of $q$ in the shrub $\pmax(T)$.  
+ Then $|H_q|\leq \frac{3}{5} |H_p|$. 
+\end{lemma}
+\begin{proof}
+ Let $w$ be the vertex in the path $p$ to which the root of the path $q$ is adjacent to in $T$, 
+ i.e. $w$ is the parent of $r_q$ in $T$.  Let $c$ be the other child of $w$, which lies in $p$.
+ As $\mathcal{P}(T)$ is a maximum path decomposition of $T$ and $c$ rather than $r_q$ is in 
+ the same path as $w$, we have $|H_q|\leq |H_c|$.  Therefore, $|H_w|\geq |H_q|+|H_c|-1\geq 2|H_q|-1$. 
+ Therefore, by \Obs{decrease}, $|H_p|+ d(r_p, w) \geq |H_w|\geq 2|H_q|-1$.  
+ Since $p$ is a light path, $d(r_p, w)\leq \sqrt{|H_p|}/100$ and so we have $|H_p|+\sqrt{|H_p|}/100\geq 2|H_q|-1$.
+ Assuming $|H_p|>11$ (and we can safely assume $|H_p|$ is bigger than any constant), this yields $\frac{3}{5} |H_p| \geq |H_q|$
+\end{proof}
+
+\ignore{
+Naturally, this geometric series behavior can be extended to more than two adjacent light paths, thus motivating 
+the following definition.
+
+\begin{definition}
+ Let $p$ be a path in a path decomposition $P(T)$, with corresponding shrub $\pathTree$.
+ Call a light path $q\in P(T)$ light reachable from $p$ if the path from $p$ to $q$ in $\pathTree$ 
+ contains only vertices which correspond to light paths.  Call the subtree of $\pathTree$ consisting 
+ of all light reachable paths from $p$ the \emph{light shrub} of $p$, and denote it $\mathcal{L}(p)$.
+\end{definition}
+
+In the following subsection we show how to exploit this geometric series behavior of light shrubs.
 
@@
 two adjacent paths in $P(T)$ we can define a parent/child relationship based on this relationship in the shrub. 
+}
 
@@
 
-\subsection{Lower Bound by Path Decomposition}
-
-In this section we show that any algorithm for computing the merge tree $T$ must pay at least $f_{path}(P(T))$ (see \Def{path}) for any path decomposition $P(T)$.
-
-We say two shrubs $S_1$ and $S_2$ (of merge trees of potentially different surfaces) are equivalent if there is an isomorphism, $f$,
-between the two shrubs such that for any node $v\in S_1$ corresponding to a path of size $|v|$, $f(v)$ also corresponds to a path of size $|v|$.
-For the purposes of the next lemma, we say two $2$-manifolds are different, if their corresponding contour trees are the different.
-
-\begin{figure}[h]\centering
-    \includegraphics[width=.15\linewidth]{figs/diamond}\hspace{2cm}
-    \includegraphics[width=.125\linewidth]{figs/topView}\hspace{2cm}
-    \includegraphics[width=.095\linewidth]{figs/sideView}
-    \caption{From left to right: the cone portion of a diamond; the top down view and then the side view of a cone face with a platform added.}
-    \label{fig:diamond}
-\end{figure}
-
-\begin{lemma}
-\label{lem:pathDecomp}
-Let $T$ be a merge tree, $P = P(T)$ any path decomposition of $T$, and $\pathTree$ the corresponding shrub.  
-Then there are at least $\Pi_{p_i\in P} |p_i|!$ different 2-manifolds with an equivalent shrub.
-\end{lemma}
-\begin{proof}
- We describe a simple input mesh, with heights specified for all vertices except for the saddles.  The main idea is that 
- this mesh will be flexible enough to allow almost any assignment of heights to the saddles, thus defining a class of merge trees 
- whose size is that described in the lemma statement.
-
- First, we describe the basic building block of this input mesh, which roughly speaking corresponds to a fixed path $p\in P$.  
- Specifically, we say a \emph{diamond} of size $m$ is the following mesh.  To start, construct a cycle 
- of length $m$, where the height of the $i$th vertex in the cycle is say $i/2m$.  Next, create an apex vertex at height 
- $\infty$ (or any appropriately large finite height), and add an edge between this vertex and each vertex in the cycle.  At this point 
- we have built an upside down cone with $m$ triangular faces (see \Fig{diamond}). To make this cone a closed surface we add a triangular cap at the bottom. 
- Specifically, add a $3$ cycle of vertices at heights $0$, $1/20m$, and $1/10m$, and connect each vertex of this $3$-cycle to an arbitrary distinct 
- vertex on the $m$ cycle (with matching cyclic ordering), and triangulate all remaining faces.  For consistency, if the size $m$ is $\leq 3$, the diamond 
- will be just a tetrahedron.
-
- Now above each face of the cone of the diamond, we will create a triangular ``platform'', which will act as the location to glue on other (shrunken) diamonds, 
- and whose highest point of connection to the face below will be a saddle vertex.  Specifically the heights of the vertices of the triangle will be $h$, $h-\eps$, and $h-2\eps$, where 
- $h$ is some unspecified height value.  This near horizontal platform is then connected to its respective face with vertical edges down to the face. See \Fig{diamond}.
- %
- %The vertex at height $h$ will lie in the plane of the respective face of the cone, and the other two will lie above the face, 
- %with edges connecting down to the face, hence creating an actual ``platform''.
-
- At this point we can construct the entire input mesh, by gluing together diamonds.  Specifically, for each path $p_i\in P$ we create a diamond of size $p_i$.
- Now we glue together diamonds of different paths in the same way the paths are connected in $\pathTree$.  Specially, for two paths $p,q\in P$ where 
- $p$ is the parent of $q$ in $\pathTree$, we glue the bottom triangular cap of the diamond for $q$ onto the triangular ``platform'' of some face of the diamond for $p$.
- (Naturally for this construction to work, diamonds for a given path will be scaled down relative to the size of the diamond of their parent).
-
- Observe now that the only saddle points in this construction are the highest connecting point on a face of an edge from the platform above.  Moreover, the only maxima are the apexes of the 
- diamonds, and there is exactly one minima corresponding to the bottom cap of the diamond representing the root of $\pathTree$.  Therefore, the saddles on a given 
- diamond will appear contiguously on a root to leaf path in the merge tree of the mesh, where the leaf corresponds to the maxima of the diamond 
- (since all these saddles have a direct line of sight to this apex).  In particular, 
- this implies that, regardless of the heights assigned to the platforms, the merge tree of this mesh has a path decomposition whose corresponding shrub is equivalent to $\pathTree$.
-
- Now there is a valid instantiation of this described mesh for any relative ordering of the heights of the saddles on a given diamond.  In particular, 
- there are $|p|!$ possible orderings of the heights of the saddles on the diamond for $p$, and hence $\Pi_{p_i\in P} |p_i|!$ possible instantiations of 
- the mesh overall.  Also, each one of these instantiations will result in a different merge.  Specifically, a permutation of the heights on a given diamond 
- corresponds to a permutation of the vertices along a path in $P$, as by construction all saddles on a given diamond will appear in sorted order in the merge tree.
-\end{proof}
-
-\begin{corollary}
- Let $A(S)$ be any algorithm to compute the merge tree of any input surface $S$ (whose contour tree is a merge tree).  
- Then the running time of $A(S)$ is $\Omega(f_{path}(P))=\Omega(\sum_{p_i\in P} |p_i|\log|p_i|)$ where $P$ is any path decomposition of the 
- merge tree of $S$.
-\end{corollary}
-\begin{proof}
- Given an input surface $S$ whose merge tree has a particular path decomposition $P$, by the above lemma, there are $\Pi_{p_i\in P} |p_i|!$ other surfaces 
- with distinct merge trees, all of which have a shrub equivalent to $\pathTree$.  Therefore, any algorithm which computes the merge tree must distinguish between 
- these possibilities, and therefore its corresponding decision tree must have $\Pi_{p_i\in P} |p_i|!$ leaves.  Hence by Stirling's approximation, 
- this decision tree must have depth $\Omega(\sum_{p_i\in P} |p_i|\log|p_i|)$.
-\end{proof}
-
-\subsection{Upper Bound by Path Decomposition}
-\Ben{This section is a work in progress.}
-
-From the previous section we now have a lower bound of $\Omega(f_{path}(P))$ for computing the merge tree in terms of any path decomposition $P$. 
-We now show that the running time of our algorithm for computing the merge tree is upper bounded by this expression for a specific path decomposition, 
-which we will call the maximum path decomposition (defined below).
-Specifically, for a given valid coloring $\chi$ of a tree $T$, we will prove that for the maximum path decomposition $f_{alg}(\chi(T)) = O(f_{path}(P(T)))$. 
-%using the observations about path decompositions from 
-%\Sec{background}, which will then imply $f_{alg}(\chi(T)) = \Theta(f_{path}(P(T)))$.
-
-Throughout this section, for a path $p\in P(T)$, we use the notation $H_p$ to refer to the heap at the root of $p$, i.e. $H_p = H_{r_p}$.
-
-\subsubsection{Heavy and light paths}
-
-Before describing the maximum path decomposition, we introduce the notion of heavy and light paths 
-(which will help motivate what the decomposition should be).
-
-Recall we are ultimately trying to show $f_{alg}(\chi(T)) = O(f_{path}(P(T)))$ for some path decomposition $P=P(T)$, 
-or more specifically $\sum_{v\in T} \log |H_v| = O(\sum_{p\in P} |p| \log |p|)$.  In particular, if we could 
-show for each path $p\in P$ that $\sum_{v\in p} \log |H_v|= |p|\log|p|$ then we would be done.
-In other words we can think of each path as having a budget of $|p|\log|p|$, and ideally the budget 
-of each path will pay for the algorithm's cost due to this path.  This is too good to hope for, so instead 
-we will divide paths into two categories, called heavy and light, which as we show below in \Lem{pathBounds} roughly corresponds 
-to whether a path can pay for itself or not.
-
-\begin{definition}
- Let T be a binary tree and let $\chi(T)$ and $P(T)$ be a valid coloring and path decomposition of $T$, respectively.  
- For $p\in P(T)$, we say that $p$ is \emph{light} if $|p| < \sqrt{|H_{p}|}/100$, and we say it is \emph{heavy} otherwise.
-\end{definition}
-
-\begin{observation}
-\label{obs:decrease}
- Let $v$ be a vertex in $T$ and let $w$ be its parent.  Then $|H_w| \geq |H_v| -1$, as $L(T_v) \subseteq L(T_w)$ and the path from 
- $w$ to the root has one less vertex than the path from $v$ to the root.
- In particular, we have the following more general property.
- Let $v$ and $u$ be any two vertices in the same root to leaf path of $T$, such that $v$ is a descendant of $u$.  
- Then $|H_v| \leq |H_u| + d(u,v)$. 
-\end{observation}
-
-The following can be thought of as a generalization of the above observation, and will be useful in later sections.
-
-\begin{lemma}
-\label{lem:adjacent}
- Let $T$ be a binary tree and let $\chi(T)$ and $P(T)$ be a valid coloring and path decomposition of $T$, respectively.
- Let $p$ be any path in $P(T)$ and let $\{q_1, \dots q_k\}$ be the set of all adjacent child paths of $p$ in $\pathTree(T)$.
- For $1\leq i\leq k$ let $r_i = r_{q_i}$ and let $H_i = H_{r_i}$.
- Also, let $H_0$ denote the heap at the leaf of $p$.
-
- Then $\sum_{i=0}^k |H_i| \leq |H_p|+|p|$, where $H_p = H_{r_p}$.
-\end{lemma}
-\begin{proof} 
- First observe that for all $i$, $r_{i}$ lies above $r_p$ on some root to leaf path. 
- Let $H_i'$ be the subset of $H_i$ that lies below $r_p$.  Observe that $H_i'\subseteq H_p$, as $L(T_{r_i})\subseteq L(T_{r_p})$.
-
- Now for any $i\neq j$, $H_i \cap H_j = \emptyset$, as $r_{i}$ and $r_{j}$ are not on the same root to leaf path (i.e. $L(T_{r_i}) \cap L(T_{r_j}) = \emptyset$).
- In particular, $(H_i\setminus H_i') \cap (H_j \setminus H_j') = \emptyset$, and so $\sum_{i=0}^k |H_i\setminus H_i'| \leq |p|$ 
- as $H_i\setminus H_i'$ is precisely the subset of $H_i$ that lies on $p$.
- Moreover, for any $i\neq j$, $ H_i' \cap H_j' = \emptyset$, and so since $H_i'\subseteq H_p$ we have $\sum_{i=0}^k |H_i'|\leq |H_p|$.
- Putting these together gives
- \[
-  \sum_{i=0}^k |H_i| = \sum_{i=0}^k |H_i'| + \sum_{i=0}^k |H_i\setminus H_i'| \leq |H_p|+|p|.
- \]
-\end{proof}
-
-The following lemma demonstrates that heavy paths can ``pay'' for themselves.
-
-\begin{lemma}
-\label{lem:pathBounds}
- Let T be a binary tree and let $\chi(T)$ and $P(T)$ be a valid coloring and path decomposition of $T$, respectively.
- Then, 
- \begin{compactenum}
-  \item for any heavy path $p\in P(T)$, $\sum_{v\in p} \log |H_v| = O(|p| \log |p|)$.
-  \item for any light path $p\in P(T)$, $\sum_{v\in p} \log |H_v| = O(|p| \log |H_{p}|)$.
- \end{compactenum}
-\end{lemma}
-\begin{proof}
- By $\Obs{decrease}$ we know that for any vertex $v\in p$, $|H_v|\leq |H_{p}| + |p|$ (as $v$ is a descendant of $r_p$ along $p$).
- As such, 
- \[
- \sum_{v\in p} \log |H_v| \leq \sum_{v\in p} \log(|H_{p}| + |p|)  = |p| \log (|H_{p}| + |p|).
- \]
- If $p$ is a heavy path then $|p| \log (|H_{p}| + |p|) = O(|p| \log |p|)$, and if $p$ is light then 
- $|p| \log (|H_{p}| + |p|) = O(|p| \log |H_{p}| )$.
-\end{proof}
-
-For a light path $p$ we can think of the quantity $|p|(\log |H_p| - \log |p|)$ as the excess of $p$, i.e. what was not ``paid for'' locally by $p$.
-Our eventual goal will be to show that by choosing the right path decomposition, this excess 
-can be recharged to heavy paths. In the next couple sections we introduce the appropriate path decomposition and prove 
-some useful facts about it.  Then given these facts, in \Sec{} we are able to make the recharging argument.
-
-\subsubsection{The decomposition and its main property}
-
-\begin{definition}
- Let $\chi(T)$ be a valid coloring of a binary tree $T$.  We define a set of disjoint paths over the vertices of 
- $T$ as follows.  For each internal vertex $v\in T$ add the edge from $v$ to $\arg \max_{v_l, v_r} \{|H_{v_l}|, |H_{v_r}|\}$ 
- where $v_l$ and $v_r$ are the children of $v$ (if $|H_{v_l}|=|H_{v_r}|$ then pick one arbitrarily).
- The corresponding collection of paths that all such edges correspond to is a path decomposition of $T$, which 
- we refer to as a \emph{maximum} path decomposition, and is denoted by $\mathcal{P}(T)$.  
-\end{definition}
-
-We now prove that in a maximum path decomposition, when going between adjacent light paths, the heap size decreases by a constant fraction.
-
-\begin{lemma}
-\label{lem:geometric}
- Let $p$ and $q$ be two light paths such that $p$ is the parent of $q$ in the shrub $\pathTreeA$.  
- Then $|H_q|\leq \frac{3}{5} |H_p|$, where $H_p$ and $H_q$ are the heaps at the roots of $p$ and $q$, respectively. 
-\end{lemma}
-\begin{proof}
- Let $w$ be the vertex in the path $p$ to which the root of the path $q$ is adjacent to in $T$, 
- i.e. $w$ is the parent of $r_q$ in $T$.  Let $c$ be the other child of $w$, which lies in $p$.
- As $\mathcal{P}(T)$ is a maximum path decomposition of $T$ and $c$ rather than $r_q$ is in 
- the same path as $w$, we have $|H_q|\leq |H_c|$.  Therefore, $|H_w|\geq |H_q|+|H_c|-1\geq 2|H_q|-1$. 
- Therefore, by \Obs{decrease}, $|H_p|+ d(r_p, w) \geq |H_w|\geq 2|H_q|-1$.  
- Since $p$ is a light path, $d(r_p, w)\leq \sqrt{|H_p|}/100$ and so we have $|H_p|+\sqrt{|H_p|}/100\geq 2|H_q|-1$.
- Assuming $|H_p|>11$ (and we can safely assume $|H_p|$ is bigger than any constant), this yields $\frac{3}{5} |H_p| \geq |H_q|$
-\end{proof}
-
-Naturally, this geometric series behavior can be extended to more than two adjacent light paths, thus motivating 
-the following definition.
-
-\begin{definition}
- Let $p$ be a path in a path decomposition $P(T)$, with corresponding shrub $\pathTree$.
- Call a light path $q\in P(T)$ light reachable from $p$ if the path from $p$ to $q$ in $\pathTree$ 
- contains only vertices which correspond to light paths.  Call the subtree of $\pathTree$ consisting 
- of all light reachable paths from $p$ the \emph{light shrub} of $p$, and denote it $\mathcal{L}(p)$.
-\end{definition}
-
-In the following subsection we show how to exploit this geometric series behavior of light shrubs.
-
 \subsubsection{An asymptotically equivalent weight function}
@@
 
-As the weight $w(p$ includes the sizes of the heaps of all contributing heavy paths and leaves in the subtree rooted at $p$, we have the following.
+As the weight $w(p)$ includes the sizes of the heaps of all contributing heavy paths and leaves in the subtree rooted at $p$, we have the following.
 
@@
 \end{proof}
+
+\section{Lower Bound by Path Decomposition}
+
+In this section we show that any algorithm for computing the merge tree $T$ must pay at least $f_{path}(P(T))$ (see \Def{path}) for any path decomposition $P(T)$.
+
+We say two shrubs $S_1$ and $S_2$ (of merge trees of potentially different surfaces) are equivalent if there is an isomorphism, $f$,
+between the two shrubs such that for any node $v\in S_1$ corresponding to a path of size $|v|$, $f(v)$ also corresponds to a path of size $|v|$.
+For the purposes of the next lemma, we say two $2$-manifolds are different, if their corresponding contour trees are the different.
+
+\begin{figure}[h]\centering
+    \includegraphics[width=.15\linewidth]{figs/diamond}\hspace{2cm}
+    \includegraphics[width=.125\linewidth]{figs/topView}\hspace{2cm}
+    \includegraphics[width=.095\linewidth]{figs/sideView}
+    \caption{From left to right: the cone portion of a diamond; the top down view and then the side view of a cone face with a platform added.}
+    \label{fig:diamond}
+\end{figure}
+
+\begin{lemma}
+\label{lem:pathDecomp}
+Let $T$ be a merge tree, $P = P(T)$ any path decomposition of $T$, and $\pathTree$ the corresponding shrub.  
+Then there are at least $\Pi_{p_i\in P} |p_i|!$ different 2-manifolds with an equivalent shrub.
+\end{lemma}
+\begin{proof}
+ We describe a simple input mesh, with heights specified for all vertices except for the saddles.  The main idea is that 
+ this mesh will be flexible enough to allow almost any assignment of heights to the saddles, thus defining a class of merge trees 
+ whose size is that described in the lemma statement.
+
+ First, we describe the basic building block of this input mesh, which roughly speaking corresponds to a fixed path $p\in P$.  
+ Specifically, we say a \emph{diamond} of size $m$ is the following mesh.  To start, construct a cycle 
+ of length $m$, where the height of the $i$th vertex in the cycle is say $i/2m$.  Next, create an apex vertex at height 
+ $\infty$ (or any appropriately large finite height), and add an edge between this vertex and each vertex in the cycle.  At this point 
+ we have built an upside down cone with $m$ triangular faces (see \Fig{diamond}). To make this cone a closed surface we add a triangular cap at the bottom. 
+ Specifically, add a $3$ cycle of vertices at heights $0$, $1/20m$, and $1/10m$, and connect each vertex of this $3$-cycle to an arbitrary distinct 
+ vertex on the $m$ cycle (with matching cyclic ordering), and triangulate all remaining faces.  For consistency, if the size $m$ is $\leq 3$, the diamond 
+ will be just a tetrahedron.
+
+ Now above each face of the cone of the diamond, we will create a triangular ``platform'', which will act as the location to glue on other (shrunken) diamonds, 
+ and whose highest point of connection to the face below will be a saddle vertex.  Specifically the heights of the vertices of the triangle will be $h$, $h-\eps$, and $h-2\eps$, where 
+ $h$ is some unspecified height value.  This near horizontal platform is then connected to its respective face with vertical edges down to the face. See \Fig{diamond}.
+ %
+ %The vertex at height $h$ will lie in the plane of the respective face of the cone, and the other two will lie above the face, 
+ %with edges connecting down to the face, hence creating an actual ``platform''.
+
+ At this point we can construct the entire input mesh, by gluing together diamonds.  Specifically, for each path $p_i\in P$ we create a diamond of size $p_i$.
+ Now we glue together diamonds of different paths in the same way the paths are connected in $\pathTree$.  Specially, for two paths $p,q\in P$ where 
+ $p$ is the parent of $q$ in $\pathTree$, we glue the bottom triangular cap of the diamond for $q$ onto the triangular ``platform'' of some face of the diamond for $p$.
+ (Naturally for this construction to work, diamonds for a given path will be scaled down relative to the size of the diamond of their parent).
+
+ Observe now that the only saddle points in this construction are the highest connecting point on a face of an edge from the platform above.  Moreover, the only maxima are the apexes of the 
+ diamonds, and there is exactly one minima corresponding to the bottom cap of the diamond representing the root of $\pathTree$.  Therefore, the saddles on a given 
+ diamond will appear contiguously on a root to leaf path in the merge tree of the mesh, where the leaf corresponds to the maxima of the diamond 
+ (since all these saddles have a direct line of sight to this apex).  In particular, 
+ this implies that, regardless of the heights assigned to the platforms, the merge tree of this mesh has a path decomposition whose corresponding shrub is equivalent to $\pathTree$.
+
+ Now there is a valid instantiation of this described mesh for any relative ordering of the heights of the saddles on a given diamond.  In particular, 
+ there are $|p|!$ possible orderings of the heights of the saddles on the diamond for $p$, and hence $\Pi_{p_i\in P} |p_i|!$ possible instantiations of 
+ the mesh overall.  Also, each one of these instantiations will result in a different merge.  Specifically, a permutation of the heights on a given diamond 
+ corresponds to a permutation of the vertices along a path in $P$, as by construction all saddles on a given diamond will appear in sorted order in the merge tree.
+\end{proof}
+
+\begin{corollary}
+ Let $A(S)$ be any algorithm to compute the merge tree of any input surface $S$ (whose contour tree is a merge tree).  
+ Then the running time of $A(S)$ is $\Omega(f_{path}(P))=\Omega(\sum_{p_i\in P} |p_i|\log|p_i|)$ where $P$ is any path decomposition of the 
+ merge tree of $S$.
+\end{corollary}
+\begin{proof}
+ Given an input surface $S$ whose merge tree has a particular path decomposition $P$, by the above lemma, there are $\Pi_{p_i\in P} |p_i|!$ other surfaces 
+ with distinct merge trees, all of which have a shrub equivalent to $\pathTree$.  Therefore, any algorithm which computes the merge tree must distinguish between 
+ these possibilities, and therefore its corresponding decision tree must have $\Pi_{p_i\in P} |p_i|!$ leaves.  Hence by Stirling's approximation, 
+ this decision tree must have depth $\Omega(\sum_{p_i\in P} |p_i|\log|p_i|)$.
+\end{proof}
+
+\bibliographystyle{alpha}
+\bibliography{contour}
+
+\appendix
+
+\section{Old stuff}
 
@@
 
-\bibliographystyle{alpha}
-\bibliography{contour}
-
 \end{document}

### 2014-09-18 paint-contour-tree.tex
more changes to upper bound

@@
 
-We have a claim that also bounds heap sizes for special set of paths.
-
-\begin{claim} \label{clm:above} Let $Q$ be a set of incomparable nodes in the shrub $P(T)$ (which are paths in $T$).
-Suppose all paths in $A(Q)$ are heavy.
-Then $\sum_{q \in Q} |H_q| \log |H_q| = O(\sum_{p \in A(Q)} |p|\log |p|)$.
+There are some light paths that we can also ``pay" for.
+
+\begin{definition} \label{def:support} A light path $p$ is called \emph{supported} if there exists
+a contiguous (from $p$) chain of ancestors $q_1, q_2, \ldots, q_k$ (in $P(T)$) such that $|H_{q_k}| \leq 99|H_p|/100$
+and all $q_i$'s are heavy.
+\end{definition}
+
+Let $\cL$ be the set of light paths, $\cL'$ be the set of supported light paths, and $\cH$ be the set of heavy paths.
+%We have a claim that also bounds heap sizes for special set of paths.
+
+\begin{claim} \label{clm:above} $\sum_{p \in \cL'} |H_p| \log |H_p| = O(\sum_{p \in \cH} |p|\log |p|)$.
 \end{claim}
 
-\begin{proof} Pick $q \in Q$ and some $v \in H_q$. As we traverse up the tree passing
-through $A(q)$, we must encounter some $p \in A(q)$ such that $v \in p$. Furthermore,
-there is an order in which these vertices (of $H_q$) occur. Let $H'$ be the first half
-of vertices that occur. For all $v \in H'$, $|H_v| \geq |H_q|/2$. Hence,
-$|H_q| \log |H_q| \leq 4 \sum_{v \in H'} \log |H_v| \leq 4 \sum_{v \in H_q} |H_v|$.
-So $\sum_{q \in Q} |H_q| \log |H_q| \leq 4 \sum_{q \in Q} \sum_{v \in H_q} |H_v|$.
-
-Since nodes in $Q$ are incomparable, all the $H_q$'s (for $q \in Q$) are disjoint.
-Each such $v$ must occur in some path $p \in A(Q)$. Thus,
-$\sum_{q \in Q} |H_q| \log |H_q| \leq 4 \sum_{p \in A(Q)} \sum_{v \in p} |H_v|$.
+\begin{proof} Pick $p \in \cL'$. As we traverse up $q_1, q_2, \ldots, q_k$, there
+are certain vertices of $H_p$ that lie in these paths. There must be at least $|H_p|/100$
+such vertices, since $H_{q_k} \leq 99H_p/100$. These are encountered in
+a fixed order. Let $H'_p$ be the first half of these vertices. For all $v \in H'_p$
+$|H_v| \geq |H_p|/200$. Hence, $|H_p|\log |H_p| = O(\sum_{v \in H'_p} \log |H_v|)$
+and $\sum_{p \in L'} |H_p|\log |H_p| = O(\sum_{p \in L'} \sum_{v \in H'_p} \log |H_v|)$.
+
+Pick another $\widetilde{p} \in \cL'$. If $p$ and $\widetilde{p}$ are ancestor-descendant,
+then their corresponding supporting ancestor chains are disjoint. Any $v \in H'_p$
+can only occur in one other $H'_q$ (for $q \in \cL'$), because $v$ has exactly two colors.
+In other words, any $v$ appears at most twice in the above double summation.
+These vertices are all present in heavy paths, so we can bound
+the double summation by $\sum_{p \in \cH} \sum_{v \in p} \log |H_v|$.
+%
+%Pick $q \in Q$ and some $v \in H_q$. As we traverse up the tree passing
+%through $A(q)$, we must encounter some $p \in A(q)$ such that $v \in p$. Furthermore,
+%there is an order in which these vertices (of $H_q$) occur. Let $H'$ be the first half
+%of vertices that occur. For all $v \in H'$, $|H_v| \geq |H_q|/2$. Hence,
+%$|H_q| \log |H_q| \leq 4 \sum_{v \in H'} \log |H_v| \leq 4 \sum_{v \in H_q} |H_v|$.
+%So $\sum_{q \in Q} |H_q| \log |H_q| \leq 4 \sum_{q \in Q} \sum_{v \in H_q} |H_v|$.
+%
+%Since nodes in $Q$ are incomparable, all the $H_q$'s (for $q \in Q$) are disjoint.
+%Each such $v$ must occur in some path $p \in A(Q)$. Thus,
+%$\sum_{q \in Q} |H_q| \log |H_q| \leq 4 \sum_{p \in A(Q)} \sum_{v \in p} |H_v|$.
 We apply the bound of \Lem{pathBounds} to complete the proof.
@@
 
-\begin{lemma} \label{lem:light} $\sum_{p \in L} \sum_{v \in p} \log |H_v| = O(\sum_{p \in L'} |H_p| \log |H_p|)$.
+\begin{lemma} \label{lem:light} $\sum_{p \in \cL} \sum_{v \in p} \log |H_v| = O(\sum_{p \in \cL'} |H_p| \log |H_p|)$.
 \end{lemma}
@@
 \begin{proof} (of \Thm{runtime}) We split into heavy and light paths.
-We have $\sum_{v \in T} \log |H_v| = \sum_{p \in L} \sum_{v \in p} \log |H_v| + \sum_{p \notin L} \sum_{v \in p} \log |H_v|$.
-For the first term, we apply \Lem{light} to bound by $O(\sum_{p \in L'} |H_p| \log |H_p|)$.
-Now, we apply \Clm{above} to the set $L'$, noting that it satifies the claim conditions and that $A(Q) \cap L = \emptyset$.
-Hence, the first term is $O(\sum_{p \notin L} |p|\log |p|)$.
+We have $\sum_{v \in T} \log |H_v| =$  $\sum_{p \in \cL} \sum_{v \in p} \log |H_v| + $ $\sum_{p \in \cH} \sum_{v \in p} \log |H_v|$.
+For the first term, we apply \Lem{light} to bound by $O(\sum_{p \in \cL'} |H_p| \log |H_p|)$.
+Then we apply \Clm{above} to bound by $O(\sum_{p \in \cH} |p|\log |p|)$.
 The latter is also bounded by the same expression, by \Lem{pathBounds}. By definition,
-$\sum_{p \notin L} |p|\log |p| \leq \cost(\pmax(T))$.
+$\sum_{p \in \cH} |p|\log |p| \leq \cost(\pmax(T))$.
 \end{proof}
@@
 
-\subsection{Upper Bound by Path Decomposition}
+\subsection{Proving \Lem{light}}
+
+We can simplify a little. We split the summation and apply \Lem{pathBounds}. Finally,
+we note that for $p \in \cL$, $|p| \leq |H_p|$.
+\begin{eqnarray*} 
+\sum_{p \in \cL} \sum_{v \in p} \log |H_v| & = & \sum_{p \in \cL'} \sum_{v \in p} \log |H_v| + \sum_{p \in \cL \setminus \cL'} \sum_{v \in p} \log |H_v| \\
+& = & \sum_{p \in \cL \setminus \cL'} \sum_{v \in p} \log |H_v| + O(|p|\log |H_p|) = \sum_{p \in \cL \setminus \cL'} \sum_{v \in p} \log |H_v| + O(|H_p|\log |H_p|)
+\end{eqnarray*}
 

### 2014-09-18 paint-contour-tree.tex
more changes

@@
 
-\begin{lemma} \label{lem:light} $\sum_{p \in \cL} \sum_{v \in p} \log |H_v| = O(\sum_{p \in \cL'} |H_p| \log |H_p|)$.
+\begin{lemma} \label{lem:light} For the path decomposition $\pmax(T)$, $\sum_{p \in \cL} \sum_{v \in p} \log |H_v| = O(\sum_{p \in \cL'} |H_p| \log |H_p|)$.
 \end{lemma}
@@
 
-\begin{proof} (of \Thm{runtime}) We split into heavy and light paths.
+\begin{proof} (of \Thm{runtime}) We focus on $\pmax(T)$ and split into heavy and light paths.
 We have $\sum_{v \in T} \log |H_v| =$  $\sum_{p \in \cL} \sum_{v \in p} \log |H_v| + $ $\sum_{p \in \cH} \sum_{v \in p} \log |H_v|$.
@@
 
-\subsection{Proving \Lem{light}}
-
+\subsection{Proving \Lem{light}} \label{sec:light}
+
+Our focus is on $\pmax(T)$.
 We can simplify a little. We split the summation and apply \Lem{pathBounds}. Finally,
@@
 \begin{eqnarray*} 
-\sum_{p \in \cL} \sum_{v \in p} \log |H_v| & = & \sum_{p \in \cL'} \sum_{v \in p} \log |H_v| + \sum_{p \in \cL \setminus \cL'} \sum_{v \in p} \log |H_v| \\
-& = & \sum_{p \in \cL \setminus \cL'} \sum_{v \in p} \log |H_v| + O(|p|\log |H_p|) = \sum_{p \in \cL \setminus \cL'} \sum_{v \in p} \log |H_v| + O(|H_p|\log |H_p|)
+\sum_{p \in \cL} \sum_{v \in p} \log |H_v| & = & \sum_{p \in \cL \setminus \cL'} \sum_{v \in p} \log |H_v| + \sum_{p \in \cL'} \sum_{v \in p} \log |H_v|\\
+& = & \sum_{p \in \cL \setminus \cL'} \sum_{v \in p} \log |H_v| + O(\sum_{p \in \cL'} |p|\log |H_p|) = \sum_{p \in \cL \setminus \cL'} \sum_{v \in p} \log |H_v| + O(\sum_{p \in \cL'} |H_p|\log |H_p|)
 \end{eqnarray*}
@@
 }
+
+It suffices to focus on unsupported light paths. \Lem{light} can be reduced to the following.
+
+\begin{lemma} \label{lem:support} $\sum_{p \in \cL \setminus \cL'} \sum_{v \in p} \log |H_v| = O(\sum_{p \in \cL'} |H_p| \log |H_p|)$.
+\end{lemma}
+
+We now construct the shrub of unsupported light paths $\cU$. Consider $p \in \cL \setminus \cL'$,
+and traverse the chain of ancestors
 

### 2014-09-20 paint-contour-tree.tex
more changes

@@
 \newcommand{\cP}{{\cal P}}
+\newcommand{\cU}{{\cal U}}
 \newcommand{\cV}{{\cal V}}
@@
 
-\subsection{Proving \Lem{light}} \label{sec:light}
+\subsection{The shrub of unsupported paths} \label{sec:light}
 
@@
 
-We now construct the shrub of unsupported light paths $\cU$. Consider $p \in \cL \setminus \cL'$,
-and traverse the chain of ancestors
-
-We focus on $\pmax(T)$, and prove the key property of this path decomposition.
+We now construct the shrub of unsupported light paths. Consider $p \in \cL \setminus \cL'$,
+and traverse the chain of ancestors from $p$. Eventually, we must reach another light path $q$.
+(If not, we have reached the root $r$ of $\pmax(T)$. But $|H_r| = 0$, implying $p$ is supported.)
+Insert edge from $p$ to $q$, so $q$ is the parent of $p$ in $\cU$. This construction leads
+the shrub forest of $\cL \setminus \cL'$, where every unsupported light path is an internal node, and all
+roots are supported light paths. \Lem{support} is a direct consequence of the following,
+which is the really the main lemma.
+
+\begin{lemma} \label{lem:cu} Let $\cU$ denote a connected component (shrub) in the shrub forest of $\cL \setminus \cL'$ and let $r$
+be the root of $\cU$. Then $\sum_{p \in \cU} \sum_{v \in p} \log |H_v| = O(|H_r| \log |H_r|)$.
+\end{lemma}
+
+We proceed with the setup to prove \Lem{cu}.
+We focus our attention on some shrub $\cU$.
+We prove a technical lemma about $\cU$, from which \Lem{cu} follows directly.
+
+\begin{lemma} \label{lem:cu-root} (i) For any $v \in p$ such that $p \in \cU$,
+$|H_v| = O(|H_r|)$. (ii) $\sum_{p \in \cU} |p| = O(|H_r|)$.
+\end{lemma}
+
+\begin{proof} (of \Lem{cu}) We have $\sum_{p \in \cU} \sum_{v \in p} \log |H_v| = O((\log |H_r|) \sum_{p \in \cU} |p|) = O(|H_r|\log |H_r|)$.
+\end{proof}
+
+\subsection{The root is everything in $\cU$} \label{sec:weight}
+
+\Lem{cu-root} asserts the root $r$ in $\cU$ pretty much encompasses all sizes and heaps in $\cU$.
+Consider $q \in \cL \setminus \cL'$ that is a child of $p$ in the shrub $\cU$.
+By construction, there is a heavy path (denoted $h(q)$) that is a child of $p$ in $\pmax(T)$,
+and an ancestor of $q$. Furthermore, $|H_{h(q)}| > 99|H_q|/100$.
+We begin with a key property, which is where the construction of $\pmax(T)$ enters the picture.
+
+%
+%
+%
+%We focus on $\pmax(T)$, and prove the key property of this path decomposition.
 
@@
 \label{lem:geometric}
- Let $p$ be the parent of $q$ in the shrub $\pmax(T)$.  
- Then $|H_q|\leq \frac{3}{5} |H_p|$. 
-\end{lemma}
-\begin{proof}
- Let $w$ be the vertex in the path $p$ to which the root of the path $q$ is adjacent to in $T$, 
- i.e. $w$ is the parent of $r_q$ in $T$.  Let $c$ be the other child of $w$, which lies in $p$.
- As $\mathcal{P}(T)$ is a maximum path decomposition of $T$ and $c$ rather than $r_q$ is in 
- the same path as $w$, we have $|H_q|\leq |H_c|$.  Therefore, $|H_w|\geq |H_q|+|H_c|-1\geq 2|H_q|-1$. 
- Therefore, by \Obs{decrease}, $|H_p|+ d(r_p, w) \geq |H_w|\geq 2|H_q|-1$.  
- Since $p$ is a light path, $d(r_p, w)\leq \sqrt{|H_p|}/100$ and so we have $|H_p|+\sqrt{|H_p|}/100\geq 2|H_q|-1$.
- Assuming $|H_p|>11$ (and we can safely assume $|H_p|$ is bigger than any constant), this yields $\frac{3}{5} |H_p| \geq |H_q|$
-\end{proof}
-
-\ignore{
-Naturally, this geometric series behavior can be extended to more than two adjacent light paths, thus motivating 
-the following definition.
-
-\begin{definition}
- Let $p$ be a path in a path decomposition $P(T)$, with corresponding shrub $\pathTree$.
- Call a light path $q\in P(T)$ light reachable from $p$ if the path from $p$ to $q$ in $\pathTree$ 
- contains only vertices which correspond to light paths.  Call the subtree of $\pathTree$ consisting 
- of all light reachable paths from $p$ the \emph{light shrub} of $p$, and denote it $\mathcal{L}(p)$.
-\end{definition}
-
-In the following subsection we show how to exploit this geometric series behavior of light shrubs.
-
-We say two paths in $P(T)$ are adjacent if their corresponding vertices are adjacent in the shrub.  Similarly, for 
-two adjacent paths in $P(T)$ we can define a parent/child relationship based on this relationship in the shrub. 
-}
-
-\remove{
-\Ben{Maybe move the remainder of this subsection to 5.3 since heavy and light paths and not used until then}
-
-\begin{definition}
- Let T be a binary tree and let $\chi(T)$ and $P(T)$ be a valid coloring and path decomposition of $T$, respectively.  
- For $p\in P(T)$, we say that $p$ is \emph{light} if $|p| < \sqrt{|H_{r_p}|}/100$, and we say it is \emph{heavy} otherwise.
-\end{definition}
-
-\begin{observation}
-\label{obs:decrease}
- Let $v$ be a vertex in $T$ and let $w$ be its parent.  Then $|H_w| \geq |H_v| -1$, as $L(T_v) \subseteq L(T_w)$ and the path from 
- $w$ to the root has one less vertex than the path from $v$ to the root.
-
- In particular, we have the following more general property.
- Let $v$ and $u$ be any two vertices in the same root to leaf path of $T$, such that $u$ is lower than $v$.  
- Then $|H_v| \leq |H_u| + d(u,v)$. 
-\end{observation}
-
-\begin{lemma}
-\label{lem:pathBounds}
- Let T be a binary tree and let $\chi(T)$ and $P(T)$ be a valid coloring and path decomposition of $T$, respectively.
- Then, 
- \begin{compactenum}
-  \item for any heavy path $p\in P(T)$, $\sum_{v\in p} \log |H_v| = O(|p| \log |p|)$.
-  \item for any light path $p\in P(T)$, $\sum_{v\in p} \log |H_v| = O(|p| \log |H_{r_p}|)$.
- \end{compactenum}
-\end{lemma}
-\begin{proof}
- By $\Obs{decrease}$ we know that for any vertex $v\in p$, $|H_v|\leq |H_{r_p}| + |p|$ (as $r_p$ lies below $v$ on $p$).
- As such, 
- \[
- \sum_{v\in p} \log |H_v| \leq \sum_{v\in p} \log(|H_{r_p}| + |p|)  = |p| \log (|H_{r_p}| + |p|).
- \]
- If $p$ is a heavy path then $|p| \log (|H_{r_p}| + |p|) = O(|p| \log |p|)$, and if $p$ is light then 
- $|p| \log (|H_{r_p}| + |p|) = O(|p| \log |H_{r_p}| )$.
-\end{proof}
-
-The following lemma can be thought of as a generalization of $\Obs{decrease}$.
-
-\begin{lemma}
-\label{lem:adjacent}
- Let $T$ be a binary tree and let $\chi(T)$ and $P(T)$ be a valid coloring and path decomposition of $T$, respectively.
- Let $p$ be any path in $P(T)$ and let $\{q_1, \dots q_k\}$ be the set of all adjacent child paths of $p$.
- For $1\leq i\leq k$ let $r_i = r_{q_i}$ and let $H_i = H_{r_i}$.
- Also, let $H_0$ denote the heap at the leaf of $p$.
-
- Then $\sum_{i=0}^k |H_i| \leq |H_p|+|p|$, where $H_p = H_{r_p}$.
-\end{lemma}
-\begin{proof} 
- First observe that for all $i$, $r_{i}$ lies above $r_p$ on some root to leaf path. 
- Let $H_i'$ be the subset of $H_i$ that lies below $r_p$.  Observe that $H_i'\subseteq H_p$, as $L(T_{r_i})\subseteq L(T_{r_p})$.
-
- Now for any $i\neq j$, $H_i \cap H_j = \emptyset$, as $r_{i}$ and $r_{j}$ are not on the same root to leaf path (i.e. $L(T_{r_i}) \cap L(T_{r_j}) = \emptyset$).
- In particular, $(H_i\setminus H_i') \cap (H_j \setminus H_j') = \emptyset$, and so $\sum_{i=0}^k |H_i\setminus H_i'| \leq |p|$ 
- as $H_i\setminus H_i'$ is precisely the subset of $H_i$ that lies on $p$.
- Moreover, for any $i\neq j$, $ H_i' \cap H_j' = \emptyset$, and so since $H_i'\subseteq H_p$ we have $\sum_{i=0}^k |H_i'|\leq |H_p|$.
- Putting these together gives
- \[
-  \sum_{i=0}^k |H_i| = \sum_{i=0}^k |H_i'| + \sum_{i=0}^k |H_i\setminus H_i'| \leq |H_p|+|p|.
- \]
-\end{proof}
-
-% \Ben{There is a fundamental flaw in the following lemma and corollary.  
-% As such, the remainder of this subsection will likely be deleted.}
-% 
-% Using the notation defined in the above lemma, we have the following crucial lemma, which we will use to 
-% argue we can charge that cost of all paths in $P(T)$ to the cost of the heavy paths.
-% 
-% \begin{lemma}
-% Let $p$ be a heavy path in $P(T)$ and let $\{q_1, \dots q_k\}$ be any set of light paths in $P(T)$ such that 
-% $\sum_{i=1}^k |H_i| = O(H_p)$.  Then $\sum_{i=1}^k \sum_{v\in q_i} \log |H_v| = O(|p|\log |p|)$
-% \end{lemma}
-% \begin{proof}
-% By \Lem{pathBounds}, for all $1\leq i\leq k$, 
-%  $\sum_{v\in q_i} \log |H_v| = O(|q_i|\log |H_i|) = O(|H_i|\log |H_i|)$, as $q_i$ is a light path.
-%  By assumption $\sum_{i=1}^k |H_i| = O(|H_p|)$, 
-%  As $x\log x$ is a convex function, Jensen's inequality then implies 
-%  $\sum_{i=1}^{k} |H_i|\log |H_i| = O(|H_p| \log |H_p|) = O(|p|\log |p|)$, since $p$ is a heavy path.
-% \end{proof}
-% 
-% 
-% \begin{corollary}
-% \label{cor:heavyLight}
-%  Let $p$ be a heavy path in $P(T)$ and let $\{q_1, \dots q_k\}$ be the set of all light paths adjacent to $p$.
-%  Then $\sum_{i=1}^k \sum_{v\in q_i} \log |H_v| = O(|p|\log |p|)$.
-% \end{corollary}
-% \begin{proof}
-%  Let $H_i$ be as defined in \Lem{adjacent}.  By \Lem{pathBounds}, for all $1\leq i\leq k$, 
-%  $\sum_{v\in q_i} \log |H_v| = O(|q_i|\log |H_i|) = O(|H_i|\log |H_i|)$, as $q_i$ is a light path.
-%  Now by \Lem{adjacent}, $\sum_{i=1}^k |H_i| \leq |H_p|+|p| = O(|p|)$, as $p$ is a heavy path. 
-%  Therefore, as $x\log x$ is a convex function, Jensen's inequality implies the claim.
-%  
-% \end{proof}
-
-}
-
-\subsubsection{An asymptotically equivalent weight function}
-\label{sec:weight}
+ Let $p$ be the parent of $q$ in the shrub $\cU$. Then $|H_p|\geq \frac{3}{2}|H_q|$. 
+\end{lemma}
+\begin{proof} We know that $h(q)$ is a child of $p$ in $\pmax(T)$. Consider the base
+of $h(q)$, which is a node $w$ in $T$. Let $v$ denote the sibling of $w$ in $T$.
+Their parent is called $u$. Note that both $u$ and $v$ are nodes in the path $p$.
+Now, the decomposition $\pmax(T)$ put $u$ and $v$ in the same path $p$. This
+implies $|H_v| \geq |H_w|$. Since $|H_u| \geq |H_v| + |H_w| - 1$,
+$|H_u| \geq 2|H_w| - 1$. Let $b$ be the base of $p$.
+We have $|H_p| = |H_b| \geq |H_u| - |p| \geq 2|H_w| - |p| - 2$. Since $p$ is a light path, $|p| < \sqrt{|H_p|}/100$.
+Applying this bound, we get $|H_p| \geq (2-\delta)|H_w|$ (for small constant $\delta > 0$).
+Since $w$ is the base of $h(q)$, $H_w = H_{h(q)}$. We apply the bound $|H_{h(q)}| > 99|H_q|/100$,
+to complete the proof.
+
+%
+%
+% Let $w$ be the vertex in the path $p$ to which the root of the path $q$ is adjacent to in $T$, 
+% i.e. $w$ is the parent of $r_q$ in $T$.  Let $c$ be the other child of $w$, which lies in $p$.
+% As $\mathcal{P}(T)$ is a maximum path decomposition of $T$ and $c$ rather than $r_q$ is in 
+% the same path as $w$, we have $|H_q|\leq |H_c|$.  Therefore, $|H_w|\geq |H_q|+|H_c|-1\geq 2|H_q|-1$. 
+% Therefore, by \Obs{decrease}, $|H_p|+ d(r_p, w) \geq |H_w|\geq 2|H_q|-1$.  
+% Since $p$ is a light path, $d(r_p, w)\leq \sqrt{|H_p|}/100$ and so we have $|H_p|+\sqrt{|H_p|}/100\geq 2|H_q|-1$.
+% Assuming $|H_p|>11$ (and we can safely assume $|H_p|$ is bigger than any constant), this yields $\frac{3}{5} |H_p| \geq |H_q|$
+\end{proof}
+
+To prove this, we need to understand how $H_p$ is constructed. The path $p$ has some children
+$q_1, q_2, \ldots, q_k$ in $\pmax(T)$. Furthermore, as a leaf path in $p$, some leaf
+$\ell$ of $T$ in one of the ends of $p$. The heap $H_p$ is essentially the
+union of $H_\ell$ and $H_{q_1}, H_{q_2}, \ldots, H_{q_k}$, with the vertices in $p$
+removed. Note that some of the $q_i$'s are the corresponding $h(q')$ vertices,
+for a child $q'$ in $\cU$.
+
 Let $\mathcal{L} = \mathcal{L}(r)$ be a light shrub of some light path $r$ in a maximum path decomposition $\mathcal{P}(T)$.
@@
 
+\ignore{
+Naturally, this geometric series behavior can be extended to more than two adjacent light paths, thus motivating 
+the following definition.
+
+\begin{definition}
+ Let $p$ be a path in a path decomposition $P(T)$, with corresponding shrub $\pathTree$.
+ Call a light path $q\in P(T)$ light reachable from $p$ if the path from $p$ to $q$ in $\pathTree$ 
+ contains only vertices which correspond to light paths.  Call the subtree of $\pathTree$ consisting 
+ of all light reachable paths from $p$ the \emph{light shrub} of $p$, and denote it $\mathcal{L}(p)$.
+\end{definition}
+
+In the following subsection we show how to exploit this geometric series behavior of light shrubs.
+
+We say two paths in $P(T)$ are adjacent if their corresponding vertices are adjacent in the shrub.  Similarly, for 
+two adjacent paths in $P(T)$ we can define a parent/child relationship based on this relationship in the shrub. 
+}
+
+\remove{
+\Ben{Maybe move the remainder of this subsection to 5.3 since heavy and light paths and not used until then}
+
+\begin{definition}
+ Let T be a binary tree and let $\chi(T)$ and $P(T)$ be a valid coloring and path decomposition of $T$, respectively.  
+ For $p\in P(T)$, we say that $p$ is \emph{light} if $|p| < \sqrt{|H_{r_p}|}/100$, and we say it is \emph{heavy} otherwise.
+\end{definition}
+
+\begin{observation}
+\label{obs:decrease}
+ Let $v$ be a vertex in $T$ and let $w$ be its parent.  Then $|H_w| \geq |H_v| -1$, as $L(T_v) \subseteq L(T_w)$ and the path from 
+ $w$ to the root has one less vertex than the path from $v$ to the root.
+
+ In particular, we have the following more general property.
+ Let $v$ and $u$ be any two vertices in the same root to leaf path of $T$, such that $u$ is lower than $v$.  
+ Then $|H_v| \leq |H_u| + d(u,v)$. 
+\end{observation}
+
+\begin{lemma}
+\label{lem:pathBounds}
+ Let T be a binary tree and let $\chi(T)$ and $P(T)$ be a valid coloring and path decomposition of $T$, respectively.
+ Then, 
+ \begin{compactenum}
+  \item for any heavy path $p\in P(T)$, $\sum_{v\in p} \log |H_v| = O(|p| \log |p|)$.
+  \item for any light path $p\in P(T)$, $\sum_{v\in p} \log |H_v| = O(|p| \log |H_{r_p}|)$.
+ \end{compactenum}
+\end{lemma}
+\begin{proof}
+ By $\Obs{decrease}$ we know that for any vertex $v\in p$, $|H_v|\leq |H_{r_p}| + |p|$ (as $r_p$ lies below $v$ on $p$).
+ As such, 
+ \[
+ \sum_{v\in p} \log |H_v| \leq \sum_{v\in p} \log(|H_{r_p}| + |p|)  = |p| \log (|H_{r_p}| + |p|).
+ \]
+ If $p$ is a heavy path then $|p| \log (|H_{r_p}| + |p|) = O(|p| \log |p|)$, and if $p$ is light then 
+ $|p| \log (|H_{r_p}| + |p|) = O(|p| \log |H_{r_p}| )$.
+\end{proof}
+
+The following lemma can be thought of as a generalization of $\Obs{decrease}$.
+
+\begin{lemma}
+\label{lem:adjacent}
+ Let $T$ be a binary tree and let $\chi(T)$ and $P(T)$ be a valid coloring and path decomposition of $T$, respectively.
+ Let $p$ be any path in $P(T)$ and let $\{q_1, \dots q_k\}$ be the set of all adjacent child paths of $p$.
+ For $1\leq i\leq k$ let $r_i = r_{q_i}$ and let $H_i = H_{r_i}$.
+ Also, let $H_0$ denote the heap at the leaf of $p$.
+
+ Then $\sum_{i=0}^k |H_i| \leq |H_p|+|p|$, where $H_p = H_{r_p}$.
+\end{lemma}
+\begin{proof} 
+ First observe that for all $i$, $r_{i}$ lies above $r_p$ on some root to leaf path. 
+ Let $H_i'$ be the subset of $H_i$ that lies below $r_p$.  Observe that $H_i'\subseteq H_p$, as $L(T_{r_i})\subseteq L(T_{r_p})$.
+
+ Now for any $i\neq j$, $H_i \cap H_j = \emptyset$, as $r_{i}$ and $r_{j}$ are not on the same root to leaf path (i.e. $L(T_{r_i}) \cap L(T_{r_j}) = \emptyset$).
+ In particular, $(H_i\setminus H_i') \cap (H_j \setminus H_j') = \emptyset$, and so $\sum_{i=0}^k |H_i\setminus H_i'| \leq |p|$ 
+ as $H_i\setminus H_i'$ is precisely the subset of $H_i$ that lies on $p$.
+ Moreover, for any $i\neq j$, $ H_i' \cap H_j' = \emptyset$, and so since $H_i'\subseteq H_p$ we have $\sum_{i=0}^k |H_i'|\leq |H_p|$.
+ Putting these together gives
+ \[
+  \sum_{i=0}^k |H_i| = \sum_{i=0}^k |H_i'| + \sum_{i=0}^k |H_i\setminus H_i'| \leq |H_p|+|p|.
+ \]
+\end{proof}
+
+% \Ben{There is a fundamental flaw in the following lemma and corollary.  
+% As such, the remainder of this subsection will likely be deleted.}
+% 
+% Using the notation defined in the above lemma, we have the following crucial lemma, which we will use to 
+% argue we can charge that cost of all paths in $P(T)$ to the cost of the heavy paths.
+% 
+% \begin{lemma}
+% Let $p$ be a heavy path in $P(T)$ and let $\{q_1, \dots q_k\}$ be any set of light paths in $P(T)$ such that 
+% $\sum_{i=1}^k |H_i| = O(H_p)$.  Then $\sum_{i=1}^k \sum_{v\in q_i} \log |H_v| = O(|p|\log |p|)$
+% \end{lemma}
+% \begin{proof}
+% By \Lem{pathBounds}, for all $1\leq i\leq k$, 
+%  $\sum_{v\in q_i} \log |H_v| = O(|q_i|\log |H_i|) = O(|H_i|\log |H_i|)$, as $q_i$ is a light path.
+%  By assumption $\sum_{i=1}^k |H_i| = O(|H_p|)$, 
+%  As $x\log x$ is a convex function, Jensen's inequality then implies 
+%  $\sum_{i=1}^{k} |H_i|\log |H_i| = O(|H_p| \log |H_p|) = O(|p|\log |p|)$, since $p$ is a heavy path.
+% \end{proof}
+% 
+% 
+% \begin{corollary}
+% \label{cor:heavyLight}
+%  Let $p$ be a heavy path in $P(T)$ and let $\{q_1, \dots q_k\}$ be the set of all light paths adjacent to $p$.
+%  Then $\sum_{i=1}^k \sum_{v\in q_i} \log |H_v| = O(|p|\log |p|)$.
+% \end{corollary}
+% \begin{proof}
+%  Let $H_i$ be as defined in \Lem{adjacent}.  By \Lem{pathBounds}, for all $1\leq i\leq k$, 
+%  $\sum_{v\in q_i} \log |H_v| = O(|q_i|\log |H_i|) = O(|H_i|\log |H_i|)$, as $q_i$ is a light path.
+%  Now by \Lem{adjacent}, $\sum_{i=1}^k |H_i| \leq |H_p|+|p| = O(|p|)$, as $p$ is a heavy path. 
+%  Therefore, as $x\log x$ is a convex function, Jensen's inequality implies the claim.
+%  
+% \end{proof}
+
+}
+
 \end{document}

### 2014-09-22 paint-contour-tree.tex
more changes

@@
 
-\begin{definition} \label{def:support} A light path $p$ is called \emph{supported} if there exists
-a contiguous (from $p$) chain of ancestors $q_1, q_2, \ldots, q_k$ (in $P(T)$) such that $|H_{q_k}| \leq 99|H_p|/100$
-and all $q_i$'s are heavy.
+\begin{definition} \label{def:support} A light path $p$ is called \emph{supported} if: 
+at least $|H_p|/100$ vertices in $p$ lie in heavy ancestors $q$ such that $|H_q| > |H_p|/100$.
+%
+%
+% there exists
+%a contiguous (from $p$) chain of ancestors $q_1, q_2, \ldots, q_k$ (in $P(T)$) such that $|H_{q_k}| \leq 99|H_p|/100$
+%and all $q_i$'s are heavy.
 \end{definition}
@@
 Insert edge from $p$ to $q$, so $q$ is the parent of $p$ in $\cU$. This construction leads
-the shrub forest of $\cL \setminus \cL'$, where every unsupported light path is an internal node, and all
+to the shrub forest of $\cL \setminus \cL'$, where every unsupported light path is an internal node, and all
 roots are supported light paths. \Lem{support} is a direct consequence of the following,
@@
 
-\begin{lemma} \label{lem:cu} Let $\cU$ denote a connected component (shrub) in the shrub forest of $\cL \setminus \cL'$ and let $r$
-be the root of $\cU$. Then $\sum_{p \in \cU} \sum_{v \in p} \log |H_v| = O(|H_r| \log |H_r|)$.
-\end{lemma}
-
-We proceed with the setup to prove \Lem{cu}.
-We focus our attention on some shrub $\cU$.
-We prove a technical lemma about $\cU$, from which \Lem{cu} follows directly.
-
-\begin{lemma} \label{lem:cu-root} (i) For any $v \in p$ such that $p \in \cU$,
+%\begin{lemma} \label{lem:cu} Let $\cU$ denote a connected component (shrub) in the shrub forest of $\cL \setminus \cL'$ and let $r$
+%be the root of $\cU$. Then $\sum_{p \in \cU} \sum_{v \in p} \log |H_v| = O(|H_r| \log |H_r|)$.
+%\end{lemma}
+%
+%We proceed with the setup to prove \Lem{cu}.
+%We focus our attention on some shrub $\cU$.
+%We prove a technical lemma about $\cU$, from which \Lem{cu} follows directly.
+
+\begin{lemma} \label{lem:cu-root} Let $\cU$ denote a connected component (shrub) in the shrub forest of $\cL \setminus \cL'$ and let $r$
+be the root of $\cU$. (i) For any $v \in p$ such that $p \in \cU$,
 $|H_v| = O(|H_r|)$. (ii) $\sum_{p \in \cU} |p| = O(|H_r|)$.
@@
 
-\begin{proof} (of \Lem{cu}) We have $\sum_{p \in \cU} \sum_{v \in p} \log |H_v| = O((\log |H_r|) \sum_{p \in \cU} |p|) = O(|H_r|\log |H_r|)$.
-\end{proof}
-
-\subsection{The root is everything in $\cU$} \label{sec:weight}
+\begin{proof} (of \Lem{support}) We have $\sum_{p \in \cU} \sum_{v \in p} \log |H_v| = O((\log |H_r|) \sum_{p \in \cU} |p|) = O(|H_r|\log |H_r|)$.
+We sum this over all $\cU$ in the shrub forest, and note that roots in the shrub forest are supported light paths.
+\end{proof}
+
+\subsection{Proving \Lem{cu-root}: the root is everything in $\cU$} \label{sec:weight}
 
@@
 % Assuming $|H_p|>11$ (and we can safely assume $|H_p|$ is bigger than any constant), this yields $\frac{3}{5} |H_p| \geq |H_q|$
+\end{proof}
+
+This immediately proves part (i) of \Lem{cu-root}. Part (ii) requires some more work.
+
+\begin{proof} (part (ii) of \Lem{cu-root}) Let level $\ell$ of $\cU$ denote all nodes at distance $\ell$ from the root $r$.
+These set of these nodes is designated as $\cU_\ell$. We set $\sigma_\ell = \sum_{p \in \cU_\ell} |p|$
+and $\sigma_{\leq \ell} = \sum_{k \leq \ell} \sigma_k$.
+By \Lem{geometric}, for any $p \in \cU_\ell$, $|H_p| \leq (2/3)^\ell |H_r|$.
+
+Since all nodes in $\cU$ represent light paths, $\forall p \in \cU$, $|p| < \sqrt{|H_p|}/100$.
+So $\sigma_\ell < \sum_{p \in \cU_\ell} \sqrt{|H_p|}/100$.
+
 \end{proof}

### 2014-09-22 paint-contour-tree.tex
working on upper bounds

@@
  a set of vertex disjoint leaf paths.  Note that the number of paths in $P(T)$ must equal the number of leaves in $T$.
-
- The \emph{cost} of $P(T)$ is $\cost(P(T)) = \sum_{p\in P(T)} |p| \log |p|$. (We use $|p|$ to denote the number of vertices in $p$.)
+% 
+% The \emph{cost} of $P(T)$ is $\cost(P(T)) = \sum_{p\in P(T)} |p| \log |p|$. (We use $|p|$ to denote the number of vertices in $p$.)
 \end{definition}
@@
 
-Our main theorem is the following.
-
-\begin{theorem} \label{thm:runtime} $\sum_{v \in T} \log |H_v| = O(\cost(\pmax(T)))$.
+Our main theorem is the following. We use $|p|$ to denote the number of vertices in $p$.
+
+\begin{theorem} \label{thm:runtime} $\sum_{v \in T} \log |H_v| = O(\sum_{p\in P(T)} |p| \log |p|)$.
 \end{theorem}
@@
 \label{lem:pathBounds} If $p$ is heavy, $\sum_{v\in p} \log |H_v| = O(|p| \log |p|)$.
-If $p$ is light, $\sum_{v\in p} \log |H_v| = O(|p| \log |H_p|)$.
+If $p$ is light, $\sum_{v\in p} \log |H_v| = O(|H_p| \log |H_p|)$.
 % Let T be a binary tree and let $\chi(T)$ and $P(T)$ be a valid coloring and path decomposition of $T$, respectively.
@@
  If $p$ is a heavy path, then $|p| \log (|H_{p}| + |p|) = O(|p| \log |p|)$. If $p$ is light, then 
- $|p| \log (|H_{p}| + |p|) = O(|p| \log |H_{p}| )$.
+ $|p| \log (|H_{p}| + |p|) = O(|p| \log |H_{p}| )$. For light paths, $|p| = O(|H_p|)$.
 \end{proof}
@@
 
-\begin{definition} \label{def:support} A light path $p$ is called \emph{supported} if: 
-at least $|H_p|/100$ vertices in $p$ lie in heavy ancestors $q$ such that $|H_q| > |H_p|/100$.
+\begin{definition} \label{def:support} A light path $p$ is \emph{supported} if
+at least $|H_p|/100$ vertices $v$ in $H_p$ lie in heavy ancestors.
 %
@@
 
-Let $\cL$ be the set of light paths, $\cL'$ be the set of supported light paths, and $\cH$ be the set of heavy paths.
+Let $\cL$ be the set of light paths, $\cL'$ be the set of supported light paths, and $\cH$ be the set of heavy paths
+given by $\pmax(T)$.
+We now construct the shrub of unsupported light paths. Consider $p \in \cL \setminus \cL'$,
+and traverse the chain of ancestors from $p$. Eventually, we must reach another light path $q$.
+(If not, we have reached the root $r$ of $\pmax(T)$. Hence, $p$ is supported.)
+Insert edge from $p$ to $q$, so $q$ is the parent of $p$ in $\cU$. This construction leads
+to the shrub forest of $\cL \setminus \cL'$, where every unsupported light path is an internal node, and all
+roots are supported light paths.
+
+Most of work goes into proving the following technical lemma.
+
+\begin{lemma} \label{lem:cu-root} Let $\cU$ denote a connected component (shrub) in the shrub forest of $\cL \setminus \cL'$ and let $r$
+be the root of $\cU$. (i) For any $v \in p$ such that $p \in \cU$,
+$|H_v| = O(|H_r|)$. (ii) $\sum_{p \in \cU} |p| = O(|H_r|)$.
+\end{lemma}
+
+We split the remaining argument into two subsections. We first prove \Thm{runtime} from \Lem{cu-root},
+which mostly imvolves routing calculations. Then we prove \Lem{cu-root}, where most
+of the interesting work happens.
+
+\subsection{Proving \Thm{runtime}} \label{sec:thm-runtime}
+
+We split the summation into heavy, light, and unsupported light paths.
+\begin{eqnarray*} 
+\sum_{p \in \cL} \sum_{v \in p} \log |H_v| & = & \sum_{p \in \cL \setminus \cL'} \sum_{v \in p} \log |H_v| + \sum_{p \in \cL'} \sum_{v \in p} \log |H_v| + \sum_{p \in \cH} \sum_{v \in p} \log |H_v|
+%& = & \sum_{p \in \cL \setminus \cL'} \sum_{v \in p} \log |H_v| + O(\sum_{p \in \cL'} |p|\log |H_p|) = \sum_{p \in \cL \setminus \cL'} \sum_{v \in p} \log |H_v| + O(\sum_{p \in \cL'} |H_p|\log |H_p|)
+\end{eqnarray*}
+The last term can be bounded by $O(\sum_{p \in \pmax(T)} |p|\log |p|)$, by \Lem{pathBounds}.
+The second term can be bounded by $O(\sum_{p \in \cL'} |H_p| \log |H_p|)$, by \Lem{pathBounds} again.
+The following claim shows that in turn is at most the last term.
+
 %We have a claim that also bounds heap sizes for special set of paths.
 
-\begin{claim} \label{clm:above} $\sum_{p \in \cL'} |H_p| \log |H_p| = O(\sum_{p \in \cH} |p|\log |p|)$.
+\begin{claim} \label{clm:above} $\sum_{p \in \cL'} |H_p| \log |H_p| = O(\sum_{q \in \cH} \sum_{v \in q} \log |H_v|)$.
 \end{claim}
 
-\begin{proof} Pick $p \in \cL'$. As we traverse up $q_1, q_2, \ldots, q_k$, there
-are certain vertices of $H_p$ that lie in these paths. There must be at least $|H_p|/100$
-such vertices, since $H_{q_k} \leq 99H_p/100$. These are encountered in
-a fixed order. Let $H'_p$ be the first half of these vertices. For all $v \in H'_p$
-$|H_v| \geq |H_p|/200$. Hence, $|H_p|\log |H_p| = O(\sum_{v \in H'_p} \log |H_v|)$
-and $\sum_{p \in L'} |H_p|\log |H_p| = O(\sum_{p \in L'} \sum_{v \in H'_p} \log |H_v|)$.
+\begin{proof} 
+Pick $p \in \cL'$. 
+%Let the first $|H_p|/200$ vertices the lie in heavy paths be denoted as $H'_p$. 
+%Observe that for any $v \in H'_p$, 
+%We have $|H_p| \log |H_p| = O(\sum_{v \in H'_p} \log |H_p|)$
+%$= O(\sum_{v \in H'_p} \log |H_v|)$. Since all the vertices lie in heavy paths,
+%we can write this as $O(\sum_{q \in \cH} \sum_{v \in H'_p \cap q} \log |H_v|)$.
+%Summing over all $p$, the expression is $\sum_{q \in \cH} \sum_{p \in \cL'} \sum_{v \in H'_p \cap q} \log |H_v|$.
+%
+%
+As we traverse the heavy ancestors, there
+are at least $|H_p|/100$ vertices of $H_p$ that lie in these paths. These are encountered in
+a fixed order. Let $H'_p$ be the first $|H_p|/200$ of these vertices. When $v \in H'_p$
+is encountered, there are $|H_p|/200$ vertices of $H_p$ not yet encountered. Hence,
+$|H_v| \geq |H_p|/200$. Hence, $|H_p|\log |H_p| = O(\sum_{v \in H'_p} \log |H_v|)$.
+Since all the vertices lie in heavy paths,
+we can write this as $O(\sum_{q \in \cH} \sum_{v \in H'_p \cap q} \log |H_v|)$.
+Summing over all $p$, the expression is $\sum_{q \in \cH} \sum_{p \in \cL'} \sum_{v \in H'_p \cap q} \log |H_v|$.
+%and $\sum_{p \in \cL'} |H_p|\log |H_p| = O(\sum_{p \in \cL'} \sum_{v \in H'_p} \log |H_v|)$.
 
@@
 then their corresponding supporting ancestor chains are disjoint. Any $v \in H'_p$
-can only occur in one other $H'_q$ (for $q \in \cL'$), because $v$ has exactly two colors.
-In other words, any $v$ appears at most twice in the above double summation.
-These vertices are all present in heavy paths, so we can bound
-the double summation by $\sum_{p \in \cH} \sum_{v \in p} \log |H_v|$.
+can only occur in one other $H'_{\hat{p}}$ (for $\hat{p} \in \cL'$), because $v$ has exactly two colors.
+In other words, any $\log |H_v|$ appears at most twice in the above triple summation.
+Hence, we can bound it by $O(\sum_{q \in \cH} \sum_{v \in q} \log |H_v|)$.
 %
@@
 %$\sum_{q \in Q} |H_q| \log |H_q| \leq 4 \sum_{p \in A(Q)} \sum_{v \in p} |H_v|$.
-We apply the bound of \Lem{pathBounds} to complete the proof.
-\end{proof}
-
+%We apply the bound of \Lem{pathBounds} to complete the proof.
+\end{proof}
+
+The first term (unsupported light paths) can be charged to the second term (supported light paths).
+This is where the critical \Lem{cu-root} plays a role.
+
+\begin{claim} \label{clm:first} $\sum_{p \in \cL \setminus \cL'} \sum_{v \in p} \log |H_v| = O(\sum_{p \in \cL'} |H_p| \log |H_p|)$.
+\end{claim}
+
+\begin{proof} Let $\cU$ denote a connected component of the shrub forest.
+We have $\sum_{p \in \cL \setminus \cL'} \leq \sum_{\cU} \sum_{p \in \cU} \sum_{v \in p} \log |H_v|$.
+By \Lem{cu-root}, $|H_v| = O(|H_r|)$, where $r$ is the root of $\cU$.
+Furthermore, $\sum_{p \in \cU} |p| = O(|H_r|)$. 
+We have $\sum_{p \in \cU} \sum_{v \in p} \log |H_v| = O((\log |H_r|) \sum_{p \in \cU} |p|) = O(|H_r|\log |H_r|)$.
+We sum this over all $\cU$ in the shrub forest, and note that roots in the shrub forest are supported light paths.
+\end{proof}
+
+\ignore{
 Now, for the critical piece of the approach. Focus on $\pmax(T)$. Let $L$ be the set of light paths, 
@@
 The main challenge is proving \Lem{light}, which we defer to the next section.
-
-\ignore{
 \begin{definition}
@@
 \end{definition}
-}
-
-\subsection{The shrub of unsupported paths} \label{sec:light}
-
-Our focus is on $\pmax(T)$.
-We can simplify a little. We split the summation and apply \Lem{pathBounds}. Finally,
-we note that for $p \in \cL$, $|p| \leq |H_p|$.
-\begin{eqnarray*} 
-\sum_{p \in \cL} \sum_{v \in p} \log |H_v| & = & \sum_{p \in \cL \setminus \cL'} \sum_{v \in p} \log |H_v| + \sum_{p \in \cL'} \sum_{v \in p} \log |H_v|\\
-& = & \sum_{p \in \cL \setminus \cL'} \sum_{v \in p} \log |H_v| + O(\sum_{p \in \cL'} |p|\log |H_p|) = \sum_{p \in \cL \setminus \cL'} \sum_{v \in p} \log |H_v| + O(\sum_{p \in \cL'} |H_p|\log |H_p|)
-\end{eqnarray*}
-
-\ignore{
+
 From the previous section we now have a lower bound of $\Omega(f_{path}(P))$ for computing the merge tree in terms of any path decomposition $P$. 
@@
 
-It suffices to focus on unsupported light paths. \Lem{light} can be reduced to the following.
-
-\begin{lemma} \label{lem:support} $\sum_{p \in \cL \setminus \cL'} \sum_{v \in p} \log |H_v| = O(\sum_{p \in \cL'} |H_p| \log |H_p|)$.
-\end{lemma}
-
-We now construct the shrub of unsupported light paths. Consider $p \in \cL \setminus \cL'$,
-and traverse the chain of ancestors from $p$. Eventually, we must reach another light path $q$.
-(If not, we have reached the root $r$ of $\pmax(T)$. But $|H_r| = 0$, implying $p$ is supported.)
-Insert edge from $p$ to $q$, so $q$ is the parent of $p$ in $\cU$. This construction leads
-to the shrub forest of $\cL \setminus \cL'$, where every unsupported light path is an internal node, and all
-roots are supported light paths. \Lem{support} is a direct consequence of the following,
-which is the really the main lemma.
-
-%\begin{lemma} \label{lem:cu} Let $\cU$ denote a connected component (shrub) in the shrub forest of $\cL \setminus \cL'$ and let $r$
-%be the root of $\cU$. Then $\sum_{p \in \cU} \sum_{v \in p} \log |H_v| = O(|H_r| \log |H_r|)$.
-%\end{lemma}
-%
-%We proceed with the setup to prove \Lem{cu}.
-%We focus our attention on some shrub $\cU$.
-%We prove a technical lemma about $\cU$, from which \Lem{cu} follows directly.
-
-\begin{lemma} \label{lem:cu-root} Let $\cU$ denote a connected component (shrub) in the shrub forest of $\cL \setminus \cL'$ and let $r$
-be the root of $\cU$. (i) For any $v \in p$ such that $p \in \cU$,
-$|H_v| = O(|H_r|)$. (ii) $\sum_{p \in \cU} |p| = O(|H_r|)$.
-\end{lemma}
-
-\begin{proof} (of \Lem{support}) We have $\sum_{p \in \cU} \sum_{v \in p} \log |H_v| = O((\log |H_r|) \sum_{p \in \cU} |p|) = O(|H_r|\log |H_r|)$.
-We sum this over all $\cU$ in the shrub forest, and note that roots in the shrub forest are supported light paths.
-\end{proof}
-
 \subsection{Proving \Lem{cu-root}: the root is everything in $\cU$} \label{sec:weight}
@@
 \Lem{cu-root} asserts the root $r$ in $\cU$ pretty much encompasses all sizes and heaps in $\cU$.
-Consider $q \in \cL \setminus \cL'$ that is a child of $p$ in the shrub $\cU$.
-By construction, there is a heavy path (denoted $h(q)$) that is a child of $p$ in $\pmax(T)$,
-and an ancestor of $q$. Furthermore, $|H_{h(q)}| > 99|H_q|/100$.
+%Consider $q \in \cL \setminus \cL'$ that is a child of $p$ in the shrub $\cU$.
 We begin with a key property, which is where the construction of $\pmax(T)$ enters the picture.
@@
 \end{lemma}
-\begin{proof} We know that $h(q)$ is a child of $p$ in $\pmax(T)$. Consider the base
+\begin{proof} By construction, there is a heavy path (denoted $h(q)$) that is a child of $p$ in $\pmax(T)$,
+and an ancestor of $q$. The chain of ancestors from $q$ to $h(q)$ consists only of heavy paths.
+These paths contain at most $|H_p|/100$ vertices of $H_p$, so $|H_{h(q)}| \geq 99|H_p|/100$.
+
+Consider the base
 of $h(q)$, which is a node $w$ in $T$. Let $v$ denote the sibling of $w$ in $T$.
@@
 
-This immediately proves part (i) of \Lem{cu-root}. Part (ii) requires some more work.
+This immediately proves part (i) of \Lem{cu-root}. Part (ii) requires much more work.
+It is helpful to spell out how the various heaps evolve. Consider $p \in \cU$,
+which has children $q_1, q_2, \ldots, q_k$ in $\pmax(T)$.
+Furthermore, as a leaf path in $p$, some leaf
+$\ell$ of $T$ in one of the ends of $p$. The heap $H_p$ is the
+union of $H_\ell$ and $H_{q_1}, H_{q_2}, \ldots, H_{q_k}$, with the vertices in $p$
+removed. Note that some of the $q_i$'s are the corresponding $h(q')$ vertices,
+for a child $q'$ in $\cU$.
 
@@
 
+\subsection{The shrub of unsupported paths} \label{sec:light}
+
+It suffices to focus on unsupported light paths. \Lem{light} can be reduced to the following.
+
+\begin{lemma} \label{lem:support} $\sum_{p \in \cL \setminus \cL'} \sum_{v \in p} \log |H_v| = O(\sum_{p \in \cL'} |H_p| \log |H_p|)$.
+\end{lemma}
+
+We now construct the shrub of unsupported light paths. Consider $p \in \cL \setminus \cL'$,
+and traverse the chain of ancestors from $p$. Eventually, we must reach another light path $q$.
+(If not, we have reached the root $r$ of $\pmax(T)$. Hence, $p$ is supported.)
+Insert edge from $p$ to $q$, so $q$ is the parent of $p$ in $\cU$. This construction leads
+to the shrub forest of $\cL \setminus \cL'$, where every unsupported light path is an internal node, and all
+roots are supported light paths.
+
+We list out some important claims about this shrub forest.
+We focus on $\cU$, a connected component (shrub) in the shrub forest and denote the 
+the root by $r(\cU)$.
+
+\begin{claim} \label{clm:loss} Consider $p \in \cL \setminus \cL'$ that is in $\cU$. Let $Q$
+be the set of heavy ancestors of $p$ that are descendants of $r(\cU)$. Then $\sum_{q \in Q} |q \cap H_p| \leq |H_p|/50$.
+\end{claim}
+
+\begin{proof} Suppose not, for contradictions sake. As we traverse the heavy ancestors of $p$,
+we encounter various vertices of $H_p$ in the ancestors $q$. Consider the first $|H_p|/100$
+such vertices of $p$ encountered. At each such $v$, there are at least $|H_p|/100$
+vertices \emph{not yet} encountered, so $|H_v| > |H_p|/100$. Each of these vertices $v$
+is supported, implying that $p$ is a supported light path.
+\end{proof}
+
+\Lem{support} is a direct consequence of the following,
+which is the really the main lemma.
+
+%\begin{lemma} \label{lem:cu} Let $\cU$ denote a connected component (shrub) in the shrub forest of $\cL \setminus \cL'$ and let $r$
+%be the root of $\cU$. Then $\sum_{p \in \cU} \sum_{v \in p} \log |H_v| = O(|H_r| \log |H_r|)$.
+%\end{lemma}
+%
+%We proceed with the setup to prove \Lem{cu}.
+%We focus our attention on some shrub $\cU$.
+%We prove a technical lemma about $\cU$, from which \Lem{cu} follows directly.
+
+\begin{proof} (of \Lem{support})
+\end{proof}
+
 \end{document}

### 2014-09-23 paint-contour-tree.tex
more changes

@@
 
+We define a \emph{residue} $R_p$ for each $p \in \cU$. Suppose $p$ has children
+$q_1, q_2, \ldots, q_k$ in $\cU$. Then $R_p = |H_p| - \sum_i |H_{q_i}|$. 
+By definition, $|H_p| = \sum_{q \in \cU_p} R_p$. 
+Note that $R_p$ can be negative. 
+Now, define $R^+_p = \max(R_p,0)$, and set $W_p = \sum_{q \in \cU_p} R^+_p$.
+Observe that $W_p \geq |H_p|$. We also get an approximate converse.
+
+\begin{claim} \label{clm:loss} $|H_p| \geq W_p - 2\sum_{q \in \cU_p} |q| - 2\sum_{q \in \cU_p} \sum_{h \in \cH} |h \cap R_q|$
+\end{claim}
+
+\begin{proof} Consider $q \in \cU_p$, and traverse $\pmax(T)$ up to $p$. Suppose we encounter
+the paths $q_0 = q, q_1, q_2, \ldots, q_k = p$. The vertices in the residue $R_q$ that 
+are removed from $H_p$ are exactly the vertices that lie in these paths. 
+We split this contribution into heavy and light paths, observing that the latter
+are all unsupported. Hence, we can (crudely) upper bound the number of vertices
+removed by $\sum_{q \in \cU_p} |q| + \sum_{q \in \cU_p} \sum_{h \in \cH} |h \cap q|$.
+Each vertex can be removed from at most two distinct $R_q$'s, completing the proof.
+\end{proof}
+
+The main challenge of the entire proof is bounding the sum of path lengths, which is done next.
+We stress that the all the previous work is mostly setup to get to this claim.
+
+\begin{claim} \label{clm:charge} Consider $p \in \cU$ such that for any $q,q' \in \cU_p$
+such that $q$ is a parent of $q'$, $W_q \geq (4/3)W_{q'}$. Then $\sum_{q \in \cU_p} |q| \leq W_p/20$.
+\end{claim}
+
+\begin{proof} By the definition of light paths, $|q| < \sqrt{|H_q|}/100 \leq \sqrt{|W_q|}/100$.
+We prove that $\sum_{q \in \cU_p} \sqrt{|W_q|}/100 \leq W_p/20$
+by a charge redistribution scheme. Assume that each $q \in \cU_p$
+starts with $\sqrt{W_p}/100$ units of charge. We redistribute this charge over all
+nodes in $\cU_p$, and then calculate the total charge. For $q \in \cU_p$, spread
+its charge to all nodes in $\cU_q$ proportional to $R^+$ values. In other words,
+give $(\sqrt{W_q}/100)\cdot(R^+_{q'}/W_q)$ units of charge to each $q' \in \cU_q$.
+
+After the redistribution, let us compute the charge deposited at $q$. Every ancestor in $\cU_p$.
+$q = a_0, a_1, a_2, \ldots, a_k$ contributes to the charge at $q$. The charge is expressed in the following
+equation. We use the assumption that $W_{a_i} \geq (4/3) W_{a_{i-1}}$ and hence $W_{a_i} \geq (4/3)^i$.
+$$ ({R^+_q}/100)\sum_{a_i} 1/\sqrt{W_{a_i}} \leq ({R^+_q}/100)\sum_{a_i} (3/4)^{i/2} \leq R^+_q/20 $$
+The total charge is $\sum_{q \in \cU_p} R^+_p/20 = W_p/20$.
+\end{proof}
+
+We wrap it all up by proving part (ii) of \Lem{cu-root}.
+
+\begin{claim} $\sum_{p \in \cU} |p| \leq H_{r(\cU)}/10$.
+\end{claim}
+
+\begin{proof} We use $r$ for $r(\cU)$. Our aim is to simply apply \Clm{charge} from $r(\cU)$. 
+Suppose $W_q \geq (4/3)W_{q'}$ (for $q$ parent of $q'$). By \Clm{charge}, $\sum_{p \in \cU} |p| \leq W_{r}/20$.
+By \Clm{loss}, $|H_r| \geq W_r - 2\sum_{p \in \cU} |p| - 2\sum_{h \in \cH} |h \cap H_p|$.
+
+We first need to prove the claim assumption,
+namely that of $W_q \geq (4/3)W_{q'}$ (for $q$ parent of $q'$). We do know from \Lem{geometric}
+that $|H_q| \geq (3/2)|H_{q'}|$. We will apply \Clm{loss} and \Clm{charge} to prove this.
+
+Pick the node furthest from $r(\cU)$
+
+\end{proof}
+
 \begin{proof} (part (ii) of \Lem{cu-root}) Let level $\ell$ of $\cU$ denote all nodes at distance $\ell$ from the root $r$.

### 2014-09-24 paint-contour-tree.tex
First pass done for upper bound

--- 2014-09-22 paint-contour-tree.tex
@@
 \newcommand{\reeb}{R}
+\newcommand{\redH}{\widetilde{H}}
 \newcommand{\rep}{rep}
@@
  A \emph{leaf assignment} $\chi$ of a tree $T$ assigns \emph{two} distinct leaves in each internal vertex $v$,
- one from the left and one from the right subtree of $v$.
+ one from the left child and one from the right child subtree of $v$.
 \end{definition}
@@
 \label{def:path} A \emph{path decomposition}, $P(T)$, is a partition of $T$ into 
- a set of vertex disjoint leaf paths.  Note that the number of paths in $P(T)$ must equal the number of leaves in $T$.
+ a set of vertex disjoint leaf paths.  
 % 
@@
 
-Fix $\chi(T)$, which fixes all the heaps $H_v$. We choose a special path decomposition, which can be related
-to the running time of the algorithm. This is defined best as a subset of edges in $T$ such that each internal
-vertex has degree exactly $1$. This naturally gives as path decomposition.
+Note that the number of paths in $P(T)$ must equal the number of leaves in $T$.
+Fix $\chi$, which fixes all the heaps $H_v$. We choose a special path decomposition that is best defined as a subset of edges in $T$ such that each internal
+vertex has degree at most $2$. This naturally gives a path decomposition.
 For each internal vertex $v\in T$, add the edge from $v$ to $\arg \max_{v_l, v_r} \{|H_{v_l}|, |H_{v_r}|\}$ 
@@
 
-\begin{theorem} \label{thm:runtime} $\sum_{v \in T} \log |H_v| = O(\sum_{p\in P(T)} |p| \log |p|)$.
+\begin{theorem} \label{thm:runtime} $\sum_{v \in T} \log |H_v| = O(\sum_{p\in \pmax(T)} |p| \log |p|)$.
 \end{theorem}
@@
 
-Most of work goes into proving the following technical lemma.
+Most of the work goes into proving the following technical lemma.
 
@@
 We split the remaining argument into two subsections. We first prove \Thm{runtime} from \Lem{cu-root},
-which mostly imvolves routing calculations. Then we prove \Lem{cu-root}, where most
-of the interesting work happens.
+which involves routine calculations. Then we prove \Lem{cu-root}, where the interesting work happens.
 
@@
 \Lem{cu-root} asserts the root $r$ in $\cU$ pretty much encompasses all sizes and heaps in $\cU$.
+We will work with the \emph{reduced} heap $\redH_p$. This is the subset of $H_p$
+of vertices that only appear in light paths. By definition, for any unsupported light path
+(hence, any non-root $p \in \cU$), $|H'_p| \geq 99|H_p|/100$. For notational convenience, we simply
+set $H'_r = H_r$.
 %Consider $q \in \cL \setminus \cL'$ that is a child of $p$ in the shrub $\cU$.
@@
 \label{lem:geometric}
- Let $p$ be the parent of $q$ in the shrub $\cU$. Then $|H_p|\geq \frac{3}{2}|H_q|$. 
+ Let $p$ be the parent of $q$ in the shrub $\cU$. Then $|\redH_p|\geq \frac{3}{2}|\redH_q|$. 
 \end{lemma}
@@
 and an ancestor of $q$. The chain of ancestors from $q$ to $h(q)$ consists only of heavy paths.
-These paths contain at most $|H_p|/100$ vertices of $H_p$, so $|H_{h(q)}| \geq 99|H_p|/100$.
+Since $q$ is unsupported, these paths contain at most $|H_p|/100$ vertices of $H_p$. Thus, $|H_{h(q)}| \geq 99|H_p|/100$.
 
@@
 Applying this bound, we get $|H_p| \geq (2-\delta)|H_w|$ (for small constant $\delta > 0$).
-Since $w$ is the base of $h(q)$, $H_w = H_{h(q)}$. We apply the bound $|H_{h(q)}| > 99|H_q|/100$,
-to complete the proof.
-
+Since $w$ is the base of $h(q)$, $H_w = H_{h(q)}$. We apply the bound $|H_{h(q)}| > 99|H_q|/100$
+to get $|H_p| \geq 197|H_q|/100$. The proof is completed by noting that $|\redH_p| \geq 99|H_p|/100$ and $|H_q| \geq |\redH_q|$.
 %
@@
 Furthermore, as a leaf path in $p$, some leaf
-$\ell$ of $T$ in one of the ends of $p$. The heap $H_p$ is the
-union of $H_\ell$ and $H_{q_1}, H_{q_2}, \ldots, H_{q_k}$, with the vertices in $p$
+$\ell$ of $T$ in one of the ends of $p$. The reduced heap $\redH_p$ is the
+union of $\redH_\ell$ and $\redH_{q_1}, \redH_{q_2}, \ldots, \redH_{q_k}$, with the vertices in $p$
 removed. Note that some of the $q_i$'s are the corresponding $h(q')$ vertices,
@@
 
-\begin{proof} (part (ii) of \Lem{cu-root}) Let level $\ell$ of $\cU$ denote all nodes at distance $\ell$ from the root $r$.
-These set of these nodes is designated as $\cU_\ell$. We set $\sigma_\ell = \sum_{p \in \cU_\ell} |p|$
-and $\sigma_{\leq \ell} = \sum_{k \leq \ell} \sigma_k$.
-By \Lem{geometric}, for any $p \in \cU_\ell$, $|H_p| \leq (2/3)^\ell |H_r|$.
-
-Since all nodes in $\cU$ represent light paths, $\forall p \in \cU$, $|p| < \sqrt{|H_p|}/100$.
-So $\sigma_\ell < \sum_{p \in \cU_\ell} \sqrt{|H_p|}/100$.
-
-\end{proof}
-
-To prove this, we need to understand how $H_p$ is constructed. The path $p$ has some children
-$q_1, q_2, \ldots, q_k$ in $\pmax(T)$. Furthermore, as a leaf path in $p$, some leaf
-$\ell$ of $T$ in one of the ends of $p$. The heap $H_p$ is essentially the
-union of $H_\ell$ and $H_{q_1}, H_{q_2}, \ldots, H_{q_k}$, with the vertices in $p$
-removed. Note that some of the $q_i$'s are the corresponding $h(q')$ vertices,
-for a child $q'$ in $\cU$.
-
-Let $\mathcal{L} = \mathcal{L}(r)$ be a light shrub of some light path $r$ in a maximum path decomposition $\mathcal{P}(T)$.
-The intuitive idea is that due to the geometric decreasing behavior of light shrubs described in the previous subsection, 
-one can show that for the entire light shrub our algorithm pays roughly the cost due to the root path $r$.  This argument 
-is hard to make directly, so instead we define an alternative quantity for which we can make the argument.
-Specifically, we now define a weight function $w(v)$ for all $v\in \mathcal{L}$, and show that $w(v)=\Theta(|H_v|)$.
-
-For a path $v\in \mathcal{L}$, let $H_v$ denote the heap at the root of $v$, and let $H_{v_f}$ denote the heap 
-at the leaf of $v$.  Let $S_l(v)$ (resp. $S_h(v)$) be the subset of light (resp. heavy) paths adjacent to $v$ in $\pathTreeA(T)$.
-
-We want to understand how the heap size changes as we move around the paths in $\mathcal{L}$.  Specifically, there are 
-three places the vertices in the heap along a light path can come from; adjacent light paths, adjacent heavy paths, and the initial heap at the leaf of the path.  
-Given this observation we now define a weight function which upper bounds the size of the heap at the root of any path in $\mathcal{L}$.
-
-First, for all $v\in \mathcal{L}$ we define the local cost function $c(v) = H_{v_f}+\sum_{i\in S_h(v)} H_i$, this is the addition to the heap at $v$ which 
-was not already charged to descendant light paths.  We now define a weighted version of $\mathcal{L}$ based on these costs.  
-To do so, we first augment $\mathcal{L}$ by attaching one 
-leaf to each vertex in $\mathcal{L}$.  Now for each newly created leaf (which are the only leaves now) assign the cost 
-$c(v)$ defined above, where $v$ is that leaf's corresponding parent.  
-Now we can define a weight function $w(u)$ over any node $u$ in the tree, which is defined to be the total cost of all leaves in its subtree.   
-To simplify notation, we now use $\mathcal{L}$ to refer to this new weighted tree (rather than its unweighted counterpart).
-
-As the weight $w(p)$ includes the sizes of the heaps of all contributing heavy paths and leaves in the subtree rooted at $p$, we have the following.
-
-\begin{claim}
-For any $p\in \mathcal{L}$, $w(p)\geq |H_p|$.  Moreover, for any $v\in p$, $w(p)\geq |H_v|$.
+We define a \emph{residue} $R_p$ for each $p \in \cU$. Suppose $p$ has children
+$q_1, q_2, \ldots, q_k$ in $\cU$. Then $R_p = |\redH_p| - \sum_i |\redH_{q_i}|$. 
+By definition, $|\redH_p| = \sum_{q \in \cU_p} R_p$. 
+Note that $R_p$ can be negative. 
+Now, define $R^+_p = \max(R_p,0)$, and set $W_p = \sum_{q \in \cU_p} R^+_p$.
+Observe that $W_p \geq |\redH_p|$. We also get an approximate converse.
+
+\begin{claim} \label{clm:loss} $|\redH_p| \geq W_p - 2\sum_{q \in \cU_p} |q|$
 \end{claim}
 
-%Let $W= \sum_{v\in \mathcal{L}} w(v) \geq \sum_{v\in \mathcal{L}} |H_v|$.  
-%Suppose we could show that conversely $W=O(H_r)$.  Then we would have 
-%$\sum_{v\in \mathcal{L}} H_v = O(H_r)$, which was our original goal.  
-%So proving $W=O(H_r)$ is now our new goal.
-
-We now would like to show conversely that $w(v)= O(|H_v|)$.  
-To do so we first introduce some notation.  
-Let $C\subseteq \mathcal{P}$ be the subset of paths that are leaves in $\mathcal{L}$.
-For any path $v$, let $\mathcal{L}_v$ (resp. $C_v$) be the subset of paths in $\mathcal{L}$ (resp $C$) 
-that are in the subtree of $\mathcal{L}$ which is rooted at $v$.
-
-First, observe that $|H_v|\geq w(v)-\sum_{u\in \mathcal{L}_v} |u|$.  
-This holds by the same argument as in \Lem{adjacent}, namely, $w(v)$ by definition 
-counts all the contributing heavy path heaps and initial leaf heaps that contribute 
-to the heap at $v$ and $\sum_{u\in \mathcal{L}_v} |u|$ is an upper bound on the number 
-of vertices that could have been removed from one of these heaps before reaching the 
-root of $v$ (since a vertex in a heap can only get removed if it appears on one of 
-the paths).
-
-As all paths in $\mathcal{L}$ are light, we thus have the following.
-
-\begin{claim}
-$|H_v|\geq w(v)-\sum_{u\in \mathcal{L}_v} \sqrt{|H_u|}/100 \geq w(v)-\sum_{u\in \mathcal{L}_v} \sqrt{w(u)}/100$
+\begin{proof} We write $W_p - |\redH_p| = \sum_{q \in \cU_p} R^+_q - R_q$ $= -\sum_{q \in \cU_p: R_q < 0} R_q$.
+Consider $q \in \cU_p$ and denote the children in $\cU_p$ by $q'_1, q'_2, \ldots$.
+Note that $R_q$ is negative exactly when $|\redH_q| < \sum_i |\redH_{q'_i}|$.
+Traverse $\pmax(T)$ from $q'_i$ to $q$.
+Other than $q$, all other nodes encountered are heavy paths and hence
+do not affect reduced heaps.
+The vertices of $\redH_{q'_i}$ that are deleted
+are exactly those present in the path $q$ (which are paths in $T$) traversed. 
+Any vertex in $q$ can be deleted from at most two reduced heaps $\redH_{q'_i}$.
+When $R_q$ is negative, it is at most $2|q|$.
+We sum over all $q$ to complete the proof.
+\end{proof}
+
+\ignore{Let $D_q = H_q \setminus \bigcup_i H_{q'_i}$. In other words, as we traverse $\cU$,
+$D_q$ is the set of vertices ``added" at $q$. Suppose we traverse $\pmax(T)$ from $q'_i$
+to $q$. All vertices of $H_{q'_i}$ that lie on these paths are removed.
+
+ and traverse $\pmax(T)$ up to $p$. Suppose we encounter
+the paths $q_0 = q, q_1, q_2, \ldots, q_k = p$. The vertices in the residue $R_q$ that 
+are removed from $H_p$ are exactly the vertices that lie in these paths. 
+We split this contribution into heavy and light paths, observing that the latter
+are all unsupported. Hence, we can (crudely) upper bound the number of vertices
+removed by $\sum_{q \in \cU_p} |q| + \sum_{q \in \cU_p} \sum_{h \in \cH} |h \cap q|$.
+Each vertex can be removed from at most two distinct $R_q$'s, completing the proof.
+}
+
+The main challenge of the entire proof is bounding the sum of path lengths, which is done next.
+We stress that the all the previous work is mostly the setup for this claim.
+
+\begin{claim} \label{clm:charge} Fix $p \in \cU$. Suppose for any $q,q' \in \cU_p$
+where $q$ is a parent of $q'$ in $\cU_p$, $W_q \geq (4/3)W_{q'}$. Then $\sum_{q \in \cU_p} |q| \leq W_p/20$.
 \end{claim}
 
-By the above claim, if we can prove $\sum_{u\in \mathcal{L}_v} \sqrt{w(u)}/100 \leq w(v)/c$ for some constant $c>1$, 
-then we will have proven that $|H_v| = \Theta(w(v))$.
-
-Recall that \Lem{geometric} told us that that $|H_{u}|\leq (3/5) |H_{u'}|$, if $u'$ is the parent $u$ in $\mathcal{L}$. 
-The following lemma shows that if we had a similar statement in terms of the weights rather than the heap sizes, 
-then we get the desired bound on the sum.
-
-\begin{lemma}
-\label{lem:distribution}
-Let $v$ be a path in $\mathcal{L}$.  Suppose that for any two paths $x,y\in \mathcal{L}_v$ 
-such that $y$ is the parent of $x$, $w(x)\leq (4/5) w(y)$.
-Then $\sum_{u\in \mathcal{P}_v} \sqrt{w(u)}/100 \leq \sum_{c\in C_v} \sqrt{w(c)}/10$
-\end{lemma}
-
-\begin{proof}
-The idea is to recursively redistribute the contribution to the sum of each internal vertex to its children, 
-until all the weight sits on the leaves.  Specifically, let $v$ be some path in $\mathcal{L}$.  Its 
-contribution to the sum is $\sqrt{w(v)}/100$, which we will call the value at $v$.  We redistribute this 
-value to its children, in proportion by the weight of each child.  For example, if $v$ has a child $x$ then 
-the fraction it gets assigned is $w(x)/w(v)$ (recall that by definition $w(v)$ is the sum of the weights of 
-its children and so the total value will get reassigned to the children).
-
-Suppose we keep redistributing the value of $v$ until it reaches the leaves of its subtree in $\mathcal{L}$.  
-Let's see how much gets assigned to a given leaf.  Specifically, let $u_0, u_1, \ldots, u_k$ 
-be a path in $\mathcal{L}$ (of paths from the set $\mathcal{P}_v$), such that $v=u_0$, $u_{i+1}$ 
-is child of $u_i$ for all $i$, and $u_k$ is a leaf of $\mathcal{L}$.
-
-Then the total value from $u_0$ that ends up at $u_k$ is 
-\[
-\frac{\sqrt{w(u_0)}}{100} \pth{ \frac{w(u_1)}{w(u_0)} \frac{w(u_2)}{w(u_1)} \cdots \frac{w(u_k)}{w(u_{k-1})} }
-= \frac{\sqrt{w(u_0)}}{100} \pth{ \frac{w(u_k)}{w(u_0)} } = \frac{w(u_k)}{100\sqrt{w(u_0)}}
-\]
-
-For a given leaf in $\mathcal{L}$, we can calculate the total value reassigned to it from it its ancestors.
-Specifically, let $u_0, u_1, u_2, \dots, u_m$ be a root to leaf path in $\mathcal{L}$ 
-(i.e. $u_0$ is the root $r$ of $\mathcal{L}$).
-
-By assumption from the lemma statement we have $w(u_{i+1})\leq (4/5) w(u_i)$.  Therefore by the above calculation, 
-the total reassigned to $u_m$ is
-
-\[
- \sum_{i=0}^m \frac{w(u_m)}{100\sqrt{w(u_i)}} \leq \frac{w(u_m)}{100} \sum_{i=0}^m 1 \Big/ \sqrt{\pth{\frac{5}{4}}^iw(u_m)} 
- \leq \frac{\sqrt{w(u_m)}}{100} \sum_{i=0}^{\infty} \pth{\frac{4}{5}}^{i/2} \leq \frac{\sqrt{w(u_m)}}{10}
-\]
-
-The claim then follows by summing this quantity over the children of $\mathcal{L}$.
-\end{proof}
-
-By combining the above lemma and claim we have the following.
-
-\begin{claim}
-For any path $v\in \mathcal{L}$ if it holds that 
-$w(x)\leq (4/5) w(y)$ for any two paths $x,y\in \mathcal{L}_v$ such that $y$ is the parent of $x$, then
-
-\[
-|H_v|\geq w(v)-\sum_{u\in \mathcal{L}_v} \sqrt{w(u)}/100 \geq w(v)- \sum_{c\in C_v} \sqrt{w(c)}/10 \geq \frac{4}{5} w(v)
-\]
+\begin{proof} Since $q$ is an unsupported light path, $|q| < \sqrt{|H_q|}/100 \leq \sqrt{|\redH_q|}/99 \leq \sqrt{|W_q|}/99$.
+We prove that $\sum_{q \in \cU_p} \sqrt{|W_q|}/99 \leq W_p/20$
+by a charge redistribution scheme. Assume that each $q \in \cU_p$
+starts with $\sqrt{W_p}/99$ units of charge. We redistribute this charge over all
+nodes in $\cU_p$, and then calculate the total charge. For $q \in \cU_p$, spread
+its charge to all nodes in $\cU_q$ proportional to $R^+$ values. In other words,
+give $(\sqrt{W_q}/99)\cdot(R^+_{q'}/W_q)$ units of charge to each $q' \in \cU_q$.
+
+After the redistribution, let us compute the charge deposited at $q$. Every ancestor in $\cU_p$ 
+$q = a_0, a_1, a_2, \ldots, a_k$ contributes to the charge at $q$. The charge is expressed in the following
+equation. We use the assumption that $W_{a_i} \geq (4/3) W_{a_{i-1}}$ and hence $W_{a_i} \geq (4/3)^i$.
+$$ ({R^+_q}/99)\sum_{a_i} 1/\sqrt{W_{a_i}} \leq ({R^+_q}/99)\sum_{a_i} (3/4)^{i/2} \leq R^+_q/20 $$
+The total charge is $\sum_{q \in \cU_p} R^+_p/20 = W_p/20$. 
+\end{proof}
+
+We wrap it all up by proving part (ii) of \Lem{cu-root}.
+
+\begin{claim} $\sum_{p \in \cU} |p| \leq |H_{r(\cU)}|/10$.
 \end{claim}
 
-\begin{lemma}
-Let $\mathcal{L}$ be the light shrub of some path $r$.  
-Then for any $v\in \mathcal{L}$, we have $|H_v|\geq \frac{4}{5} w(v)$.
-\end{lemma}
-% \begin{proof}
-% We will actually prove the stronger statement that for any $v\in \mathcal{L}$, 
-% we have $|H_v|\geq \frac{4}{5} w(v)$ and $w(v)\leq (4/5) w(p)$ where $p$ is the parent 
-% of $v$ in $\mathcal{L}$.
-% 
-% Suppose otherwise, and let $v$ be the lowest path in $\mathcal{L}$ such that this fails 
-% (i.e. the statement holds for all descendants of $v$ in $\mathcal{L}$).  Then for any 
-% two paths $x,y\in \mathcal{L}_v$ such that $y$ is the parent of $x$, 
-% we have that $w(x)\leq (4/5) w(y)$ and so by \Lem{distribution}, $|H_v|\geq \frac{4}{5} w(v)$.
-% Combining this fact with \Lem{geometric} gives,
-% \[
-% w(v) \leq \frac{5}{4}|H_v| \leq \frac{5}{4} \frac{3}{5} |H_p| \leq \frac{3}{4} w(p).
-% \]
-% 
-% \end{proof}
-%
-\begin{proof}
-By the above claim, we know that $|H_v|\geq \frac{4}{5} w(v)$ if it holds that 
-$w(x)\leq (4/5) w(y)$ for any two paths $x,y\in \mathcal{L}_v$ such that $y$ is the parent of $x$.
-So let $v$ be the lowest path in $\mathcal{L}$ such that this fractional decrease fails to hold, i.e. $w(c)> (4/5) w(v)$ 
-where $c$ is a child of $v$ in $\mathcal{L}$, and conversely all vertices in $\mathcal{L}_c$ 
-the fractional decrease holds.  As the fractional decrease holds for a paths in $\mathcal{L}_c$, 
-we have $|H_c|\geq \frac{4}{5} w(c)$.  Therefore,
-\[
-w(c) \leq \frac{5}{4}|H_c| \leq \frac{5}{4} \frac{3}{5} |H_v| \leq \frac{3}{4} w(v).
-\]
-However, this is a contradiction with the assumption $w(c)> (4/5) w(v)$ , and hence the 
-fractional decrease always holds implying the claim always holds.
-\end{proof}
-
-\begin{corollary}
-\label{cor:equivalent}
- Let $\mathcal{L}$ be a light shrub of a light path $r$.  Then for any $v\in \mathcal{L}$, $|H_v| \leq w(v) \leq \frac{5}{4} |H_v|$.
-\end{corollary}
-
-Now we are finally ready to make use of the properties we have proven about the weight function we introduced.
-
-\Ben{The constant in the lemmas below, but it is no more than twice that.}
-
-\begin{lemma}
-\label{lem:lightCost}
- Let $\mathcal{L}$ be a light shrub of a light path $r$.  Let $V$ be the set of heavy paths from the full shrub which are adjacent to paths in $\mathcal{L}$.  
-  Then $\sum_{p\in \mathcal{L}}\sum_{v\in p} \log |H_v| + \sum_{q\in V} |H_q|\log |H_q| \leq \frac{4}{3} |H_r|\log |H_r|$
-\end{lemma}
-\begin{proof}
-\Ben{
-The following is a proof for $\frac{3}{2}+\eps$, rather than $\frac{4}{3}$.  The discrepancy is due to a corrected calculation error.  
-4/3 can be achieved by tightening constants above, and is left in the lemma statement as it is used later.
-}
-
-Recall that in a proper coloring of a binary tree $T$, the color of any vertex must be the color of one of its descendants.  
-Therefore, any vertex in the light path shrub must appear in the heap of either the root of an adjacent 
-path from the full shrub or from the heap of a leaf of a path in $\mathcal{L}$.  
-Also note $|H_r|$ precisely counts sum of the number of vertices from these heaps that do not appear in $\mathcal{L}$.
-As there are no adjacent light paths to a light shrub, the total number of vertices over all paths in $\mathcal{L}$ is at most 
-$m = \sum_{p\in \mathcal{L}} |H_{p_f}| +\sum_{q\in V} |H_q| - |H_r| = w(r)-|H_r|$, where $H_{p_f}$ is the heap at the leaf of $p$.
-
-Using this observation and \Cor{equivalent} we then have, 
-\begin{align*}
- \sum_{p\in \mathcal{L}}\sum_{v\in p} \log |H_v| + \sum_{q\in V} |H_q|\log |H_q|
- &\leq \sum_{p\in \mathcal{L}}\sum_{v\in p} \log(w(r)) + \sum_{q\in V} |H_q|\log(w(r))\\
- &\leq \log(w(r))\pth{ w(r)-|H_r| } + \log(w(r))w(r)\\ 
- &= (2w(r)-|H_r|)\log(w(r)) \leq \pth{\frac{5}{2}|H_r|-|H_r|}\log\pth{\frac{5}{4}|H_r|}\\
- &= \frac{3}{2}|H_r|\log\pth{\frac{5}{4}|H_r|}
-\end{align*}
+\begin{proof} We use $r$ for $r(\cU)$. Our aim is to simply apply \Clm{charge} from $r(\cU)$. 
+Suppose $W_q \geq (4/3)W_{q'}$ (for any choice in $\cU$ of $q$ parent of $q'$). By \Clm{charge}, $\sum_{p \in \cU} |p| \leq W_{r}/20$.
+By \Clm{loss}, $|\redH_r| \geq W_r - 2\sum_{p \in \cU} |p|$. By simple algebra, we deduce, $\sum_{p \in \cU} |p| \leq |\redH_r|/10 = |H_r|/10$.
+
+We now prove that for any $q$ parent of $q'$, $W_q \geq (4/3)W_{q'}$.
+Suppose not. Let $p, p'$ be the counterexample furthest from the root,
+where $p$ is the parent of $p'$. 
+Note that for $q$ and child $q'$ in $\cU_{p'}$, $W_q \geq (4/3)W_{q'}$.
+We will apply \Clm{charge} for $\cU_{p'}$ to deduce that $\sum_{q \in \cU_{p'}} |q| \leq W_{p'}/20$.
+By \Clm{loss}, $|\redH_{p'}| \geq 19W_{p'}/20$. By \Lem{geometric}, $|\redH_p| \geq (3/2)|\redH_{p'}|$.
+Noting that $W_p \geq |\redH_p|$, we deduce that $W_p \geq (4/3)W_{p'}$.
+Hence, $p, p'$ is not a counterexample, and more generally, this is no counterexample.
+That completes the whole proof.
 \end{proof}
@@
 
-\begin{proof} (of \Lem{support})
+\section{Ben's old stuff}
+
+\begin{proof} (part (ii) of \Lem{cu-root}) Let level $\ell$ of $\cU$ denote all nodes at distance $\ell$ from the root $r$.
+These set of these nodes is designated as $\cU_\ell$. We set $\sigma_\ell = \sum_{p \in \cU_\ell} |p|$
+and $\sigma_{\leq \ell} = \sum_{k \leq \ell} \sigma_k$.
+By \Lem{geometric}, for any $p \in \cU_\ell$, $|H_p| \leq (2/3)^\ell |H_r|$.
+
+Since all nodes in $\cU$ represent light paths, $\forall p \in \cU$, $|p| < \sqrt{|H_p|}/100$.
+So $\sigma_\ell < \sum_{p \in \cU_\ell} \sqrt{|H_p|}/100$.
+
+\end{proof}
+
+To prove this, we need to understand how $H_p$ is constructed. The path $p$ has some children
+$q_1, q_2, \ldots, q_k$ in $\pmax(T)$. Furthermore, as a leaf path in $p$, some leaf
+$\ell$ of $T$ in one of the ends of $p$. The heap $H_p$ is essentially the
+union of $H_\ell$ and $H_{q_1}, H_{q_2}, \ldots, H_{q_k}$, with the vertices in $p$
+removed. Note that some of the $q_i$'s are the corresponding $h(q')$ vertices,
+for a child $q'$ in $\cU$.
+
+Let $\mathcal{L} = \mathcal{L}(r)$ be a light shrub of some light path $r$ in a maximum path decomposition $\mathcal{P}(T)$.
+The intuitive idea is that due to the geometric decreasing behavior of light shrubs described in the previous subsection, 
+one can show that for the entire light shrub our algorithm pays roughly the cost due to the root path $r$.  This argument 
+is hard to make directly, so instead we define an alternative quantity for which we can make the argument.
+Specifically, we now define a weight function $w(v)$ for all $v\in \mathcal{L}$, and show that $w(v)=\Theta(|H_v|)$.
+
+For a path $v\in \mathcal{L}$, let $H_v$ denote the heap at the root of $v$, and let $H_{v_f}$ denote the heap 
+at the leaf of $v$.  Let $S_l(v)$ (resp. $S_h(v)$) be the subset of light (resp. heavy) paths adjacent to $v$ in $\pathTreeA(T)$.
+
+We want to understand how the heap size changes as we move around the paths in $\mathcal{L}$.  Specifically, there are 
+three places the vertices in the heap along a light path can come from; adjacent light paths, adjacent heavy paths, and the initial heap at the leaf of the path.  
+Given this observation we now define a weight function which upper bounds the size of the heap at the root of any path in $\mathcal{L}$.
+
+First, for all $v\in \mathcal{L}$ we define the local cost function $c(v) = H_{v_f}+\sum_{i\in S_h(v)} H_i$, this is the addition to the heap at $v$ which 
+was not already charged to descendant light paths.  We now define a weighted version of $\mathcal{L}$ based on these costs.  
+To do so, we first augment $\mathcal{L}$ by attaching one 
+leaf to each vertex in $\mathcal{L}$.  Now for each newly created leaf (which are the only leaves now) assign the cost 
+$c(v)$ defined above, where $v$ is that leaf's corresponding parent.  
+Now we can define a weight function $w(u)$ over any node $u$ in the tree, which is defined to be the total cost of all leaves in its subtree.   
+To simplify notation, we now use $\mathcal{L}$ to refer to this new weighted tree (rather than its unweighted counterpart).
+
+As the weight $w(p)$ includes the sizes of the heaps of all contributing heavy paths and leaves in the subtree rooted at $p$, we have the following.
+
+\begin{claim}
+For any $p\in \mathcal{L}$, $w(p)\geq |H_p|$.  Moreover, for any $v\in p$, $w(p)\geq |H_v|$.
+\end{claim}
+
+%Let $W= \sum_{v\in \mathcal{L}} w(v) \geq \sum_{v\in \mathcal{L}} |H_v|$.  
+%Suppose we could show that conversely $W=O(H_r)$.  Then we would have 
+%$\sum_{v\in \mathcal{L}} H_v = O(H_r)$, which was our original goal.  
+%So proving $W=O(H_r)$ is now our new goal.
+
+We now would like to show conversely that $w(v)= O(|H_v|)$.  
+To do so we first introduce some notation.  
+Let $C\subseteq \mathcal{P}$ be the subset of paths that are leaves in $\mathcal{L}$.
+For any path $v$, let $\mathcal{L}_v$ (resp. $C_v$) be the subset of paths in $\mathcal{L}$ (resp $C$) 
+that are in the subtree of $\mathcal{L}$ which is rooted at $v$.
+
+First, observe that $|H_v|\geq w(v)-\sum_{u\in \mathcal{L}_v} |u|$.  
+This holds by the same argument as in \Lem{adjacent}, namely, $w(v)$ by definition 
+counts all the contributing heavy path heaps and initial leaf heaps that contribute 
+to the heap at $v$ and $\sum_{u\in \mathcal{L}_v} |u|$ is an upper bound on the number 
+of vertices that could have been removed from one of these heaps before reaching the 
+root of $v$ (since a vertex in a heap can only get removed if it appears on one of 
+the paths).
+
+As all paths in $\mathcal{L}$ are light, we thus have the following.
+
+\begin{claim}
+$|H_v|\geq w(v)-\sum_{u\in \mathcal{L}_v} \sqrt{|H_u|}/100 \geq w(v)-\sum_{u\in \mathcal{L}_v} \sqrt{w(u)}/100$
+\end{claim}
+
+By the above claim, if we can prove $\sum_{u\in \mathcal{L}_v} \sqrt{w(u)}/100 \leq w(v)/c$ for some constant $c>1$, 
+then we will have proven that $|H_v| = \Theta(w(v))$.
+
+Recall that \Lem{geometric} told us that that $|H_{u}|\leq (3/5) |H_{u'}|$, if $u'$ is the parent $u$ in $\mathcal{L}$. 
+The following lemma shows that if we had a similar statement in terms of the weights rather than the heap sizes, 
+then we get the desired bound on the sum.
+
+\begin{lemma}
+\label{lem:distribution}
+Let $v$ be a path in $\mathcal{L}$.  Suppose that for any two paths $x,y\in \mathcal{L}_v$ 
+such that $y$ is the parent of $x$, $w(x)\leq (4/5) w(y)$.
+Then $\sum_{u\in \mathcal{P}_v} \sqrt{w(u)}/100 \leq \sum_{c\in C_v} \sqrt{w(c)}/10$
+\end{lemma}
+
+\begin{proof}
+The idea is to recursively redistribute the contribution to the sum of each internal vertex to its children, 
+until all the weight sits on the leaves.  Specifically, let $v$ be some path in $\mathcal{L}$.  Its 
+contribution to the sum is $\sqrt{w(v)}/100$, which we will call the value at $v$.  We redistribute this 
+value to its children, in proportion by the weight of each child.  For example, if $v$ has a child $x$ then 
+the fraction it gets assigned is $w(x)/w(v)$ (recall that by definition $w(v)$ is the sum of the weights of 
+its children and so the total value will get reassigned to the children).
+
+Suppose we keep redistributing the value of $v$ until it reaches the leaves of its subtree in $\mathcal{L}$.  
+Let's see how much gets assigned to a given leaf.  Specifically, let $u_0, u_1, \ldots, u_k$ 
+be a path in $\mathcal{L}$ (of paths from the set $\mathcal{P}_v$), such that $v=u_0$, $u_{i+1}$ 
+is child of $u_i$ for all $i$, and $u_k$ is a leaf of $\mathcal{L}$.
+
+Then the total value from $u_0$ that ends up at $u_k$ is 
+\[
+\frac{\sqrt{w(u_0)}}{100} \pth{ \frac{w(u_1)}{w(u_0)} \frac{w(u_2)}{w(u_1)} \cdots \frac{w(u_k)}{w(u_{k-1})} }
+= \frac{\sqrt{w(u_0)}}{100} \pth{ \frac{w(u_k)}{w(u_0)} } = \frac{w(u_k)}{100\sqrt{w(u_0)}}
+\]
+
+For a given leaf in $\mathcal{L}$, we can calculate the total value reassigned to it from it its ancestors.
+Specifically, let $u_0, u_1, u_2, \dots, u_m$ be a root to leaf path in $\mathcal{L}$ 
+(i.e. $u_0$ is the root $r$ of $\mathcal{L}$).
+
+By assumption from the lemma statement we have $w(u_{i+1})\leq (4/5) w(u_i)$.  Therefore by the above calculation, 
+the total reassigned to $u_m$ is
+
+\[
+ \sum_{i=0}^m \frac{w(u_m)}{100\sqrt{w(u_i)}} \leq \frac{w(u_m)}{100} \sum_{i=0}^m 1 \Big/ \sqrt{\pth{\frac{5}{4}}^iw(u_m)} 
+ \leq \frac{\sqrt{w(u_m)}}{100} \sum_{i=0}^{\infty} \pth{\frac{4}{5}}^{i/2} \leq \frac{\sqrt{w(u_m)}}{10}
+\]
+
+The claim then follows by summing this quantity over the children of $\mathcal{L}$.
+\end{proof}
+
+By combining the above lemma and claim we have the following.
+
+\begin{claim}
+For any path $v\in \mathcal{L}$ if it holds that 
+$w(x)\leq (4/5) w(y)$ for any two paths $x,y\in \mathcal{L}_v$ such that $y$ is the parent of $x$, then
+
+\[
+|H_v|\geq w(v)-\sum_{u\in \mathcal{L}_v} \sqrt{w(u)}/100 \geq w(v)- \sum_{c\in C_v} \sqrt{w(c)}/10 \geq \frac{4}{5} w(v)
+\]
+\end{claim}
+
+\begin{lemma}
+Let $\mathcal{L}$ be the light shrub of some path $r$.  
+Then for any $v\in \mathcal{L}$, we have $|H_v|\geq \frac{4}{5} w(v)$.
+\end{lemma}
+% \begin{proof}
+% We will actually prove the stronger statement that for any $v\in \mathcal{L}$, 
+% we have $|H_v|\geq \frac{4}{5} w(v)$ and $w(v)\leq (4/5) w(p)$ where $p$ is the parent 
+% of $v$ in $\mathcal{L}$.
+% 
+% Suppose otherwise, and let $v$ be the lowest path in $\mathcal{L}$ such that this fails 
+% (i.e. the statement holds for all descendants of $v$ in $\mathcal{L}$).  Then for any 
+% two paths $x,y\in \mathcal{L}_v$ such that $y$ is the parent of $x$, 
+% we have that $w(x)\leq (4/5) w(y)$ and so by \Lem{distribution}, $|H_v|\geq \frac{4}{5} w(v)$.
+% Combining this fact with \Lem{geometric} gives,
+% \[
+% w(v) \leq \frac{5}{4}|H_v| \leq \frac{5}{4} \frac{3}{5} |H_p| \leq \frac{3}{4} w(p).
+% \]
+% 
+% \end{proof}
+%
+\begin{proof}
+By the above claim, we know that $|H_v|\geq \frac{4}{5} w(v)$ if it holds that 
+$w(x)\leq (4/5) w(y)$ for any two paths $x,y\in \mathcal{L}_v$ such that $y$ is the parent of $x$.
+So let $v$ be the lowest path in $\mathcal{L}$ such that this fractional decrease fails to hold, i.e. $w(c)> (4/5) w(v)$ 
+where $c$ is a child of $v$ in $\mathcal{L}$, and conversely all vertices in $\mathcal{L}_c$ 
+the fractional decrease holds.  As the fractional decrease holds for a paths in $\mathcal{L}_c$, 
+we have $|H_c|\geq \frac{4}{5} w(c)$.  Therefore,
+\[
+w(c) \leq \frac{5}{4}|H_c| \leq \frac{5}{4} \frac{3}{5} |H_v| \leq \frac{3}{4} w(v).
+\]
+However, this is a contradiction with the assumption $w(c)> (4/5) w(v)$ , and hence the 
+fractional decrease always holds implying the claim always holds.
+\end{proof}
+
+\begin{corollary}
+\label{cor:equivalent}
+ Let $\mathcal{L}$ be a light shrub of a light path $r$.  Then for any $v\in \mathcal{L}$, $|H_v| \leq w(v) \leq \frac{5}{4} |H_v|$.
+\end{corollary}
+
+Now we are finally ready to make use of the properties we have proven about the weight function we introduced.
+
+\Ben{The constant in the lemmas below, but it is no more than twice that.}
+
+\begin{lemma}
+\label{lem:lightCost}
+ Let $\mathcal{L}$ be a light shrub of a light path $r$.  Let $V$ be the set of heavy paths from the full shrub which are adjacent to paths in $\mathcal{L}$.  
+  Then $\sum_{p\in \mathcal{L}}\sum_{v\in p} \log |H_v| + \sum_{q\in V} |H_q|\log |H_q| \leq \frac{4}{3} |H_r|\log |H_r|$
+\end{lemma}
+\begin{proof}
+\Ben{
+The following is a proof for $\frac{3}{2}+\eps$, rather than $\frac{4}{3}$.  The discrepancy is due to a corrected calculation error.  
+4/3 can be achieved by tightening constants above, and is left in the lemma statement as it is used later.
+}
+
+Recall that in a proper coloring of a binary tree $T$, the color of any vertex must be the color of one of its descendants.  
+Therefore, any vertex in the light path shrub must appear in the heap of either the root of an adjacent 
+path from the full shrub or from the heap of a leaf of a path in $\mathcal{L}$.  
+Also note $|H_r|$ precisely counts sum of the number of vertices from these heaps that do not appear in $\mathcal{L}$.
+As there are no adjacent light paths to a light shrub, the total number of vertices over all paths in $\mathcal{L}$ is at most 
+$m = \sum_{p\in \mathcal{L}} |H_{p_f}| +\sum_{q\in V} |H_q| - |H_r| = w(r)-|H_r|$, where $H_{p_f}$ is the heap at the leaf of $p$.
+
+Using this observation and \Cor{equivalent} we then have, 
+\begin{align*}
+ \sum_{p\in \mathcal{L}}\sum_{v\in p} \log |H_v| + \sum_{q\in V} |H_q|\log |H_q|
+ &\leq \sum_{p\in \mathcal{L}}\sum_{v\in p} \log(w(r)) + \sum_{q\in V} |H_q|\log(w(r))\\
+ &\leq \log(w(r))\pth{ w(r)-|H_r| } + \log(w(r))w(r)\\ 
+ &= (2w(r)-|H_r|)\log(w(r)) \leq \pth{\frac{5}{2}|H_r|-|H_r|}\log\pth{\frac{5}{4}|H_r|}\\
+ &= \frac{3}{2}|H_r|\log\pth{\frac{5}{4}|H_r|}
+\end{align*}
 \end{proof}

### 2014-10-03 paint-contour-tree.tex
Section 5 updated

@@
 We introduce the concept of path decompositions. A \emph{leaf path} is simply a contiguous path in $T$
-containing a leaf.
+containing a leaf, which is monotonic in the height values of its vertices.
 % Then in the following two subsections 
@@
 \begin{proof} For convenience, denote $H_i = H_{q_i}$ and $H_0 = H_{\ell(p)}$.
-Consider $v \in \bigcup_{i} H_i$ that lies below
-$b(p)$ in $T$. Note that $v \in H_p$ and has only one color in $\bigcup_{i} L(T_(b(q_i))) \cup \ell(p)$.
-Since the latter sets are all disjoint, any $u \in H_i \cap H_j$ must lie in $p$.
-Any $u \in p$ can be in at most two of the $H_i$'s. Hence, $\sum_i |H_i| \leq |H_p| + 2|p|$.
+Consider $v \in \bigcup_{i} H_i$ that lies below $b(p)$ in $T$. 
+Note that such a vertex has only one of its two colors in $L(b(p))$.  
+Since the colors tracked by $H_i$ and $H_j$ for $i\neq j$ are disjoint, 
+such a vertex can appear in only one of the $H_i$'s.
+On the other hand, a vertex $u\in p$ can appear in more than one $H_i$, 
+but since any vertex has exactly two colors it can appear in at most two 
+such heaps.  Hence, $\sum_i |H_i| \leq |H_p| + 2|p|$.
 \ignore{
@@
 
-There are some light paths that we can also ``pay" for.
+There are some light paths that we can also ``pay" for.  Consider any light path $p$ in the shrub.  We will refer to the 
+\emph{heavy support chain} of $p$ as the heavy ancestors of $p$ in the shrub which have a path to $p$ which does not use any 
+light path (i.e. it is a chain of paths adjacent to $p$).
 
 \begin{definition} \label{def:support} A light path $p$ is \emph{supported} if
-at least $|H_p|/100$ vertices $v$ in $H_p$ lie in heavy ancestors.
+at least $|H_p|/100$ vertices $v$ in $H_p$ lie in paths in the heavy support chain of $p$.
 %
@@
 Insert edge from $p$ to $q$, so $q$ is the parent of $p$ in $\cU$. This construction leads
-to the shrub forest of $\cL \setminus \cL'$, where every unsupported light path is an internal node, and all
-roots are supported light paths.
+to the shrub forest of $\cL \setminus \cL'$, where all the roots are supported light paths, 
+and the remaining nodes are the unsupported light paths.
 
@@
 The second term can be bounded by $O(\sum_{p \in \cL'} |H_p| \log |H_p|)$, by \Lem{pathBounds} again.
-The following claim shows that in turn is at most the last term.
+The following claim shows that this in turn is at most the last term.
 
@@
 %
-As we traverse the heavy ancestors, there
+As we traverse the heavy support chain of $p$, there
 are at least $|H_p|/100$ vertices of $H_p$ that lie in these paths. These are encountered in
@@
 
-Pick another $\widetilde{p} \in \cL'$. If $p$ and $\widetilde{p}$ are ancestor-descendant,
-then their corresponding supporting ancestor chains are disjoint. Any $v \in H'_p$
-can only occur in one other $H'_{\hat{p}}$ (for $\hat{p} \in \cL'$), because $v$ has exactly two colors.
+Consider any $v \in H'_p$.  Let $S$ be the set of paths $\widetilde{p}\in \cL'$ such that 
+$v\in H'_{\widetilde{p}}$.  We now show $|S|\leq 2$ (i.e. it contains at most one path other than $p$).
+First observe that any two paths in $S$ must be unrelated (i.e. $S$ is an anti-chain), 
+since paths which have an ancestor-descendant relationship have disjoint heavy support chains.  
+However, any vertex $v$ receives exactly one color from each of its two subtrees (in $T$), and therefore $|S|\leq 2$ since any two 
+paths which share descendant leaves in $T$ (i.e. their heaps are tracking the same color) must have an ancestor-descendant relationship.
+
 In other words, any $\log |H_v|$ appears at most twice in the above triple summation.
@@
 \begin{proof} Let $\cU$ denote a connected component of the shrub forest.
-We have $\sum_{p \in \cL \setminus \cL'} \leq \sum_{\cU} \sum_{p \in \cU} \sum_{v \in p} \log |H_v|$.
+We have $\sum_{p \in \cL \setminus \cL'} \sum_{v \in p} \log |H_v| \leq \sum_{\cU} \sum_{p \in \cU} \sum_{v \in p} \log |H_v|$.
 By \Lem{cu-root}, $|H_v| = O(|H_r|)$, where $r$ is the root of $\cU$.
@@
 \Lem{cu-root} asserts the root $r$ in $\cU$ pretty much encompasses all sizes and heaps in $\cU$.
-We will work with the \emph{reduced} heap $\redH_p$. This is the subset of $H_p$
-of vertices that only appear in light paths. By definition, for any unsupported light path
-(hence, any non-root $p \in \cU$), $|H'_p| \geq 99|H_p|/100$. For notational convenience, we simply
-set $H'_r = H_r$.
+We will work with the \emph{reduced} heap $\redH_p$. This is the subset of vertices of $H_p$ that 
+do not appear on the heavy support chain of $p$.
+%only appear in light paths. 
+By definition, for any unsupported light path
+(hence, any non-root $p \in \cU$), $|\redH_p| \geq 99|H_p|/100$. 
+% For notational convenience, we simply set $\redH_r = H_r$.
 %Consider $q \in \cL \setminus \cL'$ that is a child of $p$ in the shrub $\cU$.
@@
 \label{lem:geometric}
- Let $p$ be the parent of $q$ in the shrub $\cU$. Then $|\redH_p|\geq \frac{3}{2}|\redH_q|$. 
+ Let $q$ be the child of some path $p$ in $\cU$, then $|H_p|\geq \frac{3}{2}|H_q|$. 
+ Moreover, if $p\neq r(\cU)$, then $|\redH_p|\geq \frac{3}{2}|\redH_q|$. 
 \end{lemma}
-\begin{proof} By construction, there is a heavy path (denoted $h(q)$) that is a child of $p$ in $\pmax(T)$,
-and an ancestor of $q$. The chain of ancestors from $q$ to $h(q)$ consists only of heavy paths.
-Since $q$ is unsupported, these paths contain at most $|H_p|/100$ vertices of $H_p$. Thus, $|H_{h(q)}| \geq 99|H_p|/100$.
+\begin{proof} Let $h(q)$ denote the heavy path that is a child of $p$ in $\pmax(T)$,
+and an ancestor of $q$. If no such heavy path exists, then by construction $p$ is the parent 
+of $q$ in $\pmax(T)$, and the following argument will go through by setting $h(q)=q$.
+
+The chain of ancestors from $q$ to $h(q)$ consists only of heavy paths.
+Since $q$ is unsupported, these paths contain at most $|H_q|/100$ vertices of $H_q$. 
+Thus, $|H_{h(q)}| \geq 99|H_q|/100$.
 
@@
 Now, the decomposition $\pmax(T)$ put $u$ and $v$ in the same path $p$. This
-implies $|H_v| \geq |H_w|$. Since $|H_u| \geq |H_v| + |H_w| - 1$,
-$|H_u| \geq 2|H_w| - 1$. Let $b$ be the base of $p$.
+implies $|H_v| \geq |H_w|$. Since $|H_u| \geq |H_v| + |H_w| - 2$,
+$|H_u| \geq 2|H_w| - 2$. Let $b$ be the base of $p$.
 We have $|H_p| = |H_b| \geq |H_u| - |p| \geq 2|H_w| - |p| - 2$. Since $p$ is a light path, $|p| < \sqrt{|H_p|}/100$.
-Applying this bound, we get $|H_p| \geq (2-\delta)|H_w|$ (for small constant $\delta > 0$).
-Since $w$ is the base of $h(q)$, $H_w = H_{h(q)}$. We apply the bound $|H_{h(q)}| > 99|H_q|/100$
-to get $|H_p| \geq 197|H_q|/100$. The proof is completed by noting that $|\redH_p| \geq 99|H_p|/100$ and $|H_q| \geq |\redH_q|$.
+Applying this bound, we get $|H_p| \geq (2-\delta)|H_w|$ (for a small constant $\delta > 0$).
+Since $w$ is the base of $h(q)$, $H_w = H_{h(q)}$. We apply the bound $|H_{h(q)}| \geq 99|H_q|/100$
+to get $|H_p| \geq 197|H_q|/100$, implying the first part of the lemma.  For the second part, observe that 
+if $p\neq r(\cU)$, then $p$ is unsupported and so $|\redH_p| \geq 99|H_p|/100$, and therefore the second part follows since $|H_q| \geq |\redH_q|$.
 %
@@
 This immediately proves part (i) of \Lem{cu-root}. Part (ii) requires much more work.
-It is helpful to spell out how the various heaps evolve. Consider $p \in \cU$,
-which has children $q_1, q_2, \ldots, q_k$ in $\pmax(T)$.
-Furthermore, as a leaf path in $p$, some leaf
-$\ell$ of $T$ in one of the ends of $p$. The reduced heap $\redH_p$ is the
-union of $\redH_\ell$ and $\redH_{q_1}, \redH_{q_2}, \ldots, \redH_{q_k}$, with the vertices in $p$
-removed. Note that some of the $q_i$'s are the corresponding $h(q')$ vertices,
-for a child $q'$ in $\cU$.
+% It is helpful to spell out how the various heaps evolve. Consider $p \in \cU$,
+% which has children $q_1, q_2, \ldots, q_k$ in $\pmax(T)$, and a leaf vertex $\ell$ in $T$. 
+% The reduced heap $\redH_p$ is the
+% union of $\redH_\ell$ and $\redH_{q_1}, \redH_{q_2}, \ldots, \redH_{q_k}$, with the vertices in $p$
+% removed. Note that some of the $q_i$'s are the corresponding $h(q')$ vertices,
+% for a child $q'$ in $\cU$.
 
@@
 
-\begin{claim} \label{clm:loss} $|\redH_p| \geq W_p - 2\sum_{q \in \cU_p} |q|$
+\begin{claim} \label{clm:loss} For any path $p\in \cU$, $|\redH_p| \geq W_p - 2\sum_{q \in \cU_p} |q|$.
 \end{claim}
@@
 Traverse $\pmax(T)$ from $q'_i$ to $q$.
-Other than $q$, all other nodes encountered are heavy paths and hence
-do not affect reduced heaps.
+Other than $q$, all other nodes encountered are in the heavy support chain of $q'_i$ and hence
+do not affect its reduced heap.
 The vertices of $\redH_{q'_i}$ that are deleted
-are exactly those present in the path $q$ (which are paths in $T$) traversed. 
-Any vertex in $q$ can be deleted from at most two reduced heaps $\redH_{q'_i}$.
-When $R_q$ is negative, it is at most $2|q|$.
+are exactly those present in the path $q$. 
+Any vertex in $q$ can be deleted from at most two of the reduced heaps (of the children of $q$ in $\cU_p$), since 
+theses reduced heaps do not have an ancestor-descendant relationship.
+Therefore when $R_q$ is negative, it is at most by $2|q|$.
 We sum over all $q$ to complete the proof.
@@
 
-\begin{claim} \label{clm:charge} Fix $p \in \cU$. Suppose for any $q,q' \in \cU_p$
+\begin{claim} \label{clm:charge} Fix any path $p \in \cU\setminus\{r(\cU)\}$. Suppose for any $q,q' \in \cU_p$
 where $q$ is a parent of $q'$ in $\cU_p$, $W_q \geq (4/3)W_{q'}$. Then $\sum_{q \in \cU_p} |q| \leq W_p/20$.
@@
 We prove that $\sum_{q \in \cU_p} \sqrt{|W_q|}/99 \leq W_p/20$
-by a charge redistribution scheme. Assume that each $q \in \cU_p$
-starts with $\sqrt{W_p}/99$ units of charge. We redistribute this charge over all
-nodes in $\cU_p$, and then calculate the total charge. For $q \in \cU_p$, spread
+by a charge redistribution scheme. Assume that each $q \in \cU_p^-$
+starts with $\sqrt{W_q}/99$ units of charge. We redistribute this charge over all
+nodes in $\cU_q$, and then calculate the total charge. For $q \in \cU_p$, spread
 its charge to all nodes in $\cU_q$ proportional to $R^+$ values. In other words,
@@
 
-After the redistribution, let us compute the charge deposited at $q$. Every ancestor in $\cU_p$ 
+After the redistribution, let us compute the charge deposited at $q$. Every ancestor in $\cU_p^-$ 
 $q = a_0, a_1, a_2, \ldots, a_k$ contributes to the charge at $q$. The charge is expressed in the following
-equation. We use the assumption that $W_{a_i} \geq (4/3) W_{a_{i-1}}$ and hence $W_{a_i} \geq (4/3)^i$.
+equation. We use the assumption that $W_{a_i} \geq (4/3) W_{a_{i-1}}$ and hence $W_{a_i} \geq (4/3)^i W_{a_0} \geq (4/3)^i$, 
+as $a_0$ is an unsupported light path and hence $W_{a_0}\geq 1$. 
 $$ ({R^+_q}/99)\sum_{a_i} 1/\sqrt{W_{a_i}} \leq ({R^+_q}/99)\sum_{a_i} (3/4)^{i/2} \leq R^+_q/20 $$
@@
 
+\begin{corollary} \label{cor:chargeCor}
+ Let $r$ be the root of $\cU$, and suppose that for any paths $q,q' \in \cU\setminus \{r\}$, where $q$ is a parent of $q'$ in $\cU$, $W_q \geq (4/3)W_{q'}$.
+ Then $\sum_{p \in \cU} |p| \leq W_r/20+|r|$.
+\end{corollary}
+\begin{proof}
+ Let $c_1, \dots, c_m$ be the children of $r$ in $\cU$.  By definition, $W_r = \sum_i W_{c_i}+R^+_r \geq \sum_i W_{c_i}$.
+ By \Clm{charge}, for each $c_i$ we have $W_{c_i}/20\geq \sum_{p \in \cU_{c_i}} |p|$.  Combining these to facts yields the claim.
+\end{proof}
+
+% \begin{claim} \label{clm:charge} Fix $p \in \cU$. Suppose for any $q,q' \in \cU_p$
+% where $q$ is a parent of $q'$ in $\cU_p$, $W_q \geq (4/3)W_{q'}$. Then $\sum_{q \in \cU_p} |q| \leq W_p/20$.
+% \end{claim}
+% 
+% \begin{proof} Since $q$ is an unsupported light path, $|q| < \sqrt{|H_q|}/100 \leq \sqrt{|\redH_q|}/99 \leq \sqrt{|W_q|}/99$.
+% We prove that $\sum_{q \in \cU_p} \sqrt{|W_q|}/99 \leq W_p/20$
+% by a charge redistribution scheme. Assume that each $q \in \cU_p$
+% starts with $\sqrt{W_p}/99$ units of charge. We redistribute this charge over all
+% nodes in $\cU_p$, and then calculate the total charge. For $q \in \cU_p$, spread
+% its charge to all nodes in $\cU_q$ proportional to $R^+$ values. In other words,
+% give $(\sqrt{W_q}/99)\cdot(R^+_{q'}/W_q)$ units of charge to each $q' \in \cU_q$.
+% 
+% After the redistribution, let us compute the charge deposited at $q$. Every ancestor in $\cU_p$ 
+% $q = a_0, a_1, a_2, \ldots, a_k$ contributes to the charge at $q$. The charge is expressed in the following
+% equation. We use the assumption that $W_{a_i} \geq (4/3) W_{a_{i-1}}$ and hence $W_{a_i} \geq (4/3)^i W_{a_0} \geq (4/3)^i$, 
+% as $a_0$ is an unsupported light path and hence $W_{a_0}\geq 1$. 
+% $$ ({R^+_q}/99)\sum_{a_i} 1/\sqrt{W_{a_i}} \leq ({R^+_q}/99)\sum_{a_i} (3/4)^{i/2} \leq R^+_q/20 $$
+% The total charge is $\sum_{q \in \cU_p} R^+_p/20 = W_p/20$. 
+% \end{proof}
+
 We wrap it all up by proving part (ii) of \Lem{cu-root}.
@@
 
-\begin{proof} We use $r$ for $r(\cU)$. Our aim is to simply apply \Clm{charge} from $r(\cU)$. 
-Suppose $W_q \geq (4/3)W_{q'}$ (for any choice in $\cU$ of $q$ parent of $q'$). By \Clm{charge}, $\sum_{p \in \cU} |p| \leq W_{r}/20$.
-By \Clm{loss}, $|\redH_r| \geq W_r - 2\sum_{p \in \cU} |p|$. By simple algebra, we deduce, $\sum_{p \in \cU} |p| \leq |\redH_r|/10 = |H_r|/10$.
-
-We now prove that for any $q$ parent of $q'$, $W_q \geq (4/3)W_{q'}$.
+% \begin{proof} We use $r$ for $r(\cU)$. Our aim is to simply apply \Clm{charge} from $r(\cU)$. 
+% Suppose $W_q \geq (4/3)W_{q'}$ (for any choice in $\cU$ of $q$ parent of $q'$). By \Clm{charge}, $\sum_{p \in \cU} |p| \leq W_{r}/20$.
+% By \Clm{loss}, $|\redH_r| \geq W_r - 2\sum_{p \in \cU} |p|$. By simple algebra, we deduce, $\sum_{p \in \cU} |p| \leq |\redH_r|/10 = |H_r|/10$.
+% 
+% We now prove that for any $q$ parent of $q'$, $W_q \geq (4/3)W_{q'}$.
+% Suppose not. Let $p, p'$ be the counterexample furthest from the root,
+% where $p$ is the parent of $p'$. 
+% Note that for $q$ and child $q'$ in $\cU_{p'}$, $W_q \geq (4/3)W_{q'}$.
+% We will apply \Clm{charge} for $\cU_{p'}$ to deduce that $\sum_{q \in \cU_{p'}} |q| \leq W_{p'}/20$.
+% By \Clm{loss}, $|\redH_{p'}| \geq 19W_{p'}/20$. By \Lem{geometric}, $|\redH_p| \geq (3/2)|\redH_{p'}|$.
+% Noting that $W_p \geq |\redH_p|$, we deduce that $W_p \geq (4/3)W_{p'}$.
+% Hence, $p, p'$ is not a counterexample, and more generally, this is no counterexample.
+% That completes the whole proof.
+% \end{proof}
+
+\begin{proof} We use $r$ for $r(\cU)$. %Our aim is to simply apply \Clm{charge} from $r(\cU)$. 
+Suppose $W_q \geq (4/3)W_{q'}$ (for any choice in $\cU\setminus\{r\}$ of $q$ parent of $q'$), 
+then by \Cor{chargeCor}, $\sum_{p \in \cU} |p| \leq W_{r}/20+|r|$.
+By \Clm{loss}, $|\redH_r| \geq W_r - 2\sum_{p \in \cU} |p|$, and so combining these inequalities gives,
+\[
+% \sum_{p \in \cU} |p| \leq \pth{|\redH_r|+2\sum_{p \in \cU} |p|}/20+|r| \Rightarrow 
+ \sum_{p \in \cU} |p|\leq \frac{10}{9} \pth{|\redH_r|/20+|r|} \leq 
+ \frac{10}{9}\pth{|H_r|/20+\sqrt{|H_r|}/100} \leq |H_r|/10.
+\]
+
+We now prove that for any $q$ parent of $q'$ (other than $r$), $W_q \geq (4/3)W_{q'}$.
 Suppose not. Let $p, p'$ be the counterexample furthest from the root,
@@
 We will apply \Clm{charge} for $\cU_{p'}$ to deduce that $\sum_{q \in \cU_{p'}} |q| \leq W_{p'}/20$.
-By \Clm{loss}, $|\redH_{p'}| \geq 19W_{p'}/20$. By \Lem{geometric}, $|\redH_p| \geq (3/2)|\redH_{p'}|$.
+Combining this with \Clm{loss} gives, $|\redH_{p'}| \geq 19W_{p'}/20$. By \Lem{geometric}, $|\redH_p| \geq (3/2)|\redH_{p'}|$.
 Noting that $W_p \geq |\redH_p|$, we deduce that $W_p \geq (4/3)W_{p'}$.
-Hence, $p, p'$ is not a counterexample, and more generally, this is no counterexample.
+Hence, $p, p'$ is not a counterexample, and more generally, there is no counterexample.
 That completes the whole proof.
@@
 
-\begin{claim} \label{clm:loss} Consider $p \in \cL \setminus \cL'$ that is in $\cU$. Let $Q$
+\begin{claim} \label{clm:loss2} Consider $p \in \cL \setminus \cL'$ that is in $\cU$. Let $Q$
 be the set of heavy ancestors of $p$ that are descendants of $r(\cU)$. Then $\sum_{q \in Q} |q \cap H_p| \leq |H_p|/50$.

### 2014-10-15 paint-contour-tree.tex
Section 6 updated

@@
 
-Any tree $T$ considered in this section will be a rooted full binary tree where the height of a vertex is its distance 
+Any tree $T$ considered in this section will be a rooted full binary tree\footnote{Note 
+that technically the trees considered should have a leaf vertex hanging below the root of this 
+full binary tree in order to represent the global minimum of the surface.  This vertex is 
+(safely) ignored to simplify the presentation.} where the height of a vertex is its distance 
 from the root $r$.  As such, the children of a vertex $v\in T$ are the adjacent vertices of larger height
@@
 \section{Lower Bound by Path Decomposition}
-
+In this section we provide a lower bound for the time it takes to compute a contour tree in terms the cost of path decompositions.  Specially, 
+we first consider the easier case of merge trees, after which the lower bound for more general contour trees will easily follow.
+
+\subsection{Merge Trees}
 In this section we show that any algorithm for computing the merge tree $T$ must pay at least $f_{path}(P(T))$ (see \Def{path}) for any path decomposition $P(T)$.
 
-We say two shrubs $S_1$ and $S_2$ (of merge trees of potentially different surfaces) are equivalent if there is an isomorphism, $f$,
-between the two shrubs such that for any node $v\in S_1$ corresponding to a path of size $|v|$, $f(v)$ also corresponds to a path of size $|v|$.
-For the purposes of the next lemma, we say two $2$-manifolds are different, if their corresponding contour trees are the different.
+%We say two shrubs $S_1$ and $S_2$ (of merge trees of potentially different surfaces) are equivalent if there is an isomorphism, $f$,
+%between the two shrubs such that for any node $v\in S_1$ corresponding to a path of size $|v|$, $f(v)$ also corresponds to a path of size $|v|$.
+%For the purposes of the next lemma, we say two $2$-manifolds are different, if their corresponding contour trees are the different.
+
+In the following note that two distinct PL surfaces can have the same graph structure defining their meshes, but different height values associated with their vertices.
 
@@
 \label{lem:pathDecomp}
-Let $T$ be a merge tree, $P = P(T)$ any path decomposition of $T$, and $\pathTree$ the corresponding shrub.  
-Then there are at least $\Pi_{p_i\in P} |p_i|!$ different 2-manifolds with an equivalent shrub.
+Let $T$ be a merge tree, and let $P = P(T)$ any path decomposition of $T$.  %, and $\pathTree$ the corresponding shrub.  
+Then there is a set of at least $\Pi_{p_i\in P} (|p_i|-1)!$ distinct PL $2$-manifolds, all of which have the same defining mesh.
 \end{lemma}
 \begin{proof}
- We describe a simple input mesh, with heights specified for all vertices except for the saddles.  The main idea is that 
+ We describe a simple mesh, with heights specified for all vertices except for the saddles.  The main idea is that 
  this mesh will be flexible enough to allow almost any assignment of heights to the saddles, thus defining a class of merge trees 
@@
 
- At this point we can construct the entire input mesh, by gluing together diamonds.  Specifically, for each path $p_i\in P$ we create a diamond of size $p_i$.
- Now we glue together diamonds of different paths in the same way the paths are connected in $\pathTree$.  Specially, for two paths $p,q\in P$ where 
+ At this point we can construct the entire input mesh, by gluing together diamonds.  
+ Specifically, for each path $p_i\in P$ we create a diamond of size $|p_i|-1$ (we subtract one because the tip of the diamond already corresponds to the leaf of the path).
+ Now we glue together diamonds of different paths in the same way the paths are connected in the shrub $\pathTree$.  Specially, for two paths $p,q\in P$ where 
  $p$ is the parent of $q$ in $\pathTree$, we glue the bottom triangular cap of the diamond for $q$ onto the triangular ``platform'' of some face of the diamond for $p$.
@@
  Now there is a valid instantiation of this described mesh for any relative ordering of the heights of the saddles on a given diamond.  In particular, 
- there are $|p|!$ possible orderings of the heights of the saddles on the diamond for $p$, and hence $\Pi_{p_i\in P} |p_i|!$ possible instantiations of 
- the mesh overall.  Also, each one of these instantiations will result in a different merge.  Specifically, a permutation of the heights on a given diamond 
+ there are $(|p|-1)!$ possible orderings of the heights of the saddles on the diamond for $p$, and hence $\Pi_{p_i\in P} (|p_i|-1)!$ possible instantiations of 
+ the mesh overall.  Also, each one of these instantiations will result in a different merge tree.  Specifically, a permutation of the heights on a given diamond 
  corresponds to a permutation of the vertices along a path in $P$, as by construction all saddles on a given diamond will appear in sorted order in the merge tree.
@@
 \begin{corollary}
- Let $A(S)$ be any algorithm to compute the merge tree of any input surface $S$ (whose contour tree is a merge tree).  
+ \label{cor:mergeLB}
+ Let $A(S)$ be any comparison based algorithm to compute the merge tree of any input surface $S$ (whose contour tree is a merge tree).  
  Then the running time of $A(S)$ is $\Omega(f_{path}(P))=\Omega(\sum_{p_i\in P} |p_i|\log|p_i|)$ where $P$ is any path decomposition of the 
@@
 \begin{proof}
- Given an input surface $S$ whose merge tree has a particular path decomposition $P$, by the above lemma, there are $\Pi_{p_i\in P} |p_i|!$ other surfaces 
- with distinct merge trees, all of which have a shrub equivalent to $\pathTree$.  Therefore, any algorithm which computes the merge tree must distinguish between 
- these possibilities, and therefore its corresponding decision tree must have $\Pi_{p_i\in P} |p_i|!$ leaves.  Hence by Stirling's approximation, 
- this decision tree must have depth $\Omega(\sum_{p_i\in P} |p_i|\log|p_i|)$.
-\end{proof}
+ Given an input surface $S$, by the above lemma, any path decomposition $P$ of its merge tree defines a set of $\Pi_{p_i\in P} |p_i|!$ surfaces 
+ with distinct merge trees, all of which have the same mesh.  Specifically, for each sorted ordering of the relative height values of the saddles 
+ on each diamond in the mesh, yields a different merge tree.
+ Therefore, any comparison based algorithm which computes the merge tree must distinguish between these possibilities, and therefore its corresponding decision tree must 
+ have $\Pi_{p_i\in P} |p_i|!$ leaves.  Hence by Stirling's approximation, this decision tree must have depth $\Omega(\sum_{p_i\in P} |p_i|\log|p_i|)$.
+\end{proof}
+
+\subsection{Contour Trees}
+We first generalize previous terms to the case of contour trees.
+In this section $T$ will denote an arbitrary contour tree.  Specifically, every internal vertex in $T$ has degree $3$ 
+and the height function defined over the vertex set is such that every internal vertex in either a split or merge 
+(i.e. at most $2$ neighbors lie above or below it).
+
+Again we define a path decomposition as a partition of the vertices into 
+disjoint leaf paths (i.e. each path is monotonic in height and contains a leaf). 
+However, for simplicity we now restrict our attention to path decompositions which 
+roughly speaking are consistent with the raining procedure described in \Sec{rain} 
+(more general decompositions can work, but it is not needed for our purposes).
+
+\begin{definition}
+\label{def:path2} A path decomposition, $P(T)$, is called \emph{rain consistent} if its paths can be obtained as follows.
+Perform an upward BFS from an arbitrary minima $v$ in $T$, and mark all vertices encountered.  
+Now recursively run a directional BFS from all vertices adjacent to the current marked set.  
+Specifically, for each BFS run, make it an upward BFS if it is at an odd height in the recursion tree and downward otherwise.
+
+This procedure partitions the vertex set into disjoint rooted subtrees of $T$, based on which BFS marked a vertex.  
+For each such subtree, now take any partition of the vertices into leaf paths.\footnote{Note that the subtree  
+of the initial vertex is rooted at is a minima.  For simplicity we require that the path this vertex belongs to 
+also contains a maxima.}
+\end{definition}
+
+The following is analogous to \Lem{pathDecomp}, and in particular uses it as a subroutine.
+
+\begin{lemma}
+\label{lem:pathDecomp2}
+Let $T$ be a contour tree, and let $P = P(T)$ be any rain consistent path decomposition of $T$.  
+Then there is a set of at least $\Pi_{p_i\in P} (|p_i|-1)!$ distinct PL $2$-manifolds, all of which have the same defining mesh.
+\end{lemma}
+\begin{proof}
+ As $P$ is rain consistent, the paths can be partitioned into sets $P_1, \dots, P_m$, where $P_i$ is the set of all paths with 
+ vertices from a given BFS, as described in \Def{path2}.  Specifically, let $T_i$ be the subtree of $T$ corresponding to $P_i$ and 
+ let $r_i$ be the root vertex of this subtree.  
+ Note that the $P_i$ sets naturally define a tree where $P_i$ is the parent of $P_j$ if $r_i$ is adjacent to a vertex in $P_j$.
+
+ As the set $P_i$ is a path decomposition of a rooted binary tree $T_i$, the surface mesh construction of \Lem{pathDecomp} for $P_i$ is well defined.  
+ Actually the only difference is that here the rooted tree is not a full binary tree, and so some of the faces of the constructed diamonds will be blank.  
+ Specifically, these blank faces correspond to the adjacent children of $P_i$, and they tell us how to connect the meshes of the different $P_i$'s.
+
+\begin{figure}[h]\centering
+    \includegraphics[width=.5\linewidth]{figs/bridge2}
+    \caption{On the left, the root of $P_i$ merges into a diamond of its parent $P_j$.  On the right, the root of $P_i$ splits off from a diamond of $P_j$}
+    \label{fig:diamond}
+\end{figure}
+
+ So for each $P_i$ construct a surface mesh as described in \Def{path2}.  By construction, the diamonds corresponding to the paths in $P_i$ are connected 
+ into a tree structure (i.e. corresponding to the shrub of $P_i$).  Therefore the bottoms of all these diamonds are covered except for the one corresponding 
+ to the path containing the root $r_i$.  If $r_i$ is the initial minima that the rain consistent path decomposition was defined from,  
+ we add a tetrahedral cap to the uncovered bottom of this diamond (i.e. we construct the minima).  Otherwise, $P_i$ has some parent $P_j$ in which case 
+ we connect the bottom of the diamond for $r_i$ to a free face of a diamond in the construction for $P_j$, specifically, the face corresponding to the 
+ vertex in $T$ which $r_i$ is adjacent to. 
+ If $T_i$ was created by an upward BFS, then the connection to this face should be a merge vertex, and otherwise it should be a split vertex.  
+ See \Fig{diamond}.
+
+ Just as in \Lem{pathDecomp} we now have one fixed mesh graph structure, such that each different relative ordering of the heights of the merge and split vertices 
+ on each diamond produces a surface with a distinct merge tree.  The specific bound on the number of such distinct manifolds follows by applying the 
+ bound from \Lem{pathDecomp} to each $P_i$.
+\end{proof}
+
+For the same reason that \Cor{mergeLB} followed from \Lem{pathDecomp}, we now have the following.
+
+\begin{theorem}
+ Let $A(S)$ be any comparison based algorithm to compute the contour tree of an input surface $S$.  
+ Then the running time of $A(S)$ is $\Omega(f_{path}(P))=\Omega(\sum_{p_i\in P} |p_i|\log|p_i|)$ where $P$ 
+ is any rain consistent path decomposition of the contour tree of $S$.
+\end{theorem}
+
+% \section{Temp}
+% Given a contour tree we allow any path decomp, as long as each path contains a maxima or minima (or both).
+% We call a path $p$ in the decomp a min (resp. max) path if it contains a minima (resp. maxima) (note a path can be both).
+% For every path we create a cylinder with $m$ faces for $p$, where $m$ is the number of non-leaf vertices in $p$ (either $|p|-1$ or $|p|-2$).  
+% If $p$ has a maxima we put a top cone cap on the cylinder, and if it contains a manima then put a bottom cone cap (i.e. the cap represents the maxima or minima).
+% 
+% Every edge $e$ in $T$, but not in a path decomp. tells us how to glue together cylinders.  Let $u$ be the upper endpoint of $e$ and $l$ the lower endpoint.
+% 
+% First we ``remove'' all edges $e$ such that $u$ is adjacent to a max path and $l$ adjacent to a min path by joining the free ends of the respective cylinder together.
+% Hence we can assume no such edges exist.  Now we have three cases:
+% 
+% \begin{enumerate}[{Case} 1]
+%  \item 
+% $l$ is adjacent to a min path:\newline
+% Note in this case the path containing $u$ must be a min path.
+% In this case connect the bottom of the cylinder for the path containing $l$ into a face of the cylinder for the path containing $u$.
+% \item
+% $l$ is adjacent to a max path and $u$ is adjacent to a max path:\newline
+% In this case connect the bottom of the cylinder for the path containing $u$ into a face of the cylinder for the path containing $l$.
+% \item
+% $l$ is adjacent to a max path and $u$ is adjacent to a min path:\newline
+% Connect a face of the cylinder for the $u$ path (using a split) to a face of the cylinder for the $l$ path (using a merge).
+% \end{enumerate}
 

### 2014-10-20 paint-contour-tree.tex
updates to surgery theorem

@@
 \newcommand{\rain}{{\tt rain}}
-\newcommand{\reeb}{R}
+\newcommand{\reeb}{\cC}
 \newcommand{\redH}{\widetilde{H}}
@@
 optimal in worst case).
+
+\section{Contour tree basics} \label{sec:basics}
+
+We detail the basic definitions about contour trees. Most of this is taken from Chapter 6 of Carr's thesis~\cite{Carr}.
+Our input is a continuous PL function $f:\MM \mapsto \RR$, where $\MM$ is a simplicial complex in $\RR^d$. 
+All simplices are isomorphic to a standard $d$-dimensional simplex, except for specially designated
+\emph{boundary} simplices. \Sesh{Some mention of representation and data structures here.}
+It is standard to assume
+that $f$ is Morse (we explain later what this entails).
+
+Observe that $f:\MM \rightarrow \RR$ can be thought of as a $d$-dimensional simplicial complex living in $\RR^{d+1}$.
+We can think of $f(x)$ as the ``height" of a point $x \in \MM$.
+We will often omit $f$, and assume that $f$ is encoded in the representation of $\MM$.
+
+\begin{definition} \label{def:level} The level set at value $h$ is the set $\{x \in \MM| f(x) = h\}$.
+A contour is a maximal connected component of a level set. An $h$-contour is a contour where $f$-values are $h$.
+\end{definition}
+
+Not that a contour is itself a simplicial complex of one dimension lower, and is represented (in our algorithms) as such.
+We let $\delta, \eps$ denote infinitesimals. Let $B_\eps(x)$ denote a ball of radius $\eps$ around $x$.
+
+\begin{definition} \label{def:deg} The up-degree of $x \in \MM$ is the number of $(f(x) + \delta)$-contours that intersect $B_\eps(x)$
+as $\delta, \eps \rightarrow 0^+$. The down-degree is the number of $(f(x) - \delta)$-contours that intersect $B_\eps(x)$
+as $\delta, \eps \rightarrow 0^+$.
+\end{definition}
+
+The definition allows us to categorize points in $\MM$ depending on the local changes in topology.
+
+\begin{definition} \label{def:points} A point $x \in \MM$ is categorized as follows.
+\begin{asparaitem}
+	\item Regular: Both up-degree and down-degrees are $1$.
+	\item Maximum: Up-degree is $0$.
+	\item Minimum: Down-degree is $0$.
+	\item Join: Up-degree is strictly greater than $1$.
+	\item Split: Down-degree is strictly greater than $1$.
+\end{asparaitem}
+Non-regular points are called \emph{critical}. Joins and splits are called \emph{saddles}.
+Maxima and minima are called \emph{extrema}.
+\end{definition}
+
+We will enforce a special constraint on boundaries, encapsulated in the following definition.
+
+\begin{definition} \label{def:bound} $f$ is \emph{boundary critical} if the following holds.
+Consider a boundary simplex $F$. All vertices of $F$ have the same function value. Furthermore, all
+neighbors of $F$ not in $F$ either have function values strictly greater or strictly less
+than this value.
+\end{definition}
+
+We use $\cV$ to denote the set of critical points in $\MM$.
+Because $f$ is piecewise-linear, all critical points are vertices in $\MM$. 
+A value $h$ is called critical, if $f(x) = h$, for some $x \in \cV$. The Morse assumption
+implies that all critical points have distinct values. A contour that involves a critical
+point is also denoted as critical. Note that the vertices in a boundary face collectively behaves like an extremum.
+Abusing notation, the term ``critical point", ``maxima", and ``minima" will include boundaries.
+Critical values include these boundary values. By boundary criticality, all contours are themselves
+represented as $d-1$-dimensional simplicial complices.
+
+\begin{definition} \label{def:equiv} Two contours $\psi$ and $\psi'$ are equivalent 
+if the following holds. There exists an $f$-monotone path $p$ connecting a point in $\psi$ to $\psi'$,
+such that no $x \in p$ belongs to a critical contour.
+\end{definition}
+
+These give the \emph{contour classes}. By continuity, a regular point is present on a single contour,
+and we can partition the regular contours into these classes. Every such class maps to intervals
+of the form $(f(x_i),f(x_j))$, where $x_i, x_j$ are critical points. Such a class is said
+to be created at $x_i$ and destroyed at $x_j$.
+
+\begin{definition} \label{def:tree} The contour tree is the graph on vertex set $\cV$, where
+edges are formed as follows. For every contour class that is created at $v_i$ and destroyed $v_j$,
+there is an edge $(v_i,v_j)$. (Conventially, edges are directed from higher function value to lower.)
+\end{definition}
+
+(It is not immediately obvious that this graph is a tree. But it is equivalent to the Reeb graph
+of $f:\MM \rightarrow \RR$, which is known to be a tree.) We denote the contour tree of $\MM$
+by $\reeb(\MM,f)$. The corresponding node and edge sets are denoted as $\cV(\cdot)$ and $\cE(\cdot)$.
+
+We use a key theorem from~\cite{Carr} (Theorem 6.6) that maps paths in $\cC(\MM)$ and $\MM$.
+
+\begin{theorem} \label{thm:carr-path} For every path $P$ in $\MM$, there exists a path $Q$ in the contour tree to which $P$
+maps, and for every path $Q$ in the contour tree, there exists at least one path $P$ in $\MM$ that maps to $Q$.
+\end{theorem}
+
+\section{Divide and conquer through contour surgery} \label{sec:surgery}
+
+{\bf The cutting operation:} We define a ``cut" operation on $f:\MM \rightarrow \RR$ that cuts along a regular contour to create
+a new simplicial complex with an added boundary. Given contour $\phi$, this is constructing the simplicial complex $\MM \setminus \phi$. 
+We will always enforce the condition that $\phi$ never passes through a vertex of $\MM$.
+Again, we use $\eps$ for an infinitesimally small value. We denote $\phi^+$ (resp. $\phi^-$) to be
+the contour at value $f(\phi^+) + \eps$ (resp. $f(\phi^-) - \eps$), which is at distance $\eps$
+from $\phi$.
+
+An $h$-contour is achieved by intersecting $\MM$ with the hyperplane $x_{d+1} = h$. (Think of the $d+1$-dimension
+as height.) Given some point $x$
+on an $h$-contour $\phi$ (with some simplex that $x$ is present in), we can walk along $\MM$ from $x$ and determine $\phi$.
+We can ``cut" along $\phi$ to get a new (possibly) disconnected simplicial complex $\MM'$. This is achieved
+by splitting every facet $F$ that $\phi$ intersects into an ``upper" facet and ``lower" facet. Algorithmically,
+we cut facet $F$ with $\phi^+$ and take everything above $\phi^+$ in $F$ to make the upper facet. Analogously, we cut with $\phi^-$ to get the lower facet.
+The facets are then triangulated to ensure that they are all simplices.
+Given that $\phi$ cannot cut a boundary and all non-boundary facets
+have constant size, we omit the algorithmic description for this process. We only note that this
+creates the two new boundaries $\phi^+$ and $\phi^-$, and we maintain the property of constant $f$-value at a boundary.
+This new simplicial complex is denoted by $\cut(\phi,\MM)$.
+This can be constructed in time linear in $|\phi|$.
+
+%We denote this new manifold by $\cut(\phi,\MM)$, and the new boundaries by $\phi^+$ and $\phi^-$.
+%Abusing notation, we use $\cut(\phi,\MM)$
+%for a set $\Psi$ of contours to mean the manifold obtained by cutting along all contours in $\Phi$.
+
+\ignore{
+\medskip
+\fbox{
+\begin{minipage}{0.9\textwidth}
+{\bf $\cut(\phi,\MM)$}
+
+\smallskip
+\begin{asparaenum}
+%	\item Insert the vertices of $\phi$ in $\MM$.
+	\item For every edge $e$ that intersects $\phi$ at point $x$: add edges $(e,x^+)$ and $(e,x^-)$
+	where $x^+$ (resp. $x^-$) is infinitesimally above (resp. below) $x$.
+	\item Take a point $x$ (resp. $y$) on $e$ at distance $\eps$ above (resp. below) $\phi$.
+	\item Let $\phi_x$ be the contour through $x$, and analogously define $\phi_y$. Insert a vertex
+	at each intersection point of these contours with edges of $\MM$. Add  
+	\item Delete the edges between $\phi_x$ and $\phi_y$.
+\end{asparaenum}
+\end{minipage}}
+
+\medskip
+This results in a new boundary critical simplicial complex. There are two new disjoint
+boundary simplices added. The manifold may be disconnected. 
+}
+
+We now describe a high-level approach to construct $\reeb(\MM)$.
+
+\medskip
+\fbox{
+\begin{minipage}{0.9\textwidth}
+{\bf $\surgery(\MM,\phi)$}
+
+\smallskip
+\begin{asparaenum}
+	\item Let $\MM' = \cut(\phi,\MM)$.
+	\item Construct $\reeb(\MM')$ and let $A, B$ be the nodes corresponding to the new boundaries created
+	in $\MM'$. (One is a minima and the other is maxima.)
+	\item Since $A, B$ are leaves, they each have unique neighbors $A'$ and $B'$, respectively. Insert
+	edge $(A',B')$ and delete $A, B$ to obtain $\reeb(\MM)$.
+\end{asparaenum}
+\end{minipage}}
+
+\medskip
+\begin{theorem} \label{thm:surgery} For any regular contour $\phi$, the output of $\surgery(\MM,\phi)$ is $\reeb(\MM)$.
+\end{theorem}
+
+The main theorem is a direct consequence of the following lemma.
+
+\begin{lemma} \label{lem:cut} Consider a regular contour $\phi$ contain in contour class (edge of $\reeb(\MM))$
+$(u,v)$ and denote $\MM' = \cut(\phi,\MM)$. Then $\cV(\reeb(\MM')) = \{\phi^+,\phi^-\} \cup (\cV(\MM) \setminus \phi)$
+and $\cE(\reeb(\MM')) = \{(u,\phi^+),(\phi^-,v)\} \cup (\cE(\MM) \setminus (u,v))$.
+\end{lemma}
+
+\begin{proof} The contours of $\MM'$ are exactly the contours of $\MM$ with $\phi^+$ and $\phi^-$, without $\phi$.
+All new vertices created in $\MM'$ are present on $\phi^+$ and $\phi^-$. That proves the first part.
+
+Any contour class in $\MM'$ (edge in $\cC(\MM')$) that does not involve $\phi^+$ or $\phi^-$ is also
+a contour class in $\MM$. Furthermore, a maximal contour class satisfying these properties is also
+maximal in $\MM$. So all edges of $\cC(\MM')$ that do not involve $\phi^+$ or $\phi^-$ are edges of $\cC(\MM)$.
+Analogously, every edge of $\cC(\MM)$ not involving $\phi$ is an edge of $\cC(\MM)$.
+
+Consider the contour class corresponding to edge $(u,v)$ of $\cC(\MM)$. There is a natural ordering
+of the contours by function value, ranging from $f(u)$ to $f(v)$. All contours in this class ``above" $\phi$
+form a maximal contour class in $\MM'$, represented by edge $(u,\phi^+)$. Analogously, there is another
+contour class represented by edge $(\phi^-,v)$. We have now accounted for all contours in $\cC(\MM')$,
+completing the proof.
+\end{proof}
+
+A useful corollary of this lemma shows that a contour actually splits the simplicial complex into
+two disconnected complices.
+
+\begin{theorem} \label{thm:jordan} $\cut(\phi,\MM)$ consists of two
+disconnected simplicial complices.
+\end{theorem}
+
+\begin{proof} Denote (as in \Lem{cut}) the edge containing $\phi$ to be $(u,v)$. Suppose for contradiction that there is a path between vertices $u$ and $v$
+in $\MM' = \cut(\phi,\MM)$. By \Thm{carr-path}, there is a path in $\cC(\MM')$ between $u$ and $v$. Since $\phi^+$ and $\phi^-$
+are leaves in $\cC(\MM')$, this path obviously cannot use edges incident to them. By \Lem{cut},
+all the edges of this path are in $\cE(\cC(\MM)) \setminus (u,v)$. So we get a cycle in $\cC(\MM)$. Contradiction.
+To show that there are exactly two connected components in $\cut(\phi,\MM)$, it suffices
+to see that $\cC(\MM')$ has two connected components (by \Lem{cut}) and apply \Thm{carr-path}.
+\end{proof}
+
+\section{Raining to partition $\MM$} \label{sec:rain}
+
+In this section, we describe a linear time procedure that partitions $\MM$ into special
+\emph{extremum dominant} simplicial complices.
+
+\begin{definition} \label{def:dom} A simplicial complex is \emph{minimum dominant} if there exists
+a minimum $x$ such that every non-minimal \emph{vertex} in the manifold has decreasing path to $x$.
+Analogously define \emph{maximum dominant}. 
+\end{definition}
+
+The first aspect of the partitioning is ``raining". Start at some point $x \in \MM$ and imagine rain at $x$.
+The water will flow downwards along descending paths and ``wet" all the points encountered. Note that this procedure considers all points
+of the manifold, not just vertices.
+
+\begin{definition} \label{def:wet} Fix $\MM$ and $x \in \MM$. The set of points $y \in \MM$ such that there is a non-ascending path from $x$ to $y$
+is denoted by $\wet(x,\MM)$ (which is in turn represented as a simplicial complex). A point $z$ is at the \emph{interface} of $\wet(x,\MM)$ if every neighborhood of $z$
+has non-trivial intersection with $\wet(x,\MM)$.
+\end{definition}
+
+The following claim gives a description of the interface. While its meaning is quite intuitive, the proof is tedious.
+
+\begin{claim} \label{clm:inter} For any $x$, the interface of $\wet(x,\MM)$ is a set of contours,
+each containing a join vertex.
+\end{claim}
+
+\begin{proof} If $y \in \wet(x,\MM)$, all the points in any contour containing $y$ are also in $\wet(x,\MM)$.
+ (Follow the non-ascending path to from $x$ to $y$ and then walk along the contour.) The converse is also true,
+ so $\wet(x,\MM)$ contains entire contours.
+
+Let $\eps, \delta$ be sufficiently small as usual. Fix some $y$ at the interface.
+Note that $y \in \wet(x,\MM)$. (Otherwise, $B_\eps(y)$ is dry.)
+The points in $B_\eps(y)$ that lie below $y$ have a descending path from $y$ and hence must be wet.
+There must be a dry point in $B_\eps(y)$ that is above $y$, and hence,
+there exists a dry, regular $(f(y)+\delta)$-contour $\phi$ intersecting $B_\eps(y)$.
+
+Suppose $y$ is in the interface. Consider the set $\Gamma_y$ of points connected to $y$
+by a path with all points at height $f(y)$. (This is a set of $f(y)$-contours.) Suppose for contradiction
+that $\forall z \in \Gamma_y$, $z$ has up-degree $1$. Consider the non-ascending path from $x$ to $y$ and let $z$
+be the first point of $\Gamma_y$ encountered. There exists a wet, regular $(f(y) + \delta)$-contour $\psi$ 
+intersecting $B_\eps(z)$. Now, walk from $z$ to $y$ along $\Gamma_y$. If all points $w$ in this walk
+have up-degree $1$ (ref. to \Def{deg}), then $\psi$ is the unique $f(y)+\delta$-contour
+intersecting $B_\eps(w)$. This would imply that $\phi = \psi$, contradicting the fact that $\phi$ is wet
+and $\psi$ is dry.
+
+Therefore, there must exist a join vertex $w$ in $\Gamma_y$, such that $B_\eps(w)$ intersects
+both $\phi$ and $\psi$. As $\delta, \eps \rightarrow 0^+$, the limit of $\phi$ (which is dry)
+lies in $\Gamma_y$ (which is wet). Hence, this gives the interface, which is a contour containing
+the join $w$.
+\end{proof}
+
+\ignore{
+\begin{proof} If $y \in \wet(x,\MM)$, all the points in any contour containing $y$ are also in $\wet(x,\MM)$.
+ (Follow the non-ascending path to from $x$ to $y$ and then walk along the contour.)
+ First observe that any point which in the interface must be wet.  Second, for $y \in \wet(x,\MM)$, all the points in any contour containing $y$ are also in $\wet(x,\MM)$, 
+ as one can follow the non-ascending path to $y$ and then walk along the contour.
+
+ So let $y$ be a point in the interface.
+
+ However, since $y$ is in the interface, it is also wet.  Therefore for any contour $\phi$ which contains $y$, for any arbitrarily small $\eps$ there is 
+ a wet point which lies above $\phi$ and has a descending path of length at most $\eps$ to $\phi$.
+ This implies there is some contour which lies just 
+ above $y$ which is wet.  So any contour $\phi$ containing $y$ has two contours which lie just above it, one which is dry, and one which is wet.  
+ This can only happen if $\phi$ contains a merge vertex, because otherwise all contours which lie just above would be in the same L-homotopy equivalence class and so by 
+ \Clm{mono}, if one is wet, all below it are wet.
+
+ By similar logic one can argue that of the two contours through a given merge vertex at most one can contain interface points other than the merge vertex, 
+ and that if one of the contours does contain such interface points, then the entire contour consists of interface points.
+\end{proof}
+}
+
+Note that $\wet(x,\MM)$ (and its interface) can be computed in time linear in the size of the wet simplicial complex.
+We perform a non-ascending walk from $x$. Any facet $F$ of $\MM$ encountered has some (if not all) portion of in $\wet(x,\MM)$.
+This portion is determined by cutting $F$ along the interface. Since the interface is a contour, this is equivalent
+to cutting $F$ by a hyperplane. All these operations can be performed to output $\wet(x,\MM)$ in time linear in $|\wet(x,\MM)|$.
+
+% \begin{proof} If $y \in \wet(x,\MM)$, but not in the interface, then any contour containing $y$ is also in $\wet(x,\MM)$.
+% So consider a point $y$ in the interface. We argue that all points in a contour $\phi$ containing $y$
+% are also in the interface. If such a contour is regular, we can find a contour just above it that
+% is also wet. 
+% 
+% If two contours intersect, the intersection point $y$ is critical. We can argue that the only
+% way to reach $y$ from a descending path is reach one of the contours and then follow it to get to $y$.
+% But that would imply the existence of another critical point on this contour, violating the Morse condition.
+% 
+% Let us argue that $y$ is a merge vertex. Consider the descending path from $x$ to $y$. The final edge
+% that reaches $y$ is obviously all wet, and is an increasing edge from $y$. Consider an opposite increasing
+% edge. Take two point in these edges that are infinitesimally higher than $y$ and show that the contours
+% through them as disjoint. So $y$ is a merge.
+% \end{proof}
+
+We define a simple \lift{} operation on the interface contours. Consider such a contour $\phi$ containing
+a merge vertex $y$. Take any dry increasing edge incident to $y$, and pick the point $z$ on this edge at height
+$f(y) + \delta$. Let $\lift(\phi)$ be the unique contour through regular point $z$. Note that $\lift(\phi)$ is dry.
+
+\begin{claim} \label{clm:cut-int} Let $\phi$ be an interface contour. Then $\cut(\MM,\lift(\phi))$
+results in two disjoint simplicial complices, one consisting entirely of dry points.
+\end{claim}
+
+\begin{proof} By \Thm{jordan}, $\cut(\MM,\lift(\phi))$ results in two disjoint simplicial complices. Let $\NN$ be the complex containing
+the point $x$ (the argument in $\wet(x,\MM)$), and let $\NN'$ be the other complex. 
+Any path from $x$ to $\NN'$ must intersect $\lift(\phi)$, which is dry. Hence $\NN'$ is dry.
+%
+%
+%If there exists a wet point $z$ in $\NN'$, then there exists a descending path from $x$ to $z$. 
+%As $x$ is in $\NN$ and $z$ is in $\NN'$, again by the Jordan Curve Theorem this path must intersect $\phi$, 
+%contradicting the fact that $\phi$ is dry.
+\end{proof}
+
+We describe the main partitioning procedure that cuts a simplicial complex $\NN$ into extremum
+dominant complices. It takes an additional input of a maximum $x$. To initialize,
+we begin with $\NN$ set to $\MM$ and $x$ as an arbitrary maximum. One of the critical aspects
+of the procedure is that the rain alternately flows downwards and upwards. In other words, when
+we start, rain flows downwards. In each recursive call, the direction of rain is switched to the 
+opposite direction. While this is way of implementing $\rain$, it is conceptually easier
+to think of \emph{inverting} a complex $\NN'$ when a recursive call is made. Inversion
+is easily achieved by just negating the height values. (From a running time standpoint, it suffices
+to maintain a single bit associated with $\NN'$ that determines whether heights are inverted or not.)
+We can now let rain flow downwards, as it usually does in our world.
+
+%
+%We critically perform
+%an \emph{inversion} step, which is simply achieved by negating
+%
+\medskip
+\fbox{
+\begin{minipage}{0.9\textwidth}
+{\bf $\rain(x,\NN)$}
+
+\smallskip
+\begin{compactenum}
+	\item Determine interface of $\wet(x,\NN)$. 
+	\item If the interface is empty, simply output $\NN$. Otherwise, denote the contours by $\phi_1, \phi_2, \ldots, \phi_k$ and set $\phi'_i = \lift(\phi_i)$.
+	\item Initialize $\NN_1 = \NN$.
+	\item For $i$ from $1$ to $k$: 
+	\begin{compactenum}
+		\item Construct $\cut(\phi'_i,\NN_i)$, consisting of dry complex $\LL_i$ and remainder $\NN_{i+1}$.
+		\item Let the newly created boundary of $\LL_i$ be $B_i$. Invert $\LL_i$ so that $B_i$ is a maximum. Recursively
+		call $\rain(B_i,\LL_i)$.
+	\end{compactenum}
+	\item Output $\NN_{k+1}$ together with any complices output by recursive calls.
+\end{compactenum}
+\end{minipage}}
+
+\medskip
+For convenience, denote the total output of $\rain(x,\MM)$ by $\MM_1, \MM_2, \ldots, \MM_r$.
+We first argue the correctness of $\rain$.
+
+\begin{lemma} \label{lem:rain-1} Each output $\MM_i$ is extremum dominant.
+\end{lemma}
+
+\begin{proof} Consider a call to $\rain(x,\NN)$. If the interface is empty, then all of $\NN$
+is in $\wet(x,\NN)$, so $\NN$ is trivially extremum dominant. So suppose the interface
+is non-empty and consists of $\phi_1, \phi_2, \ldots, \phi_k$ (as denoted in the procedure).
+By repeated applications of \Clm{cut-int}, $\NN_{k+1}$ contains $\wet(x,\MM)$. 
+Consider $\wet(x,\NN_{k+1})$. The interface must exactly be $\phi_1, \phi_2, \ldots, \phi_k$.
+So the only dry vertices are those in the boundaries $B_1, B_2, \ldots, B_k$. But these
+boundaries are maxima.
+\end{proof}
+
+Focus on $\rain(x,\MM)$. As $\rain$ proceeds, new facets/simplices are created because of repeated cutting. Indeed, any facet $F$
+in any simplicial complex $\NN$ that is recursively invoked, is achieved by cutting facet $F'$ in $\MM$.
+The key to the running time of $\rain(x,\MM)$ is bounding the number of newly created facets, for which we have the following lemma.
+
+\begin{lemma}\label{lem:new-verts}
+A facet $F \in \MM$ is cut at most once during the $\rain(x,\MM)$.
+\end{lemma}
+\begin{proof} All notation here follows that in the pseudocode of $\rain$.
+Firstly, by \Thm{jordan}, all the pieces of which $\rain$ is invoked are disjoint.
+Secondly, all recursive calls are made on dry complices.
+
+Consider the first time that $F$ is cut, say, during the call to $\rain(x,\NN)$.
+Specially, say this happens when $\cut(\phi'_i,\NN_i)$ is constructed. 
+So $F$ is cut by some hyperplane into lower and upper portions (which are then triangulated).
+Since the cut is infinitesimally above an interface, the lower portion is basically wet.
+It can never be part of $\LL_1, \LL_2, \ldots$, which go into recursive calls. This portion
+is in $\NN_{k+1}$. The upper portion/facet (call it $U$) is in $\LL_i$. Note that the lower
+boundary of $U$ is in the boundary $B_i$. Since a recursive call is made to $\rain(B_i,\LL_i)$
+(and $\LL_i$ is inverted), $U$ becomes wet. Hence, $U$ will not be cut subsequently.
+All in all, $F$ is cut at most once.
+\ignore{
+As the number of edges is $O(n)$, arguing that there is at most one boundary vertex created along any given edge will imply the claim.
+Consider some edge $e$ and take the first time a boundary vertex is created in $e$.
+This happens when some contour $\phi$ intersects $e$. On applying $\cut(\NN,\lift(\phi))$,
+consider the dry manifold $\LL$ and wet manifold $\MM$ constructed. The edge $e$ is split into two parts, the dry and wet part.
+The wet part is in $\MM$ and will never be involved in any recursive call.
+
+The dry part of $e$ (call it $e'$) needs to be considered.
+A boundary vertex $v$ (an new endpoint of $e$) is created, which is part of the boundary $B$ in $\LL$,
+such that $\rain(B,\LL)$ is called to the inverted $\LL$. In this call, since the water starts from $v$,
+all of $e'$ becomes wet. So there is no further interface contour that can intersect this portion of $e'$.
+}
+\end{proof}
+
+\begin{theorem} \label{thm:rain-time} The total running time of $\rain(x,\MM)$ is $O(|\MM|)$.
+\end{theorem}
+
+\begin{proof} The operations performed are the raining and cutting. We can bound this by the total
+complexities of the wet complices and the new boundaries created. No facet is every wet twice,
+and by \Lem{new-verts}, the total number of facets that every appear is $O(|\MM|)$. This also 
+bounds the complexities of the boundaries created.
+\end{proof}
+
+\ignore{
+As each original vertex in $\NN$ appears in only one of the sub-manifolds output by $\rain(x,\NN)$, we have the following corollary.
+
+\begin{corollary}
+The total size of a sub-manifolds output by $\rain(x,\MM)$ is $O(n)$.
+\end{corollary}
+
+\begin{corollary} \label{cor:rain-time} The total running time of $\rain(x,\MM)$ is $O(n)$.
+\end{corollary}
+
+\begin{proof} Consider some invocation to $\rain(x,\MM)$. The only non-trivial work done in $\rain$ is by the calls to $\wet$.
+We know that each invocation to $\wet$ runs in time linear 
+to the portion of the manifold which is wet (including the new boundary vertices created).  
+As the output of $\rain$ is precisely the distinct sub-manifolds wet by each distinct call to $\wet$, the claim follows.
+\end{proof}
+
+%Consider one of the manifolds $\MM_i$ output by $\rain$.  This manifold has a set of boundary maxima/minima that are the result 
+%of cutting along interface contours.  For each such contour there is exactly one other output manifold $\MM_j$ for which 
+%the contour appears as a boundary maxima/minima. 
+}
+
+\begin{claim} \label{clm:rain-reeb} Given $\reeb(\MM_1), \reeb(\MM_2), \ldots, \reeb(\MM_r)$, 
+%be the output of $\rain(x,\MM)$. Suppose that for each reeb graph we know its corresponding vertex in the Manifold Ordering Tree. 
+$\reeb(\MM)$ can be constructed in $O(|\MM|)$ time.
+\end{claim}
+
+\begin{proof} Consider the tree of recursive calls in $\rain(x,\MM)$, with each node
+labeled with some $\MM_i$.
+%
+%There is a natural tree ordering defined over these manifolds where the 
+%root corresponds the first manifold that was wet when we rained from $x$, and its children correspond to the recursive calls 
+%(i.e. where we rained from the interface contours of the first raining).  Call this the \emph{Manifold Ordering Tree}.
+%$\rain$ can easily be modified to also output this tree (without affecting the asymptotic running time).  
+Walk through this tree in a leaf first ordering.  Each time we visit a node we connect its contour tree to the 
+contour tree of its children in the tree using the $\surgery$ procedure. 
+Each $\surgery$ call is constant time, and the total time is the size of the recursion tree.
+%
+%
+%This takes $O(n)$ time overall by \Lem{new-verts} and since $\surgery$ takes constant time (given the appropriate vertices in each reeb graph).
+The correctness follows by induction and  \Lem{surgery}.
+\end{proof}
+
+\section{Painting to compute contour trees} \label{sec:paint}
+
+The previous section allows us to restrict attention to extremum dominant manifolds.
+We will orient so that the extremum in question is always a \emph{minimum}.
+We will fix such a simplicial complex $\MM$, with the dominant minimum $u$.
+
+We begin by associating a color with each maximum. If the set of maxima is $S$,
+this is simply an arbitrary bijection $\chi:S \mapsto [|S|]$. Imagine there being a large
+can of paint of color $\chi(y)$ at maxima $y$. We will spill different paint from each maximum and watch it flow down.
+This is analogous to the raining in the previous section, but paint is a much more viscous liquid.
+\emph{So paint only flows down edges, and it does not color the interior of faces.} Furthermore, paints
+do not mix, so eventually, every vertex and edge of $\MM$ gets a unique color.
+%
+\begin{definition} \label{def:paint} A  \emph{painting} of $\MM$ is a map $\chi:S \cup E \mapsto [|S|]$
+with the following properties. (Remember that $G(\MM) = (V,E)$.)
+\begin{asparaitem}
+	\item The restriction $\chi:S \mapsto [|S|]$ is a bijection.
+	\item Consider an edge $e$. There exists a descending path from some maximum $x$ to $e$
+	consisting of edges in $E$, such that all edges along this path have the same color as $x$. (The maximum $x$
+	is well-defined to be the unique maximum with the same color as $e$.)
+\end{asparaitem}
+\end{definition}
+%
+Note that paintings are not unique. It is straightforward to construct a painting of $\MM$ in $O(n)$ time, using a descending BFS from each maximum that does not explore previously colored edges.
+
+\subsection{The data structures} \label{sec:struct}
+
+\begin{definition} \label{def:color-set} Fix a painting $\chi$, and a color $c$.
+A critical point $x$ is \emph{touched by $c$} if $x$ is incident to a $c$-colored
+increasing edge. The set of non-extremal critical points touched by $c$ is denoted $\touch(c)$.
+For $x$, $\col(x)$ is the set of colors that touch $x$.
+\end{definition}
+%
+%We will store each set $\touch(c)$ in a binomial heap, keyed by the heights. 
+Abusing notation, $\touch(c)$ refers
+both to the set and the data structure used to store it.
+
+\medskip
+\noindent
+{\bf The binomial heaps $\touch(c)$:} For each color $c$, we store a binomial
+heap $T(c)$ of non-extremal critical points, keyed by vertex heights.
+
+\medskip
+\noindent
+{\bf The union-find data structure on colors:} We will repeatedly perform unions
+of classes of colors, and this will be maintained as a standard union-find data structure.
+For any color $c$, $\rep(c)$ denotes the representative of its class. We initialize
+by simply setting $\rep(c) = c$ for all colors.
+
+\medskip
+\noindent
+{\bf The stack $\st$:} This consists of non-extremal critical points with the following structure.
+Each point $x \in \st$ has an associated subset of $\col(x)$, denoted $\mcol(x)$.
+Both $\mcol(x)$ and its complement are stored as hash table. So lookups, inserts, and deletes
+are in these sets are all constant time operations. The stack is guaranteed to satisfy 
+the following invariants.
+\begin{asparaitem}
+	\item For every $x \in \st$: For every $c \in \mcol(x)$, $x$ is the highest element
+	in $T(c)$. Furthermore, $c = \rep(c)$.
+	\item Consider $x, y \in \st$ such that $y$ was pushed on $x$. There exists $c \in \col(x) \setminus
+	\mcol(x)$ such that $x$ is not highest in $T(c)$ but $y$ is highest in $T(c)$.
+\end{asparaitem}
+
+\medskip
+\noindent
+{\bf Highest points $\h(c)$:} For each color $c$, we maintain a special critical point $\h(c)$ of this color.
+It is supposed to denote the ``highest" point of color $c$, but the notion of highest evolves
+with the algorithm.
+
+\subsection{Updating the stack} \label{sec:stack}
+
+We have a simple, but critical, procedure that updates the stack $\st$ by pushing
+points on to it.
+
+\medskip
+\fbox{
+\begin{minipage}{0.9\textwidth}
+{\bf $\update(\st)$}
+
+\smallskip
+\begin{compactenum}
+	\item If $K$ is empty, pick arbitrary $c$. Let $x$ be the highest point in $T(c)$.
+	Push $x$ onto stack and set $\mcol(x) = \{c\}$.
+	\item Let $h$ be head of $\st$.
+	\item While $\col(h) \setminus \mcol(h)$ non-empty:
+	\begin{compactenum}
+		\item Pick some $c \in \col(h) \setminus \mcol(h)$. Replace it by $c' = \rep(c)$.
+		\item Obtain highest point $x$ in $T(c')$. 
+		\item If $x = h$, put $c'$ in $\mcol(h)$. Otherwise ($x \neq h$), push $x$ onto stack
+		and add $c'$ to $\mcol(x)$.
+		\item Let $h$ be new head of $\st$.
+	\end{compactenum}
+\end{compactenum}
+\end{minipage}}
+
+\medskip
+When $\update(\st)$ terminates, the head $h$ has the following property. For every $c \in \col(h)$,
+$h$ is the highest point in $\touch(c)$.
+
+\begin{claim} \label{clm:cont} Consider an increasing path from $h$ to some maxima $x$.
+For any $y$ in this path, $\cont(y)$ only contains points of color $c(x)$.
+Furthermore, for all vertices on $\cont(y)$, their increasing edges are colored $c(x)$.
+\end{claim}
+
+\begin{proof} Consider this increasing path, and think of moving $y$ continuously downwards from $x$
+to $h$. Initially (when $y = x$), the lemma is trivially true. Consider the first $y$
+when the condition fails. We have $\cont(y)$ only containing points of color $c(x)$,
+but there is a vertex $z \in \cont(y)$ that has increasing edges of a different color $c'$.
+All the increasing edges of $z$ that are colored $c(x)$ are contiguous (since we can construct
+a contour intersecting all of them). Thus, there are two disjoint increasing neighborhoods for $z$,
+and $z$ is a saddle point. But that contradicts the fact that $h$ is the highest saddle point
+of color $c(x)$.
+
+So, as $y$ decreases, for all vertices on $\cont(y)$, their increasing edges are colored $c(x)$.
+This also means that $\cont(y)$ only contains points of color $c(x)$.
+\end{proof}
+
+We have important lemmas characterizing $|\col(h)|$.
+
+\begin{lemma} \label{lem:mono} If $h$ is a merge vertex, then $|\col(h)| > 1$.
+\end{lemma}
+
+\begin{proof} Suppose not. Consider the edges incident to $h$ in circular order. There exist (at least) $2$
+disjoint contiguous stretches of edges that are increasing. 
+All these edges have the same color $c$, associated with some maximum $x$. Consider two neighbors $y,z$
+that lie in different stretches. By the definition of a painting, there exist descending paths (along
+edges) from $x$ to $y$ and $x$ to $z$. We can construct a Jordan curve $C$ from $x \rightarrow y
+\rightarrow h \rightarrow z \rightarrow x$. All points in this curve except $h$ have 
+heights larger than $f(h)$. By the Jordan Curve Theorem, this partitions $\MM$ into
+an ``inside" and an ``outside".
+
+One of the lower neighbors of $h$ (say $p$) is inside, and another (say $q$) is outside.
+Because $h$ is a merge, neither of these can a minimum other than the dominant minimum $u$.
+Hence, both $p$ and $q$ have descending paths to $u$. So there is a path from $p$
+to $q$ where all heights are strictly less than $f(h)$. But this path must intersect $C$.
+Contradiction.
+\end{proof}
+
+\begin{lemma} \label{lem:col} If $h$ is a merge vertex, $|\col(h)| = 2$.
+If $h$ is a split vertex, $|\col(h)| = 1$.
+\end{lemma}
+
+\begin{proof} Consider one contiguous sequence of increasing edges from $h$.
+There exists a contour intersecting all of these edges. By \Clm{cont}, they all
+must have the same color. There are two such contiguous sequences.
+
+If $h$ is a merge vertex, then $|\col(h)| = 2$ by \Lem{mono}. If $h$ is a split vertex,
+there is a single contour passing through all increasing edges. By \Clm{cont}, they
+all have the same color.
+\end{proof}
+
+\subsection{Constructing $\reeb(\MM)$} \label{sec:const}
+
+The main construction procedure is $\onestep(\MM)$. It pieces together $\reeb(\MM)$
+incrementally.
+
+\medskip
+\fbox{
+\begin{minipage}{0.9\textwidth}
+{\bf $\onestep(\MM)$}
+
+\smallskip
+\begin{compactenum}
+	\item Run $\update(\st)$. Pop $\st$ to get $h$.
+	\item Suppose $\mcol(h) = \{c\}$ (so $h$ is split). 
+	\begin{compactenum}
+		\item Connect $h$ (in $\reeb(\MM)$) to $\h(c)$ and the unique minimum corresponding to split $h$. 
+		\item Delete $h$ from $T(c)$, set $\h(c) = h$.
+		\item End procedure.
+	\end{compactenum}
+	\item Let $\mcol(h) = \{c_1, c_2\}$ (so $h$ is merge).
+	\item Connect $h$ in $\reeb(\MM)$ to $\h(c_1)$ and $\h(c_2)$.
+	\item Delete all copies of $h$ from $T(c_1)$ and $T(c_2)$.
+	\item Perform union of colors $c_1$ and $c_2$ (denote merged color as $c$). Merge heaps
+	to get $T(c) = T(c_1) \cup T(c_2)$.
+	\item Set $\h(c) = h$.	
+\end{compactenum}
+\end{minipage}}
+
+\medskip
+\begin{theorem} \label{thm:piece} Consider the very first call to $\onestep(\MM)$. Let $y$ be a point
+infinitesimally below $h$ on a decreasing edge. Consider $\cut(\MM,\cont(y)) = \{\MM_u,\MM_\ell\}$ (the upper and lower manifolds). Then $\onestep(\MM)$
+outputs $\reeb(\MM_u)$ and implicitly gives a valid painting of $\MM_\ell$.
+\end{theorem}
+
+We have an important lemma that helps to bound the running time.
+
+\begin{lemma} \label{lem:tree} For every color $c$, all points in $T(c)$ lie along a monotone path in 
+$\reeb(\MM)$.
+\end{lemma}
+
+\begin{proof} Prove by induction of the number of edges in $\reeb(\MM)$. Consider the head $h$ of $\st$ in the first call to $\onestep(\MM)$,
+and the contour $\cont(y)$ as defined in \Thm{piece}. Take $\cut(\MM,\cont(y)) = \{\MM_u,\MM_\ell\}$. Apply induction
+hypothesis on $\MM_\ell$, and use that to argue about $T(c_1)$ and $T(c_2)$ in the original manifold.
+\end{proof}
+
+\pagebreak
+\newcommand{\pathTree}{P_{\mathcal{S}}}
+\newcommand{\pathTreeA}{\mathcal{P}_{\mathcal{S}}}
+\section{Leaf assignments and path decompositions}
+
+In this section, we set up a framework to analyze the time taken to compute a merge tree.
+
+\begin{definition}
+ We say that a contour tree of a $2$-manifold is a \emph{merge tree} if it has a well defined root vertex such that height values 
+ strictly increase along any root to leaf path.
+\end{definition}
+
+\begin{observation}
+ The contour tree of a minimum dominant manifold (see \Def{dom}) is close to a merge tree.  Specifically, there is a root such that on any root 
+ to leaf path height values strictly increase, with the only exception being potentially at the leaves themselves.  As such, up to constant 
+ factors, any lower bound for computing the merge tree is a lower bound for computing the contour tree of a minimum dominant
+ manifold (i.e. intuitively just clip off the leaves or shift them to lie above their parent).
+\end{observation}
+
+Any tree $T$ considered in this section will be a rooted full binary tree\footnote{Note 
+that technically the trees considered should have a leaf vertex hanging below the root of this 
+full binary tree in order to represent the global minimum of the surface.  This vertex is 
+(safely) ignored to simplify the presentation.} where the height of a vertex is its distance 
+from the root $r$.  As such, the children of a vertex $v\in T$ are the adjacent vertices of larger height
+(and $v$ is the parent of such vertices).  Then the subtree rooted at $v$, denoted $T_v$ consists of the graph induced on all 
+vertices which are descendants of $v$ (including $v$ itself).  For two vertices $v$ and $w$ in $T$ let $d(v,w)$ denote the 
+length of the path between $v$ and $w$.
+We use $A(v)$ to denote the set of ancestors of $v$.
+For set of nodes $U$, $A(U) = \bigcup_{u \in U} A(u)$.
+
+\begin{definition}
+ A \emph{leaf assignment} $\chi$ of a tree $T$ assigns \emph{two} distinct leaves in each internal vertex $v$,
+ one from the left child and one from the right child subtree of $v$.
+\end{definition}
+
+For a vertex $v\in T$, we use $H_v$ to denote the \emph{heap} at $v$.
+Formally, $H_v = \{u | u \in A(v), \chi(u) \cap L(T_v) \neq \emptyset\}$.
+In words, $H_v$ is the set of ancestors of $v$ which are colored by some leaf in $T_v$.
+The following lemma should justify these technical definitions.
+
+\begin{lemma}
+The running time of our algorithm is $O(\sum_{v \in T} \log |H_v|)$.
+% where $T$ is the merge tree of the given input, 
+%and $\chi$ is any valid coloring of $T$.  (maybe add one in the definition of $H_v$ so that $\log |H_v|$ is always $>1$).
+\end{lemma}
+
+\begin{proof} \Sesh{add}
+
+\end{proof}
+
+%vertices along the path from the root to $v$ such that one of the two colors of each vertex is in $L(T_v)$.
+%We will refer to $H_v$ as the \emph{heap} at $v$.
+%Then define the \emph{algorithm} cost function as $f_{alg}(\chi(T)) = \sum_{v\in T} \log |H_v|$.
+%
+%\begin{lemma}
+%\label{lem:algCost}
+%NEEDS TO BE WRITTEN:
+%Here we need a lemma showing that the running time of our algorithm is $O(f_{alg}(\chi(T)))$, where $T$ is the merge tree of the given input, 
+%and $\chi$ is any valid coloring of $T$.  (maybe add one in the definition of $H_v$ so that $\log |H_v|$ is always $>1$).
+%\end{lemma}
+%
+%Now that we have a simple expression upper bounding the running time of our algorithm, we spend the remainder 
+%of this section proving that this running time is optimal.
+
+We introduce the concept of path decompositions. A \emph{leaf path} is simply a contiguous path in $T$
+containing a leaf, which is monotonic in the height values of its vertices.
+% Then in the following two subsections 
+%we relate path decomposition costs (defined below) to the running time of our algorithm using $\Lem{algCost}$.
+
+\begin{definition}
+\label{def:path} A \emph{path decomposition}, $P(T)$, is a partition of $T$ into 
+ a set of vertex disjoint leaf paths.  
+% 
+% The \emph{cost} of $P(T)$ is $\cost(P(T)) = \sum_{p\in P(T)} |p| \log |p|$. (We use $|p|$ to denote the number of vertices in $p$.)
+\end{definition}
+
+Note that the number of paths in $P(T)$ must equal the number of leaves in $T$.
+Fix $\chi$, which fixes all the heaps $H_v$. We choose a special path decomposition that is best defined as a subset of edges in $T$ such that each internal
+vertex has degree at most $2$. This naturally gives a path decomposition.
+For each internal vertex $v\in T$, add the edge from $v$ to $\arg \max_{v_l, v_r} \{|H_{v_l}|, |H_{v_r}|\}$ 
+ where $v_l$ and $v_r$ are the children of $v$ (if $|H_{v_l}|=|H_{v_r}|$ then pick one arbitrarily).
+This is called the \emph{maximum} path decomposition, denoted by $\pmax(T)$.
+
+Our main theorem is the following. We use $|p|$ to denote the number of vertices in $p$.
+
+\begin{theorem} \label{thm:runtime} $\sum_{v \in T} \log |H_v| = O(\sum_{p\in \pmax(T)} |p| \log |p|)$.
+\end{theorem}
+
+\subsection{Shrubs, heavy paths, and light paths}
+
+The paths in $P(T)$ naturally define a tree\footnote{Please excuse the 
+overloading of the term 'tree', it is the most natural term to use here.} of their own.  Specifically, in the original 
+tree $T$ contract each path down to its root.  Call the resulting tree the \emph{shrub} of $T$ corresponding to 
+the path decomposition $P(T)$. Abusing notation, we simply use $P(T)$ to denote the shrub.
+As a result, we use terms like `parent', `child', `sibling', etc. for paths as well.
+The shrub gives a handle on the heaps of a path.
+We use $b(p)$ to denote the \emph{base} of the path, which is vertex in $p$ closest
+to root of $T$. We use $\ell(p)$ to denote the leaf in $p$.
+We use $H_p$ to denote the $H_{b(p)}$.
+% and denote it by $\pathTree(T)$.
+
+\begin{lemma}
+\label{lem:adjacent}
+ Let $p$ be any path in $P(T)$ and let $\{q_1, \dots q_k\}$ be the children on $p$.
+Then $H_{\ell(p)} + \sum_{i=1}^k |H_{q_i}| \leq |H_p|+2|p|$. 
+\end{lemma}
+
+\begin{proof} For convenience, denote $H_i = H_{q_i}$ and $H_0 = H_{\ell(p)}$.
+Consider $v \in \bigcup_{i} H_i$ that lies below $b(p)$ in $T$. 
+Note that such a vertex has only one of its two colors in $L(b(p))$.  
+Since the colors tracked by $H_i$ and $H_j$ for $i\neq j$ are disjoint, 
+such a vertex can appear in only one of the $H_i$'s.
+On the other hand, a vertex $u\in p$ can appear in more than one $H_i$, 
+but since any vertex has exactly two colors it can appear in at most two 
+such heaps.  Hence, $\sum_i |H_i| \leq |H_p| + 2|p|$.
+\ignore{
+ First observe that for all $i$, $r_{i}$ lies above $r_p$ on some root to leaf path. 
+ Let $H_i'$ be the subset of $H_i$ that lies below $r_p$.  Observe that $H_i'\subseteq H_p$, as $L(T_{r_i})\subseteq L(T_{r_p})$.
+
+ Now for any $i\neq j$, $H_i \cap H_j = \emptyset$, as $r_{i}$ and $r_{j}$ are not on the same root to leaf path (i.e. $L(T_{r_i}) \cap L(T_{r_j}) = \emptyset$).
+ In particular, $(H_i\setminus H_i') \cap (H_j \setminus H_j') = \emptyset$, and so $\sum_{i=0}^k |H_i\setminus H_i'| \leq |p|$ 
+ as $H_i\setminus H_i'$ is precisely the subset of $H_i$ that lies on $p$.
+ Moreover, for any $i\neq j$, $ H_i' \cap H_j' = \emptyset$, and so since $H_i'\subseteq H_p$ we have $\sum_{i=0}^k |H_i'|\leq |H_p|$.
+ Putting these together gives
+ \[
+  \sum_{i=0}^k |H_i| = \sum_{i=0}^k |H_i'| + \sum_{i=0}^k |H_i\setminus H_i'| \leq |H_p|+|p|.
+ \]
+}
+\end{proof}
+
+We wish to prove $\sum_{v\in T} \log |H_v| = O(\sum_{p\in P} |p| \log |p|)$. 
+The simplest approach is to prove $\forall p\in P$, $\sum_{v\in p} \log |H_v|= O(|p|\log|p|)$.
+This is unfortunately not true, which is why we divide paths into two categories.
+
+\begin{definition}
+ For $p\in P(T)$, $p$ is \emph{light} if $|p| < \sqrt{|H_{p}|}/100$, and \emph{heavy} otherwise.
+\end{definition}
+
+\ignore{
+\begin{observation}
+\label{obs:decrease}
+ Let $v$ be a vertex in $T$ and let $w$ be its parent.  Then $|H_w| \geq |H_v| -1$, as $L(T_v) \subseteq L(T_w)$ and the path from 
+ $w$ to the root has one less vertex than the path from $v$ to the root.
+ In particular, we have the following more general property.
+ Let $v$ and $u$ be any two vertices in the same root to leaf path of $T$, such that $v$ is a descendant of $u$.  
+ Then $|H_v| \leq |H_u| + d(u,v)$. 
+\end{observation}
+
+The following can be thought of as a generalization of the above observation, and will be useful in later sections.
+}
+
+The following lemma demonstrates that heavy paths can ``pay'' for themselves.
+
+\begin{lemma}
+\label{lem:pathBounds} If $p$ is heavy, $\sum_{v\in p} \log |H_v| = O(|p| \log |p|)$.
+If $p$ is light, $\sum_{v\in p} \log |H_v| = O(|H_p| \log |H_p|)$.
+% Let T be a binary tree and let $\chi(T)$ and $P(T)$ be a valid coloring and path decomposition of $T$, respectively.
+% Then, 
+% \begin{compactenum}
+%  \item for any heavy path $p\in P(T)$, $\sum_{v\in p} \log |H_v| = O(|p| \log |p|)$.
+%  \item for any light path $p\in P(T)$, $\sum_{v\in p} \log |H_v| = O(|p| \log |H_{p}|)$.
+% \end{compactenum}
+\end{lemma}
+\begin{proof} \ignore{
+ By $\Obs{decrease}$ we know that for any vertex $v\in p$, $|H_v|\leq |H_{p}| + |p|$ (as $v$ is a descendant of $r_p$ along $p$).
+}
+For $v \in p$, $|H_v|\leq |H_{p}| + |p|$ (as $v$ is a descendant of $b(p)$ along $p$).
+Hence, $\sum_{v\in p} \log |H_v| \leq \sum_{v\in p} \log(|H_{p}| + |p|)  = |p| \log (|H_{p}| + |p|)$.
+ If $p$ is a heavy path, then $|p| \log (|H_{p}| + |p|) = O(|p| \log |p|)$. If $p$ is light, then 
+ $|p| \log (|H_{p}| + |p|) = O(|p| \log |H_{p}| )$. For light paths, $|p| = O(|H_p|)$.
+\end{proof}
+
+\ignore{
+For a light path $p$ we can think of the quantity $|p|(\log |H_p| - \log |p|)$ as the excess of $p$, i.e. what was not ``paid for'' locally by $p$.
+Our eventual goal will be to show that by choosing the right path decomposition, this excess 
+can be recharged to heavy paths. In the next couple sections we introduce the appropriate path decomposition and prove 
+some useful facts about it.  Then given these facts, in \Sec{} we are able to make the recharging argument.
+}
+
+There are some light paths that we can also ``pay" for.  Consider any light path $p$ in the shrub.  We will refer to the 
+\emph{heavy support chain} of $p$ as the heavy ancestors of $p$ in the shrub which have a path to $p$ which does not use any 
+light path (i.e. it is a chain of paths adjacent to $p$).
+
+\begin{definition} \label{def:support} A light path $p$ is \emph{supported} if
+at least $|H_p|/100$ vertices $v$ in $H_p$ lie in paths in the heavy support chain of $p$.
+%
+%
+% there exists
+%a contiguous (from $p$) chain of ancestors $q_1, q_2, \ldots, q_k$ (in $P(T)$) such that $|H_{q_k}| \leq 99|H_p|/100$
+%and all $q_i$'s are heavy.
+\end{definition}
+
+Let $\cL$ be the set of light paths, $\cL'$ be the set of supported light paths, and $\cH$ be the set of heavy paths
+given by $\pmax(T)$.
+We now construct the shrub of unsupported light paths. Consider $p \in \cL \setminus \cL'$,
+and traverse the chain of ancestors from $p$. Eventually, we must reach another light path $q$.
+(If not, we have reached the root $r$ of $\pmax(T)$. Hence, $p$ is supported.)
+Insert edge from $p$ to $q$, so $q$ is the parent of $p$ in $\cU$. This construction leads
+to the shrub forest of $\cL \setminus \cL'$, where all the roots are supported light paths, 
+and the remaining nodes are the unsupported light paths.
+
+Most of the work goes into proving the following technical lemma.
+
+\begin{lemma} \label{lem:cu-root} Let $\cU$ denote a connected component (shrub) in the shrub forest of $\cL \setminus \cL'$ and let $r$
+be the root of $\cU$. (i) For any $v \in p$ such that $p \in \cU$,
+$|H_v| = O(|H_r|)$. (ii) $\sum_{p \in \cU} |p| = O(|H_r|)$.
+\end{lemma}
+
+We split the remaining argument into two subsections. We first prove \Thm{runtime} from \Lem{cu-root},
+which involves routine calculations. Then we prove \Lem{cu-root}, where the interesting work happens.
+
+\subsection{Proving \Thm{runtime}} \label{sec:thm-runtime}
+
+We split the summation into heavy, light, and unsupported light paths.
+\begin{eqnarray*} 
+\sum_{p \in \cL} \sum_{v \in p} \log |H_v| & = & \sum_{p \in \cL \setminus \cL'} \sum_{v \in p} \log |H_v| + \sum_{p \in \cL'} \sum_{v \in p} \log |H_v| + \sum_{p \in \cH} \sum_{v \in p} \log |H_v|
+%& = & \sum_{p \in \cL \setminus \cL'} \sum_{v \in p} \log |H_v| + O(\sum_{p \in \cL'} |p|\log |H_p|) = \sum_{p \in \cL \setminus \cL'} \sum_{v \in p} \log |H_v| + O(\sum_{p \in \cL'} |H_p|\log |H_p|)
+\end{eqnarray*}
+The last term can be bounded by $O(\sum_{p \in \pmax(T)} |p|\log |p|)$, by \Lem{pathBounds}.
+The second term can be bounded by $O(\sum_{p \in \cL'} |H_p| \log |H_p|)$, by \Lem{pathBounds} again.
+The following claim shows that this in turn is at most the last term.
+
+%We have a claim that also bounds heap sizes for special set of paths.
+
+\begin{claim} \label{clm:above} $\sum_{p \in \cL'} |H_p| \log |H_p| = O(\sum_{q \in \cH} \sum_{v \in q} \log |H_v|)$.
+\end{claim}
+
+\begin{proof} 
+Pick $p \in \cL'$. 
+%Let the first $|H_p|/200$ vertices the lie in heavy paths be denoted as $H'_p$. 
+%Observe that for any $v \in H'_p$, 
+%We have $|H_p| \log |H_p| = O(\sum_{v \in H'_p} \log |H_p|)$
+%$= O(\sum_{v \in H'_p} \log |H_v|)$. Since all the vertices lie in heavy paths,
+%we can write this as $O(\sum_{q \in \cH} \sum_{v \in H'_p \cap q} \log |H_v|)$.
+%Summing over all $p$, the expression is $\sum_{q \in \cH} \sum_{p \in \cL'} \sum_{v \in H'_p \cap q} \log |H_v|$.
+%
+%
+As we traverse the heavy support chain of $p$, there
+are at least $|H_p|/100$ vertices of $H_p$ that lie in these paths. These are encountered in
+a fixed order. Let $H'_p$ be the first $|H_p|/200$ of these vertices. When $v \in H'_p$
+is encountered, there are $|H_p|/200$ vertices of $H_p$ not yet encountered. Hence,
+$|H_v| \geq |H_p|/200$. Hence, $|H_p|\log |H_p| = O(\sum_{v \in H'_p} \log |H_v|)$.
+Since all the vertices lie in heavy paths,
+we can write this as $O(\sum_{q \in \cH} \sum_{v \in H'_p \cap q} \log |H_v|)$.
+Summing over all $p$, the expression is $\sum_{q \in \cH} \sum_{p \in \cL'} \sum_{v \in H'_p \cap q} \log |H_v|$.
+%and $\sum_{p \in \cL'} |H_p|\log |H_p| = O(\sum_{p \in \cL'} \sum_{v \in H'_p} \log |H_v|)$.
+
+Consider any $v \in H'_p$.  Let $S$ be the set of paths $\widetilde{p}\in \cL'$ such that 
+$v\in H'_{\widetilde{p}}$.  We now show $|S|\leq 2$ (i.e. it contains at most one path other than $p$).
+First observe that any two paths in $S$ must be unrelated (i.e. $S$ is an anti-chain), 
+since paths which have an ancestor-descendant relationship have disjoint heavy support chains.  
+However, any vertex $v$ receives exactly one color from each of its two subtrees (in $T$), and therefore $|S|\leq 2$ since any two 
+paths which share descendant leaves in $T$ (i.e. their heaps are tracking the same color) must have an ancestor-descendant relationship.
+
+In other words, any $\log |H_v|$ appears at most twice in the above triple summation.
+Hence, we can bound it by $O(\sum_{q \in \cH} \sum_{v \in q} \log |H_v|)$.
+%
+%Pick $q \in Q$ and some $v \in H_q$. As we traverse up the tree passing
+%through $A(q)$, we must encounter some $p \in A(q)$ such that $v \in p$. Furthermore,
+%there is an order in which these vertices (of $H_q$) occur. Let $H'$ be the first half
+%of vertices that occur. For all $v \in H'$, $|H_v| \geq |H_q|/2$. Hence,
+%$|H_q| \log |H_q| \leq 4 \sum_{v \in H'} \log |H_v| \leq 4 \sum_{v \in H_q} |H_v|$.
+%So $\sum_{q \in Q} |H_q| \log |H_q| \leq 4 \sum_{q \in Q} \sum_{v \in H_q} |H_v|$.
+%
+%Since nodes in $Q$ are incomparable, all the $H_q$'s (for $q \in Q$) are disjoint.
+%Each such $v$ must occur in some path $p \in A(Q)$. Thus,
+%$\sum_{q \in Q} |H_q| \log |H_q| \leq 4 \sum_{p \in A(Q)} \sum_{v \in p} |H_v|$.
+%We apply the bound of \Lem{pathBounds} to complete the proof.
+\end{proof}
+
+The first term (unsupported light paths) can be charged to the second term (supported light paths).
+This is where the critical \Lem{cu-root} plays a role.
+
+\begin{claim} \label{clm:first} $\sum_{p \in \cL \setminus \cL'} \sum_{v \in p} \log |H_v| = O(\sum_{p \in \cL'} |H_p| \log |H_p|)$.
+\end{claim}
+
+\begin{proof} Let $\cU$ denote a connected component of the shrub forest.
+We have $\sum_{p \in \cL \setminus \cL'} \sum_{v \in p} \log |H_v| \leq \sum_{\cU} \sum_{p \in \cU} \sum_{v \in p} \log |H_v|$.
+By \Lem{cu-root}, $|H_v| = O(|H_r|)$, where $r$ is the root of $\cU$.
+Furthermore, $\sum_{p \in \cU} |p| = O(|H_r|)$. 
+We have $\sum_{p \in \cU} \sum_{v \in p} \log |H_v| = O((\log |H_r|) \sum_{p \in \cU} |p|) = O(|H_r|\log |H_r|)$.
+We sum this over all $\cU$ in the shrub forest, and note that roots in the shrub forest are supported light paths.
+\end{proof}
+
+\ignore{
+Now, for the critical piece of the approach. Focus on $\pmax(T)$. Let $L$ be the set of light paths, 
+and $\hat{L} \subseteq L$ be the set of paths such that no ancestor of $p \in L'$
+is light. Obviously, $\hat{L}$ is an anti-chain in $P(T)$.
+
+\begin{lemma} \label{lem:light} For the path decomposition $\pmax(T)$, $\sum_{p \in \cL} \sum_{v \in p} \log |H_v| = O(\sum_{p \in \cL'} |H_p| \log |H_p|)$.
+\end{lemma}
+
+With this lemma, the proof of \Thm{runtime} is straightforward.
+
+\begin{proof} (of \Thm{runtime}) We focus on $\pmax(T)$ and split into heavy and light paths.
+We have $\sum_{v \in T} \log |H_v| =$  $\sum_{p \in \cL} \sum_{v \in p} \log |H_v| + $ $\sum_{p \in \cH} \sum_{v \in p} \log |H_v|$.
+For the first term, we apply \Lem{light} to bound by $O(\sum_{p \in \cL'} |H_p| \log |H_p|)$.
+Then we apply \Clm{above} to bound by $O(\sum_{p \in \cH} |p|\log |p|)$.
+The latter is also bounded by the same expression, by \Lem{pathBounds}. By definition,
+$\sum_{p \in \cH} |p|\log |p| \leq \cost(\pmax(T))$.
+\end{proof}
+
+The main challenge is proving \Lem{light}, which we defer to the next section.
+\begin{definition}
+ Let $\chi(T)$ be a valid coloring of a binary tree $T$.  We define a set of disjoint paths over the vertices of 
+ $T$ as follows.  
+ The corresponding collection of paths that all such edges correspond to is a path decomposition of $T$, which 
+ we refer to as a \emph{maximum} path decomposition, and is denoted by $\mathcal{P}(T)$.  
+\end{definition}
+
+From the previous section we now have a lower bound of $\Omega(f_{path}(P))$ for computing the merge tree in terms of any path decomposition $P$. 
+We now show that the running time of our algorithm for computing the merge tree is upper bounded by this expression for a specific path decomposition, 
+which we will call the maximum path decomposition (defined below).
+Specifically, for a given valid coloring $\chi$ of a tree $T$, we will prove that for the maximum path decomposition $f_{alg}(\chi(T)) = O(f_{path}(P(T)))$. 
+%using the observations about path decompositions from 
+%\Sec{background}, which will then imply $f_{alg}(\chi(T)) = \Theta(f_{path}(P(T)))$.
+
+Throughout this section, for a path $p\in P(T)$, we use the notation $H_p$ to refer to the heap at the root of $p$, i.e. $H_p = H_{r_p}$.
+}
+
+\subsection{Proving \Lem{cu-root}: the root is everything in $\cU$} \label{sec:weight}
+
+\Lem{cu-root} asserts the root $r$ in $\cU$ pretty much encompasses all sizes and heaps in $\cU$.
+We will work with the \emph{reduced} heap $\redH_p$. This is the subset of vertices of $H_p$ that 
+do not appear on the heavy support chain of $p$.
+%only appear in light paths. 
+By definition, for any unsupported light path
+(hence, any non-root $p \in \cU$), $|\redH_p| \geq 99|H_p|/100$. 
+% For notational convenience, we simply set $\redH_r = H_r$.
+%Consider $q \in \cL \setminus \cL'$ that is a child of $p$ in the shrub $\cU$.
+We begin with a key property, which is where the construction of $\pmax(T)$ enters the picture.
+
+%
+%
+%
+%We focus on $\pmax(T)$, and prove the key property of this path decomposition.
+
+\begin{lemma}
+\label{lem:geometric}
+ Let $q$ be the child of some path $p$ in $\cU$, then $|H_p|\geq \frac{3}{2}|H_q|$. 
+ Moreover, if $p\neq r(\cU)$, then $|\redH_p|\geq \frac{3}{2}|\redH_q|$. 
+\end{lemma}
+\begin{proof} Let $h(q)$ denote the heavy path that is a child of $p$ in $\pmax(T)$,
+and an ancestor of $q$. If no such heavy path exists, then by construction $p$ is the parent 
+of $q$ in $\pmax(T)$, and the following argument will go through by setting $h(q)=q$.
+
+The chain of ancestors from $q$ to $h(q)$ consists only of heavy paths.
+Since $q$ is unsupported, these paths contain at most $|H_q|/100$ vertices of $H_q$. 
+Thus, $|H_{h(q)}| \geq 99|H_q|/100$.
+
+Consider the base
+of $h(q)$, which is a node $w$ in $T$. Let $v$ denote the sibling of $w$ in $T$.
+Their parent is called $u$. Note that both $u$ and $v$ are nodes in the path $p$.
+Now, the decomposition $\pmax(T)$ put $u$ and $v$ in the same path $p$. This
+implies $|H_v| \geq |H_w|$. Since $|H_u| \geq |H_v| + |H_w| - 2$,
+$|H_u| \geq 2|H_w| - 2$. Let $b$ be the base of $p$.
+We have $|H_p| = |H_b| \geq |H_u| - |p| \geq 2|H_w| - |p| - 2$. Since $p$ is a light path, $|p| < \sqrt{|H_p|}/100$.
+Applying this bound, we get $|H_p| \geq (2-\delta)|H_w|$ (for a small constant $\delta > 0$).
+Since $w$ is the base of $h(q)$, $H_w = H_{h(q)}$. We apply the bound $|H_{h(q)}| \geq 99|H_q|/100$
+to get $|H_p| \geq 197|H_q|/100$, implying the first part of the lemma.  For the second part, observe that 
+if $p\neq r(\cU)$, then $p$ is unsupported and so $|\redH_p| \geq 99|H_p|/100$, and therefore the second part follows since $|H_q| \geq |\redH_q|$.
+%
+%
+% Let $w$ be the vertex in the path $p$ to which the root of the path $q$ is adjacent to in $T$, 
+% i.e. $w$ is the parent of $r_q$ in $T$.  Let $c$ be the other child of $w$, which lies in $p$.
+% As $\mathcal{P}(T)$ is a maximum path decomposition of $T$ and $c$ rather than $r_q$ is in 
+% the same path as $w$, we have $|H_q|\leq |H_c|$.  Therefore, $|H_w|\geq |H_q|+|H_c|-1\geq 2|H_q|-1$. 
+% Therefore, by \Obs{decrease}, $|H_p|+ d(r_p, w) \geq |H_w|\geq 2|H_q|-1$.  
+% Since $p$ is a light path, $d(r_p, w)\leq \sqrt{|H_p|}/100$ and so we have $|H_p|+\sqrt{|H_p|}/100\geq 2|H_q|-1$.
+% Assuming $|H_p|>11$ (and we can safely assume $|H_p|$ is bigger than any constant), this yields $\frac{3}{5} |H_p| \geq |H_q|$
+\end{proof}
+
+This immediately proves part (i) of \Lem{cu-root}. Part (ii) requires much more work.
+% It is helpful to spell out how the various heaps evolve. Consider $p \in \cU$,
+% which has children $q_1, q_2, \ldots, q_k$ in $\pmax(T)$, and a leaf vertex $\ell$ in $T$. 
+% The reduced heap $\redH_p$ is the
+% union of $\redH_\ell$ and $\redH_{q_1}, \redH_{q_2}, \ldots, \redH_{q_k}$, with the vertices in $p$
+% removed. Note that some of the $q_i$'s are the corresponding $h(q')$ vertices,
+% for a child $q'$ in $\cU$.
+
+We define a \emph{residue} $R_p$ for each $p \in \cU$. Suppose $p$ has children
+$q_1, q_2, \ldots, q_k$ in $\cU$. Then $R_p = |\redH_p| - \sum_i |\redH_{q_i}|$. 
+By definition, $|\redH_p| = \sum_{q \in \cU_p} R_p$. 
+Note that $R_p$ can be negative. 
+Now, define $R^+_p = \max(R_p,0)$, and set $W_p = \sum_{q \in \cU_p} R^+_p$.
+Observe that $W_p \geq |\redH_p|$. We also get an approximate converse.
+
+\begin{claim} \label{clm:loss} For any path $p\in \cU$, $|\redH_p| \geq W_p - 2\sum_{q \in \cU_p} |q|$.
+\end{claim}
+
+\begin{proof} We write $W_p - |\redH_p| = \sum_{q \in \cU_p} R^+_q - R_q$ $= -\sum_{q \in \cU_p: R_q < 0} R_q$.
+Consider $q \in \cU_p$ and denote the children in $\cU_p$ by $q'_1, q'_2, \ldots$.
+Note that $R_q$ is negative exactly when $|\redH_q| < \sum_i |\redH_{q'_i}|$.
+Traverse $\pmax(T)$ from $q'_i$ to $q$.
+Other than $q$, all other nodes encountered are in the heavy support chain of $q'_i$ and hence
+do not affect its reduced heap.
+The vertices of $\redH_{q'_i}$ that are deleted
+are exactly those present in the path $q$. 
+Any vertex in $q$ can be deleted from at most two of the reduced heaps (of the children of $q$ in $\cU_p$), since 
+theses reduced heaps do not have an ancestor-descendant relationship.
+Therefore when $R_q$ is negative, it is at most by $2|q|$.
+We sum over all $q$ to complete the proof.
+\end{proof}
+
+\ignore{Let $D_q = H_q \setminus \bigcup_i H_{q'_i}$. In other words, as we traverse $\cU$,
+$D_q$ is the set of vertices ``added" at $q$. Suppose we traverse $\pmax(T)$ from $q'_i$
+to $q$. All vertices of $H_{q'_i}$ that lie on these paths are removed.
+
+ and traverse $\pmax(T)$ up to $p$. Suppose we encounter
+the paths $q_0 = q, q_1, q_2, \ldots, q_k = p$. The vertices in the residue $R_q$ that 
+are removed from $H_p$ are exactly the vertices that lie in these paths. 
+We split this contribution into heavy and light paths, observing that the latter
+are all unsupported. Hence, we can (crudely) upper bound the number of vertices
+removed by $\sum_{q \in \cU_p} |q| + \sum_{q \in \cU_p} \sum_{h \in \cH} |h \cap q|$.
+Each vertex can be removed from at most two distinct $R_q$'s, completing the proof.
+}
+
+The main challenge of the entire proof is bounding the sum of path lengths, which is done next.
+We stress that the all the previous work is mostly the setup for this claim.
+
+\begin{claim} \label{clm:charge} Fix any path $p \in \cU\setminus\{r(\cU)\}$. Suppose for any $q,q' \in \cU_p$
+where $q$ is a parent of $q'$ in $\cU_p$, $W_q \geq (4/3)W_{q'}$. Then $\sum_{q \in \cU_p} |q| \leq W_p/20$.
+\end{claim}
+
+\begin{proof} Since $q$ is an unsupported light path, $|q| < \sqrt{|H_q|}/100 \leq \sqrt{|\redH_q|}/99 \leq \sqrt{|W_q|}/99$.
+We prove that $\sum_{q \in \cU_p} \sqrt{|W_q|}/99 \leq W_p/20$
+by a charge redistribution scheme. Assume that each $q \in \cU_p^-$
+starts with $\sqrt{W_q}/99$ units of charge. We redistribute this charge over all
+nodes in $\cU_q$, and then calculate the total charge. For $q \in \cU_p$, spread
+its charge to all nodes in $\cU_q$ proportional to $R^+$ values. In other words,
+give $(\sqrt{W_q}/99)\cdot(R^+_{q'}/W_q)$ units of charge to each $q' \in \cU_q$.
+
+After the redistribution, let us compute the charge deposited at $q$. Every ancestor in $\cU_p^-$ 
+$q = a_0, a_1, a_2, \ldots, a_k$ contributes to the charge at $q$. The charge is expressed in the following
+equation. We use the assumption that $W_{a_i} \geq (4/3) W_{a_{i-1}}$ and hence $W_{a_i} \geq (4/3)^i W_{a_0} \geq (4/3)^i$, 
+as $a_0$ is an unsupported light path and hence $W_{a_0}\geq 1$. 
+$$ ({R^+_q}/99)\sum_{a_i} 1/\sqrt{W_{a_i}} \leq ({R^+_q}/99)\sum_{a_i} (3/4)^{i/2} \leq R^+_q/20 $$
+The total charge is $\sum_{q \in \cU_p} R^+_p/20 = W_p/20$. 
+\end{proof}
+
+\begin{corollary} \label{cor:chargeCor}
+ Let $r$ be the root of $\cU$, and suppose that for any paths $q,q' \in \cU\setminus \{r\}$, where $q$ is a parent of $q'$ in $\cU$, $W_q \geq (4/3)W_{q'}$.
+ Then $\sum_{p \in \cU} |p| \leq W_r/20+|r|$.
+\end{corollary}
+\begin{proof}
+ Let $c_1, \dots, c_m$ be the children of $r$ in $\cU$.  By definition, $W_r = \sum_i W_{c_i}+R^+_r \geq \sum_i W_{c_i}$.
+ By \Clm{charge}, for each $c_i$ we have $W_{c_i}/20\geq \sum_{p \in \cU_{c_i}} |p|$.  Combining these to facts yields the claim.
+\end{proof}
+
+% \begin{claim} \label{clm:charge} Fix $p \in \cU$. Suppose for any $q,q' \in \cU_p$
+% where $q$ is a parent of $q'$ in $\cU_p$, $W_q \geq (4/3)W_{q'}$. Then $\sum_{q \in \cU_p} |q| \leq W_p/20$.
+% \end{claim}
+% 
+% \begin{proof} Since $q$ is an unsupported light path, $|q| < \sqrt{|H_q|}/100 \leq \sqrt{|\redH_q|}/99 \leq \sqrt{|W_q|}/99$.
+% We prove that $\sum_{q \in \cU_p} \sqrt{|W_q|}/99 \leq W_p/20$
+% by a charge redistribution scheme. Assume that each $q \in \cU_p$
+% starts with $\sqrt{W_p}/99$ units of charge. We redistribute this charge over all
+% nodes in $\cU_p$, and then calculate the total charge. For $q \in \cU_p$, spread
+% its charge to all nodes in $\cU_q$ proportional to $R^+$ values. In other words,
+% give $(\sqrt{W_q}/99)\cdot(R^+_{q'}/W_q)$ units of charge to each $q' \in \cU_q$.
+% 
+% After the redistribution, let us compute the charge deposited at $q$. Every ancestor in $\cU_p$ 
+% $q = a_0, a_1, a_2, \ldots, a_k$ contributes to the charge at $q$. The charge is expressed in the following
+% equation. We use the assumption that $W_{a_i} \geq (4/3) W_{a_{i-1}}$ and hence $W_{a_i} \geq (4/3)^i W_{a_0} \geq (4/3)^i$, 
+% as $a_0$ is an unsupported light path and hence $W_{a_0}\geq 1$. 
+% $$ ({R^+_q}/99)\sum_{a_i} 1/\sqrt{W_{a_i}} \leq ({R^+_q}/99)\sum_{a_i} (3/4)^{i/2} \leq R^+_q/20 $$
+% The total charge is $\sum_{q \in \cU_p} R^+_p/20 = W_p/20$. 
+% \end{proof}
+
+We wrap it all up by proving part (ii) of \Lem{cu-root}.
+
+\begin{claim} $\sum_{p \in \cU} |p| \leq |H_{r(\cU)}|/10$.
+\end{claim}
+
+% \begin{proof} We use $r$ for $r(\cU)$. Our aim is to simply apply \Clm{charge} from $r(\cU)$. 
+% Suppose $W_q \geq (4/3)W_{q'}$ (for any choice in $\cU$ of $q$ parent of $q'$). By \Clm{charge}, $\sum_{p \in \cU} |p| \leq W_{r}/20$.
+% By \Clm{loss}, $|\redH_r| \geq W_r - 2\sum_{p \in \cU} |p|$. By simple algebra, we deduce, $\sum_{p \in \cU} |p| \leq |\redH_r|/10 = |H_r|/10$.
+% 
+% We now prove that for any $q$ parent of $q'$, $W_q \geq (4/3)W_{q'}$.
+% Suppose not. Let $p, p'$ be the counterexample furthest from the root,
+% where $p$ is the parent of $p'$. 
+% Note that for $q$ and child $q'$ in $\cU_{p'}$, $W_q \geq (4/3)W_{q'}$.
+% We will apply \Clm{charge} for $\cU_{p'}$ to deduce that $\sum_{q \in \cU_{p'}} |q| \leq W_{p'}/20$.
+% By \Clm{loss}, $|\redH_{p'}| \geq 19W_{p'}/20$. By \Lem{geometric}, $|\redH_p| \geq (3/2)|\redH_{p'}|$.
+% Noting that $W_p \geq |\redH_p|$, we deduce that $W_p \geq (4/3)W_{p'}$.
+% Hence, $p, p'$ is not a counterexample, and more generally, this is no counterexample.
+% That completes the whole proof.
+% \end{proof}
+
+\begin{proof} We use $r$ for $r(\cU)$. %Our aim is to simply apply \Clm{charge} from $r(\cU)$. 
+Suppose $W_q \geq (4/3)W_{q'}$ (for any choice in $\cU\setminus\{r\}$ of $q$ parent of $q'$), 
+then by \Cor{chargeCor}, $\sum_{p \in \cU} |p| \leq W_{r}/20+|r|$.
+By \Clm{loss}, $|\redH_r| \geq W_r - 2\sum_{p \in \cU} |p|$, and so combining these inequalities gives,
+\[
+% \sum_{p \in \cU} |p| \leq \pth{|\redH_r|+2\sum_{p \in \cU} |p|}/20+|r| \Rightarrow 
+ \sum_{p \in \cU} |p|\leq \frac{10}{9} \pth{|\redH_r|/20+|r|} \leq 
+ \frac{10}{9}\pth{|H_r|/20+\sqrt{|H_r|}/100} \leq |H_r|/10.
+\]
+
+We now prove that for any $q$ parent of $q'$ (other than $r$), $W_q \geq (4/3)W_{q'}$.
+Suppose not. Let $p, p'$ be the counterexample furthest from the root,
+where $p$ is the parent of $p'$. 
+Note that for $q$ and child $q'$ in $\cU_{p'}$, $W_q \geq (4/3)W_{q'}$.
+We will apply \Clm{charge} for $\cU_{p'}$ to deduce that $\sum_{q \in \cU_{p'}} |q| \leq W_{p'}/20$.
+Combining this with \Clm{loss} gives, $|\redH_{p'}| \geq 19W_{p'}/20$. By \Lem{geometric}, $|\redH_p| \geq (3/2)|\redH_{p'}|$.
+Noting that $W_p \geq |\redH_p|$, we deduce that $W_p \geq (4/3)W_{p'}$.
+Hence, $p, p'$ is not a counterexample, and more generally, there is no counterexample.
+That completes the whole proof.
+\end{proof}
+
+\section{Lower Bound by Path Decomposition}
+In this section we provide a lower bound for the time it takes to compute a contour tree in terms the cost of path decompositions.  Specially, 
+we first consider the easier case of merge trees, after which the lower bound for more general contour trees will easily follow.
+
+\subsection{Merge Trees}
+In this section we show that any algorithm for computing the merge tree $T$ must pay at least $f_{path}(P(T))$ (see \Def{path}) for any path decomposition $P(T)$.
+
+%We say two shrubs $S_1$ and $S_2$ (of merge trees of potentially different surfaces) are equivalent if there is an isomorphism, $f$,
+%between the two shrubs such that for any node $v\in S_1$ corresponding to a path of size $|v|$, $f(v)$ also corresponds to a path of size $|v|$.
+%For the purposes of the next lemma, we say two $2$-manifolds are different, if their corresponding contour trees are the different.
+
+In the following note that two distinct PL surfaces can have the same graph structure defining their meshes, but different height values associated with their vertices.
+
+\begin{figure}[h]\centering
+    \includegraphics[width=.15\linewidth]{figs/diamond}\hspace{2cm}
+    \includegraphics[width=.125\linewidth]{figs/topView}\hspace{2cm}
+    \includegraphics[width=.095\linewidth]{figs/sideView}
+    \caption{From left to right: the cone portion of a diamond; the top down view and then the side view of a cone face with a platform added.}
+    \label{fig:diamond}
+\end{figure}
+
+\begin{lemma}
+\label{lem:pathDecomp}
+Let $T$ be a merge tree, and let $P = P(T)$ any path decomposition of $T$.  %, and $\pathTree$ the corresponding shrub.  
+Then there is a set of at least $\Pi_{p_i\in P} (|p_i|-1)!$ distinct PL $2$-manifolds, all of which have the same defining mesh.
+\end{lemma}
+\begin{proof}
+ We describe a simple mesh, with heights specified for all vertices except for the saddles.  The main idea is that 
+ this mesh will be flexible enough to allow almost any assignment of heights to the saddles, thus defining a class of merge trees 
+ whose size is that described in the lemma statement.
+
+ First, we describe the basic building block of this input mesh, which roughly speaking corresponds to a fixed path $p\in P$.  
+ Specifically, we say a \emph{diamond} of size $m$ is the following mesh.  To start, construct a cycle 
+ of length $m$, where the height of the $i$th vertex in the cycle is say $i/2m$.  Next, create an apex vertex at height 
+ $\infty$ (or any appropriately large finite height), and add an edge between this vertex and each vertex in the cycle.  At this point 
+ we have built an upside down cone with $m$ triangular faces (see \Fig{diamond}). To make this cone a closed surface we add a triangular cap at the bottom. 
+ Specifically, add a $3$ cycle of vertices at heights $0$, $1/20m$, and $1/10m$, and connect each vertex of this $3$-cycle to an arbitrary distinct 
+ vertex on the $m$ cycle (with matching cyclic ordering), and triangulate all remaining faces.  For consistency, if the size $m$ is $\leq 3$, the diamond 
+ will be just a tetrahedron.
+
+ Now above each face of the cone of the diamond, we will create a triangular ``platform'', which will act as the location to glue on other (shrunken) diamonds, 
+ and whose highest point of connection to the face below will be a saddle vertex.  Specifically the heights of the vertices of the triangle will be $h$, $h-\eps$, and $h-2\eps$, where 
+ $h$ is some unspecified height value.  This near horizontal platform is then connected to its respective face with vertical edges down to the face. See \Fig{diamond}.
+ %
+ %The vertex at height $h$ will lie in the plane of the respective face of the cone, and the other two will lie above the face, 
+ %with edges connecting down to the face, hence creating an actual ``platform''.
+
+ At this point we can construct the entire input mesh, by gluing together diamonds.  
+ Specifically, for each path $p_i\in P$ we create a diamond of size $|p_i|-1$ (we subtract one because the tip of the diamond already corresponds to the leaf of the path).
+ Now we glue together diamonds of different paths in the same way the paths are connected in the shrub $\pathTree$.  Specially, for two paths $p,q\in P$ where 
+ $p$ is the parent of $q$ in $\pathTree$, we glue the bottom triangular cap of the diamond for $q$ onto the triangular ``platform'' of some face of the diamond for $p$.
+ (Naturally for this construction to work, diamonds for a given path will be scaled down relative to the size of the diamond of their parent).
+
+ Observe now that the only saddle points in this construction are the highest connecting point on a face of an edge from the platform above.  Moreover, the only maxima are the apexes of the 
+ diamonds, and there is exactly one minima corresponding to the bottom cap of the diamond representing the root of $\pathTree$.  Therefore, the saddles on a given 
+ diamond will appear contiguously on a root to leaf path in the merge tree of the mesh, where the leaf corresponds to the maxima of the diamond 
+ (since all these saddles have a direct line of sight to this apex).  In particular, 
+ this implies that, regardless of the heights assigned to the platforms, the merge tree of this mesh has a path decomposition whose corresponding shrub is equivalent to $\pathTree$.
+
+ Now there is a valid instantiation of this described mesh for any relative ordering of the heights of the saddles on a given diamond.  In particular, 
+ there are $(|p|-1)!$ possible orderings of the heights of the saddles on the diamond for $p$, and hence $\Pi_{p_i\in P} (|p_i|-1)!$ possible instantiations of 
+ the mesh overall.  Also, each one of these instantiations will result in a different merge tree.  Specifically, a permutation of the heights on a given diamond 
+ corresponds to a permutation of the vertices along a path in $P$, as by construction all saddles on a given diamond will appear in sorted order in the merge tree.
+\end{proof}
+
+\begin{corollary}
+ \label{cor:mergeLB}
+ Let $A(S)$ be any comparison based algorithm to compute the merge tree of any input surface $S$ (whose contour tree is a merge tree).  
+ Then the running time of $A(S)$ is $\Omega(f_{path}(P))=\Omega(\sum_{p_i\in P} |p_i|\log|p_i|)$ where $P$ is any path decomposition of the 
+ merge tree of $S$.
+\end{corollary}
+\begin{proof}
+ Given an input surface $S$, by the above lemma, any path decomposition $P$ of its merge tree defines a set of $\Pi_{p_i\in P} |p_i|!$ surfaces 
+ with distinct merge trees, all of which have the same mesh.  Specifically, for each sorted ordering of the relative height values of the saddles 
+ on each diamond in the mesh, yields a different merge tree.
+ Therefore, any comparison based algorithm which computes the merge tree must distinguish between these possibilities, and therefore its corresponding decision tree must 
+ have $\Pi_{p_i\in P} |p_i|!$ leaves.  Hence by Stirling's approximation, this decision tree must have depth $\Omega(\sum_{p_i\in P} |p_i|\log|p_i|)$.
+\end{proof}
+
+\subsection{Contour Trees}
+We first generalize previous terms to the case of contour trees.
+In this section $T$ will denote an arbitrary contour tree.  Specifically, every internal vertex in $T$ has degree $3$ 
+and the height function defined over the vertex set is such that every internal vertex in either a split or merge 
+(i.e. at most $2$ neighbors lie above or below it).
+
+Again we define a path decomposition as a partition of the vertices into 
+disjoint leaf paths (i.e. each path is monotonic in height and contains a leaf). 
+However, for simplicity we now restrict our attention to path decompositions which 
+roughly speaking are consistent with the raining procedure described in \Sec{rain} 
+(more general decompositions can work, but it is not needed for our purposes).
+
+\begin{definition}
+\label{def:path2} A path decomposition, $P(T)$, is called \emph{rain consistent} if its paths can be obtained as follows.
+Perform an upward BFS from an arbitrary minima $v$ in $T$, and mark all vertices encountered.  
+Now recursively run a directional BFS from all vertices adjacent to the current marked set.  
+Specifically, for each BFS run, make it an upward BFS if it is at an odd height in the recursion tree and downward otherwise.
+
+This procedure partitions the vertex set into disjoint rooted subtrees of $T$, based on which BFS marked a vertex.  
+For each such subtree, now take any partition of the vertices into leaf paths.\footnote{Note that the subtree  
+of the initial vertex is rooted at is a minima.  For simplicity we require that the path this vertex belongs to 
+also contains a maxima.}
+\end{definition}
+
+The following is analogous to \Lem{pathDecomp}, and in particular uses it as a subroutine.
+
+\begin{lemma}
+\label{lem:pathDecomp2}
+Let $T$ be a contour tree, and let $P = P(T)$ be any rain consistent path decomposition of $T$.  
+Then there is a set of at least $\Pi_{p_i\in P} (|p_i|-1)!$ distinct PL $2$-manifolds, all of which have the same defining mesh.
+\end{lemma}
+\begin{proof}
+ As $P$ is rain consistent, the paths can be partitioned into sets $P_1, \dots, P_m$, where $P_i$ is the set of all paths with 
+ vertices from a given BFS, as described in \Def{path2}.  Specifically, let $T_i$ be the subtree of $T$ corresponding to $P_i$ and 
+ let $r_i$ be the root vertex of this subtree.  
+ Note that the $P_i$ sets naturally define a tree where $P_i$ is the parent of $P_j$ if $r_i$ is adjacent to a vertex in $P_j$.
+
+ As the set $P_i$ is a path decomposition of a rooted binary tree $T_i$, the surface mesh construction of \Lem{pathDecomp} for $P_i$ is well defined.  
+ Actually the only difference is that here the rooted tree is not a full binary tree, and so some of the faces of the constructed diamonds will be blank.  
+ Specifically, these blank faces correspond to the adjacent children of $P_i$, and they tell us how to connect the meshes of the different $P_i$'s.
+
+\begin{figure}[h]\centering
+    \includegraphics[width=.5\linewidth]{figs/bridge2}
+    \caption{On the left, the root of $P_i$ merges into a diamond of its parent $P_j$.  On the right, the root of $P_i$ splits off from a diamond of $P_j$}
+    \label{fig:diamond}
+\end{figure}
+
+ So for each $P_i$ construct a surface mesh as described in \Def{path2}.  By construction, the diamonds corresponding to the paths in $P_i$ are connected 
+ into a tree structure (i.e. corresponding to the shrub of $P_i$).  Therefore the bottoms of all these diamonds are covered except for the one corresponding 
+ to the path containing the root $r_i$.  If $r_i$ is the initial minima that the rain consistent path decomposition was defined from,  
+ we add a tetrahedral cap to the uncovered bottom of this diamond (i.e. we construct the minima).  Otherwise, $P_i$ has some parent $P_j$ in which case 
+ we connect the bottom of the diamond for $r_i$ to a free face of a diamond in the construction for $P_j$, specifically, the face corresponding to the 
+ vertex in $T$ which $r_i$ is adjacent to. 
+ If $T_i$ was created by an upward BFS, then the connection to this face should be a merge vertex, and otherwise it should be a split vertex.  
+ See \Fig{diamond}.
+
+ Just as in \Lem{pathDecomp} we now have one fixed mesh graph structure, such that each different relative ordering of the heights of the merge and split vertices 
+ on each diamond produces a surface with a distinct merge tree.  The specific bound on the number of such distinct manifolds follows by applying the 
+ bound from \Lem{pathDecomp} to each $P_i$.
+\end{proof}
+
+For the same reason that \Cor{mergeLB} followed from \Lem{pathDecomp}, we now have the following.
+
+\begin{theorem}
+ Let $A(S)$ be any comparison based algorithm to compute the contour tree of an input surface $S$.  
+ Then the running time of $A(S)$ is $\Omega(f_{path}(P))=\Omega(\sum_{p_i\in P} |p_i|\log|p_i|)$ where $P$ 
+ is any rain consistent path decomposition of the contour tree of $S$.
+\end{theorem}
+
+% \section{Temp}
+% Given a contour tree we allow any path decomp, as long as each path contains a maxima or minima (or both).
+% We call a path $p$ in the decomp a min (resp. max) path if it contains a minima (resp. maxima) (note a path can be both).
+% For every path we create a cylinder with $m$ faces for $p$, where $m$ is the number of non-leaf vertices in $p$ (either $|p|-1$ or $|p|-2$).  
+% If $p$ has a maxima we put a top cone cap on the cylinder, and if it contains a manima then put a bottom cone cap (i.e. the cap represents the maxima or minima).
+% 
+% Every edge $e$ in $T$, but not in a path decomp. tells us how to glue together cylinders.  Let $u$ be the upper endpoint of $e$ and $l$ the lower endpoint.
+% 
+% First we ``remove'' all edges $e$ such that $u$ is adjacent to a max path and $l$ adjacent to a min path by joining the free ends of the respective cylinder together.
+% Hence we can assume no such edges exist.  Now we have three cases:
+% 
+% \begin{enumerate}[{Case} 1]
+%  \item 
+% $l$ is adjacent to a min path:\newline
+% Note in this case the path containing $u$ must be a min path.
+% In this case connect the bottom of the cylinder for the path containing $l$ into a face of the cylinder for the path containing $u$.
+% \item
+% $l$ is adjacent to a max path and $u$ is adjacent to a max path:\newline
+% In this case connect the bottom of the cylinder for the path containing $u$ into a face of the cylinder for the path containing $l$.
+% \item
+% $l$ is adjacent to a max path and $u$ is adjacent to a min path:\newline
+% Connect a face of the cylinder for the $u$ path (using a split) to a face of the cylinder for the $l$ path (using a merge).
+% \end{enumerate}
+
+\bibliographystyle{alpha}
+\bibliography{contour}
+
+\appendix
+
+\section{Old stuff}
+
+\subsubsection{Paying it down the line}
+
+% \begin{claim}
+% \label{clm:sizes}
+%  Let $p$ be a path in a maximum path decomposition, such that $|H_p|>c|p|$ for some constant $c> 6$.
+%  Let $H_f$ denote the heap at the leaf of $p$.  We then have the following
+%  \begin{compactenum}
+%   \item $|H_f|\leq (1+1/c)|H_p|$
+%   \item $|H_f|\geq |H_p|/3$
+%   \item For any path $q$ adjacent to $p$ in the shrub, $|H_q|<\frac{2}{3}|H_p|$
+%  \end{compactenum} 
+% \end{claim}
+% \begin{proof}
+%   For any vertex $v\in p$ \Obs{decrease} implies that $|H_v|\leq |H_p| + |p| < (1+1/c) |H_p|$.
+%   First observe that by setting $v=f$ this immediately implies the first part of the claim.
+%  % By a similar argument, $|H_f|\geq |H_p| - |p| > (1-1/c)|H_p|$.
+%   
+%   To prove the last part of the claim, suppose otherwise that $|H_q|\geq \frac{2}{3}|H_p|$ for some 
+%   adjacent path $q$.  As this is a maximum path decomposition, this implies that there is some 
+%   $v\in p$ such that $|H_v|\geq \frac{4}{3}|H_p|-1$.
+%   By \Obs{decrease} we then have, 
+%    $|H_p| \geq |H_v| - |p|\geq \frac{4}{3}|H_p| - 1 - |H_p|/c \geq (4/3-1/c)|H_p|-1$.
+%   When rearranged this gives $1\geq (1/3-1/c)|H_p|$.  As $c>6$ we have $1\geq |H_p|/6$, 
+%   which is a contradiction as $|H_p|>c|p|>6$.
+% \end{proof}
+
+\Ben{I had originally thought the following claim would be useful in proving the lemma below.}
+\begin{claim}
+\label{clm:sizes}
+ Let $p$ be a path in a maximum path decomposition, such that $|H_p|>c|p|$ for some constant $c> 6$.
+ We then have the following
+% \begin{compactenum}
+%  \item 
+  For any path $q$ adjacent to $p$ in the shrub, $|H_q|<\frac{2}{3}|H_p|$
+%  \item If there does not exist and adjacent path $q$ such $H_q\geq H_p/3$, then $|H_f|\geq |H_p|/3$
+% \end{compactenum} 
+\end{claim}
+\begin{proof}
+  To prove the claim, suppose otherwise that $|H_q|\geq \frac{2}{3}|H_p|$ for some 
+  adjacent path $q$.  As this is a maximum path decomposition, this implies that there is some 
+  $v\in p$ such that $|H_v|\geq \frac{4}{3}|H_p|-1$.
+  By \Obs{decrease} we then have, 
+   $|H_p| \geq |H_v| - |p|\geq \frac{4}{3}|H_p| - 1 - |H_p|/c \geq (4/3-1/c)|H_p|-1$.
+  When rearranged this gives $1\geq (1/3-1/c)|H_p|$.  As $c>6$ we have $1\geq |H_p|/6$, 
+  which is a contradiction as $|H_p|>c|p|>6$.  
+\end{proof}
+
+\begin{lemma}
+\label{lem:excess}
+ Let $p$ be a heavy path, and let $S_1, \dots S_k$ and $R_1, \dots, R_l$ be the sizes of the heaps at the roots of the light and heavy children of $p$, respectively, in the shrub.  
+ Let $H_f$ be the heap at the leaf of the path $p$.
+ Then,  $|H_p| \log |H_p|   \geq  \sum_{v\in p} \log|H_v| + \frac{4}{3}\sum_{i=1}^k S_i \log S_i + \sum_{i=1}^l R_i \log R_i  -  c|p|\log |p|$, for some sufficiently large constant $c$
+ \end{lemma}
+\begin{proof}
+ First observe that the lemma holds for $|H_p|$ less than any fixed constant, since otherwise \Lem{adjacent} implies we can choose $c$ 
+ large enough such that $c|p|\log|p|$ will dominate all other terms.
+ Next, observe that $\sum_{v\in p} \log|H_v| \leq \sum_{v\in p} \log(|H_v|+|p|) = |p|\log(|H_v|+|p|) = O(|p|\log|p|)$, as $p$ is a heavy path.  
+ Therefore, in the remainder of the proof  we can safely ignore the $\sum_{v\in p} \log|H_v|$ term as we can choose $c$ large enough to cover it.  
+ That is we now instead prove the inequality,
+ \[
+  |H_p| \log |H_p|   \geq \frac{4}{3}\sum_{i=1}^k S_i \log S_i + \sum_{i=1}^l R_i \log R_i  -  c|p|\log |p|
+ \]
+ To do so we  break the analysis into two case based on the relative size of $|p|$ and $|H_p|$. 
+ In the following, $\alpha>10$ is some yet to be determined constant.
+
+\paragraph{Case 1, $|p|\geq |H_p|/\alpha$:\\}
+First observe that by \Lem{adjacent}, $|H_p|+|p| \geq \sum_{i=1}^k S_i+ \sum_{i=1}^l R_i$.
+By choosing $c$ large enough we therefore have,
+\begin{align*}
+ c|p|\log|p| &\geq 2(1+\alpha)^2 |p|\log |p| \geq 2(1+\alpha)|p| \log((1+\alpha)|p|)\\
+             &\geq 2(|H_p|+|p|)\log(|H_p|+|p|) \geq 2\log(|H_p|+|p|) \pth{\sum_{i=1}^k S_i + \sum_{i=1}^l R_i}\\
+             &\geq \frac{4}{3}\sum_{i=1}^k S_i \log S_i + \sum_{i=1}^l R_i \log R_i 
+\end{align*}
+
+In this case $\log(|H_p|+|p|)\leq \log(2|p|) \leq 1+\log(|p|)$ (where $\log$ is base $2$).
+Therefore, assuming $|p|>1$, we have 
+\[
+(|H_p|+|p|)\log(|H_p|+|p|)\leq (1+c)|p|\log((1+c)|p|) %(|H_p|+|p|)(1+\log |p|) \leq 2(1+c)|p|\log |p|
+\]
+which when combined with the inequality from \Lem{adjacent} gives, 
+\[
+ 4(1+c)|p|\log |p| \geq 2(|H_p|+|p|)\log(|H_p|+|p|)\geq \pth{|H_f| + \frac{4}{3}\sum_{i=1}^k S_i + \sum_{i=1}^l R_i}\log(|H_p|+|p|).
+\]
+This easily implies the desired inequality as $\log(H_p+p)$ is larger than any other $\log$ term in the inequality.
+
+\paragraph{Case 2, $|p|< |H_p|/\alpha$:\\}
+
+\Ben{Was not able to handle this case.}
+
+%  First, lets assume we can show $\sum_{i=1}^k S_i + \sum_{i=1}^l R_i \leq \frac{3}{4}|H_p|$.  
+%  In this case, using the last part of \Clm{sizes}, we can prove the desired inequality as follows.
+%  
+%  \begin{align*}
+%   \frac{4}{3}\sum_{i=1}^k S_i \log S_i + \sum_{i=1}^l R_i \log R_i  &\leq  \frac{4}{3} \pth{\sum_{i=1}^k S_i \log S_i + \sum_{i=1}^l R_i \log R_i} \\
+%   &\leq  \frac{4}{3} \log |H_p| \pth{\sum_{i=1}^k S_i + \sum_{i=1}^l R_i}  \leq  |H_p| \log |H_p|
+%  \end{align*}
+%  
+%  Finally, to complete the proof we must prove our assumption that $\sum_{i=1}^k S_i + \sum_{i=1}^l R_i \leq \frac{3}{4}|H_p|$.
+%  By \Lem{adjacent} and \Clm{sizes} we have, 
+%  \[
+%  \sum_{i=1}^k S_i + \sum_{i=1}^l R_i 
+%  \leq |H_p|+|p|-|H_f| 
+%  \leq (1+1/\alpha-1/3)|H_p| = (2/3+1/\alpha)|H_p|
+%  \]
+%  
+%  Therefore $\sum_{i=1}^k S_i + \sum_{i=1}^l R_i \leq \frac{3}{4}|H_p|$ if $(2/3 + 1/\alpha)\leq 3/4$, which rearranged gives $\alpha \geq 12$
+\end{proof}
+
+\remove{
+% \subsubsection{Paying it down the line}
+% 
+% \begin{lemma}
+% \label{lem:excess}
+%  Let $p$ be a heavy path, and let $S_1, \dots S_k$ and $R_1, \dots, R_l$ be the sizes of the heaps at the roots of the light and heavy children of $p$, respectively, in the shrub.  
+%  Let $H_f$ be the size of the heap at the leaf of the path $p$, and let $X = \sum_{i=0}^k H_i \log H_i$.  
+%  Then,  $|H_p| \log |H_p|   \geq  H_f\log H_f + \sum_{i=0}^k S_i \log S_i + \sum_{i=0}^l R_i \log R_i  -  c|p|\log |p|$, for some sufficiently large constant $c$
+%  \end{lemma}
+% \begin{proof}
+% By \Lem{adjacent} we know that $|H_p|+|p| \geq  H_f + \sum_{i=0}^k S_i + \sum_{i=0}^l R_i$.  
+% %Now as $p$ is a heavy path we have $|p|\geq \sqrt{|H_p|}/100$, and so, $(|H_p|+|p|)\log(|H_p|+|p|)$
+% 
+% We  break the analysis into two case based on the relative size of $|p|$ and $|H_p|$.  In the following $c$ is some yet to be determined constant greater than $1$.
+% 
+% \paragraph{Case 1, $|p|\geq |H_p|/c$:}  
+% In this case $\log(|H_p|+|p|)\leq \log(2|p|) \leq 1+\log(|p|)$ (where $\log$ is base $2$).
+% Therefore, assuming $|p|>1$, we have 
+% \[
+% (|H_p|+|p|)\log(H_p+p)\leq (|H_p|+|p|)(1+\log |p|) \leq (2|p|)(2\log |p|)\leq 4|p|\log |p|
+% \]
+% which when combined with the inequality from \Lem{adjacent} gives, 
+% \[
+%  4|p|\log |p| \geq (|H_p|+|p|)\log(|H_p|+|p|)\geq \pth{H_f + \sum_{i=0}^k S_i + \sum_{i=0}^l R_i}\log(|H_p|+|p|).
+% \]
+% This easily implies the claim as $\log(H_p+p)$ is larger than any other $\log$ term in the inequality.
+% 
+% \paragraph{Case 2, $|p|< |H_p|/c$:}
+% \end{proof}
+}
+
+Define the crust of the shrub to be the first (i.e. closest to the root) layer of light paths in the shrub.  Apply \Lem{lightCost} to each of the light paths in the crust and then 
+say by \Lem{excess} and since the root heap goes to zero we are done.
+
+\ignore{
+Naturally, this geometric series behavior can be extended to more than two adjacent light paths, thus motivating 
+the following definition.
+
+\begin{definition}
+ Let $p$ be a path in a path decomposition $P(T)$, with corresponding shrub $\pathTree$.
+ Call a light path $q\in P(T)$ light reachable from $p$ if the path from $p$ to $q$ in $\pathTree$ 
+ contains only vertices which correspond to light paths.  Call the subtree of $\pathTree$ consisting 
+ of all light reachable paths from $p$ the \emph{light shrub} of $p$, and denote it $\mathcal{L}(p)$.
+\end{definition}
+
+In the following subsection we show how to exploit this geometric series behavior of light shrubs.
+
+We say two paths in $P(T)$ are adjacent if their corresponding vertices are adjacent in the shrub.  Similarly, for 
+two adjacent paths in $P(T)$ we can define a parent/child relationship based on this relationship in the shrub. 
+}
+
+\remove{
+\Ben{Maybe move the remainder of this subsection to 5.3 since heavy and light paths and not used until then}
+
+\begin{definition}
+ Let T be a binary tree and let $\chi(T)$ and $P(T)$ be a valid coloring and path decomposition of $T$, respectively.  
+ For $p\in P(T)$, we say that $p$ is \emph{light} if $|p| < \sqrt{|H_{r_p}|}/100$, and we say it is \emph{heavy} otherwise.
+\end{definition}
+
+\begin{observation}
+\label{obs:decrease}
+ Let $v$ be a vertex in $T$ and let $w$ be its parent.  Then $|H_w| \geq |H_v| -1$, as $L(T_v) \subseteq L(T_w)$ and the path from 
+ $w$ to the root has one less vertex than the path from $v$ to the root.
+
+ In particular, we have the following more general property.
+ Let $v$ and $u$ be any two vertices in the same root to leaf path of $T$, such that $u$ is lower than $v$.  
+ Then $|H_v| \leq |H_u| + d(u,v)$. 
+\end{observation}
+
+\begin{lemma}
+\label{lem:pathBounds}
+ Let T be a binary tree and let $\chi(T)$ and $P(T)$ be a valid coloring and path decomposition of $T$, respectively.
+ Then, 
+ \begin{compactenum}
+  \item for any heavy path $p\in P(T)$, $\sum_{v\in p} \log |H_v| = O(|p| \log |p|)$.
+  \item for any light path $p\in P(T)$, $\sum_{v\in p} \log |H_v| = O(|p| \log |H_{r_p}|)$.
+ \end{compactenum}
+\end{lemma}
+\begin{proof}
+ By $\Obs{decrease}$ we know that for any vertex $v\in p$, $|H_v|\leq |H_{r_p}| + |p|$ (as $r_p$ lies below $v$ on $p$).
+ As such, 
+ \[
+ \sum_{v\in p} \log |H_v| \leq \sum_{v\in p} \log(|H_{r_p}| + |p|)  = |p| \log (|H_{r_p}| + |p|).
+ \]
+ If $p$ is a heavy path then $|p| \log (|H_{r_p}| + |p|) = O(|p| \log |p|)$, and if $p$ is light then 
+ $|p| \log (|H_{r_p}| + |p|) = O(|p| \log |H_{r_p}| )$.
+\end{proof}
+
+The following lemma can be thought of as a generalization of $\Obs{decrease}$.
+
+\begin{lemma}
+\label{lem:adjacent}
+ Let $T$ be a binary tree and let $\chi(T)$ and $P(T)$ be a valid coloring and path decomposition of $T$, respectively.
+ Let $p$ be any path in $P(T)$ and let $\{q_1, \dots q_k\}$ be the set of all adjacent child paths of $p$.
+ For $1\leq i\leq k$ let $r_i = r_{q_i}$ and let $H_i = H_{r_i}$.
+ Also, let $H_0$ denote the heap at the leaf of $p$.
+
+ Then $\sum_{i=0}^k |H_i| \leq |H_p|+|p|$, where $H_p = H_{r_p}$.
+\end{lemma}
+\begin{proof} 
+ First observe that for all $i$, $r_{i}$ lies above $r_p$ on some root to leaf path. 
+ Let $H_i'$ be the subset of $H_i$ that lies below $r_p$.  Observe that $H_i'\subseteq H_p$, as $L(T_{r_i})\subseteq L(T_{r_p})$.
+
+ Now for any $i\neq j$, $H_i \cap H_j = \emptyset$, as $r_{i}$ and $r_{j}$ are not on the same root to leaf path (i.e. $L(T_{r_i}) \cap L(T_{r_j}) = \emptyset$).
+ In particular, $(H_i\setminus H_i') \cap (H_j \setminus H_j') = \emptyset$, and so $\sum_{i=0}^k |H_i\setminus H_i'| \leq |p|$ 
+ as $H_i\setminus H_i'$ is precisely the subset of $H_i$ that lies on $p$.
+ Moreover, for any $i\neq j$, $ H_i' \cap H_j' = \emptyset$, and so since $H_i'\subseteq H_p$ we have $\sum_{i=0}^k |H_i'|\leq |H_p|$.
+ Putting these together gives
+ \[
+  \sum_{i=0}^k |H_i| = \sum_{i=0}^k |H_i'| + \sum_{i=0}^k |H_i\setminus H_i'| \leq |H_p|+|p|.
+ \]
+\end{proof}
+
+% \Ben{There is a fundamental flaw in the following lemma and corollary.  
+% As such, the remainder of this subsection will likely be deleted.}
+% 
+% Using the notation defined in the above lemma, we have the following crucial lemma, which we will use to 
+% argue we can charge that cost of all paths in $P(T)$ to the cost of the heavy paths.
+% 
+% \begin{lemma}
+% Let $p$ be a heavy path in $P(T)$ and let $\{q_1, \dots q_k\}$ be any set of light paths in $P(T)$ such that 
+% $\sum_{i=1}^k |H_i| = O(H_p)$.  Then $\sum_{i=1}^k \sum_{v\in q_i} \log |H_v| = O(|p|\log |p|)$
+% \end{lemma}
+% \begin{proof}
+% By \Lem{pathBounds}, for all $1\leq i\leq k$, 
+%  $\sum_{v\in q_i} \log |H_v| = O(|q_i|\log |H_i|) = O(|H_i|\log |H_i|)$, as $q_i$ is a light path.
+%  By assumption $\sum_{i=1}^k |H_i| = O(|H_p|)$, 
+%  As $x\log x$ is a convex function, Jensen's inequality then implies 
+%  $\sum_{i=1}^{k} |H_i|\log |H_i| = O(|H_p| \log |H_p|) = O(|p|\log |p|)$, since $p$ is a heavy path.
+% \end{proof}
+% 
+% 
+% \begin{corollary}
+% \label{cor:heavyLight}
+%  Let $p$ be a heavy path in $P(T)$ and let $\{q_1, \dots q_k\}$ be the set of all light paths adjacent to $p$.
+%  Then $\sum_{i=1}^k \sum_{v\in q_i} \log |H_v| = O(|p|\log |p|)$.
+% \end{corollary}
+% \begin{proof}
+%  Let $H_i$ be as defined in \Lem{adjacent}.  By \Lem{pathBounds}, for all $1\leq i\leq k$, 
+%  $\sum_{v\in q_i} \log |H_v| = O(|q_i|\log |H_i|) = O(|H_i|\log |H_i|)$, as $q_i$ is a light path.
+%  Now by \Lem{adjacent}, $\sum_{i=1}^k |H_i| \leq |H_p|+|p| = O(|p|)$, as $p$ is a heavy path. 
+%  Therefore, as $x\log x$ is a convex function, Jensen's inequality implies the claim.
+%  
+% \end{proof}
+
+}
+
+\subsection{The shrub of unsupported paths} \label{sec:light}
+
+It suffices to focus on unsupported light paths. \Lem{light} can be reduced to the following.
+
+\begin{lemma} \label{lem:support} $\sum_{p \in \cL \setminus \cL'} \sum_{v \in p} \log |H_v| = O(\sum_{p \in \cL'} |H_p| \log |H_p|)$.
+\end{lemma}
+
+We now construct the shrub of unsupported light paths. Consider $p \in \cL \setminus \cL'$,
+and traverse the chain of ancestors from $p$. Eventually, we must reach another light path $q$.
+(If not, we have reached the root $r$ of $\pmax(T)$. Hence, $p$ is supported.)
+Insert edge from $p$ to $q$, so $q$ is the parent of $p$ in $\cU$. This construction leads
+to the shrub forest of $\cL \setminus \cL'$, where every unsupported light path is an internal node, and all
+roots are supported light paths.
+
+We list out some important claims about this shrub forest.
+We focus on $\cU$, a connected component (shrub) in the shrub forest and denote the 
+the root by $r(\cU)$.
+
+\begin{claim} \label{clm:loss2} Consider $p \in \cL \setminus \cL'$ that is in $\cU$. Let $Q$
+be the set of heavy ancestors of $p$ that are descendants of $r(\cU)$. Then $\sum_{q \in Q} |q \cap H_p| \leq |H_p|/50$.
+\end{claim}
+
+\begin{proof} Suppose not, for contradictions sake. As we traverse the heavy ancestors of $p$,
+we encounter various vertices of $H_p$ in the ancestors $q$. Consider the first $|H_p|/100$
+such vertices of $p$ encountered. At each such $v$, there are at least $|H_p|/100$
+vertices \emph{not yet} encountered, so $|H_v| > |H_p|/100$. Each of these vertices $v$
+is supported, implying that $p$ is a supported light path.
+\end{proof}
+
+\Lem{support} is a direct consequence of the following,
+which is the really the main lemma.
+
+%\begin{lemma} \label{lem:cu} Let $\cU$ denote a connected component (shrub) in the shrub forest of $\cL \setminus \cL'$ and let $r$
+%be the root of $\cU$. Then $\sum_{p \in \cU} \sum_{v \in p} \log |H_v| = O(|H_r| \log |H_r|)$.
+%\end{lemma}
+%
+%We proceed with the setup to prove \Lem{cu}.
+%We focus our attention on some shrub $\cU$.
+%We prove a technical lemma about $\cU$, from which \Lem{cu} follows directly.
+
+\section{Ben's old stuff}
+
+\begin{proof} (part (ii) of \Lem{cu-root}) Let level $\ell$ of $\cU$ denote all nodes at distance $\ell$ from the root $r$.
+These set of these nodes is designated as $\cU_\ell$. We set $\sigma_\ell = \sum_{p \in \cU_\ell} |p|$
+and $\sigma_{\leq \ell} = \sum_{k \leq \ell} \sigma_k$.
+By \Lem{geometric}, for any $p \in \cU_\ell$, $|H_p| \leq (2/3)^\ell |H_r|$.
+
+Since all nodes in $\cU$ represent light paths, $\forall p \in \cU$, $|p| < \sqrt{|H_p|}/100$.
+So $\sigma_\ell < \sum_{p \in \cU_\ell} \sqrt{|H_p|}/100$.
+
+\end{proof}
+
+To prove this, we need to understand how $H_p$ is constructed. The path $p$ has some children
+$q_1, q_2, \ldots, q_k$ in $\pmax(T)$. Furthermore, as a leaf path in $p$, some leaf
+$\ell$ of $T$ in one of the ends of $p$. The heap $H_p$ is essentially the
+union of $H_\ell$ and $H_{q_1}, H_{q_2}, \ldots, H_{q_k}$, with the vertices in $p$
+removed. Note that some of the $q_i$'s are the corresponding $h(q')$ vertices,
+for a child $q'$ in $\cU$.
+
+Let $\mathcal{L} = \mathcal{L}(r)$ be a light shrub of some light path $r$ in a maximum path decomposition $\mathcal{P}(T)$.
+The intuitive idea is that due to the geometric decreasing behavior of light shrubs described in the previous subsection, 
+one can show that for the entire light shrub our algorithm pays roughly the cost due to the root path $r$.  This argument 
+is hard to make directly, so instead we define an alternative quantity for which we can make the argument.
+Specifically, we now define a weight function $w(v)$ for all $v\in \mathcal{L}$, and show that $w(v)=\Theta(|H_v|)$.
+
+For a path $v\in \mathcal{L}$, let $H_v$ denote the heap at the root of $v$, and let $H_{v_f}$ denote the heap 
+at the leaf of $v$.  Let $S_l(v)$ (resp. $S_h(v)$) be the subset of light (resp. heavy) paths adjacent to $v$ in $\pathTreeA(T)$.
+
+We want to understand how the heap size changes as we move around the paths in $\mathcal{L}$.  Specifically, there are 
+three places the vertices in the heap along a light path can come from; adjacent light paths, adjacent heavy paths, and the initial heap at the leaf of the path.  
+Given this observation we now define a weight function which upper bounds the size of the heap at the root of any path in $\mathcal{L}$.
+
+First, for all $v\in \mathcal{L}$ we define the local cost function $c(v) = H_{v_f}+\sum_{i\in S_h(v)} H_i$, this is the addition to the heap at $v$ which 
+was not already charged to descendant light paths.  We now define a weighted version of $\mathcal{L}$ based on these costs.  
+To do so, we first augment $\mathcal{L}$ by attaching one 
+leaf to each vertex in $\mathcal{L}$.  Now for each newly created leaf (which are the only leaves now) assign the cost 
+$c(v)$ defined above, where $v$ is that leaf's corresponding parent.  
+Now we can define a weight function $w(u)$ over any node $u$ in the tree, which is defined to be the total cost of all leaves in its subtree.   
+To simplify notation, we now use $\mathcal{L}$ to refer to this new weighted tree (rather than its unweighted counterpart).
+
+As the weight $w(p)$ includes the sizes of the heaps of all contributing heavy paths and leaves in the subtree rooted at $p$, we have the following.
+
+\begin{claim}
+For any $p\in \mathcal{L}$, $w(p)\geq |H_p|$.  Moreover, for any $v\in p$, $w(p)\geq |H_v|$.
+\end{claim}
+
+%Let $W= \sum_{v\in \mathcal{L}} w(v) \geq \sum_{v\in \mathcal{L}} |H_v|$.  
+%Suppose we could show that conversely $W=O(H_r)$.  Then we would have 
+%$\sum_{v\in \mathcal{L}} H_v = O(H_r)$, which was our original goal.  
+%So proving $W=O(H_r)$ is now our new goal.
+
+We now would like to show conversely that $w(v)= O(|H_v|)$.  
+To do so we first introduce some notation.  
+Let $C\subseteq \mathcal{P}$ be the subset of paths that are leaves in $\mathcal{L}$.
+For any path $v$, let $\mathcal{L}_v$ (resp. $C_v$) be the subset of paths in $\mathcal{L}$ (resp $C$) 
+that are in the subtree of $\mathcal{L}$ which is rooted at $v$.
+
+First, observe that $|H_v|\geq w(v)-\sum_{u\in \mathcal{L}_v} |u|$.  
+This holds by the same argument as in \Lem{adjacent}, namely, $w(v)$ by definition 
+counts all the contributing heavy path heaps and initial leaf heaps that contribute 
+to the heap at $v$ and $\sum_{u\in \mathcal{L}_v} |u|$ is an upper bound on the number 
+of vertices that could have been removed from one of these heaps before reaching the 
+root of $v$ (since a vertex in a heap can only get removed if it appears on one of 
+the paths).
+
+As all paths in $\mathcal{L}$ are light, we thus have the following.
+
+\begin{claim}
+$|H_v|\geq w(v)-\sum_{u\in \mathcal{L}_v} \sqrt{|H_u|}/100 \geq w(v)-\sum_{u\in \mathcal{L}_v} \sqrt{w(u)}/100$
+\end{claim}
+
+By the above claim, if we can prove $\sum_{u\in \mathcal{L}_v} \sqrt{w(u)}/100 \leq w(v)/c$ for some constant $c>1$, 
+then we will have proven that $|H_v| = \Theta(w(v))$.
+
+Recall that \Lem{geometric} told us that that $|H_{u}|\leq (3/5) |H_{u'}|$, if $u'$ is the parent $u$ in $\mathcal{L}$. 
+The following lemma shows that if we had a similar statement in terms of the weights rather than the heap sizes, 
+then we get the desired bound on the sum.
+
+\begin{lemma}
+\label{lem:distribution}
+Let $v$ be a path in $\mathcal{L}$.  Suppose that for any two paths $x,y\in \mathcal{L}_v$ 
+such that $y$ is the parent of $x$, $w(x)\leq (4/5) w(y)$.
+Then $\sum_{u\in \mathcal{P}_v} \sqrt{w(u)}/100 \leq \sum_{c\in C_v} \sqrt{w(c)}/10$
+\end{lemma}
+
+\begin{proof}
+The idea is to recursively redistribute the contribution to the sum of each internal vertex to its children, 
+until all the weight sits on the leaves.  Specifically, let $v$ be some path in $\mathcal{L}$.  Its 
+contribution to the sum is $\sqrt{w(v)}/100$, which we will call the value at $v$.  We redistribute this 
+value to its children, in proportion by the weight of each child.  For example, if $v$ has a child $x$ then 
+the fraction it gets assigned is $w(x)/w(v)$ (recall that by definition $w(v)$ is the sum of the weights of 
+its children and so the total value will get reassigned to the children).
+
+Suppose we keep redistributing the value of $v$ until it reaches the leaves of its subtree in $\mathcal{L}$.  
+Let's see how much gets assigned to a given leaf.  Specifically, let $u_0, u_1, \ldots, u_k$ 
+be a path in $\mathcal{L}$ (of paths from the set $\mathcal{P}_v$), such that $v=u_0$, $u_{i+1}$ 
+is child of $u_i$ for all $i$, and $u_k$ is a leaf of $\mathcal{L}$.
+
+Then the total value from $u_0$ that ends up at $u_k$ is 
+\[
+\frac{\sqrt{w(u_0)}}{100} \pth{ \frac{w(u_1)}{w(u_0)} \frac{w(u_2)}{w(u_1)} \cdots \frac{w(u_k)}{w(u_{k-1})} }
+= \frac{\sqrt{w(u_0)}}{100} \pth{ \frac{w(u_k)}{w(u_0)} } = \frac{w(u_k)}{100\sqrt{w(u_0)}}
+\]
+
+For a given leaf in $\mathcal{L}$, we can calculate the total value reassigned to it from it its ancestors.
+Specifically, let $u_0, u_1, u_2, \dots, u_m$ be a root to leaf path in $\mathcal{L}$ 
+(i.e. $u_0$ is the root $r$ of $\mathcal{L}$).
+
+By assumption from the lemma statement we have $w(u_{i+1})\leq (4/5) w(u_i)$.  Therefore by the above calculation, 
+the total reassigned to $u_m$ is
+
+\[
+ \sum_{i=0}^m \frac{w(u_m)}{100\sqrt{w(u_i)}} \leq \frac{w(u_m)}{100} \sum_{i=0}^m 1 \Big/ \sqrt{\pth{\frac{5}{4}}^iw(u_m)} 
+ \leq \frac{\sqrt{w(u_m)}}{100} \sum_{i=0}^{\infty} \pth{\frac{4}{5}}^{i/2} \leq \frac{\sqrt{w(u_m)}}{10}
+\]
+
+The claim then follows by summing this quantity over the children of $\mathcal{L}$.
+\end{proof}
+
+By combining the above lemma and claim we have the following.
+
+\begin{claim}
+For any path $v\in \mathcal{L}$ if it holds that 
+$w(x)\leq (4/5) w(y)$ for any two paths $x,y\in \mathcal{L}_v$ such that $y$ is the parent of $x$, then
+
+\[
+|H_v|\geq w(v)-\sum_{u\in \mathcal{L}_v} \sqrt{w(u)}/100 \geq w(v)- \sum_{c\in C_v} \sqrt{w(c)}/10 \geq \frac{4}{5} w(v)
+\]
+\end{claim}
+
+\begin{lemma}
+Let $\mathcal{L}$ be the light shrub of some path $r$.  
+Then for any $v\in \mathcal{L}$, we have $|H_v|\geq \frac{4}{5} w(v)$.
+\end{lemma}
+% \begin{proof}
+% We will actually prove the stronger statement that for any $v\in \mathcal{L}$, 
+% we have $|H_v|\geq \frac{4}{5} w(v)$ and $w(v)\leq (4/5) w(p)$ where $p$ is the parent 
+% of $v$ in $\mathcal{L}$.
+% 
+% Suppose otherwise, and let $v$ be the lowest path in $\mathcal{L}$ such that this fails 
+% (i.e. the statement holds for all descendants of $v$ in $\mathcal{L}$).  Then for any 
+% two paths $x,y\in \mathcal{L}_v$ such that $y$ is the parent of $x$, 
+% we have that $w(x)\leq (4/5) w(y)$ and so by \Lem{distribution}, $|H_v|\geq \frac{4}{5} w(v)$.
+% Combining this fact with \Lem{geometric} gives,
+% \[
+% w(v) \leq \frac{5}{4}|H_v| \leq \frac{5}{4} \frac{3}{5} |H_p| \leq \frac{3}{4} w(p).
+% \]
+% 
+% \end{proof}
+%
+\begin{proof}
+By the above claim, we know that $|H_v|\geq \frac{4}{5} w(v)$ if it holds that 
+$w(x)\leq (4/5) w(y)$ for any two paths $x,y\in \mathcal{L}_v$ such that $y$ is the parent of $x$.
+So let $v$ be the lowest path in $\mathcal{L}$ such that this fractional decrease fails to hold, i.e. $w(c)> (4/5) w(v)$ 
+where $c$ is a child of $v$ in $\mathcal{L}$, and conversely all vertices in $\mathcal{L}_c$ 
+the fractional decrease holds.  As the fractional decrease holds for a paths in $\mathcal{L}_c$, 
+we have $|H_c|\geq \frac{4}{5} w(c)$.  Therefore,
+\[
+w(c) \leq \frac{5}{4}|H_c| \leq \frac{5}{4} \frac{3}{5} |H_v| \leq \frac{3}{4} w(v).
+\]
+However, this is a contradiction with the assumption $w(c)> (4/5) w(v)$ , and hence the 
+fractional decrease always holds implying the claim always holds.
+\end{proof}
+
+\begin{corollary}
+\label{cor:equivalent}
+ Let $\mathcal{L}$ be a light shrub of a light path $r$.  Then for any $v\in \mathcal{L}$, $|H_v| \leq w(v) \leq \frac{5}{4} |H_v|$.
+\end{corollary}
+
+Now we are finally ready to make use of the properties we have proven about the weight function we introduced.
+
+\Ben{The constant in the lemmas below, but it is no more than twice that.}
+
+\begin{lemma}
+\label{lem:lightCost}
+ Let $\mathcal{L}$ be a light shrub of a light path $r$.  Let $V$ be the set of heavy paths from the full shrub which are adjacent to paths in $\mathcal{L}$.  
+  Then $\sum_{p\in \mathcal{L}}\sum_{v\in p} \log |H_v| + \sum_{q\in V} |H_q|\log |H_q| \leq \frac{4}{3} |H_r|\log |H_r|$
+\end{lemma}
+\begin{proof}
+\Ben{
+The following is a proof for $\frac{3}{2}+\eps$, rather than $\frac{4}{3}$.  The discrepancy is due to a corrected calculation error.  
+4/3 can be achieved by tightening constants above, and is left in the lemma statement as it is used later.
+}
+
+Recall that in a proper coloring of a binary tree $T$, the color of any vertex must be the color of one of its descendants.  
+Therefore, any vertex in the light path shrub must appear in the heap of either the root of an adjacent 
+path from the full shrub or from the heap of a leaf of a path in $\mathcal{L}$.  
+Also note $|H_r|$ precisely counts sum of the number of vertices from these heaps that do not appear in $\mathcal{L}$.
+As there are no adjacent light paths to a light shrub, the total number of vertices over all paths in $\mathcal{L}$ is at most 
+$m = \sum_{p\in \mathcal{L}} |H_{p_f}| +\sum_{q\in V} |H_q| - |H_r| = w(r)-|H_r|$, where $H_{p_f}$ is the heap at the leaf of $p$.
+
+Using this observation and \Cor{equivalent} we then have, 
+\begin{align*}
+ \sum_{p\in \mathcal{L}}\sum_{v\in p} \log |H_v| + \sum_{q\in V} |H_q|\log |H_q|
+ &\leq \sum_{p\in \mathcal{L}}\sum_{v\in p} \log(w(r)) + \sum_{q\in V} |H_q|\log(w(r))\\
+ &\leq \log(w(r))\pth{ w(r)-|H_r| } + \log(w(r))w(r)\\ 
+ &= (2w(r)-|H_r|)\log(w(r)) \leq \pth{\frac{5}{2}|H_r|-|H_r|}\log\pth{\frac{5}{4}|H_r|}\\
+ &= \frac{3}{2}|H_r|\log\pth{\frac{5}{4}|H_r|}
+\end{align*}
+\end{proof}
 
@@
 
-We define a ``cut" operation on manifolds that cuts along a contour to create
-a manifold with a new boundary. Intuitively, any cut creates two disjoint boundaries, one
-above and one below.
-
-\medskip
-\fbox{
-\begin{minipage}{0.9\textwidth}
-{\bf $\cut(\MM,\phi)$}
-
-\smallskip
-\begin{asparaenum}
-%	\item Insert the vertices of $\phi$ in $\MM$.
-	\item Take an arbitrary edge $e$ whose interior intersect $\phi$.
-	\item Take a point $x$ (resp. $y$) on $e$ at distance $\eps$ above (resp. below) $\phi$.
-	\item Let $\phi_x$ be the contour through $x$, and analogously define $\phi_y$. Insert a vertex
-	at each intersection point of these contours with edges of the manifold. Triangulate all new faces.
-	\item Delete the vertices (and incident edges, faces, etc.) of $\phi$ from the manifold.
-\end{asparaenum}
-\end{minipage}}
-
-\medskip
-This results in a new manifold which is boundary critical if the input manifold was. There are two new disjoint
-boundary faces added. The manifold may be disconnected. Abusing notation, we use $\cut(\MM,\Psi)$
-for a set $\Psi$ of contours to mean the manifold obtained by cutting along all contours in $\Psi$.
-
-We now describe a procedure that converts the Reeb Graph of $\cut(\MM,\phi)$ to the Reeb Graph of $\MM$.
-
-\medskip
-\fbox{
-\begin{minipage}{0.9\textwidth}
-{\bf $\surgery(\MM,\phi)$}
-
-\smallskip
-\begin{asparaenum}
-	\item Let $\MM' = \cut(\MM,\phi)$.
-	\item Construct $\reeb(\MM')$ and let $B, C$ be the vertices corresponding to the new boundaries created
-	in $\MM'$. (One is a minima and the other is maxima.)
-	\item Since $B, C$ are leaves, they each have unique neighbors $B'$ and $C'$, respectively. Insert
-	edge $(B',C')$ and delete $B,C$ to obtain a new graph.
-\end{asparaenum}
-\end{minipage}}
-
-\medskip
-\begin{lemma} \label{lem:surgery} For any regular contour $\phi$, the output of $\surgery(\MM,\phi)$ is $\reeb(\MM)$.
-\end{lemma}
-
-\begin{proof} The only difference between $\reeb(\MM)$ and $\reeb(\cut(\MM,\phi))$ is in the equivalence
-class corresponding to $\phi$. This class is split into two classes in $\cut(\MM,\phi)$, which are joined
-back by $\surgery(\MM,\phi)$.
-\end{proof}
-
-\section{Raining to partition $\MM$} \label{sec:rain}
-
-In this section, we assume that $\MM$ is homeomorphic to $\SSS^2$ with boundaries.
-
-\begin{definition} \label{def:dom} A manifold is \emph{extremum dominant} if there exists
-a maximum (resp. minimum) $x$ such that every non-maximal (resp. non-minimal) \emph{vertex} in the manifold has an increasing (resp. decreasing) path to $x$.
-\end{definition}
-
-We will describe a linear time algorithm that partitions $\MM$ into a set of extremum dominant manifolds.
-The first part of this requires a ``raining" procedure. Start at some point $x \in \MM$ and imagine rain at $x$.
-The water will flow downwards along descending paths and ``wet" all the points encountered.
-
-\begin{definition} \label{def:wet} The set of points $y$ such that there is a descending path from $x$ to $y$
-is denoted by $\wet(x,\MM)$. A point $z$ is on the \emph{interface} of $\wet(x,\MM)$ if every neighborhood of $z$
-has non-trivial intersection with $\wet(x,\MM)$.
-\end{definition}
-
-Note that $\wet(x,\MM)$ (and its interface) can easily be computed in linear time in the size of the wet sub-manifold, using a descending BFS from $x$.
-
-\begin{claim} \label{clm:inter} For any $x$, the interface of $\wet(x,\MM)$ is a set of disjoint, non-regular contours.
-Furthermore, each such contour contains a merge vertex.
-\end{claim}
-
-\begin{proof}
- First observe that any point which in the interface must be wet.  Second, for $y \in \wet(x,\MM)$, all the points in any contour containing $y$ are also in $\wet(x,\MM)$, 
- as one can follow the descending path to $y$ and then walk along the contour.
-
- So let $y$ be a point in the interface.
- For $\eps$ sufficiently small, the points in any $\eps$ neighborhood of $y$ that lie below $y$ must have a descending path from $y$ and hence must be wet.
- Therefore in order for $y$ to have non-trivial intersection with $\wet(x,\MM)$ for any arbitrarily small $\eps$ there must be a dry point which lies 
- strictly above $y$.  In particular, this implies there must be an entire dry contour which lies just above $y$.
-
- However, since $y$ is in the interface, it is also wet.  Therefore for any contour $\phi$ which contains $y$, for any arbitrarily small $\eps$ there is 
- a wet point which lies above $\phi$ and has a descending path of length at most $\eps$ to $\phi$.
- This implies there is some contour which lies just 
- above $y$ which is wet.  So any contour $\phi$ containing $y$ has two contours which lie just above it, one which is dry, and one which is wet.  
- This can only happen if $\phi$ contains a merge vertex, because otherwise all contours which lie just above would be in the same L-homotopy equivalence class and so by 
- \Clm{mono}, if one is wet, all below it are wet.
-
- By similar logic one can argue that of the two contours through a given merge vertex at most one can contain interface points other than the merge vertex, 
- and that if one of the contours does contain such interface points, then the entire contour consists of interface points.
-\end{proof}
-
-% \begin{proof} If $y \in \wet(x,\MM)$, but not in the interface, then any contour containing $y$ is also in $\wet(x,\MM)$.
-% So consider a point $y$ in the interface. We argue that all points in a contour $\phi$ containing $y$
-% are also in the interface. If such a contour is regular, we can find a contour just above it that
-% is also wet. 
-% 
-% If two contours intersect, the intersection point $y$ is critical. We can argue that the only
-% way to reach $y$ from a descending path is reach one of the contours and then follow it to get to $y$.
-% But that would imply the existence of another critical point on this contour, violating the Morse condition.
-% 
-% Let us argue that $y$ is a merge vertex. Consider the descending path from $x$ to $y$. The final edge
-% that reaches $y$ is obviously all wet, and is an increasing edge from $y$. Consider an opposite increasing
-% edge. Take two point in these edges that are infinitesimally higher than $y$ and show that the contours
-% through them as disjoint. So $y$ is a merge.
-% \end{proof}
-
-We will define a \lift{} operation on the interface contours. Consider such a contour $\phi$ containing
-a merge vertex $y$. Take any dry increasing edge incident to $y$, and pick the point $z$ on this edge at height
-$f(y) + \eps$. Let $\lift(y) = \cont(z)$.
-
-\begin{claim} \label{clm:cut-int} Let $\phi$ be an interface contour. Then $\cut(\MM,\lift(\phi))$
-results in two disjoint manifolds, one consisting entirely of dry points.
-\end{claim}
-
-\begin{proof} Note that $\lift(\phi)$ is a closed, dry curve. By the Jordan Curve Theorem,
-$\cut(\MM,\lift(\phi))$ results in two disjoint manifolds. Let $\NN$ be the manifold containing
-the point $x$ that we called $\wet(x,\MM)$ from, and let $\NN'$ be the other manifold. 
-If there exists a wet point $z$ in $\NN'$, then there exists a descending path from $x$ to $z$. 
-As $x$ is in $\NN$ and $z$ is in $\NN'$, again by the Jordan Curve Theorem this path must intersect $\phi$, 
-contradicting the fact that $\phi$ is dry.
-\end{proof}
-
-We now define the main partitioning procedure that cuts up a manifold $\NN$ homeomorphic to $\SSS^2$ with boundaries into a series of extremum
-dominant manifolds. It takes as input the manifold and a maximum $x$. To initialize,
-we begin with $\NN$ set to $\MM$ and $x$ as an arbitrary maximum.
-
-\medskip
-\fbox{
-\begin{minipage}{0.9\textwidth}
-{\bf $\rain(x,\NN)$}
-
-\smallskip
-\begin{compactenum}
-	\item Compute the interface of $\wet(x,\NN)$. 
-	\item If the interface is empty, simply output $\NN$. Otherwise, denote the contours by $\phi_1, \phi_2, \ldots, \phi_k$, and set $\phi'_i = \lift(\phi_i)$.
-	\item Initialize $\NN_1 = \NN$.
-	\item For $i$ from $1$ to $k$: 
-	\begin{compactenum}
-		\item Call $\cut(\phi'_i,\NN_1)$ to get the dry manifold $\LL_i$ and wet manifold $\NN_{i+1}$.
-		\item Let the new boundary of $\LL_i$ be $B_i$. Invert $\LL_i$ so that $B_i$ is a maximum. Recursively
-		call $\rain(B_i,\LL_i)$.
-	\end{compactenum}
-	\item Output $\NN_{k+1}$.
-\end{compactenum}
-\end{minipage}}
-
-\medskip
-For convenience, denote the total output of $\rain(x,\MM)$ by $\MM_1, \MM_2, \ldots, \MM_r$.
-
-\begin{lemma} \label{lem:rain-1} Each output manifold $\MM_i$ is extremum dominant.
-\end{lemma}
-
-\begin{proof} Consider a call to $\rain(x,\NN)$. If the interface is empty, then all of $\NN$
-is in $\wet(x,\NN)$, so $\NN$ is trivially extremum dominant. So suppose the interface
-is non-empty and consists of $\phi_1, \phi_2, \ldots, \phi_k$ (as denoted in the procedure).
-By repeated applications of \Clm{cut-int}, $\NN_{k+1}$ consists of wet points. 
-Consider $\wet(x,\NN_{k+1})$. The interface must exactly be $\phi_1, \phi_2, \ldots, \phi_k$.
-So the only dry vertices are those in the boundaries $B_1, B_2, \ldots, B_k$. But these
-boundaries are maxima.
-\end{proof}
-
-The procedure $\rain$ creates new boundary vertices on the edges it cuts each time it call the 
-subroutine $\cut$.  The key to understanding the running time of this procedure is to bound the 
-number of newly created vertices, for which we have the following lemma.
-
-\begin{lemma}\label{lem:new-verts}
-The total number of boundary vertices created by a call to $\rain(x, \NN)$ is $O(n)$, where $n=|\NN|$
-\end{lemma}
-\begin{proof}
-As the number of edges is $O(n)$, arguing that there is at most one boundary vertex created along any given edge will imply the claim.
-Consider some edge $e$ and take the first time a boundary vertex is created in $e$.
-This happens when some contour $\phi$ intersects $e$. On applying $\cut(\NN,\lift(\phi))$,
-consider the dry manifold $\LL$ and wet manifold $\MM$ constructed. The edge $e$ is split into two parts, the dry and wet part.
-The wet part is in $\MM$ and will never be involved in any recursive call.
-
-The dry part of $e$ (call it $e'$) needs to be considered.
-A boundary vertex $v$ (an new endpoint of $e$) is created, which is part of the boundary $B$ in $\LL$,
-such that $\rain(B,\LL)$ is called to the inverted $\LL$. In this call, since the water starts from $v$,
-all of $e'$ becomes wet. So there is no further interface contour that can intersect this portion of $e'$.
-\end{proof}
-
-As each original vertex in $\NN$ appears in only one of the sub-manifolds output by $\rain(x,\NN)$, we have the following corollary.
-
-\begin{corollary}
-The total size of a sub-manifolds output by $\rain(x,\MM)$ is $O(n)$.
-\end{corollary}
-
-\begin{corollary} \label{cor:rain-time} The total running time of $\rain(x,\MM)$ is $O(n)$.
-\end{corollary}
-
-\begin{proof} Consider some invocation to $\rain(x,\MM)$. The only non-trivial work done in $\rain$ is by the calls to $\wet$.
-We know that each invocation to $\wet$ runs in time linear 
-to the portion of the manifold which is wet (including the new boundary vertices created).  
-As the output of $\rain$ is precisely the distinct sub-manifolds wet by each distinct call to $\wet$, the claim follows.
-\end{proof}
-
-Consider the set of manifolds output by $\rain(x,\MM)$, there is a natural tree ordering defined over these manifolds where the 
-root corresponds the first manifold that was wet when we rained from $x$, and its children correspond to the recursive calls 
-(i.e. where we rained from the interface contours of the first raining).  Call this the \emph{Manifold Ordering Tree}.
-$\rain$ can easily be modified to also output this tree (without affecting the asymptotic running time).
-
-%Consider one of the manifolds $\MM_i$ output by $\rain$.  This manifold has a set of boundary maxima/minima that are the result 
-%of cutting along interface contours.  For each such contour there is exactly one other output manifold $\MM_j$ for which 
-%the contour appears as a boundary maxima/minima.
-
-\begin{claim} \label{clm:rain-reeb} Let $\reeb(\MM_1), \reeb(\MM_2), \ldots, \reeb(\MM_r)$, 
-be the output of $\rain(x,\MM)$. Suppose that for each reeb graph we know its corresponding vertex in the Manifold Ordering Tree. 
-Then $\reeb(\MM)$ can be computed in $O(n)$ time.
-\end{claim}
-
-\begin{proof}
-We will walk through the Manifold Ordering Tree in a leaf first ordering.  Each time we visit a node we connect its reeb graph to the 
-reeb graph of its children in the tree using the $\surgery$ procedure.
-This takes $O(n)$ time overall by \Lem{new-verts} and since $\surgery$ takes constant time (given the appropriate vertices in each reeb graph).
-The correctness follows by induction and  \Lem{surgery}.
-\end{proof}
-
-\section{Painting to compute contour trees} \label{sec:paint}
-The previous section allows us to restrict attention to extremum dominant manifolds.
-We will orient so that the extremum in question is a \emph{minimum}.
-We will fix such a manifold $\MM$, with the dominant minimum $u$.
-
-We begin by associating a color with each maximum. If the set of maxima is $S$,
-this is simply an arbitrary bijection $\chi:S \mapsto [|S|]$. Imagine there being a large
-can of paint of color $\chi(y)$ at maxima $y$. We will spill different paint from each maximum and watch it flow down.
-This is analogous to the raining in the previous section, but paint is a much more viscous liquid.
-\emph{So paint only flows down edges, and it does not color the interior of faces.} Furthermore, paints
-do not mix, so eventually, every vertex and edge of $\MM$ gets a unique color.
-%
-\begin{definition} \label{def:paint} A  \emph{painting} of $\MM$ is a map $\chi:S \cup E \mapsto [|S|]$
-with the following properties. (Remember that $G(\MM) = (V,E)$.)
-\begin{asparaitem}
-	\item The restriction $\chi:S \mapsto [|S|]$ is a bijection.
-	\item Consider an edge $e$. There exists a descending path from some maximum $x$ to $e$
-	consisting of edges in $E$, such that all edges along this path have the same color as $x$. (The maximum $x$
-	is well-defined to be the unique maximum with the same color as $e$.)
-\end{asparaitem}
-\end{definition}
-%
-Note that paintings are not unique. It is straightforward to construct a painting of $\MM$ in $O(n)$ time, using a descending BFS from each maximum that does not explore previously colored edges.
-
-\subsection{The data structures} \label{sec:struct}
-
-\begin{definition} \label{def:color-set} Fix a painting $\chi$, and a color $c$.
-A critical point $x$ is \emph{touched by $c$} if $x$ is incident to a $c$-colored
-increasing edge. The set of non-extremal critical points touched by $c$ is denoted $\touch(c)$.
-For $x$, $\col(x)$ is the set of colors that touch $x$.
-\end{definition}
-%
-%We will store each set $\touch(c)$ in a binomial heap, keyed by the heights. 
-Abusing notation, $\touch(c)$ refers
-both to the set and the data structure used to store it.
-
-\medskip
-\noindent
-{\bf The binomial heaps $\touch(c)$:} For each color $c$, we store a binomial
-heap $T(c)$ of non-extremal critical points, keyed by vertex heights.
-
-\medskip
-\noindent
-{\bf The union-find data structure on colors:} We will repeatedly perform unions
-of classes of colors, and this will be maintained as a standard union-find data structure.
-For any color $c$, $\rep(c)$ denotes the representative of its class. We initialize
-by simply setting $\rep(c) = c$ for all colors.
-
-\medskip
-\noindent
-{\bf The stack $\st$:} This consists of non-extremal critical points with the following structure.
-Each point $x \in \st$ has an associated subset of $\col(x)$, denoted $\mcol(x)$.
-Both $\mcol(x)$ and its complement are stored as hash table. So lookups, inserts, and deletes
-are in these sets are all constant time operations. The stack is guaranteed to satisfy 
-the following invariants.
-\begin{asparaitem}
-	\item For every $x \in \st$: For every $c \in \mcol(x)$, $x$ is the highest element
-	in $T(c)$. Furthermore, $c = \rep(c)$.
-	\item Consider $x, y \in \st$ such that $y$ was pushed on $x$. There exists $c \in \col(x) \setminus
-	\mcol(x)$ such that $x$ is not highest in $T(c)$ but $y$ is highest in $T(c)$.
-\end{asparaitem}
-
-\medskip
-\noindent
-{\bf Highest points $\h(c)$:} For each color $c$, we maintain a special critical point $\h(c)$ of this color.
-It is supposed to denote the ``highest" point of color $c$, but the notion of highest evolves
-with the algorithm.
-
-\subsection{Updating the stack} \label{sec:stack}
-
-We have a simple, but critical, procedure that updates the stack $\st$ by pushing
-points on to it.
-
-\medskip
-\fbox{
-\begin{minipage}{0.9\textwidth}
-{\bf $\update(\st)$}
-
-\smallskip
-\begin{compactenum}
-	\item If $K$ is empty, pick arbitrary $c$. Let $x$ be the highest point in $T(c)$.
-	Push $x$ onto stack and set $\mcol(x) = \{c\}$.
-	\item Let $h$ be head of $\st$.
-	\item While $\col(h) \setminus \mcol(h)$ non-empty:
-	\begin{compactenum}
-		\item Pick some $c \in \col(h) \setminus \mcol(h)$. Replace it by $c' = \rep(c)$.
-		\item Obtain highest point $x$ in $T(c')$. 
-		\item If $x = h$, put $c'$ in $\mcol(h)$. Otherwise ($x \neq h$), push $x$ onto stack
-		and add $c'$ to $\mcol(x)$.
-		\item Let $h$ be new head of $\st$.
-	\end{compactenum}
-\end{compactenum}
-\end{minipage}}
-
-\medskip
-When $\update(\st)$ terminates, the head $h$ has the following property. For every $c \in \col(h)$,
-$h$ is the highest point in $\touch(c)$.
-
-\begin{claim} \label{clm:cont} Consider an increasing path from $h$ to some maxima $x$.
-For any $y$ in this path, $\cont(y)$ only contains points of color $c(x)$.
-Furthermore, for all vertices on $\cont(y)$, their increasing edges are colored $c(x)$.
-\end{claim}
-
-\begin{proof} Consider this increasing path, and think of moving $y$ continuously downwards from $x$
-to $h$. Initially (when $y = x$), the lemma is trivially true. Consider the first $y$
-when the condition fails. We have $\cont(y)$ only containing points of color $c(x)$,
-but there is a vertex $z \in \cont(y)$ that has increasing edges of a different color $c'$.
-All the increasing edges of $z$ that are colored $c(x)$ are contiguous (since we can construct
-a contour intersecting all of them). Thus, there are two disjoint increasing neighborhoods for $z$,
-and $z$ is a saddle point. But that contradicts the fact that $h$ is the highest saddle point
-of color $c(x)$.
-
-So, as $y$ decreases, for all vertices on $\cont(y)$, their increasing edges are colored $c(x)$.
-This also means that $\cont(y)$ only contains points of color $c(x)$.
-\end{proof}
-
-We have important lemmas characterizing $|\col(h)|$.
-
-\begin{lemma} \label{lem:mono} If $h$ is a merge vertex, then $|\col(h)| > 1$.
-\end{lemma}
-
-\begin{proof} Suppose not. Consider the edges incident to $h$ in circular order. There exist (at least) $2$
-disjoint contiguous stretches of edges that are increasing. 
-All these edges have the same color $c$, associated with some maximum $x$. Consider two neighbors $y,z$
-that lie in different stretches. By the definition of a painting, there exist descending paths (along
-edges) from $x$ to $y$ and $x$ to $z$. We can construct a Jordan curve $C$ from $x \rightarrow y
-\rightarrow h \rightarrow z \rightarrow x$. All points in this curve except $h$ have 
-heights larger than $f(h)$. By the Jordan Curve Theorem, this partitions $\MM$ into
-an ``inside" and an ``outside".
-
-One of the lower neighbors of $h$ (say $p$) is inside, and another (say $q$) is outside.
-Because $h$ is a merge, neither of these can a minimum other than the dominant minimum $u$.
-Hence, both $p$ and $q$ have descending paths to $u$. So there is a path from $p$
-to $q$ where all heights are strictly less than $f(h)$. But this path must intersect $C$.
-Contradiction.
-\end{proof}
-
-\begin{lemma} \label{lem:col} If $h$ is a merge vertex, $|\col(h)| = 2$.
-If $h$ is a split vertex, $|\col(h)| = 1$.
-\end{lemma}
-
-\begin{proof} Consider one contiguous sequence of increasing edges from $h$.
-There exists a contour intersecting all of these edges. By \Clm{cont}, they all
-must have the same color. There are two such contiguous sequences.
-
-If $h$ is a merge vertex, then $|\col(h)| = 2$ by \Lem{mono}. If $h$ is a split vertex,
-there is a single contour passing through all increasing edges. By \Clm{cont}, they
-all have the same color.
-\end{proof}
-
-\subsection{Constructing $\reeb(\MM)$} \label{sec:const}
-
-The main construction procedure is $\onestep(\MM)$. It pieces together $\reeb(\MM)$
-incrementally.
-
-\medskip
-\fbox{
-\begin{minipage}{0.9\textwidth}
-{\bf $\onestep(\MM)$}
-
-\smallskip
-\begin{compactenum}
-	\item Run $\update(\st)$. Pop $\st$ to get $h$.
-	\item Suppose $\mcol(h) = \{c\}$ (so $h$ is split). 
-	\begin{compactenum}
-		\item Connect $h$ (in $\reeb(\MM)$) to $\h(c)$ and the unique minimum corresponding to split $h$. 
-		\item Delete $h$ from $T(c)$, set $\h(c) = h$.
-		\item End procedure.
-	\end{compactenum}
-	\item Let $\mcol(h) = \{c_1, c_2\}$ (so $h$ is merge).
-	\item Connect $h$ in $\reeb(\MM)$ to $\h(c_1)$ and $\h(c_2)$.
-	\item Delete all copies of $h$ from $T(c_1)$ and $T(c_2)$.
-	\item Perform union of colors $c_1$ and $c_2$ (denote merged color as $c$). Merge heaps
-	to get $T(c) = T(c_1) \cup T(c_2)$.
-	\item Set $\h(c) = h$.	
-\end{compactenum}
-\end{minipage}}
-
-\medskip
-\begin{theorem} \label{thm:piece} Consider the very first call to $\onestep(\MM)$. Let $y$ be a point
-infinitesimally below $h$ on a decreasing edge. Consider $\cut(\MM,\cont(y)) = \{\MM_u,\MM_\ell\}$ (the upper and lower manifolds). Then $\onestep(\MM)$
-outputs $\reeb(\MM_u)$ and implicitly gives a valid painting of $\MM_\ell$.
-\end{theorem}
-
-We have an important lemma that helps to bound the running time.
-
-\begin{lemma} \label{lem:tree} For every color $c$, all points in $T(c)$ lie along a monotone path in 
-$\reeb(\MM)$.
-\end{lemma}
-
-\begin{proof} Prove by induction of the number of edges in $\reeb(\MM)$. Consider the head $h$ of $\st$ in the first call to $\onestep(\MM)$,
-and the contour $\cont(y)$ as defined in \Thm{piece}. Take $\cut(\MM,\cont(y)) = \{\MM_u,\MM_\ell\}$. Apply induction
-hypothesis on $\MM_\ell$, and use that to argue about $T(c_1)$ and $T(c_2)$ in the original manifold.
-\end{proof}
-
-\pagebreak
-\newcommand{\pathTree}{P_{\mathcal{S}}}
-\newcommand{\pathTreeA}{\mathcal{P}_{\mathcal{S}}}
-\section{Leaf assignments and path decompositions}
-
-In this section, we set up a framework to analyze the time taken to compute a merge tree.
-
-\begin{definition}
- We say that a contour tree of a $2$-manifold is a \emph{merge tree} if it has a well defined root vertex such that height values 
- strictly increase along any root to leaf path.
-\end{definition}
-
-\begin{observation}
- The contour tree of a minimum dominant manifold (see \Def{dom}) is close to a merge tree.  Specifically, there is a root such that on any root 
- to leaf path height values strictly increase, with the only exception being potentially at the leaves themselves.  As such, up to constant 
- factors, any lower bound for computing the merge tree is a lower bound for computing the contour tree of a minimum dominant
- manifold (i.e. intuitively just clip off the leaves or shift them to lie above their parent).
-\end{observation}
-
-Any tree $T$ considered in this section will be a rooted full binary tree\footnote{Note 
-that technically the trees considered should have a leaf vertex hanging below the root of this 
-full binary tree in order to represent the global minimum of the surface.  This vertex is 
-(safely) ignored to simplify the presentation.} where the height of a vertex is its distance 
-from the root $r$.  As such, the children of a vertex $v\in T$ are the adjacent vertices of larger height
-(and $v$ is the parent of such vertices).  Then the subtree rooted at $v$, denoted $T_v$ consists of the graph induced on all 
-vertices which are descendants of $v$ (including $v$ itself).  For two vertices $v$ and $w$ in $T$ let $d(v,w)$ denote the 
-length of the path between $v$ and $w$.
-We use $A(v)$ to denote the set of ancestors of $v$.
-For set of nodes $U$, $A(U) = \bigcup_{u \in U} A(u)$.
-
-\begin{definition}
- A \emph{leaf assignment} $\chi$ of a tree $T$ assigns \emph{two} distinct leaves in each internal vertex $v$,
- one from the left child and one from the right child subtree of $v$.
-\end{definition}
-
-For a vertex $v\in T$, we use $H_v$ to denote the \emph{heap} at $v$.
-Formally, $H_v = \{u | u \in A(v), \chi(u) \cap L(T_v) \neq \emptyset\}$.
-In words, $H_v$ is the set of ancestors of $v$ which are colored by some leaf in $T_v$.
-The following lemma should justify these technical definitions.
-
-\begin{lemma}
-The running time of our algorithm is $O(\sum_{v \in T} \log |H_v|)$.
-% where $T$ is the merge tree of the given input, 
-%and $\chi$ is any valid coloring of $T$.  (maybe add one in the definition of $H_v$ so that $\log |H_v|$ is always $>1$).
-\end{lemma}
-
-\begin{proof} \Sesh{add}
-
-\end{proof}
-
-%vertices along the path from the root to $v$ such that one of the two colors of each vertex is in $L(T_v)$.
-%We will refer to $H_v$ as the \emph{heap} at $v$.
-%Then define the \emph{algorithm} cost function as $f_{alg}(\chi(T)) = \sum_{v\in T} \log |H_v|$.
-%
-%\begin{lemma}
-%\label{lem:algCost}
-%NEEDS TO BE WRITTEN:
-%Here we need a lemma showing that the running time of our algorithm is $O(f_{alg}(\chi(T)))$, where $T$ is the merge tree of the given input, 
-%and $\chi$ is any valid coloring of $T$.  (maybe add one in the definition of $H_v$ so that $\log |H_v|$ is always $>1$).
-%\end{lemma}
-%
-%Now that we have a simple expression upper bounding the running time of our algorithm, we spend the remainder 
-%of this section proving that this running time is optimal.
-
-We introduce the concept of path decompositions. A \emph{leaf path} is simply a contiguous path in $T$
-containing a leaf, which is monotonic in the height values of its vertices.
-% Then in the following two subsections 
-%we relate path decomposition costs (defined below) to the running time of our algorithm using $\Lem{algCost}$.
-
-\begin{definition}
-\label{def:path} A \emph{path decomposition}, $P(T)$, is a partition of $T$ into 
- a set of vertex disjoint leaf paths.  
-% 
-% The \emph{cost} of $P(T)$ is $\cost(P(T)) = \sum_{p\in P(T)} |p| \log |p|$. (We use $|p|$ to denote the number of vertices in $p$.)
-\end{definition}
-
-Note that the number of paths in $P(T)$ must equal the number of leaves in $T$.
-Fix $\chi$, which fixes all the heaps $H_v$. We choose a special path decomposition that is best defined as a subset of edges in $T$ such that each internal
-vertex has degree at most $2$. This naturally gives a path decomposition.
-For each internal vertex $v\in T$, add the edge from $v$ to $\arg \max_{v_l, v_r} \{|H_{v_l}|, |H_{v_r}|\}$ 
- where $v_l$ and $v_r$ are the children of $v$ (if $|H_{v_l}|=|H_{v_r}|$ then pick one arbitrarily).
-This is called the \emph{maximum} path decomposition, denoted by $\pmax(T)$.
-
-Our main theorem is the following. We use $|p|$ to denote the number of vertices in $p$.
-
-\begin{theorem} \label{thm:runtime} $\sum_{v \in T} \log |H_v| = O(\sum_{p\in \pmax(T)} |p| \log |p|)$.
-\end{theorem}
-
-\subsection{Shrubs, heavy paths, and light paths}
-
-The paths in $P(T)$ naturally define a tree\footnote{Please excuse the 
-overloading of the term 'tree', it is the most natural term to use here.} of their own.  Specifically, in the original 
-tree $T$ contract each path down to its root.  Call the resulting tree the \emph{shrub} of $T$ corresponding to 
-the path decomposition $P(T)$. Abusing notation, we simply use $P(T)$ to denote the shrub.
-As a result, we use terms like `parent', `child', `sibling', etc. for paths as well.
-The shrub gives a handle on the heaps of a path.
-We use $b(p)$ to denote the \emph{base} of the path, which is vertex in $p$ closest
-to root of $T$. We use $\ell(p)$ to denote the leaf in $p$.
-We use $H_p$ to denote the $H_{b(p)}$.
-% and denote it by $\pathTree(T)$.
-
-\begin{lemma}
-\label{lem:adjacent}
- Let $p$ be any path in $P(T)$ and let $\{q_1, \dots q_k\}$ be the children on $p$.
-Then $H_{\ell(p)} + \sum_{i=1}^k |H_{q_i}| \leq |H_p|+2|p|$. 
-\end{lemma}
-
-\begin{proof} For convenience, denote $H_i = H_{q_i}$ and $H_0 = H_{\ell(p)}$.
-Consider $v \in \bigcup_{i} H_i$ that lies below $b(p)$ in $T$. 
-Note that such a vertex has only one of its two colors in $L(b(p))$.  
-Since the colors tracked by $H_i$ and $H_j$ for $i\neq j$ are disjoint, 
-such a vertex can appear in only one of the $H_i$'s.
-On the other hand, a vertex $u\in p$ can appear in more than one $H_i$, 
-but since any vertex has exactly two colors it can appear in at most two 
-such heaps.  Hence, $\sum_i |H_i| \leq |H_p| + 2|p|$.
-\ignore{
- First observe that for all $i$, $r_{i}$ lies above $r_p$ on some root to leaf path. 
- Let $H_i'$ be the subset of $H_i$ that lies below $r_p$.  Observe that $H_i'\subseteq H_p$, as $L(T_{r_i})\subseteq L(T_{r_p})$.
-
- Now for any $i\neq j$, $H_i \cap H_j = \emptyset$, as $r_{i}$ and $r_{j}$ are not on the same root to leaf path (i.e. $L(T_{r_i}) \cap L(T_{r_j}) = \emptyset$).
- In particular, $(H_i\setminus H_i') \cap (H_j \setminus H_j') = \emptyset$, and so $\sum_{i=0}^k |H_i\setminus H_i'| \leq |p|$ 
- as $H_i\setminus H_i'$ is precisely the subset of $H_i$ that lies on $p$.
- Moreover, for any $i\neq j$, $ H_i' \cap H_j' = \emptyset$, and so since $H_i'\subseteq H_p$ we have $\sum_{i=0}^k |H_i'|\leq |H_p|$.
- Putting these together gives
- \[
-  \sum_{i=0}^k |H_i| = \sum_{i=0}^k |H_i'| + \sum_{i=0}^k |H_i\setminus H_i'| \leq |H_p|+|p|.
- \]
-}
-\end{proof}
-
-We wish to prove $\sum_{v\in T} \log |H_v| = O(\sum_{p\in P} |p| \log |p|)$. 
-The simplest approach is to prove $\forall p\in P$, $\sum_{v\in p} \log |H_v|= O(|p|\log|p|)$.
-This is unfortunately not true, which is why we divide paths into two categories.
-
-\begin{definition}
- For $p\in P(T)$, $p$ is \emph{light} if $|p| < \sqrt{|H_{p}|}/100$, and \emph{heavy} otherwise.
-\end{definition}
-
-\ignore{
-\begin{observation}
-\label{obs:decrease}
- Let $v$ be a vertex in $T$ and let $w$ be its parent.  Then $|H_w| \geq |H_v| -1$, as $L(T_v) \subseteq L(T_w)$ and the path from 
- $w$ to the root has one less vertex than the path from $v$ to the root.
- In particular, we have the following more general property.
- Let $v$ and $u$ be any two vertices in the same root to leaf path of $T$, such that $v$ is a descendant of $u$.  
- Then $|H_v| \leq |H_u| + d(u,v)$. 
-\end{observation}
-
-The following can be thought of as a generalization of the above observation, and will be useful in later sections.
-}
-
-The following lemma demonstrates that heavy paths can ``pay'' for themselves.
-
-\begin{lemma}
-\label{lem:pathBounds} If $p$ is heavy, $\sum_{v\in p} \log |H_v| = O(|p| \log |p|)$.
-If $p$ is light, $\sum_{v\in p} \log |H_v| = O(|H_p| \log |H_p|)$.
-% Let T be a binary tree and let $\chi(T)$ and $P(T)$ be a valid coloring and path decomposition of $T$, respectively.
-% Then, 
-% \begin{compactenum}
-%  \item for any heavy path $p\in P(T)$, $\sum_{v\in p} \log |H_v| = O(|p| \log |p|)$.
-%  \item for any light path $p\in P(T)$, $\sum_{v\in p} \log |H_v| = O(|p| \log |H_{p}|)$.
-% \end{compactenum}
-\end{lemma}
-\begin{proof} \ignore{
- By $\Obs{decrease}$ we know that for any vertex $v\in p$, $|H_v|\leq |H_{p}| + |p|$ (as $v$ is a descendant of $r_p$ along $p$).
-}
-For $v \in p$, $|H_v|\leq |H_{p}| + |p|$ (as $v$ is a descendant of $b(p)$ along $p$).
-Hence, $\sum_{v\in p} \log |H_v| \leq \sum_{v\in p} \log(|H_{p}| + |p|)  = |p| \log (|H_{p}| + |p|)$.
- If $p$ is a heavy path, then $|p| \log (|H_{p}| + |p|) = O(|p| \log |p|)$. If $p$ is light, then 
- $|p| \log (|H_{p}| + |p|) = O(|p| \log |H_{p}| )$. For light paths, $|p| = O(|H_p|)$.
-\end{proof}
-
-\ignore{
-For a light path $p$ we can think of the quantity $|p|(\log |H_p| - \log |p|)$ as the excess of $p$, i.e. what was not ``paid for'' locally by $p$.
-Our eventual goal will be to show that by choosing the right path decomposition, this excess 
-can be recharged to heavy paths. In the next couple sections we introduce the appropriate path decomposition and prove 
-some useful facts about it.  Then given these facts, in \Sec{} we are able to make the recharging argument.
-}
-
-There are some light paths that we can also ``pay" for.  Consider any light path $p$ in the shrub.  We will refer to the 
-\emph{heavy support chain} of $p$ as the heavy ancestors of $p$ in the shrub which have a path to $p$ which does not use any 
-light path (i.e. it is a chain of paths adjacent to $p$).
-
-\begin{definition} \label{def:support} A light path $p$ is \emph{supported} if
-at least $|H_p|/100$ vertices $v$ in $H_p$ lie in paths in the heavy support chain of $p$.
-%
-%
-% there exists
-%a contiguous (from $p$) chain of ancestors $q_1, q_2, \ldots, q_k$ (in $P(T)$) such that $|H_{q_k}| \leq 99|H_p|/100$
-%and all $q_i$'s are heavy.
-\end{definition}
-
-Let $\cL$ be the set of light paths, $\cL'$ be the set of supported light paths, and $\cH$ be the set of heavy paths
-given by $\pmax(T)$.
-We now construct the shrub of unsupported light paths. Consider $p \in \cL \setminus \cL'$,
-and traverse the chain of ancestors from $p$. Eventually, we must reach another light path $q$.
-(If not, we have reached the root $r$ of $\pmax(T)$. Hence, $p$ is supported.)
-Insert edge from $p$ to $q$, so $q$ is the parent of $p$ in $\cU$. This construction leads
-to the shrub forest of $\cL \setminus \cL'$, where all the roots are supported light paths, 
-and the remaining nodes are the unsupported light paths.
-
-Most of the work goes into proving the following technical lemma.
-
-\begin{lemma} \label{lem:cu-root} Let $\cU$ denote a connected component (shrub) in the shrub forest of $\cL \setminus \cL'$ and let $r$
-be the root of $\cU$. (i) For any $v \in p$ such that $p \in \cU$,
-$|H_v| = O(|H_r|)$. (ii) $\sum_{p \in \cU} |p| = O(|H_r|)$.
-\end{lemma}
-
-We split the remaining argument into two subsections. We first prove \Thm{runtime} from \Lem{cu-root},
-which involves routine calculations. Then we prove \Lem{cu-root}, where the interesting work happens.
-
-\subsection{Proving \Thm{runtime}} \label{sec:thm-runtime}
-
-We split the summation into heavy, light, and unsupported light paths.
-\begin{eqnarray*} 
-\sum_{p \in \cL} \sum_{v \in p} \log |H_v| & = & \sum_{p \in \cL \setminus \cL'} \sum_{v \in p} \log |H_v| + \sum_{p \in \cL'} \sum_{v \in p} \log |H_v| + \sum_{p \in \cH} \sum_{v \in p} \log |H_v|
-%& = & \sum_{p \in \cL \setminus \cL'} \sum_{v \in p} \log |H_v| + O(\sum_{p \in \cL'} |p|\log |H_p|) = \sum_{p \in \cL \setminus \cL'} \sum_{v \in p} \log |H_v| + O(\sum_{p \in \cL'} |H_p|\log |H_p|)
-\end{eqnarray*}
-The last term can be bounded by $O(\sum_{p \in \pmax(T)} |p|\log |p|)$, by \Lem{pathBounds}.
-The second term can be bounded by $O(\sum_{p \in \cL'} |H_p| \log |H_p|)$, by \Lem{pathBounds} again.
-The following claim shows that this in turn is at most the last term.
-
-%We have a claim that also bounds heap sizes for special set of paths.
-
-\begin{claim} \label{clm:above} $\sum_{p \in \cL'} |H_p| \log |H_p| = O(\sum_{q \in \cH} \sum_{v \in q} \log |H_v|)$.
-\end{claim}
-
-\begin{proof} 
-Pick $p \in \cL'$. 
-%Let the first $|H_p|/200$ vertices the lie in heavy paths be denoted as $H'_p$. 
-%Observe that for any $v \in H'_p$, 
-%We have $|H_p| \log |H_p| = O(\sum_{v \in H'_p} \log |H_p|)$
-%$= O(\sum_{v \in H'_p} \log |H_v|)$. Since all the vertices lie in heavy paths,
-%we can write this as $O(\sum_{q \in \cH} \sum_{v \in H'_p \cap q} \log |H_v|)$.
-%Summing over all $p$, the expression is $\sum_{q \in \cH} \sum_{p \in \cL'} \sum_{v \in H'_p \cap q} \log |H_v|$.
-%
-%
-As we traverse the heavy support chain of $p$, there
-are at least $|H_p|/100$ vertices of $H_p$ that lie in these paths. These are encountered in
-a fixed order. Let $H'_p$ be the first $|H_p|/200$ of these vertices. When $v \in H'_p$
-is encountered, there are $|H_p|/200$ vertices of $H_p$ not yet encountered. Hence,
-$|H_v| \geq |H_p|/200$. Hence, $|H_p|\log |H_p| = O(\sum_{v \in H'_p} \log |H_v|)$.
-Since all the vertices lie in heavy paths,
-we can write this as $O(\sum_{q \in \cH} \sum_{v \in H'_p \cap q} \log |H_v|)$.
-Summing over all $p$, the expression is $\sum_{q \in \cH} \sum_{p \in \cL'} \sum_{v \in H'_p \cap q} \log |H_v|$.
-%and $\sum_{p \in \cL'} |H_p|\log |H_p| = O(\sum_{p \in \cL'} \sum_{v \in H'_p} \log |H_v|)$.
-
-Consider any $v \in H'_p$.  Let $S$ be the set of paths $\widetilde{p}\in \cL'$ such that 
-$v\in H'_{\widetilde{p}}$.  We now show $|S|\leq 2$ (i.e. it contains at most one path other than $p$).
-First observe that any two paths in $S$ must be unrelated (i.e. $S$ is an anti-chain), 
-since paths which have an ancestor-descendant relationship have disjoint heavy support chains.  
-However, any vertex $v$ receives exactly one color from each of its two subtrees (in $T$), and therefore $|S|\leq 2$ since any two 
-paths which share descendant leaves in $T$ (i.e. their heaps are tracking the same color) must have an ancestor-descendant relationship.
-
-In other words, any $\log |H_v|$ appears at most twice in the above triple summation.
-Hence, we can bound it by $O(\sum_{q \in \cH} \sum_{v \in q} \log |H_v|)$.
-%
-%Pick $q \in Q$ and some $v \in H_q$. As we traverse up the tree passing
-%through $A(q)$, we must encounter some $p \in A(q)$ such that $v \in p$. Furthermore,
-%there is an order in which these vertices (of $H_q$) occur. Let $H'$ be the first half
-%of vertices that occur. For all $v \in H'$, $|H_v| \geq |H_q|/2$. Hence,
-%$|H_q| \log |H_q| \leq 4 \sum_{v \in H'} \log |H_v| \leq 4 \sum_{v \in H_q} |H_v|$.
-%So $\sum_{q \in Q} |H_q| \log |H_q| \leq 4 \sum_{q \in Q} \sum_{v \in H_q} |H_v|$.
-%
-%Since nodes in $Q$ are incomparable, all the $H_q$'s (for $q \in Q$) are disjoint.
-%Each such $v$ must occur in some path $p \in A(Q)$. Thus,
-%$\sum_{q \in Q} |H_q| \log |H_q| \leq 4 \sum_{p \in A(Q)} \sum_{v \in p} |H_v|$.
-%We apply the bound of \Lem{pathBounds} to complete the proof.
-\end{proof}
-
-The first term (unsupported light paths) can be charged to the second term (supported light paths).
-This is where the critical \Lem{cu-root} plays a role.
-
-\begin{claim} \label{clm:first} $\sum_{p \in \cL \setminus \cL'} \sum_{v \in p} \log |H_v| = O(\sum_{p \in \cL'} |H_p| \log |H_p|)$.
-\end{claim}
-
-\begin{proof} Let $\cU$ denote a connected component of the shrub forest.
-We have $\sum_{p \in \cL \setminus \cL'} \sum_{v \in p} \log |H_v| \leq \sum_{\cU} \sum_{p \in \cU} \sum_{v \in p} \log |H_v|$.
-By \Lem{cu-root}, $|H_v| = O(|H_r|)$, where $r$ is the root of $\cU$.
-Furthermore, $\sum_{p \in \cU} |p| = O(|H_r|)$. 
-We have $\sum_{p \in \cU} \sum_{v \in p} \log |H_v| = O((\log |H_r|) \sum_{p \in \cU} |p|) = O(|H_r|\log |H_r|)$.
-We sum this over all $\cU$ in the shrub forest, and note that roots in the shrub forest are supported light paths.
-\end{proof}
-
-\ignore{
-Now, for the critical piece of the approach. Focus on $\pmax(T)$. Let $L$ be the set of light paths, 
-and $\hat{L} \subseteq L$ be the set of paths such that no ancestor of $p \in L'$
-is light. Obviously, $\hat{L}$ is an anti-chain in $P(T)$.
-
-\begin{lemma} \label{lem:light} For the path decomposition $\pmax(T)$, $\sum_{p \in \cL} \sum_{v \in p} \log |H_v| = O(\sum_{p \in \cL'} |H_p| \log |H_p|)$.
-\end{lemma}
-
-With this lemma, the proof of \Thm{runtime} is straightforward.
-
-\begin{proof} (of \Thm{runtime}) We focus on $\pmax(T)$ and split into heavy and light paths.
-We have $\sum_{v \in T} \log |H_v| =$  $\sum_{p \in \cL} \sum_{v \in p} \log |H_v| + $ $\sum_{p \in \cH} \sum_{v \in p} \log |H_v|$.
-For the first term, we apply \Lem{light} to bound by $O(\sum_{p \in \cL'} |H_p| \log |H_p|)$.
-Then we apply \Clm{above} to bound by $O(\sum_{p \in \cH} |p|\log |p|)$.
-The latter is also bounded by the same expression, by \Lem{pathBounds}. By definition,
-$\sum_{p \in \cH} |p|\log |p| \leq \cost(\pmax(T))$.
-\end{proof}
-
-The main challenge is proving \Lem{light}, which we defer to the next section.
-\begin{definition}
- Let $\chi(T)$ be a valid coloring of a binary tree $T$.  We define a set of disjoint paths over the vertices of 
- $T$ as follows.  
- The corresponding collection of paths that all such edges correspond to is a path decomposition of $T$, which 
- we refer to as a \emph{maximum} path decomposition, and is denoted by $\mathcal{P}(T)$.  
-\end{definition}
-
-From the previous section we now have a lower bound of $\Omega(f_{path}(P))$ for computing the merge tree in terms of any path decomposition $P$. 
-We now show that the running time of our algorithm for computing the merge tree is upper bounded by this expression for a specific path decomposition, 
-which we will call the maximum path decomposition (defined below).
-Specifically, for a given valid coloring $\chi$ of a tree $T$, we will prove that for the maximum path decomposition $f_{alg}(\chi(T)) = O(f_{path}(P(T)))$. 
-%using the observations about path decompositions from 
-%\Sec{background}, which will then imply $f_{alg}(\chi(T)) = \Theta(f_{path}(P(T)))$.
-
-Throughout this section, for a path $p\in P(T)$, we use the notation $H_p$ to refer to the heap at the root of $p$, i.e. $H_p = H_{r_p}$.
-}
-
-\subsection{Proving \Lem{cu-root}: the root is everything in $\cU$} \label{sec:weight}
-
-\Lem{cu-root} asserts the root $r$ in $\cU$ pretty much encompasses all sizes and heaps in $\cU$.
-We will work with the \emph{reduced} heap $\redH_p$. This is the subset of vertices of $H_p$ that 
-do not appear on the heavy support chain of $p$.
-%only appear in light paths. 
-By definition, for any unsupported light path
-(hence, any non-root $p \in \cU$), $|\redH_p| \geq 99|H_p|/100$. 
-% For notational convenience, we simply set $\redH_r = H_r$.
-%Consider $q \in \cL \setminus \cL'$ that is a child of $p$ in the shrub $\cU$.
-We begin with a key property, which is where the construction of $\pmax(T)$ enters the picture.
-
-%
-%
-%
-%We focus on $\pmax(T)$, and prove the key property of this path decomposition.
-
-\begin{lemma}
-\label{lem:geometric}
- Let $q$ be the child of some path $p$ in $\cU$, then $|H_p|\geq \frac{3}{2}|H_q|$. 
- Moreover, if $p\neq r(\cU)$, then $|\redH_p|\geq \frac{3}{2}|\redH_q|$. 
-\end{lemma}
-\begin{proof} Let $h(q)$ denote the heavy path that is a child of $p$ in $\pmax(T)$,
-and an ancestor of $q$. If no such heavy path exists, then by construction $p$ is the parent 
-of $q$ in $\pmax(T)$, and the following argument will go through by setting $h(q)=q$.
-
-The chain of ancestors from $q$ to $h(q)$ consists only of heavy paths.
-Since $q$ is unsupported, these paths contain at most $|H_q|/100$ vertices of $H_q$. 
-Thus, $|H_{h(q)}| \geq 99|H_q|/100$.
-
-Consider the base
-of $h(q)$, which is a node $w$ in $T$. Let $v$ denote the sibling of $w$ in $T$.
-Their parent is called $u$. Note that both $u$ and $v$ are nodes in the path $p$.
-Now, the decomposition $\pmax(T)$ put $u$ and $v$ in the same path $p$. This
-implies $|H_v| \geq |H_w|$. Since $|H_u| \geq |H_v| + |H_w| - 2$,
-$|H_u| \geq 2|H_w| - 2$. Let $b$ be the base of $p$.
-We have $|H_p| = |H_b| \geq |H_u| - |p| \geq 2|H_w| - |p| - 2$. Since $p$ is a light path, $|p| < \sqrt{|H_p|}/100$.
-Applying this bound, we get $|H_p| \geq (2-\delta)|H_w|$ (for a small constant $\delta > 0$).
-Since $w$ is the base of $h(q)$, $H_w = H_{h(q)}$. We apply the bound $|H_{h(q)}| \geq 99|H_q|/100$
-to get $|H_p| \geq 197|H_q|/100$, implying the first part of the lemma.  For the second part, observe that 
-if $p\neq r(\cU)$, then $p$ is unsupported and so $|\redH_p| \geq 99|H_p|/100$, and therefore the second part follows since $|H_q| \geq |\redH_q|$.
-%
-%
-% Let $w$ be the vertex in the path $p$ to which the root of the path $q$ is adjacent to in $T$, 
-% i.e. $w$ is the parent of $r_q$ in $T$.  Let $c$ be the other child of $w$, which lies in $p$.
-% As $\mathcal{P}(T)$ is a maximum path decomposition of $T$ and $c$ rather than $r_q$ is in 
-% the same path as $w$, we have $|H_q|\leq |H_c|$.  Therefore, $|H_w|\geq |H_q|+|H_c|-1\geq 2|H_q|-1$. 
-% Therefore, by \Obs{decrease}, $|H_p|+ d(r_p, w) \geq |H_w|\geq 2|H_q|-1$.  
-% Since $p$ is a light path, $d(r_p, w)\leq \sqrt{|H_p|}/100$ and so we have $|H_p|+\sqrt{|H_p|}/100\geq 2|H_q|-1$.
-% Assuming $|H_p|>11$ (and we can safely assume $|H_p|$ is bigger than any constant), this yields $\frac{3}{5} |H_p| \geq |H_q|$
-\end{proof}
-
-This immediately proves part (i) of \Lem{cu-root}. Part (ii) requires much more work.
-% It is helpful to spell out how the various heaps evolve. Consider $p \in \cU$,
-% which has children $q_1, q_2, \ldots, q_k$ in $\pmax(T)$, and a leaf vertex $\ell$ in $T$. 
-% The reduced heap $\redH_p$ is the
-% union of $\redH_\ell$ and $\redH_{q_1}, \redH_{q_2}, \ldots, \redH_{q_k}$, with the vertices in $p$
-% removed. Note that some of the $q_i$'s are the corresponding $h(q')$ vertices,
-% for a child $q'$ in $\cU$.
-
-We define a \emph{residue} $R_p$ for each $p \in \cU$. Suppose $p$ has children
-$q_1, q_2, \ldots, q_k$ in $\cU$. Then $R_p = |\redH_p| - \sum_i |\redH_{q_i}|$. 
-By definition, $|\redH_p| = \sum_{q \in \cU_p} R_p$. 
-Note that $R_p$ can be negative. 
-Now, define $R^+_p = \max(R_p,0)$, and set $W_p = \sum_{q \in \cU_p} R^+_p$.
-Observe that $W_p \geq |\redH_p|$. We also get an approximate converse.
-
-\begin{claim} \label{clm:loss} For any path $p\in \cU$, $|\redH_p| \geq W_p - 2\sum_{q \in \cU_p} |q|$.
-\end{claim}
-
-\begin{proof} We write $W_p - |\redH_p| = \sum_{q \in \cU_p} R^+_q - R_q$ $= -\sum_{q \in \cU_p: R_q < 0} R_q$.
-Consider $q \in \cU_p$ and denote the children in $\cU_p$ by $q'_1, q'_2, \ldots$.
-Note that $R_q$ is negative exactly when $|\redH_q| < \sum_i |\redH_{q'_i}|$.
-Traverse $\pmax(T)$ from $q'_i$ to $q$.
-Other than $q$, all other nodes encountered are in the heavy support chain of $q'_i$ and hence
-do not affect its reduced heap.
-The vertices of $\redH_{q'_i}$ that are deleted
-are exactly those present in the path $q$. 
-Any vertex in $q$ can be deleted from at most two of the reduced heaps (of the children of $q$ in $\cU_p$), since 
-theses reduced heaps do not have an ancestor-descendant relationship.
-Therefore when $R_q$ is negative, it is at most by $2|q|$.
-We sum over all $q$ to complete the proof.
-\end{proof}
-
-\ignore{Let $D_q = H_q \setminus \bigcup_i H_{q'_i}$. In other words, as we traverse $\cU$,
-$D_q$ is the set of vertices ``added" at $q$. Suppose we traverse $\pmax(T)$ from $q'_i$
-to $q$. All vertices of $H_{q'_i}$ that lie on these paths are removed.
-
- and traverse $\pmax(T)$ up to $p$. Suppose we encounter
-the paths $q_0 = q, q_1, q_2, \ldots, q_k = p$. The vertices in the residue $R_q$ that 
-are removed from $H_p$ are exactly the vertices that lie in these paths. 
-We split this contribution into heavy and light paths, observing that the latter
-are all unsupported. Hence, we can (crudely) upper bound the number of vertices
-removed by $\sum_{q \in \cU_p} |q| + \sum_{q \in \cU_p} \sum_{h \in \cH} |h \cap q|$.
-Each vertex can be removed from at most two distinct $R_q$'s, completing the proof.
-}
-
-The main challenge of the entire proof is bounding the sum of path lengths, which is done next.
-We stress that the all the previous work is mostly the setup for this claim.
-
-\begin{claim} \label{clm:charge} Fix any path $p \in \cU\setminus\{r(\cU)\}$. Suppose for any $q,q' \in \cU_p$
-where $q$ is a parent of $q'$ in $\cU_p$, $W_q \geq (4/3)W_{q'}$. Then $\sum_{q \in \cU_p} |q| \leq W_p/20$.
-\end{claim}
-
-\begin{proof} Since $q$ is an unsupported light path, $|q| < \sqrt{|H_q|}/100 \leq \sqrt{|\redH_q|}/99 \leq \sqrt{|W_q|}/99$.
-We prove that $\sum_{q \in \cU_p} \sqrt{|W_q|}/99 \leq W_p/20$
-by a charge redistribution scheme. Assume that each $q \in \cU_p^-$
-starts with $\sqrt{W_q}/99$ units of charge. We redistribute this charge over all
-nodes in $\cU_q$, and then calculate the total charge. For $q \in \cU_p$, spread
-its charge to all nodes in $\cU_q$ proportional to $R^+$ values. In other words,
-give $(\sqrt{W_q}/99)\cdot(R^+_{q'}/W_q)$ units of charge to each $q' \in \cU_q$.
-
-After the redistribution, let us compute the charge deposited at $q$. Every ancestor in $\cU_p^-$ 
-$q = a_0, a_1, a_2, \ldots, a_k$ contributes to the charge at $q$. The charge is expressed in the following
-equation. We use the assumption that $W_{a_i} \geq (4/3) W_{a_{i-1}}$ and hence $W_{a_i} \geq (4/3)^i W_{a_0} \geq (4/3)^i$, 
-as $a_0$ is an unsupported light path and hence $W_{a_0}\geq 1$. 
-$$ ({R^+_q}/99)\sum_{a_i} 1/\sqrt{W_{a_i}} \leq ({R^+_q}/99)\sum_{a_i} (3/4)^{i/2} \leq R^+_q/20 $$
-The total charge is $\sum_{q \in \cU_p} R^+_p/20 = W_p/20$. 
-\end{proof}
-
-\begin{corollary} \label{cor:chargeCor}
- Let $r$ be the root of $\cU$, and suppose that for any paths $q,q' \in \cU\setminus \{r\}$, where $q$ is a parent of $q'$ in $\cU$, $W_q \geq (4/3)W_{q'}$.
- Then $\sum_{p \in \cU} |p| \leq W_r/20+|r|$.
-\end{corollary}
-\begin{proof}
- Let $c_1, \dots, c_m$ be the children of $r$ in $\cU$.  By definition, $W_r = \sum_i W_{c_i}+R^+_r \geq \sum_i W_{c_i}$.
- By \Clm{charge}, for each $c_i$ we have $W_{c_i}/20\geq \sum_{p \in \cU_{c_i}} |p|$.  Combining these to facts yields the claim.
-\end{proof}
-
-% \begin{claim} \label{clm:charge} Fix $p \in \cU$. Suppose for any $q,q' \in \cU_p$
-% where $q$ is a parent of $q'$ in $\cU_p$, $W_q \geq (4/3)W_{q'}$. Then $\sum_{q \in \cU_p} |q| \leq W_p/20$.
-% \end{claim}
-% 
-% \begin{proof} Since $q$ is an unsupported light path, $|q| < \sqrt{|H_q|}/100 \leq \sqrt{|\redH_q|}/99 \leq \sqrt{|W_q|}/99$.
-% We prove that $\sum_{q \in \cU_p} \sqrt{|W_q|}/99 \leq W_p/20$
-% by a charge redistribution scheme. Assume that each $q \in \cU_p$
-% starts with $\sqrt{W_p}/99$ units of charge. We redistribute this charge over all
-% nodes in $\cU_p$, and then calculate the total charge. For $q \in \cU_p$, spread
-% its charge to all nodes in $\cU_q$ proportional to $R^+$ values. In other words,
-% give $(\sqrt{W_q}/99)\cdot(R^+_{q'}/W_q)$ units of charge to each $q' \in \cU_q$.
-% 
-% After the redistribution, let us compute the charge deposited at $q$. Every ancestor in $\cU_p$ 
-% $q = a_0, a_1, a_2, \ldots, a_k$ contributes to the charge at $q$. The charge is expressed in the following
-% equation. We use the assumption that $W_{a_i} \geq (4/3) W_{a_{i-1}}$ and hence $W_{a_i} \geq (4/3)^i W_{a_0} \geq (4/3)^i$, 
-% as $a_0$ is an unsupported light path and hence $W_{a_0}\geq 1$. 
-% $$ ({R^+_q}/99)\sum_{a_i} 1/\sqrt{W_{a_i}} \leq ({R^+_q}/99)\sum_{a_i} (3/4)^{i/2} \leq R^+_q/20 $$
-% The total charge is $\sum_{q \in \cU_p} R^+_p/20 = W_p/20$. 
-% \end{proof}
-
-We wrap it all up by proving part (ii) of \Lem{cu-root}.
-
-\begin{claim} $\sum_{p \in \cU} |p| \leq |H_{r(\cU)}|/10$.
-\end{claim}
-
-% \begin{proof} We use $r$ for $r(\cU)$. Our aim is to simply apply \Clm{charge} from $r(\cU)$. 
-% Suppose $W_q \geq (4/3)W_{q'}$ (for any choice in $\cU$ of $q$ parent of $q'$). By \Clm{charge}, $\sum_{p \in \cU} |p| \leq W_{r}/20$.
-% By \Clm{loss}, $|\redH_r| \geq W_r - 2\sum_{p \in \cU} |p|$. By simple algebra, we deduce, $\sum_{p \in \cU} |p| \leq |\redH_r|/10 = |H_r|/10$.
-% 
-% We now prove that for any $q$ parent of $q'$, $W_q \geq (4/3)W_{q'}$.
-% Suppose not. Let $p, p'$ be the counterexample furthest from the root,
-% where $p$ is the parent of $p'$. 
-% Note that for $q$ and child $q'$ in $\cU_{p'}$, $W_q \geq (4/3)W_{q'}$.
-% We will apply \Clm{charge} for $\cU_{p'}$ to deduce that $\sum_{q \in \cU_{p'}} |q| \leq W_{p'}/20$.
-% By \Clm{loss}, $|\redH_{p'}| \geq 19W_{p'}/20$. By \Lem{geometric}, $|\redH_p| \geq (3/2)|\redH_{p'}|$.
-% Noting that $W_p \geq |\redH_p|$, we deduce that $W_p \geq (4/3)W_{p'}$.
-% Hence, $p, p'$ is not a counterexample, and more generally, this is no counterexample.
-% That completes the whole proof.
-% \end{proof}
-
-\begin{proof} We use $r$ for $r(\cU)$. %Our aim is to simply apply \Clm{charge} from $r(\cU)$. 
-Suppose $W_q \geq (4/3)W_{q'}$ (for any choice in $\cU\setminus\{r\}$ of $q$ parent of $q'$), 
-then by \Cor{chargeCor}, $\sum_{p \in \cU} |p| \leq W_{r}/20+|r|$.
-By \Clm{loss}, $|\redH_r| \geq W_r - 2\sum_{p \in \cU} |p|$, and so combining these inequalities gives,
-\[
-% \sum_{p \in \cU} |p| \leq \pth{|\redH_r|+2\sum_{p \in \cU} |p|}/20+|r| \Rightarrow 
- \sum_{p \in \cU} |p|\leq \frac{10}{9} \pth{|\redH_r|/20+|r|} \leq 
- \frac{10}{9}\pth{|H_r|/20+\sqrt{|H_r|}/100} \leq |H_r|/10.
-\]
-
-We now prove that for any $q$ parent of $q'$ (other than $r$), $W_q \geq (4/3)W_{q'}$.
-Suppose not. Let $p, p'$ be the counterexample furthest from the root,
-where $p$ is the parent of $p'$. 
-Note that for $q$ and child $q'$ in $\cU_{p'}$, $W_q \geq (4/3)W_{q'}$.
-We will apply \Clm{charge} for $\cU_{p'}$ to deduce that $\sum_{q \in \cU_{p'}} |q| \leq W_{p'}/20$.
-Combining this with \Clm{loss} gives, $|\redH_{p'}| \geq 19W_{p'}/20$. By \Lem{geometric}, $|\redH_p| \geq (3/2)|\redH_{p'}|$.
-Noting that $W_p \geq |\redH_p|$, we deduce that $W_p \geq (4/3)W_{p'}$.
-Hence, $p, p'$ is not a counterexample, and more generally, there is no counterexample.
-That completes the whole proof.
-\end{proof}
-
-\section{Lower Bound by Path Decomposition}
-In this section we provide a lower bound for the time it takes to compute a contour tree in terms the cost of path decompositions.  Specially, 
-we first consider the easier case of merge trees, after which the lower bound for more general contour trees will easily follow.
-
-\subsection{Merge Trees}
-In this section we show that any algorithm for computing the merge tree $T$ must pay at least $f_{path}(P(T))$ (see \Def{path}) for any path decomposition $P(T)$.
-
-%We say two shrubs $S_1$ and $S_2$ (of merge trees of potentially different surfaces) are equivalent if there is an isomorphism, $f$,
-%between the two shrubs such that for any node $v\in S_1$ corresponding to a path of size $|v|$, $f(v)$ also corresponds to a path of size $|v|$.
-%For the purposes of the next lemma, we say two $2$-manifolds are different, if their corresponding contour trees are the different.
-
-In the following note that two distinct PL surfaces can have the same graph structure defining their meshes, but different height values associated with their vertices.
-
-\begin{figure}[h]\centering
-    \includegraphics[width=.15\linewidth]{figs/diamond}\hspace{2cm}
-    \includegraphics[width=.125\linewidth]{figs/topView}\hspace{2cm}
-    \includegraphics[width=.095\linewidth]{figs/sideView}
-    \caption{From left to right: the cone portion of a diamond; the top down view and then the side view of a cone face with a platform added.}
-    \label{fig:diamond}
-\end{figure}
-
-\begin{lemma}
-\label{lem:pathDecomp}
-Let $T$ be a merge tree, and let $P = P(T)$ any path decomposition of $T$.  %, and $\pathTree$ the corresponding shrub.  
-Then there is a set of at least $\Pi_{p_i\in P} (|p_i|-1)!$ distinct PL $2$-manifolds, all of which have the same defining mesh.
-\end{lemma}
-\begin{proof}
- We describe a simple mesh, with heights specified for all vertices except for the saddles.  The main idea is that 
- this mesh will be flexible enough to allow almost any assignment of heights to the saddles, thus defining a class of merge trees 
- whose size is that described in the lemma statement.
-
- First, we describe the basic building block of this input mesh, which roughly speaking corresponds to a fixed path $p\in P$.  
- Specifically, we say a \emph{diamond} of size $m$ is the following mesh.  To start, construct a cycle 
- of length $m$, where the height of the $i$th vertex in the cycle is say $i/2m$.  Next, create an apex vertex at height 
- $\infty$ (or any appropriately large finite height), and add an edge between this vertex and each vertex in the cycle.  At this point 
- we have built an upside down cone with $m$ triangular faces (see \Fig{diamond}). To make this cone a closed surface we add a triangular cap at the bottom. 
- Specifically, add a $3$ cycle of vertices at heights $0$, $1/20m$, and $1/10m$, and connect each vertex of this $3$-cycle to an arbitrary distinct 
- vertex on the $m$ cycle (with matching cyclic ordering), and triangulate all remaining faces.  For consistency, if the size $m$ is $\leq 3$, the diamond 
- will be just a tetrahedron.
-
- Now above each face of the cone of the diamond, we will create a triangular ``platform'', which will act as the location to glue on other (shrunken) diamonds, 
- and whose highest point of connection to the face below will be a saddle vertex.  Specifically the heights of the vertices of the triangle will be $h$, $h-\eps$, and $h-2\eps$, where 
- $h$ is some unspecified height value.  This near horizontal platform is then connected to its respective face with vertical edges down to the face. See \Fig{diamond}.
- %
- %The vertex at height $h$ will lie in the plane of the respective face of the cone, and the other two will lie above the face, 
- %with edges connecting down to the face, hence creating an actual ``platform''.
-
- At this point we can construct the entire input mesh, by gluing together diamonds.  
- Specifically, for each path $p_i\in P$ we create a diamond of size $|p_i|-1$ (we subtract one because the tip of the diamond already corresponds to the leaf of the path).
- Now we glue together diamonds of different paths in the same way the paths are connected in the shrub $\pathTree$.  Specially, for two paths $p,q\in P$ where 
- $p$ is the parent of $q$ in $\pathTree$, we glue the bottom triangular cap of the diamond for $q$ onto the triangular ``platform'' of some face of the diamond for $p$.
- (Naturally for this construction to work, diamonds for a given path will be scaled down relative to the size of the diamond of their parent).
-
- Observe now that the only saddle points in this construction are the highest connecting point on a face of an edge from the platform above.  Moreover, the only maxima are the apexes of the 
- diamonds, and there is exactly one minima corresponding to the bottom cap of the diamond representing the root of $\pathTree$.  Therefore, the saddles on a given 
- diamond will appear contiguously on a root to leaf path in the merge tree of the mesh, where the leaf corresponds to the maxima of the diamond 
- (since all these saddles have a direct line of sight to this apex).  In particular, 
- this implies that, regardless of the heights assigned to the platforms, the merge tree of this mesh has a path decomposition whose corresponding shrub is equivalent to $\pathTree$.
-
- Now there is a valid instantiation of this described mesh for any relative ordering of the heights of the saddles on a given diamond.  In particular, 
- there are $(|p|-1)!$ possible orderings of the heights of the saddles on the diamond for $p$, and hence $\Pi_{p_i\in P} (|p_i|-1)!$ possible instantiations of 
- the mesh overall.  Also, each one of these instantiations will result in a different merge tree.  Specifically, a permutation of the heights on a given diamond 
- corresponds to a permutation of the vertices along a path in $P$, as by construction all saddles on a given diamond will appear in sorted order in the merge tree.
-\end{proof}
-
-\begin{corollary}
- \label{cor:mergeLB}
- Let $A(S)$ be any comparison based algorithm to compute the merge tree of any input surface $S$ (whose contour tree is a merge tree).  
- Then the running time of $A(S)$ is $\Omega(f_{path}(P))=\Omega(\sum_{p_i\in P} |p_i|\log|p_i|)$ where $P$ is any path decomposition of the 
- merge tree of $S$.
-\end{corollary}
-\begin{proof}
- Given an input surface $S$, by the above lemma, any path decomposition $P$ of its merge tree defines a set of $\Pi_{p_i\in P} |p_i|!$ surfaces 
- with distinct merge trees, all of which have the same mesh.  Specifically, for each sorted ordering of the relative height values of the saddles 
- on each diamond in the mesh, yields a different merge tree.
- Therefore, any comparison based algorithm which computes the merge tree must distinguish between these possibilities, and therefore its corresponding decision tree must 
- have $\Pi_{p_i\in P} |p_i|!$ leaves.  Hence by Stirling's approximation, this decision tree must have depth $\Omega(\sum_{p_i\in P} |p_i|\log|p_i|)$.
-\end{proof}
-
-\subsection{Contour Trees}
-We first generalize previous terms to the case of contour trees.
-In this section $T$ will denote an arbitrary contour tree.  Specifically, every internal vertex in $T$ has degree $3$ 
-and the height function defined over the vertex set is such that every internal vertex in either a split or merge 
-(i.e. at most $2$ neighbors lie above or below it).
-
-Again we define a path decomposition as a partition of the vertices into 
-disjoint leaf paths (i.e. each path is monotonic in height and contains a leaf). 
-However, for simplicity we now restrict our attention to path decompositions which 
-roughly speaking are consistent with the raining procedure described in \Sec{rain} 
-(more general decompositions can work, but it is not needed for our purposes).
-
-\begin{definition}
-\label{def:path2} A path decomposition, $P(T)$, is called \emph{rain consistent} if its paths can be obtained as follows.
-Perform an upward BFS from an arbitrary minima $v$ in $T$, and mark all vertices encountered.  
-Now recursively run a directional BFS from all vertices adjacent to the current marked set.  
-Specifically, for each BFS run, make it an upward BFS if it is at an odd height in the recursion tree and downward otherwise.
-
-This procedure partitions the vertex set into disjoint rooted subtrees of $T$, based on which BFS marked a vertex.  
-For each such subtree, now take any partition of the vertices into leaf paths.\footnote{Note that the subtree  
-of the initial vertex is rooted at is a minima.  For simplicity we require that the path this vertex belongs to 
-also contains a maxima.}
-\end{definition}
-
-The following is analogous to \Lem{pathDecomp}, and in particular uses it as a subroutine.
-
-\begin{lemma}
-\label{lem:pathDecomp2}
-Let $T$ be a contour tree, and let $P = P(T)$ be any rain consistent path decomposition of $T$.  
-Then there is a set of at least $\Pi_{p_i\in P} (|p_i|-1)!$ distinct PL $2$-manifolds, all of which have the same defining mesh.
-\end{lemma}
-\begin{proof}
- As $P$ is rain consistent, the paths can be partitioned into sets $P_1, \dots, P_m$, where $P_i$ is the set of all paths with 
- vertices from a given BFS, as described in \Def{path2}.  Specifically, let $T_i$ be the subtree of $T$ corresponding to $P_i$ and 
- let $r_i$ be the root vertex of this subtree.  
- Note that the $P_i$ sets naturally define a tree where $P_i$ is the parent of $P_j$ if $r_i$ is adjacent to a vertex in $P_j$.
-
- As the set $P_i$ is a path decomposition of a rooted binary tree $T_i$, the surface mesh construction of \Lem{pathDecomp} for $P_i$ is well defined.  
- Actually the only difference is that here the rooted tree is not a full binary tree, and so some of the faces of the constructed diamonds will be blank.  
- Specifically, these blank faces correspond to the adjacent children of $P_i$, and they tell us how to connect the meshes of the different $P_i$'s.
-
-\begin{figure}[h]\centering
-    \includegraphics[width=.5\linewidth]{figs/bridge2}
-    \caption{On the left, the root of $P_i$ merges into a diamond of its parent $P_j$.  On the right, the root of $P_i$ splits off from a diamond of $P_j$}
-    \label{fig:diamond}
-\end{figure}
-
- So for each $P_i$ construct a surface mesh as described in \Def{path2}.  By construction, the diamonds corresponding to the paths in $P_i$ are connected 
- into a tree structure (i.e. corresponding to the shrub of $P_i$).  Therefore the bottoms of all these diamonds are covered except for the one corresponding 
- to the path containing the root $r_i$.  If $r_i$ is the initial minima that the rain consistent path decomposition was defined from,  
- we add a tetrahedral cap to the uncovered bottom of this diamond (i.e. we construct the minima).  Otherwise, $P_i$ has some parent $P_j$ in which case 
- we connect the bottom of the diamond for $r_i$ to a free face of a diamond in the construction for $P_j$, specifically, the face corresponding to the 
- vertex in $T$ which $r_i$ is adjacent to. 
- If $T_i$ was created by an upward BFS, then the connection to this face should be a merge vertex, and otherwise it should be a split vertex.  
- See \Fig{diamond}.
-
- Just as in \Lem{pathDecomp} we now have one fixed mesh graph structure, such that each different relative ordering of the heights of the merge and split vertices 
- on each diamond produces a surface with a distinct merge tree.  The specific bound on the number of such distinct manifolds follows by applying the 
- bound from \Lem{pathDecomp} to each $P_i$.
-\end{proof}
-
-For the same reason that \Cor{mergeLB} followed from \Lem{pathDecomp}, we now have the following.
-
-\begin{theorem}
- Let $A(S)$ be any comparison based algorithm to compute the contour tree of an input surface $S$.  
- Then the running time of $A(S)$ is $\Omega(f_{path}(P))=\Omega(\sum_{p_i\in P} |p_i|\log|p_i|)$ where $P$ 
- is any rain consistent path decomposition of the contour tree of $S$.
-\end{theorem}
-
-% \section{Temp}
-% Given a contour tree we allow any path decomp, as long as each path contains a maxima or minima (or both).
-% We call a path $p$ in the decomp a min (resp. max) path if it contains a minima (resp. maxima) (note a path can be both).
-% For every path we create a cylinder with $m$ faces for $p$, where $m$ is the number of non-leaf vertices in $p$ (either $|p|-1$ or $|p|-2$).  
-% If $p$ has a maxima we put a top cone cap on the cylinder, and if it contains a manima then put a bottom cone cap (i.e. the cap represents the maxima or minima).
-% 
-% Every edge $e$ in $T$, but not in a path decomp. tells us how to glue together cylinders.  Let $u$ be the upper endpoint of $e$ and $l$ the lower endpoint.
-% 
-% First we ``remove'' all edges $e$ such that $u$ is adjacent to a max path and $l$ adjacent to a min path by joining the free ends of the respective cylinder together.
-% Hence we can assume no such edges exist.  Now we have three cases:
-% 
-% \begin{enumerate}[{Case} 1]
-%  \item 
-% $l$ is adjacent to a min path:\newline
-% Note in this case the path containing $u$ must be a min path.
-% In this case connect the bottom of the cylinder for the path containing $l$ into a face of the cylinder for the path containing $u$.
-% \item
-% $l$ is adjacent to a max path and $u$ is adjacent to a max path:\newline
-% In this case connect the bottom of the cylinder for the path containing $u$ into a face of the cylinder for the path containing $l$.
-% \item
-% $l$ is adjacent to a max path and $u$ is adjacent to a min path:\newline
-% Connect a face of the cylinder for the $u$ path (using a split) to a face of the cylinder for the $l$ path (using a merge).
-% \end{enumerate}
-
-\bibliographystyle{alpha}
-\bibliography{contour}
-
-\appendix
-
-\section{Old stuff}
-
-\subsubsection{Paying it down the line}
-
-% \begin{claim}
-% \label{clm:sizes}
-%  Let $p$ be a path in a maximum path decomposition, such that $|H_p|>c|p|$ for some constant $c> 6$.
-%  Let $H_f$ denote the heap at the leaf of $p$.  We then have the following
-%  \begin{compactenum}
-%   \item $|H_f|\leq (1+1/c)|H_p|$
-%   \item $|H_f|\geq |H_p|/3$
-%   \item For any path $q$ adjacent to $p$ in the shrub, $|H_q|<\frac{2}{3}|H_p|$
-%  \end{compactenum} 
-% \end{claim}
-% \begin{proof}
-%   For any vertex $v\in p$ \Obs{decrease} implies that $|H_v|\leq |H_p| + |p| < (1+1/c) |H_p|$.
-%   First observe that by setting $v=f$ this immediately implies the first part of the claim.
-%  % By a similar argument, $|H_f|\geq |H_p| - |p| > (1-1/c)|H_p|$.
-%   
-%   To prove the last part of the claim, suppose otherwise that $|H_q|\geq \frac{2}{3}|H_p|$ for some 
-%   adjacent path $q$.  As this is a maximum path decomposition, this implies that there is some 
-%   $v\in p$ such that $|H_v|\geq \frac{4}{3}|H_p|-1$.
-%   By \Obs{decrease} we then have, 
-%    $|H_p| \geq |H_v| - |p|\geq \frac{4}{3}|H_p| - 1 - |H_p|/c \geq (4/3-1/c)|H_p|-1$.
-%   When rearranged this gives $1\geq (1/3-1/c)|H_p|$.  As $c>6$ we have $1\geq |H_p|/6$, 
-%   which is a contradiction as $|H_p|>c|p|>6$.
-% \end{proof}
-
-\Ben{I had originally thought the following claim would be useful in proving the lemma below.}
-\begin{claim}
-\label{clm:sizes}
- Let $p$ be a path in a maximum path decomposition, such that $|H_p|>c|p|$ for some constant $c> 6$.
- We then have the following
-% \begin{compactenum}
-%  \item 
-  For any path $q$ adjacent to $p$ in the shrub, $|H_q|<\frac{2}{3}|H_p|$
-%  \item If there does not exist and adjacent path $q$ such $H_q\geq H_p/3$, then $|H_f|\geq |H_p|/3$
-% \end{compactenum} 
-\end{claim}
-\begin{proof}
-  To prove the claim, suppose otherwise that $|H_q|\geq \frac{2}{3}|H_p|$ for some 
-  adjacent path $q$.  As this is a maximum path decomposition, this implies that there is some 
-  $v\in p$ such that $|H_v|\geq \frac{4}{3}|H_p|-1$.
-  By \Obs{decrease} we then have, 
-   $|H_p| \geq |H_v| - |p|\geq \frac{4}{3}|H_p| - 1 - |H_p|/c \geq (4/3-1/c)|H_p|-1$.
-  When rearranged this gives $1\geq (1/3-1/c)|H_p|$.  As $c>6$ we have $1\geq |H_p|/6$, 
-  which is a contradiction as $|H_p|>c|p|>6$.  
-\end{proof}
-
-\begin{lemma}
-\label{lem:excess}
- Let $p$ be a heavy path, and let $S_1, \dots S_k$ and $R_1, \dots, R_l$ be the sizes of the heaps at the roots of the light and heavy children of $p$, respectively, in the shrub.  
- Let $H_f$ be the heap at the leaf of the path $p$.
- Then,  $|H_p| \log |H_p|   \geq  \sum_{v\in p} \log|H_v| + \frac{4}{3}\sum_{i=1}^k S_i \log S_i + \sum_{i=1}^l R_i \log R_i  -  c|p|\log |p|$, for some sufficiently large constant $c$
- \end{lemma}
-\begin{proof}
- First observe that the lemma holds for $|H_p|$ less than any fixed constant, since otherwise \Lem{adjacent} implies we can choose $c$ 
- large enough such that $c|p|\log|p|$ will dominate all other terms.
- Next, observe that $\sum_{v\in p} \log|H_v| \leq \sum_{v\in p} \log(|H_v|+|p|) = |p|\log(|H_v|+|p|) = O(|p|\log|p|)$, as $p$ is a heavy path.  
- Therefore, in the remainder of the proof  we can safely ignore the $\sum_{v\in p} \log|H_v|$ term as we can choose $c$ large enough to cover it.  
- That is we now instead prove the inequality,
- \[
-  |H_p| \log |H_p|   \geq \frac{4}{3}\sum_{i=1}^k S_i \log S_i + \sum_{i=1}^l R_i \log R_i  -  c|p|\log |p|
- \]
- To do so we  break the analysis into two case based on the relative size of $|p|$ and $|H_p|$. 
- In the following, $\alpha>10$ is some yet to be determined constant.
-
-\paragraph{Case 1, $|p|\geq |H_p|/\alpha$:\\}
-First observe that by \Lem{adjacent}, $|H_p|+|p| \geq \sum_{i=1}^k S_i+ \sum_{i=1}^l R_i$.
-By choosing $c$ large enough we therefore have,
-\begin{align*}
- c|p|\log|p| &\geq 2(1+\alpha)^2 |p|\log |p| \geq 2(1+\alpha)|p| \log((1+\alpha)|p|)\\
-             &\geq 2(|H_p|+|p|)\log(|H_p|+|p|) \geq 2\log(|H_p|+|p|) \pth{\sum_{i=1}^k S_i + \sum_{i=1}^l R_i}\\
-             &\geq \frac{4}{3}\sum_{i=1}^k S_i \log S_i + \sum_{i=1}^l R_i \log R_i 
-\end{align*}
-
-In this case $\log(|H_p|+|p|)\leq \log(2|p|) \leq 1+\log(|p|)$ (where $\log$ is base $2$).
-Therefore, assuming $|p|>1$, we have 
-\[
-(|H_p|+|p|)\log(|H_p|+|p|)\leq (1+c)|p|\log((1+c)|p|) %(|H_p|+|p|)(1+\log |p|) \leq 2(1+c)|p|\log |p|
-\]
-which when combined with the inequality from \Lem{adjacent} gives, 
-\[
- 4(1+c)|p|\log |p| \geq 2(|H_p|+|p|)\log(|H_p|+|p|)\geq \pth{|H_f| + \frac{4}{3}\sum_{i=1}^k S_i + \sum_{i=1}^l R_i}\log(|H_p|+|p|).
-\]
-This easily implies the desired inequality as $\log(H_p+p)$ is larger than any other $\log$ term in the inequality.
-
-\paragraph{Case 2, $|p|< |H_p|/\alpha$:\\}
-
-\Ben{Was not able to handle this case.}
-
-%  First, lets assume we can show $\sum_{i=1}^k S_i + \sum_{i=1}^l R_i \leq \frac{3}{4}|H_p|$.  
-%  In this case, using the last part of \Clm{sizes}, we can prove the desired inequality as follows.
-%  
-%  \begin{align*}
-%   \frac{4}{3}\sum_{i=1}^k S_i \log S_i + \sum_{i=1}^l R_i \log R_i  &\leq  \frac{4}{3} \pth{\sum_{i=1}^k S_i \log S_i + \sum_{i=1}^l R_i \log R_i} \\
-%   &\leq  \frac{4}{3} \log |H_p| \pth{\sum_{i=1}^k S_i + \sum_{i=1}^l R_i}  \leq  |H_p| \log |H_p|
-%  \end{align*}
-%  
-%  Finally, to complete the proof we must prove our assumption that $\sum_{i=1}^k S_i + \sum_{i=1}^l R_i \leq \frac{3}{4}|H_p|$.
-%  By \Lem{adjacent} and \Clm{sizes} we have, 
-%  \[
-%  \sum_{i=1}^k S_i + \sum_{i=1}^l R_i 
-%  \leq |H_p|+|p|-|H_f| 
-%  \leq (1+1/\alpha-1/3)|H_p| = (2/3+1/\alpha)|H_p|
-%  \]
-%  
-%  Therefore $\sum_{i=1}^k S_i + \sum_{i=1}^l R_i \leq \frac{3}{4}|H_p|$ if $(2/3 + 1/\alpha)\leq 3/4$, which rearranged gives $\alpha \geq 12$
-\end{proof}
-
-\remove{
-% \subsubsection{Paying it down the line}
-% 
-% \begin{lemma}
-% \label{lem:excess}
-%  Let $p$ be a heavy path, and let $S_1, \dots S_k$ and $R_1, \dots, R_l$ be the sizes of the heaps at the roots of the light and heavy children of $p$, respectively, in the shrub.  
-%  Let $H_f$ be the size of the heap at the leaf of the path $p$, and let $X = \sum_{i=0}^k H_i \log H_i$.  
-%  Then,  $|H_p| \log |H_p|   \geq  H_f\log H_f + \sum_{i=0}^k S_i \log S_i + \sum_{i=0}^l R_i \log R_i  -  c|p|\log |p|$, for some sufficiently large constant $c$
-%  \end{lemma}
-% \begin{proof}
-% By \Lem{adjacent} we know that $|H_p|+|p| \geq  H_f + \sum_{i=0}^k S_i + \sum_{i=0}^l R_i$.  
-% %Now as $p$ is a heavy path we have $|p|\geq \sqrt{|H_p|}/100$, and so, $(|H_p|+|p|)\log(|H_p|+|p|)$
-% 
-% We  break the analysis into two case based on the relative size of $|p|$ and $|H_p|$.  In the following $c$ is some yet to be determined constant greater than $1$.
-% 
-% \paragraph{Case 1, $|p|\geq |H_p|/c$:}  
-% In this case $\log(|H_p|+|p|)\leq \log(2|p|) \leq 1+\log(|p|)$ (where $\log$ is base $2$).
-% Therefore, assuming $|p|>1$, we have 
-% \[
-% (|H_p|+|p|)\log(H_p+p)\leq (|H_p|+|p|)(1+\log |p|) \leq (2|p|)(2\log |p|)\leq 4|p|\log |p|
-% \]
-% which when combined with the inequality from \Lem{adjacent} gives, 
-% \[
-%  4|p|\log |p| \geq (|H_p|+|p|)\log(|H_p|+|p|)\geq \pth{H_f + \sum_{i=0}^k S_i + \sum_{i=0}^l R_i}\log(|H_p|+|p|).
-% \]
-% This easily implies the claim as $\log(H_p+p)$ is larger than any other $\log$ term in the inequality.
-% 
-% \paragraph{Case 2, $|p|< |H_p|/c$:}
-% \end{proof}
-}
-
-Define the crust of the shrub to be the first (i.e. closest to the root) layer of light paths in the shrub.  Apply \Lem{lightCost} to each of the light paths in the crust and then 
-say by \Lem{excess} and since the root heap goes to zero we are done.
-
-\ignore{
-Naturally, this geometric series behavior can be extended to more than two adjacent light paths, thus motivating 
-the following definition.
-
-\begin{definition}
- Let $p$ be a path in a path decomposition $P(T)$, with corresponding shrub $\pathTree$.
- Call a light path $q\in P(T)$ light reachable from $p$ if the path from $p$ to $q$ in $\pathTree$ 
- contains only vertices which correspond to light paths.  Call the subtree of $\pathTree$ consisting 
- of all light reachable paths from $p$ the \emph{light shrub} of $p$, and denote it $\mathcal{L}(p)$.
-\end{definition}
-
-In the following subsection we show how to exploit this geometric series behavior of light shrubs.
-
-We say two paths in $P(T)$ are adjacent if their corresponding vertices are adjacent in the shrub.  Similarly, for 
-two adjacent paths in $P(T)$ we can define a parent/child relationship based on this relationship in the shrub. 
-}
-
-\remove{
-\Ben{Maybe move the remainder of this subsection to 5.3 since heavy and light paths and not used until then}
-
-\begin{definition}
- Let T be a binary tree and let $\chi(T)$ and $P(T)$ be a valid coloring and path decomposition of $T$, respectively.  
- For $p\in P(T)$, we say that $p$ is \emph{light} if $|p| < \sqrt{|H_{r_p}|}/100$, and we say it is \emph{heavy} otherwise.
-\end{definition}
-
-\begin{observation}
-\label{obs:decrease}
- Let $v$ be a vertex in $T$ and let $w$ be its parent.  Then $|H_w| \geq |H_v| -1$, as $L(T_v) \subseteq L(T_w)$ and the path from 
- $w$ to the root has one less vertex than the path from $v$ to the root.
-
- In particular, we have the following more general property.
- Let $v$ and $u$ be any two vertices in the same root to leaf path of $T$, such that $u$ is lower than $v$.  
- Then $|H_v| \leq |H_u| + d(u,v)$. 
-\end{observation}
-
-\begin{lemma}
-\label{lem:pathBounds}
- Let T be a binary tree and let $\chi(T)$ and $P(T)$ be a valid coloring and path decomposition of $T$, respectively.
- Then, 
- \begin{compactenum}
-  \item for any heavy path $p\in P(T)$, $\sum_{v\in p} \log |H_v| = O(|p| \log |p|)$.
-  \item for any light path $p\in P(T)$, $\sum_{v\in p} \log |H_v| = O(|p| \log |H_{r_p}|)$.
- \end{compactenum}
-\end{lemma}
-\begin{proof}
- By $\Obs{decrease}$ we know that for any vertex $v\in p$, $|H_v|\leq |H_{r_p}| + |p|$ (as $r_p$ lies below $v$ on $p$).
- As such, 
- \[
- \sum_{v\in p} \log |H_v| \leq \sum_{v\in p} \log(|H_{r_p}| + |p|)  = |p| \log (|H_{r_p}| + |p|).
- \]
- If $p$ is a heavy path then $|p| \log (|H_{r_p}| + |p|) = O(|p| \log |p|)$, and if $p$ is light then 
- $|p| \log (|H_{r_p}| + |p|) = O(|p| \log |H_{r_p}| )$.
-\end{proof}
-
-The following lemma can be thought of as a generalization of $\Obs{decrease}$.
-
-\begin{lemma}
-\label{lem:adjacent}
- Let $T$ be a binary tree and let $\chi(T)$ and $P(T)$ be a valid coloring and path decomposition of $T$, respectively.
- Let $p$ be any path in $P(T)$ and let $\{q_1, \dots q_k\}$ be the set of all adjacent child paths of $p$.
- For $1\leq i\leq k$ let $r_i = r_{q_i}$ and let $H_i = H_{r_i}$.
- Also, let $H_0$ denote the heap at the leaf of $p$.
-
- Then $\sum_{i=0}^k |H_i| \leq |H_p|+|p|$, where $H_p = H_{r_p}$.
-\end{lemma}
-\begin{proof} 
- First observe that for all $i$, $r_{i}$ lies above $r_p$ on some root to leaf path. 
- Let $H_i'$ be the subset of $H_i$ that lies below $r_p$.  Observe that $H_i'\subseteq H_p$, as $L(T_{r_i})\subseteq L(T_{r_p})$.
-
- Now for any $i\neq j$, $H_i \cap H_j = \emptyset$, as $r_{i}$ and $r_{j}$ are not on the same root to leaf path (i.e. $L(T_{r_i}) \cap L(T_{r_j}) = \emptyset$).
- In particular, $(H_i\setminus H_i') \cap (H_j \setminus H_j') = \emptyset$, and so $\sum_{i=0}^k |H_i\setminus H_i'| \leq |p|$ 
- as $H_i\setminus H_i'$ is precisely the subset of $H_i$ that lies on $p$.
- Moreover, for any $i\neq j$, $ H_i' \cap H_j' = \emptyset$, and so since $H_i'\subseteq H_p$ we have $\sum_{i=0}^k |H_i'|\leq |H_p|$.
- Putting these together gives
- \[
-  \sum_{i=0}^k |H_i| = \sum_{i=0}^k |H_i'| + \sum_{i=0}^k |H_i\setminus H_i'| \leq |H_p|+|p|.
- \]
-\end{proof}
-
-% \Ben{There is a fundamental flaw in the following lemma and corollary.  
-% As such, the remainder of this subsection will likely be deleted.}
-% 
-% Using the notation defined in the above lemma, we have the following crucial lemma, which we will use to 
-% argue we can charge that cost of all paths in $P(T)$ to the cost of the heavy paths.
-% 
-% \begin{lemma}
-% Let $p$ be a heavy path in $P(T)$ and let $\{q_1, \dots q_k\}$ be any set of light paths in $P(T)$ such that 
-% $\sum_{i=1}^k |H_i| = O(H_p)$.  Then $\sum_{i=1}^k \sum_{v\in q_i} \log |H_v| = O(|p|\log |p|)$
-% \end{lemma}
-% \begin{proof}
-% By \Lem{pathBounds}, for all $1\leq i\leq k$, 
-%  $\sum_{v\in q_i} \log |H_v| = O(|q_i|\log |H_i|) = O(|H_i|\log |H_i|)$, as $q_i$ is a light path.
-%  By assumption $\sum_{i=1}^k |H_i| = O(|H_p|)$, 
-%  As $x\log x$ is a convex function, Jensen's inequality then implies 
-%  $\sum_{i=1}^{k} |H_i|\log |H_i| = O(|H_p| \log |H_p|) = O(|p|\log |p|)$, since $p$ is a heavy path.
-% \end{proof}
-% 
-% 
-% \begin{corollary}
-% \label{cor:heavyLight}
-%  Let $p$ be a heavy path in $P(T)$ and let $\{q_1, \dots q_k\}$ be the set of all light paths adjacent to $p$.
-%  Then $\sum_{i=1}^k \sum_{v\in q_i} \log |H_v| = O(|p|\log |p|)$.
-% \end{corollary}
-% \begin{proof}
-%  Let $H_i$ be as defined in \Lem{adjacent}.  By \Lem{pathBounds}, for all $1\leq i\leq k$, 
-%  $\sum_{v\in q_i} \log |H_v| = O(|q_i|\log |H_i|) = O(|H_i|\log |H_i|)$, as $q_i$ is a light path.
-%  Now by \Lem{adjacent}, $\sum_{i=1}^k |H_i| \leq |H_p|+|p| = O(|p|)$, as $p$ is a heavy path. 
-%  Therefore, as $x\log x$ is a convex function, Jensen's inequality implies the claim.
-%  
-% \end{proof}
-
-}
-
-\subsection{The shrub of unsupported paths} \label{sec:light}
-
-It suffices to focus on unsupported light paths. \Lem{light} can be reduced to the following.
-
-\begin{lemma} \label{lem:support} $\sum_{p \in \cL \setminus \cL'} \sum_{v \in p} \log |H_v| = O(\sum_{p \in \cL'} |H_p| \log |H_p|)$.
-\end{lemma}
-
-We now construct the shrub of unsupported light paths. Consider $p \in \cL \setminus \cL'$,
-and traverse the chain of ancestors from $p$. Eventually, we must reach another light path $q$.
-(If not, we have reached the root $r$ of $\pmax(T)$. Hence, $p$ is supported.)
-Insert edge from $p$ to $q$, so $q$ is the parent of $p$ in $\cU$. This construction leads
-to the shrub forest of $\cL \setminus \cL'$, where every unsupported light path is an internal node, and all
-roots are supported light paths.
-
-We list out some important claims about this shrub forest.
-We focus on $\cU$, a connected component (shrub) in the shrub forest and denote the 
-the root by $r(\cU)$.
-
-\begin{claim} \label{clm:loss2} Consider $p \in \cL \setminus \cL'$ that is in $\cU$. Let $Q$
-be the set of heavy ancestors of $p$ that are descendants of $r(\cU)$. Then $\sum_{q \in Q} |q \cap H_p| \leq |H_p|/50$.
-\end{claim}
-
-\begin{proof} Suppose not, for contradictions sake. As we traverse the heavy ancestors of $p$,
-we encounter various vertices of $H_p$ in the ancestors $q$. Consider the first $|H_p|/100$
-such vertices of $p$ encountered. At each such $v$, there are at least $|H_p|/100$
-vertices \emph{not yet} encountered, so $|H_v| > |H_p|/100$. Each of these vertices $v$
-is supported, implying that $p$ is a supported light path.
-\end{proof}
-
-\Lem{support} is a direct consequence of the following,
-which is the really the main lemma.
-
-%\begin{lemma} \label{lem:cu} Let $\cU$ denote a connected component (shrub) in the shrub forest of $\cL \setminus \cL'$ and let $r$
-%be the root of $\cU$. Then $\sum_{p \in \cU} \sum_{v \in p} \log |H_v| = O(|H_r| \log |H_r|)$.
-%\end{lemma}
-%
-%We proceed with the setup to prove \Lem{cu}.
-%We focus our attention on some shrub $\cU$.
-%We prove a technical lemma about $\cU$, from which \Lem{cu} follows directly.
-
-\section{Ben's old stuff}
-
-\begin{proof} (part (ii) of \Lem{cu-root}) Let level $\ell$ of $\cU$ denote all nodes at distance $\ell$ from the root $r$.
-These set of these nodes is designated as $\cU_\ell$. We set $\sigma_\ell = \sum_{p \in \cU_\ell} |p|$
-and $\sigma_{\leq \ell} = \sum_{k \leq \ell} \sigma_k$.
-By \Lem{geometric}, for any $p \in \cU_\ell$, $|H_p| \leq (2/3)^\ell |H_r|$.
-
-Since all nodes in $\cU$ represent light paths, $\forall p \in \cU$, $|p| < \sqrt{|H_p|}/100$.
-So $\sigma_\ell < \sum_{p \in \cU_\ell} \sqrt{|H_p|}/100$.
-
-\end{proof}
-
-To prove this, we need to understand how $H_p$ is constructed. The path $p$ has some children
-$q_1, q_2, \ldots, q_k$ in $\pmax(T)$. Furthermore, as a leaf path in $p$, some leaf
-$\ell$ of $T$ in one of the ends of $p$. The heap $H_p$ is essentially the
-union of $H_\ell$ and $H_{q_1}, H_{q_2}, \ldots, H_{q_k}$, with the vertices in $p$
-removed. Note that some of the $q_i$'s are the corresponding $h(q')$ vertices,
-for a child $q'$ in $\cU$.
-
-Let $\mathcal{L} = \mathcal{L}(r)$ be a light shrub of some light path $r$ in a maximum path decomposition $\mathcal{P}(T)$.
-The intuitive idea is that due to the geometric decreasing behavior of light shrubs described in the previous subsection, 
-one can show that for the entire light shrub our algorithm pays roughly the cost due to the root path $r$.  This argument 
-is hard to make directly, so instead we define an alternative quantity for which we can make the argument.
-Specifically, we now define a weight function $w(v)$ for all $v\in \mathcal{L}$, and show that $w(v)=\Theta(|H_v|)$.
-
-For a path $v\in \mathcal{L}$, let $H_v$ denote the heap at the root of $v$, and let $H_{v_f}$ denote the heap 
-at the leaf of $v$.  Let $S_l(v)$ (resp. $S_h(v)$) be the subset of light (resp. heavy) paths adjacent to $v$ in $\pathTreeA(T)$.
-
-We want to understand how the heap size changes as we move around the paths in $\mathcal{L}$.  Specifically, there are 
-three places the vertices in the heap along a light path can come from; adjacent light paths, adjacent heavy paths, and the initial heap at the leaf of the path.  
-Given this observation we now define a weight function which upper bounds the size of the heap at the root of any path in $\mathcal{L}$.
-
-First, for all $v\in \mathcal{L}$ we define the local cost function $c(v) = H_{v_f}+\sum_{i\in S_h(v)} H_i$, this is the addition to the heap at $v$ which 
-was not already charged to descendant light paths.  We now define a weighted version of $\mathcal{L}$ based on these costs.  
-To do so, we first augment $\mathcal{L}$ by attaching one 
-leaf to each vertex in $\mathcal{L}$.  Now for each newly created leaf (which are the only leaves now) assign the cost 
-$c(v)$ defined above, where $v$ is that leaf's corresponding parent.  
-Now we can define a weight function $w(u)$ over any node $u$ in the tree, which is defined to be the total cost of all leaves in its subtree.   
-To simplify notation, we now use $\mathcal{L}$ to refer to this new weighted tree (rather than its unweighted counterpart).
-
-As the weight $w(p)$ includes the sizes of the heaps of all contributing heavy paths and leaves in the subtree rooted at $p$, we have the following.
-
-\begin{claim}
-For any $p\in \mathcal{L}$, $w(p)\geq |H_p|$.  Moreover, for any $v\in p$, $w(p)\geq |H_v|$.
-\end{claim}
-
-%Let $W= \sum_{v\in \mathcal{L}} w(v) \geq \sum_{v\in \mathcal{L}} |H_v|$.  
-%Suppose we could show that conversely $W=O(H_r)$.  Then we would have 
-%$\sum_{v\in \mathcal{L}} H_v = O(H_r)$, which was our original goal.  
-%So proving $W=O(H_r)$ is now our new goal.
-
-We now would like to show conversely that $w(v)= O(|H_v|)$.  
-To do so we first introduce some notation.  
-Let $C\subseteq \mathcal{P}$ be the subset of paths that are leaves in $\mathcal{L}$.
-For any path $v$, let $\mathcal{L}_v$ (resp. $C_v$) be the subset of paths in $\mathcal{L}$ (resp $C$) 
-that are in the subtree of $\mathcal{L}$ which is rooted at $v$.
-
-First, observe that $|H_v|\geq w(v)-\sum_{u\in \mathcal{L}_v} |u|$.  
-This holds by the same argument as in \Lem{adjacent}, namely, $w(v)$ by definition 
-counts all the contributing heavy path heaps and initial leaf heaps that contribute 
-to the heap at $v$ and $\sum_{u\in \mathcal{L}_v} |u|$ is an upper bound on the number 
-of vertices that could have been removed from one of these heaps before reaching the 
-root of $v$ (since a vertex in a heap can only get removed if it appears on one of 
-the paths).
-
-As all paths in $\mathcal{L}$ are light, we thus have the following.
-
-\begin{claim}
-$|H_v|\geq w(v)-\sum_{u\in \mathcal{L}_v} \sqrt{|H_u|}/100 \geq w(v)-\sum_{u\in \mathcal{L}_v} \sqrt{w(u)}/100$
-\end{claim}
-
-By the above claim, if we can prove $\sum_{u\in \mathcal{L}_v} \sqrt{w(u)}/100 \leq w(v)/c$ for some constant $c>1$, 
-then we will have proven that $|H_v| = \Theta(w(v))$.
-
-Recall that \Lem{geometric} told us that that $|H_{u}|\leq (3/5) |H_{u'}|$, if $u'$ is the parent $u$ in $\mathcal{L}$. 
-The following lemma shows that if we had a similar statement in terms of the weights rather than the heap sizes, 
-then we get the desired bound on the sum.
-
-\begin{lemma}
-\label{lem:distribution}
-Let $v$ be a path in $\mathcal{L}$.  Suppose that for any two paths $x,y\in \mathcal{L}_v$ 
-such that $y$ is the parent of $x$, $w(x)\leq (4/5) w(y)$.
-Then $\sum_{u\in \mathcal{P}_v} \sqrt{w(u)}/100 \leq \sum_{c\in C_v} \sqrt{w(c)}/10$
-\end{lemma}
-
-\begin{proof}
-The idea is to recursively redistribute the contribution to the sum of each internal vertex to its children, 
-until all the weight sits on the leaves.  Specifically, let $v$ be some path in $\mathcal{L}$.  Its 
-contribution to the sum is $\sqrt{w(v)}/100$, which we will call the value at $v$.  We redistribute this 
-value to its children, in proportion by the weight of each child.  For example, if $v$ has a child $x$ then 
-the fraction it gets assigned is $w(x)/w(v)$ (recall that by definition $w(v)$ is the sum of the weights of 
-its children and so the total value will get reassigned to the children).
-
-Suppose we keep redistributing the value of $v$ until it reaches the leaves of its subtree in $\mathcal{L}$.  
-Let's see how much gets assigned to a given leaf.  Specifically, let $u_0, u_1, \ldots, u_k$ 
-be a path in $\mathcal{L}$ (of paths from the set $\mathcal{P}_v$), such that $v=u_0$, $u_{i+1}$ 
-is child of $u_i$ for all $i$, and $u_k$ is a leaf of $\mathcal{L}$.
-
-Then the total value from $u_0$ that ends up at $u_k$ is 
-\[
-\frac{\sqrt{w(u_0)}}{100} \pth{ \frac{w(u_1)}{w(u_0)} \frac{w(u_2)}{w(u_1)} \cdots \frac{w(u_k)}{w(u_{k-1})} }
-= \frac{\sqrt{w(u_0)}}{100} \pth{ \frac{w(u_k)}{w(u_0)} } = \frac{w(u_k)}{100\sqrt{w(u_0)}}
-\]
-
-For a given leaf in $\mathcal{L}$, we can calculate the total value reassigned to it from it its ancestors.
-Specifically, let $u_0, u_1, u_2, \dots, u_m$ be a root to leaf path in $\mathcal{L}$ 
-(i.e. $u_0$ is the root $r$ of $\mathcal{L}$).
-
-By assumption from the lemma statement we have $w(u_{i+1})\leq (4/5) w(u_i)$.  Therefore by the above calculation, 
-the total reassigned to $u_m$ is
-
-\[
- \sum_{i=0}^m \frac{w(u_m)}{100\sqrt{w(u_i)}} \leq \frac{w(u_m)}{100} \sum_{i=0}^m 1 \Big/ \sqrt{\pth{\frac{5}{4}}^iw(u_m)} 
- \leq \frac{\sqrt{w(u_m)}}{100} \sum_{i=0}^{\infty} \pth{\frac{4}{5}}^{i/2} \leq \frac{\sqrt{w(u_m)}}{10}
-\]
-
-The claim then follows by summing this quantity over the children of $\mathcal{L}$.
-\end{proof}
-
-By combining the above lemma and claim we have the following.
-
-\begin{claim}
-For any path $v\in \mathcal{L}$ if it holds that 
-$w(x)\leq (4/5) w(y)$ for any two paths $x,y\in \mathcal{L}_v$ such that $y$ is the parent of $x$, then
-
-\[
-|H_v|\geq w(v)-\sum_{u\in \mathcal{L}_v} \sqrt{w(u)}/100 \geq w(v)- \sum_{c\in C_v} \sqrt{w(c)}/10 \geq \frac{4}{5} w(v)
-\]
-\end{claim}
-
-\begin{lemma}
-Let $\mathcal{L}$ be the light shrub of some path $r$.  
-Then for any $v\in \mathcal{L}$, we have $|H_v|\geq \frac{4}{5} w(v)$.
-\end{lemma}
-% \begin{proof}
-% We will actually prove the stronger statement that for any $v\in \mathcal{L}$, 
-% we have $|H_v|\geq \frac{4}{5} w(v)$ and $w(v)\leq (4/5) w(p)$ where $p$ is the parent 
-% of $v$ in $\mathcal{L}$.
-% 
-% Suppose otherwise, and let $v$ be the lowest path in $\mathcal{L}$ such that this fails 
-% (i.e. the statement holds for all descendants of $v$ in $\mathcal{L}$).  Then for any 
-% two paths $x,y\in \mathcal{L}_v$ such that $y$ is the parent of $x$, 
-% we have that $w(x)\leq (4/5) w(y)$ and so by \Lem{distribution}, $|H_v|\geq \frac{4}{5} w(v)$.
-% Combining this fact with \Lem{geometric} gives,
-% \[
-% w(v) \leq \frac{5}{4}|H_v| \leq \frac{5}{4} \frac{3}{5} |H_p| \leq \frac{3}{4} w(p).
-% \]
-% 
-% \end{proof}
-%
-\begin{proof}
-By the above claim, we know that $|H_v|\geq \frac{4}{5} w(v)$ if it holds that 
-$w(x)\leq (4/5) w(y)$ for any two paths $x,y\in \mathcal{L}_v$ such that $y$ is the parent of $x$.
-So let $v$ be the lowest path in $\mathcal{L}$ such that this fractional decrease fails to hold, i.e. $w(c)> (4/5) w(v)$ 
-where $c$ is a child of $v$ in $\mathcal{L}$, and conversely all vertices in $\mathcal{L}_c$ 
-the fractional decrease holds.  As the fractional decrease holds for a paths in $\mathcal{L}_c$, 
-we have $|H_c|\geq \frac{4}{5} w(c)$.  Therefore,
-\[
-w(c) \leq \frac{5}{4}|H_c| \leq \frac{5}{4} \frac{3}{5} |H_v| \leq \frac{3}{4} w(v).
-\]
-However, this is a contradiction with the assumption $w(c)> (4/5) w(v)$ , and hence the 
-fractional decrease always holds implying the claim always holds.
-\end{proof}
-
-\begin{corollary}
-\label{cor:equivalent}
- Let $\mathcal{L}$ be a light shrub of a light path $r$.  Then for any $v\in \mathcal{L}$, $|H_v| \leq w(v) \leq \frac{5}{4} |H_v|$.
-\end{corollary}
-
-Now we are finally ready to make use of the properties we have proven about the weight function we introduced.
-
-\Ben{The constant in the lemmas below, but it is no more than twice that.}
-
-\begin{lemma}
-\label{lem:lightCost}
- Let $\mathcal{L}$ be a light shrub of a light path $r$.  Let $V$ be the set of heavy paths from the full shrub which are adjacent to paths in $\mathcal{L}$.  
-  Then $\sum_{p\in \mathcal{L}}\sum_{v\in p} \log |H_v| + \sum_{q\in V} |H_q|\log |H_q| \leq \frac{4}{3} |H_r|\log |H_r|$
-\end{lemma}
-\begin{proof}
-\Ben{
-The following is a proof for $\frac{3}{2}+\eps$, rather than $\frac{4}{3}$.  The discrepancy is due to a corrected calculation error.  
-4/3 can be achieved by tightening constants above, and is left in the lemma statement as it is used later.
-}
-
-Recall that in a proper coloring of a binary tree $T$, the color of any vertex must be the color of one of its descendants.  
-Therefore, any vertex in the light path shrub must appear in the heap of either the root of an adjacent 
-path from the full shrub or from the heap of a leaf of a path in $\mathcal{L}$.  
-Also note $|H_r|$ precisely counts sum of the number of vertices from these heaps that do not appear in $\mathcal{L}$.
-As there are no adjacent light paths to a light shrub, the total number of vertices over all paths in $\mathcal{L}$ is at most 
-$m = \sum_{p\in \mathcal{L}} |H_{p_f}| +\sum_{q\in V} |H_q| - |H_r| = w(r)-|H_r|$, where $H_{p_f}$ is the heap at the leaf of $p$.
-
-Using this observation and \Cor{equivalent} we then have, 
-\begin{align*}
- \sum_{p\in \mathcal{L}}\sum_{v\in p} \log |H_v| + \sum_{q\in V} |H_q|\log |H_q|
- &\leq \sum_{p\in \mathcal{L}}\sum_{v\in p} \log(w(r)) + \sum_{q\in V} |H_q|\log(w(r))\\
- &\leq \log(w(r))\pth{ w(r)-|H_r| } + \log(w(r))w(r)\\ 
- &= (2w(r)-|H_r|)\log(w(r)) \leq \pth{\frac{5}{2}|H_r|-|H_r|}\log\pth{\frac{5}{4}|H_r|}\\
- &= \frac{3}{2}|H_r|\log\pth{\frac{5}{4}|H_r|}
-\end{align*}
-\end{proof}
-
 \end{document}

### 2014-10-21 paint-contour-tree.tex
updates to section 2,3,4,5

@@
 \newcommand{\bT}{{\bf T}}
-\newcommand{\cS}{{\cal S}}
+
+\newcommand{\cA}{{\cal A}}
 \newcommand{\cB}{{\cal B}}
@@
 \newcommand{\cH}{{\cal H}}
+\newcommand{\cJ}{{\cal J}}
 \newcommand{\cL}{{\cal L}}
 \newcommand{\cP}{{\cal P}}
+\newcommand{\cS}{{\cal S}}
 \newcommand{\cU}{{\cal U}}
@@
 \newcommand{\mcol}{mcol}
+\newcommand{\merge}{{\tt merge}}
 \newcommand{\onestep}{{\tt piece}}
@@
 
-\section{Painting to compute contour trees} \label{sec:paint}
+\section{Contour trees of extremum dominant manifolds} \label{sec:paint}
 
@@
 We will orient so that the extremum in question is always a \emph{minimum}.
-We will fix such a simplicial complex $\MM$, with the dominant minimum $u$.
+We will fix such a simplicial complex $\MM$, with the dominant minimum $m^*$. 
+We observe that the contours trees of minimum dominant manifolds have a relatively simple description.
+This description is directly inspired by (and proven using) the \emph{join tree} of~\cite{csa-cctad-00}.
+We first explain what these trees are.
+
+For vertex $v$, we use $\MM^+_v$ to denote the simplicial complex obtained by only
+keeping vertices $u$ such that $f(u) > f(v)$. Analogously, define $\MM^-_v$. Note that $\MM^+$ may contain
+numerous connected components. The following definition of join and split trees were given by~\cite{csa-cctad-00}.
+Conventially, all edges are directed from higher to lower function value.
+
+\begin{definition} \label{join} The join tree $\cJ(\MM)$ of $\MM$ is built on vertex set $V(\MM)$.
+The directed edge $(u,v)$ is present when $u$ is the smallest valued vertex is a connected component of $\MM^+_v$
+\emph{and} $v$ is connected to this component (in $\MM$). The split tree $\cS(\MM)$ is obtained
+by looking at $\MM^-_v$ (or alternately, by taking the join tree of the inversion of $\MM$). 
+\end{definition}
+
+Note that unlike the contour tree, all vertices in $\MM$ are present in $\cJ(\MM)$ and $\cS(\MM)$. 
+Some basic facts about these trees. All outdegrees in $\cJ(\MM)$ and all indegrees in $\cS(\MM)$ are at most $1$.
+Indegree $2$ vertices in $\cJ(\MM)$ are joins, and outdegree $2$ vertices in $\cS(\MM)$ are splits.
+Leaves in $\cJ(\MM)$ are maxima, and leaves in $\cS(\MM)$ are minima.
+
+The key observation is that $\cS(\MM)$ is trivial for minimum dominant $\MM$.
+
+\begin{lemma} \label{lem:split} $\cS(\MM)$ consists of:
+\begin{asparaitem}
+	\item A single path (in sorted order) with all vertices except non-dominant minima.
+	\item Each non-dominant minimum is attached to a unique split.
+\end{asparaitem}
+\end{lemma}
+
+\begin{proof} First, let us complete the proof assuming that each split $v$ has one child that is just a leaf, which
+is a non-dominant minimum. Any minimum is a leaf in $\cS(\MM)$, and thereby attached to a split. If we removed
+all non-dominant minima, we must end up with a path, as asserted above.
+
+Consider a split $v$. For sufficiently small $\eps, \delta$, there are exactly two $(f(v) - \delta)$-contours
+$\phi$ and $\psi$ intersecting $B_\eps(v)$. Both of these are regular contours. There must be a non-ascending
+path from $v$ to the dominant minimum $m^*$. Consider the first edge (necessarily decreasing from $v$)
+on this path. It must intersect some contour, say $\phi$. By \Thm{jordan}, $\cut(\MM,\phi)$ has
+two connected components, with one (call it $\LL$) having $\phi^-$ as a boundary maximum. This complex
+contains $m^*$ as the non-ascending path $\phi$ intersects $\phi$ only once. Let the other
+component be called $\MM'$.
+
+Consider $\cut(\MM',\psi)$ with connected component $\NN$ having $\psi^-$ as a boundary. $\NN$
+does not contain $m^*$, so any path from the interior of $\NN$ to $m^*$ must intersect the boundary $\psi^-$.
+But the latter is a maximum in $\NN$, so there can be no non-ascending path from the interior to $m^*$.
+Since $\MM$ is overall minimum dominant, the interior of $\NN$ can only contain a single vertex $w$, a non-dominant
+minimum.
+
+The split $v$ has two children in $\cS(\MM)$, one in $\NN$ and one in $\LL$. The child in $\NN$ can only
+be the non-dominant minimum $w$, which is a leaf. 
+\end{proof}
+
+It is convenient to denote the non-dominant minima by $m_1, m_2, \ldots, m_k$, and the associated
+splits (as given by \Lem{split}) by $s_1, s_2, \ldots, s_k$.
+
+To prove our main theorem, we use the merging procedure from~\cite{csa-cctad-00} that constructs the contour tree
+from the join and split trees. 
+It actually construsts the \emph{augmented contour tree} $\cA(\MM)$, which
+is the contour tree with regular points inserted into the edges. 
+For any tree $T$ with vertex $v$, we define \emph{erasing} $v$ from $T$,
+when both in and outdegrees are at most $1$. If $v$ is a leaf, just delete $v$. Otherwise, smooth $v$ out
+(delete $v$ and connect its neighbors by an edge). This tree is denoted by $T \ominus v$.
+
+\medskip
+\fbox{
+\begin{minipage}{0.9\textwidth}
+{\bf $\merge(\cJ(\MM),\cS(\MM))$}
+
+\smallskip
+\begin{compactenum}
+	\item Set $\cJ = \cJ(\MM)$ and $\cS = \cS(\MM)$.
+	\item Denote $v$ as \emph{candidate} if sum of indegree of $v$ in $\cJ$ and outdegree of $v$ in $\cS$ is $1$. 
+	\item Add all candidates to queue.
+	\item While candidate queue is non-empty: 
+	\begin{compactenum}
+		\item Let $v$ be head of queue. If $v$ is leaf in $\cJ$, consider its edge in $\cJ$. Otherwise
+		consider edge in $\cS$. In either case, let edge by $(v,w)$.
+		\item Insert $(v,w)$ in $\cA(\MM)$. 
+		\item Set $\cJ = \cJ \ominus v$ and $\cS = \cS \ominus v$. Enqueue any new candidates.
+	\end{compactenum}
+	\item Smooth out all regular vertices in $\cA(\MM)$ to get $\cC(\MM)$.
+\end{compactenum}
+\end{minipage}}
+
+\medskip
+
+\begin{theorem} \label{thm:contour-tree} The contour tree $\cC(\MM)$ consists of all edges $\{(s_i, m_i)\}$
+as well as: for any critical point $v$, edge $(w,v)$ where $w$ is the lowest non-minimal critical point
+in a connected component of $\MM^+_v$ connected to $v$.
+\end{theorem}
+
+\begin{proof} We first show that $\cA(\MM)$ is $\cJ(\MM) \ominus \{m_i\}$ with edges $\{(s_i,m_i)\}$.
+We have flexibility in choosing the order of processing in $\merge$. We first put the non-dominant
+maxima $m_1, \ldots, m_k$ into the queue. As these are processed, the edges $\{(s_i,m_i)\}$ are inserted 
+into $\cA(\MM)$. Once all the $m_i$'s are erased, $\cS$ becomes a path, so all outdegrees are at most $1$.
+The join tree is now $\cJ(\MM) \ominus \{m_i\}$. We can now process $\cJ$ leaf by leaf, and all
+edges of $\cJ$ are inserted into $\cA(\MM)$.
+
+Note that $\cC(\MM)$ is obtained by smoothing out all regular points from $\cA(\MM)$. Smoothing out regular
+points from $\cJ(\MM) \ominus \{m_i\}$ exactly yields the edges described in the theorem.
+\end{proof}
+
+\section{Painting to compute contour trees}
 

### 2014-10-22 paint-contour-tree.tex
New figures

--- 2014-10-15 paint-contour-tree.tex
@@
 
-\title{Follow the Flow of Rain and Paint: \break An Output Sensitive Contour Tree Algorithm}
+\title{A Mountaintop View Requires Minimal Sorting: \break An Instance Optimal Contour Tree Algorithm}
 \date{}
@@
 Reeb graphs are a fundamental topological structure which tracks the evolution of levels sets of a real valued function over a manifold.  
-In the case when this graph is loop free it is called a contour tree.
-Due to the use of the contour trees in practice on large data sets, there has been significant previous work on algorithms for their efficient computation.
-
-Some of these previous results are describe as optimal, however they are only worst case optimal.
-Here we present an output sensitive approach to computing contour trees.
-Our approach leads to a faster algorithm when the contour tree is nicely behaved (i.e.
-more balanced), yet is still optimal in the worst case.
-Our approach is novel and we believe will lead to faster algorithms for other applications as well.
+In the case when this graph is loop free it is called a contour tree.  Due to the use of the contour trees in practice on large data sets, 
+there has been significant previous work in developing algorithms for their efficient computation.
+
+Here we present the first contour tree algorithm which takes into account the shape of the contour tree, 
+and in doing so yields instance optimal results (on all instances requiring $\Omega(n\alpha(n))$ time).
+In particular, our algorithm runs in $O(n\alpha(n))$ time when the tree is balanced, yet still 
+matches the previous worst case $O(n\log n)$ bound when the tree is long and skinny.
+
+In order to produce our instance optimal results we introduce a novel linear time surface partitioning scheme which is likely to have applications elsewhere.
+Moreover, in order to prove the optimality we use a non-trivial combinatorial discharging argument which we believe is of independent interest and in particular uses 
+such tools as heavy-light path decompositions.
 \end{abstract}
@@
 \section{Introduction}
-Describe what reeb graphs and contour trees are.
-
-\medskip
-\noindent
-Explain why useful:  Lets you know the ``shape of your date''.
-Map of the critical points where contours change.
-Fast extraction of seed sets in a small sized structure, which can let you do threshold queries.
-
-\medskip
-\noindent
-Paragraph on specific applications.  Scientific visualization, in particular combustion research.  GIS.
+
+Let $f$ be a real valued function defined over a manifold embedded in $\RR^d$.  
+For any real value $\alpha$, the level set $f^{-1}(\alpha)$ can be decomposed into its connected components, or ``contours''.
+The Reeb graph is a fundamental topological structure which represents the evolution of these contours as one sweeps the surface 
+from $-\infty$ to $\infty$ according to $f$.  Often this graph is acyclic, in which case it is 
+called a \emph{contour tree}.  This is the case if, for example, the input is a surface mesh of a mountain range 
+(and the contour tree would encode when mountain peeks form and how they merge together).
+
+There is a long list of practical applications of the study of contour trees (see for example \cite{c-tmi-04} and references therein).
+Often these applications involve massive data sets, and so consequently there has also been a long series 
+of results on efficient algorithms for their computation \cite{kobps-ctsssit-97,tv-cct-98,csa-cctad-00,pc-ectls-02,cllr-sooscctmp-05}.
+As an initial step these algorithms require that all $n$ contour tree vertices be sorted.  
+In particular, \cite{cllr-sooscctmp-05} provide an $O(n\log n)$ time algorithm, and since in the worst case computing the contour tree 
+requires $\Omega(n\log n)$ time, this algorithm is optimal with respect to the size of the output.  
+Here we present the first contour tree algorithm which is optimal (for instances requiring $\Omega(n\alpha(n))$ 
+time\footnote{Where $\alpha(n)$ is the inverse Ackermann function, which is $< 5$ for computation in this universe.}) 
+with respect not just to the size of the output, but also its structure.  This is accomplished by introducing a simple linear time 
+surface partitioning scheme to avoid sorting, which is likely to have applications elsewhere.  
+For all instances the bound for our algorithm either matches the best previous bound, or is faster.
+For, example when the contour tree is a balanced, our algorithm runs in $O(n\alpha(n))$ time, whereas all 
+previous algorithms had a running time of $\Omega(n\log n)$
+
+%In particular, the algorithm in \cite{cllr-sooscctmp-05} is worst case optimal with respect to the size of the output.  
+%Here we present the first contour tree algorithm which is (near\footnote{Our algorithm is optimal on all instance requiring $\Omega(n\alpha(n))$ time.}) 
+%optimal with respect not just to the size of the output, but also its structure.  
+%Specifically, the bound for our algorithm either matches \cite{cllr-sooscctmp-05}, or is faster.
+%For, example when the contour tree is a balanced, our algorithm runs in $O(n\alpha(n))$ time, whereas all 
+%previous algorithms had a running time of $\Omega(n\log n)$.
+
+This work can be seen as a continuation of the study of instance-optimal geometric algorithms introduced by Afshani \etal \cite{abc-ioga-09}.
+In order to prove instance optimality we use a non-trivial combinatorial discharging argument, which we believe is of independent interest.
+Of particular note is the use of heavy-light path decompositions, first introduced by Sleator and Tarjan \cite{st-dsdt-83} in their analysis of link-cut trees.
+
+% Explain why useful:  Lets you know the ``shape of your date''.  Map of the critical points where contours change.  
+%Fast extraction of seed sets in a small sized structure, which can let you do threshold queries.
+% 
+% \medskip\noindent
+% Paragraph on specific applications.  Scientific visualization, in particular combustion research.  GIS.
 
@@
 
-\paragraph{Contour Trees.}
-As the application areas of contour trees usually involve very large data sets, highly efficient algorithms are not only desirable but often necessary.
-As such there have been a long sequence of papers with improved running times for computing contour trees.
-Kreveld \etal \cite{kobps-ctsssit-97} presented an $O(N\log N)$ time algorithm for functions over 2D meshes and an $O(N^2)$ algorithm for higher dimensions, where $N$ is the number of simplices in the mesh.
-Tarasov and Vyalya \cite{tv-cct-98} improved the running time to $O(N\log N)$ for the 3D case by making several passes over the input.
-
-Carr \etal \cite{csa-cctad-00} improved the running time for all dimensions to $O(n\log n + N)$, where $n$ is the number of mesh vertices.
-Their algorithm works in 3 simple stages.
-The first stage consist of a top down sweep to compute what is known as the merge (or join) tree, which tracks the evolution of super-level sets (whereas the contour tree tracks level sets).
+\subsubsection{Worst Case Alternatives}
+In worst case analysis an algorithm's running time is judged in terms of its worst 
+case behavior over all inputs of a given size.  While there are benefits to such a 
+pessimistic view, it fails to capture the whole story.  As such, many alternatives 
+have been considered.  For example, one can make assumptions on the distribution of 
+inputs, which has led to a number interesting results on average case and probabilistic 
+analysis, and also smoothed analysis \cite{st-saa-04}.  However, even without making 
+assumptions on the distribution of the input, there is still more to the story.
+For example, in the study of parametrized complexity, one takes into account other 
+parameters determined by the input other than just its size (for example the maximum 
+degree in a graph or whether it is planar).
+
+\paragraph{Output Sensitive Analysis.}
+In geometric settings often the goal is to compute a (potentially complex) geometric structure determined 
+by the input, and so it is natural to also consider how the form of this output structure affects the running time.
+One example is the standard sweep line algorithm of Bentley and Ottmann \cite{bo-arcgi-79} for line segment intersection, 
+which runs $O((n+k)\log n)$ time where k is the number of intersections, which in the worst case is quadratic.
+Another example is Chan's algorithm for computing the convex hull \cite{c-ooscha-96} which runs in $O(n\log h)$ 
+where $h$ is the number of vertices on the hull, which in the worst case is $n$.  While such algorithms represent 
+significant improvements over their worst case counterparts, their running times are only measured with respect 
+to the size of the output and not any structure it may have.
+
+\paragraph{Instance Optimality.}  
+Here we are concerned with a more refined analysis than the typical output sensitive analysis.
+Specifically, we investigate an algorithm which is instance optimal as defined by Afshani \etal \cite{abc-ioga-09}.  
+In order to prove such a result, one considers a family of allowable partitions of the input, 
+each of which is defined in terms of the structure of the output.  Each such partition then has 
+an associated entropy type cost representing the number of possible instantiations of such a partition.
+Our work differs slightly from Afshani \etal \cite{abc-ioga-09} however, in that they show any algorithm 
+(for say the convex hull) requires the minimum such cost, where as in our case it is shown that 
+any algorithm requires the maximum over these costs.
+
+\subsubsection{Contour Trees}
+
+\Ben{Maybe discuss contour tree applications and why they are useful here? Or maybe earlier?}
+
+\paragraph{A History of Contour Trees.}
+As discussed above, there has been a long series of improved algorithms for computing contour trees.
+Kreveld \etal \cite{kobps-ctsssit-97} presented an $O(N\log N)$ time algorithm for functions over 2D 
+meshes and an $O(N^2)$ algorithm for higher dimensions, where $N$ is the number of simplices in the mesh. 
+Tarasov and Vyalya \cite{tv-cct-98} improved the running time to $O(N\log N)$ for the 3D case by making 
+several passes over the input.
+
+Carr \etal \cite{csa-cctad-00} improved the running time for all dimensions to $O(n\log n + N)$, 
+where $n$ is the number of mesh vertices.  Their algorithm works in 3 simple stages.  
+The first stage consist of a top down sweep to compute what is known as the join (or merge) tree, 
+which tracks the evolution of super-level sets (whereas the contour tree tracks level sets).  
 Similarly, the second stage does a bottom up sweep to compute the split tree, which tracks sub-level sets.  
-Their main contribution is the third and final step where
-they present an $O(n)$ time algorithm to merge together the merge and split trees to get the contour tree.  Chiang \etal \cite{cllr-sooscctmp-05} 
-have since
-
-%The most expensive part of the Carr \etal algorithm comes from sorting.  They sort all vertices of the mesh to facilitate an easing sweeping of the mesh.
-%asdfas \cite{} improve the running to $O()$ by sorting only critical points of the mesh.  However, without looking at the non-critical points in the mesh 
-%one does not know how the critical points connect to each other in the contour tree.  To overcome this \cite{} use ``mountain climbers'' from each saddle 
-%to determine which critical points they belong to above.
-
-\paragraph{Output Sensitivity.}
-Let $t$ be the number of critical points in the input mesh.
-There is room to improve the results of Carr \etal, as $t$ could potentially be much smaller than $n$.
-In particular, Cole-McLaughlin and Pascucci \cite{pc-ectls-02} provide an $O(n+t\log n)$ time algorithm for the case of 3 dimensional structured meshes.
-Chiang \etal \cite{cllr-sooscctmp-05} improve upon this result, and provide an $O(N+t\log t)$ which works for structured or unstructured meshes in any dimension.
-
-As the the authors point out in \cite{cllr-sooscctmp-05}, by reduction from sorting, one can show that any algorithm to compute the contour tree in the worst case takes $\Omega(N+t\log t)$ time.
-However, this lower bound only holds when the contour tree is very unbalanced (i.e.
-long and skinny) and therefore contains a sorted path with a large fraction of the vertices.
-Therefore the bound of $\Omega(N+t\log t)$ is only worst case optimal.
-In this paper we will consider an output sensitive approach which takes into account the shape of the contour tree.
-
-\paragraph{Reeb Graphs.} UNDER CONSTRUCTION.
-There have also been a number of result for the more general Reeb graph, i.e.
-when the loopless assumption is removed.
-Shinagawa and Kunii \cite{sk-crgacs-91} provided the first provably correct algorithm for the case of triangulated $2$-manifolds.
-Cole-McLaughlin \etal \cite{cehnp-lrbm-03} later improved the running time to $O(n\log n)$.
-
-\subsection{Our Approach and Contribution.}
-
-\paragraph{To Sort or not to Sort.}
-As discussed above, the algorithm of Carr \etal \cite{csa-cctad-00} computes the contour tree by first constructing the merge and split trees. 
-As vertices appear in sorted order in any root to leaf path in either tree, sorting the input vertices is a natural step.
-However, not all vertices must be sorted together.
-In particular, Chiang \etal \cite{cllr-sooscctmp-05} improved the running time by observing that as the merge and split tree only consists of critical points, it suffices to only sort critical points.
-%
-%However, without looking at the non-critical points in the mesh 
-%one does not know how the critical points connect to each other in the contour tree.
-\cite{} easily overcome this by sending out increasing monotone paths (i.e.
-``mountain climbers'') from each critical point to determine which critical points they connect to above.
-%
-If a large fraction of the vertices appear in a single root to leaf path in the merge or split tree, paying to sort all the critical 
-points is unavoidable in order to compute the merge and split trees.  However, if the merge and split trees are balanced then sorting all the critical points is potentially very costly.
-
-There is also a more fundamental problem with the now standard approach of first computing the merge and split trees in order to compute the contour tree.
-In particular, this approach can never lead to an algorithm that is optimal with respect to the shape of the contour tree.
-This is because it is possible to have merge and split trees which are long and skinny (i.e.
-costly to compute) whereas the contour tree is balanced.
-For example consider the case when the mesh looks like two (thickened) balanced binary trees, one upside down, and joined at their roots.
-In this case the contour tree is potentially cheap to compute as there is no long sorted path.
-However, the merge and split trees will have long sorted tails (consisting of the vertices in the bottom and top balanced trees, respectively), and therefore require $\Omega(n\log n)$ time to compute.
-
-\paragraph{Our Contribution.} UNDER CONSTRUCTION.
-Give a high level overview.
-Emphasize the spilling partitioning scheme used to avoid sorting, and say it may be useful else where.
-State our running time.
-Say it may be optimal (we still need to investigate).
-It matches previous on worst case (i.e.
-optimal in worst case).
+Their main contribution is the third and final step where they present an $O(n)$ time algorithm to merge 
+together the join and split trees to get the contour tree.
+
+There have also been some output sensitive results for computing the contour tree.  Let $t'$ be the number 
+of critical points in the input mesh, which could potentially be much smaller than $n$.  In particular, 
+Cole-McLaughlin and Pascucci \cite{pc-ectls-02} provide an $O(n+t\log n)$ time algorithm for the case of 
+$3$ dimensional structured meshes.  \Ben{Update distinction between critical and component-critical}
+Chiang \etal \cite{cllr-sooscctmp-05} improve upon this result, and provide an $O(N+t\log t)$ algorithm 
+which works for structured or unstructured meshes in any dimension, which is provably optimal with respect 
+to the output size but not the output structure, as discussed below.
+
+%As the the authors point out in \cite{cllr-sooscctmp-05}, by reduction from sorting, 
+%one can show that any algorithm to compute the contour tree in the worst case takes 
+%$\Omega(N+t\log t)$ time.  However, this lower bound only holds when the contour tree 
+%is very unbalanced (i.e. long and skinny) and therefore contains a sorted path with a 
+%large fraction of the vertices.  Therefore the bound of $\Omega(N+t\log t)$ is only 
+%worst case optimal.  In this paper we will consider an output sensitive approach which 
+%takes into account the shape of the contour tree.
+
+\begin{figure}[h]\centering
+    \includegraphics[width=.27\linewidth]{figs/spike3}%
+    \hspace{.35in}
+    \includegraphics[width=.27\linewidth]{figs/spike4}%
+    \hspace{.4in}
+    \includegraphics[width=.22\linewidth]{figs/spikeTree2}
+    \caption{Two surfaces with different orderings of the maxima, but the same contour tree.}
+    \label{fig:relative}
+\end{figure}
+
+\paragraph{The Issue of Sorting.}
+As discussed above, the algorithm of Carr \etal \cite{csa-cctad-00} computes the contour 
+tree by first constructing the join and split trees.  As vertices appear in sorted order 
+in any root to leaf path in the join and split trees, and in any maxima to minima path 
+in the contour tree, sorting the input vertices is a natural first step.  However, not 
+all vertices from the input need to be sorted.  In particular, Chiang \etal 
+\cite{cllr-sooscctmp-05} improved the running time by observing that as the join and 
+split trees only consist of critical points, it suffices to only sort critical points.  
+If a large fraction of the vertices appear in a single path in the join or split tree, 
+then paying to sort all the critical points is unavoidable, and therefore in the worst 
+case \cite{cllr-sooscctmp-05} is optimal.  However, if the trees are more balanced then 
+sorting all critical points together can be very costly.  Specifically, the contour tree 
+does not care about the relative ordering of vertices which do not have an 
+ancestor-descendant relationship, and so this information does not need to be computed.
+See \Fig{relative} for an example.
+
+\begin{figure}[h]\centering
+    \includegraphics[width=.3\linewidth,height=.42\linewidth]{figs/shaded}%
+    \hspace{.5in}
+    \includegraphics[width=.3\linewidth,height=.42\linewidth]{figs/trees2}
+    \caption{On the left, a surface will a balanced contour tree, but whose merge and split trees have long tails.  
+    On the right (in clockwise order from the top) the contour, split and merge trees.}
+    \label{fig:sorting}
+\end{figure}
+
+\paragraph{Suboptimal Join and Split Trees.}
+There is also a more fundamental problem with the now standard approach of first computing 
+the join and split trees in order to compute the contour tree.  In particular, this approach 
+can never lead to an algorithm that is optimal with respect to the shape of the contour tree.  
+This is because it is possible to have join and split trees which are long and skinny (i.e. 
+costly to compute) whereas the contour tree is balanced.  For example consider the case shown 
+in \Fig{sorting}, where the mesh looks like two (thickened) balanced binary trees, one upside 
+down, and joined at their roots.  In this case the contour tree is potentially cheap to 
+compute as there is no long sorted path.  However, the join and split trees will have long 
+sorted tails (consisting of the vertices in the bottom and top balanced trees, respectively), 
+and therefore require $\Omega(n\log n)$ time to compute.
+
+\subsection{Our Contribution}
+\Ben{To be Written}
+
+\paragraph{Organization.}
+
+\begin{figure}[h]\centering
+    \includegraphics[width=.18\linewidth]{figs/orderGray1}%
+    \hspace{.8in}
+    \includegraphics[width=.18\linewidth]{figs/orderGray2}
+    \caption{On left, downward paint spilling only (each shade of gray represents a different color).  On right, flipping the direction of paint spilling.}
+    \label{fig:order}
+\end{figure}
 
@@
 
-\subsection{Shrubs, heavy paths, and light paths}
+\subsection{Shrubs, tall paths, and short paths}
 
@@
 \begin{definition}
- For $p\in P(T)$, $p$ is \emph{light} if $|p| < \sqrt{|H_{p}|}/100$, and \emph{heavy} otherwise.
+ For $p\in P(T)$, $p$ is \emph{short} if $|p| < \sqrt{|H_{p}|}/100$, and \emph{tall} otherwise.
 \end{definition}
@@
 
-The following lemma demonstrates that heavy paths can ``pay'' for themselves.
+The following lemma demonstrates that tall paths can ``pay'' for themselves.
 
 \begin{lemma}
-\label{lem:pathBounds} If $p$ is heavy, $\sum_{v\in p} \log |H_v| = O(|p| \log |p|)$.
-If $p$ is light, $\sum_{v\in p} \log |H_v| = O(|H_p| \log |H_p|)$.
+\label{lem:pathBounds} If $p$ is tall, $\sum_{v\in p} \log |H_v| = O(|p| \log |p|)$.
+If $p$ is short, $\sum_{v\in p} \log |H_v| = O(|H_p| \log |H_p|)$.
 % Let T be a binary tree and let $\chi(T)$ and $P(T)$ be a valid coloring and path decomposition of $T$, respectively.
@@
 % \begin{compactenum}
-%  \item for any heavy path $p\in P(T)$, $\sum_{v\in p} \log |H_v| = O(|p| \log |p|)$.
-%  \item for any light path $p\in P(T)$, $\sum_{v\in p} \log |H_v| = O(|p| \log |H_{p}|)$.
+%  \item for any tall path $p\in P(T)$, $\sum_{v\in p} \log |H_v| = O(|p| \log |p|)$.
+%  \item for any short path $p\in P(T)$, $\sum_{v\in p} \log |H_v| = O(|p| \log |H_{p}|)$.
 % \end{compactenum}
@@
 Hence, $\sum_{v\in p} \log |H_v| \leq \sum_{v\in p} \log(|H_{p}| + |p|)  = |p| \log (|H_{p}| + |p|)$.
- If $p$ is a heavy path, then $|p| \log (|H_{p}| + |p|) = O(|p| \log |p|)$. If $p$ is light, then 
- $|p| \log (|H_{p}| + |p|) = O(|p| \log |H_{p}| )$. For light paths, $|p| = O(|H_p|)$.
+ If $p$ is a tall path, then $|p| \log (|H_{p}| + |p|) = O(|p| \log |p|)$. If $p$ is short, then 
+ $|p| \log (|H_{p}| + |p|) = O(|p| \log |H_{p}| )$. For short paths, $|p| = O(|H_p|)$.
 \end{proof}
@@
 \ignore{
-For a light path $p$ we can think of the quantity $|p|(\log |H_p| - \log |p|)$ as the excess of $p$, i.e. what was not ``paid for'' locally by $p$.
+For a short path $p$ we can think of the quantity $|p|(\log |H_p| - \log |p|)$ as the excess of $p$, i.e. what was not ``paid for'' locally by $p$.
 Our eventual goal will be to show that by choosing the right path decomposition, this excess 
-can be recharged to heavy paths. In the next couple sections we introduce the appropriate path decomposition and prove 
+can be recharged to tall paths. In the next couple sections we introduce the appropriate path decomposition and prove 
 some useful facts about it.  Then given these facts, in \Sec{} we are able to make the recharging argument.
@@
 
-There are some light paths that we can also ``pay" for.  Consider any light path $p$ in the shrub.  We will refer to the 
-\emph{heavy support chain} of $p$ as the heavy ancestors of $p$ in the shrub which have a path to $p$ which does not use any 
-light path (i.e. it is a chain of paths adjacent to $p$).
-
-\begin{definition} \label{def:support} A light path $p$ is \emph{supported} if
-at least $|H_p|/100$ vertices $v$ in $H_p$ lie in paths in the heavy support chain of $p$.
+There are some short paths that we can also ``pay" for.  Consider any short path $p$ in the shrub.  We will refer to the 
+\emph{tall support chain} of $p$ as the tall ancestors of $p$ in the shrub which have a path to $p$ which does not use any 
+short path (i.e. it is a chain of paths adjacent to $p$).
+
+\begin{definition} \label{def:support} A short path $p$ is \emph{supported} if
+at least $|H_p|/100$ vertices $v$ in $H_p$ lie in paths in the tall support chain of $p$.
 %
@@
 %a contiguous (from $p$) chain of ancestors $q_1, q_2, \ldots, q_k$ (in $P(T)$) such that $|H_{q_k}| \leq 99|H_p|/100$
-%and all $q_i$'s are heavy.
+%and all $q_i$'s are tall.
 \end{definition}
 
-Let $\cL$ be the set of light paths, $\cL'$ be the set of supported light paths, and $\cH$ be the set of heavy paths
+Let $\cL$ be the set of short paths, $\cL'$ be the set of supported short paths, and $\cH$ be the set of tall paths
 given by $\pmax(T)$.
-We now construct the shrub of unsupported light paths. Consider $p \in \cL \setminus \cL'$,
-and traverse the chain of ancestors from $p$. Eventually, we must reach another light path $q$.
+We now construct the shrub of unsupported short paths. Consider $p \in \cL \setminus \cL'$,
+and traverse the chain of ancestors from $p$. Eventually, we must reach another short path $q$.
 (If not, we have reached the root $r$ of $\pmax(T)$. Hence, $p$ is supported.)
 Insert edge from $p$ to $q$, so $q$ is the parent of $p$ in $\cU$. This construction leads
-to the shrub forest of $\cL \setminus \cL'$, where all the roots are supported light paths, 
-and the remaining nodes are the unsupported light paths.
+to the shrub forest of $\cL \setminus \cL'$, where all the roots are supported short paths, 
+and the remaining nodes are the unsupported short paths.
 
@@
 
-We split the summation into heavy, light, and unsupported light paths.
+We split the summation into tall, short, and unsupported short paths.
 \begin{eqnarray*} 
@@
 Pick $p \in \cL'$. 
-%Let the first $|H_p|/200$ vertices the lie in heavy paths be denoted as $H'_p$. 
+%Let the first $|H_p|/200$ vertices the lie in tall paths be denoted as $H'_p$. 
 %Observe that for any $v \in H'_p$, 
 %We have $|H_p| \log |H_p| = O(\sum_{v \in H'_p} \log |H_p|)$
-%$= O(\sum_{v \in H'_p} \log |H_v|)$. Since all the vertices lie in heavy paths,
+%$= O(\sum_{v \in H'_p} \log |H_v|)$. Since all the vertices lie in tall paths,
 %we can write this as $O(\sum_{q \in \cH} \sum_{v \in H'_p \cap q} \log |H_v|)$.
@@
 %
-As we traverse the heavy support chain of $p$, there
+As we traverse the tall support chain of $p$, there
 are at least $|H_p|/100$ vertices of $H_p$ that lie in these paths. These are encountered in
@@
 $|H_v| \geq |H_p|/200$. Hence, $|H_p|\log |H_p| = O(\sum_{v \in H'_p} \log |H_v|)$.
-Since all the vertices lie in heavy paths,
+Since all the vertices lie in tall paths,
 we can write this as $O(\sum_{q \in \cH} \sum_{v \in H'_p \cap q} \log |H_v|)$.
@@
 First observe that any two paths in $S$ must be unrelated (i.e. $S$ is an anti-chain), 
-since paths which have an ancestor-descendant relationship have disjoint heavy support chains.  
+since paths which have an ancestor-descendant relationship have disjoint tall support chains.  
 However, any vertex $v$ receives exactly one color from each of its two subtrees (in $T$), and therefore $|S|\leq 2$ since any two 
@@
 
-The first term (unsupported light paths) can be charged to the second term (supported light paths).
+The first term (unsupported short paths) can be charged to the second term (supported short paths).
 This is where the critical \Lem{cu-root} plays a role.
@@
 We have $\sum_{p \in \cU} \sum_{v \in p} \log |H_v| = O((\log |H_r|) \sum_{p \in \cU} |p|) = O(|H_r|\log |H_r|)$.
-We sum this over all $\cU$ in the shrub forest, and note that roots in the shrub forest are supported light paths.
+We sum this over all $\cU$ in the shrub forest, and note that roots in the shrub forest are supported short paths.
 \end{proof}
@@
 \ignore{
-Now, for the critical piece of the approach. Focus on $\pmax(T)$. Let $L$ be the set of light paths, 
+Now, for the critical piece of the approach. Focus on $\pmax(T)$. Let $L$ be the set of short paths, 
 and $\hat{L} \subseteq L$ be the set of paths such that no ancestor of $p \in L'$
-is light. Obviously, $\hat{L}$ is an anti-chain in $P(T)$.
-
-\begin{lemma} \label{lem:light} For the path decomposition $\pmax(T)$, $\sum_{p \in \cL} \sum_{v \in p} \log |H_v| = O(\sum_{p \in \cL'} |H_p| \log |H_p|)$.
+is short. Obviously, $\hat{L}$ is an anti-chain in $P(T)$.
+
+\begin{lemma} \label{lem:short} For the path decomposition $\pmax(T)$, $\sum_{p \in \cL} \sum_{v \in p} \log |H_v| = O(\sum_{p \in \cL'} |H_p| \log |H_p|)$.
 \end{lemma}
@@
 
-\begin{proof} (of \Thm{runtime}) We focus on $\pmax(T)$ and split into heavy and light paths.
+\begin{proof} (of \Thm{runtime}) We focus on $\pmax(T)$ and split into tall and short paths.
 We have $\sum_{v \in T} \log |H_v| =$  $\sum_{p \in \cL} \sum_{v \in p} \log |H_v| + $ $\sum_{p \in \cH} \sum_{v \in p} \log |H_v|$.
-For the first term, we apply \Lem{light} to bound by $O(\sum_{p \in \cL'} |H_p| \log |H_p|)$.
+For the first term, we apply \Lem{short} to bound by $O(\sum_{p \in \cL'} |H_p| \log |H_p|)$.
 Then we apply \Clm{above} to bound by $O(\sum_{p \in \cH} |p|\log |p|)$.
@@
 
-The main challenge is proving \Lem{light}, which we defer to the next section.
+The main challenge is proving \Lem{short}, which we defer to the next section.
 \begin{definition}
@@
 We will work with the \emph{reduced} heap $\redH_p$. This is the subset of vertices of $H_p$ that 
-do not appear on the heavy support chain of $p$.
-%only appear in light paths. 
-By definition, for any unsupported light path
+do not appear on the tall support chain of $p$.
+%only appear in short paths. 
+By definition, for any unsupported short path
 (hence, any non-root $p \in \cU$), $|\redH_p| \geq 99|H_p|/100$. 
@@
 \end{lemma}
-\begin{proof} Let $h(q)$ denote the heavy path that is a child of $p$ in $\pmax(T)$,
-and an ancestor of $q$. If no such heavy path exists, then by construction $p$ is the parent 
+\begin{proof} Let $h(q)$ denote the tall path that is a child of $p$ in $\pmax(T)$,
+and an ancestor of $q$. If no such tall path exists, then by construction $p$ is the parent 
 of $q$ in $\pmax(T)$, and the following argument will go through by setting $h(q)=q$.
 
-The chain of ancestors from $q$ to $h(q)$ consists only of heavy paths.
+The chain of ancestors from $q$ to $h(q)$ consists only of tall paths.
 Since $q$ is unsupported, these paths contain at most $|H_q|/100$ vertices of $H_q$. 
@@
 $|H_u| \geq 2|H_w| - 2$. Let $b$ be the base of $p$.
-We have $|H_p| = |H_b| \geq |H_u| - |p| \geq 2|H_w| - |p| - 2$. Since $p$ is a light path, $|p| < \sqrt{|H_p|}/100$.
+We have $|H_p| = |H_b| \geq |H_u| - |p| \geq 2|H_w| - |p| - 2$. Since $p$ is a short path, $|p| < \sqrt{|H_p|}/100$.
 Applying this bound, we get $|H_p| \geq (2-\delta)|H_w|$ (for a small constant $\delta > 0$).
@@
 % Therefore, by \Obs{decrease}, $|H_p|+ d(r_p, w) \geq |H_w|\geq 2|H_q|-1$.  
-% Since $p$ is a light path, $d(r_p, w)\leq \sqrt{|H_p|}/100$ and so we have $|H_p|+\sqrt{|H_p|}/100\geq 2|H_q|-1$.
+% Since $p$ is a short path, $d(r_p, w)\leq \sqrt{|H_p|}/100$ and so we have $|H_p|+\sqrt{|H_p|}/100\geq 2|H_q|-1$.
 % Assuming $|H_p|>11$ (and we can safely assume $|H_p|$ is bigger than any constant), this yields $\frac{3}{5} |H_p| \geq |H_q|$
@@
 Traverse $\pmax(T)$ from $q'_i$ to $q$.
-Other than $q$, all other nodes encountered are in the heavy support chain of $q'_i$ and hence
+Other than $q$, all other nodes encountered are in the tall support chain of $q'_i$ and hence
 do not affect its reduced heap.
@@
 are removed from $H_p$ are exactly the vertices that lie in these paths. 
-We split this contribution into heavy and light paths, observing that the latter
+We split this contribution into tall and short paths, observing that the latter
 are all unsupported. Hence, we can (crudely) upper bound the number of vertices
@@
 
-\begin{proof} Since $q$ is an unsupported light path, $|q| < \sqrt{|H_q|}/100 \leq \sqrt{|\redH_q|}/99 \leq \sqrt{|W_q|}/99$.
+\begin{proof} Since $q$ is an unsupported short path, $|q| < \sqrt{|H_q|}/100 \leq \sqrt{|\redH_q|}/99 \leq \sqrt{|W_q|}/99$.
 We prove that $\sum_{q \in \cU_p} \sqrt{|W_q|}/99 \leq W_p/20$
@@
 equation. We use the assumption that $W_{a_i} \geq (4/3) W_{a_{i-1}}$ and hence $W_{a_i} \geq (4/3)^i W_{a_0} \geq (4/3)^i$, 
-as $a_0$ is an unsupported light path and hence $W_{a_0}\geq 1$. 
+as $a_0$ is an unsupported short path and hence $W_{a_0}\geq 1$. 
 $$ ({R^+_q}/99)\sum_{a_i} 1/\sqrt{W_{a_i}} \leq ({R^+_q}/99)\sum_{a_i} (3/4)^{i/2} \leq R^+_q/20 $$
@@
 % 
-% \begin{proof} Since $q$ is an unsupported light path, $|q| < \sqrt{|H_q|}/100 \leq \sqrt{|\redH_q|}/99 \leq \sqrt{|W_q|}/99$.
+% \begin{proof} Since $q$ is an unsupported short path, $|q| < \sqrt{|H_q|}/100 \leq \sqrt{|\redH_q|}/99 \leq \sqrt{|W_q|}/99$.
 % We prove that $\sum_{q \in \cU_p} \sqrt{|W_q|}/99 \leq W_p/20$
@@
 % equation. We use the assumption that $W_{a_i} \geq (4/3) W_{a_{i-1}}$ and hence $W_{a_i} \geq (4/3)^i W_{a_0} \geq (4/3)^i$, 
-% as $a_0$ is an unsupported light path and hence $W_{a_0}\geq 1$. 
+% as $a_0$ is an unsupported short path and hence $W_{a_0}\geq 1$. 
 % $$ ({R^+_q}/99)\sum_{a_i} 1/\sqrt{W_{a_i}} \leq ({R^+_q}/99)\sum_{a_i} (3/4)^{i/2} \leq R^+_q/20 $$
@@
 \label{lem:excess}
- Let $p$ be a heavy path, and let $S_1, \dots S_k$ and $R_1, \dots, R_l$ be the sizes of the heaps at the roots of the light and heavy children of $p$, respectively, in the shrub.  
+ Let $p$ be a tall path, and let $S_1, \dots S_k$ and $R_1, \dots, R_l$ be the sizes of the heaps at the roots of the short and tall children of $p$, respectively, in the shrub.  
  Let $H_f$ be the heap at the leaf of the path $p$.
@@
  large enough such that $c|p|\log|p|$ will dominate all other terms.
- Next, observe that $\sum_{v\in p} \log|H_v| \leq \sum_{v\in p} \log(|H_v|+|p|) = |p|\log(|H_v|+|p|) = O(|p|\log|p|)$, as $p$ is a heavy path.  
+ Next, observe that $\sum_{v\in p} \log|H_v| \leq \sum_{v\in p} \log(|H_v|+|p|) = |p|\log(|H_v|+|p|) = O(|p|\log|p|)$, as $p$ is a tall path.  
  Therefore, in the remainder of the proof  we can safely ignore the $\sum_{v\in p} \log|H_v|$ term as we can choose $c$ large enough to cover it.  
@@
 % \label{lem:excess}
-%  Let $p$ be a heavy path, and let $S_1, \dots S_k$ and $R_1, \dots, R_l$ be the sizes of the heaps at the roots of the light and heavy children of $p$, respectively, in the shrub.  
+%  Let $p$ be a tall path, and let $S_1, \dots S_k$ and $R_1, \dots, R_l$ be the sizes of the heaps at the roots of the short and tall children of $p$, respectively, in the shrub.  
 %  Let $H_f$ be the size of the heap at the leaf of the path $p$, and let $X = \sum_{i=0}^k H_i \log H_i$.  
@@
 % By \Lem{adjacent} we know that $|H_p|+|p| \geq  H_f + \sum_{i=0}^k S_i + \sum_{i=0}^l R_i$.  
-% %Now as $p$ is a heavy path we have $|p|\geq \sqrt{|H_p|}/100$, and so, $(|H_p|+|p|)\log(|H_p|+|p|)$
+% %Now as $p$ is a tall path we have $|p|\geq \sqrt{|H_p|}/100$, and so, $(|H_p|+|p|)\log(|H_p|+|p|)$
 % 
@@
 
-Define the crust of the shrub to be the first (i.e. closest to the root) layer of light paths in the shrub.  Apply \Lem{lightCost} to each of the light paths in the crust and then 
+Define the crust of the shrub to be the first (i.e. closest to the root) layer of short paths in the shrub.  Apply \Lem{shortCost} to each of the short paths in the crust and then 
 say by \Lem{excess} and since the root heap goes to zero we are done.
@@
 \ignore{
-Naturally, this geometric series behavior can be extended to more than two adjacent light paths, thus motivating 
+Naturally, this geometric series behavior can be extended to more than two adjacent short paths, thus motivating 
 the following definition.
@@
  Let $p$ be a path in a path decomposition $P(T)$, with corresponding shrub $\pathTree$.
- Call a light path $q\in P(T)$ light reachable from $p$ if the path from $p$ to $q$ in $\pathTree$ 
- contains only vertices which correspond to light paths.  Call the subtree of $\pathTree$ consisting 
- of all light reachable paths from $p$ the \emph{light shrub} of $p$, and denote it $\mathcal{L}(p)$.
+ Call a short path $q\in P(T)$ short reachable from $p$ if the path from $p$ to $q$ in $\pathTree$ 
+ contains only vertices which correspond to short paths.  Call the subtree of $\pathTree$ consisting 
+ of all short reachable paths from $p$ the \emph{short shrub} of $p$, and denote it $\mathcal{L}(p)$.
 \end{definition}
 
-In the following subsection we show how to exploit this geometric series behavior of light shrubs.
+In the following subsection we show how to exploit this geometric series behavior of short shrubs.
 
@@
 \remove{
-\Ben{Maybe move the remainder of this subsection to 5.3 since heavy and light paths and not used until then}
+\Ben{Maybe move the remainder of this subsection to 5.3 since tall and short paths and not used until then}
 
@@
  Let T be a binary tree and let $\chi(T)$ and $P(T)$ be a valid coloring and path decomposition of $T$, respectively.  
- For $p\in P(T)$, we say that $p$ is \emph{light} if $|p| < \sqrt{|H_{r_p}|}/100$, and we say it is \emph{heavy} otherwise.
+ For $p\in P(T)$, we say that $p$ is \emph{short} if $|p| < \sqrt{|H_{r_p}|}/100$, and we say it is \emph{tall} otherwise.
 \end{definition}
@@
  \begin{compactenum}
-  \item for any heavy path $p\in P(T)$, $\sum_{v\in p} \log |H_v| = O(|p| \log |p|)$.
-  \item for any light path $p\in P(T)$, $\sum_{v\in p} \log |H_v| = O(|p| \log |H_{r_p}|)$.
+  \item for any tall path $p\in P(T)$, $\sum_{v\in p} \log |H_v| = O(|p| \log |p|)$.
+  \item for any short path $p\in P(T)$, $\sum_{v\in p} \log |H_v| = O(|p| \log |H_{r_p}|)$.
  \end{compactenum}
@@
  \]
- If $p$ is a heavy path then $|p| \log (|H_{r_p}| + |p|) = O(|p| \log |p|)$, and if $p$ is light then 
+ If $p$ is a tall path then $|p| \log (|H_{r_p}| + |p|) = O(|p| \log |p|)$, and if $p$ is short then 
  $|p| \log (|H_{r_p}| + |p|) = O(|p| \log |H_{r_p}| )$.
@@
 % Using the notation defined in the above lemma, we have the following crucial lemma, which we will use to 
-% argue we can charge that cost of all paths in $P(T)$ to the cost of the heavy paths.
+% argue we can charge that cost of all paths in $P(T)$ to the cost of the tall paths.
 % 
 % \begin{lemma}
-% Let $p$ be a heavy path in $P(T)$ and let $\{q_1, \dots q_k\}$ be any set of light paths in $P(T)$ such that 
+% Let $p$ be a tall path in $P(T)$ and let $\{q_1, \dots q_k\}$ be any set of short paths in $P(T)$ such that 
 % $\sum_{i=1}^k |H_i| = O(H_p)$.  Then $\sum_{i=1}^k \sum_{v\in q_i} \log |H_v| = O(|p|\log |p|)$
@@
 % By \Lem{pathBounds}, for all $1\leq i\leq k$, 
-%  $\sum_{v\in q_i} \log |H_v| = O(|q_i|\log |H_i|) = O(|H_i|\log |H_i|)$, as $q_i$ is a light path.
+%  $\sum_{v\in q_i} \log |H_v| = O(|q_i|\log |H_i|) = O(|H_i|\log |H_i|)$, as $q_i$ is a short path.
 %  By assumption $\sum_{i=1}^k |H_i| = O(|H_p|)$, 
 %  As $x\log x$ is a convex function, Jensen's inequality then implies 
-%  $\sum_{i=1}^{k} |H_i|\log |H_i| = O(|H_p| \log |H_p|) = O(|p|\log |p|)$, since $p$ is a heavy path.
+%  $\sum_{i=1}^{k} |H_i|\log |H_i| = O(|H_p| \log |H_p|) = O(|p|\log |p|)$, since $p$ is a tall path.
 % \end{proof}
@@
 % \begin{corollary}
-% \label{cor:heavyLight}
-%  Let $p$ be a heavy path in $P(T)$ and let $\{q_1, \dots q_k\}$ be the set of all light paths adjacent to $p$.
+% \label{cor:tallLight}
+%  Let $p$ be a tall path in $P(T)$ and let $\{q_1, \dots q_k\}$ be the set of all short paths adjacent to $p$.
 %  Then $\sum_{i=1}^k \sum_{v\in q_i} \log |H_v| = O(|p|\log |p|)$.
@@
 %  Let $H_i$ be as defined in \Lem{adjacent}.  By \Lem{pathBounds}, for all $1\leq i\leq k$, 
-%  $\sum_{v\in q_i} \log |H_v| = O(|q_i|\log |H_i|) = O(|H_i|\log |H_i|)$, as $q_i$ is a light path.
-%  Now by \Lem{adjacent}, $\sum_{i=1}^k |H_i| \leq |H_p|+|p| = O(|p|)$, as $p$ is a heavy path. 
+%  $\sum_{v\in q_i} \log |H_v| = O(|q_i|\log |H_i|) = O(|H_i|\log |H_i|)$, as $q_i$ is a short path.
+%  Now by \Lem{adjacent}, $\sum_{i=1}^k |H_i| \leq |H_p|+|p| = O(|p|)$, as $p$ is a tall path. 
 %  Therefore, as $x\log x$ is a convex function, Jensen's inequality implies the claim.
@@
 
-\subsection{The shrub of unsupported paths} \label{sec:light}
-
-It suffices to focus on unsupported light paths. \Lem{light} can be reduced to the following.
+\subsection{The shrub of unsupported paths} \label{sec:short}
+
+It suffices to focus on unsupported short paths. \Lem{short} can be reduced to the following.
 
@@
 
-We now construct the shrub of unsupported light paths. Consider $p \in \cL \setminus \cL'$,
-and traverse the chain of ancestors from $p$. Eventually, we must reach another light path $q$.
+We now construct the shrub of unsupported short paths. Consider $p \in \cL \setminus \cL'$,
+and traverse the chain of ancestors from $p$. Eventually, we must reach another short path $q$.
 (If not, we have reached the root $r$ of $\pmax(T)$. Hence, $p$ is supported.)
 Insert edge from $p$ to $q$, so $q$ is the parent of $p$ in $\cU$. This construction leads
-to the shrub forest of $\cL \setminus \cL'$, where every unsupported light path is an internal node, and all
-roots are supported light paths.
+to the shrub forest of $\cL \setminus \cL'$, where every unsupported short path is an internal node, and all
+roots are supported short paths.
 
@@
 \begin{claim} \label{clm:loss2} Consider $p \in \cL \setminus \cL'$ that is in $\cU$. Let $Q$
-be the set of heavy ancestors of $p$ that are descendants of $r(\cU)$. Then $\sum_{q \in Q} |q \cap H_p| \leq |H_p|/50$.
+be the set of tall ancestors of $p$ that are descendants of $r(\cU)$. Then $\sum_{q \in Q} |q \cap H_p| \leq |H_p|/50$.
 \end{claim}
 
-\begin{proof} Suppose not, for contradictions sake. As we traverse the heavy ancestors of $p$,
+\begin{proof} Suppose not, for contradictions sake. As we traverse the tall ancestors of $p$,
 we encounter various vertices of $H_p$ in the ancestors $q$. Consider the first $|H_p|/100$
@@
 vertices \emph{not yet} encountered, so $|H_v| > |H_p|/100$. Each of these vertices $v$
-is supported, implying that $p$ is a supported light path.
+is supported, implying that $p$ is a supported short path.
 \end{proof}
@@
 
-Since all nodes in $\cU$ represent light paths, $\forall p \in \cU$, $|p| < \sqrt{|H_p|}/100$.
+Since all nodes in $\cU$ represent short paths, $\forall p \in \cU$, $|p| < \sqrt{|H_p|}/100$.
 So $\sigma_\ell < \sum_{p \in \cU_\ell} \sqrt{|H_p|}/100$.
@@
 
-Let $\mathcal{L} = \mathcal{L}(r)$ be a light shrub of some light path $r$ in a maximum path decomposition $\mathcal{P}(T)$.
-The intuitive idea is that due to the geometric decreasing behavior of light shrubs described in the previous subsection, 
-one can show that for the entire light shrub our algorithm pays roughly the cost due to the root path $r$.  This argument 
+Let $\mathcal{L} = \mathcal{L}(r)$ be a short shrub of some short path $r$ in a maximum path decomposition $\mathcal{P}(T)$.
+The intuitive idea is that due to the geometric decreasing behavior of short shrubs described in the previous subsection, 
+one can show that for the entire short shrub our algorithm pays roughly the cost due to the root path $r$.  This argument 
 is hard to make directly, so instead we define an alternative quantity for which we can make the argument.
@@
 For a path $v\in \mathcal{L}$, let $H_v$ denote the heap at the root of $v$, and let $H_{v_f}$ denote the heap 
-at the leaf of $v$.  Let $S_l(v)$ (resp. $S_h(v)$) be the subset of light (resp. heavy) paths adjacent to $v$ in $\pathTreeA(T)$.
+at the leaf of $v$.  Let $S_l(v)$ (resp. $S_h(v)$) be the subset of short (resp. tall) paths adjacent to $v$ in $\pathTreeA(T)$.
 
 We want to understand how the heap size changes as we move around the paths in $\mathcal{L}$.  Specifically, there are 
-three places the vertices in the heap along a light path can come from; adjacent light paths, adjacent heavy paths, and the initial heap at the leaf of the path.  
+three places the vertices in the heap along a short path can come from; adjacent short paths, adjacent tall paths, and the initial heap at the leaf of the path.  
 Given this observation we now define a weight function which upper bounds the size of the heap at the root of any path in $\mathcal{L}$.
@@
 First, for all $v\in \mathcal{L}$ we define the local cost function $c(v) = H_{v_f}+\sum_{i\in S_h(v)} H_i$, this is the addition to the heap at $v$ which 
-was not already charged to descendant light paths.  We now define a weighted version of $\mathcal{L}$ based on these costs.  
+was not already charged to descendant short paths.  We now define a weighted version of $\mathcal{L}$ based on these costs.  
 To do so, we first augment $\mathcal{L}$ by attaching one 
@@
 
-As the weight $w(p)$ includes the sizes of the heaps of all contributing heavy paths and leaves in the subtree rooted at $p$, we have the following.
+As the weight $w(p)$ includes the sizes of the heaps of all contributing tall paths and leaves in the subtree rooted at $p$, we have the following.
 
@@
 This holds by the same argument as in \Lem{adjacent}, namely, $w(v)$ by definition 
-counts all the contributing heavy path heaps and initial leaf heaps that contribute 
+counts all the contributing tall path heaps and initial leaf heaps that contribute 
 to the heap at $v$ and $\sum_{u\in \mathcal{L}_v} |u|$ is an upper bound on the number 
@@
 
-As all paths in $\mathcal{L}$ are light, we thus have the following.
+As all paths in $\mathcal{L}$ are short, we thus have the following.
 
@@
 \begin{lemma}
-Let $\mathcal{L}$ be the light shrub of some path $r$.  
+Let $\mathcal{L}$ be the short shrub of some path $r$.  
 Then for any $v\in \mathcal{L}$, we have $|H_v|\geq \frac{4}{5} w(v)$.
@@
 \label{cor:equivalent}
- Let $\mathcal{L}$ be a light shrub of a light path $r$.  Then for any $v\in \mathcal{L}$, $|H_v| \leq w(v) \leq \frac{5}{4} |H_v|$.
+ Let $\mathcal{L}$ be a short shrub of a short path $r$.  Then for any $v\in \mathcal{L}$, $|H_v| \leq w(v) \leq \frac{5}{4} |H_v|$.
 \end{corollary}
@@
 \begin{lemma}
-\label{lem:lightCost}
- Let $\mathcal{L}$ be a light shrub of a light path $r$.  Let $V$ be the set of heavy paths from the full shrub which are adjacent to paths in $\mathcal{L}$.  
+\label{lem:shortCost}
+ Let $\mathcal{L}$ be a short shrub of a short path $r$.  Let $V$ be the set of tall paths from the full shrub which are adjacent to paths in $\mathcal{L}$.  
   Then $\sum_{p\in \mathcal{L}}\sum_{v\in p} \log |H_v| + \sum_{q\in V} |H_q|\log |H_q| \leq \frac{4}{3} |H_r|\log |H_r|$
@@
 Recall that in a proper coloring of a binary tree $T$, the color of any vertex must be the color of one of its descendants.  
-Therefore, any vertex in the light path shrub must appear in the heap of either the root of an adjacent 
+Therefore, any vertex in the short path shrub must appear in the heap of either the root of an adjacent 
 path from the full shrub or from the heap of a leaf of a path in $\mathcal{L}$.  
 Also note $|H_r|$ precisely counts sum of the number of vertices from these heaps that do not appear in $\mathcal{L}$.
-As there are no adjacent light paths to a light shrub, the total number of vertices over all paths in $\mathcal{L}$ is at most 
+As there are no adjacent short paths to a short shrub, the total number of vertices over all paths in $\mathcal{L}$ is at most 
 $m = \sum_{p\in \mathcal{L}} |H_{p_f}| +\sum_{q\in V} |H_q| - |H_r| = w(r)-|H_r|$, where $H_{p_f}$ is the heap at the leaf of $p$.

### 2014-10-22 paint-contour-tree.tex

@@
 \end{figure}
+
+\section{Contour tree basics} \label{sec:basics}
+
+We detail the basic definitions about contour trees. Most of this is taken from Chapter 6 of Carr's thesis~\cite{Carr}.
+Our input is a continuous PL function $f:\MM \mapsto \RR$, where $\MM$ is a simplicial complex in $\RR^d$. 
+All simplices are isomorphic to a standard $d$-dimensional simplex, except for specially designated
+\emph{boundary} simplices. \Sesh{Some mention of representation and data structures here.}
+It is standard to assume
+that $f$ is Morse (we explain later what this entails).
+
+Observe that $f:\MM \rightarrow \RR$ can be thought of as a $d$-dimensional simplicial complex living in $\RR^{d+1}$.
+We can think of $f(x)$ as the ``height" of a point $x \in \MM$.
+We will often omit $f$, and assume that $f$ is encoded in the representation of $\MM$.
+
+\begin{definition} \label{def:level} The level set at value $h$ is the set $\{x \in \MM| f(x) = h\}$.
+A contour is a maximal connected component of a level set. An $h$-contour is a contour where $f$-values are $h$.
+\end{definition}
+
+Not that a contour is itself a simplicial complex of one dimension lower, and is represented (in our algorithms) as such.
+We let $\delta, \eps$ denote infinitesimals. Let $B_\eps(x)$ denote a ball of radius $\eps$ around $x$.
+
+\begin{definition} \label{def:deg} The up-degree of $x \in \MM$ is the number of $(f(x) + \delta)$-contours that intersect $B_\eps(x)$
+as $\delta, \eps \rightarrow 0^+$. The down-degree is the number of $(f(x) - \delta)$-contours that intersect $B_\eps(x)$
+as $\delta, \eps \rightarrow 0^+$.
+\end{definition}
+
+The definition allows us to categorize points in $\MM$ depending on the local changes in topology.
+
+\begin{definition} \label{def:points} A point $x \in \MM$ is categorized as follows.
+\begin{asparaitem}
+	\item Regular: Both up-degree and down-degrees are $1$.
+	\item Maximum: Up-degree is $0$.
+	\item Minimum: Down-degree is $0$.
+	\item Join: Up-degree is strictly greater than $1$.
+	\item Split: Down-degree is strictly greater than $1$.
+\end{asparaitem}
+Non-regular points are called \emph{critical}. Joins and splits are called \emph{saddles}.
+Maxima and minima are called \emph{extrema}.
+\end{definition}
+
+We will enforce a special constraint on boundaries, encapsulated in the following definition.
+
+\begin{definition} \label{def:bound} $f$ is \emph{boundary critical} if the following holds.
+Consider a boundary simplex $F$. All vertices of $F$ have the same function value. Furthermore, all
+neighbors of $F$ not in $F$ either have function values strictly greater or strictly less
+than this value.
+\end{definition}
+
+We use $\cV$ to denote the set of critical points in $\MM$.
+Because $f$ is piecewise-linear, all critical points are vertices in $\MM$. 
+A value $h$ is called critical, if $f(x) = h$, for some $x \in \cV$. The Morse assumption
+implies that all critical points have distinct values. A contour that involves a critical
+point is also denoted as critical. Note that the vertices in a boundary face collectively behaves like an extremum.
+Abusing notation, the term ``critical point", ``maxima", and ``minima" will include boundaries.
+Critical values include these boundary values. By boundary criticality, all contours are themselves
+represented as $d-1$-dimensional simplicial complices.
+
+\begin{definition} \label{def:equiv} Two contours $\psi$ and $\psi'$ are equivalent 
+if the following holds. There exists an $f$-monotone path $p$ connecting a point in $\psi$ to $\psi'$,
+such that no $x \in p$ belongs to a critical contour.
+\end{definition}
+
+These give the \emph{contour classes}. By continuity, a regular point is present on a single contour,
+and we can partition the regular contours into these classes. Every such class maps to intervals
+of the form $(f(x_i),f(x_j))$, where $x_i, x_j$ are critical points. Such a class is said
+to be created at $x_i$ and destroyed at $x_j$.
+
+\begin{definition} \label{def:tree} The contour tree is the graph on vertex set $\cV$, where
+edges are formed as follows. For every contour class that is created at $v_i$ and destroyed $v_j$,
+there is an edge $(v_i,v_j)$. (Conventially, edges are directed from higher function value to lower.)
+\end{definition}
+
+(It is not immediately obvious that this graph is a tree. But it is equivalent to the Reeb graph
+of $f:\MM \rightarrow \RR$, which is known to be a tree.) We denote the contour tree of $\MM$
+by $\reeb(\MM,f)$. The corresponding node and edge sets are denoted as $\cV(\cdot)$ and $\cE(\cdot)$.
+
+We use a key theorem from~\cite{Carr} (Theorem 6.6) that maps paths in $\cC(\MM)$ and $\MM$.
+
+\begin{theorem} \label{thm:carr-path} For every path $P$ in $\MM$, there exists a path $Q$ in the contour tree to which $P$
+maps, and for every path $Q$ in the contour tree, there exists at least one path $P$ in $\MM$ that maps to $Q$.
+\end{theorem}
+
+\section{Divide and conquer through contour surgery} \label{sec:surgery}
+
+{\bf The cutting operation:} We define a ``cut" operation on $f:\MM \rightarrow \RR$ that cuts along a regular contour to create
+a new simplicial complex with an added boundary. Given contour $\phi$, this is constructing the simplicial complex $\MM \setminus \phi$. 
+We will always enforce the condition that $\phi$ never passes through a vertex of $\MM$.
+Again, we use $\eps$ for an infinitesimally small value. We denote $\phi^+$ (resp. $\phi^-$) to be
+the contour at value $f(\phi^+) + \eps$ (resp. $f(\phi^-) - \eps$), which is at distance $\eps$
+from $\phi$.
+
+An $h$-contour is achieved by intersecting $\MM$ with the hyperplane $x_{d+1} = h$. (Think of the $d+1$-dimension
+as height.) Given some point $x$
+on an $h$-contour $\phi$ (with some simplex that $x$ is present in), we can walk along $\MM$ from $x$ and determine $\phi$.
+We can ``cut" along $\phi$ to get a new (possibly) disconnected simplicial complex $\MM'$. This is achieved
+by splitting every facet $F$ that $\phi$ intersects into an ``upper" facet and ``lower" facet. Algorithmically,
+we cut facet $F$ with $\phi^+$ and take everything above $\phi^+$ in $F$ to make the upper facet. Analogously, we cut with $\phi^-$ to get the lower facet.
+The facets are then triangulated to ensure that they are all simplices.
+Given that $\phi$ cannot cut a boundary and all non-boundary facets
+have constant size, we omit the algorithmic description for this process. We only note that this
+creates the two new boundaries $\phi^+$ and $\phi^-$, and we maintain the property of constant $f$-value at a boundary.
+This new simplicial complex is denoted by $\cut(\phi,\MM)$.
+This can be constructed in time linear in $|\phi|$.
+
+%We denote this new manifold by $\cut(\phi,\MM)$, and the new boundaries by $\phi^+$ and $\phi^-$.
+%Abusing notation, we use $\cut(\phi,\MM)$
+%for a set $\Psi$ of contours to mean the manifold obtained by cutting along all contours in $\Phi$.
+
+\ignore{
+\medskip
+\fbox{
+\begin{minipage}{0.9\textwidth}
+{\bf $\cut(\phi,\MM)$}
+
+\smallskip
+\begin{asparaenum}
+%	\item Insert the vertices of $\phi$ in $\MM$.
+	\item For every edge $e$ that intersects $\phi$ at point $x$: add edges $(e,x^+)$ and $(e,x^-)$
+	where $x^+$ (resp. $x^-$) is infinitesimally above (resp. below) $x$.
+	\item Take a point $x$ (resp. $y$) on $e$ at distance $\eps$ above (resp. below) $\phi$.
+	\item Let $\phi_x$ be the contour through $x$, and analogously define $\phi_y$. Insert a vertex
+	at each intersection point of these contours with edges of $\MM$. Add  
+	\item Delete the edges between $\phi_x$ and $\phi_y$.
+\end{asparaenum}
+\end{minipage}}
+
+\medskip
+This results in a new boundary critical simplicial complex. There are two new disjoint
+boundary simplices added. The manifold may be disconnected. 
+}
+
+We now describe a high-level approach to construct $\reeb(\MM)$.
+
+\medskip
+\fbox{
+\begin{minipage}{0.9\textwidth}
+{\bf $\surgery(\MM,\phi)$}
+
+\smallskip
+\begin{asparaenum}
+	\item Let $\MM' = \cut(\phi,\MM)$.
+	\item Construct $\reeb(\MM')$ and let $A, B$ be the nodes corresponding to the new boundaries created
+	in $\MM'$. (One is a minima and the other is maxima.)
+	\item Since $A, B$ are leaves, they each have unique neighbors $A'$ and $B'$, respectively. Insert
+	edge $(A',B')$ and delete $A, B$ to obtain $\reeb(\MM)$.
+\end{asparaenum}
+\end{minipage}}
+
+\medskip
+\begin{theorem} \label{thm:surgery} For any regular contour $\phi$, the output of $\surgery(\MM,\phi)$ is $\reeb(\MM)$.
+\end{theorem}
+
+The main theorem is a direct consequence of the following lemma.
+
+\begin{lemma} \label{lem:cut} Consider a regular contour $\phi$ contain in contour class (edge of $\reeb(\MM))$
+$(u,v)$ and denote $\MM' = \cut(\phi,\MM)$. Then $\cV(\reeb(\MM')) = \{\phi^+,\phi^-\} \cup (\cV(\MM) \setminus \phi)$
+and $\cE(\reeb(\MM')) = \{(u,\phi^+),(\phi^-,v)\} \cup (\cE(\MM) \setminus (u,v))$.
+\end{lemma}
+
+\begin{proof} The contours of $\MM'$ are exactly the contours of $\MM$ with $\phi^+$ and $\phi^-$, without $\phi$.
+All new vertices created in $\MM'$ are present on $\phi^+$ and $\phi^-$. That proves the first part.
+
+Any contour class in $\MM'$ (edge in $\cC(\MM')$) that does not involve $\phi^+$ or $\phi^-$ is also
+a contour class in $\MM$. Furthermore, a maximal contour class satisfying these properties is also
+maximal in $\MM$. So all edges of $\cC(\MM')$ that do not involve $\phi^+$ or $\phi^-$ are edges of $\cC(\MM)$.
+Analogously, every edge of $\cC(\MM)$ not involving $\phi$ is an edge of $\cC(\MM)$.
+
+Consider the contour class corresponding to edge $(u,v)$ of $\cC(\MM)$. There is a natural ordering
+of the contours by function value, ranging from $f(u)$ to $f(v)$. All contours in this class ``above" $\phi$
+form a maximal contour class in $\MM'$, represented by edge $(u,\phi^+)$. Analogously, there is another
+contour class represented by edge $(\phi^-,v)$. We have now accounted for all contours in $\cC(\MM')$,
+completing the proof.
+\end{proof}
+
+A useful corollary of this lemma shows that a contour actually splits the simplicial complex into
+two disconnected complices.
+
+\begin{theorem} \label{thm:jordan} $\cut(\phi,\MM)$ consists of two
+disconnected simplicial complices.
+\end{theorem}
+
+\begin{proof} Denote (as in \Lem{cut}) the edge containing $\phi$ to be $(u,v)$. Suppose for contradiction that there is a path between vertices $u$ and $v$
+in $\MM' = \cut(\phi,\MM)$. By \Thm{carr-path}, there is a path in $\cC(\MM')$ between $u$ and $v$. Since $\phi^+$ and $\phi^-$
+are leaves in $\cC(\MM')$, this path obviously cannot use edges incident to them. By \Lem{cut},
+all the edges of this path are in $\cE(\cC(\MM)) \setminus (u,v)$. So we get a cycle in $\cC(\MM)$. Contradiction.
+To show that there are exactly two connected components in $\cut(\phi,\MM)$, it suffices
+to see that $\cC(\MM')$ has two connected components (by \Lem{cut}) and apply \Thm{carr-path}.
+\end{proof}
+
+\section{Raining to partition $\MM$} \label{sec:rain}
+
+In this section, we describe a linear time procedure that partitions $\MM$ into special
+\emph{extremum dominant} simplicial complices.
+
+\begin{definition} \label{def:dom} A simplicial complex is \emph{minimum dominant} if there exists
+a minimum $x$ such that every non-minimal \emph{vertex} in the manifold has decreasing path to $x$.
+Analogously define \emph{maximum dominant}. 
+\end{definition}
+
+The first aspect of the partitioning is ``raining". Start at some point $x \in \MM$ and imagine rain at $x$.
+The water will flow downwards along descending paths and ``wet" all the points encountered. Note that this procedure considers all points
+of the manifold, not just vertices.
+
+\begin{definition} \label{def:wet} Fix $\MM$ and $x \in \MM$. The set of points $y \in \MM$ such that there is a non-ascending path from $x$ to $y$
+is denoted by $\wet(x,\MM)$ (which is in turn represented as a simplicial complex). A point $z$ is at the \emph{interface} of $\wet(x,\MM)$ if every neighborhood of $z$
+has non-trivial intersection with $\wet(x,\MM)$.
+\end{definition}
+
+The following claim gives a description of the interface. While its meaning is quite intuitive, the proof is tedious.
+
+\begin{claim} \label{clm:inter} For any $x$, the interface of $\wet(x,\MM)$ is a set of contours,
+each containing a join vertex.
+\end{claim}
+
+\begin{proof} If $y \in \wet(x,\MM)$, all the points in any contour containing $y$ are also in $\wet(x,\MM)$.
+ (Follow the non-ascending path to from $x$ to $y$ and then walk along the contour.) The converse is also true,
+ so $\wet(x,\MM)$ contains entire contours.
+
+Let $\eps, \delta$ be sufficiently small as usual. Fix some $y$ at the interface.
+Note that $y \in \wet(x,\MM)$. (Otherwise, $B_\eps(y)$ is dry.)
+The points in $B_\eps(y)$ that lie below $y$ have a descending path from $y$ and hence must be wet.
+There must be a dry point in $B_\eps(y)$ that is above $y$, and hence,
+there exists a dry, regular $(f(y)+\delta)$-contour $\phi$ intersecting $B_\eps(y)$.
+
+Suppose $y$ is in the interface. Consider the set $\Gamma_y$ of points connected to $y$
+by a path with all points at height $f(y)$. (This is a set of $f(y)$-contours.) Suppose for contradiction
+that $\forall z \in \Gamma_y$, $z$ has up-degree $1$. Consider the non-ascending path from $x$ to $y$ and let $z$
+be the first point of $\Gamma_y$ encountered. There exists a wet, regular $(f(y) + \delta)$-contour $\psi$ 
+intersecting $B_\eps(z)$. Now, walk from $z$ to $y$ along $\Gamma_y$. If all points $w$ in this walk
+have up-degree $1$ (ref. to \Def{deg}), then $\psi$ is the unique $f(y)+\delta$-contour
+intersecting $B_\eps(w)$. This would imply that $\phi = \psi$, contradicting the fact that $\phi$ is wet
+and $\psi$ is dry.
+
+Therefore, there must exist a join vertex $w$ in $\Gamma_y$, such that $B_\eps(w)$ intersects
+both $\phi$ and $\psi$. As $\delta, \eps \rightarrow 0^+$, the limit of $\phi$ (which is dry)
+lies in $\Gamma_y$ (which is wet). Hence, this gives the interface, which is a contour containing
+the join $w$.
+\end{proof}
+
+\ignore{
+\begin{proof} If $y \in \wet(x,\MM)$, all the points in any contour containing $y$ are also in $\wet(x,\MM)$.
+ (Follow the non-ascending path to from $x$ to $y$ and then walk along the contour.)
+ First observe that any point which in the interface must be wet.  Second, for $y \in \wet(x,\MM)$, all the points in any contour containing $y$ are also in $\wet(x,\MM)$, 
+ as one can follow the non-ascending path to $y$ and then walk along the contour.
+
+ So let $y$ be a point in the interface.
+
+ However, since $y$ is in the interface, it is also wet.  Therefore for any contour $\phi$ which contains $y$, for any arbitrarily small $\eps$ there is 
+ a wet point which lies above $\phi$ and has a descending path of length at most $\eps$ to $\phi$.
+ This implies there is some contour which lies just 
+ above $y$ which is wet.  So any contour $\phi$ containing $y$ has two contours which lie just above it, one which is dry, and one which is wet.  
+ This can only happen if $\phi$ contains a merge vertex, because otherwise all contours which lie just above would be in the same L-homotopy equivalence class and so by 
+ \Clm{mono}, if one is wet, all below it are wet.
+
+ By similar logic one can argue that of the two contours through a given merge vertex at most one can contain interface points other than the merge vertex, 
+ and that if one of the contours does contain such interface points, then the entire contour consists of interface points.
+\end{proof}
+}
+
+Note that $\wet(x,\MM)$ (and its interface) can be computed in time linear in the size of the wet simplicial complex.
+We perform a non-ascending walk from $x$. Any facet $F$ of $\MM$ encountered has some (if not all) portion of in $\wet(x,\MM)$.
+This portion is determined by cutting $F$ along the interface. Since the interface is a contour, this is equivalent
+to cutting $F$ by a hyperplane. All these operations can be performed to output $\wet(x,\MM)$ in time linear in $|\wet(x,\MM)|$.
+
+% \begin{proof} If $y \in \wet(x,\MM)$, but not in the interface, then any contour containing $y$ is also in $\wet(x,\MM)$.
+% So consider a point $y$ in the interface. We argue that all points in a contour $\phi$ containing $y$
+% are also in the interface. If such a contour is regular, we can find a contour just above it that
+% is also wet. 
+% 
+% If two contours intersect, the intersection point $y$ is critical. We can argue that the only
+% way to reach $y$ from a descending path is reach one of the contours and then follow it to get to $y$.
+% But that would imply the existence of another critical point on this contour, violating the Morse condition.
+% 
+% Let us argue that $y$ is a merge vertex. Consider the descending path from $x$ to $y$. The final edge
+% that reaches $y$ is obviously all wet, and is an increasing edge from $y$. Consider an opposite increasing
+% edge. Take two point in these edges that are infinitesimally higher than $y$ and show that the contours
+% through them as disjoint. So $y$ is a merge.
+% \end{proof}
+
+We define a simple \lift{} operation on the interface contours. Consider such a contour $\phi$ containing
+a merge vertex $y$. Take any dry increasing edge incident to $y$, and pick the point $z$ on this edge at height
+$f(y) + \delta$. Let $\lift(\phi)$ be the unique contour through regular point $z$. Note that $\lift(\phi)$ is dry.
+
+\begin{claim} \label{clm:cut-int} Let $\phi$ be an interface contour. Then $\cut(\MM,\lift(\phi))$
+results in two disjoint simplicial complices, one consisting entirely of dry points.
+\end{claim}
+
+\begin{proof} By \Thm{jordan}, $\cut(\MM,\lift(\phi))$ results in two disjoint simplicial complices. Let $\NN$ be the complex containing
+the point $x$ (the argument in $\wet(x,\MM)$), and let $\NN'$ be the other complex. 
+Any path from $x$ to $\NN'$ must intersect $\lift(\phi)$, which is dry. Hence $\NN'$ is dry.
+%
+%
+%If there exists a wet point $z$ in $\NN'$, then there exists a descending path from $x$ to $z$. 
+%As $x$ is in $\NN$ and $z$ is in $\NN'$, again by the Jordan Curve Theorem this path must intersect $\phi$, 
+%contradicting the fact that $\phi$ is dry.
+\end{proof}
+
+We describe the main partitioning procedure that cuts a simplicial complex $\NN$ into extremum
+dominant complices. It takes an additional input of a maximum $x$. To initialize,
+we begin with $\NN$ set to $\MM$ and $x$ as an arbitrary maximum. One of the critical aspects
+of the procedure is that the rain alternately flows downwards and upwards. In other words, when
+we start, rain flows downwards. In each recursive call, the direction of rain is switched to the 
+opposite direction. While this is way of implementing $\rain$, it is conceptually easier
+to think of \emph{inverting} a complex $\NN'$ when a recursive call is made. Inversion
+is easily achieved by just negating the height values. (From a running time standpoint, it suffices
+to maintain a single bit associated with $\NN'$ that determines whether heights are inverted or not.)
+We can now let rain flow downwards, as it usually does in our world.
+
+%
+%We critically perform
+%an \emph{inversion} step, which is simply achieved by negating
+%
+\medskip
+\fbox{
+\begin{minipage}{0.9\textwidth}
+{\bf $\rain(x,\NN)$}
+
+\smallskip
+\begin{compactenum}
+	\item Determine interface of $\wet(x,\NN)$. 
+	\item If the interface is empty, simply output $\NN$. Otherwise, denote the contours by $\phi_1, \phi_2, \ldots, \phi_k$ and set $\phi'_i = \lift(\phi_i)$.
+	\item Initialize $\NN_1 = \NN$.
+	\item For $i$ from $1$ to $k$: 
+	\begin{compactenum}
+		\item Construct $\cut(\phi'_i,\NN_i)$, consisting of dry complex $\LL_i$ and remainder $\NN_{i+1}$.
+		\item Let the newly created boundary of $\LL_i$ be $B_i$. Invert $\LL_i$ so that $B_i$ is a maximum. Recursively
+		call $\rain(B_i,\LL_i)$.
+	\end{compactenum}
+	\item Output $\NN_{k+1}$ together with any complices output by recursive calls.
+\end{compactenum}
+\end{minipage}}
+
+\medskip
+For convenience, denote the total output of $\rain(x,\MM)$ by $\MM_1, \MM_2, \ldots, \MM_r$.
+We first argue the correctness of $\rain$.
+
+\begin{lemma} \label{lem:rain-1} Each output $\MM_i$ is extremum dominant.
+\end{lemma}
+
+\begin{proof} Consider a call to $\rain(x,\NN)$. If the interface is empty, then all of $\NN$
+is in $\wet(x,\NN)$, so $\NN$ is trivially extremum dominant. So suppose the interface
+is non-empty and consists of $\phi_1, \phi_2, \ldots, \phi_k$ (as denoted in the procedure).
+By repeated applications of \Clm{cut-int}, $\NN_{k+1}$ contains $\wet(x,\MM)$. 
+Consider $\wet(x,\NN_{k+1})$. The interface must exactly be $\phi_1, \phi_2, \ldots, \phi_k$.
+So the only dry vertices are those in the boundaries $B_1, B_2, \ldots, B_k$. But these
+boundaries are maxima.
+\end{proof}
+
+Focus on $\rain(x,\MM)$. As $\rain$ proceeds, new facets/simplices are created because of repeated cutting. Indeed, any facet $F$
+in any simplicial complex $\NN$ that is recursively invoked, is achieved by cutting facet $F'$ in $\MM$.
+The key to the running time of $\rain(x,\MM)$ is bounding the number of newly created facets, for which we have the following lemma.
+
+\begin{lemma}\label{lem:new-verts}
+A facet $F \in \MM$ is cut at most once during the $\rain(x,\MM)$.
+\end{lemma}
+\begin{proof} All notation here follows that in the pseudocode of $\rain$.
+Firstly, by \Thm{jordan}, all the pieces of which $\rain$ is invoked are disjoint.
+Secondly, all recursive calls are made on dry complices.
+
+Consider the first time that $F$ is cut, say, during the call to $\rain(x,\NN)$.
+Specially, say this happens when $\cut(\phi'_i,\NN_i)$ is constructed. 
+So $F$ is cut by some hyperplane into lower and upper portions (which are then triangulated).
+Since the cut is infinitesimally above an interface, the lower portion is basically wet.
+It can never be part of $\LL_1, \LL_2, \ldots$, which go into recursive calls. This portion
+is in $\NN_{k+1}$. The upper portion/facet (call it $U$) is in $\LL_i$. Note that the lower
+boundary of $U$ is in the boundary $B_i$. Since a recursive call is made to $\rain(B_i,\LL_i)$
+(and $\LL_i$ is inverted), $U$ becomes wet. Hence, $U$ will not be cut subsequently.
+All in all, $F$ is cut at most once.
+\ignore{
+As the number of edges is $O(n)$, arguing that there is at most one boundary vertex created along any given edge will imply the claim.
+Consider some edge $e$ and take the first time a boundary vertex is created in $e$.
+This happens when some contour $\phi$ intersects $e$. On applying $\cut(\NN,\lift(\phi))$,
+consider the dry manifold $\LL$ and wet manifold $\MM$ constructed. The edge $e$ is split into two parts, the dry and wet part.
+The wet part is in $\MM$ and will never be involved in any recursive call.
+
+The dry part of $e$ (call it $e'$) needs to be considered.
+A boundary vertex $v$ (an new endpoint of $e$) is created, which is part of the boundary $B$ in $\LL$,
+such that $\rain(B,\LL)$ is called to the inverted $\LL$. In this call, since the water starts from $v$,
+all of $e'$ becomes wet. So there is no further interface contour that can intersect this portion of $e'$.
+}
+\end{proof}
+
+\begin{theorem} \label{thm:rain-time} The total running time of $\rain(x,\MM)$ is $O(|\MM|)$.
+\end{theorem}
+
+\begin{proof} The operations performed are the raining and cutting. We can bound this by the total
+complexities of the wet complices and the new boundaries created. No facet is every wet twice,
+and by \Lem{new-verts}, the total number of facets that every appear is $O(|\MM|)$. This also 
+bounds the complexities of the boundaries created.
+\end{proof}
+
+\ignore{
+As each original vertex in $\NN$ appears in only one of the sub-manifolds output by $\rain(x,\NN)$, we have the following corollary.
+
+\begin{corollary}
+The total size of a sub-manifolds output by $\rain(x,\MM)$ is $O(n)$.
+\end{corollary}
+
+\begin{corollary} \label{cor:rain-time} The total running time of $\rain(x,\MM)$ is $O(n)$.
+\end{corollary}
+
+\begin{proof} Consider some invocation to $\rain(x,\MM)$. The only non-trivial work done in $\rain$ is by the calls to $\wet$.
+We know that each invocation to $\wet$ runs in time linear 
+to the portion of the manifold which is wet (including the new boundary vertices created).  
+As the output of $\rain$ is precisely the distinct sub-manifolds wet by each distinct call to $\wet$, the claim follows.
+\end{proof}
+
+%Consider one of the manifolds $\MM_i$ output by $\rain$.  This manifold has a set of boundary maxima/minima that are the result 
+%of cutting along interface contours.  For each such contour there is exactly one other output manifold $\MM_j$ for which 
+%the contour appears as a boundary maxima/minima. 
+}
+
+\begin{claim} \label{clm:rain-reeb} Given $\reeb(\MM_1), \reeb(\MM_2), \ldots, \reeb(\MM_r)$, 
+%be the output of $\rain(x,\MM)$. Suppose that for each reeb graph we know its corresponding vertex in the Manifold Ordering Tree. 
+$\reeb(\MM)$ can be constructed in $O(|\MM|)$ time.
+\end{claim}
+
+\begin{proof} Consider the tree of recursive calls in $\rain(x,\MM)$, with each node
+labeled with some $\MM_i$.
+%
+%There is a natural tree ordering defined over these manifolds where the 
+%root corresponds the first manifold that was wet when we rained from $x$, and its children correspond to the recursive calls 
+%(i.e. where we rained from the interface contours of the first raining).  Call this the \emph{Manifold Ordering Tree}.
+%$\rain$ can easily be modified to also output this tree (without affecting the asymptotic running time).  
+Walk through this tree in a leaf first ordering.  Each time we visit a node we connect its contour tree to the 
+contour tree of its children in the tree using the $\surgery$ procedure. 
+Each $\surgery$ call is constant time, and the total time is the size of the recursion tree.
+%
+%
+%This takes $O(n)$ time overall by \Lem{new-verts} and since $\surgery$ takes constant time (given the appropriate vertices in each reeb graph).
+The correctness follows by induction and  \Lem{surgery}.
+\end{proof}
+
+\section{Contour trees of extremum dominant manifolds} \label{sec:paint}
+
+The previous section allows us to restrict attention to extremum dominant manifolds.
+We will orient so that the extremum in question is always a \emph{minimum}.
+We will fix such a simplicial complex $\MM$, with the dominant minimum $m^*$. 
+We observe that the contours trees of minimum dominant manifolds have a relatively simple description.
+This description is directly inspired by (and proven using) the \emph{join tree} of~\cite{csa-cctad-00}.
+We first explain what these trees are.
+
+For vertex $v$, we use $\MM^+_v$ to denote the simplicial complex obtained by only
+keeping vertices $u$ such that $f(u) > f(v)$. Analogously, define $\MM^-_v$. Note that $\MM^+$ may contain
+numerous connected components. The following definition of join and split trees were given by~\cite{csa-cctad-00}.
+Conventially, all edges are directed from higher to lower function value.
+
+\begin{definition} \label{join} The join tree $\cJ(\MM)$ of $\MM$ is built on vertex set $V(\MM)$.
+The directed edge $(u,v)$ is present when $u$ is the smallest valued vertex is a connected component of $\MM^+_v$
+\emph{and} $v$ is connected to this component (in $\MM$). The split tree $\cS(\MM)$ is obtained
+by looking at $\MM^-_v$ (or alternately, by taking the join tree of the inversion of $\MM$). 
+\end{definition}
+
+Note that unlike the contour tree, all vertices in $\MM$ are present in $\cJ(\MM)$ and $\cS(\MM)$. 
+Some basic facts about these trees. All outdegrees in $\cJ(\MM)$ and all indegrees in $\cS(\MM)$ are at most $1$.
+Indegree $2$ vertices in $\cJ(\MM)$ are joins, and outdegree $2$ vertices in $\cS(\MM)$ are splits.
+Leaves in $\cJ(\MM)$ are maxima, and leaves in $\cS(\MM)$ are minima.
+
+The key observation is that $\cS(\MM)$ is trivial for minimum dominant $\MM$.
+
+\begin{lemma} \label{lem:split} $\cS(\MM)$ consists of:
+\begin{asparaitem}
+	\item A single path (in sorted order) with all vertices except non-dominant minima.
+	\item Each non-dominant minimum is attached to a unique split.
+\end{asparaitem}
+\end{lemma}
+
+\begin{proof} First, let us complete the proof assuming that each split $v$ has one child that is just a leaf, which
+is a non-dominant minimum. Any minimum is a leaf in $\cS(\MM)$, and thereby attached to a split. If we removed
+all non-dominant minima, we must end up with a path, as asserted above.
+
+Consider a split $v$. For sufficiently small $\eps, \delta$, there are exactly two $(f(v) - \delta)$-contours
+$\phi$ and $\psi$ intersecting $B_\eps(v)$. Both of these are regular contours. There must be a non-ascending
+path from $v$ to the dominant minimum $m^*$. Consider the first edge (necessarily decreasing from $v$)
+on this path. It must intersect some contour, say $\phi$. By \Thm{jordan}, $\cut(\MM,\phi)$ has
+two connected components, with one (call it $\LL$) having $\phi^-$ as a boundary maximum. This complex
+contains $m^*$ as the non-ascending path $\phi$ intersects $\phi$ only once. Let the other
+component be called $\MM'$.
+
+Consider $\cut(\MM',\psi)$ with connected component $\NN$ having $\psi^-$ as a boundary. $\NN$
+does not contain $m^*$, so any path from the interior of $\NN$ to $m^*$ must intersect the boundary $\psi^-$.
+But the latter is a maximum in $\NN$, so there can be no non-ascending path from the interior to $m^*$.
+Since $\MM$ is overall minimum dominant, the interior of $\NN$ can only contain a single vertex $w$, a non-dominant
+minimum.
+
+The split $v$ has two children in $\cS(\MM)$, one in $\NN$ and one in $\LL$. The child in $\NN$ can only
+be the non-dominant minimum $w$, which is a leaf. 
+\end{proof}
+
+It is convenient to denote the non-dominant minima by $m_1, m_2, \ldots, m_k$, and the associated
+splits (as given by \Lem{split}) by $s_1, s_2, \ldots, s_k$.
+
+To prove our main theorem, we use the merging procedure from~\cite{csa-cctad-00} that constructs the contour tree
+from the join and split trees. 
+It actually construsts the \emph{augmented contour tree} $\cA(\MM)$, which
+is the contour tree with regular points inserted into the edges. 
+For any tree $T$ with vertex $v$, we define \emph{erasing} $v$ from $T$,
+when both in and outdegrees are at most $1$. If $v$ is a leaf, just delete $v$. Otherwise, smooth $v$ out
+(delete $v$ and connect its neighbors by an edge). This tree is denoted by $T \ominus v$.
+
+\medskip
+\fbox{
+\begin{minipage}{0.9\textwidth}
+{\bf $\merge(\cJ(\MM),\cS(\MM))$}
+
+\smallskip
+\begin{compactenum}
+	\item Set $\cJ = \cJ(\MM)$ and $\cS = \cS(\MM)$.
+	\item Denote $v$ as \emph{candidate} if sum of indegree of $v$ in $\cJ$ and outdegree of $v$ in $\cS$ is $1$. 
+	\item Add all candidates to queue.
+	\item While candidate queue is non-empty: 
+	\begin{compactenum}
+		\item Let $v$ be head of queue. If $v$ is leaf in $\cJ$, consider its edge in $\cJ$. Otherwise
+		consider edge in $\cS$. In either case, let edge by $(v,w)$.
+		\item Insert $(v,w)$ in $\cA(\MM)$. 
+		\item Set $\cJ = \cJ \ominus v$ and $\cS = \cS \ominus v$. Enqueue any new candidates.
+	\end{compactenum}
+	\item Smooth out all regular vertices in $\cA(\MM)$ to get $\cC(\MM)$.
+\end{compactenum}
+\end{minipage}}
+
+\medskip
+
+\begin{theorem} \label{thm:contour-tree} The contour tree $\cC(\MM)$ consists of all edges $\{(s_i, m_i)\}$
+as well as: for any critical point $v$, edge $(w,v)$ where $w$ is the lowest non-minimal critical point
+in a connected component of $\MM^+_v$ connected to $v$.
+\end{theorem}
+
+\begin{proof} We first show that $\cA(\MM)$ is $\cJ(\MM) \ominus \{m_i\}$ with edges $\{(s_i,m_i)\}$.
+We have flexibility in choosing the order of processing in $\merge$. We first put the non-dominant
+maxima $m_1, \ldots, m_k$ into the queue. As these are processed, the edges $\{(s_i,m_i)\}$ are inserted 
+into $\cA(\MM)$. Once all the $m_i$'s are erased, $\cS$ becomes a path, so all outdegrees are at most $1$.
+The join tree is now $\cJ(\MM) \ominus \{m_i\}$. We can now process $\cJ$ leaf by leaf, and all
+edges of $\cJ$ are inserted into $\cA(\MM)$.
+
+Note that $\cC(\MM)$ is obtained by smoothing out all regular points from $\cA(\MM)$. Smoothing out regular
+points from $\cJ(\MM) \ominus \{m_i\}$ exactly yields the edges described in the theorem.
+\end{proof}
+
+\section{Painting to compute contour trees}
+
+We begin by associating a color with each maximum. If the set of maxima is $S$,
+this is simply an arbitrary bijection $\chi:S \mapsto [|S|]$. Imagine there being a large
+can of paint of color $\chi(y)$ at maxima $y$. We will spill different paint from each maximum and watch it flow down.
+This is analogous to the raining in the previous section, but paint is a much more viscous liquid.
+\emph{So paint only flows down edges, and it does not color the interior of faces.} Furthermore, paints
+do not mix, so eventually, every vertex and edge of $\MM$ gets a unique color.
+%
+\begin{definition} \label{def:paint} A  \emph{painting} of $\MM$ is a map $\chi:S \cup E \mapsto [|S|]$
+with the following properties. (Remember that $G(\MM) = (V,E)$.)
+\begin{asparaitem}
+	\item The restriction $\chi:S \mapsto [|S|]$ is a bijection.
+	\item Consider an edge $e$. There exists a descending path from some maximum $x$ to $e$
+	consisting of edges in $E$, such that all edges along this path have the same color as $x$. (The maximum $x$
+	is well-defined to be the unique maximum with the same color as $e$.)
+\end{asparaitem}
+\end{definition}
+%
+Note that paintings are not unique. It is straightforward to construct a painting of $\MM$ in $O(n)$ time, using a descending BFS from each maximum that does not explore previously colored edges.
+
+\subsection{The data structures} \label{sec:struct}
+
+\begin{definition} \label{def:color-set} Fix a painting $\chi$, and a color $c$.
+A critical point $x$ is \emph{touched by $c$} if $x$ is incident to a $c$-colored
+increasing edge. The set of non-extremal critical points touched by $c$ is denoted $\touch(c)$.
+For $x$, $\col(x)$ is the set of colors that touch $x$.
+\end{definition}
+%
+%We will store each set $\touch(c)$ in a binomial heap, keyed by the heights. 
+Abusing notation, $\touch(c)$ refers
+both to the set and the data structure used to store it.
+
+\medskip
+\noindent
+{\bf The binomial heaps $\touch(c)$:} For each color $c$, we store a binomial
+heap $T(c)$ of non-extremal critical points, keyed by vertex heights.
+
+\medskip
+\noindent
+{\bf The union-find data structure on colors:} We will repeatedly perform unions
+of classes of colors, and this will be maintained as a standard union-find data structure.
+For any color $c$, $\rep(c)$ denotes the representative of its class. We initialize
+by simply setting $\rep(c) = c$ for all colors.
+
+\medskip
+\noindent
+{\bf The stack $\st$:} This consists of non-extremal critical points with the following structure.
+Each point $x \in \st$ has an associated subset of $\col(x)$, denoted $\mcol(x)$.
+Both $\mcol(x)$ and its complement are stored as hash table. So lookups, inserts, and deletes
+are in these sets are all constant time operations. The stack is guaranteed to satisfy 
+the following invariants.
+\begin{asparaitem}
+	\item For every $x \in \st$: For every $c \in \mcol(x)$, $x$ is the highest element
+	in $T(c)$. Furthermore, $c = \rep(c)$.
+	\item Consider $x, y \in \st$ such that $y$ was pushed on $x$. There exists $c \in \col(x) \setminus
+	\mcol(x)$ such that $x$ is not highest in $T(c)$ but $y$ is highest in $T(c)$.
+\end{asparaitem}
+
+\medskip
+\noindent
+{\bf Highest points $\h(c)$:} For each color $c$, we maintain a special critical point $\h(c)$ of this color.
+It is supposed to denote the ``highest" point of color $c$, but the notion of highest evolves
+with the algorithm.
+
+\subsection{Updating the stack} \label{sec:stack}
+
+We have a simple, but critical, procedure that updates the stack $\st$ by pushing
+points on to it.
+
+\medskip
+\fbox{
+\begin{minipage}{0.9\textwidth}
+{\bf $\update(\st)$}
+
+\smallskip
+\begin{compactenum}
+	\item If $K$ is empty, pick arbitrary $c$. Let $x$ be the highest point in $T(c)$.
+	Push $x$ onto stack and set $\mcol(x) = \{c\}$.
+	\item Let $h$ be head of $\st$.
+	\item While $\col(h) \setminus \mcol(h)$ non-empty:
+	\begin{compactenum}
+		\item Pick some $c \in \col(h) \setminus \mcol(h)$. Replace it by $c' = \rep(c)$.
+		\item Obtain highest point $x$ in $T(c')$. 
+		\item If $x = h$, put $c'$ in $\mcol(h)$. Otherwise ($x \neq h$), push $x$ onto stack
+		and add $c'$ to $\mcol(x)$.
+		\item Let $h$ be new head of $\st$.
+	\end{compactenum}
+\end{compactenum}
+\end{minipage}}
+
+\medskip
+When $\update(\st)$ terminates, the head $h$ has the following property. For every $c \in \col(h)$,
+$h$ is the highest point in $\touch(c)$.
+
+\begin{claim} \label{clm:cont} Consider an increasing path from $h$ to some maxima $x$.
+For any $y$ in this path, $\cont(y)$ only contains points of color $c(x)$.
+Furthermore, for all vertices on $\cont(y)$, their increasing edges are colored $c(x)$.
+\end{claim}
+
+\begin{proof} Consider this increasing path, and think of moving $y$ continuously downwards from $x$
+to $h$. Initially (when $y = x$), the lemma is trivially true. Consider the first $y$
+when the condition fails. We have $\cont(y)$ only containing points of color $c(x)$,
+but there is a vertex $z \in \cont(y)$ that has increasing edges of a different color $c'$.
+All the increasing edges of $z$ that are colored $c(x)$ are contiguous (since we can construct
+a contour intersecting all of them). Thus, there are two disjoint increasing neighborhoods for $z$,
+and $z$ is a saddle point. But that contradicts the fact that $h$ is the highest saddle point
+of color $c(x)$.
+
+So, as $y$ decreases, for all vertices on $\cont(y)$, their increasing edges are colored $c(x)$.
+This also means that $\cont(y)$ only contains points of color $c(x)$.
+\end{proof}
+
+We have important lemmas characterizing $|\col(h)|$.
+
+\begin{lemma} \label{lem:mono} If $h$ is a merge vertex, then $|\col(h)| > 1$.
+\end{lemma}
+
+\begin{proof} Suppose not. Consider the edges incident to $h$ in circular order. There exist (at least) $2$
+disjoint contiguous stretches of edges that are increasing. 
+All these edges have the same color $c$, associated with some maximum $x$. Consider two neighbors $y,z$
+that lie in different stretches. By the definition of a painting, there exist descending paths (along
+edges) from $x$ to $y$ and $x$ to $z$. We can construct a Jordan curve $C$ from $x \rightarrow y
+\rightarrow h \rightarrow z \rightarrow x$. All points in this curve except $h$ have 
+heights larger than $f(h)$. By the Jordan Curve Theorem, this partitions $\MM$ into
+an ``inside" and an ``outside".
+
+One of the lower neighbors of $h$ (say $p$) is inside, and another (say $q$) is outside.
+Because $h$ is a merge, neither of these can a minimum other than the dominant minimum $u$.
+Hence, both $p$ and $q$ have descending paths to $u$. So there is a path from $p$
+to $q$ where all heights are strictly less than $f(h)$. But this path must intersect $C$.
+Contradiction.
+\end{proof}
+
+\begin{lemma} \label{lem:col} If $h$ is a merge vertex, $|\col(h)| = 2$.
+If $h$ is a split vertex, $|\col(h)| = 1$.
+\end{lemma}
+
+\begin{proof} Consider one contiguous sequence of increasing edges from $h$.
+There exists a contour intersecting all of these edges. By \Clm{cont}, they all
+must have the same color. There are two such contiguous sequences.
+
+If $h$ is a merge vertex, then $|\col(h)| = 2$ by \Lem{mono}. If $h$ is a split vertex,
+there is a single contour passing through all increasing edges. By \Clm{cont}, they
+all have the same color.
+\end{proof}
+
+\subsection{Constructing $\reeb(\MM)$} \label{sec:const}
+
+The main construction procedure is $\onestep(\MM)$. It pieces together $\reeb(\MM)$
+incrementally.
+
+\medskip
+\fbox{
+\begin{minipage}{0.9\textwidth}
+{\bf $\onestep(\MM)$}
+
+\smallskip
+\begin{compactenum}
+	\item Run $\update(\st)$. Pop $\st$ to get $h$.
+	\item Suppose $\mcol(h) = \{c\}$ (so $h$ is split). 
+	\begin{compactenum}
+		\item Connect $h$ (in $\reeb(\MM)$) to $\h(c)$ and the unique minimum corresponding to split $h$. 
+		\item Delete $h$ from $T(c)$, set $\h(c) = h$.
+		\item End procedure.
+	\end{compactenum}
+	\item Let $\mcol(h) = \{c_1, c_2\}$ (so $h$ is merge).
+	\item Connect $h$ in $\reeb(\MM)$ to $\h(c_1)$ and $\h(c_2)$.
+	\item Delete all copies of $h$ from $T(c_1)$ and $T(c_2)$.
+	\item Perform union of colors $c_1$ and $c_2$ (denote merged color as $c$). Merge heaps
+	to get $T(c) = T(c_1) \cup T(c_2)$.
+	\item Set $\h(c) = h$.	
+\end{compactenum}
+\end{minipage}}
+
+\medskip
+\begin{theorem} \label{thm:piece} Consider the very first call to $\onestep(\MM)$. Let $y$ be a point
+infinitesimally below $h$ on a decreasing edge. Consider $\cut(\MM,\cont(y)) = \{\MM_u,\MM_\ell\}$ (the upper and lower manifolds). Then $\onestep(\MM)$
+outputs $\reeb(\MM_u)$ and implicitly gives a valid painting of $\MM_\ell$.
+\end{theorem}
+
+We have an important lemma that helps to bound the running time.
+
+\begin{lemma} \label{lem:tree} For every color $c$, all points in $T(c)$ lie along a monotone path in 
+$\reeb(\MM)$.
+\end{lemma}
+
+\begin{proof} Prove by induction of the number of edges in $\reeb(\MM)$. Consider the head $h$ of $\st$ in the first call to $\onestep(\MM)$,
+and the contour $\cont(y)$ as defined in \Thm{piece}. Take $\cut(\MM,\cont(y)) = \{\MM_u,\MM_\ell\}$. Apply induction
+hypothesis on $\MM_\ell$, and use that to argue about $T(c_1)$ and $T(c_2)$ in the original manifold.
+\end{proof}
+
+\pagebreak
+\newcommand{\pathTree}{P_{\mathcal{S}}}
+\newcommand{\pathTreeA}{\mathcal{P}_{\mathcal{S}}}
+\section{Leaf assignments and path decompositions}
+
+In this section, we set up a framework to analyze the time taken to compute a merge tree.
+
+\begin{definition}
+ We say that a contour tree of a $2$-manifold is a \emph{merge tree} if it has a well defined root vertex such that height values 
+ strictly increase along any root to leaf path.
+\end{definition}
+
+\begin{observation}
+ The contour tree of a minimum dominant manifold (see \Def{dom}) is close to a merge tree.  Specifically, there is a root such that on any root 
+ to leaf path height values strictly increase, with the only exception being potentially at the leaves themselves.  As such, up to constant 
+ factors, any lower bound for computing the merge tree is a lower bound for computing the contour tree of a minimum dominant
+ manifold (i.e. intuitively just clip off the leaves or shift them to lie above their parent).
+\end{observation}
+
+Any tree $T$ considered in this section will be a rooted full binary tree\footnote{Note 
+that technically the trees considered should have a leaf vertex hanging below the root of this 
+full binary tree in order to represent the global minimum of the surface.  This vertex is 
+(safely) ignored to simplify the presentation.} where the height of a vertex is its distance 
+from the root $r$.  As such, the children of a vertex $v\in T$ are the adjacent vertices of larger height
+(and $v$ is the parent of such vertices).  Then the subtree rooted at $v$, denoted $T_v$ consists of the graph induced on all 
+vertices which are descendants of $v$ (including $v$ itself).  For two vertices $v$ and $w$ in $T$ let $d(v,w)$ denote the 
+length of the path between $v$ and $w$.
+We use $A(v)$ to denote the set of ancestors of $v$.
+For set of nodes $U$, $A(U) = \bigcup_{u \in U} A(u)$.
+
+\begin{definition}
+ A \emph{leaf assignment} $\chi$ of a tree $T$ assigns \emph{two} distinct leaves in each internal vertex $v$,
+ one from the left child and one from the right child subtree of $v$.
+\end{definition}
+
+For a vertex $v\in T$, we use $H_v$ to denote the \emph{heap} at $v$.
+Formally, $H_v = \{u | u \in A(v), \chi(u) \cap L(T_v) \neq \emptyset\}$.
+In words, $H_v$ is the set of ancestors of $v$ which are colored by some leaf in $T_v$.
+The following lemma should justify these technical definitions.
+
+\begin{lemma}
+The running time of our algorithm is $O(\sum_{v \in T} \log |H_v|)$.
+% where $T$ is the merge tree of the given input, 
+%and $\chi$ is any valid coloring of $T$.  (maybe add one in the definition of $H_v$ so that $\log |H_v|$ is always $>1$).
+\end{lemma}
+
+\begin{proof} \Sesh{add}
+
+\end{proof}
+
+%vertices along the path from the root to $v$ such that one of the two colors of each vertex is in $L(T_v)$.
+%We will refer to $H_v$ as the \emph{heap} at $v$.
+%Then define the \emph{algorithm} cost function as $f_{alg}(\chi(T)) = \sum_{v\in T} \log |H_v|$.
+%
+%\begin{lemma}
+%\label{lem:algCost}
+%NEEDS TO BE WRITTEN:
+%Here we need a lemma showing that the running time of our algorithm is $O(f_{alg}(\chi(T)))$, where $T$ is the merge tree of the given input, 
+%and $\chi$ is any valid coloring of $T$.  (maybe add one in the definition of $H_v$ so that $\log |H_v|$ is always $>1$).
+%\end{lemma}
+%
+%Now that we have a simple expression upper bounding the running time of our algorithm, we spend the remainder 
+%of this section proving that this running time is optimal.
+
+We introduce the concept of path decompositions. A \emph{leaf path} is simply a contiguous path in $T$
+containing a leaf, which is monotonic in the height values of its vertices.
+% Then in the following two subsections 
+%we relate path decomposition costs (defined below) to the running time of our algorithm using $\Lem{algCost}$.
+
+\begin{definition}
+\label{def:path} A \emph{path decomposition}, $P(T)$, is a partition of $T$ into 
+ a set of vertex disjoint leaf paths.  
+% 
+% The \emph{cost} of $P(T)$ is $\cost(P(T)) = \sum_{p\in P(T)} |p| \log |p|$. (We use $|p|$ to denote the number of vertices in $p$.)
+\end{definition}
+
+Note that the number of paths in $P(T)$ must equal the number of leaves in $T$.
+Fix $\chi$, which fixes all the heaps $H_v$. We choose a special path decomposition that is best defined as a subset of edges in $T$ such that each internal
+vertex has degree at most $2$. This naturally gives a path decomposition.
+For each internal vertex $v\in T$, add the edge from $v$ to $\arg \max_{v_l, v_r} \{|H_{v_l}|, |H_{v_r}|\}$ 
+ where $v_l$ and $v_r$ are the children of $v$ (if $|H_{v_l}|=|H_{v_r}|$ then pick one arbitrarily).
+This is called the \emph{maximum} path decomposition, denoted by $\pmax(T)$.
+
+Our main theorem is the following. We use $|p|$ to denote the number of vertices in $p$.
+
+\begin{theorem} \label{thm:runtime} $\sum_{v \in T} \log |H_v| = O(\sum_{p\in \pmax(T)} |p| \log |p|)$.
+\end{theorem}
+
+\subsection{Shrubs, tall paths, and short paths}
+
+The paths in $P(T)$ naturally define a tree\footnote{Please excuse the 
+overloading of the term 'tree', it is the most natural term to use here.} of their own.  Specifically, in the original 
+tree $T$ contract each path down to its root.  Call the resulting tree the \emph{shrub} of $T$ corresponding to 
+the path decomposition $P(T)$. Abusing notation, we simply use $P(T)$ to denote the shrub.
+As a result, we use terms like `parent', `child', `sibling', etc. for paths as well.
+The shrub gives a handle on the heaps of a path.
+We use $b(p)$ to denote the \emph{base} of the path, which is vertex in $p$ closest
+to root of $T$. We use $\ell(p)$ to denote the leaf in $p$.
+We use $H_p$ to denote the $H_{b(p)}$.
+% and denote it by $\pathTree(T)$.
+
+\begin{lemma}
+\label{lem:adjacent}
+ Let $p$ be any path in $P(T)$ and let $\{q_1, \dots q_k\}$ be the children on $p$.
+Then $H_{\ell(p)} + \sum_{i=1}^k |H_{q_i}| \leq |H_p|+2|p|$. 
+\end{lemma}
+
+\begin{proof} For convenience, denote $H_i = H_{q_i}$ and $H_0 = H_{\ell(p)}$.
+Consider $v \in \bigcup_{i} H_i$ that lies below $b(p)$ in $T$. 
+Note that such a vertex has only one of its two colors in $L(b(p))$.  
+Since the colors tracked by $H_i$ and $H_j$ for $i\neq j$ are disjoint, 
+such a vertex can appear in only one of the $H_i$'s.
+On the other hand, a vertex $u\in p$ can appear in more than one $H_i$, 
+but since any vertex has exactly two colors it can appear in at most two 
+such heaps.  Hence, $\sum_i |H_i| \leq |H_p| + 2|p|$.
+\ignore{
+ First observe that for all $i$, $r_{i}$ lies above $r_p$ on some root to leaf path. 
+ Let $H_i'$ be the subset of $H_i$ that lies below $r_p$.  Observe that $H_i'\subseteq H_p$, as $L(T_{r_i})\subseteq L(T_{r_p})$.
+
+ Now for any $i\neq j$, $H_i \cap H_j = \emptyset$, as $r_{i}$ and $r_{j}$ are not on the same root to leaf path (i.e. $L(T_{r_i}) \cap L(T_{r_j}) = \emptyset$).
+ In particular, $(H_i\setminus H_i') \cap (H_j \setminus H_j') = \emptyset$, and so $\sum_{i=0}^k |H_i\setminus H_i'| \leq |p|$ 
+ as $H_i\setminus H_i'$ is precisely the subset of $H_i$ that lies on $p$.
+ Moreover, for any $i\neq j$, $ H_i' \cap H_j' = \emptyset$, and so since $H_i'\subseteq H_p$ we have $\sum_{i=0}^k |H_i'|\leq |H_p|$.
+ Putting these together gives
+ \[
+  \sum_{i=0}^k |H_i| = \sum_{i=0}^k |H_i'| + \sum_{i=0}^k |H_i\setminus H_i'| \leq |H_p|+|p|.
+ \]
+}
+\end{proof}
+
+We wish to prove $\sum_{v\in T} \log |H_v| = O(\sum_{p\in P} |p| \log |p|)$. 
+The simplest approach is to prove $\forall p\in P$, $\sum_{v\in p} \log |H_v|= O(|p|\log|p|)$.
+This is unfortunately not true, which is why we divide paths into two categories.
+
+\begin{definition}
+ For $p\in P(T)$, $p$ is \emph{short} if $|p| < \sqrt{|H_{p}|}/100$, and \emph{tall} otherwise.
+\end{definition}
+
+\ignore{
+\begin{observation}
+\label{obs:decrease}
+ Let $v$ be a vertex in $T$ and let $w$ be its parent.  Then $|H_w| \geq |H_v| -1$, as $L(T_v) \subseteq L(T_w)$ and the path from 
+ $w$ to the root has one less vertex than the path from $v$ to the root.
+ In particular, we have the following more general property.
+ Let $v$ and $u$ be any two vertices in the same root to leaf path of $T$, such that $v$ is a descendant of $u$.  
+ Then $|H_v| \leq |H_u| + d(u,v)$. 
+\end{observation}
+
+The following can be thought of as a generalization of the above observation, and will be useful in later sections.
+}
+
+The following lemma demonstrates that tall paths can ``pay'' for themselves.
+
+\begin{lemma}
+\label{lem:pathBounds} If $p$ is tall, $\sum_{v\in p} \log |H_v| = O(|p| \log |p|)$.
+If $p$ is short, $\sum_{v\in p} \log |H_v| = O(|H_p| \log |H_p|)$.
+% Let T be a binary tree and let $\chi(T)$ and $P(T)$ be a valid coloring and path decomposition of $T$, respectively.
+% Then, 
+% \begin{compactenum}
+%  \item for any tall path $p\in P(T)$, $\sum_{v\in p} \log |H_v| = O(|p| \log |p|)$.
+%  \item for any short path $p\in P(T)$, $\sum_{v\in p} \log |H_v| = O(|p| \log |H_{p}|)$.
+% \end{compactenum}
+\end{lemma}
+\begin{proof} \ignore{
+ By $\Obs{decrease}$ we know that for any vertex $v\in p$, $|H_v|\leq |H_{p}| + |p|$ (as $v$ is a descendant of $r_p$ along $p$).
+}
+For $v \in p$, $|H_v|\leq |H_{p}| + |p|$ (as $v$ is a descendant of $b(p)$ along $p$).
+Hence, $\sum_{v\in p} \log |H_v| \leq \sum_{v\in p} \log(|H_{p}| + |p|)  = |p| \log (|H_{p}| + |p|)$.
+ If $p$ is a tall path, then $|p| \log (|H_{p}| + |p|) = O(|p| \log |p|)$. If $p$ is short, then 
+ $|p| \log (|H_{p}| + |p|) = O(|p| \log |H_{p}| )$. For short paths, $|p| = O(|H_p|)$.
+\end{proof}
+
+\ignore{
+For a short path $p$ we can think of the quantity $|p|(\log |H_p| - \log |p|)$ as the excess of $p$, i.e. what was not ``paid for'' locally by $p$.
+Our eventual goal will be to show that by choosing the right path decomposition, this excess 
+can be recharged to tall paths. In the next couple sections we introduce the appropriate path decomposition and prove 
+some useful facts about it.  Then given these facts, in \Sec{} we are able to make the recharging argument.
+}
+
+There are some short paths that we can also ``pay" for.  Consider any short path $p$ in the shrub.  We will refer to the 
+\emph{tall support chain} of $p$ as the tall ancestors of $p$ in the shrub which have a path to $p$ which does not use any 
+short path (i.e. it is a chain of paths adjacent to $p$).
+
+\begin{definition} \label{def:support} A short path $p$ is \emph{supported} if
+at least $|H_p|/100$ vertices $v$ in $H_p$ lie in paths in the tall support chain of $p$.
+%
+%
+% there exists
+%a contiguous (from $p$) chain of ancestors $q_1, q_2, \ldots, q_k$ (in $P(T)$) such that $|H_{q_k}| \leq 99|H_p|/100$
+%and all $q_i$'s are tall.
+\end{definition}
+
+Let $\cL$ be the set of short paths, $\cL'$ be the set of supported short paths, and $\cH$ be the set of tall paths
+given by $\pmax(T)$.
+We now construct the shrub of unsupported short paths. Consider $p \in \cL \setminus \cL'$,
+and traverse the chain of ancestors from $p$. Eventually, we must reach another short path $q$.
+(If not, we have reached the root $r$ of $\pmax(T)$. Hence, $p$ is supported.)
+Insert edge from $p$ to $q$, so $q$ is the parent of $p$ in $\cU$. This construction leads
+to the shrub forest of $\cL \setminus \cL'$, where all the roots are supported short paths, 
+and the remaining nodes are the unsupported short paths.
+
+Most of the work goes into proving the following technical lemma.
+
+\begin{lemma} \label{lem:cu-root} Let $\cU$ denote a connected component (shrub) in the shrub forest of $\cL \setminus \cL'$ and let $r$
+be the root of $\cU$. (i) For any $v \in p$ such that $p \in \cU$,
+$|H_v| = O(|H_r|)$. (ii) $\sum_{p \in \cU} |p| = O(|H_r|)$.
+\end{lemma}
+
+We split the remaining argument into two subsections. We first prove \Thm{runtime} from \Lem{cu-root},
+which involves routine calculations. Then we prove \Lem{cu-root}, where the interesting work happens.
+
+\subsection{Proving \Thm{runtime}} \label{sec:thm-runtime}
+
+We split the summation into tall, short, and unsupported short paths.
+\begin{eqnarray*} 
+\sum_{p \in \cL} \sum_{v \in p} \log |H_v| & = & \sum_{p \in \cL \setminus \cL'} \sum_{v \in p} \log |H_v| + \sum_{p \in \cL'} \sum_{v \in p} \log |H_v| + \sum_{p \in \cH} \sum_{v \in p} \log |H_v|
+%& = & \sum_{p \in \cL \setminus \cL'} \sum_{v \in p} \log |H_v| + O(\sum_{p \in \cL'} |p|\log |H_p|) = \sum_{p \in \cL \setminus \cL'} \sum_{v \in p} \log |H_v| + O(\sum_{p \in \cL'} |H_p|\log |H_p|)
+\end{eqnarray*}
+The last term can be bounded by $O(\sum_{p \in \pmax(T)} |p|\log |p|)$, by \Lem{pathBounds}.
+The second term can be bounded by $O(\sum_{p \in \cL'} |H_p| \log |H_p|)$, by \Lem{pathBounds} again.
+The following claim shows that this in turn is at most the last term.
+
+%We have a claim that also bounds heap sizes for special set of paths.
+
+\begin{claim} \label{clm:above} $\sum_{p \in \cL'} |H_p| \log |H_p| = O(\sum_{q \in \cH} \sum_{v \in q} \log |H_v|)$.
+\end{claim}
+
+\begin{proof} 
+Pick $p \in \cL'$. 
+%Let the first $|H_p|/200$ vertices the lie in tall paths be denoted as $H'_p$. 
+%Observe that for any $v \in H'_p$, 
+%We have $|H_p| \log |H_p| = O(\sum_{v \in H'_p} \log |H_p|)$
+%$= O(\sum_{v \in H'_p} \log |H_v|)$. Since all the vertices lie in tall paths,
+%we can write this as $O(\sum_{q \in \cH} \sum_{v \in H'_p \cap q} \log |H_v|)$.
+%Summing over all $p$, the expression is $\sum_{q \in \cH} \sum_{p \in \cL'} \sum_{v \in H'_p \cap q} \log |H_v|$.
+%
+%
+As we traverse the tall support chain of $p$, there
+are at least $|H_p|/100$ vertices of $H_p$ that lie in these paths. These are encountered in
+a fixed order. Let $H'_p$ be the first $|H_p|/200$ of these vertices. When $v \in H'_p$
+is encountered, there are $|H_p|/200$ vertices of $H_p$ not yet encountered. Hence,
+$|H_v| \geq |H_p|/200$. Hence, $|H_p|\log |H_p| = O(\sum_{v \in H'_p} \log |H_v|)$.
+Since all the vertices lie in tall paths,
+we can write this as $O(\sum_{q \in \cH} \sum_{v \in H'_p \cap q} \log |H_v|)$.
+Summing over all $p$, the expression is $\sum_{q \in \cH} \sum_{p \in \cL'} \sum_{v \in H'_p \cap q} \log |H_v|$.
+%and $\sum_{p \in \cL'} |H_p|\log |H_p| = O(\sum_{p \in \cL'} \sum_{v \in H'_p} \log |H_v|)$.
+
+Consider any $v \in H'_p$.  Let $S$ be the set of paths $\widetilde{p}\in \cL'$ such that 
+$v\in H'_{\widetilde{p}}$.  We now show $|S|\leq 2$ (i.e. it contains at most one path other than $p$).
+First observe that any two paths in $S$ must be unrelated (i.e. $S$ is an anti-chain), 
+since paths which have an ancestor-descendant relationship have disjoint tall support chains.  
+However, any vertex $v$ receives exactly one color from each of its two subtrees (in $T$), and therefore $|S|\leq 2$ since any two 
+paths which share descendant leaves in $T$ (i.e. their heaps are tracking the same color) must have an ancestor-descendant relationship.
+
+In other words, any $\log |H_v|$ appears at most twice in the above triple summation.
+Hence, we can bound it by $O(\sum_{q \in \cH} \sum_{v \in q} \log |H_v|)$.
+%
+%Pick $q \in Q$ and some $v \in H_q$. As we traverse up the tree passing
+%through $A(q)$, we must encounter some $p \in A(q)$ such that $v \in p$. Furthermore,
+%there is an order in which these vertices (of $H_q$) occur. Let $H'$ be the first half
+%of vertices that occur. For all $v \in H'$, $|H_v| \geq |H_q|/2$. Hence,
+%$|H_q| \log |H_q| \leq 4 \sum_{v \in H'} \log |H_v| \leq 4 \sum_{v \in H_q} |H_v|$.
+%So $\sum_{q \in Q} |H_q| \log |H_q| \leq 4 \sum_{q \in Q} \sum_{v \in H_q} |H_v|$.
+%
+%Since nodes in $Q$ are incomparable, all the $H_q$'s (for $q \in Q$) are disjoint.
+%Each such $v$ must occur in some path $p \in A(Q)$. Thus,
+%$\sum_{q \in Q} |H_q| \log |H_q| \leq 4 \sum_{p \in A(Q)} \sum_{v \in p} |H_v|$.
+%We apply the bound of \Lem{pathBounds} to complete the proof.
+\end{proof}
+
+The first term (unsupported short paths) can be charged to the second term (supported short paths).
+This is where the critical \Lem{cu-root} plays a role.
+
+\begin{claim} \label{clm:first} $\sum_{p \in \cL \setminus \cL'} \sum_{v \in p} \log |H_v| = O(\sum_{p \in \cL'} |H_p| \log |H_p|)$.
+\end{claim}
+
+\begin{proof} Let $\cU$ denote a connected component of the shrub forest.
+We have $\sum_{p \in \cL \setminus \cL'} \sum_{v \in p} \log |H_v| \leq \sum_{\cU} \sum_{p \in \cU} \sum_{v \in p} \log |H_v|$.
+By \Lem{cu-root}, $|H_v| = O(|H_r|)$, where $r$ is the root of $\cU$.
+Furthermore, $\sum_{p \in \cU} |p| = O(|H_r|)$. 
+We have $\sum_{p \in \cU} \sum_{v \in p} \log |H_v| = O((\log |H_r|) \sum_{p \in \cU} |p|) = O(|H_r|\log |H_r|)$.
+We sum this over all $\cU$ in the shrub forest, and note that roots in the shrub forest are supported short paths.
+\end{proof}
+
+\ignore{
+Now, for the critical piece of the approach. Focus on $\pmax(T)$. Let $L$ be the set of short paths, 
+and $\hat{L} \subseteq L$ be the set of paths such that no ancestor of $p \in L'$
+is short. Obviously, $\hat{L}$ is an anti-chain in $P(T)$.
+
+\begin{lemma} \label{lem:short} For the path decomposition $\pmax(T)$, $\sum_{p \in \cL} \sum_{v \in p} \log |H_v| = O(\sum_{p \in \cL'} |H_p| \log |H_p|)$.
+\end{lemma}
+
+With this lemma, the proof of \Thm{runtime} is straightforward.
+
+\begin{proof} (of \Thm{runtime}) We focus on $\pmax(T)$ and split into tall and short paths.
+We have $\sum_{v \in T} \log |H_v| =$  $\sum_{p \in \cL} \sum_{v \in p} \log |H_v| + $ $\sum_{p \in \cH} \sum_{v \in p} \log |H_v|$.
+For the first term, we apply \Lem{short} to bound by $O(\sum_{p \in \cL'} |H_p| \log |H_p|)$.
+Then we apply \Clm{above} to bound by $O(\sum_{p \in \cH} |p|\log |p|)$.
+The latter is also bounded by the same expression, by \Lem{pathBounds}. By definition,
+$\sum_{p \in \cH} |p|\log |p| \leq \cost(\pmax(T))$.
+\end{proof}
+
+The main challenge is proving \Lem{short}, which we defer to the next section.
+\begin{definition}
+ Let $\chi(T)$ be a valid coloring of a binary tree $T$.  We define a set of disjoint paths over the vertices of 
+ $T$ as follows.  
+ The corresponding collection of paths that all such edges correspond to is a path decomposition of $T$, which 
+ we refer to as a \emph{maximum} path decomposition, and is denoted by $\mathcal{P}(T)$.  
+\end{definition}
+
+From the previous section we now have a lower bound of $\Omega(f_{path}(P))$ for computing the merge tree in terms of any path decomposition $P$. 
+We now show that the running time of our algorithm for computing the merge tree is upper bounded by this expression for a specific path decomposition, 
+which we will call the maximum path decomposition (defined below).
+Specifically, for a given valid coloring $\chi$ of a tree $T$, we will prove that for the maximum path decomposition $f_{alg}(\chi(T)) = O(f_{path}(P(T)))$. 
+%using the observations about path decompositions from 
+%\Sec{background}, which will then imply $f_{alg}(\chi(T)) = \Theta(f_{path}(P(T)))$.
+
+Throughout this section, for a path $p\in P(T)$, we use the notation $H_p$ to refer to the heap at the root of $p$, i.e. $H_p = H_{r_p}$.
+}
+
+\subsection{Proving \Lem{cu-root}: the root is everything in $\cU$} \label{sec:weight}
+
+\Lem{cu-root} asserts the root $r$ in $\cU$ pretty much encompasses all sizes and heaps in $\cU$.
+We will work with the \emph{reduced} heap $\redH_p$. This is the subset of vertices of $H_p$ that 
+do not appear on the tall support chain of $p$.
+%only appear in short paths. 
+By definition, for any unsupported short path
+(hence, any non-root $p \in \cU$), $|\redH_p| \geq 99|H_p|/100$. 
+% For notational convenience, we simply set $\redH_r = H_r$.
+%Consider $q \in \cL \setminus \cL'$ that is a child of $p$ in the shrub $\cU$.
+We begin with a key property, which is where the construction of $\pmax(T)$ enters the picture.
+
+%
+%
+%
+%We focus on $\pmax(T)$, and prove the key property of this path decomposition.
+
+\begin{lemma}
+\label{lem:geometric}
+ Let $q$ be the child of some path $p$ in $\cU$, then $|H_p|\geq \frac{3}{2}|H_q|$. 
+ Moreover, if $p\neq r(\cU)$, then $|\redH_p|\geq \frac{3}{2}|\redH_q|$. 
+\end{lemma}
+\begin{proof} Let $h(q)$ denote the tall path that is a child of $p$ in $\pmax(T)$,
+and an ancestor of $q$. If no such tall path exists, then by construction $p$ is the parent 
+of $q$ in $\pmax(T)$, and the following argument will go through by setting $h(q)=q$.
+
+The chain of ancestors from $q$ to $h(q)$ consists only of tall paths.
+Since $q$ is unsupported, these paths contain at most $|H_q|/100$ vertices of $H_q$. 
+Thus, $|H_{h(q)}| \geq 99|H_q|/100$.
+
+Consider the base
+of $h(q)$, which is a node $w$ in $T$. Let $v$ denote the sibling of $w$ in $T$.
+Their parent is called $u$. Note that both $u$ and $v$ are nodes in the path $p$.
+Now, the decomposition $\pmax(T)$ put $u$ and $v$ in the same path $p$. This
+implies $|H_v| \geq |H_w|$. Since $|H_u| \geq |H_v| + |H_w| - 2$,
+$|H_u| \geq 2|H_w| - 2$. Let $b$ be the base of $p$.
+We have $|H_p| = |H_b| \geq |H_u| - |p| \geq 2|H_w| - |p| - 2$. Since $p$ is a short path, $|p| < \sqrt{|H_p|}/100$.
+Applying this bound, we get $|H_p| \geq (2-\delta)|H_w|$ (for a small constant $\delta > 0$).
+Since $w$ is the base of $h(q)$, $H_w = H_{h(q)}$. We apply the bound $|H_{h(q)}| \geq 99|H_q|/100$
+to get $|H_p| \geq 197|H_q|/100$, implying the first part of the lemma.  For the second part, observe that 
+if $p\neq r(\cU)$, then $p$ is unsupported and so $|\redH_p| \geq 99|H_p|/100$, and therefore the second part follows since $|H_q| \geq |\redH_q|$.
+%
+%
+% Let $w$ be the vertex in the path $p$ to which the root of the path $q$ is adjacent to in $T$, 
+% i.e. $w$ is the parent of $r_q$ in $T$.  Let $c$ be the other child of $w$, which lies in $p$.
+% As $\mathcal{P}(T)$ is a maximum path decomposition of $T$ and $c$ rather than $r_q$ is in 
+% the same path as $w$, we have $|H_q|\leq |H_c|$.  Therefore, $|H_w|\geq |H_q|+|H_c|-1\geq 2|H_q|-1$. 
+% Therefore, by \Obs{decrease}, $|H_p|+ d(r_p, w) \geq |H_w|\geq 2|H_q|-1$.  
+% Since $p$ is a short path, $d(r_p, w)\leq \sqrt{|H_p|}/100$ and so we have $|H_p|+\sqrt{|H_p|}/100\geq 2|H_q|-1$.
+% Assuming $|H_p|>11$ (and we can safely assume $|H_p|$ is bigger than any constant), this yields $\frac{3}{5} |H_p| \geq |H_q|$
+\end{proof}
+
+This immediately proves part (i) of \Lem{cu-root}. Part (ii) requires much more work.
+% It is helpful to spell out how the various heaps evolve. Consider $p \in \cU$,
+% which has children $q_1, q_2, \ldots, q_k$ in $\pmax(T)$, and a leaf vertex $\ell$ in $T$. 
+% The reduced heap $\redH_p$ is the
+% union of $\redH_\ell$ and $\redH_{q_1}, \redH_{q_2}, \ldots, \redH_{q_k}$, with the vertices in $p$
+% removed. Note that some of the $q_i$'s are the corresponding $h(q')$ vertices,
+% for a child $q'$ in $\cU$.
+
+We define a \emph{residue} $R_p$ for each $p \in \cU$. Suppose $p$ has children
+$q_1, q_2, \ldots, q_k$ in $\cU$. Then $R_p = |\redH_p| - \sum_i |\redH_{q_i}|$. 
+By definition, $|\redH_p| = \sum_{q \in \cU_p} R_p$. 
+Note that $R_p$ can be negative. 
+Now, define $R^+_p = \max(R_p,0)$, and set $W_p = \sum_{q \in \cU_p} R^+_p$.
+Observe that $W_p \geq |\redH_p|$. We also get an approximate converse.
+
+\begin{claim} \label{clm:loss} For any path $p\in \cU$, $|\redH_p| \geq W_p - 2\sum_{q \in \cU_p} |q|$.
+\end{claim}
+
+\begin{proof} We write $W_p - |\redH_p| = \sum_{q \in \cU_p} R^+_q - R_q$ $= -\sum_{q \in \cU_p: R_q < 0} R_q$.
+Consider $q \in \cU_p$ and denote the children in $\cU_p$ by $q'_1, q'_2, \ldots$.
+Note that $R_q$ is negative exactly when $|\redH_q| < \sum_i |\redH_{q'_i}|$.
+Traverse $\pmax(T)$ from $q'_i$ to $q$.
+Other than $q$, all other nodes encountered are in the tall support chain of $q'_i$ and hence
+do not affect its reduced heap.
+The vertices of $\redH_{q'_i}$ that are deleted
+are exactly those present in the path $q$. 
+Any vertex in $q$ can be deleted from at most two of the reduced heaps (of the children of $q$ in $\cU_p$), since 
+theses reduced heaps do not have an ancestor-descendant relationship.
+Therefore when $R_q$ is negative, it is at most by $2|q|$.
+We sum over all $q$ to complete the proof.
+\end{proof}
+
+\ignore{Let $D_q = H_q \setminus \bigcup_i H_{q'_i}$. In other words, as we traverse $\cU$,
+$D_q$ is the set of vertices ``added" at $q$. Suppose we traverse $\pmax(T)$ from $q'_i$
+to $q$. All vertices of $H_{q'_i}$ that lie on these paths are removed.
+
+ and traverse $\pmax(T)$ up to $p$. Suppose we encounter
+the paths $q_0 = q, q_1, q_2, \ldots, q_k = p$. The vertices in the residue $R_q$ that 
+are removed from $H_p$ are exactly the vertices that lie in these paths. 
+We split this contribution into tall and short paths, observing that the latter
+are all unsupported. Hence, we can (crudely) upper bound the number of vertices
+removed by $\sum_{q \in \cU_p} |q| + \sum_{q \in \cU_p} \sum_{h \in \cH} |h \cap q|$.
+Each vertex can be removed from at most two distinct $R_q$'s, completing the proof.
+}
+
+The main challenge of the entire proof is bounding the sum of path lengths, which is done next.
+We stress that the all the previous work is mostly the setup for this claim.
+
+\begin{claim} \label{clm:charge} Fix any path $p \in \cU\setminus\{r(\cU)\}$. Suppose for any $q,q' \in \cU_p$
+where $q$ is a parent of $q'$ in $\cU_p$, $W_q \geq (4/3)W_{q'}$. Then $\sum_{q \in \cU_p} |q| \leq W_p/20$.
+\end{claim}
+
+\begin{proof} Since $q$ is an unsupported short path, $|q| < \sqrt{|H_q|}/100 \leq \sqrt{|\redH_q|}/99 \leq \sqrt{|W_q|}/99$.
+We prove that $\sum_{q \in \cU_p} \sqrt{|W_q|}/99 \leq W_p/20$
+by a charge redistribution scheme. Assume that each $q \in \cU_p^-$
+starts with $\sqrt{W_q}/99$ units of charge. We redistribute this charge over all
+nodes in $\cU_q$, and then calculate the total charge. For $q \in \cU_p$, spread
+its charge to all nodes in $\cU_q$ proportional to $R^+$ values. In other words,
+give $(\sqrt{W_q}/99)\cdot(R^+_{q'}/W_q)$ units of charge to each $q' \in \cU_q$.
+
+After the redistribution, let us compute the charge deposited at $q$. Every ancestor in $\cU_p^-$ 
+$q = a_0, a_1, a_2, \ldots, a_k$ contributes to the charge at $q$. The charge is expressed in the following
+equation. We use the assumption that $W_{a_i} \geq (4/3) W_{a_{i-1}}$ and hence $W_{a_i} \geq (4/3)^i W_{a_0} \geq (4/3)^i$, 
+as $a_0$ is an unsupported short path and hence $W_{a_0}\geq 1$. 
+$$ ({R^+_q}/99)\sum_{a_i} 1/\sqrt{W_{a_i}} \leq ({R^+_q}/99)\sum_{a_i} (3/4)^{i/2} \leq R^+_q/20 $$
+The total charge is $\sum_{q \in \cU_p} R^+_p/20 = W_p/20$. 
+\end{proof}
+
+\begin{corollary} \label{cor:chargeCor}
+ Let $r$ be the root of $\cU$, and suppose that for any paths $q,q' \in \cU\setminus \{r\}$, where $q$ is a parent of $q'$ in $\cU$, $W_q \geq (4/3)W_{q'}$.
+ Then $\sum_{p \in \cU} |p| \leq W_r/20+|r|$.
+\end{corollary}
+\begin{proof}
+ Let $c_1, \dots, c_m$ be the children of $r$ in $\cU$.  By definition, $W_r = \sum_i W_{c_i}+R^+_r \geq \sum_i W_{c_i}$.
+ By \Clm{charge}, for each $c_i$ we have $W_{c_i}/20\geq \sum_{p \in \cU_{c_i}} |p|$.  Combining these to facts yields the claim.
+\end{proof}
+
+% \begin{claim} \label{clm:charge} Fix $p \in \cU$. Suppose for any $q,q' \in \cU_p$
+% where $q$ is a parent of $q'$ in $\cU_p$, $W_q \geq (4/3)W_{q'}$. Then $\sum_{q \in \cU_p} |q| \leq W_p/20$.
+% \end{claim}
+% 
+% \begin{proof} Since $q$ is an unsupported short path, $|q| < \sqrt{|H_q|}/100 \leq \sqrt{|\redH_q|}/99 \leq \sqrt{|W_q|}/99$.
+% We prove that $\sum_{q \in \cU_p} \sqrt{|W_q|}/99 \leq W_p/20$
+% by a charge redistribution scheme. Assume that each $q \in \cU_p$
+% starts with $\sqrt{W_p}/99$ units of charge. We redistribute this charge over all
+% nodes in $\cU_p$, and then calculate the total charge. For $q \in \cU_p$, spread
+% its charge to all nodes in $\cU_q$ proportional to $R^+$ values. In other words,
+% give $(\sqrt{W_q}/99)\cdot(R^+_{q'}/W_q)$ units of charge to each $q' \in \cU_q$.
+% 
+% After the redistribution, let us compute the charge deposited at $q$. Every ancestor in $\cU_p$ 
+% $q = a_0, a_1, a_2, \ldots, a_k$ contributes to the charge at $q$. The charge is expressed in the following
+% equation. We use the assumption that $W_{a_i} \geq (4/3) W_{a_{i-1}}$ and hence $W_{a_i} \geq (4/3)^i W_{a_0} \geq (4/3)^i$, 
+% as $a_0$ is an unsupported short path and hence $W_{a_0}\geq 1$. 
+% $$ ({R^+_q}/99)\sum_{a_i} 1/\sqrt{W_{a_i}} \leq ({R^+_q}/99)\sum_{a_i} (3/4)^{i/2} \leq R^+_q/20 $$
+% The total charge is $\sum_{q \in \cU_p} R^+_p/20 = W_p/20$. 
+% \end{proof}
+
+We wrap it all up by proving part (ii) of \Lem{cu-root}.
+
+\begin{claim} $\sum_{p \in \cU} |p| \leq |H_{r(\cU)}|/10$.
+\end{claim}
+
+% \begin{proof} We use $r$ for $r(\cU)$. Our aim is to simply apply \Clm{charge} from $r(\cU)$. 
+% Suppose $W_q \geq (4/3)W_{q'}$ (for any choice in $\cU$ of $q$ parent of $q'$). By \Clm{charge}, $\sum_{p \in \cU} |p| \leq W_{r}/20$.
+% By \Clm{loss}, $|\redH_r| \geq W_r - 2\sum_{p \in \cU} |p|$. By simple algebra, we deduce, $\sum_{p \in \cU} |p| \leq |\redH_r|/10 = |H_r|/10$.
+% 
+% We now prove that for any $q$ parent of $q'$, $W_q \geq (4/3)W_{q'}$.
+% Suppose not. Let $p, p'$ be the counterexample furthest from the root,
+% where $p$ is the parent of $p'$. 
+% Note that for $q$ and child $q'$ in $\cU_{p'}$, $W_q \geq (4/3)W_{q'}$.
+% We will apply \Clm{charge} for $\cU_{p'}$ to deduce that $\sum_{q \in \cU_{p'}} |q| \leq W_{p'}/20$.
+% By \Clm{loss}, $|\redH_{p'}| \geq 19W_{p'}/20$. By \Lem{geometric}, $|\redH_p| \geq (3/2)|\redH_{p'}|$.
+% Noting that $W_p \geq |\redH_p|$, we deduce that $W_p \geq (4/3)W_{p'}$.
+% Hence, $p, p'$ is not a counterexample, and more generally, this is no counterexample.
+% That completes the whole proof.
+% \end{proof}
+
+\begin{proof} We use $r$ for $r(\cU)$. %Our aim is to simply apply \Clm{charge} from $r(\cU)$. 
+Suppose $W_q \geq (4/3)W_{q'}$ (for any choice in $\cU\setminus\{r\}$ of $q$ parent of $q'$), 
+then by \Cor{chargeCor}, $\sum_{p \in \cU} |p| \leq W_{r}/20+|r|$.
+By \Clm{loss}, $|\redH_r| \geq W_r - 2\sum_{p \in \cU} |p|$, and so combining these inequalities gives,
+\[
+% \sum_{p \in \cU} |p| \leq \pth{|\redH_r|+2\sum_{p \in \cU} |p|}/20+|r| \Rightarrow 
+ \sum_{p \in \cU} |p|\leq \frac{10}{9} \pth{|\redH_r|/20+|r|} \leq 
+ \frac{10}{9}\pth{|H_r|/20+\sqrt{|H_r|}/100} \leq |H_r|/10.
+\]
+
+We now prove that for any $q$ parent of $q'$ (other than $r$), $W_q \geq (4/3)W_{q'}$.
+Suppose not. Let $p, p'$ be the counterexample furthest from the root,
+where $p$ is the parent of $p'$. 
+Note that for $q$ and child $q'$ in $\cU_{p'}$, $W_q \geq (4/3)W_{q'}$.
+We will apply \Clm{charge} for $\cU_{p'}$ to deduce that $\sum_{q \in \cU_{p'}} |q| \leq W_{p'}/20$.
+Combining this with \Clm{loss} gives, $|\redH_{p'}| \geq 19W_{p'}/20$. By \Lem{geometric}, $|\redH_p| \geq (3/2)|\redH_{p'}|$.
+Noting that $W_p \geq |\redH_p|$, we deduce that $W_p \geq (4/3)W_{p'}$.
+Hence, $p, p'$ is not a counterexample, and more generally, there is no counterexample.
+That completes the whole proof.
+\end{proof}
+
+\section{Lower Bound by Path Decomposition}
+In this section we provide a lower bound for the time it takes to compute a contour tree in terms the cost of path decompositions.  Specially, 
+we first consider the easier case of merge trees, after which the lower bound for more general contour trees will easily follow.
+
+\subsection{Merge Trees}
+In this section we show that any algorithm for computing the merge tree $T$ must pay at least $f_{path}(P(T))$ (see \Def{path}) for any path decomposition $P(T)$.
+
+%We say two shrubs $S_1$ and $S_2$ (of merge trees of potentially different surfaces) are equivalent if there is an isomorphism, $f$,
+%between the two shrubs such that for any node $v\in S_1$ corresponding to a path of size $|v|$, $f(v)$ also corresponds to a path of size $|v|$.
+%For the purposes of the next lemma, we say two $2$-manifolds are different, if their corresponding contour trees are the different.
+
+In the following note that two distinct PL surfaces can have the same graph structure defining their meshes, but different height values associated with their vertices.
+
+\begin{figure}[h]\centering
+    \includegraphics[width=.15\linewidth]{figs/diamond}\hspace{2cm}
+    \includegraphics[width=.125\linewidth]{figs/topView}\hspace{2cm}
+    \includegraphics[width=.095\linewidth]{figs/sideView}
+    \caption{From left to right: the cone portion of a diamond; the top down view and then the side view of a cone face with a platform added.}
+    \label{fig:diamond}
+\end{figure}
+
+\begin{lemma}
+\label{lem:pathDecomp}
+Let $T$ be a merge tree, and let $P = P(T)$ any path decomposition of $T$.  %, and $\pathTree$ the corresponding shrub.  
+Then there is a set of at least $\Pi_{p_i\in P} (|p_i|-1)!$ distinct PL $2$-manifolds, all of which have the same defining mesh.
+\end{lemma}
+\begin{proof}
+ We describe a simple mesh, with heights specified for all vertices except for the saddles.  The main idea is that 
+ this mesh will be flexible enough to allow almost any assignment of heights to the saddles, thus defining a class of merge trees 
+ whose size is that described in the lemma statement.
+
+ First, we describe the basic building block of this input mesh, which roughly speaking corresponds to a fixed path $p\in P$.  
+ Specifically, we say a \emph{diamond} of size $m$ is the following mesh.  To start, construct a cycle 
+ of length $m$, where the height of the $i$th vertex in the cycle is say $i/2m$.  Next, create an apex vertex at height 
+ $\infty$ (or any appropriately large finite height), and add an edge between this vertex and each vertex in the cycle.  At this point 
+ we have built an upside down cone with $m$ triangular faces (see \Fig{diamond}). To make this cone a closed surface we add a triangular cap at the bottom. 
+ Specifically, add a $3$ cycle of vertices at heights $0$, $1/20m$, and $1/10m$, and connect each vertex of this $3$-cycle to an arbitrary distinct 
+ vertex on the $m$ cycle (with matching cyclic ordering), and triangulate all remaining faces.  For consistency, if the size $m$ is $\leq 3$, the diamond 
+ will be just a tetrahedron.
+
+ Now above each face of the cone of the diamond, we will create a triangular ``platform'', which will act as the location to glue on other (shrunken) diamonds, 
+ and whose highest point of connection to the face below will be a saddle vertex.  Specifically the heights of the vertices of the triangle will be $h$, $h-\eps$, and $h-2\eps$, where 
+ $h$ is some unspecified height value.  This near horizontal platform is then connected to its respective face with vertical edges down to the face. See \Fig{diamond}.
+ %
+ %The vertex at height $h$ will lie in the plane of the respective face of the cone, and the other two will lie above the face, 
+ %with edges connecting down to the face, hence creating an actual ``platform''.
+
+ At this point we can construct the entire input mesh, by gluing together diamonds.  
+ Specifically, for each path $p_i\in P$ we create a diamond of size $|p_i|-1$ (we subtract one because the tip of the diamond already corresponds to the leaf of the path).
+ Now we glue together diamonds of different paths in the same way the paths are connected in the shrub $\pathTree$.  Specially, for two paths $p,q\in P$ where 
+ $p$ is the parent of $q$ in $\pathTree$, we glue the bottom triangular cap of the diamond for $q$ onto the triangular ``platform'' of some face of the diamond for $p$.
+ (Naturally for this construction to work, diamonds for a given path will be scaled down relative to the size of the diamond of their parent).
+
+ Observe now that the only saddle points in this construction are the highest connecting point on a face of an edge from the platform above.  Moreover, the only maxima are the apexes of the 
+ diamonds, and there is exactly one minima corresponding to the bottom cap of the diamond representing the root of $\pathTree$.  Therefore, the saddles on a given 
+ diamond will appear contiguously on a root to leaf path in the merge tree of the mesh, where the leaf corresponds to the maxima of the diamond 
+ (since all these saddles have a direct line of sight to this apex).  In particular, 
+ this implies that, regardless of the heights assigned to the platforms, the merge tree of this mesh has a path decomposition whose corresponding shrub is equivalent to $\pathTree$.
+
+ Now there is a valid instantiation of this described mesh for any relative ordering of the heights of the saddles on a given diamond.  In particular, 
+ there are $(|p|-1)!$ possible orderings of the heights of the saddles on the diamond for $p$, and hence $\Pi_{p_i\in P} (|p_i|-1)!$ possible instantiations of 
+ the mesh overall.  Also, each one of these instantiations will result in a different merge tree.  Specifically, a permutation of the heights on a given diamond 
+ corresponds to a permutation of the vertices along a path in $P$, as by construction all saddles on a given diamond will appear in sorted order in the merge tree.
+\end{proof}
+
+\begin{corollary}
+ \label{cor:mergeLB}
+ Let $A(S)$ be any comparison based algorithm to compute the merge tree of any input surface $S$ (whose contour tree is a merge tree).  
+ Then the running time of $A(S)$ is $\Omega(f_{path}(P))=\Omega(\sum_{p_i\in P} |p_i|\log|p_i|)$ where $P$ is any path decomposition of the 
+ merge tree of $S$.
+\end{corollary}
+\begin{proof}
+ Given an input surface $S$, by the above lemma, any path decomposition $P$ of its merge tree defines a set of $\Pi_{p_i\in P} |p_i|!$ surfaces 
+ with distinct merge trees, all of which have the same mesh.  Specifically, for each sorted ordering of the relative height values of the saddles 
+ on each diamond in the mesh, yields a different merge tree.
+ Therefore, any comparison based algorithm which computes the merge tree must distinguish between these possibilities, and therefore its corresponding decision tree must 
+ have $\Pi_{p_i\in P} |p_i|!$ leaves.  Hence by Stirling's approximation, this decision tree must have depth $\Omega(\sum_{p_i\in P} |p_i|\log|p_i|)$.
+\end{proof}
+
+\subsection{Contour Trees}
+We first generalize previous terms to the case of contour trees.
+In this section $T$ will denote an arbitrary contour tree.  Specifically, every internal vertex in $T$ has degree $3$ 
+and the height function defined over the vertex set is such that every internal vertex in either a split or merge 
+(i.e. at most $2$ neighbors lie above or below it).
+
+Again we define a path decomposition as a partition of the vertices into 
+disjoint leaf paths (i.e. each path is monotonic in height and contains a leaf). 
+However, for simplicity we now restrict our attention to path decompositions which 
+roughly speaking are consistent with the raining procedure described in \Sec{rain} 
+(more general decompositions can work, but it is not needed for our purposes).
+
+\begin{definition}
+\label{def:path2} A path decomposition, $P(T)$, is called \emph{rain consistent} if its paths can be obtained as follows.
+Perform an upward BFS from an arbitrary minima $v$ in $T$, and mark all vertices encountered.  
+Now recursively run a directional BFS from all vertices adjacent to the current marked set.  
+Specifically, for each BFS run, make it an upward BFS if it is at an odd height in the recursion tree and downward otherwise.
+
+This procedure partitions the vertex set into disjoint rooted subtrees of $T$, based on which BFS marked a vertex.  
+For each such subtree, now take any partition of the vertices into leaf paths.\footnote{Note that the subtree  
+of the initial vertex is rooted at is a minima.  For simplicity we require that the path this vertex belongs to 
+also contains a maxima.}
+\end{definition}
+
+The following is analogous to \Lem{pathDecomp}, and in particular uses it as a subroutine.
+
+\begin{lemma}
+\label{lem:pathDecomp2}
+Let $T$ be a contour tree, and let $P = P(T)$ be any rain consistent path decomposition of $T$.  
+Then there is a set of at least $\Pi_{p_i\in P} (|p_i|-1)!$ distinct PL $2$-manifolds, all of which have the same defining mesh.
+\end{lemma}
+\begin{proof}
+ As $P$ is rain consistent, the paths can be partitioned into sets $P_1, \dots, P_m$, where $P_i$ is the set of all paths with 
+ vertices from a given BFS, as described in \Def{path2}.  Specifically, let $T_i$ be the subtree of $T$ corresponding to $P_i$ and 
+ let $r_i$ be the root vertex of this subtree.  
+ Note that the $P_i$ sets naturally define a tree where $P_i$ is the parent of $P_j$ if $r_i$ is adjacent to a vertex in $P_j$.
+
+ As the set $P_i$ is a path decomposition of a rooted binary tree $T_i$, the surface mesh construction of \Lem{pathDecomp} for $P_i$ is well defined.  
+ Actually the only difference is that here the rooted tree is not a full binary tree, and so some of the faces of the constructed diamonds will be blank.  
+ Specifically, these blank faces correspond to the adjacent children of $P_i$, and they tell us how to connect the meshes of the different $P_i$'s.
+
+\begin{figure}[h]\centering
+    \includegraphics[width=.5\linewidth]{figs/bridge2}
+    \caption{On the left, the root of $P_i$ merges into a diamond of its parent $P_j$.  On the right, the root of $P_i$ splits off from a diamond of $P_j$}
+    \label{fig:diamond}
+\end{figure}
+
+ So for each $P_i$ construct a surface mesh as described in \Def{path2}.  By construction, the diamonds corresponding to the paths in $P_i$ are connected 
+ into a tree structure (i.e. corresponding to the shrub of $P_i$).  Therefore the bottoms of all these diamonds are covered except for the one corresponding 
+ to the path containing the root $r_i$.  If $r_i$ is the initial minima that the rain consistent path decomposition was defined from,  
+ we add a tetrahedral cap to the uncovered bottom of this diamond (i.e. we construct the minima).  Otherwise, $P_i$ has some parent $P_j$ in which case 
+ we connect the bottom of the diamond for $r_i$ to a free face of a diamond in the construction for $P_j$, specifically, the face corresponding to the 
+ vertex in $T$ which $r_i$ is adjacent to. 
+ If $T_i$ was created by an upward BFS, then the connection to this face should be a merge vertex, and otherwise it should be a split vertex.  
+ See \Fig{diamond}.
+
+ Just as in \Lem{pathDecomp} we now have one fixed mesh graph structure, such that each different relative ordering of the heights of the merge and split vertices 
+ on each diamond produces a surface with a distinct merge tree.  The specific bound on the number of such distinct manifolds follows by applying the 
+ bound from \Lem{pathDecomp} to each $P_i$.
+\end{proof}
+
+For the same reason that \Cor{mergeLB} followed from \Lem{pathDecomp}, we now have the following.
+
+\begin{theorem}
+ Let $A(S)$ be any comparison based algorithm to compute the contour tree of an input surface $S$.  
+ Then the running time of $A(S)$ is $\Omega(f_{path}(P))=\Omega(\sum_{p_i\in P} |p_i|\log|p_i|)$ where $P$ 
+ is any rain consistent path decomposition of the contour tree of $S$.
+\end{theorem}
+
+% \section{Temp}
+% Given a contour tree we allow any path decomp, as long as each path contains a maxima or minima (or both).
+% We call a path $p$ in the decomp a min (resp. max) path if it contains a minima (resp. maxima) (note a path can be both).
+% For every path we create a cylinder with $m$ faces for $p$, where $m$ is the number of non-leaf vertices in $p$ (either $|p|-1$ or $|p|-2$).  
+% If $p$ has a maxima we put a top cone cap on the cylinder, and if it contains a manima then put a bottom cone cap (i.e. the cap represents the maxima or minima).
+% 
+% Every edge $e$ in $T$, but not in a path decomp. tells us how to glue together cylinders.  Let $u$ be the upper endpoint of $e$ and $l$ the lower endpoint.
+% 
+% First we ``remove'' all edges $e$ such that $u$ is adjacent to a max path and $l$ adjacent to a min path by joining the free ends of the respective cylinder together.
+% Hence we can assume no such edges exist.  Now we have three cases:
+% 
+% \begin{enumerate}[{Case} 1]
+%  \item 
+% $l$ is adjacent to a min path:\newline
+% Note in this case the path containing $u$ must be a min path.
+% In this case connect the bottom of the cylinder for the path containing $l$ into a face of the cylinder for the path containing $u$.
+% \item
+% $l$ is adjacent to a max path and $u$ is adjacent to a max path:\newline
+% In this case connect the bottom of the cylinder for the path containing $u$ into a face of the cylinder for the path containing $l$.
+% \item
+% $l$ is adjacent to a max path and $u$ is adjacent to a min path:\newline
+% Connect a face of the cylinder for the $u$ path (using a split) to a face of the cylinder for the $l$ path (using a merge).
+% \end{enumerate}
+
+\bibliographystyle{alpha}
+\bibliography{contour}
+
+\appendix
+
+\section{Old stuff}
+
+\subsubsection{Paying it down the line}
+
+% \begin{claim}
+% \label{clm:sizes}
+%  Let $p$ be a path in a maximum path decomposition, such that $|H_p|>c|p|$ for some constant $c> 6$.
+%  Let $H_f$ denote the heap at the leaf of $p$.  We then have the following
+%  \begin{compactenum}
+%   \item $|H_f|\leq (1+1/c)|H_p|$
+%   \item $|H_f|\geq |H_p|/3$
+%   \item For any path $q$ adjacent to $p$ in the shrub, $|H_q|<\frac{2}{3}|H_p|$
+%  \end{compactenum} 
+% \end{claim}
+% \begin{proof}
+%   For any vertex $v\in p$ \Obs{decrease} implies that $|H_v|\leq |H_p| + |p| < (1+1/c) |H_p|$.
+%   First observe that by setting $v=f$ this immediately implies the first part of the claim.
+%  % By a similar argument, $|H_f|\geq |H_p| - |p| > (1-1/c)|H_p|$.
+%   
+%   To prove the last part of the claim, suppose otherwise that $|H_q|\geq \frac{2}{3}|H_p|$ for some 
+%   adjacent path $q$.  As this is a maximum path decomposition, this implies that there is some 
+%   $v\in p$ such that $|H_v|\geq \frac{4}{3}|H_p|-1$.
+%   By \Obs{decrease} we then have, 
+%    $|H_p| \geq |H_v| - |p|\geq \frac{4}{3}|H_p| - 1 - |H_p|/c \geq (4/3-1/c)|H_p|-1$.
+%   When rearranged this gives $1\geq (1/3-1/c)|H_p|$.  As $c>6$ we have $1\geq |H_p|/6$, 
+%   which is a contradiction as $|H_p|>c|p|>6$.
+% \end{proof}
+
+\Ben{I had originally thought the following claim would be useful in proving the lemma below.}
+\begin{claim}
+\label{clm:sizes}
+ Let $p$ be a path in a maximum path decomposition, such that $|H_p|>c|p|$ for some constant $c> 6$.
+ We then have the following
+% \begin{compactenum}
+%  \item 
+  For any path $q$ adjacent to $p$ in the shrub, $|H_q|<\frac{2}{3}|H_p|$
+%  \item If there does not exist and adjacent path $q$ such $H_q\geq H_p/3$, then $|H_f|\geq |H_p|/3$
+% \end{compactenum} 
+\end{claim}
+\begin{proof}
+  To prove the claim, suppose otherwise that $|H_q|\geq \frac{2}{3}|H_p|$ for some 
+  adjacent path $q$.  As this is a maximum path decomposition, this implies that there is some 
+  $v\in p$ such that $|H_v|\geq \frac{4}{3}|H_p|-1$.
+  By \Obs{decrease} we then have, 
+   $|H_p| \geq |H_v| - |p|\geq \frac{4}{3}|H_p| - 1 - |H_p|/c \geq (4/3-1/c)|H_p|-1$.
+  When rearranged this gives $1\geq (1/3-1/c)|H_p|$.  As $c>6$ we have $1\geq |H_p|/6$, 
+  which is a contradiction as $|H_p|>c|p|>6$.  
+\end{proof}
+
+\begin{lemma}
+\label{lem:excess}
+ Let $p$ be a tall path, and let $S_1, \dots S_k$ and $R_1, \dots, R_l$ be the sizes of the heaps at the roots of the short and tall children of $p$, respectively, in the shrub.  
+ Let $H_f$ be the heap at the leaf of the path $p$.
+ Then,  $|H_p| \log |H_p|   \geq  \sum_{v\in p} \log|H_v| + \frac{4}{3}\sum_{i=1}^k S_i \log S_i + \sum_{i=1}^l R_i \log R_i  -  c|p|\log |p|$, for some sufficiently large constant $c$
+ \end{lemma}
+\begin{proof}
+ First observe that the lemma holds for $|H_p|$ less than any fixed constant, since otherwise \Lem{adjacent} implies we can choose $c$ 
+ large enough such that $c|p|\log|p|$ will dominate all other terms.
+ Next, observe that $\sum_{v\in p} \log|H_v| \leq \sum_{v\in p} \log(|H_v|+|p|) = |p|\log(|H_v|+|p|) = O(|p|\log|p|)$, as $p$ is a tall path.  
+ Therefore, in the remainder of the proof  we can safely ignore the $\sum_{v\in p} \log|H_v|$ term as we can choose $c$ large enough to cover it.  
+ That is we now instead prove the inequality,
+ \[
+  |H_p| \log |H_p|   \geq \frac{4}{3}\sum_{i=1}^k S_i \log S_i + \sum_{i=1}^l R_i \log R_i  -  c|p|\log |p|
+ \]
+ To do so we  break the analysis into two case based on the relative size of $|p|$ and $|H_p|$. 
+ In the following, $\alpha>10$ is some yet to be determined constant.
+
+\paragraph{Case 1, $|p|\geq |H_p|/\alpha$:\\}
+First observe that by \Lem{adjacent}, $|H_p|+|p| \geq \sum_{i=1}^k S_i+ \sum_{i=1}^l R_i$.
+By choosing $c$ large enough we therefore have,
+\begin{align*}
+ c|p|\log|p| &\geq 2(1+\alpha)^2 |p|\log |p| \geq 2(1+\alpha)|p| \log((1+\alpha)|p|)\\
+             &\geq 2(|H_p|+|p|)\log(|H_p|+|p|) \geq 2\log(|H_p|+|p|) \pth{\sum_{i=1}^k S_i + \sum_{i=1}^l R_i}\\
+             &\geq \frac{4}{3}\sum_{i=1}^k S_i \log S_i + \sum_{i=1}^l R_i \log R_i 
+\end{align*}
+
+In this case $\log(|H_p|+|p|)\leq \log(2|p|) \leq 1+\log(|p|)$ (where $\log$ is base $2$).
+Therefore, assuming $|p|>1$, we have 
+\[
+(|H_p|+|p|)\log(|H_p|+|p|)\leq (1+c)|p|\log((1+c)|p|) %(|H_p|+|p|)(1+\log |p|) \leq 2(1+c)|p|\log |p|
+\]
+which when combined with the inequality from \Lem{adjacent} gives, 
+\[
+ 4(1+c)|p|\log |p| \geq 2(|H_p|+|p|)\log(|H_p|+|p|)\geq \pth{|H_f| + \frac{4}{3}\sum_{i=1}^k S_i + \sum_{i=1}^l R_i}\log(|H_p|+|p|).
+\]
+This easily implies the desired inequality as $\log(H_p+p)$ is larger than any other $\log$ term in the inequality.
+
+\paragraph{Case 2, $|p|< |H_p|/\alpha$:\\}
+
+\Ben{Was not able to handle this case.}
+
+%  First, lets assume we can show $\sum_{i=1}^k S_i + \sum_{i=1}^l R_i \leq \frac{3}{4}|H_p|$.  
+%  In this case, using the last part of \Clm{sizes}, we can prove the desired inequality as follows.
+%  
+%  \begin{align*}
+%   \frac{4}{3}\sum_{i=1}^k S_i \log S_i + \sum_{i=1}^l R_i \log R_i  &\leq  \frac{4}{3} \pth{\sum_{i=1}^k S_i \log S_i + \sum_{i=1}^l R_i \log R_i} \\
+%   &\leq  \frac{4}{3} \log |H_p| \pth{\sum_{i=1}^k S_i + \sum_{i=1}^l R_i}  \leq  |H_p| \log |H_p|
+%  \end{align*}
+%  
+%  Finally, to complete the proof we must prove our assumption that $\sum_{i=1}^k S_i + \sum_{i=1}^l R_i \leq \frac{3}{4}|H_p|$.
+%  By \Lem{adjacent} and \Clm{sizes} we have, 
+%  \[
+%  \sum_{i=1}^k S_i + \sum_{i=1}^l R_i 
+%  \leq |H_p|+|p|-|H_f| 
+%  \leq (1+1/\alpha-1/3)|H_p| = (2/3+1/\alpha)|H_p|
+%  \]
+%  
+%  Therefore $\sum_{i=1}^k S_i + \sum_{i=1}^l R_i \leq \frac{3}{4}|H_p|$ if $(2/3 + 1/\alpha)\leq 3/4$, which rearranged gives $\alpha \geq 12$
+\end{proof}
+
+\remove{
+% \subsubsection{Paying it down the line}
+% 
+% \begin{lemma}
+% \label{lem:excess}
+%  Let $p$ be a tall path, and let $S_1, \dots S_k$ and $R_1, \dots, R_l$ be the sizes of the heaps at the roots of the short and tall children of $p$, respectively, in the shrub.  
+%  Let $H_f$ be the size of the heap at the leaf of the path $p$, and let $X = \sum_{i=0}^k H_i \log H_i$.  
+%  Then,  $|H_p| \log |H_p|   \geq  H_f\log H_f + \sum_{i=0}^k S_i \log S_i + \sum_{i=0}^l R_i \log R_i  -  c|p|\log |p|$, for some sufficiently large constant $c$
+%  \end{lemma}
+% \begin{proof}
+% By \Lem{adjacent} we know that $|H_p|+|p| \geq  H_f + \sum_{i=0}^k S_i + \sum_{i=0}^l R_i$.  
+% %Now as $p$ is a tall path we have $|p|\geq \sqrt{|H_p|}/100$, and so, $(|H_p|+|p|)\log(|H_p|+|p|)$
+% 
+% We  break the analysis into two case based on the relative size of $|p|$ and $|H_p|$.  In the following $c$ is some yet to be determined constant greater than $1$.
+% 
+% \paragraph{Case 1, $|p|\geq |H_p|/c$:}  
+% In this case $\log(|H_p|+|p|)\leq \log(2|p|) \leq 1+\log(|p|)$ (where $\log$ is base $2$).
+% Therefore, assuming $|p|>1$, we have 
+% \[
+% (|H_p|+|p|)\log(H_p+p)\leq (|H_p|+|p|)(1+\log |p|) \leq (2|p|)(2\log |p|)\leq 4|p|\log |p|
+% \]
+% which when combined with the inequality from \Lem{adjacent} gives, 
+% \[
+%  4|p|\log |p| \geq (|H_p|+|p|)\log(|H_p|+|p|)\geq \pth{H_f + \sum_{i=0}^k S_i + \sum_{i=0}^l R_i}\log(|H_p|+|p|).
+% \]
+% This easily implies the claim as $\log(H_p+p)$ is larger than any other $\log$ term in the inequality.
+% 
+% \paragraph{Case 2, $|p|< |H_p|/c$:}
+% \end{proof}
+}
+
+Define the crust of the shrub to be the first (i.e. closest to the root) layer of short paths in the shrub.  Apply \Lem{shortCost} to each of the short paths in the crust and then 
+say by \Lem{excess} and since the root heap goes to zero we are done.
+
+\ignore{
+Naturally, this geometric series behavior can be extended to more than two adjacent short paths, thus motivating 
+the following definition.
+
+\begin{definition}
+ Let $p$ be a path in a path decomposition $P(T)$, with corresponding shrub $\pathTree$.
+ Call a short path $q\in P(T)$ short reachable from $p$ if the path from $p$ to $q$ in $\pathTree$ 
+ contains only vertices which correspond to short paths.  Call the subtree of $\pathTree$ consisting 
+ of all short reachable paths from $p$ the \emph{short shrub} of $p$, and denote it $\mathcal{L}(p)$.
+\end{definition}
+
+In the following subsection we show how to exploit this geometric series behavior of short shrubs.
+
+We say two paths in $P(T)$ are adjacent if their corresponding vertices are adjacent in the shrub.  Similarly, for 
+two adjacent paths in $P(T)$ we can define a parent/child relationship based on this relationship in the shrub. 
+}
+
+\remove{
+\Ben{Maybe move the remainder of this subsection to 5.3 since tall and short paths and not used until then}
+
+\begin{definition}
+ Let T be a binary tree and let $\chi(T)$ and $P(T)$ be a valid coloring and path decomposition of $T$, respectively.  
+ For $p\in P(T)$, we say that $p$ is \emph{short} if $|p| < \sqrt{|H_{r_p}|}/100$, and we say it is \emph{tall} otherwise.
+\end{definition}
+
+\begin{observation}
+\label{obs:decrease}
+ Let $v$ be a vertex in $T$ and let $w$ be its parent.  Then $|H_w| \geq |H_v| -1$, as $L(T_v) \subseteq L(T_w)$ and the path from 
+ $w$ to the root has one less vertex than the path from $v$ to the root.
+
+ In particular, we have the following more general property.
+ Let $v$ and $u$ be any two vertices in the same root to leaf path of $T$, such that $u$ is lower than $v$.  
+ Then $|H_v| \leq |H_u| + d(u,v)$. 
+\end{observation}
+
+\begin{lemma}
+\label{lem:pathBounds}
+ Let T be a binary tree and let $\chi(T)$ and $P(T)$ be a valid coloring and path decomposition of $T$, respectively.
+ Then, 
+ \begin{compactenum}
+  \item for any tall path $p\in P(T)$, $\sum_{v\in p} \log |H_v| = O(|p| \log |p|)$.
+  \item for any short path $p\in P(T)$, $\sum_{v\in p} \log |H_v| = O(|p| \log |H_{r_p}|)$.
+ \end{compactenum}
+\end{lemma}
+\begin{proof}
+ By $\Obs{decrease}$ we know that for any vertex $v\in p$, $|H_v|\leq |H_{r_p}| + |p|$ (as $r_p$ lies below $v$ on $p$).
+ As such, 
+ \[
+ \sum_{v\in p} \log |H_v| \leq \sum_{v\in p} \log(|H_{r_p}| + |p|)  = |p| \log (|H_{r_p}| + |p|).
+ \]
+ If $p$ is a tall path then $|p| \log (|H_{r_p}| + |p|) = O(|p| \log |p|)$, and if $p$ is short then 
+ $|p| \log (|H_{r_p}| + |p|) = O(|p| \log |H_{r_p}| )$.
+\end{proof}
+
+The following lemma can be thought of as a generalization of $\Obs{decrease}$.
+
+\begin{lemma}
+\label{lem:adjacent}
+ Let $T$ be a binary tree and let $\chi(T)$ and $P(T)$ be a valid coloring and path decomposition of $T$, respectively.
+ Let $p$ be any path in $P(T)$ and let $\{q_1, \dots q_k\}$ be the set of all adjacent child paths of $p$.
+ For $1\leq i\leq k$ let $r_i = r_{q_i}$ and let $H_i = H_{r_i}$.
+ Also, let $H_0$ denote the heap at the leaf of $p$.
+
+ Then $\sum_{i=0}^k |H_i| \leq |H_p|+|p|$, where $H_p = H_{r_p}$.
+\end{lemma}
+\begin{proof} 
+ First observe that for all $i$, $r_{i}$ lies above $r_p$ on some root to leaf path. 
+ Let $H_i'$ be the subset of $H_i$ that lies below $r_p$.  Observe that $H_i'\subseteq H_p$, as $L(T_{r_i})\subseteq L(T_{r_p})$.
+
+ Now for any $i\neq j$, $H_i \cap H_j = \emptyset$, as $r_{i}$ and $r_{j}$ are not on the same root to leaf path (i.e. $L(T_{r_i}) \cap L(T_{r_j}) = \emptyset$).
+ In particular, $(H_i\setminus H_i') \cap (H_j \setminus H_j') = \emptyset$, and so $\sum_{i=0}^k |H_i\setminus H_i'| \leq |p|$ 
+ as $H_i\setminus H_i'$ is precisely the subset of $H_i$ that lies on $p$.
+ Moreover, for any $i\neq j$, $ H_i' \cap H_j' = \emptyset$, and so since $H_i'\subseteq H_p$ we have $\sum_{i=0}^k |H_i'|\leq |H_p|$.
+ Putting these together gives
+ \[
+  \sum_{i=0}^k |H_i| = \sum_{i=0}^k |H_i'| + \sum_{i=0}^k |H_i\setminus H_i'| \leq |H_p|+|p|.
+ \]
+\end{proof}
+
+% \Ben{There is a fundamental flaw in the following lemma and corollary.  
+% As such, the remainder of this subsection will likely be deleted.}
+% 
+% Using the notation defined in the above lemma, we have the following crucial lemma, which we will use to 
+% argue we can charge that cost of all paths in $P(T)$ to the cost of the tall paths.
+% 
+% \begin{lemma}
+% Let $p$ be a tall path in $P(T)$ and let $\{q_1, \dots q_k\}$ be any set of short paths in $P(T)$ such that 
+% $\sum_{i=1}^k |H_i| = O(H_p)$.  Then $\sum_{i=1}^k \sum_{v\in q_i} \log |H_v| = O(|p|\log |p|)$
+% \end{lemma}
+% \begin{proof}
+% By \Lem{pathBounds}, for all $1\leq i\leq k$, 
+%  $\sum_{v\in q_i} \log |H_v| = O(|q_i|\log |H_i|) = O(|H_i|\log |H_i|)$, as $q_i$ is a short path.
+%  By assumption $\sum_{i=1}^k |H_i| = O(|H_p|)$, 
+%  As $x\log x$ is a convex function, Jensen's inequality then implies 
+%  $\sum_{i=1}^{k} |H_i|\log |H_i| = O(|H_p| \log |H_p|) = O(|p|\log |p|)$, since $p$ is a tall path.
+% \end{proof}
+% 
+% 
+% \begin{corollary}
+% \label{cor:tallLight}
+%  Let $p$ be a tall path in $P(T)$ and let $\{q_1, \dots q_k\}$ be the set of all short paths adjacent to $p$.
+%  Then $\sum_{i=1}^k \sum_{v\in q_i} \log |H_v| = O(|p|\log |p|)$.
+% \end{corollary}
+% \begin{proof}
+%  Let $H_i$ be as defined in \Lem{adjacent}.  By \Lem{pathBounds}, for all $1\leq i\leq k$, 
+%  $\sum_{v\in q_i} \log |H_v| = O(|q_i|\log |H_i|) = O(|H_i|\log |H_i|)$, as $q_i$ is a short path.
+%  Now by \Lem{adjacent}, $\sum_{i=1}^k |H_i| \leq |H_p|+|p| = O(|p|)$, as $p$ is a tall path. 
+%  Therefore, as $x\log x$ is a convex function, Jensen's inequality implies the claim.
+%  
+% \end{proof}
+
+}
+
+\subsection{The shrub of unsupported paths} \label{sec:short}
+
+It suffices to focus on unsupported short paths. \Lem{short} can be reduced to the following.
+
+\begin{lemma} \label{lem:support} $\sum_{p \in \cL \setminus \cL'} \sum_{v \in p} \log |H_v| = O(\sum_{p \in \cL'} |H_p| \log |H_p|)$.
+\end{lemma}
+
+We now construct the shrub of unsupported short paths. Consider $p \in \cL \setminus \cL'$,
+and traverse the chain of ancestors from $p$. Eventually, we must reach another short path $q$.
+(If not, we have reached the root $r$ of $\pmax(T)$. Hence, $p$ is supported.)
+Insert edge from $p$ to $q$, so $q$ is the parent of $p$ in $\cU$. This construction leads
+to the shrub forest of $\cL \setminus \cL'$, where every unsupported short path is an internal node, and all
+roots are supported short paths.
+
+We list out some important claims about this shrub forest.
+We focus on $\cU$, a connected component (shrub) in the shrub forest and denote the 
+the root by $r(\cU)$.
+
+\begin{claim} \label{clm:loss2} Consider $p \in \cL \setminus \cL'$ that is in $\cU$. Let $Q$
+be the set of tall ancestors of $p$ that are descendants of $r(\cU)$. Then $\sum_{q \in Q} |q \cap H_p| \leq |H_p|/50$.
+\end{claim}
+
+\begin{proof} Suppose not, for contradictions sake. As we traverse the tall ancestors of $p$,
+we encounter various vertices of $H_p$ in the ancestors $q$. Consider the first $|H_p|/100$
+such vertices of $p$ encountered. At each such $v$, there are at least $|H_p|/100$
+vertices \emph{not yet} encountered, so $|H_v| > |H_p|/100$. Each of these vertices $v$
+is supported, implying that $p$ is a supported short path.
+\end{proof}
+
+\Lem{support} is a direct consequence of the following,
+which is the really the main lemma.
+
+%\begin{lemma} \label{lem:cu} Let $\cU$ denote a connected component (shrub) in the shrub forest of $\cL \setminus \cL'$ and let $r$
+%be the root of $\cU$. Then $\sum_{p \in \cU} \sum_{v \in p} \log |H_v| = O(|H_r| \log |H_r|)$.
+%\end{lemma}
+%
+%We proceed with the setup to prove \Lem{cu}.
+%We focus our attention on some shrub $\cU$.
+%We prove a technical lemma about $\cU$, from which \Lem{cu} follows directly.
+
+\section{Ben's old stuff}
+
+\begin{proof} (part (ii) of \Lem{cu-root}) Let level $\ell$ of $\cU$ denote all nodes at distance $\ell$ from the root $r$.
+These set of these nodes is designated as $\cU_\ell$. We set $\sigma_\ell = \sum_{p \in \cU_\ell} |p|$
+and $\sigma_{\leq \ell} = \sum_{k \leq \ell} \sigma_k$.
+By \Lem{geometric}, for any $p \in \cU_\ell$, $|H_p| \leq (2/3)^\ell |H_r|$.
+
+Since all nodes in $\cU$ represent short paths, $\forall p \in \cU$, $|p| < \sqrt{|H_p|}/100$.
+So $\sigma_\ell < \sum_{p \in \cU_\ell} \sqrt{|H_p|}/100$.
+
+\end{proof}
+
+To prove this, we need to understand how $H_p$ is constructed. The path $p$ has some children
+$q_1, q_2, \ldots, q_k$ in $\pmax(T)$. Furthermore, as a leaf path in $p$, some leaf
+$\ell$ of $T$ in one of the ends of $p$. The heap $H_p$ is essentially the
+union of $H_\ell$ and $H_{q_1}, H_{q_2}, \ldots, H_{q_k}$, with the vertices in $p$
+removed. Note that some of the $q_i$'s are the corresponding $h(q')$ vertices,
+for a child $q'$ in $\cU$.
+
+Let $\mathcal{L} = \mathcal{L}(r)$ be a short shrub of some short path $r$ in a maximum path decomposition $\mathcal{P}(T)$.
+The intuitive idea is that due to the geometric decreasing behavior of short shrubs described in the previous subsection, 
+one can show that for the entire short shrub our algorithm pays roughly the cost due to the root path $r$.  This argument 
+is hard to make directly, so instead we define an alternative quantity for which we can make the argument.
+Specifically, we now define a weight function $w(v)$ for all $v\in \mathcal{L}$, and show that $w(v)=\Theta(|H_v|)$.
+
+For a path $v\in \mathcal{L}$, let $H_v$ denote the heap at the root of $v$, and let $H_{v_f}$ denote the heap 
+at the leaf of $v$.  Let $S_l(v)$ (resp. $S_h(v)$) be the subset of short (resp. tall) paths adjacent to $v$ in $\pathTreeA(T)$.
+
+We want to understand how the heap size changes as we move around the paths in $\mathcal{L}$.  Specifically, there are 
+three places the vertices in the heap along a short path can come from; adjacent short paths, adjacent tall paths, and the initial heap at the leaf of the path.  
+Given this observation we now define a weight function which upper bounds the size of the heap at the root of any path in $\mathcal{L}$.
+
+First, for all $v\in \mathcal{L}$ we define the local cost function $c(v) = H_{v_f}+\sum_{i\in S_h(v)} H_i$, this is the addition to the heap at $v$ which 
+was not already charged to descendant short paths.  We now define a weighted version of $\mathcal{L}$ based on these costs.  
+To do so, we first augment $\mathcal{L}$ by attaching one 
+leaf to each vertex in $\mathcal{L}$.  Now for each newly created leaf (which are the only leaves now) assign the cost 
+$c(v)$ defined above, where $v$ is that leaf's corresponding parent.  
+Now we can define a weight function $w(u)$ over any node $u$ in the tree, which is defined to be the total cost of all leaves in its subtree.   
+To simplify notation, we now use $\mathcal{L}$ to refer to this new weighted tree (rather than its unweighted counterpart).
+
+As the weight $w(p)$ includes the sizes of the heaps of all contributing tall paths and leaves in the subtree rooted at $p$, we have the following.
+
+\begin{claim}
+For any $p\in \mathcal{L}$, $w(p)\geq |H_p|$.  Moreover, for any $v\in p$, $w(p)\geq |H_v|$.
+\end{claim}
+
+%Let $W= \sum_{v\in \mathcal{L}} w(v) \geq \sum_{v\in \mathcal{L}} |H_v|$.  
+%Suppose we could show that conversely $W=O(H_r)$.  Then we would have 
+%$\sum_{v\in \mathcal{L}} H_v = O(H_r)$, which was our original goal.  
+%So proving $W=O(H_r)$ is now our new goal.
+
+We now would like to show conversely that $w(v)= O(|H_v|)$.  
+To do so we first introduce some notation.  
+Let $C\subseteq \mathcal{P}$ be the subset of paths that are leaves in $\mathcal{L}$.
+For any path $v$, let $\mathcal{L}_v$ (resp. $C_v$) be the subset of paths in $\mathcal{L}$ (resp $C$) 
+that are in the subtree of $\mathcal{L}$ which is rooted at $v$.
+
+First, observe that $|H_v|\geq w(v)-\sum_{u\in \mathcal{L}_v} |u|$.  
+This holds by the same argument as in \Lem{adjacent}, namely, $w(v)$ by definition 
+counts all the contributing tall path heaps and initial leaf heaps that contribute 
+to the heap at $v$ and $\sum_{u\in \mathcal{L}_v} |u|$ is an upper bound on the number 
+of vertices that could have been removed from one of these heaps before reaching the 
+root of $v$ (since a vertex in a heap can only get removed if it appears on one of 
+the paths).
+
+As all paths in $\mathcal{L}$ are short, we thus have the following.
+
+\begin{claim}
+$|H_v|\geq w(v)-\sum_{u\in \mathcal{L}_v} \sqrt{|H_u|}/100 \geq w(v)-\sum_{u\in \mathcal{L}_v} \sqrt{w(u)}/100$
+\end{claim}
+
+By the above claim, if we can prove $\sum_{u\in \mathcal{L}_v} \sqrt{w(u)}/100 \leq w(v)/c$ for some constant $c>1$, 
+then we will have proven that $|H_v| = \Theta(w(v))$.
+
+Recall that \Lem{geometric} told us that that $|H_{u}|\leq (3/5) |H_{u'}|$, if $u'$ is the parent $u$ in $\mathcal{L}$. 
+The following lemma shows that if we had a similar statement in terms of the weights rather than the heap sizes, 
+then we get the desired bound on the sum.
+
+\begin{lemma}
+\label{lem:distribution}
+Let $v$ be a path in $\mathcal{L}$.  Suppose that for any two paths $x,y\in \mathcal{L}_v$ 
+such that $y$ is the parent of $x$, $w(x)\leq (4/5) w(y)$.
+Then $\sum_{u\in \mathcal{P}_v} \sqrt{w(u)}/100 \leq \sum_{c\in C_v} \sqrt{w(c)}/10$
+\end{lemma}
+
+\begin{proof}
+The idea is to recursively redistribute the contribution to the sum of each internal vertex to its children, 
+until all the weight sits on the leaves.  Specifically, let $v$ be some path in $\mathcal{L}$.  Its 
+contribution to the sum is $\sqrt{w(v)}/100$, which we will call the value at $v$.  We redistribute this 
+value to its children, in proportion by the weight of each child.  For example, if $v$ has a child $x$ then 
+the fraction it gets assigned is $w(x)/w(v)$ (recall that by definition $w(v)$ is the sum of the weights of 
+its children and so the total value will get reassigned to the children).
+
+Suppose we keep redistributing the value of $v$ until it reaches the leaves of its subtree in $\mathcal{L}$.  
+Let's see how much gets assigned to a given leaf.  Specifically, let $u_0, u_1, \ldots, u_k$ 
+be a path in $\mathcal{L}$ (of paths from the set $\mathcal{P}_v$), such that $v=u_0$, $u_{i+1}$ 
+is child of $u_i$ for all $i$, and $u_k$ is a leaf of $\mathcal{L}$.
+
+Then the total value from $u_0$ that ends up at $u_k$ is 
+\[
+\frac{\sqrt{w(u_0)}}{100} \pth{ \frac{w(u_1)}{w(u_0)} \frac{w(u_2)}{w(u_1)} \cdots \frac{w(u_k)}{w(u_{k-1})} }
+= \frac{\sqrt{w(u_0)}}{100} \pth{ \frac{w(u_k)}{w(u_0)} } = \frac{w(u_k)}{100\sqrt{w(u_0)}}
+\]
+
+For a given leaf in $\mathcal{L}$, we can calculate the total value reassigned to it from it its ancestors.
+Specifically, let $u_0, u_1, u_2, \dots, u_m$ be a root to leaf path in $\mathcal{L}$ 
+(i.e. $u_0$ is the root $r$ of $\mathcal{L}$).
+
+By assumption from the lemma statement we have $w(u_{i+1})\leq (4/5) w(u_i)$.  Therefore by the above calculation, 
+the total reassigned to $u_m$ is
+
+\[
+ \sum_{i=0}^m \frac{w(u_m)}{100\sqrt{w(u_i)}} \leq \frac{w(u_m)}{100} \sum_{i=0}^m 1 \Big/ \sqrt{\pth{\frac{5}{4}}^iw(u_m)} 
+ \leq \frac{\sqrt{w(u_m)}}{100} \sum_{i=0}^{\infty} \pth{\frac{4}{5}}^{i/2} \leq \frac{\sqrt{w(u_m)}}{10}
+\]
+
+The claim then follows by summing this quantity over the children of $\mathcal{L}$.
+\end{proof}
+
+By combining the above lemma and claim we have the following.
+
+\begin{claim}
+For any path $v\in \mathcal{L}$ if it holds that 
+$w(x)\leq (4/5) w(y)$ for any two paths $x,y\in \mathcal{L}_v$ such that $y$ is the parent of $x$, then
+
+\[
+|H_v|\geq w(v)-\sum_{u\in \mathcal{L}_v} \sqrt{w(u)}/100 \geq w(v)- \sum_{c\in C_v} \sqrt{w(c)}/10 \geq \frac{4}{5} w(v)
+\]
+\end{claim}
+
+\begin{lemma}
+Let $\mathcal{L}$ be the short shrub of some path $r$.  
+Then for any $v\in \mathcal{L}$, we have $|H_v|\geq \frac{4}{5} w(v)$.
+\end{lemma}
+% \begin{proof}
+% We will actually prove the stronger statement that for any $v\in \mathcal{L}$, 
+% we have $|H_v|\geq \frac{4}{5} w(v)$ and $w(v)\leq (4/5) w(p)$ where $p$ is the parent 
+% of $v$ in $\mathcal{L}$.
+% 
+% Suppose otherwise, and let $v$ be the lowest path in $\mathcal{L}$ such that this fails 
+% (i.e. the statement holds for all descendants of $v$ in $\mathcal{L}$).  Then for any 
+% two paths $x,y\in \mathcal{L}_v$ such that $y$ is the parent of $x$, 
+% we have that $w(x)\leq (4/5) w(y)$ and so by \Lem{distribution}, $|H_v|\geq \frac{4}{5} w(v)$.
+% Combining this fact with \Lem{geometric} gives,
+% \[
+% w(v) \leq \frac{5}{4}|H_v| \leq \frac{5}{4} \frac{3}{5} |H_p| \leq \frac{3}{4} w(p).
+% \]
+% 
+% \end{proof}
+%
+\begin{proof}
+By the above claim, we know that $|H_v|\geq \frac{4}{5} w(v)$ if it holds that 
+$w(x)\leq (4/5) w(y)$ for any two paths $x,y\in \mathcal{L}_v$ such that $y$ is the parent of $x$.
+So let $v$ be the lowest path in $\mathcal{L}$ such that this fractional decrease fails to hold, i.e. $w(c)> (4/5) w(v)$ 
+where $c$ is a child of $v$ in $\mathcal{L}$, and conversely all vertices in $\mathcal{L}_c$ 
+the fractional decrease holds.  As the fractional decrease holds for a paths in $\mathcal{L}_c$, 
+we have $|H_c|\geq \frac{4}{5} w(c)$.  Therefore,
+\[
+w(c) \leq \frac{5}{4}|H_c| \leq \frac{5}{4} \frac{3}{5} |H_v| \leq \frac{3}{4} w(v).
+\]
+However, this is a contradiction with the assumption $w(c)> (4/5) w(v)$ , and hence the 
+fractional decrease always holds implying the claim always holds.
+\end{proof}
+
+\begin{corollary}
+\label{cor:equivalent}
+ Let $\mathcal{L}$ be a short shrub of a short path $r$.  Then for any $v\in \mathcal{L}$, $|H_v| \leq w(v) \leq \frac{5}{4} |H_v|$.
+\end{corollary}
+
+Now we are finally ready to make use of the properties we have proven about the weight function we introduced.
+
+\Ben{The constant in the lemmas below, but it is no more than twice that.}
+
+\begin{lemma}
+\label{lem:shortCost}
+ Let $\mathcal{L}$ be a short shrub of a short path $r$.  Let $V$ be the set of tall paths from the full shrub which are adjacent to paths in $\mathcal{L}$.  
+  Then $\sum_{p\in \mathcal{L}}\sum_{v\in p} \log |H_v| + \sum_{q\in V} |H_q|\log |H_q| \leq \frac{4}{3} |H_r|\log |H_r|$
+\end{lemma}
+\begin{proof}
+\Ben{
+The following is a proof for $\frac{3}{2}+\eps$, rather than $\frac{4}{3}$.  The discrepancy is due to a corrected calculation error.  
+4/3 can be achieved by tightening constants above, and is left in the lemma statement as it is used later.
+}
+
+Recall that in a proper coloring of a binary tree $T$, the color of any vertex must be the color of one of its descendants.  
+Therefore, any vertex in the short path shrub must appear in the heap of either the root of an adjacent 
+path from the full shrub or from the heap of a leaf of a path in $\mathcal{L}$.  
+Also note $|H_r|$ precisely counts sum of the number of vertices from these heaps that do not appear in $\mathcal{L}$.
+As there are no adjacent short paths to a short shrub, the total number of vertices over all paths in $\mathcal{L}$ is at most 
+$m = \sum_{p\in \mathcal{L}} |H_{p_f}| +\sum_{q\in V} |H_q| - |H_r| = w(r)-|H_r|$, where $H_{p_f}$ is the heap at the leaf of $p$.
+
+Using this observation and \Cor{equivalent} we then have, 
+\begin{align*}
+ \sum_{p\in \mathcal{L}}\sum_{v\in p} \log |H_v| + \sum_{q\in V} |H_q|\log |H_q|
+ &\leq \sum_{p\in \mathcal{L}}\sum_{v\in p} \log(w(r)) + \sum_{q\in V} |H_q|\log(w(r))\\
+ &\leq \log(w(r))\pth{ w(r)-|H_r| } + \log(w(r))w(r)\\ 
+ &= (2w(r)-|H_r|)\log(w(r)) \leq \pth{\frac{5}{2}|H_r|-|H_r|}\log\pth{\frac{5}{4}|H_r|}\\
+ &= \frac{3}{2}|H_r|\log\pth{\frac{5}{4}|H_r|}
+\end{align*}
+\end{proof}
 
@@
 
-We define a ``cut" operation on manifolds that cuts along a contour to create
-a manifold with a new boundary. Intuitively, any cut creates two disjoint boundaries, one
-above and one below.
-
-\medskip
-\fbox{
-\begin{minipage}{0.9\textwidth}
-{\bf $\cut(\MM,\phi)$}
-
-\smallskip
-\begin{asparaenum}
-%	\item Insert the vertices of $\phi$ in $\MM$.
-	\item Take an arbitrary edge $e$ whose interior intersect $\phi$.
-	\item Take a point $x$ (resp. $y$) on $e$ at distance $\eps$ above (resp. below) $\phi$.
-	\item Let $\phi_x$ be the contour through $x$, and analogously define $\phi_y$. Insert a vertex
-	at each intersection point of these contours with edges of the manifold. Triangulate all new faces.
-	\item Delete the vertices (and incident edges, faces, etc.) of $\phi$ from the manifold.
-\end{asparaenum}
-\end{minipage}}
-
-\medskip
-This results in a new manifold which is boundary critical if the input manifold was. There are two new disjoint
-boundary faces added. The manifold may be disconnected. Abusing notation, we use $\cut(\MM,\Psi)$
-for a set $\Psi$ of contours to mean the manifold obtained by cutting along all contours in $\Psi$.
-
-We now describe a procedure that converts the Reeb Graph of $\cut(\MM,\phi)$ to the Reeb Graph of $\MM$.
-
-\medskip
-\fbox{
-\begin{minipage}{0.9\textwidth}
-{\bf $\surgery(\MM,\phi)$}
-
-\smallskip
-\begin{asparaenum}
-	\item Let $\MM' = \cut(\MM,\phi)$.
-	\item Construct $\reeb(\MM')$ and let $B, C$ be the vertices corresponding to the new boundaries created
-	in $\MM'$. (One is a minima and the other is maxima.)
-	\item Since $B, C$ are leaves, they each have unique neighbors $B'$ and $C'$, respectively. Insert
-	edge $(B',C')$ and delete $B,C$ to obtain a new graph.
-\end{asparaenum}
-\end{minipage}}
-
-\medskip
-\begin{lemma} \label{lem:surgery} For any regular contour $\phi$, the output of $\surgery(\MM,\phi)$ is $\reeb(\MM)$.
-\end{lemma}
-
-\begin{proof} The only difference between $\reeb(\MM)$ and $\reeb(\cut(\MM,\phi))$ is in the equivalence
-class corresponding to $\phi$. This class is split into two classes in $\cut(\MM,\phi)$, which are joined
-back by $\surgery(\MM,\phi)$.
-\end{proof}
-
-\section{Raining to partition $\MM$} \label{sec:rain}
-
-In this section, we assume that $\MM$ is homeomorphic to $\SSS^2$ with boundaries.
-
-\begin{definition} \label{def:dom} A manifold is \emph{extremum dominant} if there exists
-a maximum (resp. minimum) $x$ such that every non-maximal (resp. non-minimal) \emph{vertex} in the manifold has an increasing (resp. decreasing) path to $x$.
-\end{definition}
-
-We will describe a linear time algorithm that partitions $\MM$ into a set of extremum dominant manifolds.
-The first part of this requires a ``raining" procedure. Start at some point $x \in \MM$ and imagine rain at $x$.
-The water will flow downwards along descending paths and ``wet" all the points encountered.
-
-\begin{definition} \label{def:wet} The set of points $y$ such that there is a descending path from $x$ to $y$
-is denoted by $\wet(x,\MM)$. A point $z$ is on the \emph{interface} of $\wet(x,\MM)$ if every neighborhood of $z$
-has non-trivial intersection with $\wet(x,\MM)$.
-\end{definition}
-
-Note that $\wet(x,\MM)$ (and its interface) can easily be computed in linear time in the size of the wet sub-manifold, using a descending BFS from $x$.
-
-\begin{claim} \label{clm:inter} For any $x$, the interface of $\wet(x,\MM)$ is a set of disjoint, non-regular contours.
-Furthermore, each such contour contains a merge vertex.
-\end{claim}
-
-\begin{proof}
- First observe that any point which in the interface must be wet.  Second, for $y \in \wet(x,\MM)$, all the points in any contour containing $y$ are also in $\wet(x,\MM)$, 
- as one can follow the descending path to $y$ and then walk along the contour.
-
- So let $y$ be a point in the interface.
- For $\eps$ sufficiently small, the points in any $\eps$ neighborhood of $y$ that lie below $y$ must have a descending path from $y$ and hence must be wet.
- Therefore in order for $y$ to have non-trivial intersection with $\wet(x,\MM)$ for any arbitrarily small $\eps$ there must be a dry point which lies 
- strictly above $y$.  In particular, this implies there must be an entire dry contour which lies just above $y$.
-
- However, since $y$ is in the interface, it is also wet.  Therefore for any contour $\phi$ which contains $y$, for any arbitrarily small $\eps$ there is 
- a wet point which lies above $\phi$ and has a descending path of length at most $\eps$ to $\phi$.
- This implies there is some contour which lies just 
- above $y$ which is wet.  So any contour $\phi$ containing $y$ has two contours which lie just above it, one which is dry, and one which is wet.  
- This can only happen if $\phi$ contains a merge vertex, because otherwise all contours which lie just above would be in the same L-homotopy equivalence class and so by 
- \Clm{mono}, if one is wet, all below it are wet.
-
- By similar logic one can argue that of the two contours through a given merge vertex at most one can contain interface points other than the merge vertex, 
- and that if one of the contours does contain such interface points, then the entire contour consists of interface points.
-\end{proof}
-
-% \begin{proof} If $y \in \wet(x,\MM)$, but not in the interface, then any contour containing $y$ is also in $\wet(x,\MM)$.
-% So consider a point $y$ in the interface. We argue that all points in a contour $\phi$ containing $y$
-% are also in the interface. If such a contour is regular, we can find a contour just above it that
-% is also wet. 
-% 
-% If two contours intersect, the intersection point $y$ is critical. We can argue that the only
-% way to reach $y$ from a descending path is reach one of the contours and then follow it to get to $y$.
-% But that would imply the existence of another critical point on this contour, violating the Morse condition.
-% 
-% Let us argue that $y$ is a merge vertex. Consider the descending path from $x$ to $y$. The final edge
-% that reaches $y$ is obviously all wet, and is an increasing edge from $y$. Consider an opposite increasing
-% edge. Take two point in these edges that are infinitesimally higher than $y$ and show that the contours
-% through them as disjoint. So $y$ is a merge.
-% \end{proof}
-
-We will define a \lift{} operation on the interface contours. Consider such a contour $\phi$ containing
-a merge vertex $y$. Take any dry increasing edge incident to $y$, and pick the point $z$ on this edge at height
-$f(y) + \eps$. Let $\lift(y) = \cont(z)$.
-
-\begin{claim} \label{clm:cut-int} Let $\phi$ be an interface contour. Then $\cut(\MM,\lift(\phi))$
-results in two disjoint manifolds, one consisting entirely of dry points.
-\end{claim}
-
-\begin{proof} Note that $\lift(\phi)$ is a closed, dry curve. By the Jordan Curve Theorem,
-$\cut(\MM,\lift(\phi))$ results in two disjoint manifolds. Let $\NN$ be the manifold containing
-the point $x$ that we called $\wet(x,\MM)$ from, and let $\NN'$ be the other manifold. 
-If there exists a wet point $z$ in $\NN'$, then there exists a descending path from $x$ to $z$. 
-As $x$ is in $\NN$ and $z$ is in $\NN'$, again by the Jordan Curve Theorem this path must intersect $\phi$, 
-contradicting the fact that $\phi$ is dry.
-\end{proof}
-
-We now define the main partitioning procedure that cuts up a manifold $\NN$ homeomorphic to $\SSS^2$ with boundaries into a series of extremum
-dominant manifolds. It takes as input the manifold and a maximum $x$. To initialize,
-we begin with $\NN$ set to $\MM$ and $x$ as an arbitrary maximum.
-
-\medskip
-\fbox{
-\begin{minipage}{0.9\textwidth}
-{\bf $\rain(x,\NN)$}
-
-\smallskip
-\begin{compactenum}
-	\item Compute the interface of $\wet(x,\NN)$. 
-	\item If the interface is empty, simply output $\NN$. Otherwise, denote the contours by $\phi_1, \phi_2, \ldots, \phi_k$, and set $\phi'_i = \lift(\phi_i)$.
-	\item Initialize $\NN_1 = \NN$.
-	\item For $i$ from $1$ to $k$: 
-	\begin{compactenum}
-		\item Call $\cut(\phi'_i,\NN_1)$ to get the dry manifold $\LL_i$ and wet manifold $\NN_{i+1}$.
-		\item Let the new boundary of $\LL_i$ be $B_i$. Invert $\LL_i$ so that $B_i$ is a maximum. Recursively
-		call $\rain(B_i,\LL_i)$.
-	\end{compactenum}
-	\item Output $\NN_{k+1}$.
-\end{compactenum}
-\end{minipage}}
-
-\medskip
-For convenience, denote the total output of $\rain(x,\MM)$ by $\MM_1, \MM_2, \ldots, \MM_r$.
-
-\begin{lemma} \label{lem:rain-1} Each output manifold $\MM_i$ is extremum dominant.
-\end{lemma}
-
-\begin{proof} Consider a call to $\rain(x,\NN)$. If the interface is empty, then all of $\NN$
-is in $\wet(x,\NN)$, so $\NN$ is trivially extremum dominant. So suppose the interface
-is non-empty and consists of $\phi_1, \phi_2, \ldots, \phi_k$ (as denoted in the procedure).
-By repeated applications of \Clm{cut-int}, $\NN_{k+1}$ consists of wet points. 
-Consider $\wet(x,\NN_{k+1})$. The interface must exactly be $\phi_1, \phi_2, \ldots, \phi_k$.
-So the only dry vertices are those in the boundaries $B_1, B_2, \ldots, B_k$. But these
-boundaries are maxima.
-\end{proof}
-
-The procedure $\rain$ creates new boundary vertices on the edges it cuts each time it call the 
-subroutine $\cut$.  The key to understanding the running time of this procedure is to bound the 
-number of newly created vertices, for which we have the following lemma.
-
-\begin{lemma}\label{lem:new-verts}
-The total number of boundary vertices created by a call to $\rain(x, \NN)$ is $O(n)$, where $n=|\NN|$
-\end{lemma}
-\begin{proof}
-As the number of edges is $O(n)$, arguing that there is at most one boundary vertex created along any given edge will imply the claim.
-Consider some edge $e$ and take the first time a boundary vertex is created in $e$.
-This happens when some contour $\phi$ intersects $e$. On applying $\cut(\NN,\lift(\phi))$,
-consider the dry manifold $\LL$ and wet manifold $\MM$ constructed. The edge $e$ is split into two parts, the dry and wet part.
-The wet part is in $\MM$ and will never be involved in any recursive call.
-
-The dry part of $e$ (call it $e'$) needs to be considered.
-A boundary vertex $v$ (an new endpoint of $e$) is created, which is part of the boundary $B$ in $\LL$,
-such that $\rain(B,\LL)$ is called to the inverted $\LL$. In this call, since the water starts from $v$,
-all of $e'$ becomes wet. So there is no further interface contour that can intersect this portion of $e'$.
-\end{proof}
-
-As each original vertex in $\NN$ appears in only one of the sub-manifolds output by $\rain(x,\NN)$, we have the following corollary.
-
-\begin{corollary}
-The total size of a sub-manifolds output by $\rain(x,\MM)$ is $O(n)$.
-\end{corollary}
-
-\begin{corollary} \label{cor:rain-time} The total running time of $\rain(x,\MM)$ is $O(n)$.
-\end{corollary}
-
-\begin{proof} Consider some invocation to $\rain(x,\MM)$. The only non-trivial work done in $\rain$ is by the calls to $\wet$.
-We know that each invocation to $\wet$ runs in time linear 
-to the portion of the manifold which is wet (including the new boundary vertices created).  
-As the output of $\rain$ is precisely the distinct sub-manifolds wet by each distinct call to $\wet$, the claim follows.
-\end{proof}
-
-Consider the set of manifolds output by $\rain(x,\MM)$, there is a natural tree ordering defined over these manifolds where the 
-root corresponds the first manifold that was wet when we rained from $x$, and its children correspond to the recursive calls 
-(i.e. where we rained from the interface contours of the first raining).  Call this the \emph{Manifold Ordering Tree}.
-$\rain$ can easily be modified to also output this tree (without affecting the asymptotic running time).
-
-%Consider one of the manifolds $\MM_i$ output by $\rain$.  This manifold has a set of boundary maxima/minima that are the result 
-%of cutting along interface contours.  For each such contour there is exactly one other output manifold $\MM_j$ for which 
-%the contour appears as a boundary maxima/minima.
-
-\begin{claim} \label{clm:rain-reeb} Let $\reeb(\MM_1), \reeb(\MM_2), \ldots, \reeb(\MM_r)$, 
-be the output of $\rain(x,\MM)$. Suppose that for each reeb graph we know its corresponding vertex in the Manifold Ordering Tree. 
-Then $\reeb(\MM)$ can be computed in $O(n)$ time.
-\end{claim}
-
-\begin{proof}
-We will walk through the Manifold Ordering Tree in a leaf first ordering.  Each time we visit a node we connect its reeb graph to the 
-reeb graph of its children in the tree using the $\surgery$ procedure.
-This takes $O(n)$ time overall by \Lem{new-verts} and since $\surgery$ takes constant time (given the appropriate vertices in each reeb graph).
-The correctness follows by induction and  \Lem{surgery}.
-\end{proof}
-
-\section{Painting to compute contour trees} \label{sec:paint}
-The previous section allows us to restrict attention to extremum dominant manifolds.
-We will orient so that the extremum in question is a \emph{minimum}.
-We will fix such a manifold $\MM$, with the dominant minimum $u$.
-
-We begin by associating a color with each maximum. If the set of maxima is $S$,
-this is simply an arbitrary bijection $\chi:S \mapsto [|S|]$. Imagine there being a large
-can of paint of color $\chi(y)$ at maxima $y$. We will spill different paint from each maximum and watch it flow down.
-This is analogous to the raining in the previous section, but paint is a much more viscous liquid.
-\emph{So paint only flows down edges, and it does not color the interior of faces.} Furthermore, paints
-do not mix, so eventually, every vertex and edge of $\MM$ gets a unique color.
-%
-\begin{definition} \label{def:paint} A  \emph{painting} of $\MM$ is a map $\chi:S \cup E \mapsto [|S|]$
-with the following properties. (Remember that $G(\MM) = (V,E)$.)
-\begin{asparaitem}
-	\item The restriction $\chi:S \mapsto [|S|]$ is a bijection.
-	\item Consider an edge $e$. There exists a descending path from some maximum $x$ to $e$
-	consisting of edges in $E$, such that all edges along this path have the same color as $x$. (The maximum $x$
-	is well-defined to be the unique maximum with the same color as $e$.)
-\end{asparaitem}
-\end{definition}
-%
-Note that paintings are not unique. It is straightforward to construct a painting of $\MM$ in $O(n)$ time, using a descending BFS from each maximum that does not explore previously colored edges.
-
-\subsection{The data structures} \label{sec:struct}
-
-\begin{definition} \label{def:color-set} Fix a painting $\chi$, and a color $c$.
-A critical point $x$ is \emph{touched by $c$} if $x$ is incident to a $c$-colored
-increasing edge. The set of non-extremal critical points touched by $c$ is denoted $\touch(c)$.
-For $x$, $\col(x)$ is the set of colors that touch $x$.
-\end{definition}
-%
-%We will store each set $\touch(c)$ in a binomial heap, keyed by the heights. 
-Abusing notation, $\touch(c)$ refers
-both to the set and the data structure used to store it.
-
-\medskip
-\noindent
-{\bf The binomial heaps $\touch(c)$:} For each color $c$, we store a binomial
-heap $T(c)$ of non-extremal critical points, keyed by vertex heights.
-
-\medskip
-\noindent
-{\bf The union-find data structure on colors:} We will repeatedly perform unions
-of classes of colors, and this will be maintained as a standard union-find data structure.
-For any color $c$, $\rep(c)$ denotes the representative of its class. We initialize
-by simply setting $\rep(c) = c$ for all colors.
-
-\medskip
-\noindent
-{\bf The stack $\st$:} This consists of non-extremal critical points with the following structure.
-Each point $x \in \st$ has an associated subset of $\col(x)$, denoted $\mcol(x)$.
-Both $\mcol(x)$ and its complement are stored as hash table. So lookups, inserts, and deletes
-are in these sets are all constant time operations. The stack is guaranteed to satisfy 
-the following invariants.
-\begin{asparaitem}
-	\item For every $x \in \st$: For every $c \in \mcol(x)$, $x$ is the highest element
-	in $T(c)$. Furthermore, $c = \rep(c)$.
-	\item Consider $x, y \in \st$ such that $y$ was pushed on $x$. There exists $c \in \col(x) \setminus
-	\mcol(x)$ such that $x$ is not highest in $T(c)$ but $y$ is highest in $T(c)$.
-\end{asparaitem}
-
-\medskip
-\noindent
-{\bf Highest points $\h(c)$:} For each color $c$, we maintain a special critical point $\h(c)$ of this color.
-It is supposed to denote the ``highest" point of color $c$, but the notion of highest evolves
-with the algorithm.
-
-\subsection{Updating the stack} \label{sec:stack}
-
-We have a simple, but critical, procedure that updates the stack $\st$ by pushing
-points on to it.
-
-\medskip
-\fbox{
-\begin{minipage}{0.9\textwidth}
-{\bf $\update(\st)$}
-
-\smallskip
-\begin{compactenum}
-	\item If $K$ is empty, pick arbitrary $c$. Let $x$ be the highest point in $T(c)$.
-	Push $x$ onto stack and set $\mcol(x) = \{c\}$.
-	\item Let $h$ be head of $\st$.
-	\item While $\col(h) \setminus \mcol(h)$ non-empty:
-	\begin{compactenum}
-		\item Pick some $c \in \col(h) \setminus \mcol(h)$. Replace it by $c' = \rep(c)$.
-		\item Obtain highest point $x$ in $T(c')$. 
-		\item If $x = h$, put $c'$ in $\mcol(h)$. Otherwise ($x \neq h$), push $x$ onto stack
-		and add $c'$ to $\mcol(x)$.
-		\item Let $h$ be new head of $\st$.
-	\end{compactenum}
-\end{compactenum}
-\end{minipage}}
-
-\medskip
-When $\update(\st)$ terminates, the head $h$ has the following property. For every $c \in \col(h)$,
-$h$ is the highest point in $\touch(c)$.
-
-\begin{claim} \label{clm:cont} Consider an increasing path from $h$ to some maxima $x$.
-For any $y$ in this path, $\cont(y)$ only contains points of color $c(x)$.
-Furthermore, for all vertices on $\cont(y)$, their increasing edges are colored $c(x)$.
-\end{claim}
-
-\begin{proof} Consider this increasing path, and think of moving $y$ continuously downwards from $x$
-to $h$. Initially (when $y = x$), the lemma is trivially true. Consider the first $y$
-when the condition fails. We have $\cont(y)$ only containing points of color $c(x)$,
-but there is a vertex $z \in \cont(y)$ that has increasing edges of a different color $c'$.
-All the increasing edges of $z$ that are colored $c(x)$ are contiguous (since we can construct
-a contour intersecting all of them). Thus, there are two disjoint increasing neighborhoods for $z$,
-and $z$ is a saddle point. But that contradicts the fact that $h$ is the highest saddle point
-of color $c(x)$.
-
-So, as $y$ decreases, for all vertices on $\cont(y)$, their increasing edges are colored $c(x)$.
-This also means that $\cont(y)$ only contains points of color $c(x)$.
-\end{proof}
-
-We have important lemmas characterizing $|\col(h)|$.
-
-\begin{lemma} \label{lem:mono} If $h$ is a merge vertex, then $|\col(h)| > 1$.
-\end{lemma}
-
-\begin{proof} Suppose not. Consider the edges incident to $h$ in circular order. There exist (at least) $2$
-disjoint contiguous stretches of edges that are increasing. 
-All these edges have the same color $c$, associated with some maximum $x$. Consider two neighbors $y,z$
-that lie in different stretches. By the definition of a painting, there exist descending paths (along
-edges) from $x$ to $y$ and $x$ to $z$. We can construct a Jordan curve $C$ from $x \rightarrow y
-\rightarrow h \rightarrow z \rightarrow x$. All points in this curve except $h$ have 
-heights larger than $f(h)$. By the Jordan Curve Theorem, this partitions $\MM$ into
-an ``inside" and an ``outside".
-
-One of the lower neighbors of $h$ (say $p$) is inside, and another (say $q$) is outside.
-Because $h$ is a merge, neither of these can a minimum other than the dominant minimum $u$.
-Hence, both $p$ and $q$ have descending paths to $u$. So there is a path from $p$
-to $q$ where all heights are strictly less than $f(h)$. But this path must intersect $C$.
-Contradiction.
-\end{proof}
-
-\begin{lemma} \label{lem:col} If $h$ is a merge vertex, $|\col(h)| = 2$.
-If $h$ is a split vertex, $|\col(h)| = 1$.
-\end{lemma}
-
-\begin{proof} Consider one contiguous sequence of increasing edges from $h$.
-There exists a contour intersecting all of these edges. By \Clm{cont}, they all
-must have the same color. There are two such contiguous sequences.
-
-If $h$ is a merge vertex, then $|\col(h)| = 2$ by \Lem{mono}. If $h$ is a split vertex,
-there is a single contour passing through all increasing edges. By \Clm{cont}, they
-all have the same color.
-\end{proof}
-
-\subsection{Constructing $\reeb(\MM)$} \label{sec:const}
-
-The main construction procedure is $\onestep(\MM)$. It pieces together $\reeb(\MM)$
-incrementally.
-
-\medskip
-\fbox{
-\begin{minipage}{0.9\textwidth}
-{\bf $\onestep(\MM)$}
-
-\smallskip
-\begin{compactenum}
-	\item Run $\update(\st)$. Pop $\st$ to get $h$.
-	\item Suppose $\mcol(h) = \{c\}$ (so $h$ is split). 
-	\begin{compactenum}
-		\item Connect $h$ (in $\reeb(\MM)$) to $\h(c)$ and the unique minimum corresponding to split $h$. 
-		\item Delete $h$ from $T(c)$, set $\h(c) = h$.
-		\item End procedure.
-	\end{compactenum}
-	\item Let $\mcol(h) = \{c_1, c_2\}$ (so $h$ is merge).
-	\item Connect $h$ in $\reeb(\MM)$ to $\h(c_1)$ and $\h(c_2)$.
-	\item Delete all copies of $h$ from $T(c_1)$ and $T(c_2)$.
-	\item Perform union of colors $c_1$ and $c_2$ (denote merged color as $c$). Merge heaps
-	to get $T(c) = T(c_1) \cup T(c_2)$.
-	\item Set $\h(c) = h$.	
-\end{compactenum}
-\end{minipage}}
-
-\medskip
-\begin{theorem} \label{thm:piece} Consider the very first call to $\onestep(\MM)$. Let $y$ be a point
-infinitesimally below $h$ on a decreasing edge. Consider $\cut(\MM,\cont(y)) = \{\MM_u,\MM_\ell\}$ (the upper and lower manifolds). Then $\onestep(\MM)$
-outputs $\reeb(\MM_u)$ and implicitly gives a valid painting of $\MM_\ell$.
-\end{theorem}
-
-We have an important lemma that helps to bound the running time.
-
-\begin{lemma} \label{lem:tree} For every color $c$, all points in $T(c)$ lie along a monotone path in 
-$\reeb(\MM)$.
-\end{lemma}
-
-\begin{proof} Prove by induction of the number of edges in $\reeb(\MM)$. Consider the head $h$ of $\st$ in the first call to $\onestep(\MM)$,
-and the contour $\cont(y)$ as defined in \Thm{piece}. Take $\cut(\MM,\cont(y)) = \{\MM_u,\MM_\ell\}$. Apply induction
-hypothesis on $\MM_\ell$, and use that to argue about $T(c_1)$ and $T(c_2)$ in the original manifold.
-\end{proof}
-
-\pagebreak
-\newcommand{\pathTree}{P_{\mathcal{S}}}
-\newcommand{\pathTreeA}{\mathcal{P}_{\mathcal{S}}}
-\section{Leaf assignments and path decompositions}
-
-In this section, we set up a framework to analyze the time taken to compute a merge tree.
-
-\begin{definition}
- We say that a contour tree of a $2$-manifold is a \emph{merge tree} if it has a well defined root vertex such that height values 
- strictly increase along any root to leaf path.
-\end{definition}
-
-\begin{observation}
- The contour tree of a minimum dominant manifold (see \Def{dom}) is close to a merge tree.  Specifically, there is a root such that on any root 
- to leaf path height values strictly increase, with the only exception being potentially at the leaves themselves.  As such, up to constant 
- factors, any lower bound for computing the merge tree is a lower bound for computing the contour tree of a minimum dominant
- manifold (i.e. intuitively just clip off the leaves or shift them to lie above their parent).
-\end{observation}
-
-Any tree $T$ considered in this section will be a rooted full binary tree\footnote{Note 
-that technically the trees considered should have a leaf vertex hanging below the root of this 
-full binary tree in order to represent the global minimum of the surface.  This vertex is 
-(safely) ignored to simplify the presentation.} where the height of a vertex is its distance 
-from the root $r$.  As such, the children of a vertex $v\in T$ are the adjacent vertices of larger height
-(and $v$ is the parent of such vertices).  Then the subtree rooted at $v$, denoted $T_v$ consists of the graph induced on all 
-vertices which are descendants of $v$ (including $v$ itself).  For two vertices $v$ and $w$ in $T$ let $d(v,w)$ denote the 
-length of the path between $v$ and $w$.
-We use $A(v)$ to denote the set of ancestors of $v$.
-For set of nodes $U$, $A(U) = \bigcup_{u \in U} A(u)$.
-
-\begin{definition}
- A \emph{leaf assignment} $\chi$ of a tree $T$ assigns \emph{two} distinct leaves in each internal vertex $v$,
- one from the left child and one from the right child subtree of $v$.
-\end{definition}
-
-For a vertex $v\in T$, we use $H_v$ to denote the \emph{heap} at $v$.
-Formally, $H_v = \{u | u \in A(v), \chi(u) \cap L(T_v) \neq \emptyset\}$.
-In words, $H_v$ is the set of ancestors of $v$ which are colored by some leaf in $T_v$.
-The following lemma should justify these technical definitions.
-
-\begin{lemma}
-The running time of our algorithm is $O(\sum_{v \in T} \log |H_v|)$.
-% where $T$ is the merge tree of the given input, 
-%and $\chi$ is any valid coloring of $T$.  (maybe add one in the definition of $H_v$ so that $\log |H_v|$ is always $>1$).
-\end{lemma}
-
-\begin{proof} \Sesh{add}
-
-\end{proof}
-
-%vertices along the path from the root to $v$ such that one of the two colors of each vertex is in $L(T_v)$.
-%We will refer to $H_v$ as the \emph{heap} at $v$.
-%Then define the \emph{algorithm} cost function as $f_{alg}(\chi(T)) = \sum_{v\in T} \log |H_v|$.
-%
-%\begin{lemma}
-%\label{lem:algCost}
-%NEEDS TO BE WRITTEN:
-%Here we need a lemma showing that the running time of our algorithm is $O(f_{alg}(\chi(T)))$, where $T$ is the merge tree of the given input, 
-%and $\chi$ is any valid coloring of $T$.  (maybe add one in the definition of $H_v$ so that $\log |H_v|$ is always $>1$).
-%\end{lemma}
-%
-%Now that we have a simple expression upper bounding the running time of our algorithm, we spend the remainder 
-%of this section proving that this running time is optimal.
-
-We introduce the concept of path decompositions. A \emph{leaf path} is simply a contiguous path in $T$
-containing a leaf, which is monotonic in the height values of its vertices.
-% Then in the following two subsections 
-%we relate path decomposition costs (defined below) to the running time of our algorithm using $\Lem{algCost}$.
-
-\begin{definition}
-\label{def:path} A \emph{path decomposition}, $P(T)$, is a partition of $T$ into 
- a set of vertex disjoint leaf paths.  
-% 
-% The \emph{cost} of $P(T)$ is $\cost(P(T)) = \sum_{p\in P(T)} |p| \log |p|$. (We use $|p|$ to denote the number of vertices in $p$.)
-\end{definition}
-
-Note that the number of paths in $P(T)$ must equal the number of leaves in $T$.
-Fix $\chi$, which fixes all the heaps $H_v$. We choose a special path decomposition that is best defined as a subset of edges in $T$ such that each internal
-vertex has degree at most $2$. This naturally gives a path decomposition.
-For each internal vertex $v\in T$, add the edge from $v$ to $\arg \max_{v_l, v_r} \{|H_{v_l}|, |H_{v_r}|\}$ 
- where $v_l$ and $v_r$ are the children of $v$ (if $|H_{v_l}|=|H_{v_r}|$ then pick one arbitrarily).
-This is called the \emph{maximum} path decomposition, denoted by $\pmax(T)$.
-
-Our main theorem is the following. We use $|p|$ to denote the number of vertices in $p$.
-
-\begin{theorem} \label{thm:runtime} $\sum_{v \in T} \log |H_v| = O(\sum_{p\in \pmax(T)} |p| \log |p|)$.
-\end{theorem}
-
-\subsection{Shrubs, tall paths, and short paths}
-
-The paths in $P(T)$ naturally define a tree\footnote{Please excuse the 
-overloading of the term 'tree', it is the most natural term to use here.} of their own.  Specifically, in the original 
-tree $T$ contract each path down to its root.  Call the resulting tree the \emph{shrub} of $T$ corresponding to 
-the path decomposition $P(T)$. Abusing notation, we simply use $P(T)$ to denote the shrub.
-As a result, we use terms like `parent', `child', `sibling', etc. for paths as well.
-The shrub gives a handle on the heaps of a path.
-We use $b(p)$ to denote the \emph{base} of the path, which is vertex in $p$ closest
-to root of $T$. We use $\ell(p)$ to denote the leaf in $p$.
-We use $H_p$ to denote the $H_{b(p)}$.
-% and denote it by $\pathTree(T)$.
-
-\begin{lemma}
-\label{lem:adjacent}
- Let $p$ be any path in $P(T)$ and let $\{q_1, \dots q_k\}$ be the children on $p$.
-Then $H_{\ell(p)} + \sum_{i=1}^k |H_{q_i}| \leq |H_p|+2|p|$. 
-\end{lemma}
-
-\begin{proof} For convenience, denote $H_i = H_{q_i}$ and $H_0 = H_{\ell(p)}$.
-Consider $v \in \bigcup_{i} H_i$ that lies below $b(p)$ in $T$. 
-Note that such a vertex has only one of its two colors in $L(b(p))$.  
-Since the colors tracked by $H_i$ and $H_j$ for $i\neq j$ are disjoint, 
-such a vertex can appear in only one of the $H_i$'s.
-On the other hand, a vertex $u\in p$ can appear in more than one $H_i$, 
-but since any vertex has exactly two colors it can appear in at most two 
-such heaps.  Hence, $\sum_i |H_i| \leq |H_p| + 2|p|$.
-\ignore{
- First observe that for all $i$, $r_{i}$ lies above $r_p$ on some root to leaf path. 
- Let $H_i'$ be the subset of $H_i$ that lies below $r_p$.  Observe that $H_i'\subseteq H_p$, as $L(T_{r_i})\subseteq L(T_{r_p})$.
-
- Now for any $i\neq j$, $H_i \cap H_j = \emptyset$, as $r_{i}$ and $r_{j}$ are not on the same root to leaf path (i.e. $L(T_{r_i}) \cap L(T_{r_j}) = \emptyset$).
- In particular, $(H_i\setminus H_i') \cap (H_j \setminus H_j') = \emptyset$, and so $\sum_{i=0}^k |H_i\setminus H_i'| \leq |p|$ 
- as $H_i\setminus H_i'$ is precisely the subset of $H_i$ that lies on $p$.
- Moreover, for any $i\neq j$, $ H_i' \cap H_j' = \emptyset$, and so since $H_i'\subseteq H_p$ we have $\sum_{i=0}^k |H_i'|\leq |H_p|$.
- Putting these together gives
- \[
-  \sum_{i=0}^k |H_i| = \sum_{i=0}^k |H_i'| + \sum_{i=0}^k |H_i\setminus H_i'| \leq |H_p|+|p|.
- \]
-}
-\end{proof}
-
-We wish to prove $\sum_{v\in T} \log |H_v| = O(\sum_{p\in P} |p| \log |p|)$. 
-The simplest approach is to prove $\forall p\in P$, $\sum_{v\in p} \log |H_v|= O(|p|\log|p|)$.
-This is unfortunately not true, which is why we divide paths into two categories.
-
-\begin{definition}
- For $p\in P(T)$, $p$ is \emph{short} if $|p| < \sqrt{|H_{p}|}/100$, and \emph{tall} otherwise.
-\end{definition}
-
-\ignore{
-\begin{observation}
-\label{obs:decrease}
- Let $v$ be a vertex in $T$ and let $w$ be its parent.  Then $|H_w| \geq |H_v| -1$, as $L(T_v) \subseteq L(T_w)$ and the path from 
- $w$ to the root has one less vertex than the path from $v$ to the root.
- In particular, we have the following more general property.
- Let $v$ and $u$ be any two vertices in the same root to leaf path of $T$, such that $v$ is a descendant of $u$.  
- Then $|H_v| \leq |H_u| + d(u,v)$. 
-\end{observation}
-
-The following can be thought of as a generalization of the above observation, and will be useful in later sections.
-}
-
-The following lemma demonstrates that tall paths can ``pay'' for themselves.
-
-\begin{lemma}
-\label{lem:pathBounds} If $p$ is tall, $\sum_{v\in p} \log |H_v| = O(|p| \log |p|)$.
-If $p$ is short, $\sum_{v\in p} \log |H_v| = O(|H_p| \log |H_p|)$.
-% Let T be a binary tree and let $\chi(T)$ and $P(T)$ be a valid coloring and path decomposition of $T$, respectively.
-% Then, 
-% \begin{compactenum}
-%  \item for any tall path $p\in P(T)$, $\sum_{v\in p} \log |H_v| = O(|p| \log |p|)$.
-%  \item for any short path $p\in P(T)$, $\sum_{v\in p} \log |H_v| = O(|p| \log |H_{p}|)$.
-% \end{compactenum}
-\end{lemma}
-\begin{proof} \ignore{
- By $\Obs{decrease}$ we know that for any vertex $v\in p$, $|H_v|\leq |H_{p}| + |p|$ (as $v$ is a descendant of $r_p$ along $p$).
-}
-For $v \in p$, $|H_v|\leq |H_{p}| + |p|$ (as $v$ is a descendant of $b(p)$ along $p$).
-Hence, $\sum_{v\in p} \log |H_v| \leq \sum_{v\in p} \log(|H_{p}| + |p|)  = |p| \log (|H_{p}| + |p|)$.
- If $p$ is a tall path, then $|p| \log (|H_{p}| + |p|) = O(|p| \log |p|)$. If $p$ is short, then 
- $|p| \log (|H_{p}| + |p|) = O(|p| \log |H_{p}| )$. For short paths, $|p| = O(|H_p|)$.
-\end{proof}
-
-\ignore{
-For a short path $p$ we can think of the quantity $|p|(\log |H_p| - \log |p|)$ as the excess of $p$, i.e. what was not ``paid for'' locally by $p$.
-Our eventual goal will be to show that by choosing the right path decomposition, this excess 
-can be recharged to tall paths. In the next couple sections we introduce the appropriate path decomposition and prove 
-some useful facts about it.  Then given these facts, in \Sec{} we are able to make the recharging argument.
-}
-
-There are some short paths that we can also ``pay" for.  Consider any short path $p$ in the shrub.  We will refer to the 
-\emph{tall support chain} of $p$ as the tall ancestors of $p$ in the shrub which have a path to $p$ which does not use any 
-short path (i.e. it is a chain of paths adjacent to $p$).
-
-\begin{definition} \label{def:support} A short path $p$ is \emph{supported} if
-at least $|H_p|/100$ vertices $v$ in $H_p$ lie in paths in the tall support chain of $p$.
-%
-%
-% there exists
-%a contiguous (from $p$) chain of ancestors $q_1, q_2, \ldots, q_k$ (in $P(T)$) such that $|H_{q_k}| \leq 99|H_p|/100$
-%and all $q_i$'s are tall.
-\end{definition}
-
-Let $\cL$ be the set of short paths, $\cL'$ be the set of supported short paths, and $\cH$ be the set of tall paths
-given by $\pmax(T)$.
-We now construct the shrub of unsupported short paths. Consider $p \in \cL \setminus \cL'$,
-and traverse the chain of ancestors from $p$. Eventually, we must reach another short path $q$.
-(If not, we have reached the root $r$ of $\pmax(T)$. Hence, $p$ is supported.)
-Insert edge from $p$ to $q$, so $q$ is the parent of $p$ in $\cU$. This construction leads
-to the shrub forest of $\cL \setminus \cL'$, where all the roots are supported short paths, 
-and the remaining nodes are the unsupported short paths.
-
-Most of the work goes into proving the following technical lemma.
-
-\begin{lemma} \label{lem:cu-root} Let $\cU$ denote a connected component (shrub) in the shrub forest of $\cL \setminus \cL'$ and let $r$
-be the root of $\cU$. (i) For any $v \in p$ such that $p \in \cU$,
-$|H_v| = O(|H_r|)$. (ii) $\sum_{p \in \cU} |p| = O(|H_r|)$.
-\end{lemma}
-
-We split the remaining argument into two subsections. We first prove \Thm{runtime} from \Lem{cu-root},
-which involves routine calculations. Then we prove \Lem{cu-root}, where the interesting work happens.
-
-\subsection{Proving \Thm{runtime}} \label{sec:thm-runtime}
-
-We split the summation into tall, short, and unsupported short paths.
-\begin{eqnarray*} 
-\sum_{p \in \cL} \sum_{v \in p} \log |H_v| & = & \sum_{p \in \cL \setminus \cL'} \sum_{v \in p} \log |H_v| + \sum_{p \in \cL'} \sum_{v \in p} \log |H_v| + \sum_{p \in \cH} \sum_{v \in p} \log |H_v|
-%& = & \sum_{p \in \cL \setminus \cL'} \sum_{v \in p} \log |H_v| + O(\sum_{p \in \cL'} |p|\log |H_p|) = \sum_{p \in \cL \setminus \cL'} \sum_{v \in p} \log |H_v| + O(\sum_{p \in \cL'} |H_p|\log |H_p|)
-\end{eqnarray*}
-The last term can be bounded by $O(\sum_{p \in \pmax(T)} |p|\log |p|)$, by \Lem{pathBounds}.
-The second term can be bounded by $O(\sum_{p \in \cL'} |H_p| \log |H_p|)$, by \Lem{pathBounds} again.
-The following claim shows that this in turn is at most the last term.
-
-%We have a claim that also bounds heap sizes for special set of paths.
-
-\begin{claim} \label{clm:above} $\sum_{p \in \cL'} |H_p| \log |H_p| = O(\sum_{q \in \cH} \sum_{v \in q} \log |H_v|)$.
-\end{claim}
-
-\begin{proof} 
-Pick $p \in \cL'$. 
-%Let the first $|H_p|/200$ vertices the lie in tall paths be denoted as $H'_p$. 
-%Observe that for any $v \in H'_p$, 
-%We have $|H_p| \log |H_p| = O(\sum_{v \in H'_p} \log |H_p|)$
-%$= O(\sum_{v \in H'_p} \log |H_v|)$. Since all the vertices lie in tall paths,
-%we can write this as $O(\sum_{q \in \cH} \sum_{v \in H'_p \cap q} \log |H_v|)$.
-%Summing over all $p$, the expression is $\sum_{q \in \cH} \sum_{p \in \cL'} \sum_{v \in H'_p \cap q} \log |H_v|$.
-%
-%
-As we traverse the tall support chain of $p$, there
-are at least $|H_p|/100$ vertices of $H_p$ that lie in these paths. These are encountered in
-a fixed order. Let $H'_p$ be the first $|H_p|/200$ of these vertices. When $v \in H'_p$
-is encountered, there are $|H_p|/200$ vertices of $H_p$ not yet encountered. Hence,
-$|H_v| \geq |H_p|/200$. Hence, $|H_p|\log |H_p| = O(\sum_{v \in H'_p} \log |H_v|)$.
-Since all the vertices lie in tall paths,
-we can write this as $O(\sum_{q \in \cH} \sum_{v \in H'_p \cap q} \log |H_v|)$.
-Summing over all $p$, the expression is $\sum_{q \in \cH} \sum_{p \in \cL'} \sum_{v \in H'_p \cap q} \log |H_v|$.
-%and $\sum_{p \in \cL'} |H_p|\log |H_p| = O(\sum_{p \in \cL'} \sum_{v \in H'_p} \log |H_v|)$.
-
-Consider any $v \in H'_p$.  Let $S$ be the set of paths $\widetilde{p}\in \cL'$ such that 
-$v\in H'_{\widetilde{p}}$.  We now show $|S|\leq 2$ (i.e. it contains at most one path other than $p$).
-First observe that any two paths in $S$ must be unrelated (i.e. $S$ is an anti-chain), 
-since paths which have an ancestor-descendant relationship have disjoint tall support chains.  
-However, any vertex $v$ receives exactly one color from each of its two subtrees (in $T$), and therefore $|S|\leq 2$ since any two 
-paths which share descendant leaves in $T$ (i.e. their heaps are tracking the same color) must have an ancestor-descendant relationship.
-
-In other words, any $\log |H_v|$ appears at most twice in the above triple summation.
-Hence, we can bound it by $O(\sum_{q \in \cH} \sum_{v \in q} \log |H_v|)$.
-%
-%Pick $q \in Q$ and some $v \in H_q$. As we traverse up the tree passing
-%through $A(q)$, we must encounter some $p \in A(q)$ such that $v \in p$. Furthermore,
-%there is an order in which these vertices (of $H_q$) occur. Let $H'$ be the first half
-%of vertices that occur. For all $v \in H'$, $|H_v| \geq |H_q|/2$. Hence,
-%$|H_q| \log |H_q| \leq 4 \sum_{v \in H'} \log |H_v| \leq 4 \sum_{v \in H_q} |H_v|$.
-%So $\sum_{q \in Q} |H_q| \log |H_q| \leq 4 \sum_{q \in Q} \sum_{v \in H_q} |H_v|$.
-%
-%Since nodes in $Q$ are incomparable, all the $H_q$'s (for $q \in Q$) are disjoint.
-%Each such $v$ must occur in some path $p \in A(Q)$. Thus,
-%$\sum_{q \in Q} |H_q| \log |H_q| \leq 4 \sum_{p \in A(Q)} \sum_{v \in p} |H_v|$.
-%We apply the bound of \Lem{pathBounds} to complete the proof.
-\end{proof}
-
-The first term (unsupported short paths) can be charged to the second term (supported short paths).
-This is where the critical \Lem{cu-root} plays a role.
-
-\begin{claim} \label{clm:first} $\sum_{p \in \cL \setminus \cL'} \sum_{v \in p} \log |H_v| = O(\sum_{p \in \cL'} |H_p| \log |H_p|)$.
-\end{claim}
-
-\begin{proof} Let $\cU$ denote a connected component of the shrub forest.
-We have $\sum_{p \in \cL \setminus \cL'} \sum_{v \in p} \log |H_v| \leq \sum_{\cU} \sum_{p \in \cU} \sum_{v \in p} \log |H_v|$.
-By \Lem{cu-root}, $|H_v| = O(|H_r|)$, where $r$ is the root of $\cU$.
-Furthermore, $\sum_{p \in \cU} |p| = O(|H_r|)$. 
-We have $\sum_{p \in \cU} \sum_{v \in p} \log |H_v| = O((\log |H_r|) \sum_{p \in \cU} |p|) = O(|H_r|\log |H_r|)$.
-We sum this over all $\cU$ in the shrub forest, and note that roots in the shrub forest are supported short paths.
-\end{proof}
-
-\ignore{
-Now, for the critical piece of the approach. Focus on $\pmax(T)$. Let $L$ be the set of short paths, 
-and $\hat{L} \subseteq L$ be the set of paths such that no ancestor of $p \in L'$
-is short. Obviously, $\hat{L}$ is an anti-chain in $P(T)$.
-
-\begin{lemma} \label{lem:short} For the path decomposition $\pmax(T)$, $\sum_{p \in \cL} \sum_{v \in p} \log |H_v| = O(\sum_{p \in \cL'} |H_p| \log |H_p|)$.
-\end{lemma}
-
-With this lemma, the proof of \Thm{runtime} is straightforward.
-
-\begin{proof} (of \Thm{runtime}) We focus on $\pmax(T)$ and split into tall and short paths.
-We have $\sum_{v \in T} \log |H_v| =$  $\sum_{p \in \cL} \sum_{v \in p} \log |H_v| + $ $\sum_{p \in \cH} \sum_{v \in p} \log |H_v|$.
-For the first term, we apply \Lem{short} to bound by $O(\sum_{p \in \cL'} |H_p| \log |H_p|)$.
-Then we apply \Clm{above} to bound by $O(\sum_{p \in \cH} |p|\log |p|)$.
-The latter is also bounded by the same expression, by \Lem{pathBounds}. By definition,
-$\sum_{p \in \cH} |p|\log |p| \leq \cost(\pmax(T))$.
-\end{proof}
-
-The main challenge is proving \Lem{short}, which we defer to the next section.
-\begin{definition}
- Let $\chi(T)$ be a valid coloring of a binary tree $T$.  We define a set of disjoint paths over the vertices of 
- $T$ as follows.  
- The corresponding collection of paths that all such edges correspond to is a path decomposition of $T$, which 
- we refer to as a \emph{maximum} path decomposition, and is denoted by $\mathcal{P}(T)$.  
-\end{definition}
-
-From the previous section we now have a lower bound of $\Omega(f_{path}(P))$ for computing the merge tree in terms of any path decomposition $P$. 
-We now show that the running time of our algorithm for computing the merge tree is upper bounded by this expression for a specific path decomposition, 
-which we will call the maximum path decomposition (defined below).
-Specifically, for a given valid coloring $\chi$ of a tree $T$, we will prove that for the maximum path decomposition $f_{alg}(\chi(T)) = O(f_{path}(P(T)))$. 
-%using the observations about path decompositions from 
-%\Sec{background}, which will then imply $f_{alg}(\chi(T)) = \Theta(f_{path}(P(T)))$.
-
-Throughout this section, for a path $p\in P(T)$, we use the notation $H_p$ to refer to the heap at the root of $p$, i.e. $H_p = H_{r_p}$.
-}
-
-\subsection{Proving \Lem{cu-root}: the root is everything in $\cU$} \label{sec:weight}
-
-\Lem{cu-root} asserts the root $r$ in $\cU$ pretty much encompasses all sizes and heaps in $\cU$.
-We will work with the \emph{reduced} heap $\redH_p$. This is the subset of vertices of $H_p$ that 
-do not appear on the tall support chain of $p$.
-%only appear in short paths. 
-By definition, for any unsupported short path
-(hence, any non-root $p \in \cU$), $|\redH_p| \geq 99|H_p|/100$. 
-% For notational convenience, we simply set $\redH_r = H_r$.
-%Consider $q \in \cL \setminus \cL'$ that is a child of $p$ in the shrub $\cU$.
-We begin with a key property, which is where the construction of $\pmax(T)$ enters the picture.
-
-%
-%
-%
-%We focus on $\pmax(T)$, and prove the key property of this path decomposition.
-
-\begin{lemma}
-\label{lem:geometric}
- Let $q$ be the child of some path $p$ in $\cU$, then $|H_p|\geq \frac{3}{2}|H_q|$. 
- Moreover, if $p\neq r(\cU)$, then $|\redH_p|\geq \frac{3}{2}|\redH_q|$. 
-\end{lemma}
-\begin{proof} Let $h(q)$ denote the tall path that is a child of $p$ in $\pmax(T)$,
-and an ancestor of $q$. If no such tall path exists, then by construction $p$ is the parent 
-of $q$ in $\pmax(T)$, and the following argument will go through by setting $h(q)=q$.
-
-The chain of ancestors from $q$ to $h(q)$ consists only of tall paths.
-Since $q$ is unsupported, these paths contain at most $|H_q|/100$ vertices of $H_q$. 
-Thus, $|H_{h(q)}| \geq 99|H_q|/100$.
-
-Consider the base
-of $h(q)$, which is a node $w$ in $T$. Let $v$ denote the sibling of $w$ in $T$.
-Their parent is called $u$. Note that both $u$ and $v$ are nodes in the path $p$.
-Now, the decomposition $\pmax(T)$ put $u$ and $v$ in the same path $p$. This
-implies $|H_v| \geq |H_w|$. Since $|H_u| \geq |H_v| + |H_w| - 2$,
-$|H_u| \geq 2|H_w| - 2$. Let $b$ be the base of $p$.
-We have $|H_p| = |H_b| \geq |H_u| - |p| \geq 2|H_w| - |p| - 2$. Since $p$ is a short path, $|p| < \sqrt{|H_p|}/100$.
-Applying this bound, we get $|H_p| \geq (2-\delta)|H_w|$ (for a small constant $\delta > 0$).
-Since $w$ is the base of $h(q)$, $H_w = H_{h(q)}$. We apply the bound $|H_{h(q)}| \geq 99|H_q|/100$
-to get $|H_p| \geq 197|H_q|/100$, implying the first part of the lemma.  For the second part, observe that 
-if $p\neq r(\cU)$, then $p$ is unsupported and so $|\redH_p| \geq 99|H_p|/100$, and therefore the second part follows since $|H_q| \geq |\redH_q|$.
-%
-%
-% Let $w$ be the vertex in the path $p$ to which the root of the path $q$ is adjacent to in $T$, 
-% i.e. $w$ is the parent of $r_q$ in $T$.  Let $c$ be the other child of $w$, which lies in $p$.
-% As $\mathcal{P}(T)$ is a maximum path decomposition of $T$ and $c$ rather than $r_q$ is in 
-% the same path as $w$, we have $|H_q|\leq |H_c|$.  Therefore, $|H_w|\geq |H_q|+|H_c|-1\geq 2|H_q|-1$. 
-% Therefore, by \Obs{decrease}, $|H_p|+ d(r_p, w) \geq |H_w|\geq 2|H_q|-1$.  
-% Since $p$ is a short path, $d(r_p, w)\leq \sqrt{|H_p|}/100$ and so we have $|H_p|+\sqrt{|H_p|}/100\geq 2|H_q|-1$.
-% Assuming $|H_p|>11$ (and we can safely assume $|H_p|$ is bigger than any constant), this yields $\frac{3}{5} |H_p| \geq |H_q|$
-\end{proof}
-
-This immediately proves part (i) of \Lem{cu-root}. Part (ii) requires much more work.
-% It is helpful to spell out how the various heaps evolve. Consider $p \in \cU$,
-% which has children $q_1, q_2, \ldots, q_k$ in $\pmax(T)$, and a leaf vertex $\ell$ in $T$. 
-% The reduced heap $\redH_p$ is the
-% union of $\redH_\ell$ and $\redH_{q_1}, \redH_{q_2}, \ldots, \redH_{q_k}$, with the vertices in $p$
-% removed. Note that some of the $q_i$'s are the corresponding $h(q')$ vertices,
-% for a child $q'$ in $\cU$.
-
-We define a \emph{residue} $R_p$ for each $p \in \cU$. Suppose $p$ has children
-$q_1, q_2, \ldots, q_k$ in $\cU$. Then $R_p = |\redH_p| - \sum_i |\redH_{q_i}|$. 
-By definition, $|\redH_p| = \sum_{q \in \cU_p} R_p$. 
-Note that $R_p$ can be negative. 
-Now, define $R^+_p = \max(R_p,0)$, and set $W_p = \sum_{q \in \cU_p} R^+_p$.
-Observe that $W_p \geq |\redH_p|$. We also get an approximate converse.
-
-\begin{claim} \label{clm:loss} For any path $p\in \cU$, $|\redH_p| \geq W_p - 2\sum_{q \in \cU_p} |q|$.
-\end{claim}
-
-\begin{proof} We write $W_p - |\redH_p| = \sum_{q \in \cU_p} R^+_q - R_q$ $= -\sum_{q \in \cU_p: R_q < 0} R_q$.
-Consider $q \in \cU_p$ and denote the children in $\cU_p$ by $q'_1, q'_2, \ldots$.
-Note that $R_q$ is negative exactly when $|\redH_q| < \sum_i |\redH_{q'_i}|$.
-Traverse $\pmax(T)$ from $q'_i$ to $q$.
-Other than $q$, all other nodes encountered are in the tall support chain of $q'_i$ and hence
-do not affect its reduced heap.
-The vertices of $\redH_{q'_i}$ that are deleted
-are exactly those present in the path $q$. 
-Any vertex in $q$ can be deleted from at most two of the reduced heaps (of the children of $q$ in $\cU_p$), since 
-theses reduced heaps do not have an ancestor-descendant relationship.
-Therefore when $R_q$ is negative, it is at most by $2|q|$.
-We sum over all $q$ to complete the proof.
-\end{proof}
-
-\ignore{Let $D_q = H_q \setminus \bigcup_i H_{q'_i}$. In other words, as we traverse $\cU$,
-$D_q$ is the set of vertices ``added" at $q$. Suppose we traverse $\pmax(T)$ from $q'_i$
-to $q$. All vertices of $H_{q'_i}$ that lie on these paths are removed.
-
- and traverse $\pmax(T)$ up to $p$. Suppose we encounter
-the paths $q_0 = q, q_1, q_2, \ldots, q_k = p$. The vertices in the residue $R_q$ that 
-are removed from $H_p$ are exactly the vertices that lie in these paths. 
-We split this contribution into tall and short paths, observing that the latter
-are all unsupported. Hence, we can (crudely) upper bound the number of vertices
-removed by $\sum_{q \in \cU_p} |q| + \sum_{q \in \cU_p} \sum_{h \in \cH} |h \cap q|$.
-Each vertex can be removed from at most two distinct $R_q$'s, completing the proof.
-}
-
-The main challenge of the entire proof is bounding the sum of path lengths, which is done next.
-We stress that the all the previous work is mostly the setup for this claim.
-
-\begin{claim} \label{clm:charge} Fix any path $p \in \cU\setminus\{r(\cU)\}$. Suppose for any $q,q' \in \cU_p$
-where $q$ is a parent of $q'$ in $\cU_p$, $W_q \geq (4/3)W_{q'}$. Then $\sum_{q \in \cU_p} |q| \leq W_p/20$.
-\end{claim}
-
-\begin{proof} Since $q$ is an unsupported short path, $|q| < \sqrt{|H_q|}/100 \leq \sqrt{|\redH_q|}/99 \leq \sqrt{|W_q|}/99$.
-We prove that $\sum_{q \in \cU_p} \sqrt{|W_q|}/99 \leq W_p/20$
-by a charge redistribution scheme. Assume that each $q \in \cU_p^-$
-starts with $\sqrt{W_q}/99$ units of charge. We redistribute this charge over all
-nodes in $\cU_q$, and then calculate the total charge. For $q \in \cU_p$, spread
-its charge to all nodes in $\cU_q$ proportional to $R^+$ values. In other words,
-give $(\sqrt{W_q}/99)\cdot(R^+_{q'}/W_q)$ units of charge to each $q' \in \cU_q$.
-
-After the redistribution, let us compute the charge deposited at $q$. Every ancestor in $\cU_p^-$ 
-$q = a_0, a_1, a_2, \ldots, a_k$ contributes to the charge at $q$. The charge is expressed in the following
-equation. We use the assumption that $W_{a_i} \geq (4/3) W_{a_{i-1}}$ and hence $W_{a_i} \geq (4/3)^i W_{a_0} \geq (4/3)^i$, 
-as $a_0$ is an unsupported short path and hence $W_{a_0}\geq 1$. 
-$$ ({R^+_q}/99)\sum_{a_i} 1/\sqrt{W_{a_i}} \leq ({R^+_q}/99)\sum_{a_i} (3/4)^{i/2} \leq R^+_q/20 $$
-The total charge is $\sum_{q \in \cU_p} R^+_p/20 = W_p/20$. 
-\end{proof}
-
-\begin{corollary} \label{cor:chargeCor}
- Let $r$ be the root of $\cU$, and suppose that for any paths $q,q' \in \cU\setminus \{r\}$, where $q$ is a parent of $q'$ in $\cU$, $W_q \geq (4/3)W_{q'}$.
- Then $\sum_{p \in \cU} |p| \leq W_r/20+|r|$.
-\end{corollary}
-\begin{proof}
- Let $c_1, \dots, c_m$ be the children of $r$ in $\cU$.  By definition, $W_r = \sum_i W_{c_i}+R^+_r \geq \sum_i W_{c_i}$.
- By \Clm{charge}, for each $c_i$ we have $W_{c_i}/20\geq \sum_{p \in \cU_{c_i}} |p|$.  Combining these to facts yields the claim.
-\end{proof}
-
-% \begin{claim} \label{clm:charge} Fix $p \in \cU$. Suppose for any $q,q' \in \cU_p$
-% where $q$ is a parent of $q'$ in $\cU_p$, $W_q \geq (4/3)W_{q'}$. Then $\sum_{q \in \cU_p} |q| \leq W_p/20$.
-% \end{claim}
-% 
-% \begin{proof} Since $q$ is an unsupported short path, $|q| < \sqrt{|H_q|}/100 \leq \sqrt{|\redH_q|}/99 \leq \sqrt{|W_q|}/99$.
-% We prove that $\sum_{q \in \cU_p} \sqrt{|W_q|}/99 \leq W_p/20$
-% by a charge redistribution scheme. Assume that each $q \in \cU_p$
-% starts with $\sqrt{W_p}/99$ units of charge. We redistribute this charge over all
-% nodes in $\cU_p$, and then calculate the total charge. For $q \in \cU_p$, spread
-% its charge to all nodes in $\cU_q$ proportional to $R^+$ values. In other words,
-% give $(\sqrt{W_q}/99)\cdot(R^+_{q'}/W_q)$ units of charge to each $q' \in \cU_q$.
-% 
-% After the redistribution, let us compute the charge deposited at $q$. Every ancestor in $\cU_p$ 
-% $q = a_0, a_1, a_2, \ldots, a_k$ contributes to the charge at $q$. The charge is expressed in the following
-% equation. We use the assumption that $W_{a_i} \geq (4/3) W_{a_{i-1}}$ and hence $W_{a_i} \geq (4/3)^i W_{a_0} \geq (4/3)^i$, 
-% as $a_0$ is an unsupported short path and hence $W_{a_0}\geq 1$. 
-% $$ ({R^+_q}/99)\sum_{a_i} 1/\sqrt{W_{a_i}} \leq ({R^+_q}/99)\sum_{a_i} (3/4)^{i/2} \leq R^+_q/20 $$
-% The total charge is $\sum_{q \in \cU_p} R^+_p/20 = W_p/20$. 
-% \end{proof}
-
-We wrap it all up by proving part (ii) of \Lem{cu-root}.
-
-\begin{claim} $\sum_{p \in \cU} |p| \leq |H_{r(\cU)}|/10$.
-\end{claim}
-
-% \begin{proof} We use $r$ for $r(\cU)$. Our aim is to simply apply \Clm{charge} from $r(\cU)$. 
-% Suppose $W_q \geq (4/3)W_{q'}$ (for any choice in $\cU$ of $q$ parent of $q'$). By \Clm{charge}, $\sum_{p \in \cU} |p| \leq W_{r}/20$.
-% By \Clm{loss}, $|\redH_r| \geq W_r - 2\sum_{p \in \cU} |p|$. By simple algebra, we deduce, $\sum_{p \in \cU} |p| \leq |\redH_r|/10 = |H_r|/10$.
-% 
-% We now prove that for any $q$ parent of $q'$, $W_q \geq (4/3)W_{q'}$.
-% Suppose not. Let $p, p'$ be the counterexample furthest from the root,
-% where $p$ is the parent of $p'$. 
-% Note that for $q$ and child $q'$ in $\cU_{p'}$, $W_q \geq (4/3)W_{q'}$.
-% We will apply \Clm{charge} for $\cU_{p'}$ to deduce that $\sum_{q \in \cU_{p'}} |q| \leq W_{p'}/20$.
-% By \Clm{loss}, $|\redH_{p'}| \geq 19W_{p'}/20$. By \Lem{geometric}, $|\redH_p| \geq (3/2)|\redH_{p'}|$.
-% Noting that $W_p \geq |\redH_p|$, we deduce that $W_p \geq (4/3)W_{p'}$.
-% Hence, $p, p'$ is not a counterexample, and more generally, this is no counterexample.
-% That completes the whole proof.
-% \end{proof}
-
-\begin{proof} We use $r$ for $r(\cU)$. %Our aim is to simply apply \Clm{charge} from $r(\cU)$. 
-Suppose $W_q \geq (4/3)W_{q'}$ (for any choice in $\cU\setminus\{r\}$ of $q$ parent of $q'$), 
-then by \Cor{chargeCor}, $\sum_{p \in \cU} |p| \leq W_{r}/20+|r|$.
-By \Clm{loss}, $|\redH_r| \geq W_r - 2\sum_{p \in \cU} |p|$, and so combining these inequalities gives,
-\[
-% \sum_{p \in \cU} |p| \leq \pth{|\redH_r|+2\sum_{p \in \cU} |p|}/20+|r| \Rightarrow 
- \sum_{p \in \cU} |p|\leq \frac{10}{9} \pth{|\redH_r|/20+|r|} \leq 
- \frac{10}{9}\pth{|H_r|/20+\sqrt{|H_r|}/100} \leq |H_r|/10.
-\]
-
-We now prove that for any $q$ parent of $q'$ (other than $r$), $W_q \geq (4/3)W_{q'}$.
-Suppose not. Let $p, p'$ be the counterexample furthest from the root,
-where $p$ is the parent of $p'$. 
-Note that for $q$ and child $q'$ in $\cU_{p'}$, $W_q \geq (4/3)W_{q'}$.
-We will apply \Clm{charge} for $\cU_{p'}$ to deduce that $\sum_{q \in \cU_{p'}} |q| \leq W_{p'}/20$.
-Combining this with \Clm{loss} gives, $|\redH_{p'}| \geq 19W_{p'}/20$. By \Lem{geometric}, $|\redH_p| \geq (3/2)|\redH_{p'}|$.
-Noting that $W_p \geq |\redH_p|$, we deduce that $W_p \geq (4/3)W_{p'}$.
-Hence, $p, p'$ is not a counterexample, and more generally, there is no counterexample.
-That completes the whole proof.
-\end{proof}
-
-\section{Lower Bound by Path Decomposition}
-In this section we provide a lower bound for the time it takes to compute a contour tree in terms the cost of path decompositions.  Specially, 
-we first consider the easier case of merge trees, after which the lower bound for more general contour trees will easily follow.
-
-\subsection{Merge Trees}
-In this section we show that any algorithm for computing the merge tree $T$ must pay at least $f_{path}(P(T))$ (see \Def{path}) for any path decomposition $P(T)$.
-
-%We say two shrubs $S_1$ and $S_2$ (of merge trees of potentially different surfaces) are equivalent if there is an isomorphism, $f$,
-%between the two shrubs such that for any node $v\in S_1$ corresponding to a path of size $|v|$, $f(v)$ also corresponds to a path of size $|v|$.
-%For the purposes of the next lemma, we say two $2$-manifolds are different, if their corresponding contour trees are the different.
-
-In the following note that two distinct PL surfaces can have the same graph structure defining their meshes, but different height values associated with their vertices.
-
-\begin{figure}[h]\centering
-    \includegraphics[width=.15\linewidth]{figs/diamond}\hspace{2cm}
-    \includegraphics[width=.125\linewidth]{figs/topView}\hspace{2cm}
-    \includegraphics[width=.095\linewidth]{figs/sideView}
-    \caption{From left to right: the cone portion of a diamond; the top down view and then the side view of a cone face with a platform added.}
-    \label{fig:diamond}
-\end{figure}
-
-\begin{lemma}
-\label{lem:pathDecomp}
-Let $T$ be a merge tree, and let $P = P(T)$ any path decomposition of $T$.  %, and $\pathTree$ the corresponding shrub.  
-Then there is a set of at least $\Pi_{p_i\in P} (|p_i|-1)!$ distinct PL $2$-manifolds, all of which have the same defining mesh.
-\end{lemma}
-\begin{proof}
- We describe a simple mesh, with heights specified for all vertices except for the saddles.  The main idea is that 
- this mesh will be flexible enough to allow almost any assignment of heights to the saddles, thus defining a class of merge trees 
- whose size is that described in the lemma statement.
-
- First, we describe the basic building block of this input mesh, which roughly speaking corresponds to a fixed path $p\in P$.  
- Specifically, we say a \emph{diamond} of size $m$ is the following mesh.  To start, construct a cycle 
- of length $m$, where the height of the $i$th vertex in the cycle is say $i/2m$.  Next, create an apex vertex at height 
- $\infty$ (or any appropriately large finite height), and add an edge between this vertex and each vertex in the cycle.  At this point 
- we have built an upside down cone with $m$ triangular faces (see \Fig{diamond}). To make this cone a closed surface we add a triangular cap at the bottom. 
- Specifically, add a $3$ cycle of vertices at heights $0$, $1/20m$, and $1/10m$, and connect each vertex of this $3$-cycle to an arbitrary distinct 
- vertex on the $m$ cycle (with matching cyclic ordering), and triangulate all remaining faces.  For consistency, if the size $m$ is $\leq 3$, the diamond 
- will be just a tetrahedron.
-
- Now above each face of the cone of the diamond, we will create a triangular ``platform'', which will act as the location to glue on other (shrunken) diamonds, 
- and whose highest point of connection to the face below will be a saddle vertex.  Specifically the heights of the vertices of the triangle will be $h$, $h-\eps$, and $h-2\eps$, where 
- $h$ is some unspecified height value.  This near horizontal platform is then connected to its respective face with vertical edges down to the face. See \Fig{diamond}.
- %
- %The vertex at height $h$ will lie in the plane of the respective face of the cone, and the other two will lie above the face, 
- %with edges connecting down to the face, hence creating an actual ``platform''.
-
- At this point we can construct the entire input mesh, by gluing together diamonds.  
- Specifically, for each path $p_i\in P$ we create a diamond of size $|p_i|-1$ (we subtract one because the tip of the diamond already corresponds to the leaf of the path).
- Now we glue together diamonds of different paths in the same way the paths are connected in the shrub $\pathTree$.  Specially, for two paths $p,q\in P$ where 
- $p$ is the parent of $q$ in $\pathTree$, we glue the bottom triangular cap of the diamond for $q$ onto the triangular ``platform'' of some face of the diamond for $p$.
- (Naturally for this construction to work, diamonds for a given path will be scaled down relative to the size of the diamond of their parent).
-
- Observe now that the only saddle points in this construction are the highest connecting point on a face of an edge from the platform above.  Moreover, the only maxima are the apexes of the 
- diamonds, and there is exactly one minima corresponding to the bottom cap of the diamond representing the root of $\pathTree$.  Therefore, the saddles on a given 
- diamond will appear contiguously on a root to leaf path in the merge tree of the mesh, where the leaf corresponds to the maxima of the diamond 
- (since all these saddles have a direct line of sight to this apex).  In particular, 
- this implies that, regardless of the heights assigned to the platforms, the merge tree of this mesh has a path decomposition whose corresponding shrub is equivalent to $\pathTree$.
-
- Now there is a valid instantiation of this described mesh for any relative ordering of the heights of the saddles on a given diamond.  In particular, 
- there are $(|p|-1)!$ possible orderings of the heights of the saddles on the diamond for $p$, and hence $\Pi_{p_i\in P} (|p_i|-1)!$ possible instantiations of 
- the mesh overall.  Also, each one of these instantiations will result in a different merge tree.  Specifically, a permutation of the heights on a given diamond 
- corresponds to a permutation of the vertices along a path in $P$, as by construction all saddles on a given diamond will appear in sorted order in the merge tree.
-\end{proof}
-
-\begin{corollary}
- \label{cor:mergeLB}
- Let $A(S)$ be any comparison based algorithm to compute the merge tree of any input surface $S$ (whose contour tree is a merge tree).  
- Then the running time of $A(S)$ is $\Omega(f_{path}(P))=\Omega(\sum_{p_i\in P} |p_i|\log|p_i|)$ where $P$ is any path decomposition of the 
- merge tree of $S$.
-\end{corollary}
-\begin{proof}
- Given an input surface $S$, by the above lemma, any path decomposition $P$ of its merge tree defines a set of $\Pi_{p_i\in P} |p_i|!$ surfaces 
- with distinct merge trees, all of which have the same mesh.  Specifically, for each sorted ordering of the relative height values of the saddles 
- on each diamond in the mesh, yields a different merge tree.
- Therefore, any comparison based algorithm which computes the merge tree must distinguish between these possibilities, and therefore its corresponding decision tree must 
- have $\Pi_{p_i\in P} |p_i|!$ leaves.  Hence by Stirling's approximation, this decision tree must have depth $\Omega(\sum_{p_i\in P} |p_i|\log|p_i|)$.
-\end{proof}
-
-\subsection{Contour Trees}
-We first generalize previous terms to the case of contour trees.
-In this section $T$ will denote an arbitrary contour tree.  Specifically, every internal vertex in $T$ has degree $3$ 
-and the height function defined over the vertex set is such that every internal vertex in either a split or merge 
-(i.e. at most $2$ neighbors lie above or below it).
-
-Again we define a path decomposition as a partition of the vertices into 
-disjoint leaf paths (i.e. each path is monotonic in height and contains a leaf). 
-However, for simplicity we now restrict our attention to path decompositions which 
-roughly speaking are consistent with the raining procedure described in \Sec{rain} 
-(more general decompositions can work, but it is not needed for our purposes).
-
-\begin{definition}
-\label{def:path2} A path decomposition, $P(T)$, is called \emph{rain consistent} if its paths can be obtained as follows.
-Perform an upward BFS from an arbitrary minima $v$ in $T$, and mark all vertices encountered.  
-Now recursively run a directional BFS from all vertices adjacent to the current marked set.  
-Specifically, for each BFS run, make it an upward BFS if it is at an odd height in the recursion tree and downward otherwise.
-
-This procedure partitions the vertex set into disjoint rooted subtrees of $T$, based on which BFS marked a vertex.  
-For each such subtree, now take any partition of the vertices into leaf paths.\footnote{Note that the subtree  
-of the initial vertex is rooted at is a minima.  For simplicity we require that the path this vertex belongs to 
-also contains a maxima.}
-\end{definition}
-
-The following is analogous to \Lem{pathDecomp}, and in particular uses it as a subroutine.
-
-\begin{lemma}
-\label{lem:pathDecomp2}
-Let $T$ be a contour tree, and let $P = P(T)$ be any rain consistent path decomposition of $T$.  
-Then there is a set of at least $\Pi_{p_i\in P} (|p_i|-1)!$ distinct PL $2$-manifolds, all of which have the same defining mesh.
-\end{lemma}
-\begin{proof}
- As $P$ is rain consistent, the paths can be partitioned into sets $P_1, \dots, P_m$, where $P_i$ is the set of all paths with 
- vertices from a given BFS, as described in \Def{path2}.  Specifically, let $T_i$ be the subtree of $T$ corresponding to $P_i$ and 
- let $r_i$ be the root vertex of this subtree.  
- Note that the $P_i$ sets naturally define a tree where $P_i$ is the parent of $P_j$ if $r_i$ is adjacent to a vertex in $P_j$.
-
- As the set $P_i$ is a path decomposition of a rooted binary tree $T_i$, the surface mesh construction of \Lem{pathDecomp} for $P_i$ is well defined.  
- Actually the only difference is that here the rooted tree is not a full binary tree, and so some of the faces of the constructed diamonds will be blank.  
- Specifically, these blank faces correspond to the adjacent children of $P_i$, and they tell us how to connect the meshes of the different $P_i$'s.
-
-\begin{figure}[h]\centering
-    \includegraphics[width=.5\linewidth]{figs/bridge2}
-    \caption{On the left, the root of $P_i$ merges into a diamond of its parent $P_j$.  On the right, the root of $P_i$ splits off from a diamond of $P_j$}
-    \label{fig:diamond}
-\end{figure}
-
- So for each $P_i$ construct a surface mesh as described in \Def{path2}.  By construction, the diamonds corresponding to the paths in $P_i$ are connected 
- into a tree structure (i.e. corresponding to the shrub of $P_i$).  Therefore the bottoms of all these diamonds are covered except for the one corresponding 
- to the path containing the root $r_i$.  If $r_i$ is the initial minima that the rain consistent path decomposition was defined from,  
- we add a tetrahedral cap to the uncovered bottom of this diamond (i.e. we construct the minima).  Otherwise, $P_i$ has some parent $P_j$ in which case 
- we connect the bottom of the diamond for $r_i$ to a free face of a diamond in the construction for $P_j$, specifically, the face corresponding to the 
- vertex in $T$ which $r_i$ is adjacent to. 
- If $T_i$ was created by an upward BFS, then the connection to this face should be a merge vertex, and otherwise it should be a split vertex.  
- See \Fig{diamond}.
-
- Just as in \Lem{pathDecomp} we now have one fixed mesh graph structure, such that each different relative ordering of the heights of the merge and split vertices 
- on each diamond produces a surface with a distinct merge tree.  The specific bound on the number of such distinct manifolds follows by applying the 
- bound from \Lem{pathDecomp} to each $P_i$.
-\end{proof}
-
-For the same reason that \Cor{mergeLB} followed from \Lem{pathDecomp}, we now have the following.
-
-\begin{theorem}
- Let $A(S)$ be any comparison based algorithm to compute the contour tree of an input surface $S$.  
- Then the running time of $A(S)$ is $\Omega(f_{path}(P))=\Omega(\sum_{p_i\in P} |p_i|\log|p_i|)$ where $P$ 
- is any rain consistent path decomposition of the contour tree of $S$.
-\end{theorem}
-
-% \section{Temp}
-% Given a contour tree we allow any path decomp, as long as each path contains a maxima or minima (or both).
-% We call a path $p$ in the decomp a min (resp. max) path if it contains a minima (resp. maxima) (note a path can be both).
-% For every path we create a cylinder with $m$ faces for $p$, where $m$ is the number of non-leaf vertices in $p$ (either $|p|-1$ or $|p|-2$).  
-% If $p$ has a maxima we put a top cone cap on the cylinder, and if it contains a manima then put a bottom cone cap (i.e. the cap represents the maxima or minima).
-% 
-% Every edge $e$ in $T$, but not in a path decomp. tells us how to glue together cylinders.  Let $u$ be the upper endpoint of $e$ and $l$ the lower endpoint.
-% 
-% First we ``remove'' all edges $e$ such that $u$ is adjacent to a max path and $l$ adjacent to a min path by joining the free ends of the respective cylinder together.
-% Hence we can assume no such edges exist.  Now we have three cases:
-% 
-% \begin{enumerate}[{Case} 1]
-%  \item 
-% $l$ is adjacent to a min path:\newline
-% Note in this case the path containing $u$ must be a min path.
-% In this case connect the bottom of the cylinder for the path containing $l$ into a face of the cylinder for the path containing $u$.
-% \item
-% $l$ is adjacent to a max path and $u$ is adjacent to a max path:\newline
-% In this case connect the bottom of the cylinder for the path containing $u$ into a face of the cylinder for the path containing $l$.
-% \item
-% $l$ is adjacent to a max path and $u$ is adjacent to a min path:\newline
-% Connect a face of the cylinder for the $u$ path (using a split) to a face of the cylinder for the $l$ path (using a merge).
-% \end{enumerate}
-
-\bibliographystyle{alpha}
-\bibliography{contour}
-
-\appendix
-
-\section{Old stuff}
-
-\subsubsection{Paying it down the line}
-
-% \begin{claim}
-% \label{clm:sizes}
-%  Let $p$ be a path in a maximum path decomposition, such that $|H_p|>c|p|$ for some constant $c> 6$.
-%  Let $H_f$ denote the heap at the leaf of $p$.  We then have the following
-%  \begin{compactenum}
-%   \item $|H_f|\leq (1+1/c)|H_p|$
-%   \item $|H_f|\geq |H_p|/3$
-%   \item For any path $q$ adjacent to $p$ in the shrub, $|H_q|<\frac{2}{3}|H_p|$
-%  \end{compactenum} 
-% \end{claim}
-% \begin{proof}
-%   For any vertex $v\in p$ \Obs{decrease} implies that $|H_v|\leq |H_p| + |p| < (1+1/c) |H_p|$.
-%   First observe that by setting $v=f$ this immediately implies the first part of the claim.
-%  % By a similar argument, $|H_f|\geq |H_p| - |p| > (1-1/c)|H_p|$.
-%   
-%   To prove the last part of the claim, suppose otherwise that $|H_q|\geq \frac{2}{3}|H_p|$ for some 
-%   adjacent path $q$.  As this is a maximum path decomposition, this implies that there is some 
-%   $v\in p$ such that $|H_v|\geq \frac{4}{3}|H_p|-1$.
-%   By \Obs{decrease} we then have, 
-%    $|H_p| \geq |H_v| - |p|\geq \frac{4}{3}|H_p| - 1 - |H_p|/c \geq (4/3-1/c)|H_p|-1$.
-%   When rearranged this gives $1\geq (1/3-1/c)|H_p|$.  As $c>6$ we have $1\geq |H_p|/6$, 
-%   which is a contradiction as $|H_p|>c|p|>6$.
-% \end{proof}
-
-\Ben{I had originally thought the following claim would be useful in proving the lemma below.}
-\begin{claim}
-\label{clm:sizes}
- Let $p$ be a path in a maximum path decomposition, such that $|H_p|>c|p|$ for some constant $c> 6$.
- We then have the following
-% \begin{compactenum}
-%  \item 
-  For any path $q$ adjacent to $p$ in the shrub, $|H_q|<\frac{2}{3}|H_p|$
-%  \item If there does not exist and adjacent path $q$ such $H_q\geq H_p/3$, then $|H_f|\geq |H_p|/3$
-% \end{compactenum} 
-\end{claim}
-\begin{proof}
-  To prove the claim, suppose otherwise that $|H_q|\geq \frac{2}{3}|H_p|$ for some 
-  adjacent path $q$.  As this is a maximum path decomposition, this implies that there is some 
-  $v\in p$ such that $|H_v|\geq \frac{4}{3}|H_p|-1$.
-  By \Obs{decrease} we then have, 
-   $|H_p| \geq |H_v| - |p|\geq \frac{4}{3}|H_p| - 1 - |H_p|/c \geq (4/3-1/c)|H_p|-1$.
-  When rearranged this gives $1\geq (1/3-1/c)|H_p|$.  As $c>6$ we have $1\geq |H_p|/6$, 
-  which is a contradiction as $|H_p|>c|p|>6$.  
-\end{proof}
-
-\begin{lemma}
-\label{lem:excess}
- Let $p$ be a tall path, and let $S_1, \dots S_k$ and $R_1, \dots, R_l$ be the sizes of the heaps at the roots of the short and tall children of $p$, respectively, in the shrub.  
- Let $H_f$ be the heap at the leaf of the path $p$.
- Then,  $|H_p| \log |H_p|   \geq  \sum_{v\in p} \log|H_v| + \frac{4}{3}\sum_{i=1}^k S_i \log S_i + \sum_{i=1}^l R_i \log R_i  -  c|p|\log |p|$, for some sufficiently large constant $c$
- \end{lemma}
-\begin{proof}
- First observe that the lemma holds for $|H_p|$ less than any fixed constant, since otherwise \Lem{adjacent} implies we can choose $c$ 
- large enough such that $c|p|\log|p|$ will dominate all other terms.
- Next, observe that $\sum_{v\in p} \log|H_v| \leq \sum_{v\in p} \log(|H_v|+|p|) = |p|\log(|H_v|+|p|) = O(|p|\log|p|)$, as $p$ is a tall path.  
- Therefore, in the remainder of the proof  we can safely ignore the $\sum_{v\in p} \log|H_v|$ term as we can choose $c$ large enough to cover it.  
- That is we now instead prove the inequality,
- \[
-  |H_p| \log |H_p|   \geq \frac{4}{3}\sum_{i=1}^k S_i \log S_i + \sum_{i=1}^l R_i \log R_i  -  c|p|\log |p|
- \]
- To do so we  break the analysis into two case based on the relative size of $|p|$ and $|H_p|$. 
- In the following, $\alpha>10$ is some yet to be determined constant.
-
-\paragraph{Case 1, $|p|\geq |H_p|/\alpha$:\\}
-First observe that by \Lem{adjacent}, $|H_p|+|p| \geq \sum_{i=1}^k S_i+ \sum_{i=1}^l R_i$.
-By choosing $c$ large enough we therefore have,
-\begin{align*}
- c|p|\log|p| &\geq 2(1+\alpha)^2 |p|\log |p| \geq 2(1+\alpha)|p| \log((1+\alpha)|p|)\\
-             &\geq 2(|H_p|+|p|)\log(|H_p|+|p|) \geq 2\log(|H_p|+|p|) \pth{\sum_{i=1}^k S_i + \sum_{i=1}^l R_i}\\
-             &\geq \frac{4}{3}\sum_{i=1}^k S_i \log S_i + \sum_{i=1}^l R_i \log R_i 
-\end{align*}
-
-In this case $\log(|H_p|+|p|)\leq \log(2|p|) \leq 1+\log(|p|)$ (where $\log$ is base $2$).
-Therefore, assuming $|p|>1$, we have 
-\[
-(|H_p|+|p|)\log(|H_p|+|p|)\leq (1+c)|p|\log((1+c)|p|) %(|H_p|+|p|)(1+\log |p|) \leq 2(1+c)|p|\log |p|
-\]
-which when combined with the inequality from \Lem{adjacent} gives, 
-\[
- 4(1+c)|p|\log |p| \geq 2(|H_p|+|p|)\log(|H_p|+|p|)\geq \pth{|H_f| + \frac{4}{3}\sum_{i=1}^k S_i + \sum_{i=1}^l R_i}\log(|H_p|+|p|).
-\]
-This easily implies the desired inequality as $\log(H_p+p)$ is larger than any other $\log$ term in the inequality.
-
-\paragraph{Case 2, $|p|< |H_p|/\alpha$:\\}
-
-\Ben{Was not able to handle this case.}
-
-%  First, lets assume we can show $\sum_{i=1}^k S_i + \sum_{i=1}^l R_i \leq \frac{3}{4}|H_p|$.  
-%  In this case, using the last part of \Clm{sizes}, we can prove the desired inequality as follows.
-%  
-%  \begin{align*}
-%   \frac{4}{3}\sum_{i=1}^k S_i \log S_i + \sum_{i=1}^l R_i \log R_i  &\leq  \frac{4}{3} \pth{\sum_{i=1}^k S_i \log S_i + \sum_{i=1}^l R_i \log R_i} \\
-%   &\leq  \frac{4}{3} \log |H_p| \pth{\sum_{i=1}^k S_i + \sum_{i=1}^l R_i}  \leq  |H_p| \log |H_p|
-%  \end{align*}
-%  
-%  Finally, to complete the proof we must prove our assumption that $\sum_{i=1}^k S_i + \sum_{i=1}^l R_i \leq \frac{3}{4}|H_p|$.
-%  By \Lem{adjacent} and \Clm{sizes} we have, 
-%  \[
-%  \sum_{i=1}^k S_i + \sum_{i=1}^l R_i 
-%  \leq |H_p|+|p|-|H_f| 
-%  \leq (1+1/\alpha-1/3)|H_p| = (2/3+1/\alpha)|H_p|
-%  \]
-%  
-%  Therefore $\sum_{i=1}^k S_i + \sum_{i=1}^l R_i \leq \frac{3}{4}|H_p|$ if $(2/3 + 1/\alpha)\leq 3/4$, which rearranged gives $\alpha \geq 12$
-\end{proof}
-
-\remove{
-% \subsubsection{Paying it down the line}
-% 
-% \begin{lemma}
-% \label{lem:excess}
-%  Let $p$ be a tall path, and let $S_1, \dots S_k$ and $R_1, \dots, R_l$ be the sizes of the heaps at the roots of the short and tall children of $p$, respectively, in the shrub.  
-%  Let $H_f$ be the size of the heap at the leaf of the path $p$, and let $X = \sum_{i=0}^k H_i \log H_i$.  
-%  Then,  $|H_p| \log |H_p|   \geq  H_f\log H_f + \sum_{i=0}^k S_i \log S_i + \sum_{i=0}^l R_i \log R_i  -  c|p|\log |p|$, for some sufficiently large constant $c$
-%  \end{lemma}
-% \begin{proof}
-% By \Lem{adjacent} we know that $|H_p|+|p| \geq  H_f + \sum_{i=0}^k S_i + \sum_{i=0}^l R_i$.  
-% %Now as $p$ is a tall path we have $|p|\geq \sqrt{|H_p|}/100$, and so, $(|H_p|+|p|)\log(|H_p|+|p|)$
-% 
-% We  break the analysis into two case based on the relative size of $|p|$ and $|H_p|$.  In the following $c$ is some yet to be determined constant greater than $1$.
-% 
-% \paragraph{Case 1, $|p|\geq |H_p|/c$:}  
-% In this case $\log(|H_p|+|p|)\leq \log(2|p|) \leq 1+\log(|p|)$ (where $\log$ is base $2$).
-% Therefore, assuming $|p|>1$, we have 
-% \[
-% (|H_p|+|p|)\log(H_p+p)\leq (|H_p|+|p|)(1+\log |p|) \leq (2|p|)(2\log |p|)\leq 4|p|\log |p|
-% \]
-% which when combined with the inequality from \Lem{adjacent} gives, 
-% \[
-%  4|p|\log |p| \geq (|H_p|+|p|)\log(|H_p|+|p|)\geq \pth{H_f + \sum_{i=0}^k S_i + \sum_{i=0}^l R_i}\log(|H_p|+|p|).
-% \]
-% This easily implies the claim as $\log(H_p+p)$ is larger than any other $\log$ term in the inequality.
-% 
-% \paragraph{Case 2, $|p|< |H_p|/c$:}
-% \end{proof}
-}
-
-Define the crust of the shrub to be the first (i.e. closest to the root) layer of short paths in the shrub.  Apply \Lem{shortCost} to each of the short paths in the crust and then 
-say by \Lem{excess} and since the root heap goes to zero we are done.
-
-\ignore{
-Naturally, this geometric series behavior can be extended to more than two adjacent short paths, thus motivating 
-the following definition.
-
-\begin{definition}
- Let $p$ be a path in a path decomposition $P(T)$, with corresponding shrub $\pathTree$.
- Call a short path $q\in P(T)$ short reachable from $p$ if the path from $p$ to $q$ in $\pathTree$ 
- contains only vertices which correspond to short paths.  Call the subtree of $\pathTree$ consisting 
- of all short reachable paths from $p$ the \emph{short shrub} of $p$, and denote it $\mathcal{L}(p)$.
-\end{definition}
-
-In the following subsection we show how to exploit this geometric series behavior of short shrubs.
-
-We say two paths in $P(T)$ are adjacent if their corresponding vertices are adjacent in the shrub.  Similarly, for 
-two adjacent paths in $P(T)$ we can define a parent/child relationship based on this relationship in the shrub. 
-}
-
-\remove{
-\Ben{Maybe move the remainder of this subsection to 5.3 since tall and short paths and not used until then}
-
-\begin{definition}
- Let T be a binary tree and let $\chi(T)$ and $P(T)$ be a valid coloring and path decomposition of $T$, respectively.  
- For $p\in P(T)$, we say that $p$ is \emph{short} if $|p| < \sqrt{|H_{r_p}|}/100$, and we say it is \emph{tall} otherwise.
-\end{definition}
-
-\begin{observation}
-\label{obs:decrease}
- Let $v$ be a vertex in $T$ and let $w$ be its parent.  Then $|H_w| \geq |H_v| -1$, as $L(T_v) \subseteq L(T_w)$ and the path from 
- $w$ to the root has one less vertex than the path from $v$ to the root.
-
- In particular, we have the following more general property.
- Let $v$ and $u$ be any two vertices in the same root to leaf path of $T$, such that $u$ is lower than $v$.  
- Then $|H_v| \leq |H_u| + d(u,v)$. 
-\end{observation}
-
-\begin{lemma}
-\label{lem:pathBounds}
- Let T be a binary tree and let $\chi(T)$ and $P(T)$ be a valid coloring and path decomposition of $T$, respectively.
- Then, 
- \begin{compactenum}
-  \item for any tall path $p\in P(T)$, $\sum_{v\in p} \log |H_v| = O(|p| \log |p|)$.
-  \item for any short path $p\in P(T)$, $\sum_{v\in p} \log |H_v| = O(|p| \log |H_{r_p}|)$.
- \end{compactenum}
-\end{lemma}
-\begin{proof}
- By $\Obs{decrease}$ we know that for any vertex $v\in p$, $|H_v|\leq |H_{r_p}| + |p|$ (as $r_p$ lies below $v$ on $p$).
- As such, 
- \[
- \sum_{v\in p} \log |H_v| \leq \sum_{v\in p} \log(|H_{r_p}| + |p|)  = |p| \log (|H_{r_p}| + |p|).
- \]
- If $p$ is a tall path then $|p| \log (|H_{r_p}| + |p|) = O(|p| \log |p|)$, and if $p$ is short then 
- $|p| \log (|H_{r_p}| + |p|) = O(|p| \log |H_{r_p}| )$.
-\end{proof}
-
-The following lemma can be thought of as a generalization of $\Obs{decrease}$.
-
-\begin{lemma}
-\label{lem:adjacent}
- Let $T$ be a binary tree and let $\chi(T)$ and $P(T)$ be a valid coloring and path decomposition of $T$, respectively.
- Let $p$ be any path in $P(T)$ and let $\{q_1, \dots q_k\}$ be the set of all adjacent child paths of $p$.
- For $1\leq i\leq k$ let $r_i = r_{q_i}$ and let $H_i = H_{r_i}$.
- Also, let $H_0$ denote the heap at the leaf of $p$.
-
- Then $\sum_{i=0}^k |H_i| \leq |H_p|+|p|$, where $H_p = H_{r_p}$.
-\end{lemma}
-\begin{proof} 
- First observe that for all $i$, $r_{i}$ lies above $r_p$ on some root to leaf path. 
- Let $H_i'$ be the subset of $H_i$ that lies below $r_p$.  Observe that $H_i'\subseteq H_p$, as $L(T_{r_i})\subseteq L(T_{r_p})$.
-
- Now for any $i\neq j$, $H_i \cap H_j = \emptyset$, as $r_{i}$ and $r_{j}$ are not on the same root to leaf path (i.e. $L(T_{r_i}) \cap L(T_{r_j}) = \emptyset$).
- In particular, $(H_i\setminus H_i') \cap (H_j \setminus H_j') = \emptyset$, and so $\sum_{i=0}^k |H_i\setminus H_i'| \leq |p|$ 
- as $H_i\setminus H_i'$ is precisely the subset of $H_i$ that lies on $p$.
- Moreover, for any $i\neq j$, $ H_i' \cap H_j' = \emptyset$, and so since $H_i'\subseteq H_p$ we have $\sum_{i=0}^k |H_i'|\leq |H_p|$.
- Putting these together gives
- \[
-  \sum_{i=0}^k |H_i| = \sum_{i=0}^k |H_i'| + \sum_{i=0}^k |H_i\setminus H_i'| \leq |H_p|+|p|.
- \]
-\end{proof}
-
-% \Ben{There is a fundamental flaw in the following lemma and corollary.  
-% As such, the remainder of this subsection will likely be deleted.}
-% 
-% Using the notation defined in the above lemma, we have the following crucial lemma, which we will use to 
-% argue we can charge that cost of all paths in $P(T)$ to the cost of the tall paths.
-% 
-% \begin{lemma}
-% Let $p$ be a tall path in $P(T)$ and let $\{q_1, \dots q_k\}$ be any set of short paths in $P(T)$ such that 
-% $\sum_{i=1}^k |H_i| = O(H_p)$.  Then $\sum_{i=1}^k \sum_{v\in q_i} \log |H_v| = O(|p|\log |p|)$
-% \end{lemma}
-% \begin{proof}
-% By \Lem{pathBounds}, for all $1\leq i\leq k$, 
-%  $\sum_{v\in q_i} \log |H_v| = O(|q_i|\log |H_i|) = O(|H_i|\log |H_i|)$, as $q_i$ is a short path.
-%  By assumption $\sum_{i=1}^k |H_i| = O(|H_p|)$, 
-%  As $x\log x$ is a convex function, Jensen's inequality then implies 
-%  $\sum_{i=1}^{k} |H_i|\log |H_i| = O(|H_p| \log |H_p|) = O(|p|\log |p|)$, since $p$ is a tall path.
-% \end{proof}
-% 
-% 
-% \begin{corollary}
-% \label{cor:tallLight}
-%  Let $p$ be a tall path in $P(T)$ and let $\{q_1, \dots q_k\}$ be the set of all short paths adjacent to $p$.
-%  Then $\sum_{i=1}^k \sum_{v\in q_i} \log |H_v| = O(|p|\log |p|)$.
-% \end{corollary}
-% \begin{proof}
-%  Let $H_i$ be as defined in \Lem{adjacent}.  By \Lem{pathBounds}, for all $1\leq i\leq k$, 
-%  $\sum_{v\in q_i} \log |H_v| = O(|q_i|\log |H_i|) = O(|H_i|\log |H_i|)$, as $q_i$ is a short path.
-%  Now by \Lem{adjacent}, $\sum_{i=1}^k |H_i| \leq |H_p|+|p| = O(|p|)$, as $p$ is a tall path. 
-%  Therefore, as $x\log x$ is a convex function, Jensen's inequality implies the claim.
-%  
-% \end{proof}
-
-}
-
-\subsection{The shrub of unsupported paths} \label{sec:short}
-
-It suffices to focus on unsupported short paths. \Lem{short} can be reduced to the following.
-
-\begin{lemma} \label{lem:support} $\sum_{p \in \cL \setminus \cL'} \sum_{v \in p} \log |H_v| = O(\sum_{p \in \cL'} |H_p| \log |H_p|)$.
-\end{lemma}
-
-We now construct the shrub of unsupported short paths. Consider $p \in \cL \setminus \cL'$,
-and traverse the chain of ancestors from $p$. Eventually, we must reach another short path $q$.
-(If not, we have reached the root $r$ of $\pmax(T)$. Hence, $p$ is supported.)
-Insert edge from $p$ to $q$, so $q$ is the parent of $p$ in $\cU$. This construction leads
-to the shrub forest of $\cL \setminus \cL'$, where every unsupported short path is an internal node, and all
-roots are supported short paths.
-
-We list out some important claims about this shrub forest.
-We focus on $\cU$, a connected component (shrub) in the shrub forest and denote the 
-the root by $r(\cU)$.
-
-\begin{claim} \label{clm:loss2} Consider $p \in \cL \setminus \cL'$ that is in $\cU$. Let $Q$
-be the set of tall ancestors of $p$ that are descendants of $r(\cU)$. Then $\sum_{q \in Q} |q \cap H_p| \leq |H_p|/50$.
-\end{claim}
-
-\begin{proof} Suppose not, for contradictions sake. As we traverse the tall ancestors of $p$,
-we encounter various vertices of $H_p$ in the ancestors $q$. Consider the first $|H_p|/100$
-such vertices of $p$ encountered. At each such $v$, there are at least $|H_p|/100$
-vertices \emph{not yet} encountered, so $|H_v| > |H_p|/100$. Each of these vertices $v$
-is supported, implying that $p$ is a supported short path.
-\end{proof}
-
-\Lem{support} is a direct consequence of the following,
-which is the really the main lemma.
-
-%\begin{lemma} \label{lem:cu} Let $\cU$ denote a connected component (shrub) in the shrub forest of $\cL \setminus \cL'$ and let $r$
-%be the root of $\cU$. Then $\sum_{p \in \cU} \sum_{v \in p} \log |H_v| = O(|H_r| \log |H_r|)$.
-%\end{lemma}
-%
-%We proceed with the setup to prove \Lem{cu}.
-%We focus our attention on some shrub $\cU$.
-%We prove a technical lemma about $\cU$, from which \Lem{cu} follows directly.
-
-\section{Ben's old stuff}
-
-\begin{proof} (part (ii) of \Lem{cu-root}) Let level $\ell$ of $\cU$ denote all nodes at distance $\ell$ from the root $r$.
-These set of these nodes is designated as $\cU_\ell$. We set $\sigma_\ell = \sum_{p \in \cU_\ell} |p|$
-and $\sigma_{\leq \ell} = \sum_{k \leq \ell} \sigma_k$.
-By \Lem{geometric}, for any $p \in \cU_\ell$, $|H_p| \leq (2/3)^\ell |H_r|$.
-
-Since all nodes in $\cU$ represent short paths, $\forall p \in \cU$, $|p| < \sqrt{|H_p|}/100$.
-So $\sigma_\ell < \sum_{p \in \cU_\ell} \sqrt{|H_p|}/100$.
-
-\end{proof}
-
-To prove this, we need to understand how $H_p$ is constructed. The path $p$ has some children
-$q_1, q_2, \ldots, q_k$ in $\pmax(T)$. Furthermore, as a leaf path in $p$, some leaf
-$\ell$ of $T$ in one of the ends of $p$. The heap $H_p$ is essentially the
-union of $H_\ell$ and $H_{q_1}, H_{q_2}, \ldots, H_{q_k}$, with the vertices in $p$
-removed. Note that some of the $q_i$'s are the corresponding $h(q')$ vertices,
-for a child $q'$ in $\cU$.
-
-Let $\mathcal{L} = \mathcal{L}(r)$ be a short shrub of some short path $r$ in a maximum path decomposition $\mathcal{P}(T)$.
-The intuitive idea is that due to the geometric decreasing behavior of short shrubs described in the previous subsection, 
-one can show that for the entire short shrub our algorithm pays roughly the cost due to the root path $r$.  This argument 
-is hard to make directly, so instead we define an alternative quantity for which we can make the argument.
-Specifically, we now define a weight function $w(v)$ for all $v\in \mathcal{L}$, and show that $w(v)=\Theta(|H_v|)$.
-
-For a path $v\in \mathcal{L}$, let $H_v$ denote the heap at the root of $v$, and let $H_{v_f}$ denote the heap 
-at the leaf of $v$.  Let $S_l(v)$ (resp. $S_h(v)$) be the subset of short (resp. tall) paths adjacent to $v$ in $\pathTreeA(T)$.
-
-We want to understand how the heap size changes as we move around the paths in $\mathcal{L}$.  Specifically, there are 
-three places the vertices in the heap along a short path can come from; adjacent short paths, adjacent tall paths, and the initial heap at the leaf of the path.  
-Given this observation we now define a weight function which upper bounds the size of the heap at the root of any path in $\mathcal{L}$.
-
-First, for all $v\in \mathcal{L}$ we define the local cost function $c(v) = H_{v_f}+\sum_{i\in S_h(v)} H_i$, this is the addition to the heap at $v$ which 
-was not already charged to descendant short paths.  We now define a weighted version of $\mathcal{L}$ based on these costs.  
-To do so, we first augment $\mathcal{L}$ by attaching one 
-leaf to each vertex in $\mathcal{L}$.  Now for each newly created leaf (which are the only leaves now) assign the cost 
-$c(v)$ defined above, where $v$ is that leaf's corresponding parent.  
-Now we can define a weight function $w(u)$ over any node $u$ in the tree, which is defined to be the total cost of all leaves in its subtree.   
-To simplify notation, we now use $\mathcal{L}$ to refer to this new weighted tree (rather than its unweighted counterpart).
-
-As the weight $w(p)$ includes the sizes of the heaps of all contributing tall paths and leaves in the subtree rooted at $p$, we have the following.
-
-\begin{claim}
-For any $p\in \mathcal{L}$, $w(p)\geq |H_p|$.  Moreover, for any $v\in p$, $w(p)\geq |H_v|$.
-\end{claim}
-
-%Let $W= \sum_{v\in \mathcal{L}} w(v) \geq \sum_{v\in \mathcal{L}} |H_v|$.  
-%Suppose we could show that conversely $W=O(H_r)$.  Then we would have 
-%$\sum_{v\in \mathcal{L}} H_v = O(H_r)$, which was our original goal.  
-%So proving $W=O(H_r)$ is now our new goal.
-
-We now would like to show conversely that $w(v)= O(|H_v|)$.  
-To do so we first introduce some notation.  
-Let $C\subseteq \mathcal{P}$ be the subset of paths that are leaves in $\mathcal{L}$.
-For any path $v$, let $\mathcal{L}_v$ (resp. $C_v$) be the subset of paths in $\mathcal{L}$ (resp $C$) 
-that are in the subtree of $\mathcal{L}$ which is rooted at $v$.
-
-First, observe that $|H_v|\geq w(v)-\sum_{u\in \mathcal{L}_v} |u|$.  
-This holds by the same argument as in \Lem{adjacent}, namely, $w(v)$ by definition 
-counts all the contributing tall path heaps and initial leaf heaps that contribute 
-to the heap at $v$ and $\sum_{u\in \mathcal{L}_v} |u|$ is an upper bound on the number 
-of vertices that could have been removed from one of these heaps before reaching the 
-root of $v$ (since a vertex in a heap can only get removed if it appears on one of 
-the paths).
-
-As all paths in $\mathcal{L}$ are short, we thus have the following.
-
-\begin{claim}
-$|H_v|\geq w(v)-\sum_{u\in \mathcal{L}_v} \sqrt{|H_u|}/100 \geq w(v)-\sum_{u\in \mathcal{L}_v} \sqrt{w(u)}/100$
-\end{claim}
-
-By the above claim, if we can prove $\sum_{u\in \mathcal{L}_v} \sqrt{w(u)}/100 \leq w(v)/c$ for some constant $c>1$, 
-then we will have proven that $|H_v| = \Theta(w(v))$.
-
-Recall that \Lem{geometric} told us that that $|H_{u}|\leq (3/5) |H_{u'}|$, if $u'$ is the parent $u$ in $\mathcal{L}$. 
-The following lemma shows that if we had a similar statement in terms of the weights rather than the heap sizes, 
-then we get the desired bound on the sum.
-
-\begin{lemma}
-\label{lem:distribution}
-Let $v$ be a path in $\mathcal{L}$.  Suppose that for any two paths $x,y\in \mathcal{L}_v$ 
-such that $y$ is the parent of $x$, $w(x)\leq (4/5) w(y)$.
-Then $\sum_{u\in \mathcal{P}_v} \sqrt{w(u)}/100 \leq \sum_{c\in C_v} \sqrt{w(c)}/10$
-\end{lemma}
-
-\begin{proof}
-The idea is to recursively redistribute the contribution to the sum of each internal vertex to its children, 
-until all the weight sits on the leaves.  Specifically, let $v$ be some path in $\mathcal{L}$.  Its 
-contribution to the sum is $\sqrt{w(v)}/100$, which we will call the value at $v$.  We redistribute this 
-value to its children, in proportion by the weight of each child.  For example, if $v$ has a child $x$ then 
-the fraction it gets assigned is $w(x)/w(v)$ (recall that by definition $w(v)$ is the sum of the weights of 
-its children and so the total value will get reassigned to the children).
-
-Suppose we keep redistributing the value of $v$ until it reaches the leaves of its subtree in $\mathcal{L}$.  
-Let's see how much gets assigned to a given leaf.  Specifically, let $u_0, u_1, \ldots, u_k$ 
-be a path in $\mathcal{L}$ (of paths from the set $\mathcal{P}_v$), such that $v=u_0$, $u_{i+1}$ 
-is child of $u_i$ for all $i$, and $u_k$ is a leaf of $\mathcal{L}$.
-
-Then the total value from $u_0$ that ends up at $u_k$ is 
-\[
-\frac{\sqrt{w(u_0)}}{100} \pth{ \frac{w(u_1)}{w(u_0)} \frac{w(u_2)}{w(u_1)} \cdots \frac{w(u_k)}{w(u_{k-1})} }
-= \frac{\sqrt{w(u_0)}}{100} \pth{ \frac{w(u_k)}{w(u_0)} } = \frac{w(u_k)}{100\sqrt{w(u_0)}}
-\]
-
-For a given leaf in $\mathcal{L}$, we can calculate the total value reassigned to it from it its ancestors.
-Specifically, let $u_0, u_1, u_2, \dots, u_m$ be a root to leaf path in $\mathcal{L}$ 
-(i.e. $u_0$ is the root $r$ of $\mathcal{L}$).
-
-By assumption from the lemma statement we have $w(u_{i+1})\leq (4/5) w(u_i)$.  Therefore by the above calculation, 
-the total reassigned to $u_m$ is
-
-\[
- \sum_{i=0}^m \frac{w(u_m)}{100\sqrt{w(u_i)}} \leq \frac{w(u_m)}{100} \sum_{i=0}^m 1 \Big/ \sqrt{\pth{\frac{5}{4}}^iw(u_m)} 
- \leq \frac{\sqrt{w(u_m)}}{100} \sum_{i=0}^{\infty} \pth{\frac{4}{5}}^{i/2} \leq \frac{\sqrt{w(u_m)}}{10}
-\]
-
-The claim then follows by summing this quantity over the children of $\mathcal{L}$.
-\end{proof}
-
-By combining the above lemma and claim we have the following.
-
-\begin{claim}
-For any path $v\in \mathcal{L}$ if it holds that 
-$w(x)\leq (4/5) w(y)$ for any two paths $x,y\in \mathcal{L}_v$ such that $y$ is the parent of $x$, then
-
-\[
-|H_v|\geq w(v)-\sum_{u\in \mathcal{L}_v} \sqrt{w(u)}/100 \geq w(v)- \sum_{c\in C_v} \sqrt{w(c)}/10 \geq \frac{4}{5} w(v)
-\]
-\end{claim}
-
-\begin{lemma}
-Let $\mathcal{L}$ be the short shrub of some path $r$.  
-Then for any $v\in \mathcal{L}$, we have $|H_v|\geq \frac{4}{5} w(v)$.
-\end{lemma}
-% \begin{proof}
-% We will actually prove the stronger statement that for any $v\in \mathcal{L}$, 
-% we have $|H_v|\geq \frac{4}{5} w(v)$ and $w(v)\leq (4/5) w(p)$ where $p$ is the parent 
-% of $v$ in $\mathcal{L}$.
-% 
-% Suppose otherwise, and let $v$ be the lowest path in $\mathcal{L}$ such that this fails 
-% (i.e. the statement holds for all descendants of $v$ in $\mathcal{L}$).  Then for any 
-% two paths $x,y\in \mathcal{L}_v$ such that $y$ is the parent of $x$, 
-% we have that $w(x)\leq (4/5) w(y)$ and so by \Lem{distribution}, $|H_v|\geq \frac{4}{5} w(v)$.
-% Combining this fact with \Lem{geometric} gives,
-% \[
-% w(v) \leq \frac{5}{4}|H_v| \leq \frac{5}{4} \frac{3}{5} |H_p| \leq \frac{3}{4} w(p).
-% \]
-% 
-% \end{proof}
-%
-\begin{proof}
-By the above claim, we know that $|H_v|\geq \frac{4}{5} w(v)$ if it holds that 
-$w(x)\leq (4/5) w(y)$ for any two paths $x,y\in \mathcal{L}_v$ such that $y$ is the parent of $x$.
-So let $v$ be the lowest path in $\mathcal{L}$ such that this fractional decrease fails to hold, i.e. $w(c)> (4/5) w(v)$ 
-where $c$ is a child of $v$ in $\mathcal{L}$, and conversely all vertices in $\mathcal{L}_c$ 
-the fractional decrease holds.  As the fractional decrease holds for a paths in $\mathcal{L}_c$, 
-we have $|H_c|\geq \frac{4}{5} w(c)$.  Therefore,
-\[
-w(c) \leq \frac{5}{4}|H_c| \leq \frac{5}{4} \frac{3}{5} |H_v| \leq \frac{3}{4} w(v).
-\]
-However, this is a contradiction with the assumption $w(c)> (4/5) w(v)$ , and hence the 
-fractional decrease always holds implying the claim always holds.
-\end{proof}
-
-\begin{corollary}
-\label{cor:equivalent}
- Let $\mathcal{L}$ be a short shrub of a short path $r$.  Then for any $v\in \mathcal{L}$, $|H_v| \leq w(v) \leq \frac{5}{4} |H_v|$.
-\end{corollary}
-
-Now we are finally ready to make use of the properties we have proven about the weight function we introduced.
-
-\Ben{The constant in the lemmas below, but it is no more than twice that.}
-
-\begin{lemma}
-\label{lem:shortCost}
- Let $\mathcal{L}$ be a short shrub of a short path $r$.  Let $V$ be the set of tall paths from the full shrub which are adjacent to paths in $\mathcal{L}$.  
-  Then $\sum_{p\in \mathcal{L}}\sum_{v\in p} \log |H_v| + \sum_{q\in V} |H_q|\log |H_q| \leq \frac{4}{3} |H_r|\log |H_r|$
-\end{lemma}
-\begin{proof}
-\Ben{
-The following is a proof for $\frac{3}{2}+\eps$, rather than $\frac{4}{3}$.  The discrepancy is due to a corrected calculation error.  
-4/3 can be achieved by tightening constants above, and is left in the lemma statement as it is used later.
-}
-
-Recall that in a proper coloring of a binary tree $T$, the color of any vertex must be the color of one of its descendants.  
-Therefore, any vertex in the short path shrub must appear in the heap of either the root of an adjacent 
-path from the full shrub or from the heap of a leaf of a path in $\mathcal{L}$.  
-Also note $|H_r|$ precisely counts sum of the number of vertices from these heaps that do not appear in $\mathcal{L}$.
-As there are no adjacent short paths to a short shrub, the total number of vertices over all paths in $\mathcal{L}$ is at most 
-$m = \sum_{p\in \mathcal{L}} |H_{p_f}| +\sum_{q\in V} |H_q| - |H_r| = w(r)-|H_r|$, where $H_{p_f}$ is the heap at the leaf of $p$.
-
-Using this observation and \Cor{equivalent} we then have, 
-\begin{align*}
- \sum_{p\in \mathcal{L}}\sum_{v\in p} \log |H_v| + \sum_{q\in V} |H_q|\log |H_q|
- &\leq \sum_{p\in \mathcal{L}}\sum_{v\in p} \log(w(r)) + \sum_{q\in V} |H_q|\log(w(r))\\
- &\leq \log(w(r))\pth{ w(r)-|H_r| } + \log(w(r))w(r)\\ 
- &= (2w(r)-|H_r|)\log(w(r)) \leq \pth{\frac{5}{2}|H_r|-|H_r|}\log\pth{\frac{5}{4}|H_r|}\\
- &= \frac{3}{2}|H_r|\log\pth{\frac{5}{4}|H_r|}
-\end{align*}
-\end{proof}
-
 \end{document}

### 2014-10-22 paint-contour-tree.tex
section 6 proofs

--- 2014-10-21 paint-contour-tree.tex
@@
 \newcommand{\col}{col}
+\newcommand{\fcol}{full}
 \newcommand{\cut}{{\tt cut}}
@@
 \newcommand{\cost}{\mathop{cost}}
-\newcommand{\h}{high}
+\newcommand{\h}{att}
 \newcommand{\lift}{{\tt lift}}
@@
 
-We begin by associating a color with each maximum. If the set of maxima is $S$,
-this is simply an arbitrary bijection $\chi:S \mapsto [|S|]$. Imagine there being a large
-can of paint of color $\chi(y)$ at maxima $y$. We will spill different paint from each maximum and watch it flow down.
+Now we come to the main algorithmic contribution. We design a new algorithm for computing contour
+trees of a minimum dominant simplicial complice $\MM$. As part of the initialization, we determine
+all critical points in $\MM$. For any vertex $v$, we can look at $f(v) \pm \delta$-contours 
+in $f|_{B_\eps(v)}$ to determine the up and down degrees. This determines the criticality (and type)
+of $v$. This requires $O(|\MM|)$ time, since every facet is looked at proportional to the number of its
+vertices.
+
+{\bf Painting:} The central tool is a notion 
+of \emph{painting} of $\MM$. Initially associate a color with each maximum. Denoting the set of maxima is $X$,
+this is an arbitrary bijection $\chi:X \mapsto [|X|]$. Imagine there being a large
+can of paint of color $\chi(x)$ at maxima $x$. We will spill different paint from each maximum and watch it flow down.
 This is analogous to the raining in the previous section, but paint is a much more viscous liquid.
-\emph{So paint only flows down edges, and it does not color the interior of faces.} Furthermore, paints
-do not mix, so eventually, every vertex and edge of $\MM$ gets a unique color.
-%
-\begin{definition} \label{def:paint} A  \emph{painting} of $\MM$ is a map $\chi:S \cup E \mapsto [|S|]$
-with the following properties. (Remember that $G(\MM) = (V,E)$.)
+\emph{So paint only flows down edges, and it does not color the interior of facets.} Furthermore, paints
+do not mix, so eventually, every vertex and edge of $\MM$ gets a unique color. This process (and indeed the entire algorithm)
+works purely on the 1-skeleton of $\MM$, which is just a graph.
+%
+\begin{definition} \label{def:paint} Let the 1-skeleton of $\MM$ be denoted $G(\MM) = (\cV,E)$.
+A  \emph{painting} of $\MM$ is a map $\chi:X \cup E \mapsto [|X|]$ with the following properties. 
 \begin{asparaitem}
-	\item The restriction $\chi:S \mapsto [|S|]$ is a bijection.
+	\item The restriction $\chi:X \mapsto [|X|]$ is a bijection.
 	\item Consider an edge $e$. There exists a descending path from some maximum $x$ to $e$
@@
 %
-Note that paintings are not unique. It is straightforward to construct a painting of $\MM$ in $O(n)$ time, using a descending BFS from each maximum that does not explore previously colored edges.
+Note that paintings are not unique. It is straightforward to construct a painting of $\MM$ in $O(|\MM|)$ time, using a descending BFS from each maximum that does not explore previously colored edges.
+
+{\bf Colors of vertices:} Note that a painting only colors edges, not vertices. We associate certain
+sets of colors with vertices.
+
+\begin{definition} \label{def:color-set} Fix a painting $\chi$, and a color $c$.
+A vertex $v$ is \emph{touched by $c$} if $x$ is incident to a $c$-colored
+edge with $v$ at the lower endpoint. For $v$, $\col(v)$ is the set of colors that touch $v$.
+\end{definition}
+
+Abusing notation, we say a contour $\phi$ is of color $c$ if all edges that intersect $\phi$ have color $c$.
+
+\begin{definition} \label{def:color-full} Consider vertex $v$, and let $\phi$ be a $(f(v)+\delta)$-contour in
+$f|_{B_\eps(v)}$. If $\phi$ is monochromatic of color $c$, then $v$ is \emph{fully touched by $c$}. The set of colors
+that fully touch $v$ is denoted $\fcol(v)$.
+\end{definition}
+
+Note that $\fcol(v)$ has size at most $2$, by the Morse condition. Note that $\col(v)$ and $\fcol(v)$ can
+be initially determined in $O(|\MM|)$ time, by a simple linear scan.
 
@@
 
-\begin{definition} \label{def:color-set} Fix a painting $\chi$, and a color $c$.
-A critical point $x$ is \emph{touched by $c$} if $x$ is incident to a $c$-colored
-increasing edge. The set of non-extremal critical points touched by $c$ is denoted $\touch(c)$.
-For $x$, $\col(x)$ is the set of colors that touch $x$.
-\end{definition}
 %
@@
 \noindent
-{\bf The binomial heaps $\touch(c)$:} For each color $c$, we store a binomial
-heap $T(c)$ of non-extremal critical points, keyed by vertex heights.
+{\bf The binomial heaps $\touch(c)$:} For each color $c$, $T(c) = \{v | c \in \fcol(v)\}$.
+This is stored as a \emph{binomial max-heap} $T(c)$, keyed by vertex heights.
 
@@
 \noindent
-{\bf The stack $\st$:} This consists of non-extremal critical points with the following structure.
+{\bf The stack $\st$:} This consists of non-extremal critical points, with monotonically increasing
+heights as we go from the base to the head.
+\ignore{
 Each point $x \in \st$ has an associated subset of $\col(x)$, denoted $\mcol(x)$.
@@
 \end{asparaitem}
+}
 
@@
 \noindent
-{\bf Highest points $\h(c)$:} For each color $c$, we maintain a special critical point $\h(c)$ of this color.
-It is supposed to denote the ``highest" point of color $c$, but the notion of highest evolves
-with the algorithm.
+{\bf Attachment vertex $\h(c)$:} For each color $c$, we maintain a critical point $\h(c)$ of this color.
+We will maintain the guarantee that the portion of the contour tree above (and including) $\h(c)$ has already been constructed.
+
+\subsection{Properties of a painted simplicial complex} \label{sec:prop}
+
+\begin{lemma} \label{lem:full} Consider vertex $v$ and take $(f(v)+\delta)$-contour $\phi$ intersecting
+$B_\eps(v)$. Suppose the edges incident to $v$ intersecting $\phi$ do not have the same color. For any such
+color $c$, there exists a critical vertex $w$ strictly higher than $v$ such that $c \in \fcol(v)$.
+\end{lemma}
+
+\begin{proof} Consider the highest
+vertex $w$ that is touched by $c$ and some other color. Clearly such a vertex exists, since the condition
+is also satisfied by $v$. Take the $(f(w)+\delta)$-contour $\phi$ that intersects $B_\eps(v)$
+and intersect some $c$-colored edge incident to $w$. Note that $\phi$ is also colored $c$,
+since $w$ is the highest vertex to be touched by $c$ and some other color. (Take the path of $c$-colored
+edges from the maximum to $w$. For any point on this path, the contour passing through this point must
+be colored $c$.) Hence, $c \in \fcol(w)$ and $w \neq v$. Furthermore, $w$ is a critical point,
+since $w$ is touched by another color. The corresponding edge cannot intersect $\phi$, so $w$
+must have up-degree $2$.
+\end{proof}
+
+\begin{definition} \label{def:ripe} A vertex $v$ is \emph{ripe} if: for all $c \in \fcol(v)$, $v$
+is the highest vertex in $T(c)$. Also, for all $c \in \col(v) \setminus \fcol(v)$, $T(c)$ is empty.
+\end{definition}
 
@@
 
-We have a simple, but critical, procedure that updates the stack $\st$ by pushing
-points on to it.
+We have an imporant procedure that updates the stack $\st$.
 
@@
 \begin{compactenum}
-	\item If $K$ is empty, pick arbitrary $c$. Let $x$ be the highest point in $T(c)$.
-	Push $x$ onto stack and set $\mcol(x) = \{c\}$.
+	\item If $K$ is empty, push arbitrary unprocessed critical point $v$.
 	\item Let $h$ be head of $\st$.
-	\item While $\col(h) \setminus \mcol(h)$ non-empty:
+	\item While $h$ is not ripe:
 	\begin{compactenum}
-		\item Pick some $c \in \col(h) \setminus \mcol(h)$. Replace it by $c' = \rep(c)$.
-		\item Obtain highest point $x$ in $T(c')$. 
-		\item If $x = h$, put $c'$ in $\mcol(h)$. Otherwise ($x \neq h$), push $x$ onto stack
-		and add $c'$ to $\mcol(x)$.
-		\item Let $h$ be new head of $\st$.
+		\item (Either (i) there is $c \in \fcol(v)$ such that $v$ is not the highest in $T(c)$ or (ii) there is 
+		$c \in \col(v) \setminus \fcol(v)$, such that $T(c)$ is non-empty.)
+		\item In either case, push the highest of $T(c)$ onto $\st$, and update head $h$.
 	\end{compactenum}
@@
 \medskip
-When $\update(\st)$ terminates, the head $h$ has the following property. For every $c \in \col(h)$,
-$h$ is the highest point in $\touch(c)$.
+Clearly, when $\update(\st)$ terminates, the head $h$ is ripe. The following crucial claim demonstrates
+the utility of the stack.
+
+\begin{claim} \label{clm:stack} Vertices are always pushed onto $\st$ in increasing order of height.
+\end{claim}
+
+\begin{proof} Consider a single push onto the stack. In Case (i), the increase in height is obvious.
+Consider Case (ii). There is a vertex $v$ touched by color $c \notin \fcol(v)$. Consider the highest
+vertex $w$ that is touched by $c$ and some other color. Clearly such a vertex exists, since the condition
+is also satisfied by $v$. Take the $(f(w)+\delta)$-contour $\phi$ that intersects $B_\eps(v)$
+and intersect some $c$-colored edge incident to $w$. Note that $\phi$ is also colored $c$,
+since $w$ is the highest vertex to be touched by $c$ and some other color. (Take the path of $c$-colored
+edges from the maximum to $w$. For any point on this path, the contour passing through this point must
+be colored $c$.) Hence, $c \in \fcol(w)$ and $w \neq v$. Furthermore, $w$ is a critical point,
+since $w$ is touched by another color. The corresponding edge cannot intersect $\phi$, so $w$
+must have up-degree $2$.
+
+All in all, when a new vertex is pushed onto the stack, it is always higher than the previous head.
+\end{proof}
+
+The painting evolves over time, and is primarily updated through \emph{color merges}. The best way
+of thinking about this is to imagine that two colors are suddenly replaced by one new common color.
+Of course, it is not possible to explicitly redo the painting, but this is implictly done
+through the union-find data structure on colors. So, when we process some vertex $v$,
+which has (say) $\col(v) = \{c_1, c_2, \ldots\}$, it may be the case that all these colors
+have merged.
 

### 2014-10-24 paint-contour-tree.tex
proof in Section 7

@@
 \newcommand{\fcol}{full}
+\newcommand{\cur}{cur}
 \newcommand{\cut}{{\tt cut}}
@@
 \newcommand{\paint}{{\tt paint}}
+\newcommand{\pal}{P}
 \newcommand{\pmax}{P_{\max}}
@@
 
-We use a key theorem from~\cite{Carr} (Theorem 6.6) that maps paths in $\cC(\MM)$ and $\MM$.
-
-\begin{theorem} \label{thm:carr-path} For every path $P$ in $\MM$, there exists a path $Q$ in the contour tree to which $P$
-maps, and for every path $Q$ in the contour tree, there exists at least one path $P$ in $\MM$ that maps to $Q$.
+We use key theorems from~\cite{Carr} (Theorem 6.6) that maps paths in $\cC(\MM)$ and $\MM$.
+
+\begin{theorem} \label{thm:carr-path} For every path $P$ in $\MM$, there exists a path $Q$ in the contour tree corresponding
+to the contours passing through points in $P$.
+For every path $Q$ in the contour tree, there exists at least one path $P$ in $\MM$ through points present
+in contours involving $Q$.
+\end{theorem}
+
+\begin{theorem} \label{thm:carr-mono} For every monotone path $P$ in $\MM$, there exists a monotone path $Q$ in the contour tree to which $P$
+maps (as in the previous theorem), and vice versa.
 \end{theorem}
@@
 \begin{definition} \label{def:paint} Let the 1-skeleton of $\MM$ be denoted $G(\MM) = (\cV,E)$.
-A  \emph{painting} of $\MM$ is a map $\chi:X \cup E \mapsto [|X|]$ with the following properties. 
-\begin{asparaitem}
-	\item The restriction $\chi:X \mapsto [|X|]$ is a bijection.
-	\item Consider an edge $e$. There exists a descending path from some maximum $x$ to $e$
+A  \emph{painting} of $\MM$ is a map $\chi:X \cup E \mapsto [|X|]$ with the following property. 
+ Consider an edge $e$. There exists a descending path from some maximum $x$ to $e$
 	consisting of edges in $E$, such that all edges along this path have the same color as $x$. (The maximum $x$
 	is well-defined to be the unique maximum with the same color as $e$.)
-\end{asparaitem}
-\end{definition}
-%
-Note that paintings are not unique. It is straightforward to construct a painting of $\MM$ in $O(|\MM|)$ time, using a descending BFS from each maximum that does not explore previously colored edges.
+
+An \emph{initial} painting has the additional restriction that the restriction $\chi:X \mapsto [|X|]$ is a bijection.
+\end{definition}
+%
+Paintings are not unique. It is straightforward to construct an initial painting of $\MM$ in $O(|\MM|)$ time, 
+using a descending BFS from each maximum that does not explore previously colored edges. 
+As the algorithm proceeds, the bijection property will not hold, though we always maintain a valid painting.
 
@@
 
-\begin{definition} \label{def:color-full} Consider vertex $v$, and let $\phi$ be a $(f(v)+\delta)$-contour in
-$f|_{B_\eps(v)}$. If $\phi$ is monochromatic of color $c$, then $v$ is \emph{fully touched by $c$}. The set of colors
-that fully touch $v$ is denoted $\fcol(v)$.
-\end{definition}
-
-Note that $\fcol(v)$ has size at most $2$, by the Morse condition. Note that $\col(v)$ and $\fcol(v)$ can
+\begin{definition} \label{def:color-full} Consider vertex $v$. An \emph{up-star} of $v$ is the set 
+of edges that all connected to a fixed component of $\MM^+_v$. (Each $v$
+participates in at mot $2$ up-stars. A color $c \in \col(v)$ \emph{fully touches} $v$
+if all edges in an up-star are colored $v$.
+%and let $\phi$ be a $(f(v)+\delta)$-contour in
+%$f|_{B_\eps(v)}$. If $\phi$ is monochromatic of color $c$, then $v$ is \emph{fully touched by $c$}. The set of colors
+%that fully touch $v$ is denoted $\fcol(v)$.
+\end{definition}
+
+%Note that $\fcol(v)$ has size at most $2$, by the Morse condition. 
+Note that $\col(v)$ 
+%and $\fcol(v)$ can
 be initially determined in $O(|\MM|)$ time, by a simple linear scan.
 
-\subsection{The data structures} \label{sec:struct}
-
-%
-%We will store each set $\touch(c)$ in a binomial heap, keyed by the heights. 
-Abusing notation, $\touch(c)$ refers
-both to the set and the data structure used to store it.
+\begin{lemma} \label{lem:full} Suppose vertex $v$ is connected to a component of $\MM^+_v$
+by edge $e$ of color $c$. Either all edges in $\MM^+_v$ are colored $c$, or there
+exists critical vertex $w \in \MM^+_v$ fully touched by $c$ and touched by another color.
+%
+%and take $(f(v)+\delta)$-contour $\phi$ intersecting
+%$B_\eps(v)$. Suppose the edges incident to $v$ intersecting $\phi$ do not have the same color. For any such
+%color $c$, there exists a critical vertex $w$ strictly higher than $v$ such that $c \in \fcol(v)$.
+\end{lemma}
+
+\begin{proof} Since $e$ has color $c$,
+there must exist vertices in $\MM^+_v$ touched by $c$. Consider the highest
+vertex $w$ in $\MM^+_v$ that is touched by $c$ and some other color. If no such vertex exists,
+this means all edges incident the a vertex touched by $c$ are colored $c$. By walking through
+$\MM^+_v$, we deduce that all edges are colored $c$.
+
+So assume $w$ exists. Take the $(f(w)+\delta)$-contour $\phi$ that intersects $B_\eps(v)$
+and intersect some $c$-colored edge incident to $w$. Note that all edges intersecting $\phi$ are also colored $c$,
+since $w$ is the highest vertex to be touched by $c$ and some other color. (Take the path of $c$-colored
+edges from the maximum to $w$. For any point on this path, the contour passing through this point must
+be colored $c$.) Hence, $c$ fully touches $w$. 
+But $w$ is touched by another color, and the corresponding edge cannot intersect $\phi$. So $w$
+must have up-degree $2$ and is critical.
+\end{proof}
+
+\subsection{The overall approach and the data structures} \label{sec:struct}
+
+We first list out the basic data structures used, and then summarize our approach.
 
@@
 \noindent
-{\bf The binomial heaps $\touch(c)$:} For each color $c$, $T(c) = \{v | c \in \fcol(v)\}$.
-This is stored as a \emph{binomial max-heap} $T(c)$, keyed by vertex heights.
+{\bf The binomial heaps $\touch(c)$:} For each color $c$, $\touch(c)$ is a subset of vertices touched by $c$,
+This is stored as a \emph{binomial max-heap} keyed by vertex heights. Abusing notation, $\touch(c)$ refers
+both to the set and the data structure used to store it.
+
+We initialize as follows. For each critical vertex $v$ and from each of its up-stars, we pick an arbitrary color $c$ touching $v$. 
+We insert $v$ into $T(c)$. Observe that $v$ participates in at least $1$ and at most $2$ heaps.
 
@@
 We will maintain the guarantee that the portion of the contour tree above (and including) $\h(c)$ has already been constructed.
-
-\subsection{Properties of a painted simplicial complex} \label{sec:prop}
-
-\begin{lemma} \label{lem:full} Consider vertex $v$ and take $(f(v)+\delta)$-contour $\phi$ intersecting
-$B_\eps(v)$. Suppose the edges incident to $v$ intersecting $\phi$ do not have the same color. For any such
-color $c$, there exists a critical vertex $w$ strictly higher than $v$ such that $c \in \fcol(v)$.
-\end{lemma}
-
-\begin{proof} Consider the highest
-vertex $w$ that is touched by $c$ and some other color. Clearly such a vertex exists, since the condition
-is also satisfied by $v$. Take the $(f(w)+\delta)$-contour $\phi$ that intersects $B_\eps(v)$
-and intersect some $c$-colored edge incident to $w$. Note that $\phi$ is also colored $c$,
-since $w$ is the highest vertex to be touched by $c$ and some other color. (Take the path of $c$-colored
-edges from the maximum to $w$. For any point on this path, the contour passing through this point must
-be colored $c$.) Hence, $c \in \fcol(w)$ and $w \neq v$. Furthermore, $w$ is a critical point,
-since $w$ is touched by another color. The corresponding edge cannot intersect $\phi$, so $w$
-must have up-degree $2$.
-\end{proof}
-
-\begin{definition} \label{def:ripe} A vertex $v$ is \emph{ripe} if: for all $c \in \fcol(v)$, $v$
-is the highest vertex in $T(c)$. Also, for all $c \in \col(v) \setminus \fcol(v)$, $T(c)$ is empty.
-\end{definition}
-
-\subsection{Updating the stack} \label{sec:stack}
-
-We have an imporant procedure that updates the stack $\st$.
+This is initialized to the maximum corresponding to $c$.
+
+\medskip
+As the algorithm proceeds, it is convenient to imagine
+it maintaining a set of contours $\Phi = \{\phi_1, \phi_2, \ldots, \phi_k\}$, which are contained
+in incomparable edges of $\cC(\MM)$. Suppose we cut $\MM$ by each of these contours, and took
+the simplicial complex $\NN$ where $\phi^-_i$'s are all boundary maxima. The algorithm will
+have already determined the contour trees of the simplicial complices
+where $\phi^+_i$ is a boundary maxima.
+
+During this process, various colors will be \emph{merged}. This involves merging the respective $T(c)$'s and
+updating the union-find data structure. The only operation performed on $T(c)$'s are delete-max and merge.
+No new vertex is ever inserted into the heaps. As colors merge, the initial painting is obviously modified.
+But it would be too expensive to explicitly update colors. So the new painting is given by $\rep(c)$,
+while we only have access to the original painting. By union-find bounds, the only adds an $\alpha(|\MM|)$
+overhead to the data structure operations. The new coloring is guaranteed to be a painting of $\NN$.
+
+While this is the intuitive viewpoint, we describe how this information is actually maintained.
+The vertices $\{\h(\rep(c))\}$ represent the set of contours $\Phi$. In terms of the contour tree, 
+we have already constructed the contour trees rooted at each $h(\rep(c))$. This is trivially true at the initialization,
+when each $h(c)$ is a maximum. One can now treat each $h(\rep(c))$ as a maximum, corresponding to (a possibly new) color $\rep(c)$.
+To reduce confusion between the original painting and the new colors, we will assume that $\col(v)$ and $\fcol(v)$
+are as given by the original painting.
+
+\subsection{The algorithm} \label{sec:algo}
+
+We formally describe the algorithm below. We require a technical definition of \emph{ripe} vertices.
+
+\begin{definition} \label{def:ripe} A vertex $v$ is \emph{ripe} if: for all $c \in \col(v)$, $v$
+is present in $T(\rep(c))$ and is also highest vertex in the heap. 
+%Also, for all $c \in \col(v) \setminus \fcol(v)$, $T(\rep(c))$ is empty.
+\end{definition}
+
+\medskip
+\fbox{
+\begin{minipage}{0.9\textwidth}
+{\bf $\build(\MM)$}
+
+\smallskip
+\begin{compactenum}
+	\item While there are unprocessed critical points:
+	\begin{compactenum}
+		\item Run $\update(\st)$. Pop $\st$ to get $h$.
+		\item Let $\cur(h) = \{\rep(c) | c \in \col(h)\}$.
+		\item For all $c' \in \cur(h)$:
+		\begin{compactenum}
+			\item Add edge $(\h(c'),h)$ to $\cC(\MM)$.
+			\item Delete $h$ from $T(c')$ (if present).
+		\end{compactenum}
+	\item Merge heaps $\{T(c') | c' \in \cur(h)\}$.
+	\item Take union of $\cur(h)$ and denote resulting color by $\widehat{c}$.
+	\item Set $\h(\widehat{c}) = h$ and mark $h$ as processed.
+	\end{compactenum}
+\end{compactenum}
+\end{minipage}}
 
@@
 	\begin{compactenum}
-		\item (Either (i) there is $c \in \fcol(v)$ such that $v$ is not the highest in $T(c)$ or (ii) there is 
-		$c \in \col(v) \setminus \fcol(v)$, such that $T(c)$ is non-empty.)
-		\item In either case, push the highest of $T(c)$ onto $\st$, and update head $h$.
+		\item Find $c \in \col(h)$ such that $h$ is not present or not the highest in $T(\rep(c))$.
+		\item Push the highest of $T(\rep(c))$ onto $\st$, and update head $h$.
 	\end{compactenum}
@@
 
+A few simple facts:
+\begin{asparaitem}
+	\item When $\update(\st)$ terminates, the head $h$ is ripe. 
+	\item After $v$ is processed, all edges incident to $v$ has the same color.
+\end{asparaitem}
+
 \medskip
-Clearly, when $\update(\st)$ terminates, the head $h$ is ripe. The following crucial claim demonstrates
-the utility of the stack.
-
-\begin{claim} \label{clm:stack} Vertices are always pushed onto $\st$ in increasing order of height.
-\end{claim}
-
-\begin{proof} Consider a single push onto the stack. In Case (i), the increase in height is obvious.
-Consider Case (ii). There is a vertex $v$ touched by color $c \notin \fcol(v)$. Consider the highest
-vertex $w$ that is touched by $c$ and some other color. Clearly such a vertex exists, since the condition
-is also satisfied by $v$. Take the $(f(w)+\delta)$-contour $\phi$ that intersects $B_\eps(v)$
-and intersect some $c$-colored edge incident to $w$. Note that $\phi$ is also colored $c$,
-since $w$ is the highest vertex to be touched by $c$ and some other color. (Take the path of $c$-colored
-edges from the maximum to $w$. For any point on this path, the contour passing through this point must
-be colored $c$.) Hence, $c \in \fcol(w)$ and $w \neq v$. Furthermore, $w$ is a critical point,
-since $w$ is touched by another color. The corresponding edge cannot intersect $\phi$, so $w$
-must have up-degree $2$.
-
-All in all, when a new vertex is pushed onto the stack, it is always higher than the previous head.
-\end{proof}
-
-The painting evolves over time, and is primarily updated through \emph{color merges}. The best way
-of thinking about this is to imagine that two colors are suddenly replaced by one new common color.
-Of course, it is not possible to explicitly redo the painting, but this is implictly done
-through the union-find data structure on colors. So, when we process some vertex $v$,
-which has (say) $\col(v) = \{c_1, c_2, \ldots\}$, it may be the case that all these colors
-have merged.
-
-\begin{claim} \label{clm:cont} Consider an increasing path from $h$ to some maxima $x$.
-For any $y$ in this path, $\cont(y)$ only contains points of color $c(x)$.
-Furthermore, for all vertices on $\cont(y)$, their increasing edges are colored $c(x)$.
-\end{claim}
-
-\begin{proof} Consider this increasing path, and think of moving $y$ continuously downwards from $x$
-to $h$. Initially (when $y = x$), the lemma is trivially true. Consider the first $y$
-when the condition fails. We have $\cont(y)$ only containing points of color $c(x)$,
-but there is a vertex $z \in \cont(y)$ that has increasing edges of a different color $c'$.
-All the increasing edges of $z$ that are colored $c(x)$ are contiguous (since we can construct
-a contour intersecting all of them). Thus, there are two disjoint increasing neighborhoods for $z$,
-and $z$ is a saddle point. But that contradicts the fact that $h$ is the highest saddle point
-of color $c(x)$.
-
-So, as $y$ decreases, for all vertices on $\cont(y)$, their increasing edges are colored $c(x)$.
-This also means that $\cont(y)$ only contains points of color $c(x)$.
-\end{proof}
-
-We have important lemmas characterizing $|\col(h)|$.
-
-\begin{lemma} \label{lem:mono} If $h$ is a merge vertex, then $|\col(h)| > 1$.
-\end{lemma}
-
-\begin{proof} Suppose not. Consider the edges incident to $h$ in circular order. There exist (at least) $2$
-disjoint contiguous stretches of edges that are increasing. 
-All these edges have the same color $c$, associated with some maximum $x$. Consider two neighbors $y,z$
-that lie in different stretches. By the definition of a painting, there exist descending paths (along
-edges) from $x$ to $y$ and $x$ to $z$. We can construct a Jordan curve $C$ from $x \rightarrow y
-\rightarrow h \rightarrow z \rightarrow x$. All points in this curve except $h$ have 
-heights larger than $f(h)$. By the Jordan Curve Theorem, this partitions $\MM$ into
-an ``inside" and an ``outside".
-
-One of the lower neighbors of $h$ (say $p$) is inside, and another (say $q$) is outside.
-Because $h$ is a merge, neither of these can a minimum other than the dominant minimum $u$.
-Hence, both $p$ and $q$ have descending paths to $u$. So there is a path from $p$
-to $q$ where all heights are strictly less than $f(h)$. But this path must intersect $C$.
-Contradiction.
-\end{proof}
-
-\begin{lemma} \label{lem:col} If $h$ is a merge vertex, $|\col(h)| = 2$.
-If $h$ is a split vertex, $|\col(h)| = 1$.
-\end{lemma}
-
-\begin{proof} Consider one contiguous sequence of increasing edges from $h$.
-There exists a contour intersecting all of these edges. By \Clm{cont}, they all
-must have the same color. There are two such contiguous sequences.
-
-If $h$ is a merge vertex, then $|\col(h)| = 2$ by \Lem{mono}. If $h$ is a split vertex,
-there is a single contour passing through all increasing edges. By \Clm{cont}, they
-all have the same color.
-\end{proof}
-
-\subsection{Constructing $\reeb(\MM)$} \label{sec:const}
-
-The main construction procedure is $\onestep(\MM)$. It pieces together $\reeb(\MM)$
-incrementally.
-
-\medskip
-\fbox{
-\begin{minipage}{0.9\textwidth}
-{\bf $\onestep(\MM)$}
-
-\smallskip
-\begin{compactenum}
-	\item Run $\update(\st)$. Pop $\st$ to get $h$.
-	\item Suppose $\mcol(h) = \{c\}$ (so $h$ is split). 
+\ignore{	
+	Suppose $\mcol(h) = \{c\}$ (so $h$ is split). 
 	\begin{compactenum}
@@
 	\item Set $\h(c) = h$.
+
+We state the primary invariant below.
+First, some notation. For any $v$, let $\psi^-_v$ denote the $(f(v)-\delta)$-contour intersecting
+$B_\eps(v)$. If there are two such contours (so $v$ is a split), choose the one that contains
+the dominant minimum. The \emph{palette} $\pal$ is the set of colors currently used,
+which is $\{T(\rep(c)) | c \in |X|\}$.
+
+Observe that $\build(\MM)$ loops over all unprocessed critical point.
+The invariant is true at the starting point of each such iteration.
+
+\medskip
+\textbf{Invariant:} 
+\begin{compactenum}
+	\item For every $c \in \pal$, the subtree of $\cJ(\MM)$ rooted at $\h(c)$ has been constructed.
+	\item Fix $c \in \pal$. Consider the set $S$ of maxima of $\MM$ in the subtree of $\cJ(\MM)$ rooted at $\h(c)$.
+	Then $\rep(c) = \bigcup_{s \in S} \chi(s)$.
+	\item Let $\Psi = \{\psi^-_{\h(c)} | c \in \pal\}$. The coloring given by $\rep(\cdot)$
+	is a valid painting of $\cut(\MM,\Psi)$, where $\psi^-_{\h(c)}$ has color $c$.
 \end{compactenum}
-\end{minipage}}
 
 \medskip
-\begin{theorem} \label{thm:piece} Consider the very first call to $\onestep(\MM)$. Let $y$ be a point
-infinitesimally below $h$ on a decreasing edge. Consider $\cut(\MM,\cont(y)) = \{\MM_u,\MM_\ell\}$ (the upper and lower manifolds). Then $\onestep(\MM)$
-outputs $\reeb(\MM_u)$ and implicitly gives a valid painting of $\MM_\ell$.
-\end{theorem}
-
-We have an important lemma that helps to bound the running time.
-
-\begin{lemma} \label{lem:tree} For every color $c$, all points in $T(c)$ lie along a monotone path in 
-$\reeb(\MM)$.
-\end{lemma}
-
-\begin{proof} Prove by induction of the number of edges in $\reeb(\MM)$. Consider the head $h$ of $\st$ in the first call to $\onestep(\MM)$,
-and the contour $\cont(y)$ as defined in \Thm{piece}. Take $\cut(\MM,\cont(y)) = \{\MM_u,\MM_\ell\}$. Apply induction
-hypothesis on $\MM_\ell$, and use that to argue about $T(c_1)$ and $T(c_2)$ in the original manifold.
-\end{proof}
-
-\pagebreak
+It is easy to see that the invariant is true at the very beginning of the algorithm.
+Each $\h(c)$ is simply the maximum colored with $c$, and we have a valid painting of $\MM$.
+}
+
+\section{Proving correctness} \label{sec:correct}
+
+\ignore{
+\begin{claim} \label{clm:process} Assume the invariant. Any vertex with a non-increasing path to some $\h(c)$
+for $c \in \pal$ has been processed.
+\end{claim}
+
+\begin{proof} This is a direct consequence of \Thm{carr-mono}, which relates monotone paths in $\MM$ to $\cC(\MM)$.
+Since the subtree of $\cJ(\MM)$ (which is basically the subtree of $\cC(\MM)$) rooted at $\h(c)$ 
+has been found, all vertices in this subtree must be processed. These are all the vertices with non-increasing paths
+to $\h(c)$ in $\cC(\MM)$, which by \Thm{carr-mono} is the same as those in $\cC(\MM)$.
+\end{proof}
+}
+
+The following crucial claim demonstrates the utility of the stack.
+
+\begin{claim} \label{clm:stack} Vertices are always pushed onto $\st$ in increasing order of height.
+\end{claim}
+
+\begin{proof} Consider a single push onto the stack and the current painting. We have $c \in \col(h)$, as given in pseudocode of $\update$.
+If $h \in T(\rep(c))$, then obviously the next head is higher. So assume that $h \notin T(\rep(c))$.
+There is obviously an edge of color $\rep(c)$ touching $h$ connected to a component $\SS$ of $\MM^+_v$. 
+From the initialization, $h \in T(\rep(c'))$, for some $c'$ touching $h$ (where $c'$ is the color
+of some edge connecting $h$ to $\SS$).
+Applying \Lem{full},  either all edges in $\SS$ are colored $\rep(c)$ or 
+there exists critical vertex $w \in \SS$ fully touched by $\rep(c)$ and another color.
+In the former case, all edges connecting $\SS$ to $h$ are also colored $\rep(c)$
+(since this is a painting). Hence, $\rep(c') = \rep(c)$ and $h \in T(\rep(c))$.
+As we assumed this is not the case, we conclude the existence of critical vertex $w$.
+Since $w$ is fully touched by $\rep(c)$, $w \in T(\rep(c))$. Since $w$
+is touched by another color, $w$ has not been processed. So the highest of $T(\rep(c))$
+is higher than $h$, as desired.
+\ignore{
+If all edges touching $h$ that connect to this component are also colored $\rep(c)$, then
+In Case (i), the increase in height is obvious.
+Consider Case (ii). There is a vertex $v$ touched by color $c \notin \fcol(v)$. Consider the highest
+vertex $w$ that is touched by $c$ and some other color. Clearly such a vertex exists, since the condition
+is also satisfied by $v$. Take the $(f(w)+\delta)$-contour $\phi$ that intersects $B_\eps(v)$
+and intersect some $c$-colored edge incident to $w$. Note that $\phi$ is also colored $c$,
+since $w$ is the highest vertex to be touched by $c$ and some other color. (Take the path of $c$-colored
+edges from the maximum to $w$. For any point on this path, the contour passing through this point must
+be colored $c$.) Hence, $c \in \fcol(w)$ and $w \neq v$. Furthermore, $w$ is a critical point,
+since $w$ is touched by another color. The corresponding edge cannot intersect $\phi$, so $w$
+must have up-degree $2$.
+All in all, when a new vertex is pushed onto the stack, it is always higher than the previous head.
+}
+\end{proof}
+
+\begin{claim} \label{clm:ripe} For ripe vertex $v$, $\{\h(\rep(c)) | c \in \col(v)\}$ is exactly
+the set of parents of $v$.
+\end{claim}
+
+\begin{proof} \Sesh{proof is a little sketchy.}
+Suppose $v$ is ripe. For any $c$ touching $v$, $v$ is the highest in $T(\rep(c))$.
+Consider some such $\rep(c)$, for an edge connecting to component $\SS$ of $\MM^+_v$.
+Either all edges in $\SS$ are colored $\rep(c)$ or there exists critical vertex $w \in \SS$
+fully touched by $\rep(c)$ and another color. In the latter case, $w$ has not been processed,
+so $w \in T(\rep(c))$ (contradiction). Therefore, all edges in $\SS$ are colored $\rep(c)$.
+Suppose there is some critical vertex in $\SS$. If it is not processed yet, it will
+be in $T(\rep(c))$, which is not possible. Thus, all critical vertices in $\SS$ have been processed.
+
+Consider $w$, the lowest critical vertex in $\SS$. When $w$ was processed, it was set to $\h(c')$, for
+some color $c'$. At this stage, all edges in $\SS$ above $w$ were also colored $c'$ (just apply the argument
+above for $w$). Since there are no critical points below $w$, all edges in $\SS$ below $w$
+are also colored $c'$ (the painting property). This means that the edges connecting $\SS$
+are also colored $c'$. So $c' = \rep(c)$, and $\h(\rep(c)) = w$ is a parent of $v$.
+
+This argument can be repeated for all such components $\SS$ to complete the proof.
+\end{proof}
+
 \newcommand{\pathTree}{P_{\mathcal{S}}}
@@
 
+\section{Old proof stuff}
+
+\begin{claim}
+Suppose ripe vertex $v$ is connected to a component of $\MM^+_v$. For every edge $e$ incident to a vertex of $\MM^+_v$,
+$\rep(\chi(e))$ is identical. Furthermore, all vertices in $\MM^+_v$ have been processed.
+\end{claim}
+
+\begin{proof} Consider the painting of the lower simplicial complex $\cut(\MM,\Psi)$ (call it $\NN$). 
+By the invariant, $\rep(\cdot)$ provides a valid coloring. Note that $v$ is not processed (it is ripe),
+so if we take $(f(v)+\delta)$-contours intersecting $B_\eps(v)$, these contours could not have been
+processed. Therefore, there must exist a non-trivial component of $\NN^+_v$ that $v$
+is connected to. By \Lem{full}, either all edges in this component have the same color
+or
+
+Suppose ripe vertex $v$
+is connected to some component in $\NN^+_v$. 
+Consider edge $e$ incident to $v$ intersecting $\phi$ of color $c$. 
+By \Lem{full}, there exists critical vertex $w \in \MM^+_v$ such that $w \in \fcol(c)$. 
+Regardless of whether $v \in \fcol(c)$ or not, the vertex $w$ must have been processed.
+
+\end{proof}
+
+\subsection{Updating the stack} \label{sec:stack}
+
+We have an imporant procedure that updates the stack $\st$.
+
+The painting evolves over time, and is primarily updated through \emph{color merges}. The best way
+of thinking about this is to imagine that two colors are suddenly replaced by one new common color.
+Of course, it is not possible to explicitly redo the painting, but this is implictly done
+through the union-find data structure on colors. So, when we process some vertex $v$,
+which has (say) $\col(v) = \{c_1, c_2, \ldots\}$, it may be the case that all these colors
+have merged.
+
+\begin{claim} \label{clm:cont} Consider an increasing path from $h$ to some maxima $x$.
+For any $y$ in this path, $\cont(y)$ only contains points of color $c(x)$.
+Furthermore, for all vertices on $\cont(y)$, their increasing edges are colored $c(x)$.
+\end{claim}
+
+\begin{proof} Consider this increasing path, and think of moving $y$ continuously downwards from $x$
+to $h$. Initially (when $y = x$), the lemma is trivially true. Consider the first $y$
+when the condition fails. We have $\cont(y)$ only containing points of color $c(x)$,
+but there is a vertex $z \in \cont(y)$ that has increasing edges of a different color $c'$.
+All the increasing edges of $z$ that are colored $c(x)$ are contiguous (since we can construct
+a contour intersecting all of them). Thus, there are two disjoint increasing neighborhoods for $z$,
+and $z$ is a saddle point. But that contradicts the fact that $h$ is the highest saddle point
+of color $c(x)$.
+
+So, as $y$ decreases, for all vertices on $\cont(y)$, their increasing edges are colored $c(x)$.
+This also means that $\cont(y)$ only contains points of color $c(x)$.
+\end{proof}
+
+We have important lemmas characterizing $|\col(h)|$.
+
+\begin{lemma} \label{lem:mono} If $h$ is a merge vertex, then $|\col(h)| > 1$.
+\end{lemma}
+
+\begin{proof} Suppose not. Consider the edges incident to $h$ in circular order. There exist (at least) $2$
+disjoint contiguous stretches of edges that are increasing. 
+All these edges have the same color $c$, associated with some maximum $x$. Consider two neighbors $y,z$
+that lie in different stretches. By the definition of a painting, there exist descending paths (along
+edges) from $x$ to $y$ and $x$ to $z$. We can construct a Jordan curve $C$ from $x \rightarrow y
+\rightarrow h \rightarrow z \rightarrow x$. All points in this curve except $h$ have 
+heights larger than $f(h)$. By the Jordan Curve Theorem, this partitions $\MM$ into
+an ``inside" and an ``outside".
+
+One of the lower neighbors of $h$ (say $p$) is inside, and another (say $q$) is outside.
+Because $h$ is a merge, neither of these can a minimum other than the dominant minimum $u$.
+Hence, both $p$ and $q$ have descending paths to $u$. So there is a path from $p$
+to $q$ where all heights are strictly less than $f(h)$. But this path must intersect $C$.
+Contradiction.
+\end{proof}
+
+\begin{lemma} \label{lem:col} If $h$ is a merge vertex, $|\col(h)| = 2$.
+If $h$ is a split vertex, $|\col(h)| = 1$.
+\end{lemma}
+
+\begin{proof} Consider one contiguous sequence of increasing edges from $h$.
+There exists a contour intersecting all of these edges. By \Clm{cont}, they all
+must have the same color. There are two such contiguous sequences.
+
+If $h$ is a merge vertex, then $|\col(h)| = 2$ by \Lem{mono}. If $h$ is a split vertex,
+there is a single contour passing through all increasing edges. By \Clm{cont}, they
+all have the same color.
+\end{proof}
+
+\subsection{Constructing $\reeb(\MM)$} \label{sec:const}
+
+The main construction procedure is $\onestep(\MM)$. It pieces together $\reeb(\MM)$
+incrementally.
+
+\medskip
+\begin{theorem} \label{thm:piece} Consider the very first call to $\onestep(\MM)$. Let $y$ be a point
+infinitesimally below $h$ on a decreasing edge. Consider $\cut(\MM,\cont(y)) = \{\MM_u,\MM_\ell\}$ (the upper and lower manifolds). Then $\onestep(\MM)$
+outputs $\reeb(\MM_u)$ and implicitly gives a valid painting of $\MM_\ell$.
+\end{theorem}
+
+We have an important lemma that helps to bound the running time.
+
+\begin{lemma} \label{lem:tree} For every color $c$, all points in $T(c)$ lie along a monotone path in 
+$\reeb(\MM)$.
+\end{lemma}
+
+\begin{proof} Prove by induction of the number of edges in $\reeb(\MM)$. Consider the head $h$ of $\st$ in the first call to $\onestep(\MM)$,
+and the contour $\cont(y)$ as defined in \Thm{piece}. Take $\cut(\MM,\cont(y)) = \{\MM_u,\MM_\ell\}$. Apply induction
+hypothesis on $\MM_\ell$, and use that to argue about $T(c_1)$ and $T(c_2)$ in the original manifold.
+\end{proof}
+
+\pagebreak
+
 \end{document}

### 2014-10-24 paint-contour-tree.tex
small changes

--- 2014-10-22 paint-contour-tree.tex
@@
 \newcommand{\col}{col}
-\newcommand{\fcol}{full}
 \newcommand{\cut}{{\tt cut}}
@@
 \newcommand{\cost}{\mathop{cost}}
-\newcommand{\h}{att}
+\newcommand{\h}{high}
 \newcommand{\lift}{{\tt lift}}
@@
 There have also been some output sensitive results for computing the contour tree.  Let $t'$ be the number 
-of critical points in the input mesh, which could potentially be much smaller than $n$.  In particular, 
-Cole-McLaughlin and Pascucci \cite{pc-ectls-02} provide an $O(n+t\log n)$ time algorithm for the case of 
-$3$ dimensional structured meshes.  \Ben{Update distinction between critical and component-critical}
+of critical points in the input mesh, which could potentially be much smaller than $n$.  
+Cole-McLaughlin and Pascucci \cite{pc-ectls-02} provided an $O(n+t'\log n)$ time algorithm for the case of 
+$3$-dimensional structured meshes (note $N$ does not show up in the running time as the mesh is structured).
 Chiang \etal \cite{cllr-sooscctmp-05} improve upon this result, and provide an $O(N+t\log t)$ algorithm 
-which works for structured or unstructured meshes in any dimension, which is provably optimal with respect 
-to the output size but not the output structure, as discussed below.
+which works for structured or unstructured meshes in any dimension, where $t$ is the number of component 
+critical vertices.  This is provably optimal with respect to the output size but not the output structure, 
+as discussed below.
 
@@
 
-\subsection{Our Contribution}
+\subsection{Overview and Contribution}
+
+\paragraph{Algorithm Overview.}
+At a high level, the technique is rather simple.  Break the input manifold into nicely 
+behaved pieces, compute the contour tree on each piece, and then stitch together the computed 
+contour trees according to the cuts we made to break up the manifold.  Below we informally 
+describe how this is accomplished.
+\begin{compactitem}
+ \item {\bf Raining to partition.}  To start with, pick an arbitrary maxima.  Now imagine 
+ ``raining'' water down on the manifold from this maxima, i.e. mark all points which are reachable 
+ by non-ascending paths from the maxima.  Let the boundary between wet and dry portions of the 
+ manifold be called the interface (which may have different connected components).  
+ Remove the wet submanifold, and invert the remaining pieces
+ of the original manifold.  Now recursively rain in each one of these pieces, but rather than 
+ picking an arbitrary maxima, start from the portion of the interface that cut that piece off.
+
+ The benefit of this partitioning is that each resulting piece 
+ is maxima dominant, i.e. there is a single maxima that has a non-ascending path to each point.  
+ This implies the contour of this submanifold is the same as its split tree, or in other words 
+ its contour tree is a rooted binary tree.
+
+ %
+ \item {\bf Painting trees.}  For each maxima dominant piece we want to compute the split 
+ tree without sorting all the vertices.  For conceptual ease lets first flip the submanifold 
+ (i.e., it is now minima dominant and we compute the join tree).  Now arbitrarily pick a maxima 
+ and spill paint from it, specifically color all edges and vertices of the mesh that are reachable 
+ from the maxima by descending paths.  Now go to each other maxima and spill a different color, 
+ but be aware that paints do not mix, i.e. if we reach a vertex that was previosly explored 
+ by another color then we do not explore it.
+
+ This procedure partitions the set of critical points into color classes.  Clearly each color 
+ class is connected in the manifold.  Suppose each class was also connected in the join tree.  
+ In this case we partitioned the join tree into a set of descending paths, each of which 
+ can be constructed by sorting the critical points in a class.  Roughly speaking, 
+ these paths can then be joined together by what amounts to doing a post-order traversal of the tree.
+ However, we may not be so lucky and the color classes may not be connected in the 
+ tree.  This turns out not to be a problem, but it requires proper data-structure use. 
+%  
+%  Now suppose we were lucky and the spilling of paint in the manifold corresponded with a paint 
+%  spilling in the join tree, i.e. the color classes of the critical points partition the join 
+%  tree into disjoint descending paths, each containing one leaf.   In this case 
+%  we have partitioned the tree into branches, and so the tree can be formed by gluing these 
+%  branches together.  Intuitively (i.e. not exactly true), if two branches glue together then 
+%  it means there is a join vertex which got colored by the respective colors of each branch 
+%  (only one of these colors actually explored the vertex in the mesh and this is the ``main'' 
+%  branch you are gluing into).  Therefore, the tree can be constructed by starting at the leaf 
+%  of the trunk (i.e. the branch with the global minima) and recursively building the subtrees 
+%  while walking towards the root.  In other words we are building the tree by doing a post-order traversal.
+%  Unfortunately, we may not be so lucky and the color classes may not appear contiguously in 
+%  the tree.  In this case we can still build the tree in this leaf first fashion, but doing so 
+%  efficiently requires the use of data-structures.
+\end{compactitem}
+
+\paragraph{Lower Bound Overview.}
 \Ben{To be Written}
 
+\paragraph{Our Contribution.}
+Our contribution can be broken into three main aspects.  The first two are algorithmic.  Namely, 
+first we describe a new linear time manifold partitioning scheme which breaks the manifold into 
+``extremum'' dominant submanifolds.  
+Second, we describe how to compute join trees optimally with respect to shape of the tree.
+As the contour tree of each submanifold is (roughly) equal to its join tree, this leads to an 
+optimal algorithm for the contour tree of the input manifold.  
+Therefore, our third and final contribution is proving the optimality of this result, which 
+requires a non-trivial amount of work, and uses techniques we believe to be of independent interest.
+
 \paragraph{Organization.}
-
-\begin{figure}[h]\centering
-    \includegraphics[width=.18\linewidth]{figs/orderGray1}%
-    \hspace{.8in}
-    \includegraphics[width=.18\linewidth]{figs/orderGray2}
-    \caption{On left, downward paint spilling only (each shade of gray represents a different color).  On right, flipping the direction of paint spilling.}
-    \label{fig:order}
-\end{figure}
+\Ben{To be Written}
 
@@
 
-We detail the basic definitions about contour trees. Most of this is taken from Chapter 6 of Carr's thesis~\cite{Carr}.
-Our input is a continuous PL function $f:\MM \mapsto \RR$, where $\MM$ is a simplicial complex in $\RR^d$. 
-All simplices are isomorphic to a standard $d$-dimensional simplex, except for specially designated
-\emph{boundary} simplices. \Sesh{Some mention of representation and data structures here.}
-It is standard to assume
-that $f$ is Morse (we explain later what this entails).
+We detail the basic definitions about contour trees. The following is based on Chapter 6 of Carr's thesis \cite{c-tmi-04}.
+Our input is a continuous PL function $f:\MM \mapsto \RR$, where $\MM$ is a fully triangulated simplicial complex in $\RR^d$. 
+%All simplices are isomorphic to a standard $d$-dimensional simplex, except for specially designated \emph{boundary} simplices. 
+\Sesh{Some mention of representation and data structures here.}
+We use the standard assumption that $f$ is Morse (we explain later what this entails).
 
@@
 
-Not that a contour is itself a simplicial complex of one dimension lower, and is represented (in our algorithms) as such.
+Note that a contour is itself a simplicial complex of one dimension lower, and is represented (in our algorithms) as such.
 We let $\delta, \eps$ denote infinitesimals. Let $B_\eps(x)$ denote a ball of radius $\eps$ around $x$.
@@
 
+\begin{remark}
+ For two manifolds, at each saddle point either two contours join together or split apart, and hence the terminology.
+In higher dimensions this may not be the case, however, it always holds that at a join or split vertex the topology of the level set changes, and hence such vertices are still 'critical'.
+Note that determining which category each vertex falls into according to the above definition can be determined for all vertices simultaneously in linear time, as it is a local neighborhood condition.
+\end{remark}
+
 We will enforce a special constraint on boundaries, encapsulated in the following definition.
@@
 Consider a boundary simplex $F$. All vertices of $F$ have the same function value. Furthermore, all
-neighbors of $F$ not in $F$ either have function values strictly greater or strictly less
+neighbors of $F$ not in $F$ either have all function values strictly greater or strictly less
 than this value.
 \end{definition}
+
+\begin{remark}
+Note that by assuming that our manifold is boundary critical, the vertices on a given boundary are either all maxima or all minima.  
+Since all such vertices on a boundary component are in the same contour, we will abuse notation and refer to this entire set of vertices as a maxima (resp. minima).
+In particular, we do not view this as a violation of the Morse condition on $f$, which requires distinct heights for critical points.
+\end{remark}
 
@@
 edges are formed as follows. For every contour class that is created at $v_i$ and destroyed $v_j$,
-there is an edge $(v_i,v_j)$. (Conventially, edges are directed from higher function value to lower.)
-\end{definition}
-
-(It is not immediately obvious that this graph is a tree. But it is equivalent to the Reeb graph
-of $f:\MM \rightarrow \RR$, which is known to be a tree.) We denote the contour tree of $\MM$
-by $\reeb(\MM,f)$. The corresponding node and edge sets are denoted as $\cV(\cdot)$ and $\cE(\cdot)$.
-
-We use a key theorem from~\cite{Carr} (Theorem 6.6) that maps paths in $\cC(\MM)$ and $\MM$.
+there is an edge $(v_i,v_j)$. (Conventionally, edges are directed from higher function value to lower.)
+\end{definition}
+
+We denote the contour tree of $\MM$
+by $\reeb(\MM) = \reeb(\MM,f)$. The corresponding node and edge sets are denoted as $\cV(\cdot)$ and $\cE(\cdot)$.
+%
+It is not immediately obvious that this graph is a tree, but the above definition is equivalent to that of the Reeb graph of $f:\MM \rightarrow \RR$ on $\MM$, and this is known to be a tree for PL functions defined over PL manifolds.
+
+\begin{remark}
+Note that the contour tree as defined above contains vertices for all critical points, 
+which differs slightly from the usual definition which only contains critical points corresponding to a change in the number of contours.
+Specifically, if a vertex in the tree we defined has degree $2$ then only the topology of a contour is changing rather than the number of contour.
+As such the standard contour tree can be computed from ours by a linear time pass to contract maximal degree 2 vertex paths.
+
+For simplicity we will also make the standard assumption that vertices in the contour 
+tree have degree at most $3$, i.e. there are no multi-saddles.
+\end{remark}
+
+We use a key theorem from \cite{c-tmi-04} (Theorem 6.6) that maps paths in $\cC(\MM)$ and $\MM$.
 
@@
 Again, we use $\eps$ for an infinitesimally small value. We denote $\phi^+$ (resp. $\phi^-$) to be
-the contour at value $f(\phi^+) + \eps$ (resp. $f(\phi^-) - \eps$), which is at distance $\eps$
+the contour at value $f(\phi) + \eps$ (resp. $f(\phi) - \eps$), which is at distance $\eps$
 from $\phi$.
 
-An $h$-contour is achieved by intersecting $\MM$ with the hyperplane $x_{d+1} = h$. (Think of the $d+1$-dimension
+An $h$-contour is achieved by intersecting $\MM$ with the hyperplane $x_{d+1} = h$ and taking a connected component. (Think of the $d+1$-dimension
 as height.) Given some point $x$
@@
 
-\begin{lemma} \label{lem:cut} Consider a regular contour $\phi$ contain in contour class (edge of $\reeb(\MM))$
-$(u,v)$ and denote $\MM' = \cut(\phi,\MM)$. Then $\cV(\reeb(\MM')) = \{\phi^+,\phi^-\} \cup (\cV(\MM) \setminus \phi)$
+\begin{lemma} \label{lem:cut} Consider a regular contour $\phi$ contained in a contour class (of an edge of $\reeb(\MM))$
+$(u,v)$ and denote $\MM' = \cut(\phi,\MM)$. Then $\cV(\reeb(\MM')) = \{\phi^+,\phi^-\} \cup \cV(\MM)$
 and $\cE(\reeb(\MM')) = \{(u,\phi^+),(\phi^-,v)\} \cup (\cE(\MM) \setminus (u,v))$.
@@
 maximal in $\MM$. So all edges of $\cC(\MM')$ that do not involve $\phi^+$ or $\phi^-$ are edges of $\cC(\MM)$.
-Analogously, every edge of $\cC(\MM)$ not involving $\phi$ is an edge of $\cC(\MM)$.
+Analogously, every edge of $\cC(\MM)$ not involving $\phi$ is an edge of $\cC(\MM')$.
 
@@
 are leaves in $\cC(\MM')$, this path obviously cannot use edges incident to them. By \Lem{cut},
-all the edges of this path are in $\cE(\cC(\MM)) \setminus (u,v)$. So we get a cycle in $\cC(\MM)$. Contradiction.
+all the edges of this path are in $\cE(\cC(\MM)) \setminus (u,v)$. So we get a cycle in $\cC(\MM)$, a contradiction.
 To show that there are exactly two connected components in $\cut(\phi,\MM)$, it suffices
@@
 \begin{definition} \label{def:dom} A simplicial complex is \emph{minimum dominant} if there exists
-a minimum $x$ such that every non-minimal \emph{vertex} in the manifold has decreasing path to $x$.
+a minimum $x$ such that every non-minimal \emph{vertex} in the manifold has a decreasing path to $x$.
 Analogously define \emph{maximum dominant}. 
@@
 \begin{definition} \label{def:wet} Fix $\MM$ and $x \in \MM$. The set of points $y \in \MM$ such that there is a non-ascending path from $x$ to $y$
-is denoted by $\wet(x,\MM)$ (which is in turn represented as a simplicial complex). A point $z$ is at the \emph{interface} of $\wet(x,\MM)$ if every neighborhood of $z$
+is denoted by $\wet(x,\MM)$ (which is in turn is represented as a simplicial complex). A point $z$ is at the \emph{interface} of $\wet(x,\MM)$ if every neighborhood of $z$
 has non-trivial intersection with $\wet(x,\MM)$.
@@
 
-\begin{proof} If $y \in \wet(x,\MM)$, all the points in any contour containing $y$ are also in $\wet(x,\MM)$.
- (Follow the non-ascending path to from $x$ to $y$ and then walk along the contour.) The converse is also true,
+\begin{proof} If $p \in \wet(x,\MM)$, all the points in any contour containing $p$ are also in $\wet(x,\MM)$.
+ (Follow the non-ascending path to from $x$ to $p$ and then walk along the contour.) The converse is also true,
  so $\wet(x,\MM)$ contains entire contours.
@@
 The points in $B_\eps(y)$ that lie below $y$ have a descending path from $y$ and hence must be wet.
-There must be a dry point in $B_\eps(y)$ that is above $y$, and hence,
+There must also be a dry point in $B_\eps(y)$ that is above $y$, and hence,
 there exists a dry, regular $(f(y)+\delta)$-contour $\phi$ intersecting $B_\eps(y)$.
 
-Suppose $y$ is in the interface. Consider the set $\Gamma_y$ of points connected to $y$
-by a path with all points at height $f(y)$. (This is a set of $f(y)$-contours.) Suppose for contradiction
-that $\forall z \in \Gamma_y$, $z$ has up-degree $1$. Consider the non-ascending path from $x$ to $y$ and let $z$
+Let $\Gamma_y$ be the contour containing $y$.  Suppose for contradiction
+that $\forall p \in \Gamma_y$, $p$ has up-degree $1$ (see \Def{deg}). Consider the non-ascending path from $x$ to $y$ and let $z$
 be the first point of $\Gamma_y$ encountered. There exists a wet, regular $(f(y) + \delta)$-contour $\psi$ 
 intersecting $B_\eps(z)$. Now, walk from $z$ to $y$ along $\Gamma_y$. If all points $w$ in this walk
-have up-degree $1$ (ref. to \Def{deg}), then $\psi$ is the unique $f(y)+\delta$-contour
-intersecting $B_\eps(w)$. This would imply that $\phi = \psi$, contradicting the fact that $\phi$ is wet
-and $\psi$ is dry.
+have up-degree $1$, then $\psi$ is the unique $f(y)+\delta$-contour
+intersecting $B_\eps(w)$. This would imply that $\phi = \psi$, contradicting the fact that $\psi$ is wet
+and $\phi$ is dry.
 
@@
 both $\phi$ and $\psi$. As $\delta, \eps \rightarrow 0^+$, the limit of $\phi$ (which is dry)
-lies in $\Gamma_y$ (which is wet). Hence, this gives the interface, which is a contour containing
-the join $w$.
+lies in $\Gamma_y$ (which is wet). 
+Hence this limit gives a contour of the interface containing the join $w$.
 \end{proof}
@@
 Note that $\wet(x,\MM)$ (and its interface) can be computed in time linear in the size of the wet simplicial complex.
-We perform a non-ascending walk from $x$. Any facet $F$ of $\MM$ encountered has some (if not all) portion of in $\wet(x,\MM)$.
+We perform a non-ascending search from $x$. Any facet $F$ of $\MM$ encountered is partially (if not entirely) in $\wet(x,\MM)$.
 This portion is determined by cutting $F$ along the interface. Since the interface is a contour, this is equivalent
@@
 a merge vertex $y$. Take any dry increasing edge incident to $y$, and pick the point $z$ on this edge at height
-$f(y) + \delta$. Let $\lift(\phi)$ be the unique contour through regular point $z$. Note that $\lift(\phi)$ is dry.
+$f(y) + \delta$. Let $\lift(\phi)$ be the unique contour through the regular point $z$. Note that $\lift(\phi)$ is dry.
 
@@
 
+\begin{figure}[h]\centering
+    \includegraphics[width=.18\linewidth,height=.35\linewidth]{figs/orderGray1}%
+    \hspace{.8in}
+    \includegraphics[width=.18\linewidth,height=.35\linewidth]{figs/orderGray2}
+    \caption{On left, downward paint spilling only (each shade of gray represents a different color), producing a grid.  On right, flipping the direction of paint spilling.}
+    \label{fig:order}
+\end{figure}
+
 We describe the main partitioning procedure that cuts a simplicial complex $\NN$ into extremum
@@
 we begin with $\NN$ set to $\MM$ and $x$ as an arbitrary maximum. One of the critical aspects
-of the procedure is that the rain alternately flows downwards and upwards. In other words, when
+of the procedure is that the rain alternately flows downwards and upwards, 
+since otherwise faces may be cut a super constant number of times (see \Fig{order}).
+In other words, when
 we start, rain flows downwards. In each recursive call, the direction of rain is switched to the 
-opposite direction. While this is way of implementing $\rain$, it is conceptually easier
+opposite direction. While one can implement $\rain$ this way, it is conceptually easier
 to think of \emph{inverting} a complex $\NN'$ when a recursive call is made. Inversion
@@
 \begin{lemma}\label{lem:new-verts}
-A facet $F \in \MM$ is cut at most once during the $\rain(x,\MM)$.
+A facet $F \in \MM$ is cut at most once during $\rain(x,\MM)$.
 \end{lemma}
 \begin{proof} All notation here follows that in the pseudocode of $\rain$.
-Firstly, by \Thm{jordan}, all the pieces of which $\rain$ is invoked are disjoint.
-Secondly, all recursive calls are made on dry complices.
+First, by \Thm{jordan}, all the pieces on which $\rain$ is invoked are disjoint.
+Second, all recursive calls are made on dry complices.
 
 Consider the first time that $F$ is cut, say, during the call to $\rain(x,\NN)$.
-Specially, say this happens when $\cut(\phi'_i,\NN_i)$ is constructed. 
+Specifically, say this happens when $\cut(\phi'_i,\NN_i)$ is constructed. 
 So $F$ is cut by some hyperplane into lower and upper portions (which are then triangulated).
@@
 \begin{proof} The operations performed are the raining and cutting. We can bound this by the total
-complexities of the wet complices and the new boundaries created. No facet is every wet twice,
+complexities of the wet complices and the new boundaries created. No facet is ever wet twice,
 and by \Lem{new-verts}, the total number of facets that every appear is $O(|\MM|)$. This also 
@@
 
-Now we come to the main algorithmic contribution. We design a new algorithm for computing contour
-trees of a minimum dominant simplicial complice $\MM$. As part of the initialization, we determine
-all critical points in $\MM$. For any vertex $v$, we can look at $f(v) \pm \delta$-contours 
-in $f|_{B_\eps(v)}$ to determine the up and down degrees. This determines the criticality (and type)
-of $v$. This requires $O(|\MM|)$ time, since every facet is looked at proportional to the number of its
-vertices.
-
-{\bf Painting:} The central tool is a notion 
-of \emph{painting} of $\MM$. Initially associate a color with each maximum. Denoting the set of maxima is $X$,
-this is an arbitrary bijection $\chi:X \mapsto [|X|]$. Imagine there being a large
-can of paint of color $\chi(x)$ at maxima $x$. We will spill different paint from each maximum and watch it flow down.
+We begin by associating a color with each maximum. If the set of maxima is $S$,
+this is simply an arbitrary bijection $\chi:S \mapsto [|S|]$. Imagine there being a large
+can of paint of color $\chi(y)$ at maxima $y$. We will spill different paint from each maximum and watch it flow down.
 This is analogous to the raining in the previous section, but paint is a much more viscous liquid.
-\emph{So paint only flows down edges, and it does not color the interior of facets.} Furthermore, paints
-do not mix, so eventually, every vertex and edge of $\MM$ gets a unique color. This process (and indeed the entire algorithm)
-works purely on the 1-skeleton of $\MM$, which is just a graph.
-%
-\begin{definition} \label{def:paint} Let the 1-skeleton of $\MM$ be denoted $G(\MM) = (\cV,E)$.
-A  \emph{painting} of $\MM$ is a map $\chi:X \cup E \mapsto [|X|]$ with the following properties. 
+\emph{So paint only flows down edges, and it does not color the interior of faces.} Furthermore, paints
+do not mix, so eventually, every vertex and edge of $\MM$ gets a unique color.
+%
+\begin{definition} \label{def:paint} A  \emph{painting} of $\MM$ is a map $\chi:S \cup E \mapsto [|S|]$
+with the following properties. (Remember that $G(\MM) = (V,E)$.)
 \begin{asparaitem}
-	\item The restriction $\chi:X \mapsto [|X|]$ is a bijection.
+	\item The restriction $\chi:S \mapsto [|S|]$ is a bijection.
 	\item Consider an edge $e$. There exists a descending path from some maximum $x$ to $e$
@@
 %
-Note that paintings are not unique. It is straightforward to construct a painting of $\MM$ in $O(|\MM|)$ time, using a descending BFS from each maximum that does not explore previously colored edges.
-
-{\bf Colors of vertices:} Note that a painting only colors edges, not vertices. We associate certain
-sets of colors with vertices.
+Note that paintings are not unique. It is straightforward to construct a painting of $\MM$ in $O(n)$ time, using a descending BFS from each maximum that does not explore previously colored edges.
+
+\subsection{The data structures} \label{sec:struct}
 
 \begin{definition} \label{def:color-set} Fix a painting $\chi$, and a color $c$.
-A vertex $v$ is \emph{touched by $c$} if $x$ is incident to a $c$-colored
-edge with $v$ at the lower endpoint. For $v$, $\col(v)$ is the set of colors that touch $v$.
-\end{definition}
-
-Abusing notation, we say a contour $\phi$ is of color $c$ if all edges that intersect $\phi$ have color $c$.
-
-\begin{definition} \label{def:color-full} Consider vertex $v$, and let $\phi$ be a $(f(v)+\delta)$-contour in
-$f|_{B_\eps(v)}$. If $\phi$ is monochromatic of color $c$, then $v$ is \emph{fully touched by $c$}. The set of colors
-that fully touch $v$ is denoted $\fcol(v)$.
-\end{definition}
-
-Note that $\fcol(v)$ has size at most $2$, by the Morse condition. Note that $\col(v)$ and $\fcol(v)$ can
-be initially determined in $O(|\MM|)$ time, by a simple linear scan.
-
-\subsection{The data structures} \label{sec:struct}
-
+A critical point $x$ is \emph{touched by $c$} if $x$ is incident to a $c$-colored
+increasing edge. The set of non-extremal critical points touched by $c$ is denoted $\touch(c)$.
+For $x$, $\col(x)$ is the set of colors that touch $x$.
+\end{definition}
 %
@@
 \noindent
-{\bf The binomial heaps $\touch(c)$:} For each color $c$, $T(c) = \{v | c \in \fcol(v)\}$.
-This is stored as a \emph{binomial max-heap} $T(c)$, keyed by vertex heights.
+{\bf The binomial heaps $\touch(c)$:} For each color $c$, we store a binomial
+heap $T(c)$ of non-extremal critical points, keyed by vertex heights.
 
@@
 \noindent
-{\bf The stack $\st$:} This consists of non-extremal critical points, with monotonically increasing
-heights as we go from the base to the head.
-\ignore{
+{\bf The stack $\st$:} This consists of non-extremal critical points with the following structure.
 Each point $x \in \st$ has an associated subset of $\col(x)$, denoted $\mcol(x)$.
@@
 \end{asparaitem}
-}
 
@@
 \noindent
-{\bf Attachment vertex $\h(c)$:} For each color $c$, we maintain a critical point $\h(c)$ of this color.
-We will maintain the guarantee that the portion of the contour tree above (and including) $\h(c)$ has already been constructed.
-
-\subsection{Properties of a painted simplicial complex} \label{sec:prop}
-
-\begin{lemma} \label{lem:full} Consider vertex $v$ and take $(f(v)+\delta)$-contour $\phi$ intersecting
-$B_\eps(v)$. Suppose the edges incident to $v$ intersecting $\phi$ do not have the same color. For any such
-color $c$, there exists a critical vertex $w$ strictly higher than $v$ such that $c \in \fcol(v)$.
-\end{lemma}
-
-\begin{proof} Consider the highest
-vertex $w$ that is touched by $c$ and some other color. Clearly such a vertex exists, since the condition
-is also satisfied by $v$. Take the $(f(w)+\delta)$-contour $\phi$ that intersects $B_\eps(v)$
-and intersect some $c$-colored edge incident to $w$. Note that $\phi$ is also colored $c$,
-since $w$ is the highest vertex to be touched by $c$ and some other color. (Take the path of $c$-colored
-edges from the maximum to $w$. For any point on this path, the contour passing through this point must
-be colored $c$.) Hence, $c \in \fcol(w)$ and $w \neq v$. Furthermore, $w$ is a critical point,
-since $w$ is touched by another color. The corresponding edge cannot intersect $\phi$, so $w$
-must have up-degree $2$.
-\end{proof}
-
-\begin{definition} \label{def:ripe} A vertex $v$ is \emph{ripe} if: for all $c \in \fcol(v)$, $v$
-is the highest vertex in $T(c)$. Also, for all $c \in \col(v) \setminus \fcol(v)$, $T(c)$ is empty.
-\end{definition}
+{\bf Highest points $\h(c)$:} For each color $c$, we maintain a special critical point $\h(c)$ of this color.
+It is supposed to denote the ``highest" point of color $c$, but the notion of highest evolves
+with the algorithm.
 
@@
 
-We have an imporant procedure that updates the stack $\st$.
+We have a simple, but critical, procedure that updates the stack $\st$ by pushing
+points on to it.
 
@@
 \begin{compactenum}
-	\item If $K$ is empty, push arbitrary unprocessed critical point $v$.
+	\item If $K$ is empty, pick arbitrary $c$. Let $x$ be the highest point in $T(c)$.
+	Push $x$ onto stack and set $\mcol(x) = \{c\}$.
 	\item Let $h$ be head of $\st$.
-	\item While $h$ is not ripe:
+	\item While $\col(h) \setminus \mcol(h)$ non-empty:
 	\begin{compactenum}
-		\item (Either (i) there is $c \in \fcol(v)$ such that $v$ is not the highest in $T(c)$ or (ii) there is 
-		$c \in \col(v) \setminus \fcol(v)$, such that $T(c)$ is non-empty.)
-		\item In either case, push the highest of $T(c)$ onto $\st$, and update head $h$.
+		\item Pick some $c \in \col(h) \setminus \mcol(h)$. Replace it by $c' = \rep(c)$.
+		\item Obtain highest point $x$ in $T(c')$. 
+		\item If $x = h$, put $c'$ in $\mcol(h)$. Otherwise ($x \neq h$), push $x$ onto stack
+		and add $c'$ to $\mcol(x)$.
+		\item Let $h$ be new head of $\st$.
 	\end{compactenum}
@@
 \medskip
-Clearly, when $\update(\st)$ terminates, the head $h$ is ripe. The following crucial claim demonstrates
-the utility of the stack.
-
-\begin{claim} \label{clm:stack} Vertices are always pushed onto $\st$ in increasing order of height.
-\end{claim}
-
-\begin{proof} Consider a single push onto the stack. In Case (i), the increase in height is obvious.
-Consider Case (ii). There is a vertex $v$ touched by color $c \notin \fcol(v)$. Consider the highest
-vertex $w$ that is touched by $c$ and some other color. Clearly such a vertex exists, since the condition
-is also satisfied by $v$. Take the $(f(w)+\delta)$-contour $\phi$ that intersects $B_\eps(v)$
-and intersect some $c$-colored edge incident to $w$. Note that $\phi$ is also colored $c$,
-since $w$ is the highest vertex to be touched by $c$ and some other color. (Take the path of $c$-colored
-edges from the maximum to $w$. For any point on this path, the contour passing through this point must
-be colored $c$.) Hence, $c \in \fcol(w)$ and $w \neq v$. Furthermore, $w$ is a critical point,
-since $w$ is touched by another color. The corresponding edge cannot intersect $\phi$, so $w$
-must have up-degree $2$.
-
-All in all, when a new vertex is pushed onto the stack, it is always higher than the previous head.
-\end{proof}
-
-The painting evolves over time, and is primarily updated through \emph{color merges}. The best way
-of thinking about this is to imagine that two colors are suddenly replaced by one new common color.
-Of course, it is not possible to explicitly redo the painting, but this is implictly done
-through the union-find data structure on colors. So, when we process some vertex $v$,
-which has (say) $\col(v) = \{c_1, c_2, \ldots\}$, it may be the case that all these colors
-have merged.
+When $\update(\st)$ terminates, the head $h$ has the following property. For every $c \in \col(h)$,
+$h$ is the highest point in $\touch(c)$.
 
@@
 \bibliography{contour}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%
+%
+%end of writup
+%stuff below should be deleted
+%
+%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\remove{
 
@@
 
+}
+
 \end{document}

### 2014-10-24 paint-contour-tree.tex

--- 2014-10-24 paint-contour-tree.tex
@@
 edges are formed as follows. For every contour class that is created at $v_i$ and destroyed $v_j$,
-there is an edge $(v_i,v_j)$. (Conventially, edges are directed from higher function value to lower.)
-\end{definition}
-
-(It is not immediately obvious that this graph is a tree. But it is equivalent to the Reeb graph
-of $f:\MM \rightarrow \RR$, which is known to be a tree.) We denote the contour tree of $\MM$
-by $\reeb(\MM,f)$. The corresponding node and edge sets are denoted as $\cV(\cdot)$ and $\cE(\cdot)$.
-
-We use key theorems from~\cite{Carr} (Theorem 6.6) that maps paths in $\cC(\MM)$ and $\MM$.
+there is an edge $(v_i,v_j)$. (Conventionally, edges are directed from higher function value to lower.)
+\end{definition}
+
+We denote the contour tree of $\MM$
+by $\reeb(\MM) = \reeb(\MM,f)$. The corresponding node and edge sets are denoted as $\cV(\cdot)$ and $\cE(\cdot)$.
+%
+It is not immediately obvious that this graph is a tree, but the above definition is equivalent to that of the Reeb graph of $f:\MM \rightarrow \RR$ on $\MM$, and this is known to be a tree for PL functions defined over PL manifolds.
+
+\begin{remark}
+Note that the contour tree as defined above contains vertices for all critical points, 
+which differs slightly from the usual definition which only contains critical points corresponding to a change in the number of contours.
+Specifically, if a vertex in the tree we defined has degree $2$ then only the topology of a contour is changing rather than the number of contour.
+As such the standard contour tree can be computed from ours by a linear time pass to contract maximal degree 2 vertex paths.
+
+For simplicity we will also make the standard assumption that vertices in the contour 
+tree have degree at most $3$, i.e. there are no multi-saddles.
+\end{remark}
+
+We use a key theorem from \cite{c-tmi-04} (Theorem 6.6) that maps paths in $\cC(\MM)$ and $\MM$.
 
@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%
+%
+%end of writup
+%stuff below should be deleted
+%
+%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\remove{
+
 \appendix
@@
 \pagebreak
+}
 

### 2014-10-27 paint-contour-tree.tex
updates to section 5

--- 2014-10-24 paint-contour-tree.tex
@@
 \newcommand{\NN}{\mathbb{N}}
+\newcommand{\PP}{\mathbb{P}}
 \newcommand{\RR}{\mathbb{R}}
@@
 \newcommand{\h}{att}
+\newcommand{\jc}{\cJ_C}
 \newcommand{\lift}{{\tt lift}}
@@
 We will fix such a simplicial complex $\MM$, with the dominant minimum $m^*$. 
-We observe that the contours trees of minimum dominant manifolds have a relatively simple description.
-This description is directly inspired by (and proven using) the \emph{join tree} of~\cite{csa-cctad-00}.
+The set of non-dominant minima is denoted by $M$.
+For vertex $v$, we use $\MM^+_v$ to denote the simplicial complex obtained by only
+keeping vertices $u$ such that $f(u) > f(v)$. Analogously, define $\MM^-_v$. Note that $\MM^+_v$ may contain
+numerous connected components.
+
+The main theorem of this section asserts that the contours trees of minimum dominant manifolds have a simple description.
+This description is based on \emph{join} and \emph{split} trees, as given by~\cite{csa-cctad-00}.
+Conventially, all edges are directed from higher to lower function value.
 We first explain what these trees are.
 
-For vertex $v$, we use $\MM^+_v$ to denote the simplicial complex obtained by only
-keeping vertices $u$ such that $f(u) > f(v)$. Analogously, define $\MM^-_v$. Note that $\MM^+$ may contain
-numerous connected components. The following definition of join and split trees were given by~\cite{csa-cctad-00}.
-Conventially, all edges are directed from higher to lower function value.
-
 \begin{definition} \label{join} The join tree $\cJ(\MM)$ of $\MM$ is built on vertex set $V(\MM)$.
-The directed edge $(u,v)$ is present when $u$ is the smallest valued vertex is a connected component of $\MM^+_v$
+The directed edge $(u,v)$ is present when $u$ is the smallest valued vertex in a connected component of $\MM^+_v$
 \emph{and} $v$ is connected to this component (in $\MM$). The split tree $\cS(\MM)$ is obtained
 by looking at $\MM^-_v$ (or alternately, by taking the join tree of the inversion of $\MM$).
-\end{definition}
-
-Note that unlike the contour tree, all vertices in $\MM$ are present in $\cJ(\MM)$ and $\cS(\MM)$. 
-Some basic facts about these trees. All outdegrees in $\cJ(\MM)$ and all indegrees in $\cS(\MM)$ are at most $1$.
+
+The critical join tree $\jc(\MM)$ is built on $V(\MM) \setminus M$. 
+The directed edge $(u,v)$ is present when $u$ is the smallest valued vertex in $V(\MM) \setminus M$ in a connected component of $\MM^+_v$
+and $v$ is connected to this component (in $\MM$). 
+\end{definition}
+
+Some basic facts about these trees. 
+%Note that unlike the contour tree, all vertices in $\MM$ are present in $\cJ(\MM)$ and $\cS(\MM)$. 
+All outdegrees in $\cJ(\MM)$ and all indegrees in $\cS(\MM)$ are at most $1$.
 Indegree $2$ vertices in $\cJ(\MM)$ are joins, and outdegree $2$ vertices in $\cS(\MM)$ are splits.
 Leaves in $\cJ(\MM)$ are maxima, and leaves in $\cS(\MM)$ are minima.
-
 The key observation is that $\cS(\MM)$ is trivial for minimum dominant $\MM$.
@@
 
-It is convenient to denote the non-dominant minima by $m_1, m_2, \ldots, m_k$, and the associated
-splits (as given by \Lem{split}) by $s_1, s_2, \ldots, s_k$.
-
 To prove our main theorem, we use the merging procedure from~\cite{csa-cctad-00} that constructs the contour tree
@@
 is the contour tree with regular points inserted into the edges. 
-For any tree $T$ with vertex $v$, we define \emph{erasing} $v$ from $T$,
-when both in and outdegrees are at most $1$. If $v$ is a leaf, just delete $v$. Otherwise, smooth $v$ out
+Consider tree $T$ with vertex $v$ of in and outdegree at most $1$.
+\emph{Erasing} $v$ from $T$ is the following operation: if $v$ is a leaf, just delete $v$. Otherwise, smooth $v$ out
 (delete $v$ and connect its neighbors by an edge). This tree is denoted by $T \ominus v$.
@@
 
-\begin{theorem} \label{thm:contour-tree} The contour tree $\cC(\MM)$ consists of all edges $\{(s_i, m_i)\}$
-as well as: for any critical point $v$, edge $(w,v)$ where $w$ is the lowest non-minimal critical point
-in a connected component of $\MM^+_v$ connected to $v$.
+\begin{theorem} \label{thm:contour-tree} Let $\MM$ have a dominant minimum. 
+The contour tree $\cC(\MM)$ consists of all edges $\{(s_i, m_i)\}$ and $\jc(\MM)$.
 \end{theorem}
@@
 
-Now we come to the main algorithmic contribution. We design a new algorithm for computing contour
-trees of a minimum dominant simplicial complice $\MM$. As part of the initialization, we determine
+The main algorithmic contribution is a new algorithm for computing join trees of any triangulated simplicial
+complex $\MM$.
+
+As part of the initialization, we determine
 all critical points in $\MM$. For any vertex $v$, we can look at $f(v) \pm \delta$-contours 
@@
 {\bf Painting:} The central tool is a notion 
-of \emph{painting} of $\MM$. Initially associate a color with each maximum. Denoting the set of maxima is $X$,
-this is an arbitrary bijection $\chi:X \mapsto [|X|]$. Imagine there being a large
-can of paint of color $\chi(x)$ at maxima $x$. We will spill different paint from each maximum and watch it flow down.
+of \emph{painting} of $\MM$. Initially associate a color with each maximum. Imagine there being a large
+can of paint of a distinct color at each maxima $x$. We will spill different paint from each maximum and watch it flow down.
 This is analogous to the raining in the previous section, but paint is a much more viscous liquid.
 \emph{So paint only flows down edges, and it does not color the interior of facets.} Furthermore, paints
-do not mix, so eventually, every vertex and edge of $\MM$ gets a unique color. This process (and indeed the entire algorithm)
+do not mix, so every edge of $\MM$ gets a unique color. This process (and indeed the entire algorithm)
 works purely on the 1-skeleton of $\MM$, which is just a graph.
 %
-\begin{definition} \label{def:paint} Let the 1-skeleton of $\MM$ be denoted $G(\MM) = (\cV,E)$.
+\begin{definition} \label{def:paint} Let the 1-skeleton of $\MM$ have edge set $E$ and maxima $X$.
 A  \emph{painting} of $\MM$ is a map $\chi:X \cup E \mapsto [|X|]$ with the following property. 
@@
 If $h \in T(\rep(c))$, then obviously the next head is higher. So assume that $h \notin T(\rep(c))$.
-There is obviously an edge of color $\rep(c)$ touching $h$ connected to a component $\SS$ of $\MM^+_v$. 
+There is obviously an edge of color $\rep(c)$ touching $h$ connected to a component $\PP$ of $\MM^+_v$. 
 From the initialization, $h \in T(\rep(c'))$, for some $c'$ touching $h$ (where $c'$ is the color
-of some edge connecting $h$ to $\SS$).
-Applying \Lem{full},  either all edges in $\SS$ are colored $\rep(c)$ or 
-there exists critical vertex $w \in \SS$ fully touched by $\rep(c)$ and another color.
-In the former case, all edges connecting $\SS$ to $h$ are also colored $\rep(c)$
+of some edge connecting $h$ to $\PP$).
+Applying \Lem{full},  either all edges in $\PP$ are colored $\rep(c)$ or 
+there exists critical vertex $w \in \PP$ fully touched by $\rep(c)$ and another color.
+In the former case, all edges connecting $\PP$ to $h$ are also colored $\rep(c)$
 (since this is a painting). Hence, $\rep(c') = \rep(c)$ and $h \in T(\rep(c))$.
@@
 Suppose $v$ is ripe. For any $c$ touching $v$, $v$ is the highest in $T(\rep(c))$.
-Consider some such $\rep(c)$, for an edge connecting to component $\SS$ of $\MM^+_v$.
-Either all edges in $\SS$ are colored $\rep(c)$ or there exists critical vertex $w \in \SS$
+Consider some such $\rep(c)$, for an edge connecting to component $\PP$ of $\MM^+_v$.
+Either all edges in $\PP$ are colored $\rep(c)$ or there exists critical vertex $w \in \PP$
 fully touched by $\rep(c)$ and another color. In the latter case, $w$ has not been processed,
-so $w \in T(\rep(c))$ (contradiction). Therefore, all edges in $\SS$ are colored $\rep(c)$.
-Suppose there is some critical vertex in $\SS$. If it is not processed yet, it will
-be in $T(\rep(c))$, which is not possible. Thus, all critical vertices in $\SS$ have been processed.
-
-Consider $w$, the lowest critical vertex in $\SS$. When $w$ was processed, it was set to $\h(c')$, for
-some color $c'$. At this stage, all edges in $\SS$ above $w$ were also colored $c'$ (just apply the argument
-above for $w$). Since there are no critical points below $w$, all edges in $\SS$ below $w$
-are also colored $c'$ (the painting property). This means that the edges connecting $\SS$
+so $w \in T(\rep(c))$ (contradiction). Therefore, all edges in $\PP$ are colored $\rep(c)$.
+Suppose there is some critical vertex in $\PP$. If it is not processed yet, it will
+be in $T(\rep(c))$, which is not possible. Thus, all critical vertices in $\PP$ have been processed.
+
+Consider $w$, the lowest critical vertex in $\PP$. Note that $w$ is a parent of $v$. When $w$ was processed, it was set to $\h(c')$, for
+some color $c'$. At this stage, all edges in $\PP$ above $w$ were also colored $c'$ (just apply the argument
+above for $w$). Since there are no critical points below $w$, all edges in $\PP$ below $w$
+are also colored $c'$ (the painting property). This means that the edges connecting $\PP$
 are also colored $c'$. So $c' = \rep(c)$, and $\h(\rep(c)) = w$ is a parent of $v$.
 
-This argument can be repeated for all such components $\SS$ to complete the proof.
+This argument can be repeated for all such components $\PP$ to complete the proof.
 \end{proof}

### 2014-10-27 paint-contour-tree.tex
intro changes

--- 2014-10-24 paint-contour-tree.tex
@@
 Here we present the first contour tree algorithm which takes into account the shape of the contour tree, 
-and in doing so yields instance optimal results (on all instances requiring $\Omega(n\alpha(n))$ time).
-In particular, our algorithm runs in $O(n\alpha(n))$ time when the tree is balanced, yet still 
-matches the previous worst case $O(n\log n)$ bound when the tree is long and skinny.
+and in doing so yields instance optimal results.
+Specifically, we prove that for any partition of the contour tree into a set of descending paths, $P$, 
+any comparison based algorithm requires $\sum_{p\in P} |p|\log|p|$ time.  Moreover, we show there   
+exists some partition $P'$ such that the running time of our algorithm is 
+$O(\sum_{p\in P'} |p|\log|p| + N\alpha(N))$, where $N$ is total complexity of the input manifold 
+(and hence this is optimal on all instances requiring $\Omega(N\alpha(N))$ time).
+In particular, our algorithm runs in $O(N\alpha(N))$ time when the tree is balanced, 
+while previous algorithms ran in $O(N+t\log t)$ (where $t$ is the size of the contour tree).
+%In particular, our algorithm runs in $O(N\alpha(N))$ time when the tree is balanced, yet still 
+%matches the previous worst case $O(N+n\log n)$ bound when the tree is long and skinny.
 
@@
 of results on efficient algorithms for their computation \cite{kobps-ctsssit-97,tv-cct-98,csa-cctad-00,pc-ectls-02,cllr-sooscctmp-05}.
-As an initial step these algorithms require that all $n$ contour tree vertices be sorted.  
-In particular, \cite{cllr-sooscctmp-05} provide an $O(n\log n)$ time algorithm, and since in the worst case computing the contour tree 
-requires $\Omega(n\log n)$ time, this algorithm is optimal with respect to the size of the output.  
-Here we present the first contour tree algorithm which is optimal (for instances requiring $\Omega(n\alpha(n))$ 
-time\footnote{Where $\alpha(n)$ is the inverse Ackermann function, which is $< 5$ for computation in this universe.}) 
+Let $N$ be the total size of the input manifold, $n$ the number of vertices, and $t$ the number of critical points. (Note that in many applications $N=O(n)$.)
+As an initial step these algorithms require that all $t$ contour tree vertices be sorted.  
+In particular, \cite{cllr-sooscctmp-05} provide an $O(N+t\log t)$ time algorithm, and since in the worst case computing the contour tree 
+requires $\Omega(N+t\log t)$ time, this algorithm is optimal with respect to the size of the output.  
+Here we present the first contour tree algorithm which is optimal (for instances requiring $\Omega(N\alpha(N))$ 
+time\footnote{Where $\alpha(\cdot)$ is the inverse Ackermann function, which is $< 5$ for computation in this universe.}) 
 with respect not just to the size of the output, but also its structure.  This is accomplished by introducing a simple linear time 
@@
 For all instances the bound for our algorithm either matches the best previous bound, or is faster.
-For, example when the contour tree is a balanced, our algorithm runs in $O(n\alpha(n))$ time, whereas all 
-previous algorithms had a running time of $\Omega(n\log n)$
+For, example when the contour tree is a balanced, our algorithm runs in $O(N\alpha(N))$ time, whereas all 
+previous algorithms had a running time of $\Omega(N+t\log t)$
 
@@
 
-There have also been some output sensitive results for computing the contour tree.  Let $t'$ be the number 
+There have also been some output sensitive results for computing the contour tree.  Let $t$ be the number 
 of critical points in the input mesh, which could potentially be much smaller than $n$.  
-Cole-McLaughlin and Pascucci \cite{pc-ectls-02} provided an $O(n+t'\log n)$ time algorithm for the case of 
+Cole-McLaughlin and Pascucci \cite{pc-ectls-02} provided an $O(n+t\log n)$ time algorithm for the case of 
 $3$-dimensional structured meshes (note $N$ does not show up in the running time as the mesh is structured).
-Chiang \etal \cite{cllr-sooscctmp-05} improve upon this result, and provide an $O(N+t\log t)$ algorithm 
-which works for structured or unstructured meshes in any dimension, where $t$ is the number of component 
+Chiang \etal \cite{cllr-sooscctmp-05} improve upon this result, and provide an $O(N+t'\log t')$ algorithm 
+which works for structured or unstructured meshes in any dimension, where $t'$ is the number of component 
 critical vertices.  This is provably optimal with respect to the output size but not the output structure, 
@@
  is maxima dominant, i.e. there is a single maxima that has a non-ascending path to each point.  
- This implies the contour of this submanifold is the same as its split tree, or in other words 
+ This implies the contour tree of this submanifold is the same as its split tree, or in other words 
  its contour tree is a rooted binary tree.
@@
 \paragraph{Lower Bound Overview.}
-\Ben{To be Written}
+In order to show that our algorithm for computing the join tree is optimal we consider paritioing the tree into disjoint 
+descending paths, each containing a leaf.  It turns out that each such partition into paths, $P$, 
+defines a lower bound of $\sum_{p\in P} |p|\log|p|$ for computing the contour tree, as you need to know the 
+sorted order of the vertices along each path in order to determine the tree.
+Specifically, we show that there is a set of at least $\Pi_{p_i\in P} (|p_i|-1)!$ distinct PL $2$-manifolds, 
+all of which have the same defining mesh, and for which each sorted order of each group of saddles determines a different join tree.
+
+This instance optimality of our algorithm then follows, as we also show that there is a particular path decomposition 
+whose cost upper bounds our.  Specifically, we assign a weight to each node in the tree which is determined by 
+the painting procedure described above (which represents the time it takes to process this node).  
+The path decomposition then turns out to be the heavy-light path decomposition 
+described by Sleator and Tarjan \cite{st-dsdt-83}, according to this weight function.
+However, this decomposition is just the first step in the analysis, which 
+requires further case analysis depending on the relative size of the length of each path 
+and the weights of its vertices.
 
@@
 Critical values include these boundary values. By boundary criticality, all contours are themselves
-represented as $d-1$-dimensional simplicial complices.
+represented as $d-1$-dimensional simplicial complexes.
 
@@
 \begin{definition} \label{join} The join tree $\cJ(\MM)$ of $\MM$ is built on vertex set $V(\MM)$.
-The directed edge $(u,v)$ is present when $u$ is the smallest valued vertex is a connected component of $\MM^+_v$
+The directed edge $(u,v)$ is present when $u$ is the smallest valued vertex in a connected component of $\MM^+_v$
 \emph{and} $v$ is connected to this component (in $\MM$). The split tree $\cS(\MM)$ is obtained
@@
 
-\begin{proof} First, let us complete the proof assuming that each split $v$ has one child that is just a leaf, which
-is a non-dominant minimum. Any minimum is a leaf in $\cS(\MM)$, and thereby attached to a split. If we removed
-all non-dominant minima, we must end up with a path, as asserted above.
+\begin{proof} It suffices to prove that each split $v$ has one child that is just a leaf, which
+is a non-dominant minimum.  Specifically, any minimum is a leaf in $\cS(\MM)$ and thereby attached to a split, 
+which implies that if we removed all non-dominant minima, we must end up with a path, as asserted above.
 
@@
 path from $v$ to the dominant minimum $m^*$. Consider the first edge (necessarily decreasing from $v$)
-on this path. It must intersect some contour, say $\phi$. By \Thm{jordan}, $\cut(\MM,\phi)$ has
+on this path. It must intersect one of the $(f(v) - \delta)$-contours, say $\phi$. By \Thm{jordan}, $\cut(\MM,\phi)$ has
 two connected components, with one (call it $\LL$) having $\phi^-$ as a boundary maximum. This complex
-contains $m^*$ as the non-ascending path $\phi$ intersects $\phi$ only once. Let the other
+contains $m^*$ as the non-ascending path intersects $\phi$ only once. Let the other
 component be called $\MM'$.
@@
 Now we come to the main algorithmic contribution. We design a new algorithm for computing contour
-trees of a minimum dominant simplicial complice $\MM$. As part of the initialization, we determine
+trees of a minimum dominant simplicial complex $\MM$. As part of the initialization, we determine
 all critical points in $\MM$. For any vertex $v$, we can look at $f(v) \pm \delta$-contours 
@@
 {\bf Painting:} The central tool is a notion 
-of \emph{painting} of $\MM$. Initially associate a color with each maximum. Denoting the set of maxima is $X$,
+of \emph{painting} of $\MM$. Initially associate a color with each maximum. Denoting the set of maxima by $X$,
 this is an arbitrary bijection $\chi:X \mapsto [|X|]$. Imagine there being a large
@@
 
-An \emph{initial} painting has the additional restriction that the restriction $\chi:X \mapsto [|X|]$ is a bijection.
+An \emph{initial} painting has the additional property that the restriction $\chi:X \mapsto [|X|]$ is a bijection.
 \end{definition}
@@
 \begin{definition} \label{def:color-full} Consider vertex $v$. An \emph{up-star} of $v$ is the set 
-of edges that all connected to a fixed component of $\MM^+_v$. (Each $v$
-participates in at mot $2$ up-stars. A color $c \in \col(v)$ \emph{fully touches} $v$
+of edges that all connect to a fixed component of $\MM^+_v$. (Each $v$
+participates in at mot $2$ up-stars.) A color $c \in \col(v)$ \emph{fully touches} $v$
 if all edges in an up-star are colored $v$.
@@
 %and $\fcol(v)$ can
-be initially determined in $O(|\MM|)$ time, by a simple linear scan.
+can be initially determined in $O(|\MM|)$ time, by a simple linear scan.
 
@@
 vertex $w$ in $\MM^+_v$ that is touched by $c$ and some other color. If no such vertex exists,
-this means all edges incident the a vertex touched by $c$ are colored $c$. By walking through
+this means all edges incident to a vertex touched by $c$ are colored $c$. By walking through
 $\MM^+_v$, we deduce that all edges are colored $c$.
@@
 edges from the maximum to $w$. For any point on this path, the contour passing through this point must
-be colored $c$.) Hence, $c$ fully touches $w$. 
+be colored $c$.) Hence $c$ fully touches $w$. 
 But $w$ is touched by another color, and the corresponding edge cannot intersect $\phi$. So $w$
@@
 But it would be too expensive to explicitly update colors. So the new painting is given by $\rep(c)$,
-while we only have access to the original painting. By union-find bounds, the only adds an $\alpha(|\MM|)$
+while we only have access to the original painting. By union-find bounds, this only adds an $\alpha(|\MM|)$
 overhead to the data structure operations. The new coloring is guaranteed to be a painting of $\NN$.
 
-While this is the intuitive viewpoint, we describe how this information is actually maintained.
+%While this is the intuitive viewpoint, we describe how this information is actually maintained.
 The vertices $\{\h(\rep(c))\}$ represent the set of contours $\Phi$. In terms of the contour tree, 

### 2014-10-27 paint-contour-tree.tex
more changes to section 6, 7

--- 2014-10-27 paint-contour-tree.tex
@@
 \newcommand{\h}{att}
+\newcommand{\init}{{\tt init}}
 \newcommand{\jc}{\cJ_C}
@@
 The main theorem of this section asserts that the contours trees of minimum dominant manifolds have a simple description.
-This description is based on \emph{join} and \emph{split} trees, as given by~\cite{csa-cctad-00}.
-Conventially, all edges are directed from higher to lower function value.
-We first explain what these trees are.
+The exact statement will require some definitions and notation.
+We required the notions of \emph{join} and \emph{split} trees, as given by~\cite{csa-cctad-00}.
+Conventially, all edges are directed from higher to lower function value. We will use the terminology
+``parent" for an adjacent node that is higher. Because of the orientation, a node
+may have multiple parents.
+%We first explain what these trees are.
 
@@
 by looking at $\MM^-_v$ (or alternately, by taking the join tree of the inversion of $\MM$). 
-
-The critical join tree $\jc(\MM)$ is built on $V(\MM) \setminus M$. 
-The directed edge $(u,v)$ is present when $u$ is the smallest valued vertex in $V(\MM) \setminus M$ in a connected component of $\MM^+_v$
-and $v$ is connected to this component (in $\MM$). 
 \end{definition}
@@
 %Note that unlike the contour tree, all vertices in $\MM$ are present in $\cJ(\MM)$ and $\cS(\MM)$. 
-All outdegrees in $\cJ(\MM)$ and all indegrees in $\cS(\MM)$ are at most $1$.
-Indegree $2$ vertices in $\cJ(\MM)$ are joins, and outdegree $2$ vertices in $\cS(\MM)$ are splits.
-Leaves in $\cJ(\MM)$ are maxima, and leaves in $\cS(\MM)$ are minima.
+All outdegrees in $\cJ(\MM)$ are at most $1$, all indegree $2$ vertices are joins, all leaves are maxima,
+and the global minimum is the root. All indegrees in $\cS(\MM)$ are at most $1$, all outdegree $2$
+vertices are splits, leaves are minima, and the global maximum is the root.
 The key observation is that $\cS(\MM)$ is trivial for minimum dominant $\MM$.
@@
 	\item A single path (in sorted order) with all vertices except non-dominant minima.
-	\item Each non-dominant minimum is attached to a unique split.
+	\item Each non-dominant minimum is attached to a unique split (which is adjacent to it).
 \end{asparaitem}
@@
 \end{proof}
+
+It is convenient to denote the non-dominant minima as $m_1, m_2, \ldots, m_k$
+and the corresponding splits (as given by the lemma above) as $s_1, s_2, \ldots, s_k$.
 
@@
 
+\begin{definition} The critical join tree $\jc(\MM)$ is built on the set $V'$ of all
+critical points that are \emph{not} non-dominant minima. 
+The directed edge $(u,v)$ is present when $u$ is the smallest valued vertex in $V'$ in a connected component of $\MM^+_v$
+and $v$ is connected to this component (in $\MM$). 
+\end{definition}
+
 \begin{theorem} \label{thm:contour-tree} Let $\MM$ have a dominant minimum. 
@@
 complex $\MM$.
-
-As part of the initialization, we determine
-all critical points in $\MM$. For any vertex $v$, we can look at $f(v) \pm \delta$-contours 
-in $f|_{B_\eps(v)}$ to determine the up and down degrees. This determines the criticality (and type)
-of $v$. This requires $O(|\MM|)$ time, since every facet is looked at proportional to the number of its
-vertices.
+%
+%
+%
+%As part of the initialization, we determine
+%all critical points in $\MM$. For any vertex $v$, we can look at $f(v) \pm \delta$-contours 
+%in $f|_{B_\eps(v)}$ to determine the up and down degrees. This determines the criticality (and type)
+%of $v$. This requires $O(|\MM|)$ time, since every facet is looked at proportional to the number of its
+%vertices.
 
 {\bf Painting:} The central tool is a notion 
-of \emph{painting} of $\MM$. Initially associate a color with each maximum. Imagine there being a large
+of \emph{painting} $\MM$. Initially associate a color with each maximum. Imagine there being a large
 can of paint of a distinct color at each maxima $x$. We will spill different paint from each maximum and watch it flow down.
@@
  Consider an edge $e$. There exists a descending path from some maximum $x$ to $e$
-	consisting of edges in $E$, such that all edges along this path have the same color as $x$. (The maximum $x$
-	is well-defined to be the unique maximum with the same color as $e$.)
-
-An \emph{initial} painting has the additional restriction that the restriction $\chi:X \mapsto [|X|]$ is a bijection.
-\end{definition}
-%
-Paintings are not unique. It is straightforward to construct an initial painting of $\MM$ in $O(|\MM|)$ time, 
-using a descending BFS from each maximum that does not explore previously colored edges. 
-As the algorithm proceeds, the bijection property will not hold, though we always maintain a valid painting.
-
-{\bf Colors of vertices:} Note that a painting only colors edges, not vertices. We associate certain
+	consisting of edges in $E$, such that all edges along this path have the same color as $x$.
+
+An \emph{initial} painting has the additional property that the restriction $\chi:X \mapsto [|X|]$ is a bijection.
+\end{definition}
+%
+%Paintings are not unique. It is straightforward to construct an initial painting of $\MM$ in $O(|\MM|)$ time, 
+%using a descending BFS from each maximum that does not explore previously colored edges. 
+%As the algorithm proceeds, the bijection property will not hold, though we always maintain a valid painting.
+
+A painting only colors edges, not vertices. We associate certain
 sets of colors with vertices.
 
-\begin{definition} \label{def:color-set} Fix a painting $\chi$, and a color $c$.
-A vertex $v$ is \emph{touched by $c$} if $x$ is incident to a $c$-colored
+\begin{definition} \label{def:color-set} Fix a painting $\chi$ and vertex $v$.
+\begin{asparaitem}
+	 \item An \emph{up-star} of $v$ is the set of edges that all connected to a fixed component of $\MM^+_v$.
+	 \item A vertex $v$ is \emph{touched by color $c$} if $x$ is incident to a $c$-colored
 	 edge with $v$ at the lower endpoint. For $v$, $\col(v)$ is the set of colors that touch $v$.
-\end{definition}
-
+	 \item A color $c \in \col(v)$ \emph{fully touches} $v$ if all edges in an up-star are colored $v$.
+\end{asparaitem}
+\end{definition}
+
+Each non-maxima $v$ participates in at least $1$ and at most $2$ up-stars (the latter iff
+$v$ is a join).
+
+\ignore{
 Abusing notation, we say a contour $\phi$ is of color $c$ if all edges that intersect $\phi$ have color $c$.
@@
 \begin{definition} \label{def:color-full} Consider vertex $v$. An \emph{up-star} of $v$ is the set 
-of edges that all connected to a fixed component of $\MM^+_v$. (Each $v$
-participates in at mot $2$ up-stars. A color $c \in \col(v)$ \emph{fully touches} $v$
-if all edges in an up-star are colored $v$.
+of edges that all connected to a fixed component of $\MM^+_v$. 
 %and let $\phi$ be a $(f(v)+\delta)$-contour in
@@
 be initially determined in $O(|\MM|)$ time, by a simple linear scan.
-
-\begin{lemma} \label{lem:full} Suppose vertex $v$ is connected to a component of $\MM^+_v$
-by edge $e$ of color $c$. Either all edges in $\MM^+_v$ are colored $c$, or there
-exists critical vertex $w \in \MM^+_v$ fully touched by $c$ and touched by another color.
-%
-%and take $(f(v)+\delta)$-contour $\phi$ intersecting
-%$B_\eps(v)$. Suppose the edges incident to $v$ intersecting $\phi$ do not have the same color. For any such
-%color $c$, there exists a critical vertex $w$ strictly higher than $v$ such that $c \in \fcol(v)$.
-\end{lemma}
-
-\begin{proof} Since $e$ has color $c$,
-there must exist vertices in $\MM^+_v$ touched by $c$. Consider the highest
-vertex $w$ in $\MM^+_v$ that is touched by $c$ and some other color. If no such vertex exists,
-this means all edges incident the a vertex touched by $c$ are colored $c$. By walking through
-$\MM^+_v$, we deduce that all edges are colored $c$.
-
-So assume $w$ exists. Take the $(f(w)+\delta)$-contour $\phi$ that intersects $B_\eps(v)$
-and intersect some $c$-colored edge incident to $w$. Note that all edges intersecting $\phi$ are also colored $c$,
-since $w$ is the highest vertex to be touched by $c$ and some other color. (Take the path of $c$-colored
-edges from the maximum to $w$. For any point on this path, the contour passing through this point must
-be colored $c$.) Hence, $c$ fully touches $w$. 
-But $w$ is touched by another color, and the corresponding edge cannot intersect $\phi$. So $w$
-must have up-degree $2$ and is critical.
-\end{proof}
+}
 
@@
 
-We first list out the basic data structures used, and then summarize our approach.
+We give an informal description of the algorithm, which might motivate the data structures.
+
+Our algorithm incrementally builds $\jc(\MM)$ from the leaves (maxima) to the root (dominant minimum). 
+Let us focus on an initial painting, where each colors has 1-1 correspondence with the maxima.
+Consider two sibling leaves $\ell_1, \ell_2$ and their common parent $v$. The leaves are maxima,
+and $v$ is a join that ``merges" $\ell_1, \ell_2$. A nice mental picture is given when $\MM$ is 2-dimensional.
+In that case, there are ``mounds" corresponding to each $\ell_1$ and $\ell_2$ that merge
+at a valley $v$. Suppose this was the entire input, and $\ell_1$ was colored blue and $\ell_2$ was colored red. 
+Both the mounds are colored completely blue or red, while $v$ is touched by both colors.
+Moreover, each up-star of $v$ (one connecting to each mound) is fully colored. So this
+indicates that $v$ joins the blue maxima and red maxima in $\jc(\MM)$.
+
+This is precisely how we hope to exploit the information in the painting. We prove later that
+when some join $v$ has both up-stars fully colored, the corresponding maxima (of those colors) are exactly
+the parents of $v$ in $\jc(\MM)$. To proceed further, we ``merge" the colors red and blue into a new color, purple. In other words,
+we replace all red and blue edges by purple edges. This indicates that the red and blue maxima have been
+handled. Imagine flattening the red and blue mounds till $v$, so now that erstwhile join $v$
+is a new maxima. Suppose purple paint was poured from $v$. In terms on $\jc(\MM)$, this
+is equivalent to remove leaves $\ell_1$ and $\ell_2$, now making $v$ as a leaf.
+Alternately, $\jc(\MM)$ has been constructed up to $v$, and it remains to determine $v$'s parent.
+
+Of course, things are more complicated when there are other mounds. There may be a yellow mound,
+corresponding to $\ell_3$ that joins with the blue mound higher up (at vertex $u$). In $\jc(\MM)$, $\ell_2$ and $\ell_3$
+are sibling leaves, and $\ell_1$ is a sibling of some ancestor of these leaves. So we cannot
+merge red and blue, until yellow and blue merge. Naturally, we use priority queues to handle this issue.
+We know that $u$ must also be (fully) touched by blue. So all critical vertices touched by blue
+are put into a priority queue keyed by height, and vertices are handled in that order.
+
+What happens when finally blue and red join at $v$? We merge the two colors, but now have
+a blue heap and red queues of critical vertices. We also need to merge the priority queues
+to get a consistent painting. And that decides the use of \emph{binomial heaps} to store the
+critical points of one color. Merges can be done in logarithmic time, the same time
+required for deletes.
+
+In this discussion, we have ignored a somewhat annoying problem. Up-stars of critical vertices
+may actually be touched by numerous colors, not just one as we assumed above. A simple solution
+would be to insert vertices into heaps corresponding to all colors touching it. But there could
+be super-constant numbers of copies of a vertex, and handling all these copies would lead
+to an extra overhead. We show that it suffices to simply put each vertex $v$ into at most two heaps,
+one for each up-star. We are guaranteed that when $v$ needs to processed, its up-stars
+automatically have one color, because of all the color merges that previously occured.
+
+With this preamble, we are ready to discuss our data structures. We discuss all the initializations
+in the next section.
 
@@
 
-We initialize as follows. For each critical vertex $v$ and from each of its up-stars, we pick an arbitrary color $c$ touching $v$. 
-We insert $v$ into $T(c)$. Observe that $v$ participates in at least $1$ and at most $2$ heaps.
+%We initialize as follows. For each critical vertex $v$ and from each of its up-stars, we pick an arbitrary color $c$ touching $v$. 
+%We insert $v$ into $T(c)$. Observe that $v$ participates in at least $1$ and at most $2$ heaps.
+%
 
@@
 of classes of colors, and this will be maintained as a standard union-find data structure.
-For any color $c$, $\rep(c)$ denotes the representative of its class. We initialize
-by simply setting $\rep(c) = c$ for all colors.
+For any color $c$, $\rep(c)$ denotes the representative of its class.
 
@@
 We will maintain the guarantee that the portion of the contour tree above (and including) $\h(c)$ has already been constructed.
-This is initialized to the maximum corresponding to $c$.
-
+%This is initialized to the maximum corresponding to $c$.
+
+\ignore{
 \medskip
@@
 are as given by the original painting. 
+}
 
@@
 
-We formally describe the algorithm below. We require a technical definition of \emph{ripe} vertices.
+We formally describe the algorithm below. 
+We require a technical definition of \emph{ripe} vertices.
 
@@
 \end{definition}
+
+\medskip
+\fbox{
+\begin{minipage}{0.9\textwidth}
+{\bf $\init(\MM)$}
+
+\smallskip
+\begin{compactenum}
+	\item Construct an initial painting of $\MM$ using a descending BFS from maxima that does not explore previously colored edges.
+	\item Determine all critical points in $\MM$. For each $v$, look at $(f(v) \pm \delta)$-contours in $f|_{B_\eps(v)}$ 
+	to determine the up and down degrees. 
+	\item Mark each critical $v$ as unprocessed.
+	\item For each critical $v$ and each up-star, pick arbitrary color $c$ touching $v$. Insert $v$ into $T(c)$.
+	\item Initialize $\rep(c) = c$ and set $\h(c)$ to be the unique maximum colored $c$.
+\end{compactenum}
+\end{minipage}}
 
@@
 \begin{compactenum}
+	\item Run $\init(\MM)$.
 	\item While there are unprocessed critical points:
@@
 		\begin{compactenum}
-			\item Add edge $(\h(c'),h)$ to $\cC(\MM)$.
+			\item Add edge $(\h(c'),h)$ to $\jc(\MM)$.
 			\item Delete $h$ from $T(c')$ (if present).
@@
 
+\bigskip
+
 A few simple facts:
 \begin{asparaitem}
+	\item At all times, the colors form a valid painting.
+	\item Each vertex is present in at most $2$ heaps. After processing, it is removed from all heaps.
+	\item After $v$ is processed, all edges incident to $v$ have the same color (technically, same representative). 
 	\item When $\update(\st)$ terminates, the head $h$ is ripe. 
-	\item After $v$ is processed, all edges incident to $v$ has the same color.
 \end{asparaitem}
@@
 }
+
+Our main workhorse is the following technical lemma.
+
+\begin{lemma} \label{lem:full} Suppose vertex $v$ is connected to a component of $\MM^+_v$
+by edge $e$ of color $c$. Either all edges in $\MM^+_v$ are colored $c$, or there
+exists critical vertex $w \in \MM^+_v$ fully touched by $c$ and touched by another color.
+%
+%and take $(f(v)+\delta)$-contour $\phi$ intersecting
+%$B_\eps(v)$. Suppose the edges incident to $v$ intersecting $\phi$ do not have the same color. For any such
+%color $c$, there exists a critical vertex $w$ strictly higher than $v$ such that $c \in \fcol(v)$.
+\end{lemma}
+
+\begin{proof} Since $e$ has color $c$,
+there must exist vertices in $\MM^+_v$ touched by $c$. Consider the highest
+vertex $w$ in $\MM^+_v$ that is touched by $c$ and some other color. If no such vertex exists,
+this means all edges incident the a vertex touched by $c$ are colored $c$. By walking through
+$\MM^+_v$, we deduce that all edges are colored $c$.
+
+So assume $w$ exists. Take the $(f(w)+\delta)$-contour $\phi$ that intersects $B_\eps(v)$
+and intersect some $c$-colored edge incident to $w$. Note that all edges intersecting $\phi$ are also colored $c$,
+since $w$ is the highest vertex to be touched by $c$ and some other color. (Take the path of $c$-colored
+edges from the maximum to $w$. For any point on this path, the contour passing through this point must
+be colored $c$.) Hence, $c$ fully touches $w$. 
+But $w$ is touched by another color, and the corresponding edge cannot intersect $\phi$. So $w$
+must have up-degree $2$ and is critical.
+\end{proof}
+
+We prove a series of claims that will lead to the correctness proof.
+
+\begin{claim} \label{clm:upstar} Consider ripe vertex $v$ and take the up-star connecting
+to some component of $\MM^+_v$. All edges in this component and the up-star have the same color.
+\end{claim}
+
+\begin{proof} Let $c$ be the color of some edge in this up-star
+By ripeness, $v$ is the highest in $T(\rep(c))$.
+Denote the component of $\MM^+_v$ by $\PP$.
+By \Lem{full}, either all edges in $\PP$ are colored $\rep(c)$ or there exists critical vertex $w \in \PP$
+fully touched by $\rep(c)$ and another color. In the latter case, $w$ has not been processed,
+so $w \in T(\rep(c))$ (contradiction to ripeness). Therefore, all edges in $\PP$ are colored $\rep(c)$.
+\end{proof}
+
+\begin{claim} \label{clm:process} The partial output on the processed vertices is exactly
+the restriction of $\jc(\MM)$ to these vertices.
+\end{claim}
+
+\begin{proof} More generally, we prove the following: all outputs on processed vertices
+are edges of $\cC(\MM)$ and for any color $c$, $\h(c)$ is the lowest processed vertex
+of that color. We prove by induction on the processing order. Initially, this is trivially
+true, handling these base case. For the induction step, consider the situation
+when $v$ is being processed.
+
+The vertex $v$ is ripe. Take any up-star, and the corresponding component $\PP$
+of $\MM^+_v$ that it connects to. By \Clm{upstar}, all edges in $\PP$ and the up-star
+have the same color (say $c$). If some critical vertex in $\PP$ is not processed,
+it must be in $T(c)$, which violates the ripeness of $v$.
+Thus, all critical vertices in $\PP$ have been processed.
+
+Consider $w$, the lowest critical vertex in $\PP$. Note that $w$ is a parent of $v$. When $w$ was processed, it was set to $\h(c')$, for
+some color $c'$. At this stage, all edges in $\PP$ above $w$ were also colored $c'$ (just apply the argument
+above for $w$). Since there are no critical points below $w$, all edges in $\PP$ below $w$
+are also colored $c'$ (the painting property). This means that the edges connecting $\PP$
+are also colored $c'$. So $c' = \rep(c)$, and $\h(\rep(c)) = w$ is a parent of $v$.
+
+This argument can be repeated for all such components $\PP$ to complete the proof.
+\end{proof}
 
@@
 
-\begin{claim} \label{clm:ripe} For ripe vertex $v$, $\{\h(\rep(c)) | c \in \col(v)\}$ is exactly
-the set of parents of $v$.
-\end{claim}
-
-\begin{proof} \Sesh{proof is a little sketchy.}
-Suppose $v$ is ripe. For any $c$ touching $v$, $v$ is the highest in $T(\rep(c))$.
-Consider some such $\rep(c)$, for an edge connecting to component $\PP$ of $\MM^+_v$.
-Either all edges in $\PP$ are colored $\rep(c)$ or there exists critical vertex $w \in \PP$
-fully touched by $\rep(c)$ and another color. In the latter case, $w$ has not been processed,
-so $w \in T(\rep(c))$ (contradiction). Therefore, all edges in $\PP$ are colored $\rep(c)$.
-Suppose there is some critical vertex in $\PP$. If it is not processed yet, it will
-be in $T(\rep(c))$, which is not possible. Thus, all critical vertices in $\PP$ have been processed.
-
-Consider $w$, the lowest critical vertex in $\PP$. Note that $w$ is a parent of $v$. When $w$ was processed, it was set to $\h(c')$, for
-some color $c'$. At this stage, all edges in $\PP$ above $w$ were also colored $c'$ (just apply the argument
-above for $w$). Since there are no critical points below $w$, all edges in $\PP$ below $w$
-are also colored $c'$ (the painting property). This means that the edges connecting $\PP$
-are also colored $c'$. So $c' = \rep(c)$, and $\h(\rep(c)) = w$ is a parent of $v$.
-
-This argument can be repeated for all such components $\PP$ to complete the proof.
-\end{proof}
-
 \newcommand{\pathTree}{P_{\mathcal{S}}}

### 2014-10-28 paint-contour-tree.tex

--- 2014-10-27 paint-contour-tree.tex
@@
 exists some partition $P'$ such that the running time of our algorithm is 
-$O(\sum_{p\in P'} |p|\log|p| + N\alpha(N))$, where $N$ is total complexity of the input manifold 
+$O(\sum_{p\in P'} |p|\log|p| + N\alpha(N))$, where $N$ is the total complexity of the input manifold 
 (and hence this is optimal on all instances requiring $\Omega(N\alpha(N))$ time).
@@
 In worst case analysis an algorithm's running time is judged in terms of its worst 
-case behavior over all inputs of a given size.  While there are benefits to such a 
-pessimistic view, it fails to capture the whole story.  As such, many alternatives 
+case behavior over all inputs of a given size.  While there are benefits to this viewpoint, 
+such as robustness,  it fails to capture the whole story.  As such, many alternatives 
 have been considered.  For example, one can make assumptions on the distribution of 
@@
 We will fix such a simplicial complex $\MM$, with the dominant minimum $m^*$. 
-We observe that the contours trees of minimum dominant manifolds have a relatively simple description.
-This description is directly inspired by (and proven using) the \emph{join tree} of~\cite{csa-cctad-00}.
+The set of non-dominant minima is denoted by $M$.
+For vertex $v$, we use $\MM^+_v$ to denote the simplicial complex obtained by only
+keeping vertices $u$ such that $f(u) > f(v)$. Analogously, define $\MM^-_v$. Note that $\MM^+_v$ may contain
+numerous connected components.
+
+The main theorem of this section asserts that the contours trees of minimum dominant manifolds have a simple description.
+This description is based on \emph{join} and \emph{split} trees, as given by~\cite{csa-cctad-00}.
+Conventially, all edges are directed from higher to lower function value.
 We first explain what these trees are.
-
-For vertex $v$, we use $\MM^+_v$ to denote the simplicial complex obtained by only
-keeping vertices $u$ such that $f(u) > f(v)$. Analogously, define $\MM^-_v$. Note that $\MM^+$ may contain
-numerous connected components. The following definition of join and split trees were given by~\cite{csa-cctad-00}.
-Conventially, all edges are directed from higher to lower function value.
 
@@
 by looking at $\MM^-_v$ (or alternately, by taking the join tree of the inversion of $\MM$).
-\end{definition}
-
-Note that unlike the contour tree, all vertices in $\MM$ are present in $\cJ(\MM)$ and $\cS(\MM)$. 
-Some basic facts about these trees. All outdegrees in $\cJ(\MM)$ and all indegrees in $\cS(\MM)$ are at most $1$.
+
+The critical join tree $\jc(\MM)$ is built on $V(\MM) \setminus M$. 
+The directed edge $(u,v)$ is present when $u$ is the smallest valued vertex in $V(\MM) \setminus M$ in a connected component of $\MM^+_v$
+and $v$ is connected to this component (in $\MM$). 
+\end{definition}
+
+Some basic facts about these trees. 
+%Note that unlike the contour tree, all vertices in $\MM$ are present in $\cJ(\MM)$ and $\cS(\MM)$. 
+All outdegrees in $\cJ(\MM)$ and all indegrees in $\cS(\MM)$ are at most $1$.
 Indegree $2$ vertices in $\cJ(\MM)$ are joins, and outdegree $2$ vertices in $\cS(\MM)$ are splits.
 Leaves in $\cJ(\MM)$ are maxima, and leaves in $\cS(\MM)$ are minima.
-
 The key observation is that $\cS(\MM)$ is trivial for minimum dominant $\MM$.
@@
 
-Now we come to the main algorithmic contribution. We design a new algorithm for computing contour
-trees of a minimum dominant simplicial complex $\MM$. As part of the initialization, we determine
+The main algorithmic contribution is a new algorithm for computing join trees of any triangulated simplicial
+complex $\MM$.
+
+As part of the initialization, we determine
 all critical points in $\MM$. For any vertex $v$, we can look at $f(v) \pm \delta$-contours 
@@
 {\bf Painting:} The central tool is a notion 
-of \emph{painting} of $\MM$. Initially associate a color with each maximum. Denoting the set of maxima by $X$,
-this is an arbitrary bijection $\chi:X \mapsto [|X|]$. Imagine there being a large
-can of paint of color $\chi(x)$ at maxima $x$. We will spill different paint from each maximum and watch it flow down.
+of \emph{painting} of $\MM$. Initially associate a color with each maximum. Imagine there being a large
+can of paint of a distinct color at each maxima $x$. We will spill different paint from each maximum and watch it flow down.
 This is analogous to the raining in the previous section, but paint is a much more viscous liquid.
 \emph{So paint only flows down edges, and it does not color the interior of facets.} Furthermore, paints
-do not mix, so eventually, every vertex and edge of $\MM$ gets a unique color. This process (and indeed the entire algorithm)
+do not mix, so every edge of $\MM$ gets a unique color. This process (and indeed the entire algorithm)
 works purely on the 1-skeleton of $\MM$, which is just a graph.
 %
-\begin{definition} \label{def:paint} Let the 1-skeleton of $\MM$ be denoted $G(\MM) = (\cV,E)$.
+\begin{definition} \label{def:paint} Let the 1-skeleton of $\MM$ have edge set $E$ and maxima $X$.
 A  \emph{painting} of $\MM$ is a map $\chi:X \cup E \mapsto [|X|]$ with the following property. 

### 2014-10-28 paint-contour-tree.tex

@@
 
-As part of the initialization, we determine
-all critical points in $\MM$. For any vertex $v$, we can look at $f(v) \pm \delta$-contours 
-in $f|_{B_\eps(v)}$ to determine the up and down degrees. This determines the criticality (and type)
-of $v$. This requires $O(|\MM|)$ time, since every facet is looked at proportional to the number of its
-vertices.
+%
+%As part of the initialization, we determine
+%all critical points in $\MM$. For any vertex $v$, we can look at $f(v) \pm \delta$-contours 
+%in $f|_{B_\eps(v)}$ to determine the up and down degrees. This determines the criticality (and type)
+%of $v$. This requires $O(|\MM|)$ time, since every facet is looked at proportional to the number of its
+%vertices.
 
 {\bf Painting:} The central tool is a notion 
-of \emph{painting} of $\MM$. Initially associate a color with each maximum. Imagine there being a large
+of \emph{painting} $\MM$. Initially associate a color with each maximum. Imagine there being a large
 can of paint of a distinct color at each maxima $x$. We will spill different paint from each maximum and watch it flow down.
@@
  Consider an edge $e$. There exists a descending path from some maximum $x$ to $e$
-	consisting of edges in $E$, such that all edges along this path have the same color as $x$. (The maximum $x$
-	is well-defined to be the unique maximum with the same color as $e$.)
+	consisting of edges in $E$, such that all edges along this path have the same color as $x$.
 
@@
 %
-Paintings are not unique. It is straightforward to construct an initial painting of $\MM$ in $O(|\MM|)$ time, 
-using a descending BFS from each maximum that does not explore previously colored edges. 
-As the algorithm proceeds, the bijection property will not hold, though we always maintain a valid painting.
-
-{\bf Colors of vertices:} Note that a painting only colors edges, not vertices. We associate certain
+%Paintings are not unique. It is straightforward to construct an initial painting of $\MM$ in $O(|\MM|)$ time, 
+%using a descending BFS from each maximum that does not explore previously colored edges. 
+%As the algorithm proceeds, the bijection property will not hold, though we always maintain a valid painting.
+
+A painting only colors edges, not vertices. We associate certain
 sets of colors with vertices.
 
-\begin{definition} \label{def:color-set} Fix a painting $\chi$, and a color $c$.
-A vertex $v$ is \emph{touched by $c$} if $x$ is incident to a $c$-colored
+\begin{definition} \label{def:color-set} Fix a painting $\chi$ and vertex $v$.
+\begin{asparaitem}
+	 \item An \emph{up-star} of $v$ is the set of edges that all connected to a fixed component of $\MM^+_v$.
+	 \item A vertex $v$ is \emph{touched by color $c$} if $x$ is incident to a $c$-colored
 	 edge with $v$ at the lower endpoint. For $v$, $\col(v)$ is the set of colors that touch $v$.
-\end{definition}
-
+	 \item A color $c \in \col(v)$ \emph{fully touches} $v$ if all edges in an up-star are colored $v$.
+\end{asparaitem}
+\end{definition}
+
+Each non-maxima $v$ participates in at least $1$ and at most $2$ up-stars (the latter iff
+$v$ is a join).
+
+\ignore{
 Abusing notation, we say a contour $\phi$ is of color $c$ if all edges that intersect $\phi$ have color $c$.
@@
 \begin{definition} \label{def:color-full} Consider vertex $v$. An \emph{up-star} of $v$ is the set 
-of edges that all connect to a fixed component of $\MM^+_v$. (Each $v$
-participates in at mot $2$ up-stars.) A color $c \in \col(v)$ \emph{fully touches} $v$
-if all edges in an up-star are colored $v$.
+of edges that all connected to a fixed component of $\MM^+_v$. 
 %and let $\phi$ be a $(f(v)+\delta)$-contour in
@@
 %and $\fcol(v)$ can
-can be initially determined in $O(|\MM|)$ time, by a simple linear scan.
-
-\begin{lemma} \label{lem:full} Suppose vertex $v$ is connected to a component of $\MM^+_v$
-by edge $e$ of color $c$. Either all edges in $\MM^+_v$ are colored $c$, or there
-exists critical vertex $w \in \MM^+_v$ fully touched by $c$ and touched by another color.
-%
-%and take $(f(v)+\delta)$-contour $\phi$ intersecting
-%$B_\eps(v)$. Suppose the edges incident to $v$ intersecting $\phi$ do not have the same color. For any such
-%color $c$, there exists a critical vertex $w$ strictly higher than $v$ such that $c \in \fcol(v)$.
-\end{lemma}
-
-\begin{proof} Since $e$ has color $c$,
-there must exist vertices in $\MM^+_v$ touched by $c$. Consider the highest
-vertex $w$ in $\MM^+_v$ that is touched by $c$ and some other color. If no such vertex exists,
-this means all edges incident to a vertex touched by $c$ are colored $c$. By walking through
-$\MM^+_v$, we deduce that all edges are colored $c$.
-
-So assume $w$ exists. Take the $(f(w)+\delta)$-contour $\phi$ that intersects $B_\eps(v)$
-and intersect some $c$-colored edge incident to $w$. Note that all edges intersecting $\phi$ are also colored $c$,
-since $w$ is the highest vertex to be touched by $c$ and some other color. (Take the path of $c$-colored
-edges from the maximum to $w$. For any point on this path, the contour passing through this point must
-be colored $c$.) Hence $c$ fully touches $w$. 
-But $w$ is touched by another color, and the corresponding edge cannot intersect $\phi$. So $w$
-must have up-degree $2$ and is critical.
-\end{proof}
+be initially determined in $O(|\MM|)$ time, by a simple linear scan.
+}
 

### 2014-10-28 paint-contour-tree.tex
minor section 1 changes
running time proof added

@@
 
-This instance optimality of our algorithm then follows, as we also show that there is a particular path decomposition 
+The instance optimality of our algorithm then follows, as we also show that there is a particular path decomposition 
 whose cost upper bounds our.  Specifically, we assign a weight to each node in the tree which is determined by 
@@
 vertices are splits, leaves are minima, and the global maximum is the root.
-The key observation is that $\cS(\MM)$ is trivial for minimum dominant $\MM$.
+The key observation is that $\cS(\MM)$ is trivial for a minimum dominant $\MM$.
 
@@
 		\item Let $v$ be head of queue. If $v$ is leaf in $\cJ$, consider its edge in $\cJ$. Otherwise
-		consider edge in $\cS$. In either case, let edge by $(v,w)$.
+		consider its edge in $\cS$. In either case, denote the edge by $(v,w)$.
 		\item Insert $(v,w)$ in $\cA(\MM)$. 
@@
 
-\begin{definition} The critical join tree $\jc(\MM)$ is built on the set $V'$ of all
-critical points that are \emph{not} non-dominant minima. 
+\begin{definition} 
+\label{def:criticalJoin}
+The critical join tree $\jc(\MM)$ is built on the set $V'$ of all
+critical points other than the non-dominant minima. 
 The directed edge $(u,v)$ is present when $u$ is the smallest valued vertex in $V'$ in a connected component of $\MM^+_v$
@@
 Our algorithm incrementally builds $\jc(\MM)$ from the leaves (maxima) to the root (dominant minimum). 
-Let us focus on an initial painting, where each colors has 1-1 correspondence with the maxima.
+Let us focus on an initial painting, where the colors have 1-1 correspondence with the maxima.
 Consider two sibling leaves $\ell_1, \ell_2$ and their common parent $v$. The leaves are maxima,
@@
 we replace all red and blue edges by purple edges. This indicates that the red and blue maxima have been
-handled. Imagine flattening the red and blue mounds till $v$, so now that erstwhile join $v$
-is a new maxima. Suppose purple paint was poured from $v$. In terms on $\jc(\MM)$, this
-is equivalent to remove leaves $\ell_1$ and $\ell_2$, now making $v$ as a leaf.
+handled. Imagine flattening the red and blue mounds until reaching $v$, so that the former join $v$
+is now a new maxima. Suppose purple paint was poured from $v$. In terms of $\jc(\MM)$, this
+is equivalent to removing leaves $\ell_1$ and $\ell_2$, making $v$ a new leaf.
 Alternately, $\jc(\MM)$ has been constructed up to $v$, and it remains to determine $v$'s parent.
@@
 What happens when finally blue and red join at $v$? We merge the two colors, but now have
-a blue heap and red queues of critical vertices. We also need to merge the priority queues
+blue and red queues of critical vertices. We also need to merge the priority queues
 to get a consistent painting. And that decides the use of \emph{binomial heaps} to store the
@@
 be super-constant numbers of copies of a vertex, and handling all these copies would lead
-to an extra overhead. We show that it suffices to simply put each vertex $v$ into at most two heaps,
-one for each up-star. We are guaranteed that when $v$ needs to processed, its up-stars
+to extra overhead. We show that it suffices to simply put each vertex $v$ into at most two heaps,
+one for each up-star. We are guaranteed that when $v$ needs to be processed, its up-stars
 automatically have one color, because of all the color merges that previously occured.
@@
 
-\begin{claim} \label{clm:upstar} Consider ripe vertex $v$ and take the up-star connecting
+\begin{claim} \label{clm:upstar} Consider a ripe vertex $v$ and take the up-star connecting
 to some component of $\MM^+_v$. All edges in this component and the up-star have the same color.
@@
 
-\begin{proof} Let $c$ be the color of some edge in this up-star
+\begin{proof} Let $c$ be the color of some edge in this up-star.
 By ripeness, $v$ is the highest in $T(\rep(c))$.
@@
 
-In this section, we set up a framework to analyze the time taken to compute a merge tree.
-
-\begin{definition}
- We say that a contour tree of a $2$-manifold is a \emph{merge tree} if it has a well defined root vertex such that height values 
- strictly increase along any root to leaf path.
-\end{definition}
-
-\begin{observation}
- The contour tree of a minimum dominant manifold (see \Def{dom}) is close to a merge tree.  Specifically, there is a root such that on any root 
- to leaf path height values strictly increase, with the only exception being potentially at the leaves themselves.  As such, up to constant 
- factors, any lower bound for computing the merge tree is a lower bound for computing the contour tree of a minimum dominant
- manifold (i.e. intuitively just clip off the leaves or shift them to lie above their parent).
-\end{observation}
-
-Any tree $T$ considered in this section will be a rooted full binary tree\footnote{Note 
+\Ben{All merge tree references need to be switched to join.  Also, need to make sure that not assuming the bin trees are full was safe to drop.
+Also, the parent and child relationship is the opposite in this section as in earlier ones.}
+
+In this section, we set up a framework to analyze the time taken to compute a critical join tree $\jc(\MM)$ (see \Def{criticalJoin}).
+%
+% \begin{definition}
+%  We say that a contour tree of a $2$-manifold is a \emph{merge tree} if it has a well defined root vertex such that height values 
+%  strictly increase along any root to leaf path.
+% \end{definition}
+% 
+% \begin{observation}
+%  The contour tree of a minimum dominant manifold (see \Def{dom}) is close to a merge tree.  Specifically, there is a root such that on any root 
+%  to leaf path height values strictly increase, with the only exception being potentially at the leaves themselves.  As such, up to constant 
+%  factors, any lower bound for computing the merge tree is a lower bound for computing the contour tree of a minimum dominant
+%  manifold (i.e. intuitively just clip off the leaves or shift them to lie above their parent).
+% \end{observation}
+%
+As such any tree $T$ considered in this section will be a rooted binary tree\footnote{Note 
 that technically the trees considered should have a leaf vertex hanging below the root of this 
-full binary tree in order to represent the global minimum of the surface.  This vertex is 
+in order to represent the global minimum of the complex.  This vertex is 
 (safely) ignored to simplify the presentation.} where the height of a vertex is its distance 
@@
 
+Note that the subroutine $\init(\MM)$ from \Sec{algo} naturally defines a leaf assignment to $\jc(\MM)$ 
+according to the priority queue for each upstar we put a given vertex in.
+
 \begin{lemma}
-The running time of our algorithm is $O(\sum_{v \in T} \log |H_v|)$.
+Let $\MM$ be a simplicial complex with $m$ edges.  For every vertex in $\jc(\MM)$, 
+let $H_v$ be defined by the leaf assignment determined by $\init(\MM)$.
+The running time of our algorithm is $O(m\alpha(m) + \sum_{v \in \jc(\MM)} \log |H_v|)$.
 % where $T$ is the merge tree of the given input, 
@@
 
-\begin{proof} \Sesh{add}
-
+\begin{proof}
+First we look at the initialization procedure $\init(\MM)$.  This procedure runs in $O(m)$ time (note that as the input complex is connected, $N=O(m)$).
+Indeed, the painting procedure consists of several BFS's but as each vertex is only explored by one of the BFS's, it is linear time overall.
+Determining the critical points is a local computation on the neighborhood of each vertex as so is linear (i.e. each edge is viewed at most twice).
+Finally, each vertex is inserted into at most two heaps and so initializing the heaps takes linear time in the number of vertices.
+
+Now consider the union-find operations performed by $\init$ and $\update$.  
+Initially the union find data structure has a singleton component for each leaf (and no new components are ever created), 
+and so each union-find operation takes $O(m)$ time.
+The $\update(K)$ procedure performs at most one find for each edge and no unions.  
+$\init(\MM)$ performs one union and at most two finds for each vertex.  Therefore the total number of union find 
+operations is $O(m)$.
+
+For the remaining operations, observe that every time $\update$ is called a vertex is pushed onto the stack and each
+vertex can only be pushed onto the stack once (since the only way it leaves the stack is by being processed). 
+Therefore the total running time due to $\update$ is linear (ignoring the find operations).
+
+What remains is the time it takes process a vertex $v$ in $\init(\MM)$.  
+In order to process a vertex there are a few constant time operations, union-find operations, and queue operations.
+Therefore the only thing we must bound are the queue operations.
+Let $v$ be a vertex in $\jc(\MM)$, and let $p_1$ and $p_2$ be its parents (the same argument holds if $v$ has only one parent).
+At the time $v$ is handled, the colors and queues of all vertices in a given component of $\MM^+_v$ have merged together.
+In particular, when $v$ is handled we know it is ripe and so all vertices above $v$ in each component of $\MM^+_v$ have been processed, implying 
+these merged queues are the queues of the current colors of $p_1$ and $p_2$.  Again since $v$ is ripe, it must be on the top of these queues and so 
+the only vertices of left in these queues are those in $H_{p_1}$ and $H_{p_2}$.
+
+Now when $v$ is handled, three queue operations are performed.  Specifically, $v$ is removed from the queues of $p_1$ and $p_2$, and then the queues are 
+are merged together.  By the above arguments the sizes of the queues for each of these operations are $H_{p_1}$, $H_{p_2}$, and $H_v$, respectively.  
+As merging and deleting takes logarithmic time in the heap size for binomial heaps, the claim now follows.
 \end{proof}

### 2014-10-28 paint-contour-tree.tex
correctness updated

@@
 We required the notions of \emph{join} and \emph{split} trees, as given by~\cite{csa-cctad-00}.
-Conventially, all edges are directed from higher to lower function value. We will use the terminology
-``parent" for an adjacent node that is higher. Because of the orientation, a node
-may have multiple parents.
-%We first explain what these trees are.
+Conventially, all edges are directed from higher to lower function value.
 
@@
 vertices are splits, leaves are minima, and the global maximum is the root.
+%
+As these trees rooted, we can use ancestor-descendant terminology.  
+Specifically, for two adjacent vertices $u$ and $v$, $u$ is the parent of $v$ if $u$ is closer to the root 
+(i.e. each node can have at most one parent, but can have two children).
+
 The key observation is that $\cS(\MM)$ is trivial for a minimum dominant $\MM$.
@@
 when some join $v$ has both up-stars fully colored, the corresponding maxima (of those colors) are exactly
-the parents of $v$ in $\jc(\MM)$. To proceed further, we ``merge" the colors red and blue into a new color, purple. In other words,
+the children of $v$ in $\jc(\MM)$. To proceed further, we ``merge" the colors red and blue into a new color, purple. In other words,
 we replace all red and blue edges by purple edges. This indicates that the red and blue maxima have been
@@
 			\item Add edge $(\h(c'),h)$ to $\jc(\MM)$.
-			\item Delete $h$ from $T(c')$ (if present).
+			\item Delete $h$ from $T(c')$.
 		\end{compactenum}
@@
 	\begin{compactenum}
-		\item Find $c \in \col(h)$ such that $h$ is not present or not the highest in $T(\rep(c))$.
+		\item Find $c \in \col(h)$ such that $h$ is not the highest in $T(\rep(c))$.
 		\item Push the highest of $T(\rep(c))$ onto $\st$, and update head $h$.
@@
 A few simple facts:
-\begin{asparaitem}
+\begin{compactitem}
 	\item At all times, the colors form a valid painting.
@@
 	\item After $v$ is processed, all edges incident to $v$ have the same color (technically, same representative). 
-	\item When $\update(\st)$ terminates, the head $h$ is ripe. 
-\end{asparaitem}
+%	\item When $\update(\st)$ terminates, the head $h$ is ripe. 
+	\item Vertices on the stack are in increasing height order. 
+\end{compactitem}
 
@@
 
-\section{Proving correctness} \label{sec:correct}
+\subsection{Proving correctness} \label{sec:correct}
 
@@
 
-\begin{lemma} \label{lem:full} Suppose vertex $v$ is connected to a component of $\MM^+_v$
-by edge $e$ of color $c$. Either all edges in $\MM^+_v$ are colored $c$, or there
-exists critical vertex $w \in \MM^+_v$ fully touched by $c$ and touched by another color.
+\begin{lemma} \label{lem:full} Suppose vertex $v$ is connected to a component $\PP$ of $\MM^+_v$
+by edge $e$ of color $c$. Either all edges in $\PP$ are colored $c$, or there
+exists critical vertex $w \in \PP$ fully touched by $c$ and touched by another color.
 %
@@
 \begin{proof} Since $e$ has color $c$,
-there must exist vertices in $\MM^+_v$ touched by $c$. Consider the highest
-vertex $w$ in $\MM^+_v$ that is touched by $c$ and some other color. If no such vertex exists,
-this means all edges incident the a vertex touched by $c$ are colored $c$. By walking through
-$\MM^+_v$, we deduce that all edges are colored $c$.
+there must exist vertices in $\PP$ touched by $c$. Consider the highest
+vertex $w$ in $\PP$ that is touched by $c$ and some other color. If no such vertex exists,
+this means all edges incident to a vertex touched by $c$ are colored $c$. By walking through
+$\PP$, we deduce that all edges are colored $c$.
 
@@
 
+\begin{corollary}
+\label{cor:terminate}
+ Each time $\update(K)$ is called, it terminates with a ripe vertex on top of the stack.
+\end{corollary}
+\begin{proof}
+ $\update(K)$ is only called if there are unprocessed vertices remaining, and so by the time we reach step 3 in 
+ $\update(K)$, the stack has some unprocessed vertex $h$ on it.  
+ If $h$ is ripe, then we are done, so suppose otherwise.
+
+ Let $\PP$ be one of the components of $\MM^+_h$.  By construction, $h$ was put in the heap of some initial 
+ adjacent color $c$.  Therefore, $h$ must be in the current heap of $\rep(c)$ (since elements are only removed 
+ from heaps when being processed).  Now by \Lem{full}, either all edges in $\PP$ are colored $\rep(c)$ or  
+ there is some vertex $w$ fully touched by $\rep(c)$ and some other color.  
+ The former case implies that if there are any unprocessed vertices in $\PP$ then they are all in $T(\rep(c))$, 
+ implying that $h$ is not the highest vertex.  Otherwise, all the vertices in $\PP$ have been processed.  
+ However, it cannot be the case that all vertices in all components of $\MM^+_h$ have already been processed, 
+ since this would imply that $h$ was ripe and so $h$ so can apply the same argument to the other non-fully processed component.
+
+ Now consider the latter case, where we have a non-monochromatic vertex $w$.
+ In this case $w$ cannot have been processed (since after being processed it is touched only by one color), 
+ and so it must be in $T(\rep(c))$ since it must be in some heap of a color in each upstar (and one upstar is entirely colored $\rep(c)$).
+ As $w$ lies above $h$ in $\MM$, this implies $h$ is not on the top of this heap.
+\end{proof}
+
 We prove a series of claims that will lead to the correctness proof.
@@
 \begin{proof} More generally, we prove the following: all outputs on processed vertices
-are edges of $\cC(\MM)$ and for any color $c$, $\h(c)$ is the lowest processed vertex
-of that color. We prove by induction on the processing order. Initially, this is trivially
-true, handling these base case. For the induction step, consider the situation
+are edges of $\jc(\MM)$ and for any color $c$, $\h(c)$ is the lowest processed vertex
+of that color. We prove this by induction on the processing order. The base case is is trivially
+true, as initially the processed vertices and attachments of the color classes are the set of maxima.
+For the induction step, consider the situation
 when $v$ is being processed.
 
-The vertex $v$ is ripe. Take any up-star, and the corresponding component $\PP$
+Since $v$ is being processed, we know by \Cor{terminate} that it is ripe. Take any up-star of $v$, and the corresponding component $\PP$
 of $\MM^+_v$ that it connects to. By \Clm{upstar}, all edges in $\PP$ and the up-star
@@
 it must be in $T(c)$, which violates the ripeness of $v$.
-Thus, all critical vertices in $\PP$ have been processed.
-
-Consider $w$, the lowest critical vertex in $\PP$. Note that $w$ is a parent of $v$. When $w$ was processed, it was set to $\h(c')$, for
-some color $c'$. At this stage, all edges in $\PP$ above $w$ were also colored $c'$ (just apply the argument
-above for $w$). Since there are no critical points below $w$, all edges in $\PP$ below $w$
-are also colored $c'$ (the painting property). This means that the edges connecting $\PP$
-are also colored $c'$. So $c' = \rep(c)$, and $\h(\rep(c)) = w$ is a parent of $v$.
-
-This argument can be repeated for all such components $\PP$ to complete the proof.
-\end{proof}
-
-The following crucial claim demonstrates the utility of the stack.
-
-\begin{claim} \label{clm:stack} Vertices are always pushed onto $\st$ in increasing order of height.
-\end{claim}
-
-\begin{proof} Consider a single push onto the stack and the current painting. We have $c \in \col(h)$, as given in pseudocode of $\update$.
-If $h \in T(\rep(c))$, then obviously the next head is higher. So assume that $h \notin T(\rep(c))$.
-There is obviously an edge of color $\rep(c)$ touching $h$ connected to a component $\PP$ of $\MM^+_v$. 
-From the initialization, $h \in T(\rep(c'))$, for some $c'$ touching $h$ (where $c'$ is the color
-of some edge connecting $h$ to $\PP$).
-Applying \Lem{full},  either all edges in $\PP$ are colored $\rep(c)$ or 
-there exists critical vertex $w \in \PP$ fully touched by $\rep(c)$ and another color.
-In the former case, all edges connecting $\PP$ to $h$ are also colored $\rep(c)$
-(since this is a painting). Hence, $\rep(c') = \rep(c)$ and $h \in T(\rep(c))$.
-As we assumed this is not the case, we conclude the existence of critical vertex $w$.
-Since $w$ is fully touched by $\rep(c)$, $w \in T(\rep(c))$. Since $w$
-is touched by another color, $w$ has not been processed. So the highest of $T(\rep(c))$
-is higher than $h$, as desired.
-\ignore{
-If all edges touching $h$ that connect to this component are also colored $\rep(c)$, then
-In Case (i), the increase in height is obvious.
-Consider Case (ii). There is a vertex $v$ touched by color $c \notin \fcol(v)$. Consider the highest
-vertex $w$ that is touched by $c$ and some other color. Clearly such a vertex exists, since the condition
-is also satisfied by $v$. Take the $(f(w)+\delta)$-contour $\phi$ that intersects $B_\eps(v)$
-and intersect some $c$-colored edge incident to $w$. Note that $\phi$ is also colored $c$,
-since $w$ is the highest vertex to be touched by $c$ and some other color. (Take the path of $c$-colored
-edges from the maximum to $w$. For any point on this path, the contour passing through this point must
-be colored $c$.) Hence, $c \in \fcol(w)$ and $w \neq v$. Furthermore, $w$ is a critical point,
-since $w$ is touched by another color. The corresponding edge cannot intersect $\phi$, so $w$
-must have up-degree $2$.
-All in all, when a new vertex is pushed onto the stack, it is always higher than the previous head.
-}
-\end{proof}
+Thus, all critical vertices in $\PP$ have been processed, and so by the induction hypothesis, the restriction of $\jc(\MM)$ to $\PP$ has been correctly computed.
+Additionally, since all critical vertices in $\PP$ have processed, they all have the same color $c$ of the lowest critical vertex in $\PP$.
+Thus by the strengthened induction hypothesis, this lowest critical vertex is $\h(c)$.
+
+If there is another component of $\MM^+_v$, the same argument implies 
+the lowest critical vertex in this component is $\h(c')$ (where $c'$ is the color of edges in the respective component).
+Now by the definition of $\jc(\MM)$, the critical vertex $v$ connects to the lowest critical vertex in each component of $\MM^+_v$, 
+and so by the above $v$ should connect to $\h(c)$ and $\h(c')$, which is precisely what $v$ is connected to by $\build(\MM)$.
+Moreover, $\build$ merges the colors $c$ and $c'$ and correctly sets $v$ to be the attachment, 
+as $v$ is the lowest processed vertex of this merged color (as by induction $\h(c)$ and $\h(c')$ where the lowest vertices before merging colors).
+
+% Consider $w$, the lowest critical vertex in $\PP$. Note that $w$ is a parent of $v$. When $w$ was processed, it was set to $\h(c')$, for
+% some color $c'$. At this stage, all edges in $\PP$ above $w$ were also colored $c'$ (just apply the argument
+% above for $w$). Since there are no critical points below $w$, all edges in $\PP$ below $w$
+% are also colored $c'$ (the painting property). This means that the edges connecting $\PP$
+% are also colored $c'$. So $c' = \rep(c)$, and $\h(\rep(c)) = w$ is a parent of $v$. 
+%
+%This argument can be repeated for all such components $\PP$ to complete the proof.
+\end{proof}
+
+\begin{theorem}
+ Given an input complex $\MM$, $\build(\MM)$ terminates and outputs $\jc(\MM)$.
+\end{theorem}
+\begin{proof}
+ First observe that each vertex can be processed at most once by $\build(\MM)$.  By \Cor{terminate}, we know that as long as there 
+ is an unprocessed vertex, $\update(K)$ will be called and will terminate with a ripe vertex which is ready to be processed. 
+ Therefore, eventually all vertices will be processed, and so by \Clm{process} the algorithm will terminate having 
+ computed $\jc(\MM)$.
+\end{proof}
+
+% The following crucial claim demonstrates the utility of the stack.
+% 
+% \begin{claim} \label{clm:stack} Vertices are always pushed onto $\st$ in increasing order of height.
+% \end{claim}
+% 
+% \begin{proof} Consider a single push onto the stack and the current painting. We have $c \in \col(h)$, as given in pseudocode of $\update$.
+% If $h \in T(\rep(c))$, then obviously the next head is higher. So assume that $h \notin T(\rep(c))$.
+% There is obviously an edge of color $\rep(c)$ touching $h$ connected to a component $\PP$ of $\MM^+_v$. 
+% From the initialization, $h \in T(\rep(c'))$, for some $c'$ touching $h$ (where $c'$ is the color
+% of some edge connecting $h$ to $\PP$).
+% Applying \Lem{full},  either all edges in $\PP$ are colored $\rep(c)$ or 
+% there exists critical vertex $w \in \PP$ fully touched by $\rep(c)$ and another color.
+% In the former case, all edges connecting $\PP$ to $h$ are also colored $\rep(c)$
+% (since this is a painting). Hence, $\rep(c') = \rep(c)$ and $h \in T(\rep(c))$.
+% As we assumed this is not the case, we conclude the existence of critical vertex $w$.
+% Since $w$ is fully touched by $\rep(c)$, $w \in T(\rep(c))$. Since $w$
+% is touched by another color, $w$ has not been processed. So the highest of $T(\rep(c))$
+% is higher than $h$, as desired.
+% \ignore{
+% If all edges touching $h$ that connect to this component are also colored $\rep(c)$, then
+% In Case (i), the increase in height is obvious.
+% Consider Case (ii). There is a vertex $v$ touched by color $c \notin \fcol(v)$. Consider the highest
+% vertex $w$ that is touched by $c$ and some other color. Clearly such a vertex exists, since the condition
+% is also satisfied by $v$. Take the $(f(w)+\delta)$-contour $\phi$ that intersects $B_\eps(v)$
+% and intersect some $c$-colored edge incident to $w$. Note that $\phi$ is also colored $c$,
+% since $w$ is the highest vertex to be touched by $c$ and some other color. (Take the path of $c$-colored
+% edges from the maximum to $w$. For any point on this path, the contour passing through this point must
+% be colored $c$.) Hence, $c \in \fcol(w)$ and $w \neq v$. Furthermore, $w$ is a critical point,
+% since $w$ is touched by another color. The corresponding edge cannot intersect $\phi$, so $w$
+% must have up-degree $2$.
+% All in all, when a new vertex is pushed onto the stack, it is always higher than the previous head.
+% }
+% \end{proof}
 
@@
 
-\Ben{All merge tree references need to be switched to join.  Also, need to make sure that not assuming the bin trees are full was safe to drop.
-Also, the parent and child relationship is the opposite in this section as in earlier ones.}
+\Ben{All merge tree references need to be switched to join.  Also, need to make sure that not assuming the bin trees are full was safe to drop.}
 

### 2014-10-29 paint-contour-tree.tex
making changes to intro

@@
 
+\newcommand{\DD}{\mathbb{D}}
 \newcommand{\LL}{\mathbb{L}}
@@
 \begin{abstract}
+Consider a scalar field $f:\MM \mapsto \RR$, where $\MM$ is a simplicial mesh in $\RR^d$. 
+A level set, or contour, at value $v$ is a connected component of $f^{-1}(v)$. 
+As $v$ is changed, these contours change topology, merge into each other, or split.
+\emph{Contour trees} are concise representations of $f$ that track this contour
+behavior. The vertices of these trees are the critical points of $f$, where the gradient is zero.
+The edges represent changes in the topology of contours.
+It is a fundamental data structure in data analysis and visualization,
+and there is much past work (theoretical and practical) on algorithms for construct contour trees.
+
+Suppose $\MM$ has $n$ vertices, $N$ facets, and $t$ critical points.
+A classic result of Carr, Snoeyink, and Axen (2000) give an algorithm that takes $O(n\log n + N\alpha(N))$ time
+(where $\alpha(\cdot)$ is the inverse Ackermann function). A further improvement to $O(t\log t + N\alpha(N))$ time
+was given by Chiang et al. All these algorithms involve a global sort of the critical points,
+which is a major bottleneck for large datasets. Unfortunately, lower bounds of $\Omega(t\log t)$ also exist.
+
+We present the first algorithm that can avoid the global sort and has a refined time complexity that
+depends on the contour tree structure. Intuitively, if the tree is short and fat, we get significant improvements 
+in running time. For a partition of the contour tree into a set of descending paths, $P$,
+our algorithm runs in $O(\sum_{p\in P} |p|\log|p| + N\alpha(N))$. If the contour tree is a balanced
+binary tree, our running time is $O(n + N\alpha(N))$. Moreover, we prove a lower bound of
+exactly $\Omega(\sum_{p \in P} |p|\log |p|)$ for computing contour trees, thereby showing
+that the tree structure must play a role in the time complexity.
+
+\ignore{
+Moreover, we show there   
+exists some partition $P'$ such that the running time of our algorithm is 
+$O(\sum_{p\in P'} |p|\log|p| + N\alpha(N))$, where $N$ is the total complexity of the input manifold
+
+All previous contour tree algorithms involve a global sort 
+of the heights of all points or critical vertices. Thereby, these algorithms always pay $\Omega(t \log t)$.
+This is a major roadblock for efficiency, since sorting is the most expensive operation
+
 Reeb graphs are a fundamental topological structure which tracks the evolution of levels sets of a real valued function over a manifold.  
@@
 %matches the previous worst case $O(N+n\log n)$ bound when the tree is long and skinny.
+}
 
@@
 \section{Introduction}
+
+A large class of data, such as satellite terrain maps, physical simulations, combustion data, can
+be thought as a function $f: \RR^d \mapsto \RR$. Typically, a finite representation is given
+by consider $f$ to be piecewise linear over some mesh (simplicial complex) $\MM$ in $\RR^d$.
+\emph{Contour trees} are a specific topological data structure used to represent and visualize
+the function $f$.
+
+It is convenient to think of $f$ as a $(d+1)$-dimensional manifold, or alternately consider
+$\MM$ in $\RR^{d+1}$ with the last coordinate (height) given by $f$. Imagine sweeping the hyperplane $x_{d+1} = h$
+with $h$ going from $+\infty$ to $-\infty$. At every instance, the intersection of this plane 
+with $f$ gives a set of connected components, the \emph{contours} at height $h$. At the sweeping 
+proceeds various events occur: new contours are created or destroyed, contours merge into each other,
+split into new components, contours acquire or lose handles. These are the relevant events where 
+the topology of the level set changes. The contour tree is a concise representation of all these events.
+
+For convenience, imagine $f$ to be smooth. All points where the gradient of $f$ is zero are \emph{critical point},
+exactly where the topology of the contours change. The vertices of the contour tree are exactly the critical points.
+
+Contour trees and related topological data structures have been applied in analysis of fluid mixing, combustion simulations,
+and studying chemical systems~\cite{LaBe+06,BrWe+10,BeWe+11,BrWe+11,MaGr+11}.
 

### 2014-10-29 paint-contour-tree.tex
various changes

--- 2014-10-28 paint-contour-tree.tex
@@
     \includegraphics[width=.3\linewidth,height=.42\linewidth]{figs/trees2}
-    \caption{On the left, a surface will a balanced contour tree, but whose merge and split trees have long tails.  
-    On the right (in clockwise order from the top) the contour, split and merge trees.}
+    \caption{On the left, a surface will a balanced contour tree, but whose join and split trees have long tails.  
+    On the right (in clockwise order from the top) the contour, split and join trees.}
     \label{fig:sorting}
@@
 A useful corollary of this lemma shows that a contour actually splits the simplicial complex into
-two disconnected complices.
+two disconnected complexes.
 
 \begin{theorem} \label{thm:jordan} $\cut(\phi,\MM)$ consists of two
-disconnected simplicial complices.
+disconnected simplicial complexes.
 \end{theorem}
@@
 In this section, we describe a linear time procedure that partitions $\MM$ into special
-\emph{extremum dominant} simplicial complices.
+\emph{extremum dominant} simplicial complexes.
 
@@
 We define a simple \lift{} operation on the interface contours. Consider such a contour $\phi$ containing
-a merge vertex $y$. Take any dry increasing edge incident to $y$, and pick the point $z$ on this edge at height
+a join vertex $y$. Take any dry increasing edge incident to $y$, and pick the point $z$ on this edge at height
 $f(y) + \delta$. Let $\lift(\phi)$ be the unique contour through the regular point $z$. Note that $\lift(\phi)$ is dry.
@@
 \begin{claim} \label{clm:cut-int} Let $\phi$ be an interface contour. Then $\cut(\MM,\lift(\phi))$
-results in two disjoint simplicial complices, one consisting entirely of dry points.
+results in two disjoint simplicial complexes, one consisting entirely of dry points.
 \end{claim}
 
-\begin{proof} By \Thm{jordan}, $\cut(\MM,\lift(\phi))$ results in two disjoint simplicial complices. Let $\NN$ be the complex containing
+\begin{proof} By \Thm{jordan}, $\cut(\MM,\lift(\phi))$ results in two disjoint simplicial complexes. Let $\NN$ be the complex containing
 the point $x$ (the argument in $\wet(x,\MM)$), and let $\NN'$ be the other complex. 
@@
 We describe the main partitioning procedure that cuts a simplicial complex $\NN$ into extremum
-dominant complices. It takes an additional input of a maximum $x$. To initialize,
+dominant complexes. It takes an additional input of a maximum $x$. To initialize,
 we begin with $\NN$ set to $\MM$ and $x$ as an arbitrary maximum. One of the critical aspects
@@
 	\end{compactenum}
-	\item Output $\NN_{k+1}$ together with any complices output by recursive calls.
+	\item Output $\NN_{k+1}$ together with any complexes output by recursive calls.
 \end{compactenum}
@@
 First, by \Thm{jordan}, all the pieces on which $\rain$ is invoked are disjoint.
-Second, all recursive calls are made on dry complices.
+Second, all recursive calls are made on dry complexes.
 
@@
 \begin{proof} The operations performed are the raining and cutting. We can bound this by the total
-complexities of the wet complices and the new boundaries created. No facet is ever wet twice,
+complexities of the wet complexes and the new boundaries created. No facet is ever wet twice,
 and by \Lem{new-verts}, the total number of facets that every appear is $O(|\MM|)$. This also 
@@
 
-To prove our main theorem, we use the merging procedure from~\cite{csa-cctad-00} that constructs the contour tree
-from the join and split trees. 
+Using the above lemma we can now prove that computing the contour tree for a minimum dominant manifold 
+amounts to computing its join tree.  Specifically, to prove our main theorem, we rely on the correctness of the
+merging procedure from~\cite{csa-cctad-00} that constructs the contour tree from the join and split trees. 
 It actually construsts the \emph{augmented contour tree} $\cA(\MM)$, which
@@
 are put into a priority queue keyed by height, and vertices are handled in that order.
+
+\Ben{This figure will not show up in black and white printout.}
+
+\begin{figure}[h]\centering
+    \includegraphics[width=.27\linewidth]{figs/mounds}%
+    \hspace{.35in}
+    \includegraphics[width=.27\linewidth]{figs/mound2}%
+    \hspace{.4in}
+    \includegraphics[width=.22\linewidth]{figs/mound3}
+    \caption{On the left, three initial colors.  In the middle, yellow and blue merge into green.  On the right, green and red merge into purple.}
+    \label{fig:colors}
+\end{figure}
 
@@
 the simplicial complex $\NN$ where $\phi^-_i$'s are all boundary maxima. The algorithm will
-have already determined the contour trees of the simplicial complices
+have already determined the contour trees of the simplicial complexes
 where $\phi^+_i$ is a boundary maxima.
@@
 
+\begin{observation}
+\label{obs:twoQueues}
+ Each unprocessed vertex is always in exactly one queue of the colors in each up-star.  Specifically, 
+ for a given up-star of a vertex $v$, $\init(\MM)$ puts $v$ into the queue of exactly one color, say $c$.  As time goes on this queue 
+ may merge with other queues, but while unprocessed $v$ is only ever (and always) in the queue of $\rep(c)$,  
+ since $v$ is never added to a new queue and is not removed until it is processed.  
+ In particular, finding the queues of a vertex in $\update(K)$ requires at most two union find operations (assuming each vertex records its two colors from $\init(\MM)$)
+\end{observation}
+
 \medskip
@@
 \begin{lemma} \label{lem:full} Suppose vertex $v$ is connected to a component $\PP$ of $\MM^+_v$
-by edge $e$ of color $c$. Either all edges in $\PP$ are colored $c$, or there
+by edge $e$ which is currently colored $c$. Either all edges in $\PP$ are currently colored $c$, or there
 exists critical vertex $w \in \PP$ fully touched by $c$ and touched by another color.
@@
  Let $\PP$ be one of the components of $\MM^+_h$.  By construction, $h$ was put in the heap of some initial 
- adjacent color $c$.  Therefore, $h$ must be in the current heap of $\rep(c)$ (since elements are only removed 
- from heaps when being processed).  Now by \Lem{full}, either all edges in $\PP$ are colored $\rep(c)$ or  
+ adjacent color $c$.  Therefore, $h$ must be in the current heap of $\rep(c)$ (see \Obs{twoQueues}).  
+ Now by \Lem{full}, either all edges in $\PP$ are colored $\rep(c)$ or  
  there is some vertex $w$ fully touched by $\rep(c)$ and some other color.  
  The former case implies that if there are any unprocessed vertices in $\PP$ then they are all in $T(\rep(c))$, 
- implying that $h$ is not the highest vertex.  Otherwise, all the vertices in $\PP$ have been processed.  
+ implying that $h$ is not the highest vertex and a new higher up unprocessed vertex will be put on the stack for the next iteration of the while loop.  
+ Otherwise, all the vertices in $\PP$ have been processed.  
  However, it cannot be the case that all vertices in all components of $\MM^+_h$ have already been processed, 
- since this would imply that $h$ was ripe and so $h$ so can apply the same argument to the other non-fully processed component.
+ since this would imply that $h$ was ripe, and so one can apply the same argument to the other non-fully processed component.
 
@@
  In this case $w$ cannot have been processed (since after being processed it is touched only by one color), 
- and so it must be in $T(\rep(c))$ since it must be in some heap of a color in each upstar (and one upstar is entirely colored $\rep(c)$).
+ and so it must be in $T(\rep(c))$ since it must be in some heap of a color in each up-star (and one up-star is entirely colored $\rep(c)$).
  As $w$ lies above $h$ in $\MM$, this implies $h$ is not on the top of this heap.
@@
 \begin{proof} More generally, we prove the following: all outputs on processed vertices
-are edges of $\jc(\MM)$ and for any color $c$, $\h(c)$ is the lowest processed vertex
+are edges of $\jc(\MM)$ and for any current color $c$, $\h(c)$ is the lowest processed vertex
 of that color. We prove this by induction on the processing order. The base case is is trivially
@@
 Moreover, $\build$ merges the colors $c$ and $c'$ and correctly sets $v$ to be the attachment, 
-as $v$ is the lowest processed vertex of this merged color (as by induction $\h(c)$ and $\h(c')$ where the lowest vertices before merging colors).
+as $v$ is the lowest processed vertex of this merged color (as by induction $\h(c)$ and $\h(c')$ were the lowest vertices before merging colors).
 
@@
 
-\newcommand{\pathTree}{P_{\mathcal{S}}}
-\newcommand{\pathTreeA}{\mathcal{P}_{\mathcal{S}}}
-\section{Leaf assignments and path decompositions}
-
-\Ben{All merge tree references need to be switched to join.  Also, need to make sure that not assuming the bin trees are full was safe to drop.}
-
-In this section, we set up a framework to analyze the time taken to compute a critical join tree $\jc(\MM)$ (see \Def{criticalJoin}).
-%
-% \begin{definition}
-%  We say that a contour tree of a $2$-manifold is a \emph{merge tree} if it has a well defined root vertex such that height values 
-%  strictly increase along any root to leaf path.
-% \end{definition}
-% 
-% \begin{observation}
-%  The contour tree of a minimum dominant manifold (see \Def{dom}) is close to a merge tree.  Specifically, there is a root such that on any root 
-%  to leaf path height values strictly increase, with the only exception being potentially at the leaves themselves.  As such, up to constant 
-%  factors, any lower bound for computing the merge tree is a lower bound for computing the contour tree of a minimum dominant
-%  manifold (i.e. intuitively just clip off the leaves or shift them to lie above their parent).
-% \end{observation}
-%
-As such any tree $T$ considered in this section will be a rooted binary tree\footnote{Note 
+\subsection{Running Time}
+\label{sec:runTime}
+
+We now bound the running time of the above algorithm.  In the subsequent sections we will then provide a 
+matching lower bound for the running time.  Therefore, it will be useful to set up some terminology that can 
+be used consistently in both places.  Specifically, the lower bound proof will be a purely combinatorial 
+statement on colored rooted trees, and so the terminology is of this form.
+
+Any tree $T$ considered in following will be a rooted binary tree\footnote{Note 
 that technically the trees considered should have a leaf vertex hanging below the root of this 
@@
 (safely) ignored to simplify the presentation.} where the height of a vertex is its distance 
-from the root $r$.  As such, the children of a vertex $v\in T$ are the adjacent vertices of larger height
+from the root $r$ (i.e. conceptually $T$ will be a join tree with $r$ at the bottom).  
+As such, the children of a vertex $v\in T$ are the adjacent vertices of larger height
 (and $v$ is the parent of such vertices).  Then the subtree rooted at $v$, denoted $T_v$ consists of the graph induced on all 
@@
 We use $A(v)$ to denote the set of ancestors of $v$.
-For set of nodes $U$, $A(U) = \bigcup_{u \in U} A(u)$.
+For a set of nodes $U$, $A(U) = \bigcup_{u \in U} A(u)$.
 
 \begin{definition}
- A \emph{leaf assignment} $\chi$ of a tree $T$ assigns \emph{two} distinct leaves in each internal vertex $v$,
- one from the left child and one from the right child subtree of $v$.
+ A \emph{leaf assignment} $\chi$ of a tree $T$ assigns \emph{two} distinct leaves to each internal vertex $v$,
+ one from the left child and one from the right child subtree of $v$ (naturally if $v$ has only one child it is assigned 
+ only one color).
 \end{definition}
@@
 Note that the subroutine $\init(\MM)$ from \Sec{algo} naturally defines a leaf assignment to $\jc(\MM)$ 
-according to the priority queue for each upstar we put a given vertex in.
+according to the priority queue for each up-star we put a given vertex in.
 
 \begin{lemma}
-Let $\MM$ be a simplicial complex with $m$ edges.  For every vertex in $\jc(\MM)$, 
+\label{lem:runTimeUpper}
+Let $\MM$ be a simplicial complex with $t$ critical points.  For every vertex in $\jc(\MM)$, 
 let $H_v$ be defined by the leaf assignment determined by $\init(\MM)$.
-The running time of our algorithm is $O(m\alpha(m) + \sum_{v \in \jc(\MM)} \log |H_v|)$.
-% where $T$ is the merge tree of the given input, 
-%and $\chi$ is any valid coloring of $T$.  (maybe add one in the definition of $H_v$ so that $\log |H_v|$ is always $>1$).
+The running time of $\build(\MM)$ is $O(N+t\alpha(t) + \sum_{v \in \jc(\MM)} \log |H_v|)$.
 \end{lemma}
@@
 \begin{proof}
-First we look at the initialization procedure $\init(\MM)$.  This procedure runs in $O(m)$ time (note that as the input complex is connected, $N=O(m)$).
+First we look at the initialization procedure $\init(\MM)$.  This procedure runs in $O(N)$ time.
 Indeed, the painting procedure consists of several BFS's but as each vertex is only explored by one of the BFS's, it is linear time overall.
@@
 
-Now consider the union-find operations performed by $\init$ and $\update$.  
+Now consider the union-find operations performed by $\build$ and $\update$.  
 Initially the union find data structure has a singleton component for each leaf (and no new components are ever created), 
-and so each union-find operation takes $O(m)$ time.
-The $\update(K)$ procedure performs at most one find for each edge and no unions.  
-$\init(\MM)$ performs one union and at most two finds for each vertex.  Therefore the total number of union find 
-operations is $O(m)$.
-
-For the remaining operations, observe that every time $\update$ is called a vertex is pushed onto the stack and each
+and so each union-find operation takes $O(\alpha(t))$ time.
+For $\update$, by \Obs{twoQueues}, each iteration of the while loop requires a constant number of finds (and no unions).
+Specifically, if a vertex is found to be ripe (and hence processed next) then these can be charged to that vertex.
+If a vertex is not ripe, then these can be charged to the vertex put on the stack.
+As each vertex is put on the stack or processed at most once, $\update$ performs $O(t)$ finds overall.
+Finally, $\build(\MM)$ performs one union and at most two finds for each vertex.  Therefore the total number of union find 
+operations is $O(t)$.
+
+For the remaining operations, observe that for every iteration of the loop in $\update$, a vertex is pushed onto the stack and each
 vertex can only be pushed onto the stack once (since the only way it leaves the stack is by being processed). 
@@
 
-What remains is the time it takes process a vertex $v$ in $\init(\MM)$.  
+What remains is the time it takes to process a vertex $v$ in $\build(\MM)$.  
 In order to process a vertex there are a few constant time operations, union-find operations, and queue operations.
-Therefore the only thing we must bound are the queue operations.
-Let $v$ be a vertex in $\jc(\MM)$, and let $p_1$ and $p_2$ be its parents (the same argument holds if $v$ has only one parent).
-At the time $v$ is handled, the colors and queues of all vertices in a given component of $\MM^+_v$ have merged together.
-In particular, when $v$ is handled we know it is ripe and so all vertices above $v$ in each component of $\MM^+_v$ have been processed, implying 
-these merged queues are the queues of the current colors of $p_1$ and $p_2$.  Again since $v$ is ripe, it must be on the top of these queues and so 
-the only vertices of left in these queues are those in $H_{p_1}$ and $H_{p_2}$.
-
-Now when $v$ is handled, three queue operations are performed.  Specifically, $v$ is removed from the queues of $p_1$ and $p_2$, and then the queues are 
-are merged together.  By the above arguments the sizes of the queues for each of these operations are $H_{p_1}$, $H_{p_2}$, and $H_v$, respectively.  
+Therefore the only thing left to bound are the queue operations.
+Let $v$ be a vertex in $\jc(\MM)$, and let $c_1$ and $c_2$ be its children (the same argument holds if $v$ has only one child).
+At the time $v$ is processed, the colors and queues of all vertices in a given component of $\MM^+_v$ have merged together.
+In particular, when $v$ is processed we know it is ripe and so all vertices above $v$ in each component of $\MM^+_v$ have been processed, implying 
+these merged queues are the queues of the current colors of $c_1$ and $c_2$.  Again since $v$ is ripe, it must be on the top of these queues and so 
+the only vertices left in these queues are those in $H_{c_1}$ and $H_{c_2}$.
+
+Now when $v$ is handled, three queue operations are performed.  Specifically, $v$ is removed from the queues of $c_1$ and $c_2$, and then the queues are 
+are merged together.  By the above arguments the sizes of the queues for each of these operations are $H_{c_1}$, $H_{c_2}$, and $H_v$, respectively.  
 As merging and deleting takes logarithmic time in the heap size for binomial heaps, the claim now follows.
@@
 
-%vertices along the path from the root to $v$ such that one of the two colors of each vertex is in $L(T_v)$.
-%We will refer to $H_v$ as the \emph{heap} at $v$.
-%Then define the \emph{algorithm} cost function as $f_{alg}(\chi(T)) = \sum_{v\in T} \log |H_v|$.
-%
-%\begin{lemma}
-%\label{lem:algCost}
-%NEEDS TO BE WRITTEN:
-%Here we need a lemma showing that the running time of our algorithm is $O(f_{alg}(\chi(T)))$, where $T$ is the merge tree of the given input, 
-%and $\chi$ is any valid coloring of $T$.  (maybe add one in the definition of $H_v$ so that $\log |H_v|$ is always $>1$).
-%\end{lemma}
-%
-%Now that we have a simple expression upper bounding the running time of our algorithm, we spend the remainder 
-%of this section proving that this running time is optimal.
+\newcommand{\pathTree}{P_{\mathcal{S}}}
+\newcommand{\pathTreeA}{\mathcal{P}_{\mathcal{S}}}
+\section{Leaf assignments and path decompositions}
+
+In this section, we set up a framework to analyze the time taken to compute a critical join tree $\jc(\MM)$ (see \Def{criticalJoin}).
+We adopt all notation already defined in \Sec{runTime}.  From here forward we will often assume binary trees are full binary trees 
+(this assumption simplifies the presentation but is not necessary).
+
+\Ben{For now we are keeping the above full bin tree assumption. At some point this should be carefully removed.}
 
@@
 
-Our main theorem is the following. We use $|p|$ to denote the number of vertices in $p$.
+Our main theorem is the following (which by \Lem{runTimeUpper} upper bounds the running time of our algorithm). We use $|p|$ to denote the number of vertices in $p$.
 
@@
 In this section we provide a lower bound for the time it takes to compute a contour tree in terms the cost of path decompositions.  Specially, 
-we first consider the easier case of merge trees, after which the lower bound for more general contour trees will easily follow.
-
-\subsection{Merge Trees}
-In this section we show that any algorithm for computing the merge tree $T$ must pay at least $f_{path}(P(T))$ (see \Def{path}) for any path decomposition $P(T)$.
-
-%We say two shrubs $S_1$ and $S_2$ (of merge trees of potentially different surfaces) are equivalent if there is an isomorphism, $f$,
+we first consider the easier case of join trees, after which the lower bound for more general contour trees will easily follow.
+
+\subsection{Join Trees}
+In this section we show that any algorithm for computing the join tree $T$ must pay at least $f_{path}(P(T))$ (see \Def{path}) for any path decomposition $P(T)$.
+
+%We say two shrubs $S_1$ and $S_2$ (of join trees of potentially different surfaces) are equivalent if there is an isomorphism, $f$,
 %between the two shrubs such that for any node $v\in S_1$ corresponding to a path of size $|v|$, $f(v)$ also corresponds to a path of size $|v|$.
@@
 \label{lem:pathDecomp}
-Let $T$ be a merge tree, and let $P = P(T)$ any path decomposition of $T$.  %, and $\pathTree$ the corresponding shrub.  
+Let $T$ be a join tree, and let $P = P(T)$ any path decomposition of $T$.  %, and $\pathTree$ the corresponding shrub.  
 Then there is a set of at least $\Pi_{p_i\in P} (|p_i|-1)!$ distinct PL $2$-manifolds, all of which have the same defining mesh.
@@
  We describe a simple mesh, with heights specified for all vertices except for the saddles.  The main idea is that 
- this mesh will be flexible enough to allow almost any assignment of heights to the saddles, thus defining a class of merge trees 
+ this mesh will be flexible enough to allow almost any assignment of heights to the saddles, thus defining a class of join trees 
  whose size is that described in the lemma statement.
@@
  diamonds, and there is exactly one minima corresponding to the bottom cap of the diamond representing the root of $\pathTree$.  Therefore, the saddles on a given 
- diamond will appear contiguously on a root to leaf path in the merge tree of the mesh, where the leaf corresponds to the maxima of the diamond 
+ diamond will appear contiguously on a root to leaf path in the join tree of the mesh, where the leaf corresponds to the maxima of the diamond 
  (since all these saddles have a direct line of sight to this apex).  In particular, 
- this implies that, regardless of the heights assigned to the platforms, the merge tree of this mesh has a path decomposition whose corresponding shrub is equivalent to $\pathTree$.
+ this implies that, regardless of the heights assigned to the platforms, the join tree of this mesh has a path decomposition whose corresponding shrub is equivalent to $\pathTree$.
 
@@
  there are $(|p|-1)!$ possible orderings of the heights of the saddles on the diamond for $p$, and hence $\Pi_{p_i\in P} (|p_i|-1)!$ possible instantiations of 
- the mesh overall.  Also, each one of these instantiations will result in a different merge tree.  Specifically, a permutation of the heights on a given diamond 
- corresponds to a permutation of the vertices along a path in $P$, as by construction all saddles on a given diamond will appear in sorted order in the merge tree.
+ the mesh overall.  Also, each one of these instantiations will result in a different join tree.  Specifically, a permutation of the heights on a given diamond 
+ corresponds to a permutation of the vertices along a path in $P$, as by construction all saddles on a given diamond will appear in sorted order in the join tree.
 \end{proof}
@@
 \begin{corollary}
- \label{cor:mergeLB}
- Let $A(S)$ be any comparison based algorithm to compute the merge tree of any input surface $S$ (whose contour tree is a merge tree).  
+ \label{cor:joinLB}
+ Let $A(S)$ be any comparison based algorithm to compute the join tree of any input surface $S$ (whose contour tree is a join tree).  
  Then the running time of $A(S)$ is $\Omega(f_{path}(P))=\Omega(\sum_{p_i\in P} |p_i|\log|p_i|)$ where $P$ is any path decomposition of the 
- merge tree of $S$.
+ join tree of $S$.
 \end{corollary}
 \begin{proof}
- Given an input surface $S$, by the above lemma, any path decomposition $P$ of its merge tree defines a set of $\Pi_{p_i\in P} |p_i|!$ surfaces 
- with distinct merge trees, all of which have the same mesh.  Specifically, for each sorted ordering of the relative height values of the saddles 
- on each diamond in the mesh, yields a different merge tree.
- Therefore, any comparison based algorithm which computes the merge tree must distinguish between these possibilities, and therefore its corresponding decision tree must 
+ Given an input surface $S$, by the above lemma, any path decomposition $P$ of its join tree defines a set of $\Pi_{p_i\in P} |p_i|!$ surfaces 
+ with distinct join trees, all of which have the same mesh.  Specifically, for each sorted ordering of the relative height values of the saddles 
+ on each diamond in the mesh, yields a different join tree.
+ Therefore, any comparison based algorithm which computes the join tree must distinguish between these possibilities, and therefore its corresponding decision tree must 
  have $\Pi_{p_i\in P} |p_i|!$ leaves.  Hence by Stirling's approximation, this decision tree must have depth $\Omega(\sum_{p_i\in P} |p_i|\log|p_i|)$.
@@
 In this section $T$ will denote an arbitrary contour tree.  Specifically, every internal vertex in $T$ has degree $3$ 
-and the height function defined over the vertex set is such that every internal vertex in either a split or merge 
+and the height function defined over the vertex set is such that every internal vertex in either a split or join 
 (i.e. at most $2$ neighbors lie above or below it).
@@
  vertex in $T$ which $r_i$ is adjacent to. 
- If $T_i$ was created by an upward BFS, then the connection to this face should be a merge vertex, and otherwise it should be a split vertex.  
+ If $T_i$ was created by an upward BFS, then the connection to this face should be a join vertex, and otherwise it should be a split vertex.  
  See \Fig{diamond}.
 
- Just as in \Lem{pathDecomp} we now have one fixed mesh graph structure, such that each different relative ordering of the heights of the merge and split vertices 
- on each diamond produces a surface with a distinct merge tree.  The specific bound on the number of such distinct manifolds follows by applying the 
+ Just as in \Lem{pathDecomp} we now have one fixed mesh graph structure, such that each different relative ordering of the heights of the join and split vertices 
+ on each diamond produces a surface with a distinct join tree.  The specific bound on the number of such distinct manifolds follows by applying the 
  bound from \Lem{pathDecomp} to each $P_i$.
@@
 
-For the same reason that \Cor{mergeLB} followed from \Lem{pathDecomp}, we now have the following.
+For the same reason that \Cor{joinLB} followed from \Lem{pathDecomp}, we now have the following.
 
@@
 \end{theorem}
+
+\begin{remark}
+ Note that for the surfaces described in this section, the number of critical points is within a constant factor of 
+ the total number of vertices.  In particular, for this family of surfaces, all previous algorithms required $\Omega(n\log n)$ time.
+\end{remark}
 
@@
 % $l$ is adjacent to a max path and $u$ is adjacent to a min path:\newline
-% Connect a face of the cylinder for the $u$ path (using a split) to a face of the cylinder for the $l$ path (using a merge).
+% Connect a face of the cylinder for the $u$ path (using a split) to a face of the cylinder for the $l$ path (using a join).
 % \end{enumerate}

### 2014-10-29 paint-contour-tree.tex
sanity commit

--- 2014-10-29 paint-contour-tree.tex
@@
 
-A large class of data, such as satellite terrain maps, physical simulations, combustion data, can
+A large class of data, such as satellite terrain maps, physical simulation output, and combustion data, can
 be thought as a function $f: \RR^d \mapsto \RR$. Typically, a finite representation is given
-by consider $f$ to be piecewise linear over some mesh (simplicial complex) $\MM$ in $\RR^d$.
+by considering $f$ to be piecewise linear over some mesh (simplicial complex) $\MM$ in $\RR^d$.
 \emph{Contour trees} are a specific topological data structure used to represent and visualize
@@
 For convenience, imagine $f$ to be smooth. All points where the gradient of $f$ is zero are \emph{critical point},
-exactly where the topology of the contours change. The vertices of the contour tree are exactly the critical points.
+exactly where the topology of the contours change. The vertices of the contour tree are exactly these critical points,
+representing the important topological events. An edge connects two critical points if
 

### 2014-10-30 paint-contour-tree.tex
figs updated
updates to intro

@@
 exactly where the topology of the contours change. The vertices of the contour tree are exactly these critical points,
-representing the important topological events. An edge connects two critical points if
-
-Contour trees and related topological data structures have been applied in analysis of fluid mixing, combustion simulations,
-and studying chemical systems~\cite{LaBe+06,BrWe+10,BeWe+11,BrWe+11,MaGr+11}.
+representing the important topological events. An edge connects two critical points if one event immediately ``follows"
+the other as the sweep plane makes its pass. (We provide formal definitions later.)
+
+\Fig{relative} and \Fig{sorting} show examples of simplicial complices with heights and their contour trees.
+Think of the edges as point downwards, so we can talk of the ``up-degree" and ``down-degree" of vertices.
+Leaves of down-degree $1$ are maxima, because a new contour is created at such point. Leaves of up-degree
+$1$ are minimum, because a contour is destroyed there. Up-degree $2$ vertices are ``joins", where contours
+merge, and down-degree $2$ vertices are ``splits". In these examples ($d=2$), contours are closed loops.
+In dimensions $d \geq 3$, we may have internal
+vertices of both up and down-degree $1$. The correspond to contours (which are themselves manifolds) gaining or losing handles.
+
+Suppose we have $f: \MM \mapsto \RR$, where $\MM$ has $n$ vertices, $N$ facets in total, and $t \leq n$ critical points.
+It is standard to assume that $f$ is PL-Morse (a well-behaved assumption) and that the maximum degree in the contour tree
+is $3$. A fundamental result in this area is the algorithm of Carr, Snoeyink, and Axen to compute contour trees,
+which runs in $O(n\log n + N\alpha(N)$ time~\cite{csa-cctad-00}. (Henceforth, $\alpha(\cdot)$ denotes the inverse
+Ackermann function.) One cannot overstate the importance of this result, since 
+subsequent practical and theoretical work in contour trees uses their ideas. It is natural to think of $N$ as $\Theta(n)$
+in practical applications (certainly true for $d=2$). The most expensive operation is an initial sort of all the vertex
+heights. Chiang et al build on this approach to get a faster algorithm that only sorts the critical vertices,
+yielding a running time of $O(t\log t + N)$~\cite{cllr-sooscctmp-05}.
+Common applications for contour trees involve turbulent combustion or noisy data, where the number of critical points is likely $\Omega(n)$.
+So we are still left with the bottleneck of sorting a large set of numbers. Unfortunately, there is a worst-case
+lower bound of $\Omega(t\log t)$ from Chiang et al~\cite{cllr-sooscctmp-05}, based on a construction of Bajaj et al~\cite{BaKr+98}.
+
+But this is only a worst-case bound. All previous algorithm begin by sorting (at least) the critical points, and
+pay $\Omega(t\log t)$ for all instances. Can we beat this sorting bound? Is there a more refined version of the complexity
+of computing contour trees? Our main result gives an affirmative answer. First, we require some technical definitions.
+
+In a contour tree $T$, a \emph{leaf path} is simply a contiguous path in $T$
+containing a leaf, which is also monotonic in the height values of its vertices. A leaf path runs either upwards
+or downwards, never both.
+% Then in the following two subsections 
+%we relate path decomposition costs (defined below) to the running time of our algorithm using $\Lem{algCost}$.
+
+\begin{definition}
+\label{def:path} For a contour tree $T$, a \emph{path decomposition}, $P(T)$, is a partition of $T$ into a set of vertex disjoint leaf paths.  
+\end{definition}
+
+\begin{theorem} \label{thm:alg-main} Consider $f:\MM \mapsto \RR$ that is PL-Morse, where the contour tree $T$
+has maximum degree\footnote{The maximum degree $3$ assumption is primarily there to simplify the presentation and proof. Our algorithm
+does not need the assumption.} $3$. There is an algorithm whose running time is $O(\sum_{p \in P(T)} |p|\log |p| + t\alpha(t) + N)$,
+where $P(T)$ is a specific path decomposition (constructed implictly by the algorithm).
+\end{theorem}
+
+Let us interpret the running time. 
+For any path decomposition $P(T)$ and any $p \in P(T)$, obviously $|p| \leq |T| = t$.
+So the running time is clearly $O(t\log t + N)$. But this is only the case when $T$ is tall and thin.  
+An obvious upper bound is $O(t\log D + t\alpha(t) + N)$, where $D$ is the diameter of $P(T)$. 
+But if $P(T)$ is a rooted, balanced, binary tree, then $\sum_{p \in P(T)} |p|\log |p| = O(t)$.
+Our running time is now $O(t\alpha(t))$, a significant improvement from $O(t\log t)$.
+Consider the example of \Fig{sorting} extended to a larger binary tree both above and below.
+(In the example, $t = \Omega(n)$.) The running time promised by \Thm{alg-main}
+is $O(n\alpha(n))$, whereas all previous algorithms require $\Omega(n\log n)$.
 
@@
 \subsection{Previous Work}
+
+Contour trees were (unsurprisingly) used to study terrain maps by Boyell and Ruston, and Freeman and Morse~\cite{BoRu63,FrMo67}.
+They were used for isoline extraction in geometric data by van Krevald et al~\cite{kobps-ctsssit-97}, who provided
+the first formal algorithm. 
+Contour trees and related topological data structures have been applied in analysis of fluid mixing, combustion simulations,
+and studying chemical systems~\cite{LaBe+06,BrWe+10,BeWe+11,BrWe+11,MaGr+11}. Carr's thesis gives various
+applications of contour trees for data visualization~\cite{c-tmi-04}.
 
@@
 
-We introduce the concept of path decompositions. A \emph{leaf path} is simply a contiguous path in $T$
-containing a leaf, which is monotonic in the height values of its vertices.
-% Then in the following two subsections 
-%we relate path decomposition costs (defined below) to the running time of our algorithm using $\Lem{algCost}$.
-
-\begin{definition}
-\label{def:path} A \emph{path decomposition}, $P(T)$, is a partition of $T$ into 
- a set of vertex disjoint leaf paths.  
-% 
-% The \emph{cost} of $P(T)$ is $\cost(P(T)) = \sum_{p\in P(T)} |p| \log |p|$. (We use $|p|$ to denote the number of vertices in $p$.)
-\end{definition}
-
 Note that the number of paths in $P(T)$ must equal the number of leaves in $T$.

### 2014-10-30 paint-contour-tree.tex
small section 2 changes

@@
 \Sesh{Some mention of representation and data structures here.}
-We use the standard assumption that $f$ is Morse (we explain later what this entails).
+We use the standard assumption that $f$ is Morse.  In particular, we are assuming no two critical points have the same height, 
+and that there are no multi-saddles.
 
@@
 Note that a contour is itself a simplicial complex of one dimension lower, and is represented (in our algorithms) as such.
-We let $\delta, \eps$ denote infinitesimals. Let $B_\eps(x)$ denote a ball of radius $\eps$ around $x$.
-
-\begin{definition} \label{def:deg} The up-degree of $x \in \MM$ is the number of $(f(x) + \delta)$-contours that intersect $B_\eps(x)$
-as $\delta, \eps \rightarrow 0^+$. The down-degree is the number of $(f(x) - \delta)$-contours that intersect $B_\eps(x)$
+We let $\delta, \eps$ denote infinitesimals. Let $B_\eps(x)$ denote a ball of radius $\eps$ around $x$, and let 
+$f|B_\eps(x)$ be the restriction of $f$ to $B_\eps(x)$.
+
+% \begin{definition} \label{def:deg} The up-degree of $x \in \MM$ is the number of $(f(x) + \delta)$-contours that intersect $B_\eps(x)$
+% as $\delta, \eps \rightarrow 0^+$. The down-degree is the number of $(f(x) - \delta)$-contours that intersect $B_\eps(x)$
+% as $\delta, \eps \rightarrow 0^+$.
+% \end{definition}
+
+\begin{definition} \label{def:deg} The up-degree of $x \in \MM$ is the number of $(f(x) + \delta)$-contours in $f|B_\eps(x)$
+as $\delta, \eps \rightarrow 0^+$. The down-degree is the number of $(f(x) - \delta)$-contours in $f|B_\eps(x)$
 as $\delta, \eps \rightarrow 0^+$.
@@
 
-The definition allows us to categorize points in $\MM$ depending on the local changes in topology.
+This definition allows us to categorize points in $\MM$ depending on the local changes in topology.
 
@@
 \begin{remark}
- For two manifolds, at each saddle point either two contours join together or split apart, and hence the terminology.
-In higher dimensions this may not be the case, however, it always holds that at a join or split vertex the topology of the level set changes, and hence such vertices are still 'critical'.
-Note that determining which category each vertex falls into according to the above definition can be determined for all vertices simultaneously in linear time, as it is a local neighborhood condition.
+ Note that \Def{deg} can be changed so that the up-degree (resp. down-degree) is the number of $(f(x) + \delta)$-contours (resp. $(f(x) - \delta)$-contours) that intersect $B_\eps(x)$.
+ For $2$-manifolds, the two definitions are equivalent, and at each saddle point either two contours join together or split apart, and hence the terminology.  
+ In higher dimensions though these definitions differ.  Specifically, this alternative definition only includes points at which the number of contours in a level set change.
+ These points are a subset of our defined critical points which also include points at which the topology of the contours change (and hence the term `critical' is still applicable).
+ Note that our critical point definition allows one to classify all vertices (according to \Def{points}) simultaneously in linear time, as it is a local neighborhood condition.
 \end{remark}
@@
     \hspace{.35in}
-    \includegraphics[width=.27\linewidth]{figs/mound2}%
+    \includegraphics[width=.27\linewidth]{figs/mounds2}%
     \hspace{.4in}
-    \includegraphics[width=.22\linewidth]{figs/mound3}
+    \includegraphics[width=.22\linewidth]{figs/mounds3}
     \caption{On the left, three initial colors.  In the middle, yellow and blue merge into green.  On the right, green and red merge into purple.}

### 2014-10-30 paint-contour-tree.tex
changes to intro

@@
 
+\newcommand{\bF}{{\bf F}}
 \newcommand{\bT}{{\bf T}}
@@
 A classic result of Carr, Snoeyink, and Axen (2000) give an algorithm that takes $O(n\log n + N\alpha(N))$ time
-(where $\alpha(\cdot)$ is the inverse Ackermann function). A further improvement to $O(t\log t + N\alpha(N))$ time
+(where $\alpha(\cdot)$ is the inverse Ackermann function). A further improvement to $O(t\log t + N)$ time
 was given by Chiang et al. All these algorithms involve a global sort of the critical points,
@@
 in running time. For a partition of the contour tree into a set of descending paths, $P$,
-our algorithm runs in $O(\sum_{p\in P} |p|\log|p| + N\alpha(N))$. If the contour tree is a balanced
-binary tree, our running time is $O(n + N\alpha(N))$. Moreover, we prove a lower bound of
-exactly $\Omega(\sum_{p \in P} |p|\log |p|)$ for computing contour trees, thereby showing
+our algorithm runs in $O(\sum_{p\in P} |p|\log|p| + t\alpha(t))$. Moreover, we prove a general lower bound of
+$\Omega(\sum_{p \in P} |p|\log |p|)$ for computing contour trees, showing
 that the tree structure must play a role in the time complexity.
@@
 %matches the previous worst case $O(N+n\log n)$ bound when the tree is long and skinny.
-}
 
@@
 such tools as heavy-light path decompositions.
+}
 \end{abstract}
@@
 
-\begin{theorem} \label{thm:alg-main} Consider $f:\MM \mapsto \RR$ that is PL-Morse, where the contour tree $T$
+\begin{theorem} \label{thm:main-alg} Consider $f:\MM \mapsto \RR$ that is PL-Morse, where the contour tree $T$
 has maximum degree\footnote{The maximum degree $3$ assumption is primarily there to simplify the presentation and proof. Our algorithm
@@
 Consider the example of \Fig{sorting} extended to a larger binary tree both above and below.
-(In the example, $t = \Omega(n)$.) The running time promised by \Thm{alg-main}
+(In the example, $t = \Omega(n)$.) The running time promised by \Thm{main-alg}
 is $O(n\alpha(n))$, whereas all previous algorithms require $\Omega(n\log n)$.
+
+We complement \Thm{main-alg} with an almost matching lower bound. We show that the complexity
+of $\sum_{p \in P(T)} |p|\log |p|$ is required for any contour tree algorithm. The exact phrasing
+of the lower bound requires some care. Consider the $d=2$ (terrain) case. For any $f$,
+we use $\cost(f)$ to denote $\sum_{p \in P(T)} |p|\log |p|$ for the specific $P(T)$
+given in \Thm{main-alg}. Ideally, we would like to argue that any contour-tree algorithm
+must pay $\cost(f)$ for $f$. But this is obviously false, since one could consider an algorithm
+that hard codes the answer for $f$ and merely checks this answer.
+
+Instead, call a possible cost $C$ \emph{valid} if $C = \cost(f)$ for some $f$\footnote{If one is concerned
+that $C$ might be irrational, define valid if $2^C = 2^{\cost(f)}$.}. We prove
+a refined lower bound matching \Thm{main-alg}. A \emph{height decision tree} is an algorithm
+that is allowed aribtrary computation on the domain $\MM$, and is an algebraic decision
+tree over the heights. The running time is simply the number of queries made.
+
+\begin{theorem} \label{thm:main-lb} For any valid cost $C$, there exists a set $\bF$ of functions over 2-dimensions
+with the following properties.
+For all $f \in \bF$, $\cost(f) = C$. Any height decision tree that correctly computes
+the contour tree on all of $\bF$ has an average case (uniform over $\bF$) running time of $\Omega(C)$.
+\end{theorem}
+
+In other words, for any valid cost $C$, we can find a family of inputs where \Thm{main-alg}
+guarantees a running time of $C + t\alpha(t)$, and any height decision tree requires $\Omega(C)$ time.
+This shows that the cost function is inherent to the computation of contour trees.
 

### 2014-10-30 paint-contour-tree.tex
fig 2 changed

@@
 
+% \begin{figure}[h]\centering
+%     \includegraphics[width=.3\linewidth,height=.42\linewidth]{figs/shaded}%
+%     \hspace{.5in}
+%     \includegraphics[width=.3\linewidth,height=.42\linewidth]{figs/trees2}
+%     \caption{On the left, a surface will a balanced contour tree, but whose join and split trees have long tails.  
+%     On the right (in clockwise order from the top) the contour, split and join trees.}
+%     \label{fig:sorting}
+% \end{figure}
 \begin{figure}[h]\centering
-    \includegraphics[width=.3\linewidth,height=.42\linewidth]{figs/shaded}%
-    \hspace{.5in}
-    \includegraphics[width=.3\linewidth,height=.42\linewidth]{figs/trees2}
-    \caption{On the left, a surface will a balanced contour tree, but whose join and split trees have long tails.  
-    On the right (in clockwise order from the top) the contour, split and join trees.}
+    \includegraphics[width=.7\linewidth,height=.28\linewidth]{figs/balanced2}
+    \includegraphics[width=.7\linewidth,height=.25\linewidth]{figs/trees}
+    \caption{On top, a surface will a balanced contour tree, but whose join and split trees have long tails.  
+    On bottom (from left to right), the contour, join and split trees.}
     \label{fig:sorting}

### 2014-10-30 paint-contour-tree.tex
changed figures
updates to intro

@@
 
-\title{A Mountaintop View Requires Minimal Sorting: \break An Instance Optimal Contour Tree Algorithm}
+\title{A Mountaintop View Requires Minimal Sorting: \break A Faster Contour Tree Algorithm}
 \date{}
@@
 
-A large class of data, such as satellite terrain maps, physical simulation output, and combustion data, can
-be thought as a function $f: \RR^d \mapsto \RR$. Typically, a finite representation is given
+Massive geometric data is often represented as a function $f: \RR^d \mapsto \RR$. Typically, a finite representation is given
 by considering $f$ to be piecewise linear over some mesh (simplicial complex) $\MM$ in $\RR^d$.
@@
 
-For convenience, imagine $f$ to be smooth. All points where the gradient of $f$ is zero are \emph{critical point},
-exactly where the topology of the contours change. The vertices of the contour tree are exactly these critical points,
+For convenience, imagine $f$ to be smooth. All points where the gradient of $f$ is zero are \emph{critical points},
+where the topology of the contours change. The vertices of the contour tree are exactly these critical points,
 representing the important topological events. An edge connects two critical points if one event immediately ``follows"
@@
 \Fig{relative} and \Fig{sorting} show examples of simplicial complices with heights and their contour trees.
-Think of the edges as point downwards, so we can talk of the ``up-degree" and ``down-degree" of vertices.
+Think of the edges pointing downwards, so we can talk of the ``up-degree" and ``down-degree" of vertices.
 Leaves of down-degree $1$ are maxima, because a new contour is created at such point. Leaves of up-degree
@@
 In dimensions $d \geq 3$, we may have internal
-vertices of both up and down-degree $1$. The correspond to contours (which are themselves manifolds) gaining or losing handles.
-
-Suppose we have $f: \MM \mapsto \RR$, where $\MM$ has $n$ vertices, $N$ facets in total, and $t \leq n$ critical points.
+vertices of both up and down-degree $1$. They correspond to contours (which are $2$-manifolds) gaining or losing handles.
+
+Suppose we have $f: \MM \mapsto \RR$, where $\MM$ is a triangulated mesh with $n$ vertices, $N$ facets in total, and $t \leq n$ critical points.
 It is standard to assume that $f$ is PL-Morse (a well-behaved assumption) and that the maximum degree in the contour tree
@@
 which runs in $O(n\log n + N\alpha(N)$ time~\cite{csa-cctad-00}. (Henceforth, $\alpha(\cdot)$ denotes the inverse
-Ackermann function.) One cannot overstate the importance of this result, since 
-subsequent practical and theoretical work in contour trees uses their ideas. It is natural to think of $N$ as $\Theta(n)$
+Ackermann function.) It is natural to think of $N$ as $\Theta(n)$
 in practical applications (certainly true for $d=2$). The most expensive operation is an initial sort of all the vertex
@@
 But this is only a worst-case bound. All previous algorithm begin by sorting (at least) the critical points, and
-pay $\Omega(t\log t)$ for all instances. Can we beat this sorting bound? Is there a more refined version of the complexity
+pay must $\Omega(t\log t)$ for all instances. Can we beat this sorting bound? Is there a more refined version of the complexity
 of computing contour trees? Our main result gives an affirmative answer. First, we require some technical definitions.
@@
 
-\begin{theorem} \label{thm:main-alg} Consider $f:\MM \mapsto \RR$ that is PL-Morse, where the contour tree $T$
-has maximum degree\footnote{The maximum degree $3$ assumption is primarily there to simplify the presentation and proof. Our algorithm
-does not need the assumption.} $3$. There is an algorithm whose running time is $O(\sum_{p \in P(T)} |p|\log |p| + t\alpha(t) + N)$,
+\begin{theorem} \label{thm:main-alg} Consider PL-Morse\footnote{This is a standard ``well-behaved" assumption that implies
+that all critical points have distinct values, and all the contour tree has max-degree $3$. Our algorithm
+does not need this, but it simplifies the proof and presentation.} $f:\MM \mapsto \RR$, where the contour tree $T$
+has maximum degree $3$. There is an algorithm whose running time is $O(\sum_{p \in P(T)} |p|\log |p| + t\alpha(t) + N)$,
 where $P(T)$ is a specific path decomposition (constructed implictly by the algorithm).
@@
 An obvious upper bound is $O(t\log D + t\alpha(t) + N)$, where $D$ is the diameter of $P(T)$. 
+This is clearly an improvement when the tree is short and fat.
 But if $P(T)$ is a rooted, balanced, binary tree, then $\sum_{p \in P(T)} |p|\log |p| = O(t)$.
@@
 
-\begin{theorem} \label{thm:main-lb} For any valid cost $C$, there exists a set $\bF$ of functions over 2-dimensions
-with the following properties.
-For all $f \in \bF$, $\cost(f) = C$. Any height decision tree that correctly computes
+\begin{theorem} \label{thm:main-lb} Consider the set of functions $\bF_C$ that all have the same (valid) cost $C$.
+Any height decision tree that correctly computes
 the contour tree on all of $\bF$ has an average case (uniform over $\bF$) running time of $\Omega(C)$.
@@
 
-In other words, for any valid cost $C$, we can find a family of inputs where \Thm{main-alg}
-guarantees a running time of $C + t\alpha(t)$, and any height decision tree requires $\Omega(C)$ time.
-This shows that the cost function is inherent to the computation of contour trees.
-
-Let $f$ be a real valued function defined over a manifold embedded in $\RR^d$.  
-For any real value $\alpha$, the level set $f^{-1}(\alpha)$ can be decomposed into its connected components, or ``contours''.
-The Reeb graph is a fundamental topological structure which represents the evolution of these contours as one sweeps the surface 
-from $-\infty$ to $\infty$ according to $f$.  Often this graph is acyclic, in which case it is 
-called a \emph{contour tree}.  This is the case if, for example, the input is a surface mesh of a mountain range 
-(and the contour tree would encode when mountain peeks form and how they merge together).
-
-There is a long list of practical applications of the study of contour trees (see for example \cite{c-tmi-04} and references therein).
-Often these applications involve massive data sets, and so consequently there has also been a long series 
-of results on efficient algorithms for their computation \cite{kobps-ctsssit-97,tv-cct-98,csa-cctad-00,pc-ectls-02,cllr-sooscctmp-05}.
-Let $N$ be the total size of the input manifold, $n$ the number of vertices, and $t$ the number of critical points. (Note that in many applications $N=O(n)$.)
-As an initial step these algorithms require that all $t$ contour tree vertices be sorted.  
-In particular, \cite{cllr-sooscctmp-05} provide an $O(N+t\log t)$ time algorithm, and since in the worst case computing the contour tree 
-requires $\Omega(N+t\log t)$ time, this algorithm is optimal with respect to the size of the output.  
-Here we present the first contour tree algorithm which is optimal (for instances requiring $\Omega(N\alpha(N))$ 
-time\footnote{Where $\alpha(\cdot)$ is the inverse Ackermann function, which is $< 5$ for computation in this universe.}) 
-with respect not just to the size of the output, but also its structure.  This is accomplished by introducing a simple linear time 
-surface partitioning scheme to avoid sorting, which is likely to have applications elsewhere.  
-For all instances the bound for our algorithm either matches the best previous bound, or is faster.
-For, example when the contour tree is a balanced, our algorithm runs in $O(N\alpha(N))$ time, whereas all 
-previous algorithms had a running time of $\Omega(N+t\log t)$
-
-%In particular, the algorithm in \cite{cllr-sooscctmp-05} is worst case optimal with respect to the size of the output.  
-%Here we present the first contour tree algorithm which is (near\footnote{Our algorithm is optimal on all instance requiring $\Omega(n\alpha(n))$ time.}) 
-%optimal with respect not just to the size of the output, but also its structure.  
-%Specifically, the bound for our algorithm either matches \cite{cllr-sooscctmp-05}, or is faster.
-%For, example when the contour tree is a balanced, our algorithm runs in $O(n\alpha(n))$ time, whereas all 
-%previous algorithms had a running time of $\Omega(n\log n)$.
-
-This work can be seen as a continuation of the study of instance-optimal geometric algorithms introduced by Afshani \etal \cite{abc-ioga-09}.
-In order to prove instance optimality we use a non-trivial combinatorial discharging argument, which we believe is of independent interest.
-Of particular note is the use of heavy-light path decompositions, first introduced by Sleator and Tarjan \cite{st-dsdt-83} in their analysis of link-cut trees.
-
-% Explain why useful:  Lets you know the ``shape of your date''.  Map of the critical points where contours change.  
-%Fast extraction of seed sets in a small sized structure, which can let you do threshold queries.
-% 
-% \medskip\noindent
-% Paragraph on specific applications.  Scientific visualization, in particular combustion research.  GIS.
+In other words, it is not possible to improve on \Thm{main-alg} even when restricted to the subset
+of inputs of fixed cost.
 
@@
 Contour trees and related topological data structures have been applied in analysis of fluid mixing, combustion simulations,
-and studying chemical systems~\cite{LaBe+06,BrWe+10,BeWe+11,BrWe+11,MaGr+11}. Carr's thesis gives various
-applications of contour trees for data visualization~\cite{c-tmi-04}.
-
-\subsubsection{Worst Case Alternatives}
-In worst case analysis an algorithm's running time is judged in terms of its worst 
-case behavior over all inputs of a given size.  While there are benefits to this viewpoint, 
-such as robustness,  it fails to capture the whole story.  As such, many alternatives 
-have been considered.  For example, one can make assumptions on the distribution of 
-inputs, which has led to a number interesting results on average case and probabilistic 
-analysis, and also smoothed analysis \cite{st-saa-04}.  However, even without making 
-assumptions on the distribution of the input, there is still more to the story.
-For example, in the study of parametrized complexity, one takes into account other 
-parameters determined by the input other than just its size (for example the maximum 
-degree in a graph or whether it is planar).
-
-\paragraph{Output Sensitive Analysis.}
-In geometric settings often the goal is to compute a (potentially complex) geometric structure determined 
-by the input, and so it is natural to also consider how the form of this output structure affects the running time.
-One example is the standard sweep line algorithm of Bentley and Ottmann \cite{bo-arcgi-79} for line segment intersection, 
-which runs $O((n+k)\log n)$ time where k is the number of intersections, which in the worst case is quadratic.
-Another example is Chan's algorithm for computing the convex hull \cite{c-ooscha-96} which runs in $O(n\log h)$ 
-where $h$ is the number of vertices on the hull, which in the worst case is $n$.  While such algorithms represent 
-significant improvements over their worst case counterparts, their running times are only measured with respect 
-to the size of the output and not any structure it may have.
-
-\paragraph{Instance Optimality.}  
-Here we are concerned with a more refined analysis than the typical output sensitive analysis.
-Specifically, we investigate an algorithm which is instance optimal as defined by Afshani \etal \cite{abc-ioga-09}.  
-In order to prove such a result, one considers a family of allowable partitions of the input, 
-each of which is defined in terms of the structure of the output.  Each such partition then has 
-an associated entropy type cost representing the number of possible instantiations of such a partition.
-Our work differs slightly from Afshani \etal \cite{abc-ioga-09} however, in that they show any algorithm 
-(for say the convex hull) requires the minimum such cost, where as in our case it is shown that 
-any algorithm requires the maximum over these costs.
-
-\subsubsection{Contour Trees}
-
-\Ben{Maybe discuss contour tree applications and why they are useful here? Or maybe earlier?}
-
-\paragraph{A History of Contour Trees.}
-As discussed above, there has been a long series of improved algorithms for computing contour trees.
-Kreveld \etal \cite{kobps-ctsssit-97} presented an $O(N\log N)$ time algorithm for functions over 2D 
-meshes and an $O(N^2)$ algorithm for higher dimensions, where $N$ is the number of simplices in the mesh. 
-Tarasov and Vyalya \cite{tv-cct-98} improved the running time to $O(N\log N)$ for the 3D case by making 
-several passes over the input.
-
-Carr \etal \cite{csa-cctad-00} improved the running time for all dimensions to $O(n\log n + N)$, 
-where $n$ is the number of mesh vertices.  Their algorithm works in 3 simple stages.  
-The first stage consist of a top down sweep to compute what is known as the join (or merge) tree, 
-which tracks the evolution of super-level sets (whereas the contour tree tracks level sets).  
-Similarly, the second stage does a bottom up sweep to compute the split tree, which tracks sub-level sets.  
-Their main contribution is the third and final step where they present an $O(n)$ time algorithm to merge 
-together the join and split trees to get the contour tree.
-
-There have also been some output sensitive results for computing the contour tree.  Let $t$ be the number 
-of critical points in the input mesh, which could potentially be much smaller than $n$.  
-Cole-McLaughlin and Pascucci \cite{pc-ectls-02} provided an $O(n+t\log n)$ time algorithm for the case of 
-$3$-dimensional structured meshes (note $N$ does not show up in the running time as the mesh is structured).
-Chiang \etal \cite{cllr-sooscctmp-05} improve upon this result, and provide an $O(N+t'\log t')$ algorithm 
-which works for structured or unstructured meshes in any dimension, where $t'$ is the number of component 
-critical vertices.  This is provably optimal with respect to the output size but not the output structure, 
-as discussed below.
-
-%As the the authors point out in \cite{cllr-sooscctmp-05}, by reduction from sorting, 
-%one can show that any algorithm to compute the contour tree in the worst case takes 
-%$\Omega(N+t\log t)$ time.  However, this lower bound only holds when the contour tree 
-%is very unbalanced (i.e. long and skinny) and therefore contains a sorted path with a 
-%large fraction of the vertices.  Therefore the bound of $\Omega(N+t\log t)$ is only 
-%worst case optimal.  In this paper we will consider an output sensitive approach which 
-%takes into account the shape of the contour tree.
+and studying chemical systems~\cite{LaBe+06,BrWe+10,BeWe+11,BrWe+11,MaGr+11}. Carr's thesis~\cite{c-tmi-04} gives various
+applications of contour trees for data visualization and is an excellent reference for contour trees definitions and algorithms.
+
+There has been a long series of improved algorithms for computing contour trees.
+van Kreveld \etal \cite{kobps-ctsssit-97} presented an $O(N\log N)$ time algorithm for functions over 2D 
+meshes and an $O(N^2)$ algorithm for higher dimensions. 
+Tarasov and Vyalya \cite{tv-cct-98} improved the running time to $O(N\log N)$ for the 3D case
+The influential paper of Carr \etal \cite{csa-cctad-00} improved the running time for all dimensions to $O(n\log n + N)$.
+One cannot overstate the importance of this result, since it forms the basis 
+of subsequent practical and theoretical work in computational topology applications. 
+Cole-McLaughlin and Pascucci \cite{pc-ectls-02} provided an $O(n+t\log n)$ time algorithm for 
+$3$-dimensional structured meshes. Chiang \etal \cite{cllr-sooscctmp-05} improve upon this result, and provide an unconditional $O(N+t\log t)$ algorithm.
+
+Contour trees are a special case of Reeb graphs, a general topological representation for real-valued function
+on any manifold. We refer the reader to Chapter 6 in~\cite{HaEd10} for more details. Algorithms for computing Reeb graphs
+is an active topic of research~\cite{sk-crgacs-91,cehnp-lrbm-03,PaScBr07,DoNa09,HaWaWe10,Pa12}, where
+two results explicitly reduce to computing contour trees~\cite{TiGySi09,DoNa13}.
 
@@
 
-\paragraph{The Issue of Sorting.}
-As discussed above, the algorithm of Carr \etal \cite{csa-cctad-00} computes the contour 
-tree by first constructing the join and split trees.  As vertices appear in sorted order 
-in any root to leaf path in the join and split trees, and in any maxima to minima path 
-in the contour tree, sorting the input vertices is a natural first step.  However, not 
-all vertices from the input need to be sorted.  In particular, Chiang \etal 
-\cite{cllr-sooscctmp-05} improved the running time by observing that as the join and 
-split trees only consist of critical points, it suffices to only sort critical points.  
-If a large fraction of the vertices appear in a single path in the join or split tree, 
-then paying to sort all the critical points is unavoidable, and therefore in the worst 
-case \cite{cllr-sooscctmp-05} is optimal.  However, if the trees are more balanced then 
-sorting all critical points together can be very costly.  Specifically, the contour tree 
-does not care about the relative ordering of vertices which do not have an 
-ancestor-descendant relationship, and so this information does not need to be computed.
-See \Fig{relative} for an example.
-
-% \begin{figure}[h]\centering
-%     \includegraphics[width=.3\linewidth,height=.42\linewidth]{figs/shaded}%
-%     \hspace{.5in}
-%     \includegraphics[width=.3\linewidth,height=.42\linewidth]{figs/trees2}
-%     \caption{On the left, a surface will a balanced contour tree, but whose join and split trees have long tails.  
-%     On the right (in clockwise order from the top) the contour, split and join trees.}
-%     \label{fig:sorting}
-% \end{figure}
 \begin{figure}[h]\centering
@@
 
-\paragraph{Suboptimal Join and Split Trees.}
-There is also a more fundamental problem with the now standard approach of first computing 
-the join and split trees in order to compute the contour tree.  In particular, this approach 
-can never lead to an algorithm that is optimal with respect to the shape of the contour tree.  
-This is because it is possible to have join and split trees which are long and skinny (i.e. 
-costly to compute) whereas the contour tree is balanced.  For example consider the case shown 
-in \Fig{sorting}, where the mesh looks like two (thickened) balanced binary trees, one upside 
-down, and joined at their roots.  In this case the contour tree is potentially cheap to 
-compute as there is no long sorted path.  However, the join and split trees will have long 
-sorted tails (consisting of the vertices in the bottom and top balanced trees, respectively), 
-and therefore require $\Omega(n\log n)$ time to compute.
-
-\subsection{Overview and Contribution}
-
-\paragraph{Algorithm Overview.}
-At a high level, the technique is rather simple.  Break the input manifold into nicely 
-behaved pieces, compute the contour tree on each piece, and then stitch together the computed 
-contour trees according to the cuts we made to break up the manifold.  Below we informally 
-describe how this is accomplished.
-\begin{compactitem}
- \item {\bf Raining to partition.}  To start with, pick an arbitrary maxima.  Now imagine 
- ``raining'' water down on the manifold from this maxima, i.e. mark all points which are reachable 
- by non-ascending paths from the maxima.  Let the boundary between wet and dry portions of the 
- manifold be called the interface (which may have different connected components).  
- Remove the wet submanifold, and invert the remaining pieces
- of the original manifold.  Now recursively rain in each one of these pieces, but rather than 
- picking an arbitrary maxima, start from the portion of the interface that cut that piece off.
-
- The benefit of this partitioning is that each resulting piece 
- is maxima dominant, i.e. there is a single maxima that has a non-ascending path to each point.  
- This implies the contour tree of this submanifold is the same as its split tree, or in other words 
- its contour tree is a rooted binary tree.
-
- %
- \item {\bf Painting trees.}  For each maxima dominant piece we want to compute the split 
- tree without sorting all the vertices.  For conceptual ease lets first flip the submanifold 
- (i.e., it is now minima dominant and we compute the join tree).  Now arbitrarily pick a maxima 
- and spill paint from it, specifically color all edges and vertices of the mesh that are reachable 
- from the maxima by descending paths.  Now go to each other maxima and spill a different color, 
- but be aware that paints do not mix, i.e. if we reach a vertex that was previosly explored 
- by another color then we do not explore it.
-
- This procedure partitions the set of critical points into color classes.  Clearly each color 
- class is connected in the manifold.  Suppose each class was also connected in the join tree.  
- In this case we partitioned the join tree into a set of descending paths, each of which 
- can be constructed by sorting the critical points in a class.  Roughly speaking, 
- these paths can then be joined together by what amounts to doing a post-order traversal of the tree.
- However, we may not be so lucky and the color classes may not be connected in the 
- tree.  This turns out not to be a problem, but it requires proper data-structure use. 
-%  
-%  Now suppose we were lucky and the spilling of paint in the manifold corresponded with a paint 
-%  spilling in the join tree, i.e. the color classes of the critical points partition the join 
-%  tree into disjoint descending paths, each containing one leaf.   In this case 
-%  we have partitioned the tree into branches, and so the tree can be formed by gluing these 
-%  branches together.  Intuitively (i.e. not exactly true), if two branches glue together then 
-%  it means there is a join vertex which got colored by the respective colors of each branch 
-%  (only one of these colors actually explored the vertex in the mesh and this is the ``main'' 
-%  branch you are gluing into).  Therefore, the tree can be constructed by starting at the leaf 
-%  of the trunk (i.e. the branch with the global minima) and recursively building the subtrees 
-%  while walking towards the root.  In other words we are building the tree by doing a post-order traversal.
-%  Unfortunately, we may not be so lucky and the color classes may not appear contiguously in 
-%  the tree.  In this case we can still build the tree in this leaf first fashion, but doing so 
-%  efficiently requires the use of data-structures.
-\end{compactitem}
-
-\paragraph{Lower Bound Overview.}
-In order to show that our algorithm for computing the join tree is optimal we consider paritioing the tree into disjoint 
-descending paths, each containing a leaf.  It turns out that each such partition into paths, $P$, 
-defines a lower bound of $\sum_{p\in P} |p|\log|p|$ for computing the contour tree, as you need to know the 
-sorted order of the vertices along each path in order to determine the tree.
-Specifically, we show that there is a set of at least $\Pi_{p_i\in P} (|p_i|-1)!$ distinct PL $2$-manifolds, 
-all of which have the same defining mesh, and for which each sorted order of each group of saddles determines a different join tree.
-
-The instance optimality of our algorithm then follows, as we also show that there is a particular path decomposition 
-whose cost upper bounds our.  Specifically, we assign a weight to each node in the tree which is determined by 
-the painting procedure described above (which represents the time it takes to process this node).  
-The path decomposition then turns out to be the heavy-light path decomposition 
-described by Sleator and Tarjan \cite{st-dsdt-83}, according to this weight function.
-However, this decomposition is just the first step in the analysis, which 
-requires further case analysis depending on the relative size of the length of each path 
-and the weights of its vertices.
-
-\paragraph{Our Contribution.}
-Our contribution can be broken into three main aspects.  The first two are algorithmic.  Namely, 
-first we describe a new linear time manifold partitioning scheme which breaks the manifold into 
-``extremum'' dominant submanifolds.  
-Second, we describe how to compute join trees optimally with respect to shape of the tree.
-As the contour tree of each submanifold is (roughly) equal to its join tree, this leads to an 
-optimal algorithm for the contour tree of the input manifold.  
-Therefore, our third and final contribution is proving the optimality of this result, which 
-requires a non-trivial amount of work, and uses techniques we believe to be of independent interest.
-
-\paragraph{Organization.}
-\Ben{To be Written}
-
 \section{Contour tree basics} \label{sec:basics}
 
-We detail the basic definitions about contour trees. The following is based on Chapter 6 of Carr's thesis \cite{c-tmi-04}.
-Our input is a continuous PL function $f:\MM \mapsto \RR$, where $\MM$ is a fully triangulated simplicial complex in $\RR^d$. 
-%All simplices are isomorphic to a standard $d$-dimensional simplex, except for specially designated \emph{boundary} simplices. 
-\Sesh{Some mention of representation and data structures here.}
-We use the standard assumption that $f$ is Morse.  In particular, we are assuming no two critical points have the same height, 
-and that there are no multi-saddles.
-
+We detail the basic definitions about contour trees. We follow the terminology of Chapter 6 of Carr's thesis \cite{c-tmi-04},
+which provides one of the first rigorous, Morse-theoretic defintions of contour trees. All our assumptions and definitions
+are standard for results in this area, though there is some variability in notation.
+The input is a continuous piecewise-linear (PL) function $f:\MM \mapsto \RR$, where $\MM$ is a fully triangulated simplicial complex in $\RR^d$,
+except for specially designated \emph{boundary facets}. So $f$ is defined only on the vertices of $\MM$,
+and all other valued are obtained by interpolation.
+
+We assume that $f$ has distinct values on all vertices, except for boundaries,
+as expressed in the following constraint.
+
+\begin{definition} \label{def:bound} $f$ is \emph{boundary critical} if the following holds.
+Consider a boundary facet $F$. All vertices of $F$ have the same function value. Furthermore, all
+neighbors of $F$ not in $F$ either have all function values strictly greater or strictly less
+than this value.
+\end{definition}
+
+This is convenient, as we can now assume that $f$ is defined on $\RR^d$. Any point inside a boundary facet
+has a well-defined height, and this includes the infinite facet. 
+We think of $d$ as constant, and assume that $\MM$ is represented in data structures that allow constant-time traversals
+in $\MM$ (e.g.~\cite{BoMa12}). (This is analogous to a DCEL for higher dimensions.)
 Observe that $f:\MM \rightarrow \RR$ can be thought of as a $d$-dimensional simplicial complex living in $\RR^{d+1}$.
-We can think of $f(x)$ as the ``height" of a point $x \in \MM$.
-We will often omit $f$, and assume that $f$ is encoded in the representation of $\MM$.
-
-\begin{definition} \label{def:level} The level set at value $h$ is the set $\{x \in \MM| f(x) = h\}$.
+We can think of $f(x)$ as the ``height" of a point $x \in \MM$, and encoded in the representation of $\MM$. 
+%Abusing notation, $\MM$ refers to this encoding, so we do not refer to $f$.
+
+\begin{definition} \label{def:level} The level set at value $h$ is the set $\{x| f(x) = h\}$.
 A contour is a maximal connected component of a level set. An $h$-contour is a contour where $f$-values are $h$.
@@
 
-Note that a contour is itself a simplicial complex of one dimension lower, and is represented (in our algorithms) as such.
+Note that a contour that does not contain a boundary is itself a simplicial complex of one dimension lower, and is represented (in our algorithms) as such.
 We let $\delta, \eps$ denote infinitesimals. Let $B_\eps(x)$ denote a ball of radius $\eps$ around $x$, and let 
@@
 
-\begin{definition} \label{def:deg} The up-degree of $x \in \MM$ is the number of $(f(x) + \delta)$-contours in $f|B_\eps(x)$
-as $\delta, \eps \rightarrow 0^+$. The down-degree is the number of $(f(x) - \delta)$-contours in $f|B_\eps(x)$
+\begin{definition} \label{def:deg} The Morse up-degree of $x$ is the number of $(f(x) + \delta)$-contours in $f|B_\eps(x)$
+as $\delta, \eps \rightarrow 0^+$. The Morse down-degree is the number of $(f(x) - \delta)$-contours in $f|B_\eps(x)$
 as $\delta, \eps \rightarrow 0^+$.
@@
 
-This definition allows us to categorize points in $\MM$ depending on the local changes in topology.
-
-\begin{definition} \label{def:points} A point $x \in \MM$ is categorized as follows.
+We categorize points depending on the local changes in topology.
+
+\begin{definition} \label{def:points} A point $x$ is categorized as follows.
 \begin{asparaitem}
-	\item Regular: Both up-degree and down-degrees are $1$.
-	\item Maximum: Up-degree is $0$.
-	\item Minimum: Down-degree is $0$.
-	\item Join: Up-degree is strictly greater than $1$.
-	\item Split: Down-degree is strictly greater than $1$.
+	\item Regular: Both Morse up-degree and down-degrees are $1$.
+	\item Maximum: Morse up-degree is $0$.
+	\item Minimum: Morse down-degree is $0$.
+	\item Morse Join: Morse up-degree is strictly greater than $1$.
+	\item Morse Split: Morse down-degree is strictly greater than $1$.
 \end{asparaitem}
-Non-regular points are called \emph{critical}. Joins and splits are called \emph{saddles}.
-Maxima and minima are called \emph{extrema}.
+Non-regular points are \emph{critical}. Morse joins and splits are \emph{saddles}.
+Maxima and minima are \emph{extrema}.
 \end{definition}
 
+We use $\cV(f)$ to denote the set of critical points.
+Because $f$ is piecewise-linear, all critical points are vertices in $\MM$. 
+A value $h$ is called critical, if $f(v) = h$, for some $v \in \cV(f)$. 
+The critical points are exactly where the topology of level sets change.
+By assuming that our manifold is boundary critical, the vertices on a given boundary are either collectively all maxima or all minima.  
+We abuse notation and refer to this entire set of vertices as a maxima or minima.
+%In particular, we do not view this as a violation of the Morse condition on $f$, which requires distinct heights for critical points.
+
+\ignore{
+The Morse assumption
+implies that all critical points have distinct values. A contour that involves a critical
+point is also denoted as critical. Note that the vertices in a boundary face collectively behaves like an extremum.
+Abusing notation, the term ``critical point", ``maxima", and ``minima" will include boundaries.
+Critical values include these boundary values. By boundary criticality, all contours are themselves
+represented as $d-1$-dimensional simplicial complexes.
+}
+
+\begin{definition} \label{def:equiv} Two contours $\psi$ and $\psi'$ are equivalent 
+if the following holds. There exists an $f$-monotone path $p$ connecting a point in $\psi$ to $\psi'$,
+such that no $x \in p$ belongs to a critical contour.
+\end{definition}
+
+These give the \emph{contour classes}. By continuity, a regular point is present on a single contour,
+and we can partition the regular contours into these classes. Every such class maps to intervals
+of the form $(f(x_i),f(x_j))$, where $x_i, x_j$ are critical points. Such a class is said
+to be created at $x_i$ and destroyed at $x_j$.
+
+\begin{definition} \label{def:tree} The contour tree is the graph on vertex set $\cV$, where
+edges are formed as follows. For every contour class that is created at $v_i$ and destroyed $v_j$,
+there is an edge $(v_i,v_j)$. (Conventionally, edges are directed from higher function value to lower.)
+\end{definition}
+
+We denote the contour tree of $\MM$
+by $\reeb(\MM)$. The corresponding node and edge sets are denoted as $\cV(\cdot)$ and $\cE(\cdot)$.
+It is not immediately obvious that this graph is a tree, but the above definition is equivalent to that of the Reeb graph of $f:\MM \rightarrow \RR$, 
+known to be a tree for PL functions defined over PL manifolds (Chapter 6 of~\cite{HaEd10}). Alternate definitions of the contour tree
+in~\cite{csa-cctad-00} imply this is a tree. We can talk about up-degrees and down-degrees in $\reeb(\MM)$
+We assume there are no ``multi-saddles", so up and down-degrees are at most $2$, and the total degree is at most $3$. 
+This is again a standard assumption in topological algorithms and can be achieved by vertex unfolding (Section 6.3 in~\cite{HaEd10}).
+
+\subsection{Some technical remarks}
+
+The $(f(x)+\delta)$-contours in $f|_{B_\eps(x)}$ given by \Def{deg} might actually be the same
+contour in $f$. The up-degree is defined as the number of $(f(x)+\delta)$-contours
+that intersect $B_\eps(x)$, a potentially smaller number. This up-degree
+is exactly the up-degree of $x$ in $\reeb(\MM)$. (Analogously, for down-degree.)
+When the Morse up-degree is $2$ but the up-degree is $1$, then the level set gains a handle.
+When $d=2$, this distinction is not necessary, since any point with $>1$
+Morse degrees will have some $>1$ degree in $\reeb(\MM)$.
+
+As Carr points out in his thesis, the term contour tree can be used for a family of related structures.
+Every vertex in $\MM$ can be asociated with an edge in $\reeb(\MM)$, and sometimes the
+vertex is explictly placed in $\reeb(\MM)$ (subdividing the respective edge). This is referred
+to as augmenting the contour tree, and it is common to augment $\reeb(\MM)$ will all vertices.
+Alternately, one can smooth out all vertices of up-degree and down-degree $1$ to get the
+unaugmented contour tree. (For $d=2$, there are no such vertices in $\reeb(\MM)$.) 
+The contour tree of \Def{tree} is the typical definition in all results on output-sensitive
+contour trees, and is the smallest tree that contains all the topological changes
+of level sets. Our algorithm works for any augmentation of $\reeb(\MM)$ with a predefined
+set of vertices, though we will not delve into these aspects in this paper.
+
+\ignore{
 \begin{remark}
@@
 
-We will enforce a special constraint on boundaries, encapsulated in the following definition.
-
-\begin{definition} \label{def:bound} $f$ is \emph{boundary critical} if the following holds.
-Consider a boundary simplex $F$. All vertices of $F$ have the same function value. Furthermore, all
-neighbors of $F$ not in $F$ either have all function values strictly greater or strictly less
-than this value.
-\end{definition}
-
-\begin{remark}
-Note that by assuming that our manifold is boundary critical, the vertices on a given boundary are either all maxima or all minima.  
-Since all such vertices on a boundary component are in the same contour, we will abuse notation and refer to this entire set of vertices as a maxima (resp. minima).
-In particular, we do not view this as a violation of the Morse condition on $f$, which requires distinct heights for critical points.
-\end{remark}
-
-We use $\cV$ to denote the set of critical points in $\MM$.
-Because $f$ is piecewise-linear, all critical points are vertices in $\MM$. 
-A value $h$ is called critical, if $f(x) = h$, for some $x \in \cV$. The Morse assumption
-implies that all critical points have distinct values. A contour that involves a critical
-point is also denoted as critical. Note that the vertices in a boundary face collectively behaves like an extremum.
-Abusing notation, the term ``critical point", ``maxima", and ``minima" will include boundaries.
-Critical values include these boundary values. By boundary criticality, all contours are themselves
-represented as $d-1$-dimensional simplicial complexes.
-
-\begin{definition} \label{def:equiv} Two contours $\psi$ and $\psi'$ are equivalent 
-if the following holds. There exists an $f$-monotone path $p$ connecting a point in $\psi$ to $\psi'$,
-such that no $x \in p$ belongs to a critical contour.
-\end{definition}
-
-These give the \emph{contour classes}. By continuity, a regular point is present on a single contour,
-and we can partition the regular contours into these classes. Every such class maps to intervals
-of the form $(f(x_i),f(x_j))$, where $x_i, x_j$ are critical points. Such a class is said
-to be created at $x_i$ and destroyed at $x_j$.
-
-\begin{definition} \label{def:tree} The contour tree is the graph on vertex set $\cV$, where
-edges are formed as follows. For every contour class that is created at $v_i$ and destroyed $v_j$,
-there is an edge $(v_i,v_j)$. (Conventionally, edges are directed from higher function value to lower.)
-\end{definition}
-
-We denote the contour tree of $\MM$
-by $\reeb(\MM) = \reeb(\MM,f)$. The corresponding node and edge sets are denoted as $\cV(\cdot)$ and $\cE(\cdot)$.
-%
-It is not immediately obvious that this graph is a tree, but the above definition is equivalent to that of the Reeb graph of $f:\MM \rightarrow \RR$ on $\MM$, and this is known to be a tree for PL functions defined over PL manifolds.
-
 \begin{remark}
@@
 \end{remark}
-
-We use a key theorem from \cite{c-tmi-04} (Theorem 6.6) that maps paths in $\cC(\MM)$ and $\MM$.
-
-\begin{theorem} \label{thm:carr-path} For every path $P$ in $\MM$, there exists a path $Q$ in the contour tree corresponding
-to the contours passing through points in $P$.
-For every path $Q$ in the contour tree, there exists at least one path $P$ in $\MM$ through points present
-in contours involving $Q$.
-\end{theorem}
-
-\begin{theorem} \label{thm:carr-mono} For every monotone path $P$ in $\MM$, there exists a monotone path $Q$ in the contour tree to which $P$
-maps (as in the previous theorem), and vice versa.
-\end{theorem}
+}
 
@@
 \begin{theorem} \label{thm:surgery} For any regular contour $\phi$, the output of $\surgery(\MM,\phi)$ is $\reeb(\MM)$.
+\end{theorem}
+
+We require some theorems from \cite{c-tmi-04} (Theorems 6.6 and 6.7) that map paths in $\cC(\MM)$ to $\MM$.
+
+\begin{theorem} \label{thm:carr-path} For every path $P$ in $\MM$, there exists a path $Q$ in the contour tree corresponding
+to the contours passing through points in $P$.
+For every path $Q$ in the contour tree, there exists at least one path $P$ in $\MM$ through points present
+in contours involving $Q$.
+\end{theorem}
+
+\begin{theorem} \label{thm:carr-mono} For every monotone path $P$ in $\MM$, there exists a monotone path $Q$ in the contour tree to which $P$
+maps (as in the previous theorem), and vice versa.
 \end{theorem}
@@
 
+\appendix
+
+\section{Old stuff}
+
+Their algorithm works in 3 simple stages.  
+The first stage consist of a top down sweep to compute what is known as the join (or merge) tree, 
+which tracks the evolution of super-level sets (whereas the contour tree tracks level sets).  
+Similarly, the second stage does a bottom up sweep to compute the split tree, which tracks sub-level sets.  
+Their main contribution is the third and final step where they present an $O(n)$ time algorithm to merge 
+together the join and split trees to get the contour tree.
+
+There have also been some output sensitive results for computing the contour tree.  Let $t$ be the number 
+of critical points in the input mesh, which could potentially be much smaller than $n$.  
+Cole-McLaughlin and Pascucci \cite{pc-ectls-02} provided an $O(n+t\log n)$ time algorithm for the case of 
+$3$-dimensional structured meshes (note $N$ does not show up in the running time as the mesh is structured).
+Chiang \etal \cite{cllr-sooscctmp-05} improve upon this result, and provide an $O(N+t'\log t')$ algorithm 
+which works for structured or unstructured meshes in any dimension, where $t'$ is the number of component 
+critical vertices.  This is provably optimal with respect to the output size but not the output structure, 
+as discussed below.
+
+\subsubsection{Worst Case Alternatives}
+In worst case analysis an algorithm's running time is judged in terms of its worst 
+case behavior over all inputs of a given size.  While there are benefits to this viewpoint, 
+such as robustness,  it fails to capture the whole story.  As such, many alternatives 
+have been considered.  For example, one can make assumptions on the distribution of 
+inputs, which has led to a number interesting results on average case and probabilistic 
+analysis, and also smoothed analysis \cite{st-saa-04}.  However, even without making 
+assumptions on the distribution of the input, there is still more to the story.
+For example, in the study of parametrized complexity, one takes into account other 
+parameters determined by the input other than just its size (for example the maximum 
+degree in a graph or whether it is planar).
+
+\paragraph{Output Sensitive Analysis.}
+In geometric settings often the goal is to compute a (potentially complex) geometric structure determined 
+by the input, and so it is natural to also consider how the form of this output structure affects the running time.
+One example is the standard sweep line algorithm of Bentley and Ottmann \cite{bo-arcgi-79} for line segment intersection, 
+which runs $O((n+k)\log n)$ time where k is the number of intersections, which in the worst case is quadratic.
+Another example is Chan's algorithm for computing the convex hull \cite{c-ooscha-96} which runs in $O(n\log h)$ 
+where $h$ is the number of vertices on the hull, which in the worst case is $n$.  While such algorithms represent 
+significant improvements over their worst case counterparts, their running times are only measured with respect 
+to the size of the output and not any structure it may have.
+
+\paragraph{Instance Optimality.}  
+Here we are concerned with a more refined analysis than the typical output sensitive analysis.
+Specifically, we investigate an algorithm which is instance optimal as defined by Afshani \etal \cite{abc-ioga-09}.  
+In order to prove such a result, one considers a family of allowable partitions of the input, 
+each of which is defined in terms of the structure of the output.  Each such partition then has 
+an associated entropy type cost representing the number of possible instantiations of such a partition.
+Our work differs slightly from Afshani \etal \cite{abc-ioga-09} however, in that they show any algorithm 
+(for say the convex hull) requires the minimum such cost, where as in our case it is shown that 
+any algorithm requires the maximum over these costs.
+
+\subsubsection{Contour Trees}
+
+\Ben{Maybe discuss contour tree applications and why they are useful here? Or maybe earlier?}
+
+\paragraph{A History of Contour Trees.}
+
+%As the the authors point out in \cite{cllr-sooscctmp-05}, by reduction from sorting, 
+%one can show that any algorithm to compute the contour tree in the worst case takes 
+%$\Omega(N+t\log t)$ time.  However, this lower bound only holds when the contour tree 
+%is very unbalanced (i.e. long and skinny) and therefore contains a sorted path with a 
+%large fraction of the vertices.  Therefore the bound of $\Omega(N+t\log t)$ is only 
+%worst case optimal.  In this paper we will consider an output sensitive approach which 
+%takes into account the shape of the contour tree.
+
+In other words, for any valid cost $C$, we can find a family of inputs where \Thm{main-alg}
+guarantees a running time of $C + t\alpha(t)$, and any height decision tree requires $\Omega(C)$ time.
+This shows that the cost function is inherent to the computation of contour trees.
+
+Let $f$ be a real valued function defined over a manifold embedded in $\RR^d$.  
+For any real value $\alpha$, the level set $f^{-1}(\alpha)$ can be decomposed into its connected components, or ``contours''.
+The Reeb graph is a fundamental topological structure which represents the evolution of these contours as one sweeps the surface 
+from $-\infty$ to $\infty$ according to $f$.  Often this graph is acyclic, in which case it is 
+called a \emph{contour tree}.  This is the case if, for example, the input is a surface mesh of a mountain range 
+(and the contour tree would encode when mountain peeks form and how they merge together).
+
+There is a long list of practical applications of the study of contour trees (see for example \cite{c-tmi-04} and references therein).
+Often these applications involve massive data sets, and so consequently there has also been a long series 
+of results on efficient algorithms for their computation \cite{kobps-ctsssit-97,tv-cct-98,csa-cctad-00,pc-ectls-02,cllr-sooscctmp-05}.
+Let $N$ be the total size of the input manifold, $n$ the number of vertices, and $t$ the number of critical points. (Note that in many applications $N=O(n)$.)
+As an initial step these algorithms require that all $t$ contour tree vertices be sorted.  
+In particular, \cite{cllr-sooscctmp-05} provide an $O(N+t\log t)$ time algorithm, and since in the worst case computing the contour tree 
+requires $\Omega(N+t\log t)$ time, this algorithm is optimal with respect to the size of the output.  
+Here we present the first contour tree algorithm which is optimal (for instances requiring $\Omega(N\alpha(N))$ 
+time\footnote{Where $\alpha(\cdot)$ is the inverse Ackermann function, which is $< 5$ for computation in this universe.}) 
+with respect not just to the size of the output, but also its structure.  This is accomplished by introducing a simple linear time 
+surface partitioning scheme to avoid sorting, which is likely to have applications elsewhere.  
+For all instances the bound for our algorithm either matches the best previous bound, or is faster.
+For, example when the contour tree is a balanced, our algorithm runs in $O(N\alpha(N))$ time, whereas all 
+previous algorithms had a running time of $\Omega(N+t\log t)$
+
+%In particular, the algorithm in \cite{cllr-sooscctmp-05} is worst case optimal with respect to the size of the output.  
+%Here we present the first contour tree algorithm which is (near\footnote{Our algorithm is optimal on all instance requiring $\Omega(n\alpha(n))$ time.}) 
+%optimal with respect not just to the size of the output, but also its structure.  
+%Specifically, the bound for our algorithm either matches \cite{cllr-sooscctmp-05}, or is faster.
+%For, example when the contour tree is a balanced, our algorithm runs in $O(n\alpha(n))$ time, whereas all 
+%previous algorithms had a running time of $\Omega(n\log n)$.
+
+This work can be seen as a continuation of the study of instance-optimal geometric algorithms introduced by Afshani \etal \cite{abc-ioga-09}.
+In order to prove instance optimality we use a non-trivial combinatorial discharging argument, which we believe is of independent interest.
+Of particular note is the use of heavy-light path decompositions, first introduced by Sleator and Tarjan \cite{st-dsdt-83} in their analysis of link-cut trees.
+
+% Explain why useful:  Lets you know the ``shape of your date''.  Map of the critical points where contours change.  
+%Fast extraction of seed sets in a small sized structure, which can let you do threshold queries.
+% 
+% \medskip\noindent
+% Paragraph on specific applications.  Scientific visualization, in particular combustion research.  GIS.
+
+\paragraph{The Issue of Sorting.}
+As discussed above, the algorithm of Carr \etal \cite{csa-cctad-00} computes the contour 
+tree by first constructing the join and split trees.  As vertices appear in sorted order 
+in any root to leaf path in the join and split trees, and in any maxima to minima path 
+in the contour tree, sorting the input vertices is a natural first step.  However, not 
+all vertices from the input need to be sorted.  In particular, Chiang \etal 
+\cite{cllr-sooscctmp-05} improved the running time by observing that as the join and 
+split trees only consist of critical points, it suffices to only sort critical points.  
+If a large fraction of the vertices appear in a single path in the join or split tree, 
+then paying to sort all the critical points is unavoidable, and therefore in the worst 
+case \cite{cllr-sooscctmp-05} is optimal.  However, if the trees are more balanced then 
+sorting all critical points together can be very costly.  Specifically, the contour tree 
+does not care about the relative ordering of vertices which do not have an 
+ancestor-descendant relationship, and so this information does not need to be computed.
+See \Fig{relative} for an example.
+
+% \begin{figure}[h]\centering
+%     \includegraphics[width=.3\linewidth,height=.42\linewidth]{figs/shaded}%
+%     \hspace{.5in}
+%     \includegraphics[width=.3\linewidth,height=.42\linewidth]{figs/trees2}
+%     \caption{On the left, a surface will a balanced contour tree, but whose join and split trees have long tails.  
+%     On the right (in clockwise order from the top) the contour, split and join trees.}
+%     \label{fig:sorting}
+% \end{figure}
+
+\paragraph{Suboptimal Join and Split Trees.}
+There is also a more fundamental problem with the now standard approach of first computing 
+the join and split trees in order to compute the contour tree.  In particular, this approach 
+can never lead to an algorithm that is optimal with respect to the shape of the contour tree.  
+This is because it is possible to have join and split trees which are long and skinny (i.e. 
+costly to compute) whereas the contour tree is balanced.  For example consider the case shown 
+in \Fig{sorting}, where the mesh looks like two (thickened) balanced binary trees, one upside 
+down, and joined at their roots.  In this case the contour tree is potentially cheap to 
+compute as there is no long sorted path.  However, the join and split trees will have long 
+sorted tails (consisting of the vertices in the bottom and top balanced trees, respectively), 
+and therefore require $\Omega(n\log n)$ time to compute.
+
+\subsection{Overview and Contribution}
+
+\paragraph{Algorithm Overview.}
+At a high level, the technique is rather simple.  Break the input manifold into nicely 
+behaved pieces, compute the contour tree on each piece, and then stitch together the computed 
+contour trees according to the cuts we made to break up the manifold.  Below we informally 
+describe how this is accomplished.
+\begin{compactitem}
+ \item {\bf Raining to partition.}  To start with, pick an arbitrary maxima.  Now imagine 
+ ``raining'' water down on the manifold from this maxima, i.e. mark all points which are reachable 
+ by non-ascending paths from the maxima.  Let the boundary between wet and dry portions of the 
+ manifold be called the interface (which may have different connected components).  
+ Remove the wet submanifold, and invert the remaining pieces
+ of the original manifold.  Now recursively rain in each one of these pieces, but rather than 
+ picking an arbitrary maxima, start from the portion of the interface that cut that piece off.
+
+ The benefit of this partitioning is that each resulting piece 
+ is maxima dominant, i.e. there is a single maxima that has a non-ascending path to each point.  
+ This implies the contour tree of this submanifold is the same as its split tree, or in other words 
+ its contour tree is a rooted binary tree.
+
+ %
+ \item {\bf Painting trees.}  For each maxima dominant piece we want to compute the split 
+ tree without sorting all the vertices.  For conceptual ease lets first flip the submanifold 
+ (i.e., it is now minima dominant and we compute the join tree).  Now arbitrarily pick a maxima 
+ and spill paint from it, specifically color all edges and vertices of the mesh that are reachable 
+ from the maxima by descending paths.  Now go to each other maxima and spill a different color, 
+ but be aware that paints do not mix, i.e. if we reach a vertex that was previosly explored 
+ by another color then we do not explore it.
+
+ This procedure partitions the set of critical points into color classes.  Clearly each color 
+ class is connected in the manifold.  Suppose each class was also connected in the join tree.  
+ In this case we partitioned the join tree into a set of descending paths, each of which 
+ can be constructed by sorting the critical points in a class.  Roughly speaking, 
+ these paths can then be joined together by what amounts to doing a post-order traversal of the tree.
+ However, we may not be so lucky and the color classes may not be connected in the 
+ tree.  This turns out not to be a problem, but it requires proper data-structure use. 
+%  
+%  Now suppose we were lucky and the spilling of paint in the manifold corresponded with a paint 
+%  spilling in the join tree, i.e. the color classes of the critical points partition the join 
+%  tree into disjoint descending paths, each containing one leaf.   In this case 
+%  we have partitioned the tree into branches, and so the tree can be formed by gluing these 
+%  branches together.  Intuitively (i.e. not exactly true), if two branches glue together then 
+%  it means there is a join vertex which got colored by the respective colors of each branch 
+%  (only one of these colors actually explored the vertex in the mesh and this is the ``main'' 
+%  branch you are gluing into).  Therefore, the tree can be constructed by starting at the leaf 
+%  of the trunk (i.e. the branch with the global minima) and recursively building the subtrees 
+%  while walking towards the root.  In other words we are building the tree by doing a post-order traversal.
+%  Unfortunately, we may not be so lucky and the color classes may not appear contiguously in 
+%  the tree.  In this case we can still build the tree in this leaf first fashion, but doing so 
+%  efficiently requires the use of data-structures.
+\end{compactitem}
+
+\paragraph{Lower Bound Overview.}
+In order to show that our algorithm for computing the join tree is optimal we consider paritioing the tree into disjoint 
+descending paths, each containing a leaf.  It turns out that each such partition into paths, $P$, 
+defines a lower bound of $\sum_{p\in P} |p|\log|p|$ for computing the contour tree, as you need to know the 
+sorted order of the vertices along each path in order to determine the tree.
+Specifically, we show that there is a set of at least $\Pi_{p_i\in P} (|p_i|-1)!$ distinct PL $2$-manifolds, 
+all of which have the same defining mesh, and for which each sorted order of each group of saddles determines a different join tree.
+
+The instance optimality of our algorithm then follows, as we also show that there is a particular path decomposition 
+whose cost upper bounds our.  Specifically, we assign a weight to each node in the tree which is determined by 
+the painting procedure described above (which represents the time it takes to process this node).  
+The path decomposition then turns out to be the heavy-light path decomposition 
+described by Sleator and Tarjan \cite{st-dsdt-83}, according to this weight function.
+However, this decomposition is just the first step in the analysis, which 
+requires further case analysis depending on the relative size of the length of each path 
+and the weights of its vertices.
+
+\paragraph{Our Contribution.}
+Our contribution can be broken into three main aspects.  The first two are algorithmic.  Namely, 
+first we describe a new linear time manifold partitioning scheme which breaks the manifold into 
+``extremum'' dominant submanifolds.  
+Second, we describe how to compute join trees optimally with respect to shape of the tree.
+As the contour tree of each submanifold is (roughly) equal to its join tree, this leads to an 
+optimal algorithm for the contour tree of the input manifold.  
+Therefore, our third and final contribution is proving the optimality of this result, which 
+requires a non-trivial amount of work, and uses techniques we believe to be of independent interest.
+
+\paragraph{Organization.}
+\Ben{To be Written}
+
 \end{document}

### 2014-10-30 paint-contour-tree.tex
typos corrected

--- 2014-10-30 paint-contour-tree.tex
@@
 Suppose $\MM$ has $n$ vertices, $N$ facets, and $t$ critical points.
-A classic result of Carr, Snoeyink, and Axen (2000) give an algorithm that takes $O(n\log n + N\alpha(N))$ time
+A classic result of Carr, Snoeyink, and Axen (2000) gives an algorithm that takes $O(n\log n + N\alpha(N))$ time
 (where $\alpha(\cdot)$ is the inverse Ackermann function). A further improvement to $O(t\log t + N)$ time
@@
 A large class of data, such as satellite terrain maps, physical simulation output, and combustion data, can
-be thought as a function $f: \RR^d \mapsto \RR$. Typically, a finite representation is given
+be thought of as a function $f: \RR^d \mapsto \RR$. Typically, a finite representation is given
 by considering $f$ to be piecewise linear over some mesh (simplicial complex) $\MM$ in $\RR^d$.
@@
 \begin{figure}[h]\centering
-    \includegraphics[width=.27\linewidth]{figs/spike3}%
+    \includegraphics[width=.27\linewidth]{figs/spikey1}%
     \hspace{.35in}
-    \includegraphics[width=.27\linewidth]{figs/spike4}%
+    \includegraphics[width=.27\linewidth]{figs/spikey2}%
     \hspace{.4in}
-    \includegraphics[width=.22\linewidth]{figs/spikeTree2}
+    \includegraphics[width=.22\linewidth]{figs/spikeyTree}
     \caption{Two surfaces with different orderings of the maxima, but the same contour tree.}
@@
 \begin{figure}[h]\centering
-    \includegraphics[width=.7\linewidth,height=.28\linewidth]{figs/balanced2}
+    \includegraphics[width=.7\linewidth,height=.3\linewidth]{figs/balanced3}
+    \vspace{.1in}
+
     \includegraphics[width=.7\linewidth,height=.25\linewidth]{figs/trees}
@@
 are put into a priority queue keyed by height, and vertices are handled in that order.
-
-\Ben{This figure will not show up in black and white printout.}
 

### 2014-10-31 paint-contour-tree.tex
section 8 modifications

@@
 \newcommand{\cost}{\mathop{cost}}
+\newcommand{\costs}{\mathop{costs}}
 \newcommand{\h}{att}
@@
 in running time. For a partition of the contour tree into a set of descending paths, $P$,
-our algorithm runs in $O(\sum_{p\in P} |p|\log|p| + t\alpha(t))$. Moreover, we prove a general lower bound of
+our algorithm runs in $O(\sum_{p\in P} |p|\log|p| + t\alpha(t)+N)$, which is 
+always bounded by $O(t\log t + N)$ and is $O(N+t\alpha(t))$ for balanced trees. 
+Moreover, we prove a general lower bound of
 $\Omega(\sum_{p \in P} |p|\log |p|)$ for computing contour trees, showing
@@
 Massive geometric data is often represented as a function $f: \RR^d \mapsto \RR$. Typically, a finite representation is given
-by considering $f$ to be piecewise linear over some mesh (simplicial complex) $\MM$ in $\RR^d$.
+by considering $f$ to be piecewise linear over some mesh (i.e. simplicial complex) $\MM$ in $\RR^d$.
 \emph{Contour trees} are a specific topological data structure used to represent and visualize
@@
 It is convenient to think of $f$ as a $(d+1)$-dimensional manifold, or alternately consider
-$\MM$ in $\RR^{d+1}$ with the last coordinate (height) given by $f$. Imagine sweeping the hyperplane $x_{d+1} = h$
+$\MM$ in $\RR^{d+1}$ with the last coordinate (i.e. height) given by $f$. Imagine sweeping the hyperplane $x_{d+1} = h$
 with $h$ going from $+\infty$ to $-\infty$. At every instance, the intersection of this plane 
-with $f$ gives a set of connected components, the \emph{contours} at height $h$. At the sweeping 
-proceeds various events occur: new contours are created or destroyed, contours merge into each other,
+with $f$ gives a set of connected components, the \emph{contours} at height $h$. As the sweeping 
+proceeds various events occur: new contours are created or destroyed, contours merge into each other or
 split into new components, contours acquire or lose handles. These are the relevant events where 
@@
 
-\Fig{relative} and \Fig{sorting} show examples of simplicial complices with heights and their contour trees.
-Think of the edges pointing downwards, so we can talk of the ``up-degree" and ``down-degree" of vertices.
-Leaves of down-degree $1$ are maxima, because a new contour is created at such point. Leaves of up-degree
-$1$ are minimum, because a contour is destroyed there. Up-degree $2$ vertices are ``joins", where contours
-merge, and down-degree $2$ vertices are ``splits". In these examples ($d=2$), contours are closed loops.
+\Fig{relative} and \Fig{sorting} show examples of simplicial complexes, with heights and their contour trees.
+Think of the contour tree edges as pointing downwards, so we can talk of the ``up-degree" and ``down-degree" of vertices.
+Leaves of down-degree $1$ are maxima, because a new contour is created at such points. Leaves of up-degree
+$1$ are minima, because a contour is destroyed there. Up-degree $2$ vertices are ``joins", where contours
+merge, and down-degree $2$ vertices are ``splits".  When $d=2$ (as in \Fig{relative}), contours are closed loops.
 In dimensions $d \geq 3$, we may have internal
@@
 But this is only a worst-case bound. All previous algorithm begin by sorting (at least) the critical points, and
-pay must $\Omega(t\log t)$ for all instances. Can we beat this sorting bound? Is there a more refined version of the complexity
+hence must pay $\Omega(t\log t)$ for all instances. Can we beat this sorting bound? Is there a more refined version of the complexity
 of computing contour trees? Our main result gives an affirmative answer. First, we require some technical definitions.
@@
 In a contour tree $T$, a \emph{leaf path} is simply a contiguous path in $T$
-containing a leaf, which is also monotonic in the height values of its vertices. A leaf path runs either upwards
-or downwards, never both.
+containing a leaf, which is also monotonic in the height values of its vertices. 
+%A leaf path runs either upwards or downwards, never both.
+%
 % Then in the following two subsections 
@@
 
-\begin{theorem} \label{thm:main-alg} Consider PL-Morse\footnote{This is a standard ``well-behaved" assumption that implies
-that all critical points have distinct values, and all the contour tree has max-degree $3$. Our algorithm
+\begin{theorem} \label{thm:main-alg} Consider a PL-Morse\footnote{This is a standard ``well-behaved" assumption that implies
+that all critical points have distinct values, and the contour tree has max-degree $3$. Our algorithm
 does not need this, but it simplifies the proof and presentation.} $f:\MM \mapsto \RR$, where the contour tree $T$
 has maximum degree $3$. There is an algorithm whose running time is $O(\sum_{p \in P(T)} |p|\log |p| + t\alpha(t) + N)$,
-where $P(T)$ is a specific path decomposition (constructed implictly by the algorithm).
+where $P(T)$ is a specific path decomposition (constructed implicitly by the algorithm).
 \end{theorem}
@@
 Let us interpret the running time. 
-For any path decomposition $P(T)$ and any $p \in P(T)$, obviously $|p| \leq |T| = t$.
-So the running time is clearly $O(t\log t + N)$. But this is only the case when $T$ is tall and thin.  
-An obvious upper bound is $O(t\log D + t\alpha(t) + N)$, where $D$ is the diameter of $P(T)$. 
+For any path decomposition $P(T)$, obviously $\sum_{p\in P(T)}|p| = |T| = t$, and 
+so the running time is $O(t\log t + N)$. But this is tight only when $T$ is tall and thin.  
+A better but still obvious upper bound is $O(t\log D + t\alpha(t) + N)$, where $D$ is the diameter of $P(T)$. 
 This is clearly an improvement when the tree is short and fat.
-But if $P(T)$ is a rooted, balanced, binary tree, then $\sum_{p \in P(T)} |p|\log |p| = O(t)$.
-Our running time is now $O(t\alpha(t))$, a significant improvement from $O(t\log t)$.
+In particular, for any decomposition of a rooted balanced binary tree, this gives $O(t\log\log n + t\alpha(t) + N)$, a significant improvement from $O(t\log t+N)$.
+However, our actual bound in this case gives an even further improved $O(t\alpha(t)+N)$.
+%
 Consider the example of \Fig{sorting} extended to a larger binary tree both above and below.
@@
 
-We complement \Thm{main-alg} with an almost matching lower bound. We show that the complexity
+We complement \Thm{main-alg} with an almost matching lower bound, showing that the complexity
 of $\sum_{p \in P(T)} |p|\log |p|$ is required for any contour tree algorithm. The exact phrasing
-of the lower bound requires some care. Consider the $d=2$ (terrain) case. For any $f$,
-we use $\cost(f)$ to denote $\sum_{p \in P(T)} |p|\log |p|$ for the specific $P(T)$
-given in \Thm{main-alg}. Ideally, we would like to argue that any contour-tree algorithm
-must pay $\cost(f)$ for $f$. But this is obviously false, since one could consider an algorithm
+of the lower bound requires some care. Consider the $d=2$ (terrain) case, and fix a mesh $\MM$.
+As referenced in \Thm{main-alg}, for any height function $f$ our algorithm defines a 
+specific path decomposition $P(T)$\footnote{The specific decomposition used is defined in \Sec{pathDecomp}, 
+but also depends on the raining and painting procedures described in \Sec{rain} and \Sec{paint}.  
+The important thing is that our algorithm is deterministic, and so for a fixed $\MM$ and $f$ it is well defined.}. 
+We define $\cost(f) = \sum_{p \in P(T)} |p|\log |p|$ for this specific decomposition $P(T)$.
+Ideally, we would like to argue that any contour-tree algorithm
+must pay $\cost(f)$ for $f$. 
+But this is obviously false, since one could consider an algorithm
 that hard codes the answer for $f$ and merely checks this answer.
 
-Instead, call a possible cost $C$ \emph{valid} if $C = \cost(f)$ for some $f$\footnote{If one is concerned
+Instead, call a possible cost $C$ \emph{valid} if $C = \cost(f)$ for some $f$ 
+defined over the fixed mesh $\MM$\footnote{If one is concerned
 that $C$ might be irrational, define valid if $2^C = 2^{\cost(f)}$.}. We prove
-a refined lower bound matching \Thm{main-alg}. A \emph{height decision tree} is an algorithm
-that is allowed aribtrary computation on the domain $\MM$, and is an algebraic decision
+a refined lower bound matching \Thm{main-alg}.
+
+\begin{definition}\label{def:hfdt}
+A \emph{height function decision tree} is an algorithm
+that is allowed arbitrary computation on the domain $\MM$, and is an algebraic decision
 tree over the heights. The running time is simply the number of queries made.
-
-\begin{theorem} \label{thm:main-lb} Consider the set of functions $\bF_C$ that all have the same (valid) cost $C$.
-Any height decision tree that correctly computes
-the contour tree on all of $\bF$ has an average case (uniform over $\bF$) running time of $\Omega(C)$.
+\end{definition}
+
+\begin{theorem} \label{thm:main-lb} Consider the set of functions $\bF_C$ that all have the same (valid) cost $C$ 
+(defined over a fixed mesh $\MM$).  Any height function decision tree that correctly computes
+the contour tree on all of $\bF_C$ has an average case (uniform over $\bF_C$) running time of $\Omega(C)$.
 \end{theorem}
 
-In other words, it is not possible to improve on \Thm{main-alg} even when restricted to the subset
+% We complement \Thm{main-alg} with an almost matching lower bound, showing that the complexity
+% of $\sum_{p \in P(T)} |p|\log |p|$ is required for any contour tree algorithm. The exact phrasing
+% of the lower bound requires some care. Consider the $d=2$ (terrain) case, and fix a mesh $\MM$.
+% Now any height function $f$ (together with $\MM$) defines a specific contour tree $T$.
+% As referenced in \Thm{main-alg}, given $f$, our algorithm then defines some specific path 
+% decomposition $\overline{P}(T)$, such that the running time of our algorithm is bounded by 
+% $O(\cost(\overline{P}(T))+t\alpha(t)+N)$, where $\cost(P(T)) = \sum_{p \in P(T)} |p|\log |p|$.
+% A priori, which path decomposition this will be is unclear\footnote{The specific decomposition, defined in \Sec{pathDecomp}, 
+% depends on arbitrary choices made during the raining and painting procedures described in \Sec{rain} and \Sec{paint}.}, 
+% however we know there is some large set of potential path decompositions it lies in
+% (the exact set is defined in \Def{path2}, but for now one can think of the set of all decompositions).
+% Let $\costs(f)$ denote the set of costs of all these potential path decompositions that $f$ defines
+% (i.e. $\cost(\overline{P}(T))\in \costs(f)$).
+% 
+% Ideally, we would like to argue that any contour-tree algorithm must pay $\cost(\overline{P}(T))$ for $f$. 
+% But this is obviously false, since one could consider an algorithm
+% that hard codes the answer for $f$ and merely checks this answer.
+% Instead, call a possible cost $C$ \emph{valid} if $C \in \costs(f)$ for some $f$ 
+% defined over the fixed mesh $\MM$\footnote{If one is concerned
+% that $C$ might be irrational, define valid if $2^C = 2^{\cost(f)}$.}. We prove
+% a refined lower bound matching \Thm{main-alg}. 
+% 
+% \begin{definition} \label{def:hfdt}
+% A \emph{height function decision tree} is an algorithm
+% that is allowed arbitrary computation on the domain $\MM$, and is an algebraic decision
+% tree over the heights. The running time is simply the number of queries made.
+% \end{definition}
+% 
+% \begin{theorem} \label{thm:main-lb} Consider the set of functions $\bF_C$ (defined over a fixed mesh $\MM$) 
+% that all have the same value $C$ as a valid cost.  Any height function decision tree that correctly computes
+% the contour tree on all of $\bF_C$ has an average case (uniform over $\bF_C$) running time of $\Omega(C)$.
+% \end{theorem}
+
+In other words, it is not possible to improve on \Thm{main-alg} even when restricted to a subset
 of inputs of fixed cost.
@@
 meshes and an $O(N^2)$ algorithm for higher dimensions. 
-Tarasov and Vyalya \cite{tv-cct-98} improved the running time to $O(N\log N)$ for the 3D case
+Tarasov and Vyalya \cite{tv-cct-98} improved the running time to $O(N\log N)$ for the 3D case.
 The influential paper of Carr \etal \cite{csa-cctad-00} improved the running time for all dimensions to $O(n\log n + N)$.
@@
 \begin{figure}[h]\centering
-    \includegraphics[width=.7\linewidth,height=.3\linewidth]{figs/balanced3}
+    \includegraphics[width=.7\linewidth,height=.3\linewidth]{figs/balanced4}
     \vspace{.1in}
@@
 has a well-defined height, and this includes the infinite facet. 
-We think of $d$ as constant, and assume that $\MM$ is represented in data structures that allow constant-time traversals
+We think of $d$ as constant, and assume that $\MM$ is represented in a data structure that allow constant-time traversals of neighboring simplices
 in $\MM$ (e.g.~\cite{BoMa12}). (This is analogous to a DCEL for higher dimensions.)
@@
 edges are formed as follows. For every contour class that is created at $v_i$ and destroyed $v_j$,
-there is an edge $(v_i,v_j)$. (Conventionally, edges are directed from higher function value to lower.)
+there is an edge $(v_i,v_j)$. (Conventionally, edges are directed from higher to lower function value.)
 \end{definition}
@@
 by $\reeb(\MM)$. The corresponding node and edge sets are denoted as $\cV(\cdot)$ and $\cE(\cdot)$.
-It is not immediately obvious that this graph is a tree, but the above definition is equivalent to that of the Reeb graph of $f:\MM \rightarrow \RR$, 
-known to be a tree for PL functions defined over PL manifolds (Chapter 6 of~\cite{HaEd10}). Alternate definitions of the contour tree
-in~\cite{csa-cctad-00} imply this is a tree. We can talk about up-degrees and down-degrees in $\reeb(\MM)$
+It is not immediately obvious that this graph is a tree, but alternate definitions of the contour tree
+in~\cite{csa-cctad-00} imply this is a tree. 
+Since this tree has height values associated with the vertices, we can talk about up-degrees and down-degrees in $\reeb(\MM)$.
 We assume there are no ``multi-saddles", so up and down-degrees are at most $2$, and the total degree is at most $3$. 
@@
 When $d=2$, this distinction is not necessary, since any point with $>1$
-Morse degrees will have some $>1$ degree in $\reeb(\MM)$.
+Morse degree will have some $>1$ degree in $\reeb(\MM)$.
 
@@
 \begin{asparaenum}
-	\item Let $\MM' = \cut(\phi,\MM)$.
+	\item Let $\MM' = \cut(\MM,\phi)$.
 	\item Construct $\reeb(\MM')$ and let $A, B$ be the nodes corresponding to the new boundaries created
@@
 \begin{lemma} \label{lem:cut} Consider a regular contour $\phi$ contained in a contour class (of an edge of $\reeb(\MM))$
-$(u,v)$ and denote $\MM' = \cut(\phi,\MM)$. Then $\cV(\reeb(\MM')) = \{\phi^+,\phi^-\} \cup \cV(\MM)$
+$(u,v)$ and denote $\MM' = \cut(\MM,\phi)$. Then $\cV(\reeb(\MM')) = \{\phi^+,\phi^-\} \cup \cV(\MM)$
 and $\cE(\reeb(\MM')) = \{(u,\phi^+),(\phi^-,v)\} \cup (\cE(\MM) \setminus (u,v))$.
@@
 
-\begin{theorem} \label{thm:jordan} $\cut(\phi,\MM)$ consists of two
+\begin{theorem} \label{thm:jordan} $\cut(\MM,\phi)$ consists of two
 disconnected simplicial complexes.
@@
 \begin{proof} Denote (as in \Lem{cut}) the edge containing $\phi$ to be $(u,v)$. Suppose for contradiction that there is a path between vertices $u$ and $v$
-in $\MM' = \cut(\phi,\MM)$. By \Thm{carr-path}, there is a path in $\cC(\MM')$ between $u$ and $v$. Since $\phi^+$ and $\phi^-$
+in $\MM' = \cut(\MM,\phi)$. By \Thm{carr-path}, there is a path in $\cC(\MM')$ between $u$ and $v$. Since $\phi^+$ and $\phi^-$
 are leaves in $\cC(\MM')$, this path obviously cannot use edges incident to them. By \Lem{cut},
 all the edges of this path are in $\cE(\cC(\MM)) \setminus (u,v)$. So we get a cycle in $\cC(\MM)$, a contradiction.
-To show that there are exactly two connected components in $\cut(\phi,\MM)$, it suffices
+To show that there are exactly two connected components in $\cut(\MM,\phi)$, it suffices
 to see that $\cC(\MM')$ has two connected components (by \Lem{cut}) and apply \Thm{carr-path}.
@@
 \begin{figure}[h]\centering
-    \includegraphics[width=.18\linewidth,height=.35\linewidth]{figs/orderGray1}%
-    \hspace{.8in}
-    \includegraphics[width=.18\linewidth,height=.35\linewidth]{figs/orderGray2}
+    \includegraphics[width=.25\linewidth]{figs/tent2}%
+    \hspace{.9in}
+    \includegraphics[width=.25\linewidth]{figs/tent1}
     \caption{On left, downward paint spilling only (each shade of gray represents a different color), producing a grid.  On right, flipping the direction of paint spilling.}
@@
 	\begin{compactenum}
-		\item Construct $\cut(\phi'_i,\NN_i)$, consisting of dry complex $\LL_i$ and remainder $\NN_{i+1}$.
+		\item Construct $\cut(\NN_i,\phi'_i)$, consisting of dry complex $\LL_i$ and remainder $\NN_{i+1}$.
 		\item Let the newly created boundary of $\LL_i$ be $B_i$. Invert $\LL_i$ so that $B_i$ is a maximum. Recursively
@@
 Consider the first time that $F$ is cut, say, during the call to $\rain(x,\NN)$.
-Specifically, say this happens when $\cut(\phi'_i,\NN_i)$ is constructed. 
+Specifically, say this happens when $\cut(\NN_i,\phi'_i)$ is constructed. 
 So $F$ is cut by some hyperplane into lower and upper portions (which are then triangulated).
@@
 \section{Painting to compute contour trees}
+\label{sec:paint}
 
@@
 \begin{definition}
+\label{def:leafAssign}
  A \emph{leaf assignment} $\chi$ of a tree $T$ assigns \emph{two} distinct leaves to each internal vertex $v$,
@@
 \section{Leaf assignments and path decompositions}
+\label{sec:pathDecomp}
 
@@
 \subsection{Join Trees}
-In this section we show that any algorithm for computing the join tree $T$ must pay at least $f_{path}(P(T))$ (see \Def{path}) for any path decomposition $P(T)$.
-
-%We say two shrubs $S_1$ and $S_2$ (of join trees of potentially different surfaces) are equivalent if there is an isomorphism, $f$,
-%between the two shrubs such that for any node $v\in S_1$ corresponding to a path of size $|v|$, $f(v)$ also corresponds to a path of size $|v|$.
-%For the purposes of the next lemma, we say two $2$-manifolds are different, if their corresponding contour trees are the different.
-
-In the following note that two distinct PL surfaces can have the same graph structure defining their meshes, but different height values associated with their vertices.
+
+In this section we prove \Thm{main-lb}, but for join trees instead of contour trees.  Observe that given a fixed $(f,\MM)$, 
+our algorithm for computing $\jc(\MM)$ is deterministic and so the path decomposition $P=P(\jc(\MM))$ in \Thm{runtime} is fixed, 
+allowing us to define the cost, $\cost(f)=\sum_{p\in P} |p|\log|p|$.
+
+%In the following note that two distinct terrains can have the same graph structure defining their meshes, but different height values associated with their vertices.
 
@@
 \label{lem:pathDecomp}
-Let $T$ be a join tree, and let $P = P(T)$ any path decomposition of $T$.  %, and $\pathTree$ the corresponding shrub.  
-Then there is a set of at least $\Pi_{p_i\in P} (|p_i|-1)!$ distinct PL $2$-manifolds, all of which have the same defining mesh.
+Let $\cost(f)=\sum_{p\in P} |p|\log|p|$, for some height function $f$ defined over a complex $\MM$.
+Then there is a family of functions, $F$, all defined over a common triangulation in plane, 
+such that the size of $F$ is $\Pi_{p_i\in P} (|p_i|-1)!$, and no two function in $F$ define the same join tree.
 \end{lemma}
 \begin{proof}
- We describe a simple mesh, with heights specified for all vertices except for the saddles.  The main idea is that 
- this mesh will be flexible enough to allow almost any assignment of heights to the saddles, thus defining a class of join trees 
+ We describe a simple triangulation, with heights specified for all vertices except for the saddles.  The main idea is that 
+ this triangulation will be flexible enough to allow almost any assignment of heights to the saddles, thus defining a class of join trees 
  whose size is that described in the lemma statement.
 
- First, we describe the basic building block of this input mesh, which roughly speaking corresponds to a fixed path $p\in P$.  
- Specifically, we say a \emph{diamond} of size $m$ is the following mesh.  To start, construct a cycle 
+ First, we describe the basic building block of these terrains, which roughly speaking corresponds to a fixed path $p\in P$.  
+ Specifically, we say a \emph{diamond} of size $m$ is the following terrain.  To start, construct a cycle 
  of length $m$, where the height of the $i$th vertex in the cycle is say $i/2m$.  Next, create an apex vertex at height 
  $\infty$ (or any appropriately large finite height), and add an edge between this vertex and each vertex in the cycle.  At this point 
- we have built an upside down cone with $m$ triangular faces (see \Fig{diamond}). To make this cone a closed surface we add a triangular cap at the bottom. 
- Specifically, add a $3$ cycle of vertices at heights $0$, $1/20m$, and $1/10m$, and connect each vertex of this $3$-cycle to an arbitrary distinct 
- vertex on the $m$ cycle (with matching cyclic ordering), and triangulate all remaining faces.  For consistency, if the size $m$ is $\leq 3$, the diamond 
- will be just a tetrahedron.
-
- Now above each face of the cone of the diamond, we will create a triangular ``platform'', which will act as the location to glue on other (shrunken) diamonds, 
+ we have built an upside down cone with $m$ triangular faces (see \Fig{diamond}). 
+ To make the terrain fully triangulated, we add a triangular outer face around the cycle. 
+ Specifically, add a triangle with vertices at heights $0$, $1/20m$, and $1/10m$, such that in the projection to the plane 
+ this triangle contains the $m$ length cycle, and there is an arbitrary triangulation of the space between this triangle 
+ and the cycle.  For consistency, the inner cycle of a diamond will always have at length at least $3$, i.e. even if $m\leq 3$.
+
+ Now above each face of the cone of the diamond, we will create a triangular ``platform'', which will act as the location to glue on the bounding triangle of other (shrunken) diamonds, 
  and whose highest point of connection to the face below will be a saddle vertex.  Specifically the heights of the vertices of the triangle will be $h$, $h-\eps$, and $h-2\eps$, where 
- $h$ is some unspecified height value.  This near horizontal platform is then connected to its respective face with vertical edges down to the face. See \Fig{diamond}.
- %
- %The vertex at height $h$ will lie in the plane of the respective face of the cone, and the other two will lie above the face, 
- %with edges connecting down to the face, hence creating an actual ``platform''.
-
- At this point we can construct the entire input mesh, by gluing together diamonds.  
+ $h$ is some unspecified height value.  This near horizontal platform is then connected to its respective face with (nearly) vertical edges down to the face. See \Fig{diamond}.
+
+ At this point we can construct the entire terrain, by gluing together diamonds.  
  Specifically, for each path $p_i\in P$ we create a diamond of size $|p_i|-1$ (we subtract one because the tip of the diamond already corresponds to the leaf of the path).
  Now we glue together diamonds of different paths in the same way the paths are connected in the shrub $\pathTree$.  Specially, for two paths $p,q\in P$ where 
- $p$ is the parent of $q$ in $\pathTree$, we glue the bottom triangular cap of the diamond for $q$ onto the triangular ``platform'' of some face of the diamond for $p$.
+ $p$ is the parent of $q$ in $\pathTree$, we glue the outer triangular cap of the diamond for $q$ onto the triangular ``platform'' of some face of the diamond for $p$.
  (Naturally for this construction to work, diamonds for a given path will be scaled down relative to the size of the diamond of their parent).
@@
  Observe now that the only saddle points in this construction are the highest connecting point on a face of an edge from the platform above.  Moreover, the only maxima are the apexes of the 
- diamonds, and there is exactly one minima corresponding to the bottom cap of the diamond representing the root of $\pathTree$.  Therefore, the saddles on a given 
- diamond will appear contiguously on a root to leaf path in the join tree of the mesh, where the leaf corresponds to the maxima of the diamond 
+ diamonds, and there is exactly one minima corresponding to the outer cap of the diamond representing the root of $\pathTree$.  Therefore, the saddles on a given 
+ diamond will appear contiguously on a root to leaf path in the join tree of the terrain, where the leaf corresponds to the maxima of the diamond 
  (since all these saddles have a direct line of sight to this apex).  In particular, 
- this implies that, regardless of the heights assigned to the platforms, the join tree of this mesh has a path decomposition whose corresponding shrub is equivalent to $\pathTree$.
-
- Now there is a valid instantiation of this described mesh for any relative ordering of the heights of the saddles on a given diamond.  In particular, 
- there are $(|p|-1)!$ possible orderings of the heights of the saddles on the diamond for $p$, and hence $\Pi_{p_i\in P} (|p_i|-1)!$ possible instantiations of 
- the mesh overall.  Also, each one of these instantiations will result in a different join tree.  Specifically, a permutation of the heights on a given diamond 
+ this implies that, regardless of the heights assigned to the platforms, the join tree has a path decomposition whose corresponding shrub is equivalent to $\pathTree$.
+
+ Now there is a valid instance of this described construction for any relative ordering of the heights of the saddles on a given diamond.  
+ Let the set of height functions for these instances be denote $F$.
+ In particular, there are $(|p|-1)!$ possible orderings of the heights of the saddles on the diamond for $p$, and hence $\Pi_{p_i\in P} (|p_i|-1)!$ possible 
+ functions in this set.  Also, each one of these functions will result in a different join tree.  Specifically, a permutation of the heights on a given diamond 
  corresponds to a permutation of the vertices along a path in $P$, as by construction all saddles on a given diamond will appear in sorted order in the join tree.
@@
 
-\begin{corollary}
- \label{cor:joinLB}
- Let $A(S)$ be any comparison based algorithm to compute the join tree of any input surface $S$ (whose contour tree is a join tree).  
- Then the running time of $A(S)$ is $\Omega(f_{path}(P))=\Omega(\sum_{p_i\in P} |p_i|\log|p_i|)$ where $P$ is any path decomposition of the 
- join tree of $S$.
-\end{corollary}
+\begin{lemma}
+\label{lem:cost}
+ For any $f\in F$, $\cost(f) = \sum_{p\in P} |p|\log|p|$
+\end{lemma}
 \begin{proof}
- Given an input surface $S$, by the above lemma, any path decomposition $P$ of its join tree defines a set of $\Pi_{p_i\in P} |p_i|!$ surfaces 
- with distinct join trees, all of which have the same mesh.  Specifically, for each sorted ordering of the relative height values of the saddles 
- on each diamond in the mesh, yields a different join tree.
- Therefore, any comparison based algorithm which computes the join tree must distinguish between these possibilities, and therefore its corresponding decision tree must 
- have $\Pi_{p_i\in P} |p_i|!$ leaves.  Hence by Stirling's approximation, this decision tree must have depth $\Omega(\sum_{p_i\in P} |p_i|\log|p_i|)$.
+First observe that by construction, regardless of the choice of $f\in F$, 
+there is some path decomposition $\overline{P}(T)$ (of the tree $f$ defines) such that 
+each path $p$ consists exactly of the set of critical points from a given diamond. 
+We now show that regardless of how paint spilling is done, this will be the maximum 
+path decomposition of $T$.
+
+ So consider any order of paint spilling over the surface defined by $f$.
+ Consider the set of joins around any diamond.  All these joins have an edge which connects 
+ directly to the maxima of that diamond.  In particular, all these joins have 
+ an upstar which is entirely colored by the color of this diamond's maxima.  
+ Therefore, in the leaf assignment of the join tree, determined by this painting (see \Def{leafAssign})
+ all these join get assigned a common color.  
+ In particular, all vertices in each path in $\overline{P}$ get assigned a common color.
+
+ \Ben{I am stuck here.}
+%  One can now prove by induction (on the distance of a path from the root of the shrub), 
+%  that if all vertices in every path have a common color, then that color can only appear 
+%  in that path and the vertex to which that path connects (which then trivial implies this 
+%  is a maximum path decomposition).  
+%  First observe that this trivially holds for the path containing the root of $T$.  
+%  For the induction step, consider any path $p$, and let $p'$ be the path 
+%  which the root of $r_p$ of the path $p$ connect to.
+%  Specifically, let $v$ be the vertex of $p'$ at which $p$ connects.
+%  
+%  So consider any vertex $v$ in the path $p$ which contains the root of $T$.  
+%  Now $v$ has two children, one child $c_1$ in $p$ and some other child $c_2$ in another path $p'$.
+%  %In order to show this is a maximum path decomposition we must argue $H_{c_1}\geq H_{c_2}$.
+%  The claim is that $H_{c_2}$ cannot contain any vertex of $p$ other than $v$.
+%  Specifically, consider any vertex on $p$ between the root and $v$.  This vertex must 
+%  get assigned two color, one from each of it subtree.  It is already assigned a color from 
+%  the subtree of $c_1$ hence its second color cannot come from the subtree of $c_2$, 
+%  since both $c_1$ and $c_2$ are in the same subtree of this vertex.
+%  This proves the base case, and the induction step follows by the same argument.
+%  
+%  One can now prove by induction (on the distance of a path from the root of the shrub), 
+%  that if all vertices in every path have a common color, then that color can only appear 
+%  in that path and the vertex to which that path connects (which then trivial implies this 
+%  is a maximum path decomposition).  
+%  So consider any vertex $v$ in the path $p$ which contains the root of $T$.  
+%  Now $v$ has two children, one child $c_1$ in $p$ and some other child $c_2$ in another path $p'$.
+%  %In order to show this is a maximum path decomposition we must argue $H_{c_1}\geq H_{c_2}$.
+%  The claim is that $H_{c_2}$ cannot contain any vertex of $p$ other than $v$.
+%  Specifically, consider any vertex on $p$ between the root and $v$.  This vertex must 
+%  get assigned two color, one from each of it subtree.  It is already assigned a color from 
+%  the subtree of $c_1$ hence its second color cannot come from the subtree of $c_2$, 
+%  since both $c_1$ and $c_2$ are in the same subtree of this vertex.
+%  This proves the base case, and the induction step follows by the same argument.
+\end{proof}
+
+The following is the analogous statement to \Thm{main-lb}, but for join trees.
+
+\begin{theorem}\label{thm:joinLB}
+Consider the set of functions $\bF_C$ that all have the same (valid) cost $C$.  
+Any height function decision tree that correctly computes
+the join tree on all of $\bF_C$, has an average case (uniform over $\bF_C$) running time of $\Omega(C)$.
+\end{theorem}
+\begin{proof}
+ If $C$ is a valid cost then there some $f$ such that $\cost(f)=\sum_{p\in P(T)} |p|\log|p|$, where $P(T)$ is the maximum path decomposition of the join tree defined by f.
+ By the above lemmas, there is then a set $\bF$ of $\Pi_{p_i\in P} |p_i|!$ functions, such that for any $g\in \bF$, $C= \cost(g)$.  
+ (Specifically, each sorted ordering of the relative height values of the saddles on each diamond yields a function in this class.)
+ Therefore, any height function decision tree for the set $\bF$ must distinguish between these possibilities, and therefore must 
+ have $\Pi_{p_i\in P} |p_i|!$ leaves.  Hence by Stirling's approximation, the average depth in this tree is $\Omega(\sum_{p_i\in P} |p_i|\log|p_i|)$.
 \end{proof}
@@
 Let $T$ be a contour tree, and let $P = P(T)$ be any rain consistent path decomposition of $T$.  
-Then there is a set of at least $\Pi_{p_i\in P} (|p_i|-1)!$ distinct PL $2$-manifolds, all of which have the same defining mesh.
+Then there is a set of at least $\Pi_{p_i\in P} (|p_i|-1)!$ distinct terrains, all of which have the same defining mesh.
 \end{lemma}
@@
 
- As the set $P_i$ is a path decomposition of a rooted binary tree $T_i$, the surface mesh construction of \Lem{pathDecomp} for $P_i$ is well defined.  
+ As the set $P_i$ is a path decomposition of a rooted binary tree $T_i$, the terrain construction of \Lem{pathDecomp} for $P_i$ is well defined.  
  Actually the only difference is that here the rooted tree is not a full binary tree, and so some of the faces of the constructed diamonds will be blank.  
@@
 
- So for each $P_i$ construct a surface mesh as described in \Def{path2}.  By construction, the diamonds corresponding to the paths in $P_i$ are connected 
+ So for each $P_i$ construct a surface mesh as described in \Lem{pathDecomp}.  By construction, the diamonds corresponding to the paths in $P_i$ are connected 
  into a tree structure (i.e. corresponding to the shrub of $P_i$).  Therefore the bottoms of all these diamonds are covered except for the one corresponding 
@@
 
-For the same reason that \Cor{joinLB} followed from \Lem{pathDecomp}, we now have the following.
+For the same reason that \Thm{joinLB} followed from \Lem{pathDecomp} and \Lem{cost}, we now have the following.
 

### 2014-11-01 paint-contour-tree.tex
done with first pass over section 3

--- 2014-10-30 paint-contour-tree.tex
@@
 \maketitle
+\thispagestyle{empty}
 
@@
 Suppose $\MM$ has $n$ vertices, $N$ facets, and $t$ critical points.
-A classic result of Carr, Snoeyink, and Axen (2000) gives an algorithm that takes $O(n\log n + N\alpha(N))$ time
+A classic result of Carr, Snoeyink, and Axen (2000) give an algorithm that takes $O(n\log n + N\alpha(N))$ time
 (where $\alpha(\cdot)$ is the inverse Ackermann function). A further improvement to $O(t\log t + N)$ time
@@
 
+\newpage
+
+\pagenumbering{arabic}
+
 \section{Introduction}
@@
 This is clearly an improvement when the tree is short and fat.
-But if $P(T)$ is a rooted, balanced, binary tree, then $\sum_{p \in P(T)} |p|\log |p| = O(t)$.
+But if $P(T)$ is a rooted, balanced, binary tree, then a calculation yields $\sum_{p \in P(T)} |p|\log |p| = O(t)$.
 Our running time is now $O(t\alpha(t))$, a significant improvement from $O(t\log t)$.
@@
 of inputs of fixed cost.
+
+\paragraph{Beyond worst-case analysis:} Our result provides understanding of the complexity
+of contour tree construction beyond the standard worst-case analysis. Our theorem and lower bound explain \emph{when}
+computing contour trees is expensive, and provide insight into the hardness of particular instances. It is intriguing
+that the cost (as defined earlier) roughly captures the hardness of an instance.
+In that sense, it is related to the concept of instance optimal algorithms, as defined by Afshani \etal \cite{abc-ioga-09}.
+Their notion of optimality is far more refined that what we prove, but it shares the flavor of understanding the spectrum
+of easy to hard instances. From a mathematical perspective, this sort of analysis forces one to look closely
+at the relationship between sorting and the contour tree problem.
+
+\ignore{
+classic output sensitive convex hull algorithms of Kirkpatrick and Seidel~\cite{KiSe86} and Chan~\cite{c-ooscha-96}.
+
+\paragraph{Output Sensitive Analysis.}
+In geometric settings often the goal is to compute a (potentially complex) geometric structure determined 
+by the input, and so it is natural to also consider how the form of this output structure affects the running time.
+One example is the standard sweep line algorithm of Bentley and Ottmann \cite{bo-arcgi-79} for line segment intersection, 
+which runs $O((n+k)\log n)$ time where k is the number of intersections, which in the worst case is quadratic.
+Another example is Chan's algorithm for computing the convex hull \cite{c-ooscha-96} which runs in $O(n\log h)$ 
+where $h$ is the number of vertices on the hull, which in the worst case is $n$.  While such algorithms represent 
+significant improvements over their worst case counterparts, their running times are only measured with respect 
+to the size of the output and not any structure it may have.
+
+\paragraph{Instance Optimality.}  
+Here we are concerned with a more refined analysis than the typical output sensitive analysis.
+Specifically, we investigate an algorithm which is instance optimal as defined by Afshani \etal \cite{abc-ioga-09}.  
+In order to prove such a result, one considers a family of allowable partitions of the input, 
+each of which is defined in terms of the structure of the output.  Each such partition then has 
+an associated entropy type cost representing the number of possible instantiations of such a partition.
+Our work differs slightly from Afshani \etal \cite{abc-ioga-09} however, in that they show any algorithm 
+(for say the convex hull) requires the minimum such cost, where as in our case it is shown that 
+any algorithm requires the maximum over these costs.
+}
 
@@
 $3$-dimensional structured meshes. Chiang \etal \cite{cllr-sooscctmp-05} improve upon this result, and provide an unconditional $O(N+t\log t)$ algorithm.
+The latter result crucially uses monotone paths to improve the running time, and we repeatedly
+exploit numerous properties of monotone paths. Carr's thesis shows relationships between monotone paths
+in $\MM$ and in $\cC(\MM)$.
 
@@
 \begin{figure}[h]\centering
-    \includegraphics[width=.7\linewidth,height=.3\linewidth]{figs/balanced3}
-    \vspace{.1in}
-
-    \includegraphics[width=.7\linewidth,height=.25\linewidth]{figs/trees}
+    \includegraphics[width=.35\linewidth,height=.15\linewidth]{figs/balanced3}
+    \hspace{.3in}
+    \includegraphics[width=.35\linewidth,height=.15\linewidth]{figs/trees}
     \caption{On top, a surface will a balanced contour tree, but whose join and split trees have long tails.  
@@
 
-As Carr points out in his thesis, the term contour tree can be used for a family of related structures.
-Every vertex in $\MM$ can be asociated with an edge in $\reeb(\MM)$, and sometimes the
-vertex is explictly placed in $\reeb(\MM)$ (subdividing the respective edge). This is referred
+As Carr points out in Chapter 6 of his thesis, the term contour tree can be used for a family of related structures.
+Every vertex in $\MM$ is associated with an edge in $\reeb(\MM)$, and sometimes the
+vertex is explictly placed in $\reeb(\MM)$ (by subdividing the respective edge). This is referred
 to as augmenting the contour tree, and it is common to augment $\reeb(\MM)$ will all vertices.
@@
 contour trees, and is the smallest tree that contains all the topological changes
-of level sets. Our algorithm works for any augmentation of $\reeb(\MM)$ with a predefined
+of level sets. \Thm{main-alg} is applicable for any augmentation of $\reeb(\MM)$ with a predefined
 set of vertices, though we will not delve into these aspects in this paper.
@@
 }
+
+\section{A tour of the contour tree algorithm} \label{sec:approach}
+
+Our final algorithm is quite technical and has numerous moving parts. This section
+provides a high level view of the entire result, and highlights the main ideas
+and intuition. 
+It is helpful to keep the $d=2$ case in mind, so the input is just a triangulated
+terrain. In the interest of presentation, the definitions and theorem statements in this section will
+slightly differ from those in the main body. They may also differ from the original definitions proposed
+in earlier work.
+
+\paragraph{Do not globally sort:} Arguably, the starting point for this work is \Fig{relative}. We have two terrains
+with exactly the same contour tree, but different orderings of (heights of) the critical point.
+Turning it around, we cannot deduce the height ordering of critical points from the contour tree. Clearly,
+sorting the critical points is computationally reduntant for constructing the contour tree.
+In \Fig{sorting}, the contour tree basically consists of two balanced binary trees, one of the joins,
+another of the splits. Again, it is not necessary to know the relative ordering between the mounds
+on the left (or among the depressions in the right) to compute the contour tree. Yet some ordering
+information is necessary: in the left, the little valleys are higher than the big central valley,
+and this is reflected in the contour tree. More generally, leafs paths in the contour tree have points
+in sorted order, but incomparable points in the tree are unconstrained.
+How do we sort exactly what is required, without knowing the contour tree in advance?
+
+\subsection{Breaking $\MM$ into simpler pieces} \label{sec:break}
+
+Let us begin with the algorithm of Carr, Snoeyink, and Axen~\cite{csa-cctad-00}. The key insight is to build
+two different trees, called the join and split trees, and then merge them together into the contour tree.
+Consider sweeping down via the hyperplane $x_{d+1} = h$ and taking the \emph{superlevel} sets. These 
+are the connected components of $\MM$ with height greater than $h$. For a terrain, the superlevel
+sets are a collection of ``mounds". As we sweep downwards, these mounds keep joining each other,
+until finally, we end up with all of $\MM$. The join tree tracks exactly these events. 
+Formally, let $\MM^+_v$ denote the simplicial complex obtained by taking only those
+vertices higher than $v$.
+
+\begin{definition} 
+\label{def:int-criticalJoin}
+The critical join tree $\jc(\MM)$ is built on the set $V'$ of all critical points. 
+The directed edge $(u,v)$ is present when $u$ is the smallest valued vertex in $V'$ in a connected component of $\MM^+_v$
+and $v$ is connected to this component (in $\MM$). 
+\end{definition}
+
+(Abusing notation, we just call this the join tree.)
+Refer to \Fig{sorting} for the join tree of a terrain. Note that nothing happens at splits, but these
+are still put as vertices in the join tree. They simply form a long path. The split tree
+is obtained by simply inverting this procedure, sweeping upwards and tracking sublevel sets.
+
+A major insight of~\cite{csa-cctad-00} is an ingeniously simple linear time procedure to construct
+the contour tree from the join and split tree. So the bottleneck is computing these trees. And observe
+in \Fig{sorting} that the split vertices form a long path in the join tree (and vice versa). Constructing
+these trees forces a global sort of the splits, a redundant computation. Unfortunately (unlike \Fig{sorting}),
+the heights of joins and splits may be interleaved in a complex manner. The final merging of~\cite{csa-cctad-00}
+to get the contour tree requires having the split vertices in the join tree. Without
+this, it is not clear how to get a consistent view of both joins and splits, required for the contour tree.
+
+Our aim is break $\MM$ into smaller pieces, where this redundant computation can be saved.
+
+%
+%{\bf Punchline:} We give a linear time procedure that partitions $\MM$ into smaller pieces. We guarantee
+%that the contour tree of each individual piece is basically the join (or split) tree. This saves
+%some of the redundant computation. We then show to put it all together the get the full contour tree.
+%Now for more details.
+
+\paragraph{Contour surgery:} We first need a divide-and-conquer lemma. Any contour $\phi$ can be associated
+with an edge $e$ of the contour tree. Suppose we ``cut" $\MM$ along this contour. We prove that $\MM$
+is split into two disconnected pieces, such the contour trees of these pieces is obtained
+by simply cutting $e$ in $\reeb(\MM)$. Alternatively, the contour trees of these pieces can
+be glued together to get $\reeb(\MM)$. This is not particularly surprising, and is fairly easy to prove
+with the right definitions. The idea of loop surgery has been used to reduce Reeb graphs to contour trees~\cite{TiGySi09,DoNa13}.
+Nonetheless, our theorem appears to be new and works for all dimensions.
+
+\begin{figure}[h]\centering
+    \includegraphics[width=.18\linewidth,height=.2\linewidth]{figs/orderGray1}%
+    \hspace{.8in}
+    \includegraphics[width=.18\linewidth,height=.2\linewidth]{figs/orderGray2}
+    \caption{On left, downward rain only (each shade of gray represents a different iteration), producing a grid.  On right, flipping the direction of rain.}
+    \label{fig:order}
+\end{figure}
+
+\paragraph{Cutting $\MM$ into extremum dominant pieces:} We define a simplicial complex
+endowed with a height to be \emph{minimum dominant} if there only exists a single minima.
+In such a complex, there exists a non-ascending path from any point to this unique minima.
+(Our real definition is more complicated, and involves simplicial complices
+that have many minima.) Analogously, we can define maximum dominant simplicial complices,
+and both are collectively called extremum dominant.
+
+We will cut $\MM$ into disjoint extremum dominant pieces, in linear time.
+The best way to think of our procedure is a meteorology analogy. Take an arbitrary maxima $x$,
+and imagine torrential rain at the maxima. The water flows down, wetting any point that has a non-ascending
+from $x$. We end up with two portions, the wet part of $\MM$ and the dry part. This is similar
+to \emph{watershed} algorithms used for image segmentation~\cite{RoMe00}. The wet part is obviously connected,
+while there may be numerous disconnected dry parts. The interface between the dry and wet parts
+are a set of contours, given the by ``water line". The wet part is clearly maximum dominant, since all
+wet points have a non-descending path to $x$. So we can simply cut along the interface
+contours to get the wet maximum dominant piece $\MM'$. By our contour surgery theorem,
+we are left with a set of disconnected dry parts, and we could recur this procedure on them.
+
+But here's the catch. Every time we cut $\MM$ along a contour, we potentially increase
+the complexity of $\MM$. Water flows in the interior of facets, and the interface will naturally
+cut some facets. Each cut introduces new vertices, and a poor choice of repeated raining leads
+to a large increase in complexity. Consider the left of \Fig{order}. Each raining produces a single
+wet and dry piece, and each cut introduces many new vertices. If we wanted to partition this terrain into 
+maximum dominant simplicial complices, the final complexity would be forbiddingly large.
+
+A simple trick saves the day. Unlike reality, we can choose rain to flow solely downwards or solely upwards.
+Apply the procedure above to get a single wet maximum dominant $\MM'$ and a set of dry pieces.
+Observe that a single dry piece $\NN$ is boundary critical with the newly introduced boundary $\phi$ (the wet-dry interface) behaving as a minima.
+So we can rain upwards from this minima, and get a \emph{minimum dominant} portion $\NN'$.
+This ensures that the new interface (after applying the procedure on $\NN$) does not
+cut any facet previously cut by $\phi$.
+For each of the new dry pieces, the newly introduced boundary is now a maximum. So we rain downwards from there.
+More formally, we alternate between raining upwards and downwards as we go down the recursion tree.
+We can prove that an original facet of $\MM$ is cut at most once, so the final complexity can be bounded.
+In \Fig{order}, this procedure would yield two pieces, one maximum dominant, and one minimum dominant.
+
+Using the contour surgery theorem previously discussed, we can build the contour tree of $\MM$
+from the contour trees of the various pieces created. All in all, we prove the following theorem.
+
+\begin{theorem} \label{thm:int-rain} There is an $O(N)$ time procedure that cuts $\MM$ into extremum dominant simplicial complices
+$\MM_1, \MM_2, \ldots$. Furthermore, given the set of contour trees $\{\cC(\MM_i)\}$, $\cC(\MM)$ can be constructed in $O(N)$ time.
+\end{theorem}
+
+\paragraph{Extremum dominance simplifies contour trees:} We will focus on minimum dominant simplicial complices $\MM$. By \Thm{int-rain}, it suffices
+to design an algorithm for contour trees on such inputs. 
+For the $d=2$ case, it helps to visualize such an input as a terrain with no minima,
+except at a unique boundary face (think of a large boundary triangle that is the boundary).
+All the valleys in such a terrain are necessarily joins, and there can be no splits. Look
+at \Fig{sorting}. The portion on the left is minimum dominant in exactly this fashion.
+Water poured at any critical point must flow to the boundary, and the entire wet portion are connected.
+More formally, $\MM^-_v$ is connected for all $v$, so there are no splits.
+
+We can prove that the split tree is just a path, and the contour tree is exactly
+the join tree. The formal argument is a little involved, and we employ the merging procedure
+of~\cite{csa-cctad-00} to get a proof. We demonstrate that the merging procedure will
+actually just output the join tree, so we do not need to actually compute the split tree.
+(The real definition of minimum dominant is a little more complicated,
+so the contour tree is more than just the join tree. But computationally, it suffices
+to construct the join tree.)
+
+We stress the importance of this step for our approach. Given the algorithm of~\cite{csa-cctad-00},
+one may think that that it suffices to design faster algorithms for join trees. But this
+cannot give the sort of optimality we hope. Again, consider \Fig{sorting}. Any algorithm
+to construct the true join tree must construct the path of splits, which implies
+sorting all of them. It is absolutely necessary to cut $\MM$ into pieces where the cost
+of building the join trees can be related to the cost of building $\cC(\MM)$.
+
+\subsection{Join trees from painted mountaintops} \label{sec:join-paint}
+
+Arguably, everything up to this point is a preamble for the main result: a faster
+algorithm for join trees. Our algorithm does not require the input to be extremum
+dominant. This is only required to relate the join trees to the contour tree of initial input $\MM$.
+For convenience, we use $\NN$ to denote the input here. Note that as defined in \Def{int-criticalJoin},
+the join tree is defined purely combinatorially in terms of the 1-skeleton (the underlying graph) of $\NN$.
+
+The join tree $\jc(\NN)$ is a rooted tree with the dominant minimum at the root, and we direct edges downwards (towards the root). So it makes sense to talk
+of comparable vs incomparable vertices.
+We arrive at the primary challenge: how to sort only those critical points that are comparable, without
+constructing the join tree? The join tree algorithm of~\cite{csa-cctad-00} is a typical event-based
+computational geometry algorithm. We have to step away from this viewpoint to avoid the global sort.
+
+The key idea is \emph{paint spilling}. Start with each maxima having a large can of paint, with distinct
+colors for each maxima. In arbitrary order, we spill paint from each maxima, wait till it flows down,
+then spill from the next, etc. Paint is viscous, and only flows down edges. \emph{It does not
+paint the interior of facets.} Alternately, this process is restricted to the 1-skeleton of $\NN$.
+Furthermore, our paints do not mix, so each edge receives a unique color, which is decided by the first
+paint to reach it.
+
+\begin{definition} \label{def:paint} Let the 1-skeleton of $\NN$ have edge set $E$ and maxima $X$.
+A  \emph{painting} of $\NN$ is a map $\chi:X \cup E \mapsto [|X|]$ with the following property. 
+ Consider an edge $e$. There exists a descending path from some maximum $x$ to $e$
+	consisting of edges in $E$, such that all edges along this path have the same color as $x$.
+
+An \emph{initial} painting has the additional property that the restriction $\chi:X \mapsto [|X|]$ is a bijection.
+\end{definition}
+%
+%Paintings are not unique. It is straightforward to construct an initial painting of $\MM$ in $O(|\MM|)$ time, 
+%using a descending BFS from each maximum that does not explore previously colored edges. 
+%As the algorithm proceeds, the bijection property will not hold, though we always maintain a valid painting.
+
+Note that a painting only colors edges, not vertices. Our definition also does not require the timing aspect
+of iterating over colors, though that is one way of painting $\NN$. We begin with an initial painting, since all maxima
+colors are distinct. A few comments on paint vs water. The interface between two regions of different color
+is \emph{not} a contour, and is actually difficult to characterize. So we cannot apply the divide-and-conquer
+approach of contour surgery. On the other hand, painting does not cut $\NN$, so there is no increase in complexity.
+Clearly, an initial painting can be constructed in $O(N)$ time. This is the tradeoff between water and paint.
+Water allows for an easy divide-and-conquer, at the cost of more complexity in the input. For an extremum
+dominant input, using water to divide the input $\NN$ raise the complexity too much.
+
+Our algorithm incrementally builds $\jc(\MM)$ from the leaves (maxima) to the root (dominant minimum). 
+We say that vertex $v$ is \emph{touched} by color $c$, if there is a $c$-colored edge with lower endpoint $v$.
+Let us focus on an initial painting, where the colors have 1-1 correspondence with the maxima. Refer to \Fig{colors}.
+Consider two sibling leaves $\ell_1, \ell_2$ and their common parent $v$. The leaves are maxima,
+and $v$ is a join that ``merges" $\ell_1, \ell_2$. 
+In that case, there are ``mounds" corresponding to each $\ell_1$ and $\ell_2$ that merge
+at a valley $v$. Suppose this was the entire input, and $\ell_1$ was colored blue and $\ell_2$ was colored red. 
+(So ignore the yellow mound in \Fig{colors}.)
+Both the mounds are colored completely blue or red, while $v$ is touched by both colors.
+So this indicates that $v$ joins the blue maxima and red maxima in $\jc(\MM)$.
+
+This is precisely how we hope to exploit the information in the painting. We prove later that
+when some join $v$ has all incident edges with exactly two colors, the corresponding maxima (of those colors) are exactly
+the children of $v$ in $\jc(\MM)$. To proceed further, we ``merge" the colors red and blue into a new color, purple. In other words,
+we replace all red and blue edges by purple edges. This indicates that the red and blue maxima have been
+handled. Imagine flattening the red and blue mounds until reaching $v$, so that the former join $v$
+is now a new maxima. Suppose purple paint was poured from $v$. In terms of $\jc(\MM)$, this
+is equivalent to removing leaves $\ell_1$ and $\ell_2$, making $v$ a new leaf.
+Alternately, $\jc(\MM)$ has been constructed up to $v$, and it remains to determine $v$'s parent.
+The merging of the colors is not explicitly as that would be too expensive; we maintain a union-find
+data structure for that.
+
+Of course, things are more complicated when there are other mounds. There may be a yellow mound,
+corresponding to $\ell_3$ that joins with the blue mound higher up (at vertex $u$). In $\jc(\MM)$, $\ell_2$ and $\ell_3$
+are sibling leaves, and $\ell_1$ is a sibling of some ancestor of these leaves. So we cannot
+merge red and blue, until yellow and blue merge. Naturally, we use priority queues to handle this issue.
+We know that $u$ must also be touched by blue. So all critical vertices touched by blue
+are put into a priority queue keyed by height, and vertices are handled in that order.
+
+\begin{figure}[h]\centering
+    \includegraphics[width=.27\linewidth]{figs/mounds}%
+    \hspace{.35in}
+    \includegraphics[width=.27\linewidth]{figs/mounds2}%
+    \hspace{.4in}
+    \includegraphics[width=.22\linewidth]{figs/mounds3}
+    \caption{On the left, three initial colors.  In the middle, yellow and blue merge into green.  On the right, green and red merge into purple.}
+    \label{fig:colors}
+\end{figure}
+
+What happens when finally blue and red join at $v$? We merge the two colors, but now have
+blue and red queues of critical vertices. We also need to merge the priority queues
+to get a consistent painting. And that decides the use of \emph{binomial heaps}~\cite{Vu78} to store the
+critical points of one color. Merges can be done in logarithmic time, the same time
+required for deletes. We stress that the feasibility of the entire approach hinges on the use of binomial heaps.
+
+In this discussion, we ignored an annoying problem. Vertices
+may actually be touched by numerous colors, not just one or two as assumed above. A simple solution
+would be to insert vertices into heaps corresponding to all colors touching it. But there could
+be super-constant numbers of copies of a vertex, and handling all these copies would lead
+to extra overhead. We show that it suffices to simply put each vertex $v$ into at most two heaps, one for each
+``side" of a possible join. We are guaranteed that when $v$ needs to be processed, all edges have
+at most $2$ colors, because of all the color merges that previously occured.
+
+\paragraph{The running time analysis:} Relating the running time to a path decomposition is the most technical part of the paper. 
+All the non-heap operations can be easily bounded as $O(t\alpha(t) + N)$ ($t\alpha(t)$ from the union-find data structure
+for colors). It is not hard to argue that at all times, any heap always contains a subset of a leaf
+to root path. Unfortunately, this subset is \emph{not} contiguous, as \Fig{colors} shows. \Sesh{explain}
+Nonetheless, we can get a non-trivial (but non-optimal) bound. Let $d_v$ denote the distance to the root for vertex $v$ in the join tree.
+The total cost (of the heap operations) is at most $\sum_v \log d_v$. This immediately proves a bound of $O(t\log D)$, where
+$D$ is the maximum distance to the root, an improvement over previous work. But this bound is non-optimal.
+For a balanced binary tree, this bound is $O(t\log\log t)$, whereas the cost of any path decomposition is $O(t)$.
+
+The cost of heap operations depends on the sizes of the heaps, which keeps changing because
+of the repeated merging. This is a major headache for the analysis. There are situations where
+the initial heap sizes are small, but they eventually merge to create larger heaps. 
+We employ a variant
+of \emph{heavy path decompositions}, first used for Sleator and Tarjan for analyzing link/cut trees~\cite{st-dsdt-83}.
+The final analysis is basically charges expensive heap operations to long paths in the decomposition.
+
+\subsection{The lower bound}
+
+Consider a contour tree $T$ and the path decomposition $P(T)$ used to bound the running time. 
+Denoting $\cost(P(T)) = \sum_{p \in P(T)} |p|\log |p|$, we construct a set of $\prod_{p \in P(T)} |p|!$
+functions on a fixed domain such that each function has a distinct (labeled) contour tree. By a simple
+entropy argument, any height decision tree that correctly computes the contour tree on all instances
+requires average case $\Omega(\cost(P(T)))$ time. We prove that our algorithm makes $\Theta(\cost(P(T)))$
+height comparisons on all these instances. In general, any algorithm requires $\Omega(C)$ time
+to solve the set of instances that our algorithm solves with $C$ comparisons.
+
+We have a fairly simple construction that works for terrains. In $P(T)$, consider the path
+$p$ that involves the root. The base of the construction is a conical ``tent", and there
+will be $|p|$ triangular faces that will each have a saddle. The heights of these saddles
+can be varied arbitrarily, and that will give $|p|!$ different choices. Each of these saddles
+will be connected to a recursive construction involving other paths in $P(T)$. Effectively,
+one can think of tiny tents that are sticking out of each face of the main tent. The contour
+trees of these tiny tents attach to a main branch of length $|p|$. Working out the details,
+we get $\prod_{p \in P(T)} |p|!$ terrains each with a distinct contour tree. It is possible to argue
+that our algorithm requires $\Theta(\cost(P(T)))$ on all these instances.
 
@@
 \end{proof}
-
-\begin{figure}[h]\centering
-    \includegraphics[width=.18\linewidth,height=.35\linewidth]{figs/orderGray1}%
-    \hspace{.8in}
-    \includegraphics[width=.18\linewidth,height=.35\linewidth]{figs/orderGray2}
-    \caption{On left, downward paint spilling only (each shade of gray represents a different color), producing a grid.  On right, flipping the direction of paint spilling.}
-    \label{fig:order}
-\end{figure}
 
@@
 We give an informal description of the algorithm, which might motivate the data structures.
-
-Our algorithm incrementally builds $\jc(\MM)$ from the leaves (maxima) to the root (dominant minimum). 
-Let us focus on an initial painting, where the colors have 1-1 correspondence with the maxima.
-Consider two sibling leaves $\ell_1, \ell_2$ and their common parent $v$. The leaves are maxima,
-and $v$ is a join that ``merges" $\ell_1, \ell_2$. A nice mental picture is given when $\MM$ is 2-dimensional.
-In that case, there are ``mounds" corresponding to each $\ell_1$ and $\ell_2$ that merge
-at a valley $v$. Suppose this was the entire input, and $\ell_1$ was colored blue and $\ell_2$ was colored red. 
-Both the mounds are colored completely blue or red, while $v$ is touched by both colors.
-Moreover, each up-star of $v$ (one connecting to each mound) is fully colored. So this
-indicates that $v$ joins the blue maxima and red maxima in $\jc(\MM)$.
-
-This is precisely how we hope to exploit the information in the painting. We prove later that
-when some join $v$ has both up-stars fully colored, the corresponding maxima (of those colors) are exactly
-the children of $v$ in $\jc(\MM)$. To proceed further, we ``merge" the colors red and blue into a new color, purple. In other words,
-we replace all red and blue edges by purple edges. This indicates that the red and blue maxima have been
-handled. Imagine flattening the red and blue mounds until reaching $v$, so that the former join $v$
-is now a new maxima. Suppose purple paint was poured from $v$. In terms of $\jc(\MM)$, this
-is equivalent to removing leaves $\ell_1$ and $\ell_2$, making $v$ a new leaf.
-Alternately, $\jc(\MM)$ has been constructed up to $v$, and it remains to determine $v$'s parent.
-
-Of course, things are more complicated when there are other mounds. There may be a yellow mound,
-corresponding to $\ell_3$ that joins with the blue mound higher up (at vertex $u$). In $\jc(\MM)$, $\ell_2$ and $\ell_3$
-are sibling leaves, and $\ell_1$ is a sibling of some ancestor of these leaves. So we cannot
-merge red and blue, until yellow and blue merge. Naturally, we use priority queues to handle this issue.
-We know that $u$ must also be (fully) touched by blue. So all critical vertices touched by blue
-are put into a priority queue keyed by height, and vertices are handled in that order.
-
-\begin{figure}[h]\centering
-    \includegraphics[width=.27\linewidth]{figs/mounds}%
-    \hspace{.35in}
-    \includegraphics[width=.27\linewidth]{figs/mounds2}%
-    \hspace{.4in}
-    \includegraphics[width=.22\linewidth]{figs/mounds3}
-    \caption{On the left, three initial colors.  In the middle, yellow and blue merge into green.  On the right, green and red merge into purple.}
-    \label{fig:colors}
-\end{figure}
-
-What happens when finally blue and red join at $v$? We merge the two colors, but now have
-blue and red queues of critical vertices. We also need to merge the priority queues
-to get a consistent painting. And that decides the use of \emph{binomial heaps} to store the
-critical points of one color. Merges can be done in logarithmic time, the same time
-required for deletes.
-
-In this discussion, we have ignored a somewhat annoying problem. Up-stars of critical vertices
-may actually be touched by numerous colors, not just one as we assumed above. A simple solution
-would be to insert vertices into heaps corresponding to all colors touching it. But there could
-be super-constant numbers of copies of a vertex, and handling all these copies would lead
-to extra overhead. We show that it suffices to simply put each vertex $v$ into at most two heaps,
-one for each up-star. We are guaranteed that when $v$ needs to be processed, its up-stars
-automatically have one color, because of all the color merges that previously occured.
 
@@
 
-\paragraph{Output Sensitive Analysis.}
-In geometric settings often the goal is to compute a (potentially complex) geometric structure determined 
-by the input, and so it is natural to also consider how the form of this output structure affects the running time.
-One example is the standard sweep line algorithm of Bentley and Ottmann \cite{bo-arcgi-79} for line segment intersection, 
-which runs $O((n+k)\log n)$ time where k is the number of intersections, which in the worst case is quadratic.
-Another example is Chan's algorithm for computing the convex hull \cite{c-ooscha-96} which runs in $O(n\log h)$ 
-where $h$ is the number of vertices on the hull, which in the worst case is $n$.  While such algorithms represent 
-significant improvements over their worst case counterparts, their running times are only measured with respect 
-to the size of the output and not any structure it may have.
-
-\paragraph{Instance Optimality.}  
-Here we are concerned with a more refined analysis than the typical output sensitive analysis.
-Specifically, we investigate an algorithm which is instance optimal as defined by Afshani \etal \cite{abc-ioga-09}.  
-In order to prove such a result, one considers a family of allowable partitions of the input, 
-each of which is defined in terms of the structure of the output.  Each such partition then has 
-an associated entropy type cost representing the number of possible instantiations of such a partition.
-Our work differs slightly from Afshani \etal \cite{abc-ioga-09} however, in that they show any algorithm 
-(for say the convex hull) requires the minimum such cost, where as in our case it is shown that 
-any algorithm requires the maximum over these costs.
-
 \subsubsection{Contour Trees}

### 2014-11-01 paint-contour-tree.tex

--- 2014-10-31 paint-contour-tree.tex
@@
 However, our actual bound in this case gives an even further improved $O(t\alpha(t)+N)$.
-%
 Consider the example of \Fig{sorting} extended to a larger binary tree both above and below.
@@
 
+\paragraph{Beyond worst-case analysis:} Our result provides understanding of the complexity
+of contour tree construction beyond the standard worst-case analysis. Our theorem and lower bound explain \emph{when}
+computing contour trees is expensive, and provide insight into the hardness of particular instances. It is intriguing
+that the cost (as defined earlier) roughly captures the hardness of an instance.
+In that sense, it is related to the concept of instance optimal algorithms, as defined by Afshani \etal \cite{abc-ioga-09}.
+Their notion of optimality is far more refined that what we prove, but it shares the flavor of understanding the spectrum
+of easy to hard instances. From a mathematical perspective, this sort of analysis forces one to look closely
+at the relationship between sorting and the contour tree problem.
+
+\ignore{
+classic output sensitive convex hull algorithms of Kirkpatrick and Seidel~\cite{KiSe86} and Chan~\cite{c-ooscha-96}.
+
+\paragraph{Output Sensitive Analysis.}
+In geometric settings often the goal is to compute a (potentially complex) geometric structure determined 
+by the input, and so it is natural to also consider how the form of this output structure affects the running time.
+One example is the standard sweep line algorithm of Bentley and Ottmann \cite{bo-arcgi-79} for line segment intersection, 
+which runs $O((n+k)\log n)$ time where k is the number of intersections, which in the worst case is quadratic.
+Another example is Chan's algorithm for computing the convex hull \cite{c-ooscha-96} which runs in $O(n\log h)$ 
+where $h$ is the number of vertices on the hull, which in the worst case is $n$.  While such algorithms represent 
+significant improvements over their worst case counterparts, their running times are only measured with respect 
+to the size of the output and not any structure it may have.
+
+\paragraph{Instance Optimality.}  
+Here we are concerned with a more refined analysis than the typical output sensitive analysis.
+Specifically, we investigate an algorithm which is instance optimal as defined by Afshani \etal \cite{abc-ioga-09}.  
+In order to prove such a result, one considers a family of allowable partitions of the input, 
+each of which is defined in terms of the structure of the output.  Each such partition then has 
+an associated entropy type cost representing the number of possible instantiations of such a partition.
+Our work differs slightly from Afshani \etal \cite{abc-ioga-09} however, in that they show any algorithm 
+(for say the convex hull) requires the minimum such cost, where as in our case it is shown that 
+any algorithm requires the maximum over these costs.
+}
+
 \subsection{Previous Work}
@@
 \begin{figure}[h]\centering
-    \includegraphics[width=.7\linewidth,height=.3\linewidth]{figs/balanced4}
-    \vspace{.1in}
-
-    \includegraphics[width=.7\linewidth,height=.25\linewidth]{figs/trees}
+    \includegraphics[width=.35\linewidth,height=.15\linewidth]{figs/balanced3}
+    \hspace{.3in}
+    \includegraphics[width=.35\linewidth,height=.15\linewidth]{figs/trees}
     \caption{On top, a surface will a balanced contour tree, but whose join and split trees have long tails.  
@@
 }
+
+\section{A tour of the contour tree algorithm} \label{sec:approach}
+
+Our final algorithm is quite technical and has numerous moving parts. This section
+provides a high level view of the entire result, and highlights the main ideas
+and intuition. 
+It is helpful to keep the $d=2$ case in mind, so the input is just a triangulated
+terrain. In the interest of presentation, the definitions and theorem statements in this section will
+slightly differ from those in the main body. They may also differ from the original definitions proposed
+in earlier work.
+
+\paragraph{Do not globally sort:} Arguably, the starting point for this work is \Fig{relative}. We have two terrains
+with exactly the same contour tree, but different orderings of (heights of) the critical point.
+Turning it around, we cannot deduce the height ordering of critical points from the contour tree. Clearly,
+sorting the critical points is computationally reduntant for constructing the contour tree.
+In \Fig{sorting}, the contour tree basically consists of two balanced binary trees, one of the joins,
+another of the splits. Again, it is not necessary to know the relative ordering between the mounds
+on the left (or among the depressions in the right) to compute the contour tree. Yet some ordering
+information is necessary: in the left, the little valleys are higher than the big central valley,
+and this is reflected in the contour tree. More generally, leafs paths in the contour tree have points
+in sorted order, but incomparable points in the tree are unconstrained.
+How do we sort exactly what is required, without knowing the contour tree in advance?
+
+\subsection{Breaking $\MM$ into simpler pieces} \label{sec:break}
+
+Let us begin with the algorithm of Carr, Snoeyink, and Axen~\cite{csa-cctad-00}. The key insight is to build
+two different trees, called the join and split trees, and then merge them together into the contour tree.
+Consider sweeping down via the hyperplane $x_{d+1} = h$ and taking the \emph{superlevel} sets. These 
+are the connected components of $\MM$ with height greater than $h$. For a terrain, the superlevel
+sets are a collection of ``mounds". As we sweep downwards, these mounds keep joining each other,
+until finally, we end up with all of $\MM$. The join tree tracks exactly these events. 
+Formally, let $\MM^+_v$ denote the simplicial complex obtained by taking only those
+vertices higher than $v$.
+
+\begin{definition} 
+\label{def:int-criticalJoin}
+The critical join tree $\jc(\MM)$ is built on the set $V'$ of all critical points. 
+The directed edge $(u,v)$ is present when $u$ is the smallest valued vertex in $V'$ in a connected component of $\MM^+_v$
+and $v$ is connected to this component (in $\MM$). 
+\end{definition}
+
+(Abusing notation, we just call this the join tree.)
+Refer to \Fig{sorting} for the join tree of a terrain. Note that nothing happens at splits, but these
+are still put as vertices in the join tree. They simply form a long path. The split tree
+is obtained by simply inverting this procedure, sweeping upwards and tracking sublevel sets.
+
+A major insight of~\cite{csa-cctad-00} is an ingeniously simple linear time procedure to construct
+the contour tree from the join and split tree. So the bottleneck is computing these trees. And observe
+in \Fig{sorting} that the split vertices form a long path in the join tree (and vice versa). Constructing
+these trees forces a global sort of the splits, a redundant computation. Unfortunately (unlike \Fig{sorting}),
+the heights of joins and splits may be interleaved in a complex manner. The final merging of~\cite{csa-cctad-00}
+to get the contour tree requires having the split vertices in the join tree. Without
+this, it is not clear how to get a consistent view of both joins and splits, required for the contour tree.
+
+Our aim is break $\MM$ into smaller pieces, where this redundant computation can be saved.
+
+%
+%{\bf Punchline:} We give a linear time procedure that partitions $\MM$ into smaller pieces. We guarantee
+%that the contour tree of each individual piece is basically the join (or split) tree. This saves
+%some of the redundant computation. We then show to put it all together the get the full contour tree.
+%Now for more details.
+
+\paragraph{Contour surgery:} We first need a divide-and-conquer lemma. Any contour $\phi$ can be associated
+with an edge $e$ of the contour tree. Suppose we ``cut" $\MM$ along this contour. We prove that $\MM$
+is split into two disconnected pieces, such the contour trees of these pieces is obtained
+by simply cutting $e$ in $\reeb(\MM)$. Alternatively, the contour trees of these pieces can
+be glued together to get $\reeb(\MM)$. This is not particularly surprising, and is fairly easy to prove
+with the right definitions. The idea of loop surgery has been used to reduce Reeb graphs to contour trees~\cite{TiGySi09,DoNa13}.
+Nonetheless, our theorem appears to be new and works for all dimensions.
+
+\begin{figure}[h]\centering
+    \includegraphics[width=.18\linewidth]{figs/tent2}%
+    \hspace{.9in}
+    \includegraphics[width=.18\linewidth]{figs/tent1}
+    \caption{On left, downward rain spilling only (each shade of gray represents a different color), producing a grid.  On right, flipping the direction of rain spilling.}
+    \label{fig:order}
+\end{figure}
+
+\paragraph{Cutting $\MM$ into extremum dominant pieces:} We define a simplicial complex
+endowed with a height to be \emph{minimum dominant} if there only exists a single minima.
+In such a complex, there exists a non-ascending path from any point to this unique minima.
+(Our real definition is more complicated, and involves simplicial complices
+that have many minima.) Analogously, we can define maximum dominant simplicial complices,
+and both are collectively called extremum dominant.
+
+We will cut $\MM$ into disjoint extremum dominant pieces, in linear time.
+The best way to think of our procedure is a meteorology analogy. Take an arbitrary maxima $x$,
+and imagine torrential rain at the maxima. The water flows down, wetting any point that has a non-ascending
+from $x$. We end up with two portions, the wet part of $\MM$ and the dry part. This is similar
+to \emph{watershed} algorithms used for image segmentation~\cite{RoMe00}. The wet part is obviously connected,
+while there may be numerous disconnected dry parts. The interface between the dry and wet parts
+are a set of contours, given the by ``water line". The wet part is clearly maximum dominant, since all
+wet points have a non-descending path to $x$. So we can simply cut along the interface
+contours to get the wet maximum dominant piece $\MM'$. By our contour surgery theorem,
+we are left with a set of disconnected dry parts, and we could recur this procedure on them.
+
+But here's the catch. Every time we cut $\MM$ along a contour, we potentially increase
+the complexity of $\MM$. Water flows in the interior of facets, and the interface will naturally
+cut some facets. Each cut introduces new vertices, and a poor choice of repeated raining leads
+to a large increase in complexity. Consider the left of \Fig{order}. Each raining produces a single
+wet and dry piece, and each cut introduces many new vertices. If we wanted to partition this terrain into 
+maximum dominant simplicial complices, the final complexity would be forbiddingly large.
+
+A simple trick saves the day. Unlike reality, we can choose rain to flow solely downwards or solely upwards.
+Apply the procedure above to get a single wet maximum dominant $\MM'$ and a set of dry pieces.
+Observe that a single dry piece $\NN$ is boundary critical with the newly introduced boundary $\phi$ (the wet-dry interface) behaving as a minima.
+So we can rain upwards from this minima, and get a \emph{minimum dominant} portion $\NN'$.
+This ensures that the new interface (after applying the procedure on $\NN$) does not
+cut any facet previously cut by $\phi$.
+For each of the new dry pieces, the newly introduced boundary is now a maximum. So we rain downwards from there.
+More formally, we alternate between raining upwards and downwards as we go down the recursion tree.
+We can prove that an original facet of $\MM$ is cut at most once, so the final complexity can be bounded.
+In \Fig{order}, this procedure would yield two pieces, one maximum dominant, and one minimum dominant.
+
+Using the contour surgery theorem previously discussed, we can build the contour tree of $\MM$
+from the contour trees of the various pieces created. All in all, we prove the following theorem.
+
+\begin{theorem} \label{thm:int-rain} There is an $O(N)$ time procedure that cuts $\MM$ into extremum dominant simplicial complices
+$\MM_1, \MM_2, \ldots$. Furthermore, given the set of contour trees $\{\cC(\MM_i)\}$, $\cC(\MM)$ can be constructed in $O(N)$ time.
+\end{theorem}
+
+\paragraph{Extremum dominance simplifies contour trees:} We will focus on minimum dominant simplicial complices $\MM$. By \Thm{int-rain}, it suffices
+to design an algorithm for contour trees on such inputs. 
+For the $d=2$ case, it helps to visualize such an input as a terrain with no minima,
+except at a unique boundary face (think of a large boundary triangle that is the boundary).
+All the valleys in such a terrain are necessarily joins, and there can be no splits. Look
+at \Fig{sorting}. The portion on the left is minimum dominant in exactly this fashion.
+Water poured at any critical point must flow to the boundary, and the entire wet portion are connected.
+More formally, $\MM^-_v$ is connected for all $v$, so there are no splits.
+
+We can prove that the split tree is just a path, and the contour tree is exactly
+the join tree. The formal argument is a little involved, and we employ the merging procedure
+of~\cite{csa-cctad-00} to get a proof. We demonstrate that the merging procedure will
+actually just output the join tree, so we do not need to actually compute the split tree.
+(The real definition of minimum dominant is a little more complicated,
+so the contour tree is more than just the join tree. But computationally, it suffices
+to construct the join tree.)
+
+We stress the importance of this step for our approach. Given the algorithm of~\cite{csa-cctad-00},
+one may think that that it suffices to design faster algorithms for join trees. But this
+cannot give the sort of optimality we hope. Again, consider \Fig{sorting}. Any algorithm
+to construct the true join tree must construct the path of splits, which implies
+sorting all of them. It is absolutely necessary to cut $\MM$ into pieces where the cost
+of building the join trees can be related to the cost of building $\cC(\MM)$.
+
+\subsection{Join trees from painted mountaintops} \label{sec:join-paint}
+
+Arguably, everything up to this point is a preamble for the main result: a faster
+algorithm for join trees. Our algorithm does not require the input to be extremum
+dominant. This is only required to relate the join trees to the contour tree of initial input $\MM$.
+For convenience, we use $\NN$ to denote the input here. Note that as defined in \Def{int-criticalJoin},
+the join tree is defined purely combinatorially in terms of the 1-skeleton (the underlying graph) of $\NN$.
+
+The join tree $\jc(\NN)$ is a rooted tree with the dominant minimum at the root, and we direct edges downwards (towards the root). So it makes sense to talk
+of comparable vs incomparable vertices.
+We arrive at the primary challenge: how to sort only those critical points that are comparable, without
+constructing the join tree? The join tree algorithm of~\cite{csa-cctad-00} is a typical event-based
+computational geometry algorithm. We have to step away from this viewpoint to avoid the global sort.
+
+The key idea is \emph{paint spilling}. Start with each maxima having a large can of paint, with distinct
+colors for each maxima. In arbitrary order, we spill paint from each maxima, wait till it flows down,
+then spill from the next, etc. Paint is viscous, and only flows down edges. \emph{It does not
+paint the interior of facets.} Alternately, this process is restricted to the 1-skeleton of $\NN$.
+Furthermore, our paints do not mix, so each edge receives a unique color, which is decided by the first
+paint to reach it.
+
+\begin{definition} \label{def:paint} Let the 1-skeleton of $\NN$ have edge set $E$ and maxima $X$.
+A  \emph{painting} of $\NN$ is a map $\chi:X \cup E \mapsto [|X|]$ with the following property. 
+ Consider an edge $e$. There exists a descending path from some maximum $x$ to $e$
+	consisting of edges in $E$, such that all edges along this path have the same color as $x$.
+
+An \emph{initial} painting has the additional property that the restriction $\chi:X \mapsto [|X|]$ is a bijection.
+\end{definition}
+%
+%Paintings are not unique. It is straightforward to construct an initial painting of $\MM$ in $O(|\MM|)$ time, 
+%using a descending BFS from each maximum that does not explore previously colored edges. 
+%As the algorithm proceeds, the bijection property will not hold, though we always maintain a valid painting.
+
+Note that a painting only colors edges, not vertices. Our definition also does not require the timing aspect
+of iterating over colors, though that is one way of painting $\NN$. We begin with an initial painting, since all maxima
+colors are distinct. A few comments on paint vs water. The interface between two regions of different color
+is \emph{not} a contour, and is actually difficult to characterize. So we cannot apply the divide-and-conquer
+approach of contour surgery. On the other hand, painting does not cut $\NN$, so there is no increase in complexity.
+Clearly, an initial painting can be constructed in $O(N)$ time. This is the tradeoff between water and paint.
+Water allows for an easy divide-and-conquer, at the cost of more complexity in the input. For an extremum
+dominant input, using water to divide the input $\NN$ raise the complexity too much.
+
+Our algorithm incrementally builds $\jc(\MM)$ from the leaves (maxima) to the root (dominant minimum). 
+We say that vertex $v$ is \emph{touched} by color $c$, if there is a $c$-colored edge with lower endpoint $v$.
+Let us focus on an initial painting, where the colors have 1-1 correspondence with the maxima. Refer to \Fig{colors}.
+Consider two sibling leaves $\ell_1, \ell_2$ and their common parent $v$. The leaves are maxima,
+and $v$ is a join that ``merges" $\ell_1, \ell_2$. 
+In that case, there are ``mounds" corresponding to each $\ell_1$ and $\ell_2$ that merge
+at a valley $v$. Suppose this was the entire input, and $\ell_1$ was colored blue and $\ell_2$ was colored red. 
+(So ignore the yellow mound in \Fig{colors}.)
+Both the mounds are colored completely blue or red, while $v$ is touched by both colors.
+So this indicates that $v$ joins the blue maxima and red maxima in $\jc(\MM)$.
+
+This is precisely how we hope to exploit the information in the painting. We prove later that
+when some join $v$ has all incident edges with exactly two colors, the corresponding maxima (of those colors) are exactly
+the children of $v$ in $\jc(\MM)$. To proceed further, we ``merge" the colors red and blue into a new color, purple. In other words,
+we replace all red and blue edges by purple edges. This indicates that the red and blue maxima have been
+handled. Imagine flattening the red and blue mounds until reaching $v$, so that the former join $v$
+is now a new maxima. Suppose purple paint was poured from $v$. In terms of $\jc(\MM)$, this
+is equivalent to removing leaves $\ell_1$ and $\ell_2$, making $v$ a new leaf.
+Alternately, $\jc(\MM)$ has been constructed up to $v$, and it remains to determine $v$'s parent.
+The merging of the colors is not explicitly as that would be too expensive; we maintain a union-find
+data structure for that.
+
+Of course, things are more complicated when there are other mounds. There may be a yellow mound,
+corresponding to $\ell_3$ that joins with the blue mound higher up (at vertex $u$). In $\jc(\MM)$, $\ell_2$ and $\ell_3$
+are sibling leaves, and $\ell_1$ is a sibling of some ancestor of these leaves. So we cannot
+merge red and blue, until yellow and blue merge. Naturally, we use priority queues to handle this issue.
+We know that $u$ must also be touched by blue. So all critical vertices touched by blue
+are put into a priority queue keyed by height, and vertices are handled in that order.
+
+\begin{figure}[h]\centering
+    \includegraphics[width=.27\linewidth]{figs/mounds}%
+    \hspace{.35in}
+    \includegraphics[width=.27\linewidth]{figs/mounds2}%
+    \hspace{.4in}
+    \includegraphics[width=.22\linewidth]{figs/mounds3}
+    \caption{On the left, three initial colors.  In the middle, yellow and blue merge into green.  On the right, green and red merge into purple.}
+    \label{fig:colors}
+\end{figure}
+
+What happens when finally blue and red join at $v$? We merge the two colors, but now have
+blue and red queues of critical vertices. We also need to merge the priority queues
+to get a consistent painting. And that decides the use of \emph{binomial heaps}~\cite{Vu78} to store the
+critical points of one color. Merges can be done in logarithmic time, the same time
+required for deletes. We stress that the feasibility of the entire approach hinges on the use of binomial heaps.
+
+In this discussion, we ignored an annoying problem. Vertices
+may actually be touched by numerous colors, not just one or two as assumed above. A simple solution
+would be to insert vertices into heaps corresponding to all colors touching it. But there could
+be super-constant numbers of copies of a vertex, and handling all these copies would lead
+to extra overhead. We show that it suffices to simply put each vertex $v$ into at most two heaps, one for each
+``side" of a possible join. We are guaranteed that when $v$ needs to be processed, all edges have
+at most $2$ colors, because of all the color merges that previously occured.
+
+\paragraph{The running time analysis:} Relating the running time to a path decomposition is the most technical part of the paper. 
+All the non-heap operations can be easily bounded as $O(t\alpha(t) + N)$ ($t\alpha(t)$ from the union-find data structure
+for colors). It is not hard to argue that at all times, any heap always contains a subset of a leaf
+to root path. Unfortunately, this subset is \emph{not} contiguous, as \Fig{colors} shows. \Sesh{explain}
+Nonetheless, we can get a non-trivial (but non-optimal) bound. Let $d_v$ denote the distance to the root for vertex $v$ in the join tree.
+The total cost (of the heap operations) is at most $\sum_v \log d_v$. This immediately proves a bound of $O(t\log D)$, where
+$D$ is the maximum distance to the root, an improvement over previous work. But this bound is non-optimal.
+For a balanced binary tree, this bound is $O(t\log\log t)$, whereas the cost of any path decomposition is $O(t)$.
+
+The cost of heap operations depends on the sizes of the heaps, which keeps changing because
+of the repeated merging. This is a major headache for the analysis. There are situations where
+the initial heap sizes are small, but they eventually merge to create larger heaps. 
+We employ a variant
+of \emph{heavy path decompositions}, first used for Sleator and Tarjan for analyzing link/cut trees~\cite{st-dsdt-83}.
+The final analysis is basically charges expensive heap operations to long paths in the decomposition.
+
+\subsection{The lower bound}
+
+Consider a contour tree $T$ and the path decomposition $P(T)$ used to bound the running time. 
+Denoting $\cost(P(T)) = \sum_{p \in P(T)} |p|\log |p|$, we construct a set of $\prod_{p \in P(T)} |p|!$
+functions on a fixed domain such that each function has a distinct (labeled) contour tree. By a simple
+entropy argument, any height decision tree that correctly computes the contour tree on all instances
+requires average case $\Omega(\cost(P(T)))$ time. We prove that our algorithm makes $\Theta(\cost(P(T)))$
+height comparisons on all these instances. In general, any algorithm requires $\Omega(C)$ time
+to solve the set of instances that our algorithm solves with $C$ comparisons.
+
+We have a fairly simple construction that works for terrains. In $P(T)$, consider the path
+$p$ that involves the root. The base of the construction is a conical ``tent", and there
+will be $|p|$ triangular faces that will each have a saddle. The heights of these saddles
+can be varied arbitrarily, and that will give $|p|!$ different choices. Each of these saddles
+will be connected to a recursive construction involving other paths in $P(T)$. Effectively,
+one can think of tiny tents that are sticking out of each face of the main tent. The contour
+trees of these tiny tents attach to a main branch of length $|p|$. Working out the details,
+we get $\prod_{p \in P(T)} |p|!$ terrains each with a distinct contour tree. It is possible to argue
+that our algorithm requires $\Theta(\cost(P(T)))$ on all these instances.
 
@@
 \end{proof}
-
-\begin{figure}[h]\centering
-    \includegraphics[width=.25\linewidth]{figs/tent2}%
-    \hspace{.9in}
-    \includegraphics[width=.25\linewidth]{figs/tent1}
-    \caption{On left, downward paint spilling only (each shade of gray represents a different color), producing a grid.  On right, flipping the direction of paint spilling.}
-    \label{fig:order}
-\end{figure}
 
@@
 
-\subsection{The overall approach and the data structures} \label{sec:struct}
-
-We give an informal description of the algorithm, which might motivate the data structures.
-
-Our algorithm incrementally builds $\jc(\MM)$ from the leaves (maxima) to the root (dominant minimum). 
-Let us focus on an initial painting, where the colors have 1-1 correspondence with the maxima.
-Consider two sibling leaves $\ell_1, \ell_2$ and their common parent $v$. The leaves are maxima,
-and $v$ is a join that ``merges" $\ell_1, \ell_2$. A nice mental picture is given when $\MM$ is 2-dimensional.
-In that case, there are ``mounds" corresponding to each $\ell_1$ and $\ell_2$ that merge
-at a valley $v$. Suppose this was the entire input, and $\ell_1$ was colored blue and $\ell_2$ was colored red. 
-Both the mounds are colored completely blue or red, while $v$ is touched by both colors.
-Moreover, each up-star of $v$ (one connecting to each mound) is fully colored. So this
-indicates that $v$ joins the blue maxima and red maxima in $\jc(\MM)$.
-
-This is precisely how we hope to exploit the information in the painting. We prove later that
-when some join $v$ has both up-stars fully colored, the corresponding maxima (of those colors) are exactly
-the children of $v$ in $\jc(\MM)$. To proceed further, we ``merge" the colors red and blue into a new color, purple. In other words,
-we replace all red and blue edges by purple edges. This indicates that the red and blue maxima have been
-handled. Imagine flattening the red and blue mounds until reaching $v$, so that the former join $v$
-is now a new maxima. Suppose purple paint was poured from $v$. In terms of $\jc(\MM)$, this
-is equivalent to removing leaves $\ell_1$ and $\ell_2$, making $v$ a new leaf.
-Alternately, $\jc(\MM)$ has been constructed up to $v$, and it remains to determine $v$'s parent.
-
-Of course, things are more complicated when there are other mounds. There may be a yellow mound,
-corresponding to $\ell_3$ that joins with the blue mound higher up (at vertex $u$). In $\jc(\MM)$, $\ell_2$ and $\ell_3$
-are sibling leaves, and $\ell_1$ is a sibling of some ancestor of these leaves. So we cannot
-merge red and blue, until yellow and blue merge. Naturally, we use priority queues to handle this issue.
-We know that $u$ must also be (fully) touched by blue. So all critical vertices touched by blue
-are put into a priority queue keyed by height, and vertices are handled in that order.
-
-\begin{figure}[h]\centering
-    \includegraphics[width=.27\linewidth]{figs/mounds}%
-    \hspace{.35in}
-    \includegraphics[width=.27\linewidth]{figs/mounds2}%
-    \hspace{.4in}
-    \includegraphics[width=.22\linewidth]{figs/mounds3}
-    \caption{On the left, three initial colors.  In the middle, yellow and blue merge into green.  On the right, green and red merge into purple.}
-    \label{fig:colors}
-\end{figure}
-
-What happens when finally blue and red join at $v$? We merge the two colors, but now have
-blue and red queues of critical vertices. We also need to merge the priority queues
-to get a consistent painting. And that decides the use of \emph{binomial heaps} to store the
-critical points of one color. Merges can be done in logarithmic time, the same time
-required for deletes.
-
-In this discussion, we have ignored a somewhat annoying problem. Up-stars of critical vertices
-may actually be touched by numerous colors, not just one as we assumed above. A simple solution
-would be to insert vertices into heaps corresponding to all colors touching it. But there could
-be super-constant numbers of copies of a vertex, and handling all these copies would lead
-to extra overhead. We show that it suffices to simply put each vertex $v$ into at most two heaps,
-one for each up-star. We are guaranteed that when $v$ needs to be processed, its up-stars
-automatically have one color, because of all the color merges that previously occured.
-
-With this preamble, we are ready to discuss our data structures. We discuss all the initializations
-in the next section.
+\subsection{The data structures} \label{sec:struct}
+
+We discuss all the initializations in the next section.
 

### 2014-11-01 paint-contour-tree.tex
section 8 changes

--- 2014-10-31 paint-contour-tree.tex
@@
 \begin{figure}[h]\centering
-    \includegraphics[width=.15\linewidth]{figs/diamond}\hspace{2cm}
-    \includegraphics[width=.125\linewidth]{figs/topView}\hspace{2cm}
-    \includegraphics[width=.095\linewidth]{figs/sideView}
-    \caption{From left to right: the cone portion of a diamond; the top down view and then the side view of a cone face with a platform added.}
+    \includegraphics[width=.18\linewidth]{figs/diamond}\hspace{2cm}
+    \includegraphics[width=.3\linewidth]{figs/cones}
+%    \includegraphics[width=.125\linewidth]{figs/topView}\hspace{2cm}
+%    \includegraphics[width=.095\linewidth]{figs/sideView}
+    \caption{Left: angled view of a diamond / Right: a parent and child cone put together}
     \label{fig:diamond}
@@
  First, we describe the basic building block of these terrains, which roughly speaking corresponds to a fixed path $p\in P$.  
- Specifically, we say a \emph{diamond} of size $m$ is the following terrain.  To start, construct a cycle 
- of length $m$, where the height of the $i$th vertex in the cycle is say $i/2m$.  Next, create an apex vertex at height 
- $\infty$ (or any appropriately large finite height), and add an edge between this vertex and each vertex in the cycle.  At this point 
+ Specifically, we say a \emph{diamond} of size $m$ is the following terrain.  To start, construct a slightly tilted cycle 
+ of length $m$.  Specifically have two antipodal points, at heights $1$ and $0$, which we will call the anchor and trap of the diamond, respectively.
+ The remaining $m-2$ are evenly spread between on the cycle and heights decrease monotonically when going from the anchor to the trap.
+ Next, create an apex vertex at some appropriately large height, and add an edge between it and each vertex in the cycle.  At this point 
  we have built an upside down cone with $m$ triangular faces (see \Fig{diamond}). 
- To make the terrain fully triangulated, we add a triangular outer face around the cycle. 
- Specifically, add a triangle with vertices at heights $0$, $1/20m$, and $1/10m$, such that in the projection to the plane 
- this triangle contains the $m$ length cycle, and there is an arbitrary triangulation of the space between this triangle 
- and the cycle.  For consistency, the inner cycle of a diamond will always have at length at least $3$, i.e. even if $m\leq 3$.
-
- Now above each face of the cone of the diamond, we will create a triangular ``platform'', which will act as the location to glue on the bounding triangle of other (shrunken) diamonds, 
- and whose highest point of connection to the face below will be a saddle vertex.  Specifically the heights of the vertices of the triangle will be $h$, $h-\eps$, and $h-2\eps$, where 
- $h$ is some unspecified height value.  This near horizontal platform is then connected to its respective face with (nearly) vertical edges down to the face. See \Fig{diamond}.
-
- At this point we can construct the entire terrain, by gluing together diamonds.  
- Specifically, for each path $p_i\in P$ we create a diamond of size $|p_i|-1$ (we subtract one because the tip of the diamond already corresponds to the leaf of the path).
- Now we glue together diamonds of different paths in the same way the paths are connected in the shrub $\pathTree$.  Specially, for two paths $p,q\in P$ where 
- $p$ is the parent of $q$ in $\pathTree$, we glue the outer triangular cap of the diamond for $q$ onto the triangular ``platform'' of some face of the diamond for $p$.
+ %To make the terrain fully triangulated, we add a triangular outer face around the cycle. 
+ %Specifically, add a triangle with vertices at heights $0$, $1/20m$, and $1/10m$, such that in the projection to the plane 
+ %this triangle contains the $m$ length cycle, and there is an arbitrary triangulation of the space between this triangle 
+ %and the cycle.  For consistency, the inner cycle of a diamond will always have at length at least $3$, i.e. even if $m\leq 3$.
+ %
+ %Now above each face of the cone of the diamond, we will create a triangular ``platform'', which will act as the location to glue on the bounding triangle of other (shrunken) diamonds, 
+ %and whose highest point of connection to the face below will be a saddle vertex.  Specifically the heights of the vertices of the triangle will be $h$, $h-\eps$, and $h-2\eps$, where 
+ %$h$ is some unspecified height value.  This near horizontal platform is then connected to its respective face with (nearly) vertical edges down to the face. See \Fig{diamond}.
+
+ Now we describe how to attach two diamonds.  In this process, we glue the base of a scaled down ``child'' diamond on to a triangular cone face of the large  ``parent'' diamond (see \Fig{diamond}).
+ Specifically, the anchor of the child diamond is attached directly to a face of the parent diamond at some unspecified height $h$. The remained of the base of the child cone is then 
+ extended down until it hits the face of the parent.
+
+ At this point we can construct the entire terrain, by repeatedly gluing together diamonds.  
+ Specifically, for each path $p_i\in P$ we create a diamond of size $|p_i|+1$.
+ The two faces adjacent to the anchor are always empty, and the remaining faces are for gluing on other diamonds.
+ (Note that diamonds have size $|p_i|+1$, since path $|p_i|-1$ faces represent the joins of $p_i$, the apex represents the leaf, 
+ and we need two empty faces next to the anchor.)
+ Now we glue together diamonds of different paths in the same way the paths are connected in the shrub $\pathTree$.  
+ Specially, for two paths $p,q\in P$ where 
+ $p$ is the parent of $q$ in $\pathTree$, we glue $q$ onto a face of the diamond for $p$ as described above.
  (Naturally for this construction to work, diamonds for a given path will be scaled down relative to the size of the diamond of their parent).
 
- Observe now that the only saddle points in this construction are the highest connecting point on a face of an edge from the platform above.  Moreover, the only maxima are the apexes of the 
- diamonds, and there is exactly one minima corresponding to the outer cap of the diamond representing the root of $\pathTree$.  Therefore, the saddles on a given 
- diamond will appear contiguously on a root to leaf path in the join tree of the terrain, where the leaf corresponds to the maxima of the diamond 
+ Observe now that the only saddle points in this construction are the anchor points.  Moreover, the only maxima are the apexes of the 
+ diamonds.  We create a global boundary minimum by setting the vertices at the base of the diamond representing the root of $\pathTree$ all to the same height (and there are no other minima).  
+ Therefore, the saddles on a given diamond will appear contiguously on a root to leaf path in the join tree of the terrain, where the leaf corresponds to the maxima of the diamond 
  (since all these saddles have a direct line of sight to this apex).  In particular, 
@@
 First observe that by construction, regardless of the choice of $f\in F$, 
-there is some path decomposition $\overline{P}(T)$ (of the tree $f$ defines) such that 
-each path $p$ consists exactly of the set of critical points from a given diamond. 
+there is some path decomposition $\overline{P}(T)$ of the join tree such that 
+each path $p$ consists exactly of the set of critical points (except for its anchor) from a given diamond (i.e. $\overline{P}(T)$ is the same for any $f\in F$). 
 We now show that regardless of how paint spilling is done, this will be the maximum 
@@
  So consider any order of paint spilling over the surface defined by $f$.
- Consider the set of joins around any diamond.  All these joins have an edge which connects 
- directly to the maxima of that diamond.  In particular, all these joins have 
- an upstar which is entirely colored by the color of this diamond's maxima.  
- Therefore, in the leaf assignment of the join tree, determined by this painting (see \Def{leafAssign})
- all these join get assigned a common color.  
- In particular, all vertices in each path in $\overline{P}$ get assigned a common color.
+ Consider any join on a diamond.  This join is the anchor point of some connecting child diamond.  
+ The claim is that the two colors of this join (according to leaf assignment, \Def{leafAssign}, used in \Lem{runTimeUpper})
+ are the colors of the apexes of the child and parent diamond.  
+ This easily follows since this join has two up-stars, each of which has exactly one edge.  
+ Specifically, one of these edges connects directly to the apex of the parent diamond and so the join must get the color of this maxima.
+ The other edge connects directly to the apex of the child diamond (this is why we left the faces adjacent to an anchor empty) and 
+ so the second color is that of the maxima of the child diamond.
+
+ So let $c$ be the color of some maxima, $m$.  This claim implies that in the join tree, the vertices for 
+ which $c$ appears are exactly the vertices in path in $\overline{P}(T)$ corresponding to the diamond of $m$
+ and the vertex to which the root of this path connects to.
+ This in turn implies that $\overline{P}(T)$ is a maximum path decomposition and hence $\cost(f) = \sum_{p\in \overline{P}} |p|\log|p|$
+ To see this, consider a vertex ..... (remark that ties can be broken consistently, and so the there is one maximum decomp)
+
+%  All these joins have an edge which connects 
+%  directly to the maxima of that diamond.  In particular, all these joins have 
+%  an upstar which is entirely colored by the color of this diamond's maxima.  
+%  Therefore, in the leaf assignment of the join tree, determined by this painting (see \Def{leafAssign})
+%  all these join get assigned a common color.  
+%  In particular, all vertices in each path in $\overline{P}$ get assigned a common color.
+
+%  Now consider any vertex $v$ in $T$, and let $p$ be the path that contains it.  
+%  Consider the child of $v$ that is not in $p$, and let $S(v)$ be the set of all the colors in this subtree.
+%  The claim is that $v$ is the last occurence (i.e. closest to the root) of any color in $S(v)$.
+%  We prove this by induction on the distance of $p$ from the root of the shrub.
+%  For the base case, let $p$ be the root of the shrub, and let $v\in p$.
+%  Let $c$ be any color in $S(v)$.  Suppose there is some vertex $w$ in between $v$ and 
+%  r_p which is colored $c$.  This clearly cannot happen since $w$ must have the color of the leaf of $p$, 
+%  and its second color must come from S(w) which does not contain $c$.
+%  Now for the induction step, let $p$ be some path, and suppose the statement holds for $p'$ 
+%  closer to the root of the shrub.
+%  
+%  
+%  This vertx gets two colors,  e from the path p that it is in 
+%  and the from the set of all colors of the subtree of the child not in $p$.  Call this set of The claim is that regarless of which color
 

### 2014-11-02 paint-contour-tree.tex
section 9 done

@@
 
-\newtheorem{theorem}{Theorem}
+\newtheorem{theorem}{Theorem}[section]
 \newtheorem{lemmata}[theorem]{Lemmata}
@@
 \newtheorem{claim}[theorem]{Claim}
-\newtheorem{subclaim}{Subclaim}
+\newtheorem{subclaim}[theorem]{Subclaim}
 \newtheorem{proposition}[theorem]{Proposition}
@@
 \theoremstyle{definition}
-\newtheorem{remark}{Remark}
-\newtheorem{observation}{Observation}
+\newtheorem{remark}[theorem]{Remark}
+\newtheorem{observation}[theorem]{Observation}
 
@@
 
+%\usepackage{thmtools}
+%\usepackage{thm-restate}
+%\declaretheorem[name=Theorem,numberwithin=section]{thm}
+
 \author{
@@
 \begin{figure}[h]\centering
-    \includegraphics[width=.35\linewidth,height=.15\linewidth]{figs/balanced3}
+    \includegraphics[width=.35\linewidth,height=.15\linewidth]{figs/balanced4}
     \hspace{.3in}
     \includegraphics[width=.35\linewidth,height=.15\linewidth]{figs/trees}
-    \caption{On top, a surface will a balanced contour tree, but whose join and split trees have long tails.  
-    On bottom (from left to right), the contour, join and split trees.}
+    \caption{On left, a surface will a balanced contour tree, but whose join and split trees have long tails.  
+    On right (from left to right), the contour, join and split trees.}
     \label{fig:sorting}
@@
 \begin{figure}[h]\centering
-    \includegraphics[width=.27\linewidth]{figs/mounds}%
-    \hspace{.35in}
-    \includegraphics[width=.27\linewidth]{figs/mounds2}%
-    \hspace{.4in}
-    \includegraphics[width=.22\linewidth]{figs/mounds3}
-    \caption{On the left, three initial colors.  In the middle, yellow and blue merge into green.  On the right, green and red merge into purple.}
+    \includegraphics[width=.28\linewidth]{figs/mound}%
+    \hspace{.1in}
+    \includegraphics[width=.28\linewidth]{figs/mound2}%
+    \hspace{.1in}
+    \includegraphics[width=.28\linewidth]{figs/mound3}%
+    \hspace{.1in}
+    \includegraphics[width=.065\linewidth]{figs/moundTree}
+    \caption{On the left, three initial colors.  In the middle, yellow and blue merge into green.  On the right, green and red merge into purple.
+    On the far right, the contour tree.}
     \label{fig:colors}
@@
 
-\section{Contour trees of extremum dominant manifolds} \label{sec:paint}
+\section{Contour trees of extremum dominant manifolds} \label{sec:extreme}
 
@@
 %
-\begin{definition} \label{def:paint} Let the 1-skeleton of $\MM$ have edge set $E$ and maxima $X$.
+
+\Ben{This is already defined in section 3, should we use the same numbering?}
+\begin{definition} Let the 1-skeleton of $\MM$ have edge set $E$ and maxima $X$.
 A  \emph{painting} of $\MM$ is a map $\chi:X \cup E \mapsto [|X|]$ with the following property. 
@@
 In words, $H_v$ is the set of ancestors of $v$ which are colored by some leaf in $T_v$.
+
+% We will use the $H_v$ values for a given $\chi$ to define a specific path decomposition of $T$ \Def{path}.
+% Specifically, this path decomposition is naturally defined by specifying a subset of edges in $T$ such that each internal
+% vertex has degree at most $2$.
+% For each internal vertex $v\in T$, add the edge from $v$ to $\arg \max_{v_l, v_r} \{|H_{v_l}|, |H_{v_r}|\}$ 
+% where $v_l$ and $v_r$ are the children of $v$ (if $|H_{v_l}|=|H_{v_r}|$ then pick one arbitrarily).
+% This is called the \emph{maximum} path decomposition, denoted by $\pmax(T)$ (which is the decomposition referenced in \Thm{main-alg}).
+
+\begin{definition}
+\label{def:initialColoring}
+Note that the subroutine $\init(\MM)$ from \Sec{algo} naturally defines a leaf assignment to $\jc(\MM)$ 
+according to the priority queue for each up-star we put a given vertex in.  Call this the \emph{initial coloring}
+of the vertices in $\jc(\MM)$.  Note also that this initial coloring defines the $H_v$ values for all $v\in \jc(\MM)$.
+\end{definition}
+
 The following lemma should justify these technical definitions.
-
-Note that the subroutine $\init(\MM)$ from \Sec{algo} naturally defines a leaf assignment to $\jc(\MM)$ 
-according to the priority queue for each up-star we put a given vertex in.
 
@@
 Let $\MM$ be a simplicial complex with $t$ critical points.  For every vertex in $\jc(\MM)$, 
-let $H_v$ be defined by the leaf assignment determined by $\init(\MM)$.
+let $H_v$ be defined by the initial coloring of $\MM$.
 The running time of $\build(\MM)$ is $O(N+t\alpha(t) + \sum_{v \in \jc(\MM)} \log |H_v|)$.
@@
 
-\Ben{For now we are keeping the above full bin tree assumption. At some point this should be carefully removed.}
-
-Note that the number of paths in $P(T)$ must equal the number of leaves in $T$.
-Fix $\chi$, which fixes all the heaps $H_v$. We choose a special path decomposition that is best defined as a subset of edges in $T$ such that each internal
+Let $\chi$ be some fixed leaf assignment to a rooted binary tree $T$, which in turn fixes all the heaps $H_v$. 
+We choose a special path decomposition that is best defined as a subset of edges in $T$ such that each internal
 vertex has degree at most $2$. This naturally gives a path decomposition.
@@
 
-Our main theorem is the following (which by \Lem{runTimeUpper} upper bounds the running time of our algorithm). We use $|p|$ to denote the number of vertices in $p$.
+Our main goal in this section is to prove the following theorem.  We use $|p|$ to denote the number of vertices in $p$.
 
@@
 
+Specifically, proving the above theorem will imply our main result, \Thm{main-alg}, which we now restate and prove.
+
+\begin{theorem}
+Consider a PL-Morse $f:\MM \mapsto \RR$, where the contour tree $T$
+has maximum degree $3$. There is an algorithm whose running time is $O(\sum_{p \in P(T)} |p|\log |p| + t\alpha(t) + N)$,
+where $P(T)$ is a specific path decomposition (constructed implicitly by the algorithm).
+\end{theorem}
+\begin{proof}
+\Ben{Prove the theorem, and change the numbering on the theorem.}
+\end{proof}
+
 \subsection{Shrubs, tall paths, and short paths}
+\label{sec:shrubs}
 
@@
 we first consider the easier case of join trees, after which the lower bound for more general contour trees will easily follow.
+\Ben{Hopefully the following sentence clear up confusion between $\MM$ and $(f,\MM)$}
+Note earlier we usually used $\MM$ to denote the lift of a simplicial complex according to some height function $f$, 
+however in this section we make a distinction between the complex $\MM$ and the height function $f$.  
+In particular all previous definitions or subroutines which took $\MM$ as input now take $(f,\MM)$ as input.
 
@@
 
-In this section we prove \Thm{main-lb}, but for join trees instead of contour trees.  Observe that given a fixed $(f,\MM)$, 
-our algorithm for computing $\jc(\MM)$ is deterministic and so the path decomposition $P=P(\jc(\MM))$ in \Thm{runtime} is fixed, 
-allowing us to define the cost, $\cost(f)=\sum_{p\in P} |p|\log|p|$.
-
-%In the following note that two distinct terrains can have the same graph structure defining their meshes, but different height values associated with their vertices.
+In this section we prove \Thm{main-lb}, but for join trees instead of contour trees.  
+%Observe that given a fixed $(f,\MM)$, 
+%our algorithm for computing $\jc$ is deterministic and so the maximum path decomposition $P=P(\jc)$ defined by the initial coloring (\Def{initialColoring}) is fixed, 
+%allowing us to define the cost, $\cost(f)=\sum_{p\in P} |p|\log|p|$.
+%  
+Let $\MM$ be fixed.  For a given height function $f$, we can now define $\cost_{\jc}(f)$.  Specifically, consider running the $\build(f,\MM)$ procedure 
+of section \Sec{paint}, which computes the join tree $\jc(f,\MM)$.  This procedure calls $\init$, which will (uniquely) define some 
+initial coloring of $\jc$ (\Def{initialColoring}).  Let $P(f)=P(\jc(f,\MM))$ be the maximum path decomposition of the join tree according to this coloring. 
+Then $\cost_{\jc}(f) = \sum_{p\in P(f)} |p|\log|p|$
 
@@
 %    \includegraphics[width=.095\linewidth]{figs/sideView}
-    \caption{Left: angled view of a diamond / Right: a parent and child cone put together}
+    \caption{Left: angled view of a diamond / Right: a parent and child diamond put together}
     \label{fig:diamond}
@@
 \label{lem:pathDecomp}
-Let $\cost(f)=\sum_{p\in P} |p|\log|p|$, for some height function $f$ defined over a complex $\MM$.
-Then there is a family of functions, $F$, all defined over a common triangulation in plane, 
-such that the size of $F$ is $\Pi_{p_i\in P} (|p_i|-1)!$, and no two function in $F$ define the same join tree.
+Let $\mathcal{P} = P(f)$ be the maximum path decomposition of some height function $f$ defined over a complex $\MM$. 
+Then there is a family of functions, $\bF$, all defined over a common triangulation in plane, 
+such that the size of $\bF$ is $\Pi_{p_i\in P} (|p_i|-1)!$, and no two function in $\bF$ define the same join tree.
 \end{lemma}
@@
  of length $m$.  Specifically have two antipodal points, at heights $1$ and $0$, which we will call the anchor and trap of the diamond, respectively.
- The remaining $m-2$ are evenly spread between on the cycle and heights decrease monotonically when going from the anchor to the trap.
+ The remaining $m-2$ vertices are evenly spread around the cycle and heights decrease monotonically when going from the anchor to the trap.
  Next, create an apex vertex at some appropriately large height, and add an edge between it and each vertex in the cycle.  At this point 
  we have built an upside down cone with $m$ triangular faces (see \Fig{diamond}).
- %To make the terrain fully triangulated, we add a triangular outer face around the cycle. 
- %Specifically, add a triangle with vertices at heights $0$, $1/20m$, and $1/10m$, such that in the projection to the plane 
- %this triangle contains the $m$ length cycle, and there is an arbitrary triangulation of the space between this triangle 
- %and the cycle.  For consistency, the inner cycle of a diamond will always have at length at least $3$, i.e. even if $m\leq 3$.
- %
- %Now above each face of the cone of the diamond, we will create a triangular ``platform'', which will act as the location to glue on the bounding triangle of other (shrunken) diamonds, 
- %and whose highest point of connection to the face below will be a saddle vertex.  Specifically the heights of the vertices of the triangle will be $h$, $h-\eps$, and $h-2\eps$, where 
- %$h$ is some unspecified height value.  This near horizontal platform is then connected to its respective face with (nearly) vertical edges down to the face. See \Fig{diamond}.
-
- Now we describe how to attach two diamonds.  In this process, we glue the base of a scaled down ``child'' diamond on to a triangular cone face of the large  ``parent'' diamond (see \Fig{diamond}).
+
+ Now we describe how to attach two diamonds.  In this process, we glue the base of a scaled down ``child'' diamond on to a triangular cone face of the larger  ``parent'' diamond (see \Fig{diamond}).
  Specifically, the anchor of the child diamond is attached directly to a face of the parent diamond at some unspecified height $h$. The remained of the base of the child cone is then 
- extended down until it hits the face of the parent.
+ extended down (at a slight angle) until it hits the face of the parent.
 
@@
  The two faces adjacent to the anchor are always empty, and the remaining faces are for gluing on other diamonds.
- (Note that diamonds have size $|p_i|+1$, since path $|p_i|-1$ faces represent the joins of $p_i$, the apex represents the leaf, 
+ (Note that diamonds have size $|p_i|+1$ since $|p_i|-1$ faces represent the joins of $p_i$, the apex represents the leaf, 
  and we need two empty faces next to the anchor.)
- Now we glue together diamonds of different paths in the same way the paths are connected in the shrub $\pathTree$.  
+ Now we glue together diamonds of different paths in the same way the paths are connected in the shrub $\pathTree$ (see \Sec{shrubs}).  
  Specially, for two paths $p,q\in P$ where 
@@
  (since all these saddles have a direct line of sight to this apex).  In particular, 
- this implies that, regardless of the heights assigned to the platforms, the join tree has a path decomposition whose corresponding shrub is equivalent to $\pathTree$.
+ this implies that, regardless of the heights assigned to the anchors, the join tree has a path decomposition whose corresponding shrub is equivalent to $\pathTree$.
 
  Now there is a valid instance of this described construction for any relative ordering of the heights of the saddles on a given diamond.  
- Let the set of height functions for these instances be denote $F$.
+ Let the set of height functions for these instances be denote $\bF$.
  In particular, there are $(|p|-1)!$ possible orderings of the heights of the saddles on the diamond for $p$, and hence $\Pi_{p_i\in P} (|p_i|-1)!$ possible 
@@
 
+In the following we say to path decompositions $P_1$ and $P_2$ (of potentially different complexes and/or height functions) are equivalent 
+if there is an graph isomorphism, $f$, between their respective shrubs, $S_1$ and $S_2$, which has the property that for any node $v\in S_1$ corresponding to a path of 
+size $|v|$, $f(v)$ also corresponds to a path of size $|v|$. Note that for equivalent decompositions, $\sum_{p\in P_1} |p|\log|p| = \sum_{p\in P_2} |p|\log|p|$.
+
 \begin{lemma}
 \label{lem:cost}
- For any $f\in F$, $\cost(f) = \sum_{p\in P} |p|\log|p|$
+Let $\mathcal{P}$ and $\bF$ be as defined above.  Then for all $f\in \bF$, $\cost_{\jc}(f) = \sum_{p\in P(f)} |p|\log|p| = \sum_{p\in \mathcal{P}} |p|\log |p|$
 \end{lemma}
 \begin{proof}
-First observe that by construction, regardless of the choice of $f\in F$, 
-there is some path decomposition $\overline{P}(T)$ of the join tree such that 
-each path $p$ consists exactly of the set of critical points (except for its anchor) from a given diamond (i.e. $\overline{P}(T)$ is the same for any $f\in F$). 
-We now show that regardless of how paint spilling is done, this will be the maximum 
-path decomposition of $T$.
+First observe that by construction, regardless of the choice of $f\in \bF$, there is a path decomposition $\overline{P}(T)$ of the join tree which is equivalent to $\mathcal{P}$.
+Specifically, this is the decomposition where each path $p\in \overline{P}(T)$ consists exactly of the set of critical points (except for its anchor) from a given diamond. 
+We now show that regardless of how paint spilling is done, this will be the maximum path decomposition of $T$ (implying the lemma statement).
 
  So consider any order of paint spilling over the surface defined by $f$.
- Consider any join on a diamond.  This join is the anchor point of some connecting child diamond.  
- The claim is that the two colors of this join (according to leaf assignment, \Def{leafAssign}, used in \Lem{runTimeUpper})
+ Consider any join on a face of some diamond.  This join is the anchor point of some connecting child diamond.  
+ The claim is that the two colors of this join (according to \Def{initialColoring})
  are the colors of the apexes of the child and parent diamond.  
@@
 
- So let $c$ be the color of some maxima, $m$.  This claim implies that in the join tree, the vertices for 
- which $c$ appears are exactly the vertices in path in $\overline{P}(T)$ corresponding to the diamond of $m$
- and the vertex to which the root of this path connects to.
- This in turn implies that $\overline{P}(T)$ is a maximum path decomposition and hence $\cost(f) = \sum_{p\in \overline{P}} |p|\log|p|$
- To see this, consider a vertex ..... (remark that ties can be broken consistently, and so the there is one maximum decomp)
-
-%  All these joins have an edge which connects 
-%  directly to the maxima of that diamond.  In particular, all these joins have 
-%  an upstar which is entirely colored by the color of this diamond's maxima.  
-%  Therefore, in the leaf assignment of the join tree, determined by this painting (see \Def{leafAssign})
-%  all these join get assigned a common color.  
-%  In particular, all vertices in each path in $\overline{P}$ get assigned a common color.
-
-%  Now consider any vertex $v$ in $T$, and let $p$ be the path that contains it.  
-%  Consider the child of $v$ that is not in $p$, and let $S(v)$ be the set of all the colors in this subtree.
-%  The claim is that $v$ is the last occurence (i.e. closest to the root) of any color in $S(v)$.
-%  We prove this by induction on the distance of $p$ from the root of the shrub.
-%  For the base case, let $p$ be the root of the shrub, and let $v\in p$.
-%  Let $c$ be any color in $S(v)$.  Suppose there is some vertex $w$ in between $v$ and 
-%  r_p which is colored $c$.  This clearly cannot happen since $w$ must have the color of the leaf of $p$, 
-%  and its second color must come from S(w) which does not contain $c$.
-%  Now for the induction step, let $p$ be some path, and suppose the statement holds for $p'$ 
-%  closer to the root of the shrub.
-%  
-%  
-%  This vertx gets two colors,  e from the path p that it is in 
-%  and the from the set of all colors of the subtree of the child not in $p$.  Call this set of The claim is that regarless of which color
-
- \Ben{I am stuck here.}
-%  One can now prove by induction (on the distance of a path from the root of the shrub), 
-%  that if all vertices in every path have a common color, then that color can only appear 
-%  in that path and the vertex to which that path connects (which then trivial implies this 
-%  is a maximum path decomposition).  
-%  First observe that this trivially holds for the path containing the root of $T$.  
-%  For the induction step, consider any path $p$, and let $p'$ be the path 
-%  which the root of $r_p$ of the path $p$ connect to.
-%  Specifically, let $v$ be the vertex of $p'$ at which $p$ connects.
-%  
-%  So consider any vertex $v$ in the path $p$ which contains the root of $T$.  
-%  Now $v$ has two children, one child $c_1$ in $p$ and some other child $c_2$ in another path $p'$.
-%  %In order to show this is a maximum path decomposition we must argue $H_{c_1}\geq H_{c_2}$.
-%  The claim is that $H_{c_2}$ cannot contain any vertex of $p$ other than $v$.
-%  Specifically, consider any vertex on $p$ between the root and $v$.  This vertex must 
-%  get assigned two color, one from each of it subtree.  It is already assigned a color from 
-%  the subtree of $c_1$ hence its second color cannot come from the subtree of $c_2$, 
-%  since both $c_1$ and $c_2$ are in the same subtree of this vertex.
-%  This proves the base case, and the induction step follows by the same argument.
-%  
-%  One can now prove by induction (on the distance of a path from the root of the shrub), 
-%  that if all vertices in every path have a common color, then that color can only appear 
-%  in that path and the vertex to which that path connects (which then trivial implies this 
-%  is a maximum path decomposition).  
-%  So consider any vertex $v$ in the path $p$ which contains the root of $T$.  
-%  Now $v$ has two children, one child $c_1$ in $p$ and some other child $c_2$ in another path $p'$.
-%  %In order to show this is a maximum path decomposition we must argue $H_{c_1}\geq H_{c_2}$.
-%  The claim is that $H_{c_2}$ cannot contain any vertex of $p$ other than $v$.
-%  Specifically, consider any vertex on $p$ between the root and $v$.  This vertex must 
-%  get assigned two color, one from each of it subtree.  It is already assigned a color from 
-%  the subtree of $c_1$ hence its second color cannot come from the subtree of $c_2$, 
-%  since both $c_1$ and $c_2$ are in the same subtree of this vertex.
-%  This proves the base case, and the induction step follows by the same argument.
+ So let $c$ be the color of some maxima, $m$.  The above proven property of the coloring implies that in the join tree, the vertices for 
+ which $c$ appears are exactly the vertices of the path in  $\overline{P}(T)$ corresponding $m$, as well as 
+ the vertex to which the root of this path connects to (i.e. the anchor of the diamond of $m$).
+ Now we show this implies $\overline{P}(T)$ is a maximum path decomposition.
+ To see this, consider a vertex $v\in T$, and let $p$ be the path in $\overline{P}(T)$ which contain $v$. 
+ Now $v$ has two children (if it is a join).  Let $c_1$ be the child which is in $p$, and $c_2$ the child which is the root of some adjacent path $p'$.  
+ In order for $\overline{P}(T)$ to be the maximum path decomposition, we must show $|H_{c_1}|\geq |H_{c_2}|$.
+ %(Note, to ensure a unique maximum path decomposition, one can break ties consistently by assigning some arbitrary but fixed ordering to the vertices.)
+ %Now $v$ gets assigned two colors, one 
+ %of which we know is the color assigned to all vertices in $p$ and the other is the one assigned to all the vertices of $p'$. 
+ Observe that none of the colors in the subtree of $c_2$ appear below $v$.  Hence the heap at $c_2$ consists only of $v$.  
+ On the other hand the heap of $c_1$ contains $v$, any other vertex of $p$ below $v$, and the vertex to which the root of $p$ is adjacent to, 
+ and hence $|H_{c_1}|\geq |H_{c_2}|$.
 \end{proof}
@@
 The following is the analogous statement to \Thm{main-lb}, but for join trees.
+Call a cost $C$ join tree valid if $C = cost_{\jc}(f)$ for some $f$ defined over a given complex $\MM$.
 
 \begin{theorem}\label{thm:joinLB}
-Consider the set of functions $\bF_C$ that all have the same (valid) cost $C$.  
+Consider the set of functions $\bF_C$ that all have the same join tree valid cost $C$.  
 Any height function decision tree that correctly computes
@@
 \begin{proof}
- If $C$ is a valid cost then there some $f$ such that $\cost(f)=\sum_{p\in P(T)} |p|\log|p|$, where $P(T)$ is the maximum path decomposition of the join tree defined by f.
+ If $C$ is a join tree valid cost then there some $f$ such that $\cost_{\jc}(f)=\sum_{p\in P(f)} |p|\log|p|$.
  By the above lemmas, there is then a set $\bF$ of $\Pi_{p_i\in P} |p_i|!$ functions, such that for any $g\in \bF$, $C= \cost(g)$.  
  (Specifically, each sorted ordering of the relative height values of the saddles on each diamond yields a function in this class.)
- Therefore, any height function decision tree for the set $\bF$ must distinguish between these possibilities, and therefore must 
+ Hence $\bF\subseteq \bF_C$, and so any height function decision tree for the set $\bF_C$ must distinguish between these possibilities, and therefore must 
  have $\Pi_{p_i\in P} |p_i|!$ leaves.  Hence by Stirling's approximation, the average depth in this tree is $\Omega(\sum_{p_i\in P} |p_i|\log|p_i|)$.
@@
 
+We need to generalize the definition of path decompositions, which previously were only defined for rooted trees.
 Again we define a path decomposition as a partition of the vertices into 
@@
 \label{def:path2} A path decomposition, $P(T)$, is called \emph{rain consistent} if its paths can be obtained as follows.
-Perform an upward BFS from an arbitrary minima $v$ in $T$, and mark all vertices encountered.  
+Perform an downward BFS from an arbitrary maxima $v$ in $T$, and mark all vertices encountered.  
 Now recursively run a directional BFS from all vertices adjacent to the current marked set.  
-Specifically, for each BFS run, make it an upward BFS if it is at an odd height in the recursion tree and downward otherwise.
+Specifically, for each BFS run, make it an downward BFS if it is at an odd height in the recursion tree and upward otherwise.
 
@@
 For each such subtree, now take any partition of the vertices into leaf paths.\footnote{Note that the subtree  
-of the initial vertex is rooted at is a minima.  For simplicity we require that the path this vertex belongs to 
-also contains a maxima.}
+of the initial vertex is rooted at is a maxima.  For simplicity we require that the path this vertex belongs to 
+also contains a minima.}
 \end{definition}
+
+Consider our algorithm on a fixed $(f,\MM)$.  First we run $\rain(x,\MM)$ to partition the input 
+into extreme dominant complexes.  To make this partition uniquely defined lets say that the initial vertex $x$
+that rain is called from is the global maximum in the input (note that given a specific $x$ the partition is uniquely defined).
+Let $\MM_1, \dots, \MM_k$ be the set of produced subcomplexes, and let $f_i$ be the restriction of $f$ to $\MM_i$.
+Now on each of the $(f_i,\MM_i)$ we run the $\paint$ procedure to compute the join tree of $(f_i, \MM_i)$.
+As described in the previous section, this painting procedure defines a maximum path decomposition with 
+and associated cost, $\cost_{\jc}(f_i) = \sum_{p\in P_i} |p|\log |p|$.  
+It is not hard to see that the union of these paths decomposition $P= \cup_{i=1}^k P_i$ defines 
+a rain consistent path decomposition of the contour tree for $(f,\MM)$.  
+In particular for a fixed $\MM$, $f$ defines a specific path decomposition $P=P(f)$, which we call the \emph{rain consistent maximum path decomposition}, 
+and we write $\cost(f) = \sum_{p\in P} |p|\log |p|$.
 
@@
 \label{lem:pathDecomp2}
-Let $T$ be a contour tree, and let $P = P(T)$ be any rain consistent path decomposition of $T$.  
-Then there is a set of at least $\Pi_{p_i\in P} (|p_i|-1)!$ distinct terrains, all of which have the same defining mesh.
+Let $\mathcal{P} = P(f)$ be the rain consistent maximum path decomposition of some height function $f$ defined over a complex $\MM$. 
+Then there is a family of functions, $\bF$, all defined over a common triangulation in plane, 
+such that the size of $\bF$ is $\Pi_{p_i\in P} (|p_i|-1)!$, and no two function in $\bF$ define the same contour tree.
 \end{lemma}
 \begin{proof}
- As $P$ is rain consistent, the paths can be partitioned into sets $P_1, \dots, P_m$, where $P_i$ is the set of all paths with 
+ As $\mathcal{P}$ is rain consistent, the paths can be partitioned into sets $P_1, \dots, P_k$, where $P_i$ is the set of all paths with 
  vertices from a given BFS, as described in \Def{path2}.  Specifically, let $T_i$ be the subtree of $T$ corresponding to $P_i$ and 
  let $r_i$ be the root vertex of this subtree.  
- Note that the $P_i$ sets naturally define a tree where $P_i$ is the parent of $P_j$ if $r_i$ is adjacent to a vertex in $P_j$.
+ Note that the $P_i$ sets naturally define a tree where $P_i$ is the parent of $P_j$ if $r_i$ (i.e. the root of $T_i$) is adjacent to a vertex in $P_j$.
 
  As the set $P_i$ is a path decomposition of a rooted binary tree $T_i$, the terrain construction of \Lem{pathDecomp} for $P_i$ is well defined.  
- Actually the only difference is that here the rooted tree is not a full binary tree, and so some of the faces of the constructed diamonds will be blank.  
- Specifically, these blank faces correspond to the adjacent children of $P_i$, and they tell us how to connect the meshes of the different $P_i$'s.
+ Actually the only difference is that here the rooted tree is not a full binary tree, and so some of the (non-achor adjacent) faces of the constructed diamonds will be blank.  
+ Specifically, these blank faces correspond to the adjacent children of $P_i$, and they tell us how to connect the terrains of the different $P_i$'s.
+
+% \begin{figure}[h]\centering
+%     \includegraphics[width=.5\linewidth]{figs/bridge2}
+%     \caption{On the left, the root of $P_i$ merges into a diamond of its parent $P_j$.  On the right, the root of $P_i$ splits off from a diamond of $P_j$}
+%     \label{fig:diamond}
+% \end{figure}
+
+ So for each $P_i$ construct a terrain as described in \Lem{pathDecomp}.  Now each $T_i$ is (roughly speaking) a join or a split tree, depending on 
+ whether the BFS which produced it was an upward or downward BFS, respectively.  As the construction in \Lem{pathDecomp} was for join trees, for each terrain 
+ we constructed for a $P_i$ which came from a split tree, must be flipped upside down.
+ Now we must described how to glue the terrains together.
 
  \begin{figure}[h]\centering
-    \includegraphics[width=.5\linewidth]{figs/bridge2}
-    \caption{On the left, the root of $P_i$ merges into a diamond of its parent $P_j$.  On the right, the root of $P_i$ splits off from a diamond of $P_j$}
-    \label{fig:diamond}
+    \includegraphics[width=.28\linewidth]{figs/cones2}
+    \caption{A child diamond attached to a parent diamond with opposite orientation.}
+    \label{fig:diamond2}
 \end{figure}
 
- So for each $P_i$ construct a surface mesh as described in \Lem{pathDecomp}.  By construction, the diamonds corresponding to the paths in $P_i$ are connected 
+ By construction, the diamonds corresponding to the paths in $P_i$ are connected 
  into a tree structure (i.e. corresponding to the shrub of $P_i$).  Therefore the bottoms of all these diamonds are covered except for the one corresponding 
- to the path containing the root $r_i$.  If $r_i$ is the initial minima that the rain consistent path decomposition was defined from,  
- we add a tetrahedral cap to the uncovered bottom of this diamond (i.e. we construct the minima).  Otherwise, $P_i$ has some parent $P_j$ in which case 
- we connect the bottom of the diamond for $r_i$ to a free face of a diamond in the construction for $P_j$, specifically, the face corresponding to the 
+ to the path containing the root $r_i$.  
+ If $r_i$ corresponds to the initial maxima that the rain consistent path decomposition was defined from, then this will be flat and corresponds to the global outer face.
+ Otherwise, $P_i$ has some parent $P_j$ in which case we connect the bottom of the diamond for $r_i$ to a free face of a diamond in the construction for $P_j$, specifically, the face corresponding to the
  vertex in $T$ which $r_i$ is adjacent to.
- If $T_i$ was created by an upward BFS, then the connection to this face should be a join vertex, and otherwise it should be a split vertex.  
- See \Fig{diamond}.
-
- Just as in \Lem{pathDecomp} we now have one fixed mesh graph structure, such that each different relative ordering of the heights of the join and split vertices 
- on each diamond produces a surface with a distinct join tree.  The specific bound on the number of such distinct manifolds follows by applying the 
+This gluing is done in the same manner as in \Lem{pathDecomp}, attaching the anchor for the root of $P_i$ directly the corresponding face of $P_j$, except that now $P_i$ and $P_j$ have opposite orientations.
+See \Fig{diamond2}.
+
+ Just as in \Lem{pathDecomp} we now have one fixed terrain structure, such that each different relative ordering of the heights of the join and split vertices 
+ on each diamond produces a surface with a distinct join tree.  The specific bound on the size of $\bF$, defining these distinct join trees, follows by applying the 
  bound from \Lem{pathDecomp} to each $P_i$.
@@
 
+\begin{lemma}
+Let $\mathcal{P}$ and $\bF$ be as defined above.  Then for all $f\in \bF$, $\cost(f) = \sum_{p\in P(f)} |p|\log|p| = \sum_{p\in \mathcal{P}} |p|\log |p|$
+\end{lemma}
+\begin{proof}
+ This lemma follows immediately from \Lem{cost}.  Specifically, the sum $\cost(P)$, for any rain consistent maximum path decomposition $P$, can be broken 
+ into the sums $\cost_{\jc}(P_i)$ of some maximum path decompositions $P_1, \dots P_k$.  Now apply \Lem{cost} to each of the $P_i$ separately and take the sum.
+\end{proof}
+
+Call a cost $C$ valid if $C = cost(f)$ for some $f$ defined over a given complex $\MM$.
 For the same reason that \Thm{joinLB} followed from \Lem{pathDecomp} and \Lem{cost}, we now have the following.
@@
 \begin{theorem}
- Let $A(S)$ be any comparison based algorithm to compute the contour tree of an input surface $S$.  
- Then the running time of $A(S)$ is $\Omega(f_{path}(P))=\Omega(\sum_{p_i\in P} |p_i|\log|p_i|)$ where $P$ 
- is any rain consistent path decomposition of the contour tree of $S$.
+Consider the set of functions $\bF_C$ that all have the same valid cost $C$.  
+Any height function decision tree that correctly computes
+the contour tree on all of $\bF_C$, has an average case (uniform over $\bF_C$) running time of $\Omega(C)$.
 \end{theorem}
@@
 \begin{remark}
- Note that for the surfaces described in this section, the number of critical points is within a constant factor of 
- the total number of vertices.  In particular, for this family of surfaces, all previous algorithms required $\Omega(n\log n)$ time.
+ Note that for the terrains described in this section, the number of critical points is within a constant factor of 
+ the total number of vertices.  In particular, for this family of terrains, all previous algorithms required $\Omega(n\log n)$ time.
 \end{remark}

### 2014-11-03 paint-contour-tree.tex
small changes

@@
 
+\newcommand{\atgen}{\symbol{'100}}
+\newcommand{\si}[1]{#1}
+\newcommand{\BenThanks}[1]{\thanks{%
+      Department of Computer Science; %
+      University of Illinois; %
+      201 N. Goodwin Avenue; %
+      Urbana, IL, 61801, USA; %
+      {\tt \si{raichel}2\atgen{}\si{uiuc}.\si{edu}}; %
+      {\tt \url{\si{http://illinois.edu/\string~\si{raichel2}}}}%
+      . %
+      #1}}
+
 \author{
-  Benjamin Raichel
+  Benjamin Raichel\BenThanks{}
   \and
@@
 
-\title{A Mountaintop View Requires Minimal Sorting: \break A Faster Contour Tree Algorithm}
+\title{A Mountaintop View Requires Minimal Sorting: \break A Faster Contour Tree Algorithm%
+\footnote{The full updated version of this paper is available online \url{\si{http://web.engr.illinois.edu/\string~\si{raichel2}/\si{paint.pdf}}}}%
+}
 \date{}
@@
 \end{figure}
+
+\paragraph{Organization.}  
+\Sec{basics} contains basic definition for contour trees.  In \Sec{approach} we give a high level overview and intuition for our approach 
+for computing contour trees.  \Sec{surgery} and \Sec{rain} describe how we break the intial input into nicely behaved (extemum dominant) pieces. 
+\Sec{extreme} shows that computing the contour tree for these nicely behaved pieces is roughly equivalent to computing the join tree. 
+\Sec{paint} contains our algorithm for computing join trees, which along with previous sections gives our algorithm for contour trees.  
+Finally, in \Sec{pathDecomp} and \Sec{lb} we relate the running time of our algorithm (providing upper and lower bounds) to a particular decomposition of the contour tree.
 
@@
 \section{Lower Bound by Path Decomposition}
+\label{sec:lb}
 In this section we provide a lower bound for the time it takes to compute a contour tree in terms the cost of path decompositions.  Specially, 

### 2014-11-03 paint-contour-tree.tex
more small changes

@@
 It is a fundamental data structure in data analysis and visualization,
-and there is much past work (theoretical and practical) on algorithms for construct contour trees.
+and there is significant previous work (both theoretical and practical) on algorithms for constructing contour trees.
 
@@
 is $3$. A fundamental result in this area is the algorithm of Carr, Snoeyink, and Axen to compute contour trees,
-which runs in $O(n\log n + N\alpha(N)$ time~\cite{csa-cctad-00}. (Henceforth, $\alpha(\cdot)$ denotes the inverse
+which runs in $O(n\log n + N\alpha(N))$ time~\cite{csa-cctad-00}. (Henceforth, $\alpha(\cdot)$ denotes the inverse
 Ackermann function.) It is natural to think of $N$ as $\Theta(n)$
@@
 exploit numerous properties of monotone paths. Carr's thesis shows relationships between monotone paths
-in $\MM$ and in $\cC(\MM)$.
-
-Contour trees are a special case of Reeb graphs, a general topological representation for real-valued function
+in $\MM$ and in its contour tree $\cC(\MM)$.
+
+Contour trees are a special case of Reeb graphs, a general topological representation for real-valued functions
 on any manifold. We refer the reader to Chapter 6 in~\cite{HaEd10} for more details. Algorithms for computing Reeb graphs
@@
 The $(f(x)+\delta)$-contours in $f|_{B_\eps(x)}$ given by \Def{deg} might actually be the same
-contour in $f$. The up-degree is defined as the number of $(f(x)+\delta)$-contours
+contour in $f$. The up-degree (as opposed to \emph{Morse} up-degree) is defined as the number of $(f(x)+\delta)$-contours
 that intersect $B_\eps(x)$, a potentially smaller number. This up-degree
@@
 to as augmenting the contour tree, and it is common to augment $\reeb(\MM)$ will all vertices.
-Alternately, one can smooth out all vertices of up-degree and down-degree $1$ to get the
+Alternatively, one can smooth out all vertices of up-degree and down-degree $1$ to get the
 unaugmented contour tree. (For $d=2$, there are no such vertices in $\reeb(\MM)$.) 
@@
 \paragraph{Do not globally sort:} Arguably, the starting point for this work is \Fig{relative}. We have two terrains
-with exactly the same contour tree, but different orderings of (heights of) the critical point.
-Turning it around, we cannot deduce the height ordering of critical points from the contour tree. Clearly,
-sorting the critical points is computationally reduntant for constructing the contour tree.
+with exactly the same contour tree, but different orderings of (heights of) the critical points.
+Turning it around, we cannot deduce the full height ordering of critical points from the contour tree. Therefore, clearly
+sorting all critical points is computationally unnecessary for constructing the contour tree.
 In \Fig{sorting}, the contour tree basically consists of two balanced binary trees, one of the joins,
 another of the splits. Again, it is not necessary to know the relative ordering between the mounds
-on the left (or among the depressions in the right) to compute the contour tree. Yet some ordering
-information is necessary: in the left, the little valleys are higher than the big central valley,
+on the left (or among the depressions on the right) to compute the contour tree. Yet some ordering
+information is necessary: on the left, the little valleys are higher than the big central valley,
 and this is reflected in the contour tree. More generally, leafs paths in the contour tree have points
@@
 Consider sweeping down via the hyperplane $x_{d+1} = h$ and taking the \emph{superlevel} sets. These 
-are the connected components of $\MM$ with height greater than $h$. For a terrain, the superlevel
+are the connected components of the portion of $\MM$ above height $h$. For a terrain, the superlevel
 sets are a collection of ``mounds". As we sweep downwards, these mounds keep joining each other,
 until finally, we end up with all of $\MM$. The join tree tracks exactly these events. 
-Formally, let $\MM^+_v$ denote the simplicial complex obtained by taking only those
-vertices higher than $v$.
+Formally, let $\MM^+_v$ denote the simplicial complex induced on the subset of 
+vertices which are higher than $v$.
 

### 2014-11-03 paint-contour-tree.tex
small changes

@@
 proceeds various events occur: new contours are created or destroyed, contours merge into each other or
-split into new components, contours acquire or lose handles. These are the relevant events where 
+split into new components, contours acquire or lose handles. More generally, these are the events where 
 the topology of the level set changes. The contour tree is a concise representation of all these events.
@@
 In dimensions $d \geq 3$, we may have internal
-vertices of both up and down-degree $1$. They correspond to contours (which are $2$-manifolds) gaining or losing handles.
+vertices of both up and down-degree $1$. They correspond to contours (which are $(d-1)$-manifolds) gaining or losing handles.
 
@@
 so the running time is $O(t\log t + N)$. But this is tight only when $T$ is tall and thin.  
-A better but still obvious upper bound is $O(t\log D + t\alpha(t) + N)$, where $D$ is the diameter of $P(T)$. 
+A better but still obvious upper bound is $O(t\log D + t\alpha(t) + N)$, where $D$ is the diameter of $T$. 
 This is clearly an improvement when the tree is short and fat.
@@
 Tarasov and Vyalya \cite{tv-cct-98} improved the running time to $O(N\log N)$ for the 3D case.
-The influential paper of Carr \etal \cite{csa-cctad-00} improved the running time for all dimensions to $O(n\log n + N)$.
+The influential paper of Carr \etal \cite{csa-cctad-00} improved the running time for all dimensions to $O(n\log n + N\alpha(N))$.
 One cannot overstate the importance of this result, since it forms the basis 
@@
 is exactly the up-degree of $x$ in $\reeb(\MM)$. (Analogously, for down-degree.)
-When the Morse up-degree is $2$ but the up-degree is $1$, then the level set gains a handle.
+When the Morse up-degree is $2$ but the up-degree is $1$, the topology of the level set changes but not by the number of connected components changing.
+For example, when $d=3$ this is equivalent to the contour gaining a handle.
 When $d=2$, this distinction is not necessary, since any point with $>1$
@@
 \label{def:int-criticalJoin}
-The critical join tree $\jc(\MM)$ is built on the set $V'$ of all critical points. 
-The directed edge $(u,v)$ is present when $u$ is the smallest valued vertex in $V'$ in a connected component of $\MM^+_v$
+The critical join tree $\jc(\MM)$ is built on the set $\cV$ of all critical points. 
+The directed edge $(u,v)$ is present when $u$ is the smallest valued vertex in $\cV$ in a connected component of $\MM^+_v$
 and $v$ is connected to this component (in $\MM$). 
@@
 A major insight of~\cite{csa-cctad-00} is an ingeniously simple linear time procedure to construct
-the contour tree from the join and split tree. So the bottleneck is computing these trees. And observe
-in \Fig{sorting} that the split vertices form a long path in the join tree (and vice versa). Constructing
-these trees forces a global sort of the splits, a redundant computation. Unfortunately (unlike \Fig{sorting}),
-the heights of joins and splits may be interleaved in a complex manner. The final merging of~\cite{csa-cctad-00}
+the contour tree from the join and split trees. So the bottleneck is computing these trees. Observe
+in \Fig{sorting} that the split vertices form a long path in the join tree (and vice versa). Therefore, constructing
+these trees forces a global sort of the splits, an unnecessary computation for the contour tree. Unfortunately, in general (i.e. unlike \Fig{sorting}) 
+the heights of joins and splits may be interleaved in a complex manner, and hence the final merging of~\cite{csa-cctad-00}
 to get the contour tree requires having the split vertices in the join tree. Without
@@
 
-Our aim is break $\MM$ into smaller pieces, where this redundant computation can be saved.
+Our aim is to break $\MM$ into smaller pieces, where this unnecessary computation can be avoided.
 
@@
     \includegraphics[width=.18\linewidth]{figs/tent1}
-    \caption{On left, downward rain spilling only (each shade of gray represents a different color), producing a grid.  On right, flipping the direction of rain spilling.}
+    \caption{On left, downward rain spilling only (each shade of gray represents a different color), producing a grid.  
+    Note we are assuming raining was done in reverse sorted order of maxima.  
+    On right, flipping the direction of rain spilling.}
     \label{fig:order}
@@
 endowed with a height to be \emph{minimum dominant} if there only exists a single minima.
+(Our real definition is more complicated, and involves simplicial complexes
+that allow additional ``trivial'' minima.)
 In such a complex, there exists a non-ascending path from any point to this unique minima.
-(Our real definition is more complicated, and involves simplicial complices
-that have many minima.) Analogously, we can define maximum dominant simplicial complices,
+Analogously, we can define maximum dominant simplicial complexes,
 and both are collectively called extremum dominant.
@@
 The best way to think of our procedure is a meteorology analogy. Take an arbitrary maxima $x$,
-and imagine torrential rain at the maxima. The water flows down, wetting any point that has a non-ascending
+and imagine torrential rain at the maxima. The water flows down, wetting any point that has a non-ascending path 
 from $x$. We end up with two portions, the wet part of $\MM$ and the dry part. This is similar
@@
 while there may be numerous disconnected dry parts. The interface between the dry and wet parts
-are a set of contours, given the by ``water line". The wet part is clearly maximum dominant, since all
+is a set of contours, given by the ``water line". The wet part is clearly maximum dominant, since all
 wet points have a non-descending path to $x$. So we can simply cut along the interface
 contours to get the wet maximum dominant piece $\MM'$. By our contour surgery theorem,
-we are left with a set of disconnected dry parts, and we could recur this procedure on them.
+we are left with a set of disconnected dry parts, and we can recur this procedure on them.
 
@@
 wet and dry piece, and each cut introduces many new vertices. If we wanted to partition this terrain into 
-maximum dominant simplicial complices, the final complexity would be forbiddingly large.
+maximum dominant simplicial complexes, the final complexity would be forbiddingly large.
 
@@
 
-\begin{theorem} \label{thm:int-rain} There is an $O(N)$ time procedure that cuts $\MM$ into extremum dominant simplicial complices
+\begin{theorem} \label{thm:int-rain} There is an $O(N)$ time procedure that cuts $\MM$ into extremum dominant simplicial complexes
 $\MM_1, \MM_2, \ldots$. Furthermore, given the set of contour trees $\{\cC(\MM_i)\}$, $\cC(\MM)$ can be constructed in $O(N)$ time.
@@
 
-\paragraph{Extremum dominance simplifies contour trees:} We will focus on minimum dominant simplicial complices $\MM$. By \Thm{int-rain}, it suffices
+\paragraph{Extremum dominance simplifies contour trees:} We will focus on minimum dominant simplicial complexes $\MM$. By \Thm{int-rain}, it suffices
 to design an algorithm for contour trees on such inputs. 
@@
 at \Fig{sorting}. The portion on the left is minimum dominant in exactly this fashion.
-Water poured at any critical point must flow to the boundary, and the entire wet portion are connected.
+Water poured at any critical point must flow to the boundary, and the entire wet portion is connected.
 More formally, $\MM^-_v$ is connected for all $v$, so there are no splits.

### 2014-11-03 paint-contour-tree.tex
added changes to intro

--- 2014-11-03 paint-contour-tree.tex
@@
 \begin{abstract}
-Consider a scalar field $f:\MM \mapsto \RR$, where $\MM$ is a simplicial mesh in $\RR^d$. 
+Consider a scalar field $f:\MM \mapsto \RR$, where $\MM$ is a triangulated simplicial mesh in $\RR^d$. 
 A level set, or contour, at value $v$ is a connected component of $f^{-1}(v)$. 
@@
 was given by Chiang et al. All these algorithms involve a global sort of the critical points,
-which is a major bottleneck for large datasets. Unfortunately, lower bounds of $\Omega(t\log t)$ also exist.
+a significant computational bottleneck. Unfortunately, lower bounds of $\Omega(t\log t)$ also exist.
 
@@
 our algorithm runs in $O(\sum_{p\in P} |p|\log|p| + t\alpha(t)+N)$, which is 
-always bounded by $O(t\log t + N)$ and is $O(N+t\alpha(t))$ for balanced trees. 
-Moreover, we prove a general lower bound of
-$\Omega(\sum_{p \in P} |p|\log |p|)$ for computing contour trees, showing
-that the tree structure must play a role in the time complexity.
+always bounded by $O(t\log t + N)$ and is $O(t\alpha(t)+N)$ for balanced trees. 
+Moreover, we prove a strong optimality property of our algorithm, showing that $\sum_{p\in P} |p|\log|p|$ complexity
+is inherent to computing contour trees. Consider the family of instances where our algorithm
+requires $C$ comparison. Any algorithm computing contour trees on this family requires
+$\Omega(C)$ comparisons in the worst case.
+
+%
+%general lower bound of
+%$\Omega(\sum_{p \in P} |p|\log |p|)$ for computing contour trees, showing
+%that the tree structure must play a role in the time complexity.
 
@@
 exists some partition $P'$ such that the running time of our algorithm is 
-$O(\sum_{p\in P'} |p|\log|p| + N\alpha(N))$, where $N$ is the total complexity of the input manifold 
-(and hence this is optimal on all instances requiring $\Omega(N\alpha(N))$ time).
+$O(\sum_{p\in P'} |p|\log|p| + N\alpha(N))$, where $N$ is the total complexity of the input manifold.
 In particular, our algorithm runs in $O(N\alpha(N))$ time when the tree is balanced, 
@@
 Massive geometric data is often represented as a function $f: \RR^d \mapsto \RR$. Typically, a finite representation is given
-by considering $f$ to be piecewise linear over some mesh (i.e. simplicial complex) $\MM$ in $\RR^d$.
+by considering $f$ to be piecewise linear over some triangulated mesh (i.e. simplicial complex) $\MM$ in $\RR^d$.
 \emph{Contour trees} are a specific topological data structure used to represent and visualize
@@
 proceeds various events occur: new contours are created or destroyed, contours merge into each other or
-split into new components, contours acquire or lose handles. These are the relevant events where 
+split into new components, contours may acquire or lose handles. These are the relevant events where 
 the topology of the level set changes. The contour tree is a concise representation of all these events.
 
-For convenience, imagine $f$ to be smooth. All points where the gradient of $f$ is zero are \emph{critical points},
-where the topology of the contours change. The vertices of the contour tree are exactly these critical points,
-representing the important topological events. An edge connects two critical points if one event immediately ``follows"
+If $f$ is smooth, all points where the gradient of $f$ is zero are \emph{critical points}.
+These points are the ``events" where the contour topology changes and form the vertices of the contour tree. 
+An edge connects two critical points if one event immediately ``follows"
 the other as the sweep plane makes its pass. (We provide formal definitions later.)
@@
 But this is only a worst-case bound. All previous algorithm begin by sorting (at least) the critical points, and
-hence must pay $\Omega(t\log t)$ for all instances. Can we beat this sorting bound? Is there a more refined version of the complexity
-of computing contour trees? Our main result gives an affirmative answer. First, we require some technical definitions.
+hence must pay $\Omega(t\log t)$ for all instances. Can we beat this sorting bound for some instances? Is there a more refined version of the complexity
+of computing contour trees, and can we characterize what inputs are hard? Our main result gives an affirmative answer. First, we require some technical definitions.
 
@@
 does not need this, but it simplifies the proof and presentation.} $f:\MM \mapsto \RR$, where the contour tree $T$
-has maximum degree $3$. There is an algorithm whose running time is $O(\sum_{p \in P(T)} |p|\log |p| + t\alpha(t) + N)$,
+has maximum degree $3$. There is a deterministic algorithm whose running time is $O(\sum_{p \in P(T)} |p|\log |p| + t\alpha(t) + N)$,
 where $P(T)$ is a specific path decomposition (constructed implicitly by the algorithm).
@@
 so the running time is $O(t\log t + N)$. But this is tight only when $T$ is tall and thin.  
-A better but still obvious upper bound is $O(t\log D + t\alpha(t) + N)$, where $D$ is the diameter of $P(T)$. 
+A better but still obvious upper bound is $O(t\log D + t\alpha(t) + N)$, where $D$ is the diameter of $T$. 
 This is clearly an improvement when the tree is short and fat.
-In particular, for any decomposition of a rooted balanced binary tree, this gives $O(t\log\log n + t\alpha(t) + N)$, a significant improvement from $O(t\log t+N)$.
-However, our actual bound in this case gives an even further improved $O(t\alpha(t)+N)$.
+But it can be even better. A calculation yields that for any decomposition of a rooted balanced binary tree, 
+our bound is $O(t\alpha(t) + n)$. 
+%
+%
+%In particular, for any decomposition of a rooted balanced binary tree, this gives $O(t\log\log n + t\alpha(t) + N)$, a significant improvement from $O(t\log t+N)$.
+%However, our actual bound in this case gives an even further improved $O(t\alpha(t)+N)$.
 Consider the example of \Fig{sorting} extended to a larger binary tree both above and below.
@@
 of $\sum_{p \in P(T)} |p|\log |p|$ is required for any contour tree algorithm. The exact phrasing
-of the lower bound requires some care. Consider the $d=2$ (terrain) case, and fix a mesh $\MM$.
-As referenced in \Thm{main-alg}, for any height function $f$ our algorithm defines a 
+of the lower bound requires some care. Consider the $d=2$ (terrain) case.
+As referenced in \Thm{main-alg}, for any $f:\MM \mapsto \RR$ our deterministic algorithm defines a 
 specific path decomposition $P(T)$\footnote{The specific decomposition used is defined in \Sec{pathDecomp}, 
@@
 Ideally, we would like to argue that any contour-tree algorithm
-must pay $\cost(f)$ for $f$. 
+must pay $\cost(f)$ for ant $f$. 
 But this is obviously false, since one could consider an algorithm
@@
 
-Instead, call a possible cost $C$ \emph{valid} if $C = \cost(f)$ for some $f$ 
-defined over the fixed mesh $\MM$\footnote{If one is concerned
+Instead, call a possible cost $C$ \emph{valid} if $C = \cost(f)$ for some $f$\footnote{If one is concerned
 that $C$ might be irrational, define valid if $2^C = 2^{\cost(f)}$.}. We prove
@@
 
-\begin{theorem} \label{thm:main-lb} Consider the set of functions $\bF_C$ that all have the same (valid) cost $C$ 
-(defined over a fixed mesh $\MM$).  Any height function decision tree that correctly computes
-the contour tree on all of $\bF_C$ has an average case (uniform over $\bF_C$) running time of $\Omega(C)$.
+\begin{theorem} \label{thm:main-lb} 
+%Fix a sufficiently large numbers $t$ and $C$, 
+%where $t$ is the number of critical points and $C$ is a valid cost for an instance with $t$ critical point.
+Consider the set of functions $\bF_C$ that all same number of critical points and the same (valid) cost $C$.  Any algebraic decision tree that correctly computes
+the contour tree on all of $\bF_C$ has a worst case running time of $\Omega(C)$.
 \end{theorem}
@@
 
-In other words, it is not possible to improve on \Thm{main-alg} even when restricted to a subset
+Roughly speaking, it is not possible to improve on \Thm{main-alg} even when restricted to a subset
 of inputs of fixed cost.

### 2014-11-03 paint-contour-tree.tex

@@
 proceeds various events occur: new contours are created or destroyed, contours merge into each other or
-split into new components, contours may acquire or lose handles. These are the relevant events where 
+split into new components, contours acquire or lose handles. More generally, these are the events where 
 the topology of the level set changes. The contour tree is a concise representation of all these events.

### 2014-11-03 paint-contour-tree.tex
8.4 added

--- 2014-11-03 paint-contour-tree.tex
@@
 does not need this, but it simplifies the proof and presentation.} $f:\MM \mapsto \RR$, where the contour tree $T$
-has maximum degree $3$. There is an algorithm whose running time is $O(\sum_{p \in P(T)} |p|\log |p| + t\alpha(t) + N)$,
+has maximum degree $3$. There is an algorithm to compute $T$ whose running time is $O(\sum_{p \in P(T)} |p|\log |p| + t\alpha(t) + N)$,
 where $P(T)$ is a specific path decomposition (constructed implicitly by the algorithm).
@@
 We stress the importance of this step for our approach. Given the algorithm of~\cite{csa-cctad-00},
-one may think that that it suffices to design faster algorithms for join trees. But this
-cannot give the sort of optimality we hope. Again, consider \Fig{sorting}. Any algorithm
+one may think that it suffices to design faster algorithms for join trees. But this
+cannot give the sort of optimality we hope for. Again, consider \Fig{sorting}. Any algorithm
 to construct the true join tree must construct the path of splits, which implies
 sorting all of them. It is absolutely necessary to cut $\MM$ into pieces where the cost
-of building the join trees can be related to the cost of building $\cC(\MM)$.
+of building the join tree can be related to the cost of building $\cC(\MM)$.
 
@@
 then spill from the next, etc. Paint is viscous, and only flows down edges. \emph{It does not
-paint the interior of facets.} Alternately, this process is restricted to the 1-skeleton of $\NN$.
+paint the interior of facets.} That is, this process is restricted to the 1-skeleton of $\NN$.
 Furthermore, our paints do not mix, so each edge receives a unique color, which is decided by the first
@@
 Water allows for an easy divide-and-conquer, at the cost of more complexity in the input. For an extremum
-dominant input, using water to divide the input $\NN$ raise the complexity too much.
+dominant input, using water to divide the input $\NN$ raises the complexity too much.
 
@@
 We say that vertex $v$ is \emph{touched} by color $c$, if there is a $c$-colored edge with lower endpoint $v$.
-Let us focus on an initial painting, where the colors have 1-1 correspondence with the maxima. Refer to \Fig{colors}.
+Let us focus on an initial painting, where the colors have 1-1 correspondence with the maxima. Refer to the left part of \Fig{colors}.
 Consider two sibling leaves $\ell_1, \ell_2$ and their common parent $v$. The leaves are maxima,
 and $v$ is a join that ``merges" $\ell_1, \ell_2$. 
-In that case, there are ``mounds" corresponding to each $\ell_1$ and $\ell_2$ that merge
+In that case, there are ``mounds" corresponding to $\ell_1$ and $\ell_2$ that merge
 at a valley $v$. Suppose this was the entire input, and $\ell_1$ was colored blue and $\ell_2$ was colored red. 
-(So ignore the yellow mound in \Fig{colors}.)
 Both the mounds are colored completely blue or red, while $v$ is touched by both colors.
@@
 Alternately, $\jc(\MM)$ has been constructed up to $v$, and it remains to determine $v$'s parent.
-The merging of the colors is not explicitly as that would be too expensive; we maintain a union-find
+The merging of the colors is not explicitly performed as that would be too expensive; we maintain a union-find
 data structure for that.
@@
 Of course, things are more complicated when there are other mounds. There may be a yellow mound,
-corresponding to $\ell_3$ that joins with the blue mound higher up (at vertex $u$). In $\jc(\MM)$, $\ell_2$ and $\ell_3$
-are sibling leaves, and $\ell_1$ is a sibling of some ancestor of these leaves. So we cannot
+corresponding to $\ell_3$ that joins with the blue mound higher up at some vertex $u$ (see the right part of \Fig{colors}). In $\jc(\MM)$, $\ell_1$ and $\ell_3$
+are sibling leaves, and $\ell_2$ is a sibling of some ancestor of these leaves. So we cannot
 merge red and blue, until yellow and blue merge. Naturally, we use priority queues to handle this issue.
@@
 \begin{figure}[h]\centering
-    \includegraphics[width=.28\linewidth]{figs/mound}%
-    \hspace{.1in}
-    \includegraphics[width=.28\linewidth]{figs/mound2}%
-    \hspace{.1in}
-    \includegraphics[width=.28\linewidth]{figs/mound3}%
-    \hspace{.1in}
+    \includegraphics[width=.25\linewidth]{figs/redBlue}%
+    \hspace{.07in}
+    \includegraphics[width=.25\linewidth]{figs/redBlue2}%
+    \hspace{.07in}
+    \includegraphics[width=.03\linewidth]{figs/redBlueTree}
+    \hspace{.09in}
+    \includegraphics[width=.3\linewidth]{figs/mound}%
+    \hspace{.07in}
     \includegraphics[width=.065\linewidth]{figs/moundTree}
-    \caption{On the left, three initial colors.  In the middle, yellow and blue merge into green.  On the right, green and red merge into purple.
-    On the far right, the contour tree.}
+    \caption{On the left, red and blue merge to make purple, followed by the contour tree with initial colors.  On the right, additional maxima and the resulting contour tree.}
     \label{fig:colors}
@@
 We employ a variant
-of \emph{heavy path decompositions}, first used for Sleator and Tarjan for analyzing link/cut trees~\cite{st-dsdt-83}.
-The final analysis is basically charges expensive heap operations to long paths in the decomposition.
+of \emph{heavy path decompositions}, first used by Sleator and Tarjan for analyzing link/cut trees~\cite{st-dsdt-83}.
+The final analysis basically charges expensive heap operations to long paths in the decomposition.
 
@@
 we get $\prod_{p \in P(T)} |p|!$ terrains each with a distinct contour tree. It is possible to argue
-that our algorithm requires $\Theta(\cost(P(T)))$ on all these instances.
+that our algorithm requires $\Theta(\cost(P(T)))$ comparisons on all these instances.
 
@@
 
-\Ben{This is already defined in section 3, should we use the same numbering?}
+We now restate \Def{paint}.
+
 \begin{definition} Let the 1-skeleton of $\MM$ have edge set $E$ and maxima $X$.
@@
 \begin{theorem}
+\label{thm:correct}
  Given an input complex $\MM$, $\build(\MM)$ terminates and outputs $\jc(\MM)$.
@@
 
-Specifically, proving the above theorem will imply our main result, \Thm{main-alg}, which we now restate and prove.
-
-\begin{theorem}
-Consider a PL-Morse $f:\MM \mapsto \RR$, where the contour tree $T$
-has maximum degree $3$. There is an algorithm whose running time is $O(\sum_{p \in P(T)} |p|\log |p| + t\alpha(t) + N)$,
-where $P(T)$ is a specific path decomposition (constructed implicitly by the algorithm).
-\end{theorem}
-\begin{proof}
-\Ben{Prove the theorem, and change the numbering on the theorem.}
-\end{proof}
+We conclude this section in \Sec{implications} by showing that proving this theorem implies our main result \Thm{main-alg}.
 
@@
 That completes the whole proof.
+\end{proof}
+
+\subsection{Our Main Result}
+\label{sec:implications}
+
+We now show that \Thm{runtime} allows us to upper bound the running time for our join tree and contour 
+tree algorithms in terms of path decompositions.
+
+\begin{theorem}
+Consider a PL-Morse $f:\MM \mapsto \RR$, where the join tree $\jc(\MM)$ has maximum degree $3$. 
+There is an algorithm to compute the join tree whose running time is $O(\sum_{p \in \pmax(\jc)} |p|\log |p| + t\alpha(t) + N)$.
+\end{theorem}
+\begin{proof}
+By \Thm{correct} we know that $\build(\MM)$ correctly outputs $\jc(\MM)$, and by \Lem{runTimeUpper}
+we know this takes $O(\sum_{v \in \jc(\MM)} \log |H_v| + t\alpha(t) + N)$ time, where the $H_v$ values are determined as in \Def{initialColoring}.
+Therefore by \Thm{runtime}, $\build(\MM)$ takes $O(\sum_{p \in \pmax(\jc)} |p|\log |p| + t\alpha(t) + N)$ 
+time to correctly compute $\jc(\MM)$.
+\end{proof}
+
+This result for join trees easily implies our main result, \Thm{main-alg}, which we now restate and prove.
+
+\begin{theorem}
+Consider a PL-Morse $f:\MM \mapsto \RR$, where the contour tree $\cC=\cC(\MM)$
+has maximum degree $3$. There is an algorithm to compute $\cC$ whose running time is $O(\sum_{p \in P(\cC)} |p|\log |p| + t\alpha(t) + N)$,
+where $P(T)$ is a specific path decomposition (constructed implicitly by the algorithm).
+\end{theorem}
+\begin{proof}
+First, lets review the various pieces of our algorithm. 
+On a given input simplicial complex, we first we break it into extremum dominant pieces using $\rain(\MM)$ (and in $O(|\MM|)$ time by \Thm{rain-time}).  
+Specifically, \Lem{rain-1} proves that the output of $\rain(\MM)$ is a set of extremum dominant pieces, 
+$\MM_1, \dots, \MM_k$, and \Clm{rain-reeb} shows that given the contour trees, $\cC(\MM_1), \dots, \cC(\MM_k)$ , 
+the full contour tree, $\cC(\MM)$, can be constructed (in $O(|\MM|)$ time).
+
+Now one of the key observations was that for extremum dominant manifolds, computing the contour tree is roughly the same as 
+computing the join tree.  Specifically, \Thm{contour-tree} implies that given $\jc(\MM_i)$ , 
+we can obtain $\cC(\MM_i)$ by simply sticking on the non-dominant minima at their respective splits (which can easily be done in linear time).
+However, by the above theorem we know $\jc(\MM_i)$ can be computed in $O(\sum_{p \in \pmax(\jc(\MM_i))} |p|\log |p| + t_i\alpha(t_i) + N_i)$
+(where $t_i$ and $N_i$ are the number of critical points and facets when restricted to $\MM_i$).
+
+At this point we can now see what the path decomposition referenced in theorem statement should be.  It is just the union of all the maximum 
+path decomposition across the extremum dominant pieces, $\pmax(\cC(\MM)) = \cup_{i=1}^k \pmax(\jc(\MM_i))$.
+Since all procedures besides computing the join trees take linear time in the size of the input complex, we can therefore 
+compute the contour tree in time
+\[
+O\pth{N+ \sum_{i=1}^k \pth{\sum_{p \in \pmax(\jc(\MM_i))} |p|\log |p|} + t_i\alpha(t_i) + N_i} = O\pth{\pth{\sum_{p \in \pmax(\cC(\MM))} |p|\log |p|} + t\alpha(t) + N}
+\]
+
+%\Thm{rain-time} proves that $\rain$ takes $O(|\MM|)$ time
+
+%\Lem{rain-1} proves that output of $\rain$ are extemum dominant.
+
+%\Clm{rain-reeb} shows that given the contour trees of the output manifolds of $\rain$, the full contour tree can be constructed in $O(|\MM|)$ time:
+
+%\Thm{contour-tree} implies that given an $\jc(\MM)$ for an etremum dominant $\MM$, we can obtain $\cC(\MM)$ in by simply sticking on the non-dominant minima at their respective splits (which can easily be done in linear time).
 \end{proof}

### 2014-11-03 paint-contour-tree.tex

@@
 
-For convenience, imagine $f$ to be smooth. All points where the gradient of $f$ is zero are \emph{critical points},
-where the topology of the contours change. The vertices of the contour tree are exactly these critical points,
-representing the important topological events. An edge connects two critical points if one event immediately ``follows"
+If $f$ is smooth, all points where the gradient of $f$ is zero are \emph{critical points}.
+These points are the ``events" where the contour topology changes and form the vertices of the contour tree. 
+An edge connects two critical points if one event immediately ``follows"
 the other as the sweep plane makes its pass. (We provide formal definitions later.)
@@
 does not need this, but it simplifies the proof and presentation.} $f:\MM \mapsto \RR$, where the contour tree $T$
-has maximum degree $3$. There is an algorithm to compute $T$ whose running time is $O(\sum_{p \in P(T)} |p|\log |p| + t\alpha(t) + N)$,
+has maximum degree $3$. There is a deterministic algorithm to compute $T$ whose running time is $O(\sum_{p \in P(T)} |p|\log |p| + t\alpha(t) + N)$,
 where $P(T)$ is a specific path decomposition (constructed implicitly by the algorithm).
@@
 This is clearly an improvement when the tree is short and fat.
-In particular, for any decomposition of a rooted balanced binary tree, this gives $O(t\log\log n + t\alpha(t) + N)$, a significant improvement from $O(t\log t+N)$.
-However, our actual bound in this case gives an even further improved $O(t\alpha(t)+N)$.
+But it can be even better. A calculation yields that for any decomposition of a rooted balanced binary tree, 
+our bound is $O(t\alpha(t) + n)$. 
+%
+%
+%In particular, for any decomposition of a rooted balanced binary tree, this gives $O(t\log\log n + t\alpha(t) + N)$, a significant improvement from $O(t\log t+N)$.
+%However, our actual bound in this case gives an even further improved $O(t\alpha(t)+N)$.
 Consider the example of \Fig{sorting} extended to a larger binary tree both above and below.

### 2014-11-03 paint-contour-tree.tex
minor changes

--- 2014-11-03 paint-contour-tree.tex
@@
       University of Illinois; %
-      201 N. Goodwin Avenue; %
-      Urbana, IL, 61801, USA; %
+%      201 N. Goodwin Avenue; %
+      Urbana, IL, USA; %
       {\tt \si{raichel}2\atgen{}\si{uiuc}.\si{edu}}; %
@@
   \and
-  C. Seshadhri
+  C. Seshadhri\thanks{Sandia National Laboratories; Livermore, CA, USA; {\tt scomand@sandia.gov}}
 }
@@
 merge, and down-degree $2$ vertices are ``splits".  When $d=2$ (as in \Fig{relative}), contours are closed loops.
-In dimensions $d \geq 3$, we may have internal
-vertices of both up and down-degree $1$. They correspond to contours (which are $(d-1)$-manifolds) gaining or losing handles.
+In dimensions $d \geq 3$, we may have internal vertices of both up and down-degree $1$. 
+For example, these could be events where contours gain or lose handles.
 
@@
 that $C$ might be irrational, define valid if $2^C = 2^{\cost(f)}$.}. We prove
-a refined lower bound matching \Thm{main-alg}.
-
-\begin{definition}\label{def:hfdt}
-A \emph{height function decision tree} is an algorithm
-that is allowed arbitrary computation on the domain $\MM$, and is an algebraic decision
-tree over the heights. The running time is simply the number of queries made.
-\end{definition}
-
+a refined lower bound matching the performance of \Thm{main-alg}. 
+Roughly speaking, it is not possible to improve on \Thm{main-alg} even when restricted to a subset
+of inputs of fixed cost.
+
+%\begin{definition}\label{def:hfdt}
+%A \emph{height function decision tree} is an algorithm
+%that is allowed arbitrary computation on the domain $\MM$, and is an algebraic decision
+%tree over the heights. The running time is simply the number of queries made.
+%\end{definition}
+%
 \begin{theorem} \label{thm:main-lb} 
@@
 %where $t$ is the number of critical points and $C$ is a valid cost for an instance with $t$ critical point.
-Consider the set of functions $\bF_C$ that all same number of critical points and the same (valid) cost $C$.  Any algebraic decision tree that correctly computes
+Consider the set of functions $\bF_C$ that have the same number of critical points and the same (valid) cost $C$.  Any algebraic decision tree that correctly computes
 the contour tree on all of $\bF_C$ has a worst case running time of $\Omega(C)$.
@@
 
-Roughly speaking, it is not possible to improve on \Thm{main-alg} even when restricted to a subset
-of inputs of fixed cost.
-
 \paragraph{Beyond worst-case analysis:} Our result provides understanding of the complexity
@@
 that the cost (as defined earlier) roughly captures the hardness of an instance.
-In that sense, it is related to the concept of instance optimal algorithms, as defined by Afshani \etal \cite{abc-ioga-09}.
+Trees that are short and fast are easy to compute, whereas trees with long paths are hard.
+Our result is related to the concept of instance optimal algorithms, as defined by Afshani \etal \cite{abc-ioga-09}.
 Their notion of optimality is far more refined that what we prove, but it shares the flavor of understanding the spectrum
@@
 that our algorithm requires $\Theta(\cost(P(T)))$ on all these instances.
+
+\section{Acknowledgements}
+
+We thank Jeff Erickson and Yusu Wang for numerous useful discussions. This work is
+supported by the Laboratory Directed Research and Development
+(LDRD) program of Sandia National Laboratories. Sandia National
+Laboratories is a multi-program laboratory managed and operated
+by Sandia Corporation, a wholly owned subsidiary of Lockheed
+Martin Corporation, for the U.S. Department of Energy’s National
+Nuclear Security Administration under contract DE-AC04-94AL85000.
 
@@
 
-\Ben{This is already defined in section 3, should we use the same numbering?}
+%\Ben{This is already defined in section 3, should we use the same numbering?}
 \begin{definition} Let the 1-skeleton of $\MM$ have edge set $E$ and maxima $X$.
@@
 \label{sec:lb}
-In this section we provide a lower bound for the time it takes to compute a contour tree in terms the cost of path decompositions.  Specially, 
-we first consider the easier case of join trees, after which the lower bound for more general contour trees will easily follow.
-\Ben{Hopefully the following sentence clear up confusion between $\MM$ and $(f,\MM)$}
+In this section we provide a lower bound for the time it takes to compute a contour tree in terms the cost of path decompositions.  
+We first prove a lower bound for join trees, and then generalize to contour trees.
+%\Ben{Hopefully the following sentence clear up confusion between $\MM$ and $(f,\MM)$}
 Note earlier we usually used $\MM$ to denote the lift of a simplicial complex according to some height function $f$, 

### 2014-11-03 paint-contour-tree.tex

--- 2014-11-03 paint-contour-tree.tex
@@
 that our algorithm requires $\Theta(\cost(P(T)))$ comparisons on all these instances.
+
+\section{Acknowledgements}
+
+We thank Jeff Erickson and Yusu Wang for numerous useful discussions. This work is
+supported by the Laboratory Directed Research and Development
+(LDRD) program of Sandia National Laboratories. Sandia National
+Laboratories is a multi-program laboratory managed and operated
+by Sandia Corporation, a wholly owned subsidiary of Lockheed
+Martin Corporation, for the U.S. Department of Energy’s National
+Nuclear Security Administration under contract DE-AC04-94AL85000.
 

### 2014-11-04 paint-contour-tree.tex
more small changes

@@
 is inherent to computing contour trees. Consider the family of instances where our algorithm
-requires $C$ comparison. Any algorithm computing contour trees on this family requires
+requires $C$ comparisons. Any algorithm computing contour trees on this family requires
 $\Omega(C)$ comparisons in the worst case.
@@
 
-\paragraph{Organization.}  
-\Sec{basics} contains basic definition for contour trees.  In \Sec{approach} we give a high level overview and intuition for our approach 
-for computing contour trees.  \Sec{surgery} and \Sec{rain} describe how we break the intial input into nicely behaved (extemum dominant) pieces. 
-\Sec{extreme} shows that computing the contour tree for these nicely behaved pieces is roughly equivalent to computing the join tree. 
-\Sec{paint} contains our algorithm for computing join trees, which along with previous sections gives our algorithm for contour trees.  
-Finally, in \Sec{pathDecomp} and \Sec{lb} we relate the running time of our algorithm (providing upper and lower bounds) to a particular decomposition of the contour tree.
+% \paragraph{Organization.}  
+% \Sec{basics} contains basic definition for contour trees.  In \Sec{approach} we give a high level overview and intuition for our approach 
+% for computing contour trees.  \Sec{surgery} and \Sec{rain} describe how we break the intial input into nicely behaved (extemum dominant) pieces. 
+% \Sec{extreme} shows that computing the contour tree for these nicely behaved pieces is roughly equivalent to computing the join tree. 
+% \Sec{paint} contains our algorithm for computing join trees, which along with previous sections gives our algorithm for contour trees.  
+% Finally, in \Sec{pathDecomp} and \Sec{lb} we relate the running time of our algorithm (providing upper and lower bounds) to a particular decomposition of the contour tree.
 
@@
 When $d=2$, this distinction is not necessary, since any point with $>1$
-Morse degree will have some $>1$ degree in $\reeb(\MM)$.
+Morse degree will have $>1$ degree in $\reeb(\MM)$.
 
@@
     \includegraphics[width=.18\linewidth]{figs/tent1}
-    \caption{On left, downward rain spilling only (each shade of gray represents a different color), producing a grid.  
+    \caption{On left, downward rain spilling only (each shade of gray represents a piece created by each different spilling), producing a grid.  
     Note we are assuming raining was done in reverse sorted order of maxima.  
@@
 sorting all of them. It is absolutely necessary to cut $\MM$ into pieces where the cost
-of building the join tree can be related to the cost of building $\cC(\MM)$.
+of building the join tree can be related to that of building $\cC(\MM)$.
 
@@
 \paragraph{The running time analysis:} Relating the running time to a path decomposition is the most technical part of the paper. 
-All the non-heap operations can be easily bounded as $O(t\alpha(t) + N)$ ($t\alpha(t)$ from the union-find data structure
+All the non-heap operations can be easily bounded by $O(t\alpha(t) + N)$ (the $t\alpha(t)$ is from the union-find data structure
 for colors). It is not hard to argue that at all times, any heap always contains a subset of a leaf
-to root path. Unfortunately, this subset is \emph{not} contiguous, as \Fig{colors} shows. \Sesh{explain}
+to root path. Unfortunately, this subset is \emph{not} contiguous, as the right part of \Fig{colors} shows. 
+Specifically, in this figure the far left saddle (labeled $i$) is hit by blue paint.  However, there is another 
+saddle on the far right (labeled $j$) which is not hit by blue paint.  Since this far right saddle is slightly 
+higher than the far left one, it will merge into the component containing the blue mound (and also the yellow and red mounds)
+before the far left one.  Hence, the vertices initially touched by blue are not contiguous in the contour tree.
+
 Nonetheless, we can get a non-trivial (but non-optimal) bound. Let $d_v$ denote the distance to the root for vertex $v$ in the join tree.
@@
 functions on a fixed domain such that each function has a distinct (labeled) contour tree. By a simple
-entropy argument, any height decision tree that correctly computes the contour tree on all instances
-requires average case $\Omega(\cost(P(T)))$ time. We prove that our algorithm makes $\Theta(\cost(P(T)))$
-height comparisons on all these instances. In general, any algorithm requires $\Omega(C)$ time
+entropy argument, any algebraic decision tree that correctly computes the contour tree on all instances
+requires worst case $\Omega(\cost(P(T)))$ time. We prove that our algorithm makes $\Theta(\cost(P(T)))$ 
+comparisons on all these instances. In general, any algorithm requires $\Omega(C)$ time
 to solve the set of instances that our algorithm solves with $C$ comparisons.
@@
 
-\section{Acknowledgements}
-
-We thank Jeff Erickson and Yusu Wang for numerous useful discussions. This work is
+%\section{Acknowledgements}
+\paragraph{Acknowledgements.}
+We thank Hsien-Chih Chang, Jeff Erickson, and Yusu Wang for numerous useful discussions. This work is
 supported by the Laboratory Directed Research and Development
@@
 %end of writup
-%stuff below should be deleted
 %
@@
 
-\remove{
-
 \appendix
 
-\section{Old stuff}
-
-\subsubsection{Paying it down the line}
-
-% \begin{claim}
-% \label{clm:sizes}
-%  Let $p$ be a path in a maximum path decomposition, such that $|H_p|>c|p|$ for some constant $c> 6$.
-%  Let $H_f$ denote the heap at the leaf of $p$.  We then have the following
-%  \begin{compactenum}
-%   \item $|H_f|\leq (1+1/c)|H_p|$
-%   \item $|H_f|\geq |H_p|/3$
-%   \item For any path $q$ adjacent to $p$ in the shrub, $|H_q|<\frac{2}{3}|H_p|$
-%  \end{compactenum} 
-% \end{claim}
-% \begin{proof}
-%   For any vertex $v\in p$ \Obs{decrease} implies that $|H_v|\leq |H_p| + |p| < (1+1/c) |H_p|$.
-%   First observe that by setting $v=f$ this immediately implies the first part of the claim.
-%  % By a similar argument, $|H_f|\geq |H_p| - |p| > (1-1/c)|H_p|$.
-%   
-%   To prove the last part of the claim, suppose otherwise that $|H_q|\geq \frac{2}{3}|H_p|$ for some 
-%   adjacent path $q$.  As this is a maximum path decomposition, this implies that there is some 
-%   $v\in p$ such that $|H_v|\geq \frac{4}{3}|H_p|-1$.
-%   By \Obs{decrease} we then have, 
-%    $|H_p| \geq |H_v| - |p|\geq \frac{4}{3}|H_p| - 1 - |H_p|/c \geq (4/3-1/c)|H_p|-1$.
-%   When rearranged this gives $1\geq (1/3-1/c)|H_p|$.  As $c>6$ we have $1\geq |H_p|/6$, 
-%   which is a contradiction as $|H_p|>c|p|>6$.
-% \end{proof}
-
-\Ben{I had originally thought the following claim would be useful in proving the lemma below.}
-\begin{claim}
-\label{clm:sizes}
- Let $p$ be a path in a maximum path decomposition, such that $|H_p|>c|p|$ for some constant $c> 6$.
- We then have the following
-% \begin{compactenum}
-%  \item 
-  For any path $q$ adjacent to $p$ in the shrub, $|H_q|<\frac{2}{3}|H_p|$
-%  \item If there does not exist and adjacent path $q$ such $H_q\geq H_p/3$, then $|H_f|\geq |H_p|/3$
-% \end{compactenum} 
-\end{claim}
-\begin{proof}
-  To prove the claim, suppose otherwise that $|H_q|\geq \frac{2}{3}|H_p|$ for some 
-  adjacent path $q$.  As this is a maximum path decomposition, this implies that there is some 
-  $v\in p$ such that $|H_v|\geq \frac{4}{3}|H_p|-1$.
-  By \Obs{decrease} we then have, 
-   $|H_p| \geq |H_v| - |p|\geq \frac{4}{3}|H_p| - 1 - |H_p|/c \geq (4/3-1/c)|H_p|-1$.
-  When rearranged this gives $1\geq (1/3-1/c)|H_p|$.  As $c>6$ we have $1\geq |H_p|/6$, 
-  which is a contradiction as $|H_p|>c|p|>6$.  
-\end{proof}
-
-\begin{lemma}
-\label{lem:excess}
- Let $p$ be a tall path, and let $S_1, \dots S_k$ and $R_1, \dots, R_l$ be the sizes of the heaps at the roots of the short and tall children of $p$, respectively, in the shrub.  
- Let $H_f$ be the heap at the leaf of the path $p$.
- Then,  $|H_p| \log |H_p|   \geq  \sum_{v\in p} \log|H_v| + \frac{4}{3}\sum_{i=1}^k S_i \log S_i + \sum_{i=1}^l R_i \log R_i  -  c|p|\log |p|$, for some sufficiently large constant $c$
- \end{lemma}
-\begin{proof}
- First observe that the lemma holds for $|H_p|$ less than any fixed constant, since otherwise \Lem{adjacent} implies we can choose $c$ 
- large enough such that $c|p|\log|p|$ will dominate all other terms.
- Next, observe that $\sum_{v\in p} \log|H_v| \leq \sum_{v\in p} \log(|H_v|+|p|) = |p|\log(|H_v|+|p|) = O(|p|\log|p|)$, as $p$ is a tall path.  
- Therefore, in the remainder of the proof  we can safely ignore the $\sum_{v\in p} \log|H_v|$ term as we can choose $c$ large enough to cover it.  
- That is we now instead prove the inequality,
- \[
-  |H_p| \log |H_p|   \geq \frac{4}{3}\sum_{i=1}^k S_i \log S_i + \sum_{i=1}^l R_i \log R_i  -  c|p|\log |p|
- \]
- To do so we  break the analysis into two case based on the relative size of $|p|$ and $|H_p|$. 
- In the following, $\alpha>10$ is some yet to be determined constant.
-
-\paragraph{Case 1, $|p|\geq |H_p|/\alpha$:\\}
-First observe that by \Lem{adjacent}, $|H_p|+|p| \geq \sum_{i=1}^k S_i+ \sum_{i=1}^l R_i$.
-By choosing $c$ large enough we therefore have,
-\begin{align*}
- c|p|\log|p| &\geq 2(1+\alpha)^2 |p|\log |p| \geq 2(1+\alpha)|p| \log((1+\alpha)|p|)\\
-             &\geq 2(|H_p|+|p|)\log(|H_p|+|p|) \geq 2\log(|H_p|+|p|) \pth{\sum_{i=1}^k S_i + \sum_{i=1}^l R_i}\\
-             &\geq \frac{4}{3}\sum_{i=1}^k S_i \log S_i + \sum_{i=1}^l R_i \log R_i 
-\end{align*}
-
-In this case $\log(|H_p|+|p|)\leq \log(2|p|) \leq 1+\log(|p|)$ (where $\log$ is base $2$).
-Therefore, assuming $|p|>1$, we have 
-\[
-(|H_p|+|p|)\log(|H_p|+|p|)\leq (1+c)|p|\log((1+c)|p|) %(|H_p|+|p|)(1+\log |p|) \leq 2(1+c)|p|\log |p|
-\]
-which when combined with the inequality from \Lem{adjacent} gives, 
-\[
- 4(1+c)|p|\log |p| \geq 2(|H_p|+|p|)\log(|H_p|+|p|)\geq \pth{|H_f| + \frac{4}{3}\sum_{i=1}^k S_i + \sum_{i=1}^l R_i}\log(|H_p|+|p|).
-\]
-This easily implies the desired inequality as $\log(H_p+p)$ is larger than any other $\log$ term in the inequality.
-
-\paragraph{Case 2, $|p|< |H_p|/\alpha$:\\}
-
-\Ben{Was not able to handle this case.}
-
-%  First, lets assume we can show $\sum_{i=1}^k S_i + \sum_{i=1}^l R_i \leq \frac{3}{4}|H_p|$.  
-%  In this case, using the last part of \Clm{sizes}, we can prove the desired inequality as follows.
-%  
-%  \begin{align*}
-%   \frac{4}{3}\sum_{i=1}^k S_i \log S_i + \sum_{i=1}^l R_i \log R_i  &\leq  \frac{4}{3} \pth{\sum_{i=1}^k S_i \log S_i + \sum_{i=1}^l R_i \log R_i} \\
-%   &\leq  \frac{4}{3} \log |H_p| \pth{\sum_{i=1}^k S_i + \sum_{i=1}^l R_i}  \leq  |H_p| \log |H_p|
-%  \end{align*}
-%  
-%  Finally, to complete the proof we must prove our assumption that $\sum_{i=1}^k S_i + \sum_{i=1}^l R_i \leq \frac{3}{4}|H_p|$.
-%  By \Lem{adjacent} and \Clm{sizes} we have, 
-%  \[
-%  \sum_{i=1}^k S_i + \sum_{i=1}^l R_i 
-%  \leq |H_p|+|p|-|H_f| 
-%  \leq (1+1/\alpha-1/3)|H_p| = (2/3+1/\alpha)|H_p|
-%  \]
-%  
-%  Therefore $\sum_{i=1}^k S_i + \sum_{i=1}^l R_i \leq \frac{3}{4}|H_p|$ if $(2/3 + 1/\alpha)\leq 3/4$, which rearranged gives $\alpha \geq 12$
-\end{proof}
-
-\remove{
-% \subsubsection{Paying it down the line}
-% 
-% \begin{lemma}
-% \label{lem:excess}
-%  Let $p$ be a tall path, and let $S_1, \dots S_k$ and $R_1, \dots, R_l$ be the sizes of the heaps at the roots of the short and tall children of $p$, respectively, in the shrub.  
-%  Let $H_f$ be the size of the heap at the leaf of the path $p$, and let $X = \sum_{i=0}^k H_i \log H_i$.  
-%  Then,  $|H_p| \log |H_p|   \geq  H_f\log H_f + \sum_{i=0}^k S_i \log S_i + \sum_{i=0}^l R_i \log R_i  -  c|p|\log |p|$, for some sufficiently large constant $c$
-%  \end{lemma}
-% \begin{proof}
-% By \Lem{adjacent} we know that $|H_p|+|p| \geq  H_f + \sum_{i=0}^k S_i + \sum_{i=0}^l R_i$.  
-% %Now as $p$ is a tall path we have $|p|\geq \sqrt{|H_p|}/100$, and so, $(|H_p|+|p|)\log(|H_p|+|p|)$
-% 
-% We  break the analysis into two case based on the relative size of $|p|$ and $|H_p|$.  In the following $c$ is some yet to be determined constant greater than $1$.
-% 
-% \paragraph{Case 1, $|p|\geq |H_p|/c$:}  
-% In this case $\log(|H_p|+|p|)\leq \log(2|p|) \leq 1+\log(|p|)$ (where $\log$ is base $2$).
-% Therefore, assuming $|p|>1$, we have 
-% \[
-% (|H_p|+|p|)\log(H_p+p)\leq (|H_p|+|p|)(1+\log |p|) \leq (2|p|)(2\log |p|)\leq 4|p|\log |p|
-% \]
-% which when combined with the inequality from \Lem{adjacent} gives, 
-% \[
-%  4|p|\log |p| \geq (|H_p|+|p|)\log(|H_p|+|p|)\geq \pth{H_f + \sum_{i=0}^k S_i + \sum_{i=0}^l R_i}\log(|H_p|+|p|).
-% \]
-% This easily implies the claim as $\log(H_p+p)$ is larger than any other $\log$ term in the inequality.
-% 
-% \paragraph{Case 2, $|p|< |H_p|/c$:}
-% \end{proof}
-}
-
-Define the crust of the shrub to be the first (i.e. closest to the root) layer of short paths in the shrub.  Apply \Lem{shortCost} to each of the short paths in the crust and then 
-say by \Lem{excess} and since the root heap goes to zero we are done.
-
-\ignore{
-Naturally, this geometric series behavior can be extended to more than two adjacent short paths, thus motivating 
-the following definition.
-
-\begin{definition}
- Let $p$ be a path in a path decomposition $P(T)$, with corresponding shrub $\pathTree$.
- Call a short path $q\in P(T)$ short reachable from $p$ if the path from $p$ to $q$ in $\pathTree$ 
- contains only vertices which correspond to short paths.  Call the subtree of $\pathTree$ consisting 
- of all short reachable paths from $p$ the \emph{short shrub} of $p$, and denote it $\mathcal{L}(p)$.
-\end{definition}
-
-In the following subsection we show how to exploit this geometric series behavior of short shrubs.
-
-We say two paths in $P(T)$ are adjacent if their corresponding vertices are adjacent in the shrub.  Similarly, for 
-two adjacent paths in $P(T)$ we can define a parent/child relationship based on this relationship in the shrub. 
-}
-
-\remove{
-\Ben{Maybe move the remainder of this subsection to 5.3 since tall and short paths and not used until then}
-
-\begin{definition}
- Let T be a binary tree and let $\chi(T)$ and $P(T)$ be a valid coloring and path decomposition of $T$, respectively.  
- For $p\in P(T)$, we say that $p$ is \emph{short} if $|p| < \sqrt{|H_{r_p}|}/100$, and we say it is \emph{tall} otherwise.
-\end{definition}
-
-\begin{observation}
-\label{obs:decrease}
- Let $v$ be a vertex in $T$ and let $w$ be its parent.  Then $|H_w| \geq |H_v| -1$, as $L(T_v) \subseteq L(T_w)$ and the path from 
- $w$ to the root has one less vertex than the path from $v$ to the root.
-
- In particular, we have the following more general property.
- Let $v$ and $u$ be any two vertices in the same root to leaf path of $T$, such that $u$ is lower than $v$.  
- Then $|H_v| \leq |H_u| + d(u,v)$. 
-\end{observation}
-
-\begin{lemma}
-\label{lem:pathBounds}
- Let T be a binary tree and let $\chi(T)$ and $P(T)$ be a valid coloring and path decomposition of $T$, respectively.
- Then, 
- \begin{compactenum}
-  \item for any tall path $p\in P(T)$, $\sum_{v\in p} \log |H_v| = O(|p| \log |p|)$.
-  \item for any short path $p\in P(T)$, $\sum_{v\in p} \log |H_v| = O(|p| \log |H_{r_p}|)$.
- \end{compactenum}
-\end{lemma}
-\begin{proof}
- By $\Obs{decrease}$ we know that for any vertex $v\in p$, $|H_v|\leq |H_{r_p}| + |p|$ (as $r_p$ lies below $v$ on $p$).
- As such, 
- \[
- \sum_{v\in p} \log |H_v| \leq \sum_{v\in p} \log(|H_{r_p}| + |p|)  = |p| \log (|H_{r_p}| + |p|).
- \]
- If $p$ is a tall path then $|p| \log (|H_{r_p}| + |p|) = O(|p| \log |p|)$, and if $p$ is short then 
- $|p| \log (|H_{r_p}| + |p|) = O(|p| \log |H_{r_p}| )$.
-\end{proof}
-
-The following lemma can be thought of as a generalization of $\Obs{decrease}$.
-
-\begin{lemma}
-\label{lem:adjacent}
- Let $T$ be a binary tree and let $\chi(T)$ and $P(T)$ be a valid coloring and path decomposition of $T$, respectively.
- Let $p$ be any path in $P(T)$ and let $\{q_1, \dots q_k\}$ be the set of all adjacent child paths of $p$.
- For $1\leq i\leq k$ let $r_i = r_{q_i}$ and let $H_i = H_{r_i}$.
- Also, let $H_0$ denote the heap at the leaf of $p$.
-
- Then $\sum_{i=0}^k |H_i| \leq |H_p|+|p|$, where $H_p = H_{r_p}$.
-\end{lemma}
-\begin{proof} 
- First observe that for all $i$, $r_{i}$ lies above $r_p$ on some root to leaf path. 
- Let $H_i'$ be the subset of $H_i$ that lies below $r_p$.  Observe that $H_i'\subseteq H_p$, as $L(T_{r_i})\subseteq L(T_{r_p})$.
-
- Now for any $i\neq j$, $H_i \cap H_j = \emptyset$, as $r_{i}$ and $r_{j}$ are not on the same root to leaf path (i.e. $L(T_{r_i}) \cap L(T_{r_j}) = \emptyset$).
- In particular, $(H_i\setminus H_i') \cap (H_j \setminus H_j') = \emptyset$, and so $\sum_{i=0}^k |H_i\setminus H_i'| \leq |p|$ 
- as $H_i\setminus H_i'$ is precisely the subset of $H_i$ that lies on $p$.
- Moreover, for any $i\neq j$, $ H_i' \cap H_j' = \emptyset$, and so since $H_i'\subseteq H_p$ we have $\sum_{i=0}^k |H_i'|\leq |H_p|$.
- Putting these together gives
- \[
-  \sum_{i=0}^k |H_i| = \sum_{i=0}^k |H_i'| + \sum_{i=0}^k |H_i\setminus H_i'| \leq |H_p|+|p|.
- \]
-\end{proof}
-
-% \Ben{There is a fundamental flaw in the following lemma and corollary.  
-% As such, the remainder of this subsection will likely be deleted.}
-% 
-% Using the notation defined in the above lemma, we have the following crucial lemma, which we will use to 
-% argue we can charge that cost of all paths in $P(T)$ to the cost of the tall paths.
-% 
-% \begin{lemma}
-% Let $p$ be a tall path in $P(T)$ and let $\{q_1, \dots q_k\}$ be any set of short paths in $P(T)$ such that 
-% $\sum_{i=1}^k |H_i| = O(H_p)$.  Then $\sum_{i=1}^k \sum_{v\in q_i} \log |H_v| = O(|p|\log |p|)$
-% \end{lemma}
-% \begin{proof}
-% By \Lem{pathBounds}, for all $1\leq i\leq k$, 
-%  $\sum_{v\in q_i} \log |H_v| = O(|q_i|\log |H_i|) = O(|H_i|\log |H_i|)$, as $q_i$ is a short path.
-%  By assumption $\sum_{i=1}^k |H_i| = O(|H_p|)$, 
-%  As $x\log x$ is a convex function, Jensen's inequality then implies 
-%  $\sum_{i=1}^{k} |H_i|\log |H_i| = O(|H_p| \log |H_p|) = O(|p|\log |p|)$, since $p$ is a tall path.
-% \end{proof}
-% 
-% 
-% \begin{corollary}
-% \label{cor:tallLight}
-%  Let $p$ be a tall path in $P(T)$ and let $\{q_1, \dots q_k\}$ be the set of all short paths adjacent to $p$.
-%  Then $\sum_{i=1}^k \sum_{v\in q_i} \log |H_v| = O(|p|\log |p|)$.
-% \end{corollary}
-% \begin{proof}
-%  Let $H_i$ be as defined in \Lem{adjacent}.  By \Lem{pathBounds}, for all $1\leq i\leq k$, 
-%  $\sum_{v\in q_i} \log |H_v| = O(|q_i|\log |H_i|) = O(|H_i|\log |H_i|)$, as $q_i$ is a short path.
-%  Now by \Lem{adjacent}, $\sum_{i=1}^k |H_i| \leq |H_p|+|p| = O(|p|)$, as $p$ is a tall path. 
-%  Therefore, as $x\log x$ is a convex function, Jensen's inequality implies the claim.
-%  
-% \end{proof}
-
-}
-
-\subsection{The shrub of unsupported paths} \label{sec:short}
-
-It suffices to focus on unsupported short paths. \Lem{short} can be reduced to the following.
-
-\begin{lemma} \label{lem:support} $\sum_{p \in \cL \setminus \cL'} \sum_{v \in p} \log |H_v| = O(\sum_{p \in \cL'} |H_p| \log |H_p|)$.
-\end{lemma}
-
-We now construct the shrub of unsupported short paths. Consider $p \in \cL \setminus \cL'$,
-and traverse the chain of ancestors from $p$. Eventually, we must reach another short path $q$.
-(If not, we have reached the root $r$ of $\pmax(T)$. Hence, $p$ is supported.)
-Insert edge from $p$ to $q$, so $q$ is the parent of $p$ in $\cU$. This construction leads
-to the shrub forest of $\cL \setminus \cL'$, where every unsupported short path is an internal node, and all
-roots are supported short paths.
-
-We list out some important claims about this shrub forest.
-We focus on $\cU$, a connected component (shrub) in the shrub forest and denote the 
-the root by $r(\cU)$.
-
-\begin{claim} \label{clm:loss2} Consider $p \in \cL \setminus \cL'$ that is in $\cU$. Let $Q$
-be the set of tall ancestors of $p$ that are descendants of $r(\cU)$. Then $\sum_{q \in Q} |q \cap H_p| \leq |H_p|/50$.
-\end{claim}
-
-\begin{proof} Suppose not, for contradictions sake. As we traverse the tall ancestors of $p$,
-we encounter various vertices of $H_p$ in the ancestors $q$. Consider the first $|H_p|/100$
-such vertices of $p$ encountered. At each such $v$, there are at least $|H_p|/100$
-vertices \emph{not yet} encountered, so $|H_v| > |H_p|/100$. Each of these vertices $v$
-is supported, implying that $p$ is a supported short path.
-\end{proof}
-
-\Lem{support} is a direct consequence of the following,
-which is the really the main lemma.
-
-%\begin{lemma} \label{lem:cu} Let $\cU$ denote a connected component (shrub) in the shrub forest of $\cL \setminus \cL'$ and let $r$
-%be the root of $\cU$. Then $\sum_{p \in \cU} \sum_{v \in p} \log |H_v| = O(|H_r| \log |H_r|)$.
-%\end{lemma}
-%
-%We proceed with the setup to prove \Lem{cu}.
-%We focus our attention on some shrub $\cU$.
-%We prove a technical lemma about $\cU$, from which \Lem{cu} follows directly.
-
-\section{Ben's old stuff}
-
-\begin{proof} (part (ii) of \Lem{cu-root}) Let level $\ell$ of $\cU$ denote all nodes at distance $\ell$ from the root $r$.
-These set of these nodes is designated as $\cU_\ell$. We set $\sigma_\ell = \sum_{p \in \cU_\ell} |p|$
-and $\sigma_{\leq \ell} = \sum_{k \leq \ell} \sigma_k$.
-By \Lem{geometric}, for any $p \in \cU_\ell$, $|H_p| \leq (2/3)^\ell |H_r|$.
-
-Since all nodes in $\cU$ represent short paths, $\forall p \in \cU$, $|p| < \sqrt{|H_p|}/100$.
-So $\sigma_\ell < \sum_{p \in \cU_\ell} \sqrt{|H_p|}/100$.
-
-\end{proof}
-
-To prove this, we need to understand how $H_p$ is constructed. The path $p$ has some children
-$q_1, q_2, \ldots, q_k$ in $\pmax(T)$. Furthermore, as a leaf path in $p$, some leaf
-$\ell$ of $T$ in one of the ends of $p$. The heap $H_p$ is essentially the
-union of $H_\ell$ and $H_{q_1}, H_{q_2}, \ldots, H_{q_k}$, with the vertices in $p$
-removed. Note that some of the $q_i$'s are the corresponding $h(q')$ vertices,
-for a child $q'$ in $\cU$.
-
-Let $\mathcal{L} = \mathcal{L}(r)$ be a short shrub of some short path $r$ in a maximum path decomposition $\mathcal{P}(T)$.
-The intuitive idea is that due to the geometric decreasing behavior of short shrubs described in the previous subsection, 
-one can show that for the entire short shrub our algorithm pays roughly the cost due to the root path $r$.  This argument 
-is hard to make directly, so instead we define an alternative quantity for which we can make the argument.
-Specifically, we now define a weight function $w(v)$ for all $v\in \mathcal{L}$, and show that $w(v)=\Theta(|H_v|)$.
-
-For a path $v\in \mathcal{L}$, let $H_v$ denote the heap at the root of $v$, and let $H_{v_f}$ denote the heap 
-at the leaf of $v$.  Let $S_l(v)$ (resp. $S_h(v)$) be the subset of short (resp. tall) paths adjacent to $v$ in $\pathTreeA(T)$.
-
-We want to understand how the heap size changes as we move around the paths in $\mathcal{L}$.  Specifically, there are 
-three places the vertices in the heap along a short path can come from; adjacent short paths, adjacent tall paths, and the initial heap at the leaf of the path.  
-Given this observation we now define a weight function which upper bounds the size of the heap at the root of any path in $\mathcal{L}$.
-
-First, for all $v\in \mathcal{L}$ we define the local cost function $c(v) = H_{v_f}+\sum_{i\in S_h(v)} H_i$, this is the addition to the heap at $v$ which 
-was not already charged to descendant short paths.  We now define a weighted version of $\mathcal{L}$ based on these costs.  
-To do so, we first augment $\mathcal{L}$ by attaching one 
-leaf to each vertex in $\mathcal{L}$.  Now for each newly created leaf (which are the only leaves now) assign the cost 
-$c(v)$ defined above, where $v$ is that leaf's corresponding parent.  
-Now we can define a weight function $w(u)$ over any node $u$ in the tree, which is defined to be the total cost of all leaves in its subtree.   
-To simplify notation, we now use $\mathcal{L}$ to refer to this new weighted tree (rather than its unweighted counterpart).
-
-As the weight $w(p)$ includes the sizes of the heaps of all contributing tall paths and leaves in the subtree rooted at $p$, we have the following.
-
-\begin{claim}
-For any $p\in \mathcal{L}$, $w(p)\geq |H_p|$.  Moreover, for any $v\in p$, $w(p)\geq |H_v|$.
-\end{claim}
-
-%Let $W= \sum_{v\in \mathcal{L}} w(v) \geq \sum_{v\in \mathcal{L}} |H_v|$.  
-%Suppose we could show that conversely $W=O(H_r)$.  Then we would have 
-%$\sum_{v\in \mathcal{L}} H_v = O(H_r)$, which was our original goal.  
-%So proving $W=O(H_r)$ is now our new goal.
-
-We now would like to show conversely that $w(v)= O(|H_v|)$.  
-To do so we first introduce some notation.  
-Let $C\subseteq \mathcal{P}$ be the subset of paths that are leaves in $\mathcal{L}$.
-For any path $v$, let $\mathcal{L}_v$ (resp. $C_v$) be the subset of paths in $\mathcal{L}$ (resp $C$) 
-that are in the subtree of $\mathcal{L}$ which is rooted at $v$.
-
-First, observe that $|H_v|\geq w(v)-\sum_{u\in \mathcal{L}_v} |u|$.  
-This holds by the same argument as in \Lem{adjacent}, namely, $w(v)$ by definition 
-counts all the contributing tall path heaps and initial leaf heaps that contribute 
-to the heap at $v$ and $\sum_{u\in \mathcal{L}_v} |u|$ is an upper bound on the number 
-of vertices that could have been removed from one of these heaps before reaching the 
-root of $v$ (since a vertex in a heap can only get removed if it appears on one of 
-the paths).
-
-As all paths in $\mathcal{L}$ are short, we thus have the following.
-
-\begin{claim}
-$|H_v|\geq w(v)-\sum_{u\in \mathcal{L}_v} \sqrt{|H_u|}/100 \geq w(v)-\sum_{u\in \mathcal{L}_v} \sqrt{w(u)}/100$
-\end{claim}
-
-By the above claim, if we can prove $\sum_{u\in \mathcal{L}_v} \sqrt{w(u)}/100 \leq w(v)/c$ for some constant $c>1$, 
-then we will have proven that $|H_v| = \Theta(w(v))$.
-
-Recall that \Lem{geometric} told us that that $|H_{u}|\leq (3/5) |H_{u'}|$, if $u'$ is the parent $u$ in $\mathcal{L}$. 
-The following lemma shows that if we had a similar statement in terms of the weights rather than the heap sizes, 
-then we get the desired bound on the sum.
-
-\begin{lemma}
-\label{lem:distribution}
-Let $v$ be a path in $\mathcal{L}$.  Suppose that for any two paths $x,y\in \mathcal{L}_v$ 
-such that $y$ is the parent of $x$, $w(x)\leq (4/5) w(y)$.
-Then $\sum_{u\in \mathcal{P}_v} \sqrt{w(u)}/100 \leq \sum_{c\in C_v} \sqrt{w(c)}/10$
-\end{lemma}
-
-\begin{proof}
-The idea is to recursively redistribute the contribution to the sum of each internal vertex to its children, 
-until all the weight sits on the leaves.  Specifically, let $v$ be some path in $\mathcal{L}$.  Its 
-contribution to the sum is $\sqrt{w(v)}/100$, which we will call the value at $v$.  We redistribute this 
-value to its children, in proportion by the weight of each child.  For example, if $v$ has a child $x$ then 
-the fraction it gets assigned is $w(x)/w(v)$ (recall that by definition $w(v)$ is the sum of the weights of 
-its children and so the total value will get reassigned to the children).
-
-Suppose we keep redistributing the value of $v$ until it reaches the leaves of its subtree in $\mathcal{L}$.  
-Let's see how much gets assigned to a given leaf.  Specifically, let $u_0, u_1, \ldots, u_k$ 
-be a path in $\mathcal{L}$ (of paths from the set $\mathcal{P}_v$), such that $v=u_0$, $u_{i+1}$ 
-is child of $u_i$ for all $i$, and $u_k$ is a leaf of $\mathcal{L}$.
-
-Then the total value from $u_0$ that ends up at $u_k$ is 
-\[
-\frac{\sqrt{w(u_0)}}{100} \pth{ \frac{w(u_1)}{w(u_0)} \frac{w(u_2)}{w(u_1)} \cdots \frac{w(u_k)}{w(u_{k-1})} }
-= \frac{\sqrt{w(u_0)}}{100} \pth{ \frac{w(u_k)}{w(u_0)} } = \frac{w(u_k)}{100\sqrt{w(u_0)}}
-\]
-
-For a given leaf in $\mathcal{L}$, we can calculate the total value reassigned to it from it its ancestors.
-Specifically, let $u_0, u_1, u_2, \dots, u_m$ be a root to leaf path in $\mathcal{L}$ 
-(i.e. $u_0$ is the root $r$ of $\mathcal{L}$).
-
-By assumption from the lemma statement we have $w(u_{i+1})\leq (4/5) w(u_i)$.  Therefore by the above calculation, 
-the total reassigned to $u_m$ is
-
-\[
- \sum_{i=0}^m \frac{w(u_m)}{100\sqrt{w(u_i)}} \leq \frac{w(u_m)}{100} \sum_{i=0}^m 1 \Big/ \sqrt{\pth{\frac{5}{4}}^iw(u_m)} 
- \leq \frac{\sqrt{w(u_m)}}{100} \sum_{i=0}^{\infty} \pth{\frac{4}{5}}^{i/2} \leq \frac{\sqrt{w(u_m)}}{10}
-\]
-
-The claim then follows by summing this quantity over the children of $\mathcal{L}$.
-\end{proof}
-
-By combining the above lemma and claim we have the following.
-
-\begin{claim}
-For any path $v\in \mathcal{L}$ if it holds that 
-$w(x)\leq (4/5) w(y)$ for any two paths $x,y\in \mathcal{L}_v$ such that $y$ is the parent of $x$, then
-
-\[
-|H_v|\geq w(v)-\sum_{u\in \mathcal{L}_v} \sqrt{w(u)}/100 \geq w(v)- \sum_{c\in C_v} \sqrt{w(c)}/10 \geq \frac{4}{5} w(v)
-\]
-\end{claim}
-
-\begin{lemma}
-Let $\mathcal{L}$ be the short shrub of some path $r$.  
-Then for any $v\in \mathcal{L}$, we have $|H_v|\geq \frac{4}{5} w(v)$.
-\end{lemma}
-% \begin{proof}
-% We will actually prove the stronger statement that for any $v\in \mathcal{L}$, 
-% we have $|H_v|\geq \frac{4}{5} w(v)$ and $w(v)\leq (4/5) w(p)$ where $p$ is the parent 
-% of $v$ in $\mathcal{L}$.
-% 
-% Suppose otherwise, and let $v$ be the lowest path in $\mathcal{L}$ such that this fails 
-% (i.e. the statement holds for all descendants of $v$ in $\mathcal{L}$).  Then for any 
-% two paths $x,y\in \mathcal{L}_v$ such that $y$ is the parent of $x$, 
-% we have that $w(x)\leq (4/5) w(y)$ and so by \Lem{distribution}, $|H_v|\geq \frac{4}{5} w(v)$.
-% Combining this fact with \Lem{geometric} gives,
-% \[
-% w(v) \leq \frac{5}{4}|H_v| \leq \frac{5}{4} \frac{3}{5} |H_p| \leq \frac{3}{4} w(p).
-% \]
-% 
-% \end{proof}
-%
-\begin{proof}
-By the above claim, we know that $|H_v|\geq \frac{4}{5} w(v)$ if it holds that 
-$w(x)\leq (4/5) w(y)$ for any two paths $x,y\in \mathcal{L}_v$ such that $y$ is the parent of $x$.
-So let $v$ be the lowest path in $\mathcal{L}$ such that this fractional decrease fails to hold, i.e. $w(c)> (4/5) w(v)$ 
-where $c$ is a child of $v$ in $\mathcal{L}$, and conversely all vertices in $\mathcal{L}_c$ 
-the fractional decrease holds.  As the fractional decrease holds for a paths in $\mathcal{L}_c$, 
-we have $|H_c|\geq \frac{4}{5} w(c)$.  Therefore,
-\[
-w(c) \leq \frac{5}{4}|H_c| \leq \frac{5}{4} \frac{3}{5} |H_v| \leq \frac{3}{4} w(v).
-\]
-However, this is a contradiction with the assumption $w(c)> (4/5) w(v)$ , and hence the 
-fractional decrease always holds implying the claim always holds.
-\end{proof}
-
-\begin{corollary}
-\label{cor:equivalent}
- Let $\mathcal{L}$ be a short shrub of a short path $r$.  Then for any $v\in \mathcal{L}$, $|H_v| \leq w(v) \leq \frac{5}{4} |H_v|$.
-\end{corollary}
-
-Now we are finally ready to make use of the properties we have proven about the weight function we introduced.
-
-\Ben{The constant in the lemmas below, but it is no more than twice that.}
-
-\begin{lemma}
-\label{lem:shortCost}
- Let $\mathcal{L}$ be a short shrub of a short path $r$.  Let $V$ be the set of tall paths from the full shrub which are adjacent to paths in $\mathcal{L}$.  
-  Then $\sum_{p\in \mathcal{L}}\sum_{v\in p} \log |H_v| + \sum_{q\in V} |H_q|\log |H_q| \leq \frac{4}{3} |H_r|\log |H_r|$
-\end{lemma}
-\begin{proof}
-\Ben{
-The following is a proof for $\frac{3}{2}+\eps$, rather than $\frac{4}{3}$.  The discrepancy is due to a corrected calculation error.  
-4/3 can be achieved by tightening constants above, and is left in the lemma statement as it is used later.
-}
-
-Recall that in a proper coloring of a binary tree $T$, the color of any vertex must be the color of one of its descendants.  
-Therefore, any vertex in the short path shrub must appear in the heap of either the root of an adjacent 
-path from the full shrub or from the heap of a leaf of a path in $\mathcal{L}$.  
-Also note $|H_r|$ precisely counts sum of the number of vertices from these heaps that do not appear in $\mathcal{L}$.
-As there are no adjacent short paths to a short shrub, the total number of vertices over all paths in $\mathcal{L}$ is at most 
-$m = \sum_{p\in \mathcal{L}} |H_{p_f}| +\sum_{q\in V} |H_q| - |H_r| = w(r)-|H_r|$, where $H_{p_f}$ is the heap at the leaf of $p$.
-
-Using this observation and \Cor{equivalent} we then have, 
-\begin{align*}
- \sum_{p\in \mathcal{L}}\sum_{v\in p} \log |H_v| + \sum_{q\in V} |H_q|\log |H_q|
- &\leq \sum_{p\in \mathcal{L}}\sum_{v\in p} \log(w(r)) + \sum_{q\in V} |H_q|\log(w(r))\\
- &\leq \log(w(r))\pth{ w(r)-|H_r| } + \log(w(r))w(r)\\ 
- &= (2w(r)-|H_r|)\log(w(r)) \leq \pth{\frac{5}{2}|H_r|-|H_r|}\log\pth{\frac{5}{4}|H_r|}\\
- &= \frac{3}{2}|H_r|\log\pth{\frac{5}{4}|H_r|}
-\end{align*}
-\end{proof}
-
-\section{Preliminaries}
-Consider a piecewise-linear (PL) $2$-manifold $\MM$ with boundaries.
-This is represented by a triangulation (given as a DCEL) where edges
-can be oriented and the boundary of a face is always counterclockwise. Some faces will \emph{not}
-be triangular, and each of these corresponds to a \emph{boundary face}.
-Let $G(\MM) = (V,E)$ denote this triangulation. We assume this triangulation
-has bounded degree.
-We will assume that $\MM$ is embedding in $\RR^d$.
-The manifold within each face is just linear.
-We define a \emph{height function} on $\MM$, $f:\MM \mapsto \RR$. For each point in $x \in \MM$, let $f(x)$ simply be the value of $x_1$.
-We distinguish between vertices and points in $\MM$. A point simply denotes any $x \in \MM$. A vertex is a point
-corresponding to some vertex of the triangulation $G(\MM)$.
-
-We define the size of the manifold $\MM$, denoted $|\MM|$, to be the number of faces, edges, and vertices.  
-Let $n$ denote the number of vertices.  
-Since we are working with $2$-manifolds, the size of the manifold is $\Theta(n)$.
-
-For smooth manifolds the ``status" of an internal point is decided by the gradient at that point. 
-Points of zero gradient are deemed \emph{critical};
-all other points are regular.
-For PL manifolds, we need a discrete version of this definition.
-First, all non-vertex points are considered regular.
-For vertex $v$, we use $\nbd(v)$ to denote the neighborhood of $v$ in $G(\MM)$. Let $\outnbd(v) := \{ w | f(v) > f(w), w \in \Gamma(v)\}$
-and $\innbd(v) := \{w | f(v) < f(w), w \in \Gamma(v)\}$. 
-%It is convenient to define $D(\MM)$, a directed variant of $G(\MM)$,
-%where edges are directed from higher to lower $f$-value. So $\outnbd(v)$ and $\innbd(v)$ are exactly the out- and in-neighborhoods
-%of $v$. A vertex in $v \in V$ can be classified in four types.
-\smallskip
-\begin{asparaenum}
-	\item Regular: Both $\outnbd(v)$ and $\innbd(v)$ are non-empty and are contiguous in $\nbd(v)$.
-	\item Maxima: $\nbd(v) = \outnbd(v)$.
-	\item Minima: $\nbd(v) = \innbd(v)$.
-	\item Saddle: Both $\outnbd(v)$ and $\innbd(v)$ are not contiguous in $\nbd(v)$.
-\end{asparaenum}
-\smallskip
-%Because $\MM$ is PL, the gradient at all non-vertex points of $\MM$ is zero.
-%So all non-vertex points are regular.
-
-We will assume that $f$ is Morse, so the function value on all critical points is distinct. 
-We will also assume the stronger condition that all internal vertices have distinct function values.
-Note that assuming $f$ is Morse implies that $\MM$ has no multi-saddles 
-(i.e. $\outnbd(v)$ and $\innbd(v)$ can each have at most two non-contiguous components).
-A value $t$ is \emph{non-critical} if there exists no critical point $x$ such that $f(x) = t$.
-
-Furthermore, the boundary vertices have special properties, encapsulated in the following definition.
-
-\begin{definition} \label{def:bound} A manifold is \emph{boundary critical} if the following holds.
-Consider a boundary face $F$. All vertices in $F$ have the same function value. Furthermore, all
-internal neighbors of $F$ either have function values strictly greater (or strictly less)
-than this value.
-\end{definition}
-
-In other words, the vertices in a boundary face collectively behave like a maximum or minimum. 
-Abusing notation, the term ``critical point", ``maxima", and ``minima" will also refer to boundaries.
-Critical values include these boundary values.
-
-\begin{definition} \label{def:desc} A \emph{descending path} in $\MM$ is a continuous function $\lambda:[0,1] \mapsto \MM$
-such that for all $a < b$, $f(\lambda(a)) \geq f(\lambda(b))$. A \emph{PL descending path} from $x$ to $y$ is a sequence
-of points $x = s_0, s_1, s_2, \ldots, s_k = y$ with the following properties: for all $s_i$ but the first and last,
-$s_i$ lies on an edge of $\MM$. $s_i$ and $s_{i+1}$ lie in the same face of $\MM$. 
-For $i < j$, $f(s_i) \geq f(s_j)$.
-\end{definition}
-
-Since $\MM$ is PL, if there is a descending path from $x$ to $y$ then it has PL representation. So we will always refer to
-PL descending paths.
-
-\begin{definition} \label{def:cont} A \emph{contour} in $\MM$ is the image of a continuous and injective function $\phi:\SSS^1 \mapsto \MM$ such that $\forall a, b \in \SSS^1$, $f(\phi(a)) = f(\phi(b))$.
-This is called a \emph{$t$-contour} if $f(\phi(a)) = t$.
-%A contour is \emph{simple} if $\forall a, b \in \SSS^1$, $\lambda(a) \neq \lambda(b)$. 
-\end{definition}
-
-Note that just as in the case of paths, contours are also PL and have a corresponding discrete representation.
-
-\begin{remark}
-For contours which pass through critical points, our definition of a contour differs from the ``standard" definition (i.e. contours which are the result of a horizontal cutting).
-
-In particular, the injective requirement does not allow contours through maxima or minima.
-We can add these single point contours at the maxima and minima to our set of contours (though this is not really necessary).
-\end{remark}
-
-\begin{lemma} \label{lem:cont} If two distinct contours intersect, then their intersection is a saddle point.
-\end{lemma}
-
-\begin{proof}
-Consider any (closed) triangular face $f$, interior to the $2$-manifold.
-The set of points on this face at a given height $h$ forms a closed line segment, call it $l_f$.
-Let $\phi$ be any contour that intersects some point in the interior of this face at height $h$.
-Then $\phi$ must contain all of $l_f$, since otherwise continuity or simplicity would be violated.
-
-So consider two distinct intersecting contours $\phi_1$ and $\phi_2$.
-Since they are distinct, there must be a point on the boundary of their intersection, i.e.
-any (arbitrarily small) ball centered at this point contains contour points both inside and outside the intersection of the contours.
-Call this point $p$.
-
-If $p$ lies in the interior of a face $f$ then both $\phi_1$ and $\phi_2$ must contain the entire above mentioned closed lines segment $l_f$, and hence $p$ was not on the intersection boundary. 
-Similarly suppose that $p$ is in the interior of an edge (clearly such an edge would need to be interior to the manifold for $\phi_1$ and $\phi_2$ to be contours).
-Since this is a $2$-manifold each edge is adjacent to exactly two faces.
-Let $l_a$ and $l_b$ be the above described closed line segments on each face.
-Then again $\phi_1$ and $\phi_2$ contain all of $l_a$ and $l_b$, and $p$ was not on the intersection boundary.
-
-So $p$ must be a vertex $v$ of the manifold.
-Let $\phi$ be any contour through $v$.
-There must be two distinct faces $f_a$ and $f_b$ adjacent to $v$ which $\phi$ intersects.
-Any face adjacent to $v$ has exactly two bounding edges adjacent to $v$.
-Since faces are linear interpolates of the vertices, and since $\phi$ intersects $f_a$ and $f_b$, for either face one of these two edges must be increasing from $v$ and the other decreasing from $v$.
-Call such a face a middle face.
-
-Now if $v$ is on the boundary of the intersection of $\phi_1$ and $\phi_2$ then we need at least 3 middle faces adjacent to $v$ (since otherwise it again is like the case we $p$ was interior to an edge).
-However, the only vertices the have at least 3 such faces are by definition saddle points.
-As no two vertices are at the same height, this saddle point can be the only point of intersection.
-\end{proof}
-
-%\begin{proof} Consider the intersection point $x$. It cannot be internal to a face, otherwise $\MM$ self-intersects.
-%By the distinctness of the contours, there must be $4$ faces incident to $x$, where each face
-%has vertex with a larger function value.
-%It cannot be internal to an edge, otherwise this edge would participate in $4$ faces. So it must be a vertex. 
-%These $4$ faces lead to at least $2$ neighbors of $x$ with higher function value. One can also show the existence of $2$
-%interleaved neighbors with lower function value. So $x$ is critical.
-%\end{proof}
-
-\begin{corollary}
-There is a unique contour containing a regular point.
-\end{corollary}
-
-\begin{definition} \label{def:p-cont}
-For a regular point $x$, $\cont(x)$ denotes the unique contour containing $x$.
-\end{definition}
-
-%By the Morse property of distinct critical values, we get the following important lemma.
-
-\begin{corollary} \label{cor:cont} Let $t$ be a non-critical value. The set of $t$-contours is a set
-of disjoint PL contours.
-\end{corollary}
-
-We can use contours to classify saddle points. For a saddle point, we call two increasing (or decreasing) neighbors \emph{opposite}
-if they lie in disjoint parts of $\Gamma^+(w)$ (or $\Gamma^-(w)$).
-
-In the following definition and, unless otherwise stated, throughout the rest of this writeup $\eps$ is some infinitesimally small amount
-(in particular smaller than the minimum height difference between vertices in the manifold).
-
-\begin{definition} \label{def:merge} Consider two increasing (resp. decreasing) opposite edges incident to a saddle point $x$,
-and let $y$ and $z$ be points on these edges such that $f(y) = f(z) = f(x) + \eps$ (resp. $=f(x) - \eps$).
-If $\cont(y)$ and $\cont(z)$ are disjoint, then $x$ is called a \emph{merge} (resp. \emph{split}) vertex.
-\end{definition}
-
-As $f$ is a Morse function over a $2$-manifold, we have the following.
-
-\begin{claim} \label{clm:saddle} 
-  \begin{compactenum}[(1)]
-    \item Every saddle is either a merge or a split, but not both.
-    \item Every saddle has exactly two distinct contours which contain it.
-  \end{compactenum}
-\end{claim}
-
-\subsection{The L-homotopy} \label{sec:l-hom}
-We need a specific notion of homotopy between contours. The L below stands for ``level".
-
-\begin{definition} \label{def:cont-hom} Two contours $\phi_0, \phi_1$ are \emph{L-homotopic}
-if: there exists a continuous function $H:\SSS^1 \times [0,1] \mapsto \MM$
-such that $H(\SSS^1,0) = \phi_0$, $H(\SSS^1,1) = \phi_1$, and $\forall a \in [0,1]$,
-$H(\SSS^1,a)$ is a contour.
-\end{definition}
-
-\begin{claim} \label{clm:non-hom} Consider a saddle point $x$. Take an increasing edge incident to $x$
-and let $y$ be a point on this edge within an $\eps$-ball of $x$. Similarly, take a decreasing edge
-and a point $z$ on this edge. $\psi(y)$ and $\psi(z)$ are not L-homotopic.
-\end{claim}
-
-\begin{proof}
-Without loss of generality assume that $x$ is a merge vertex.  By \Clm{saddle} there are exactly two distinct contours, 
-$\phi_1$ and $\phi_2$ through $x$.  Let $E_1$ and $E_2$ be the sets of manifold edges that are not adjacent to $x$ and that $\phi_1$ and $\phi_2$ intersect, respectively.
-Observe that as contours are by definition non-trivial, both $E_1$ and $E_2$ must non-empty.  Moreover since $\phi_1$ and $\phi_2$ only intersect at $x$, $E_1\cap E_2 = \emptyset$
-
-As $x$ is a merge vertex, $\psi(y)$ cannot intersect all edges in $E_1 \cup E_2$.
-On the other hand since $x$ cannot be both a merge and a split by \Clm{saddle}, $\psi(z)$ must intersect all of $E_1 \cup E_2$.  As $x$ is the only vertex in between $z$ and $y$ 
-and no edge in $E_1\cup E_2$ is adjacent to $x$, it is not hard to see that by the continuity requirement $\psi(z)$ cannot be L-homotopic to $\psi(y)$.
-%
-%
-% Without loss of generality assume that $x$ is a merge vertex.  By \ref{clm:saddle} there are exactly two distinct contours, 
-% $\phi_1$ and $\phi_2$ through $x$.  Observe that by continuity of L-homotopy, if $\psi(z)$ is $L$-homotopic to $\psi(y)$, 
-% then $\psi(z)$ must L-homotopic to either $\phi_1$ or $\phi_2$, as these are the only two possible contours at the height of $x$ 
-% (in the connected component of the manifold in between $y$ and $z$ and containing $x$).
-% We now show that $\psi(z)$ is not L-homotopic to either $\phi_1$ or $\phi_2$.
-% 
-% First observe that since contours are non-trivial (except at extrema), by continuity $\psi(z)$ cannot be 
-% simultaneously L-homotopic to both contours through $x$.  So suppose $\psi(z)$ is L-homotopic to one of the contours, say $\phi_1$.
-% 
-% As $\eps$ is smaller than the minimum height difference of two vertices in the manifold, 
-% and $z$ is $\eps$ lower than $x$, there are no vertices in between $z$ and $x$.  In particular, whatever edges $\phi_1$ and $\phi_2$ intersect (in their interier), 
-% $\psi(z)$ must also intersect.
-\end{proof}
-
-\begin{definition} \label{def:hom-mono} Two contours $\phi_0, \phi_1$ are \emph{monotonically L-homotopic}
-if the $L$-homotopy $H$ has the following property. Consider function $\alpha:[0,1] \mapsto \RR$
-where $\alpha(a) = f^{-1}(H(\SSS^1,a))$. The function $\alpha$ is strictly increasing or strictly decreasing.
-\end{definition}
-
-\begin{claim} \label{clm:mono} If two contours are L-homotopic, then they are monotonically L-homotopic.
-\end{claim}
-
-\begin{proof} Consider some L-homotopy $H$. We can assume wlog that for all $a \neq b$, $H(\SSS^1,a) \neq H(\SSS^1,b)$.
-(Otherwise, one could simplify $H$ and ``shortcut" between $a$ and $b$.)
-By continuity of $H$, the function $\alpha$ is continuous. Hence, if it is not strictly
-increasing or decreasing, $\alpha$ attains a local maximum at some $a \in (0,1)$. So there exist $a_1 = a - \eps_1$
-and $a_2 = a + \eps_2$ such that $\alpha(a_1) = \alpha(a_2)$.
-
-By continuity, $\eps_1$ and $\eps_2$ can be chosen small enough such that there are no vertices 
-in between $f^{-1}(H(\SSS^1,a_1))$ and the local maximum. 
-Now the local maximum at $a$ may contain a vertex, but by \Clm{non-hom} this vertex cannot be a saddle.
-Therefore $H(\SSS^1,a_1)$ and $H(\SSS^1,a_2)$ must intersect the same set of edges and therefore cannot be distinct.
-\end{proof}
-
-\begin{claim} \label{clm:reg} Consider a contour $\phi$ containing only regular points. 
-Take the edge cut by $\phi$ whose upper endpoint $x$ has minimum value, and let $e$
-denote the portion above $\phi$. Then $\phi$ is L-homotopic to the contour
-through any point in the interior of $e$. If $x$ is regular, then $\phi$ is L-homotopic
-to the contour through $x$.
-\end{claim}
-
-\begin{proof} 
-Let $a$ be the height of $\phi$ and $b$ the height of $x$. Consider the (potentially disconnected) 
-manifold obtained by removing all manifold points above $b$ and below $a$.  
-Specifically consider connected component that contains $\phi$.  
-By continuity of L-homotopy the claim holds iff it holds on this manifold. 
-However, this manifold contains no vectices except at the ends, and therefore
-an L-homotopy can be constructed using the natural piecewise linear map.
-\end{proof}
-
-\subsection{Reeb graphs} \label{sec:reeb}
-
-Now we get to the real definitions. Call a contour regular if it only contains regular points.
-
-\begin{definition} \label{def:rel} Define a relation $\sim$ between regular contours, so $\phi \sim \phi'$ if $\phi$
-is L-homotopic to $\phi$'.
-\end{definition}
-
-It is easy to check that $\sim$ is an equivalence relation, so the set of regular contours
-is partitioned into equivalence classes. All contours in a class are L-homotopic to each other.
-
-\begin{claim} \label{clm:equiv} Consider such an equivalence class $\Phi$. Then $f(\Phi)$
-is an open interval $(a,b)$, where $a$ and $b$ are critical values.
-\end{claim}
-
-\begin{proof} \Clm{mono} and continuity of L-homotopy immediately imply that for $\phi, \phi' \in \Phi$,
-for any $t \in [f(\phi),f(\phi')]$, there exists $\phi'' \in \Phi$ such that $f(\phi'') = t$.
-So $f(\Phi)$ is an interval. As equivalence classes are defined over only regular contours, \Clm{reg} 
-tells us that this interval must be open with critical value endpoints.
-\end{proof}
-
-We can now define the Reeb Graph $\reeb(\MM)$. We use the fact that critical points 
-have distinct values. (Technically, the Reeb Graph depends on the function $f$.
-Because it is implicit by the embedding of $\MM$, we will remove it from the notation.)
-
-\begin{definition} \label{def:reeb} The \emph{Reeb Graph} $\reeb(\MM)$ has as a vertex set
-the set of critical points. The edge $(x,y)$ is present if there exists equivalence class
-$\Phi$ such that $f(\Phi) = (f(x),f(y))$.
-\end{definition}
-
-\section{Old proof stuff}
-
-\begin{claim}
-Suppose ripe vertex $v$ is connected to a component of $\MM^+_v$. For every edge $e$ incident to a vertex of $\MM^+_v$,
-$\rep(\chi(e))$ is identical. Furthermore, all vertices in $\MM^+_v$ have been processed.
-\end{claim}
-
-\begin{proof} Consider the painting of the lower simplicial complex $\cut(\MM,\Psi)$ (call it $\NN$). 
-By the invariant, $\rep(\cdot)$ provides a valid coloring. Note that $v$ is not processed (it is ripe),
-so if we take $(f(v)+\delta)$-contours intersecting $B_\eps(v)$, these contours could not have been
-processed. Therefore, there must exist a non-trivial component of $\NN^+_v$ that $v$
-is connected to. By \Lem{full}, either all edges in this component have the same color
-or
-
-Suppose ripe vertex $v$
-is connected to some component in $\NN^+_v$. 
-Consider edge $e$ incident to $v$ intersecting $\phi$ of color $c$. 
-By \Lem{full}, there exists critical vertex $w \in \MM^+_v$ such that $w \in \fcol(c)$. 
-Regardless of whether $v \in \fcol(c)$ or not, the vertex $w$ must have been processed.
-
-\end{proof}
-
-\subsection{Updating the stack} \label{sec:stack}
-
-We have an imporant procedure that updates the stack $\st$.
-
-The painting evolves over time, and is primarily updated through \emph{color merges}. The best way
-of thinking about this is to imagine that two colors are suddenly replaced by one new common color.
-Of course, it is not possible to explicitly redo the painting, but this is implictly done
-through the union-find data structure on colors. So, when we process some vertex $v$,
-which has (say) $\col(v) = \{c_1, c_2, \ldots\}$, it may be the case that all these colors
-have merged.
-
-\begin{claim} \label{clm:cont} Consider an increasing path from $h$ to some maxima $x$.
-For any $y$ in this path, $\cont(y)$ only contains points of color $c(x)$.
-Furthermore, for all vertices on $\cont(y)$, their increasing edges are colored $c(x)$.
-\end{claim}
-
-\begin{proof} Consider this increasing path, and think of moving $y$ continuously downwards from $x$
-to $h$. Initially (when $y = x$), the lemma is trivially true. Consider the first $y$
-when the condition fails. We have $\cont(y)$ only containing points of color $c(x)$,
-but there is a vertex $z \in \cont(y)$ that has increasing edges of a different color $c'$.
-All the increasing edges of $z$ that are colored $c(x)$ are contiguous (since we can construct
-a contour intersecting all of them). Thus, there are two disjoint increasing neighborhoods for $z$,
-and $z$ is a saddle point. But that contradicts the fact that $h$ is the highest saddle point
-of color $c(x)$.
-
-So, as $y$ decreases, for all vertices on $\cont(y)$, their increasing edges are colored $c(x)$.
-This also means that $\cont(y)$ only contains points of color $c(x)$.
-\end{proof}
-
-We have important lemmas characterizing $|\col(h)|$.
-
-\begin{lemma} \label{lem:mono} If $h$ is a merge vertex, then $|\col(h)| > 1$.
-\end{lemma}
-
-\begin{proof} Suppose not. Consider the edges incident to $h$ in circular order. There exist (at least) $2$
-disjoint contiguous stretches of edges that are increasing. 
-All these edges have the same color $c$, associated with some maximum $x$. Consider two neighbors $y,z$
-that lie in different stretches. By the definition of a painting, there exist descending paths (along
-edges) from $x$ to $y$ and $x$ to $z$. We can construct a Jordan curve $C$ from $x \rightarrow y
-\rightarrow h \rightarrow z \rightarrow x$. All points in this curve except $h$ have 
-heights larger than $f(h)$. By the Jordan Curve Theorem, this partitions $\MM$ into
-an ``inside" and an ``outside".
-
-One of the lower neighbors of $h$ (say $p$) is inside, and another (say $q$) is outside.
-Because $h$ is a merge, neither of these can a minimum other than the dominant minimum $u$.
-Hence, both $p$ and $q$ have descending paths to $u$. So there is a path from $p$
-to $q$ where all heights are strictly less than $f(h)$. But this path must intersect $C$.
-Contradiction.
-\end{proof}
-
-\begin{lemma} \label{lem:col} If $h$ is a merge vertex, $|\col(h)| = 2$.
-If $h$ is a split vertex, $|\col(h)| = 1$.
-\end{lemma}
-
-\begin{proof} Consider one contiguous sequence of increasing edges from $h$.
-There exists a contour intersecting all of these edges. By \Clm{cont}, they all
-must have the same color. There are two such contiguous sequences.
-
-If $h$ is a merge vertex, then $|\col(h)| = 2$ by \Lem{mono}. If $h$ is a split vertex,
-there is a single contour passing through all increasing edges. By \Clm{cont}, they
-all have the same color.
-\end{proof}
-
-\subsection{Constructing $\reeb(\MM)$} \label{sec:const}
-
-The main construction procedure is $\onestep(\MM)$. It pieces together $\reeb(\MM)$
-incrementally.
-
-\medskip
-\begin{theorem} \label{thm:piece} Consider the very first call to $\onestep(\MM)$. Let $y$ be a point
-infinitesimally below $h$ on a decreasing edge. Consider $\cut(\MM,\cont(y)) = \{\MM_u,\MM_\ell\}$ (the upper and lower manifolds). Then $\onestep(\MM)$
-outputs $\reeb(\MM_u)$ and implicitly gives a valid painting of $\MM_\ell$.
-\end{theorem}
-
-We have an important lemma that helps to bound the running time.
-
-\begin{lemma} \label{lem:tree} For every color $c$, all points in $T(c)$ lie along a monotone path in 
-$\reeb(\MM)$.
-\end{lemma}
-
-\begin{proof} Prove by induction of the number of edges in $\reeb(\MM)$. Consider the head $h$ of $\st$ in the first call to $\onestep(\MM)$,
-and the contour $\cont(y)$ as defined in \Thm{piece}. Take $\cut(\MM,\cont(y)) = \{\MM_u,\MM_\ell\}$. Apply induction
-hypothesis on $\MM_\ell$, and use that to argue about $T(c_1)$ and $T(c_2)$ in the original manifold.
-\end{proof}
-
-\pagebreak
-}
-
-\appendix
-
-\section{Old stuff}
-
-Their algorithm works in 3 simple stages.  
-The first stage consist of a top down sweep to compute what is known as the join (or merge) tree, 
-which tracks the evolution of super-level sets (whereas the contour tree tracks level sets).  
-Similarly, the second stage does a bottom up sweep to compute the split tree, which tracks sub-level sets.  
-Their main contribution is the third and final step where they present an $O(n)$ time algorithm to merge 
-together the join and split trees to get the contour tree.
-
-There have also been some output sensitive results for computing the contour tree.  Let $t$ be the number 
-of critical points in the input mesh, which could potentially be much smaller than $n$.  
-Cole-McLaughlin and Pascucci \cite{pc-ectls-02} provided an $O(n+t\log n)$ time algorithm for the case of 
-$3$-dimensional structured meshes (note $N$ does not show up in the running time as the mesh is structured).
-Chiang \etal \cite{cllr-sooscctmp-05} improve upon this result, and provide an $O(N+t'\log t')$ algorithm 
-which works for structured or unstructured meshes in any dimension, where $t'$ is the number of component 
-critical vertices.  This is provably optimal with respect to the output size but not the output structure, 
-as discussed below.
-
-\subsubsection{Worst Case Alternatives}
-In worst case analysis an algorithm's running time is judged in terms of its worst 
-case behavior over all inputs of a given size.  While there are benefits to this viewpoint, 
-such as robustness,  it fails to capture the whole story.  As such, many alternatives 
-have been considered.  For example, one can make assumptions on the distribution of 
-inputs, which has led to a number interesting results on average case and probabilistic 
-analysis, and also smoothed analysis \cite{st-saa-04}.  However, even without making 
-assumptions on the distribution of the input, there is still more to the story.
-For example, in the study of parametrized complexity, one takes into account other 
-parameters determined by the input other than just its size (for example the maximum 
-degree in a graph or whether it is planar).
-
-\subsubsection{Contour Trees}
-
-\Ben{Maybe discuss contour tree applications and why they are useful here? Or maybe earlier?}
-
-\paragraph{A History of Contour Trees.}
-
-%As the the authors point out in \cite{cllr-sooscctmp-05}, by reduction from sorting, 
-%one can show that any algorithm to compute the contour tree in the worst case takes 
-%$\Omega(N+t\log t)$ time.  However, this lower bound only holds when the contour tree 
-%is very unbalanced (i.e. long and skinny) and therefore contains a sorted path with a 
-%large fraction of the vertices.  Therefore the bound of $\Omega(N+t\log t)$ is only 
-%worst case optimal.  In this paper we will consider an output sensitive approach which 
-%takes into account the shape of the contour tree.
-
-In other words, for any valid cost $C$, we can find a family of inputs where \Thm{main-alg}
-guarantees a running time of $C + t\alpha(t)$, and any height decision tree requires $\Omega(C)$ time.
-This shows that the cost function is inherent to the computation of contour trees.
-
-Let $f$ be a real valued function defined over a manifold embedded in $\RR^d$.  
-For any real value $\alpha$, the level set $f^{-1}(\alpha)$ can be decomposed into its connected components, or ``contours''.
-The Reeb graph is a fundamental topological structure which represents the evolution of these contours as one sweeps the surface 
-from $-\infty$ to $\infty$ according to $f$.  Often this graph is acyclic, in which case it is 
-called a \emph{contour tree}.  This is the case if, for example, the input is a surface mesh of a mountain range 
-(and the contour tree would encode when mountain peeks form and how they merge together).
-
-There is a long list of practical applications of the study of contour trees (see for example \cite{c-tmi-04} and references therein).
-Often these applications involve massive data sets, and so consequently there has also been a long series 
-of results on efficient algorithms for their computation \cite{kobps-ctsssit-97,tv-cct-98,csa-cctad-00,pc-ectls-02,cllr-sooscctmp-05}.
-Let $N$ be the total size of the input manifold, $n$ the number of vertices, and $t$ the number of critical points. (Note that in many applications $N=O(n)$.)
-As an initial step these algorithms require that all $t$ contour tree vertices be sorted.  
-In particular, \cite{cllr-sooscctmp-05} provide an $O(N+t\log t)$ time algorithm, and since in the worst case computing the contour tree 
-requires $\Omega(N+t\log t)$ time, this algorithm is optimal with respect to the size of the output.  
-Here we present the first contour tree algorithm which is optimal (for instances requiring $\Omega(N\alpha(N))$ 
-time\footnote{Where $\alpha(\cdot)$ is the inverse Ackermann function, which is $< 5$ for computation in this universe.}) 
-with respect not just to the size of the output, but also its structure.  This is accomplished by introducing a simple linear time 
-surface partitioning scheme to avoid sorting, which is likely to have applications elsewhere.  
-For all instances the bound for our algorithm either matches the best previous bound, or is faster.
-For, example when the contour tree is a balanced, our algorithm runs in $O(N\alpha(N))$ time, whereas all 
-previous algorithms had a running time of $\Omega(N+t\log t)$
-
-%In particular, the algorithm in \cite{cllr-sooscctmp-05} is worst case optimal with respect to the size of the output.  
-%Here we present the first contour tree algorithm which is (near\footnote{Our algorithm is optimal on all instance requiring $\Omega(n\alpha(n))$ time.}) 
-%optimal with respect not just to the size of the output, but also its structure.  
-%Specifically, the bound for our algorithm either matches \cite{cllr-sooscctmp-05}, or is faster.
-%For, example when the contour tree is a balanced, our algorithm runs in $O(n\alpha(n))$ time, whereas all 
-%previous algorithms had a running time of $\Omega(n\log n)$.
-
-This work can be seen as a continuation of the study of instance-optimal geometric algorithms introduced by Afshani \etal \cite{abc-ioga-09}.
-In order to prove instance optimality we use a non-trivial combinatorial discharging argument, which we believe is of independent interest.
-Of particular note is the use of heavy-light path decompositions, first introduced by Sleator and Tarjan \cite{st-dsdt-83} in their analysis of link-cut trees.
-
-% Explain why useful:  Lets you know the ``shape of your date''.  Map of the critical points where contours change.  
-%Fast extraction of seed sets in a small sized structure, which can let you do threshold queries.
-% 
-% \medskip\noindent
-% Paragraph on specific applications.  Scientific visualization, in particular combustion research.  GIS.
-
-\paragraph{The Issue of Sorting.}
-As discussed above, the algorithm of Carr \etal \cite{csa-cctad-00} computes the contour 
-tree by first constructing the join and split trees.  As vertices appear in sorted order 
-in any root to leaf path in the join and split trees, and in any maxima to minima path 
-in the contour tree, sorting the input vertices is a natural first step.  However, not 
-all vertices from the input need to be sorted.  In particular, Chiang \etal 
-\cite{cllr-sooscctmp-05} improved the running time by observing that as the join and 
-split trees only consist of critical points, it suffices to only sort critical points.  
-If a large fraction of the vertices appear in a single path in the join or split tree, 
-then paying to sort all the critical points is unavoidable, and therefore in the worst 
-case \cite{cllr-sooscctmp-05} is optimal.  However, if the trees are more balanced then 
-sorting all critical points together can be very costly.  Specifically, the contour tree 
-does not care about the relative ordering of vertices which do not have an 
-ancestor-descendant relationship, and so this information does not need to be computed.
-See \Fig{relative} for an example.
-
-% \begin{figure}[h]\centering
-%     \includegraphics[width=.3\linewidth,height=.42\linewidth]{figs/shaded}%
-%     \hspace{.5in}
-%     \includegraphics[width=.3\linewidth,height=.42\linewidth]{figs/trees2}
-%     \caption{On the left, a surface will a balanced contour tree, but whose join and split trees have long tails.  
-%     On the right (in clockwise order from the top) the contour, split and join trees.}
-%     \label{fig:sorting}
-% \end{figure}
-
-\paragraph{Suboptimal Join and Split Trees.}
-There is also a more fundamental problem with the now standard approach of first computing 
-the join and split trees in order to compute the contour tree.  In particular, this approach 
-can never lead to an algorithm that is optimal with respect to the shape of the contour tree.  
-This is because it is possible to have join and split trees which are long and skinny (i.e. 
-costly to compute) whereas the contour tree is balanced.  For example consider the case shown 
-in \Fig{sorting}, where the mesh looks like two (thickened) balanced binary trees, one upside 
-down, and joined at their roots.  In this case the contour tree is potentially cheap to 
-compute as there is no long sorted path.  However, the join and split trees will have long 
-sorted tails (consisting of the vertices in the bottom and top balanced trees, respectively), 
-and therefore require $\Omega(n\log n)$ time to compute.
-
-\subsection{Overview and Contribution}
-
-\paragraph{Algorithm Overview.}
-At a high level, the technique is rather simple.  Break the input manifold into nicely 
-behaved pieces, compute the contour tree on each piece, and then stitch together the computed 
-contour trees according to the cuts we made to break up the manifold.  Below we informally 
-describe how this is accomplished.
-\begin{compactitem}
- \item {\bf Raining to partition.}  To start with, pick an arbitrary maxima.  Now imagine 
- ``raining'' water down on the manifold from this maxima, i.e. mark all points which are reachable 
- by non-ascending paths from the maxima.  Let the boundary between wet and dry portions of the 
- manifold be called the interface (which may have different connected components).  
- Remove the wet submanifold, and invert the remaining pieces
- of the original manifold.  Now recursively rain in each one of these pieces, but rather than 
- picking an arbitrary maxima, start from the portion of the interface that cut that piece off.
-
- The benefit of this partitioning is that each resulting piece 
- is maxima dominant, i.e. there is a single maxima that has a non-ascending path to each point.  
- This implies the contour tree of this submanifold is the same as its split tree, or in other words 
- its contour tree is a rooted binary tree.
-
- %
- \item {\bf Painting trees.}  For each maxima dominant piece we want to compute the split 
- tree without sorting all the vertices.  For conceptual ease lets first flip the submanifold 
- (i.e., it is now minima dominant and we compute the join tree).  Now arbitrarily pick a maxima 
- and spill paint from it, specifically color all edges and vertices of the mesh that are reachable 
- from the maxima by descending paths.  Now go to each other maxima and spill a different color, 
- but be aware that paints do not mix, i.e. if we reach a vertex that was previosly explored 
- by another color then we do not explore it.
-
- This procedure partitions the set of critical points into color classes.  Clearly each color 
- class is connected in the manifold.  Suppose each class was also connected in the join tree.  
- In this case we partitioned the join tree into a set of descending paths, each of which 
- can be constructed by sorting the critical points in a class.  Roughly speaking, 
- these paths can then be joined together by what amounts to doing a post-order traversal of the tree.
- However, we may not be so lucky and the color classes may not be connected in the 
- tree.  This turns out not to be a problem, but it requires proper data-structure use. 
-%  
-%  Now suppose we were lucky and the spilling of paint in the manifold corresponded with a paint 
-%  spilling in the join tree, i.e. the color classes of the critical points partition the join 
-%  tree into disjoint descending paths, each containing one leaf.   In this case 
-%  we have partitioned the tree into branches, and so the tree can be formed by gluing these 
-%  branches together.  Intuitively (i.e. not exactly true), if two branches glue together then 
-%  it means there is a join vertex which got colored by the respective colors of each branch 
-%  (only one of these colors actually explored the vertex in the mesh and this is the ``main'' 
-%  branch you are gluing into).  Therefore, the tree can be constructed by starting at the leaf 
-%  of the trunk (i.e. the branch with the global minima) and recursively building the subtrees 
-%  while walking towards the root.  In other words we are building the tree by doing a post-order traversal.
-%  Unfortunately, we may not be so lucky and the color classes may not appear contiguously in 
-%  the tree.  In this case we can still build the tree in this leaf first fashion, but doing so 
-%  efficiently requires the use of data-structures.
-\end{compactitem}
-
-\paragraph{Lower Bound Overview.}
-In order to show that our algorithm for computing the join tree is optimal we consider paritioing the tree into disjoint 
-descending paths, each containing a leaf.  It turns out that each such partition into paths, $P$, 
-defines a lower bound of $\sum_{p\in P} |p|\log|p|$ for computing the contour tree, as you need to know the 
-sorted order of the vertices along each path in order to determine the tree.
-Specifically, we show that there is a set of at least $\Pi_{p_i\in P} (|p_i|-1)!$ distinct PL $2$-manifolds, 
-all of which have the same defining mesh, and for which each sorted order of each group of saddles determines a different join tree.
-
-The instance optimality of our algorithm then follows, as we also show that there is a particular path decomposition 
-whose cost upper bounds our.  Specifically, we assign a weight to each node in the tree which is determined by 
-the painting procedure described above (which represents the time it takes to process this node).  
-The path decomposition then turns out to be the heavy-light path decomposition 
-described by Sleator and Tarjan \cite{st-dsdt-83}, according to this weight function.
-However, this decomposition is just the first step in the analysis, which 
-requires further case analysis depending on the relative size of the length of each path 
-and the weights of its vertices.
-
-\paragraph{Our Contribution.}
-Our contribution can be broken into three main aspects.  The first two are algorithmic.  Namely, 
-first we describe a new linear time manifold partitioning scheme which breaks the manifold into 
-``extremum'' dominant submanifolds.  
-Second, we describe how to compute join trees optimally with respect to shape of the tree.
-As the contour tree of each submanifold is (roughly) equal to its join tree, this leads to an 
-optimal algorithm for the contour tree of the input manifold.  
-Therefore, our third and final contribution is proving the optimality of this result, which 
-requires a non-trivial amount of work, and uses techniques we believe to be of independent interest.
-
-\paragraph{Organization.}
-\Ben{To be Written}
-
 \end{document}

### 2014-11-04 paint-contour-tree.tex
switch added

@@
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%
+%%%%%%%%%%Comment out the switch below for full version 
+%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\def \STOC {}
+
+\ifx\STOC\undefined
+  \newcommand{\FullVer}[1]{#1}
+  \newcommand{\SubmitVer}[1]{}
+\else
+  \newcommand{\FullVer}[1]{}
+  \newcommand{\SubmitVer}[1]{#1}
+\fi
+
 \documentclass[11pt]{article}
@@
 
-\Fig{relative} and \Fig{sorting} show examples of simplicial complexes, with heights and their contour trees.
+\Fig{relative} and \Fig{sorting} 
+show examples of simplicial complexes, with heights and their contour trees.
 Think of the contour tree edges as pointing downwards, so we can talk of the ``up-degree" and ``down-degree" of vertices.
@@
 \begin{definition}
-\label{def:path} For a contour tree $T$, a \emph{path decomposition}, $P(T)$, is a partition of $T$ into a set of vertex disjoint leaf paths.  
+\label{def:path} 
+For a contour tree $T$, a \emph{path decomposition}, $P(T)$, is a partition of $T$ into a set of vertex disjoint leaf paths.  
 \end{definition}
 
-\begin{theorem} \label{thm:main-alg} Consider a PL-Morse\footnote{This is a standard ``well-behaved" assumption that implies
+\begin{theorem} 
+\label{thm:main-alg} 
+Consider a PL-Morse\footnote{This is a standard ``well-behaved" assumption that implies
 that all critical points have distinct values, and the contour tree has max-degree $3$. Our algorithm
@@
 As referenced in \Thm{main-alg}, for any $f:\MM \mapsto \RR$ our deterministic algorithm defines a 
-specific path decomposition $P(T)$\footnote{The specific decomposition used is defined in \Sec{pathDecomp}, 
+specific path decomposition $P(T)$%
+\FullVer{
+\footnote{The specific decomposition used is defined in \Sec{pathDecomp}, 
 but also depends on the raining and painting procedures described in \Sec{rain} and \Sec{paint}.  
-The important thing is that our algorithm is deterministic, and so for a fixed $\MM$ and $f$ it is well defined.}. 
+The important thing is that our algorithm is deterministic, and so for a fixed $\MM$ and $f$ it is well defined.}.}
+\SubmitVer{
+\footnote{The specific decomposition used is defined in the full version of the paper.  
+The important thing is that our algorithm is deterministic, and so for a fixed $\MM$ and $f$ it is well defined.}.}
 We define $\cost(f) = \sum_{p \in P(T)} |p|\log |p|$ for this specific decomposition $P(T)$.
@@
 %
-\begin{theorem} \label{thm:main-lb} 
+\begin{theorem} 
+\label{thm:main-lb}
 %Fix a sufficiently large numbers $t$ and $C$, 
@@
 and intuition. 
+\SubmitVer{The full version with all the details can be found in the appendix.}
 It is helpful to keep the $d=2$ case in mind, so the input is just a triangulated
@@
 at \Fig{sorting}. The portion on the left is minimum dominant in exactly this fashion.
-Water poured at any critical point must flow to the boundary, and the entire wet portion is connected.
+%Water poured at any critical point must flow to the boundary, and the entire wet portion is connected.
 More formally, $\MM^-_v$ is connected for all $v$, so there are no splits.
@@
 
-%\section{Acknowledgements}
-\paragraph{Acknowledgements.}
-We thank Hsien-Chih Chang, Jeff Erickson, and Yusu Wang for numerous useful discussions. This work is
-supported by the Laboratory Directed Research and Development
-(LDRD) program of Sandia National Laboratories. Sandia National
-Laboratories is a multi-program laboratory managed and operated
-by Sandia Corporation, a wholly owned subsidiary of Lockheed
-Martin Corporation, for the U.S. Department of Energy’s National
-Nuclear Security Administration under contract DE-AC04-94AL85000.
-
+\FullVer{
 \section{Divide and conquer through contour surgery} \label{sec:surgery}
@@
 % \end{enumerate}
+}
+
+\paragraph{Acknowledgements.}
+We thank Hsien-Chih Chang, Jeff Erickson, and Yusu Wang for numerous useful discussions. This work is
+supported by the Laboratory Directed Research and Development
+(LDRD) program of Sandia National Laboratories. Sandia National
+Laboratories is a multi-program laboratory managed and operated
+by Sandia Corporation, a wholly owned subsidiary of Lockheed
+Martin Corporation, for the U.S. Department of Energy’s National
+Nuclear Security Administration under contract DE-AC04-94AL85000.
 

### 2014-11-04 paint-contour-tree.tex
lower bound changes

--- 2014-11-03 paint-contour-tree.tex
@@
 \label{sec:lb}
-In this section we provide a lower bound for the time it takes to compute a contour tree in terms the cost of path decompositions.  
 We first prove a lower bound for join trees, and then generalize to contour trees.
 %\Ben{Hopefully the following sentence clear up confusion between $\MM$ and $(f,\MM)$}
-Note earlier we usually used $\MM$ to denote the lift of a simplicial complex according to some height function $f$, 
-however in this section we make a distinction between the complex $\MM$ and the height function $f$.  
-In particular all previous definitions or subroutines which took $\MM$ as input now take $(f,\MM)$ as input.
+%Note earlier we usually used $\MM$ to denote the lift of a simplicial complex according to some height function $f$, 
+%however in this section we make a distinction between the complex $\MM$ and the height function $f$.  
+%In particular all previous definitions or subroutines which took $\MM$ as input now take $(f,\MM)$ as input.
 
@@
 
-In this section we prove \Thm{main-lb}, but for join trees instead of contour trees.  
+%In this section we prove \Thm{main-lb}, but for join trees instead of contour trees.  
 %Observe that given a fixed $(f,\MM)$, 
@@
 %  
-Let $\MM$ be fixed.  For a given height function $f$, we can now define $\cost_{\jc}(f)$.  Specifically, consider running the $\build(f,\MM)$ procedure 
-of section \Sec{paint}, which computes the join tree $\jc(f,\MM)$.  This procedure calls $\init$, which will (uniquely) define some 
-initial coloring of $\jc$ (\Def{initialColoring}).  Let $P(f)=P(\jc(f,\MM))$ be the maximum path decomposition of the join tree according to this coloring. 
-Then $\cost_{\jc}(f) = \sum_{p\in P(f)} |p|\log|p|$
+We focus on terrains, so $d=2$.
+Consider any valid path decomposition $P$ of a valid join tree (actually, for any rooted binary tree,
+there exist a terrain for this is the contour tree).
+When we say ``compute the join tree", we require the join tree to
+be labeled with the corresponding vertices of the terrain.
+
+%
+%For given $\MM$, consider running the algorithm of \Sec{also} to compute $\jc(\MM)$.
+%Let $P(\jc(\MM))$ be the maximum path decomposition of the join tree according to the initial coloring
+%of $\init(\MM)$. Define $\cost_{\jc}(f) = \sum_{p\in P(\jc(\MM))} |p|\log|p|$
 
@@
 \label{lem:pathDecomp}
-Let $\mathcal{P} = P(f)$ be the maximum path decomposition of some height function $f$ defined over a complex $\MM$. 
-Then there is a family of functions, $\bF$, all defined over a common triangulation in plane, 
-such that the size of $\bF$ is $\Pi_{p_i\in P} (|p_i|-1)!$, and no two function in $\bF$ define the same join tree.
+Fix a valid path decomposition $P$. There is a family of terrains, $\bF_P$, all with the same triangulation, 
+such that $|\bF_P| = \Pi_{p_i\in P} (|p_i|-1)!$, and no two terrains in $\bF_P$ define the same join tree.
 \end{lemma}
+
 \begin{proof}
- We describe a simple triangulation, with heights specified for all vertices except for the saddles.  The main idea is that 
- this triangulation will be flexible enough to allow almost any assignment of heights to the saddles, thus defining a class of join trees 
- whose size is that described in the lemma statement.
-
- First, we describe the basic building block of these terrains, which roughly speaking corresponds to a fixed path $p\in P$.  
- Specifically, we say a \emph{diamond} of size $m$ is the following terrain.  To start, construct a slightly tilted cycle 
- of length $m$.  Specifically have two antipodal points, at heights $1$ and $0$, which we will call the anchor and trap of the diamond, respectively.
+% We describe a simple triangulation, with heights specified for all vertices except for the saddles.
+
+%The main idea is that 
+% this triangulation will be flexible enough to allow almost any assignment of heights to the saddles, thus defining a class of join trees 
+% whose size is that described in the lemma statement. 
+ We describe the basic building block of these terrains, which corresponds to a fixed path $p\in P$.  
+ Informally, a \emph{diamond} is an upside down cone with $m$ triangular faces (see \Fig{diamond}). 
+ Construct a slightly tilted cycle 
+ of length $m$ with the two antipodal points at heights $1$ and $0$. These are called the anchor and trap of the diamond, respectively.
  The remaining $m-2$ vertices are evenly spread around the cycle and heights decrease monotonically when going from the anchor to the trap.
- Next, create an apex vertex at some appropriately large height, and add an edge between it and each vertex in the cycle.  At this point 
- we have built an upside down cone with $m$ triangular faces (see \Fig{diamond}).
-
- Now we describe how to attach two diamonds.  In this process, we glue the base of a scaled down ``child'' diamond on to a triangular cone face of the larger  ``parent'' diamond (see \Fig{diamond}).
- Specifically, the anchor of the child diamond is attached directly to a face of the parent diamond at some unspecified height $h$. The remained of the base of the child cone is then 
+ Next, create an apex vertex at some appropriately large height, and add an edge each vertex in the cycle.
+
+ Now we describe how to attach two different diamonds.
+In this process, we glue the base of a scaled down ``child'' diamond on to a triangular cone face of the larger  ``parent'' diamond (see \Fig{diamond}).
+ Specifically, the anchor of the child diamond is attached directly to a face of the parent diamond at some height $h$. The remaineder of the base of the child cone is then 
  extended down (at a slight angle) until it hits the face of the parent.
 
- At this point we can construct the entire terrain, by repeatedly gluing together diamonds.  
- Specifically, for each path $p_i\in P$ we create a diamond of size $|p_i|+1$.
+ The full terrain is obtained by repeatedly gluing diamonds. For each path $p_i\in P$, we create a diamond of size $|p_i|+1$.
  The two faces adjacent to the anchor are always empty, and the remaining faces are for gluing on other diamonds.
@@
  (Naturally for this construction to work, diamonds for a given path will be scaled down relative to the size of the diamond of their parent).
+ By varying the heights of the gluing, we get the family of terrains.
 
@@
 
- Now there is a valid instance of this described construction for any relative ordering of the heights of the saddles on a given diamond.  
- Let the set of height functions for these instances be denote $\bF$.
+ There is a valid instance of this described construction for any relative ordering of the heights of the saddles on a given diamond.  
  In particular, there are $(|p|-1)!$ possible orderings of the heights of the saddles on the diamond for $p$, and hence $\Pi_{p_i\in P} (|p_i|-1)!$ possible 
- functions in this set.  Also, each one of these functions will result in a different join tree.  Specifically, a permutation of the heights on a given diamond 
- corresponds to a permutation of the vertices along a path in $P$, as by construction all saddles on a given diamond will appear in sorted order in the join tree.
-\end{proof}
-
-In the following we say to path decompositions $P_1$ and $P_2$ (of potentially different complexes and/or height functions) are equivalent 
-if there is an graph isomorphism, $f$, between their respective shrubs, $S_1$ and $S_2$, which has the property that for any node $v\in S_1$ corresponding to a path of 
-size $|v|$, $f(v)$ also corresponds to a path of size $|v|$. Note that for equivalent decompositions, $\sum_{p\in P_1} |p|\log|p| = \sum_{p\in P_2} |p|\log|p|$.
+ terrains we can build.  Each one of these functions will result in a different (labeled) join tree. All saddles on a given diamond will appear in sorted order in the join tree.
+ So, any permutation of the heights on a given diamond corresponds to a permutation of the vertices along a path in $P$.
+\end{proof}
+
+Two path decompositions $P_1$ and $P_2$ (of potentially different complexes and/or height functions) are equivalent 
+if: there is a 1-1 correspondence between the sizes of the constituent paths, and the shrubs are isomorphic.
+%
+%
+%if their respective shrubs, $S_1$ and $S_2$, which has the property that for any node $v\in S_1$ corresponding to a path of 
+%size $|v|$, $f(v)$ also corresponds to a path of size $|v|$. Note that for equivalent decompositions, $\sum_{p\in P_1} |p|\log|p| = \sum_{p\in P_2} |p|\log|p|$.
 
@@
 \label{lem:cost}
-Let $\mathcal{P}$ and $\bF$ be as defined above.  Then for all $f\in \bF$, $\cost_{\jc}(f) = \sum_{p\in P(f)} |p|\log|p| = \sum_{p\in \mathcal{P}} |p|\log |p|$
+For all $\MM \in \bF_P$, the total number of heap operations performed by $\build(\MM)$ is $\Theta(\sum_{p\in P} |p|\log|p|)$.
 \end{lemma}
 \begin{proof}
-First observe that by construction, regardless of the choice of $f\in \bF$, there is a path decomposition $\overline{P}(T)$ of the join tree which is equivalent to $\mathcal{P}$.
-Specifically, this is the decomposition where each path $p\in \overline{P}(T)$ consists exactly of the set of critical points (except for its anchor) from a given diamond. 
-We now show that regardless of how paint spilling is done, this will be the maximum path decomposition of $T$ (implying the lemma statement).
-
- So consider any order of paint spilling over the surface defined by $f$.
- Consider any join on a face of some diamond.  This join is the anchor point of some connecting child diamond.  
- The claim is that the two colors of this join (according to \Def{initialColoring})
+%For any choice of $\MM \in \bF$, there is a path decomposition $\overline{P}(T)$ of the join tree equivalent to $\mathcal{P}$.
+%Specifically, this is the decomposition where each path $p\in \overline{P}(T)$ consists exactly of the set of critical points (except for its anchor) from a given diamond. 
+The primary ``non-determinism" of the algorithm is the initial painting constructed by $\init(\MM)$. 
+We show that regardless of how paint spilling is done, the number of heap operations is bounded as above.
+
+ Consider an arbitrary order of the initial paint spilling over the surface defined by $f$.
+ Consider any join on a face of some diamond, which is the anchor point of some connecting child diamond.  
+ The join has two up-stars, each of which has exactly one edge. Each edge connects to a maxima
+ and must be colored by that maxima.
+ Hence, two colors touching this join (according to \Def{initialColoring})
  are the colors of the apexes of the child and parent diamond.  
- This easily follows since this join has two up-stars, each of which has exactly one edge.  
- Specifically, one of these edges connects directly to the apex of the parent diamond and so the join must get the color of this maxima.
- The other edge connects directly to the apex of the child diamond (this is why we left the faces adjacent to an anchor empty) and 
- so the second color is that of the maxima of the child diamond.
-
- So let $c$ be the color of some maxima, $m$.  The above proven property of the coloring implies that in the join tree, the vertices for 
- which $c$ appears are exactly the vertices of the path in  $\overline{P}(T)$ corresponding $m$, as well as 
- the vertex to which the root of this path connects to (i.e. the anchor of the diamond of $m$).
- Now we show this implies $\overline{P}(T)$ is a maximum path decomposition.
- To see this, consider a vertex $v\in T$, and let $p$ be the path in $\overline{P}(T)$ which contain $v$. 
- Now $v$ has two children (if it is a join).  Let $c_1$ be the child which is in $p$, and $c_2$ the child which is the root of some adjacent path $p'$.  
- In order for $\overline{P}(T)$ to be the maximum path decomposition, we must show $|H_{c_1}|\geq |H_{c_2}|$.
- %(Note, to ensure a unique maximum path decomposition, one can break ties consistently by assigning some arbitrary but fixed ordering to the vertices.)
- %Now $v$ gets assigned two colors, one 
- %of which we know is the color assigned to all vertices in $p$ and the other is the one assigned to all the vertices of $p'$. 
- Observe that none of the colors in the subtree of $c_2$ appear below $v$.  Hence the heap at $c_2$ consists only of $v$.  
- On the other hand the heap of $c_1$ contains $v$, any other vertex of $p$ below $v$, and the vertex to which the root of $p$ is adjacent to, 
- and hence $|H_{c_1}|\geq |H_{c_2}|$.
+% 
+% Specifically, one of these edges connects directly to the apex of the parent diamond and so the join must get the color of this maxima.
+% The other edge connects directly to the apex of the child diamond (this is why we left the faces adjacent to an anchor empty) and 
+% so the second color is that of the maxima of the child diamond.
+
+ Take any join $v$, with two children $w_1$ and $w_2$. Suppose $w_1$ and $v$ belong to the same path
+ in the decomposition. The key is that any color from a maxima in the subtree at $w_2$ cannot touch
+ any ancestor of $v$. This subtree is exactly the contour tree of the child diamond attached at $v$.
+ The base of this diamond is completely contained in a face of the parent diamond. So all colors
+ from the child ``drain off" to the base of the parent, and do not touch any joins on the parent diamond.
+
+ Hence, $|H_v|$ is at most the size of the path in $P$ containing $v$. By \Lem{runTimeUpper}, the total number of heap operations
+ is at most $\sum_v \log |H_v|$, completing the proof.
+
+% 
+% So let $c$ be the color of some maxima, $m$.  The above proven property of the coloring implies that in the join tree, the vertices for 
+% which $c$ appears are exactly the vertices of the path in  $\overline{P}(T)$ corresponding $m$, as well as 
+% the vertex to which the root of this path connects to (i.e. the anchor of the diamond of $m$).
+% Now we show this implies $\overline{P}(T)$ is a maximum path decomposition.
+% To see this, consider a vertex $v\in T$, and let $p$ be the path in $\overline{P}(T)$ which contain $v$. 
+% Now $v$ has two children (if it is a join).  Let $c_1$ be the child which is in $p$, and $c_2$ the child which is the root of some adjacent path $p'$.  
+% In order for $\overline{P}(T)$ to be the maximum path decomposition, we must show $|H_{c_1}|\geq |H_{c_2}|$.
+% %(Note, to ensure a unique maximum path decomposition, one can break ties consistently by assigning some arbitrary but fixed ordering to the vertices.)
+% %Now $v$ gets assigned two colors, one 
+% %of which we know is the color assigned to all vertices in $p$ and the other is the one assigned to all the vertices of $p'$. 
+% Observe that none of the colors in the subtree of $c_2$ appear below $v$.  Hence the heap at $c_2$ consists only of $v$.  
+% On the other hand the heap of $c_1$ contains $v$, any other vertex of $p$ below $v$, and the vertex to which the root of $p$ is adjacent to, 
+% and hence $|H_{c_1}|\geq |H_{c_2}|$.
 \end{proof}

### 2014-11-04 paint-contour-tree.tex
done with first pass

@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\def \STOC {}
+%\def \STOC {}
 
@@
 in running time. For a partition of the contour tree into a set of descending paths, $P$,
-our algorithm runs in $O(\sum_{p\in P} |p|\log|p| + t\alpha(t)+N)$, which is 
-always bounded by $O(t\log t + N)$ and is $O(t\alpha(t)+N)$ for balanced trees. 
-Moreover, we prove a strong optimality property of our algorithm, showing that $\sum_{p\in P} |p|\log|p|$ complexity
-is inherent to computing contour trees. Consider the family of instances where our algorithm
-requires $C$ comparisons. Any algorithm computing contour trees on this family requires
-$\Omega(C)$ comparisons in the worst case.
+our algorithm runs in $O(\sum_{p\in P} |p|\log|p| + t\alpha(t)+N)$.
+This is at most $O(t\log D + N)$, where $D$ is the diameter of the contour tree.
+Moreover, it is $O(t\alpha(t)+N)$ for balanced trees, a significant improvement over the previous complexity.
+
+Our algorithm requires numerous ideas: partitioning the contour tree into join and split trees, 
+a local growing procedure to iteratively build contour trees, and the use of heavy path 
+decompositions for the time complexity analysis. There is a crucial use of a family of binomial heaps to 
+maintain priorities, ensuring that any comparison made is between comparable nodes of the contour tree.
+We also prove lower bounds showing that the $\sum_{p\in P} |p|\log|p|$ complexity
+is inherent to computing contour trees. 
+%Consider the family of instances where our algorithm
+%requires $C$ comparisons. Any algorithm computing contour trees on this family requires
+%$\Omega(C)$ comparisons in the worst case.
 
@@
 where $P(T)$ is a specific path decomposition (constructed implicitly by the algorithm).
+
+The number of comparisons made is $O(\sum_{p \in P(T)} |p|\log |p| + N)$.
 \end{theorem}
@@
 
-We complement \Thm{main-alg} with an almost matching lower bound, showing that the complexity
-of $\sum_{p \in P(T)} |p|\log |p|$ is required for any contour tree algorithm. The exact phrasing
+We complement \Thm{main-alg} with a lower bound, suggesting that the comparison complexity
+of $\sum_{p \in P(T)} |p|\log |p|$ is inherent for any contour tree algorithm. The exact phrasing
 of the lower bound requires some care. Consider the $d=2$ (terrain) case.
-As referenced in \Thm{main-alg}, for any $f:\MM \mapsto \RR$ our deterministic algorithm defines a 
-specific path decomposition $P(T)$%
-\FullVer{
+Let a path decomposition $P$ be called \emph{valid} if it is actually output by the algorithm
+on some input.
+%As referenced in \Thm{main-alg}, for any $f:\MM \mapsto \RR$ our deterministic algorithm defines a 
+%specific path decomposition $P(T)$%
+\ignore{\FullVer{
 \footnote{The specific decomposition used is defined in \Sec{pathDecomp}, 
@@
 %
+}
 \begin{theorem} 
@@
 %where $t$ is the number of critical points and $C$ is a valid cost for an instance with $t$ critical point.
-Consider the set of functions $\bF_C$ that have the same number of critical points and the same (valid) cost $C$.  Any algebraic decision tree that correctly computes
-the contour tree on all of $\bF_C$ has a worst case running time of $\Omega(C)$.
+Consider a valid path decomposition $P$. There exists a family $\bF_P$ of terrains ($d=2$) with the following properties.
+Any contour tree algorithm makes $\Omega(\sum_{p \in P} |p|\log |p|)$ comparisons in the worst case over $\bF_P$.
+Furthermore, for any terrain in $\bF_P$, the algorithm of \Thm{main-alg} makes $O(\sum_{p \in P} |p|\log |p|)$ comparisons.
+%
+%the set of functions $\bF_C$ that have the same number of critical points and the same (valid) cost $C$.  Any algebraic decision tree that correctly computes
+%the contour tree on all of $\bF_C$ has a worst case running time of $\Omega(C)$.
 \end{theorem}
+
+In other words, for any choice of $\sum_{p \in P} |p|\log |p|$, we can design a hard family where any algorithm
+requires this cost in the worst case, and our algorithm matches this bound.
 
@@
 of contour tree construction beyond the standard worst-case analysis. Our theorem and lower bound explain \emph{when}
-computing contour trees is expensive, and provide insight into the hardness of particular instances. It is intriguing
-that the cost (as defined earlier) roughly captures the hardness of an instance.
+computing contour trees is expensive, and provide insight into the hardness of particular instances. 
+%It is intriguing
+%that the cost (as defined earlier) roughly captures the hardness of an instance.
 Trees that are short and fast are easy to compute, whereas trees with long paths are hard.
-Our result is related to the concept of instance optimal algorithms, as defined by Afshani \etal \cite{abc-ioga-09}.
+This is tangentially related to the concept of instance optimal algorithms, as defined by Afshani \etal \cite{abc-ioga-09}.
 Their notion of optimality is far more refined that what we prove, but it shares the flavor of understanding the spectrum
@@
 
-The following is the analogous statement to \Thm{main-lb}, but for join trees.
-Call a cost $C$ join tree valid if $C = cost_{\jc}(f)$ for some $f$ defined over a given complex $\MM$.
+The following is the equivalent of \Thm{main-lb} for join trees, and immediately follows from
+the previous lemmas.
+%Call a cost $C$ join tree valid if $C = cost_{\jc}(f)$ for some $f$ defined over a given complex $\MM$.
 
 \begin{theorem}\label{thm:joinLB}
-Consider the set of functions $\bF_C$ that all have the same join tree valid cost $C$.  
-Any height function decision tree that correctly computes
-the join tree on all of $\bF_C$, has an average case (uniform over $\bF_C$) running time of $\Omega(C)$.
+Consider a rooted tree $T$ and an arbitrary path decomposition $P$ of $t$.
+There is a family $\bF_P$ of terrains such that any algebraic decision
+tree computing the contour tree (on $\bF_P$) requires $\Omega(\sum_{p \in P} |p|\log |p|)$ time.
+Furthermore, our algorithm makes $O(\sum_{p \in P} |p|\log |p|)$ on all these instances.
+%
+%Consider the set of functions $\bF_C$ that all have the same join tree valid cost $C$.  
+%Any height function decision tree that correctly computes
+%the join tree on all of $\bF_C$, has an average case (uniform over $\bF_C$) running time of $\Omega(C)$.
 \end{theorem}
+
 \begin{proof}
- If $C$ is a join tree valid cost then there some $f$ such that $\cost_{\jc}(f)=\sum_{p\in P(f)} |p|\log|p|$.
- By the above lemmas, there is then a set $\bF$ of $\Pi_{p_i\in P} |p_i|!$ functions, such that for any $g\in \bF$, $C= \cost(g)$.  
- (Specifically, each sorted ordering of the relative height values of the saddles on each diamond yields a function in this class.)
- Hence $\bF\subseteq \bF_C$, and so any height function decision tree for the set $\bF_C$ must distinguish between these possibilities, and therefore must 
- have $\Pi_{p_i\in P} |p_i|!$ leaves.  Hence by Stirling's approximation, the average depth in this tree is $\Omega(\sum_{p_i\in P} |p_i|\log|p_i|)$.
+The proof is a basic entropy argument. Any algebraic decision tree that is correct on all of $\bF_P$
+must distinguish all inputs in this family. By Stirling's approximation, the average depth in this tree is $\Omega(\sum_{p_i\in P} |p_i|\log|p_i|)$.
+\Lem{cost} completes the proof.
+%
+%
+% If $C$ is a join tree valid cost then there some $f$ such that $\cost_{\jc}(f)=\sum_{p\in P(f)} |p|\log|p|$.
+% By the above lemmas, there is then a set $\bF$ of $\Pi_{p_i\in P} |p_i|!$ functions, such that for any $g\in \bF$, $C= \cost(g)$.  
+% (Specifically, each sorted ordering of the relative height values of the saddles on each diamond yields a function in this class.)
+% Hence $\bF\subseteq \bF_C$, and so any height function decision tree for the set $\bF_C$ must distinguish between these possibilities, and therefore must 
+% have $\Pi_{p_i\in P} |p_i|!$ leaves.  Hence by Stirling's approximation, the average depth in this tree is $\Omega(\sum_{p_i\in P} |p_i|\log|p_i|)$.
 \end{proof}
@@
 We first generalize previous terms to the case of contour trees.
-In this section $T$ will denote an arbitrary contour tree.  Specifically, every internal vertex in $T$ has degree $3$ 
-and the height function defined over the vertex set is such that every internal vertex in either a split or join 
-(i.e. at most $2$ neighbors lie above or below it).
-
-We need to generalize the definition of path decompositions, which previously were only defined for rooted trees.
-Again we define a path decomposition as a partition of the vertices into 
-disjoint leaf paths (i.e. each path is monotonic in height and contains a leaf). 
-However, for simplicity we now restrict our attention to path decompositions which 
-roughly speaking are consistent with the raining procedure described in \Sec{rain} 
+In this section $T$ will denote an arbitrary contour tree with every internal vertex of degree $3$.
+%and the height function defined over the vertex set is such that every internal vertex in either a split or join 
+%(i.e. at most $2$ neighbors lie above or below it).
+
+%We need to generalize the definition of path decompositions, which previously were only defined for rooted trees.
+%Again we define a path decomposition as a partition of the vertices into 
+%disjoint leaf paths (i.e. each path is monotonic in height and contains a leaf). 
+For simplicity we now restrict our attention to path decompositions 
+consistent with the raining procedure described in \Sec{rain} 
 (more general decompositions can work, but it is not needed for our purposes).
@@
 
-Consider our algorithm on a fixed $(f,\MM)$.  First we run $\rain(x,\MM)$ to partition the input 
-into extreme dominant complexes.  To make this partition uniquely defined lets say that the initial vertex $x$
-that rain is called from is the global maximum in the input (note that given a specific $x$ the partition is uniquely defined).
-Let $\MM_1, \dots, \MM_k$ be the set of produced subcomplexes, and let $f_i$ be the restriction of $f$ to $\MM_i$.
-Now on each of the $(f_i,\MM_i)$ we run the $\paint$ procedure to compute the join tree of $(f_i, \MM_i)$.
-As described in the previous section, this painting procedure defines a maximum path decomposition with 
-and associated cost, $\cost_{\jc}(f_i) = \sum_{p\in P_i} |p|\log |p|$.  
-It is not hard to see that the union of these paths decomposition $P= \cup_{i=1}^k P_i$ defines 
-a rain consistent path decomposition of the contour tree for $(f,\MM)$.  
-In particular for a fixed $\MM$, $f$ defines a specific path decomposition $P=P(f)$, which we call the \emph{rain consistent maximum path decomposition}, 
-and we write $\cost(f) = \sum_{p\in P} |p|\log |p|$.
+%
+%Consider our algorithm on a fixed $(f,\MM)$.  First we run $\rain(x,\MM)$ to partition the input 
+%into extreme dominant complexes.  To make this partition uniquely defined lets say that the initial vertex $x$
+%that rain is called from is the global maximum in the input (note that given a specific $x$ the partition is uniquely defined).
+%Let $\MM_1, \dots, \MM_k$ be the set of produced subcomplexes, and let $f_i$ be the restriction of $f$ to $\MM_i$.
+%Now on each of the $(f_i,\MM_i)$ we run the $\paint$ procedure to compute the join tree of $(f_i, \MM_i)$.
+%As described in the previous section, this painting procedure defines a maximum path decomposition with 
+%and associated cost, $\cost_{\jc}(f_i) = \sum_{p\in P_i} |p|\log |p|$.  
+%It is not hard to see that the union of these paths decomposition $P= \cup_{i=1}^k P_i$ defines 
+%a rain consistent path decomposition of the contour tree for $(f,\MM)$.  
+%In particular for a fixed $\MM$, $f$ defines a specific path decomposition $P=P(f)$, which we call the \emph{rain consistent maximum path decomposition}, 
+%and we write $\cost(f) = \sum_{p\in P} |p|\log |p|$.
 
@@
 \label{lem:pathDecomp2}
-Let $\mathcal{P} = P(f)$ be the rain consistent maximum path decomposition of some height function $f$ defined over a complex $\MM$. 
-Then there is a family of functions, $\bF$, all defined over a common triangulation in plane, 
-such that the size of $\bF$ is $\Pi_{p_i\in P} (|p_i|-1)!$, and no two function in $\bF$ define the same contour tree.
+Let $P$ be a rain consistent maximum path decomposition of some contour tree. 
+There is a family of terrains, $\bF$, all with the same triangulation, 
+such that the size of $\bF$ is $\Pi_{p_i\in P} (|p_i|-1)!$, and no two functions in $\bF$ define the same contour tree.
 \end{lemma}
 \begin{proof}
- As $\mathcal{P}$ is rain consistent, the paths can be partitioned into sets $P_1, \dots, P_k$, where $P_i$ is the set of all paths with 
+ As $P$ is rain consistent, the paths can be partitioned into sets $P_1, \dots, P_k$, where $P_i$ is the set of all paths with 
  vertices from a given BFS, as described in \Def{path2}.  Specifically, let $T_i$ be the subtree of $T$ corresponding to $P_i$ and 
@@
 \begin{lemma}
-Let $\mathcal{P}$ and $\bF$ be as defined above.  Then for all $f\in \bF$, $\cost(f) = \sum_{p\in P(f)} |p|\log|p| = \sum_{p\in \mathcal{P}} |p|\log |p|$
+For all $\MM \in \bF_P$, the number of heap operations is $\Theta(\sum_{p\in P} |p|\log |p|)$
 \end{lemma}
 \begin{proof}
- This lemma follows immediately from \Lem{cost}.  Specifically, the sum $\cost(P)$, for any rain consistent maximum path decomposition $P$, can be broken 
- into the sums $\cost_{\jc}(P_i)$ of some maximum path decompositions $P_1, \dots P_k$.  Now apply \Lem{cost} to each of the $P_i$ separately and take the sum.
-\end{proof}
-
-Call a cost $C$ valid if $C = cost(f)$ for some $f$ defined over a given complex $\MM$.
-For the same reason that \Thm{joinLB} followed from \Lem{pathDecomp} and \Lem{cost}, we now have the following.
-
-\begin{theorem}
-Consider the set of functions $\bF_C$ that all have the same valid cost $C$.  
-Any height function decision tree that correctly computes
-the contour tree on all of $\bF_C$, has an average case (uniform over $\bF_C$) running time of $\Omega(C)$.
-\end{theorem}
+ This lemma follows immediately from \Lem{cost}.  The heap operations can be partitioned into the
+ operations performed in each $P_i$.
+ Apply \Lem{cost} to each of the $P_i$ separately and take the sum.
+\end{proof}
+
+\Thm{main-lb} follows immediately from an entropy argument, analogous to \Thm{joinLB}.
+%
+%
+%Call a cost $C$ valid if $C = cost(f)$ for some $f$ defined over a given complex $\MM$.
+%For the same reason that \Thm{joinLB} followed from \Lem{pathDecomp} and \Lem{cost}, we now have the following. 
+%
+%\begin{theorem}
+%Consider the set of functions $\bF_C$ that all have the same valid cost $C$.  
+%Any height function decision tree that correctly computes
+%the contour tree on all of $\bF_C$, has an average case (uniform over $\bF_C$) running time of $\Omega(C)$.
+%\end{theorem}
 

### 2014-11-04 paint-contour-tree.tex
small change

--- 2014-11-04 paint-contour-tree.tex
@@
 \label{sec:lb}
+In this section we provide a lower bound for the time it takes to compute a contour tree in terms the cost of path decompositions.  
 We first prove a lower bound for join trees, and then generalize to contour trees.
 %\Ben{Hopefully the following sentence clear up confusion between $\MM$ and $(f,\MM)$}
-%Note earlier we usually used $\MM$ to denote the lift of a simplicial complex according to some height function $f$, 
-%however in this section we make a distinction between the complex $\MM$ and the height function $f$.  
-%In particular all previous definitions or subroutines which took $\MM$ as input now take $(f,\MM)$ as input.
+Note earlier we usually used $\MM$ to denote the lift of a simplicial complex according to some height function $f$, 
+however in this section we make a distinction between the complex $\MM$ and the height function $f$.  
+In particular all previous definitions or subroutines which took $\MM$ as input now take $(f,\MM)$ as input.
 
@@
 
-%In this section we prove \Thm{main-lb}, but for join trees instead of contour trees.  
+In this section we prove \Thm{main-lb}, but for join trees instead of contour trees.  
 %Observe that given a fixed $(f,\MM)$, 
@@
 %  
-We focus on terrains, so $d=2$.
-Consider any valid path decomposition $P$ of a valid join tree (actually, for any rooted binary tree,
-there exist a terrain for this is the contour tree).
-When we say ``compute the join tree", we require the join tree to
-be labeled with the corresponding vertices of the terrain.
-
-%
-%For given $\MM$, consider running the algorithm of \Sec{also} to compute $\jc(\MM)$.
-%Let $P(\jc(\MM))$ be the maximum path decomposition of the join tree according to the initial coloring
-%of $\init(\MM)$. Define $\cost_{\jc}(f) = \sum_{p\in P(\jc(\MM))} |p|\log|p|$
+Let $\MM$ be fixed.  For a given height function $f$, we can now define $\cost_{\jc}(f)$.  Specifically, consider running the $\build(f,\MM)$ procedure 
+of section \Sec{paint}, which computes the join tree $\jc(f,\MM)$.  This procedure calls $\init$, which will (uniquely) define some 
+initial coloring of $\jc$ (\Def{initialColoring}).  Let $P(f)=P(\jc(f,\MM))$ be the maximum path decomposition of the join tree according to this coloring. 
+Then $\cost_{\jc}(f) = \sum_{p\in P(f)} |p|\log|p|$
 
@@
 \label{lem:pathDecomp}
-Fix a valid path decomposition $P$. There is a family of terrains, $\bF_P$, all with the same triangulation, 
-such that $|\bF_P| = \Pi_{p_i\in P} (|p_i|-1)!$, and no two terrains in $\bF_P$ define the same join tree.
+Let $\mathcal{P} = P(f)$ be the maximum path decomposition of some height function $f$ defined over a complex $\MM$. 
+Then there is a family of functions, $\bF$, all defined over a common triangulation in plane, 
+such that the size of $\bF$ is $\Pi_{p_i\in P} (|p_i|-1)!$, and no two function in $\bF$ define the same join tree.
 \end{lemma}
-
 \begin{proof}
-% We describe a simple triangulation, with heights specified for all vertices except for the saddles.
-
-%The main idea is that 
-% this triangulation will be flexible enough to allow almost any assignment of heights to the saddles, thus defining a class of join trees 
-% whose size is that described in the lemma statement. 
- We describe the basic building block of these terrains, which corresponds to a fixed path $p\in P$.  
- Informally, a \emph{diamond} is an upside down cone with $m$ triangular faces (see \Fig{diamond}). 
- Construct a slightly tilted cycle 
- of length $m$ with the two antipodal points at heights $1$ and $0$. These are called the anchor and trap of the diamond, respectively.
+ We describe a simple triangulation, with heights specified for all vertices except for the saddles.  The main idea is that 
+ this triangulation will be flexible enough to allow almost any assignment of heights to the saddles, thus defining a class of join trees 
+ whose size is that described in the lemma statement.
+
+ First, we describe the basic building block of these terrains, which roughly speaking corresponds to a fixed path $p\in P$.  
+ Specifically, we say a \emph{diamond} of size $m$ is the following terrain.  To start, construct a slightly tilted cycle 
+ of length $m$.  Specifically have two antipodal points, at heights $1$ and $0$, which we will call the anchor and trap of the diamond, respectively.
  The remaining $m-2$ vertices are evenly spread around the cycle and heights decrease monotonically when going from the anchor to the trap.
- Next, create an apex vertex at some appropriately large height, and add an edge each vertex in the cycle.
-
- Now we describe how to attach two different diamonds.
-In this process, we glue the base of a scaled down ``child'' diamond on to a triangular cone face of the larger  ``parent'' diamond (see \Fig{diamond}).
- Specifically, the anchor of the child diamond is attached directly to a face of the parent diamond at some height $h$. The remaineder of the base of the child cone is then 
+ Next, create an apex vertex at some appropriately large height, and add an edge between it and each vertex in the cycle.  At this point 
+ we have built an upside down cone with $m$ triangular faces (see \Fig{diamond}).
+
+ Now we describe how to attach two diamonds.  In this process, we glue the base of a scaled down ``child'' diamond on to a triangular cone face of the larger  ``parent'' diamond (see \Fig{diamond}).
+ Specifically, the anchor of the child diamond is attached directly to a face of the parent diamond at some unspecified height $h$. The remained of the base of the child cone is then 
  extended down (at a slight angle) until it hits the face of the parent.
 
- The full terrain is obtained by repeatedly gluing diamonds. For each path $p_i\in P$, we create a diamond of size $|p_i|+1$.
+ At this point we can construct the entire terrain, by repeatedly gluing together diamonds.  
+ Specifically, for each path $p_i\in P$ we create a diamond of size $|p_i|+1$.
  The two faces adjacent to the anchor are always empty, and the remaining faces are for gluing on other diamonds.
@@
  (Naturally for this construction to work, diamonds for a given path will be scaled down relative to the size of the diamond of their parent).
- By varying the heights of the gluing, we get the family of terrains.
 
@@
 
- There is a valid instance of this described construction for any relative ordering of the heights of the saddles on a given diamond.  
+ Now there is a valid instance of this described construction for any relative ordering of the heights of the saddles on a given diamond.  
+ Let the set of height functions for these instances be denote $\bF$.
  In particular, there are $(|p|-1)!$ possible orderings of the heights of the saddles on the diamond for $p$, and hence $\Pi_{p_i\in P} (|p_i|-1)!$ possible 
- terrains we can build.  Each one of these functions will result in a different (labeled) join tree. All saddles on a given diamond will appear in sorted order in the join tree.
- So, any permutation of the heights on a given diamond corresponds to a permutation of the vertices along a path in $P$.
-\end{proof}
-
-Two path decompositions $P_1$ and $P_2$ (of potentially different complexes and/or height functions) are equivalent 
-if: there is a 1-1 correspondence between the sizes of the constituent paths, and the shrubs are isomorphic.
-%
-%
-%if their respective shrubs, $S_1$ and $S_2$, which has the property that for any node $v\in S_1$ corresponding to a path of 
-%size $|v|$, $f(v)$ also corresponds to a path of size $|v|$. Note that for equivalent decompositions, $\sum_{p\in P_1} |p|\log|p| = \sum_{p\in P_2} |p|\log|p|$.
+ functions in this set.  Also, each one of these functions will result in a different join tree.  Specifically, a permutation of the heights on a given diamond 
+ corresponds to a permutation of the vertices along a path in $P$, as by construction all saddles on a given diamond will appear in sorted order in the join tree.
+\end{proof}
+
+In the following we say to path decompositions $P_1$ and $P_2$ (of potentially different complexes and/or height functions) are equivalent 
+if there is an graph isomorphism, $f$, between their respective shrubs, $S_1$ and $S_2$, which has the property that for any node $v\in S_1$ corresponding to a path of 
+size $|v|$, $f(v)$ also corresponds to a path of size $|v|$. Note that for equivalent decompositions, $\sum_{p\in P_1} |p|\log|p| = \sum_{p\in P_2} |p|\log|p|$.
 
@@
 \label{lem:cost}
-For all $\MM \in \bF_P$, the total number of heap operations performed by $\build(\MM)$ is $\Theta(\sum_{p\in P} |p|\log|p|)$.
+Let $\mathcal{P}$ and $\bF$ be as defined above.  Then for all $f\in \bF$, $\cost_{\jc}(f) = \sum_{p\in P(f)} |p|\log|p| = \sum_{p\in \mathcal{P}} |p|\log |p|$
 \end{lemma}
 \begin{proof}
-%For any choice of $\MM \in \bF$, there is a path decomposition $\overline{P}(T)$ of the join tree equivalent to $\mathcal{P}$.
-%Specifically, this is the decomposition where each path $p\in \overline{P}(T)$ consists exactly of the set of critical points (except for its anchor) from a given diamond. 
-The primary ``non-determinism" of the algorithm is the initial painting constructed by $\init(\MM)$. 
-We show that regardless of how paint spilling is done, the number of heap operations is bounded as above.
-
- Consider an arbitrary order of the initial paint spilling over the surface defined by $f$.
- Consider any join on a face of some diamond, which is the anchor point of some connecting child diamond.  
- The join has two up-stars, each of which has exactly one edge. Each edge connects to a maxima
- and must be colored by that maxima.
- Hence, two colors touching this join (according to \Def{initialColoring})
+First observe that by construction, regardless of the choice of $f\in \bF$, there is a path decomposition $\overline{P}(T)$ of the join tree which is equivalent to $\mathcal{P}$.
+Specifically, this is the decomposition where each path $p\in \overline{P}(T)$ consists exactly of the set of critical points (except for its anchor) from a given diamond. 
+We now show that regardless of how paint spilling is done, this will be the maximum path decomposition of $T$ (implying the lemma statement).
+
+ So consider any order of paint spilling over the surface defined by $f$.
+ Consider any join on a face of some diamond.  This join is the anchor point of some connecting child diamond.  
+ The claim is that the two colors of this join (according to \Def{initialColoring})
  are the colors of the apexes of the child and parent diamond.  
-% 
-% Specifically, one of these edges connects directly to the apex of the parent diamond and so the join must get the color of this maxima.
-% The other edge connects directly to the apex of the child diamond (this is why we left the faces adjacent to an anchor empty) and 
-% so the second color is that of the maxima of the child diamond.
-
- Take any join $v$, with two children $w_1$ and $w_2$. Suppose $w_1$ and $v$ belong to the same path
- in the decomposition. The key is that any color from a maxima in the subtree at $w_2$ cannot touch
- any ancestor of $v$. This subtree is exactly the contour tree of the child diamond attached at $v$.
- The base of this diamond is completely contained in a face of the parent diamond. So all colors
- from the child ``drain off" to the base of the parent, and do not touch any joins on the parent diamond.
-
- Hence, $|H_v|$ is at most the size of the path in $P$ containing $v$. By \Lem{runTimeUpper}, the total number of heap operations
- is at most $\sum_v \log |H_v|$, completing the proof.
-
-% 
-% So let $c$ be the color of some maxima, $m$.  The above proven property of the coloring implies that in the join tree, the vertices for 
-% which $c$ appears are exactly the vertices of the path in  $\overline{P}(T)$ corresponding $m$, as well as 
-% the vertex to which the root of this path connects to (i.e. the anchor of the diamond of $m$).
-% Now we show this implies $\overline{P}(T)$ is a maximum path decomposition.
-% To see this, consider a vertex $v\in T$, and let $p$ be the path in $\overline{P}(T)$ which contain $v$. 
-% Now $v$ has two children (if it is a join).  Let $c_1$ be the child which is in $p$, and $c_2$ the child which is the root of some adjacent path $p'$.  
-% In order for $\overline{P}(T)$ to be the maximum path decomposition, we must show $|H_{c_1}|\geq |H_{c_2}|$.
-% %(Note, to ensure a unique maximum path decomposition, one can break ties consistently by assigning some arbitrary but fixed ordering to the vertices.)
-% %Now $v$ gets assigned two colors, one 
-% %of which we know is the color assigned to all vertices in $p$ and the other is the one assigned to all the vertices of $p'$. 
-% Observe that none of the colors in the subtree of $c_2$ appear below $v$.  Hence the heap at $c_2$ consists only of $v$.  
-% On the other hand the heap of $c_1$ contains $v$, any other vertex of $p$ below $v$, and the vertex to which the root of $p$ is adjacent to, 
-% and hence $|H_{c_1}|\geq |H_{c_2}|$.
+ This easily follows since this join has two up-stars, each of which has exactly one edge.  
+ Specifically, one of these edges connects directly to the apex of the parent diamond and so the join must get the color of this maxima.
+ The other edge connects directly to the apex of the child diamond (this is why we left the faces adjacent to an anchor empty) and 
+ so the second color is that of the maxima of the child diamond.
+
+ So let $c$ be the color of some maxima, $m$.  The above proven property of the coloring implies that in the join tree, the vertices for 
+ which $c$ appears are exactly the vertices of the path in  $\overline{P}(T)$ corresponding $m$, as well as 
+ the vertex to which the root of this path connects to (i.e. the anchor of the diamond of $m$).
+ Now we show this implies $\overline{P}(T)$ is a maximum path decomposition.
+ To see this, consider a vertex $v\in T$, and let $p$ be the path in $\overline{P}(T)$ which contain $v$. 
+ Now $v$ has two children (if it is a join).  Let $c_1$ be the child which is in $p$, and $c_2$ the child which is the root of some adjacent path $p'$.  
+ In order for $\overline{P}(T)$ to be the maximum path decomposition, we must show $|H_{c_1}|\geq |H_{c_2}|$.
+ %(Note, to ensure a unique maximum path decomposition, one can break ties consistently by assigning some arbitrary but fixed ordering to the vertices.)
+ %Now $v$ gets assigned two colors, one 
+ %of which we know is the color assigned to all vertices in $p$ and the other is the one assigned to all the vertices of $p'$. 
+ Observe that none of the colors in the subtree of $c_2$ appear below $v$.  Hence the heap at $c_2$ consists only of $v$.  
+ On the other hand the heap of $c_1$ contains $v$, any other vertex of $p$ below $v$, and the vertex to which the root of $p$ is adjacent to, 
+ and hence $|H_{c_1}|\geq |H_{c_2}|$.
 \end{proof}

### 2014-11-04 paint-contour-tree.tex
small edits

--- 2014-11-04 paint-contour-tree.tex
@@
 in running time. For a partition of the contour tree into a set of descending paths, $P$,
-our algorithm runs in $O(\sum_{p\in P} |p|\log|p| + t\alpha(t)+N)$.
+our algorithm runs in $O(\sum_{p\in P} |p|\log|p| + t\alpha(t)+N)$ time.
 This is at most $O(t\log D + N)$, where $D$ is the diameter of the contour tree.
@@
 decompositions for the time complexity analysis. There is a crucial use of a family of binomial heaps to 
-maintain priorities, ensuring that any comparison made is between comparable nodes of the contour tree.
+maintain priorities, ensuring that any comparison made is between nodes in the same path in the contour tree.
 We also prove lower bounds showing that the $\sum_{p\in P} |p|\log|p|$ complexity
@@
 First, lets review the various pieces of our algorithm. 
-On a given input simplicial complex, we first we break it into extremum dominant pieces using $\rain(\MM)$ (and in $O(|\MM|)$ time by \Thm{rain-time}).  
+On a given input simplicial complex, we first break it into extremum dominant pieces using $\rain(\MM)$ (and in $O(|\MM|)$ time by \Thm{rain-time}).  
 Specifically, \Lem{rain-1} proves that the output of $\rain(\MM)$ is a set of extremum dominant pieces, 
-$\MM_1, \dots, \MM_k$, and \Clm{rain-reeb} shows that given the contour trees, $\cC(\MM_1), \dots, \cC(\MM_k)$ , 
+$\MM_1, \dots, \MM_k$, and \Clm{rain-reeb} shows that given the contour trees, $\cC(\MM_1), \dots, \cC(\MM_k)$, 
 the full contour tree, $\cC(\MM)$, can be constructed (in $O(|\MM|)$ time).
@@
 We focus on terrains, so $d=2$.
-Consider any valid path decomposition $P$ of a valid join tree (actually, for any rooted binary tree,
-there exist a terrain for this is the contour tree).
+Consider any path decomposition $P$ of a valid join tree (i.e. any rooted binary tree).
 When we say ``compute the join tree", we require the join tree to
@@
 \label{lem:pathDecomp}
-Fix a valid path decomposition $P$. There is a family of terrains, $\bF_P$, all with the same triangulation, 
+Fix a path decomposition $P$. There is a family of terrains, $\bF_P$, all with the same triangulation, 
 such that $|\bF_P| = \Pi_{p_i\in P} (|p_i|-1)!$, and no two terrains in $\bF_P$ define the same join tree.
@@
  The remaining $m-2$ vertices are evenly spread around the cycle and heights decrease monotonically when going from the anchor to the trap.
- Next, create an apex vertex at some appropriately large height, and add an edge each vertex in the cycle.
+ Next, create an apex vertex at some appropriately large height, and add an edge to each vertex in the cycle.
 
@@
  There is a valid instance of this described construction for any relative ordering of the heights of the saddles on a given diamond.  
- In particular, there are $(|p|-1)!$ possible orderings of the heights of the saddles on the diamond for $p$, and hence $\Pi_{p_i\in P} (|p_i|-1)!$ possible 
+ In particular, there are $(|p_i|-1)!$ possible orderings of the heights of the saddles on the diamond for $p_i$, and hence $\Pi_{p_i\in P} (|p_i|-1)!$ possible 
  terrains we can build.  Each one of these functions will result in a different (labeled) join tree. All saddles on a given diamond will appear in sorted order in the join tree.
@@
 \label{lem:cost}
-For all $\MM \in \bF_P$, the total number of heap operations performed by $\build(\MM)$ is $\Theta(\sum_{p\in P} |p|\log|p|)$.
+For all $\MM \in \bF_P$, the total number of heap operations performed by $\build(\MM)$ is $O(\sum_{p\in P} |p|\log|p|)$.
 \end{lemma}
@@
 
- Consider an arbitrary order of the initial paint spilling over the surface defined by $f$.
+ Consider an arbitrary order of the initial paint spilling over the surface.
  Consider any join on a face of some diamond, which is the anchor point of some connecting child diamond.  
@@
  and must be colored by that maxima.
- Hence, two colors touching this join (according to \Def{initialColoring})
+ Hence, the two colors touching this join (according to \Def{initialColoring})
  are the colors of the apexes of the child and parent diamond.  
@@
 \begin{theorem}\label{thm:joinLB}
-Consider a rooted tree $T$ and an arbitrary path decomposition $P$ of $t$.
+Consider a rooted tree $T$ and an arbitrary path decomposition $P$ of $T$.
 There is a family $\bF_P$ of terrains such that any algebraic decision
 tree computing the contour tree (on $\bF_P$) requires $\Omega(\sum_{p \in P} |p|\log |p|)$ time.
-Furthermore, our algorithm makes $O(\sum_{p \in P} |p|\log |p|)$ on all these instances.
+Furthermore, our algorithm makes $O(\sum_{p \in P} |p|\log |p|)$ comparisons on all these instances.
 %
@@
 The proof is a basic entropy argument. Any algebraic decision tree that is correct on all of $\bF_P$
-must distinguish all inputs in this family. By Stirling's approximation, the average depth in this tree is $\Omega(\sum_{p_i\in P} |p_i|\log|p_i|)$.
+must distinguish all inputs in this family. By Stirling's approximation, the depth of this tree is $\Omega(\sum_{p_i\in P} |p_i|\log|p_i|)$.
 \Lem{cost} completes the proof.
@@
 Let $P$ be a rain consistent maximum path decomposition of some contour tree. 
-There is a family of terrains, $\bF$, all with the same triangulation, 
-such that the size of $\bF$ is $\Pi_{p_i\in P} (|p_i|-1)!$, and no two functions in $\bF$ define the same contour tree.
+There is a family of terrains, $\bF_P$, all with the same triangulation, 
+such that the size of $\bF_P$ is $\Pi_{p_i\in P} (|p_i|-1)!$, and no two terrains in $\bF_P$ define the same contour tree.
 \end{lemma}
@@
  So for each $P_i$ construct a terrain as described in \Lem{pathDecomp}.  Now each $T_i$ is (roughly speaking) a join or a split tree, depending on 
- whether the BFS which produced it was an upward or downward BFS, respectively.  As the construction in \Lem{pathDecomp} was for join trees, for each terrain 
+ whether the BFS which produced it was an upward or downward BFS, respectively.  As the construction in \Lem{pathDecomp} was for join trees, each terrain 
  we constructed for a $P_i$ which came from a split tree, must be flipped upside down.
@@
  Just as in \Lem{pathDecomp} we now have one fixed terrain structure, such that each different relative ordering of the heights of the join and split vertices 
- on each diamond produces a surface with a distinct join tree.  The specific bound on the size of $\bF$, defining these distinct join trees, follows by applying the 
+ on each diamond produces a surface with a distinct contour tree.  The specific bound on the size of $\bF_P$, defining these distinct contour trees, follows by applying the 
  bound from \Lem{pathDecomp} to each $P_i$.
@@
 
-\Thm{main-lb} follows immediately from an entropy argument, analogous to \Thm{joinLB}.
+\Thm{main-lb} now follows immediately from an entropy argument, analogous to \Thm{joinLB}.
 %

### 2014-11-04 paint-contour-tree.tex
valid changed

@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%\def \STOC {}
+\def \STOC {}
 
@@
 decompositions for the time complexity analysis. There is a crucial use of a family of binomial heaps to 
-maintain priorities, ensuring that any comparison made is between nodes in the same path in the contour tree.
+maintain priorities, ensuring that any comparison made is between nodes in the same descending path in the contour tree.
 We also prove lower bounds showing that the $\sum_{p\in P} |p|\log|p|$ complexity
@@
 of the lower bound requires some care. Consider the $d=2$ (terrain) case.
-Let a path decomposition $P$ be called \emph{valid} if it is actually output by the algorithm
-on some input.
+%
+%
+We define a class of valid path decompositions.  This include any decomposition of rooted trees,
+and more generally is a large superset of all decomposition produced by our 
+algorithm.
+%\footnote{Our construction likely works for any decomposition, however, 
+%to keep the presentation simple we restrict to a subset relevant for our algorithm.}
+%
+%
+%Let a path decomposition $P$ be called \emph{valid} if it is actually output by the algorithm
+%on some input.
+%
+%
 %As referenced in \Thm{main-alg}, for any $f:\MM \mapsto \RR$ our deterministic algorithm defines a 

### 2014-11-04 paint-contour-tree.tex
sentences added

@@
 
+In particular, any comparisons made are only between ancestors and descendants in the contour tree, where as previous algorithm compared everything, 
+and that is where we gain.
+
 Let us interpret the running time. 
@@
 %
-%
-We define a class of valid path decompositions.  This include any decomposition of rooted trees,
-and more generally is a large superset of all decomposition produced by our 
-algorithm.
+We define a class of valid path decompositions which characterizes decompositions produced by our algorithm.
+%
+%We define a class of valid path decompositions.  This include any decomposition of rooted trees,
+%and more generally is a large superset of all decomposition produced by our 
+%algorithm.
 %\footnote{Our construction likely works for any decomposition, however, 
 %to keep the presentation simple we restrict to a subset relevant for our algorithm.}
-%
 %

### 2014-11-04 paint-contour-tree.tex
changed sentence
final version

@@
 
-In particular, any comparisons made are only between ancestors and descendants in the contour tree, where as previous algorithm compared everything, 
+In particular, any comparisons made are only between ancestors and descendants in the contour tree, where as previous algorithms compared everything, 
 and that is where we gain.
@@
 %
-We define a class of valid path decompositions which characterizes decompositions produced by our algorithm.
-%
 %We define a class of valid path decompositions.  This include any decomposition of rooted trees,
@@
 %
-%Let a path decomposition $P$ be called \emph{valid} if it is actually output by the algorithm
-%on some input.
+Let a path decomposition $P$ be called \emph{valid} if it is actually output by the algorithm
+on some input.
 %
@@
 \label{lem:pathDecomp}
-Fix a path decomposition $P$. There is a family of terrains, $\bF_P$, all with the same triangulation, 
+Fix any path decomposition $P$. There is a family of terrains, $\bF_P$, all with the same triangulation, 
 such that $|\bF_P| = \Pi_{p_i\in P} (|p_i|-1)!$, and no two terrains in $\bF_P$ define the same join tree.
@@
 \label{lem:pathDecomp2}
-Let $P$ be a rain consistent maximum path decomposition of some contour tree. 
+Let $P$ be any rain consistent path decomposition of some contour tree. 
 There is a family of terrains, $\bF_P$, all with the same triangulation, 
@@
 
-\Thm{main-lb} now follows immediately from an entropy argument, analogous to \Thm{joinLB}.
+We now restate \Thm{main-lb}, which follows immediately from an entropy argument, analogous to \Thm{joinLB}.
+
+\begin{theorem} 
+Consider any rain consistent path decomposition $P$. There exists a family $\bF_P$ of terrains ($d=2$) with the following properties.
+Any contour tree algorithm makes $\Omega(\sum_{p \in P} |p|\log |p|)$ comparisons in the worst case over $\bF_P$.
+Furthermore, for any terrain in $\bF_P$, our algorithm makes $O(\sum_{p \in P} |p|\log |p|)$ comparisons.
+\end{theorem}
+
 %

### 2015-03-31 paint-contour-tree.tex
small change

@@
 is inherent to computing contour trees. 
-%Consider the family of instances where our algorithm
-%requires $C$ comparisons. Any algorithm computing contour trees on this family requires
-%$\Omega(C)$ comparisons in the worst case.
-
-%
-%general lower bound of
-%$\Omega(\sum_{p \in P} |p|\log |p|)$ for computing contour trees, showing
-%that the tree structure must play a role in the time complexity.
-
-\ignore{
-Moreover, we show there   
-exists some partition $P'$ such that the running time of our algorithm is 
-$O(\sum_{p\in P'} |p|\log|p| + N\alpha(N))$, where $N$ is the total complexity of the input manifold
-
-All previous contour tree algorithms involve a global sort 
-of the heights of all points or critical vertices. Thereby, these algorithms always pay $\Omega(t \log t)$.
-This is a major roadblock for efficiency, since sorting is the most expensive operation
-
-Reeb graphs are a fundamental topological structure which tracks the evolution of levels sets of a real valued function over a manifold.  
-In the case when this graph is loop free it is called a contour tree.  Due to the use of the contour trees in practice on large data sets, 
-there has been significant previous work in developing algorithms for their efficient computation.
-
-Here we present the first contour tree algorithm which takes into account the shape of the contour tree, 
-and in doing so yields instance optimal results.
-Specifically, we prove that for any partition of the contour tree into a set of descending paths, $P$, 
-any comparison based algorithm requires $\sum_{p\in P} |p|\log|p|$ time.  Moreover, we show there   
-exists some partition $P'$ such that the running time of our algorithm is 
-$O(\sum_{p\in P'} |p|\log|p| + N\alpha(N))$, where $N$ is the total complexity of the input manifold.
-In particular, our algorithm runs in $O(N\alpha(N))$ time when the tree is balanced, 
-while previous algorithms ran in $O(N+t\log t)$ (where $t$ is the size of the contour tree).
-%In particular, our algorithm runs in $O(N\alpha(N))$ time when the tree is balanced, yet still 
-%matches the previous worst case $O(N+n\log n)$ bound when the tree is long and skinny.
-
-In order to produce our instance optimal results we introduce a novel linear time surface partitioning scheme which is likely to have applications elsewhere.
-Moreover, in order to prove the optimality we use a non-trivial combinatorial discharging argument which we believe is of independent interest and in particular uses 
-such tools as heavy-light path decompositions.
-}
 \end{abstract}
@@
 containing a leaf, which is also monotonic in the height values of its vertices.
-%A leaf path runs either upwards or downwards, never both.
-%
-% Then in the following two subsections 
-%we relate path decomposition costs (defined below) to the running time of our algorithm using $\Lem{algCost}$.
 
@@
 of the lower bound requires some care. Consider the $d=2$ (terrain) case.
-%
-%We define a class of valid path decompositions.  This include any decomposition of rooted trees,
-%and more generally is a large superset of all decomposition produced by our 
-%algorithm.
-%\footnote{Our construction likely works for any decomposition, however, 
-%to keep the presentation simple we restrict to a subset relevant for our algorithm.}
-%
 Let a path decomposition $P$ be called \emph{valid} if it is actually output by the algorithm
 on some input.
-%
-%
-%As referenced in \Thm{main-alg}, for any $f:\MM \mapsto \RR$ our deterministic algorithm defines a 
-%specific path decomposition $P(T)$%
-\ignore{\FullVer{
-\footnote{The specific decomposition used is defined in \Sec{pathDecomp}, 
-but also depends on the raining and painting procedures described in \Sec{rain} and \Sec{paint}.  
-The important thing is that our algorithm is deterministic, and so for a fixed $\MM$ and $f$ it is well defined.}.}
-\SubmitVer{
-\footnote{The specific decomposition used is defined in the full version of the paper.  
-The important thing is that our algorithm is deterministic, and so for a fixed $\MM$ and $f$ it is well defined.}.}
-We define $\cost(f) = \sum_{p \in P(T)} |p|\log |p|$ for this specific decomposition $P(T)$.
-Ideally, we would like to argue that any contour-tree algorithm
-must pay $\cost(f)$ for ant $f$. 
-But this is obviously false, since one could consider an algorithm
-that hard codes the answer for $f$ and merely checks this answer.
-
-Instead, call a possible cost $C$ \emph{valid} if $C = \cost(f)$ for some $f$\footnote{If one is concerned
-that $C$ might be irrational, define valid if $2^C = 2^{\cost(f)}$.}. We prove
-a refined lower bound matching the performance of \Thm{main-alg}. 
-Roughly speaking, it is not possible to improve on \Thm{main-alg} even when restricted to a subset
-of inputs of fixed cost.
-
-%\begin{definition}\label{def:hfdt}
-%A \emph{height function decision tree} is an algorithm
-%that is allowed arbitrary computation on the domain $\MM$, and is an algebraic decision
-%tree over the heights. The running time is simply the number of queries made.
-%\end{definition}
-%
-}
+
 \begin{theorem} 
@@
 requires this cost in the worst case, and our algorithm matches this bound.
-
-% We complement \Thm{main-alg} with an almost matching lower bound, showing that the complexity
-% of $\sum_{p \in P(T)} |p|\log |p|$ is required for any contour tree algorithm. The exact phrasing
-% of the lower bound requires some care. Consider the $d=2$ (terrain) case, and fix a mesh $\MM$.
-% Now any height function $f$ (together with $\MM$) defines a specific contour tree $T$.
-% As referenced in \Thm{main-alg}, given $f$, our algorithm then defines some specific path 
-% decomposition $\overline{P}(T)$, such that the running time of our algorithm is bounded by 
-% $O(\cost(\overline{P}(T))+t\alpha(t)+N)$, where $\cost(P(T)) = \sum_{p \in P(T)} |p|\log |p|$.
-% A priori, which path decomposition this will be is unclear\footnote{The specific decomposition, defined in \Sec{pathDecomp}, 
-% depends on arbitrary choices made during the raining and painting procedures described in \Sec{rain} and \Sec{paint}.}, 
-% however we know there is some large set of potential path decompositions it lies in
-% (the exact set is defined in \Def{path2}, but for now one can think of the set of all decompositions).
-% Let $\costs(f)$ denote the set of costs of all these potential path decompositions that $f$ defines
-% (i.e. $\cost(\overline{P}(T))\in \costs(f)$).
-% 
-% Ideally, we would like to argue that any contour-tree algorithm must pay $\cost(\overline{P}(T))$ for $f$. 
-% But this is obviously false, since one could consider an algorithm
-% that hard codes the answer for $f$ and merely checks this answer.
-% Instead, call a possible cost $C$ \emph{valid} if $C \in \costs(f)$ for some $f$ 
-% defined over the fixed mesh $\MM$\footnote{If one is concerned
-% that $C$ might be irrational, define valid if $2^C = 2^{\cost(f)}$.}. We prove
-% a refined lower bound matching \Thm{main-alg}. 
-% 
-% \begin{definition} \label{def:hfdt}
-% A \emph{height function decision tree} is an algorithm
-% that is allowed arbitrary computation on the domain $\MM$, and is an algebraic decision
-% tree over the heights. The running time is simply the number of queries made.
-% \end{definition}
-% 
-% \begin{theorem} \label{thm:main-lb} Consider the set of functions $\bF_C$ (defined over a fixed mesh $\MM$) 
-% that all have the same value $C$ as a valid cost.  Any height function decision tree that correctly computes
-% the contour tree on all of $\bF_C$ has an average case (uniform over $\bF_C$) running time of $\Omega(C)$.
-% \end{theorem}
 
@@
 at the relationship between sorting and the contour tree problem.
-
-\ignore{
-classic output sensitive convex hull algorithms of Kirkpatrick and Seidel~\cite{KiSe86} and Chan~\cite{c-ooscha-96}.
-
-\paragraph{Output Sensitive Analysis.}
-In geometric settings often the goal is to compute a (potentially complex) geometric structure determined 
-by the input, and so it is natural to also consider how the form of this output structure affects the running time.
-One example is the standard sweep line algorithm of Bentley and Ottmann \cite{bo-arcgi-79} for line segment intersection, 
-which runs $O((n+k)\log n)$ time where k is the number of intersections, which in the worst case is quadratic.
-Another example is Chan's algorithm for computing the convex hull \cite{c-ooscha-96} which runs in $O(n\log h)$ 
-where $h$ is the number of vertices on the hull, which in the worst case is $n$.  While such algorithms represent 
-significant improvements over their worst case counterparts, their running times are only measured with respect 
-to the size of the output and not any structure it may have.
-
-\paragraph{Instance Optimality.}  
-Here we are concerned with a more refined analysis than the typical output sensitive analysis.
-Specifically, we investigate an algorithm which is instance optimal as defined by Afshani \etal \cite{abc-ioga-09}.  
-In order to prove such a result, one considers a family of allowable partitions of the input, 
-each of which is defined in terms of the structure of the output.  Each such partition then has 
-an associated entropy type cost representing the number of possible instantiations of such a partition.
-Our work differs slightly from Afshani \etal \cite{abc-ioga-09} however, in that they show any algorithm 
-(for say the convex hull) requires the minimum such cost, where as in our case it is shown that 
-any algorithm requires the maximum over these costs.
-}
 
@@
 We can think of $f(x)$ as the ``height" of a point $x \in \MM$, and encoded in the representation of $\MM$.
-%Abusing notation, $\MM$ refers to this encoding, so we do not refer to $f$.
 
@@
 $f|B_\eps(x)$ be the restriction of $f$ to $B_\eps(x)$.
-
-% \begin{definition} \label{def:deg} The up-degree of $x \in \MM$ is the number of $(f(x) + \delta)$-contours that intersect $B_\eps(x)$
-% as $\delta, \eps \rightarrow 0^+$. The down-degree is the number of $(f(x) - \delta)$-contours that intersect $B_\eps(x)$
-% as $\delta, \eps \rightarrow 0^+$.
-% \end{definition}
 
@@
 We abuse notation and refer to this entire set of vertices as a maxima or minima.
-%In particular, we do not view this as a violation of the Morse condition on $f$, which requires distinct heights for critical points.
-
-\ignore{
-The Morse assumption
-implies that all critical points have distinct values. A contour that involves a critical
-point is also denoted as critical. Note that the vertices in a boundary face collectively behaves like an extremum.
-Abusing notation, the term ``critical point", ``maxima", and ``minima" will include boundaries.
-Critical values include these boundary values. By boundary criticality, all contours are themselves
-represented as $d-1$-dimensional simplicial complexes.
-}
 
@@
 set of vertices, though we will not delve into these aspects in this paper.
-
-\ignore{
-\begin{remark}
- Note that \Def{deg} can be changed so that the up-degree (resp. down-degree) is the number of $(f(x) + \delta)$-contours (resp. $(f(x) - \delta)$-contours) that intersect $B_\eps(x)$.
- For $2$-manifolds, the two definitions are equivalent, and at each saddle point either two contours join together or split apart, and hence the terminology.  
- In higher dimensions though these definitions differ.  Specifically, this alternative definition only includes points at which the number of contours in a level set change.
- These points are a subset of our defined critical points which also include points at which the topology of the contours change (and hence the term `critical' is still applicable).
- Note that our critical point definition allows one to classify all vertices (according to \Def{points}) simultaneously in linear time, as it is a local neighborhood condition.
-\end{remark}
-
-\begin{remark}
-Note that the contour tree as defined above contains vertices for all critical points, 
-which differs slightly from the usual definition which only contains critical points corresponding to a change in the number of contours.
-Specifically, if a vertex in the tree we defined has degree $2$ then only the topology of a contour is changing rather than the number of contour.
-As such the standard contour tree can be computed from ours by a linear time pass to contract maximal degree 2 vertex paths.
-
-For simplicity we will also make the standard assumption that vertices in the contour 
-tree have degree at most $3$, i.e. there are no multi-saddles.
-\end{remark}
-}
 
@@
 
-%
-%{\bf Punchline:} We give a linear time procedure that partitions $\MM$ into smaller pieces. We guarantee
-%that the contour tree of each individual piece is basically the join (or split) tree. This saves
-%some of the redundant computation. We then show to put it all together the get the full contour tree.
-%Now for more details.
-
 \paragraph{Contour surgery:} We first need a divide-and-conquer lemma. Any contour $\phi$ can be associated
@@
 \end{definition}
-%
-%Paintings are not unique. It is straightforward to construct an initial painting of $\MM$ in $O(|\MM|)$ time, 
-%using a descending BFS from each maximum that does not explore previously colored edges. 
-%As the algorithm proceeds, the bijection property will not hold, though we always maintain a valid painting.
 
@@
 
-\ignore{
-\medskip
-\fbox{
-\begin{minipage}{0.9\textwidth}
-{\bf $\cut(\phi,\MM)$}
-
-\smallskip
-\begin{asparaenum}
-%	\item Insert the vertices of $\phi$ in $\MM$.
-	\item For every edge $e$ that intersects $\phi$ at point $x$: add edges $(e,x^+)$ and $(e,x^-)$
-	where $x^+$ (resp. $x^-$) is infinitesimally above (resp. below) $x$.
-	\item Take a point $x$ (resp. $y$) on $e$ at distance $\eps$ above (resp. below) $\phi$.
-	\item Let $\phi_x$ be the contour through $x$, and analogously define $\phi_y$. Insert a vertex
-	at each intersection point of these contours with edges of $\MM$. Add  
-	\item Delete the edges between $\phi_x$ and $\phi_y$.
-\end{asparaenum}
-\end{minipage}}
-
-\medskip
-This results in a new boundary critical simplicial complex. There are two new disjoint
-boundary simplices added. The manifold may be disconnected. 
-}
-
 We now describe a high-level approach to construct $\reeb(\MM)$.
@@
 
-\ignore{
-\begin{proof} If $y \in \wet(x,\MM)$, all the points in any contour containing $y$ are also in $\wet(x,\MM)$.
- (Follow the non-ascending path to from $x$ to $y$ and then walk along the contour.)
- First observe that any point which in the interface must be wet.  Second, for $y \in \wet(x,\MM)$, all the points in any contour containing $y$ are also in $\wet(x,\MM)$, 
- as one can follow the non-ascending path to $y$ and then walk along the contour.
-
- So let $y$ be a point in the interface.
-
- However, since $y$ is in the interface, it is also wet.  Therefore for any contour $\phi$ which contains $y$, for any arbitrarily small $\eps$ there is 
- a wet point which lies above $\phi$ and has a descending path of length at most $\eps$ to $\phi$.
- This implies there is some contour which lies just 
- above $y$ which is wet.  So any contour $\phi$ containing $y$ has two contours which lie just above it, one which is dry, and one which is wet.  
- This can only happen if $\phi$ contains a merge vertex, because otherwise all contours which lie just above would be in the same L-homotopy equivalence class and so by 
- \Clm{mono}, if one is wet, all below it are wet.
-
- By similar logic one can argue that of the two contours through a given merge vertex at most one can contain interface points other than the merge vertex, 
- and that if one of the contours does contain such interface points, then the entire contour consists of interface points.
-\end{proof}
-}
-
 Note that $\wet(x,\MM)$ (and its interface) can be computed in time linear in the size of the wet simplicial complex.
@@
 to cutting $F$ by a hyperplane. All these operations can be performed to output $\wet(x,\MM)$ in time linear in $|\wet(x,\MM)|$.
-
-% \begin{proof} If $y \in \wet(x,\MM)$, but not in the interface, then any contour containing $y$ is also in $\wet(x,\MM)$.
-% So consider a point $y$ in the interface. We argue that all points in a contour $\phi$ containing $y$
-% are also in the interface. If such a contour is regular, we can find a contour just above it that
-% is also wet. 
-% 
-% If two contours intersect, the intersection point $y$ is critical. We can argue that the only
-% way to reach $y$ from a descending path is reach one of the contours and then follow it to get to $y$.
-% But that would imply the existence of another critical point on this contour, violating the Morse condition.
-% 
-% Let us argue that $y$ is a merge vertex. Consider the descending path from $x$ to $y$. The final edge
-% that reaches $y$ is obviously all wet, and is an increasing edge from $y$. Consider an opposite increasing
-% edge. Take two point in these edges that are infinitesimally higher than $y$ and show that the contours
-% through them as disjoint. So $y$ is a merge.
-% \end{proof}
 
@@
 
-\ignore{
-As each original vertex in $\NN$ appears in only one of the sub-manifolds output by $\rain(x,\NN)$, we have the following corollary.
-
-\begin{corollary}
-The total size of a sub-manifolds output by $\rain(x,\MM)$ is $O(n)$.
-\end{corollary}
-
-\begin{corollary} \label{cor:rain-time} The total running time of $\rain(x,\MM)$ is $O(n)$.
-\end{corollary}
-
-\begin{proof} Consider some invocation to $\rain(x,\MM)$. The only non-trivial work done in $\rain$ is by the calls to $\wet$.
-We know that each invocation to $\wet$ runs in time linear 
-to the portion of the manifold which is wet (including the new boundary vertices created).  
-As the output of $\rain$ is precisely the distinct sub-manifolds wet by each distinct call to $\wet$, the claim follows.
-\end{proof}
-
-%Consider one of the manifolds $\MM_i$ output by $\rain$.  This manifold has a set of boundary maxima/minima that are the result 
-%of cutting along interface contours.  For each such contour there is exactly one other output manifold $\MM_j$ for which 
-%the contour appears as a boundary maxima/minima. 
-}
-
 \begin{claim} \label{clm:rain-reeb} Given $\reeb(\MM_1), \reeb(\MM_2), \ldots, \reeb(\MM_r)$, 
@@
 labeled with some $\MM_i$.
-%
-%There is a natural tree ordering defined over these manifolds where the 
-%root corresponds the first manifold that was wet when we rained from $x$, and its children correspond to the recursive calls 
-%(i.e. where we rained from the interface contours of the first raining).  Call this the \emph{Manifold Ordering Tree}.
-%$\rain$ can easily be modified to also output this tree (without affecting the asymptotic running time).  
 Walk through this tree in a leaf first ordering.  Each time we visit a node we connect its contour tree to the 
@@
 Each $\surgery$ call is constant time, and the total time is the size of the recursion tree.
-%
-%
-%This takes $O(n)$ time overall by \Lem{new-verts} and since $\surgery$ takes constant time (given the appropriate vertices in each reeb graph).
-The correctness follows by induction and  \Lem{surgery}.
 \end{proof}
@@
 
-%
-%As part of the initialization, we determine
-%all critical points in $\MM$. For any vertex $v$, we can look at $f(v) \pm \delta$-contours 
-%in $f|_{B_\eps(v)}$ to determine the up and down degrees. This determines the criticality (and type)
-%of $v$. This requires $O(|\MM|)$ time, since every facet is looked at proportional to the number of its
-%vertices.
-
 {\bf Painting:} The central tool is a notion 
@@
 
-\ignore{
-Abusing notation, we say a contour $\phi$ is of color $c$ if all edges that intersect $\phi$ have color $c$.
-
-\begin{definition} \label{def:color-full} Consider vertex $v$. An \emph{up-star} of $v$ is the set 
-of edges that all connected to a fixed component of $\MM^+_v$. 
-%and let $\phi$ be a $(f(v)+\delta)$-contour in
-%$f|_{B_\eps(v)}$. If $\phi$ is monochromatic of color $c$, then $v$ is \emph{fully touched by $c$}. The set of colors
-%that fully touch $v$ is denoted $\fcol(v)$.
-\end{definition}
-
-%Note that $\fcol(v)$ has size at most $2$, by the Morse condition. 
-Note that $\col(v)$ 
-%and $\fcol(v)$ can
-be initially determined in $O(|\MM|)$ time, by a simple linear scan.
-}
-
 \subsection{The data structures} \label{sec:struct}
@@
 both to the set and the data structure used to store it.
-
-%We initialize as follows. For each critical vertex $v$ and from each of its up-stars, we pick an arbitrary color $c$ touching $v$. 
-%We insert $v$ into $T(c)$. Observe that $v$ participates in at least $1$ and at most $2$ heaps.
-%
 
@@
 %This is initialized to the maximum corresponding to $c$.
-
-\ignore{
-\medskip
-As the algorithm proceeds, it is convenient to imagine
-it maintaining a set of contours $\Phi = \{\phi_1, \phi_2, \ldots, \phi_k\}$, which are contained
-in incomparable edges of $\cC(\MM)$. Suppose we cut $\MM$ by each of these contours, and took
-the simplicial complex $\NN$ where $\phi^-_i$'s are all boundary maxima. The algorithm will
-have already determined the contour trees of the simplicial complexes
-where $\phi^+_i$ is a boundary maxima.
-
-During this process, various colors will be \emph{merged}. This involves merging the respective $T(c)$'s and
-updating the union-find data structure. The only operation performed on $T(c)$'s are delete-max and merge.
-No new vertex is ever inserted into the heaps. As colors merge, the initial painting is obviously modified.
-But it would be too expensive to explicitly update colors. So the new painting is given by $\rep(c)$,
-while we only have access to the original painting. By union-find bounds, this only adds an $\alpha(|\MM|)$
-overhead to the data structure operations. The new coloring is guaranteed to be a painting of $\NN$.
-
-%While this is the intuitive viewpoint, we describe how this information is actually maintained.
-The vertices $\{\h(\rep(c))\}$ represent the set of contours $\Phi$. In terms of the contour tree, 
-we have already constructed the contour trees rooted at each $h(\rep(c))$. This is trivially true at the initialization,
-when each $h(c)$ is a maximum. One can now treat each $h(\rep(c))$ as a maximum, corresponding to (a possibly new) color $\rep(c)$.
-To reduce confusion between the original painting and the new colors, we will assume that $\col(v)$ and $\fcol(v)$
-are as given by the original painting. 
-}
 
@@
 exists critical vertex $w \in \PP$ fully touched by $c$ and touched by another color.
-%
-%and take $(f(v)+\delta)$-contour $\phi$ intersecting
-%$B_\eps(v)$. Suppose the edges incident to $v$ intersecting $\phi$ do not have the same color. For any such
-%color $c$, there exists a critical vertex $w$ strictly higher than $v$ such that $c \in \fcol(v)$.
 \end{lemma}
@@
 If $p$ is short, $\sum_{v\in p} \log |H_v| = O(|H_p| \log |H_p|)$.
-% Let T be a binary tree and let $\chi(T)$ and $P(T)$ be a valid coloring and path decomposition of $T$, respectively.
-% Then, 
-% \begin{compactenum}
-%  \item for any tall path $p\in P(T)$, $\sum_{v\in p} \log |H_v| = O(|p| \log |p|)$.
-%  \item for any short path $p\in P(T)$, $\sum_{v\in p} \log |H_v| = O(|p| \log |H_{p}|)$.
-% \end{compactenum}
 \end{lemma}
@@
 We begin with a key property, which is where the construction of $\pmax(T)$ enters the picture.
-
-%
-%
-%
-%We focus on $\pmax(T)$, and prove the key property of this path decomposition.
 
@@
 We first prove a lower bound for join trees, and then generalize to contour trees.
-%\Ben{Hopefully the following sentence clear up confusion between $\MM$ and $(f,\MM)$}
-%Note earlier we usually used $\MM$ to denote the lift of a simplicial complex according to some height function $f$, 
-%however in this section we make a distinction between the complex $\MM$ and the height function $f$.  
-%In particular all previous definitions or subroutines which took $\MM$ as input now take $(f,\MM)$ as input.
 
@@
 
-%In this section we prove \Thm{main-lb}, but for join trees instead of contour trees.  
-%Observe that given a fixed $(f,\MM)$, 
-%our algorithm for computing $\jc$ is deterministic and so the maximum path decomposition $P=P(\jc)$ defined by the initial coloring (\Def{initialColoring}) is fixed, 
-%allowing us to define the cost, $\cost(f)=\sum_{p\in P} |p|\log|p|$.
-%  
 We focus on terrains, so $d=2$.
@@
 be labeled with the corresponding vertices of the terrain.
-
-%
-%For given $\MM$, consider running the algorithm of \Sec{also} to compute $\jc(\MM)$.
-%Let $P(\jc(\MM))$ be the maximum path decomposition of the join tree according to the initial coloring
-%of $\init(\MM)$. Define $\cost_{\jc}(f) = \sum_{p\in P(\jc(\MM))} |p|\log|p|$
 
@@
 
-% \begin{figure}[h]\centering
-%     \includegraphics[width=.5\linewidth]{figs/bridge2}
-%     \caption{On the left, the root of $P_i$ merges into a diamond of its parent $P_j$.  On the right, the root of $P_i$ splits off from a diamond of $P_j$}
-%     \label{fig:diamond}
-% \end{figure}
-
  So for each $P_i$ construct a terrain as described in \Lem{pathDecomp}.  Now each $T_i$ is (roughly speaking) a join or a split tree, depending on 

### 2015-04-01 paint-contour-tree.tex
Updated abstract

@@
   \and
-  C. Seshadhri\thanks{Sandia National Laboratories; Livermore, CA, USA; {\tt scomand@sandia.gov}}
+  C. Seshadhri\thanks{University of California, Santa Cruz; {\tt scomandu@ucsc.edu}}
 }
@@
 \begin{abstract}
-Consider a scalar field $f:\MM \mapsto \RR$, where $\MM$ is a triangulated simplicial mesh in $\RR^d$. 
-A level set, or contour, at value $v$ is a connected component of $f^{-1}(v)$. 
-As $v$ is changed, these contours change topology, merge into each other, or split.
-\emph{Contour trees} are concise representations of $f$ that track this contour
-behavior. The vertices of these trees are the critical points of $f$, where the gradient is zero.
-The edges represent changes in the topology of contours.
-It is a fundamental data structure in data analysis and visualization,
-and there is significant previous work (both theoretical and practical) on algorithms for constructing contour trees.
-
-Suppose $\MM$ has $n$ vertices, $N$ facets, and $t$ critical points.
-A classic result of Carr, Snoeyink, and Axen (2000) gives an algorithm that takes $O(n\log n + N\alpha(N))$ time
-(where $\alpha(\cdot)$ is the inverse Ackermann function). A further improvement to $O(t\log t + N)$ time
-was given by Chiang et al. All these algorithms involve a global sort of the critical points,
-a significant computational bottleneck. Unfortunately, lower bounds of $\Omega(t\log t)$ also exist.
-
-We present the first algorithm that can avoid the global sort and has a refined time complexity that
-depends on the contour tree structure. Intuitively, if the tree is short and fat, we get significant improvements 
-in running time. For a partition of the contour tree into a set of descending paths, $P$,
-our algorithm runs in $O(\sum_{p\in P} |p|\log|p| + t\alpha(t)+N)$ time.
-This is at most $O(t\log D + N)$, where $D$ is the diameter of the contour tree.
-Moreover, it is $O(t\alpha(t)+N)$ for balanced trees, a significant improvement over the previous complexity.
-
-Our algorithm requires numerous ideas: partitioning the contour tree into join and split trees, 
+We revisit the classic problem of computing the \emph{contour tree}
+of a scalar field $f:\MM \mapsto \RR$, where $\MM$ is a triangulated simplicial mesh in $\RR^d$. 
+% A level set, or contour, at value $v$ is a connected component of $f^{-1}(v)$. 
+% As $v$ is changed, these contours change topology, merge into each other, or split.
+The contour tree is a fundamental topological data structure that tracks
+the evolution of level sets of $f$ and 
+% concise representations of $f$ that track the evolution
+% of levels sets, and this contour
+% behavior. The vertices of these trees are the critical points of $f$, where the gradient is zero.
+% The edges represent changes in the topology of contours.
+has numerous applications in data analysis and visualization,
+
+Roughly speaking, all existing algorithms require $\Omega(n\log n)$ time for all instances, since
+the first step is a global sort of the values of $f$ (technically, only the critical points must be sorted,
+but these can be $\Omega(n)$). 
+Existing lower bounds show that there are pathological instances where this sort is required.
+We present the first input-sensitive algorithm whose time complexity depends
+on the contour tree struture, and avoids the global sort for non-pathological inputs.
+Our results match all existing upper bounds, but also
+characterize the amount of sorting required to compute the contour tree.
+This leads to a linear time algorithm for contour tree computation for many
+classes of natural instances. We complement with matching lower bounds for contour tree computation.
+
+Our algorithm requires several novel ideas: partitioning $\MM$ in well-behaved portions, 
 a local growing procedure to iteratively build contour trees, and the use of heavy path 
-decompositions for the time complexity analysis. There is a crucial use of a family of binomial heaps to 
+decompositions for the time complexity analysis. 
+There is a crucial use of a family of binomial heaps to 
 maintain priorities, ensuring that any comparison made is between nodes in the same descending path in the contour tree.
-We also prove lower bounds showing that the $\sum_{p\in P} |p|\log|p|$ complexity
-is inherent to computing contour trees. 
+
+% Our results demonstrate that this sort can be avoided for many instances, and
+% also characterizes these hard instances.
+% At a high level, if the tree is short and fat, we get significant improvements in running time.
+% 
+% 
+% 
+% 
+% and there is significant previous work (both theoretical and practical) on algorithms for constructing contour trees.
+% Suppose $\MM$ has $n$ vertices, $N$ facets, and $t$ critical points.
+% The best known algorithm for constructing the contour tree requires $O(t\log t + N)$ time, and all algorithms to date
+% involve a global sort of the critical points. Unfortunately, worst case lower bounds of $\Omega(t\log t)$ also exist.
+% A classic result of Carr, Snoeyink, and Axen (2000) gives an algorithm that takes $O(n\log n + N\alpha(N))$ time
+% (where $\alpha(\cdot)$ is the inverse Ackermann function). A further improvement to $O(t\log t + N)$ time
+% was given by Chiang et al. All these algorithms involve a global sort of the critical points,
+% a significant computational bottleneck. Unfortunately, lower bounds of $\Omega(t\log t)$ also exist.
+
+% We present the first algorithm that can avoid the global sort and has a refined time complexity that
+% depends on the contour tree structure. Intuitively, if the tree is short and fat, we get significant improvements 
+% in running time. For a partition of the contour tree into a set of descending paths, $P$,
+% our algorithm runs in $O(\sum_{p\in P} |p|\log|p| + t\alpha(t)+N)$ time.
+% This subsumes all existing upper bounds, and is a significant improvement 
+% This is at most $O(t\log D + N)$, where $D$ is the diameter of the contour tree.
+% Moreover, it is $O(t\alpha(t)+N)$ for balanced trees, a significant improvement over the previous complexity.
+
+% We also prove lower bounds showing that the $\sum_{p\in P} |p|\log|p|$ complexity
+% is inherent to computing contour trees. 
 \end{abstract}

### 2015-04-01 paint-contour-tree.tex
More changes

@@
 
-It is convenient to think of $f$ as a $(d+1)$-dimensional manifold, or alternately consider
-$\MM$ in $\RR^{d+1}$ with the last coordinate (i.e. height) given by $f$. Imagine sweeping the hyperplane $x_{d+1} = h$
+It is convenient to think of $f$ as a $(d+1)$-dimensional manifold
+%  or alternately consider
+% $\MM$ in $\RR^{d+1}$
+with the last coordinate (i.e. height) given by $f$. Imagine sweeping the hyperplane $x_{d+1} = h$
 with $h$ going from $+\infty$ to $-\infty$. At every instance, the intersection of this plane 
@@
 is $3$. A fundamental result in this area is the algorithm of Carr, Snoeyink, and Axen to compute contour trees,
-which runs in $O(n\log n + N\alpha(N))$ time~\cite{csa-cctad-00}. (Henceforth, $\alpha(\cdot)$ denotes the inverse
-Ackermann function.) It is natural to think of $N$ as $\Theta(n)$
-in practical applications (certainly true for $d=2$). The most expensive operation is an initial sort of all the vertex
+which runs in $O(n\log n + N\alpha(N))$ time~\cite{csa-cctad-00} (where $\alpha(\cdot)$ denotes the inverse
+Ackermann function). In practical applications, $N$ is typically $\Theta(n)$
+(certainly true for $d=2$). The most expensive operation is an initial sort of all the vertex
 heights. Chiang et al build on this approach to get a faster algorithm that only sorts the critical vertices,
@@
 
-But this is only a worst-case bound. All previous algorithm begin by sorting (at least) the critical points, and
-hence must pay $\Omega(t\log t)$ for all instances. Can we beat this sorting bound for some instances? Is there a more refined version of the complexity
-of computing contour trees, and can we characterize what inputs are hard? Our main result gives an affirmative answer. First, we require some technical definitions.
+But this is only a worst-case bound. All previous algorithms begin by sorting (at least) the critical points, and
+hence must pay $\Omega(t\log t)$ for all instances. Can we beat this sorting bound for some instances
+and characterize what inputs are hard? Our main result gives an affirmative answer. First, we require some technical definitions.
 
@@
 
-In particular, any comparisons made are only between ancestors and descendants in the contour tree, where as previous algorithms compared everything, 
-and that is where we gain.
+In particular, any comparisons made are only between ancestors and descendants in the contour tree, while previous algorithms compared everything.
 
@@
 so the running time is $O(t\log t + N)$. But this is tight only when $T$ is tall and thin.  
-A better but still obvious upper bound is $O(t\log D + t\alpha(t) + N)$, where $D$ is the diameter of $T$. 
+A better upper bound is $O(t\log D + t\alpha(t) + N)$, where $D$ is the diameter of $T$. 
 This is clearly an improvement when the tree is short and fat.
@@
 \paragraph{Beyond worst-case analysis:} Our result provides understanding of the complexity
-of contour tree construction beyond the standard worst-case analysis. Our theorem and lower bound explain \emph{when}
-computing contour trees is expensive, and provide insight into the hardness of particular instances. 
+of contour tree construction beyond the standard worst-case analysis. Our theorem and lower bound 
+characterize hard instances.
+% explain \emph{when}
+% computing contour trees is expensive, and provide insight into the hardness of particular instances. 
 %It is intriguing
@@
 
-Contour trees were (unsurprisingly) used to study terrain maps by Boyell and Ruston, and Freeman and Morse~\cite{BoRu63,FrMo67}.
+Contour trees were first used to study terrain maps by Boyell and Ruston, and Freeman and Morse~\cite{BoRu63,FrMo67}.
 They were used for isoline extraction in geometric data by van Krevald et al~\cite{kobps-ctsssit-97}, who provided
@@
 
-There has been a long series of improved algorithms for computing contour trees.
+There are numerous algorithms for computing contour trees.
 van Kreveld \etal \cite{kobps-ctsssit-97} presented an $O(N\log N)$ time algorithm for functions over 2D 
@@
 The influential paper of Carr \etal \cite{csa-cctad-00} improved the running time for all dimensions to $O(n\log n + N\alpha(N))$.
-One cannot overstate the importance of this result, since it forms the basis 
+This result forms the basis 
 of subsequent practical and theoretical work in computational topology applications. 
 Cole-McLaughlin and Pascucci \cite{pc-ectls-02} provided an $O(n+t\log n)$ time algorithm for 
-$3$-dimensional structured meshes. Chiang \etal \cite{cllr-sooscctmp-05} improve upon this result, and provide an unconditional $O(N+t\log t)$ algorithm.
+$3$-dimensional structured meshes. Chiang \etal \cite{cllr-sooscctmp-05} provide an unconditional $O(N+t\log t)$ algorithm.
 The latter result crucially uses monotone paths to improve the running time, and we repeatedly
@@
 
-We detail the basic definitions about contour trees. We follow the terminology of Chapter 6 of Carr's thesis \cite{c-tmi-04},
-which provides one of the first rigorous, Morse-theoretic defintions of contour trees. All our assumptions and definitions
+We detail the basic definitions about contour trees. We follow the terminology of Chapter 6 of Carr's thesis \cite{c-tmi-04}.
+All our assumptions and definitions
 are standard for results in this area, though there is some variability in notation.

### 2015-04-01 paint-contour-tree.tex
abstract changes, example

@@
 
-\title{A Mountaintop View Requires Minimal Sorting: \break A Faster Contour Tree Algorithm%
+\title{%A Mountaintop View Requires Minimal Sorting:
+Avoiding the Global Sort: 
+\break A Faster Contour Tree Algorithm%
 \footnote{The full updated version of this paper is available online \url{\si{http://web.engr.illinois.edu/\string~\si{raichel2}/\si{paint.pdf}}}}%
@@
 This leads to a linear time algorithm for contour tree computation for many
-classes of natural instances. We complement with matching lower bounds for contour tree computation.
+classes of natural instances. 
+%There is a crucial use of a family of binomial heaps to maintain priorities, 
+Specifically, our approach ensures that any comparison made is between nodes in the same descending path in the contour tree, 
+allowing us to complement our algorithmic results with matching lower bounds for contour tree computation.
 
@@
 decompositions for the time complexity analysis. 
-There is a crucial use of a family of binomial heaps to 
-maintain priorities, ensuring that any comparison made is between nodes in the same descending path in the contour tree.
+%There is a crucial use of a family of binomial heaps to 
+%maintain priorities, ensuring that any comparison made is between nodes in the same descending path in the contour tree.
 
@@
 the function $f$. 
+For example, consider a rocket engine during a burn.  The 3D mesh represents the locations of points in space around the engine, 
+and $f$ represents the temperature at each point.  The boundary of the flame is then represented by an isotherm, i.e. a contour.
 

### 2015-04-01 paint-contour-tree.tex
small changes

@@
 
-Massive geometric data is often represented as a function $f: \RR^d \mapsto \RR$. Typically, a finite representation is given
+Geometric data (potentially massive is size) is often represented as a function $f: \RR^d \mapsto \RR$. Typically, a finite representation is given
 by considering $f$ to be piecewise linear over some triangulated mesh (i.e. simplicial complex) $\MM$ in $\RR^d$.
@@
 (certainly true for $d=2$). The most expensive operation is an initial sort of all the vertex
-heights. Chiang et al build on this approach to get a faster algorithm that only sorts the critical vertices,
+heights. Chiang \etal build on this approach to get a faster algorithm that only sorts the critical vertices,
 yielding a running time of $O(t\log t + N)$~\cite{cllr-sooscctmp-05}.
-Common applications for contour trees involve turbulent combustion or noisy data, where the number of critical points is likely $\Omega(n)$.
+Common applications for contour trees involve turbulent combustion or noisy data, where the number of critical points is likely to be $\Omega(n)$.
 So we are still left with the bottleneck of sorting a large set of numbers. Unfortunately, there is a worst-case
-lower bound of $\Omega(t\log t)$ from Chiang et al~\cite{cllr-sooscctmp-05}, based on a construction of Bajaj et al~\cite{BaKr+98}.
+lower bound of $\Omega(t\log t)$ by Chiang \etal \cite{cllr-sooscctmp-05}, based on a construction of Bajaj \etal \cite{BaKr+98}.
 
 But this is only a worst-case bound. All previous algorithms begin by sorting (at least) the critical points, and
-hence must pay $\Omega(t\log t)$ for all instances. Can we beat this sorting bound for some instances
+hence must pay $\Omega(t\log t)$ for all instances. Can one beat this sorting bound for some instances
 and characterize what inputs are hard? Our main result gives an affirmative answer. First, we require some technical definitions.
@@
 Contour trees were first used to study terrain maps by Boyell and Ruston, and Freeman and Morse~\cite{BoRu63,FrMo67}.
-They were used for isoline extraction in geometric data by van Krevald et al~\cite{kobps-ctsssit-97}, who provided
+They were used for isoline extraction in geometric data by van Krevald \etal \cite{kobps-ctsssit-97}, who provided
 the first formal algorithm. 

### 2015-04-01 paint-contour-tree.tex
Minor edits
Minor edit

@@
 % The edges represent changes in the topology of contours.
-has numerous applications in data analysis and visualization,
+has numerous applications in data analysis and visualization.
 
@@
 characterize the amount of sorting required to compute the contour tree.
-This leads to a linear time algorithm for contour tree computation for many
+This leads to an almost linear (inverse Ackermann overhead) time algorithm for contour tree computation for many
 classes of natural instances. 
@@
 hence must pay $\Omega(t\log t)$ for all instances. Can one beat this sorting bound for some instances
-and characterize what inputs are hard? Our main result gives an affirmative answer. First, we require some technical definitions.
+and characterize what inputs are hard? Our main result gives an affirmative answer. 
+% First, we require some technical definitions.
 

### 2015-04-01 paint-contour-tree.tex
more changes

--- 2015-04-01 paint-contour-tree.tex
@@
 But this is only a worst-case bound. All previous algorithms begin by sorting (at least) the critical points, and
-hence must pay $\Omega(t\log t)$ for all instances. Can one beat this sorting bound for some instances
-and characterize what inputs are hard? Our main result gives an affirmative answer. First, we require some technical definitions.
-
-In a contour tree $T$, a \emph{leaf path} is simply a contiguous path in $T$
-containing a leaf, which is also monotonic in the height values of its vertices.
+hence must pay $\Omega(t\log t)$ for all instances. Can we beat this sorting bound for certain instances,
+and can we characterize which inputs are hard? Our main result gives an affirmative answer. First, we require a technical definition.
+
+% In a contour tree $T$, a \emph{leaf path} is simply a path in $T$
+% containing a leaf, which is also monotonic in the height values of its vertices.
 
@@
 \label{def:path} 
-For a contour tree $T$, a \emph{path decomposition}, $P(T)$, is a partition of $T$ into a set of vertex disjoint leaf paths.  
+For a contour tree $T$, a \emph{leaf path} is any path in $T$ containing a leaf, 
+which is also monotonic in the height values of its vertices. 
+Then a \emph{path decomposition}, $P(T)$, is a partition of $T$ into a set of vertex disjoint leaf paths.  
 \end{definition}
@@
 A better upper bound is $O(t\log D + t\alpha(t) + N)$, where $D$ is the diameter of $T$. 
-This is clearly an improvement when the tree is short and fat.
-But it can be even better. A calculation yields that for any decomposition of a rooted balanced binary tree, 
+This is clearly an improvement when the tree is short and fat, but it can be even better. 
+A calculation yields that for any decomposition of a rooted balanced binary tree, 
 our bound is $O(t\alpha(t) + n)$. 
@@
 Consider the example of \Fig{sorting} extended to a larger binary tree both above and below.
-(In the example, $t = \Omega(n)$.) The running time promised by \Thm{main-alg}
-is $O(n\alpha(n))$, whereas all previous algorithms require $\Omega(n\log n)$.
+(In the example, $t = \Omega(n)$.) The running time of the algorithm of \Thm{main-alg}
+is $O(n\alpha(n))$, whereas all previous algorithms require $\Omega(n\log n)$ time.
 
@@
 
-\paragraph{Beyond worst-case analysis:} Our result provides understanding of the complexity
+\paragraph{Beyond worst-case analysis.} Our result provides understanding of the complexity
 of contour tree construction beyond the standard worst-case analysis. Our theorem and lower bound 
@@
 %that the cost (as defined earlier) roughly captures the hardness of an instance.
-Trees that are short and fast are easy to compute, whereas trees with long paths are hard.
+Trees that are short and fat are easy to compute, whereas trees with long paths are hard.
 This is tangentially related to the concept of instance optimal algorithms, as defined by Afshani \etal \cite{abc-ioga-09}.
@@
 of subsequent practical and theoretical work in computational topology applications. 
-Cole-McLaughlin and Pascucci \cite{pc-ectls-02} provided an $O(n+t\log n)$ time algorithm for 
+Pascucci and Cole-McLaughlin \cite{pc-ectls-02} provided an $O(n+t\log n)$ time algorithm for 
 $3$-dimensional structured meshes. Chiang \etal \cite{cllr-sooscctmp-05} provide an unconditional $O(N+t\log t)$ algorithm.

### 2015-04-01 paint-contour-tree.tex

@@
 But this is only a worst-case bound. All previous algorithms begin by sorting (at least) the critical points, and
-hence must pay $\Omega(t\log t)$ for all instances. Can we beat this sorting bound for certain instances,
-and can we characterize which inputs are hard? Our main result gives an affirmative answer. First, we require a technical definition.
+% <<<<<<< HEAD
+% hence must pay $\Omega(t\log t)$ for all instances. Can we beat this sorting bound for certain instances,
+% and can we characterize which inputs are hard? Our main result gives an affirmative answer. First, we require a technical definition.
+% =======
+hence must pay $\Omega(t\log t)$ for all instances. Can one beat this sorting bound for some instances
+and characterize what inputs are hard? Our main result gives an affirmative answer. 
+% First, we require some technical definitions.
+%>>>>>>> bbe1b0b1cd16827b201e5bfef1d03efbc4caa55f
 

### 2015-04-01 paint-contour-tree.tex
fixed conflict

--- 2015-04-01 paint-contour-tree.tex
@@
 hence must pay $\Omega(t\log t)$ for all instances. Can we beat this sorting bound for certain instances,
-and can we characterize which inputs are hard? Our main result gives an affirmative answer. First, we require a technical definition.
+and can we characterize which inputs are hard? Our main result gives an affirmative answer. 
+% First, we require some technical definitions.
 

### 2015-04-02 paint-contour-tree.tex
ref to arxiv
done making changes

@@
 \break A Faster Contour Tree Algorithm%
-\footnote{The full updated version of this paper is available online \url{\si{http://web.engr.illinois.edu/\string~\si{raichel2}/\si{paint.pdf}}}}%
+\footnote{The full updated version of this paper is available on the arXiv \cite{rs-mvrms-14}}%
 }
@@
 
-We complement \Thm{main-alg} with a lower bound, suggesting that the comparison complexity
+We complement \Thm{main-alg} with a lower bound, suggesting that the comparison model complexity
 of $\sum_{p \in P(T)} |p|\log |p|$ is inherent for any contour tree algorithm. The exact phrasing
@@
 %where $t$ is the number of critical points and $C$ is a valid cost for an instance with $t$ critical point.
-Consider a valid path decomposition $P$. There exists a family $\bF_P$ of terrains ($d=2$) with the following properties.
+Consider a valid path decomposition $P$. There exists a family $\bF_P$ of terrains (i.e., $d=2$) with the following properties.
 Any contour tree algorithm makes $\Omega(\sum_{p \in P} |p|\log |p|)$ comparisons in the worst case over $\bF_P$.
@@
 In other words, for any choice of $\sum_{p \in P} |p|\log |p|$, we can design a hard family where any algorithm
-requires this cost in the worst case, and our algorithm matches this bound.
+requires this time in the worst case, and our algorithm matches this bound.
 
@@
 Their notion of optimality is far more refined that what we prove, but it shares the flavor of understanding the spectrum
-of easy to hard instances. From a mathematical perspective, this sort of analysis forces one to look closely
+of easy to hard instances. In general, our analysis forces one to look closely
 at the relationship between sorting and the contour tree problem.
@@
 and studying chemical systems~\cite{LaBe+06,BrWe+10,BeWe+11,BrWe+11,MaGr+11}. Carr's thesis~\cite{c-tmi-04} gives various
-applications of contour trees for data visualization and is an excellent reference for contour trees definitions and algorithms.
+applications of contour trees for data visualization and is an excellent reference for contour tree definitions and algorithms.
 
@@
 $3$-dimensional structured meshes. Chiang \etal \cite{cllr-sooscctmp-05} provide an unconditional $O(N+t\log t)$ algorithm.
-The latter result crucially uses monotone paths to improve the running time, and we repeatedly
-exploit numerous properties of monotone paths. Carr's thesis shows relationships between monotone paths
+The latter result crucially uses monotone paths to improve the running time, and our algorithm repeatedly
+exploits numerous properties of monotone paths. Carr's thesis shows relationships between monotone paths
 in $\MM$ and in its contour tree $\cC(\MM)$.
@@
 We assume that $f$ has distinct values on all vertices, except for boundaries,
-as expressed in the following constraint.
-
-\begin{definition} \label{def:bound} $f$ is \emph{boundary critical} if the following holds.
+as expressed in the following constraint, which intuitively requires that 
+all boundaries look like (local) horizontal cuts.
+
+\begin{definition} \label{def:bound} The function $f$ is \emph{boundary critical} if the following holds.
 Consider a boundary facet $F$. All vertices of $F$ have the same function value. Furthermore, all
-neighbors of $F$ not in $F$ either have all function values strictly greater or strictly less
-than this value.
+neighbors of vertices in $F$, which are not also in $F$ itself, 
+either have all function values strictly greater than or all function values strictly less than the function value at $F$.
 \end{definition}
@@
 has a well-defined height, and this includes the infinite facet. 
-We think of $d$ as constant, and assume that $\MM$ is represented in a data structure that allow constant-time traversals of neighboring simplices
+We think of the dimension, $d$, as constant, and assume that $\MM$ is represented in a data structure that allows constant-time traversals of neighboring simplices
 in $\MM$ (e.g.~\cite{BoMa12}). (This is analogous to a DCEL for higher dimensions.)
@@
 
-\begin{definition} \label{def:level} The level set at value $h$ is the set $\{x| f(x) = h\}$.
-A contour is a maximal connected component of a level set. An $h$-contour is a contour where $f$-values are $h$.
+\begin{definition} \label{def:level} The \emph{level set} at value $h$ is the set $\{x| f(x) = h\}$.
+A \emph{contour} is a maximal connected component of a level set. An \emph{$h$-contour} is a contour where $f$-values are $h$.
 \end{definition}
@@
 Note that a contour that does not contain a boundary is itself a simplicial complex of one dimension lower, and is represented (in our algorithms) as such.
-We let $\delta, \eps$ denote infinitesimals. Let $B_\eps(x)$ denote a ball of radius $\eps$ around $x$, and let 
+We let $\delta$ and $\eps$ denote infinitesimals. Let $B_\eps(x)$ denote a ball of radius $\eps$ around $x$, and let 
 $f|B_\eps(x)$ be the restriction of $f$ to $B_\eps(x)$.
@@
 
-We use $\cV(f)$ to denote the set of critical points.
+The set of critical points is denoted by $\cV(f)$.
 Because $f$ is piecewise-linear, all critical points are vertices in $\MM$. 
-A value $h$ is called critical, if $f(v) = h$, for some $v \in \cV(f)$. 
+A value $h$ is called \emph{critical}, if $f(v) = h$, for some $v \in \cV(f)$. 
 The critical points are exactly where the topology of level sets change.
@@
 \begin{definition} \label{def:equiv} Two contours $\psi$ and $\psi'$ are equivalent 
-if the following holds. There exists an $f$-monotone path $p$ connecting a point in $\psi$ to $\psi'$,
+if there exists an $f$-monotone path $p$ connecting a point in $\psi$ to $\psi'$,
 such that no $x \in p$ belongs to a critical contour.
@@
 
-These give the \emph{contour classes}. By continuity, a regular point is present on a single contour,
+This equivalence relation gives a set \emph{contour classes}. By continuity, a regular point is present on a single contour,
 and we can partition the regular contours into these classes. Every such class maps to intervals
@@
 
-\begin{definition} \label{def:tree} The contour tree is the graph on vertex set $\cV$, where
+\begin{definition} \label{def:tree} The contour tree is the graph on vertex set $\cV= \cV(f)$, where
 edges are formed as follows. For every contour class that is created at $v_i$ and destroyed $v_j$,
@@
 \subsection{Some technical remarks}
-
-The $(f(x)+\delta)$-contours in $f|_{B_\eps(x)}$ given by \Def{deg} might actually be the same
-contour in $f$. The up-degree (as opposed to \emph{Morse} up-degree) is defined as the number of $(f(x)+\delta)$-contours
+Note that if one intersects $\MM$ with given ball $B$, then a single contour in $\MM$ might be split into more than one contour in the intersection. 
+In particular, two $(f(x)+\delta)$-contours in $f|_{B_\eps(x)}$, given by \Def{deg}, might actually be the same contour in $f$. 
+Alternatively, one can define the up-degree (as opposed to \emph{Morse} up-degree) as the number of $(f(x)+\delta)$-contours (in the full $\MM$)
 that intersect $B_\eps(x)$, a potentially smaller number. This up-degree
@@
 vertex is explictly placed in $\reeb(\MM)$ (by subdividing the respective edge). This is referred
-to as augmenting the contour tree, and it is common to augment $\reeb(\MM)$ will all vertices.
+to as augmenting the contour tree, and it is common to augment $\reeb(\MM)$ with all vertices.
 Alternatively, one can smooth out all vertices of up-degree and down-degree $1$ to get the
@@
 
-\section{A tour of the contour tree algorithm} \label{sec:approach}
+\section{A tour of the new contour tree algorithm} \label{sec:approach}
 

### 2015-04-02 paint-contour-tree.tex
submitted version

@@
 \begin{abstract}
-We revisit the classic problem of computing the \emph{contour tree}
+We revisit the classical problem of computing the \emph{contour tree}
 of a scalar field $f:\MM \mapsto \RR$, where $\MM$ is a triangulated simplicial mesh in $\RR^d$. 
-% A level set, or contour, at value $v$ is a connected component of $f^{-1}(v)$. 
-% As $v$ is changed, these contours change topology, merge into each other, or split.
 The contour tree is a fundamental topological data structure that tracks
 the evolution of level sets of $f$ and 
-% concise representations of $f$ that track the evolution
-% of levels sets, and this contour
-% behavior. The vertices of these trees are the critical points of $f$, where the gradient is zero.
-% The edges represent changes in the topology of contours.
 has numerous applications in data analysis and visualization.
@@
 We present the first input-sensitive algorithm whose time complexity depends
-on the contour tree struture, and avoids the global sort for non-pathological inputs.
+on the contour tree structure, and avoids the global sort for non-pathological inputs.
 Our results match all existing upper bounds, but also
@@
 classes of natural instances. 
-%There is a crucial use of a family of binomial heaps to maintain priorities, 
 Specifically, our approach ensures that any comparison made is between nodes in the same descending path in the contour tree, 
@@
 decompositions for the time complexity analysis. 
-%There is a crucial use of a family of binomial heaps to 
-%maintain priorities, ensuring that any comparison made is between nodes in the same descending path in the contour tree.
-
-% Our results demonstrate that this sort can be avoided for many instances, and
-% also characterizes these hard instances.
-% At a high level, if the tree is short and fat, we get significant improvements in running time.
-% 
-% 
-% 
-% 
-% and there is significant previous work (both theoretical and practical) on algorithms for constructing contour trees.
-% Suppose $\MM$ has $n$ vertices, $N$ facets, and $t$ critical points.
-% The best known algorithm for constructing the contour tree requires $O(t\log t + N)$ time, and all algorithms to date
-% involve a global sort of the critical points. Unfortunately, worst case lower bounds of $\Omega(t\log t)$ also exist.
-% A classic result of Carr, Snoeyink, and Axen (2000) gives an algorithm that takes $O(n\log n + N\alpha(N))$ time
-% (where $\alpha(\cdot)$ is the inverse Ackermann function). A further improvement to $O(t\log t + N)$ time
-% was given by Chiang et al. All these algorithms involve a global sort of the critical points,
-% a significant computational bottleneck. Unfortunately, lower bounds of $\Omega(t\log t)$ also exist.
-
-% We present the first algorithm that can avoid the global sort and has a refined time complexity that
-% depends on the contour tree structure. Intuitively, if the tree is short and fat, we get significant improvements 
-% in running time. For a partition of the contour tree into a set of descending paths, $P$,
-% our algorithm runs in $O(\sum_{p\in P} |p|\log|p| + t\alpha(t)+N)$ time.
-% This subsumes all existing upper bounds, and is a significant improvement 
-% This is at most $O(t\log D + N)$, where $D$ is the diameter of the contour tree.
-% Moreover, it is $O(t\alpha(t)+N)$ for balanced trees, a significant improvement over the previous complexity.
-
-% We also prove lower bounds showing that the $\sum_{p\in P} |p|\log|p|$ complexity
-% is inherent to computing contour trees. 
 \end{abstract}

### 2015-04-02 paint-contour-tree.tex
typo corrected

@@
 \subsection{Some technical remarks}
-Note that if one intersects $\MM$ with given ball $B$, then a single contour in $\MM$ might be split into more than one contour in the intersection. 
-In particular, two $(f(x)+\delta)$-contours in $f|_{B_\eps(x)}$, given by \Def{deg}, might actually be the same contour in $f$. 
+Note that if one intersects $\MM$ with a given ball $B$, then a single contour in $\MM$ might be split into more than one contour in the intersection. 
+In particular, two $(f(x)+\delta)$-contours in $f|_{B_\eps(x)}$, given by \Def{deg}, might actually be the same contour in $\MM$. 
 Alternatively, one can define the up-degree (as opposed to \emph{Morse} up-degree) as the number of $(f(x)+\delta)$-contours (in the full $\MM$)

### 2015-04-02 paint-contour-tree.tex
Minor edits

--- 2015-04-02 paint-contour-tree.tex
@@
 \break A Faster Contour Tree Algorithm%
-\footnote{The full updated version of this paper is available on the arXiv \cite{rs-mvrms-14}}%
+\footnote{The full updated version of this paper is available online \url{\si{http://web.engr.illinois.edu/\string~\si{raichel2}/\si{paint.pdf}}}}%
 }
@@
 This is clearly an improvement when the tree is short and fat, but it can be even better. 
-A calculation yields that for any decomposition of a rooted balanced binary tree, 
+Consider a contour tree that is constant balanced, meaning that the ratio of sizes of sibling subtrees
+is $\Theta(1)$. 
+A calculation yields that for any decomposition of such a contour tree, 
 our bound is $O(t\alpha(t) + n)$. 
-%
 %
@@
 %However, our actual bound in this case gives an even further improved $O(t\alpha(t)+N)$.
-Consider the example of \Fig{sorting} extended to a larger binary tree both above and below.
-(In the example, $t = \Omega(n)$.) The running time of the algorithm of \Thm{main-alg}
+Consider the example of \Fig{sorting} extended to a larger binary tree both above and below,
+so $t = \Omega(n)$. The running time of the algorithm of \Thm{main-alg}
 is $O(n\alpha(n))$, whereas all previous algorithms require $\Omega(n\log n)$ time.
 
-We complement \Thm{main-alg} with a lower bound, suggesting that the comparison model complexity
+We complement \Thm{main-alg} with a lower bound, suggesting that the comparison complexity
 of $\sum_{p \in P(T)} |p|\log |p|$ is inherent for any contour tree algorithm. The exact phrasing
@@
 %where $t$ is the number of critical points and $C$ is a valid cost for an instance with $t$ critical point.
-Consider a valid path decomposition $P$. There exists a family $\bF_P$ of terrains (i.e., $d=2$) with the following properties.
+Consider a valid path decomposition $P$. There exists a family $\bF_P$ of terrains ($d=2$) with the following properties.
 Any contour tree algorithm makes $\Omega(\sum_{p \in P} |p|\log |p|)$ comparisons in the worst case over $\bF_P$.
@@
 In other words, for any choice of $\sum_{p \in P} |p|\log |p|$, we can design a hard family where any algorithm
-requires this time in the worst case, and our algorithm matches this bound.
+requires this cost in the worst case, and our algorithm matches this bound.
 
@@
 Their notion of optimality is far more refined that what we prove, but it shares the flavor of understanding the spectrum
-of easy to hard instances. In general, our analysis forces one to look closely
+of easy to hard instances. From a mathematical perspective, this sort of analysis forces one to look closely
 at the relationship between sorting and the contour tree problem.
@@
 and studying chemical systems~\cite{LaBe+06,BrWe+10,BeWe+11,BrWe+11,MaGr+11}. Carr's thesis~\cite{c-tmi-04} gives various
-applications of contour trees for data visualization and is an excellent reference for contour tree definitions and algorithms.
+applications of contour trees for data visualization and is an excellent reference for contour trees definitions and algorithms.
 
@@
 $3$-dimensional structured meshes. Chiang \etal \cite{cllr-sooscctmp-05} provide an unconditional $O(N+t\log t)$ algorithm.
-The latter result crucially uses monotone paths to improve the running time, and our algorithm repeatedly
-exploits numerous properties of monotone paths. Carr's thesis shows relationships between monotone paths
+The latter result crucially uses monotone paths to improve the running time, and we repeatedly
+exploit numerous properties of monotone paths. Carr's thesis shows relationships between monotone paths
 in $\MM$ and in its contour tree $\cC(\MM)$.
@@
 We assume that $f$ has distinct values on all vertices, except for boundaries,
-as expressed in the following constraint, which intuitively requires that 
-all boundaries look like (local) horizontal cuts.
-
-\begin{definition} \label{def:bound} The function $f$ is \emph{boundary critical} if the following holds.
+as expressed in the following constraint.
+
+\begin{definition} \label{def:bound} $f$ is \emph{boundary critical} if the following holds.
 Consider a boundary facet $F$. All vertices of $F$ have the same function value. Furthermore, all
-neighbors of vertices in $F$, which are not also in $F$ itself, 
-either have all function values strictly greater than or all function values strictly less than the function value at $F$.
+neighbors of $F$ not in $F$ either have all function values strictly greater or strictly less
+than this value.
 \end{definition}
@@
 has a well-defined height, and this includes the infinite facet. 
-We think of the dimension, $d$, as constant, and assume that $\MM$ is represented in a data structure that allows constant-time traversals of neighboring simplices
+We think of $d$ as constant, and assume that $\MM$ is represented in a data structure that allow constant-time traversals of neighboring simplices
 in $\MM$ (e.g.~\cite{BoMa12}). (This is analogous to a DCEL for higher dimensions.)
@@
 
-\begin{definition} \label{def:level} The \emph{level set} at value $h$ is the set $\{x| f(x) = h\}$.
-A \emph{contour} is a maximal connected component of a level set. An \emph{$h$-contour} is a contour where $f$-values are $h$.
+\begin{definition} \label{def:level} The level set at value $h$ is the set $\{x| f(x) = h\}$.
+A contour is a maximal connected component of a level set. An $h$-contour is a contour where $f$-values are $h$.
 \end{definition}
@@
 Note that a contour that does not contain a boundary is itself a simplicial complex of one dimension lower, and is represented (in our algorithms) as such.
-We let $\delta$ and $\eps$ denote infinitesimals. Let $B_\eps(x)$ denote a ball of radius $\eps$ around $x$, and let 
+We let $\delta, \eps$ denote infinitesimals. Let $B_\eps(x)$ denote a ball of radius $\eps$ around $x$, and let 
 $f|B_\eps(x)$ be the restriction of $f$ to $B_\eps(x)$.
@@
 
-The set of critical points is denoted by $\cV(f)$.
+We use $\cV(f)$ to denote the set of critical points.
 Because $f$ is piecewise-linear, all critical points are vertices in $\MM$. 
-A value $h$ is called \emph{critical}, if $f(v) = h$, for some $v \in \cV(f)$. 
+A value $h$ is called critical, if $f(v) = h$, for some $v \in \cV(f)$. 
 The critical points are exactly where the topology of level sets change.
@@
 \begin{definition} \label{def:equiv} Two contours $\psi$ and $\psi'$ are equivalent 
-if there exists an $f$-monotone path $p$ connecting a point in $\psi$ to $\psi'$,
+if the following holds. There exists an $f$-monotone path $p$ connecting a point in $\psi$ to $\psi'$,
 such that no $x \in p$ belongs to a critical contour.
@@
 
-This equivalence relation gives a set \emph{contour classes}. By continuity, a regular point is present on a single contour,
+These give the \emph{contour classes}. By continuity, a regular point is present on a single contour,
 and we can partition the regular contours into these classes. Every such class maps to intervals
@@
 
-\begin{definition} \label{def:tree} The contour tree is the graph on vertex set $\cV= \cV(f)$, where
+\begin{definition} \label{def:tree} The contour tree is the graph on vertex set $\cV$, where
 edges are formed as follows. For every contour class that is created at $v_i$ and destroyed $v_j$,
@@
 \subsection{Some technical remarks}
-Note that if one intersects $\MM$ with given ball $B$, then a single contour in $\MM$ might be split into more than one contour in the intersection. 
-In particular, two $(f(x)+\delta)$-contours in $f|_{B_\eps(x)}$, given by \Def{deg}, might actually be the same contour in $f$. 
-Alternatively, one can define the up-degree (as opposed to \emph{Morse} up-degree) as the number of $(f(x)+\delta)$-contours (in the full $\MM$)
+
+The $(f(x)+\delta)$-contours in $f|_{B_\eps(x)}$ given by \Def{deg} might actually be the same
+contour in $f$. The up-degree (as opposed to \emph{Morse} up-degree) is defined as the number of $(f(x)+\delta)$-contours
 that intersect $B_\eps(x)$, a potentially smaller number. This up-degree
@@
 vertex is explictly placed in $\reeb(\MM)$ (by subdividing the respective edge). This is referred
-to as augmenting the contour tree, and it is common to augment $\reeb(\MM)$ with all vertices.
+to as augmenting the contour tree, and it is common to augment $\reeb(\MM)$ will all vertices.
 Alternatively, one can smooth out all vertices of up-degree and down-degree $1$ to get the
@@
 
-\section{A tour of the new contour tree algorithm} \label{sec:approach}
+\section{A tour of the contour tree algorithm} \label{sec:approach}
 

### 2015-04-02 paint-contour-tree.tex

--- 2015-04-02 paint-contour-tree.tex
@@
 %However, our actual bound in this case gives an even further improved $O(t\alpha(t)+N)$.
-Consider the example of \Fig{sorting} extended to a larger binary tree both above and below.
-(In the example, $t = \Omega(n)$.) The running time of the algorithm of \Thm{main-alg}
+Consider the example of \Fig{sorting} extended to a larger binary tree both above and below,
+so $t = \Omega(n)$. The running time of the algorithm of \Thm{main-alg}
 is $O(n\alpha(n))$, whereas all previous algorithms require $\Omega(n\log n)$ time.

### 2015-07-01 paint-contour-tree.tex
Many reviewer comments handled

--- 2015-04-02 paint-contour-tree.tex
@@
 We revisit the classical problem of computing the \emph{contour tree}
-of a scalar field $f:\MM \mapsto \RR$, where $\MM$ is a triangulated simplicial mesh in $\RR^d$. 
-The contour tree is a fundamental topological data structure that tracks
+of a scalar field $f:\MM \to \RR$, where $\MM$ is a triangulated simplicial mesh in $\RR^d$. 
+The contour tree is a fundamental topological structure that tracks
 the evolution of level sets of $f$ and 
@@
 Existing lower bounds show that there are pathological instances where this sort is required.
-We present the first input-sensitive algorithm whose time complexity depends
+We present the first algorithm whose time complexity depends
 on the contour tree structure, and avoids the global sort for non-pathological inputs.
@@
 
-Geometric data (potentially massive is size) is often represented as a function $f: \RR^d \mapsto \RR$. Typically, a finite representation is given
-by considering $f$ to be piecewise linear over some triangulated mesh (i.e. simplicial complex) $\MM$ in $\RR^d$.
-\emph{Contour trees} are a specific topological data structure used to represent and visualize
+Geometric data (potentially massive in size) is often represented as a function $f: \RR^d \to \RR$. Typically, a finite representation is given
+by considering $f$ to be piecewise linear over some triangulated mesh (i.e.\ simplicial complex) $\MM$ in $\RR^d$.
+\emph{Contour trees} are a specific topological structure used to represent and visualize
 the function $f$. 
 For example, consider a rocket engine during a burn.  The 3D mesh represents the locations of points in space around the engine, 
-and $f$ represents the temperature at each point.  The boundary of the flame is then represented by an isotherm, i.e. a contour.
-
-It is convenient to think of $f$ as a $(d+1)$-dimensional manifold
+and $f$ represents the temperature at each point.  The boundary of the flame is then represented by an isotherm, i.e.\ a contour.
+
+It is convenient to think of $f$ as a manifold sitting in $\RR^{d+1}$,
 %  or alternately consider
 % $\MM$ in $\RR^{d+1}$
-with the last coordinate (i.e. height) given by $f$. Imagine sweeping the hyperplane $x_{d+1} = h$
+with the last coordinate (i.e.\ height) given by $f$. Imagine sweeping the hyperplane $x_{d+1} = h$
 with $h$ going from $+\infty$ to $-\infty$. At every instance, the intersection of this plane 
@@
 If $f$ is smooth, all points where the gradient of $f$ is zero are \emph{critical points}.
-These points are the ``events" where the contour topology changes and form the vertices of the contour tree. 
-An edge connects two critical points if one event immediately ``follows"
+These points are the ``events" where the contour topology changes and which form the vertices of the contour tree. 
+An edge of the contour tree connects two critical points if one event immediately ``follows"
 the other as the sweep plane makes its pass. (We provide formal definitions later.)
@@
 
-Suppose we have $f: \MM \mapsto \RR$, where $\MM$ is a triangulated mesh with $n$ vertices, $N$ facets in total, and $t \leq n$ critical points.
-It is standard to assume that $f$ is PL-Morse (a well-behaved assumption) and that the maximum degree in the contour tree
+Suppose we have $f: \MM \to \RR$, where $\MM$ is a triangulated mesh with $n$ vertices, $N$ faces in total, and $t \leq n$ critical points.
+It is standard to assume that $f$ is PL-Morse (a well-behaved input assumption) and that the maximum degree in the contour tree
 is $3$. A fundamental result in this area is the algorithm of Carr, Snoeyink, and Axen to compute contour trees,
@@
 which is also monotonic in the height values of its vertices. 
-Then a \emph{path decomposition}, $P(T)$, is a partition of $T$ into a set of vertex disjoint leaf paths.  
+Then a \emph{path decomposition}, $P(T)$, is a partition of the vertices of $T$ into a set of vertex disjoint leaf paths.  
 \end{definition}
@@
 \label{thm:main-alg} 
-Consider a PL-Morse\footnote{This is a standard ``well-behaved" assumption that implies
+Consider a PL-Morse\footnote{This is a standard ``well-behaved" input assumption that implies
 that all critical points have distinct values, and the contour tree has max-degree $3$. Our algorithm
-does not need this, but it simplifies the proof and presentation.} $f:\MM \mapsto \RR$, where the contour tree $T$
+does not need this, but it simplifies the proof and presentation.} $f:\MM \to \RR$, where the contour tree $T$
 has maximum degree $3$. There is a deterministic algorithm to compute $T$ whose running time is $O(\sum_{p \in P(T)} |p|\log |p| + t\alpha(t) + N)$,
@@
 %where $t$ is the number of critical points and $C$ is a valid cost for an instance with $t$ critical point.
-Consider a valid path decomposition $P$. There exists a family $\bF_P$ of terrains (i.e., $d=2$) with the following properties.
+Consider a valid path decomposition $P$. There exists a family $\bF_P$ of terrains (i.e.\ $d=2$) with the following properties.
 Any contour tree algorithm makes $\Omega(\sum_{p \in P} |p|\log |p|)$ comparisons in the worst case over $\bF_P$.
@@
 
-In other words, for any choice of $\sum_{p \in P} |p|\log |p|$, we can design a hard family where any algorithm
-requires this time in the worst case, and our algorithm matches this bound.
+In other words, for any path decomposition, $P$, we can design a hard family where any algorithm
+requires $\sum_{p \in P} |p|\log |p|$ time in the worst case, and our algorithm matches this bound.
 
@@
 Trees that are short and fat are easy to compute, whereas trees with long paths are hard.
-This is tangentially related to the concept of instance optimal algorithms, as defined by Afshani \etal \cite{abc-ioga-09}.
+This is tangentially related to the concept of instance-optimal algorithms, as defined by Afshani \etal \cite{abc-ioga-09}.
 Their notion of optimality is far more refined that what we prove, but it shares the flavor of understanding the spectrum
@@
 the first formal algorithm. 
-Contour trees and related topological data structures have been applied in analysis of fluid mixing, combustion simulations,
+Contour trees and related topological structures have been applied in analysis of fluid mixing, combustion simulations,
 and studying chemical systems~\cite{LaBe+06,BrWe+10,BeWe+11,BrWe+11,MaGr+11}. Carr's thesis~\cite{c-tmi-04} gives various
@@
 There are numerous algorithms for computing contour trees.
-van Kreveld \etal \cite{kobps-ctsssit-97} presented an $O(N\log N)$ time algorithm for functions over 2D 
-meshes and an $O(N^2)$ algorithm for higher dimensions. 
+An $O(N\log N)$ time algorithm for functions over 2D 
+meshes and an $O(N^2)$ algorithm for higher dimensions, was presented by van Kreveld \etal \cite{kobps-ctsssit-97}. 
 Tarasov and Vyalya \cite{tv-cct-98} improved the running time to $O(N\log N)$ for the 3D case.
@@
     \includegraphics[width=.35\linewidth,height=.15\linewidth]{figs/trees}
-    \caption{On left, a surface will a balanced contour tree, but whose join and split trees have long tails.  
+    \caption{On left, a surface with a balanced contour tree, but whose join and split trees have long tails.  
     On right (from left to right), the contour, join and split trees.}
@@
 are standard for results in this area, though there is some variability in notation.
-The input is a continuous piecewise-linear (PL) function $f:\MM \mapsto \RR$, where $\MM$ is a fully triangulated simplicial complex in $\RR^d$,
+The input is a continuous piecewise-linear (PL) function $f:\MM \to \RR$, where $\MM$ is a simply connected and fully triangulated simplicial complex in $\RR^d$,
 except for specially designated \emph{boundary facets}. So $f$ is defined only on the vertices of $\MM$,
-and all other valued are obtained by interpolation.
+and all other values are obtained by linear interpolation.
 
@@
 This is convenient, as we can now assume that $f$ is defined on $\RR^d$. Any point inside a boundary facet
-has a well-defined height, and this includes the infinite facet. 
+has a well-defined height, including the infinite facet, which is required to be a boundary facet.  However, 
+we allow for other boundary facets, to capture the resulting surface pieces after our algorithm make a horizontal cut.
+
 We think of the dimension, $d$, as constant, and assume that $\MM$ is represented in a data structure that allows constant-time traversals of neighboring simplices
-in $\MM$ (e.g.~\cite{BoMa12}). (This is analogous to a DCEL for higher dimensions.)
-Observe that $f:\MM \rightarrow \RR$ can be thought of as a $d$-dimensional simplicial complex living in $\RR^{d+1}$.
-We can think of $f(x)$ as the ``height" of a point $x \in \MM$, and encoded in the representation of $\MM$.
+in $\MM$ (e.g.~\cite{BoMa12}). (This is analogous to a doubly connected edge list, but for higher dimensions.)
+Observe that $f:\MM \rightarrow \RR$ can be thought of as a $d$-dimensional simplicial complex living in $\RR^{d+1}$, 
+where $f(x)$ is the ``height" of a point $x \in \MM$, which is encoded in the representation of $\MM$. 
+Specifically, rather than writing our input as $(\MM,f)$, we abuse notation and typically just write $\MM$ to denote the lifted complex.
 
@@
 
-\begin{definition} \label{def:deg} The Morse up-degree of $x$ is the number of $(f(x) + \delta)$-contours in $f|B_\eps(x)$
-as $\delta, \eps \rightarrow 0^+$. The Morse down-degree is the number of $(f(x) - \delta)$-contours in $f|B_\eps(x)$
+\begin{definition} \label{def:deg} The \emph{Morse up-degree} of $x$ is the number of $(f(x) + \delta)$-contours of $f|B_\eps(x)$
+as $\delta, \eps \rightarrow 0^+$. The \emph{Morse down-degree} is the number of $(f(x) - \delta)$-contours of $f|B_\eps(x)$
 as $\delta, \eps \rightarrow 0^+$.
@@
 \begin{asparaitem}
-	\item Regular: Both Morse up-degree and down-degrees are $1$.
-	\item Maximum: Morse up-degree is $0$.
-	\item Minimum: Morse down-degree is $0$.
-	\item Morse Join: Morse up-degree is strictly greater than $1$.
-	\item Morse Split: Morse down-degree is strictly greater than $1$.
+	\item \emph{Regular:} Both Morse up-degree and down-degrees are $1$.
+	\item \emph{Maximum:} Morse up-degree is $0$.
+	\item \emph{Minimum:} Morse down-degree is $0$.
+	\item \emph{Morse Join:} Morse up-degree is strictly greater than $1$.
+	\item \emph{Morse Split:} Morse down-degree is strictly greater than $1$.
 \end{asparaitem}
@@
 A value $h$ is called \emph{critical}, if $f(v) = h$, for some $v \in \cV(f)$. 
+A contour is called \emph{critical}, if it contains a critical point, and it is called \emph{regular} otherwise.
+
 The critical points are exactly where the topology of level sets change.
 By assuming that our manifold is boundary critical, the vertices on a given boundary are either collectively all maxima or all minima.  
-We abuse notation and refer to this entire set of vertices as a maxima or minima.
-
-\begin{definition} \label{def:equiv} Two contours $\psi$ and $\psi'$ are equivalent 
+We abuse notation and refer to this entire set of vertices as a maximum or minimum.
+
+\begin{definition} \label{def:equiv} Two contours $\psi$ and $\psi'$ are \emph{equivalent} 
 if there exists an $f$-monotone path $p$ connecting a point in $\psi$ to $\psi'$,
@@
 
-\begin{definition} \label{def:tree} The contour tree is the graph on vertex set $\cV= \cV(f)$, where
+\begin{definition} \label{def:tree} The \emph{contour tree} is the graph on vertex set $\cV= \cV(f)$, where
 edges are formed as follows. For every contour class that is created at $v_i$ and destroyed $v_j$,
@@
 Note that if one intersects $\MM$ with a given ball $B$, then a single contour in $\MM$ might be split into more than one contour in the intersection. 
-In particular, two $(f(x)+\delta)$-contours in $f|_{B_\eps(x)}$, given by \Def{deg}, might actually be the same contour in $\MM$. 
+In particular, two $(f(x)+\delta)$-contours of $f|_{B_\eps(x)}$, given by \Def{deg}, might actually be the same contour in $\MM$. 
 Alternatively, one can define the up-degree (as opposed to \emph{Morse} up-degree) as the number of $(f(x)+\delta)$-contours (in the full $\MM$)
@@
 For example, when $d=3$ this is equivalent to the contour gaining a handle.
-When $d=2$, this distinction is not necessary, since any point with $>1$
-Morse degree will have $>1$ degree in $\reeb(\MM)$.
+When $d=2$, this distinction is not necessary, since any point with Morse degree strictly greater than $1$ will have degree strictly greater than $1$ in $\reeb(\MM)$.
 
@@
 Every vertex in $\MM$ is associated with an edge in $\reeb(\MM)$, and sometimes the
-vertex is explictly placed in $\reeb(\MM)$ (by subdividing the respective edge). This is referred
+vertex is explicitly placed in $\reeb(\MM)$ (by subdividing the respective edge). This is referred
 to as augmenting the contour tree, and it is common to augment $\reeb(\MM)$ with all vertices.
@@
 \section{A tour of the new contour tree algorithm} \label{sec:approach}
-
-Our final algorithm is quite technical and has numerous moving parts. This section
-provides a high level view of the entire result, and highlights the main ideas
-and intuition. 
-\SubmitVer{The full version with all the details can be found in the appendix.}
-It is helpful to keep the $d=2$ case in mind, so the input is just a triangulated
-terrain. In the interest of presentation, the definitions and theorem statements in this section will
+Our final algorithm is quite technical and has numerous moving parts. However, for the $d=2$ case, 
+where the input is just a triangulated terrain, the main ideas of the parts of the algorithm 
+can be explained clearly.  
+\SubmitVer{Therefore, in the remaining few allotted pages, 
+rather than providing an incomplete subset of the technical content, we provide a high level view of the entire result, 
+which we view as essential for understanding our approach. The technical details can be found in the full version of the paper \cite{rs-mvrms-14}.}
+\FullVer{Therefore, here we first provide a high level view of the entire result, 
+which we view as essential for understanding our approach. }
+
+In the interest of presentation, the definitions and theorem statements in this section will
 slightly differ from those in the main body. They may also differ from the original definitions proposed
 in earlier work.
+
+% Our final algorithm is quite technical and has numerous moving parts. This section
+% provides a high level view of the entire result, and highlights the main ideas
+% and intuition. 
+% \SubmitVer{The full version with all the details can be found in the appendix.}
+% It is helpful to keep the $d=2$ case in mind, so the input is just a triangulated
+% terrain. In the interest of presentation, the definitions and theorem statements in this section will
+% slightly differ from those in the main body. They may also differ from the original definitions proposed
+% in earlier work.
 
@@
 information is necessary: on the left, the little valleys are higher than the big central valley,
-and this is reflected in the contour tree. More generally, leafs paths in the contour tree have points
+and this is reflected in the contour tree. More generally, leaf paths in the contour tree have points
 in sorted order, but incomparable points in the tree are unconstrained.
@@
 \label{def:int-criticalJoin}
-The critical join tree $\jc(\MM)$ is built on the set $\cV$ of all critical points. 
+The \emph{critical join tree} $\jc(\MM)$ is built on the set $\cV$ of all critical points. 
 The directed edge $(u,v)$ is present when $u$ is the smallest valued vertex in $\cV$ in a connected component of $\MM^+_v$
@@
 
-(Abusing notation, we just call this the join tree.)
+Abusing notation, for now we will just call this the join tree. (In Section~\ref{sec:extreme} we carefully handle the distinction 
+between the join tree and the critical join tree.)
 Refer to \Fig{sorting} for the join tree of a terrain. Note that nothing happens at splits, but these
@@
 in \Fig{sorting} that the split vertices form a long path in the join tree (and vice versa). Therefore, constructing
-these trees forces a global sort of the splits, an unnecessary computation for the contour tree. Unfortunately, in general (i.e. unlike \Fig{sorting}) 
+these trees forces a global sort of the splits, an unnecessary computation for the contour tree. Unfortunately, in general (i.e.\ unlike \Fig{sorting}) 
 the heights of joins and splits may be interleaved in a complex manner, and hence the final merging of~\cite{csa-cctad-00}
@@
 \paragraph{Cutting $\MM$ into extremum dominant pieces:} We define a simplicial complex
-endowed with a height to be \emph{minimum dominant} if there only exists a single minima.
+endowed with a height to be \emph{minimum dominant} if there exists only a single minimum.
 (Our real definition is more complicated, and involves simplicial complexes
 that allow additional ``trivial'' minima.)
-In such a complex, there exists a non-ascending path from any point to this unique minima.
+In such a complex, there exists a non-ascending path from any point to this unique minimum.
 Analogously, we can define maximum dominant simplicial complexes,
@@
 We will cut $\MM$ into disjoint extremum dominant pieces, in linear time.
-The best way to think of our procedure is a meteorology analogy. Take an arbitrary maxima $x$,
-and imagine torrential rain at the maxima. The water flows down, wetting any point that has a non-ascending path 
+The best way to think of our procedure is a meteorology analogy. Take an arbitrary maximum $x$,
+and imagine torrential rain at the maximum. The water flows down, wetting any point that has a non-ascending path 
 from $x$. We end up with two portions, the wet part of $\MM$ and the dry part. This is similar
@@
 But here's the catch. Every time we cut $\MM$ along a contour, we potentially increase
-the complexity of $\MM$. Water flows in the interior of facets, and the interface will naturally
-cut some facets. Each cut introduces new vertices, and a poor choice of repeated raining leads
+the complexity of $\MM$. Water flows in the interior of faces, and the interface will naturally
+cut some faces. Each cut introduces new vertices, and a poor choice of repeated raining leads
 to a large increase in complexity. Consider the left of \Fig{order}. Each raining produces a single
@@
 Apply the procedure above to get a single wet maximum dominant $\MM'$ and a set of dry pieces.
-Observe that a single dry piece $\NN$ is boundary critical with the newly introduced boundary $\phi$ (the wet-dry interface) behaving as a minima.
-So we can rain upwards from this minima, and get a \emph{minimum dominant} portion $\NN'$.
+Observe that a single dry piece $\NN$ is boundary critical with the newly introduced boundary $\phi$ (the wet-dry interface) behaving as a minimum.
+So we can rain upwards from this minimum, and get a \emph{minimum dominant} portion $\NN'$.
 This ensures that the new interface (after applying the procedure on $\NN$) does not
-cut any facet previously cut by $\phi$.
+cut any face previously cut by $\phi$.
 For each of the new dry pieces, the newly introduced boundary is now a maximum. So we rain downwards from there.
 More formally, we alternate between raining upwards and downwards as we go down the recursion tree.
-We can prove that an original facet of $\MM$ is cut at most once, so the final complexity can be bounded.
+We can prove that an original face of $\MM$ is cut at most once, so the final complexity can be bounded.
 In \Fig{order}, this procedure would yield two pieces, one maximum dominant, and one minimum dominant.
@@
 algorithm for join trees. Our algorithm does not require the input to be extremum
-dominant. This is only required to relate the join trees to the contour tree of initial input $\MM$.
-For convenience, we use $\NN$ to denote the input here. Note that as defined in \Def{int-criticalJoin},
+dominant. This is only required to relate the join trees to the contour tree of the initial input $\MM$.
+For convenience, we use $\NN$ to denote the input here. Note that in \Def{int-criticalJoin},
 the join tree is defined purely combinatorially in terms of the 1-skeleton (the underlying graph) of $\NN$.
@@
 
-The key idea is \emph{paint spilling}. Start with each maxima having a large can of paint, with distinct
-colors for each maxima. In arbitrary order, we spill paint from each maxima, wait till it flows down,
+The key idea is \emph{paint spilling}. Start with each maximum having a large can of paint, with distinct
+colors for each maximum. In arbitrary order, we spill paint from each maximum, wait till it flows down,
 then spill from the next, etc. Paint is viscous, and only flows down edges. \emph{It does not
-paint the interior of facets.} That is, this process is restricted to the 1-skeleton of $\NN$.
-Furthermore, our paints do not mix, so each edge receives a unique color, which is decided by the first
-paint to reach it.
+paint the interior of higher dimensional faces.} That is, this process is restricted to the 1-skeleton of $\NN$.
+Furthermore, our paints do not mix, so each edge receives a unique color, decided by the first
+paint to reach it.  In the following, $[n]$ denotes the set $\{1, \ldots, n\}$, for any natural number $n$.
 
 \begin{definition} \label{def:paint} Let the 1-skeleton of $\NN$ have edge set $E$ and maxima $X$.
-A  \emph{painting} of $\NN$ is a map $\chi:X \cup E \mapsto [|X|]$ with the following property. 
+A  \emph{painting} of $\NN$ is a map $\chi:X \cup E \to [|X|]$ with the following property. 
  Consider an edge $e$. There exists a descending path from some maximum $x$ to $e$
@@
 
-An \emph{initial} painting has the additional property that the restriction $\chi:X \mapsto [|X|]$ is a bijection.
+An \emph{initial} painting has the additional property that the restriction $\chi:X \to [|X|]$ is a bijection.
 \end{definition}
 
-Note that a painting only colors edges, not vertices. Our definition also does not require the timing aspect
-of iterating over colors, though that is one way of painting $\NN$. We begin with an initial painting, since all maxima
+Note that a painting colors edges, and not vertices (except for maxima). Our definition also does not require the timing aspect
+of iterating over colors, though that is one way of painting $\NN$. We begin with an initial painting, since all maximum
 colors are distinct. A few comments on paint vs water. The interface between two regions of different color
-is \emph{not} a contour, and is actually difficult to characterize. So we cannot apply the divide-and-conquer
+is \emph{not} a contour, and so we cannot apply the divide-and-conquer
 approach of contour surgery. On the other hand, painting does not cut $\NN$, so there is no increase in complexity.
@@
 at a valley $v$. Suppose this was the entire input, and $\ell_1$ was colored blue and $\ell_2$ was colored red. 
-Both the mounds are colored completely blue or red, while $v$ is touched by both colors.
-So this indicates that $v$ joins the blue maxima and red maxima in $\jc(\MM)$.
+Both mounds are colored completely blue or red, while $v$ is touched by both colors.
+So this indicates that $v$ joins the blue maximum and red maximum in $\jc(\MM)$.
 
@@
 handled. Imagine flattening the red and blue mounds until reaching $v$, so that the former join $v$
-is now a new maxima. Suppose purple paint was poured from $v$. In terms of $\jc(\MM)$, this
-is equivalent to removing leaves $\ell_1$ and $\ell_2$, making $v$ a new leaf.
-Alternately, $\jc(\MM)$ has been constructed up to $v$, and it remains to determine $v$'s parent.
-The merging of the colors is not explicitly performed as that would be too expensive; we maintain a union-find
+is now a new maximum, from which purple paint is poured. In terms of $\jc(\MM)$, this
+is equivalent to removing leaves $\ell_1$ and $\ell_2$, and making $v$ a new leaf.
+Alternatively, $\jc(\MM)$ has been constructed up to $v$, and it remains to determine $v$'s parent.
+The merging of the colors is not explicitly performed as that would be too expensive; instead we maintain a union-find
 data structure for that.
@@
 What happens when finally blue and red join at $v$? We merge the two colors, but now have
-blue and red queues of critical vertices. We also need to merge the priority queues
-to get a consistent painting. And that decides the use of \emph{binomial heaps}~\cite{Vu78} to store the
-critical points of one color. Merges can be done in logarithmic time, the same time
-required for deletes. We stress that the feasibility of the entire approach hinges on the use of binomial heaps.
+blue and red queues of critical vertices, which also need to be merged to get a consistent painting. 
+This necessitates using a priority queue with efficient merges.
+Specifically, we use \emph{binomial heaps}~\cite{Vu78}, as they provide logarithmic time merges and deletes (though other similar heaps work). 
+We stress that the feasibility of the entire approach hinges on the use of such an efficient heap structure.
 
@@
 ``side" of a possible join. We are guaranteed that when $v$ needs to be processed, all edges have
-at most $2$ colors, because of all the color merges that previously occured.
+at most $2$ colors, because of all the color merges that previously occurred.
 
@@
 The cost of heap operations depends on the sizes of the heaps, which keeps changing because
-of the repeated merging. This is a major headache for the analysis. There are situations where
+of the repeated merging. This causes the analysis to be technically challenging. There are situations where
 the initial heap sizes are small, but they eventually merge to create larger heaps. 
@@
 trees of these tiny tents attach to a main branch of length $|p|$. Working out the details,
-we get $\prod_{p \in P(T)} |p|!$ terrains each with a distinct contour tree. It is possible to argue
-that our algorithm requires $\Theta(\cost(P(T)))$ comparisons on all these instances.
-
-\FullVer{
+we get $\prod_{p \in P(T)} |p|!$ terrains each with a distinct contour tree. 
+Our algorithm requires $\Theta(\cost(P(T)))$ comparisons on all these instances.
+
+%\FullVer{
 \section{Divide and conquer through contour surgery} \label{sec:surgery}
@@
 {\bf The cutting operation:} We define a ``cut" operation on $f:\MM \rightarrow \RR$ that cuts along a regular contour to create
-a new simplicial complex with an added boundary. Given contour $\phi$, this is constructing the simplicial complex $\MM \setminus \phi$. 
+a new simplicial complex with an added boundary. Given a contour $\phi$, roughly speaking, this constructs the simplicial complex $\MM \setminus \phi$. 
 We will always enforce the condition that $\phi$ never passes through a vertex of $\MM$.
@@
 An $h$-contour is achieved by intersecting $\MM$ with the hyperplane $x_{d+1} = h$ and taking a connected component. (Think of the $d+1$-dimension
-as height.) Given some point $x$
-on an $h$-contour $\phi$ (with some simplex that $x$ is present in), we can walk along $\MM$ from $x$ and determine $\phi$.
+as height.) Given some point $x$ on an $h$-contour $\phi$, we can walk along $\MM$ from $x$ to determine $\phi$.
 We can ``cut" along $\phi$ to get a new (possibly) disconnected simplicial complex $\MM'$. This is achieved
-by splitting every facet $F$ that $\phi$ intersects into an ``upper" facet and ``lower" facet. Algorithmically,
-we cut facet $F$ with $\phi^+$ and take everything above $\phi^+$ in $F$ to make the upper facet. Analogously, we cut with $\phi^-$ to get the lower facet.
-The facets are then triangulated to ensure that they are all simplices.
-Given that $\phi$ cannot cut a boundary and all non-boundary facets
-have constant size, we omit the algorithmic description for this process. We only note that this
-creates the two new boundaries $\phi^+$ and $\phi^-$, and we maintain the property of constant $f$-value at a boundary.
+by splitting every face $F$ that $\phi$ intersects into an ``upper" face and ``lower" face. Algorithmically,
+we cut $F$ with $\phi^+$ and take everything above $\phi^+$ in $F$ to make the upper face. Analogously, we cut with $\phi^-$ to get the lower face.
+The faces are then triangulated to ensure that they are all simplices.
+This creates the two new boundaries $\phi^+$ and $\phi^-$, and we maintain the property of constant $f$-value at a boundary.
+
+Note that by assumption $\phi$ cannot cut a boundary face, and moreover all non-boundary faces
+have constant size. Therefore, this process takes time linear in $|\phi|$, the number of faces $\phi$ intersects.
 This new simplicial complex is denoted by $\cut(\phi,\MM)$.
-This can be constructed in time linear in $|\phi|$.
-
-%We denote this new manifold by $\cut(\phi,\MM)$, and the new boundaries by $\phi^+$ and $\phi^-$.
-%Abusing notation, we use $\cut(\phi,\MM)$
-%for a set $\Psi$ of contours to mean the manifold obtained by cutting along all contours in $\Phi$.
-
-We now describe a high-level approach to construct $\reeb(\MM)$.
+%
+We now describe a high-level approach to construct $\reeb(\MM)$ using this cutting procedure.
 
@@
 	\item Construct $\reeb(\MM')$ and let $A, B$ be the nodes corresponding to the new boundaries created
-	in $\MM'$. (One is a minima and the other is maxima.)
+	in $\MM'$. (One is a minimum and the other is maximum.)
 	\item Since $A, B$ are leaves, they each have unique neighbors $A'$ and $B'$, respectively. Insert
@@
 
-We require some theorems from \cite{c-tmi-04} (Theorems 6.6 and 6.7) that map paths in $\cC(\MM)$ to $\MM$.
+To prove the above theorem, we require some theorems from \cite{c-tmi-04} (Theorems 6.6 and 6.7) that map paths in $\cC(\MM)$ to $\MM$.
 
@@
 
-The main theorem is a direct consequence of the following lemma.
+Theorem~\ref{thm:surgery} is a direct consequence of the following lemma.
 
@@
 
-\begin{proof} The contours of $\MM'$ are exactly the contours of $\MM$ with $\phi^+$ and $\phi^-$, without $\phi$.
-All new vertices created in $\MM'$ are present on $\phi^+$ and $\phi^-$. That proves the first part.
-
-Any contour class in $\MM'$ (edge in $\cC(\MM')$) that does not involve $\phi^+$ or $\phi^-$ is also
+\begin{proof} First observe that since $\phi$ is a regular contour, the vertex set in the complex $\MM'$ is the same 
+as the vertex set in $\MM$, except with the addition of the newly created vertices on $\phi^+$ and $\phi^-$.  
+Moreover, $\cut(\MM,\phi)$ does not affect the local neighborhood of any vertex in $\MM$. 
+Therefore since a vertex being critical is a local condition, with the exception of new boundary vertices, the 
+critical vertices in $\MM$ and $\MM'$ are the same.  Finally, the new vertices on $\phi^+$ and $\phi^-$ 
+collectively behave as a minimum and maximum, respectively, and so $\cV(\reeb(\MM')) = \{\phi^+,\phi^-\} \cup \cV(\MM)$.
+
+Now consider the edge sets of the contour trees.  Any contour class in $\MM'$ (i.e.\ edge in $\cC(\MM')$) that does not involve $\phi^+$ or $\phi^-$ is also
 a contour class in $\MM$. Furthermore, a maximal contour class satisfying these properties is also
@@
 in $\MM' = \cut(\MM,\phi)$. By \Thm{carr-path}, there is a path in $\cC(\MM')$ between $u$ and $v$. Since $\phi^+$ and $\phi^-$
-are leaves in $\cC(\MM')$, this path obviously cannot use edges incident to them. By \Lem{cut},
+are leaves in $\cC(\MM')$, this path cannot use their incident edges. Therefore by \Lem{cut},
 all the edges of this path are in $\cE(\cC(\MM)) \setminus (u,v)$. So we get a cycle in $\cC(\MM)$, a contradiction.
 To show that there are exactly two connected components in $\cut(\MM,\phi)$, it suffices
-to see that $\cC(\MM')$ has two connected components (by \Lem{cut}) and apply \Thm{carr-path}.
+to see that $\cC(\MM')$ has two connected components (by \Lem{cut}) and then applying \Thm{carr-path}.
 \end{proof}
@@
 \begin{definition} \label{def:dom} A simplicial complex is \emph{minimum dominant} if there exists
-a minimum $x$ such that every non-minimal \emph{vertex} in the manifold has a decreasing path to $x$.
+a minimum $x$ such that every non-minimal \emph{vertex} in the manifold has a non-ascending path to $x$.
 Analogously define \emph{maximum dominant}. 
@@
 The first aspect of the partitioning is ``raining". Start at some point $x \in \MM$ and imagine rain at $x$.
-The water will flow downwards along descending paths and ``wet" all the points encountered. Note that this procedure considers all points
+The water will flow downwards along non-ascending paths and ``wet" all the points encountered. Note that this procedure considers all points
 of the manifold, not just vertices.
 
-\begin{definition} \label{def:wet} Fix $\MM$ and $x \in \MM$. The set of points $y \in \MM$ such that there is a non-ascending path from $x$ to $y$
-is denoted by $\wet(x,\MM)$ (which is in turn is represented as a simplicial complex). A point $z$ is at the \emph{interface} of $\wet(x,\MM)$ if every neighborhood of $z$
+\begin{definition} \label{def:wet} Fix $x \in \MM$. The set of points $y \in \MM$ such that there is a non-ascending path from $x$ to $y$
+is denoted by $\wet(x,\MM)$ (which in turn is represented as a simplicial complex). A point $z$ is at the \emph{interface} of $\wet(x,\MM)$ if every neighborhood of $z$
 has non-trivial intersection with $\wet(x,\MM)$.
@@
 \begin{proof} If $p \in \wet(x,\MM)$, all the points in any contour containing $p$ are also in $\wet(x,\MM)$.
- (Follow the non-ascending path to from $x$ to $p$ and then walk along the contour.) The converse is also true,
+ (Follow the non-ascending path from $x$ to $p$ and then walk along the contour.) The converse is also true,
  so $\wet(x,\MM)$ contains entire contours.
@@
 intersecting $B_\eps(z)$. Now, walk from $z$ to $y$ along $\Gamma_y$. If all points $w$ in this walk
-have up-degree $1$, then $\psi$ is the unique $f(y)+\delta$-contour
+have up-degree $1$, then $\psi$ is the unique $(f(y)+\delta)$-contour
 intersecting $B_\eps(w)$. This would imply that $\phi = \psi$, contradicting the fact that $\psi$ is wet
@@
 Note that $\wet(x,\MM)$ (and its interface) can be computed in time linear in the size of the wet simplicial complex.
-We perform a non-ascending search from $x$. Any facet $F$ of $\MM$ encountered is partially (if not entirely) in $\wet(x,\MM)$.
-This portion is determined by cutting $F$ along the interface. Since the interface is a contour, this is equivalent
+We perform a non-ascending search from $x$. Any face $F$ of $\MM$ encountered is partially (if not entirely) in $\wet(x,\MM)$.
+The wet portion is determined by cutting $F$ along the interface. Since the interface is a contour, this is equivalent
 to cutting $F$ by a hyperplane. All these operations can be performed to output $\wet(x,\MM)$ in time linear in $|\wet(x,\MM)|$.
@@
 a join vertex $y$. Take any dry increasing edge incident to $y$, and pick the point $z$ on this edge at height
-$f(y) + \delta$. Let $\lift(\phi)$ be the unique contour through the regular point $z$. Note that $\lift(\phi)$ is dry.
+$f(y) + \delta$ (where $\delta$ is an infinitesimal, but larger than the value $\eps$ used in the definition of $\cut$). 
+Let $\lift(\phi)$ be the unique contour through the regular point $z$. Note that $\lift(\phi)$ is dry.
 
@@
 
-Focus on $\rain(x,\MM)$. As $\rain$ proceeds, new facets/simplices are created because of repeated cutting. Indeed, any facet $F$
-in any simplicial complex $\NN$ that is recursively invoked, is achieved by cutting facet $F'$ in $\MM$.
-The key to the running time of $\rain(x,\MM)$ is bounding the number of newly created facets, for which we have the following lemma.
+As $\rain(x,\MM)$ proceeds, new faces/simplices are created because of repeated cutting. 
+%Indeed, any facet $F$ in any simplicial complex $\NN$ that is recursively invoked, is achieved by cutting facet $F'$ in $\MM$.
+The key to the running time of $\rain(x,\MM)$ is bounding the number of newly created faces, for which we have the following lemma.
 
 \begin{lemma}\label{lem:new-verts}
-A facet $F \in \MM$ is cut at most once during $\rain(x,\MM)$.
+A face $F \in \MM$ is cut\footnote{Technically what we are calling a single cut is done with two hyperplanes.} at most once during $\rain(x,\MM)$.
 \end{lemma}
@@
 Specifically, say this happens when $\cut(\NN_i,\phi'_i)$ is constructed. 
-So $F$ is cut by some hyperplane into lower and upper portions (which are then triangulated).
-Since the cut is infinitesimally above an interface, the lower portion is basically wet.
-It can never be part of $\LL_1, \LL_2, \ldots$, which go into recursive calls. This portion
-is in $\NN_{k+1}$. The upper portion/facet (call it $U$) is in $\LL_i$. Note that the lower
+$\cut(\NN_i,\phi'_i)$ will cut $F$ with two horizontal cutting planes, one $\eps$ above $\phi'_i$
+and one $\eps$ below $\phi'_i$.  This breaks $F$ into lower and upper portions which are then triangulated
+(there is also a discarded middle portion).  The lower portion, which is adjacent $\phi_i$, gets included 
+in $\NN_{k+1}$, the complex containing the wet points, and hence does not participate in any later recursive calls.  
+The upper portion (call it $U$) is in $\LL_i$. Note that the lower
 boundary of $U$ is in the boundary $B_i$. Since a recursive call is made to $\rain(B_i,\LL_i)$
-(and $\LL_i$ is inverted), $U$ becomes wet. Hence, $U$ will not be cut subsequently.
-All in all, $F$ is cut at most once.
-\ignore{
-As the number of edges is $O(n)$, arguing that there is at most one boundary vertex created along any given edge will imply the claim.
-Consider some edge $e$ and take the first time a boundary vertex is created in $e$.
-This happens when some contour $\phi$ intersects $e$. On applying $\cut(\NN,\lift(\phi))$,
-consider the dry manifold $\LL$ and wet manifold $\MM$ constructed. The edge $e$ is split into two parts, the dry and wet part.
-The wet part is in $\MM$ and will never be involved in any recursive call.
-
-The dry part of $e$ (call it $e'$) needs to be considered.
-A boundary vertex $v$ (an new endpoint of $e$) is created, which is part of the boundary $B$ in $\LL$,
-such that $\rain(B,\LL)$ is called to the inverted $\LL$. In this call, since the water starts from $v$,
-all of $e'$ becomes wet. So there is no further interface contour that can intersect this portion of $e'$.
-}
+(and $\LL_i$ is inverted), $U$ becomes wet. Hence $U$, and correspondingly $F$, will not be subsequently cut.
 \end{proof}
@@
 
-\begin{proof} The operations performed are the raining and cutting. We can bound this by the total
-complexities of the wet complexes and the new boundaries created. No facet is ever wet twice,
-and by \Lem{new-verts}, the total number of facets that every appear is $O(|\MM|)$. This also 
-bounds the complexities of the boundaries created.
+\begin{proof} The only non-trivial operations performed are $\wet$ and $\cut$.  
+Since $\cut$ is a linear time procedure, Lemma~\ref{lem:new-verts} implies the total time for all calls to $\cut$ is $O(|\MM|)$.  
+As for the $\wet$ procedure, observe that Lemma~\ref{lem:new-verts} additionally implies there are only $O(|\MM|)$ new faces created by $\rain$.
+Therefore, since $\wet$ is also a linear time procedure, and no face is ever wet twice, the total time for all calls to $\wet$ is $O(|\MM|)$.
 \end{proof}
@@
 contour tree of its children in the tree using the $\surgery$ procedure. 
-Each $\surgery$ call is constant time, and the total time is the size of the recursion tree.
+Each $\surgery$ call takes constant time, and the total time is the size of the recursion tree.
 \end{proof}
@@
 We will fix such a simplicial complex $\MM$, with the dominant minimum $m^*$. 
-The set of non-dominant minima is denoted by $M$.
+%The set of non-dominant minima is denoted by $M$.
 For vertex $v$, we use $\MM^+_v$ to denote the simplicial complex obtained by only
@@
 The exact statement will require some definitions and notation.
-We required the notions of \emph{join} and \emph{split} trees, as given by~\cite{csa-cctad-00}.
-Conventially, all edges are directed from higher to lower function value.
-
-\begin{definition} \label{join} The join tree $\cJ(\MM)$ of $\MM$ is built on vertex set $V(\MM)$.
+We require the notions of \emph{join} and \emph{split} trees, as given by~\cite{csa-cctad-00}.
+Conventionally, all edges are directed from higher to lower function value.
+
+\begin{definition} \label{join} The \emph{join tree} $\cJ(\MM)$ of $\MM$ is built on vertex set $V(\MM)$.
 The directed edge $(u,v)$ is present when $u$ is the smallest valued vertex in a connected component of $\MM^+_v$
-\emph{and} $v$ is connected to this component (in $\MM$). The split tree $\cS(\MM)$ is obtained
+\emph{and} $v$ is connected to this component (in $\MM$). The \emph{split tree} $\cS(\MM)$ is obtained
 by looking at $\MM^-_v$ (or alternately, by taking the join tree of the inversion of $\MM$). 
@@
 and the global minimum is the root. All indegrees in $\cS(\MM)$ are at most $1$, all outdegree $2$
-vertices are splits, leaves are minima, and the global maximum is the root.
-%
-As these trees rooted, we can use ancestor-descendant terminology.  
+vertices are splits, all leaves are minima, and the global maximum is the root.
+%
+As these trees are rooted, we can use ancestor-descendant terminology.  
 Specifically, for two adjacent vertices $u$ and $v$, $u$ is the parent of $v$ if $u$ is closer to the root 
-(i.e. each node can have at most one parent, but can have two children).
+(i.e.\ each node can have at most one parent, but can have two children).
 
@@
 merging procedure from~\cite{csa-cctad-00} that constructs the contour tree from the join and split trees. 
-It actually construsts the \emph{augmented contour tree} $\cA(\MM)$, which
-is the contour tree with regular points inserted into the edges. 
-Consider tree $T$ with vertex $v$ of in and outdegree at most $1$.
-\emph{Erasing} $v$ from $T$ is the following operation: if $v$ is a leaf, just delete $v$. Otherwise, smooth $v$ out
-(delete $v$ and connect its neighbors by an edge). This tree is denoted by $T \ominus v$.
+It actually constructs the \emph{augmented contour tree} $\cA(\MM)$, which
+obtained by replacing each edge in the contour tree with a path of all regular vertices 
+(sorted by height) whose corresponding contour belongs to the equivalence class of that edge.
+
+Consider a tree $T$ with a vertex $v$ of in and out degree at most $1$.
+\emph{Erasing} $v$ from $T$ is the following operation: if $v$ is a leaf, just delete $v$. Otherwise, 
+delete $v$ and connect its neighbors by an edge (i.e.\ smooth $v$ out). This tree is denoted by $T \ominus v$.
 
@@
 	\item Set $\cJ = \cJ(\MM)$ and $\cS = \cS(\MM)$.
-	\item Denote $v$ as \emph{candidate} if sum of indegree of $v$ in $\cJ$ and outdegree of $v$ in $\cS$ is $1$. 
+	\item Denote $v$ as a \emph{candidate} if the sum of its indegree in $\cJ$ and outdegree in $\cS$ is $1$. 
 	\item Add all candidates to queue.
@@
 \label{def:criticalJoin}
-The critical join tree $\jc(\MM)$ is built on the set $V'$ of all
+The \emph{critical join tree} $\jc(\MM)$ is built on the set $V'$ of all
 critical points other than the non-dominant minima. 
@@
 
-Note that $\cC(\MM)$ is obtained by smoothing out all regular points from $\cA(\MM)$. Smoothing out regular
-points from $\cJ(\MM) \ominus \{m_i\}$ exactly yields the edges described in the theorem.
+Note that $\cC(\MM)$ is obtained by smoothing out all regular points from $\cA(\MM)$. Similarly, smoothing out regular
+points from $\cJ(\MM) \ominus \{m_i\}$ yields the edges of $\jc(\MM)$.
 \end{proof}
@@
 of \emph{painting} $\MM$. Initially associate a color with each maximum. Imagine there being a large
-can of paint of a distinct color at each maxima $x$. We will spill different paint from each maximum and watch it flow down.
+can of paint of a distinct color at each maximum $x$. We will spill different paint from each maximum and watch it flow down.
 This is analogous to the raining in the previous section, but paint is a much more viscous liquid.
-\emph{So paint only flows down edges, and it does not color the interior of facets.} Furthermore, paints
+\emph{So paint only flows down edges, and it does not color the interior of higher dimensional faces.} Furthermore, paints
 do not mix, so every edge of $\MM$ gets a unique color. This process (and indeed the entire algorithm)
@@
 \begin{definition} Let the 1-skeleton of $\MM$ have edge set $E$ and maxima $X$.
-A  \emph{painting} of $\MM$ is a map $\chi:X \cup E \mapsto [|X|]$ with the following property. 
+A  \emph{painting} of $\MM$ is a map $\chi:X \cup E \to [|X|]$ with the following property. 
  Consider an edge $e$. There exists a descending path from some maximum $x$ to $e$
@@
 
-An \emph{initial} painting has the additional property that the restriction $\chi:X \mapsto [|X|]$ is a bijection.
+An \emph{initial} painting has the additional property that the restriction $\chi:X \to [|X|]$ is a bijection.
 \end{definition}
@@
 	 \item A color $c \in \col(v)$ \emph{fully touches} $v$ if all edges in an up-star are colored $v$.
+	 \item For any maximum $x\in X$, we say that $x$ is both touched and fully touched by $\chi(x)$.
 \end{asparaitem}
@@
 
-Each non-maxima $v$ participates in at least $1$ and at most $2$ up-stars (the latter iff
+Each non-maximum $v$ participates in at least $1$ and at most $2$ up-stars (the latter iff
 $v$ is a join).
@@
 \begin{definition} \label{def:ripe} A vertex $v$ is \emph{ripe} if: for all $c \in \col(v)$, $v$
-is present in $T(\rep(c))$ and is also highest vertex in the heap. 
+is present in $T(\rep(c))$ and is also the highest vertex in this heap. 
 %Also, for all $c \in \col(v) \setminus \fcol(v)$, $T(\rep(c))$ is empty.
@@
 	\item Initialize $\rep(c) = c$ and set $\h(c)$ to be the unique maximum colored $c$.
+	\item Initialize $K$ to be an empty stack.
 \end{compactenum}
@@
 	\item Each vertex is present in at most $2$ heaps. After processing, it is removed from all heaps.
-	\item After $v$ is processed, all edges incident to $v$ have the same color (technically, same representative). 
+	\item After $v$ is processed, all edges incident to $v$ have the same color (more technically, the same representative). 
 %	\item When $\update(\st)$ terminates, the head $h$ is ripe. 
@@
 \label{obs:twoQueues}
- Each unprocessed vertex is always in exactly one queue of the colors in each up-star.  Specifically, 
- for a given up-star of a vertex $v$, $\init(\MM)$ puts $v$ into the queue of exactly one color, say $c$.  As time goes on this queue 
- may merge with other queues, but while unprocessed $v$ is only ever (and always) in the queue of $\rep(c)$,  
+ Each unprocessed vertex is always in exactly one queue of the colors in each of its up-stars.  Specifically, 
+ for a given up-star of a vertex $v$, $\init(\MM)$ puts $v$ into the queue of exactly one of the colors of the up-star, say $c$.  As time goes on this queue 
+ may merge with other queues, but while $v$ remains unprocessed, it is only ever (and always) in the queue of $\rep(c)$,  
  since $v$ is never added to a new queue and is not removed until it is processed.  
- In particular, finding the queues of a vertex in $\update(K)$ requires at most two union find operations (assuming each vertex records its two colors from $\init(\MM)$)
+ In particular, finding the queues of a vertex in $\update(K)$ requires at most two union find operations (assuming each vertex records its two colors from $\init(\MM)$).
 \end{observation}
@@
 
+\Ben{Current color is never defined, but it means the the current rep(c).}
+
 \begin{lemma} \label{lem:full} Suppose vertex $v$ is connected to a component $\PP$ of $\MM^+_v$
-by edge $e$ which is currently colored $c$. Either all edges in $\PP$ are currently colored $c$, or there
-exists critical vertex $w \in \PP$ fully touched by $c$ and touched by another color.
+by an edge $e$ which is currently colored $c$. Either all edges in $\PP$ are currently colored $c$, or there
+exists a critical vertex $w \in \PP$ fully touched by $c$ and touched by another color.
 \end{lemma}
@@
 So assume $w$ exists. Take the $(f(w)+\delta)$-contour $\phi$ that intersects $B_\eps(v)$
-and intersect some $c$-colored edge incident to $w$. Note that all edges intersecting $\phi$ are also colored $c$,
+and intersects some $c$-colored edge incident to $w$. Note that all edges intersecting $\phi$ are also colored $c$,
 since $w$ is the highest vertex to be touched by $c$ and some other color. (Take the path of $c$-colored
@@
 are edges of $\jc(\MM)$ and for any current color $c$, $\h(c)$ is the lowest processed vertex
-of that color. We prove this by induction on the processing order. The base case is is trivially
+of that color. We prove this by induction on the processing order. The base case is trivially
 true, as initially the processed vertices and attachments of the color classes are the set of maxima.
@@
 (safely) ignored to simplify the presentation.} where the height of a vertex is its distance 
-from the root $r$ (i.e. conceptually $T$ will be a join tree with $r$ at the bottom).  
+from the root $r$ (i.e.\ conceptually $T$ will be a join tree with $r$ at the bottom).  
 As such, the children of a vertex $v\in T$ are the adjacent vertices of larger height
@@
 Indeed, the painting procedure consists of several BFS's but as each vertex is only explored by one of the BFS's, it is linear time overall.
-Determining the critical points is a local computation on the neighborhood of each vertex as so is linear (i.e. each edge is viewed at most twice).
+Determining the critical points is a local computation on the neighborhood of each vertex as so is linear (i.e.\ each edge is viewed at most twice).
 Finally, each vertex is inserted into at most two heaps and so initializing the heaps takes linear time in the number of vertices.
@@
 
- Now for any $i\neq j$, $H_i \cap H_j = \emptyset$, as $r_{i}$ and $r_{j}$ are not on the same root to leaf path (i.e. $L(T_{r_i}) \cap L(T_{r_j}) = \emptyset$).
+ Now for any $i\neq j$, $H_i \cap H_j = \emptyset$, as $r_{i}$ and $r_{j}$ are not on the same root to leaf path (i.e.\ $L(T_{r_i}) \cap L(T_{r_j}) = \emptyset$).
  In particular, $(H_i\setminus H_i') \cap (H_j \setminus H_j') = \emptyset$, and so $\sum_{i=0}^k |H_i\setminus H_i'| \leq |p|$ 
@@
 \ignore{
-For a short path $p$ we can think of the quantity $|p|(\log |H_p| - \log |p|)$ as the excess of $p$, i.e. what was not ``paid for'' locally by $p$.
+For a short path $p$ we can think of the quantity $|p|(\log |H_p| - \log |p|)$ as the excess of $p$, i.e.\ what was not ``paid for'' locally by $p$.
 Our eventual goal will be to show that by choosing the right path decomposition, this excess 
@@
 \emph{tall support chain} of $p$ as the tall ancestors of $p$ in the shrub which have a path to $p$ which does not use any 
-short path (i.e. it is a chain of paths adjacent to $p$).
+short path (i.e.\ it is a chain of paths adjacent to $p$).
 
@@
 Consider any $v \in H'_p$.  Let $S$ be the set of paths $\widetilde{p}\in \cL'$ such that 
-$v\in H'_{\widetilde{p}}$.  We now show $|S|\leq 2$ (i.e. it contains at most one path other than $p$).
-First observe that any two paths in $S$ must be unrelated (i.e. $S$ is an anti-chain), 
+$v\in H'_{\widetilde{p}}$.  We now show $|S|\leq 2$ (i.e.\ it contains at most one path other than $p$).
+First observe that any two paths in $S$ must be unrelated (i.e.\ $S$ is an anti-chain), 
 since paths which have an ancestor-descendant relationship have disjoint tall support chains.  
 However, any vertex $v$ receives exactly one color from each of its two subtrees (in $T$), and therefore $|S|\leq 2$ since any two 
-paths which share descendant leaves in $T$ (i.e. their heaps are tracking the same color) must have an ancestor-descendant relationship.
+paths which share descendant leaves in $T$ (i.e.\ their heaps are tracking the same color) must have an ancestor-descendant relationship.
 
@@
 
-Throughout this section, for a path $p\in P(T)$, we use the notation $H_p$ to refer to the heap at the root of $p$, i.e. $H_p = H_{r_p}$.
+Throughout this section, for a path $p\in P(T)$, we use the notation $H_p$ to refer to the heap at the root of $p$, i.e.\ $H_p = H_{r_p}$.
 }
@@
 % Let $w$ be the vertex in the path $p$ to which the root of the path $q$ is adjacent to in $T$, 
-% i.e. $w$ is the parent of $r_q$ in $T$.  Let $c$ be the other child of $w$, which lies in $p$.
+% i.e.\ $w$ is the parent of $r_q$ in $T$.  Let $c$ be the other child of $w$, which lies in $p$.
 % As $\mathcal{P}(T)$ is a maximum path decomposition of $T$ and $c$ rather than $r_q$ is in 
@@
 \begin{theorem}
-Consider a PL-Morse $f:\MM \mapsto \RR$, where the join tree $\jc(\MM)$ has maximum degree $3$. 
+Consider a PL-Morse $f:\MM \to \RR$, where the join tree $\jc(\MM)$ has maximum degree $3$. 
 There is an algorithm to compute the join tree whose running time is $O(\sum_{p \in \pmax(\jc)} |p|\log |p| + t\alpha(t) + N)$.
@@
 \begin{theorem}
-Consider a PL-Morse $f:\MM \mapsto \RR$, where the contour tree $\cC=\cC(\MM)$
+Consider a PL-Morse $f:\MM \to \RR$, where the contour tree $\cC=\cC(\MM)$
 has maximum degree $3$. There is an algorithm to compute $\cC$ whose running time is $O(\sum_{p \in P(\cC)} |p|\log |p| + t\alpha(t) + N)$,
@@
 However, by the above theorem we know $\jc(\MM_i)$ can be computed in $O(\sum_{p \in \pmax(\jc(\MM_i))} |p|\log |p| + t_i\alpha(t_i) + N_i)$
-(where $t_i$ and $N_i$ are the number of critical points and facets when restricted to $\MM_i$).
+(where $t_i$ and $N_i$ are the number of critical points and faces when restricted to $\MM_i$).
 
@@
 We focus on terrains, so $d=2$.
-Consider any path decomposition $P$ of a valid join tree (i.e. any rooted binary tree).
+Consider any path decomposition $P$ of a valid join tree (i.e.\ any rooted binary tree).
 When we say ``compute the join tree", we require the join tree to
@@
  diamonds.  We create a global boundary minimum by setting the vertices at the base of the diamond representing the root of $\pathTree$ all to the same height (and there are no other minima).  
- Therefore, the saddles on a given diamond will appear contiguously on a root to leaf path in the join tree of the terrain, where the leaf corresponds to the maxima of the diamond 
+ Therefore, the saddles on a given diamond will appear contiguously on a root to leaf path in the join tree of the terrain, where the leaf corresponds to the maximum of the diamond 
  (since all these saddles have a direct line of sight to this apex).  In particular, 
@@
  Consider any join on a face of some diamond, which is the anchor point of some connecting child diamond.  
- The join has two up-stars, each of which has exactly one edge. Each edge connects to a maxima
- and must be colored by that maxima.
+ The join has two up-stars, each of which has exactly one edge. Each edge connects to a maximum
+ and must be colored by that maximum.
  Hence, the two colors touching this join (according to \Def{initialColoring})
@@
  Take any join $v$, with two children $w_1$ and $w_2$. Suppose $w_1$ and $v$ belong to the same path
- in the decomposition. The key is that any color from a maxima in the subtree at $w_2$ cannot touch
+ in the decomposition. The key is that any color from a maximum in the subtree at $w_2$ cannot touch
  any ancestor of $v$. This subtree is exactly the contour tree of the child diamond attached at $v$.
@@
 % which $c$ appears are exactly the vertices of the path in  $\overline{P}(T)$ corresponding $m$, as well as 
-% the vertex to which the root of this path connects to (i.e. the anchor of the diamond of $m$).
+% the vertex to which the root of this path connects to (i.e.\ the anchor of the diamond of $m$).
 % Now we show this implies $\overline{P}(T)$ is a maximum path decomposition.
@@
 %and the height function defined over the vertex set is such that every internal vertex in either a split or join 
-%(i.e. at most $2$ neighbors lie above or below it).
+%(i.e.\ at most $2$ neighbors lie above or below it).
 
@@
 %Again we define a path decomposition as a partition of the vertices into 
-%disjoint leaf paths (i.e. each path is monotonic in height and contains a leaf). 
+%disjoint leaf paths (i.e.\ each path is monotonic in height and contains a leaf). 
 For simplicity we now restrict our attention to path decompositions 
@@
 \label{def:path2} A path decomposition, $P(T)$, is called \emph{rain consistent} if its paths can be obtained as follows.
-Perform an downward BFS from an arbitrary maxima $v$ in $T$, and mark all vertices encountered.  
+Perform an downward BFS from an arbitrary maximum $v$ in $T$, and mark all vertices encountered.  
 Now recursively run a directional BFS from all vertices adjacent to the current marked set.  
@@
 For each such subtree, now take any partition of the vertices into leaf paths.\footnote{Note that the subtree  
-of the initial vertex is rooted at is a maxima.  For simplicity we require that the path this vertex belongs to 
-also contains a minima.}
+of the initial vertex is rooted at a maximum.  For simplicity we require that the path this vertex belongs to 
+also contains a minimum.}
 \end{definition}
@@
  let $r_i$ be the root vertex of this subtree.  
- Note that the $P_i$ sets naturally define a tree where $P_i$ is the parent of $P_j$ if $r_i$ (i.e. the root of $T_i$) is adjacent to a vertex in $P_j$.
+ Note that the $P_i$ sets naturally define a tree where $P_i$ is the parent of $P_j$ if $r_i$ (i.e.\ the root of $T_i$) is adjacent to a vertex in $P_j$.
 
@@
  By construction, the diamonds corresponding to the paths in $P_i$ are connected 
- into a tree structure (i.e. corresponding to the shrub of $P_i$).  Therefore the bottoms of all these diamonds are covered except for the one corresponding 
+ into a tree structure (i.e.\ corresponding to the shrub of $P_i$).  Therefore the bottoms of all these diamonds are covered except for the one corresponding 
  to the path containing the root $r_i$.  
- If $r_i$ corresponds to the initial maxima that the rain consistent path decomposition was defined from, then this will be flat and corresponds to the global outer face.
+ If $r_i$ corresponds to the initial maximum that the rain consistent path decomposition was defined from, then this will be flat and corresponds to the global outer face.
  Otherwise, $P_i$ has some parent $P_j$ in which case we connect the bottom of the diamond for $r_i$ to a free face of a diamond in the construction for $P_j$, specifically, the face corresponding to the
@@
 % For every path we create a cylinder with $m$ faces for $p$, where $m$ is the number of non-leaf vertices in $p$ (either $|p|-1$ or $|p|-2$).  
-% If $p$ has a maxima we put a top cone cap on the cylinder, and if it contains a manima then put a bottom cone cap (i.e. the cap represents the maxima or minima).
+% If $p$ has a maxima we put a top cone cap on the cylinder, and if it contains a manima then put a bottom cone cap (i.e.\ the cap represents the maxima or minima).
 % 
@@
 % \end{enumerate}
-}
+%}
 

### 2015-07-02 paint-contour-tree.tex
interface fixed

@@
 while there may be numerous disconnected dry parts. The interface between the dry and wet parts
-is a set of contours, given by the ``water line". The wet part is clearly maximum dominant, since all
+is a set of contours\footnote{Technically, they are not contours, but rather the limits of sequences of contours.}, 
+given by the ``water line". The wet part is clearly maximum dominant, since all
 wet points have a non-descending path to $x$. So we can simply cut along the interface
@@
 
-\begin{claim} \label{clm:inter} For any $x$, the interface of $\wet(x,\MM)$ is a set of contours,
-each containing a join vertex.
+\begin{claim} \label{clm:inter} For any $x$, each connected component of the interface of $\wet(x,\MM)$ contains a join vertex.
 \end{claim}
@@
 and $\phi$ is dry.
-
-Therefore, there must exist a join vertex $w$ in $\Gamma_y$, such that $B_\eps(w)$ intersects
-both $\phi$ and $\psi$. As $\delta, \eps \rightarrow 0^+$, the limit of $\phi$ (which is dry)
-lies in $\Gamma_y$ (which is wet). 
-Hence this limit gives a contour of the interface containing the join $w$.
+%
+% Therefore, there must exist a join vertex $w$ in $\Gamma_y$, such that $B_\eps(w)$ intersects
+% both $\phi$ and $\psi$. As $\delta, \eps \rightarrow 0^+$, the limit of $\phi$ (which is dry)
+% lies in $\Gamma_y$ (which is wet). 
+% Hence this limit gives a contour of the interface containing the join $w$.
 \end{proof}
@@
 We perform a non-ascending search from $x$. Any face $F$ of $\MM$ encountered is partially (if not entirely) in $\wet(x,\MM)$.
-The wet portion is determined by cutting $F$ along the interface. Since the interface is a contour, this is equivalent
-to cutting $F$ by a hyperplane. All these operations can be performed to output $\wet(x,\MM)$ in time linear in $|\wet(x,\MM)|$.
-
-We define a simple \lift{} operation on the interface contours. Consider such a contour $\phi$ containing
+The wet portion is determined by cutting $F$ along the interface. Since each component of the interface connected subset of a contour, this is equivalent
+to locally cutting $F$ by a hyperplane. All these operations can be performed to output $\wet(x,\MM)$ in time linear in $|\wet(x,\MM)|$.
+
+We define a simple \lift{} operation on the interface components. Consider such a component $\phi$ containing
 a join vertex $y$. Take any dry increasing edge incident to $y$, and pick the point $z$ on this edge at height
@@
 
-\begin{claim} \label{clm:cut-int} Let $\phi$ be an interface contour. Then $\cut(\MM,\lift(\phi))$
+\begin{claim} \label{clm:cut-int} Let $\phi$ be a connected component of the interface. Then $\cut(\MM,\lift(\phi))$
 results in two disjoint simplicial complexes, one consisting entirely of dry points.
@@
 	\item Determine interface of $\wet(x,\NN)$. 
-	\item If the interface is empty, simply output $\NN$. Otherwise, denote the contours by $\phi_1, \phi_2, \ldots, \phi_k$ and set $\phi'_i = \lift(\phi_i)$.
+	\item If the interface is empty, simply output $\NN$. Otherwise, denote the connected components by $\phi_1, \phi_2, \ldots, \phi_k$ and set $\phi'_i = \lift(\phi_i)$.
 	\item Initialize $\NN_1 = \NN$.

### 2015-07-02 paint-contour-tree.tex
Minor edits

@@
 with exactly the same contour tree, but different orderings of (heights of) the critical points.
-Turning it around, we cannot deduce the full height ordering of critical points from the contour tree. Therefore, clearly
-sorting all critical points is computationally unnecessary for constructing the contour tree.
-In \Fig{sorting}, the contour tree basically consists of two balanced binary trees, one of the joins,
+Turning it around, we cannot deduce the full height ordering of critical points from the contour tree.
+Sorting all critical points is computationally unnecessary for constructing the contour tree.
+In \Fig{sorting}, the contour tree consists of two balanced binary trees, one of the joins,
 another of the splits. Again, it is not necessary to know the relative ordering between the mounds
@@
 information is necessary: on the left, the little valleys are higher than the big central valley,
-and this is reflected in the contour tree. More generally, leaf paths in the contour tree have points
+and this is reflected in the contour tree. Leaf paths in the contour tree have points
 in sorted order, but incomparable points in the tree are unconstrained.

### 2015-07-03 paint-contour-tree.tex
small changes

@@
 exploits numerous properties of monotone paths. Carr's thesis shows relationships between monotone paths
-in $\MM$ and in its contour tree $\cC(\MM)$.
+in $\MM$ and its contour tree $\cC(\MM)$.
 
@@
 \begin{figure}[h]\centering
-    \includegraphics[width=.27\linewidth]{figs/spikey1}%
+    \includegraphics[width=.26\linewidth]{figs/spikey1}%
     \hspace{.35in}
-    \includegraphics[width=.27\linewidth]{figs/spikey2}%
+    \includegraphics[width=.26\linewidth]{figs/spikey2}%
     \hspace{.4in}
-    \includegraphics[width=.22\linewidth]{figs/spikeyTree}
+    \includegraphics[width=.215\linewidth]{figs/spikeyTree}
     \caption{Two surfaces with different orderings of the maxima, but the same contour tree.}
@@
 
-\begin{definition} \label{def:equiv} Two contours $\psi$ and $\psi'$ are \emph{equivalent} 
+\begin{definition} \label{def:equiv} Two regular contours $\psi$ and $\psi'$ are \emph{equivalent} 
 if there exists an $f$-monotone path $p$ connecting a point in $\psi$ to $\psi'$,
@@
 \label{def:int-criticalJoin}
-The \emph{critical join tree} $\jc(\MM)$ is built on the set $\cV$ of all critical points. 
-The directed edge $(u,v)$ is present when $u$ is the smallest valued vertex in $\cV$ in a connected component of $\MM^+_v$
+The \emph{critical join tree} $\jc(\MM)$ is built on the set $\cV$ of all critical points.\footnote{%
+\FullVer{Later, in Definition~\ref{def:criticalJoin}, we define the critical join tree on a certain subset of $\cV$.}
+\SubmitVer{In the full version we define the critical join tree on a certain subset of $\cV$.}
+} The directed edge $(u,v)$ is present when $u$ is the smallest valued vertex in $\cV$ in a connected component of $\MM^+_v$
 and $v$ is connected to this component (in $\MM$). 
@@
 
-Abusing notation, for now we will just call this the join tree. (In Section~\ref{sec:extreme} we carefully handle the distinction 
+Abusing notation, for now we will just call this the join tree. (In \SubmitVer{the full version}\FullVer{Section~\ref{sec:extreme}} we carefully handle the distinction 
 between the join tree and the critical join tree.)
@@
 \begin{figure}[h]\centering
-    \includegraphics[width=.18\linewidth]{figs/tent2}%
-    \hspace{.9in}
-    \includegraphics[width=.18\linewidth]{figs/tent1}
+    \includegraphics[width=.172\linewidth]{figs/tent2}%
+    \hspace{1.2in}
+    \includegraphics[width=.172\linewidth]{figs/tent1}
     \caption{On left, downward rain spilling only (each shade of gray represents a piece created by each different spilling), producing a grid.  
@@
 In such a complex, there exists a non-ascending path from any point to this unique minimum.
-Analogously, we can define maximum dominant simplicial complexes,
+Analogously, we can define maximum dominant complexes,
 and both are collectively called extremum dominant.
@@
 of comparable vs incomparable vertices.
-We arrive at the primary challenge: how to sort only those critical points that are comparable, without
+We arrive at the main challenge: how to sort only the comparable critical points, without
 constructing the join tree? The join tree algorithm of~\cite{csa-cctad-00} is a typical event-based
@@
 
-%\FullVer{
+\FullVer{
 \section{Divide and conquer through contour surgery} \label{sec:surgery}
@@
 \end{remark}
+}
 

### 2015-07-05 paint-contour-tree.tex
handled more reviewer comments

@@
 
-We think of the dimension, $d$, as constant, and assume that $\MM$ is represented in a data structure that allows constant-time traversals of neighboring simplices
+We think of the dimension, $d$, as constant, and assume that $\MM$ is represented in a data structure that allows constant-time access to neighboring simplices
 in $\MM$ (e.g.~\cite{BoMa12}). (This is analogous to a doubly connected edge list, but for higher dimensions.)
@@
 
-This equivalence relation gives a set \emph{contour classes}. By continuity, a regular point is present on a single contour,
-and we can partition the regular contours into these classes. Every such class maps to intervals
+This equivalence relation gives a set \emph{contour classes}. 
+%By continuity, a regular point is present on a single contour,
+%and we can partition the regular contours into these classes. 
+Every such class maps to intervals
 of the form $(f(x_i),f(x_j))$, where $x_i, x_j$ are critical points. Such a class is said
@@
 We will cut $\MM$ into disjoint extremum dominant pieces, in linear time.
-The best way to think of our procedure is a meteorology analogy. Take an arbitrary maximum $x$,
+One way to think of our procedure is a meteorological analogy. Take an arbitrary maximum $x$,
 and imagine torrential rain at the maximum. The water flows down, wetting any point that has a non-ascending path 
@@
 
-The cost of heap operations depends on the sizes of the heaps, which keeps changing because
+The cost of heap operations depends on heap sizes, which keep changing because
 of the repeated merging. This causes the analysis to be technically challenging. There are situations where
@@
 of \emph{heavy path decompositions}, first used by Sleator and Tarjan for analyzing link/cut trees~\cite{st-dsdt-83}.
-The final analysis basically charges expensive heap operations to long paths in the decomposition.
+The final analysis charges expensive heap operations to long paths in the decomposition.
 
@@
 requires worst case $\Omega(\cost(P(T)))$ time. We prove that our algorithm makes $\Theta(\cost(P(T)))$ 
-comparisons on all these instances. In general, any algorithm requires $\Omega(C)$ time
-to solve the set of instances that our algorithm solves with $C$ comparisons.
+comparisons on all these instances. 
+%In general, any algorithm requires $\Omega(C)$ time
+%to solve the set of instances that our algorithm solves with $C$ comparisons.
 
@@
 we get $\prod_{p \in P(T)} |p|!$ terrains each with a distinct contour tree. 
-Our algorithm requires $\Theta(\cost(P(T)))$ comparisons on all these instances.
+%Our algorithm requires $\Theta(\cost(P(T)))$ comparisons on all these instances.
 
@@
 
-\Ben{Current color is never defined, but it means the the current rep(c).}
+%\Ben{Current color is never defined, but it means the the current rep(c).}
 

### 2015-07-08 paint-contour-tree.tex
typo corrected

@@
 In the interest of presentation, the definitions and theorem statements in this section will
-slightly differ from those in the main body. They may also differ from the original definitions proposed
+slightly differ from those in the main body\SubmitVer{ of the full version}. They may also differ from the original definitions proposed
 in earlier work.
@@
 
-Our main workhorse is the following technical lemma.
+Our main workhorse is the following technical lemma. In the following, the current color of an edge, $e$, is the value of $\rep(\chi(e))$, 
+where $\chi(e)$ is the color of $e$ from the initial painting.
 
@@
 
-So assume $w$ exists. Take the $(f(w)+\delta)$-contour $\phi$ that intersects $B_\eps(v)$
+So assume $w$ exists. Take the $(f(w)+\delta)$-contour $\phi$ that intersects $B_\eps(w)$
 and intersects some $c$-colored edge incident to $w$. Note that all edges intersecting $\phi$ are also colored $c$,

### 2015-11-25 paint-contour-tree.tex
major structural changes

@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%\def \STOC {}
-
-\ifx\STOC\undefined
-  \newcommand{\FullVer}[1]{#1}
-  \newcommand{\SubmitVer}[1]{}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%
+\def\SoCGVer{TRUE}%
+%%%
+%\def\HideComments{TRUE}%
+%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+%\def\ProofInAppendix{TRUE}
+
+\ifx\SoCGVer\undefined
+   \newcommand{\InSoCGVer}[1]{}%
+   \newcommand{\InNotSoCGVer}[1]{#1}%
 \else
-  \newcommand{\FullVer}[1]{}
-  \newcommand{\SubmitVer}[1]{#1}
+%   \def\ProofInAppendix{TRUE}
+   \newcommand{\InSoCGVer}[1]{#1}%
+   \newcommand{\InNotSoCGVer}[1]{}%
 \fi
 
+\newcommand{\CiteFullVer}{\cite{rs-mvrms-14}}%
+
+% \InNotSoCGVer{%
+%    \newcommand{\SeeFullVer}{}%
+%    \newcommand{\LemmaProofInFullVer}[1]%
+%    {%
+%       \begin{lemma}
+%           #1
+%       \end{lemma}%
+%    }
+% }
+% 
+% \InSoCGVer{%
+%    \newcommand{\SeeFullVer}{{{\CiteFullVer}}}%
+%    \newcommand{\LemmaProofInFullVer}[1]%
+%    {%
+%       \begin{lemma}[For proof see  \CiteFullVer{}]
+%           #1
+%       \end{lemma}
+%    }   
+% }
+
+\InSoCGVer{%
+   \documentclass[a4paper,UKenglish]{lipics}
+%   \documentclass[11pt]{article}%
+
+%  \usepackage{proofapnd}
+}
+\InNotSoCGVer{%
   \documentclass[11pt]{article}
   \usepackage[margin=2.54cm]{geometry}
-\usepackage[dvipsnames,usenames]{color}
+}
+
+\InNotSoCGVer{
+   \newcommand{\myparagraph}[1]{\paragraph{#1}}
+}
+\InSoCGVer{
+   \newcommand{\myparagraph}[1]{\smallskip\noindent{\textbf{#1}}}
+}
+
 \usepackage{amsfonts,amsmath,amssymb,amsthm,mathtools}
@@
 \usepackage{framed}
-
-%\usepackage{mathabx}
-\usepackage[pagebackref,letterpaper=true,colorlinks=true,pdfpagemode=none,urlcolor=blue,linkcolor=blue,citecolor=BrickRed,pdfstartview=FitH]{hyperref}
-\usepackage{xspace,prettyref}
+\usepackage{xspace}
+\usepackage{prettyref}
+%\usepackage[dvipsnames,usenames]{color}
 \usepackage{color}
 \usepackage{graphics}
+%\usepackage[pagebackref,colorlinks=true,urlcolor=blue,linkcolor=blue,citecolor=BrickRed,pdfstartview=FitH]{hyperref}
+\usepackage{hyperref}%
+\hypersetup{%
+   breaklinks,%
+   colorlinks=true,%
+   urlcolor=[rgb]{0.2,0.0,0.0},
+   linkcolor=[rgb]{0.5,0.0,0.0},%
+   citecolor=[rgb]{0,0,0.445},
+   filecolor=[rgb]{0,0,0.4},
+   anchorcolor=[rgb]{0.0,0.1,0.2}
+}
+
 %\usepackage{MnSymbol}
@@
 
+\InNotSoCGVer{
  \newtheorem{theorem}{Theorem}[section]
+}
 \newtheorem{lemmata}[theorem]{Lemmata}
-\newtheorem{lemma}[theorem]{Lemma}
 \newtheorem{claim}[theorem]{Claim}
@@
 \newtheorem{proposition}[theorem]{Proposition}
+\InNotSoCGVer{
+ \newtheorem{lemma}[theorem]{Lemma}
  \newtheorem{corollary}[theorem]{Corollary}
+ \newtheorem{example}[theorem]{Example}
+ \newtheorem{definition}[theorem]{Definition}
+}
 \newtheorem{fact}[theorem]{Fact}
@@
 \newtheorem{question}[theorem]{Question}
-\newtheorem{example}[theorem]{Example}
-\newtheorem{definition}[theorem]{Definition}
 \theoremstyle{definition}
+\InNotSoCGVer{ 
  \newtheorem{remark}[theorem]{Remark} 
+}
 \newtheorem{observation}[theorem]{Observation}
@@
 \newcommand{\Merge}{\mathsf{Merge}}
-\newcommand{\stack}{\mathsf{S}}
+%\newcommand{\stack}{\mathsf{S}}
 \newcommand{\AQ}{\mathsf{AQ}}
@@
 \newcommand{\rep}{rep}
-\newcommand{\st}{K}
+\newcommand{\stack}{K}
 \newcommand{\surgery}{{\tt surgery}}
@@
 \newcommand{\Init}{\Algorithm{Init}\xspace}
+
+\newcommand{\pathTree}{P_{\mathcal{S}}}
+\newcommand{\pathTreeA}{\mathcal{P}_{\mathcal{S}}}
 
@@
    }
+\ifx\HideComments\undefined%
+\else%
+\renewcommand{\XSays}[2]{}%
+\fi%
+
 \newcommand{\Ben}[1]{{\XSays{Ben}{#1}}}
@@
 
+\begin{document}
+
+\InNotSoCGVer{
 \author{
-  Benjamin Raichel\BenThanks{}
+  Benjamin Raichel%\BenThanks{}
   \and
-  C. Seshadhri\thanks{University of California, Santa Cruz; {\tt scomandu@ucsc.edu}}
+  C. Seshadhri%\thanks{University of California, Santa Cruz; {\tt scomandu@ucsc.edu}}
 }
@@
 \date{}
-
-\begin{document}
+}
+
+\InSoCGVer{
+% Author macros::begin %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\title{%A Mountaintop View Requires Minimal Sorting:
+Avoiding the Global Sort: 
+\break A Faster Contour Tree Algorithm%
+\footnote{The full updated version of this paper is available on the arXiv \cite{rs-mvrms-14}}%
+}
+\titlerunning{Avoiding the Global Sort: A Faster Contour Tree Algorithm} %optional, in case that the title is too long; the running title should fit into the top page column
+
+\author[1]{Benjamin Raichel}
+\author[2]{C. Seshadhri}
+  \affil[1]{%
+      Department of Computer Science, %
+      University of Texas at Dallas\\ %
+      Richardson, TX, 75080, USA\\ %
+      \texttt{\si{benjamin.raichel}@\si{utdallas}.\si{edu}} %
+   }%
+   \affil[2]{%
+      Department of Computer Science, %
+      University of California, Santa Cruz\\ %
+      Santa Cruz, CA, 95064, USA\\ %
+      \texttt{\si{scomandu}@\si{ucsc}.\si{edu}} %
+   }%
+
+\authorrunning{B. Raichel and C. Seshadhri} %mandatory. First: Use abbreviated first/middle names. Second (only in severe cases): Use first author plus 'et. al.'
+
+\Copyright{Benjamin Raichel and Seshadhri Comandur}%mandatory, please use full first names. LIPIcs license is "CC-BY";  http://creativecommons.org/licenses/by/3.0/
+
+\subjclass{F.2.2, I.1.2, I.3.5}% mandatory: Please choose ACM 1998 classifications from http://www.acm.org/about/class/ccs98-html . E.g., cite as "F.1.1 Models of Computation". 
+\keywords{contour trees, computational topology, computational geometry}% mandatory: Please provide 1-5 keywords
+% Author macros::end %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+% %Editor-only macros:: begin (do not touch as author)%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% \serieslogo{}%please provide filename (without suffix)
+% \volumeinfo%(easychair interface)
+%   {Billy Editor and Bill Editors}% editors
+%   {2}% number of editors: 1, 2, ....
+%   {Conference title on which this volume is based on}% event
+%   {1}% volume
+%   {1}% issue
+%   {1}% starting page number
+% \EventShortName{}
+% \DOI{10.4230/LIPIcs.xxx.yyy.p}% to be completed by the volume editor
+% % Editor-only macros::end %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+}
 
 \maketitle
+\InNotSoCGVer{
 \thispagestyle{empty}
+}
 
@@
 
+\InNotSoCGVer{
 \newpage
-
 \pagenumbering{arabic}
+}
 
@@
 split into new components, contours acquire or lose handles. More generally, these are the events where 
-the topology of the level set changes. The contour tree is a concise representation of all these events.
+the topology of the level set changes. The contour tree is a concise representation of all these events.\footnote{%
+Throughout we follow the definition of contour trees from \cite{kobps-ctsssit-97} which includes all changes in topology.  
+For $d>2$, some subsequent works, such as \cite{csa-cctad-00}, only include changes in the number of components.
+}
+\Ben{See added footnote}
 
@@
 Suppose we have $f: \MM \to \RR$, where $\MM$ is a triangulated mesh with $n$ vertices, $N$ faces in total, and $t \leq n$ critical points. 
-It is standard to assume that $f$ is PL-Morse (a well-behaved input assumption) and that the maximum degree in the contour tree
-is $3$. A fundamental result in this area is the algorithm of Carr, Snoeyink, and Axen to compute contour trees,
+\Ben{It is standard to assume that $f$ has distinct values on the vertices of $\MM$ and the value at interior points is linearly interpolated.}
+%It is standard to assume that $f$ is PL-Morse (a well-behaved input assumption) and that the maximum degree in the contour tree is $3$. 
+A fundamental result in this area is the algorithm of Carr, Snoeyink, and Axen to compute contour trees,
 which runs in $O(n\log n + N\alpha(N))$ time~\cite{csa-cctad-00} (where $\alpha(\cdot)$ denotes the inverse
@@
 \label{thm:main-alg} 
-Consider a PL-Morse\footnote{This is a standard ``well-behaved" input assumption that implies
-that all critical points have distinct values, and the contour tree has max-degree $3$. Our algorithm
-does not need this, but it simplifies the proof and presentation.} $f:\MM \to \RR$, where the contour tree $T$
-has maximum degree $3$. There is a deterministic algorithm to compute $T$ whose running time is $O(\sum_{p \in P(T)} |p|\log |p| + t\alpha(t) + N)$,
+%Consider a PL-Morse\footnote{This is a standard ``well-behaved" input assumption that implies that all critical points have distinct values. 
+%Our algorithm does not need this, but it simplifies the proof and presentation.} 
+\Ben{Let $f:\MM \to \RR$ be the linear interpolant over distinct valued vertices, where the contour tree $T$
+has maximum degree $3$}.\footnote{The degree $\leq 3$ assumption helps simplify the proof and presentation, and is often made, as in \cite{kobps-ctsssit-97}.}\Ben{See footnote} 
+There is a deterministic algorithm to compute $T$ whose running time is $O(\sum_{p \in P(T)} |p|\log |p| + t\alpha(t) + N)$,
 where $P(T)$ is a specific path decomposition (constructed implicitly by the algorithm).
@@
 
-\paragraph{Beyond worst-case analysis.} Our result provides understanding of the complexity
+\myparagraph{Beyond worst-case analysis.} Our result provides understanding of the complexity
 of contour tree construction beyond the standard worst-case analysis. Our theorem and lower bound 
@@
 are standard for results in this area, though there is some variability in notation.
-The input is a continuous piecewise-linear (PL) function $f:\MM \to \RR$, where $\MM$ is a simply connected and fully triangulated simplicial complex in $\RR^d$,
+The input is a continuous piecewise-linear function $f:\MM \to \RR$, where $\MM$ is a simply connected and fully triangulated simplicial complex in $\RR^d$,
 except for specially designated \emph{boundary facets}. So $f$ is defined only on the vertices of $\MM$,
@@
 has a well-defined height, including the infinite facet, which is required to be a boundary facet.  However, 
-we allow for other boundary facets, to capture the resulting surface pieces after our algorithm make a horizontal cut.
+we allow for other boundary facets, to capture the resulting surface pieces after our algorithm makes a horizontal cut.
 
@@
 \begin{definition} \label{def:level} The \emph{level set} at value $h$ is the set $\{x| f(x) = h\}$.
-A \emph{contour} is a maximal connected component of a level set. An \emph{$h$-contour} is a contour where $f$-values are $h$.
+A \emph{contour} is a connected component of a level set. An \emph{$h$-contour} is a contour where $f$-values are $h$.
 \end{definition}
@@
 Since this tree has height values associated with the vertices, we can talk about up-degrees and down-degrees in $\reeb(\MM)$.
-We assume there are no ``multi-saddles", so up and down-degrees are at most $2$, and the total degree is at most $3$. 
-This is again a standard assumption in topological algorithms and can be achieved by vertex unfolding (Section 6.3 in~\cite{HaEd10}).
-
-\subsection{Some technical remarks}
+%
+\Ben{Similar to \cite{kobps-ctsssit-97} (among others), multi-saddles are treated as a set of ordinary saddles, which can be realized 
+via vertex unfolding (which can increase surface complexity if multi-saddle degrees are allowed to be super-constant).
+Therefore, to simplify the presentation, for the remainder of the paper up and down-degrees are at most $2$, and the total degree is at most $3$.}
+%
+%We assume there are no ``multi-saddles", so up and down-degrees are at most $2$, and the total degree is at most $3$. 
+%This is again a standard assumption in topological algorithms and can be achieved by vertex unfolding (Section 6.3 in~\cite{HaEd10}).
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\InSoCGVer{See Appendix~\ref{sec:techmarks} for further technical remarks on the above definitions.}
+\newcommand{\technicalRemarks}{
+%\subsection{Some technical remarks}
 Note that if one intersects $\MM$ with a given ball $B$, then a single contour in $\MM$ might be split into more than one contour in the intersection. 
@@
 set of vertices, though we will not delve into these aspects in this paper.
-
+}
+\InNotSoCGVer{
+\subsection{Some technical remarks}
+\technicalRemarks
+}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\newcommand{\intuitionSection}{
 \section{A tour of the new contour tree algorithm} \label{sec:approach}
+\InNotSoCGVer{
 Our final algorithm is quite technical and has numerous moving parts. However, for the $d=2$ case, 
 where the input is just a triangulated terrain, the main ideas of the parts of the algorithm 
-can be explained clearly.  
-\SubmitVer{Therefore, in the remaining few allotted pages, 
-rather than providing an incomplete subset of the technical content, we provide a high level view of the entire result, 
-which we view as essential for understanding our approach. The technical details can be found in the full version of the paper \cite{rs-mvrms-14}.}
-\FullVer{Therefore, here we first provide a high level view of the entire result, 
-which we view as essential for understanding our approach. }
+can be explained clearly.  Therefore, here we first provide a high level view of the entire result.
+}
+\InSoCGVer{
+For the $d=2$ case, where the input is just a triangulated terrain, the main ideas of the parts of the algorithm 
+can be explained clearly.  Therefore, here we provide a high level view of the entire result.
+}
 
 In the interest of presentation, the definitions and theorem statements in this section will
-slightly differ from those in the main body\SubmitVer{ of the full version}. They may also differ from the original definitions proposed
+slightly differ from those in the main body. They may also differ from the original definitions proposed
 in earlier work.
 
-% Our final algorithm is quite technical and has numerous moving parts. This section
-% provides a high level view of the entire result, and highlights the main ideas
-% and intuition. 
-% \SubmitVer{The full version with all the details can be found in the appendix.}
-% It is helpful to keep the $d=2$ case in mind, so the input is just a triangulated
-% terrain. In the interest of presentation, the definitions and theorem statements in this section will
-% slightly differ from those in the main body. They may also differ from the original definitions proposed
-% in earlier work.
-
-\paragraph{Do not globally sort:} Arguably, the starting point for this work is \Fig{relative}. We have two terrains
+\myparagraph{Do not globally sort:} The starting point for this work is \Fig{relative}. We have two terrains
 with exactly the same contour tree, but different orderings of (heights of) the critical points.
@@
 \subsection{Breaking $\MM$ into simpler pieces} \label{sec:break}
-
 Let us begin with the algorithm of Carr, Snoeyink, and Axen~\cite{csa-cctad-00}. The key insight is to build
@@
 The \emph{critical join tree} $\jc(\MM)$ is built on the set $\cV$ of all critical points.\footnote{%
-\FullVer{Later, in Definition~\ref{def:criticalJoin}, we define the critical join tree on a certain subset of $\cV$.}
-\SubmitVer{In the full version we define the critical join tree on a certain subset of $\cV$.}
-} The directed edge $(u,v)$ is present when $u$ is the smallest valued vertex in $\cV$ in a connected component of $\MM^+_v$
-and $v$ is connected to this component (in $\MM$). 
+In Definition~\ref{def:criticalJoin}, we define the critical join tree on a certain subset of $\cV$.}
+The directed edge $(u,v)$ is present when $u$ is the smallest valued vertex in $\cV$ in a connected component of $\MM^+_v$
+and $v$ is adjacent to a vertex in this component (in $\MM$). 
 \end{definition}
 
-Abusing notation, for now we will just call this the join tree. (In \SubmitVer{the full version}\FullVer{Section~\ref{sec:extreme}} we carefully handle the distinction 
+Abusing notation, for now we will just call this the join tree. (In \Sec{extreme} we carefully handle the distinction 
 between the join tree and the critical join tree.)
@@
 
-\paragraph{Contour surgery:} We first need a divide-and-conquer lemma. Any contour $\phi$ can be associated
+\myparagraph{Contour surgery:} We first need a divide-and-conquer lemma. Any contour $\phi$ can be associated
 with an edge $e$ of the contour tree. Suppose we ``cut" $\MM$ along this contour. We prove that $\MM$
@@
     \caption{On left, downward rain spilling only (each shade of gray represents a piece created by each different spilling), producing a grid.  
-    Note we are assuming raining was done in reverse sorted order of maxima.  
+    Note we are assuming raining was done in sorted order of the maxima (i.e.\ lowest to highest).  
     On right, flipping the direction of rain spilling.}
@@
 
-\paragraph{Cutting $\MM$ into extremum dominant pieces:} We define a simplicial complex
+\myparagraph{Cutting $\MM$ into extremum dominant pieces:} We define a simplicial complex
 endowed with a height to be \emph{minimum dominant} if there exists only a single minimum.
@@
 We can prove that an original face of $\MM$ is cut at most once, so the final complexity can be bounded.
-In \Fig{order}, this procedure would yield two pieces, one maximum dominant, and one minimum dominant.
+In \Fig{order}, regardless of the choice of the starting maximum,  this procedure would yield (at most) two pieces, 
+one maximum dominant, and one minimum dominant.
 
@@
 
-\paragraph{Extremum dominance simplifies contour trees:} We will focus on minimum dominant simplicial complexes $\MM$. By \Thm{int-rain}, it suffices
+\myparagraph{Extremum dominance simplifies contour trees:} We will focus on minimum dominant simplicial complexes $\MM$. By \Thm{int-rain}, it suffices
 to design an algorithm for contour trees on such inputs. 
@@
 except at a unique boundary face (think of a large boundary triangle that is the boundary).
-All the valleys in such a terrain are necessarily joins, and there can be no splits. Look
-at \Fig{sorting}. The portion on the left is minimum dominant in exactly this fashion.
+All the saddles in such a terrain are necessarily joins, and there can be no splits.
+In \Fig{sorting}, the portion on the left is minimum dominant in exactly this fashion, albeit in one dimension lower.
 %Water poured at any critical point must flow to the boundary, and the entire wet portion is connected.
@@
 Arguably, everything up to this point is a preamble for the main result: a faster
-algorithm for join trees. Our algorithm does not require the input to be extremum
-dominant. This is only required to relate the join trees to the contour tree of the initial input $\MM$.
-For convenience, we use $\NN$ to denote the input here. Note that in \Def{int-criticalJoin},
+algorithm for join trees. Our algorithm does not require the initial input to be extremum
+dominant. This is only required to relate the join trees, of the resulting subcomplexes of Theorem~\ref{thm:int-rain}, to the contour tree of the initial input $\MM$.
+For clarity, we use $\NN$ to denote the input here. Note that in \Def{int-criticalJoin},
 the join tree is defined purely combinatorially in terms of the 1-skeleton (the underlying graph) of $\NN$.
@@
 
-\begin{definition} \label{def:paint} Let the 1-skeleton of $\NN$ have edge set $E$ and maxima $X$.
+\begin{definition} \label{def:paint1} \InSoCGVer{(Restatement of \Def{paint2}.)} Let the 1-skeleton of $\NN$ have edge set $E$ and maxima $X$.
 A  \emph{painting} of $\NN$ is a map $\chi:X \cup E \to [|X|]$ with the following property. 
@@
 
-\paragraph{The running time analysis:} Relating the running time to a path decomposition is the most technical part of the paper. 
+\myparagraph{The running time analysis:} Relating the running time to a path decomposition is the most technical part of the paper. 
 All the non-heap operations can be easily bounded by $O(t\alpha(t) + N)$ (the $t\alpha(t)$ is from the union-find data structure
@@
 %Our algorithm requires $\Theta(\cost(P(T)))$ comparisons on all these instances.
-
-\FullVer{
+}
+
+\InSoCGVer{
+\subsection{How to Read this paper}
+\Ben{New section:}
+The final algorithm is quite technical and has numerous moving parts. However, for the $d=2$ case, 
+where the input is just a triangulated terrain, the main ideas of the parts of the algorithm 
+can be explained clearly. Therefore, in Appendix~\ref{sec:approach} we provide a high level view of the entire result.
+We strongly encourage the reader to overview this section, as then the remainder of the paper can be understood 
+as a simple matter of implementing these ideas.
+
+Alternatively, the reader can view the full version of the paper \CiteFullVer{} which contains this high level overview and all other appendices 
+in the main body of the text.
+
+\myparagraph{Outline of the algorithm and the following sections.} 
+The high level idea is to partition $\MM$ into nicely behaved pieces, on which we know how to compute the contour tree optimally. 
+\Sec{surgery} describes our cutting procedure, and \Sec{rain} describes how to use this cutting procedure.
+Roughly speaking the resulting pieces are ``nice'' because the contour tree of each piece 
+looks like a rooted tree, i.e.\ the join tree of \cite{csa-cctad-00}, and the precise relation to join trees is the subject of \Sec{extreme}.
+\Sec{paint} then describes how to optimally compute the join tree (and hence the contour tree) 
+on each piece (see Appendix~\ref{sec:approach} for discussion on why the join tree should not be computed directly on the input complex). 
+\Ben{Mention lower bound an analysis sections which are going to be moved to appendix.}
+
+}
+
+\InNotSoCGVer{
+\intuitionSection
+}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
 \section{Divide and conquer through contour surgery} \label{sec:surgery}
@@
 The faces are then triangulated to ensure that they are all simplices. 
+\Ben{Reviewer wanted clarification on how triangulation is done. Maybe can use bottom-vertex triangulation of Clarkson 88, discussed in Matousek book?}
 This creates the two new boundaries $\phi^+$ and $\phi^-$, and we maintain the property of constant $f$-value at a boundary.
@@
 We now describe a high-level approach to construct $\reeb(\MM)$ using this cutting procedure.
+
+\Ben{Surgery should be rewritten to take instead take as input the output of cut, i.e. two complexes and a contour.  
+This is how it is eventually used in Claim\ref{clm:rain-reeb}. Currently it does nothing really.}
 
@@
 \medskip
+
+\InSoCGVer{The following theorems are intuitively obvious, and are proven in Appendix~\ref{sec:proofOfSurgery}.}
+
 \begin{theorem} \label{thm:surgery} For any regular contour $\phi$, the output of $\surgery(\MM,\phi)$ is $\reeb(\MM)$.
@@
 
-To prove the above theorem, we require some theorems from \cite{c-tmi-04} (Theorems 6.6 and 6.7) that map paths in $\cC(\MM)$ to $\MM$.
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\newcommand{\surgeryBody}{
+%To prove the above theorem, we require some theorems from \cite{c-tmi-04} (Theorems 6.6 and 6.7) that map paths in $\cC(\MM)$ to $\MM$.
+To prove Theorem~\ref{thm:surgery}, we require a theorem from \cite{c-tmi-04} (Theorems 6.6) that map paths in $\cC(\MM)$ to $\MM$.
 
@@
 in contours involving $Q$.
+
+In particular, for every monotone path $P$ in $\MM$, there exists a monotone path $Q$ in the contour tree to which $P$
+maps, and vice versa.
 \end{theorem}
 
-\begin{theorem} \label{thm:carr-mono} For every monotone path $P$ in $\MM$, there exists a monotone path $Q$ in the contour tree to which $P$
-maps (as in the previous theorem), and vice versa.
-\end{theorem}
+% \begin{theorem} \label{thm:carr-mono} For every monotone path $P$ in $\MM$, there exists a monotone path $Q$ in the contour tree to which $P$
+% maps (as in the previous theorem), and vice versa.
+% \end{theorem}
 
@@
 \begin{lemma} \label{lem:cut} Consider a regular contour $\phi$ contained in a contour class (of an edge of $\reeb(\MM))$
-$(u,v)$ and denote $\MM' = \cut(\MM,\phi)$. Then $\cV(\reeb(\MM')) = \{\phi^+,\phi^-\} \cup \cV(\MM)$
+$(u,v)$ and let $\MM' = \cut(\MM,\phi)$. Then $\cV(\reeb(\MM')) = \{\phi^+,\phi^-\} \cup \cV(\MM)$
 and $\cE(\reeb(\MM')) = \{(u,\phi^+),(\phi^-,v)\} \cup (\cE(\MM) \setminus (u,v))$.
@@
 two disconnected complexes.
+}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\InNotSoCGVer{
+\surgeryBody
+}
 
@@
 
+\newcommand{\proofofDisconnected}{
 \begin{proof} Denote (as in \Lem{cut}) the edge containing $\phi$ to be $(u,v)$. Suppose for contradiction that there is a path between vertices $u$ and $v$
@@
 \end{proof}
+}
+\InNotSoCGVer{\proofofDisconnected}
+
+% 
+% \begin{lemma} \label{lem:blahblah}
+%  Testing 123 \RefProofInAppendix{blahblah}
+% \end{lemma}
+% 
+% \begin{proof:in:appendix:e}{\ref{lem:blahblah}}{blahblah}
+%  blah balh
+% \end{proof:in:appendix:e}
 
@@
 
-The first aspect of the partitioning is ``raining". Start at some point $x \in \MM$ and imagine rain at $x$.
-The water will flow downwards along non-ascending paths and ``wet" all the points encountered. Note that this procedure considers all points
+The first aspect of the partitioning is ``raining''. Start at some point $x \in \MM$ and imagine rain at $x$.
+The water will flow downwards along non-ascending paths and ``wet'' all the points encountered. Note that this procedure considers all points
 of the manifold, not just vertices.
@@
 is denoted by $\wet(x,\MM)$ (which in turn is represented as a simplicial complex). A point $z$ is at the \emph{interface} of $\wet(x,\MM)$ if every neighborhood of $z$
-has non-trivial intersection with $\wet(x,\MM)$.
+has non-trivial intersection with $\wet(x,\MM)$ \Ben{(i.e.\ the intersection is neither empty nor the entire neighborhood)}.
 \end{definition}
@@
 
-\begin{claim} \label{clm:inter} For any $x$, each connected component of the interface of $\wet(x,\MM)$ contains a join vertex.
+\begin{claim} \label{clm:inter} For any $x$, each component of the interface of $\wet(x,\MM)$ contains a join vertex.
 \end{claim}
@@
 We perform a non-ascending search from $x$. Any face $F$ of $\MM$ encountered is partially (if not entirely) in $\wet(x,\MM)$.
-The wet portion is determined by cutting $F$ along the interface. Since each component of the interface connected subset of a contour, this is equivalent
+The wet portion is determined by cutting $F$ along the interface. Since each component of the interface is a contour, this is equivalent
 to locally cutting $F$ by a hyperplane. All these operations can be performed to output $\wet(x,\MM)$ in time linear in $|\wet(x,\MM)|$.
@@
 of the procedure is that the rain alternately flows downwards and upwards, 
-since otherwise faces may be cut a super constant number of times (see \Fig{order}).
+since otherwise faces may be cut a super constant number of times (see \Fig{order}\InSoCGVer{ in Appendix~\ref{sec:approach}}).
 In other words, when
@@
 \end{lemma}
-\begin{proof} All notation here follows that in the pseudocode of $\rain$.
+\begin{proof} Notation here follows the pseudocode of $\rain$.
 First, by \Thm{jordan}, all the pieces on which $\rain$ is invoked are disjoint.
@@
 and one $\eps$ below $\phi'_i$.  This breaks $F$ into lower and upper portions which are then triangulated
-(there is also a discarded middle portion).  The lower portion, which is adjacent $\phi_i$, gets included 
+(there is also a discarded middle portion).  The lower portion, which is adjacent to $\phi_i$, gets included 
 in $\NN_{k+1}$, the complex containing the wet points, and hence does not participate in any later recursive calls.  
@@
 
-The main theorem of this section asserts that the contours trees of minimum dominant manifolds have a simple description.
+\Ben{Read remainder of this section carefully, it has been cut up, and parts rewritten to fit.}
+
+\InNotSoCGVer{
+The main theorem of this section asserts that contour trees of minimum dominant manifolds have a simple description.
 The exact statement will require some definitions and notation.
@@
 Conventionally, all edges are directed from higher to lower function value. 
-
-\begin{definition} \label{join} The \emph{join tree} $\cJ(\MM)$ of $\MM$ is built on vertex set $V(\MM)$.
+}
+\InSoCGVer{
+The main theorem of this section asserts that contour trees of minimum dominant manifolds have a simple description. 
+Intuitively, the cutting procedure of \Sec{rain} introduced ``stumps'' where we made cuts, which 
+is why we allowed for non-dominant minima and maxima in the definition of extremum dominant manifolds.  
+The punchline is that, ignoring these stumps, the contour tree of an extremum dominant manifold is equivalent to its \emph{join} (or \emph{split}) tree, defined in~\cite{csa-cctad-00}.
+Hence the critical join tree below is just the join tree without these stumps.  
+Precise definitions and proofs are given Appendix~\ref{sec:jstedm}.
+}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\newcommand{\joinTreeSplitTreeMerge}{
+\begin{definition} \label{join} The \emph{join tree} $\cJ(\MM)$ of $\MM$ is built on vertex set $\cV(\MM)$.
 The directed edge $(u,v)$ is present when $u$ is the smallest valued vertex in a connected component of $\MM^+_v$
-\emph{and} $v$ is connected to this component (in $\MM$). The \emph{split tree} $\cS(\MM)$ is obtained
-by looking at $\MM^-_v$ (or alternately, by taking the join tree of the inversion of $\MM$). 
+\emph{and} $v$ is adjacent to a vertex in this component (in $\MM$). The \emph{split tree} $\cS(\MM)$ is obtained
+by looking at $\MM^-_v$ (or alternatively, by taking the join tree of the inversion of $\MM$). 
 \end{definition}
@@
 merging procedure from~\cite{csa-cctad-00} that constructs the contour tree from the join and split trees. 
-It actually constructs the \emph{augmented contour tree} $\cA(\MM)$, which
+It actually constructs the \emph{augmented contour tree} $\cA(\MM)$, which is 
 obtained by replacing each edge in the contour tree with a path of all regular vertices 
@@
 \medskip
+}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\InNotSoCGVer{\joinTreeSplitTreeMerge}
 
@@
 The directed edge $(u,v)$ is present when $u$ is the smallest valued vertex in $V'$ in a connected component of $\MM^+_v$
-and $v$ is connected to this component (in $\MM$). 
+and $v$ \Ben{is adjacent to a vertex} in this component (in $\MM$). 
 \end{definition}
+
+\InSoCGVer{Each non-dominant minimum, $m_i$, connects to the contour tree at a corresponding split $s_i$. We have the following (see Appendix~\ref{sec:jstedm} for more details).}
 
@@
 
+\newcommand{\proofofContourJoinEquiv}{
 \begin{proof} We first show that $\cA(\MM)$ is $\cJ(\MM) \ominus \{m_i\}$ with edges $\{(s_i,m_i)\}$.
@@
 \end{proof}
+}
+
+\InNotSoCGVer{\proofofContourJoinEquiv}
 
@@
 can of paint of a distinct color at each maximum $x$. We will spill different paint from each maximum and watch it flow down.
-This is analogous to the raining in the previous section, but paint is a much more viscous liquid.
+This is analogous to the raining of \Sec{rain}, but paint is a much more viscous liquid.
 \emph{So paint only flows down edges, and it does not color the interior of higher dimensional faces.} Furthermore, paints
@@
 
-We now restate \Def{paint}.
-
-\begin{definition} Let the 1-skeleton of $\MM$ have edge set $E$ and maxima $X$.
+\InNotSoCGVer{We now restate \Def{paint1}.}
+
+\begin{definition} \label{def:paint2} Let the 1-skeleton of $\MM$ have edge set $E$ and maxima $X$.
 A  \emph{painting} of $\MM$ is a map $\chi:X \cup E \to [|X|]$ with the following property. 
@@
 
-A painting only colors edges, not vertices. We associate certain
+A painting colors edges, not vertices. We associate certain
 sets of colors with vertices.
@@
 	 \item An \emph{up-star} of $v$ is the set of edges that all connected to a fixed component of $\MM^+_v$.
-	 \item A vertex $v$ is \emph{touched by color $c$} if $x$ is incident to a $c$-colored
+	 \item A vertex $v$ is \emph{touched by color $c$} if $v$ is incident to a $c$-colored
 	 edge with $v$ at the lower endpoint. For $v$, $\col(v)$ is the set of colors that touch $v$.
-	 \item A color $c \in \col(v)$ \emph{fully touches} $v$ if all edges in an up-star are colored $v$.
+	 \item A color $c \in \col(v)$ \emph{fully touches} $v$ if all edges in an up-star are colored $c$.
 	 \item For any maximum $x\in X$, we say that $x$ is both touched and fully touched by $\chi(x)$.
@@
 
-Each non-maximum $v$ participates in at least $1$ and at most $2$ up-stars (the latter iff
-$v$ is a join).
+Non-maximum $v$ participate in at least $1$ and at most $2$ up-stars (the latter iff
+$v$ a join).
 
@@
 \noindent
-{\bf The stack $\st$:} This consists of non-extremal critical points, with monotonically increasing
+{\bf The stack $\stack$:} This consists of non-extremal critical points, with monotonically increasing
 heights as we go from the base to the head.
 \ignore{
-Each point $x \in \st$ has an associated subset of $\col(x)$, denoted $\mcol(x)$.
+Each point $x \in \stack$ has an associated subset of $\col(x)$, denoted $\mcol(x)$.
 Both $\mcol(x)$ and its complement are stored as hash table. So lookups, inserts, and deletes
@@
 \begin{asparaitem}
-	\item For every $x \in \st$: For every $c \in \mcol(x)$, $x$ is the highest element
+	\item For every $x \in \stack$: For every $c \in \mcol(x)$, $x$ is the highest element
 	in $T(c)$. Furthermore, $c = \rep(c)$.
-	\item Consider $x, y \in \st$ such that $y$ was pushed on $x$. There exists $c \in \col(x) \setminus
+	\item Consider $x, y \in \stack$ such that $y$ was pushed on $x$. There exists $c \in \col(x) \setminus
 	\mcol(x)$ such that $x$ is not highest in $T(c)$ but $y$ is highest in $T(c)$.
@@
 	\begin{compactenum}
-		\item Run $\update(\st)$. Pop $\st$ to get $h$.
+		\item Run $\update(\stack)$. Pop $\stack$ to get $h$.
 		\item Let $\cur(h) = \{\rep(c) | c \in \col(h)\}$.
@@
 \begin{minipage}{0.9\textwidth}
-{\bf $\update(\st)$}
+{\bf $\update(\stack)$}
 
@@
 	\item If $K$ is empty, push arbitrary unprocessed critical point $v$.
-	\item Let $h$ be head of $\st$.
+	\item Let $h$ be head of $\stack$.
 	\item While $h$ is not ripe:
@@
 		\item Find $c \in \col(h)$ such that $h$ is not the highest in $T(\rep(c))$.
-		\item Push the highest of $T(\rep(c))$ onto $\st$, and update head $h$.
+		\item Push the highest of $T(\rep(c))$ onto $\stack$, and update head $h$.
 	\end{compactenum}
@@
 	\item After $v$ is processed, all edges incident to $v$ have the same color (more technically, the same representative). 
-%	\item When $\update(\st)$ terminates, the head $h$ is ripe. 
+%	\item When $\update(\stack)$ terminates, the head $h$ is ripe. 
 	\item Vertices on the stack are in increasing height order. 
@@
 % 
-% \begin{claim} \label{clm:stack} Vertices are always pushed onto $\st$ in increasing order of height.
+% \begin{claim} \label{clm:stack} Vertices are always pushed onto $\stack$ in increasing order of height.
 % \end{claim}
@@
 \label{sec:runTime}
+\Ben{This section needs to be reduced dramatically in size.}
 
@@
 
-\newcommand{\pathTree}{P_{\mathcal{S}}}
-\newcommand{\pathTreeA}{\mathcal{P}_{\mathcal{S}}}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\newcommand{\leafAssignPathDecomp}{
 \section{Leaf assignments and path decompositions}
@@
 \begin{theorem}
-Consider a PL-Morse $f:\MM \to \RR$, where the join tree $\jc(\MM)$ has maximum degree $3$. 
+%Consider a PL-Morse $f:\MM \to \RR$, where the join tree $\jc(\MM)$ has maximum degree $3$.
+Let $f:\MM \to \RR$ be the linear interpolant over distinct valued vertices, where the join tree $\jc(\MM)$ has maximum degree $3$.
 There is an algorithm to compute the join tree whose running time is $O(\sum_{p \in \pmax(\jc)} |p|\log |p| + t\alpha(t) + N)$.
@@
 \begin{theorem}
-Consider a PL-Morse $f:\MM \to \RR$, where the contour tree $\cC=\cC(\MM)$
-has maximum degree $3$. There is an algorithm to compute $\cC$ whose running time is $O(\sum_{p \in P(\cC)} |p|\log |p| + t\alpha(t) + N)$,
+%Consider a PL-Morse $f:\MM \to \RR$, where the contour tree $\cC=\cC(\MM)$ has maximum degree $3$. 
+Let $f:\MM \to \RR$ be the linear interpolant over distinct valued vertices, where the contour tree $\cC=\cC(\MM)$ has maximum degree $3$. 
+There is an algorithm to compute $\cC$ whose running time is $O(\sum_{p \in P(\cC)} |p|\log |p| + t\alpha(t) + N)$,
 where $P(T)$ is a specific path decomposition (constructed implicitly by the algorithm).
@@
 \end{proof}
-
+}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\InSoCGVer{
+\Ben{Summarize Appendix~\ref{sec:pathDecomp} on path decomposition upped bound here.}
+}
+
+\InNotSoCGVer{
+\leafAssignPathDecomp
+}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\newcommand{\lowerBoundbyPathDecomp}{
 \section{Lower Bound by Path Decomposition}
@@
 In this process, we glue the base of a scaled down ``child'' diamond on to a triangular cone face of the larger  ``parent'' diamond (see \Fig{diamond}).
- Specifically, the anchor of the child diamond is attached directly to a face of the parent diamond at some height $h$. The remaineder of the base of the child cone is then 
+ Specifically, the anchor of the child diamond is attached directly to a face of the parent diamond at some height $h$. The remainder of the base of the child cone is then 
  extended down (at a slight angle) until it hits the face of the parent.
@@
 
-%
-%Consider our algorithm on a fixed $(f,\MM)$.  First we run $\rain(x,\MM)$ to partition the input 
-%into extreme dominant complexes.  To make this partition uniquely defined lets say that the initial vertex $x$
-%that rain is called from is the global maximum in the input (note that given a specific $x$ the partition is uniquely defined).
-%Let $\MM_1, \dots, \MM_k$ be the set of produced subcomplexes, and let $f_i$ be the restriction of $f$ to $\MM_i$.
-%Now on each of the $(f_i,\MM_i)$ we run the $\paint$ procedure to compute the join tree of $(f_i, \MM_i)$.
-%As described in the previous section, this painting procedure defines a maximum path decomposition with 
-%and associated cost, $\cost_{\jc}(f_i) = \sum_{p\in P_i} |p|\log |p|$.  
-%It is not hard to see that the union of these paths decomposition $P= \cup_{i=1}^k P_i$ defines 
-%a rain consistent path decomposition of the contour tree for $(f,\MM)$.  
-%In particular for a fixed $\MM$, $f$ defines a specific path decomposition $P=P(f)$, which we call the \emph{rain consistent maximum path decomposition}, 
-%and we write $\cost(f) = \sum_{p\in P} |p|\log |p|$.
-
 The following is analogous to \Lem{pathDecomp}, and in particular uses it as a subroutine.
@@
 
-%
-%
-%Call a cost $C$ valid if $C = cost(f)$ for some $f$ defined over a given complex $\MM$.
-%For the same reason that \Thm{joinLB} followed from \Lem{pathDecomp} and \Lem{cost}, we now have the following. 
-%
-%\begin{theorem}
-%Consider the set of functions $\bF_C$ that all have the same valid cost $C$.  
-%Any height function decision tree that correctly computes
-%the contour tree on all of $\bF_C$, has an average case (uniform over $\bF_C$) running time of $\Omega(C)$.
-%\end{theorem}
-
 \begin{remark}
@@
 
-% \section{Temp}
-% Given a contour tree we allow any path decomp, as long as each path contains a maxima or minima (or both).
-% We call a path $p$ in the decomp a min (resp. max) path if it contains a minima (resp. maxima) (note a path can be both).
-% For every path we create a cylinder with $m$ faces for $p$, where $m$ is the number of non-leaf vertices in $p$ (either $|p|-1$ or $|p|-2$).  
-% If $p$ has a maxima we put a top cone cap on the cylinder, and if it contains a manima then put a bottom cone cap (i.e.\ the cap represents the maxima or minima).
-% 
-% Every edge $e$ in $T$, but not in a path decomp. tells us how to glue together cylinders.  Let $u$ be the upper endpoint of $e$ and $l$ the lower endpoint.
-% 
-% First we ``remove'' all edges $e$ such that $u$ is adjacent to a max path and $l$ adjacent to a min path by joining the free ends of the respective cylinder together.
-% Hence we can assume no such edges exist.  Now we have three cases:
-% 
-% \begin{enumerate}[{Case} 1]
-%  \item 
-% $l$ is adjacent to a min path:\newline
-% Note in this case the path containing $u$ must be a min path.
-% In this case connect the bottom of the cylinder for the path containing $l$ into a face of the cylinder for the path containing $u$.
-% \item
-% $l$ is adjacent to a max path and $u$ is adjacent to a max path:\newline
-% In this case connect the bottom of the cylinder for the path containing $u$ into a face of the cylinder for the path containing $l$.
-% \item
-% $l$ is adjacent to a max path and $u$ is adjacent to a min path:\newline
-% Connect a face of the cylinder for the $u$ path (using a split) to a face of the cylinder for the $l$ path (using a join).
-% \end{enumerate}
-%}
-
-\paragraph{Acknowledgements.}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\InSoCGVer{
+\Ben{summarize Appendix \ref{sec:lb}, the lower bound section, result here.}
+}
+\InNotSoCGVer{
+\lowerBoundbyPathDecomp
+}
+
+\myparagraph{Acknowledgements.}
 We thank Hsien-Chih Chang, Jeff Erickson, and Yusu Wang for numerous useful discussions. This work is
@@
 by Sandia Corporation, a wholly owned subsidiary of Lockheed
-Martin Corporation, for the U.S. Department of Energy’s National
+Martin Corporation, for the U.S. Department of Energy's National
 Nuclear Security Administration under contract DE-AC04-94AL85000.
 
+\InNotSoCGVer{
 \bibliographystyle{alpha}
+%\bibliographystyle{plain}
+}
+
+\InSoCGVer{
+\bibliographystyle{plain}
+}
+
 \bibliography{contour}
+
+\InSoCGVer{
+\appendix
+
+\intuitionSection
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\section{Some technical remarks}
+\label{sec:techmarks}
+The following is a continuation from the end of \Sec{basics}.
+
+\technicalRemarks
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\section{Proof of Surgery}
+\label{sec:proofOfSurgery}
+\surgeryBody
+That is, Theorem~\ref{thm:jordan}, which is proven below.
+\proofofDisconnected
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\section{Join and Split Trees of Extremum Dominant Manifolds}
+\label{sec:jstedm}
+In this section we prove Theorem~\ref{thm:contour-tree} (of \Sec{extreme}), which states the equivalence 
+of the contour tree on extremum dominant manifolds and the critical join tree of Definition~\ref{def:criticalJoin}.
+This proof requires some definitions and notation, specially
+the notions of \emph{join} and \emph{split} trees, as given by~\cite{csa-cctad-00}.
+Conventionally, all edges are directed from higher to lower function value.
+
+\joinTreeSplitTreeMerge
+
+% We now prove Theorem~\ref{thm:contour-tree} (of Section~\ref{sec:extreme}), which states the equivalence 
+% of the contour tree on extremum dominant manifolds and the critical join tree of Definition~\ref{def:criticalJoin}.
+We are now ready to prove Theorem~\ref{thm:contour-tree}.
+\proofofContourJoinEquiv
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\leafAssignPathDecomp
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\lowerBoundbyPathDecomp
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+%\InsertAppendixOfProofs
+}
 

### 2015-11-27 paint-contour-tree.tex
small changes

@@
 There is a deterministic algorithm to compute $T$ whose running time is $O(\sum_{p \in P(T)} |p|\log |p| + t\alpha(t) + N)$,
-where $P(T)$ is a specific path decomposition (constructed implicitly by the algorithm).
+where $P(T)$ is a specific path decomposition (constructed implicitly by the algorithm)\Ben{More than one reviewer has complained about the running 
+time being stated as a function of our algorithm.  So Rephrase, and maybe discussion about connection to sorting.}.
 
@@
 
-\InSoCGVer{See Appendix~\ref{sec:techmarks} for further technical remarks on the above definitions.}
+\InSoCGVer{See \Sec{techmarks} for further technical remarks on the above definitions.}
 \newcommand{\technicalRemarks}{
@@
 The directed edge $(u,v)$ is present when $u$ is the smallest valued vertex in $\cV$ in a connected component of $\MM^+_v$
-and $v$ is adjacent to a vertex in this component (in $\MM$). 
+and $v$ is adjacent (in $\MM$) to a vertex in this component. 
 \end{definition}
@@
 where the input is just a triangulated terrain, the main ideas of the parts of the algorithm 
-can be explained clearly. Therefore, in Appendix~\ref{sec:approach} we provide a high level view of the entire result.
+can be explained clearly. Therefore, in \Sec{approach} we provide a high level view of the entire result.
 We strongly encourage the reader to overview this section, as then the remainder of the paper can be understood 
@@
 
-\InSoCGVer{The following theorems are intuitively obvious, and are proven in Appendix~\ref{sec:proofOfSurgery}.}
+\InSoCGVer{The following theorems are intuitively obvious, and are proven in \Sec{proofOfSurgery}.}
 
@@
 of the procedure is that the rain alternately flows downwards and upwards, 
-since otherwise faces may be cut a super constant number of times (see \Fig{order}\InSoCGVer{ in Appendix~\ref{sec:approach}}).
+since otherwise faces may be cut a super constant number of times (see \Fig{order}\InSoCGVer{ in \Sec{approach}}).
 In other words, when
@@
 Hence the critical join tree below is just the join tree without these stumps.  
-Precise definitions and proofs are given Appendix~\ref{sec:jstedm}.
+Precise definitions and proofs are given \Sec{jstedm}.
 }
@@
 The directed edge $(u,v)$ is present when $u$ is the smallest valued vertex in $V'$ in a connected component of $\MM^+_v$
-and $v$ \Ben{is adjacent to a vertex} in this component (in $\MM$). 
+and $v$ \Ben{is adjacent (in $\MM$) to a vertex} in this component. 
 \end{definition}
 
-\InSoCGVer{Each non-dominant minimum, $m_i$, connects to the contour tree at a corresponding split $s_i$. We have the following (see Appendix~\ref{sec:jstedm} for more details).}
+\InSoCGVer{Each non-dominant minimum, $m_i$, connects to the contour tree at a corresponding split $s_i$. We have the following (see \Sec{jstedm} for more details).}
 
@@
 	\item Mark each critical $v$ as unprocessed.
-	\item For each critical $v$ and each up-star, pick arbitrary color $c$ touching $v$. Insert $v$ into $T(c)$.
+	\item For each critical $v$ and each up-star, pick an arbitrary color $c$ touching $v$. Insert $v$ into $T(c)$.
 	\item Initialize $\rep(c) = c$ and set $\h(c)$ to be the unique maximum colored $c$.
@@
 	\item If $K$ is empty, push arbitrary unprocessed critical point $v$.
-	\item Let $h$ be head of $\stack$.
+	\item Let $h$ be the head of $\stack$.
 	\item While $h$ is not ripe:
@@
 	\item Each vertex is present in at most $2$ heaps. After processing, it is removed from all heaps.
-	\item After $v$ is processed, all edges incident to $v$ have the same color (more technically, the same representative). 
+	\item After $v$ is processed, all edges incident to $v$ have the same (representative) color. %(more technically, the same representative). 
 %	\item When $\update(\stack)$ terminates, the head $h$ is ripe. 
@@
 
-\subsection{Running Time}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\newcommand{\algorithmRunTimeSection}{
+\InNotSoCGVer{\subsection{Running Time}}
+\InSoCGVer{\section{Running Time}}
 \label{sec:runTime}
-\Ben{This section needs to be reduced dramatically in size.}
-
-We now bound the running time of the above algorithm.  In the subsequent sections we will then provide a 
-matching lower bound for the running time.  Therefore, it will be useful to set up some terminology that can 
-be used consistently in both places.  Specifically, the lower bound proof will be a purely combinatorial 
-statement on colored rooted trees, and so the terminology is of this form.
+
+We now bound the running time of the algorithm of \Sec{algo}.  
+In subsequent sections, through a sophisticated charging argument, this bound is then related to matching upper 
+and lower bounds in terms of path decompositions. 
+%In the subsequent sections we will then provide a matching lower bound for the running time.  
+Therefore, it will be useful to set up some terminology that can 
+be used consistently in both places.  Specifically, the path decomposition bounds will be purely combinatorial 
+statements on colored rooted trees, and so the terminology is of this form.
 
 Any tree $T$ considered in following will be a rooted binary tree\footnote{Note 
-that technically the trees considered should have a leaf vertex hanging below the root of this 
+that technically the trees considered should have a leaf vertex hanging below the root  
 in order to represent the global minimum of the complex.  This vertex is 
-(safely) ignored to simplify the presentation.} where the height of a vertex is its distance 
+(safely) ignored to simplify presentation.} where the height of a vertex is its distance 
 from the root $r$ (i.e.\ conceptually $T$ will be a join tree with $r$ at the bottom).  
@@
 For a vertex $v\in T$, we use $H_v$ to denote the \emph{heap} at $v$.
-Formally, $H_v = \{u | u \in A(v), \chi(u) \cap L(T_v) \neq \emptyset\}$.
+Formally, $H_v = \{u | u \in A(v), \chi(u) \cap L(T_v) \neq \emptyset\}$, where $L(T_v)$ is the set of leaves of $T_v$.
 In words, $H_v$ is the set of ancestors of $v$ which are colored by some leaf in $T_v$.
@@
 \end{proof}
+}
+
+\InNotSoCGVer{\algorithmRunTimeSection}
+%%%%%%%%%%%%%%%%%%%%
+\InSoCGVer{
+\subsection{Upper and Lower bounds for Running Time}
+We now bound the running time of the algorithm of \Sec{algo}.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\Ben{Summarize \Sec{runTime} on the run time in terms of heap sizes.}
+
+\Ben{Summarize \Sec{pathDecomp} on path decomposition upped bound here.}
+
+\Ben{summarize \Sec{lb}, the lower bound section, result here.}
+
+% % We complement \Thm{main-alg} with a lower bound, suggesting that the comparison model complexity
+% % of $\sum_{p \in P(T)} |p|\log |p|$ is inherent for any contour tree algorithm. The exact phrasing
+% % of the lower bound requires some care. Consider the $d=2$ (terrain) case.
+% Let a path decomposition $P$ be called \emph{valid} if it is actually output by the algorithm on some input.
+% \begin{theorem} 
+% \label{thm:main-lb}
+% Consider a valid path decomposition $P$. There exists a family $\bF_P$ of terrains (i.e.\ $d=2$) with the following properties.
+% Any contour tree algorithm makes $\Omega(\sum_{p \in P} |p|\log |p|)$ comparisons in the worst case over $\bF_P$.
+% Furthermore, for any terrain in $\bF_P$, the algorithm of \Thm{main-alg} makes $O(\sum_{p \in P} |p|\log |p|)$ comparisons.
+% \end{theorem}
+% 
+% In other words, for any path decomposition, $P$, that our algorithm chooses, we can design a hard family where any 
+% algorithm requires $\sum_{p \in P} |p|\log |p|$ time in the worst case, and our algorithm matches this bound. 
+}
+%%%%%%%%%%%%%
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\InSoCGVer{
-\Ben{Summarize Appendix~\ref{sec:pathDecomp} on path decomposition upped bound here.}
-}
 
@@
 
-\InSoCGVer{
-\Ben{summarize Appendix \ref{sec:lb}, the lower bound section, result here.}
-}
 \InNotSoCGVer{
@@
 
+\algorithmRunTimeSection
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
 \leafAssignPathDecomp

### 2015-11-29 paint-contour-tree.tex
summary added

@@
 \subsection{How to Read this paper}
-\Ben{New section:}
+\Ben{New section, read carefully:}
 The final algorithm is quite technical and has numerous moving parts. However, for the $d=2$ case, 
@@
 on each piece (see Appendix~\ref{sec:approach} for discussion on why the join tree should not be computed directly on the input complex). 
-\Ben{Mention lower bound an analysis sections which are going to be moved to appendix.}
-
+Finally, \Sec{runTimeSummary} summarizes all sections relating to the running time, specifically \Sec{runTime} which upper bound the 
+running time of the algorithm of \Sec{paint}, \Sec{pathDecomp} which uses a sophisticated charging argument to express this bound 
+in terms of path decompositions, and \Sec{lb} which provides matching running time lower bounds.
 }
@@
 \InSoCGVer{
-\subsection{Upper and Lower bounds for Running Time}
-We now bound the running time of the algorithm of \Sec{algo}.
+\subsection{Upper and Lower bounds for Running Time}\label{sec:runTimeSummary}
+The full running time analysis can be found in \Sec{runTime}, \Sec{pathDecomp}, and \Sec{lb}, or in the full version \CiteFullVer.  
+The main results of these sections are summarized here.
+%
+In \Sec{runTime}, we give a tight bound on the running time of $\build(\MM)$, as a function of the choices made by the subroutine $\init(\MM)$.
+To state this bound we require the following technical definitions (which are repeated in \Sec{runTime}).
+
+\begin{definition}
+A \emph{leaf assignment} $\chi$ of a binary tree $T$ assigns \emph{two} distinct leaves to each internal vertex $v$,
+one from the left and one from the right subtree of $v$ (or only one leaf if $v$ has only one child). % naturally if $v$ has only one child it is assigned only one color). 
+The subroutine $\init(\MM)$ naturally defines a leaf assignment to $\jc(\MM)$ (which 
+is a rooted binary tree) according to the priority queue for each up-star we put a given vertex in.  
+Call this the \emph{initial coloring} of the vertices in $\jc(\MM)$, and denote it by $\chi$.
+
+For a vertex $v$ in $\jc(\MM)$, let $L(v)$ denote the set of leaves of the subtree rooted at $v$, and let $A(v)$ denote the set of 
+ancestors of $v$, i.e.\ the vertices on the $v$ to root path. For a vertex $v\in \jc(\MM)$, and an initial coloring $\chi$, 
+we use $H_v$ to denote the \emph{heap} at $v$. Formally, $H_v = \{u | u \in A(v), \chi(u) \cap L(v) \neq \emptyset\}$, i.e.\
+the set of ancestors colored by some leaf in $L(v)$.
+\end{definition}
+
+The algorithm $\build(\MM)$ processes vertices in $\jc(\MM)$ one at a time.  The main processing cost  
+comes from priority queue operations.  The heap as defined above characterizes the  
+priority queue size at the time a vertex gets processed and hence we get the following.
+
+\begin{lemma}
+Let $\MM$ be a simplicial complex with $t$ critical points.  
+The running time of $\build(\MM)$ is $O(N+t\alpha(t) + \sum_{v \in \jc(\MM)} \log |H_v|)$, where $H_v$ is defined by an initial coloring.
+%For every vertex in $\jc(\MM)$, let $H_v$ be defined by the initial coloring of $\MM$.
+%The running time of $\build(\MM)$ is $O(N+t\alpha(t) + \sum_{v \in \jc(\MM)} \log |H_v|)$.
+\end{lemma}
+
+While the above gives a full running time characterization, ideally the bound should be expressed as   
+a function of some structural property of $\jc(\MM)$.
+This is the focus of \Sec{pathDecomp}, where it is shown that for any binary tree $T$ 
+the $H_v$ sets (defined by any leaf assignment) naturally induces a path decomposition of $T$ (see \Def{path}), 
+denoted $\pmax(T)$ (see \Sec{pathDecomp} for details).
+
+% Specifically, the main result of \Sec{pathDecomp} is the following theorem, which together with the above 
+% bound in terms of heap size immediately implies our main result, Theorem~\ref{thm:main-alg} .
+
+\begin{theorem} $\sum_{v \in T} \log |H_v| = O(\sum_{p\in \pmax(T)} |p| \log |p|)$.
+\end{theorem}
+
+The above is a purely combinatorial statement on binary trees, and may find applications in other charging arguments.  
+However, for the specific when $T=\jc(\MM)$, plugging this theorem into the above lemma, and applying this lemma 
+to each subcomplex piece output by $\rain$, yields our main result, Theorem~\ref{thm:main-alg}.
+
+The path decomposition $\pmax(T)$ is a function of the choices made by the linear time $\init(\MM)$.  
+One cannot control which path decomposition $\init$ determines, as sorting can be reduced to finding the optimal path decomposition.  
+Specifically, in \Sec{lb} we exploit this connection to prove the following matching lower bound for our approach.
+In the following a rain consistent path decomposition is a decomposition which is the union of paths from 
+any decomposition on each subcomplex output by $\rain$ (see Definition~\ref{def:path2} for a formal definition).
+
+% \begin{theorem}
+% Consider a rooted tree $T$ and an arbitrary path decomposition $P$ of $T$.
+% There is a family $\bF_P$ of terrains such that any algebraic decision
+% tree computing the join tree (on $\bF_P$) requires $\Omega(\sum_{p \in P} |p|\log |p|)$ time.
+% Furthermore, our algorithm makes $O(\sum_{p \in P} |p|\log |p|)$ comparisons on all these instances.
+% \end{theorem}
+% 
+% The result can also 
+% For simplicity we now restrict our attention to path decompositions 
+% consistent with the raining procedure described in \Sec{rain} 
+% (more general decompositions can work, but it is not needed for our purposes).
+% 
+% Note that in the above $T$ is assumed to be a rooted tree, and so the contour tree is equivalent to the join tree.   
+% In general our input $\MM$ will not have a contour tree with this structure, which is why we required the cutting procedure of \Sec{rain}.  
+% The above result can be extended to such general contour trees, however, here we restrict to rain consistent path decomposition, 
+% i.e.\ decompositions which are the union of paths decompositions from each piece (see \ref{def:path2} for a formal definition)
+
+\begin{theorem} 
+Consider any rain consistent path decomposition $P$. There exists a family $\bF_P$ of terrains ($d=2$) with the following properties.
+Any contour tree algorithm makes $\Omega(\sum_{p \in P} |p|\log |p|)$ comparisons in the worst case over $\bF_P$.
+Furthermore, for any terrain in $\bF_P$, our algorithm makes $O(\sum_{p \in P} |p|\log |p|)$ comparisons.
+\end{theorem}
 
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\Ben{Summarize \Sec{runTime} on the run time in terms of heap sizes.}
-
-\Ben{Summarize \Sec{pathDecomp} on path decomposition upped bound here.}
-
-\Ben{summarize \Sec{lb}, the lower bound section, result here.}
 
@@
  in the decomposition. The key is that any color from a maximum in the subtree at $w_2$ cannot touch
- any ancestor of $v$. This subtree is exactly the contour tree of the child diamond attached at $v$.
+ any ancestor of $v$. This subtree is exactly the join tree of the child diamond attached at $v$.
  The base of this diamond is completely contained in a face of the parent diamond. So all colors
@@
 There is a family $\bF_P$ of terrains such that any algebraic decision
-tree computing the contour tree (on $\bF_P$) requires $\Omega(\sum_{p \in P} |p|\log |p|)$ time.
+tree computing the join tree\footnote{Note that for the referenced family of terrains, the join tree and contour tree are equivalent} 
+(on $\bF_P$) requires $\Omega(\sum_{p \in P} |p|\log |p|)$ time.
 Furthermore, our algorithm makes $O(\sum_{p \in P} |p|\log |p|)$ comparisons on all these instances.

### 2015-11-30 paint-contour-tree.tex
small changes

@@
 where $P(T)$ is a specific path decomposition (constructed implicitly by the algorithm)\Ben{More than one reviewer has complained about the running 
-time being stated as a function of our algorithm.  So Rephrase, and maybe discussion about connection to sorting.}.
+time being stated as a function of our algorithm.  So rephrase or add remark/footnote saying controlling which path decomp is used is like knowing the right ordering 
+before sorting, but unlike sorting, there are certain contour tree shapes such that there is no bad path decomp.  This somewhat said in the paragraph below, 
+but reviewers are not getting it.}.
 
@@
 
-Abusing notation, for now we will just call this the join tree. (In \Sec{extreme} we carefully handle the distinction 
+Abusing notation, for now we will just call this the join tree. (In \Sec{extreme} \InSoCGVer{and \Sec{jstedm}} we carefully handle the distinction 
 between the join tree and the critical join tree.)
@@
 on each piece (see Appendix~\ref{sec:approach} for discussion on why the join tree should not be computed directly on the input complex). 
-Finally, \Sec{runTimeSummary} summarizes all sections relating to the running time, specifically \Sec{runTime} which upper bound the 
+Finally, \Sec{runTimeSummary} summarizes all sections relating to the running time, specifically \Sec{runTime} which upper bounds the 
 running time of the algorithm of \Sec{paint}, \Sec{pathDecomp} which uses a sophisticated charging argument to express this bound 
@@
 Hence the critical join tree below is just the join tree without these stumps.  
-Precise definitions and proofs are given \Sec{jstedm}.
+Precise definitions and proofs are given in \Sec{jstedm}.
 }
@@
 \subsection{Upper and Lower bounds for Running Time}\label{sec:runTimeSummary}
+\Ben{This section is new, and only appears in the socg version}
 The full running time analysis can be found in \Sec{runTime}, \Sec{pathDecomp}, and \Sec{lb}, or in the full version \CiteFullVer.  
@@
 Specifically, in \Sec{lb} we exploit this connection to prove the following matching lower bound for our approach.
-In the following a rain consistent path decomposition is a decomposition which is the union of paths from 
-any decomposition on each subcomplex output by $\rain$ (see Definition~\ref{def:path2} for a formal definition).
+In the following a rain consistent path decomposition is any decomposition which is the union of paths from 
+decompositions on each subcomplex output by $\rain$ (see Definition~\ref{def:path2} for a formal definition).
 
@@
 %    \includegraphics[width=.095\linewidth]{figs/sideView}
-    \caption{Left: angled view of a diamond / Right: a parent and child diamond put together}
+    \caption{Left: angled view of a tent / Right: a parent and child tent put together}
     \label{fig:diamond}
@@
  We describe the basic building block of these terrains, which corresponds to a fixed path $p\in P$.  
- Informally, a \emph{diamond} is an upside down cone with $m$ triangular faces (see \Fig{diamond}). 
+ Informally, a \emph{tent} is an upside down cone with $m$ triangular faces (see \Fig{diamond}). 
  Construct a slightly tilted cycle 
- of length $m$ with the two antipodal points at heights $1$ and $0$. These are called the anchor and trap of the diamond, respectively.
+ of length $m$ with the two antipodal points at heights $1$ and $0$. These are called the anchor and trap of the tent, respectively.
  The remaining $m-2$ vertices are evenly spread around the cycle and heights decrease monotonically when going from the anchor to the trap.
@@
 
- Now we describe how to attach two different diamonds.
-In this process, we glue the base of a scaled down ``child'' diamond on to a triangular cone face of the larger  ``parent'' diamond (see \Fig{diamond}).
- Specifically, the anchor of the child diamond is attached directly to a face of the parent diamond at some height $h$. The remainder of the base of the child cone is then 
+ Now we describe how to attach two different tents.  In this process, we glue the base of a scaled down ``child'' tent on to a triangular cone face of the larger  ``parent'' tent (see \Fig{diamond}).
+ Specifically, the anchor of the child tent is attached directly to a face of the parent tent at some height $h$. The remainder of the base of the child cone is then 
  extended down (at a slight angle) until it hits the face of the parent.
 
- The full terrain is obtained by repeatedly gluing diamonds. For each path $p_i\in P$, we create a diamond of size $|p_i|+1$.
- The two faces adjacent to the anchor are always empty, and the remaining faces are for gluing on other diamonds.
- (Note that diamonds have size $|p_i|+1$ since $|p_i|-1$ faces represent the joins of $p_i$, the apex represents the leaf, 
+ The full terrain is obtained by repeatedly gluing tents. For each path $p_i\in P$, we create a tent of size $|p_i|+1$.
+ The two faces adjacent to the anchor are always empty, and the remaining faces are for gluing on other tents.
+ (Note that tents have size $|p_i|+1$ since $|p_i|-1$ faces represent the joins of $p_i$, the apex represents the leaf, 
  and we need two empty faces next to the anchor.)
- Now we glue together diamonds of different paths in the same way the paths are connected in the shrub $\pathTree$ (see \Sec{shrubs}).  
+ Now we glue together tents of different paths in the same way the paths are connected in the shrub $\pathTree$ (see \Sec{shrubs}).  
  Specially, for two paths $p,q\in P$ where 
- $p$ is the parent of $q$ in $\pathTree$, we glue $q$ onto a face of the diamond for $p$ as described above.
- (Naturally for this construction to work, diamonds for a given path will be scaled down relative to the size of the diamond of their parent).
+ $p$ is the parent of $q$ in $\pathTree$, we glue $q$ onto a face of the tent for $p$ as described above.
+ (Naturally for this construction to work, tents for a given path will be scaled down relative to the size of the tent of their parent).
  By varying the heights of the gluing, we get the family of terrains.
@@
  Observe now that the only saddle points in this construction are the anchor points.  Moreover, the only maxima are the apexes of the 
- diamonds.  We create a global boundary minimum by setting the vertices at the base of the diamond representing the root of $\pathTree$ all to the same height (and there are no other minima).  
- Therefore, the saddles on a given diamond will appear contiguously on a root to leaf path in the join tree of the terrain, where the leaf corresponds to the maximum of the diamond 
+ tents.  We create a global boundary minimum by setting the vertices at the base of the tent representing the root of $\pathTree$ all to the same height (and there are no other minima).  
+ Therefore, the saddles on a given tent will appear contiguously on a root to leaf path in the join tree of the terrain, where the leaf corresponds to the maximum of the tent 
  (since all these saddles have a direct line of sight to this apex).  In particular, 
@@
 
- There is a valid instance of this described construction for any relative ordering of the heights of the saddles on a given diamond.  
- In particular, there are $(|p_i|-1)!$ possible orderings of the heights of the saddles on the diamond for $p_i$, and hence $\Pi_{p_i\in P} (|p_i|-1)!$ possible 
- terrains we can build.  Each one of these functions will result in a different (labeled) join tree. All saddles on a given diamond will appear in sorted order in the join tree.
- So, any permutation of the heights on a given diamond corresponds to a permutation of the vertices along a path in $P$.
+ There is a valid instance of this described construction for any relative ordering of the heights of the saddles on a given tent.  
+ In particular, there are $(|p_i|-1)!$ possible orderings of the heights of the saddles on the tent for $p_i$, and hence $\Pi_{p_i\in P} (|p_i|-1)!$ possible 
+ terrains we can build.  Each one of these functions will result in a different (labeled) join tree. All saddles on a given tent will appear in sorted order in the join tree.
+ So, any permutation of the heights on a given tent corresponds to a permutation of the vertices along a path in $P$.
 \end{proof}
@@
 %For any choice of $\MM \in \bF$, there is a path decomposition $\overline{P}(T)$ of the join tree equivalent to $\mathcal{P}$.
-%Specifically, this is the decomposition where each path $p\in \overline{P}(T)$ consists exactly of the set of critical points (except for its anchor) from a given diamond. 
+%Specifically, this is the decomposition where each path $p\in \overline{P}(T)$ consists exactly of the set of critical points (except for its anchor) from a given tent. 
 The primary ``non-determinism" of the algorithm is the initial painting constructed by $\init(\MM)$. 
@@
  Consider an arbitrary order of the initial paint spilling over the surface.
- Consider any join on a face of some diamond, which is the anchor point of some connecting child diamond.  
+ Consider any join on a face of some tent, which is the anchor point of some connecting child tent.  
  The join has two up-stars, each of which has exactly one edge. Each edge connects to a maximum
@@
  Hence, the two colors touching this join (according to \Def{initialColoring})
- are the colors of the apexes of the child and parent diamond.  
-% 
-% Specifically, one of these edges connects directly to the apex of the parent diamond and so the join must get the color of this maxima.
-% The other edge connects directly to the apex of the child diamond (this is why we left the faces adjacent to an anchor empty) and 
-% so the second color is that of the maxima of the child diamond.
+ are the colors of the apexes of the child and parent tent.  
+% 
+% Specifically, one of these edges connects directly to the apex of the parent tent and so the join must get the color of this maxima.
+% The other edge connects directly to the apex of the child tent (this is why we left the faces adjacent to an anchor empty) and 
+% so the second color is that of the maxima of the child tent.
 
@@
  in the decomposition. The key is that any color from a maximum in the subtree at $w_2$ cannot touch
- any ancestor of $v$. This subtree is exactly the join tree of the child diamond attached at $v$.
- The base of this diamond is completely contained in a face of the parent diamond. So all colors
- from the child ``drain off" to the base of the parent, and do not touch any joins on the parent diamond.
+ any ancestor of $v$. This subtree is exactly the join tree of the child tent attached at $v$.
+ The base of this tent is completely contained in a face of the parent tent. So all colors
+ from the child ``drain off" to the base of the parent, and do not touch any joins on the parent tent.
 
@@
 % which $c$ appears are exactly the vertices of the path in  $\overline{P}(T)$ corresponding $m$, as well as 
-% the vertex to which the root of this path connects to (i.e.\ the anchor of the diamond of $m$).
+% the vertex to which the root of this path connects to (i.e.\ the anchor of the tent of $m$).
 % Now we show this implies $\overline{P}(T)$ is a maximum path decomposition.
@@
 % By the above lemmas, there is then a set $\bF$ of $\Pi_{p_i\in P} |p_i|!$ functions, such that for any $g\in \bF$, $C= \cost(g)$.  
-% (Specifically, each sorted ordering of the relative height values of the saddles on each diamond yields a function in this class.)
+% (Specifically, each sorted ordering of the relative height values of the saddles on each tent yields a function in this class.)
 % Hence $\bF\subseteq \bF_C$, and so any height function decision tree for the set $\bF_C$ must distinguish between these possibilities, and therefore must 
@@
  As the set $P_i$ is a path decomposition of a rooted binary tree $T_i$, the terrain construction of \Lem{pathDecomp} for $P_i$ is well defined.  
- Actually the only difference is that here the rooted tree is not a full binary tree, and so some of the (non-achor adjacent) faces of the constructed diamonds will be blank.  
+ Actually the only difference is that here the rooted tree is not a full binary tree, and so some of the (non-achor adjacent) faces of the constructed tents will be blank.  
  Specifically, these blank faces correspond to the adjacent children of $P_i$, and they tell us how to connect the terrains of the different $P_i$'s.
@@
     \includegraphics[width=.28\linewidth]{figs/cones2}
-    \caption{A child diamond attached to a parent diamond with opposite orientation.}
+    \caption{A child tent attached to a parent tent with opposite orientation.}
     \label{fig:diamond2}
@@
 
- By construction, the diamonds corresponding to the paths in $P_i$ are connected 
- into a tree structure (i.e.\ corresponding to the shrub of $P_i$).  Therefore the bottoms of all these diamonds are covered except for the one corresponding 
+ By construction, the tents corresponding to the paths in $P_i$ are connected 
+ into a tree structure (i.e.\ corresponding to the shrub of $P_i$).  Therefore the bottoms of all these tents are covered except for the one corresponding 
  to the path containing the root $r_i$.  
  If $r_i$ corresponds to the initial maximum that the rain consistent path decomposition was defined from, then this will be flat and corresponds to the global outer face.
- Otherwise, $P_i$ has some parent $P_j$ in which case we connect the bottom of the diamond for $r_i$ to a free face of a diamond in the construction for $P_j$, specifically, the face corresponding to the
+ Otherwise, $P_i$ has some parent $P_j$ in which case we connect the bottom of the tent for $r_i$ to a free face of a tent in the construction for $P_j$, specifically, the face corresponding to the 
  vertex in $T$ which $r_i$ is adjacent to.
@@
  Just as in \Lem{pathDecomp} we now have one fixed terrain structure, such that each different relative ordering of the heights of the join and split vertices 
- on each diamond produces a surface with a distinct contour tree.  The specific bound on the size of $\bF_P$, defining these distinct contour trees, follows by applying the 
+ on each tent produces a surface with a distinct contour tree.  The specific bound on the size of $\bF_P$, defining these distinct contour trees, follows by applying the 
  bound from \Lem{pathDecomp} to each $P_i$.

### 2015-12-02 paint-contour-tree.tex
seshs pass

@@
 Specifically, our approach ensures that any comparison made is between nodes in the same descending path in the contour tree,
-allowing us to complement our algorithmic results with matching lower bounds for contour tree computation.
+allowing us to argue strong optimality properties of our algorithm.
 
@@
 
-Geometric data (potentially massive in size) is often represented as a function $f: \RR^d \to \RR$. Typically, a finite representation is given
+Geometric data is often represented as a function $f: \RR^d \to \RR$. Typically, a finite representation is given
 by considering $f$ to be piecewise linear over some triangulated mesh (i.e.\ simplicial complex) $\MM$ in $\RR^d$.
-\emph{Contour trees} are a specific topological structure used to represent and visualize
+\emph{Contour trees} are a topological structure used to represent and visualize
 the function $f$. 
-For example, consider a rocket engine during a burn.  The 3D mesh represents the locations of points in space around the engine, 
-and $f$ represents the temperature at each point.  The boundary of the flame is then represented by an isotherm, i.e.\ a contour.
-
+% For example, consider a rocket engine during a burn.  The 3D mesh represents the locations of points in space around the engine, 
+% and $f$ represents the temperature at each point.  The boundary of the flame is then represented by an isotherm, i.e.\ a contour. 
+% 
 It is convenient to think of $f$ as a manifold sitting in $\RR^{d+1}$,
@@
 split into new components, contours acquire or lose handles. More generally, these are the events where 
-the topology of the level set changes. The contour tree is a concise representation of all these events.\footnote{%
+the topology of the level set changes. The contour tree is a concise representation of all these events.
 Throughout we follow the definition of contour trees from \cite{kobps-ctsssit-97} which includes all changes in topology.  
 For $d>2$, some subsequent works, such as \cite{csa-cctad-00}, only include changes in the number of components.
-}
-\Ben{See added footnote}
 
@@
 Suppose we have $f: \MM \to \RR$, where $\MM$ is a triangulated mesh with $n$ vertices, $N$ faces in total, and $t \leq n$ critical points. 
-\Ben{It is standard to assume that $f$ has distinct values on the vertices of $\MM$ and the value at interior points is linearly interpolated.}
+We assume the input is $f:\MM \to \RR$ a linear interpolant over distinct valued vertices, where the contour tree
+$T$ has maximum degree $3$. The degree assumption simplifies the proof and presentation, and is commonly made~\cite{kobps-ctsssit-97}.
+% \Ben{It is standard to assume that $f$ has distinct values on the vertices of $\MM$ and the value at interior points is linearly interpolated.}
 %It is standard to assume that $f$ is PL-Morse (a well-behaved input assumption) and that the maximum degree in the contour tree is $3$. 
@@
 
+% \Sesh{Almost everything below is new/reordered. Please read through.}
 But this is only a worst-case bound. All previous algorithms begin by sorting (at least) the critical points, and
 hence must pay $\Omega(t\log t)$ for all instances. Can we beat this sorting bound for certain instances,
-and can we characterize which inputs are hard? Our main result gives an affirmative answer. 
+and can we characterize which inputs are hard? Intuitively, points that are incomparable in the contour tree
+do not need to be compared. Look at \Fig{relative} to see such an example. All previous algorithms waste time
+sorting all the maxima. Also consider the surface of \Fig{sorting}. The final contour tree is basically two binary
+trees joined at their roots, and we do not need to entire sorted order of critical points
+to construct the contour tree. Is it possible to beat the $t\log t$ bound for these examples?
+
+\begin{figure}[h]\centering
+    \includegraphics[width=.26\linewidth]{figs/spikey1}%
+    \hspace{.35in}
+    \includegraphics[width=.26\linewidth]{figs/spikey2}%
+    \hspace{.4in}
+    \includegraphics[width=.215\linewidth]{figs/spikeyTree}
+    \caption{Two surfaces with different orderings of the maxima, but the same contour tree.}
+    \label{fig:relative}
+\end{figure}
+
+\begin{figure}[h]\centering
+    \includegraphics[width=.35\linewidth,height=.15\linewidth]{figs/balanced4}
+    \hspace{.3in}
+    \includegraphics[width=.35\linewidth,height=.15\linewidth]{figs/trees}
+    \caption{On left, a surface with a balanced contour tree, but whose join and split trees have long tails.  
+    On right (from left to right), the contour, join and split trees.}
+    \label{fig:sorting}
+\end{figure}
+
+Our main result gives an affirmative answer. Remember that we can consider
+the contour tree as directed from top to bottom. For any node $v$ in the tree, let $\ell_v$ denote the length of the 
+longest directed path passing through $v$.
+
+\begin{theorem} \label{thm:main-corr} Consider a simplicial complex $f:\MM \to \RR$, described as above, and denote
+the contour tree by $T$ with vertex set (the critical points) $C(T)$. There exists an algorithm to compute the contour tree $T$
+can be computed in $O(\sum_{v \in C(T)} \ell_v + t\alpha(t) + N)$ time. Moreover, this algorithm only
+compares function values at pairs of points that are ancestor-descendant in $T$.
+\end{theorem}
+
+Essentially, the ``run time per critical point" is the height/depth of the point in the contour tree.
+This bound immediately yields a run time of $O(t\log D + t\alpha(t) + N)$,
+where $D$ is the diameter of the contour tree. 
+This is a significant improvement for short and fat contour trees. For example, if the tree is balanced,
+then we get a bound of $O(t\log\log t)$.
+Even if $T$ contains a long path of length $O(t/\log t)$, but is otherwise short, we get the improved bound of $O(t\log\log t)$.
+
+\subsection{A refined bound with optimality properties}
+
+\Thm{main-corr} is a direct corollary of a more cumbersome theorem.
+
 % First, we require some technical definitions.
@@
 %Our algorithm does not need this, but it simplifies the proof and presentation.} 
-\Ben{Let $f:\MM \to \RR$ be the linear interpolant over distinct valued vertices, where the contour tree $T$
-has maximum degree $3$}.\footnote{The degree $\leq 3$ assumption helps simplify the proof and presentation, and is often made, as in \cite{kobps-ctsssit-97}.}\Ben{See footnote} 
 There is a deterministic algorithm to compute $T$ whose running time is $O(\sum_{p \in P(T)} |p|\log |p| + t\alpha(t) + N)$,
-where $P(T)$ is a specific path decomposition (constructed implicitly by the algorithm)\Ben{More than one reviewer has complained about the running 
-time being stated as a function of our algorithm.  So rephrase or add remark/footnote saying controlling which path decomp is used is like knowing the right ordering 
-before sorting, but unlike sorting, there are certain contour tree shapes such that there is no bad path decomp.  This somewhat said in the paragraph below, 
-but reviewers are not getting it.}.
-
+where $P(T)$ is a specific path decomposition (constructed implicitly by the algorithm).
 The number of comparisons made is $O(\sum_{p \in P(T)} |p|\log |p| + N)$.
+In particular, any comparisons made are only between ancestors and descendants in the contour tree.
 \end{theorem}
 
-In particular, any comparisons made are only between ancestors and descendants in the contour tree, while previous algorithms compared everything.
-
-Let us interpret the running time. 
-For any path decomposition $P(T)$, obviously $\sum_{p\in P(T)}|p| = |T| = t$, and 
-so the running time is $O(t\log t + N)$. But this is tight only when $T$ is tall and thin.  
-A better upper bound is $O(t\log D + t\alpha(t) + N)$, where $D$ is the diameter of $T$. 
-This is clearly an improvement when the tree is short and fat, but it can be even better. 
-Consider a contour tree that is constant balanced, meaning that the ratio of sizes of sibling subtrees
-is $\Theta(1)$. 
-A calculation yields that for any decomposition of such a contour tree, 
-our bound is $O(t\alpha(t) + n)$. 
-%
-%In particular, for any decomposition of a rooted balanced binary tree, this gives $O(t\log\log n + t\alpha(t) + N)$, a significant improvement from $O(t\log t+N)$.
-%However, our actual bound in this case gives an even further improved $O(t\alpha(t)+N)$.
-Consider the example of \Fig{sorting} extended to a larger binary tree both above and below,
-so $t = \Omega(n)$. The running time of the algorithm of \Thm{main-alg}
-is $O(n\alpha(n))$, whereas all previous algorithms require $\Omega(n\log n)$ time.
-
-We complement \Thm{main-alg} with a lower bound, suggesting that the comparison model complexity
-of $\sum_{p \in P(T)} |p|\log |p|$ is inherent for any contour tree algorithm. The exact phrasing
-of the lower bound requires some care. Consider the $d=2$ (terrain) case.
-Let a path decomposition $P$ be called \emph{valid} if it is actually output by the algorithm
-on some input.
+Note that \Thm{main-corr} is a direct corollary of this statement. This is strictly stronger,
+since for any balanced contour tree, the run time bound is $O(t\alpha(t) + N)$, and $O(t)$
+comparisons are made.
+
+The bound of \Thm{main-alg} may seem artificial, since it actually depends on the $P(T)$ that
+is implicitly constructed by the algorithm. Nonetheless, we prove that the algorithm of \Thm{main-alg}
+has strong optimality properties. For convenience, fix some value of $t$, and consider the set
+of terrains ($d=2$) with $t$ critical points. The bound of \Thm{main-alg} takes values ranging
+from $t$ to $t\log t$. Consider some $C \in [t,t\log t]$, and consider the set of terrains
+where the algorithm makes $C$ comparisons. Then \emph{any algorithm} must make roughly $C$
+comparisons in the worst-case over this set.
 
@@
 \label{thm:main-lb}
-%Fix a sufficiently large numbers $t$ and $C$, 
-%where $t$ is the number of critical points and $C$ is a valid cost for an instance with $t$ critical point.
-Consider a valid path decomposition $P$. There exists a family $\bF_P$ of terrains (i.e.\ $d=2$) with the following properties.
-Any contour tree algorithm makes $\Omega(\sum_{p \in P} |p|\log |p|)$ comparisons in the worst case over $\bF_P$.
-Furthermore, for any terrain in $\bF_P$, the algorithm of \Thm{main-alg} makes $O(\sum_{p \in P} |p|\log |p|)$ comparisons.
-%
-%the set of functions $\bF_C$ that have the same number of critical points and the same (valid) cost $C$.  Any algebraic decision tree that correctly computes
-%the contour tree on all of $\bF_C$ has a worst case running time of $\Omega(C)$.
+There exists some absolute constant $\alpha$ such that the following holds.
+For sufficiently large $t$ and $C$: consider the set $\bF_C$ of terrains with $t$ critical points such that
+the number of comparisons made by the algorithm of \Thm{main-alg} on these terrains is in $[C,\alpha C]$.
+Any algebraic decision tree that correctly computes
+the contour tree on all of $\bF_C$ has a worst case running time of $\Omega(C)$.
 \end{theorem}
-
-In other words, for any path decomposition, $P$, we can design a hard family where any algorithm
-requires $\sum_{p \in P} |p|\log |p|$ time in the worst case, and our algorithm matches this bound.
-
-\myparagraph{Beyond worst-case analysis.} Our result provides understanding of the complexity
-of contour tree construction beyond the standard worst-case analysis. Our theorem and lower bound 
-characterize hard instances.
-% explain \emph{when}
-% computing contour trees is expensive, and provide insight into the hardness of particular instances. 
-%It is intriguing
-%that the cost (as defined earlier) roughly captures the hardness of an instance.
-Trees that are short and fat are easy to compute, whereas trees with long paths are hard.
-This is tangentially related to the concept of instance-optimal algorithms, as defined by Afshani \etal \cite{abc-ioga-09}.
-Their notion of optimality is far more refined that what we prove, but it shares the flavor of understanding the spectrum
-of easy to hard instances. In general, our analysis forces one to look closely
-at the relationship between sorting and the contour tree problem.
+% 
+% \myparagraph{Beyond worst-case analysis.} Our result provides understanding of the complexity
+% of contour tree construction beyond the standard worst-case analysis. Our theorem and lower bound 
+% characterize hard instances.
+% % explain \emph{when}
+% % computing contour trees is expensive, and provide insight into the hardness of particular instances. 
+% %It is intriguing
+% %that the cost (as defined earlier) roughly captures the hardness of an instance.
+% Trees that are short and fat are easy to compute, whereas trees with long paths are hard.
+% This is tangentially related to the concept of instance-optimal algorithms, as defined by Afshani \etal \cite{abc-ioga-09}.
+% Their notion of optimality is far more refined that what we prove, but it shares the flavor of understanding the spectrum
+% of easy to hard instances. In general, our analysis forces one to look closely
+% at the relationship between sorting and the contour tree problem. 
+%
 
@@
 
-\begin{figure}[h]\centering
-    \includegraphics[width=.26\linewidth]{figs/spikey1}%
-    \hspace{.35in}
-    \includegraphics[width=.26\linewidth]{figs/spikey2}%
-    \hspace{.4in}
-    \includegraphics[width=.215\linewidth]{figs/spikeyTree}
-    \caption{Two surfaces with different orderings of the maxima, but the same contour tree.}
-    \label{fig:relative}
-\end{figure}
-
-\begin{figure}[h]\centering
-    \includegraphics[width=.35\linewidth,height=.15\linewidth]{figs/balanced4}
-    \hspace{.3in}
-    \includegraphics[width=.35\linewidth,height=.15\linewidth]{figs/trees}
-    \caption{On left, a surface with a balanced contour tree, but whose join and split trees have long tails.  
-    On right (from left to right), the contour, join and split trees.}
-    \label{fig:sorting}
-\end{figure}
-
 % \paragraph{Organization.}  
@@
 
-\newcommand{\intuitionSection}{
+\newcommand{\newintuitionSection}{
 \section{A tour of the new contour tree algorithm} \label{sec:approach}
-\InNotSoCGVer{
 Our final algorithm is quite technical and has numerous moving parts. However, for the $d=2$ case, 
 where the input is just a triangulated terrain, the main ideas of the parts of the algorithm 
-can be explained clearly.  Therefore, here we first provide a high level view of the entire result.
-}
-\InSoCGVer{
-For the $d=2$ case, where the input is just a triangulated terrain, the main ideas of the parts of the algorithm 
-can be explained clearly.  Therefore, here we provide a high level view of the entire result.
-}
-
-In the interest of presentation, the definitions and theorem statements in this section will
-slightly differ from those in the main body. They may also differ from the original definitions proposed
-in earlier work.
-
-\myparagraph{Do not globally sort:} The starting point for this work is \Fig{relative}. We have two terrains
-with exactly the same contour tree, but different orderings of (heights of) the critical points.
-Turning it around, we cannot deduce the full height ordering of critical points from the contour tree.
-Sorting all critical points is computationally unnecessary for constructing the contour tree.
-In \Fig{sorting}, the contour tree consists of two balanced binary trees, one of the joins,
-another of the splits. Again, it is not necessary to know the relative ordering between the mounds
-on the left (or among the depressions on the right) to compute the contour tree. Yet some ordering
-information is necessary: on the left, the little valleys are higher than the big central valley,
-and this is reflected in the contour tree. Leaf paths in the contour tree have points
-in sorted order, but incomparable points in the tree are unconstrained.
-How do we sort exactly what is required, without knowing the contour tree in advance?
-
-\subsection{Breaking $\MM$ into simpler pieces} \label{sec:break}
-Let us begin with the algorithm of Carr, Snoeyink, and Axen~\cite{csa-cctad-00}. The key insight is to build
-two different trees, called the join and split trees, and then merge them together into the contour tree.
-Consider sweeping down via the hyperplane $x_{d+1} = h$ and taking the \emph{superlevel} sets. These 
-are the connected components of the portion of $\MM$ above height $h$. For a terrain, the superlevel
+can be explained clearly.
+
+\myparagraph{Cutting $\MM$ into extremum dominant pieces:} We define a simplicial complex
+endowed with a height to be \emph{minimum dominant} if there exists only a single minimum.
+Analogously, we can define maximum dominant complexes, and both are collectively called extremum dominant.
+
+We first cut $\MM$ into disjoint extremum dominant pieces in linear time.
+Take an arbitrary maximum $x$ and imagine torrential rain at the maximum. The water flows down, wetting any point that has a non-ascending path 
+from $x$. We end up with two portions, the wet part of $\MM$ and the dry part. This is similar
+to \emph{watershed} algorithms used for image segmentation~\cite{RoMe00}. The wet part is obviously connected and
+can be shown to be extremum dominant. We can cut along the interface of the wet and dry, remove
+the wet part, and recurse of the dry part. Each additional cut increases the complexity of the surface,
+so care is required to keep the total complexity linear. This is achieved by alternately ``raining"
+upwards and downwards.  We finally get a collection of extremum dominant pieces.
+We prove a simple contour surgery theorem that builds the contour tree of $\MM$
+from the contour trees of the various pieces created.
+
+Using ideas from~\cite{csa-cctad-00}, we prove that the contour tree of an extremum dominant complex is (basically) just the \emph{join tree}.
+For the $d=2$ case, it helps to visualize such an input as a terrain with no minima,
+except at a unique boundary face (think of a large boundary triangle that is the boundary).
+Consider sweeping down via the hyperplane $x_{d+1} = h$ and taking the connected components of the portion of $\MM$ above height $h$. For a terrain, these superlevel
 sets are a collection of ``mounds". As we sweep downwards, these mounds keep joining each other,
 until finally, we end up with all of $\MM$. The join tree tracks exactly these events.
-Formally, let $\MM^+_v$ denote the simplicial complex induced on the subset of 
-vertices which are higher than $v$.
-
-\begin{definition} 
-\label{def:int-criticalJoin}
-The \emph{critical join tree} $\jc(\MM)$ is built on the set $\cV$ of all critical points.\footnote{%
-In Definition~\ref{def:criticalJoin}, we define the critical join tree on a certain subset of $\cV$.}
-The directed edge $(u,v)$ is present when $u$ is the smallest valued vertex in $\cV$ in a connected component of $\MM^+_v$
-and $v$ is adjacent (in $\MM$) to a vertex in this component. 
-\end{definition}
-
-Abusing notation, for now we will just call this the join tree. (In \Sec{extreme} \InSoCGVer{and \Sec{jstedm}} we carefully handle the distinction 
-between the join tree and the critical join tree.)
-Refer to \Fig{sorting} for the join tree of a terrain. Note that nothing happens at splits, but these
-are still put as vertices in the join tree. They simply form a long path. The split tree
-is obtained by simply inverting this procedure, sweeping upwards and tracking sublevel sets.
-
-A major insight of~\cite{csa-cctad-00} is an ingeniously simple linear time procedure to construct
-the contour tree from the join and split trees. So the bottleneck is computing these trees. Observe
-in \Fig{sorting} that the split vertices form a long path in the join tree (and vice versa). Therefore, constructing
-these trees forces a global sort of the splits, an unnecessary computation for the contour tree. Unfortunately, in general (i.e.\ unlike \Fig{sorting}) 
-the heights of joins and splits may be interleaved in a complex manner, and hence the final merging of~\cite{csa-cctad-00}
-to get the contour tree requires having the split vertices in the join tree. Without
-this, it is not clear how to get a consistent view of both joins and splits, required for the contour tree.
-
-Our aim is to break $\MM$ into smaller pieces, where this unnecessary computation can be avoided.
-
-\myparagraph{Contour surgery:} We first need a divide-and-conquer lemma. Any contour $\phi$ can be associated
-with an edge $e$ of the contour tree. Suppose we ``cut" $\MM$ along this contour. We prove that $\MM$
-is split into two disconnected pieces, such the contour trees of these pieces is obtained
-by simply cutting $e$ in $\reeb(\MM)$. Alternatively, the contour trees of these pieces can
-be glued together to get $\reeb(\MM)$. This is not particularly surprising, and is fairly easy to prove
-with the right definitions. The idea of loop surgery has been used to reduce Reeb graphs to contour trees~\cite{TiGySi09,DoNa13}.
-Nonetheless, our theorem appears to be new and works for all dimensions.
-
-\begin{figure}[h]\centering
-    \includegraphics[width=.172\linewidth]{figs/tent2}%
-    \hspace{1.2in}
-    \includegraphics[width=.172\linewidth]{figs/tent1}
-    \caption{On left, downward rain spilling only (each shade of gray represents a piece created by each different spilling), producing a grid.  
-    Note we are assuming raining was done in sorted order of the maxima (i.e.\ lowest to highest).  
-    On right, flipping the direction of rain spilling.}
-    \label{fig:order}
-\end{figure}
-
-\myparagraph{Cutting $\MM$ into extremum dominant pieces:} We define a simplicial complex
-endowed with a height to be \emph{minimum dominant} if there exists only a single minimum.
-(Our real definition is more complicated, and involves simplicial complexes
-that allow additional ``trivial'' minima.)
-In such a complex, there exists a non-ascending path from any point to this unique minimum.
-Analogously, we can define maximum dominant complexes,
-and both are collectively called extremum dominant.
-
-We will cut $\MM$ into disjoint extremum dominant pieces, in linear time.
-One way to think of our procedure is a meteorological analogy. Take an arbitrary maximum $x$,
-and imagine torrential rain at the maximum. The water flows down, wetting any point that has a non-ascending path 
-from $x$. We end up with two portions, the wet part of $\MM$ and the dry part. This is similar
-to \emph{watershed} algorithms used for image segmentation~\cite{RoMe00}. The wet part is obviously connected,
-while there may be numerous disconnected dry parts. The interface between the dry and wet parts
-is a set of contours\footnote{Technically, they are not contours, but rather the limits of sequences of contours.}, 
-given by the ``water line". The wet part is clearly maximum dominant, since all
-wet points have a non-descending path to $x$. So we can simply cut along the interface
-contours to get the wet maximum dominant piece $\MM'$. By our contour surgery theorem,
-we are left with a set of disconnected dry parts, and we can recur this procedure on them.
-
-But here's the catch. Every time we cut $\MM$ along a contour, we potentially increase
-the complexity of $\MM$. Water flows in the interior of faces, and the interface will naturally
-cut some faces. Each cut introduces new vertices, and a poor choice of repeated raining leads
-to a large increase in complexity. Consider the left of \Fig{order}. Each raining produces a single
-wet and dry piece, and each cut introduces many new vertices. If we wanted to partition this terrain into 
-maximum dominant simplicial complexes, the final complexity would be forbiddingly large.
-
-A simple trick saves the day. Unlike reality, we can choose rain to flow solely downwards or solely upwards.
-Apply the procedure above to get a single wet maximum dominant $\MM'$ and a set of dry pieces.
-Observe that a single dry piece $\NN$ is boundary critical with the newly introduced boundary $\phi$ (the wet-dry interface) behaving as a minimum.
-So we can rain upwards from this minimum, and get a \emph{minimum dominant} portion $\NN'$.
-This ensures that the new interface (after applying the procedure on $\NN$) does not
-cut any face previously cut by $\phi$.
-For each of the new dry pieces, the newly introduced boundary is now a maximum. So we rain downwards from there.
-More formally, we alternate between raining upwards and downwards as we go down the recursion tree.
-We can prove that an original face of $\MM$ is cut at most once, so the final complexity can be bounded.
-In \Fig{order}, regardless of the choice of the starting maximum,  this procedure would yield (at most) two pieces, 
-one maximum dominant, and one minimum dominant.
-
-Using the contour surgery theorem previously discussed, we can build the contour tree of $\MM$
-from the contour trees of the various pieces created. All in all, we prove the following theorem.
-
-\begin{theorem} \label{thm:int-rain} There is an $O(N)$ time procedure that cuts $\MM$ into extremum dominant simplicial complexes
-$\MM_1, \MM_2, \ldots$. Furthermore, given the set of contour trees $\{\cC(\MM_i)\}$, $\cC(\MM)$ can be constructed in $O(N)$ time.
-\end{theorem}
-
-\myparagraph{Extremum dominance simplifies contour trees:} We will focus on minimum dominant simplicial complexes $\MM$. By \Thm{int-rain}, it suffices
-to design an algorithm for contour trees on such inputs. 
-For the $d=2$ case, it helps to visualize such an input as a terrain with no minima,
-except at a unique boundary face (think of a large boundary triangle that is the boundary).
-All the saddles in such a terrain are necessarily joins, and there can be no splits.
-In \Fig{sorting}, the portion on the left is minimum dominant in exactly this fashion, albeit in one dimension lower.
-%Water poured at any critical point must flow to the boundary, and the entire wet portion is connected.
-More formally, $\MM^-_v$ is connected for all $v$, so there are no splits.
-
-We can prove that the split tree is just a path, and the contour tree is exactly
-the join tree. The formal argument is a little involved, and we employ the merging procedure
-of~\cite{csa-cctad-00} to get a proof. We demonstrate that the merging procedure will
-actually just output the join tree, so we do not need to actually compute the split tree.
-(The real definition of minimum dominant is a little more complicated,
-so the contour tree is more than just the join tree. But computationally, it suffices
-to construct the join tree.)
-
-We stress the importance of this step for our approach. Given the algorithm of~\cite{csa-cctad-00},
-one may think that it suffices to design faster algorithms for join trees. But this
-cannot give the sort of optimality we hope for. Again, consider \Fig{sorting}. Any algorithm
-to construct the true join tree must construct the path of splits, which implies
-sorting all of them. It is absolutely necessary to cut $\MM$ into pieces where the cost
-of building the join tree can be related to that of building $\cC(\MM)$.
-
-\subsection{Join trees from painted mountaintops} \label{sec:join-paint}
-
-Arguably, everything up to this point is a preamble for the main result: a faster
-algorithm for join trees. Our algorithm does not require the initial input to be extremum
-dominant. This is only required to relate the join trees, of the resulting subcomplexes of Theorem~\ref{thm:int-rain}, to the contour tree of the initial input $\MM$.
-For clarity, we use $\NN$ to denote the input here. Note that in \Def{int-criticalJoin},
-the join tree is defined purely combinatorially in terms of the 1-skeleton (the underlying graph) of $\NN$.
-
-The join tree $\jc(\NN)$ is a rooted tree with the dominant minimum at the root, and we direct edges downwards (towards the root). So it makes sense to talk
-of comparable vs incomparable vertices.
-We arrive at the main challenge: how to sort only the comparable critical points, without
-constructing the join tree? The join tree algorithm of~\cite{csa-cctad-00} is a typical event-based
-computational geometry algorithm. We have to step away from this viewpoint to avoid the global sort.
-
+
+\myparagraph{Join trees from painted mountaintops:} Our main result is a faster algorithm for join trees.
+Denote the input as a terrain $\NN$.
 The key idea is \emph{paint spilling}. Start with each maximum having a large can of paint, with distinct
@@
 Furthermore, our paints do not mix, so each edge receives a unique color, decided by the first
-paint to reach it.  In the following, $[n]$ denotes the set $\{1, \ldots, n\}$, for any natural number $n$.
-
-\begin{definition} \label{def:paint1} \InSoCGVer{(Restatement of \Def{paint2}.)} Let the 1-skeleton of $\NN$ have edge set $E$ and maxima $X$.
-A  \emph{painting} of $\NN$ is a map $\chi:X \cup E \to [|X|]$ with the following property. 
- Consider an edge $e$. There exists a descending path from some maximum $x$ to $e$
-	consisting of edges in $E$, such that all edges along this path have the same color as $x$.
-
-An \emph{initial} painting has the additional property that the restriction $\chi:X \to [|X|]$ is a bijection.
-\end{definition}
-
-Note that a painting colors edges, and not vertices (except for maxima). Our definition also does not require the timing aspect
-of iterating over colors, though that is one way of painting $\NN$. We begin with an initial painting, since all maximum
-colors are distinct. A few comments on paint vs water. The interface between two regions of different color
-is \emph{not} a contour, and so we cannot apply the divide-and-conquer
-approach of contour surgery. On the other hand, painting does not cut $\NN$, so there is no increase in complexity.
-Clearly, an initial painting can be constructed in $O(N)$ time. This is the tradeoff between water and paint.
-Water allows for an easy divide-and-conquer, at the cost of more complexity in the input. For an extremum
-dominant input, using water to divide the input $\NN$ raises the complexity too much.
+paint to reach it.  We begin with an initial painting, since all maximum
+colors are distinct.
 
@@
 Both mounds are colored completely blue or red, while $v$ is touched by both colors.
-So this indicates that $v$ joins the blue maximum and red maximum in $\jc(\MM)$.
-
-This is precisely how we hope to exploit the information in the painting. We prove later that
-when some join $v$ has all incident edges with exactly two colors, the corresponding maxima (of those colors) are exactly
-the children of $v$ in $\jc(\MM)$. To proceed further, we ``merge" the colors red and blue into a new color, purple. In other words,
-we replace all red and blue edges by purple edges. This indicates that the red and blue maxima have been
-handled. Imagine flattening the red and blue mounds until reaching $v$, so that the former join $v$
-is now a new maximum, from which purple paint is poured. In terms of $\jc(\MM)$, this
+So this indicates that $v$ joins the blue maximum and red maximum in $\jc(\MM)$, and is a critical point.
+
+To process $v$, we ``merge" the colors red and blue into a new color, purple. 
+In terms of $\jc(\MM)$, this
 is equivalent to removing leaves $\ell_1$ and $\ell_2$, and making $v$ a new leaf.
 Alternatively, $\jc(\MM)$ has been constructed up to $v$, and it remains to determine $v$'s parent.
-The merging of the colors is not explicitly performed as that would be too expensive; instead we maintain a union-find
-data structure for that.
 
@@
 merge red and blue, until yellow and blue merge. Naturally, we use priority queues to handle this issue.
-We know that $u$ must also be touched by blue. So all critical vertices touched by blue
-are put into a priority queue keyed by height, and vertices are handled in that order.
 
@@
 We stress that the feasibility of the entire approach hinges on the use of such an efficient heap structure.
+There are numerous details to handle to prove the final run time bound, but this should provide the reader
+some intuition behind our algorithm.
+}
+
+\newcommand{\intuitionSection}{
+\section{A tour of the new contour tree algorithm} \label{sec:approach}
+\InNotSoCGVer{
+Our final algorithm is quite technical and has numerous moving parts. However, for the $d=2$ case, 
+where the input is just a triangulated terrain, the main ideas of the parts of the algorithm 
+can be explained clearly.  Therefore, here we first provide a high level view of the entire result.
+}
+\InSoCGVer{
+For the $d=2$ case, where the input is just a triangulated terrain, the main ideas of the parts of the algorithm 
+can be explained clearly.  Therefore, here we provide a high level view of the entire result.
+}
+
+In the interest of presentation, the definitions and theorem statements in this section will
+slightly differ from those in the main body. They may also differ from the original definitions proposed
+in earlier work.
+
+\myparagraph{Do not globally sort:} The starting point for this work is \Fig{relative}. We have two terrains
+with exactly the same contour tree, but different orderings of (heights of) the critical points.
+Turning it around, we cannot deduce the full height ordering of critical points from the contour tree.
+Sorting all critical points is computationally unnecessary for constructing the contour tree.
+In \Fig{sorting}, the contour tree consists of two balanced binary trees, one of the joins,
+another of the splits. Again, it is not necessary to know the relative ordering between the mounds
+on the left (or among the depressions on the right) to compute the contour tree. Yet some ordering
+information is necessary: on the left, the little valleys are higher than the big central valley,
+and this is reflected in the contour tree. Leaf paths in the contour tree have points
+in sorted order, but incomparable points in the tree are unconstrained.
+How do we sort exactly what is required, without knowing the contour tree in advance?
+
+\subsection{Breaking $\MM$ into simpler pieces} \label{sec:break}
+Let us begin with the algorithm of Carr, Snoeyink, and Axen~\cite{csa-cctad-00}. The key insight is to build
+two different trees, called the join and split trees, and then merge them together into the contour tree.
+Consider sweeping down via the hyperplane $x_{d+1} = h$ and taking the \emph{superlevel} sets. These 
+are the connected components of the portion of $\MM$ above height $h$. For a terrain, the superlevel
+sets are a collection of ``mounds". As we sweep downwards, these mounds keep joining each other,
+until finally, we end up with all of $\MM$. The join tree tracks exactly these events. 
+Formally, let $\MM^+_v$ denote the simplicial complex induced on the subset of 
+vertices which are higher than $v$.
+
+\begin{definition} 
+\label{def:int-criticalJoin}
+The \emph{critical join tree} $\jc(\MM)$ is built on the set $\cV$ of all critical points.\footnote{%
+In Definition~\ref{def:criticalJoin}, we define the critical join tree on a certain subset of $\cV$.}
+The directed edge $(u,v)$ is present when $u$ is the smallest valued vertex in $\cV$ in a connected component of $\MM^+_v$
+and $v$ is adjacent (in $\MM$) to a vertex in this component. 
+\end{definition}
+
+Abusing notation, for now we will just call this the join tree. (In \Sec{extreme} \InSoCGVer{and \Sec{jstedm}} we carefully handle the distinction 
+between the join tree and the critical join tree.)
+Refer to \Fig{sorting} for the join tree of a terrain. Note that nothing happens at splits, but these
+are still put as vertices in the join tree. They simply form a long path. The split tree
+is obtained by simply inverting this procedure, sweeping upwards and tracking sublevel sets.
+
+A major insight of~\cite{csa-cctad-00} is an ingeniously simple linear time procedure to construct
+the contour tree from the join and split trees. So the bottleneck is computing these trees. Observe
+in \Fig{sorting} that the split vertices form a long path in the join tree (and vice versa). Therefore, constructing
+these trees forces a global sort of the splits, an unnecessary computation for the contour tree. Unfortunately, in general (i.e.\ unlike \Fig{sorting}) 
+the heights of joins and splits may be interleaved in a complex manner, and hence the final merging of~\cite{csa-cctad-00}
+to get the contour tree requires having the split vertices in the join tree. Without
+this, it is not clear how to get a consistent view of both joins and splits, required for the contour tree.
+
+Our aim is to break $\MM$ into smaller pieces, where this unnecessary computation can be avoided.
+
+\myparagraph{Contour surgery:} We first need a divide-and-conquer lemma. Any contour $\phi$ can be associated
+with an edge $e$ of the contour tree. Suppose we ``cut" $\MM$ along this contour. We prove that $\MM$
+is split into two disconnected pieces, such the contour trees of these pieces is obtained
+by simply cutting $e$ in $\reeb(\MM)$. Alternatively, the contour trees of these pieces can
+be glued together to get $\reeb(\MM)$. This is not particularly surprising, and is fairly easy to prove
+with the right definitions. The idea of loop surgery has been used to reduce Reeb graphs to contour trees~\cite{TiGySi09,DoNa13}.
+Nonetheless, our theorem appears to be new and works for all dimensions.
+
+\begin{figure}[h]\centering
+    \includegraphics[width=.172\linewidth]{figs/tent2}%
+    \hspace{1.2in}
+    \includegraphics[width=.172\linewidth]{figs/tent1}
+    \caption{On left, downward rain spilling only (each shade of gray represents a piece created by each different spilling), producing a grid.  
+    Note we are assuming raining was done in sorted order of the maxima (i.e.\ lowest to highest).  
+    On right, flipping the direction of rain spilling.}
+    \label{fig:order}
+\end{figure}
+
+\myparagraph{Cutting $\MM$ into extremum dominant pieces:} We define a simplicial complex
+endowed with a height to be \emph{minimum dominant} if there exists only a single minimum.
+(Our real definition is more complicated, and involves simplicial complexes
+that allow additional ``trivial'' minima.)
+In such a complex, there exists a non-ascending path from any point to this unique minimum.
+Analogously, we can define maximum dominant complexes,
+and both are collectively called extremum dominant.
+
+We will cut $\MM$ into disjoint extremum dominant pieces, in linear time.
+One way to think of our procedure is a meteorological analogy. Take an arbitrary maximum $x$,
+and imagine torrential rain at the maximum. The water flows down, wetting any point that has a non-ascending path 
+from $x$. We end up with two portions, the wet part of $\MM$ and the dry part. This is similar
+to \emph{watershed} algorithms used for image segmentation~\cite{RoMe00}. The wet part is obviously connected,
+while there may be numerous disconnected dry parts. The interface between the dry and wet parts
+is a set of contours\footnote{Technically, they are not contours, but rather the limits of sequences of contours.}, 
+given by the ``water line". The wet part is clearly maximum dominant, since all
+wet points have a non-descending path to $x$. So we can simply cut along the interface
+contours to get the wet maximum dominant piece $\MM'$. By our contour surgery theorem,
+we are left with a set of disconnected dry parts, and we can recur this procedure on them.
+
+But here's the catch. Every time we cut $\MM$ along a contour, we potentially increase
+the complexity of $\MM$. Water flows in the interior of faces, and the interface will naturally
+cut some faces. Each cut introduces new vertices, and a poor choice of repeated raining leads
+to a large increase in complexity. Consider the left of \Fig{order}. Each raining produces a single
+wet and dry piece, and each cut introduces many new vertices. If we wanted to partition this terrain into 
+maximum dominant simplicial complexes, the final complexity would be forbiddingly large.
+
+A simple trick saves the day. Unlike reality, we can choose rain to flow solely downwards or solely upwards.
+Apply the procedure above to get a single wet maximum dominant $\MM'$ and a set of dry pieces.
+Observe that a single dry piece $\NN$ is boundary critical with the newly introduced boundary $\phi$ (the wet-dry interface) behaving as a minimum.
+So we can rain upwards from this minimum, and get a \emph{minimum dominant} portion $\NN'$.
+This ensures that the new interface (after applying the procedure on $\NN$) does not
+cut any face previously cut by $\phi$.
+For each of the new dry pieces, the newly introduced boundary is now a maximum. So we rain downwards from there.
+More formally, we alternate between raining upwards and downwards as we go down the recursion tree.
+We can prove that an original face of $\MM$ is cut at most once, so the final complexity can be bounded.
+In \Fig{order}, regardless of the choice of the starting maximum,  this procedure would yield (at most) two pieces, 
+one maximum dominant, and one minimum dominant.
+
+Using the contour surgery theorem previously discussed, we can build the contour tree of $\MM$
+from the contour trees of the various pieces created. All in all, we prove the following theorem.
+
+\begin{theorem} \label{thm:int-rain} There is an $O(N)$ time procedure that cuts $\MM$ into extremum dominant simplicial complexes
+$\MM_1, \MM_2, \ldots$. Furthermore, given the set of contour trees $\{\cC(\MM_i)\}$, $\cC(\MM)$ can be constructed in $O(N)$ time.
+\end{theorem}
+
+\myparagraph{Extremum dominance simplifies contour trees:} We will focus on minimum dominant simplicial complexes $\MM$. By \Thm{int-rain}, it suffices
+to design an algorithm for contour trees on such inputs. 
+For the $d=2$ case, it helps to visualize such an input as a terrain with no minima,
+except at a unique boundary face (think of a large boundary triangle that is the boundary).
+All the saddles in such a terrain are necessarily joins, and there can be no splits.
+In \Fig{sorting}, the portion on the left is minimum dominant in exactly this fashion, albeit in one dimension lower.
+%Water poured at any critical point must flow to the boundary, and the entire wet portion is connected.
+More formally, $\MM^-_v$ is connected for all $v$, so there are no splits.
+
+We can prove that the split tree is just a path, and the contour tree is exactly
+the join tree. The formal argument is a little involved, and we employ the merging procedure
+of~\cite{csa-cctad-00} to get a proof. We demonstrate that the merging procedure will
+actually just output the join tree, so we do not need to actually compute the split tree.
+(The real definition of minimum dominant is a little more complicated,
+so the contour tree is more than just the join tree. But computationally, it suffices
+to construct the join tree.)
+
+We stress the importance of this step for our approach. Given the algorithm of~\cite{csa-cctad-00},
+one may think that it suffices to design faster algorithms for join trees. But this
+cannot give the sort of optimality we hope for. Again, consider \Fig{sorting}. Any algorithm
+to construct the true join tree must construct the path of splits, which implies
+sorting all of them. It is absolutely necessary to cut $\MM$ into pieces where the cost
+of building the join tree can be related to that of building $\cC(\MM)$.
+
+\subsection{Join trees from painted mountaintops} \label{sec:join-paint}
+
+Arguably, everything up to this point is a preamble for the main result: a faster
+algorithm for join trees. Our algorithm does not require the initial input to be extremum
+dominant. This is only required to relate the join trees, of the resulting subcomplexes of Theorem~\ref{thm:int-rain}, to the contour tree of the initial input $\MM$.
+For clarity, we use $\NN$ to denote the input here. Note that in \Def{int-criticalJoin},
+the join tree is defined purely combinatorially in terms of the 1-skeleton (the underlying graph) of $\NN$.
+
+The join tree $\jc(\NN)$ is a rooted tree with the dominant minimum at the root, and we direct edges downwards (towards the root). So it makes sense to talk
+of comparable vs incomparable vertices.
+We arrive at the main challenge: how to sort only the comparable critical points, without
+constructing the join tree? The join tree algorithm of~\cite{csa-cctad-00} is a typical event-based
+computational geometry algorithm. We have to step away from this viewpoint to avoid the global sort.
+
+The key idea is \emph{paint spilling}. Start with each maximum having a large can of paint, with distinct
+colors for each maximum. In arbitrary order, we spill paint from each maximum, wait till it flows down,
+then spill from the next, etc. Paint is viscous, and only flows down edges. \emph{It does not
+paint the interior of higher dimensional faces.} That is, this process is restricted to the 1-skeleton of $\NN$.
+Furthermore, our paints do not mix, so each edge receives a unique color, decided by the first
+paint to reach it.  In the following, $[n]$ denotes the set $\{1, \ldots, n\}$, for any natural number $n$.
+
+\begin{definition} \label{def:paint1} \InSoCGVer{(Restatement of \Def{paint2}.)} Let the 1-skeleton of $\NN$ have edge set $E$ and maxima $X$.
+A  \emph{painting} of $\NN$ is a map $\chi:X \cup E \to [|X|]$ with the following property. 
+ Consider an edge $e$. There exists a descending path from some maximum $x$ to $e$
+	consisting of edges in $E$, such that all edges along this path have the same color as $x$.
+
+An \emph{initial} painting has the additional property that the restriction $\chi:X \to [|X|]$ is a bijection.
+\end{definition}
+
+Note that a painting colors edges, and not vertices (except for maxima). Our definition also does not require the timing aspect
+of iterating over colors, though that is one way of painting $\NN$. We begin with an initial painting, since all maximum
+colors are distinct. A few comments on paint vs water. The interface between two regions of different color
+is \emph{not} a contour, and so we cannot apply the divide-and-conquer
+approach of contour surgery. On the other hand, painting does not cut $\NN$, so there is no increase in complexity.
+Clearly, an initial painting can be constructed in $O(N)$ time. This is the tradeoff between water and paint.
+Water allows for an easy divide-and-conquer, at the cost of more complexity in the input. For an extremum
+dominant input, using water to divide the input $\NN$ raises the complexity too much.
+
+Our algorithm incrementally builds $\jc(\MM)$ from the leaves (maxima) to the root (dominant minimum). 
+We say that vertex $v$ is \emph{touched} by color $c$, if there is a $c$-colored edge with lower endpoint $v$.
+Let us focus on an initial painting, where the colors have 1-1 correspondence with the maxima. Refer to the left part of \Fig{colors}.
+Consider two sibling leaves $\ell_1, \ell_2$ and their common parent $v$. The leaves are maxima,
+and $v$ is a join that ``merges" $\ell_1, \ell_2$. 
+In that case, there are ``mounds" corresponding to $\ell_1$ and $\ell_2$ that merge
+at a valley $v$. Suppose this was the entire input, and $\ell_1$ was colored blue and $\ell_2$ was colored red. 
+Both mounds are colored completely blue or red, while $v$ is touched by both colors.
+So this indicates that $v$ joins the blue maximum and red maximum in $\jc(\MM)$.
+
+This is precisely how we hope to exploit the information in the painting. We prove later that
+when some join $v$ has all incident edges with exactly two colors, the corresponding maxima (of those colors) are exactly
+the children of $v$ in $\jc(\MM)$. To proceed further, we ``merge" the colors red and blue into a new color, purple. In other words,
+we replace all red and blue edges by purple edges. This indicates that the red and blue maxima have been
+handled. Imagine flattening the red and blue mounds until reaching $v$, so that the former join $v$
+is now a new maximum, from which purple paint is poured. In terms of $\jc(\MM)$, this
+is equivalent to removing leaves $\ell_1$ and $\ell_2$, and making $v$ a new leaf.
+Alternatively, $\jc(\MM)$ has been constructed up to $v$, and it remains to determine $v$'s parent.
+The merging of the colors is not explicitly performed as that would be too expensive; instead we maintain a union-find
+data structure for that.
+
+Of course, things are more complicated when there are other mounds. There may be a yellow mound,
+corresponding to $\ell_3$ that joins with the blue mound higher up at some vertex $u$ (see the right part of \Fig{colors}). In $\jc(\MM)$, $\ell_1$ and $\ell_3$
+are sibling leaves, and $\ell_2$ is a sibling of some ancestor of these leaves. So we cannot
+merge red and blue, until yellow and blue merge. Naturally, we use priority queues to handle this issue.
+We know that $u$ must also be touched by blue. So all critical vertices touched by blue
+are put into a priority queue keyed by height, and vertices are handled in that order.
+
+\begin{figure}[h]\centering
+    \includegraphics[width=.25\linewidth]{figs/redBlue}%
+    \hspace{.07in}
+    \includegraphics[width=.25\linewidth]{figs/redBlue2}%
+    \hspace{.07in}
+    \includegraphics[width=.03\linewidth]{figs/redBlueTree}
+    \hspace{.09in}
+    \includegraphics[width=.3\linewidth]{figs/mound}%
+    \hspace{.07in}
+    \includegraphics[width=.065\linewidth]{figs/moundTree}
+    \caption{On the left, red and blue merge to make purple, followed by the contour tree with initial colors.  On the right, additional maxima and the resulting contour tree.}
+    \label{fig:colors}
+\end{figure}
+
+What happens when finally blue and red join at $v$? We merge the two colors, but now have
+blue and red queues of critical vertices, which also need to be merged to get a consistent painting. 
+This necessitates using a priority queue with efficient merges.
+Specifically, we use \emph{binomial heaps}~\cite{Vu78}, as they provide logarithmic time merges and deletes (though other similar heaps work). 
+We stress that the feasibility of the entire approach hinges on the use of such an efficient heap structure.
 
@@
 \InSoCGVer{
-\subsection{How to Read this paper}
-\Ben{New section, read carefully:}
-The final algorithm is quite technical and has numerous moving parts. However, for the $d=2$ case, 
-where the input is just a triangulated terrain, the main ideas of the parts of the algorithm 
-can be explained clearly. Therefore, in \Sec{approach} we provide a high level view of the entire result.
-We strongly encourage the reader to overview this section, as then the remainder of the paper can be understood 
-as a simple matter of implementing these ideas.
-
-Alternatively, the reader can view the full version of the paper \CiteFullVer{} which contains this high level overview and all other appendices 
-in the main body of the text.
-
-\myparagraph{Outline of the algorithm and the following sections.} 
-The high level idea is to partition $\MM$ into nicely behaved pieces, on which we know how to compute the contour tree optimally. 
-\Sec{surgery} describes our cutting procedure, and \Sec{rain} describes how to use this cutting procedure.
-Roughly speaking the resulting pieces are ``nice'' because the contour tree of each piece 
-looks like a rooted tree, i.e.\ the join tree of \cite{csa-cctad-00}, and the precise relation to join trees is the subject of \Sec{extreme}.
-\Sec{paint} then describes how to optimally compute the join tree (and hence the contour tree) 
-on each piece (see Appendix~\ref{sec:approach} for discussion on why the join tree should not be computed directly on the input complex). 
-Finally, \Sec{runTimeSummary} summarizes all sections relating to the running time, specifically \Sec{runTime} which upper bounds the 
-running time of the algorithm of \Sec{paint}, \Sec{pathDecomp} which uses a sophisticated charging argument to express this bound 
-in terms of path decompositions, and \Sec{lb} which provides matching running time lower bounds.
+
+\newintuitionSection
+
+% \subsection{Outline}
+% The first step is to partition $\MM$ into nicely behaved pieces, such that the contour tree of
+% each piece is a rooted tree.  This is done by cutting $\MM$ at certain contours. Of course, any such ``cut"
+% increases the complexity of the input, so care is required to keep the complexity down.
+% \Sec{surgery} describes our cutting procedure, and \Sec{rain} describes how to use this cutting procedure.
+% It turns out that a simpler construct, the \emph{join tree}~\cite{csa-cctad-00} of each piece is exactly
+% the contour tree. Thus, it suffices to 
+% 
+% Roughly speaking the resulting pieces are ``nice'' because the contour tree of each piece 
+% looks like a rooted tree, i.e.\ the join tree of \cite{csa-cctad-00}, and the precise relation to join trees is the subject of \Sec{extreme}.
+% \Sec{paint} then describes how to optimally compute the join tree (and hence the contour tree) 
+% on each piece (see Appendix~\ref{sec:approach} for discussion on why the join tree should not be computed directly on the input complex). 
+% Finally, \Sec{runTimeSummary} summarizes all sections relating to the running time, specifically \Sec{runTime} which upper bounds the 
+% running time of the algorithm of \Sec{paint}, \Sec{pathDecomp} which uses a sophisticated charging argument to express this bound 
+% in terms of path decompositions, and \Sec{lb} which provides matching running time lower bounds.
+% 
+% The final algorithm is quite technical and has numerous moving parts. However, for the $d=2$ case, 
+% where the input is just a triangulated terrain, the main ideas of the parts of the algorithm 
+% can be explained clearly. Therefore, in \Sec{approach} we provide a high level view of the entire result.
+% We strongly encourage the reader to overview this section, as then the remainder of the paper can be understood 
+% as a simple matter of implementing these ideas.
+% 
+% Alternatively, the reader can view the full version of the paper \CiteFullVer{} which contains this high level overview and all other appendices 
+% in the main body of the text.
+% 
 }

### 2015-12-02 paint-contour-tree.tex
done with pass

@@
 
-Roughly speaking, all existing algorithms require $\Omega(n\log n)$ time for all instances, since
-the first step is a global sort of the values of $f$ (technically, only the critical points must be sorted,
-but these can be $\Omega(n)$). 
+All existing algorithms begin with a global sort of at least all critical values of $f$,
+which can require (roughly) $\Omega(n\log n)$ time.
 Existing lower bounds show that there are pathological instances where this sort is required.
@@
 on the contour tree structure, and avoids the global sort for non-pathological inputs.
-Our results match all existing upper bounds, but also
-characterize the amount of sorting required to compute the contour tree.
-This leads to an almost linear (inverse Ackermann overhead) time algorithm for contour tree computation for many
-classes of natural instances. 
+If $C$ denotes the set of critical points in $\MM$, the running time is
+basically $O(\sum_{v \in C} \log \ell_v)$, where $\ell_v$ is the depth of $v$ in the contour tree.
+This matches al existing upper bounds, but is a significant improvement when the contour tree is short and fat.
+% Our results match all existing upper bounds, but also
+% characterize the amount of sorting required to compute the contour tree.
 Specifically, our approach ensures that any comparison made is between nodes in the same descending path in the contour tree,
@@
 proceeds various events occur: new contours are created or destroyed, contours merge into each other or
-split into new components, contours acquire or lose handles. More generally, these are the events where 
-the topology of the level set changes. The contour tree is a concise representation of all these events.
+split into new components, contours acquire or lose handles. 
+The contour tree is a concise representation of all these events.
 Throughout we follow the definition of contour trees from \cite{kobps-ctsssit-97} which includes all changes in topology.  
@@
 If $f$ is smooth, all points where the gradient of $f$ is zero are \emph{critical points}.
-These points are the ``events" where the contour topology changes and which form the vertices of the contour tree. 
+These points are the ``events" where the contour topology changes and form the vertices of the contour tree. 
 An edge of the contour tree connects two critical points if one event immediately ``follows"
 the other as the sweep plane makes its pass. (We provide formal definitions later.) 
-
 \Fig{relative} and \Fig{sorting} 
 show examples of simplicial complexes, with heights and their contour trees.
-Think of the contour tree edges as pointing downwards, so we can talk of the ``up-degree" and ``down-degree" of vertices.
-Leaves of down-degree $1$ are maxima, because a new contour is created at such points. Leaves of up-degree
-$1$ are minima, because a contour is destroyed there. Up-degree $2$ vertices are ``joins", where contours
-merge, and down-degree $2$ vertices are ``splits".  When $d=2$ (as in \Fig{relative}), contours are closed loops.
-In dimensions $d \geq 3$, we may have internal vertices of both up and down-degree $1$. 
-For example, these could be events where contours gain or lose handles.
-
-Suppose we have $f: \MM \to \RR$, where $\MM$ is a triangulated mesh with $n$ vertices, $N$ faces in total, and $t \leq n$ critical points. 
-We assume the input is $f:\MM \to \RR$ a linear interpolant over distinct valued vertices, where the contour tree
-$T$ has maximum degree $3$. The degree assumption simplifies the proof and presentation, and is commonly made~\cite{kobps-ctsssit-97}.
+Think of the contour tree edges as pointing downwards. Leaves are either maxima or minima,
+and internal nodes are either ``joins" or ``splits".
+
+% so we can talk of the ``up-degree" and ``down-degree" of vertices.
+% Leaves of down-degree $1$ are maxima, because a new contour is created at such points. Leaves of up-degree
+% $1$ are minima, because a contour is destroyed there. Up-degree $2$ vertices are ``joins", where contours
+% merge, and down-degree $2$ vertices are ``splits".  When $d=2$ (as in \Fig{relative}), contours are closed loops.
+% In dimensions $d \geq 3$, we may have internal vertices of both up and down-degree $1$. 
+% For example, these could be events where contours gain or lose handles.
+
+Consider $f: \MM \to \RR$, where $\MM$ is a triangulated mesh with $n$ vertices, $N$ faces in total, and $t \leq n$ critical points. 
+(We assume that $f:\MM \to \RR$ a linear interpolant over distinct valued vertices, where the contour tree
+$T$ has maximum degree $3$. The degree assumption simplifies the presentation, and is commonly made~\cite{kobps-ctsssit-97}.)
 % \Ben{It is standard to assume that $f$ has distinct values on the vertices of $\MM$ and the value at interior points is linearly interpolated.}
@@
 Common applications for contour trees involve turbulent combustion or noisy data, where the number of critical points is likely to be $\Omega(n)$.
-So we are still left with the bottleneck of sorting a large set of numbers. Unfortunately, there is a worst-case
+There is a worst-case
 lower bound of $\Omega(t\log t)$ by Chiang \etal \cite{cllr-sooscctmp-05}, based on a construction of Bajaj \etal \cite{BaKr+98}.
@@
 % \Sesh{Almost everything below is new/reordered. Please read through.}
-But this is only a worst-case bound. All previous algorithms begin by sorting (at least) the critical points, and
-hence must pay $\Omega(t\log t)$ for all instances. Can we beat this sorting bound for certain instances,
+All previous algorithms begin by sorting (at least) the critical points. Can we beat this sorting bound for certain instances,
 and can we characterize which inputs are hard? Intuitively, points that are incomparable in the contour tree
@@
 trees joined at their roots, and we do not need to entire sorted order of critical points
-to construct the contour tree. Is it possible to beat the $t\log t$ bound for these examples?
+to construct the contour tree.
 
@@
 the contour tree by $T$ with vertex set (the critical points) $C(T)$. There exists an algorithm to compute the contour tree $T$
-can be computed in $O(\sum_{v \in C(T)} \ell_v + t\alpha(t) + N)$ time. Moreover, this algorithm only
+can be computed in $O(\sum_{v \in C(T)} \log \ell_v + t\alpha(t) + N)$ time. Moreover, this algorithm only
 compares function values at pairs of points that are ancestor-descendant in $T$.
@@
 
-Note that \Thm{main-corr} is a direct corollary of this statement. This is strictly stronger,
-since for any balanced contour tree, the run time bound is $O(t\alpha(t) + N)$, and $O(t)$
+Note that \Thm{main-corr} is a direct corollary of this statement. For any $v$, $\ell_v$ is at most
+the length of the path in $P(T)$ that contains $v$.
+This bound is strictly stronger,
+since for any balanced contour tree, the run time bound of \Thm{main-alg} is $O(t\alpha(t) + N)$, and $O(t)$
 comparisons are made.
@@
 where the algorithm makes $C$ comparisons. Then \emph{any algorithm} must make roughly $C$
-comparisons in the worst-case over this set.
+comparisons in the worst-case over this set. (All further details are 
+in Appendix~\Sec{lb}.)
 
@@
 Contour trees were first used to study terrain maps by Boyell and Ruston, and Freeman and Morse~\cite{BoRu63,FrMo67}.
-They were used for isoline extraction in geometric data by van Krevald \etal \cite{kobps-ctsssit-97}, who provided
-the first formal algorithm. 
-Contour trees and related topological structures have been applied in analysis of fluid mixing, combustion simulations,
+Contour trees have been applied in analysis of fluid mixing, combustion simulations,
 and studying chemical systems~\cite{LaBe+06,BrWe+10,BeWe+11,BrWe+11,MaGr+11}. Carr's thesis~\cite{c-tmi-04} gives various
@@
 
-There are numerous algorithms for computing contour trees.
-An $O(N\log N)$ time algorithm for functions over 2D 
-meshes and an $O(N^2)$ algorithm for higher dimensions, was presented by van Kreveld \etal \cite{kobps-ctsssit-97}. 
+The first formal result was an $O(N\log N)$ time algorithm for functions over 2D 
+meshes and an $O(N^2)$ algorithm for higher dimensions, by van Kreveld \etal \cite{kobps-ctsssit-97}. 
 Tarasov and Vyalya \cite{tv-cct-98} improved the running time to $O(N\log N)$ for the 3D case.
 The influential paper of Carr \etal \cite{csa-cctad-00} improved the running time for all dimensions to $O(n\log n + N\alpha(N))$.
-This result forms the basis 
-of subsequent practical and theoretical work in computational topology applications. 
 Pascucci and Cole-McLaughlin \cite{pc-ectls-02} provided an $O(n+t\log n)$ time algorithm for 
 $3$-dimensional structured meshes. Chiang \etal \cite{cllr-sooscctmp-05} provide an unconditional $O(N+t\log t)$ algorithm.
-The latter result crucially uses monotone paths to improve the running time, and our algorithm repeatedly
-exploits numerous properties of monotone paths. Carr's thesis shows relationships between monotone paths
-in $\MM$ and its contour tree $\cC(\MM)$.
+% Carr's thesis shows relationships between monotone paths
+% in $\MM$ and its contour tree $\cC(\MM)$.
 
 Contour trees are a special case of Reeb graphs, a general topological representation for real-valued functions
-on any manifold. We refer the reader to Chapter 6 in~\cite{HaEd10} for more details. Algorithms for computing Reeb graphs
+on any manifold. Algorithms for computing Reeb graphs
 is an active topic of research~\cite{sk-crgacs-91,cehnp-lrbm-03,PaScBr07,DoNa09,HaWaWe10,Pa12}, where
@@
 
-We detail the basic definitions about contour trees. We follow the terminology of Chapter 6 of Carr's thesis \cite{c-tmi-04}.
+We detail the basic definitions about contour trees, following the terminology of Chapter 6 of Carr's thesis \cite{c-tmi-04}.
 All our assumptions and definitions
@@
 The input is a continuous piecewise-linear function $f:\MM \to \RR$, where $\MM$ is a simply connected and fully triangulated simplicial complex in $\RR^d$,
-except for specially designated \emph{boundary facets}. So $f$ is defined only on the vertices of $\MM$,
+except for specially designated \emph{boundary facets}. So $f$ is explicitly defined only on the vertices of $\MM$,
 and all other values are obtained by linear interpolation.
 
-We assume that $f$ has distinct values on all vertices, except for boundaries,
-as expressed in the following constraint, which intuitively requires that 
-all boundaries look like (local) horizontal cuts.
+We assume that the boundary values satisfy a special property. This is mainly for convenience
+in presentation.
 
@@
 
-We think of the dimension, $d$, as constant, and assume that $\MM$ is represented in a data structure that allows constant-time access to neighboring simplices
+We think of the dimension $d$, as constant, and assume that $\MM$ is represented in a data structure that allows constant-time access to neighboring simplices
 in $\MM$ (e.g.~\cite{BoMa12}). (This is analogous to a doubly connected edge list, but for higher dimensions.)
@@
 as $\delta, \eps \rightarrow 0^+$.
+
+A \emph{regular} point has both Morse up-degree and down-degree $1$. A \emph{maximum} has Morse up-degree $1$, while
+a \emph{minimum} has Morse up-degree $2$. A \emph{Morse Join} has Morse up-degree strictly greater than $1$,
+while a \emph{Morse Split} has Morse down-degree strictly greater than $1$. Non-regular points are called \emph{critical}.
 \end{definition}
 
-We categorize points depending on the local changes in topology.
-
-\begin{definition} \label{def:points} A point $x$ is categorized as follows.
-\begin{asparaitem}
-	\item \emph{Regular:} Both Morse up-degree and down-degrees are $1$.
-	\item \emph{Maximum:} Morse up-degree is $0$.
-	\item \emph{Minimum:} Morse down-degree is $0$.
-	\item \emph{Morse Join:} Morse up-degree is strictly greater than $1$.
-	\item \emph{Morse Split:} Morse down-degree is strictly greater than $1$.
-\end{asparaitem}
-Non-regular points are \emph{critical}. Morse joins and splits are \emph{saddles}.
-Maxima and minima are \emph{extrema}.
-\end{definition}
+% We categorize points depending on the local changes in topology.
+% 
+% \begin{definition} \label{def:points} A point $x$ is categorized as follows.
+% \begin{asparaitem}
+% 	\item \emph{Regular:} Both Morse up-degree and down-degrees are $1$.
+% 	\item \emph{Maximum:} Morse up-degree is $0$.
+% 	\item \emph{Minimum:} Morse down-degree is $0$.
+% 	\item \emph{Morse Join:} Morse up-degree is strictly greater than $1$.
+% 	\item \emph{Morse Split:} Morse down-degree is strictly greater than $1$.
+% \end{asparaitem}
+% Non-regular points are \emph{critical}. Morse joins and splits are \emph{saddles}.
+% Maxima and minima are \emph{extrema}.
+% \end{definition}
+%
 
@@
 \section{A tour of the new contour tree algorithm} \label{sec:approach}
-Our final algorithm is quite technical and has numerous moving parts. However, for the $d=2$ case, 
-where the input is just a triangulated terrain, the main ideas of the parts of the algorithm 
-can be explained clearly.
+We provide a short, high-level
+description of the main ideas.
+% However, for the $d=2$ case, 
+% where the input is just a triangulated terrain, the main ideas of the parts of the algorithm 
+% can be explained clearly.
 
 \myparagraph{Cutting $\MM$ into extremum dominant pieces:} We define a simplicial complex
-endowed with a height to be \emph{minimum dominant} if there exists only a single minimum.
-Analogously, we can define maximum dominant complexes, and both are collectively called extremum dominant.
-
+endowed with a height to be \emph{extremum dominant} if there exists only a single minimum, or only a single maximum.
 We first cut $\MM$ into disjoint extremum dominant pieces in linear time.
@@
 can be shown to be extremum dominant. We can cut along the interface of the wet and dry, remove
-the wet part, and recurse of the dry part. Each additional cut increases the complexity of the surface,
-so care is required to keep the total complexity linear. This is achieved by alternately ``raining"
-upwards and downwards.  We finally get a collection of extremum dominant pieces.
+the wet part, and recurse of the dry part. This is carefully done to ensure that the total complexity
+of the pieces is linear.
+% 
+% 
+% Each additional cut increases the complexity of the surface,
+% so care is required to keep the total complexity linear. This is achieved by alternately ``raining"
+% upwards and downwards.  We finally get a collection of extremum dominant pieces.
 We prove a simple contour surgery theorem that builds the contour tree of $\MM$
@@
 Using ideas from~\cite{csa-cctad-00}, we prove that the contour tree of an extremum dominant complex is (basically) just the \emph{join tree}.
-For the $d=2$ case, it helps to visualize such an input as a terrain with no minima,
-except at a unique boundary face (think of a large boundary triangle that is the boundary).
+% For the $d=2$ case, it helps to visualize such an input as a terrain with no minima,
+% except at a unique boundary face (think of a large boundary triangle that is the boundary).
 Consider sweeping down via the hyperplane $x_{d+1} = h$ and taking the connected components of the portion of $\MM$ above height $h$. For a terrain, these superlevel
@@
 \myparagraph{Join trees from painted mountaintops:} Our main result is a faster algorithm for join trees.
-Denote the input as a terrain $\NN$.
 The key idea is \emph{paint spilling}. Start with each maximum having a large can of paint, with distinct
 colors for each maximum. In arbitrary order, we spill paint from each maximum, wait till it flows down,
-then spill from the next, etc. Paint is viscous, and only flows down edges. \emph{It does not
-paint the interior of higher dimensional faces.} That is, this process is restricted to the 1-skeleton of $\NN$.
-Furthermore, our paints do not mix, so each edge receives a unique color, decided by the first
-paint to reach it.  We begin with an initial painting, since all maximum
-colors are distinct.
-
-Our algorithm incrementally builds $\jc(\MM)$ from the leaves (maxima) to the root (dominant minimum). 
-We say that vertex $v$ is \emph{touched} by color $c$, if there is a $c$-colored edge with lower endpoint $v$.
-Let us focus on an initial painting, where the colors have 1-1 correspondence with the maxima. Refer to the left part of \Fig{colors}.
+then spill from the next, etc. Paint is viscous, and only flows down edges. Furthermore, our paints do not mix, so each edge receives a unique color, decided by the first
+paint to reach it.
+
+Our algorithm incrementally builds the join tree from the leaves (maxima) to the root. 
+Refer to the left part of \Fig{colors}.
 Consider two sibling leaves $\ell_1, \ell_2$ and their common parent $v$. The leaves are maxima,
@@
 Both mounds are colored completely blue or red, while $v$ is touched by both colors.
-So this indicates that $v$ joins the blue maximum and red maximum in $\jc(\MM)$, and is a critical point.
-
+So this indicates that $v$ joins the blue maximum and red maximum in the join tree and is a critical point.
 To process $v$, we ``merge" the colors red and blue into a new color, purple. 
-In terms of $\jc(\MM)$, this
+In terms of join tree, this
 is equivalent to removing leaves $\ell_1$ and $\ell_2$, and making $v$ a new leaf.
-Alternatively, $\jc(\MM)$ has been constructed up to $v$, and it remains to determine $v$'s parent.
 
 Of course, things are more complicated when there are other mounds. There may be a yellow mound,
-corresponding to $\ell_3$ that joins with the blue mound higher up at some vertex $u$ (see the right part of \Fig{colors}). In $\jc(\MM)$, $\ell_1$ and $\ell_3$
+corresponding to $\ell_3$ that joins with the blue mound higher up at some vertex $u$ (see the right part of \Fig{colors}). In the join tree, $\ell_1$ and $\ell_3$
 are sibling leaves, and $\ell_2$ is a sibling of some ancestor of these leaves. So we cannot
-merge red and blue, until yellow and blue merge. Naturally, we use priority queues to handle this issue.
+merge red and blue, until yellow and blue merge. 
+A critical insight is the use of \emph{binomial heaps}~\cite{Vu78} to handle all the priorities and the merging.
+This ensures that all comparisons are only made between points that are ancestor-descendant in the join tree.
+There are numerous details to handle to prove the final run time bound, but this should provide the reader
+some intuition behind our algorithm.
 
@@
 
-What happens when finally blue and red join at $v$? We merge the two colors, but now have
-blue and red queues of critical vertices, which also need to be merged to get a consistent painting. 
-This necessitates using a priority queue with efficient merges.
-Specifically, we use \emph{binomial heaps}~\cite{Vu78}, as they provide logarithmic time merges and deletes (though other similar heaps work). 
-We stress that the feasibility of the entire approach hinges on the use of such an efficient heap structure.
-There are numerous details to handle to prove the final run time bound, but this should provide the reader
-some intuition behind our algorithm.
+% What happens when finally blue and red join at $v$? We merge the two colors, but now have
+% blue and red queues of critical vertices, which also need to be merged to get a consistent painting. 
+% This necessitates using a priority queue with efficient merges.
+% Specifically, we use \emph{binomial heaps}~\cite{Vu78}, as they provide logarithmic time merges and deletes (though other similar heaps work). 
+% We stress that the feasibility of the entire approach hinges on the use of such an efficient heap structure.
 }
@@
 
-The following claim gives a description of the interface. While its meaning is quite intuitive, the proof is tedious.
+The following claim gives a description of the interface.
 
@@
 Let $\lift(\phi)$ be the unique contour through the regular point $z$. Note that $\lift(\phi)$ is dry.
+The following claim follows directly from \Thm{jordan}.
 
@@
 
+\InNotSoCGVer{
 \begin{proof} By \Thm{jordan}, $\cut(\MM,\lift(\phi))$ results in two disjoint simplicial complexes. Let $\NN$ be the complex containing
@@
 \end{proof}
+}
 
@@
 dominant complexes. It takes an additional input of a maximum $x$. To initialize,
-we begin with $\NN$ set to $\MM$ and $x$ as an arbitrary maximum. One of the critical aspects
-of the procedure is that the rain alternately flows downwards and upwards, 
-since otherwise faces may be cut a super constant number of times (see \Fig{order}\InSoCGVer{ in \Sec{approach}}).
-In other words, when
-we start, rain flows downwards. In each recursive call, the direction of rain is switched to the 
-opposite direction. While one can implement $\rain$ this way, it is conceptually easier
-to think of \emph{inverting} a complex $\NN'$ when a recursive call is made. Inversion
-is easily achieved by just negating the height values. (From a running time standpoint, it suffices
-to maintain a single bit associated with $\NN'$ that determines whether heights are inverted or not.)
+we begin with $\NN$ set to $\MM$ and $x$ as an arbitrary maximum. 
+% One of the critical aspects
+% of the procedure is that the rain alternately flows downwards and upwards, 
+% since otherwise faces may be cut a super constant number of times (see \Fig{order}\InSoCGVer{ in \Sec{approach}}).
+When
+we start, rain flows downwards. In each recursive call, the direction of rain is \emph{switched} to the 
+opposite direction. This is crucial to ensure a linear running time. 
+The switching is easily implemented by inverting a complex $\NN'$, achieved by negating the height values.
+% While one can implement $\rain$ this way, it is conceptually easier
+% to think of \emph{inverting} a complex $\NN'$ when a recursive call is made. Inversion
+% is easily achieved by just negating the height values. (From a running time standpoint, it suffices
+% to maintain a single bit associated with $\NN'$ that determines whether heights are inverted or not.)
 We can now let rain flow downwards, as it usually does in our world.
@@
 For convenience, denote the total output of $\rain(x,\MM)$ by $\MM_1, \MM_2, \ldots, \MM_r$.
-We first argue the correctness of $\rain$.
+% We first argue the correctness of $\rain$.
 
@@
 
+The following are direct consequences of \Lem{new-verts} and the surgery procedure.
+
 \begin{theorem} \label{thm:rain-time} The total running time of $\rain(x,\MM)$ is $O(|\MM|)$.
@@
 
+\InNotSoCGVer{
 \begin{proof} The only non-trivial operations performed are $\wet$ and $\cut$.  
@@
 
+\InNotSoCGVer{
 \begin{proof} Consider the tree of recursive calls in $\rain(x,\MM)$, with each node
@@
 numerous connected components.
-
-\Ben{Read remainder of this section carefully, it has been cut up, and parts rewritten to fit.}
 
@@
 
-A painting colors edges, not vertices. We associate certain
-sets of colors with vertices.
+% A painting colors edges, not vertices. We associate certain
+% sets of colors with vertices.
 
@@
 
-Non-maximum $v$ participate in at least $1$ and at most $2$ up-stars (the latter iff
-$v$ a join).
+% Non-maximum $v$ participate in at least $1$ and at most $2$ up-stars (the latter iff
+% $v$ a join).
 
@@
 
-We discuss all the initializations in the next section.
-
-\medskip
 \noindent
@@
 
-We prove a series of claims that will lead to the correctness proof.
+% We prove a series of claims that will lead to the correctness proof.
 
@@
 \subsection{Upper and Lower bounds for Running Time}\label{sec:runTimeSummary}
-\Ben{This section is new, and only appears in the socg version}
 The full running time analysis can be found in \Sec{runTime}, \Sec{pathDecomp}, and \Sec{lb}, or in the full version \CiteFullVer.  
@@
 In \Sec{runTime}, we give a tight bound on the running time of $\build(\MM)$, as a function of the choices made by the subroutine $\init(\MM)$.
-To state this bound we require the following technical definitions (which are repeated in \Sec{runTime}).
+% To state this bound we require the following technical definitions (which are repeated in \Sec{runTime}).
 
@@
 
-The above is a purely combinatorial statement on binary trees, and may find applications in other charging arguments.  
-However, for the specific when $T=\jc(\MM)$, plugging this theorem into the above lemma, and applying this lemma 
-to each subcomplex piece output by $\rain$, yields our main result, Theorem~\ref{thm:main-alg}.
-
-The path decomposition $\pmax(T)$ is a function of the choices made by the linear time $\init(\MM)$.  
-One cannot control which path decomposition $\init$ determines, as sorting can be reduced to finding the optimal path decomposition.  
-Specifically, in \Sec{lb} we exploit this connection to prove the following matching lower bound for our approach.
-In the following a rain consistent path decomposition is any decomposition which is the union of paths from 
-decompositions on each subcomplex output by $\rain$ (see Definition~\ref{def:path2} for a formal definition).
+% The above is a purely combinatorial statement on binary trees, and may find applications in other charging arguments.  
+% However, for the specific when $T=\jc(\MM)$, plugging this theorem into the above lemma, and applying this lemma 
+% to each subcomplex piece output by $\rain$, yields our main result, Theorem~\ref{thm:main-alg}.
+% 
+% The path decomposition $\pmax(T)$ is a function of the choices made by the linear time $\init(\MM)$.  
+% One cannot control which path decomposition $\init$ determines, as sorting can be reduced to finding the optimal path decomposition.  
+% Specifically, in \Sec{lb} we exploit this connection to prove the following matching lower bound for our approach.
+% In the following a rain consistent path decomposition is any decomposition which is the union of paths from 
+% decompositions on each subcomplex output by $\rain$ (see Definition~\ref{def:path2} for a formal definition).
 
@@
 
-\begin{theorem} 
-Consider any rain consistent path decomposition $P$. There exists a family $\bF_P$ of terrains ($d=2$) with the following properties.
-Any contour tree algorithm makes $\Omega(\sum_{p \in P} |p|\log |p|)$ comparisons in the worst case over $\bF_P$.
-Furthermore, for any terrain in $\bF_P$, our algorithm makes $O(\sum_{p \in P} |p|\log |p|)$ comparisons.
-\end{theorem}
+% \begin{theorem} 
+% Consider any rain consistent path decomposition $P$. There exists a family $\bF_P$ of terrains ($d=2$) with the following properties.
+% Any contour tree algorithm makes $\Omega(\sum_{p \in P} |p|\log |p|)$ comparisons in the worst case over $\bF_P$.
+% Furthermore, for any terrain in $\bF_P$, our algorithm makes $O(\sum_{p \in P} |p|\log |p|)$ comparisons.
+% \end{theorem}
 

### 2015-12-02 paint-contour-tree.tex
various changes, fig corrected

@@
 %%%
-%\def\HideComments{TRUE}%
+\def\HideComments{TRUE}%
 %%%
@@
 the contour tree by $T$ with vertex set (the critical points) $C(T)$. There exists an algorithm to compute the contour tree $T$
-can be computed in $O(\sum_{v \in C(T)} \log \ell_v + t\alpha(t) + N)$ time. Moreover, this algorithm only
+in $O(\sum_{v \in C(T)} \log \ell_v + t\alpha(t) + N)$ time. Moreover, this algorithm only
 compares function values at pairs of points that are ancestor-descendant in $T$.
@@
 There exists some absolute constant $\alpha$ such that the following holds.
-For sufficiently large $t$ and $C$: consider the set $\bF_C$ of terrains with $t$ critical points such that
+For sufficiently large $t$ and any $C\in [t, t\log t]$, consider the set $\bF_C$ of terrains with $t$ critical points such that
 the number of comparisons made by the algorithm of \Thm{main-alg} on these terrains is in $[C,\alpha C]$.
@@
 
-A \emph{regular} point has both Morse up-degree and down-degree $1$. A \emph{maximum} has Morse up-degree $1$, while
-a \emph{minimum} has Morse up-degree $2$. A \emph{Morse Join} has Morse up-degree strictly greater than $1$,
+A \emph{regular} point has both Morse up-degree and down-degree $1$. A \emph{maximum} has Morse up-degree $0$, while
+a \emph{minimum} has Morse down-degree $0$. A \emph{Morse Join} has Morse up-degree strictly greater than $1$,
 while a \emph{Morse Split} has Morse down-degree strictly greater than $1$. Non-regular points are called \emph{critical}.
@@
 
-This equivalence relation gives a set \emph{contour classes}. 
+This equivalence relation gives a set of \emph{contour classes}. 
 %By continuity, a regular point is present on a single contour,
@@
 %
-\Ben{Similar to \cite{kobps-ctsssit-97} (among others), multi-saddles are treated as a set of ordinary saddles, which can be realized 
+Similar to \cite{kobps-ctsssit-97} (among others), multi-saddles are treated as a set of ordinary saddles, which can be realized 
 via vertex unfolding (which can increase surface complexity if multi-saddle degrees are allowed to be super-constant).
-Therefore, to simplify the presentation, for the remainder of the paper up and down-degrees are at most $2$, and the total degree is at most $3$.}
+Therefore, to simplify the presentation, for the remainder of the paper up and down-degrees are at most $2$, and total degree is at most $3$.
 %
@@
 can be shown to be extremum dominant. We can cut along the interface of the wet and dry, remove
-the wet part, and recurse of the dry part. This is carefully done to ensure that the total complexity
+the wet part, and recurse on the dry parts. This is carefully done to ensure that the total complexity
 of the pieces is linear.
@@
 To process $v$, we ``merge" the colors red and blue into a new color, purple. 
-In terms of join tree, this
+In terms of the join tree, this
 is equivalent to removing leaves $\ell_1$ and $\ell_2$, and making $v$ a new leaf.
@@
 is denoted by $\wet(x,\MM)$ (which in turn is represented as a simplicial complex). A point $z$ is at the \emph{interface} of $\wet(x,\MM)$ if every neighborhood of $z$
-has non-trivial intersection with $\wet(x,\MM)$ \Ben{(i.e.\ the intersection is neither empty nor the entire neighborhood)}.
+has non-trivial intersection with $\wet(x,\MM)$ (i.e.\ the intersection is neither empty nor the entire neighborhood).
 \end{definition}
@@
 
-The following are direct consequences of \Lem{new-verts} and the surgery procedure.
+The following are direct consequences of \Lem{new-verts} and the $\surgery$ procedure.
 
@@
 The directed edge $(u,v)$ is present when $u$ is the smallest valued vertex in $V'$ in a connected component of $\MM^+_v$
-and $v$ \Ben{is adjacent (in $\MM$) to a vertex} in this component. 
+and $v$ is adjacent (in $\MM$) to a vertex in this component. 
 \end{definition}
@@
 We first prove a lower bound for join trees, and then generalize to contour trees.
+Note that form of the theorem statements in this section differ from \Thm{main-lb}, 
+as they are stated directly in terms of path decompositions.  \Thm{main-lb} 
+is an immediate corollary of the final theorem of this section \Thm{path-main-lb}.
 
@@
 
-\begin{theorem} 
+\begin{theorem}\label{thm:path-main-lb}
 Consider any rain consistent path decomposition $P$. There exists a family $\bF_P$ of terrains ($d=2$) with the following properties.
@@
 
-\intuitionSection
+%\intuitionSection
 

### 2015-12-03 paint-contour-tree.tex
7.4 rewrite, and various small updates

@@
 
-\subsection{A refined bound with optimality properties}
-
-\Thm{main-corr} is a direct corollary of a more cumbersome theorem.
+\subsection{A refined bound with optimality properties}\label{sec:more-refined}
+
+\Thm{main-corr} is a direct corollary of a stronger but more cumbersome theorem.
 
@@
 %Our algorithm does not need this, but it simplifies the proof and presentation.} 
-There is a deterministic algorithm to compute $T$ whose running time is $O(\sum_{p \in P(T)} |p|\log |p| + t\alpha(t) + N)$,
+There is a deterministic algorithm to compute the contour tree, $T$, whose running time is $O(\sum_{p \in P(T)} |p|\log |p| + t\alpha(t) + N)$,
 where $P(T)$ is a specific path decomposition (constructed implicitly by the algorithm).
@@
 \section{A tour of the new contour tree algorithm} \label{sec:approach}
-We provide a short, high-level
-description of the main ideas.
+We provide a short, high-level description of the main ideas of subsequent sections. 
+A longer more detailed version of this high-level description can be found in the full version \CiteFullVer.
 % However, for the $d=2$ case, 
@@
 endowed with a height to be \emph{extremum dominant} if there exists only a single minimum, or only a single maximum.
+(When formally defined in \Sec{rain}, extremum dominant complexes will allow additional trivial minima or maxima which are a small complication resulting from the following cutting procedure.)
 We first cut $\MM$ into disjoint extremum dominant pieces in linear time.
@@
 
-Using ideas from~\cite{csa-cctad-00}, we prove that the contour tree of an extremum dominant complex is (basically) just the \emph{join tree}.
+Using ideas from~\cite{csa-cctad-00}, we prove that the contour tree of an extremum dominant complex is (basically) just the \emph{join tree}, 
+which unlike the contour tree tracks superlevel sets rather than level sets.
 % For the $d=2$ case, it helps to visualize such an input as a terrain with no minima,
 % except at a unique boundary face (think of a large boundary triangle that is the boundary).
-Consider sweeping down via the hyperplane $x_{d+1} = h$ and taking the connected components of the portion of $\MM$ above height $h$. For a terrain, these superlevel
+Consider sweeping down the hyperplane $x_{d+1} = h$ and taking the connected components of the portion of $\MM$ above height $h$. For a terrain, these superlevel
 sets are a collection of ``mounds". As we sweep downwards, these mounds keep joining each other,
@@
 two different trees, called the join and split trees, and then merge them together into the contour tree.
-Consider sweeping down via the hyperplane $x_{d+1} = h$ and taking the \emph{superlevel} sets. These 
+Consider sweeping down the hyperplane $x_{d+1} = h$ and taking the \emph{superlevel} sets. These 
 are the connected components of the portion of $\MM$ above height $h$. For a terrain, the superlevel
@@
 
-\newcommand{\algorithmRunTimeSection}{
-\InNotSoCGVer{\subsection{Running Time}}
-\InSoCGVer{\section{Running Time}}
+%\newcommand{\algorithmRunTimeSection}{
+%\InNotSoCGVer{\subsection{Running Time}}
+%\InSoCGVer{\section{Running Time}}
+\InNotSoCGVer{
+\subsection{Running Time}
 \label{sec:runTime}
@@
 \end{lemma}
-
+}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\newcommand{\proofofRunTime}{
 \begin{proof}
@@
 }
-
-\InNotSoCGVer{\algorithmRunTimeSection}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\InNotSoCGVer{\proofofRunTime}
+
+%\InNotSoCGVer{\algorithmRunTimeSection}
 %%%%%%%%%%%%%%%%%%%%
 \InSoCGVer{
-\subsection{Upper and Lower bounds for Running Time}\label{sec:runTimeSummary}
-The full running time analysis can be found in \Sec{runTime}, \Sec{pathDecomp}, and \Sec{lb}, or in the full version \CiteFullVer.  
-The main results of these sections are summarized here.
-%
-In \Sec{runTime}, we give a tight bound on the running time of $\build(\MM)$, as a function of the choices made by the subroutine $\init(\MM)$.
-% To state this bound we require the following technical definitions (which are repeated in \Sec{runTime}).
+\subsection{Upper Bounds for Running Time}\label{sec:runTimeSummary}
+
+% The full running time analysis can be found in \Sec{runTime}, \Sec{pathDecomp}, and \Sec{lb}, or in the full version \CiteFullVer.  
+% \Sec{runTime} is summarized here, which gives a bound immediately implying the running time of our main result \Thm{main-corr}.  
+% Using a significantly more sophisticated analysis involving path decompositions, this bound will also implies the more refined running time bounds mentioned in \Sec{more-refined} 
+% are the topics \Sec{pathDecomp} and \Sec{lb}.
+% %
+% %In \Sec{runTime}, we give a tight bound on the running time of $\build(\MM)$, as a function of the choices made by the subroutine $\init(\MM)$.
+
+The algorithm $\build(\MM)$ processes vertices in $\jc(\MM)$ one at a time.  The main processing cost  
+comes from priority queue operations.  The cost of these operations is a function of the size of the queue which is in turn a function of the choices made by the subroutine $\init(\MM)$.
+
+%The heap as defined above characterizes the priority queue size at the time a vertex gets processed and hence we get the following.
 
@@
 
-The algorithm $\build(\MM)$ processes vertices in $\jc(\MM)$ one at a time.  The main processing cost  
-comes from priority queue operations.  The heap as defined above characterizes the  
-priority queue size at the time a vertex gets processed and hence we get the following.
-
-\begin{lemma}
+Given the above technical definition, the proof of the following lemma is straightforward, though long and so has been moved to \Sec{runTime}.
+
+\begin{lemma}\label{heapRunTimeBound}
 Let $\MM$ be a simplicial complex with $t$ critical points.  
@@
 
-While the above gives a full running time characterization, ideally the bound should be expressed as   
-a function of some structural property of $\jc(\MM)$.
-This is the focus of \Sec{pathDecomp}, where it is shown that for any binary tree $T$ 
-the $H_v$ sets (defined by any leaf assignment) naturally induces a path decomposition of $T$ (see \Def{path}), 
-denoted $\pmax(T)$ (see \Sec{pathDecomp} for details).
-
-% Specifically, the main result of \Sec{pathDecomp} is the following theorem, which together with the above 
-% bound in terms of heap size immediately implies our main result, Theorem~\ref{thm:main-alg} .
-
-\begin{theorem} $\sum_{v \in T} \log |H_v| = O(\sum_{p\in \pmax(T)} |p| \log |p|)$.
-\end{theorem}
+Our main result, \Thm{main-corr}, is an easy corollary of the above lemma. Specifically, consider a critical point $v$ of the initial input 
+complex.  By \Thm{jordan} this vertex appears in exactly one of the pieces output by $\rain$.  As in the \Thm{main-corr} statement, let $\ell_v$ 
+denote the length of the longest directed path passing through $v$ in the contour tree the input complex, and let $\ell_v'$ 
+denote the longest directed path passing through $v$ in the join tree of the piece containing $v$.
+By \Thm{surgery}, ignoring non-dominant extremum introduced from cutting, the join tree on each piece output by $\rain$ is isomorphic to some 
+connected subgraph of the contour tree of the input complex, and hence $\ell_v'\leq \ell_v$.  
+Moreover, $|H_v|$ only counts vertices in a $v$ to root path and so trivially $|H_v|\leq \ell_v'$, implying \Thm{main-corr}.
+
+Note that there is fair amount of slack in this argument as $|H_v|$ may be significantly smaller than $\ell_v'$.  
+This slack allows for the more refined upper and lower bounds mentioned in \Sec{more-refined}.  Quantifying this slack however is quite challenging, 
+and requires a significantly more sophisticated analysis involving path decompositions, which is the subject of \Sec{pathDecomp} and \Sec{lb}.
+
+% The full running time analysis can be found in \Sec{runTime}, \Sec{pathDecomp}, and \Sec{lb}, or in the full version \CiteFullVer.  
+% \Sec{runTime} is summarized here, which gives a bound immediately implying the running time of our main result \Thm{main-corr}.  
+% Using a significantly more sophisticated analysis involving path decompositions, this bound will also implies the more refined running time bounds mentioned in \Sec{more-refined} 
+% are the topics \Sec{pathDecomp} and \Sec{lb}.
+
+% While the above gives a full running time characterization, ideally the bound should be expressed as   
+% a function of some structural property of $\jc(\MM)$.
+% This is the focus of \Sec{pathDecomp}, where it is shown that for any binary tree $T$ 
+% the $H_v$ sets (defined by any leaf assignment) naturally induces a path decomposition of $T$ (see \Def{path}), 
+% denoted $\pmax(T)$ (see \Sec{pathDecomp} for details).   
+% 
+% % Specifically, the main result of \Sec{pathDecomp} is the following theorem, which together with the above 
+% % bound in terms of heap size immediately implies our main result, Theorem~\ref{thm:main-alg} .
+% 
+% \begin{theorem} $\sum_{v \in T} \log |H_v| = O(\sum_{p\in \pmax(T)} |p| \log |p|)$.
+% \end{theorem}
 
@@
 
-\algorithmRunTimeSection
+%\algorithmRunTimeSection
+\section{Proof of Lemma~\ref{heapRunTimeBound}}
+\label{sec:runTime}
+\proofofRunTime
 

### 2015-12-03 paint-contour-tree.tex
critical join tree issue

@@
 The directed edge $(u,v)$ is present when $u$ is the smallest valued vertex in $V'$ in a connected component of $\MM^+_v$
-and $v$ is adjacent (in $\MM$) to a vertex in this component. 
+and $v$ is adjacent (in $\MM$) to a vertex in this component. \InSoCGVer{The \emph{join tree}, $\cJ(\MM)$, is defined analogously, but instead on $\cV(\MM)$.}
+%
+%\InSoCGVer{Have sentence defining the join tree here.}
+%
 \end{definition}
@@
 \InNotSoCGVer{\proofofContourJoinEquiv}
+
+\begin{remark}
+The above theorem, combined with the previous sections, implies that in order to get an efficient contour tree algorithm, 
+it suffices to have an efficient algorithm for computing $\jc(\MM)$.
+Due to minor technicalities, the following section is instead phrased in terms of computing $\cJ(\MM)$ efficiently.  
+Note however that for minimum dominant complexes output by $\rain$, converting between $\jc$ and $\cJ$ is trivial, as $\cJ$ is just $\jc$ 
+with each non-dominant minimum $m_i$ augmented along the edge leaving $s_i$.
+\end{remark}
 
@@
 denote the longest directed path passing through $v$ in the join tree of the piece containing $v$.
-By \Thm{surgery}, ignoring non-dominant extremum introduced from cutting, the join tree on each piece output by $\rain$ is isomorphic to some 
+By \Thm{surgery}, ignoring non-dominant extrema introduced from cutting (whose cost can be charged to a corresponding saddle), the join tree on each piece output by $\rain$ is isomorphic to some 
 connected subgraph of the contour tree of the input complex, and hence $\ell_v'\leq \ell_v$.  

### 2015-12-03 paint-contour-tree.tex
labels corrected

@@
 
-\begin{definition}
+\begin{definition}\label{def:initialColoring}
 A \emph{leaf assignment} $\chi$ of a binary tree $T$ assigns \emph{two} distinct leaves to each internal vertex $v$,
@@
 
-\begin{lemma}\label{heapRunTimeBound}
+\begin{lemma}\label{lem:runTimeUpper}
 Let $\MM$ be a simplicial complex with $t$ critical points.  
@@
 %\algorithmRunTimeSection
-\section{Proof of Lemma~\ref{heapRunTimeBound}}
+\section{Proof of \Lem{runTimeUpper}}
 \label{sec:runTime}

### 2015-12-04 paint-contour-tree.tex
critical join replaced with join

@@
 \theoremstyle{definition}
-\InNotSoCGVer{ 
- \newtheorem{remark}[theorem]{Remark} 
-}
+% \InNotSoCGVer{ 
+%  \newtheorem{remark}[theorem]{Remark} 
+% }
+\newtheorem{Remark}[theorem]{Remark}
 \newtheorem{observation}[theorem]{Observation}
@@
 
-\begin{remark}
+\begin{Remark}\label{rem:joinAndCriticalJoin}
 The above theorem, combined with the previous sections, implies that in order to get an efficient contour tree algorithm, 
 it suffices to have an efficient algorithm for computing $\jc(\MM)$.
-Due to minor technicalities, the following section is instead phrased in terms of computing $\cJ(\MM)$ efficiently.  
+Due to minor technicalities, it is easier to phrase the following section instead in terms of computing $\cJ(\MM)$ efficiently.  
 Note however that for minimum dominant complexes output by $\rain$, converting between $\jc$ and $\cJ$ is trivial, as $\cJ$ is just $\jc$ 
 with each non-dominant minimum $m_i$ augmented along the edge leaving $s_i$.
-\end{remark}
+\end{Remark}
 
@@
 		\begin{compactenum}
-			\item Add edge $(\h(c'),h)$ to $\jc(\MM)$.
+			\item Add edge $(\h(c'),h)$ to $\cJ(\MM)$.
 			\item Delete $h$ from $T(c')$.
@@
 \begin{claim} \label{clm:process} The partial output on the processed vertices is exactly
-the restriction of $\jc(\MM)$ to these vertices.
+the restriction of $\cJ(\MM)$ to these vertices.
 \end{claim}
@@
 \begin{proof} More generally, we prove the following: all outputs on processed vertices
-are edges of $\jc(\MM)$ and for any current color $c$, $\h(c)$ is the lowest processed vertex
+are edges of $\cJ(\MM)$ and for any current color $c$, $\h(c)$ is the lowest processed vertex
 of that color. We prove this by induction on the processing order. The base case is trivially
@@
 it must be in $T(c)$, which violates the ripeness of $v$.
-Thus, all critical vertices in $\PP$ have been processed, and so by the induction hypothesis, the restriction of $\jc(\MM)$ to $\PP$ has been correctly computed.
+Thus, all critical vertices in $\PP$ have been processed, and so by the induction hypothesis, the restriction of $\cJ(\MM)$ to $\PP$ has been correctly computed.
 Additionally, since all critical vertices in $\PP$ have processed, they all have the same color $c$ of the lowest critical vertex in $\PP$.
@@
 the lowest critical vertex in this component is $\h(c')$ (where $c'$ is the color of edges in the respective component).
-Now by the definition of $\jc(\MM)$, the critical vertex $v$ connects to the lowest critical vertex in each component of $\MM^+_v$, 
+Now by the definition of $\cJ(\MM)$, the critical vertex $v$ connects to the lowest critical vertex in each component of $\MM^+_v$, 
 and so by the above $v$ should connect to $\h(c)$ and $\h(c')$, which is precisely what $v$ is connected to by $\build(\MM)$.
@@
 \label{thm:correct}
- Given an input complex $\MM$, $\build(\MM)$ terminates and outputs $\jc(\MM)$.
+ Given an input complex $\MM$, $\build(\MM)$ terminates and outputs $\cJ(\MM)$.
 \end{theorem}
@@
  Therefore, eventually all vertices will be processed, and so by \Clm{process} the algorithm will terminate having 
- computed $\jc(\MM)$.
+ computed $\cJ(\MM)$.
 \end{proof}
@@
 \label{def:initialColoring}
-Note that the subroutine $\init(\MM)$ from \Sec{algo} naturally defines a leaf assignment to $\jc(\MM)$ 
+Note that the subroutine $\init(\MM)$ from \Sec{algo} naturally defines a leaf assignment to $\cJ(\MM)$ 
 according to the priority queue for each up-star we put a given vertex in.  Call this the \emph{initial coloring}
-of the vertices in $\jc(\MM)$.  Note also that this initial coloring defines the $H_v$ values for all $v\in \jc(\MM)$.
+of the vertices in $\cJ(\MM)$.  Note also that this initial coloring defines the $H_v$ values for all $v\in \cJ(\MM)$.
 \end{definition}
@@
 \label{lem:runTimeUpper}
-Let $\MM$ be a simplicial complex with $t$ critical points.  For every vertex in $\jc(\MM)$, 
+Let $\MM$ be a simplicial complex with $t$ critical points.  For every vertex in $\cJ(\MM)$, 
 let $H_v$ be defined by the initial coloring of $\MM$.
-The running time of $\build(\MM)$ is $O(N+t\alpha(t) + \sum_{v \in \jc(\MM)} \log |H_v|)$.
+The running time of $\build(\MM)$ is $O(N+t\alpha(t) + \sum_{v \in \cJ(\MM)} \log |H_v|)$.
 \end{lemma}
@@
 Therefore the only thing left to bound are the queue operations.
-Let $v$ be a vertex in $\jc(\MM)$, and let $c_1$ and $c_2$ be its children (the same argument holds if $v$ has only one child).
+Let $v$ be a vertex in $\cJ(\MM)$, and let $c_1$ and $c_2$ be its children (the same argument holds if $v$ has only one child).
 At the time $v$ is processed, the colors and queues of all vertices in a given component of $\MM^+_v$ have merged together.
@@
 
-The algorithm $\build(\MM)$ processes vertices in $\jc(\MM)$ one at a time.  The main processing cost  
+The algorithm $\build(\MM)$ processes vertices in $\cJ(\MM)$ one at a time.  The main processing cost  
 comes from priority queue operations.  The cost of these operations is a function of the size of the queue which is in turn a function of the choices made by the subroutine $\init(\MM)$.
@@
 one from the left and one from the right subtree of $v$ (or only one leaf if $v$ has only one child). % naturally if $v$ has only one child it is assigned only one color). 
-The subroutine $\init(\MM)$ naturally defines a leaf assignment to $\jc(\MM)$ (which 
+The subroutine $\init(\MM)$ naturally defines a leaf assignment to $\cJ(\MM)$ (which 
 is a rooted binary tree) according to the priority queue for each up-star we put a given vertex in.  
-Call this the \emph{initial coloring} of the vertices in $\jc(\MM)$, and denote it by $\chi$.
-
-For a vertex $v$ in $\jc(\MM)$, let $L(v)$ denote the set of leaves of the subtree rooted at $v$, and let $A(v)$ denote the set of 
-ancestors of $v$, i.e.\ the vertices on the $v$ to root path. For a vertex $v\in \jc(\MM)$, and an initial coloring $\chi$, 
+Call this the \emph{initial coloring} of the vertices in $\cJ(\MM)$, and denote it by $\chi$.
+
+For a vertex $v$ in $\cJ(\MM)$, let $L(v)$ denote the set of leaves of the subtree rooted at $v$, and let $A(v)$ denote the set of 
+ancestors of $v$, i.e.\ the vertices on the $v$ to root path. For a vertex $v\in \cJ(\MM)$, and an initial coloring $\chi$, 
 we use $H_v$ to denote the \emph{heap} at $v$. Formally, $H_v = \{u | u \in A(v), \chi(u) \cap L(v) \neq \emptyset\}$, i.e.\
@@
 Let $\MM$ be a simplicial complex with $t$ critical points.  
-The running time of $\build(\MM)$ is $O(N+t\alpha(t) + \sum_{v \in \jc(\MM)} \log |H_v|)$, where $H_v$ is defined by an initial coloring.
-%For every vertex in $\jc(\MM)$, let $H_v$ be defined by the initial coloring of $\MM$.
-%The running time of $\build(\MM)$ is $O(N+t\alpha(t) + \sum_{v \in \jc(\MM)} \log |H_v|)$.
+The running time of $\build(\MM)$ is $O(N+t\alpha(t) + \sum_{v \in \cJ(\MM)} \log |H_v|)$, where $H_v$ is defined by an initial coloring.
+%For every vertex in $\cJ(\MM)$, let $H_v$ be defined by the initial coloring of $\MM$.
+%The running time of $\build(\MM)$ is $O(N+t\alpha(t) + \sum_{v \in \cJ(\MM)} \log |H_v|)$.
 \end{lemma}
@@
 % While the above gives a full running time characterization, ideally the bound should be expressed as   
-% a function of some structural property of $\jc(\MM)$.
+% a function of some structural property of $\cJ(\MM)$.
 % This is the focus of \Sec{pathDecomp}, where it is shown that for any binary tree $T$ 
@@
 % The above is a purely combinatorial statement on binary trees, and may find applications in other charging arguments.  
-% However, for the specific when $T=\jc(\MM)$, plugging this theorem into the above lemma, and applying this lemma 
+% However, for the specific when $T=\cJ(\MM)$, plugging this theorem into the above lemma, and applying this lemma 
 % to each subcomplex piece output by $\rain$, yields our main result, Theorem~\ref{thm:main-alg}.
@@
 
-In this section, we set up a framework to analyze the time taken to compute a critical join tree $\jc(\MM)$ (see \Def{criticalJoin}).
+In this section, we set up a framework to analyze the time taken to compute a join tree $\cJ(\MM)$ (see \Def{criticalJoin}).
 We adopt all notation already defined in \Sec{runTime}.  From here forward we will often assume binary trees are full binary trees 
@@
 \begin{proof} Let $\cU$ denote a connected component of the shrub forest.
-We have $\sum_{p \in \cL \setminus \cL'} \sum_{v \in p} \log |H_v| \leq \sum_{\cU} \sum_{p \in \cU} \sum_{v \in p} \log |H_v|$.
+We have $\sum_{p \in \cL \setminus \cL'} \sum_{v \in p} \log |H_v| \allowbreak \leq \sum_{\cU} \sum_{p \in \cU} \sum_{v \in p} \log |H_v|$.
 By \Lem{cu-root}, $|H_v| = O(|H_r|)$, where $r$ is the root of $\cU$.
@@
 \begin{theorem}
-%Consider a PL-Morse $f:\MM \to \RR$, where the join tree $\jc(\MM)$ has maximum degree $3$.
-Let $f:\MM \to \RR$ be the linear interpolant over distinct valued vertices, where the join tree $\jc(\MM)$ has maximum degree $3$.
-There is an algorithm to compute the join tree whose running time is $O(\sum_{p \in \pmax(\jc)} |p|\log |p| + t\alpha(t) + N)$.
+%Consider a PL-Morse $f:\MM \to \RR$, where the join tree $\cJ(\MM)$ has maximum degree $3$.
+Let $f:\MM \to \RR$ be the linear interpolant over distinct valued vertices, where the join tree $\cJ(\MM)$ has maximum degree $3$.
+There is an algorithm to compute the join tree whose running time is $O(\sum_{p \in \pmax(\cJ)} |p|\log |p| + t\alpha(t) + N)$.
 \end{theorem}
 \begin{proof}
-By \Thm{correct} we know that $\build(\MM)$ correctly outputs $\jc(\MM)$, and by \Lem{runTimeUpper}
-we know this takes $O(\sum_{v \in \jc(\MM)} \log |H_v| + t\alpha(t) + N)$ time, where the $H_v$ values are determined as in \Def{initialColoring}.
-Therefore by \Thm{runtime}, $\build(\MM)$ takes $O(\sum_{p \in \pmax(\jc)} |p|\log |p| + t\alpha(t) + N)$ 
-time to correctly compute $\jc(\MM)$.
+By \Thm{correct} we know that $\build(\MM)$ correctly outputs $\cJ(\MM)$, and by \Lem{runTimeUpper}
+we know this takes $O(\sum_{v \in \cJ(\MM)} \log |H_v| + t\alpha(t) + N)$ time, where the $H_v$ values are determined as in \Def{initialColoring}.
+Therefore by \Thm{runtime}, $\build(\MM)$ takes $O(\sum_{p \in \pmax(\cJ)} |p|\log |p| + t\alpha(t) + N)$ 
+time to correctly compute $\cJ(\MM)$.
 \end{proof}
@@
 we can obtain $\cC(\MM_i)$ by simply sticking on the non-dominant minima at their respective splits (which can easily be done in linear time).
-However, by the above theorem we know $\jc(\MM_i)$ can be computed in $O(\sum_{p \in \pmax(\jc(\MM_i))} |p|\log |p| + t_i\alpha(t_i) + N_i)$
+Remark~\ref{rem:joinAndCriticalJoin} implies that $\jc(\MM_i)$ is trivially obtained from the $\cJ(\MM_i)$, and 
+ by the above theorem we know $\cJ(\MM_i)$ can be computed in $O(\sum_{p \in \pmax(\cJ(\MM_i))} |p|\log |p| + t_i\alpha(t_i) + N_i)$
 (where $t_i$ and $N_i$ are the number of critical points and faces when restricted to $\MM_i$).
@@
 compute the contour tree in time
+\InSoCGVer{
+\begin{align*}
+\hspace{.8in} O\pth{N+ \sum_{i=1}^k \pth{\sum_{p \in \pmax(\jc(\MM_i))} |p|\log |p|} + t_i\alpha(t_i) + N_i}& \\
+=\break O\pth{\pth{\sum_{p \in \pmax(\cC(\MM))} |p|\log |p|} + t\alpha(t) + N}&
+\end{align*}
+}
+\InNotSoCGVer{
 \[
-O\pth{N+ \sum_{i=1}^k \pth{\sum_{p \in \pmax(\jc(\MM_i))} |p|\log |p|} + t_i\alpha(t_i) + N_i} = O\pth{\pth{\sum_{p \in \pmax(\cC(\MM))} |p|\log |p|} + t\alpha(t) + N}
+O\pth{N+ \sum_{i=1}^k \pth{\sum_{p \in \pmax(\jc(\MM_i))} |p|\log |p|} + t_i\alpha(t_i) + N_i}
+=\break O\pth{\pth{\sum_{p \in \pmax(\cC(\MM))} |p|\log |p|} + t\alpha(t) + N}
 \]
+}
 
@@
 
-\begin{remark}
+\begin{Remark}
  Note that for the terrains described in this section, the number of critical points is within a constant factor of 
  the total number of vertices.  In particular, for this family of terrains, all previous algorithms required $\Omega(n\log n)$ time.
-\end{remark}
+\end{Remark}
 }

### 2015-12-04 paint-contour-tree.tex
typos corrected

@@
 sorting all the maxima. Also consider the surface of \Fig{sorting}. The final contour tree is basically two binary
-trees joined at their roots, and we do not need to entire sorted order of critical points
+trees joined at their roots, and we do not need the entire sorted order of critical points
 to construct the contour tree.
@@
 where the algorithm makes $C$ comparisons. Then \emph{any algorithm} must make roughly $C$
-comparisons in the worst-case over this set. (All further details are 
-in Appendix~\Sec{lb}.)
+comparisons in the worst-case over this set. (All further details are in \InSoCGVer{Appendix~}\Sec{lb}.)
 
@@
 Hence the critical join tree below is just the join tree without these stumps.  
-Precise definitions and proofs are given in \Sec{jstedm}.
+Additional definitions and proofs are given in \Sec{jstedm}.
 }
@@
 \InSoCGVer{
-\subsection{Upper Bounds for Running Time}\label{sec:runTimeSummary}
+\subsection{Upper Bounds for Running Time}\label{sec:runTime}
 
@@
 
-Given the above technical definition, the proof of the following lemma is straightforward, though long and so has been moved to \Sec{runTime}.
+Given the above technical definition, the proof of the following lemma is straightforward, though long and so has been moved to \Sec{runTimeProof}.
 
@@
 complex.  By \Thm{jordan} this vertex appears in exactly one of the pieces output by $\rain$.  As in the \Thm{main-corr} statement, let $\ell_v$ 
-denote the length of the longest directed path passing through $v$ in the contour tree the input complex, and let $\ell_v'$ 
+denote the length of the longest directed path passing through $v$ in the contour tree of the input complex, and let $\ell_v'$ 
 denote the longest directed path passing through $v$ in the join tree of the piece containing $v$.
@@
 At this point we can now see what the path decomposition referenced in theorem statement should be.  It is just the union of all the maximum 
-path decomposition across the extremum dominant pieces, $\pmax(\cC(\MM)) = \cup_{i=1}^k \pmax(\jc(\MM_i))$.
+path decomposition across the extremum dominant pieces, $\pmax(\cC(\MM)) = \cup_{i=1}^k \pmax(\cJ(\MM_i))$.
 Since all procedures besides computing the join trees take linear time in the size of the input complex, we can therefore 
@@
 \begin{align*}
-\hspace{.8in} O\pth{N+ \sum_{i=1}^k \pth{\sum_{p \in \pmax(\jc(\MM_i))} |p|\log |p|} + t_i\alpha(t_i) + N_i}& \\
+\hspace{.8in} O\pth{N+ \sum_{i=1}^k \pth{\sum_{p \in \pmax(\cJ(\MM_i))} |p|\log |p|} + t_i\alpha(t_i) + N_i}& \\
 =\break O\pth{\pth{\sum_{p \in \pmax(\cC(\MM))} |p|\log |p|} + t\alpha(t) + N}&
@@
 \[
-O\pth{N+ \sum_{i=1}^k \pth{\sum_{p \in \pmax(\jc(\MM_i))} |p|\log |p|} + t_i\alpha(t_i) + N_i}
+O\pth{N+ \sum_{i=1}^k \pth{\sum_{p \in \pmax(\cJ(\MM_i))} |p|\log |p|} + t_i\alpha(t_i) + N_i}
 =\break O\pth{\pth{\sum_{p \in \pmax(\cC(\MM))} |p|\log |p|} + t\alpha(t) + N}
@@
 We first prove a lower bound for join trees, and then generalize to contour trees.
-Note that form of the theorem statements in this section differ from \Thm{main-lb}, 
+Note that the form of the theorem statements in this section differ from \Thm{main-lb}, 
 as they are stated directly in terms of path decompositions.  \Thm{main-lb} 
-is an immediate corollary of the final theorem of this section \Thm{path-main-lb}.
+is an immediate corollary of the final theorem of this section, \Thm{path-main-lb}.
 
@@
 \section{Proof of \Lem{runTimeUpper}}
-\label{sec:runTime}
+\label{sec:runTimeProof}
 \proofofRunTime

### 2015-12-04 paint-contour-tree.tex
typo in abstract fixed

@@
 If $C$ denotes the set of critical points in $\MM$, the running time is
-basically $O(\sum_{v \in C} \log \ell_v)$, where $\ell_v$ is the depth of $v$ in the contour tree.
-This matches al existing upper bounds, but is a significant improvement when the contour tree is short and fat.
+roughly $O(\sum_{v \in C} \log \ell_v)$, where $\ell_v$ is the depth of $v$ in the contour tree.
+This matches all existing upper bounds, but is a significant improvement when the contour tree is short and fat.
 % Our results match all existing upper bounds, but also

### 2015-12-10 paint-contour-tree.tex
full ver updated for arxiv

@@
 %%%
-\def\SoCGVer{TRUE}%
+%\def\SoCGVer{TRUE}%
 %%%
@@
 This matches all existing upper bounds, but is a significant improvement when the contour tree is short and fat.
-% Our results match all existing upper bounds, but also
-% characterize the amount of sorting required to compute the contour tree.
 Specifically, our approach ensures that any comparison made is between nodes in the same descending path in the contour tree,
@@
 \label{def:int-criticalJoin}
-The \emph{critical join tree} $\jc(\MM)$ is built on the set $\cV$ of all critical points.\footnote{%
-In Definition~\ref{def:criticalJoin}, we define the critical join tree on a certain subset of $\cV$.}
+The \emph{join tree} $\cJ(\MM)$ is built on the set $\cV$ of all critical points.
 The directed edge $(u,v)$ is present when $u$ is the smallest valued vertex in $\cV$ in a connected component of $\MM^+_v$
@@
 
-Abusing notation, for now we will just call this the join tree. (In \Sec{extreme} \InSoCGVer{and \Sec{jstedm}} we carefully handle the distinction 
-between the join tree and the critical join tree.)
+%Abusing notation, for now we will just call this the join tree. 
+%(In \Sec{extreme} \InSoCGVer{and \Sec{jstedm}} we carefully handle the distinction between the join tree and the critical join tree.)
 Refer to \Fig{sorting} for the join tree of a terrain. Note that nothing happens at splits, but these
@@
 
-The join tree $\jc(\NN)$ is a rooted tree with the dominant minimum at the root, and we direct edges downwards (towards the root). So it makes sense to talk
+The join tree $\cJ(\NN)$ is a rooted tree with the dominant minimum at the root, and we direct edges downwards (towards the root). So it makes sense to talk
 of comparable vs incomparable vertices.
@@
 
-Our algorithm incrementally builds $\jc(\MM)$ from the leaves (maxima) to the root (dominant minimum). 
+Our algorithm incrementally builds $\cJ(\MM)$ from the leaves (maxima) to the root (dominant minimum). 
 We say that vertex $v$ is \emph{touched} by color $c$, if there is a $c$-colored edge with lower endpoint $v$.
@@
 Both mounds are colored completely blue or red, while $v$ is touched by both colors.
-So this indicates that $v$ joins the blue maximum and red maximum in $\jc(\MM)$.
+So this indicates that $v$ joins the blue maximum and red maximum in $\cJ(\MM)$.
 
@@
 when some join $v$ has all incident edges with exactly two colors, the corresponding maxima (of those colors) are exactly
-the children of $v$ in $\jc(\MM)$. To proceed further, we ``merge" the colors red and blue into a new color, purple. In other words,
+the children of $v$ in $\cJ(\MM)$. To proceed further, we ``merge" the colors red and blue into a new color, purple. In other words,
 we replace all red and blue edges by purple edges. This indicates that the red and blue maxima have been
 handled. Imagine flattening the red and blue mounds until reaching $v$, so that the former join $v$
-is now a new maximum, from which purple paint is poured. In terms of $\jc(\MM)$, this
+is now a new maximum, from which purple paint is poured. In terms of $\cJ(\MM)$, this
 is equivalent to removing leaves $\ell_1$ and $\ell_2$, and making $v$ a new leaf.
-Alternatively, $\jc(\MM)$ has been constructed up to $v$, and it remains to determine $v$'s parent.
+Alternatively, $\cJ(\MM)$ has been constructed up to $v$, and it remains to determine $v$'s parent.
 The merging of the colors is not explicitly performed as that would be too expensive; instead we maintain a union-find
@@
 Of course, things are more complicated when there are other mounds. There may be a yellow mound,
-corresponding to $\ell_3$ that joins with the blue mound higher up at some vertex $u$ (see the right part of \Fig{colors}). In $\jc(\MM)$, $\ell_1$ and $\ell_3$
+corresponding to $\ell_3$ that joins with the blue mound higher up at some vertex $u$ (see the right part of \Fig{colors}). In $\cJ(\MM)$, $\ell_1$ and $\ell_3$
 are sibling leaves, and $\ell_2$ is a sibling of some ancestor of these leaves. So we cannot
@@
 
-\myparagraph{The running time analysis:} Relating the running time to a path decomposition is the most technical part of the paper. 
+\myparagraph{The running time analysis:} 
  All the non-heap operations can be easily bounded by $O(t\alpha(t) + N)$ (the $t\alpha(t)$ is from the union-find data structure
-for colors). It is not hard to argue that at all times, any heap always contains a subset of a leaf
-to root path. Unfortunately, this subset is \emph{not} contiguous, as the right part of \Fig{colors} shows. 
+ for colors). It is not hard to argue that at all times, any heap always contains a subset of a leaf to root path.
+ This observation suffices to get a running time bound which is the analogue of \Thm{main-corr}, but for join trees. 
+ Each heap deletion and merge can be charged to a vertex in the join tree (where each vertex gets charged only a constant number of times).
+ Let $d_v$ denote the distance to the root for a vertex $v$ in the join tree, then since a heap's elements are on a single leaf to root 
+ path, the size of $v$'s heap at the time an associated heap operation is made is at most $d_v$.
+ Therefore, the total cost (of the heap operations) is at most $O(\sum_v \log d_v)$. This immediately proves a bound of $O(t\log D)$, where
+ $D$ is the maximum distance to the root, an improvement over previous work.
+
+ However, this bound is non-optimal.  For example, for a balanced binary tree, this gives a bound of $O(t\log\log t)$, however, by using 
+ an analysis involving path decompositions we can get an $O(t)$ bound.  Imagine walking from some leaf 
+ towards the root. Each vertex on this path can have at most two colors (the ones getting merged), however, as we get closer to the root 
+ the competition for which two colors a vertex gets assigned to grows, as the number of descendant leaf colors grows. This means that for some vertices, 
+ $v$, the size of the heap for an associated heap operation will be significantly smaller than $d_v$.
+
+The intuition is that the paint spilling from the maxmima in the simplicial complex, corresponds to paint spilling from the leaves in 
+the join tree, which decomposes the join tree into a set of colored paths.  Unfortunately, the situation is more complex since 
+while a given color class is confined to a single leaf to root path, it may \emph{not} appear contiguously on this path, as the right part of \Fig{colors} shows. 
 Specifically, in this figure the far left saddle (labeled $i$) is hit by blue paint.  However, there is another 
@@
 higher than the far left one, it will merge into the component containing the blue mound (and also the yellow and red mounds)
-before the far left one.  Hence, the vertices initially touched by blue are not contiguous in the contour tree.
-
-Nonetheless, we can get a non-trivial (but non-optimal) bound. Let $d_v$ denote the distance to the root for vertex $v$ in the join tree.
-The total cost (of the heap operations) is at most $\sum_v \log d_v$. This immediately proves a bound of $O(t\log D)$, where
-$D$ is the maximum distance to the root, an improvement over previous work. But this bound is non-optimal.
-For a balanced binary tree, this bound is $O(t\log\log t)$, whereas the cost of any path decomposition is $O(t)$.
-
-The cost of heap operations depends on heap sizes, which keep changing because
-of the repeated merging. This causes the analysis to be technically challenging. There are situations where
-the initial heap sizes are small, but they eventually merge to create larger heaps. 
-We employ a variant
-of \emph{heavy path decompositions}, first used by Sleator and Tarjan for analyzing link/cut trees~\cite{st-dsdt-83}.
-The final analysis charges expensive heap operations to long paths in the decomposition.
+before the far left one.  Hence, the vertices initially touched by blue are not contiguous in the join tree.
+
+This non-contiguous complication along with the fact that heap size keep changing as color classes merge, causes the analysis to be technically challenging. 
+%There are situations where the initial heap sizes are small, but they eventually merge to create larger heaps. 
+We employ a variant of \emph{heavy path decompositions}, first used by Sleator and Tarjan for analyzing link/cut trees~\cite{st-dsdt-83}.
+The final analysis charges expensive heap operations to long paths in the decomposition, resulting in the bound stated in \Thm{main-alg}.
+
+% Relating the running time to a path decomposition is the most technical part of the paper. 
+% All the non-heap operations can be easily bounded by $O(t\alpha(t) + N)$ (the $t\alpha(t)$ is from the union-find data structure
+% for colors). It is not hard to argue that at all times, any heap always contains a subset of a leaf
+% to root path. Unfortunately, this subset is \emph{not} contiguous, as the right part of \Fig{colors} shows. 
+% Specifically, in this figure the far left saddle (labeled $i$) is hit by blue paint.  However, there is another 
+% saddle on the far right (labeled $j$) which is not hit by blue paint.  Since this far right saddle is slightly 
+% higher than the far left one, it will merge into the component containing the blue mound (and also the yellow and red mounds)
+% before the far left one.  Hence, the vertices initially touched by blue are not contiguous in the contour tree.
+% 
+% Nonetheless, we can get a non-trivial (but non-optimal) bound, which is the analogue of \Thm{main-corr} for join trees. 
+% Let $d_v$ denote the distance to the root for vertex $v$ in the join tree.
+% The total cost (of the heap operations) is at most $\sum_v \log d_v$. This immediately proves a bound of $O(t\log D)$, where
+% $D$ is the maximum distance to the root, an improvement over previous work. But this bound is non-optimal.
+% For a balanced binary tree, this bound is $O(t\log\log t)$, whereas the cost of any path decomposition is $O(t)$.
+% \Thm{main-alg}
+%
+% The cost of heap operations depends on heap sizes, which keep changing because
+% of the repeated merging. This causes the analysis to be technically challenging. There are situations where
+% the initial heap sizes are small, but they eventually merge to create larger heaps. 
+% We employ a variant
+% of \emph{heavy path decompositions}, first used by Sleator and Tarjan for analyzing link/cut trees~\cite{st-dsdt-83}.
+% The final analysis charges expensive heap operations to long paths in the decomposition.
 
@@
 \newcommand{\joinTreeSplitTreeMerge}{
-\begin{definition} \label{join} The \emph{join tree} $\cJ(\MM)$ of $\MM$ is built on vertex set $\cV(\MM)$.
+\begin{definition} \label{def:join} The \emph{join tree} $\cJ(\MM)$ of $\MM$ is built on vertex set $\cV(\MM)$.
 The directed edge $(u,v)$ is present when $u$ is the smallest valued vertex in a connected component of $\MM^+_v$
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\InNotSoCGVer{\proofofRunTime}
-
-%\InNotSoCGVer{\algorithmRunTimeSection}
-%%%%%%%%%%%%%%%%%%%%
+\InNotSoCGVer{
+\proofofRunTime
+}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \InSoCGVer{
@@
 
-% The full running time analysis can be found in \Sec{runTime}, \Sec{pathDecomp}, and \Sec{lb}, or in the full version \CiteFullVer.  
-% \Sec{runTime} is summarized here, which gives a bound immediately implying the running time of our main result \Thm{main-corr}.  
-% Using a significantly more sophisticated analysis involving path decompositions, this bound will also implies the more refined running time bounds mentioned in \Sec{more-refined} 
-% are the topics \Sec{pathDecomp} and \Sec{lb}.
-% %
-% %In \Sec{runTime}, we give a tight bound on the running time of $\build(\MM)$, as a function of the choices made by the subroutine $\init(\MM)$.
-
 The algorithm $\build(\MM)$ processes vertices in $\cJ(\MM)$ one at a time.  The main processing cost  
 comes from priority queue operations.  The cost of these operations is a function of the size of the queue which is in turn a function of the choices made by the subroutine $\init(\MM)$.
-
-%The heap as defined above characterizes the priority queue size at the time a vertex gets processed and hence we get the following.
 
@@
 The running time of $\build(\MM)$ is $O(N+t\alpha(t) + \sum_{v \in \cJ(\MM)} \log |H_v|)$, where $H_v$ is defined by an initial coloring.
-%For every vertex in $\cJ(\MM)$, let $H_v$ be defined by the initial coloring of $\MM$.
-%The running time of $\build(\MM)$ is $O(N+t\alpha(t) + \sum_{v \in \cJ(\MM)} \log |H_v|)$.
 \end{lemma}
-
-Our main result, \Thm{main-corr}, is an easy corollary of the above lemma. Specifically, consider a critical point $v$ of the initial input 
+}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\InSoCGVer{Our main result, } \Thm{main-corr}, is an easy corollary of the above lemma. Specifically, consider a critical point $v$ of the initial input 
 complex.  By \Thm{jordan} this vertex appears in exactly one of the pieces output by $\rain$.  As in the \Thm{main-corr} statement, let $\ell_v$ 
@@
 
-% The full running time analysis can be found in \Sec{runTime}, \Sec{pathDecomp}, and \Sec{lb}, or in the full version \CiteFullVer.  
-% \Sec{runTime} is summarized here, which gives a bound immediately implying the running time of our main result \Thm{main-corr}.  
-% Using a significantly more sophisticated analysis involving path decompositions, this bound will also implies the more refined running time bounds mentioned in \Sec{more-refined} 
-% are the topics \Sec{pathDecomp} and \Sec{lb}.
-
-% While the above gives a full running time characterization, ideally the bound should be expressed as   
-% a function of some structural property of $\cJ(\MM)$.
-% This is the focus of \Sec{pathDecomp}, where it is shown that for any binary tree $T$ 
-% the $H_v$ sets (defined by any leaf assignment) naturally induces a path decomposition of $T$ (see \Def{path}), 
-% denoted $\pmax(T)$ (see \Sec{pathDecomp} for details).   
-% 
-% % Specifically, the main result of \Sec{pathDecomp} is the following theorem, which together with the above 
-% % bound in terms of heap size immediately implies our main result, Theorem~\ref{thm:main-alg} .
-% 
-% \begin{theorem} $\sum_{v \in T} \log |H_v| = O(\sum_{p\in \pmax(T)} |p| \log |p|)$.
-% \end{theorem}
-
-% The above is a purely combinatorial statement on binary trees, and may find applications in other charging arguments.  
-% However, for the specific when $T=\cJ(\MM)$, plugging this theorem into the above lemma, and applying this lemma 
-% to each subcomplex piece output by $\rain$, yields our main result, Theorem~\ref{thm:main-alg}.
-% 
-% The path decomposition $\pmax(T)$ is a function of the choices made by the linear time $\init(\MM)$.  
-% One cannot control which path decomposition $\init$ determines, as sorting can be reduced to finding the optimal path decomposition.  
-% Specifically, in \Sec{lb} we exploit this connection to prove the following matching lower bound for our approach.
-% In the following a rain consistent path decomposition is any decomposition which is the union of paths from 
-% decompositions on each subcomplex output by $\rain$ (see Definition~\ref{def:path2} for a formal definition).
-
-% \begin{theorem}
-% Consider a rooted tree $T$ and an arbitrary path decomposition $P$ of $T$.
-% There is a family $\bF_P$ of terrains such that any algebraic decision
-% tree computing the join tree (on $\bF_P$) requires $\Omega(\sum_{p \in P} |p|\log |p|)$ time.
-% Furthermore, our algorithm makes $O(\sum_{p \in P} |p|\log |p|)$ comparisons on all these instances.
-% \end{theorem}
-% 
-% The result can also 
-% For simplicity we now restrict our attention to path decompositions 
-% consistent with the raining procedure described in \Sec{rain} 
-% (more general decompositions can work, but it is not needed for our purposes).
-% 
-% Note that in the above $T$ is assumed to be a rooted tree, and so the contour tree is equivalent to the join tree.   
-% In general our input $\MM$ will not have a contour tree with this structure, which is why we required the cutting procedure of \Sec{rain}.  
-% The above result can be extended to such general contour trees, however, here we restrict to rain consistent path decomposition, 
-% i.e.\ decompositions which are the union of paths decompositions from each piece (see \ref{def:path2} for a formal definition)
-
-% \begin{theorem} 
-% Consider any rain consistent path decomposition $P$. There exists a family $\bF_P$ of terrains ($d=2$) with the following properties.
-% Any contour tree algorithm makes $\Omega(\sum_{p \in P} |p|\log |p|)$ comparisons in the worst case over $\bF_P$.
-% Furthermore, for any terrain in $\bF_P$, our algorithm makes $O(\sum_{p \in P} |p|\log |p|)$ comparisons.
-% \end{theorem}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-% % We complement \Thm{main-alg} with a lower bound, suggesting that the comparison model complexity
-% % of $\sum_{p \in P(T)} |p|\log |p|$ is inherent for any contour tree algorithm. The exact phrasing
-% % of the lower bound requires some care. Consider the $d=2$ (terrain) case.
-% Let a path decomposition $P$ be called \emph{valid} if it is actually output by the algorithm on some input.
-% \begin{theorem} 
-% \label{thm:main-lb}
-% Consider a valid path decomposition $P$. There exists a family $\bF_P$ of terrains (i.e.\ $d=2$) with the following properties.
-% Any contour tree algorithm makes $\Omega(\sum_{p \in P} |p|\log |p|)$ comparisons in the worst case over $\bF_P$.
-% Furthermore, for any terrain in $\bF_P$, the algorithm of \Thm{main-alg} makes $O(\sum_{p \in P} |p|\log |p|)$ comparisons.
-% \end{theorem}
-% 
-% In other words, for any path decomposition, $P$, that our algorithm chooses, we can design a hard family where any 
-% algorithm requires $\sum_{p \in P} |p|\log |p|$ time in the worst case, and our algorithm matches this bound. 
-}
-%%%%%%%%%%%%%
-
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 
-In this section, we set up a framework to analyze the time taken to compute a join tree $\cJ(\MM)$ (see \Def{criticalJoin}).
+In this section, we set up a framework to analyze the time taken to compute a join tree $\cJ(\MM)$ (see \InSoCGVer{\Def{criticalJoin}}\InNotSoCGVer{\Def{join}}).
 We adopt all notation already defined in \Sec{runTime}.  From here forward we will often assume binary trees are full binary trees 

### 2016-03-17 paint-contour-tree.tex
changes for camera ready

@@
 %%%
-%\def\SoCGVer{TRUE}%
+\def\SoCGVer{TRUE}%
 %%%
@@
 
-\newcommand{\CiteFullVer}{\cite{rs-mvrms-14}}%
+\newcommand{\CiteFullVer}{\cite{rs-mvrms-14}\xspace}%
 
@@
 \InSoCGVer{%
-   \documentclass[a4paper,UKenglish]{lipics}
+   \documentclass[a4paper,UKenglish]{lipics-v2016}
 %   \documentclass[11pt]{article}%
@@
 
+\newcommand{\thmConst}{\gamma}
+
 \newcommand{\CodeComment}[1]{\textcolor{blue}{\texttt{#1}}}
@@
 
-%\usepackage{thmtools}
-%\usepackage{thm-restate}
-%\declaretheorem[name=Theorem,numberwithin=section]{thm}
-
 \newcommand{\atgen}{\symbol{'100}}
@@
       Department of Computer Science; %
-      University of Illinois; %
-%      201 N. Goodwin Avenue; %
-      Urbana, IL, USA; %
-      {\tt \si{raichel}2\atgen{}\si{uiuc}.\si{edu}}; %
-      {\tt \url{\si{http://illinois.edu/\string~\si{raichel2}}}}%
+      University of Texas at Dallas; %
+      Richardson, TX, USA; %
+      {\tt \si{benjamin.raichel}@\si{utdallas}.\si{edu}}; %
+      {\tt \url{\si{http://utdallas.edu/\string~\si{benjamin.raichel}}}}%
       . %
@@
 
-\Copyright{Benjamin Raichel and Seshadhri Comandur}%mandatory, please use full first names. LIPIcs license is "CC-BY";  http://creativecommons.org/licenses/by/3.0/
+\Copyright{Benjamin Raichel and C. Seshadhri}%mandatory, please use full first names. LIPIcs license is "CC-BY";  http://creativecommons.org/licenses/by/3.0/
 
@@
 
-% %Editor-only macros:: begin (do not touch as author)%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% \serieslogo{}%please provide filename (without suffix)
-% \volumeinfo%(easychair interface)
-%   {Billy Editor and Bill Editors}% editors
-%   {2}% number of editors: 1, 2, ....
-%   {Conference title on which this volume is based on}% event
-%   {1}% volume
-%   {1}% issue
-%   {1}% starting page number
-% \EventShortName{}
-% \DOI{10.4230/LIPIcs.xxx.yyy.p}% to be completed by the volume editor
-% % Editor-only macros::end %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\EventEditors{S\'andor Fekete and Anna Lubiw}
+\EventNoEds{2}
+\EventLongTitle{32nd International Symposium on Computational Geometry
+(SoCG 2016)}
+\EventShortTitle{SoCG 2016}
+\EventAcronym{SoCG}
+\EventYear{2016}
+\EventDate{June 14-18, 2016}
+\EventLocation{Boston, USA}
+\EventLogo{}
+\SeriesVolume{51}
+\ArticleNo{57}
+
 }
@@
 roughly $O(\sum_{v \in C} \log \ell_v)$, where $\ell_v$ is the depth of $v$ in the contour tree.
-This matches all existing upper bounds, but is a significant improvement when the contour tree is short and fat.
-Specifically, our approach ensures that any comparison made is between nodes in the same descending path in the contour tree,
+This matches all existing upper bounds, but is a significant asymptotic improvement when the contour tree is short and fat.
+Specifically, our approach ensures that any comparison made is between nodes that are either adjacent in $\MM$ 
+or in the same descending path in the contour tree,
 allowing us to argue strong optimality properties of our algorithm.
@@
 the function $f$. 
-% For example, consider a rocket engine during a burn.  The 3D mesh represents the locations of points in space around the engine, 
-% and $f$ represents the temperature at each point.  The boundary of the flame is then represented by an isotherm, i.e.\ a contour. 
-% 
-It is convenient to think of $f$ as a manifold sitting in $\RR^{d+1}$,
-%  or alternately consider
-% $\MM$ in $\RR^{d+1}$
+It is convenient to think of $f$ as a simplicial complex sitting in $\RR^{d+1}$,
 with the last coordinate (i.e.\ height) given by $f$. Imagine sweeping the hyperplane $x_{d+1} = h$
@@
 Throughout we follow the definition of contour trees from \cite{kobps-ctsssit-97} which includes all changes in topology.  
-For $d>2$, some subsequent works, such as \cite{csa-cctad-00}, only include changes in the number of components.
+For $d>2$, some subsequent works, such as \cite{csa-cctad-03}, only include changes in the number of components.
 
@@
 A fundamental result in this area is the algorithm of Carr, Snoeyink, and Axen to compute contour trees,
-which runs in $O(n\log n + N\alpha(N))$ time~\cite{csa-cctad-00} (where $\alpha(\cdot)$ denotes the inverse
+which runs in $O(n\log n + N\alpha(N))$ time~\cite{csa-cctad-03} (where $\alpha(\cdot)$ denotes the inverse
 Ackermann function). In practical applications, $N$ is typically $\Theta(n)$
@@
 \begin{theorem} \label{thm:main-corr} Consider a simplicial complex $f:\MM \to \RR$, described as above, and denote
-the contour tree by $T$ with vertex set (the critical points) $C(T)$. There exists an algorithm to compute the contour tree $T$
+the contour tree by $T$ with vertex set (the critical points) $C(T)$. There exists an algorithm to compute $T$
 in $O(\sum_{v \in C(T)} \log \ell_v + t\alpha(t) + N)$ time. Moreover, this algorithm only
-compares function values at pairs of points that are ancestor-descendant in $T$.
+compares function values at pairs of vertices that are ancestor-descendant in $T$ or adjacent in $\MM$.
 \end{theorem}
@@
 where $D$ is the diameter of the contour tree. 
-This is a significant improvement for short and fat contour trees. For example, if the tree is balanced,
+This is a significant asymptotic improvement for short and fat contour trees. For example, if the tree is balanced,
 then we get a bound of $O(t\log\log t)$.
@@
 The number of comparisons made is $O(\sum_{p \in P(T)} |p|\log |p| + N)$.
-In particular, any comparisons made are only between ancestors and descendants in the contour tree.
+In particular, comparisons are only made between pairs of vertices that are ancestor-descendant in $T$ or adjacent in $\MM$.
 \end{theorem}
 
-Note that \Thm{main-corr} is a direct corollary of this statement. For any $v$, $\ell_v$ is at most
+Note that \Thm{main-corr} is a direct corollary of this statement. For any $v$, $\ell_v$ is at least
 the length of the path in $P(T)$ that contains $v$.
 This bound is strictly stronger,
-since for any balanced contour tree, the run time bound of \Thm{main-alg} is $O(t\alpha(t) + N)$, and $O(t)$
-comparisons are made.
+since for any balanced contour tree, the run time bound of \Thm{main-alg} is $O(t\alpha(t) + N)$, and $O(N)$
+comparisons are made. (Since for a balanced tree, one can show $\sum_{p \in P(T)} |p|\log |p|= O(t)$.)
 
@@
 is implicitly constructed by the algorithm. Nonetheless, we prove that the algorithm of \Thm{main-alg}
-has strong optimality properties. For convenience, fix some value of $t$, and consider the set
+has strong optimality properties. For convenience, fix some value $t=\Omega(N)$, and consider the set
 of terrains ($d=2$) with $t$ critical points. The bound of \Thm{main-alg} takes values ranging
-from $t$ to $t\log t$. Consider some $C \in [t,t\log t]$, and consider the set of terrains
-where the algorithm makes $C$ comparisons. Then \emph{any algorithm} must make roughly $C$
-comparisons in the worst-case over this set. (All further details are in \InSoCGVer{Appendix~}\Sec{lb}.)
+from $t$ to $t\log t$. Consider some $\thmConst \in [t,t\log t]$, and consider the set of terrains
+where the algorithm makes $\thmConst$ comparisons. Then \emph{any algorithm} must make roughly $\thmConst$
+comparisons in the worst-case over this set. (Further details are in \InSoCGVer{the full version \CiteFullVer}\InNotSoCGVer{\Sec{lb}}.)
 
@@
 There exists some absolute constant $\alpha$ such that the following holds.
-For sufficiently large $t$ and any $C\in [t, t\log t]$, consider the set $\bF_C$ of terrains with $t$ critical points such that
-the number of comparisons made by the algorithm of \Thm{main-alg} on these terrains is in $[C,\alpha C]$.
+For sufficiently large $t$ and any $\thmConst\in [t, t\log t]$, consider the set $\bF_\thmConst$ of terrains with $t$ critical points such that
+the number of comparisons made by the algorithm of \Thm{main-alg} on these terrains is in $[\thmConst,\alpha \thmConst]$.
 Any algebraic decision tree that correctly computes
-the contour tree on all of $\bF_C$ has a worst case running time of $\Omega(C)$.
+the contour tree on all of $\bF_\thmConst$ has a worst case running time of $\Omega(\thmConst)$.
 \end{theorem}
@@
 Tarasov and Vyalya \cite{tv-cct-98} improved the running time to $O(N\log N)$ for the 3D case.
-The influential paper of Carr \etal \cite{csa-cctad-00} improved the running time for all dimensions to $O(n\log n + N\alpha(N))$.
+The influential paper of Carr \etal \cite{csa-cctad-03} improved the running time for all dimensions to $O(n\log n + N\alpha(N))$.
 Pascucci and Cole-McLaughlin \cite{pc-ectls-02} provided an $O(n+t\log n)$ time algorithm for 
@@
 The critical points are exactly where the topology of level sets change.
-By assuming that our manifold is boundary critical, the vertices on a given boundary are either collectively all maxima or all minima.  
+By assuming that our simplicial complex is boundary critical, the vertices on a given boundary are either collectively all maxima or all minima.  
 We abuse notation and refer to this entire set of vertices as a maximum or minimum.
@@
 It is not immediately obvious that this graph is a tree, but alternate definitions of the contour tree
-in~\cite{csa-cctad-00} imply this is a tree. 
+in~\cite{csa-cctad-03} imply this is a tree. 
 Since this tree has height values associated with the vertices, we can talk about up-degrees and down-degrees in $\reeb(\MM)$.
@@
 
-\InSoCGVer{See \Sec{techmarks} for further technical remarks on the above definitions.}
+\InSoCGVer{See the full version \CiteFullVer for further technical remarks on the above definitions.}
 \newcommand{\technicalRemarks}{
@@
 
-Using ideas from~\cite{csa-cctad-00}, we prove that the contour tree of an extremum dominant complex is (basically) just the \emph{join tree}, 
+Using ideas from~\cite{csa-cctad-03}, we prove that the contour tree of an extremum dominant complex is (basically) just the \emph{join tree}, 
 which unlike the contour tree tracks superlevel sets rather than level sets.
@@
 merge red and blue, until yellow and blue merge. 
-A critical insight is the use of \emph{binomial heaps}~\cite{Vu78} to handle all the priorities and the merging.
-This ensures that all comparisons are only made between points that are ancestor-descendant in the join tree.
+A critical insight is the use of \emph{binomial heaps}~\cite{Vu78}, or any other efficiently mergeable heaps, to handle all the priorities and the merging.
+This ensures that all comparisons are only made between points that are ancestor-descendant in the join tree or adjacent in $\MM$.
 There are numerous details to handle to prove the final run time bound, but this should provide the reader
@@
 \subsection{Breaking $\MM$ into simpler pieces} \label{sec:break}
-Let us begin with the algorithm of Carr, Snoeyink, and Axen~\cite{csa-cctad-00}. The key insight is to build
+Let us begin with the algorithm of Carr, Snoeyink, and Axen~\cite{csa-cctad-03}. The key insight is to build
 two different trees, called the join and split trees, and then merge them together into the contour tree.
@@
 
-A major insight of~\cite{csa-cctad-00} is an ingeniously simple linear time procedure to construct
+A major insight of~\cite{csa-cctad-03} is an ingeniously simple linear time procedure to construct
 the contour tree from the join and split trees. So the bottleneck is computing these trees. Observe
@@
 these trees forces a global sort of the splits, an unnecessary computation for the contour tree. Unfortunately, in general (i.e.\ unlike \Fig{sorting}) 
-the heights of joins and splits may be interleaved in a complex manner, and hence the final merging of~\cite{csa-cctad-00}
+the heights of joins and splits may be interleaved in a complex manner, and hence the final merging of~\cite{csa-cctad-03}
 to get the contour tree requires having the split vertices in the join tree. Without
@@
 the join tree. The formal argument is a little involved, and we employ the merging procedure
-of~\cite{csa-cctad-00} to get a proof. We demonstrate that the merging procedure will
+of~\cite{csa-cctad-03} to get a proof. We demonstrate that the merging procedure will
 actually just output the join tree, so we do not need to actually compute the split tree.
@@
 
-We stress the importance of this step for our approach. Given the algorithm of~\cite{csa-cctad-00},
+We stress the importance of this step for our approach. Given the algorithm of~\cite{csa-cctad-03},
 one may think that it suffices to design faster algorithms for join trees. But this
@@
 We arrive at the main challenge: how to sort only the comparable critical points, without
-constructing the join tree? The join tree algorithm of~\cite{csa-cctad-00} is a typical event-based
+constructing the join tree? The join tree algorithm of~\cite{csa-cctad-03} is a typical event-based
 computational geometry algorithm. We have to step away from this viewpoint to avoid the global sort.
@@
 
-\InSoCGVer{The following theorems are intuitively obvious, and are proven in \Sec{proofOfSurgery}.}
+\InSoCGVer{The following theorems are intuitively obvious, and are proven in the full version \CiteFullVer.}
 
@@
 \begin{definition} \label{def:dom} A simplicial complex is \emph{minimum dominant} if there exists
-a minimum $x$ such that every non-minimal \emph{vertex} in the manifold has a non-ascending path to $x$.
+a minimum $x$ such that every non-minimal \emph{vertex} in the complex has a non-ascending path to $x$.
 Analogously define \emph{maximum dominant}. 
@@
 The water will flow downwards along non-ascending paths and ``wet'' all the points encountered. Note that this procedure considers all points
-of the manifold, not just vertices.
+of the complex, not just vertices.
 
@@
 Let $\eps, \delta$ be sufficiently small as usual. Fix some $y$ at the interface.
-Note that $y \in \wet(x,\MM)$. (Otherwise, $B_\eps(y)$ is dry.)
-The points in $B_\eps(y)$ that lie below $y$ have a descending path from $y$ and hence must be wet.
-There must also be a dry point in $B_\eps(y)$ that is above $y$, and hence,
+Note that $y \in \wet(x,\MM)$. (Otherwise, $B_\eps(y)\cap \MM$ is dry.)
+The points in $B_\eps(y)\cap \MM$ that lie below $y$ have a descending path from $y$ and hence must be wet.
+There must also be a dry point in $B_\eps(y)\cap \MM$ that is above $y$, and hence,
 there exists a dry, regular $(f(y)+\delta)$-contour $\phi$ intersecting $B_\eps(y)$.
@@
 
-\section{Contour trees of extremum dominant manifolds} \label{sec:extreme}
-
-The previous section allows us to restrict attention to extremum dominant manifolds.
+\section{Contour trees of extremum dominant complexes} \label{sec:extreme}
+
+The previous section allows us to restrict attention to extremum dominant complexes.
 We will orient so that the extremum in question is always a \emph{minimum}.
@@
 \InNotSoCGVer{
-The main theorem of this section asserts that contour trees of minimum dominant manifolds have a simple description.
+The main theorem of this section asserts that contour trees of minimum dominant complexes have a simple description.
 The exact statement will require some definitions and notation.
-We require the notions of \emph{join} and \emph{split} trees, as given by~\cite{csa-cctad-00}.
+We require the notions of \emph{join} and \emph{split} trees, as given by~\cite{csa-cctad-03}.
 Conventionally, all edges are directed from higher to lower function value. 
@@
 \InSoCGVer{
-The main theorem of this section asserts that contour trees of minimum dominant manifolds have a simple description. 
+The main theorem of this section asserts that contour trees of minimum dominant complexes have a simple description. 
 Intuitively, the cutting procedure of \Sec{rain} introduced ``stumps'' where we made cuts, which 
-is why we allowed for non-dominant minima and maxima in the definition of extremum dominant manifolds.  
-The punchline is that, ignoring these stumps, the contour tree of an extremum dominant manifold is equivalent to its \emph{join} (or \emph{split}) tree, defined in~\cite{csa-cctad-00}.
+is why we allowed for non-dominant minima and maxima in the definition of extremum dominant complexes.  
+The punchline is that, ignoring these stumps, the contour tree of an extremum dominant complex is equivalent to its \emph{join} (or \emph{split}) tree, defined in~\cite{csa-cctad-03}.
 Hence the critical join tree below is just the join tree without these stumps.  
-Additional definitions and proofs are given in \Sec{jstedm}.
+Additional definitions and proofs are given in the full version \CiteFullVer.
 }
@@
 
-\begin{lemma} \label{lem:split} $\cS(\MM)$ consists of:
+\begin{lemma} \label{lem:split} For minimum dominant $\MM$, $\cS(\MM)$ consists of:
 \begin{asparaitem}
@@
 
-Using the above lemma we can now prove that computing the contour tree for a minimum dominant manifold 
+Using the above lemma we can now prove that computing the contour tree for a minimum dominant complex 
 amounts to computing its join tree.  Specifically, to prove our main theorem, we rely on the correctness of the
-merging procedure from~\cite{csa-cctad-00} that constructs the contour tree from the join and split trees. 
+merging procedure from~\cite{csa-cctad-03} that constructs the contour tree from the join and split trees. 
 It actually constructs the \emph{augmented contour tree} $\cA(\MM)$, which is 
@@
 
-\InSoCGVer{Each non-dominant minimum, $m_i$, connects to the contour tree at a corresponding split $s_i$. We have the following (see \Sec{jstedm} for more details).}
+\InSoCGVer{Each non-dominant minimum, $m_i$, connects to the contour tree at a corresponding split $s_i$. We have the following (see the full version \CiteFullVer for more details).}
 
@@
 \begin{definition} \label{def:paint2} Let the 1-skeleton of $\MM$ have edge set $E$ and maxima $X$.
-A  \emph{painting} of $\MM$ is a map $\chi:X \cup E \to [|X|]$ with the following property. 
+A  \emph{painting} of $\MM$ is a map $\chi:X \cup E \to [|X|]$, where $\chi(z)$ is referred to as the \emph{color} of $z$,
+with the following property. 
  Consider an edge $e$. There exists a descending path from some maximum $x$ to $e$
@@
 {\bf Attachment vertex $\h(c)$:} For each color $c$, we maintain a critical point $\h(c)$ of this color.
-We will maintain the guarantee that the portion of the contour tree above (and including) $\h(c)$ has already been constructed.
+We will maintain the guarantee that the portion of the join tree above (and including) $\h(c)$ has already been constructed.
 %This is initialized to the maximum corresponding to $c$.
@@
 as $v$ is the lowest processed vertex of this merged color (as by induction $\h(c)$ and $\h(c')$ were the lowest vertices before merging colors).
-
+%
+%
 % Consider $w$, the lowest critical vertex in $\PP$. Note that $w$ is a parent of $v$. When $w$ was processed, it was set to $\h(c')$, for
@@
 Initially the union find data structure has a singleton component for each leaf (and no new components are ever created), 
-and so each union-find operation takes $O(\alpha(t))$ time.
+and so each union-find operation takes $O(\alpha(t))$ amortized time.
 For $\update$, by \Obs{twoQueues}, each iteration of the while loop requires a constant number of finds (and no unions).
@@
 Now when $v$ is handled, three queue operations are performed.  Specifically, $v$ is removed from the queues of $c_1$ and $c_2$, and then the queues are 
-are merged together.  By the above arguments the sizes of the queues for each of these operations are $H_{c_1}$, $H_{c_2}$, and $H_v$, respectively.  
+are merged together.  By the above arguments the sizes of the queues for each of these operations are $|H_{c_1}|$, $|H_{c_2}|$, and $|H_v|$, respectively.  
 As merging and deleting takes logarithmic time in the heap size for binomial heaps, the claim now follows.
@@
 
-Given the above technical definition, the proof of the following lemma is straightforward, though long and so has been moved to \Sec{runTimeProof}.
+Given the above technical definition, the proof of the following lemma is straightforward, though long and so has been moved to the full version \CiteFullVer.
 
@@
 
-\InSoCGVer{Our main result, } \Thm{main-corr}, is an easy corollary of the above lemma. Specifically, consider a critical point $v$ of the initial input 
+\InSoCGVer{Our main result, }\Thm{main-corr}\InSoCGVer{,} is an easy corollary of the above lemma.
+\begin{proof}[Proof of \Thm{main-corr}]
+Consider a critical point $v$ of the initial input 
 complex.  By \Thm{jordan} this vertex appears in exactly one of the pieces output by $\rain$.  As in the \Thm{main-corr} statement, let $\ell_v$ 
@@
 Moreover, $|H_v|$ only counts vertices in a $v$ to root path and so trivially $|H_v|\leq \ell_v'$, implying \Thm{main-corr}.
+\end{proof}
 
@@
 This slack allows for the more refined upper and lower bounds mentioned in \Sec{more-refined}.  Quantifying this slack however is quite challenging, 
-and requires a significantly more sophisticated analysis involving path decompositions, which is the subject of \Sec{pathDecomp} and \Sec{lb}.
+and requires a significantly more sophisticated analysis involving path decompositions, detailed in the full version \CiteFullVer.
 
@@
  Let $p$ be any path in $P(T)$ and let $\{q_1, \dots q_k\}$ be the children on $p$.
-Then $H_{\ell(p)} + \sum_{i=1}^k |H_{q_i}| \leq |H_p|+2|p|$. 
+Then $|H_{\ell(p)}| + \sum_{i=1}^k |H_{q_i}| \leq |H_p|+2|p|$. 
 \end{lemma}
@@
 
-Now one of the key observations was that for extremum dominant manifolds, computing the contour tree is roughly the same as 
+Now one of the key observations was that for extremum dominant complexes, computing the contour tree is roughly the same as 
 computing the join tree.  Specifically, \Thm{contour-tree} implies that given $\jc(\MM_i)$ , 
@@
 The following is the equivalent of \Thm{main-lb} for join trees, and immediately follows from
-the previous lemmas.
+the previous lemmas.  Note that the phrasing of the two theorems differ, but can be connected by observing that for any value of $\thmConst$ from \Thm{main-lb}, 
+there is a path decomposition $P$ of some tree $T$ such that $\thmConst= \Theta(\sum_{p \in P} |p|\log |p|)$.
+
 %Call a cost $C$ join tree valid if $C = cost_{\jc}(f)$ for some $f$ defined over a given complex $\MM$.
@@
 We now restate \Thm{main-lb}, which follows immediately from an entropy argument, analogous to \Thm{joinLB}.
+Again, as discussed above for \Thm{joinLB}, the phrasing differs from \Thm{main-lb}.
 
@@
 
-%\intuitionSection
-
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 
-\section{Join and Split Trees of Extremum Dominant Manifolds}
+\section{Join and Split Trees of Extremum Dominant Complexes}
 \label{sec:jstedm}
 In this section we prove Theorem~\ref{thm:contour-tree} (of \Sec{extreme}), which states the equivalence 
-of the contour tree on extremum dominant manifolds and the critical join tree of Definition~\ref{def:criticalJoin}.
+of the contour tree on extremum dominant complexes and the critical join tree of Definition~\ref{def:criticalJoin}.
 This proof requires some definitions and notation, specially
-the notions of \emph{join} and \emph{split} trees, as given by~\cite{csa-cctad-00}.
+the notions of \emph{join} and \emph{split} trees, as given by~\cite{csa-cctad-03}.
 Conventionally, all edges are directed from higher to lower function value.
@@
 
-% We now prove Theorem~\ref{thm:contour-tree} (of Section~\ref{sec:extreme}), which states the equivalence 
-% of the contour tree on extremum dominant manifolds and the critical join tree of Definition~\ref{def:criticalJoin}.
 We are now ready to prove Theorem~\ref{thm:contour-tree}.
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-%\algorithmRunTimeSection
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%reeb graph
+
 \section{Proof of \Lem{runTimeUpper}}
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-%\InsertAppendixOfProofs
 }

### 2016-03-21 paint-contour-tree.tex
done with edits for camera ready version

@@
 We think of the dimension $d$, as constant, and assume that $\MM$ is represented in a data structure that allows constant-time access to neighboring simplices
-in $\MM$ (e.g.~\cite{BoMa12}). (This is analogous to a doubly connected edge list, but for higher dimensions.)
+in $\MM$ (e.g. incidence graphs~\cite{Ed87}). (This is analogous to a doubly connected edge list, but for higher dimensions.)
 Observe that $f:\MM \rightarrow \RR$ can be thought of as a $d$-dimensional simplicial complex living in $\RR^{d+1}$, 
@@
 For any color $c$, $\rep(c)$ denotes the representative of its class.
+
+\medskip
+\noindent
+{\bf Color list $\col(v)$:} For each point $v$, we maintain $\col(v)$ as a simple list.
+In addition, we will maintain another (sub)list of colors $L$ such that $\forall c \in L$,
+$v$ is \emph{not} the highest vertex in $T(\rep(c))$. Note that given $c \in \col(v)$
+and $\rep(c)$, this property can be checked in constant time. If $c$ is ever removed
+from this sublist, it can never enter it again.
+
+For notational convenience, we will not explicitly
+maintain this sublist. We simply assume that, in constant
+time, one can determine (if it exists) an arbitrary color $c \in \col(v)$ such that $v$ is not the highest vertex in $T(\rep(c))$.
 
@@
 \begin{lemma} \label{lem:full} Suppose vertex $v$ is connected to a component $\PP$ of $\MM^+_v$
-by an edge $e$ which is currently colored $c$. Either all edges in $\PP$ are currently colored $c$, or there
+by an edge $e$ which is currently colored $c$. At all times: either all edges in $\PP$ are currently colored $c$, or there
 exists a critical vertex $w \in \PP$ fully touched by $c$ and touched by another color.

### 2016-03-22 paint-contour-tree.tex

@@
 
+% \begin{figure}[h]\centering
+%     \includegraphics[width=.26\linewidth]{figs/spikey1}%
+%     \hspace{.35in}
+%     \includegraphics[width=.26\linewidth]{figs/spikey2}%
+%     \hspace{.4in}
+%     \includegraphics[width=.215\linewidth]{figs/spikeyTree}
+%     \caption{Two surfaces with different orderings of the maxima, but the same contour tree.}
+%     \label{fig:relative}
+% \end{figure}
+
 \begin{figure}[h]\centering
-    \includegraphics[width=.26\linewidth]{figs/spikey1}%
-    \hspace{.35in}
-    \includegraphics[width=.26\linewidth]{figs/spikey2}%
-    \hspace{.4in}
-    \includegraphics[width=.215\linewidth]{figs/spikeyTree}
+    \includegraphics[width=.23\linewidth]{figs/spikey1}%
+    \hspace{.37in}
+    \includegraphics[width=.23\linewidth]{figs/spikey2}%
+    \hspace{.41in}
+    \includegraphics[width=.21\linewidth]{figs/spikeyTree}
     \caption{Two surfaces with different orderings of the maxima, but the same contour tree.}
@@
 is in $\wet(x,\NN)$, so $\NN$ is trivially extremum dominant. So suppose the interface
-is non-empty and consists of $\phi_1, \phi_2, \ldots, \phi_k$ (as denoted in the procedure).
+is non-empty and consists of $\phi_1, \ldots, \phi_k$ (as denoted in the procedure).
 By repeated applications of \Clm{cut-int}, $\NN_{k+1}$ contains $\wet(x,\MM)$. 
-Consider $\wet(x,\NN_{k+1})$. The interface must exactly be $\phi_1, \phi_2, \ldots, \phi_k$.
-So the only dry vertices are those in the boundaries $B_1, B_2, \ldots, B_k$. But these
+Consider $\wet(x,\NN_{k+1})$. The interface is exactly $\phi_1, \ldots, \phi_k$.
+So the only dry vertices are those in the boundaries $B_1, \ldots, B_k$. But these
 boundaries are maxima.
@@
 
-An \emph{initial} painting has the additional property that the restriction $\chi:X \to [|X|]$ is a bijection.
+An \emph{initial} painting also requires that the restriction $\chi:X \to [|X|]$ is a bijection.
 \end{definition}
@@
 
-\medskip
+\smallskip
 \fbox{
@@
 
-\medskip
+%\medskip
 \ignore{	
@@
 
-\InSoCGVer{
-\appendix
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\section{Some technical remarks}
-\label{sec:techmarks}
-The following is a continuation from the end of \Sec{basics}.
-
-\technicalRemarks
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\section{Proof of Surgery}
-\label{sec:proofOfSurgery}
-\surgeryBody
-That is, Theorem~\ref{thm:jordan}, which is proven below.
-\proofofDisconnected
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\section{Join and Split Trees of Extremum Dominant Complexes}
-\label{sec:jstedm}
-In this section we prove Theorem~\ref{thm:contour-tree} (of \Sec{extreme}), which states the equivalence 
-of the contour tree on extremum dominant complexes and the critical join tree of Definition~\ref{def:criticalJoin}.
-This proof requires some definitions and notation, specially
-the notions of \emph{join} and \emph{split} trees, as given by~\cite{csa-cctad-03}.
-Conventionally, all edges are directed from higher to lower function value.
-
-\joinTreeSplitTreeMerge
-
-We are now ready to prove Theorem~\ref{thm:contour-tree}.
-\proofofContourJoinEquiv
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%reeb graph
-
-\section{Proof of \Lem{runTimeUpper}}
-\label{sec:runTimeProof}
-\proofofRunTime
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\leafAssignPathDecomp
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\lowerBoundbyPathDecomp
-
-}
+% % 
+% % 
+% % \InSoCGVer{
+% % \appendix
+% % 
+% % %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% % 
+% % \section{Some technical remarks}
+% % \label{sec:techmarks}
+% % The following is a continuation from the end of \Sec{basics}.
+% % 
+% % \technicalRemarks
+% % 
+% % %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% % 
+% % \section{Proof of Surgery}
+% % \label{sec:proofOfSurgery}
+% % \surgeryBody
+% % That is, Theorem~\ref{thm:jordan}, which is proven below.
+% % \proofofDisconnected
+% % 
+% % %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% % 
+% % \section{Join and Split Trees of Extremum Dominant Complexes}
+% % \label{sec:jstedm}
+% % In this section we prove Theorem~\ref{thm:contour-tree} (of \Sec{extreme}), which states the equivalence 
+% % of the contour tree on extremum dominant complexes and the critical join tree of Definition~\ref{def:criticalJoin}.
+% % This proof requires some definitions and notation, specially
+% % the notions of \emph{join} and \emph{split} trees, as given by~\cite{csa-cctad-03}.
+% % Conventionally, all edges are directed from higher to lower function value. 
+% % 
+% % \joinTreeSplitTreeMerge
+% % 
+% % We are now ready to prove Theorem~\ref{thm:contour-tree}.
+% % \proofofContourJoinEquiv
+% % 
+% % %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%reeb graph
+% % 
+% % \section{Proof of \Lem{runTimeUpper}}
+% % \label{sec:runTimeProof}
+% % \proofofRunTime
+% % 
+% % %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% % 
+% % \leafAssignPathDecomp
+% % 
+% % %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% % 
+% % \lowerBoundbyPathDecomp
+% % 
+% % }
+% %
 

### 2016-07-08 paint-contour-tree.tex
typos corrected

@@
 
-Essentially, the ``run time per critical point" is the height/depth of the point in the contour tree.
+Essentially, the ``run time per critical point" is determined by the height/depth of the point in the contour tree.
 This bound immediately yields a run time of $O(t\log D + t\alpha(t) + N)$,
@@
 by $\reeb(\MM)$. The corresponding node and edge sets are denoted as $\cV(\cdot)$ and $\cE(\cdot)$.
-It is not immediately obvious that this graph is a tree, but alternate definitions of the contour tree
+It is not immediately obvious that this graph is a tree, but alternative definitions of the contour tree
 in~\cite{csa-cctad-03} imply this is a tree. 
@@
 the path decomposition $P(T)$. Abusing notation, we simply use $P(T)$ to denote the shrub.
-As a result, we use terms like `parent', `child', `sibling', etc. for paths as well.
+As a result, we use terms like `parent', `child', `sibling', etc.\ for paths as well.
 The shrub gives a handle on the heaps of a path.
-We use $b(p)$ to denote the \emph{base} of the path, which is vertex in $p$ closest
+We use $b(p)$ to denote the \emph{base} of the path, which is the vertex in $p$ closest
 to root of $T$. We use $\ell(p)$ to denote the leaf in $p$.
-We use $H_p$ to denote the $H_{b(p)}$.
+We use $H_p$ to denote $H_{b(p)}$.
 % and denote it by $\pathTree(T)$.
@@
 \label{lem:adjacent}
- Let $p$ be any path in $P(T)$ and let $\{q_1, \dots q_k\}$ be the children on $p$.
+ Let $p$ be any path in $P(T)$ and let $\{q_1, \dots q_k\}$ be the children of $p$.
 Then $|H_{\ell(p)}| + \sum_{i=1}^k |H_{q_i}| \leq |H_p|+2|p|$. 
@@
 (If not, we have reached the root $r$ of $\pmax(T)$. Hence, $p$ is supported.)
-Insert edge from $p$ to $q$, so $q$ is the parent of $p$ in $\cU$. This construction leads
+Insert an edge from $p$ to $q$, so $q$ is the parent of $p$. This construction leads
 to the shrub forest of $\cL \setminus \cL'$, where all the roots are supported short paths, 
@@
 
-We split the summation into tall, short, and unsupported short paths.
+We split the summation into tall, supported short, and unsupported short paths.
 \begin{eqnarray*} 
-\sum_{p \in \cL} \sum_{v \in p} \log |H_v| & = & \sum_{p \in \cL \setminus \cL'} \sum_{v \in p} \log |H_v| + \sum_{p \in \cL'} \sum_{v \in p} \log |H_v| + \sum_{p \in \cH} \sum_{v \in p} \log |H_v|
+\sum_{p \in \pmax(T)} \sum_{v \in p} \log |H_v| & = & \sum_{p \in \cL \setminus \cL'} \sum_{v \in p} \log |H_v| + \sum_{p \in \cL'} \sum_{v \in p} \log |H_v| + \sum_{p \in \cH} \sum_{v \in p} \log |H_v|
 %& = & \sum_{p \in \cL \setminus \cL'} \sum_{v \in p} \log |H_v| + O(\sum_{p \in \cL'} |p|\log |H_p|) = \sum_{p \in \cL \setminus \cL'} \sum_{v \in p} \log |H_v| + O(\sum_{p \in \cL'} |H_p|\log |H_p|)
@@
 The last term can be bounded by $O(\sum_{p \in \pmax(T)} |p|\log |p|)$, by \Lem{pathBounds}.
-The second term can be bounded by $O(\sum_{p \in \cL'} |H_p| \log |H_p|)$, by \Lem{pathBounds} again.
-The following claim shows that this in turn is at most the last term.
+The second term can be bounded by $O(\sum_{p \in \cL'} |H_p| \log |H_p|)$, again by \Lem{pathBounds}.
+The following claim shows that this in turn is asymptotically bounded by the last term.
 
@@
 We have $\sum_{p \in \cL \setminus \cL'} \sum_{v \in p} \log |H_v| \allowbreak \leq \sum_{\cU} \sum_{p \in \cU} \sum_{v \in p} \log |H_v|$.
-By \Lem{cu-root}, $|H_v| = O(|H_r|)$, where $r$ is the root of $\cU$.
-Furthermore, $\sum_{p \in \cU} |p| = O(|H_r|)$. 
+By \Lem{cu-root}, $|H_v| = O(|H_r|)$, where $r$ is the root of $\cU$, and furthermore, $\sum_{p \in \cU} |p| = O(|H_r|)$. 
 We have $\sum_{p \in \cU} \sum_{v \in p} \log |H_v| = O((\log |H_r|) \sum_{p \in \cU} |p|) = O(|H_r|\log |H_r|)$.
@@
 $q_1, q_2, \ldots, q_k$ in $\cU$. Then $R_p = |\redH_p| - \sum_i |\redH_{q_i}|$. 
-By definition, $|\redH_p| = \sum_{q \in \cU_p} R_p$. 
+By definition, $|\redH_p| = \sum_{q \in \cU_p} R_q$. 
 Note that $R_p$ can be negative. 
-Now, define $R^+_p = \max(R_p,0)$, and set $W_p = \sum_{q \in \cU_p} R^+_p$.
+Now, define $R^+_p = \max(R_p,0)$, and set $W_p = \sum_{q \in \cU_p} R^+_q$, 
+where $\cU_p$ is the subtree of $\cU$ rooted at $p$.
 Observe that $W_p \geq |\redH_p|$. We also get an approximate converse.
@@
 
-\begin{proof} We write $W_p - |\redH_p| = \sum_{q \in \cU_p} R^+_q - R_q$ $= -\sum_{q \in \cU_p: R_q < 0} R_q$.
+\begin{proof} We write $W_p - |\redH_p| = \sum_{q \in \cU_p} (R^+_q - R_q)$ $= -\sum_{q \in \cU_p: R_q < 0} R_q$.
 Consider $q \in \cU_p$ and denote the children in $\cU_p$ by $q'_1, q'_2, \ldots$.

### 2016-07-27 paint-contour-tree.tex
section 8.2 added

@@
 which involves routine calculations. Then we prove \Lem{cu-root}, where the interesting work happens.
+
+\newcommand{\buff}{\mathcal{B}}
+\newcommand{\loss}{\lambda}
+\subsection{Pruning Away Buffered Sub-shrubs}
+Let $\cU$ be a shrub in $\cL \setminus \cL'$.  For any node $q$ in the shrub $\cU$, let $TSC(q)$ denote the set of all vertices of $T$ from all paths in the tall support chain of $q$.
+For $q$ in $\cU$ we then define the \emph{loss} to be $\loss(q) = TSC(q)\cap H_q$. 
+Intuitively, if $c$ is a child node of a parent node $p$ in $\cU$, then $\loss(c)$ are the vertices ``lost'' from the heap of $c$ to its tall support chain, by the time one reaches $p$.
+
+% In the following, any connected subgraph of a shrub will be called a \emph{sub-shrub} (i.e. it may not be a proper subtree).
+% 
+% \begin{definition}
+%  Let $\cU$ denote a shrub of $\cL \setminus \cL'$.  
+%  Then a sub-shrub $\buff$ of $\cU$ is called \emph{buffered} if the following holds:
+%  \begin{enumerate}
+%   \item $\sum_{p\in \buff} \loss(p) \geq \frac{1}{100} \sum_{p\in \buff} |p|$
+%   \item $\buff$ is minimal, i.e.\ there is no sub-shrub of $\buff$ for which the above inequality holds.  
+%  \end{enumerate}
+% \end{definition}
+
+Ultimately our goal in this subsection is to prune away subtrees from $\cU$ which can be paid for by charging to their tall support chains (similarly to supported short paths).
+The remaining portion of the shrub will then be handled in a later subsection.
+We now formally define this pruning process, which is quite intuitive if one thinks of $\cU$ as an actual physical shrub.
+
+\begin{definition}
+  Let $\cU$ denote a shrub of $\cL \setminus \cL'$.
+A \emph{pruning} $\cU'$ of $\cU$ is any subgraph defined by some number of rounds of the following iterative process.
+Initially set $\cU' = \cU$.
+In each iteration select some subtree $\cU'_r$ rooted at a node $r\in \cU'$ and set $\cU' = \cU'\setminus \{\cU'_r\}$.
+\end{definition}
+
+The above definition does not specify which subtrees are pruned, which we now formally define.
+
+\begin{definition}\label{def:buff}
+ Let $\cU'$ be a pruning of a shrub of $\cU$ of $\cL \setminus \cL'$. Then a rooted subtree $\buff$ of $\cU'$ is called a \emph{buffered sub-shrub} if the following conditions hold:
+  \begin{enumerate}
+  \item $\sum_{p\in \buff} |\loss(p)| \geq \frac{1}{100} \sum_{p\in \buff} |p|$
+  \item No rooted subtree of $\buff$ satisfies the above inequality (other than $\buff$ itself). 
+ \end{enumerate}
+\end{definition}
+
+% \begin{definition}
+%  Let $\cU$ denote a shrub of $\cL \setminus \cL'$.
+We define an interative  procedure which ``prunes'' $\cU$.
+Specifically, let $\cU'$ denote the current remaining portion of the shrub, so initially, $\cU' = \cU$.
+A single iteration consists of the following: Visit the nodes in $\cU'$ in a reverse topological order (i.e.
+leaves first).
+Let $q$ be the current node being considered, and let $\cU'_q$ be the subtree rooted at $q$.
+We say $\cU'_q$ is buffered if the following condition holds
+%  \[
+%   \sum_{p\in \cU'_q} \loss(p) \geq \frac{1}{100} \sum_{p\in \cU'_q} |p|
+%  \]
+%  An iteration stops when we reach a node $q$ defining a buffered subtree, at which point we set $\cU' = \cU'\setminus{\cU'_q}$.
+% \end{definition}
+
+Ultimately we will prune buffered subtrees from $\cU$ by starting at the leaves and working towards the root, and hence the second condition of buffered will trivially be satisfied.
+
+We require the following lemma so that we can charge to the root in a buffered subtree.
+
+\begin{lemma}
+ Let $p$ be any node in a buffered sub-shrub $\buff$, and let $d$ be any descendant node of $p$ in $\buff$.  Then for any $v \in d$, $|H_v| = O(|H_p|)$.
+\end{lemma}
+\begin{proof}
+ Missing.  \Lem{cu-root} will be rephrased to handle this slightly more general case (i.e. here $p$ may not be the root of $\cU$). Note the proof will not 
+ have to be changed as \Lem{geometric} is the proof which already implies general case.
+\end{proof}
+
+\begin{lemma}
+ Let $\buff$ be a buffered sub-shrub.  Then $\sum_{p\in \buff}\sum_{v\in p} \log |H_v| = O(\sum_{p\in \buff}\sum_{v\in \loss(p)} \log |H_v| )$
+\end{lemma}
+\begin{proof}
+ ????  Ignore this ``proof'', it is just a graveyard of observations. ????
+
+ Let $r$ be the root of $\buff$.
+First observe that $\loss(r) \geq |r|/100$.
+Suppose otherwise, then in order for the first condition of \Def{buff} to be satisfied for $\buff$, the rooted subtree of some child of $r$, call it $\buff'$, must satisfy $\sum_{p\in \buff'} \loss(p) \geq \frac{1}{100} \sum_{p\in \buff'} |p|$.
+However, this would contradict the second condition of \Def{buff}.
+
+ Consider any node $q\in \buff$.
+Then for any vertex $v\in \loss(q)$, we have $|H_v| = \Omega(|H_q|)$, since $q$ is unsupported and hence by definition less than $|H_q|/100$ vertices of $H_q$ appear in $\loss(q)$.
+
+ \begin{align*}
+  \sum_{p\in \buff}\sum_{v\in p} \log |H_v| &=  O\left( \sum_{p\in \buff}\sum_{v\in p} \log |H_p| \right)   &\mbox{definition of short} \\
+  &= O\left( \sum_{p\in \buff} |p| \log |H_p| \right)   &\mbox{} \\
+  &= ????   &\mbox{buffered is not sufficient to get to next line} \\
+  &= O\left( \sum_{p\in \buff} |\loss(p)| \log |H_p| \right)   &\mbox{} \\
+  &= O\left( \sum_{p\in \buff}\sum_{v\in \loss(p)} \log |H_v| \right)
+ \end{align*}
+\end{proof}
+
+Let $p$ and $q$ be two distinct paths in $\cL \setminus \cL'$.  Note that while it is possible that $TSC(p)\cap TSC(q)\neq \emptyset$, 
+it holds that $\loss(p)\cap \loss(q) = \emptyset$.  Thus the above lemma implies that the total heap costs due to vertices in buffered sub-shrubs can be charged to those in tall paths.
+
+\begin{corollary}
+ Let $B= \{\buff_1, \ldots, \buff_k\}$ be any collection of disjoint buffered sub-shrubs from $\cL \setminus \cL'$. Then, 
+ $\sum_{i=1}^k \sum_{p\in \buff_i}\sum_{v\in p} |H_v| = O(\sum_{p\in \cH}\sum_{v\in p} \log |H_v| )$
+\end{corollary}
 

### 2016-07-28 paint-contour-tree.tex
section 8 updated

@@
 
-The paths in $P(T)$ naturally define a tree\footnote{Please excuse the 
+The paths in $\pmax(T)$ naturally define a tree\footnote{Please excuse the 
 overloading of the term 'tree', it is the most natural term to use here.} of their own.  Specifically, in the original 
 tree $T$ contract each path down to its root.  Call the resulting tree the \emph{shrub} of $T$ corresponding to 
-the path decomposition $P(T)$. Abusing notation, we simply use $P(T)$ to denote the shrub.
+the path decomposition $\pmax(T)$. Abusing notation, we simply use $\pmax(T)$ to denote the shrub.
 As a result, we use terms like `parent', `child', `sibling', etc.\ for paths as well.
@@
 \label{lem:adjacent}
- Let $p$ be any path in $P(T)$ and let $\{q_1, \dots q_k\}$ be the children of $p$.
+ Let $p$ be any path in $\pmax(T)$ and let $\{q_1, \dots q_k\}$ be the children of $p$.
 Then $|H_{\ell(p)}| + \sum_{i=1}^k |H_{q_i}| \leq |H_p|+2|p|$. 
@@
 
-We wish to prove $\sum_{v\in T} \log |H_v| = O(\sum_{p\in P} |p| \log |p|)$. 
-The simplest approach is to prove $\forall p\in P$, $\sum_{v\in p} \log |H_v|= O(|p|\log|p|)$.
+We wish to prove $\sum_{v\in T} \log |H_v| = O(\sum_{p\in \pmax} |p| \log |p|)$. 
+The simplest approach is to prove $\forall p\in \pmax$, $\sum_{v\in p} \log |H_v|= O(|p|\log|p|)$.
 This is unfortunately not true, which is why we divide paths into two categories.
@@
 \begin{definition}
- For $p\in P(T)$, $p$ is \emph{short} if $|p| < \sqrt{|H_{p}|}/100$, and \emph{tall} otherwise.
+ For $p\in \pmax(T)$, $p$ is \emph{short} if $|p| < \sqrt{|H_{p}|}/100$, and \emph{tall} otherwise.
 \end{definition}
@@
 Let $\cL$ be the set of short paths, $\cL'$ be the set of supported short paths, and $\cH$ be the set of tall paths
-given by $\pmax(T)$.
-We now construct the shrub of unsupported short paths. Consider $p \in \cL \setminus \cL'$,
+given by $\pmax(T)$. We now describe a partitioning of the unsupported paths into a forest of shrubs. Consider $p \in \cL \setminus \cL'$,
 and traverse the chain of ancestors from $p$. Eventually, we must reach another short path $q$.
@@
 Insert an edge from $p$ to $q$, so $q$ is the parent of $p$. This construction leads
-to the shrub forest of $\cL \setminus \cL'$, where all the roots are supported short paths, 
+to the \emph{shrub forest} of $\cL \setminus \cL'$, denoted $\cF$, where all the roots are supported short paths, 
 and the remaining nodes are the unsupported short paths.
 
-Most of the work goes into proving the following technical lemma.
-
-\begin{lemma} \label{lem:cu-root} Let $\cU$ denote a connected component (shrub) in the shrub forest of $\cL \setminus \cL'$ and let $r$
-be the root of $\cU$. (i) For any $v \in p$ such that $p \in \cU$,
-$|H_v| = O(|H_r|)$. (ii) $\sum_{p \in \cU} |p| = O(|H_r|)$.
-\end{lemma}
-
-We split the remaining argument into two subsections. We first prove \Thm{runtime} from \Lem{cu-root},
-which involves routine calculations. Then we prove \Lem{cu-root}, where the interesting work happens.
-
-\newcommand{\buff}{\mathcal{B}}
-\newcommand{\loss}{\lambda}
-\subsection{Pruning Away Buffered Sub-shrubs}
-Let $\cU$ be a shrub in $\cL \setminus \cL'$.  For any node $q$ in the shrub $\cU$, let $TSC(q)$ denote the set of all vertices of $T$ from all paths in the tall support chain of $q$.
-For $q$ in $\cU$ we then define the \emph{loss} to be $\loss(q) = TSC(q)\cap H_q$. 
-Intuitively, if $c$ is a child node of a parent node $p$ in $\cU$, then $\loss(c)$ are the vertices ``lost'' from the heap of $c$ to its tall support chain, by the time one reaches $p$.
-
-% In the following, any connected subgraph of a shrub will be called a \emph{sub-shrub} (i.e. it may not be a proper subtree).
-% 
-% \begin{definition}
-%  Let $\cU$ denote a shrub of $\cL \setminus \cL'$.  
-%  Then a sub-shrub $\buff$ of $\cU$ is called \emph{buffered} if the following holds:
-%  \begin{enumerate}
-%   \item $\sum_{p\in \buff} \loss(p) \geq \frac{1}{100} \sum_{p\in \buff} |p|$
-%   \item $\buff$ is minimal, i.e.\ there is no sub-shrub of $\buff$ for which the above inequality holds.  
-%  \end{enumerate}
-% \end{definition}
-
-Ultimately our goal in this subsection is to prune away subtrees from $\cU$ which can be paid for by charging to their tall support chains (similarly to supported short paths).
-The remaining portion of the shrub will then be handled in a later subsection.
-We now formally define this pruning process, which is quite intuitive if one thinks of $\cU$ as an actual physical shrub.
-
-\begin{definition}
-  Let $\cU$ denote a shrub of $\cL \setminus \cL'$.
-A \emph{pruning} $\cU'$ of $\cU$ is any subgraph defined by some number of rounds of the following iterative process.
-Initially set $\cU' = \cU$.
-In each iteration select some subtree $\cU'_r$ rooted at a node $r\in \cU'$ and set $\cU' = \cU'\setminus \{\cU'_r\}$.
-\end{definition}
-
-The above definition does not specify which subtrees are pruned, which we now formally define.
-
-\begin{definition}\label{def:buff}
- Let $\cU'$ be a pruning of a shrub of $\cU$ of $\cL \setminus \cL'$. Then a rooted subtree $\buff$ of $\cU'$ is called a \emph{buffered sub-shrub} if the following conditions hold:
-  \begin{enumerate}
-  \item $\sum_{p\in \buff} |\loss(p)| \geq \frac{1}{100} \sum_{p\in \buff} |p|$
-  \item No rooted subtree of $\buff$ satisfies the above inequality (other than $\buff$ itself). 
- \end{enumerate}
-\end{definition}
-
-% \begin{definition}
-%  Let $\cU$ denote a shrub of $\cL \setminus \cL'$.
-We define an interative  procedure which ``prunes'' $\cU$.
-Specifically, let $\cU'$ denote the current remaining portion of the shrub, so initially, $\cU' = \cU$.
-A single iteration consists of the following: Visit the nodes in $\cU'$ in a reverse topological order (i.e.
-leaves first).
-Let $q$ be the current node being considered, and let $\cU'_q$ be the subtree rooted at $q$.
-We say $\cU'_q$ is buffered if the following condition holds
-%  \[
-%   \sum_{p\in \cU'_q} \loss(p) \geq \frac{1}{100} \sum_{p\in \cU'_q} |p|
-%  \]
-%  An iteration stops when we reach a node $q$ defining a buffered subtree, at which point we set $\cU' = \cU'\setminus{\cU'_q}$.
-% \end{definition}
-
-Ultimately we will prune buffered subtrees from $\cU$ by starting at the leaves and working towards the root, and hence the second condition of buffered will trivially be satisfied.
-
-We require the following lemma so that we can charge to the root in a buffered subtree.
-
-\begin{lemma}
- Let $p$ be any node in a buffered sub-shrub $\buff$, and let $d$ be any descendant node of $p$ in $\buff$.  Then for any $v \in d$, $|H_v| = O(|H_p|)$.
-\end{lemma}
-\begin{proof}
- Missing.  \Lem{cu-root} will be rephrased to handle this slightly more general case (i.e. here $p$ may not be the root of $\cU$). Note the proof will not 
- have to be changed as \Lem{geometric} is the proof which already implies general case.
-\end{proof}
-
-\begin{lemma}
- Let $\buff$ be a buffered sub-shrub.  Then $\sum_{p\in \buff}\sum_{v\in p} \log |H_v| = O(\sum_{p\in \buff}\sum_{v\in \loss(p)} \log |H_v| )$
-\end{lemma}
-\begin{proof}
- ????  Ignore this ``proof'', it is just a graveyard of observations. ????
-
- Let $r$ be the root of $\buff$.
-First observe that $\loss(r) \geq |r|/100$.
-Suppose otherwise, then in order for the first condition of \Def{buff} to be satisfied for $\buff$, the rooted subtree of some child of $r$, call it $\buff'$, must satisfy $\sum_{p\in \buff'} \loss(p) \geq \frac{1}{100} \sum_{p\in \buff'} |p|$.
-However, this would contradict the second condition of \Def{buff}.
-
- Consider any node $q\in \buff$.
-Then for any vertex $v\in \loss(q)$, we have $|H_v| = \Omega(|H_q|)$, since $q$ is unsupported and hence by definition less than $|H_q|/100$ vertices of $H_q$ appear in $\loss(q)$.
-
- \begin{align*}
-  \sum_{p\in \buff}\sum_{v\in p} \log |H_v| &=  O\left( \sum_{p\in \buff}\sum_{v\in p} \log |H_p| \right)   &\mbox{definition of short} \\
-  &= O\left( \sum_{p\in \buff} |p| \log |H_p| \right)   &\mbox{} \\
-  &= ????   &\mbox{buffered is not sufficient to get to next line} \\
-  &= O\left( \sum_{p\in \buff} |\loss(p)| \log |H_p| \right)   &\mbox{} \\
-  &= O\left( \sum_{p\in \buff}\sum_{v\in \loss(p)} \log |H_v| \right)
- \end{align*}
-\end{proof}
-
-Let $p$ and $q$ be two distinct paths in $\cL \setminus \cL'$.  Note that while it is possible that $TSC(p)\cap TSC(q)\neq \emptyset$, 
-it holds that $\loss(p)\cap \loss(q) = \emptyset$.  Thus the above lemma implies that the total heap costs due to vertices in buffered sub-shrubs can be charged to those in tall paths.
-
-\begin{corollary}
- Let $B= \{\buff_1, \ldots, \buff_k\}$ be any collection of disjoint buffered sub-shrubs from $\cL \setminus \cL'$. Then, 
- $\sum_{i=1}^k \sum_{p\in \buff_i}\sum_{v\in p} |H_v| = O(\sum_{p\in \cH}\sum_{v\in p} \log |H_v| )$
-\end{corollary}
-
-\subsection{Proving \Thm{runtime}} \label{sec:thm-runtime}
-
-We split the summation into tall, supported short, and unsupported short paths.
-\begin{eqnarray*} 
-\sum_{p \in \pmax(T)} \sum_{v \in p} \log |H_v| & = & \sum_{p \in \cL \setminus \cL'} \sum_{v \in p} \log |H_v| + \sum_{p \in \cL'} \sum_{v \in p} \log |H_v| + \sum_{p \in \cH} \sum_{v \in p} \log |H_v|
-%& = & \sum_{p \in \cL \setminus \cL'} \sum_{v \in p} \log |H_v| + O(\sum_{p \in \cL'} |p|\log |H_p|) = \sum_{p \in \cL \setminus \cL'} \sum_{v \in p} \log |H_v| + O(\sum_{p \in \cL'} |H_p|\log |H_p|)
-\end{eqnarray*}
-The last term can be bounded by $O(\sum_{p \in \pmax(T)} |p|\log |p|)$, by \Lem{pathBounds}.
-The second term can be bounded by $O(\sum_{p \in \cL'} |H_p| \log |H_p|)$, again by \Lem{pathBounds}.
-The following claim shows that this in turn is asymptotically bounded by the last term.
-
-%We have a claim that also bounds heap sizes for special set of paths.
-
-\begin{claim} \label{clm:above} $\sum_{p \in \cL'} |H_p| \log |H_p| = O(\sum_{q \in \cH} \sum_{v \in q} \log |H_v|)$.
-\end{claim}
-
-\begin{proof} 
-Pick $p \in \cL'$. 
-%Let the first $|H_p|/200$ vertices the lie in tall paths be denoted as $H'_p$. 
-%Observe that for any $v \in H'_p$, 
-%We have $|H_p| \log |H_p| = O(\sum_{v \in H'_p} \log |H_p|)$
-%$= O(\sum_{v \in H'_p} \log |H_v|)$. Since all the vertices lie in tall paths,
-%we can write this as $O(\sum_{q \in \cH} \sum_{v \in H'_p \cap q} \log |H_v|)$.
-%Summing over all $p$, the expression is $\sum_{q \in \cH} \sum_{p \in \cL'} \sum_{v \in H'_p \cap q} \log |H_v|$.
-%
-%
-As we traverse the tall support chain of $p$, there
-are at least $|H_p|/100$ vertices of $H_p$ that lie in these paths. These are encountered in
-a fixed order. Let $H'_p$ be the first $|H_p|/200$ of these vertices. When $v \in H'_p$
-is encountered, there are $|H_p|/200$ vertices of $H_p$ not yet encountered. Hence,
-$|H_v| \geq |H_p|/200$. Hence, $|H_p|\log |H_p| = O(\sum_{v \in H'_p} \log |H_v|)$.
-Since all the vertices lie in tall paths,
-we can write this as $O(\sum_{q \in \cH} \sum_{v \in H'_p \cap q} \log |H_v|)$.
-Summing over all $p$, the expression is $\sum_{q \in \cH} \sum_{p \in \cL'} \sum_{v \in H'_p \cap q} \log |H_v|$.
-%and $\sum_{p \in \cL'} |H_p|\log |H_p| = O(\sum_{p \in \cL'} \sum_{v \in H'_p} \log |H_v|)$.
-
-Consider any $v \in H'_p$.  Let $S$ be the set of paths $\widetilde{p}\in \cL'$ such that 
-$v\in H'_{\widetilde{p}}$.  We now show $|S|\leq 2$ (i.e.\ it contains at most one path other than $p$).
-First observe that any two paths in $S$ must be unrelated (i.e.\ $S$ is an anti-chain), 
-since paths which have an ancestor-descendant relationship have disjoint tall support chains.  
-However, any vertex $v$ receives exactly one color from each of its two subtrees (in $T$), and therefore $|S|\leq 2$ since any two 
-paths which share descendant leaves in $T$ (i.e.\ their heaps are tracking the same color) must have an ancestor-descendant relationship.
-
-In other words, any $\log |H_v|$ appears at most twice in the above triple summation.
-Hence, we can bound it by $O(\sum_{q \in \cH} \sum_{v \in q} \log |H_v|)$.
-%
-%Pick $q \in Q$ and some $v \in H_q$. As we traverse up the tree passing
-%through $A(q)$, we must encounter some $p \in A(q)$ such that $v \in p$. Furthermore,
-%there is an order in which these vertices (of $H_q$) occur. Let $H'$ be the first half
-%of vertices that occur. For all $v \in H'$, $|H_v| \geq |H_q|/2$. Hence,
-%$|H_q| \log |H_q| \leq 4 \sum_{v \in H'} \log |H_v| \leq 4 \sum_{v \in H_q} |H_v|$.
-%So $\sum_{q \in Q} |H_q| \log |H_q| \leq 4 \sum_{q \in Q} \sum_{v \in H_q} |H_v|$.
-%
-%Since nodes in $Q$ are incomparable, all the $H_q$'s (for $q \in Q$) are disjoint.
-%Each such $v$ must occur in some path $p \in A(Q)$. Thus,
-%$\sum_{q \in Q} |H_q| \log |H_q| \leq 4 \sum_{p \in A(Q)} \sum_{v \in p} |H_v|$.
-%We apply the bound of \Lem{pathBounds} to complete the proof.
-\end{proof}
-
-The first term (unsupported short paths) can be charged to the second term (supported short paths).
-This is where the critical \Lem{cu-root} plays a role.
-
-\begin{claim} \label{clm:first} $\sum_{p \in \cL \setminus \cL'} \sum_{v \in p} \log |H_v| = O(\sum_{p \in \cL'} |H_p| \log |H_p|)$.
-\end{claim}
-
-\begin{proof} Let $\cU$ denote a connected component of the shrub forest.
-We have $\sum_{p \in \cL \setminus \cL'} \sum_{v \in p} \log |H_v| \allowbreak \leq \sum_{\cU} \sum_{p \in \cU} \sum_{v \in p} \log |H_v|$.
-By \Lem{cu-root}, $|H_v| = O(|H_r|)$, where $r$ is the root of $\cU$, and furthermore, $\sum_{p \in \cU} |p| = O(|H_r|)$. 
-We have $\sum_{p \in \cU} \sum_{v \in p} \log |H_v| = O((\log |H_r|) \sum_{p \in \cU} |p|) = O(|H_r|\log |H_r|)$.
-We sum this over all $\cU$ in the shrub forest, and note that roots in the shrub forest are supported short paths.
-\end{proof}
-
-\ignore{
-Now, for the critical piece of the approach. Focus on $\pmax(T)$. Let $L$ be the set of short paths, 
-and $\hat{L} \subseteq L$ be the set of paths such that no ancestor of $p \in L'$
-is short. Obviously, $\hat{L}$ is an anti-chain in $P(T)$.
-
-\begin{lemma} \label{lem:short} For the path decomposition $\pmax(T)$, $\sum_{p \in \cL} \sum_{v \in p} \log |H_v| = O(\sum_{p \in \cL'} |H_p| \log |H_p|)$.
-\end{lemma}
-
-With this lemma, the proof of \Thm{runtime} is straightforward.
-
-\begin{proof} (of \Thm{runtime}) We focus on $\pmax(T)$ and split into tall and short paths.
-We have $\sum_{v \in T} \log |H_v| =$  $\sum_{p \in \cL} \sum_{v \in p} \log |H_v| + $ $\sum_{p \in \cH} \sum_{v \in p} \log |H_v|$.
-For the first term, we apply \Lem{short} to bound by $O(\sum_{p \in \cL'} |H_p| \log |H_p|)$.
-Then we apply \Clm{above} to bound by $O(\sum_{p \in \cH} |p|\log |p|)$.
-The latter is also bounded by the same expression, by \Lem{pathBounds}. By definition,
-$\sum_{p \in \cH} |p|\log |p| \leq \cost(\pmax(T))$.
-\end{proof}
-
-The main challenge is proving \Lem{short}, which we defer to the next section.
-\begin{definition}
- Let $\chi(T)$ be a valid coloring of a binary tree $T$.  We define a set of disjoint paths over the vertices of 
- $T$ as follows.  
- The corresponding collection of paths that all such edges correspond to is a path decomposition of $T$, which 
- we refer to as a \emph{maximum} path decomposition, and is denoted by $\mathcal{P}(T)$.  
-\end{definition}
-
-From the previous section we now have a lower bound of $\Omega(f_{path}(P))$ for computing the merge tree in terms of any path decomposition $P$. 
-We now show that the running time of our algorithm for computing the merge tree is upper bounded by this expression for a specific path decomposition, 
-which we will call the maximum path decomposition (defined below).
-Specifically, for a given valid coloring $\chi$ of a tree $T$, we will prove that for the maximum path decomposition $f_{alg}(\chi(T)) = O(f_{path}(P(T)))$. 
-%using the observations about path decompositions from 
-%\Sec{background}, which will then imply $f_{alg}(\chi(T)) = \Theta(f_{path}(P(T)))$.
-
-Throughout this section, for a path $p\in P(T)$, we use the notation $H_p$ to refer to the heap at the root of $p$, i.e.\ $H_p = H_{r_p}$.
-}
-
-\subsection{Proving \Lem{cu-root}: the root is everything in $\cU$} \label{sec:weight}
-
-\Lem{cu-root} asserts the root $r$ in $\cU$ pretty much encompasses all sizes and heaps in $\cU$.
-We will work with the \emph{reduced} heap $\redH_p$. This is the subset of vertices of $H_p$ that 
-do not appear on the tall support chain of $p$.
-%only appear in short paths. 
-By definition, for any unsupported short path
-(hence, any non-root $p \in \cU$), $|\redH_p| \geq 99|H_p|/100$. 
-% For notational convenience, we simply set $\redH_r = H_r$.
-%Consider $q \in \cL \setminus \cL'$ that is a child of $p$ in the shrub $\cU$.
-We begin with a key property, which is where the construction of $\pmax(T)$ enters the picture.
+So far we know that tall paths and supported short paths can be ``paid for.''  We will also be able to pay for 
+unsupported paths, though doing so is much trickier.  The following is a key property of unsupported paths which 
+we will use multiple times. It is worth noting that the proof of this lemma is where the construction of $\pmax(T)$ enters the picture.
 
@@
 \label{lem:geometric}
- Let $q$ be the child of some path $p$ in $\cU$, then $|H_p|\geq \frac{3}{2}|H_q|$. 
- Moreover, if $p\neq r(\cU)$, then $|\redH_p|\geq \frac{3}{2}|\redH_q|$. 
+ Let $\cU$ denote a connected component of $\cF$, and let $q$ be the child of some node $p$ in $\cU$. 
+ Then $|H_p|\geq \frac{3}{2}|H_q|$. 
 \end{lemma}
@@
 Since $w$ is the base of $h(q)$, $H_w = H_{h(q)}$. We apply the bound $|H_{h(q)}| \geq 99|H_q|/100$
-to get $|H_p| \geq 197|H_q|/100$, implying the first part of the lemma.  For the second part, observe that 
-if $p\neq r(\cU)$, then $p$ is unsupported and so $|\redH_p| \geq 99|H_p|/100$, and therefore the second part follows since $|H_q| \geq |\redH_q|$.
-%
-%
-% Let $w$ be the vertex in the path $p$ to which the root of the path $q$ is adjacent to in $T$, 
-% i.e.\ $w$ is the parent of $r_q$ in $T$.  Let $c$ be the other child of $w$, which lies in $p$.
-% As $\mathcal{P}(T)$ is a maximum path decomposition of $T$ and $c$ rather than $r_q$ is in 
-% the same path as $w$, we have $|H_q|\leq |H_c|$.  Therefore, $|H_w|\geq |H_q|+|H_c|-1\geq 2|H_q|-1$. 
-% Therefore, by \Obs{decrease}, $|H_p|+ d(r_p, w) \geq |H_w|\geq 2|H_q|-1$.  
-% Since $p$ is a short path, $d(r_p, w)\leq \sqrt{|H_p|}/100$ and so we have $|H_p|+\sqrt{|H_p|}/100\geq 2|H_q|-1$.
-% Assuming $|H_p|>11$ (and we can safely assume $|H_p|$ is bigger than any constant), this yields $\frac{3}{5} |H_p| \geq |H_q|$
-\end{proof}
-
-This immediately proves part (i) of \Lem{cu-root}. Part (ii) requires much more work.
+to get $|H_p| \geq 197|H_q|/100$, implying the lemma.  
+%For the second part, observe that if $p\neq r(\cU)$, then $p$ is unsupported and so 
+%$|\redH_p| \geq 99|H_p|/100$, and therefore the second part follows since $|H_q| \geq |\redH_q|$.
+\end{proof}
+
+\begin{corollary}\label{cor:geometric}
+ Let $\cU$ denote a connected component of $\cF$. Let $p$ and $d$ be any two nodes in $\cU$ such that $d$ is a descendant of $p$.
+ Then for any $v \in d$, $|H_v| = O(|H_p|)$.
+\end{corollary}
+
+To handle the unsupported paths, for every shrub $\cU$ in $\cF$ we first prune away what we call buffered sub-shrubs.  In the following subsection, 
+we show how to charge away the cost of these sub-shrubs to their tall support chains.  In \Sec{weight} we then show how to pay for the remaining pruned shrubs, 
+which is the most difficult part of the argument.  Afterwards in \Sec{thm-runtime}, it is then a straightforward task to put everything together to prove \Thm{runtime}.
+
+%We split the remaining argument into two subsections. We first prove \Thm{runtime} from \Lem{cu-root},
+%which involves routine calculations. Then we prove \Lem{cu-root}, where the interesting work happens.
+
+\newcommand{\buff}{\mathcal{B}}
+\newcommand{\loss}{\lambda}
+\subsection{Pruning Away Buffered Sub-shrubs}\label{sec:buffSub}
+Let $\cU$ be a shrub in $\cF$.  For any node $q$ in the shrub $\cU$, let $TSC(q)$ denote the set of all vertices of $T$ from all paths in the tall support chain of $q$.
+For $q$ in $\cU$ we then define the \emph{loss} to be $\loss(q) = TSC(q)\cap H_q$. 
+Intuitively, if $c$ is a child node of a parent node $p$ in $\cU$, then $\loss(c)$ are the vertices ``lost'' from the heap of $c$ to its tall support chain, by the time one reaches $p$.
+
+Ultimately our goal in this subsection is to prune away subtrees from $\cU$ which can be paid for by charging to their tall support chains (similarly to supported short paths).
+The remaining portion of the shrub will then be handled in a later subsection.
+We now formally define this pruning process, which is quite intuitive if one thinks of $\cU$ as an actual physical shrub.
+
+\begin{definition}\label{def:prune}
+  Let $\cU$ denote a shrub in $\cF$.
+A \emph{pruning} $\cU'$ of $\cU$ is any subgraph defined by some number of rounds of the following iterative process.
+Initially set $\cU' = \cU$.
+In each iteration select some subtree $\cU'_r$ rooted at a node $r\in \cU'$ and set $\cU' = \cU'\setminus \{\cU'_r\}$.
+\end{definition}
+
+The above definition does not specify which subtrees are pruned, which we now formally define.
+
+\begin{definition}\label{def:buff}
+ Let $\cU'$ be a pruning of a shrub of $\cU$ of $\cF$. Then a rooted subtree $\buff$ of $\cU'$ is called a \emph{buffered sub-shrub} if the following conditions hold:
+  \begin{enumerate}
+  \item $\sum_{p\in \buff} |\loss(p)| \geq \frac{1}{100} \sum_{p\in \buff} |p|$
+  \item No rooted subtree of $\buff$ satisfies the above inequality (other than $\buff$ itself). 
+ \end{enumerate}
+\end{definition}
+
+Ultimately we will prune buffered subtrees from $\cU$ by starting at the leaves and working towards the root, and hence the second condition of buffered will trivially be satisfied.
+
+The following will allow buffered sub-shrubs to be charged to their tall support chains.
+
+%We require the following lemma so that we can charge to the root in a buffered subtree.
+%
+% \begin{lemma}
+%  Let $p$ be any node in a buffered sub-shrub $\buff$, and let $d$ be any descendant node of $p$ in $\buff$.  Then for any $v \in d$, $|H_v| = O(|H_p|)$.
+% \end{lemma}
+% \begin{proof}
+%  Missing.  \Lem{cu-root} will be rephrased to handle this slightly more general case (i.e. here $p$ may not be the root of $\cU$). Note the proof will not 
+%  have to be changed as \Lem{geometric} is the proof which already implies general case.
+% \end{proof}
+
+\begin{lemma}
+ Let $\buff$ be a buffered sub-shrub.  Then $\sum_{p\in \buff}\sum_{v\in p} \log |H_v| = O(\sum_{p\in \buff}\sum_{v\in \loss(p)} \log |H_v| )$
+\end{lemma}
+\begin{proof}
+\textcolor{red}{\huge Proof missing. Potentially useful observations below}
+
+ Let $r$ be the root of $\buff$.
+First observe that $\loss(r) \geq |r|/100$.
+Suppose otherwise, then in order for the first condition of \Def{buff} to be satisfied for $\buff$, the rooted subtree of some child of $r$, call it $\buff'$, must satisfy $\sum_{p\in \buff'} \loss(p) \geq \frac{1}{100} \sum_{p\in \buff'} |p|$.
+However, this would contradict the second condition of \Def{buff}.
+
+ Consider any node $q\in \buff$.
+Then for any vertex $v\in \loss(q)$, we have $|H_v| = \Omega(|H_q|)$, since $q$ is unsupported and hence by definition less than $|H_q|/100$ vertices of $H_q$ appear in $\loss(q)$.
+
+ \begin{align*}
+  \sum_{p\in \buff}\sum_{v\in p} \log |H_v| &=  O\left( \sum_{p\in \buff}\sum_{v\in p} \log |H_p| \right)   &\mbox{definition of short} \\
+  &= O\left( \sum_{p\in \buff} |p| \log |H_p| \right)   &\mbox{} \\
+  &= ????   &\mbox{buffered is not sufficient to get to next line} \\
+  &= O\left( \sum_{p\in \buff} |\loss(p)| \log |H_p| \right)   &\mbox{} \\
+  &= O\left( \sum_{p\in \buff}\sum_{v\in \loss(p)} \log |H_v| \right)
+ \end{align*}
+
+ \Cor{geometric} should be used somewhere.
+\end{proof}
+
+Let $p$ and $q$ be two distinct paths in $\cF$.  Note that while it is possible that $TSC(p)\cap TSC(q)\neq \emptyset$, 
+it holds that $\loss(p)\cap \loss(q) = \emptyset$.  Thus the above lemma implies that the total heap costs due to vertices in buffered sub-shrubs can be charged to those in tall paths.
+
+\begin{corollary}
+ Let $B= \{\buff_1, \ldots, \buff_k\}$ be any collection of disjoint buffered sub-shrubs from $\cF$. Then, 
+ $\sum_{i=1}^k \sum_{p\in \buff_i}\sum_{v\in p} |H_v| = O(\sum_{p\in \cH}\sum_{v\in p} \log |H_v| )$
+\end{corollary}
+
+\newcommand{\pruned}{\cU/\cB}
+\subsection{Paying for Pruned Shrubs: the root is everything} \label{sec:weight}
+
+\begin{definition}
+ Let $\cU$ be a shrub in $\cF$. Consider a reverse topological ordering (i.e.\ leaves first) of the nodes in $\cU$, and use it to define a total ordering.  
+ We compute a pruning of $\cU$ (see \Def{prune}), by iteratively removing all buffered sub-shrubs (see \Def{buff}), where buffered sub-shrubs whose roots are earlier in the total ordering are pruned first.
+We use $\pruned$ to denote the resulting sub-shrub, which we refer to as the \emph{buffer-pruned} sub-shrub of $\cU$.
+\end{definition}
+
+The goal of this section is to prove the following lemma, which ultimately allows us to pay for pruned shrubs.
+
+\begin{lemma} \label{lem:cu-root} Let $\pruned$ denote a buffer-pruned sub-shrub of a shrub in $\cF$ and let $r$
+be the root of $\pruned$. (i) For any $v \in p$ such that $p \in \pruned$,
+$|H_v| = O(|H_r|)$. (ii) $\sum_{p \in \pruned} |p| = O(|H_r|)$.
+\end{lemma}
+
+\Lem{cu-root} asserts the root $r$ in $\pruned$ pretty much encompasses all sizes and heaps in $\pruned$.
+Observe that part (i) of \Lem{cu-root} is immediate, as it is a special case of \Cor{geometric}. Proving part (ii) requires much more work.
 % It is helpful to spell out how the various heaps evolve. Consider $p \in \cU$,
@@
 % for a child $q'$ in $\cU$.
+
+\textcolor{red}{\huge The rest of this subsection needs updating}
 
@@
 
-We wrap it all up by proving part (ii) of \Lem{cu-root}.
-
-\begin{claim} $\sum_{p \in \cU} |p| \leq |H_{r(\cU)}|/10$.
-\end{claim}
-
-% \begin{proof} We use $r$ for $r(\cU)$. Our aim is to simply apply \Clm{charge} from $r(\cU)$. 
-% Suppose $W_q \geq (4/3)W_{q'}$ (for any choice in $\cU$ of $q$ parent of $q'$). By \Clm{charge}, $\sum_{p \in \cU} |p| \leq W_{r}/20$.
-% By \Clm{loss}, $|\redH_r| \geq W_r - 2\sum_{p \in \cU} |p|$. By simple algebra, we deduce, $\sum_{p \in \cU} |p| \leq |\redH_r|/10 = |H_r|/10$.
-% 
-% We now prove that for any $q$ parent of $q'$, $W_q \geq (4/3)W_{q'}$.
-% Suppose not. Let $p, p'$ be the counterexample furthest from the root,
-% where $p$ is the parent of $p'$. 
-% Note that for $q$ and child $q'$ in $\cU_{p'}$, $W_q \geq (4/3)W_{q'}$.
-% We will apply \Clm{charge} for $\cU_{p'}$ to deduce that $\sum_{q \in \cU_{p'}} |q| \leq W_{p'}/20$.
-% By \Clm{loss}, $|\redH_{p'}| \geq 19W_{p'}/20$. By \Lem{geometric}, $|\redH_p| \geq (3/2)|\redH_{p'}|$.
-% Noting that $W_p \geq |\redH_p|$, we deduce that $W_p \geq (4/3)W_{p'}$.
-% Hence, $p, p'$ is not a counterexample, and more generally, this is no counterexample.
-% That completes the whole proof.
-% \end{proof}
-
-\begin{proof} We use $r$ for $r(\cU)$. %Our aim is to simply apply \Clm{charge} from $r(\cU)$. 
+% We wrap it all up by proving part (ii) of \Lem{cu-root}.
+%
+% \begin{claim} $\sum_{p \in \cU} |p| \leq |H_{r(\cU)}|/10$.
+% \end{claim}
+% 
+% % \begin{proof} We use $r$ for $r(\cU)$. Our aim is to simply apply \Clm{charge} from $r(\cU)$. 
+% % Suppose $W_q \geq (4/3)W_{q'}$ (for any choice in $\cU$ of $q$ parent of $q'$). By \Clm{charge}, $\sum_{p \in \cU} |p| \leq W_{r}/20$.
+% % By \Clm{loss}, $|\redH_r| \geq W_r - 2\sum_{p \in \cU} |p|$. By simple algebra, we deduce, $\sum_{p \in \cU} |p| \leq |\redH_r|/10 = |H_r|/10$.
+% % 
+% % We now prove that for any $q$ parent of $q'$, $W_q \geq (4/3)W_{q'}$.
+% % Suppose not. Let $p, p'$ be the counterexample furthest from the root,
+% % where $p$ is the parent of $p'$. 
+% % Note that for $q$ and child $q'$ in $\cU_{p'}$, $W_q \geq (4/3)W_{q'}$.
+% % We will apply \Clm{charge} for $\cU_{p'}$ to deduce that $\sum_{q \in \cU_{p'}} |q| \leq W_{p'}/20$.
+% % By \Clm{loss}, $|\redH_{p'}| \geq 19W_{p'}/20$. By \Lem{geometric}, $|\redH_p| \geq (3/2)|\redH_{p'}|$.
+% % Noting that $W_p \geq |\redH_p|$, we deduce that $W_p \geq (4/3)W_{p'}$.
+% % Hence, $p, p'$ is not a counterexample, and more generally, this is no counterexample.
+% % That completes the whole proof.
+% % \end{proof}
+
+\begin{proof}[Proof of \Lem{cu-root}]
+We use $r$ for $r(\cU)$. %Our aim is to simply apply \Clm{charge} from $r(\cU)$. 
 Suppose $W_q \geq (4/3)W_{q'}$ (for any choice in $\cU\setminus\{r\}$ of $q$ parent of $q'$), 
@@
 That completes the whole proof.
+\end{proof}
+
+\subsection{Proving \Thm{runtime}} \label{sec:thm-runtime}
+\textcolor{red}{\huge Ben will update this subsection to be consistent with new changes}
+
+We split the summation into tall, supported short, and unsupported short paths.
+\begin{eqnarray*} 
+\sum_{p \in \pmax(T)} \sum_{v \in p} \log |H_v| & = & \sum_{p \in \cL \setminus \cL'} \sum_{v \in p} \log |H_v| + \sum_{p \in \cL'} \sum_{v \in p} \log |H_v| + \sum_{p \in \cH} \sum_{v \in p} \log |H_v|
+\end{eqnarray*}
+The last term can be bounded by $O(\sum_{p \in \pmax(T)} |p|\log |p|)$, by \Lem{pathBounds}.
+The second term can be bounded by $O(\sum_{p \in \cL'} |H_p| \log |H_p|)$, again by \Lem{pathBounds}.
+The following claim shows that this in turn is asymptotically bounded by the last term.
+
+\begin{claim} \label{clm:above} $\sum_{p \in \cL'} |H_p| \log |H_p| = O(\sum_{q \in \cH} \sum_{v \in q} \log |H_v|)$.
+\end{claim}
+
+\begin{proof} 
+Pick $p \in \cL'$. 
+As we traverse the tall support chain of $p$, there
+are at least $|H_p|/100$ vertices of $H_p$ that lie in these paths. These are encountered in
+a fixed order. Let $H'_p$ be the first $|H_p|/200$ of these vertices. When $v \in H'_p$
+is encountered, there are $|H_p|/200$ vertices of $H_p$ not yet encountered. Hence,
+$|H_v| \geq |H_p|/200$. Hence, $|H_p|\log |H_p| = O(\sum_{v \in H'_p} \log |H_v|)$.
+Since all the vertices lie in tall paths,
+we can write this as $O(\sum_{q \in \cH} \sum_{v \in H'_p \cap q} \log |H_v|)$.
+Summing over all $p$, the expression is $\sum_{q \in \cH} \sum_{p \in \cL'} \sum_{v \in H'_p \cap q} \log |H_v|$.
+%and $\sum_{p \in \cL'} |H_p|\log |H_p| = O(\sum_{p \in \cL'} \sum_{v \in H'_p} \log |H_v|)$.
+
+Consider any $v \in H'_p$.  Let $S$ be the set of paths $\widetilde{p}\in \cL'$ such that 
+$v\in H'_{\widetilde{p}}$.  We now show $|S|\leq 2$ (i.e.\ it contains at most one path other than $p$).
+First observe that any two paths in $S$ must be unrelated (i.e.\ $S$ is an anti-chain), 
+since paths which have an ancestor-descendant relationship have disjoint tall support chains.  
+However, any vertex $v$ receives exactly one color from each of its two subtrees (in $T$), and therefore $|S|\leq 2$ since any two 
+paths which share descendant leaves in $T$ (i.e.\ their heaps are tracking the same color) must have an ancestor-descendant relationship.
+
+In other words, any $\log |H_v|$ appears at most twice in the above triple summation.
+Hence, we can bound it by $O(\sum_{q \in \cH} \sum_{v \in q} \log |H_v|)$.
+\end{proof}
+
+The first term (unsupported short paths) can be charged to the second term (supported short paths).
+This is where the critical \Lem{cu-root} plays a role.
+
+\begin{claim} \label{clm:first} $\sum_{p \in \cL \setminus \cL'} \sum_{v \in p} \log |H_v| = O(\sum_{p \in \cL'} |H_p| \log |H_p|)$.
+\end{claim}
+
+\begin{proof} Let $\cU$ denote a connected component of the shrub forest.
+We have $\sum_{p \in \cL \setminus \cL'} \sum_{v \in p} \log |H_v| \allowbreak \leq \sum_{\cU} \sum_{p \in \cU} \sum_{v \in p} \log |H_v|$.
+By \Lem{cu-root}, $|H_v| = O(|H_r|)$, where $r$ is the root of $\cU$, and furthermore, $\sum_{p \in \cU} |p| = O(|H_r|)$. 
+We have $\sum_{p \in \cU} \sum_{v \in p} \log |H_v| = O((\log |H_r|) \sum_{p \in \cU} |p|) = O(|H_r|\log |H_r|)$.
+We sum this over all $\cU$ in the shrub forest, and note that roots in the shrub forest are supported short paths.
 \end{proof}
@@
 There is an algorithm to compute $\cC$ whose running time is $O(\sum_{p \in P(\cC)} |p|\log |p| + t\alpha(t) + N)$,
-where $P(T)$ is a specific path decomposition (constructed implicitly by the algorithm).
+where $P(\cC)$ is a specific path decomposition (constructed implicitly by the algorithm).
 \end{theorem}

### 2016-07-29 paint-contour-tree.tex
section 8 fixes

@@
 
-\subsection{Shrubs, tall paths, and short paths}
+\subsection{Tall and short paths}
 \label{sec:shrubs}
@@
 
-\begin{lemma}
-\label{lem:adjacent}
- Let $p$ be any path in $\pmax(T)$ and let $\{q_1, \dots q_k\}$ be the children of $p$.
-Then $|H_{\ell(p)}| + \sum_{i=1}^k |H_{q_i}| \leq |H_p|+2|p|$. 
-\end{lemma}
-
-\begin{proof} For convenience, denote $H_i = H_{q_i}$ and $H_0 = H_{\ell(p)}$.
-Consider $v \in \bigcup_{i} H_i$ that lies below $b(p)$ in $T$. 
-Note that such a vertex has only one of its two colors in $L(b(p))$.  
-Since the colors tracked by $H_i$ and $H_j$ for $i\neq j$ are disjoint, 
-such a vertex can appear in only one of the $H_i$'s.
-On the other hand, a vertex $u\in p$ can appear in more than one $H_i$, 
-but since any vertex has exactly two colors it can appear in at most two 
-such heaps.  Hence, $\sum_i |H_i| \leq |H_p| + 2|p|$.
-\ignore{
- First observe that for all $i$, $r_{i}$ lies above $r_p$ on some root to leaf path. 
- Let $H_i'$ be the subset of $H_i$ that lies below $r_p$.  Observe that $H_i'\subseteq H_p$, as $L(T_{r_i})\subseteq L(T_{r_p})$.
-
- Now for any $i\neq j$, $H_i \cap H_j = \emptyset$, as $r_{i}$ and $r_{j}$ are not on the same root to leaf path (i.e.\ $L(T_{r_i}) \cap L(T_{r_j}) = \emptyset$).
- In particular, $(H_i\setminus H_i') \cap (H_j \setminus H_j') = \emptyset$, and so $\sum_{i=0}^k |H_i\setminus H_i'| \leq |p|$ 
- as $H_i\setminus H_i'$ is precisely the subset of $H_i$ that lies on $p$.
- Moreover, for any $i\neq j$, $ H_i' \cap H_j' = \emptyset$, and so since $H_i'\subseteq H_p$ we have $\sum_{i=0}^k |H_i'|\leq |H_p|$.
- Putting these together gives
- \[
-  \sum_{i=0}^k |H_i| = \sum_{i=0}^k |H_i'| + \sum_{i=0}^k |H_i\setminus H_i'| \leq |H_p|+|p|.
- \]
-}
-\end{proof}
+% \begin{lemma}
+% \label{lem:adjacent}
+%  Let $p$ be any path in $\pmax(T)$ and let $\{q_1, \dots q_k\}$ be the children of $p$.
+% Then $|H_{\ell(p)}| + \sum_{i=1}^k |H_{q_i}| \leq |H_p|+2|p|$. 
+% \end{lemma}
+% 
+% \begin{proof} For convenience, denote $H_i = H_{q_i}$ and $H_0 = H_{\ell(p)}$.
+% Consider $v \in \bigcup_{i} H_i$ that lies below $b(p)$ in $T$. 
+% Note that such a vertex has only one of its two colors in $L(b(p))$.  
+% Since the colors tracked by $H_i$ and $H_j$ for $i\neq j$ are disjoint, 
+% such a vertex can appear in only one of the $H_i$'s.
+% On the other hand, a vertex $u\in p$ can appear in more than one $H_i$, 
+% but since any vertex has exactly two colors it can appear in at most two 
+% such heaps.  Hence, $\sum_i |H_i| \leq |H_p| + 2|p|$.
+% \ignore{
+%  First observe that for all $i$, $r_{i}$ lies above $r_p$ on some root to leaf path. 
+%  Let $H_i'$ be the subset of $H_i$ that lies below $r_p$.  Observe that $H_i'\subseteq H_p$, as $L(T_{r_i})\subseteq L(T_{r_p})$.
+%   
+%  Now for any $i\neq j$, $H_i \cap H_j = \emptyset$, as $r_{i}$ and $r_{j}$ are not on the same root to leaf path (i.e.\ $L(T_{r_i}) \cap L(T_{r_j}) = \emptyset$).
+%  In particular, $(H_i\setminus H_i') \cap (H_j \setminus H_j') = \emptyset$, and so $\sum_{i=0}^k |H_i\setminus H_i'| \leq |p|$ 
+%  as $H_i\setminus H_i'$ is precisely the subset of $H_i$ that lies on $p$.
+%  Moreover, for any $i\neq j$, $ H_i' \cap H_j' = \emptyset$, and so since $H_i'\subseteq H_p$ we have $\sum_{i=0}^k |H_i'|\leq |H_p|$.
+%  Putting these together gives
+%  \[
+%   \sum_{i=0}^k |H_i| = \sum_{i=0}^k |H_i'| + \sum_{i=0}^k |H_i\setminus H_i'| \leq |H_p|+|p|.
+%  \]
+% }
+% \end{proof}
 
@@
 Let $\cL$ be the set of short paths, $\cL'$ be the set of supported short paths, and $\cH$ be the set of tall paths
+given by $\pmax(T)$ (notation which we be used for the remainder of this section).
+
+\begin{claim} \label{clm:above} 
+$\sum_{p \in \cL'} |H_p| \log |H_p| = O(\sum_{q \in \cH} \sum_{v \in q} \log |H_v|)$.
+\end{claim}
+\begin{proof} 
+Pick $p \in \cL'$. 
+As we traverse the tall support chain of $p$, there
+are at least $|H_p|/100$ vertices of $H_p$ that lie in these paths. These are encountered in
+a fixed order. Let $H'_p$ be the first $|H_p|/200$ of these vertices. When $v \in H'_p$
+is encountered, there are $|H_p|/200$ vertices of $H_p$ not yet encountered. Hence,
+$|H_v| \geq |H_p|/200$. Hence, $|H_p|\log |H_p| = O(\sum_{v \in H'_p} \log |H_v|)$.
+Since all the vertices lie in tall paths,
+we can write this as $O(\sum_{q \in \cH} \sum_{v \in H'_p \cap q} \log |H_v|)$.
+Summing over all $p$, the expression is $\sum_{q \in \cH} \sum_{p \in \cL'} \sum_{v \in H'_p \cap q} \log |H_v|$.
+%and $\sum_{p \in \cL'} |H_p|\log |H_p| = O(\sum_{p \in \cL'} \sum_{v \in H'_p} \log |H_v|)$.
+
+Consider any $v \in H'_p$.  Let $S$ be the set of paths $\widetilde{p}\in \cL'$ such that 
+$v\in H'_{\widetilde{p}}$.  We now show $|S|\leq 2$ (i.e.\ it contains at most one path other than $p$).
+First observe that any two paths in $S$ must be unrelated (i.e.\ $S$ is an anti-chain), 
+since paths which have an ancestor-descendant relationship have disjoint tall support chains.  
+However, any vertex $v$ receives exactly one color from each of its two subtrees (in $T$), and therefore $|S|\leq 2$ since any two 
+paths which share descendant leaves in $T$ (i.e.\ their heaps are tracking the same color) must have an ancestor-descendant relationship.
+
+In other words, any $\log |H_v|$ appears at most twice in the above triple summation.
+Hence, we can bound it by $O(\sum_{q \in \cH} \sum_{v \in q} \log |H_v|)$.
+\end{proof}
+
+\begin{corollary}\label{cor:supported}
+  $\sum_{p \in \cL'} \sum_{v \in p} \log |H_v|= O(\sum_{p\in \pmax} |p|\log |p|)$
+\end{corollary}
+\begin{proof}
+ By applying \Lem{pathBounds}, \Clm{above}, and then \Lem{pathBounds} again, we have the following.
+ \begin{align*}
+  \sum_{p \in \cL'} \sum_{v \in p} \log |H_v| 
+  =  O\left(\sum_{p \in \cL'} |H_p| \log |H_p|\right)
+  = O\left(\sum_{p \in \cH} \sum_{v \in p} \log |H_v|\right)
+  = O\left(\sum_{p\in \pmax} |p|\log |p|\right)
+ \end{align*}
+%
+% By \Lem{pathBounds}, $\sum_{p \in \cL'} \sum_{v \in p} \log |H_v|$
+\end{proof}
+
+\subsection{Unsupported shrubs}
+
+So far we know that tall paths and supported short paths can be ``paid for.''  We will also be able to pay for 
+unsupported paths, though doing so is much trickier.
+
+Recall that $\cL$ denotes the set of short paths, $\cL'$ the set of supported short paths, and $\cH$ the set of tall paths
 given by $\pmax(T)$. We now describe a partitioning of the unsupported paths into a forest of shrubs. Consider $p \in \cL \setminus \cL'$,
@@
 
-So far we know that tall paths and supported short paths can be ``paid for.''  We will also be able to pay for 
-unsupported paths, though doing so is much trickier.  The following is a key property of unsupported paths which 
+The following is a key property of unsupported paths which 
 we will use multiple times. It is worth noting that the proof of this lemma is where the construction of $\pmax(T)$ enters the picture.
@@
 \newcommand{\loss}{\lambda}
-\subsection{Pruning Away Buffered Sub-shrubs}\label{sec:buffSub}
+\subsection{Pruning away buffered sub-shrubs}\label{sec:buffSub}
 Let $\cU$ be a shrub in $\cF$.  For any node $q$ in the shrub $\cU$, let $TSC(q)$ denote the set of all vertices of $T$ from all paths in the tall support chain of $q$.
@@
 
-%We require the following lemma so that we can charge to the root in a buffered subtree.
-%
-% \begin{lemma}
-%  Let $p$ be any node in a buffered sub-shrub $\buff$, and let $d$ be any descendant node of $p$ in $\buff$.  Then for any $v \in d$, $|H_v| = O(|H_p|)$.
-% \end{lemma}
-% \begin{proof}
-%  Missing.  \Lem{cu-root} will be rephrased to handle this slightly more general case (i.e. here $p$ may not be the root of $\cU$). Note the proof will not 
-%  have to be changed as \Lem{geometric} is the proof which already implies general case.
-% \end{proof}
-
 \begin{lemma}
@@
 
-Let $p$ and $q$ be two distinct paths in $\cF$.  Note that while it is possible that $TSC(p)\cap TSC(q)\neq \emptyset$, 
-it holds that $\loss(p)\cap \loss(q) = \emptyset$.  Thus the above lemma implies that the total heap costs due to vertices in buffered sub-shrubs can be charged to those in tall paths.
-
-\begin{corollary}
+For any fixed vertex $v$ on a tall path, there are at most two unsupported short paths such that $v$ is contained in $\loss(\cdot)$. Specifically, 
+$v$ is assigned exactly two colors, one from each of its child subtrees in $T$.  Pick one of these two colors, which corresponds to some leaf in $T$.  
+Any two unsupported short paths which are tracking the color of this leaf, must be ancestor-descendant in $\pmax$, however such paths have disjoint tall support chains, 
+and hence only one such path can count $v$ in $\loss(\cdot)$.
+(Note this is the same argument as in \Clm{above}.)  Thus by summing over all tall support chains we have the following.
+
+%Let $p$ and $q$ be two distinct paths in $\cF$.  Note that while it is possible that $TSC(p)\cap TSC(q)\neq \emptyset$, 
+%it holds that $\loss(p)\cap \loss(q) = \emptyset$.  Thus the above lemma implies that the total heap costs due to vertices in buffered sub-shrubs can be charged to those in tall paths.
+
+\begin{corollary}\label{cor:disjointBuff}
  Let $B= \{\buff_1, \ldots, \buff_k\}$ be any collection of disjoint buffered sub-shrubs from $\cF$. Then, 
- $\sum_{i=1}^k \sum_{p\in \buff_i}\sum_{v\in p} |H_v| = O(\sum_{p\in \cH}\sum_{v\in p} \log |H_v| )$
+ $\sum_{i=1}^k \sum_{p\in \buff_i}\sum_{v\in p} |H_v| = O(\sum_{p\in \cH}\sum_{v\in p} \log |H_v| )$.
 \end{corollary}
@@
 
-\begin{definition}
+\begin{definition}\label{def:pruningAShrub}
  Let $\cU$ be a shrub in $\cF$. Consider a reverse topological ordering (i.e.\ leaves first) of the nodes in $\cU$, and use it to define a total ordering.  
@@
 \subsection{Proving \Thm{runtime}} \label{sec:thm-runtime}
-\textcolor{red}{\huge Ben will update this subsection to be consistent with new changes}
 
@@
 \end{eqnarray*}
-The last term can be bounded by $O(\sum_{p \in \pmax(T)} |p|\log |p|)$, by \Lem{pathBounds}.
-The second term can be bounded by $O(\sum_{p \in \cL'} |H_p| \log |H_p|)$, again by \Lem{pathBounds}.
-The following claim shows that this in turn is asymptotically bounded by the last term.
-
-\begin{claim} \label{clm:above} $\sum_{p \in \cL'} |H_p| \log |H_p| = O(\sum_{q \in \cH} \sum_{v \in q} \log |H_v|)$.
+By \Cor{supported} and \Lem{pathBounds}, the second and last term can be bounded by $O(\sum_{p \in \pmax(T)} |p|\log |p|)$.
+The following claim applies the results of the previous two subsections to give the same bound on the first term, 
+thus completing the proof of \Thm{runtime}.
+
+% \begin{claim} \label{clm:above} $\sum_{p \in \cL'} |H_p| \log |H_p| = O(\sum_{q \in \cH} \sum_{v \in q} \log |H_v|)$.
+% \end{claim}
+% 
+% \begin{proof} 
+% Pick $p \in \cL'$. 
+% As we traverse the tall support chain of $p$, there
+% are at least $|H_p|/100$ vertices of $H_p$ that lie in these paths. These are encountered in
+% a fixed order. Let $H'_p$ be the first $|H_p|/200$ of these vertices. When $v \in H'_p$
+% is encountered, there are $|H_p|/200$ vertices of $H_p$ not yet encountered. Hence,
+% $|H_v| \geq |H_p|/200$. Hence, $|H_p|\log |H_p| = O(\sum_{v \in H'_p} \log |H_v|)$.
+% Since all the vertices lie in tall paths,
+% we can write this as $O(\sum_{q \in \cH} \sum_{v \in H'_p \cap q} \log |H_v|)$.
+% Summing over all $p$, the expression is $\sum_{q \in \cH} \sum_{p \in \cL'} \sum_{v \in H'_p \cap q} \log |H_v|$.
+% %and $\sum_{p \in \cL'} |H_p|\log |H_p| = O(\sum_{p \in \cL'} \sum_{v \in H'_p} \log |H_v|)$.
+% 
+% Consider any $v \in H'_p$.  Let $S$ be the set of paths $\widetilde{p}\in \cL'$ such that 
+% $v\in H'_{\widetilde{p}}$.  We now show $|S|\leq 2$ (i.e.\ it contains at most one path other than $p$).
+% First observe that any two paths in $S$ must be unrelated (i.e.\ $S$ is an anti-chain), 
+% since paths which have an ancestor-descendant relationship have disjoint tall support chains.  
+% However, any vertex $v$ receives exactly one color from each of its two subtrees (in $T$), and therefore $|S|\leq 2$ since any two 
+% paths which share descendant leaves in $T$ (i.e.\ their heaps are tracking the same color) must have an ancestor-descendant relationship.
+% 
+% In other words, any $\log |H_v|$ appears at most twice in the above triple summation.
+% Hence, we can bound it by $O(\sum_{q \in \cH} \sum_{v \in q} \log |H_v|)$.
+% \end{proof}
+
+% The first term (unsupported short paths) can be charged to the second term (supported short paths).
+% This is where the critical \Lem{cu-root} plays a role.
+
+\begin{claim} \label{clm:first} 
+$\sum_{p \in \cL \setminus \cL'} \sum_{v \in p} \log |H_v| = O(\sum_{p\in \pmax} |p|\log |p|)$.%= O(\sum_{p \in \cL'} |H_p| \log |H_p|)$.
 \end{claim}
@@
 \begin{proof} 
-Pick $p \in \cL'$. 
-As we traverse the tall support chain of $p$, there
-are at least $|H_p|/100$ vertices of $H_p$ that lie in these paths. These are encountered in
-a fixed order. Let $H'_p$ be the first $|H_p|/200$ of these vertices. When $v \in H'_p$
-is encountered, there are $|H_p|/200$ vertices of $H_p$ not yet encountered. Hence,
-$|H_v| \geq |H_p|/200$. Hence, $|H_p|\log |H_p| = O(\sum_{v \in H'_p} \log |H_v|)$.
-Since all the vertices lie in tall paths,
-we can write this as $O(\sum_{q \in \cH} \sum_{v \in H'_p \cap q} \log |H_v|)$.
-Summing over all $p$, the expression is $\sum_{q \in \cH} \sum_{p \in \cL'} \sum_{v \in H'_p \cap q} \log |H_v|$.
-%and $\sum_{p \in \cL'} |H_p|\log |H_p| = O(\sum_{p \in \cL'} \sum_{v \in H'_p} \log |H_v|)$.
-
-Consider any $v \in H'_p$.  Let $S$ be the set of paths $\widetilde{p}\in \cL'$ such that 
-$v\in H'_{\widetilde{p}}$.  We now show $|S|\leq 2$ (i.e.\ it contains at most one path other than $p$).
-First observe that any two paths in $S$ must be unrelated (i.e.\ $S$ is an anti-chain), 
-since paths which have an ancestor-descendant relationship have disjoint tall support chains.  
-However, any vertex $v$ receives exactly one color from each of its two subtrees (in $T$), and therefore $|S|\leq 2$ since any two 
-paths which share descendant leaves in $T$ (i.e.\ their heaps are tracking the same color) must have an ancestor-descendant relationship.
-
-In other words, any $\log |H_v|$ appears at most twice in the above triple summation.
-Hence, we can bound it by $O(\sum_{q \in \cH} \sum_{v \in q} \log |H_v|)$.
-\end{proof}
-
-The first term (unsupported short paths) can be charged to the second term (supported short paths).
-This is where the critical \Lem{cu-root} plays a role.
-
-\begin{claim} \label{clm:first} $\sum_{p \in \cL \setminus \cL'} \sum_{v \in p} \log |H_v| = O(\sum_{p \in \cL'} |H_p| \log |H_p|)$.
-\end{claim}
-
-\begin{proof} Let $\cU$ denote a connected component of the shrub forest.
-We have $\sum_{p \in \cL \setminus \cL'} \sum_{v \in p} \log |H_v| \allowbreak \leq \sum_{\cU} \sum_{p \in \cU} \sum_{v \in p} \log |H_v|$.
-By \Lem{cu-root}, $|H_v| = O(|H_r|)$, where $r$ is the root of $\cU$, and furthermore, $\sum_{p \in \cU} |p| = O(|H_r|)$. 
-We have $\sum_{p \in \cU} \sum_{v \in p} \log |H_v| = O((\log |H_r|) \sum_{p \in \cU} |p|) = O(|H_r|\log |H_r|)$.
-We sum this over all $\cU$ in the shrub forest, and note that roots in the shrub forest are supported short paths.
+For any component $\cU$ of the shrub forest $\cF$ of $\cL\setminus \cL'$, let $B(\cU)$ denote the collection of buffered sub-shrubs that we 
+removed in the pruning process of \Def{pruningAShrub} that produced the buffer-pruned sub-shrub $\pruned$
+(i.e.\ these are disjoint sub-shrubs whose union cover $\cU\setminus (\pruned)$). 
+Thus we can decompose the sum as follows, 
+\begin{align*}
+ \sum_{p \in \cL \setminus \cL'} \sum_{v \in p} \log |H_v| 
+ &\leq \sum_{\cU\in \cF} \sum_{p \in \cU} \sum_{v \in p} \log |H_v|
+ = \sum_{\cU\in \cF}\sum_{p \in \pruned} \sum_{v \in p} \log |H_v|  +  \sum_{\cU\in \cF}\sum_{\buff\in B(\cU)} \sum_{p \in \buff} \sum_{v \in p} \log |H_v|\\
+ &= \sum_{\cU\in \cF}\sum_{p \in \pruned} \sum_{v \in p} \log |H_v|  +  O\left(\sum_{p\in \cH}\sum_{v\in p} \log |H_v| \right)\\
+ &= \sum_{\cU\in \cF}\sum_{p \in \pruned} \sum_{v \in p} \log |H_v|  +  O\left(\sum_{p\in \pmax} |p|\log |p|\right),
+\end{align*}
+where the last two lines follow from \Cor{disjointBuff} and \Lem{pathBounds}, respectively.
+
+So what remains is to bound the first term.  
+%We have $\sum_{p \in \cL \setminus \cL'} \sum_{v \in p} \log |H_v| \allowbreak \leq \sum_{\cU} \sum_{p \in \cU} \sum_{v \in p} \log |H_v|$.
+By \Lem{cu-root}, $|H_v| = O(|H_r|)$, where $r$ is the root of $\pruned$, and furthermore, $\sum_{p \in \pruned} |p| = O(|H_r|)$. 
+We have $\sum_{p \in \pruned} \sum_{v \in p} \log |H_v| = O((\log |H_r|) \sum_{p \in \pruned} |p|) = O(|H_r|\log |H_r|)$.
+Note that the roots in the shrub forest are supported short paths, and hence when we sum this over all $\cU$ in the shrub forest we get, 
+\[
+\sum_{\cU\in \cF} O\left(|H_{r(\pruned)}|\log |H_{r(\pruned)}|\right)
+=\sum_{p\in \cL'} O(|H_p|\log |H_p|) = O\left(\sum_{p\in \pmax} |p|\log |p|\right),
+\]
+where the last equality follows by \Clm{above} (via the same argument as in \Cor{supported}).
 \end{proof}

### 2016-08-01 paint-contour-tree.tex
updating with proof of buffered shrubs, and cleaning up reweighting argument

@@
 
+\newcommand{\loss}{\lambda}
+
 \subsection{Unsupported shrubs}
@@
 and the remaining nodes are the unsupported short paths.
+
+Let $\cU$ be a shrub in $\cF$.  For any node $q$ in the shrub $\cU$, let $TSC(q)$ denote the set of all vertices of $T$ from all paths in the tall support chain of $q$.
+For $q$ in $\cU$ we then define the \emph{loss} to be $\loss(q) = TSC(q)\cap H_q$. 
+Intuitively, if $c$ is a child node of a parent node $p$ in $\cU$, then $\loss(c)$ are the vertices ``lost'' from the heap of $c$ to its tall support chain, by the time one reaches $p$.
 
@@
  Let $\cU$ denote a connected component of $\cF$. Let $p$ and $d$ be any two nodes in $\cU$ such that $d$ is a descendant of $p$.
- Then for any $v \in d$, $|H_v| = O(|H_p|)$.
+ Then for any $w \in d$ and $v \in \lambda(p)$, $|H_w| = O(|H_v|)$.
 \end{corollary}
+
+\begin{proof} By the previous lemma, $|H_w| = O(|H_p|)$. Note that there are at most
+$|H_p|/100$ nodes in $\lambda(p)$, since these are all in the tall support chain of $v$
+(which is unsupported). Thus, for any $v \in \lambda(p)$, $H_v \geq 99|H_p|/100$,
+completing the proof.
+\end{proof}
 
@@
 \newcommand{\buff}{\mathcal{B}}
-\newcommand{\loss}{\lambda}
 \subsection{Pruning away buffered sub-shrubs}\label{sec:buffSub}
-Let $\cU$ be a shrub in $\cF$.  For any node $q$ in the shrub $\cU$, let $TSC(q)$ denote the set of all vertices of $T$ from all paths in the tall support chain of $q$.
-For $q$ in $\cU$ we then define the \emph{loss} to be $\loss(q) = TSC(q)\cap H_q$. 
-Intuitively, if $c$ is a child node of a parent node $p$ in $\cU$, then $\loss(c)$ are the vertices ``lost'' from the heap of $c$ to its tall support chain, by the time one reaches $p$.
-
 Ultimately our goal in this subsection is to prune away subtrees from $\cU$ which can be paid for by charging to their tall support chains (similarly to supported short paths).
@@
   \begin{enumerate}
-  \item $\sum_{p\in \buff} |\loss(p)| \geq \frac{1}{100} \sum_{p\in \buff} |p|$
+  \item $\sum_{p\in \buff} |\loss(p)| > \frac{1}{100} \sum_{p\in \buff} |p|$
   \item No rooted subtree of $\buff$ satisfies the above inequality (other than $\buff$ itself). 
@@
 
-Ultimately we will prune buffered subtrees from $\cU$ by starting at the leaves and working towards the root, and hence the second condition of buffered will trivially be satisfied.
-
-The following will allow buffered sub-shrubs to be charged to their tall support chains.
-
-\begin{lemma}
+The following is quite straightforward. By pruning from the leaves towards
+the root, we can remove all buffered sub-shrubs from a given $\cU$.
+
+\begin{lemma} \label{lem:split-buff} Any shrub $\cU$ can be partitioned
+into $\{\cU', \cB_1, \cB_2, \ldots \cB_r\}$ where each $\cB_i$ is buffered.
+Furthermore, no subtree of $\cU'$ is buffered.
+\end{lemma}
+
+\begin{proof} Take any reverse topological order of the nodes in $\cU$
+(so descendants occur before ancestors). Initialize $\cU'$ to $\cU$.
+We process the nodes in order. For node $p$, let $\cU'_p$ denote the rooted
+subtree of $p$ in $\cU'$. We simply check if $\sum_{q \in \cU'_p} |\loss(p)| \geq \sum_{q \in \cU'_p} |q|/100$. If so, we prune off $\cU'_p$. (Note that $\cU'$ is redefined
+to be the remainder.)
+
+At the end of this process, we have the desired decomposition. Since we process
+in reverse topological order, the minimality condition of \Def{buff}
+automatically holds. Thus, each pruned sub-shrub is buffered and the remaining $\cU'$
+has the desired property.
+\end{proof}
+
+We prove an important lemma. The cost of heap operations corresponding 
+to a buffered sub-shrub to be charged to tall support chains.
+
+\begin{lemma} \label{lem:buff}
  Let $\buff$ be a buffered sub-shrub.  Then $\sum_{p\in \buff}\sum_{v\in p} \log |H_v| = O(\sum_{p\in \buff}\sum_{v\in \loss(p)} \log |H_v| )$
 \end{lemma}
-\begin{proof}
-\textcolor{red}{\huge Proof missing. Potentially useful observations below}
-
- Let $r$ be the root of $\buff$.
+\begin{proof} The proof is done by a charging argument. For every $p \in \buff$
+and $v \in \loss(p)$, we create 100 tokens. Each of these tokens is labeled $\log |H_v|$,
+and this is called the ``value" of the token. Observe that the sum of token values is, up to a constant
+factor, the bound we wish to show. The set of tokens created for all $v \in \loss(p)$ is denoted
+$T_p$.
+
+There is a grand total of $100 \sum_{p \in \buff} |\loss(p)|$ tokens, which
+(by the buffered property of $\buff$) is strictly more than $\sum_{p \in \buff} |p|$.
+We will place all these tokens on the nodes in $\buff$, using the following process.
+We reserve topologically sort (descendants before ancestors) the nodes in $\buff$.
+For each $p$ in order, we will place tokens in $T_p$ on the nodes of the subtree $\buff_p$.
+Take an arbitrary token in $T_p$, and find the smallest (in the ordering) descendant $q$
+of $p$ that currently has strictly less than $|q|$ tokens. Place the token on $q$,
+and repeat. This is repeated until all descendants $q$ have $|q|$ tokens, or
+all tokens in $T_p$ are used up. When a node $q$ has $|q|$ tokens, we use
+``$q$ is full".
+
+Now for a useful claim. \emph{For non-root $p$, all tokens of $T_p$ are used up.}
+Suppose not. Consider the first such $p$ in the ordering, and look at the situation
+when we finish processing $T_p$. Every descendant $q$ of $p$ must be full. Furthermore,
+all the tokens on $q \in \buff_p$ must have come from some $T_{p'}$, where $p' \in \buff_p$.
+Thus, $\sum_{q \in \buff_p} |q| < |\bigcup_{p' \in \buff_p} T_{p'}|$ $=100\sum_{p' \in \buff_p} |\loss(p')|$. This violates the second condition of buffered shrubs.
+
+Let $r$ be the root of $\buff$. We argue that $T_r$ is \emph{not used up}.
+Suppose the contrary, so $T_r$ is used up. By the previous paragraph,
+all tokens are used up. But this implies that $\sum_{p \in \buff} |p| \geq |\bigcup_{p \in \buff} T_p|$
+$ = 100\sum_{p \in \buff} |\loss(p)|$, violating the buffered condition.
+
+Since $T_r$ is not used up, it must mean that all $p$s are full. Consider any $p$,
+and let $K_p$ be the set of tokens placed at $p$. Each token is labeled $\log |H_v|$,
+for some $v \in \loss(\hat{p})$, for some ancestor $\hat{p}$ of $p$. By \Cor{geometric},
+for every $w \in p$, $\log |H_w| = O(\log |H_v|)$. Because $p$ is full,
+there are exactly $|p|$ tokens on $p$. Thus, $\sum_{w \in p} \log |H_w|$
+is $O(V_p)$, where $V_p$ is the sum of token values in $K_p$. We sum
+over all $p$ to complete the proof.
+
+% \textcolor{red}{\huge Proof missing. Potentially useful observations below}
+%  
+%  Let $r$ be the root of $\buff$.
 First observe that $\loss(r) \geq |r|/100$.
@@
 However, this would contradict the second condition of \Def{buff}.
-
- Consider any node $q\in \buff$.
+%  
+%  Consider any node $q\in \buff$.
 Then for any vertex $v\in \loss(q)$, we have $|H_v| = \Omega(|H_q|)$, since $q$ is unsupported and hence by definition less than $|H_q|/100$ vertices of $H_q$ appear in $\loss(q)$.
-
- \begin{align*}
-  \sum_{p\in \buff}\sum_{v\in p} \log |H_v| &=  O\left( \sum_{p\in \buff}\sum_{v\in p} \log |H_p| \right)   &\mbox{definition of short} \\
-  &= O\left( \sum_{p\in \buff} |p| \log |H_p| \right)   &\mbox{} \\
-  &= ????   &\mbox{buffered is not sufficient to get to next line} \\
-  &= O\left( \sum_{p\in \buff} |\loss(p)| \log |H_p| \right)   &\mbox{} \\
-  &= O\left( \sum_{p\in \buff}\sum_{v\in \loss(p)} \log |H_v| \right)
- \end{align*}
-
- \Cor{geometric} should be used somewhere.
+%  
+%  \begin{align*}
+%   \sum_{p\in \buff}\sum_{v\in p} \log |H_v| &=  O\left( \sum_{p\in \buff}\sum_{v\in p} \log |H_p| \right)   &\mbox{definition of short} \\
+%   &= O\left( \sum_{p\in \buff} |p| \log |H_p| \right)   &\mbox{} \\
+%   &= ????   &\mbox{buffered is not sufficient to get to next line} \\
+%   &= O\left( \sum_{p\in \buff} |\loss(p)| \log |H_p| \right)   &\mbox{} \\
+%   &= O\left( \sum_{p\in \buff}\sum_{v\in \loss(p)} \log |H_v| \right)
+%  \end{align*}
+%  
+%  \Cor{geometric} should be used somewhere.
 \end{proof}
@@
 
-\newcommand{\pruned}{\cU/\cB}
+\newcommand{\pruned}{\cU'}
 \subsection{Paying for Pruned Shrubs: the root is everything} \label{sec:weight}
 
-\begin{definition}\label{def:pruningAShrub}
- Let $\cU$ be a shrub in $\cF$. Consider a reverse topological ordering (i.e.\ leaves first) of the nodes in $\cU$, and use it to define a total ordering.  
- We compute a pruning of $\cU$ (see \Def{prune}), by iteratively removing all buffered sub-shrubs (see \Def{buff}), where buffered sub-shrubs whose roots are earlier in the total ordering are pruned first.
+We focus attention on the remaining shrub left by \Lem{split-buff}: this is 
+a shrub $\cU$ such that no subtree of $\cU$ is buffered. Refer to such
+a shrub as \emph{buffer-pruned}.
+
+% 
+% \begin{definition}\label{def:pruningAShrub}
+%  Let $\cU$ be a shrub in $\cF$. Consider a reverse topological ordering (i.e.\ leaves first) of the nodes in $\cU$, and use it to define a total ordering.  
+%  We compute a pruning of $\cU$ (see \Def{prune}), by iteratively removing all buffered sub-shrubs (see \Def{buff}), where buffered sub-shrubs whose roots are earlier in the total ordering are pruned first.
 We use $\pruned$ to denote the resulting sub-shrub, which we refer to as the \emph{buffer-pruned} sub-shrub of $\cU$.
-\end{definition}
-
-The goal of this section is to prove the following lemma, which ultimately allows us to pay for pruned shrubs.
+% \end{definition}
+% 
+The goal of this section is to prove the following lemma.
+% , which ultimately allows us to pay for pruned shrubs.
 
@@
 
-\textcolor{red}{\huge The rest of this subsection needs updating}
-
-We define a \emph{residue} $R_p$ for each $p \in \cU$. Suppose $p$ has children
-$q_1, q_2, \ldots, q_k$ in $\cU$. Then $R_p = |\redH_p| - \sum_i |\redH_{q_i}|$. 
-By definition, $|\redH_p| = \sum_{q \in \cU_p} R_q$. 
-Note that $R_p$ can be negative. 
-Now, define $R^+_p = \max(R_p,0)$, and set $W_p = \sum_{q \in \cU_p} R^+_q$, 
-where $\cU_p$ is the subtree of $\cU$ rooted at $p$.
-Observe that $W_p \geq |\redH_p|$. We also get an approximate converse.
-
-\begin{claim} \label{clm:loss} For any path $p\in \cU$, $|\redH_p| \geq W_p - 2\sum_{q \in \cU_p} |q|$.
+We need the following technical claim. Let $N(\pruned)$ denote the set of nodes in $\pruned$.
+
+\begin{claim} \label{clm:weights} There exists a function $\alpha:N(\pruned) \mapsto \RR^+$
+with the following properties. (We use $\alpha_p$ to denote the function value.)
+Let $A_p = \sum_{q \in \pruned_p} \alpha_q$.
+Then, $A_p - 3\sum_{q \in \pruned_p, q \neq p} |q| \leq |H_p| \leq A_p$.
 \end{claim}
 
-\begin{proof} We write $W_p - |\redH_p| = \sum_{q \in \cU_p} (R^+_q - R_q)$ $= -\sum_{q \in \cU_p: R_q < 0} R_q$.
-Consider $q \in \cU_p$ and denote the children in $\cU_p$ by $q'_1, q'_2, \ldots$.
-Note that $R_q$ is negative exactly when $|\redH_q| < \sum_i |\redH_{q'_i}|$.
-Traverse $\pmax(T)$ from $q'_i$ to $q$.
-Other than $q$, all other nodes encountered are in the tall support chain of $q'_i$ and hence
-do not affect its reduced heap.
-The vertices of $\redH_{q'_i}$ that are deleted
-are exactly those present in the path $q$. 
-Any vertex in $q$ can be deleted from at most two of the reduced heaps (of the children of $q$ in $\cU_p$), since 
-theses reduced heaps do not have an ancestor-descendant relationship.
-Therefore when $R_q$ is negative, it is at most by $2|q|$.
-We sum over all $q$ to complete the proof.
-\end{proof}
-
-\ignore{Let $D_q = H_q \setminus \bigcup_i H_{q'_i}$. In other words, as we traverse $\cU$,
-$D_q$ is the set of vertices ``added" at $q$. Suppose we traverse $\pmax(T)$ from $q'_i$
-to $q$. All vertices of $H_{q'_i}$ that lie on these paths are removed.
-
- and traverse $\pmax(T)$ up to $p$. Suppose we encounter
-the paths $q_0 = q, q_1, q_2, \ldots, q_k = p$. The vertices in the residue $R_q$ that 
-are removed from $H_p$ are exactly the vertices that lie in these paths. 
-We split this contribution into tall and short paths, observing that the latter
-are all unsupported. Hence, we can (crudely) upper bound the number of vertices
-removed by $\sum_{q \in \cU_p} |q| + \sum_{q \in \cU_p} \sum_{h \in \cH} |h \cap q|$.
-Each vertex can be removed from at most two distinct $R_q$'s, completing the proof.
-}
-
-The main challenge of the entire proof is bounding the sum of path lengths, which is done next.
-We stress that the all the previous work is mostly the setup for this claim.
-
-\begin{claim} \label{clm:charge} Fix any path $p \in \cU\setminus\{r(\cU)\}$. Suppose for any $q,q' \in \cU_p$
-where $q$ is a parent of $q'$ in $\cU_p$, $W_q \geq (4/3)W_{q'}$. Then $\sum_{q \in \cU_p} |q| \leq W_p/20$.
+\begin{proof} We define $\alpha$ inductively, from leaves upwards.
+For a leaf $\ell$ in $\pruned$, set $\alpha_\ell = |H_\ell|$. Consider internal node $p$
+with children $q_1, q_2, \ldots, q_k$. The heap $H(p)$ is formed by the union
+of the $H(q_i)$ heaps, after deleting certain vertices from them, and potentially
+adding some new vertices. The new vertices are potentially those that come
+from other nodes that have been pruned off.
+The vertices in each $q_i$ as well as $\lambda(q_i)$ are deleted.
+Define $\alpha_p$ to be the number of new vertices added to $H_p$.
+
+Thus, we have $|H_p| = \sum_{i \leq k} |H_q| + \alpha_p$
+$- \textrm{\# vertices deleted from $H_q$s}$.
+The number of vertices deleted is at most $2\sum_{i \leq k} (|q_i| + |\lambda(q_i)|)$.
+There is a factor $2$ because each vertex appears in at most $2$ heaps.
+
+It is evident that $|H_p| \leq \sum_{q \in \pruned_p, q \neq p} \alpha_q = A_p$.
+Note that $|H_p| \geq A_p - 2\sum_{q \in \pruned_p, q \neq p} (|q| + |\lambda(q)|)$.
+Crucially, since $\pruned$ is buffer-pruned, $\sum_{q \in \pruned_p, q \neq p} |\lambda(p)|
+\leq \sum_{q \in \pruned_p} |q|/100$. This completes the proof.
+\end{proof}
+
+We have the main charging claim. This assumes that $A_p$ is increasing exponentially
+as we go up the tree. That property is proven later by induction, using
+the following claim for sub-shrubs.
+
+\begin{claim} \label{clm:charge} Fix any path $p \in \pruned\setminus r$. Suppose for any $q,q' \in \pruned_p$
+where $q$ is a parent of $q'$ in $\pruned_p$, $A_q \geq (5/4)A_{q'}$. Then $\sum_{q \in \pruned_p} |q| \leq A_p/20$.
 \end{claim}
 
-\begin{proof} Since $q$ is an unsupported short path, $|q| < \sqrt{|H_q|}/100 \leq \sqrt{|\redH_q|}/99 \leq \sqrt{|W_q|}/99$.
-We prove that $\sum_{q \in \cU_p} \sqrt{|W_q|}/99 \leq W_p/20$
-by a charge redistribution scheme. Assume that each $q \in \cU_p^-$
-starts with $\sqrt{W_q}/99$ units of charge. We redistribute this charge over all
-nodes in $\cU_q$, and then calculate the total charge. For $q \in \cU_p$, spread
-its charge to all nodes in $\cU_q$ proportional to $R^+$ values. In other words,
-give $(\sqrt{W_q}/99)\cdot(R^+_{q'}/W_q)$ units of charge to each $q' \in \cU_q$.
-
-After the redistribution, let us compute the charge deposited at $q$. Every ancestor in $\cU_p^-$ 
-$q = a_0, a_1, a_2, \ldots, a_k$ contributes to the charge at $q$. The charge is expressed in the following
-equation. We use the assumption that $W_{a_i} \geq (4/3) W_{a_{i-1}}$ and hence $W_{a_i} \geq (4/3)^i W_{a_0} \geq (4/3)^i$, 
-as $a_0$ is an unsupported short path and hence $W_{a_0}\geq 1$. 
-$$ ({R^+_q}/99)\sum_{a_i} 1/\sqrt{W_{a_i}} \leq ({R^+_q}/99)\sum_{a_i} (3/4)^{i/2} \leq R^+_q/20 $$
-The total charge is $\sum_{q \in \cU_p} R^+_p/20 = W_p/20$. 
-\end{proof}
-
-\begin{corollary} \label{cor:chargeCor}
- Let $r$ be the root of $\cU$, and suppose that for any paths $q,q' \in \cU\setminus \{r\}$, where $q$ is a parent of $q'$ in $\cU$, $W_q \geq (4/3)W_{q'}$.
- Then $\sum_{p \in \cU} |p| \leq W_r/20+|r|$.
-\end{corollary}
-\begin{proof}
- Let $c_1, \dots, c_m$ be the children of $r$ in $\cU$.  By definition, $W_r = \sum_i W_{c_i}+R^+_r \geq \sum_i W_{c_i}$.
- By \Clm{charge}, for each $c_i$ we have $W_{c_i}/20\geq \sum_{p \in \cU_{c_i}} |p|$.  Combining these to facts yields the claim.
-\end{proof}
+\begin{proof} Since $q$ is an unsupported short path, $|q| < \sqrt{|H_q|}/100 \leq \sqrt{|A_q|}/99$.
+We prove that $\sum_{q \in \pruned_p} \sqrt{|A_q|}/99 \leq A_p/20$
+by a charge redistribution scheme. Assume that each $q \in \pruned_p$
+starts with $\sqrt{A_q}/99$ units of charge. We redistribute this charge over all
+nodes in $\pruned_q$, and then calculate the total charge. For $q \in \pruned_p$, spread
+its charge to all nodes in $\pruned_q$ proportional to $\alpha$ values. In other words,
+give $(\sqrt{A_q}/99)\cdot(\alpha_{q'}/A_q)$ units of charge to each $q' \in \pruned_q$.
+
+After the redistribution, let us compute the charge deposited at $q$. Every ancestor in $\pruned_p$,
+denoted $a_0, a_1, a_2, \ldots, a_k$, contributes to the charge at $q$. The charge is expressed in the following
+equation. Ae use the assumption that $A_{a_i} \geq (5/4) A_{a_{i-1}}$ and hence $A_{a_i} \geq (5/4)^i A_{a_0} \geq (5/4)^i$, 
+as $a_0$ is an unsupported short path and hence $A_{a_0}\geq 1$. 
+$$ ({\alpha_q}/99)\sum_{a_i} 1/\sqrt{A_{a_i}} \leq ({\alpha_q}/99)\sum_{a_i} (5/4)^{i/2} \leq \alpha_q/20 $$
+The total charge is $\sum_{q \in \pruned_p} \alpha_p/20 = A_p/20$. 
+\end{proof}
+
+\begin{claim} \label{clm:exp-A} For any $q, q' \in \pruned_p$ where $q$
+is a parent of $q'$, $A_q \geq (4/3)A_{q'}$.
+\end{claim}
+
+\begin{proof} We prove this by induction over the depth of $q$.
+When $q$ is a leaf, this is vacuously tree. Consider some internal node $q$,
+with children $q'_1, q'_2, \ldots, q'_k$. Applying the induction
+hypothesis to the subshrubs rooted at $q'_i$, we can invoke \Clm{charge}
+on each subshrub $\pruned_{q'_i}$. Thus, for all $i$,
+$\sum_{\hat{q} \in \pruned_{q'_i}} |\hat{q}| \leq A_{q'_i}/20$. 
+By \Clm{weights}, $|H_{q'_i}| \leq 17A_{q'_i}/20$ and $A_q \geq |H_q|$.
+By \Lem{geometric}, $|H_q| \geq (3/2)|H_{q'_i}|$. We combine these bounds
+to complete the proof.
+\end{proof}
+
+The proof of \Lem{cu-root} just combines all these claims, and essentially
+repeats the argument in the previous claim. We apply \Clm{exp-A} and \Clm{charge}
+to every child of the root $r$. Denoting the children as $q_1, q_2, \ldots$,
+we get that $\sum_{q' \in \pruned_{q_i}} |q| \leq A_{q_i}/20$.
+Summing this over the children, $\sum_{p \in \pruned, p \neq r} \leq A_r/20$.
+By \Clm{weights}, $A_r \leq 17|H_r|/20$. Furthermore $|r| \leq |H_r$.
+We combine these bounds to show that $\sum_{p \in \pruned} |p| = O(|H_r|)$.
+% 
+% 
+% 
+% \begin{proof}[Proof of \Lem{cu-root}] We first pro 
+% We use $r$ for $r(\cU)$. %Our aim is to simply apply \Clm{charge} from $r(\cU)$. 
+% Suppose $W_q \geq (4/3)W_{q'}$ (for any choice in $\cU\setminus\{r\}$ of $q$ parent of $q'$), 
+% then by \Cor{chargeCor}, $\sum_{p \in \cU} |p| \leq W_{r}/20+|r|$.
+% By \Clm{loss}, $|\redH_r| \geq W_r - 2\sum_{p \in \cU} |p|$, and so combining these inequalities gives,
+% \[
+% % \sum_{p \in \cU} |p| \leq \pth{|\redH_r|+2\sum_{p \in \cU} |p|}/20+|r| \Rightarrow 
+%  \sum_{p \in \cU} |p|\leq \frac{10}{9} \pth{|\redH_r|/20+|r|} \leq 
+%  \frac{10}{9}\pth{|H_r|/20+\sqrt{|H_r|}/100} \leq |H_r|/10.
+% \]
+% 
+% We now prove that for any $q$ parent of $q'$ (other than $r$), $W_q \geq (4/3)W_{q'}$.
+% Suppose not. Let $p, p'$ be the counterexample furthest from the root,
+% where $p$ is the parent of $p'$. 
+% Note that for $q$ and child $q'$ in $\cU_{p'}$, $W_q \geq (4/3)W_{q'}$.
+% We will apply \Clm{charge} for $\cU_{p'}$ to deduce that $\sum_{q \in \cU_{p'}} |q| \leq W_{p'}/20$.
+% Combining this with \Clm{loss} gives, $|\redH_{p'}| \geq 19W_{p'}/20$. By \Lem{geometric}, $|\redH_p| \geq (3/2)|\redH_{p'}|$.
+% Noting that $W_p \geq |\redH_p|$, we deduce that $W_p \geq (4/3)W_{p'}$.
+% Hence, $p, p'$ is not a counterexample, and more generally, there is no counterexample.
+% That completes the whole proof.
+% \end{proof}
+% 
+% 
+% \textcolor{red}{\huge The rest of this subsection needs updating}
+% 
+% We define a \emph{residue} $R_p$ for each $p \in \cU$. Suppose $p$ has children
+% $q_1, q_2, \ldots, q_k$ in $\cU$. Then $R_p = |\redH_p| - \sum_i |\redH_{q_i}|$. 
+% By definition, $|\redH_p| = \sum_{q \in \cU_p} R_q$. 
+% Note that $R_p$ can be negative. 
+% Now, define $R^+_p = \max(R_p,0)$, and set $W_p = \sum_{q \in \cU_p} R^+_q$, 
+% where $\cU_p$ is the subtree of $\cU$ rooted at $p$.
+% Observe that $W_p \geq |\redH_p|$. We also get an approximate converse.
+% 
+% 
+% \begin{claim} \label{clm:loss} For any path $p\in \cU$, $|\redH_p| \geq W_p - 2\sum_{q \in \cU_p} |q|$.
+% \end{claim}
+% 
+% \begin{proof} We write $W_p - |\redH_p| = \sum_{q \in \cU_p} (R^+_q - R_q)$ $= -\sum_{q \in \cU_p: R_q < 0} R_q$.
+% Consider $q \in \cU_p$ and denote the children in $\cU_p$ by $q'_1, q'_2, \ldots$.
+% Note that $R_q$ is negative exactly when $|\redH_q| < \sum_i |\redH_{q'_i}|$.
+% Traverse $\pmax(T)$ from $q'_i$ to $q$.
+% Other than $q$, all other nodes encountered are in the tall support chain of $q'_i$ and hence
+% do not affect its reduced heap.
+% The vertices of $\redH_{q'_i}$ that are deleted
+% are exactly those present in the path $q$. 
+% Any vertex in $q$ can be deleted from at most two of the reduced heaps (of the children of $q$ in $\cU_p$), since 
+% theses reduced heaps do not have an ancestor-descendant relationship.
+% Therefore when $R_q$ is negative, it is at most by $2|q|$.
+% We sum over all $q$ to complete the proof.
+% \end{proof}
+% 
+% \ignore{Let $D_q = H_q \setminus \bigcup_i H_{q'_i}$. In other words, as we traverse $\cU$,
+% $D_q$ is the set of vertices ``added" at $q$. Suppose we traverse $\pmax(T)$ from $q'_i$
+% to $q$. All vertices of $H_{q'_i}$ that lie on these paths are removed.
+% 
+% 
+%  and traverse $\pmax(T)$ up to $p$. Suppose we encounter
+% the paths $q_0 = q, q_1, q_2, \ldots, q_k = p$. The vertices in the residue $R_q$ that 
+% are removed from $H_p$ are exactly the vertices that lie in these paths. 
+% We split this contribution into tall and short paths, observing that the latter
+% are all unsupported. Hence, we can (crudely) upper bound the number of vertices
+% removed by $\sum_{q \in \cU_p} |q| + \sum_{q \in \cU_p} \sum_{h \in \cH} |h \cap q|$.
+% Each vertex can be removed from at most two distinct $R_q$'s, completing the proof.
+% }
+% 
+% 
+% The main challenge of the entire proof is bounding the sum of path lengths, which is done next.
+% We stress that the all the previous work is mostly the setup for this claim.
+% 
+% 
+% \begin{corollary} \label{cor:chargeCor}
+%  Let $r$ be the root of $\cU$, and suppose that for any paths $q,q' \in \cU\setminus \{r\}$, where $q$ is a parent of $q'$ in $\cU$, $W_q \geq (4/3)W_{q'}$.
+%  Then $\sum_{p \in \cU} |p| \leq W_r/20+|r|$.
+% \end{corollary}
+% \begin{proof}
+%  Let $c_1, \dots, c_m$ be the children of $r$ in $\cU$.  By definition, $W_r = \sum_i W_{c_i}+R^+_r \geq \sum_i W_{c_i}$.
+%  By \Clm{charge}, for each $c_i$ we have $W_{c_i}/20\geq \sum_{p \in \cU_{c_i}} |p|$.  Combining these to facts yields the claim.
+% \end{proof}
+% 
+%
 
@@
 
-\begin{proof}[Proof of \Lem{cu-root}]
-We use $r$ for $r(\cU)$. %Our aim is to simply apply \Clm{charge} from $r(\cU)$. 
-Suppose $W_q \geq (4/3)W_{q'}$ (for any choice in $\cU\setminus\{r\}$ of $q$ parent of $q'$), 
-then by \Cor{chargeCor}, $\sum_{p \in \cU} |p| \leq W_{r}/20+|r|$.
-By \Clm{loss}, $|\redH_r| \geq W_r - 2\sum_{p \in \cU} |p|$, and so combining these inequalities gives,
-\[
-% \sum_{p \in \cU} |p| \leq \pth{|\redH_r|+2\sum_{p \in \cU} |p|}/20+|r| \Rightarrow 
- \sum_{p \in \cU} |p|\leq \frac{10}{9} \pth{|\redH_r|/20+|r|} \leq 
- \frac{10}{9}\pth{|H_r|/20+\sqrt{|H_r|}/100} \leq |H_r|/10.
-\]
-
-We now prove that for any $q$ parent of $q'$ (other than $r$), $W_q \geq (4/3)W_{q'}$.
-Suppose not. Let $p, p'$ be the counterexample furthest from the root,
-where $p$ is the parent of $p'$. 
-Note that for $q$ and child $q'$ in $\cU_{p'}$, $W_q \geq (4/3)W_{q'}$.
-We will apply \Clm{charge} for $\cU_{p'}$ to deduce that $\sum_{q \in \cU_{p'}} |q| \leq W_{p'}/20$.
-Combining this with \Clm{loss} gives, $|\redH_{p'}| \geq 19W_{p'}/20$. By \Lem{geometric}, $|\redH_p| \geq (3/2)|\redH_{p'}|$.
-Noting that $W_p \geq |\redH_p|$, we deduce that $W_p \geq (4/3)W_{p'}$.
-Hence, $p, p'$ is not a counterexample, and more generally, there is no counterexample.
-That completes the whole proof.
-\end{proof}
-
 \subsection{Proving \Thm{runtime}} \label{sec:thm-runtime}
@@
 
-\begin{proof} 
-For any component $\cU$ of the shrub forest $\cF$ of $\cL\setminus \cL'$, let $B(\cU)$ denote the collection of buffered sub-shrubs that we 
-removed in the pruning process of \Def{pruningAShrub} that produced the buffer-pruned sub-shrub $\pruned$
-(i.e.\ these are disjoint sub-shrubs whose union cover $\cU\setminus (\pruned)$). 
-Thus we can decompose the sum as follows, 
+\begin{proof} Apply \Lem{split-buff} to $\cU$ to get the family of buffered
+sub-shrubs $B(\cU)$ and buffer-pruned $\pruned$.
+% For any component $\cU$ of the shrub forest $\cF$ of $\cL\setminus \cL'$, let $B(\cU)$ denote the collection of buffered sub-shrubs that we 
+% remo that produced the buffer-pruned sub-shrub $\pruned$
+% (i.e.\ these are disjoint sub-shrubs whose union cover $\cU\setminus (\pruned)$). 
+% Thus we can decompose the sum as follows, 
 \begin{align*}

### 2016-08-01 paint-contour-tree.tex
typos fixed

@@
 We prove an important lemma. The cost of heap operations corresponding 
-to a buffered sub-shrub to be charged to tall support chains.
+to a buffered sub-shrub can be charged to tall support chains.
 
@@
 We will place all these tokens on the nodes in $\buff$, using the following process.
-We reserve topologically sort (descendants before ancestors) the nodes in $\buff$.
+We reverse topologically sort (descendants before ancestors) the nodes in $\buff$.
 For each $p$ in order, we will place tokens in $T_p$ on the nodes of the subtree $\buff_p$.
@@
 \begin{proof} We define $\alpha$ inductively, from leaves upwards.
-For a leaf $\ell$ in $\pruned$, set $\alpha_\ell = |H_\ell|$. Consider internal node $p$
+For a leaf $\ell$ in $\pruned$, set $\alpha_\ell = |H_\ell|$. Consider an internal node $p$
 with children $q_1, q_2, \ldots, q_k$. The heap $H(p)$ is formed by the union
@@
 denoted $a_0, a_1, a_2, \ldots, a_k$, contributes to the charge at $q$. The charge is expressed in the following
-equation. Ae use the assumption that $A_{a_i} \geq (5/4) A_{a_{i-1}}$ and hence $A_{a_i} \geq (5/4)^i A_{a_0} \geq (5/4)^i$, 
+equation. We use the assumption that $A_{a_i} \geq (5/4) A_{a_{i-1}}$ and hence $A_{a_i} \geq (5/4)^i A_{a_0} \geq (5/4)^i$, 
 as $a_0$ is an unsupported short path and hence $A_{a_0}\geq 1$. 

### 2016-08-02 paint-contour-tree.tex
small changes

@@
 \begin{corollary}\label{cor:geometric}
- Let $\cU$ denote a connected component of $\cF$. Let $p$ and $d$ be any two nodes in $\cU$ such that $d$ is a descendant of $p$.
- Then for any $w \in d$ and $v \in \lambda(p)$, $|H_w| = O(|H_v|)$.
+ Let $\cU$ denote a connected component of $\cF$. Let $p$ and $d$ be any two nodes in $\cU$ such that $d$ is a descendant of $p$, and $p$ 
+ is not the root of $\cU$.  Then for any $w \in d$ and $v \in \lambda(p)$, $|H_w| = O(|H_v|)$.
 \end{corollary}
@@
 \begin{proof} By the previous lemma, $|H_w| = O(|H_p|)$. Note that there are at most
-$|H_p|/100$ nodes in $\lambda(p)$, since these are all in the tall support chain of $v$
-(which is unsupported). Thus, for any $v \in \lambda(p)$, $H_v \geq 99|H_p|/100$,
+$|H_p|/100$ nodes in $\lambda(p)$, since these are all in the tall support chain of $p$
+(which is unsupported). Thus, for any $v \in \lambda(p)$, $|H_v| \geq 99|H_p|/100$,
 completing the proof.
@@
 \begin{definition}\label{def:buff}
- Let $\cU'$ be a pruning of a shrub of $\cU$ of $\cF$. Then a rooted subtree $\buff$ of $\cU'$ is called a \emph{buffered sub-shrub} if the following conditions hold:
+ Let $\cU'$ be a pruning of a shrub of $\cU$ of $\cF$.
+Then a rooted subtree $\buff$ of $\cU'$ is called a \emph{buffered sub-shrub}, if it does not contain the root of $\cU'$ (i.e.\ it is a proper subtree), and the following conditions hold:
   \begin{enumerate}
@@
 We process the nodes in order. For node $p$, let $\cU'_p$ denote the rooted
-subtree of $p$ in $\cU'$. We simply check if $\sum_{q \in \cU'_p} |\loss(p)| \geq \sum_{q \in \cU'_p} |q|/100$. If so, we prune off $\cU'_p$. (Note that $\cU'$ is redefined
+subtree of $p$ in $\cU'$. We simply check if $\sum_{q \in \cU'_p} |\loss(p)| \geq \sum_{q \in \cU'_p} |q|/100$. If so, we prune off $\cU'_p$, unless $p$ is the root. 
+(Note that $\cU'$ is redefined
 to be the remainder.)
@@
 \Lem{cu-root} asserts the root $r$ in $\pruned$ pretty much encompasses all sizes and heaps in $\pruned$.
-Observe that part (i) of \Lem{cu-root} is immediate, as it is a special case of \Cor{geometric}. Proving part (ii) requires much more work.
+Observe that part (i) of \Lem{cu-root} is immediate, as it follows directly from \Lem{geometric}. Proving part (ii) requires much more work.
 % It is helpful to spell out how the various heaps evolve. Consider $p \in \cU$,
@@
 For a leaf $\ell$ in $\pruned$, set $\alpha_\ell = |H_\ell|$. Consider an internal node $p$
-with children $q_1, q_2, \ldots, q_k$. The heap $H(p)$ is formed by the union
-of the $H(q_i)$ heaps, after deleting certain vertices from them, and potentially
+with children $q_1, q_2, \ldots, q_k$. The heap $H_p$ is formed by the union
+of the $H_{q_i}$ heaps, after deleting certain vertices from them, and potentially
 adding some new vertices. The new vertices are potentially those that come
@@
 Thus, we have $|H_p| = \sum_{i \leq k} |H_q| + \alpha_p$
-$- \textrm{\# vertices deleted from $H_q$s}$.
+$- \textrm{\# vertices deleted from $H_q$'s}$.
 The number of vertices deleted is at most $2\sum_{i \leq k} (|q_i| + |\lambda(q_i)|)$.
@@
 as $a_0$ is an unsupported short path and hence $A_{a_0}\geq 1$. 
-$$ ({\alpha_q}/99)\sum_{a_i} 1/\sqrt{A_{a_i}} \leq ({\alpha_q}/99)\sum_{a_i} (5/4)^{i/2} \leq \alpha_q/20 $$
+$$ ({\alpha_q}/99)\sum_{a_i} 1/\sqrt{A_{a_i}} \leq ({\alpha_q}/99)\sum_{a_i} (4/5)^{i/2} \leq \alpha_q/20 $$
 The total charge is $\sum_{q \in \pruned_p} \alpha_p/20 = A_p/20$. 
@@
 \begin{proof} We prove this by induction over the depth of $q$.
-When $q$ is a leaf, this is vacuously tree. Consider some internal node $q$,
+When $q$ is a leaf, this is vacuously true. Consider some internal node $q$,
 with children $q'_1, q'_2, \ldots, q'_k$. Applying the induction
@@
 Summing this over the children, $\sum_{p \in \pruned, p \neq r} \leq A_r/20$.
-By \Clm{weights}, $A_r \leq 17|H_r|/20$. Furthermore $|r| \leq |H_r$.
+By \Clm{weights}, $A_r \leq 17|H_r|/20$. Furthermore $|r| \leq |H_r|$.
 We combine these bounds to show that $\sum_{p \in \pruned} |p| = O(|H_r|)$.

### 2016-08-03 paint-contour-tree.tex
small changes
fixed issue with Claim 8.15
claim 8.15 modified

@@
 Let $A_p = \sum_{q \in \pruned_p} \alpha_q$.
-Then, $A_p - 3\sum_{q \in \pruned_p, q \neq p} |q| \leq |H_p| \leq A_p$.
+Then, $A_p - 3\sum_{q \in \pruned_p} |q| \leq |H_p| \leq A_p$.
 \end{claim}
@@
 of the $H_{q_i}$ heaps, after deleting certain vertices from them, and potentially
-adding some new vertices. The new vertices are potentially those that come
-from other nodes that have been pruned off.
-The vertices in each $q_i$ as well as $\lambda(q_i)$ are deleted.
+adding some new vertices. 
+The vertices in $p$ and those in the $\lambda(q_i)$ are deleted.
 Define $\alpha_p$ to be the number of new vertices added to $H_p$.
 
-Thus, we have $|H_p| = \sum_{i \leq k} |H_q| + \alpha_p$
-$- \textrm{\# vertices deleted from $H_q$'s}$.
-The number of vertices deleted is at most $2\sum_{i \leq k} (|q_i| + |\lambda(q_i)|)$.
+Thus, we have $|H_p| = \sum_{i \leq k} |H_{q_i}| + \alpha_p$
+$- \textrm{\# vertices deleted from $H_{q_i}$'s}$.
+The number of vertices deleted is at most $2(|p|+\sum_{i \leq k} |\lambda(q_i)|)$.
 There is a factor $2$ because each vertex appears in at most $2$ heaps.
 
-It is evident that $|H_p| \leq \sum_{q \in \pruned_p, q \neq p} \alpha_q = A_p$.
-Note that $|H_p| \geq A_p - 2\sum_{q \in \pruned_p, q \neq p} (|q| + |\lambda(q)|)$.
+It is evident that $|H_p| \leq \sum_{q \in \pruned_p} \alpha_q = A_p$.
+Note that $|H_p| \geq A_p - 2\sum_{q \in \pruned_p, q \neq p} |\lambda(q)|$ $-2\sum_{q \in \pruned_p} |q|$.
 Crucially, since $\pruned$ is buffer-pruned, $\sum_{q \in \pruned_p, q \neq p} |\lambda(p)|
-\leq \sum_{q \in \pruned_p} |q|/100$. This completes the proof.
+\leq \sum_{q \in \pruned_p, q \neq p} |q|/100$ $\leq \sum_{q \in \pruned_p} |q|/100$. This completes the proof.
 \end{proof}
@@
 
-\begin{claim} \label{clm:exp-A} For any $q, q' \in \pruned_p$ where $q$
-is a parent of $q'$, $A_q \geq (4/3)A_{q'}$.
+\begin{claim} \label{clm:exp-A} Fix any path $p \in \pruned \setminus r$.
+For any $q, q' \in \pruned_p$ where $q$
+is a parent of $q'$, $A_q \geq (5/4)A_{q'}$.
 \end{claim}
@@
 $\sum_{\hat{q} \in \pruned_{q'_i}} |\hat{q}| \leq A_{q'_i}/20$. 
-By \Clm{weights}, $|H_{q'_i}| \leq 17A_{q'_i}/20$ and $A_q \geq |H_q|$.
+By \Clm{weights}, 
+$|H_{q'_i}| \geq 17A_{q'_i}/20$ and $A_q \geq |H_q|$.
 By \Lem{geometric}, $|H_q| \geq (3/2)|H_{q'_i}|$. We combine these bounds
@@
 to every child of the root $r$. Denoting the children as $q_1, q_2, \ldots$,
-we get that $\sum_{q' \in \pruned_{q_i}} |q| \leq A_{q_i}/20$.
-Summing this over the children, $\sum_{p \in \pruned, p \neq r} \leq A_r/20$.
-By \Clm{weights}, $A_r \leq 17|H_r|/20$. Furthermore $|r| \leq |H_r|$.
+we get that $\sum_{p \in \pruned_{q_i}} |p| \leq A_{q_i}/20$.
+Summing this over the children, $\sum_{p \in \pruned, p \neq r} |p| \leq A_r/20$.
+By \Clm{weights}, $|H_r| \geq 17A_r/20 - 3|r|$. Since $r$ represented a short path, 
+we have the bound $|r| \leq \sqrt{|H_r|}/100$. Thus, $|H_r| = \Omega(A_r)$.
 We combine these bounds to show that $\sum_{p \in \pruned} |p| = O(|H_r|)$.

### 2016-12-20 paint-contour-tree.tex
update for dcg

@@
 \author{
-  Benjamin Raichel%\BenThanks{}
+  Benjamin Raichel\thanks{%
+      Department of Computer Science; %
+      University of Texas at Dallas; %
+      Richardson, TX, 75080, USA; %
+      \texttt{\si{benjamin.raichel}@\si{utdallas}.\si{edu}} %
+  }
   \and
-  C. Seshadhri%\thanks{University of California, Santa Cruz; {\tt scomandu@ucsc.edu}}
+  C. Seshadhri\thanks{%
+        Department of Computer Science; %
+      University of California, Santa Cruz; %
+      Santa Cruz, CA, 95064, USA; %
+      \texttt{\si{scomandu}@\si{ucsc}.\si{edu}} %
+  }
 }
@@
 If $C$ denotes the set of critical points in $\MM$, the running time is
-roughly $O(\sum_{v \in C} \log \ell_v)$, where $\ell_v$ is the depth of $v$ in the contour tree.
+roughly $O(N+\sum_{v \in C} \log \ell_v)$, where $\ell_v$ is the depth of $v$ in the contour tree and $N$ is the total complexity of $\MM$.
 This matches all existing upper bounds, but is a significant asymptotic improvement when the contour tree is short and fat.
@@
 An edge of the contour tree connects two critical points if one event immediately ``follows"
-the other as the sweep plane makes its pass. (We provide formal definitions later.) 
+the other as the sweep plane makes its pass. We provide formal definitions later. 
 \Fig{relative} and \Fig{sorting} 
@@
 This bound immediately yields a run time of $O(t\log D + t\alpha(t) + N)$,
-where $D$ is the diameter of the contour tree. 
+where $D$ is the directed diameter of the contour tree. 
 This is a significant asymptotic improvement for short and fat contour trees. For example, if the tree is balanced,
@@
 of terrains ($d=2$) with $t$ critical points. The bound of \Thm{main-alg} takes values ranging
-from $t$ to $t\log t$. Consider some $\thmConst \in [t,t\log t]$, and consider the set of terrains
-where the algorithm makes $\thmConst$ comparisons. Then \emph{any algorithm} must make roughly $\thmConst$
+from $t$ to $c t\log t$, for some constant $c$. Consider some $\thmConst \in [t,t\log t]$, and consider the set of terrains
+where the algorithm makes $\alpha \thmConst$ comparisons, for any constant $\alpha$. Then \emph{any algorithm} must make roughly $\thmConst$
 comparisons in the worst-case over this set. (Further details are in \InSoCGVer{the full version \CiteFullVer}\InNotSoCGVer{\Sec{lb}}.)
@@
 The input is a continuous piecewise-linear function $f:\MM \to \RR$, where $\MM$ is a simply connected and fully triangulated simplicial complex in $\RR^d$,
-except for specially designated \emph{boundary facets}. So $f$ is explicitly defined only on the vertices of $\MM$,
-and all other values are obtained by linear interpolation.
+except for specially designated \emph{boundary facets}. We assume that all vertices of $\MM$ have distinct function values, 
+and that $f$ is only explicitly defined on the vertices, and all other values are obtained by linear interpolation.
 
@@
 
-\begin{definition} \label{def:level} The \emph{level set} at value $h$ is the set $\{x| f(x) = h\}$.
+\begin{definition} \label{def:level} The \emph{level set} at value $h$ is the set $\{x\in \MM| f(x) = h\}$.
 A \emph{contour} is a connected component of a level set. An \emph{$h$-contour} is a contour where $f$-values are $h$.
@@
 As Carr points out in Chapter 6 of his thesis, the term contour tree can be used for a family of related structures.
-Every vertex in $\MM$ is associated with an edge in $\reeb(\MM)$, and sometimes the
+Every regular vertex in $\MM$ is associated with an edge in $\reeb(\MM)$, and sometimes the
 vertex is explicitly placed in $\reeb(\MM)$ (by subdividing the respective edge). This is referred
@@
 \InNotSoCGVer{
-\subsection{Some technical remarks}
+\subsection{Some remarks}
 \technicalRemarks
@@
 \section{A tour of the new contour tree algorithm} \label{sec:approach}
-\InNotSoCGVer{
 Our final algorithm is quite technical and has numerous moving parts. However, for the $d=2$ case, 
@@
 can be explained clearly.  Therefore, here we first provide a high level view of the entire result.
-}
-\InSoCGVer{
-For the $d=2$ case, where the input is just a triangulated terrain, the main ideas of the parts of the algorithm 
-can be explained clearly.  Therefore, here we provide a high level view of the entire result.
-}
 
@@
 In such a complex, there exists a non-ascending path from any point to this unique minimum.
-Analogously, we can define maximum dominant complexes,
-and both are collectively called extremum dominant.
+Analogously, we can define \emph{maximum dominant} complexes,
+and both are collectively called \emph{extremum dominant}.
 
@@
 while there may be numerous disconnected dry parts. The interface between the dry and wet parts
-is a set of contours\footnote{Technically, they are not contours, but rather the limits of sequences of contours.}, 
+%is a set of contours\footnote{Technically, they are not contours, but rather the limits of sequences of contours.}, 
+is a set of limits of sequences of contours, which informally we call just a set of contours, which are
 given by the ``water line". The wet part is clearly maximum dominant, since all
@@
 to construct the true join tree must construct the path of splits, which implies
-sorting all of them. It is absolutely necessary to cut $\MM$ into pieces where the cost
+sorting all of them. Thus it is necessary to cut $\MM$ into pieces where the cost
 of building the join tree can be related to that of building $\cC(\MM)$.
@@
 then spill from the next, etc. Paint is viscous, and only flows down edges. \emph{It does not
-paint the interior of higher dimensional faces.} That is, this process is restricted to the 1-skeleton of $\NN$.
+paint the interior of faces with dimension strictly larger than 1.} That is, this process is restricted to the 1-skeleton of $\NN$.
 Furthermore, our paints do not mix, so each edge receives a unique color, decided by the first
@@
 This is precisely how we hope to exploit the information in the painting. We prove later that
-when some join $v$ has all incident edges with exactly two colors, the corresponding maxima (of those colors) are exactly
+when some join $v$ has exactly two colors among all incident edges with $v$ as the lower endpoint, the corresponding maxima (of those colors) are exactly
 the children of $v$ in $\cJ(\MM)$. To proceed further, we ``merge" the colors red and blue into a new color, purple. In other words,
@@
 The merging of the colors is not explicitly performed as that would be too expensive; instead we maintain a union-find
-data structure for that.
+data structure for that. So the red and blue mounds are not actually recolored purple, 
+and instead say red now points to blue in the union find data structure.
 
@@
 to extra overhead. We show that it suffices to simply put each vertex $v$ into at most two heaps, one for each
-``side" of a possible join. We are guaranteed that when $v$ needs to be processed, all edges have
+``side" of a possible join. We are guaranteed that when $v$ needs to be processed, all edges incident from above have
 at most $2$ colors, because of all the color merges that previously occurred.
@@
 saddle on the far right (labeled $j$) which is not hit by blue paint.  Since this far right saddle is slightly 
-higher than the far left one, it will merge into the component containing the blue mound (and also the yellow and red mounds)
-before the far left one.  Hence, the vertices initially touched by blue are not contiguous in the join tree.
+higher than the far left one, the merge event involving the component containing the blue mound (and also the yellow and red mounds) and corresponding to the far right saddle
+will occur before the one corresponding to the far left saddle.  Hence, the vertices initially touched by blue are not contiguous in the join tree.
 
@@
 Consider $\wet(x,\NN_{k+1})$. The interface is exactly $\phi_1, \ldots, \phi_k$.
-So the only dry vertices are those in the boundaries $B_1, \ldots, B_k$. But these
-boundaries are maxima.
+So the only dry vertices of $\NN_{k+1}$ are those on the boundaries created by calling $\cut$ for each $\phi_i$, 
+but such boundaries are all maxima.
+%$B_1, \ldots, B_k$. But these boundaries are maxima.
 \end{proof}
@@
 \begin{lemma}\label{lem:new-verts}
-A face $F \in \MM$ is cut\footnote{Technically what we are calling a single cut is done with two hyperplanes.} at most once during $\rain(x,\MM)$.
+A face $F \in \MM$ is cut\footnote{What we are calling a single cut is actually done with two hyperplanes.} at most once during $\rain(x,\MM)$.
 \end{lemma}
@@
 merging procedure from~\cite{csa-cctad-03} that constructs the contour tree from the join and split trees. 
-It actually constructs the \emph{augmented contour tree} $\cA(\MM)$, which is 
+It actually first constructs the \emph{augmented contour tree} $\cA(\MM)$, which is 
 obtained by replacing each edge in the contour tree with a path of all regular vertices 
@@
 delete $v$ and connect its neighbors by an edge (i.e.\ smooth $v$ out). This tree is denoted by $T \ominus v$.
+%
+The merging procedure of \cite{csa-cctad-03} is shown below as $\merge(\cJ(\MM),\cS(\MM))$.  
+This procedure builds $\cA(\MM)$ by iteratively adding edges that are adjacent to leaves of either $\cJ(\MM)$ or $\cS(\MM)$, 
+erasing the corresponding vertex from both $\cJ(\MM)$ and $\cS(\MM)$, and repeating.  $\cC(\MM)$ is then produced by erasing all regular vertices in $\cA(\MM)$.
+We use the fact that this procedure correctly produces the contour tree (regardless of the order leaves are processed), 
+and refer the interested reader to \cite{csa-cctad-03} for the proof of this fact, as it is not needed for our purposes.
 
@@
 	\end{compactenum}
-	\item Smooth out all regular vertices in $\cA(\MM)$ to get $\cC(\MM)$.
+	\item Erase all regular vertices in $\cA(\MM)$ to get $\cC(\MM)$.
 \end{compactenum}
@@
 and $v$ is adjacent (in $\MM$) to a vertex in this component. \InSoCGVer{The \emph{join tree}, $\cJ(\MM)$, is defined analogously, but instead on $\cV(\MM)$.}
-%
-%\InSoCGVer{Have sentence defining the join tree here.}
-%
 \end{definition}
@@
 \newcommand{\proofofContourJoinEquiv}{
-\begin{proof} We first show that $\cA(\MM)$ is $\cJ(\MM) \ominus \{m_i\}$ with edges $\{(s_i,m_i)\}$.
+\begin{proof} %We first show that $\cA(\MM)$ is $\cJ(\MM) \ominus \{m_i\}$ with edges $\{(s_i,m_i)\}$.
 We have flexibility in choosing the order of processing in $\merge$. We first put the non-dominant
-maxima $m_1, \ldots, m_k$ into the queue. As these are processed, the edges $\{(s_i,m_i)\}$ are inserted 
+minima $m_1, \ldots, m_k$ into the queue. As these are processed, the edges $\{(s_i,m_i)\}$ are inserted 
 into $\cA(\MM)$. Once all the $m_i$'s are erased, $\cS$ becomes a path, so all outdegrees are at most $1$.
-The join tree is now $\cJ(\MM) \ominus \{m_i\}$. We can now process $\cJ$ leaf by leaf, and all
-edges of $\cJ$ are inserted into $\cA(\MM)$.
-
-Note that $\cC(\MM)$ is obtained by smoothing out all regular points from $\cA(\MM)$. Similarly, smoothing out regular
-points from $\cJ(\MM) \ominus \{m_i\}$ yields the edges of $\jc(\MM)$.
+Moreover the join tree is now $\cJ(\MM) \ominus \{m_i\}$, and so we can now process $\cJ$ leaf by leaf, and all
+remaining edges of $\cJ$ are inserted into $\cA(\MM)$.
+%
+Note that smoothing out regular points from $\cJ(\MM) \ominus \{m_i\}$ yields the edges of $\jc(\MM)$.
+Thus after smoothing out all regular points from $\cA(\MM)$ we recover the edges of $\jc(\MM)$.  
+Hence the resulting contour tree contains these edges and the $\{(s_i,m_i)\}$ as claimed.
+%Note that $\cC(\MM)$ is obtained by smoothing out all regular points from $\cA(\MM)$. Similarly, smoothing out regular
+%points from $\cJ(\MM) \ominus \{m_i\}$ yields the edges of $\jc(\MM)$.
 \end{proof}
@@
 it suffices to have an efficient algorithm for computing $\jc(\MM)$.
-Due to minor technicalities, it is easier to phrase the following section instead in terms of computing $\cJ(\MM)$ efficiently.  
-Note however that for minimum dominant complexes output by $\rain$, converting between $\jc$ and $\cJ$ is trivial, as $\cJ$ is just $\jc$ 
+In the other sections of the paper it is better to phrase things instead in terms of computing $\cJ(\MM)$ efficiently.  
+%Note however that for minimum dominant complexes output by $\rain$, converting between $\jc$ and $\cJ$ is trivial, as $\cJ$ is just $\jc$ 
+%with each non-dominant minimum $m_i$ augmented along the edge leaving $s_i$.
+Note however that for minimum dominant complexes, converting between $\jc$ and $\cJ$ is trivial, as $\cJ$ is just $\jc$ 
 with each non-dominant minimum $m_i$ augmented along the edge leaving $s_i$.  
+(Indeed $\jc(\MM)$ was only defined to make the above theorem statement clearer.)
 \end{Remark}
@@
 \noindent
-{\bf Color list $\col(v)$:} For each point $v$, we maintain $\col(v)$ as a simple list.
+{\bf Color list $\col(v)$:} For each vertex $v$, we maintain $\col(v)$ as a simple list.
 In addition, we will maintain another (sub)list of colors $L$ such that $\forall c \in L$,
@@
 We formally describe the algorithm below. 
-We require a technical definition of \emph{ripe} vertices.
+We require the following definition of \emph{ripe} vertices.
 
@@
 	\item Merge heaps $\{T(c') | c' \in \cur(h)\}$.
-	\item Take union of $\cur(h)$ and denote resulting color by $\widehat{c}$.
+	\item Take union of $\cur(h)$ and denote resulting representative color by $\widehat{c}$.
 	\item Set $\h(\widehat{c}) = h$ and mark $h$ as processed.
@@
 	\item Each vertex is present in at most $2$ heaps. After processing, it is removed from all heaps.
-	\item After $v$ is processed, all edges incident to $v$ have the same (representative) color. %(more technically, the same representative). 
+	\item After $v$ is processed, all edges incident to $v$ from above have the same (representative) color. %(more technically, the same representative). 
 %	\item When $\update(\stack)$ terminates, the head $h$ is ripe. 
@@
 
-Our main workhorse is the following technical lemma. In the following, the current color of an edge, $e$, is the value of $\rep(\chi(e))$, 
+Our main workhorse is the following lemma. The \emph{current color} of an edge, $e$, refers to the value of $\rep(\chi(e))$, 
 where $\chi(e)$ is the color of $e$ from the initial painting.
@@
 by an edge $e$ which is currently colored $c$. At all times: either all edges in $\PP$ are currently colored $c$, or there
-exists a critical vertex $w \in \PP$ fully touched by $c$ and touched by another color.
+exists a critical vertex $w \in \PP$ currently fully touched by $c$ and currently touched by another color.
 \end{lemma}
 
-\begin{proof} Since $e$ has color $c$,
-there must exist vertices in $\PP$ touched by $c$. Consider the highest
-vertex $w$ in $\PP$ that is touched by $c$ and some other color. If no such vertex exists,
-this means all edges incident to a vertex touched by $c$ are colored $c$. By walking through
-$\PP$, we deduce that all edges are colored $c$.
+\begin{proof} Since $e$ has current color $c$,
+there must exist vertices in $\PP$ currently touched by $c$. Consider the highest
+vertex $w$ in $\PP$ that is currently touched by $c$ and some other color. If no such vertex exists,
+this means all edges incident to a vertex currently touched by $c$ are currently colored $c$. By walking through
+$\PP$, we deduce that all edges are currently colored $c$.
 
 So assume $w$ exists. Take the $(f(w)+\delta)$-contour $\phi$ that intersects $B_\eps(w)$
-and intersects some $c$-colored edge incident to $w$. Note that all edges intersecting $\phi$ are also colored $c$,
-since $w$ is the highest vertex to be touched by $c$ and some other color. (Take the path of $c$-colored
+and intersects some currently $c$-colored edge incident to $w$. Note that all edges intersecting $\phi$ are also currently colored $c$,
+since $w$ is the highest vertex to be currently touched by $c$ and some other color. (Take the path of currently $c$-colored
 edges from the maximum to $w$. For any point on this path, the contour passing through this point must
-be colored $c$.) Hence, $c$ fully touches $w$. 
-But $w$ is touched by another color, and the corresponding edge cannot intersect $\phi$. So $w$
+be currently colored $c$.) Hence, $c$ currently fully touches $w$. 
+But $w$ is currently touched by another color, and the corresponding edge cannot intersect $\phi$. So $w$
 must have up-degree $2$ and is critical.
@@
 
- Let $\PP$ be one of the components of $\MM^+_h$.  By construction, $h$ was put in the heap of some initial 
+ Let $\PP$ be one of the components of $\MM^+_h$.  By construction, $h$ was put in the heap of some initially  
  adjacent color $c$.  Therefore, $h$ must be in the current heap of $\rep(c)$ (see \Obs{twoQueues}).  
  Now by \Lem{full}, either all edges in $\PP$ are colored $\rep(c)$ or  
- there is some vertex $w$ fully touched by $\rep(c)$ and some other color.  
+ there is some vertex $w$ fully touched by $\rep(c)$ and touched by some other color.  
  The former case implies that if there are any unprocessed vertices in $\PP$ then they are all in $T(\rep(c))$, 
@@
 \begin{claim} \label{clm:upstar} Consider a ripe vertex $v$ and take the up-star connecting
-to some component of $\MM^+_v$. All edges in this component and the up-star have the same color.
+to some component of $\MM^+_v$. All edges in this component and the up-star have the same current color.
 \end{claim}
@@
 By \Lem{full}, either all edges in $\PP$ are colored $\rep(c)$ or there exists critical vertex $w \in \PP$
-fully touched by $\rep(c)$ and another color. In the latter case, $w$ has not been processed,
+fully touched by $\rep(c)$ and touched by another color. In the latter case, $w$ has not been processed,
 so $w \in T(\rep(c))$ (contradiction to ripeness). Therefore, all edges in $\PP$ are colored $\rep(c)$.
@@
 \begin{proof} More generally, we prove the following: all outputs on processed vertices
-are edges of $\cJ(\MM)$ and for any current color $c$, $\h(c)$ is the lowest processed vertex
-of that color. We prove this by induction on the processing order. The base case is trivially
+are edges of $\cJ(\MM)$ and for any representative color $c$, $\h(c)$ is the lowest processed vertex
+of that representative color. We prove this by induction on the processing order. The base case is trivially
 true, as initially the processed vertices and attachments of the color classes are the set of maxima.
@@
 of $\MM^+_v$ that it connects to. By \Clm{upstar}, all edges in $\PP$ and the up-star
-have the same color (say $c$). If some critical vertex in $\PP$ is not processed,
+have the same current color, say $c$. If some critical vertex in $\PP$ is not processed,
 it must be in $T(c)$, which violates the ripeness of $v$.
 Thus, all critical vertices in $\PP$ have been processed, and so by the induction hypothesis, the restriction of $\cJ(\MM)$ to $\PP$ has been correctly computed.
-Additionally, since all critical vertices in $\PP$ have processed, they all have the same color $c$ of the lowest critical vertex in $\PP$.
+Additionally, since all critical vertices in $\PP$ have been processed, they all have the same representative color $c$ of the lowest critical vertex in $\PP$.
 Thus by the strengthened induction hypothesis, this lowest critical vertex is $\h(c)$.
@@
 If there is another component of $\MM^+_v$, the same argument implies 
-the lowest critical vertex in this component is $\h(c')$ (where $c'$ is the color of edges in the respective component).
+the lowest critical vertex in this component is $\h(c')$ (where $c'$ is the current color of edges in the respective component).
 Now by the definition of $\cJ(\MM)$, the critical vertex $v$ connects to the lowest critical vertex in each component of $\MM^+_v$, 
 and so by the above $v$ should connect to $\h(c)$ and $\h(c')$, which is precisely what $v$ is connected to by $\build(\MM)$.
-Moreover, $\build$ merges the colors $c$ and $c'$ and correctly sets $v$ to be the attachment, 
+Moreover, $\build$ merges the representative colors $c$ and $c'$ and correctly sets $v$ to be the attachment, 
 as $v$ is the lowest processed vertex of this merged color (as by induction $\h(c)$ and $\h(c')$ were the lowest vertices before merging colors).
@@
 
-Any tree $T$ considered in following will be a rooted binary tree\footnote{Note 
+Any tree $T$ considered in the following will be a rooted binary tree\footnote{Note 
 that technically the trees considered should have a leaf vertex hanging below the root  
@@
  one from the left child and one from the right child subtree of $v$ (naturally if $v$ has only one child it is assigned 
- only one color).
+ only one leaf).
 \end{definition}
@@
 \label{def:initialColoring}
-Note that the subroutine $\init(\MM)$ from \Sec{algo} naturally defines a leaf assignment to $\cJ(\MM)$ 
+The subroutine $\init(\MM)$ from \Sec{algo} naturally defines a leaf assignment to $\cJ(\MM)$ 
 according to the priority queue for each up-star we put a given vertex in.  Call this the \emph{initial coloring}
-of the vertices in $\cJ(\MM)$.  Note also that this initial coloring defines the $H_v$ values for all $v\in \cJ(\MM)$.
+of the vertices in $\cJ(\MM)$.  This initial coloring also defines the $H_v$ values for all $v\in \cJ(\MM)$.
 \end{definition}
 
-The following lemma should justify these technical definitions.
+The following lemma should justify these definitions.
 
@@
 let $H_v$ be defined by the initial coloring of $\MM$.
-The running time of $\build(\MM)$ is $O(N+t\alpha(t) + \sum_{v \in \cJ(\MM)} \log |H_v|)$.
+The total running time of $\build(\MM)$, including the subroutine call to $\init(\MM)$ and all calls to $\update$, is $O(N+t\alpha(t) + \sum_{v \in \cJ(\MM)} \log |H_v|)$.
 \end{lemma}
@@
 Indeed, the painting procedure consists of several BFS's but as each vertex is only explored by one of the BFS's, it is linear time overall.
-Determining the critical points is a local computation on the neighborhood of each vertex as so is linear (i.e.\ each edge is viewed at most twice).
+Determining the critical points is a local computation on the neighborhood of each vertex and so is linear (i.e.\ each edge is viewed at most twice).
 Finally, each vertex is inserted into at most two heaps and so initializing the heaps takes linear time in the number of vertices.
@@
 Now when $v$ is handled, three queue operations are performed.  Specifically, $v$ is removed from the queues of $c_1$ and $c_2$, and then the queues are 
-are merged together.  By the above arguments the sizes of the queues for each of these operations are $|H_{c_1}|$, $|H_{c_2}|$, and $|H_v|$, respectively.  
+merged together.  By the above arguments the sizes of the queues for each of these operations are $|H_{c_1}|$, $|H_{c_2}|$, and $|H_v|$, respectively.  
 As merging and deleting takes logarithmic time in the heap size for binomial heaps, the claim now follows.
@@
 This slack allows for the more refined upper and lower bounds mentioned in \Sec{more-refined}.  Quantifying this slack however is quite challenging, 
-and requires a significantly more sophisticated analysis involving path decompositions, detailed in the full version \CiteFullVer.
+and requires a significantly more sophisticated analysis involving path decompositions, which is provided in the next section.
 
@@
 
-Our main goal in this section is to prove the following theorem.  We use $|p|$ to denote the number of vertices in $p$.
+Our main goal in this section is to prove the following theorem.  We use $|p|$ to denote the number of vertices in a path $p$.
 
@@
 
-The paths in $\pmax(T)$ naturally define a tree\footnote{Please excuse the 
-overloading of the term 'tree', it is the most natural term to use here.} of their own.  Specifically, in the original 
+The paths in $\pmax(T)$ naturally define a tree\footnote{
+%Please excuse the overloading of the term 'tree', it is the most natural term to use here.
+The term 'tree' is the most natural term to use here, but the reader should be careful not to confuse this tree on paths from $T$ with the tree $T$ itself.
+} of their own.  Specifically, in the original 
 tree $T$ contract each path down to its root.  Call the resulting tree the \emph{shrub} of $T$ corresponding to 
@@
 We use $H_p$ to denote $H_{b(p)}$.
-% and denote it by $\pathTree(T)$.
-
-% \begin{lemma}
-% \label{lem:adjacent}
-%  Let $p$ be any path in $\pmax(T)$ and let $\{q_1, \dots q_k\}$ be the children of $p$.
-% Then $|H_{\ell(p)}| + \sum_{i=1}^k |H_{q_i}| \leq |H_p|+2|p|$. 
-% \end{lemma}
-% 
-% \begin{proof} For convenience, denote $H_i = H_{q_i}$ and $H_0 = H_{\ell(p)}$.
-% Consider $v \in \bigcup_{i} H_i$ that lies below $b(p)$ in $T$. 
-% Note that such a vertex has only one of its two colors in $L(b(p))$.  
-% Since the colors tracked by $H_i$ and $H_j$ for $i\neq j$ are disjoint, 
-% such a vertex can appear in only one of the $H_i$'s.
-% On the other hand, a vertex $u\in p$ can appear in more than one $H_i$, 
-% but since any vertex has exactly two colors it can appear in at most two 
-% such heaps.  Hence, $\sum_i |H_i| \leq |H_p| + 2|p|$.
-% \ignore{
-%  First observe that for all $i$, $r_{i}$ lies above $r_p$ on some root to leaf path. 
-%  Let $H_i'$ be the subset of $H_i$ that lies below $r_p$.  Observe that $H_i'\subseteq H_p$, as $L(T_{r_i})\subseteq L(T_{r_p})$.
-%   
-%  Now for any $i\neq j$, $H_i \cap H_j = \emptyset$, as $r_{i}$ and $r_{j}$ are not on the same root to leaf path (i.e.\ $L(T_{r_i}) \cap L(T_{r_j}) = \emptyset$).
-%  In particular, $(H_i\setminus H_i') \cap (H_j \setminus H_j') = \emptyset$, and so $\sum_{i=0}^k |H_i\setminus H_i'| \leq |p|$ 
-%  as $H_i\setminus H_i'$ is precisely the subset of $H_i$ that lies on $p$.
-%  Moreover, for any $i\neq j$, $ H_i' \cap H_j' = \emptyset$, and so since $H_i'\subseteq H_p$ we have $\sum_{i=0}^k |H_i'|\leq |H_p|$.
-%  Putting these together gives
-%  \[
-%   \sum_{i=0}^k |H_i| = \sum_{i=0}^k |H_i'| + \sum_{i=0}^k |H_i\setminus H_i'| \leq |H_p|+|p|.
-%  \]
-% }
-% \end{proof}
 
@@
 \end{definition}
-
-\ignore{
-\begin{observation}
-\label{obs:decrease}
- Let $v$ be a vertex in $T$ and let $w$ be its parent.  Then $|H_w| \geq |H_v| -1$, as $L(T_v) \subseteq L(T_w)$ and the path from 
- $w$ to the root has one less vertex than the path from $v$ to the root.
- In particular, we have the following more general property.
- Let $v$ and $u$ be any two vertices in the same root to leaf path of $T$, such that $v$ is a descendant of $u$.  
- Then $|H_v| \leq |H_u| + d(u,v)$. 
-\end{observation}
-
-The following can be thought of as a generalization of the above observation, and will be useful in later sections.
-}
 
@@
 \end{lemma}
-\begin{proof} \ignore{
- By $\Obs{decrease}$ we know that for any vertex $v\in p$, $|H_v|\leq |H_{p}| + |p|$ (as $v$ is a descendant of $r_p$ along $p$).
-}
+\begin{proof}
 For $v \in p$, $|H_v|\leq |H_{p}| + |p|$ (as $v$ is a descendant of $b(p)$ along $p$).
@@
 
-\ignore{
-For a short path $p$ we can think of the quantity $|p|(\log |H_p| - \log |p|)$ as the excess of $p$, i.e.\ what was not ``paid for'' locally by $p$.
-Our eventual goal will be to show that by choosing the right path decomposition, this excess 
-can be recharged to tall paths. In the next couple sections we introduce the appropriate path decomposition and prove 
-some useful facts about it.  Then given these facts, in \Sec{} we are able to make the recharging argument.
-}
-
 There are some short paths that we can also ``pay" for.  Consider any short path $p$ in the shrub.  We will refer to the 
 \emph{tall support chain} of $p$ as the tall ancestors of $p$ in the shrub which have a path to $p$ which does not use any 
-short path (i.e.\ it is a chain of paths adjacent to $p$).
+short path (i.e.\ it is a chain of tall paths adjacent to $p$).
 
@@
 at least $|H_p|/100$ vertices $v$ in $H_p$ lie in paths in the tall support chain of $p$.
-%
-%
-% there exists
-%a contiguous (from $p$) chain of ancestors $q_1, q_2, \ldots, q_k$ (in $P(T)$) such that $|H_{q_k}| \leq 99|H_p|/100$
-%and all $q_i$'s are tall.
 \end{definition}
@@
 Let $\cL$ be the set of short paths, $\cL'$ be the set of supported short paths, and $\cH$ be the set of tall paths
-given by $\pmax(T)$ (notation which we be used for the remainder of this section).
+given by $\pmax(T)$ (notation which will be used for the remainder of this section).
 
@@
  \end{align*}
-%
-% By \Lem{pathBounds}, $\sum_{p \in \cL'} \sum_{v \in p} \log |H_v|$
 \end{proof}
@@
 We process the nodes in order. For node $p$, let $\cU'_p$ denote the rooted
-subtree of $p$ in $\cU'$. We simply check if $\sum_{q \in \cU'_p} |\loss(p)| \geq \sum_{q \in \cU'_p} |q|/100$. If so, we prune off $\cU'_p$, unless $p$ is the root. 
+subtree of $p$ in $\cU'$. We simply check if $\sum_{q \in \cU'_p} |\loss(q)| \geq \sum_{q \in \cU'_p} |q|/100$. If so, we prune off $\cU'_p$, unless $p$ is the root. 
 (Note that $\cU'$ is redefined
@@
 \begin{lemma} \label{lem:buff}
- Let $\buff$ be a buffered sub-shrub.  Then $\sum_{p\in \buff}\sum_{v\in p} \log |H_v| = O(\sum_{p\in \buff}\sum_{v\in \loss(p)} \log |H_v| )$
+ Let $\buff$ be a buffered sub-shrub.  Then $\sum_{p\in \buff}\sum_{w\in p} \log |H_w| = O(\sum_{p\in \buff}\sum_{v\in \loss(p)} \log |H_v| )$
 \end{lemma}
@@
 
-Since $T_r$ is not used up, it must mean that all $p$s are full. Consider any $p$,
-and let $K_p$ be the set of tokens placed at $p$. Each token is labeled $\log |H_v|$,
-for some $v \in \loss(\hat{p})$, for some ancestor $\hat{p}$ of $p$. By \Cor{geometric},
+Since $T_r$ is not used up, it must mean that all $p$'s are full, as $r$ is the root and so $T_r$ tokens can be placed on any path which is not full. 
+Consider any $p$, and let $K_p$ be the set of tokens placed at $p$. Each token is labeled $\log |H_v|$,
+for some $v \in \loss(\hat{p})$, where $\hat{p}$ is an ancestor of $p$ or $p$ itself (as descendants do not place tokens on ancestors). By \Cor{geometric},
 for every $w \in p$, $\log |H_w| = O(\log |H_v|)$. Because $p$ is full,
 there are exactly $|p|$ tokens on $p$. Thus, $\sum_{w \in p} \log |H_w|$
-is $O(V_p)$, where $V_p$ is the sum of token values in $K_p$. We sum
-over all $p$ to complete the proof.
+is $O(V_p)$, where $V_p$ is the sum of token values in $K_p$. 
+Thus summing over all $p$ in $\buff$ completes the proof. 
+Specifically, $\sum_{p\in \buff} \sum_{w \in p} \log |H_w| = O(\sum_{p\in \buff} V_p) = O(\sum_{p\in \buff}\sum_{v\in \loss(p)} \log |H_v| )$
 
@@
  Let $B= \{\buff_1, \ldots, \buff_k\}$ be any collection of disjoint buffered sub-shrubs from $\cF$. Then, 
- $\sum_{i=1}^k \sum_{p\in \buff_i}\sum_{v\in p} |H_v| = O(\sum_{p\in \cH}\sum_{v\in p} \log |H_v| )$.
+ $\sum_{i=1}^k \sum_{p\in \buff_i}\sum_{v\in p} \log |H_v| = O(\sum_{p\in \cH}\sum_{v\in p} \log |H_v| )$.
 \end{corollary}
@@
 Note that $|H_p| \geq A_p - 2\sum_{q \in \pruned_p, q \neq p} |\lambda(q)|$ $-2\sum_{q \in \pruned_p} |q|$.
-Crucially, since $\pruned$ is buffer-pruned, $\sum_{q \in \pruned_p, q \neq p} |\lambda(p)|
+Crucially, since $\pruned$ is buffer-pruned, $\sum_{q \in \pruned_p, q \neq p} |\lambda(q)|
 \leq \sum_{q \in \pruned_p, q \neq p} |q|/100$ $\leq \sum_{q \in \pruned_p} |q|/100$. This completes the proof.
@@
 $$ ({\alpha_q}/99)\sum_{a_i} 1/\sqrt{A_{a_i}} \leq ({\alpha_q}/99)\sum_{a_i} (4/5)^{i/2} \leq \alpha_q/20 $$
-The total charge is $\sum_{q \in \pruned_p} \alpha_p/20 = A_p/20$. 
+The total charge is $\sum_{q \in \pruned_p} \alpha_qt/20 = A_p/20$. 
 \end{proof}
@@
 \begin{claim} \label{clm:exp-A} Fix any path $p \in \pruned \setminus r$.
-For any $q, q' \in \pruned_p$ where $q$
-is a parent of $q'$, $A_q \geq (5/4)A_{q'}$.
+%For any $q, q' \in \pruned_p$ where $q$ is a parent of $q'$, $A_q \geq (5/4)A_{q'}$.
+Let $q$ be any path in $\pruned_p$, then for any child path $q'$ of $q$ it holds that $A_q \geq (5/4)A_{q'}$
 \end{claim}
@@
 At this point we can now see what the path decomposition referenced in theorem statement should be.  It is just the union of all the maximum 
-path decomposition across the extremum dominant pieces, $\pmax(\cC(\MM)) = \cup_{i=1}^k \pmax(\cJ(\MM_i))$.
+path decompositions across the extremum dominant pieces, $\pmax(\cC(\MM)) = \cup_{i=1}^k \pmax(\cJ(\MM_i))$.
 Since all procedures besides computing the join trees take linear time in the size of the input complex, we can therefore 
@@
 Fix any path decomposition $P$. There is a family of terrains, $\bF_P$, all with the same triangulation, 
-such that $|\bF_P| = \Pi_{p_i\in P} (|p_i|-1)!$, and no two terrains in $\bF_P$ define the same join tree.
+such that $|\bF_P| = \Pi_{p_i\in P} (|p_i|-1)!$, and no two terrains in $\bF_P$ define the same labeled join tree.
 \end{lemma}
@@
 \begin{proof}
-% We describe a simple triangulation, with heights specified for all vertices except for the saddles.
-
-%The main idea is that 
-% this triangulation will be flexible enough to allow almost any assignment of heights to the saddles, thus defining a class of join trees 
-% whose size is that described in the lemma statement. 
  We describe the basic building block of these terrains, which corresponds to a fixed path $p\in P$.  
@@
  Next, create an apex vertex at some appropriately large height, and add an edge to each vertex in the cycle. 
+ Note that in order to obtain a terrain, a cycle here means the vertices lie on a circle 
+ but are connected with straight line segments rather than by the curved edges of the circle.
 
@@
  extended down (at a slight angle) until it hits the face of the parent.
+ Note that this gluing process introduces some non-triangular faces, and so these must be triangulated in order to maintain a proper triangulated terrain.
 
@@
 
-Two path decompositions $P_1$ and $P_2$ (of potentially different complexes and/or height functions) are equivalent 
-if: there is a 1-1 correspondence between the sizes of the constituent paths, and the shrubs are isomorphic.
-%
-%
-%if their respective shrubs, $S_1$ and $S_2$, which has the property that for any node $v\in S_1$ corresponding to a path of 
-%size $|v|$, $f(v)$ also corresponds to a path of size $|v|$. Note that for equivalent decompositions, $\sum_{p\in P_1} |p|\log|p| = \sum_{p\in P_2} |p|\log|p|$.
+% Two path decompositions $P_1$ and $P_2$ (of potentially different complexes and/or height functions) are equivalent 
+% if: there is a 1-1 correspondence between the sizes of the constituent paths, and the shrubs are isomorphic.
 

### 2016-12-21 paint-contour-tree.tex
more changes for dcg

@@
 
-%\def\ProofInAppendix{TRUE}
-
-\ifx\SoCGVer\undefined
-   \newcommand{\InSoCGVer}[1]{}%
-   \newcommand{\InNotSoCGVer}[1]{#1}%
-\else
-%   \def\ProofInAppendix{TRUE}
-   \newcommand{\InSoCGVer}[1]{#1}%
-   \newcommand{\InNotSoCGVer}[1]{}%
-\fi
-
-\newcommand{\CiteFullVer}{\cite{rs-mvrms-14}\xspace}%
-
-% \InNotSoCGVer{%
-%    \newcommand{\SeeFullVer}{}%
-%    \newcommand{\LemmaProofInFullVer}[1]%
-%    {%
-%       \begin{lemma}
-%           #1
-%       \end{lemma}%
-%    }
-% }
-% 
-% \InSoCGVer{%
-%    \newcommand{\SeeFullVer}{{{\CiteFullVer}}}%
-%    \newcommand{\LemmaProofInFullVer}[1]%
-%    {%
-%       \begin{lemma}[For proof see  \CiteFullVer{}]
-%           #1
-%       \end{lemma}
-%    }   
-% }
-
-\InSoCGVer{%
-   \documentclass[a4paper,UKenglish]{lipics-v2016}
-%   \documentclass[11pt]{article}%
-
-%  \usepackage{proofapnd}
-}
-\InNotSoCGVer{%
+%\newcommand{\CiteFullVer}{\cite{rs-mvrms-14}\xspace}%
+
   \documentclass[11pt]{article}
   \usepackage[margin=2.54cm]{geometry}
-}
-
-\InNotSoCGVer{
+
    \newcommand{\myparagraph}[1]{\paragraph{#1}}
-}
-\InSoCGVer{
-   \newcommand{\myparagraph}[1]{\smallskip\noindent{\textbf{#1}}}
-}
 
@@
 
-\InNotSoCGVer{
 \newtheorem{theorem}{Theorem}[section]
-}
 \newtheorem{lemmata}[theorem]{Lemmata}
@@
 \newtheorem{proposition}[theorem]{Proposition}
-\InNotSoCGVer{
  \newtheorem{lemma}[theorem]{Lemma}
@@
  \newtheorem{definition}[theorem]{Definition}
-}
 \newtheorem{fact}[theorem]{Fact}
@@
 \theoremstyle{definition}
-% \InNotSoCGVer{ 
-%  \newtheorem{remark}[theorem]{Remark} 
-% }
 \newtheorem{Remark}[theorem]{Remark}
@@
 
-\InNotSoCGVer{
 \author{
@@
 \date{}
-}
-
-\InSoCGVer{
-% Author macros::begin %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\title{%A Mountaintop View Requires Minimal Sorting:
-Avoiding the Global Sort: 
-\break A Faster Contour Tree Algorithm%
-\footnote{The full updated version of this paper is available on the arXiv \cite{rs-mvrms-14}}%
-}
-\titlerunning{Avoiding the Global Sort: A Faster Contour Tree Algorithm} %optional, in case that the title is too long; the running title should fit into the top page column
-
-\author[1]{Benjamin Raichel}
-\author[2]{C. Seshadhri}
-  \affil[1]{%
-      Department of Computer Science, %
-      University of Texas at Dallas\\ %
-      Richardson, TX, 75080, USA\\ %
-      \texttt{\si{benjamin.raichel}@\si{utdallas}.\si{edu}} %
-   }%
-   \affil[2]{%
-      Department of Computer Science, %
-      University of California, Santa Cruz\\ %
-      Santa Cruz, CA, 95064, USA\\ %
-      \texttt{\si{scomandu}@\si{ucsc}.\si{edu}} %
-   }%
-
-\authorrunning{B. Raichel and C. Seshadhri} %mandatory. First: Use abbreviated first/middle names. Second (only in severe cases): Use first author plus 'et. al.'
-
-\Copyright{Benjamin Raichel and C. Seshadhri}%mandatory, please use full first names. LIPIcs license is "CC-BY";  http://creativecommons.org/licenses/by/3.0/
-
-\subjclass{F.2.2, I.1.2, I.3.5}% mandatory: Please choose ACM 1998 classifications from http://www.acm.org/about/class/ccs98-html . E.g., cite as "F.1.1 Models of Computation". 
-\keywords{contour trees, computational topology, computational geometry}% mandatory: Please provide 1-5 keywords
-% Author macros::end %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\EventEditors{S\'andor Fekete and Anna Lubiw}
-\EventNoEds{2}
-\EventLongTitle{32nd International Symposium on Computational Geometry
-(SoCG 2016)}
-\EventShortTitle{SoCG 2016}
-\EventAcronym{SoCG}
-\EventYear{2016}
-\EventDate{June 14-18, 2016}
-\EventLocation{Boston, USA}
-\EventLogo{}
-\SeriesVolume{51}
-\ArticleNo{57}
-
-}
 
 \maketitle
-\InNotSoCGVer{
 \thispagestyle{empty}
-}
 
@@
 
-\InNotSoCGVer{
 \newpage
 \pagenumbering{arabic}
-}
 
@@
 where the algorithm makes $\alpha \thmConst$ comparisons, for any constant $\alpha$. Then \emph{any algorithm} must make roughly $\thmConst$
-comparisons in the worst-case over this set. (Further details are in \InSoCGVer{the full version \CiteFullVer}\InNotSoCGVer{\Sec{lb}}.)
+comparisons in the worst-case over this set. (Further details are in \Sec{lb}.)
 
@@
 
-\InSoCGVer{See the full version \CiteFullVer for further technical remarks on the above definitions.}
-\newcommand{\technicalRemarks}{
-%\subsection{Some technical remarks}
+\subsection{Some remarks}
 Note that if one intersects $\MM$ with a given ball $B$, then a single contour in $\MM$ might be split into more than one contour in the intersection. 
@@
 set of vertices, though we will not delve into these aspects in this paper.
-}
-\InNotSoCGVer{
-\subsection{Some remarks}
-\technicalRemarks
-}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\newcommand{\newintuitionSection}{
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+%\pagebreak
 \section{A tour of the new contour tree algorithm} \label{sec:approach}
-We provide a short, high-level description of the main ideas of subsequent sections. 
-A longer more detailed version of this high-level description can be found in the full version \CiteFullVer.
-% However, for the $d=2$ case, 
-% where the input is just a triangulated terrain, the main ideas of the parts of the algorithm 
-% can be explained clearly.
+We provide a high-level description of the main ideas of subsequent sections.
+
+\myparagraph{Do not globally sort:} The starting point for this work is \Fig{relative}. We have two terrains
+with exactly the same contour tree, but different orderings of (heights of) the critical points.
+Turning it around, we cannot deduce the full height ordering of critical points from the contour tree, and so 
+sorting all critical points is computationally unnecessary.
+In \Fig{sorting}, the contour tree consists of two balanced binary trees, one of the joins,
+another of the splits. Again, it is not necessary to know the relative ordering between the mounds
+on the left (or among the depressions on the right) to compute the contour tree. Yet some ordering
+information is necessary: on the left, the little valleys are higher than the big central valley,
+and this is reflected in the contour tree. Leaf paths in the contour tree have points
+in sorted order, but incomparable points in the tree are unconstrained.
+How do we sort exactly what is required, without knowing the contour tree in advance?
+
+\myparagraph{Join and split trees:}
+The standard contour tree algorithm of Carr, Snoeyink, and Axen~\cite{csa-cctad-03}, builds
+two different trees, called the join and split trees, and then merges them together to get the contour tree.
+Consider sweeping down the hyperplane $x_{d+1} = h$ and taking the \emph{superlevel} sets. These 
+are the connected components of the portion of $\MM$ above height $h$. For a terrain, the superlevel
+sets are a collection of ``mounds". As we sweep downwards, these mounds keep joining each other,
+until finally, we end up with all of $\MM$. The join tree, $\cJ(\MM)$, tracks exactly these events. 
+Formally, let $\MM^+_v$ denote the simplicial complex induced on the subset of 
+vertices higher than $v$. 
+Refer to \Fig{sorting} for the join tree of a terrain. Note that nothing happens at splits, but these
+are still put as vertices in the join tree. The split tree
+is obtained by simply inverting this procedure, sweeping upwards and tracking sublevel sets.
+
+A major insight of~\cite{csa-cctad-03} is an ingeniously simple linear time procedure to construct
+the contour tree from the join and split trees. So the bottleneck is computing these trees. Observe
+in \Fig{sorting} that the split vertices form a long path in the join tree (and vice versa). Therefore, constructing
+these trees forces a global sort of the splits, an unnecessary computation for the contour tree. Unfortunately, in general (i.e.\ unlike \Fig{sorting}) 
+the heights of joins and splits may be interleaved in a complex manner, and hence the final merging of~\cite{csa-cctad-03}
+to get the contour tree requires having the split vertices in the join tree. Without
+this, it is not clear how to get a consistent view of both joins and splits, required for the contour tree.
+
+Our aim is to break $\MM$ into smaller pieces, where this unnecessary computation can be avoided.
 
@@
 the wet part, and recurse on the dry parts. This is carefully done to ensure that the total complexity
-of the pieces is linear.
+of the pieces is linear (see \Sec{rain} for details).
 % 
@@
 % upwards and downwards.  We finally get a collection of extremum dominant pieces.
-We prove a simple contour surgery theorem that builds the contour tree of $\MM$
-from the contour trees of the various pieces created.
-
-Using ideas from~\cite{csa-cctad-03}, we prove that the contour tree of an extremum dominant complex is (basically) just the \emph{join tree}, 
-which unlike the contour tree tracks superlevel sets rather than level sets.
-% For the $d=2$ case, it helps to visualize such an input as a terrain with no minima,
-% except at a unique boundary face (think of a large boundary triangle that is the boundary).
-Consider sweeping down the hyperplane $x_{d+1} = h$ and taking the connected components of the portion of $\MM$ above height $h$. For a terrain, these superlevel
-sets are a collection of ``mounds". As we sweep downwards, these mounds keep joining each other,
-until finally, we end up with all of $\MM$. The join tree tracks exactly these events.
+% We prove a simple contour surgery theorem that builds the contour tree of $\MM$
+% from the contour trees of the various pieces created.
+
+We now have carved $\MM$ up into a collection of extremum dominant pieces. 
+In \Sec{surgery} we prove a simple contour surgery theorem allowing us to build the contour tree of $\MM$
+from the contour trees of these various pieces.  Specifically, we argue our cutting in $\MM$ was done in such a way that 
+each cut corresponds to cutting only a single edge in the contour tree.  Informally this means gluing $\MM$ back together along these 
+cuts corresponds to gluing the contour tree together.
+
+Ultimately, the reason for cutting $\MM$ into extremum dominant pieces is that such pieces have simpler contour tree structure. 
+Specifically, using ideas from~\cite{csa-cctad-03}, in \Sec{extreme} we prove that the contour tree of an extremum dominant complex is (basically) just the join tree. 
+Thus there is no longer a disconnect between the amount of sorting required for the contour tree versus the join tree, 
+and so the problem of computing contour trees efficiently reduces to that for join trees.
 
@@
 The key idea is \emph{paint spilling}. Start with each maximum having a large can of paint, with distinct
-colors for each maximum. In arbitrary order, we spill paint from each maximum, wait till it flows down,
-then spill from the next, etc. Paint is viscous, and only flows down edges. Furthermore, our paints do not mix, so each edge receives a unique color, decided by the first
-paint to reach it.
-
-Our algorithm incrementally builds the join tree from the leaves (maxima) to the root. 
-Refer to the left part of \Fig{colors}.
+colors for each maximum. In arbitrary order, spill paint from each maximum, wait till it flows down,
+then spill from the next, etc. Paint is viscous, and only flows down edges. \emph{It does not
+paint the interior of faces with dimension strictly larger than 1.} 
+Furthermore, paints do not mix, so each edge receives a unique color, decided by the first paint to reach it.  
+%
+Note that the raining procedure, used above to carve the input into extremum dominant pieces, 
+cannot be used here as it wets interiors of higher dimensional faces and so raining from each maximum may significantly increase the input complexity (see \Fig{order}).
+On the other hand, the interface between two different colors is \emph{not} a contour, and so the join trees of each color class cannot just be simply glued together.
+
+% In the following, $[n]$ denotes $\{1, \ldots, n\}$, for any natural number $n$.
+% \begin{definition} \label{def:paint1} \InSoCGVer{(Restatement of \Def{paint2}.)} Let the 1-skeleton of $\NN$ have edge set $E$ and maxima $X$.
+% A  \emph{painting} of $\NN$ is a map $\chi:X \cup E \to [|X|]$ with the following property. 
+%  Consider an edge $e$. There exists a descending path from some maximum $x$ to $e$
+% 	consisting of edges in $E$, such that all edges along this path have the same color as $x$. 
+% 
+% An \emph{initial} painting has the additional property that the restriction $\chi:X \to [|X|]$ is a bijection.
+% \end{definition}
+
+% Note that a painting colors edges, and not vertices (except for maxima). Our definition also does not require the timing aspect
+% of iterating over colors, though that is one way of painting $\NN$. We begin with an initial painting, since all maximum
+% colors are distinct. A few comments on paint vs water. The interface between two regions of different color
+% is \emph{not} a contour, and so we cannot apply the divide-and-conquer
+% approach of contour surgery. On the other hand, painting does not cut $\NN$, so there is no increase in complexity.
+% Clearly, an initial painting can be constructed in $O(N)$ time. This is the tradeoff between water and paint.
+% Water allows for an easy divide-and-conquer, at the cost of more complexity in the input. For an extremum
+% dominant input, using water to divide the input $\NN$ raises the complexity too much.
+
+Our algorithm incrementally builds $\cJ(\MM)$ from the leaves (maxima) to the root (dominant minimum), by repeatedly merging colors and components together.  
+We say that vertex $v$ is \emph{touched} by color $c$, if there is a $c$-colored edge with lower endpoint $v$.
+Let us focus on the initial painting, where the colors have 1-1 correspondence with the maxima. Refer to the left part of \Fig{colors}.
 Consider two sibling leaves $\ell_1, \ell_2$ and their common parent $v$. The leaves are maxima,
@@
 Both mounds are colored completely blue or red, while $v$ is touched by both colors.
-So this indicates that $v$ joins the blue maximum and red maximum in the join tree and is a critical point.
-To process $v$, we ``merge" the colors red and blue into a new color, purple. 
-In terms of the join tree, this
+So this indicates that $v$ joins the blue maximum and red maximum in $\cJ(\MM)$.
+
+This is precisely how we hope to exploit the information in the painting. We prove later that
+when some join $v$ has exactly two colors among all incident edges with $v$ as the lower endpoint, the corresponding maxima (of those colors) are exactly
+the children of $v$ in $\cJ(\MM)$. To proceed further, we ``merge" the colors red and blue into a new color, purple. In other words,
+we replace all red and blue edges by purple edges. This indicates that the red and blue maxima have been
+handled. Imagine flattening the red and blue mounds until reaching $v$, so that the former join $v$
+is now a new maximum, from which purple paint is poured. In terms of $\cJ(\MM)$, this
 is equivalent to removing leaves $\ell_1$ and $\ell_2$, and making $v$ a new leaf.
+Alternatively, $\cJ(\MM)$ has been constructed up to $v$, and it remains to determine $v$'s parent.
+The merging of the colors is not explicitly performed as that would be too expensive; instead we maintain a union-find
+data structure for that. So the red and blue mounds are not actually recolored purple, 
+and instead say red now points to blue in the union find data structure.
 
 Of course, things are more complicated when there are other mounds. There may be a yellow mound,
-corresponding to $\ell_3$ that joins with the blue mound higher up at some vertex $u$ (see the right part of \Fig{colors}). In the join tree, $\ell_1$ and $\ell_3$
+corresponding to $\ell_3$ that joins with the blue mound higher up at some vertex $u$ (see the right part of \Fig{colors}). In $\cJ(\MM)$, $\ell_1$ and $\ell_3$
 are sibling leaves, and $\ell_2$ is a sibling of some ancestor of these leaves. So we cannot
-merge red and blue, until yellow and blue merge. 
-A critical insight is the use of \emph{binomial heaps}~\cite{Vu78}, or any other efficiently mergeable heaps, to handle all the priorities and the merging.
-This ensures that all comparisons are only made between points that are ancestor-descendant in the join tree or adjacent in $\MM$.
-There are numerous details to handle to prove the final run time bound, but this should provide the reader
-some intuition behind our algorithm.
+merge red and blue, until yellow and blue merge. Naturally, we use priority queues to handle this.
+We know $u$ must also be touched by blue, so all critical vertices touched by blue
+are put in a priority queue keyed by height, and vertices are handled in that order.
 
@@
 
-% What happens when finally blue and red join at $v$? We merge the two colors, but now have
-% blue and red queues of critical vertices, which also need to be merged to get a consistent painting. 
-% This necessitates using a priority queue with efficient merges.
-% Specifically, we use \emph{binomial heaps}~\cite{Vu78}, as they provide logarithmic time merges and deletes (though other similar heaps work). 
-% We stress that the feasibility of the entire approach hinges on the use of such an efficient heap structure.
-}
-
-\newcommand{\intuitionSection}{
+What happens when finally blue and red join at $v$? We merge the two colors, but now have
+blue and red queues of critical vertices, which also need to be merged to get a consistent painting. 
+This necessitates using a priority queue with efficient merges.
+Specifically, we use \emph{binomial heaps}~\cite{Vu78}, as they provide logarithmic time merges and deletes (though other similar heaps work). 
+We stress that the feasibility of the entire approach hinges on the use of such an efficient heap structure.
+
+In this discussion, we ignored an annoying problem. Vertices
+may actually be touched by numerous colors, not just one or two as assumed above. A simple solution
+would be to insert vertices into heaps corresponding to all colors touching it. But there could
+be super-constant numbers of copies of a vertex, and handling all these copies would lead
+to extra overhead. We show that it suffices to simply put each vertex $v$ into at most two heaps, one for each
+``side" of a possible join. We are guaranteed that when $v$ needs to be processed, all edges incident from above have
+at most $2$ colors, because of all the color merges that previously occurred.
+
+\myparagraph{Running time intuition:} 
+ All non-heap operations can be easily bounded by $O(t\alpha(t) + N)$. 
+ One can argue any heap always contains a subset of a single leaf to root path.  
+ Bounding each heap by the size of this path suffices to prove the bound of \Thm{main-corr}, stated for join trees. 
+ However, this may grossly overestimate heap sizes, as each vertex on a path can have at most two colors (the ones getting merged).
+ Thus as we get closer to the root the competition for which two colors a vertex gets assigned to grows, since the number of descendant leaf colors grows. 
+ This means for some vertices the size of the heap will be significantly smaller during an associated heap operation.
+
+The intuition is that the paint spilling from the maxmima in the simplicial complex, corresponds to paint spilling from the leaves in 
+the join tree, which decomposes the join tree into a set of colored paths.  Unfortunately, the situation is more complex since 
+while a given color class is confined to a single leaf to root path, it may \emph{not} appear contiguously on this path, as the right part of \Fig{colors} shows. 
+Specifically, in this figure the far left saddle (labeled $i$) is hit by blue paint.  However, there is another 
+saddle on the far right (labeled $j$) which is not hit by blue paint.  Since this far right saddle is slightly 
+higher than the far left one, the merge event involving the component containing the blue mound (and also the yellow and red mounds) and corresponding to the far right saddle
+will occur before the one corresponding to the far left saddle.  Hence, the vertices initially touched by blue are not contiguous in the join tree.
+%
+This non-contiguous complication along with the fact that heap sizes keep changing as color classes merge, causes the analysis to be technically challenging. 
+We employ a variant of \emph{heavy path decompositions}, first used by Sleator and Tarjan for analyzing link/cut trees~\cite{st-dsdt-83}.
+The final analysis charges expensive heap operations to long paths in the decomposition, resulting in the bound stated in \Thm{main-alg}.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\newcommand{\newintuitionSection}{
 \section{A tour of the new contour tree algorithm} \label{sec:approach}
-Our final algorithm is quite technical and has numerous moving parts. However, for the $d=2$ case, 
-where the input is just a triangulated terrain, the main ideas of the parts of the algorithm 
-can be explained clearly.  Therefore, here we first provide a high level view of the entire result.
-
-In the interest of presentation, the definitions and theorem statements in this section will
-slightly differ from those in the main body. They may also differ from the original definitions proposed
-in earlier work.
-
-\myparagraph{Do not globally sort:} The starting point for this work is \Fig{relative}. We have two terrains
-with exactly the same contour tree, but different orderings of (heights of) the critical points.
-Turning it around, we cannot deduce the full height ordering of critical points from the contour tree.
-Sorting all critical points is computationally unnecessary for constructing the contour tree.
-In \Fig{sorting}, the contour tree consists of two balanced binary trees, one of the joins,
-another of the splits. Again, it is not necessary to know the relative ordering between the mounds
-on the left (or among the depressions on the right) to compute the contour tree. Yet some ordering
-information is necessary: on the left, the little valleys are higher than the big central valley,
-and this is reflected in the contour tree. Leaf paths in the contour tree have points
-in sorted order, but incomparable points in the tree are unconstrained.
-How do we sort exactly what is required, without knowing the contour tree in advance?
-
-\subsection{Breaking $\MM$ into simpler pieces} \label{sec:break}
-Let us begin with the algorithm of Carr, Snoeyink, and Axen~\cite{csa-cctad-03}. The key insight is to build
-two different trees, called the join and split trees, and then merge them together into the contour tree.
-Consider sweeping down the hyperplane $x_{d+1} = h$ and taking the \emph{superlevel} sets. These 
-are the connected components of the portion of $\MM$ above height $h$. For a terrain, the superlevel
+We provide a short, high-level description of the main ideas of subsequent sections. 
+A longer more detailed version of this high-level description can be found in the full version \CiteFullVer.
+
+\myparagraph{Cutting $\MM$ into extremum dominant pieces:} We define a simplicial complex
+endowed with a height to be \emph{extremum dominant} if there exists only a single minimum, or only a single maximum.
+(When formally defined in \Sec{rain}, extremum dominant complexes will allow additional trivial minima or maxima which are a small complication resulting from the following cutting procedure.)
+We first cut $\MM$ into disjoint extremum dominant pieces in linear time.
+Take an arbitrary maximum $x$ and imagine torrential rain at the maximum. The water flows down, wetting any point that has a non-ascending path 
+from $x$. We end up with two portions, the wet part of $\MM$ and the dry part. This is similar
+to \emph{watershed} algorithms used for image segmentation~\cite{RoMe00}. The wet part is obviously connected and
+can be shown to be extremum dominant. We can cut along the interface of the wet and dry, remove
+the wet part, and recurse on the dry parts. This is carefully done to ensure that the total complexity
+of the pieces is linear.
+We prove a simple contour surgery theorem that builds the contour tree of $\MM$
+from the contour trees of the various pieces created.
+
+Using ideas from~\cite{csa-cctad-03}, we prove that the contour tree of an extremum dominant complex is (basically) just the \emph{join tree}, 
+which unlike the contour tree tracks superlevel sets rather than level sets.
+Consider sweeping down the hyperplane $x_{d+1} = h$ and taking the connected components of the portion of $\MM$ above height $h$. For a terrain, these superlevel
 sets are a collection of ``mounds". As we sweep downwards, these mounds keep joining each other,
 until finally, we end up with all of $\MM$. The join tree tracks exactly these events.
-Formally, let $\MM^+_v$ denote the simplicial complex induced on the subset of 
-vertices which are higher than $v$.
-
-\begin{definition} 
-\label{def:int-criticalJoin}
-The \emph{join tree} $\cJ(\MM)$ is built on the set $\cV$ of all critical points.
-The directed edge $(u,v)$ is present when $u$ is the smallest valued vertex in $\cV$ in a connected component of $\MM^+_v$
-and $v$ is adjacent (in $\MM$) to a vertex in this component. 
-\end{definition}
-
-%Abusing notation, for now we will just call this the join tree. 
-%(In \Sec{extreme} \InSoCGVer{and \Sec{jstedm}} we carefully handle the distinction between the join tree and the critical join tree.)
-Refer to \Fig{sorting} for the join tree of a terrain. Note that nothing happens at splits, but these
-are still put as vertices in the join tree. They simply form a long path. The split tree
-is obtained by simply inverting this procedure, sweeping upwards and tracking sublevel sets.
-
-A major insight of~\cite{csa-cctad-03} is an ingeniously simple linear time procedure to construct
-the contour tree from the join and split trees. So the bottleneck is computing these trees. Observe
-in \Fig{sorting} that the split vertices form a long path in the join tree (and vice versa). Therefore, constructing
-these trees forces a global sort of the splits, an unnecessary computation for the contour tree. Unfortunately, in general (i.e.\ unlike \Fig{sorting}) 
-the heights of joins and splits may be interleaved in a complex manner, and hence the final merging of~\cite{csa-cctad-03}
-to get the contour tree requires having the split vertices in the join tree. Without
-this, it is not clear how to get a consistent view of both joins and splits, required for the contour tree.
-
-Our aim is to break $\MM$ into smaller pieces, where this unnecessary computation can be avoided.
-
-\myparagraph{Contour surgery:} We first need a divide-and-conquer lemma. Any contour $\phi$ can be associated
-with an edge $e$ of the contour tree. Suppose we ``cut" $\MM$ along this contour. We prove that $\MM$
-is split into two disconnected pieces, such the contour trees of these pieces is obtained
-by simply cutting $e$ in $\reeb(\MM)$. Alternatively, the contour trees of these pieces can
-be glued together to get $\reeb(\MM)$. This is not particularly surprising, and is fairly easy to prove
-with the right definitions. The idea of loop surgery has been used to reduce Reeb graphs to contour trees~\cite{TiGySi09,DoNa13}.
-Nonetheless, our theorem appears to be new and works for all dimensions.
-
-\begin{figure}[h]\centering
-    \includegraphics[width=.172\linewidth]{figs/tent2}%
-    \hspace{1.2in}
-    \includegraphics[width=.172\linewidth]{figs/tent1}
-    \caption{On left, downward rain spilling only (each shade of gray represents a piece created by each different spilling), producing a grid.  
-    Note we are assuming raining was done in sorted order of the maxima (i.e.\ lowest to highest).  
-    On right, flipping the direction of rain spilling.}
-    \label{fig:order}
-\end{figure}
-
-\myparagraph{Cutting $\MM$ into extremum dominant pieces:} We define a simplicial complex
-endowed with a height to be \emph{minimum dominant} if there exists only a single minimum.
-(Our real definition is more complicated, and involves simplicial complexes
-that allow additional ``trivial'' minima.)
-In such a complex, there exists a non-ascending path from any point to this unique minimum.
-Analogously, we can define \emph{maximum dominant} complexes,
-and both are collectively called \emph{extremum dominant}.
-
-We will cut $\MM$ into disjoint extremum dominant pieces, in linear time.
-One way to think of our procedure is a meteorological analogy. Take an arbitrary maximum $x$,
-and imagine torrential rain at the maximum. The water flows down, wetting any point that has a non-ascending path 
-from $x$. We end up with two portions, the wet part of $\MM$ and the dry part. This is similar
-to \emph{watershed} algorithms used for image segmentation~\cite{RoMe00}. The wet part is obviously connected,
-while there may be numerous disconnected dry parts. The interface between the dry and wet parts
-%is a set of contours\footnote{Technically, they are not contours, but rather the limits of sequences of contours.}, 
-is a set of limits of sequences of contours, which informally we call just a set of contours, which are
-given by the ``water line". The wet part is clearly maximum dominant, since all
-wet points have a non-descending path to $x$. So we can simply cut along the interface
-contours to get the wet maximum dominant piece $\MM'$. By our contour surgery theorem,
-we are left with a set of disconnected dry parts, and we can recur this procedure on them.
-
-But here's the catch. Every time we cut $\MM$ along a contour, we potentially increase
-the complexity of $\MM$. Water flows in the interior of faces, and the interface will naturally
-cut some faces. Each cut introduces new vertices, and a poor choice of repeated raining leads
-to a large increase in complexity. Consider the left of \Fig{order}. Each raining produces a single
-wet and dry piece, and each cut introduces many new vertices. If we wanted to partition this terrain into 
-maximum dominant simplicial complexes, the final complexity would be forbiddingly large.
-
-A simple trick saves the day. Unlike reality, we can choose rain to flow solely downwards or solely upwards.
-Apply the procedure above to get a single wet maximum dominant $\MM'$ and a set of dry pieces.
-Observe that a single dry piece $\NN$ is boundary critical with the newly introduced boundary $\phi$ (the wet-dry interface) behaving as a minimum.
-So we can rain upwards from this minimum, and get a \emph{minimum dominant} portion $\NN'$.
-This ensures that the new interface (after applying the procedure on $\NN$) does not
-cut any face previously cut by $\phi$.
-For each of the new dry pieces, the newly introduced boundary is now a maximum. So we rain downwards from there.
-More formally, we alternate between raining upwards and downwards as we go down the recursion tree.
-We can prove that an original face of $\MM$ is cut at most once, so the final complexity can be bounded.
-In \Fig{order}, regardless of the choice of the starting maximum,  this procedure would yield (at most) two pieces, 
-one maximum dominant, and one minimum dominant.
-
-Using the contour surgery theorem previously discussed, we can build the contour tree of $\MM$
-from the contour trees of the various pieces created. All in all, we prove the following theorem.
-
-\begin{theorem} \label{thm:int-rain} There is an $O(N)$ time procedure that cuts $\MM$ into extremum dominant simplicial complexes
-$\MM_1, \MM_2, \ldots$. Furthermore, given the set of contour trees $\{\cC(\MM_i)\}$, $\cC(\MM)$ can be constructed in $O(N)$ time.
-\end{theorem}
-
-\myparagraph{Extremum dominance simplifies contour trees:} We will focus on minimum dominant simplicial complexes $\MM$. By \Thm{int-rain}, it suffices
-to design an algorithm for contour trees on such inputs. 
-For the $d=2$ case, it helps to visualize such an input as a terrain with no minima,
-except at a unique boundary face (think of a large boundary triangle that is the boundary).
-All the saddles in such a terrain are necessarily joins, and there can be no splits.
-In \Fig{sorting}, the portion on the left is minimum dominant in exactly this fashion, albeit in one dimension lower.
-%Water poured at any critical point must flow to the boundary, and the entire wet portion is connected.
-More formally, $\MM^-_v$ is connected for all $v$, so there are no splits.
-
-We can prove that the split tree is just a path, and the contour tree is exactly
-the join tree. The formal argument is a little involved, and we employ the merging procedure
-of~\cite{csa-cctad-03} to get a proof. We demonstrate that the merging procedure will
-actually just output the join tree, so we do not need to actually compute the split tree.
-(The real definition of minimum dominant is a little more complicated,
-so the contour tree is more than just the join tree. But computationally, it suffices
-to construct the join tree.)
-
-We stress the importance of this step for our approach. Given the algorithm of~\cite{csa-cctad-03},
-one may think that it suffices to design faster algorithms for join trees. But this
-cannot give the sort of optimality we hope for. Again, consider \Fig{sorting}. Any algorithm
-to construct the true join tree must construct the path of splits, which implies
-sorting all of them. Thus it is necessary to cut $\MM$ into pieces where the cost
-of building the join tree can be related to that of building $\cC(\MM)$.
-
-\subsection{Join trees from painted mountaintops} \label{sec:join-paint}
-
-Arguably, everything up to this point is a preamble for the main result: a faster
-algorithm for join trees. Our algorithm does not require the initial input to be extremum
-dominant. This is only required to relate the join trees, of the resulting subcomplexes of Theorem~\ref{thm:int-rain}, to the contour tree of the initial input $\MM$.
-For clarity, we use $\NN$ to denote the input here. Note that in \Def{int-criticalJoin},
-the join tree is defined purely combinatorially in terms of the 1-skeleton (the underlying graph) of $\NN$.
-
-The join tree $\cJ(\NN)$ is a rooted tree with the dominant minimum at the root, and we direct edges downwards (towards the root). So it makes sense to talk
-of comparable vs incomparable vertices.
-We arrive at the main challenge: how to sort only the comparable critical points, without
-constructing the join tree? The join tree algorithm of~\cite{csa-cctad-03} is a typical event-based
-computational geometry algorithm. We have to step away from this viewpoint to avoid the global sort.
-
+
+\myparagraph{Join trees from painted mountaintops:} Our main result is a faster algorithm for join trees.
 The key idea is \emph{paint spilling}. Start with each maximum having a large can of paint, with distinct
 colors for each maximum. In arbitrary order, we spill paint from each maximum, wait till it flows down,
-then spill from the next, etc. Paint is viscous, and only flows down edges. \emph{It does not
-paint the interior of faces with dimension strictly larger than 1.} That is, this process is restricted to the 1-skeleton of $\NN$.
-Furthermore, our paints do not mix, so each edge receives a unique color, decided by the first
-paint to reach it.  In the following, $[n]$ denotes the set $\{1, \ldots, n\}$, for any natural number $n$.
-
-\begin{definition} \label{def:paint1} \InSoCGVer{(Restatement of \Def{paint2}.)} Let the 1-skeleton of $\NN$ have edge set $E$ and maxima $X$.
-A  \emph{painting} of $\NN$ is a map $\chi:X \cup E \to [|X|]$ with the following property. 
- Consider an edge $e$. There exists a descending path from some maximum $x$ to $e$
-	consisting of edges in $E$, such that all edges along this path have the same color as $x$.
-
-An \emph{initial} painting has the additional property that the restriction $\chi:X \to [|X|]$ is a bijection.
-\end{definition}
-
-Note that a painting colors edges, and not vertices (except for maxima). Our definition also does not require the timing aspect
-of iterating over colors, though that is one way of painting $\NN$. We begin with an initial painting, since all maximum
-colors are distinct. A few comments on paint vs water. The interface between two regions of different color
-is \emph{not} a contour, and so we cannot apply the divide-and-conquer
-approach of contour surgery. On the other hand, painting does not cut $\NN$, so there is no increase in complexity.
-Clearly, an initial painting can be constructed in $O(N)$ time. This is the tradeoff between water and paint.
-Water allows for an easy divide-and-conquer, at the cost of more complexity in the input. For an extremum
-dominant input, using water to divide the input $\NN$ raises the complexity too much.
-
-Our algorithm incrementally builds $\cJ(\MM)$ from the leaves (maxima) to the root (dominant minimum). 
-We say that vertex $v$ is \emph{touched} by color $c$, if there is a $c$-colored edge with lower endpoint $v$.
-Let us focus on an initial painting, where the colors have 1-1 correspondence with the maxima. Refer to the left part of \Fig{colors}.
+then spill from the next, etc. Paint is viscous, and only flows down edges. Furthermore, our paints do not mix, so each edge receives a unique color, decided by the first
+paint to reach it.
+
+Our algorithm incrementally builds the join tree from the leaves (maxima) to the root. 
+Refer to the left part of \Fig{colors}.
 Consider two sibling leaves $\ell_1, \ell_2$ and their common parent $v$. The leaves are maxima,
@@
 Both mounds are colored completely blue or red, while $v$ is touched by both colors.
-So this indicates that $v$ joins the blue maximum and red maximum in $\cJ(\MM)$.
-
-This is precisely how we hope to exploit the information in the painting. We prove later that
-when some join $v$ has exactly two colors among all incident edges with $v$ as the lower endpoint, the corresponding maxima (of those colors) are exactly
-the children of $v$ in $\cJ(\MM)$. To proceed further, we ``merge" the colors red and blue into a new color, purple. In other words,
-we replace all red and blue edges by purple edges. This indicates that the red and blue maxima have been
-handled. Imagine flattening the red and blue mounds until reaching $v$, so that the former join $v$
-is now a new maximum, from which purple paint is poured. In terms of $\cJ(\MM)$, this
+So this indicates that $v$ joins the blue maximum and red maximum in the join tree and is a critical point.
+To process $v$, we ``merge" the colors red and blue into a new color, purple. 
+In terms of the join tree, this
 is equivalent to removing leaves $\ell_1$ and $\ell_2$, and making $v$ a new leaf.
-Alternatively, $\cJ(\MM)$ has been constructed up to $v$, and it remains to determine $v$'s parent.
-The merging of the colors is not explicitly performed as that would be too expensive; instead we maintain a union-find
-data structure for that. So the red and blue mounds are not actually recolored purple, 
-and instead say red now points to blue in the union find data structure.
 
 Of course, things are more complicated when there are other mounds. There may be a yellow mound,
-corresponding to $\ell_3$ that joins with the blue mound higher up at some vertex $u$ (see the right part of \Fig{colors}). In $\cJ(\MM)$, $\ell_1$ and $\ell_3$
+corresponding to $\ell_3$ that joins with the blue mound higher up at some vertex $u$ (see the right part of \Fig{colors}). In the join tree, $\ell_1$ and $\ell_3$
 are sibling leaves, and $\ell_2$ is a sibling of some ancestor of these leaves. So we cannot
-merge red and blue, until yellow and blue merge. Naturally, we use priority queues to handle this issue.
-We know that $u$ must also be touched by blue. So all critical vertices touched by blue
-are put into a priority queue keyed by height, and vertices are handled in that order.
+merge red and blue, until yellow and blue merge. 
+A critical insight is the use of \emph{binomial heaps}~\cite{Vu78}, or any other efficiently mergeable heaps, to handle all the priorities and the merging.
+This ensures that all comparisons are only made between points that are ancestor-descendant in the join tree or adjacent in $\MM$.
+There are numerous details to handle to prove the final run time bound, but this should provide the reader
+some intuition behind our algorithm.
 
@@
 \end{figure}
+}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\newcommand{\intuitionSection}{
+\section{A tour of the new contour tree algorithm} \label{sec:approach}
+Our final algorithm is quite technical and has numerous moving parts. However, for the $d=2$ case, 
+where the input is just a triangulated terrain, the main ideas of the parts of the algorithm 
+can be explained clearly.  Therefore, here we first provide a high level view of the entire result.
+
+In the interest of presentation, the definitions and theorem statements in this section will
+slightly differ from those in the main body. They may also differ from the original definitions proposed
+in earlier work.
+
+\myparagraph{Do not globally sort:} The starting point for this work is \Fig{relative}. We have two terrains
+with exactly the same contour tree, but different orderings of (heights of) the critical points.
+Turning it around, we cannot deduce the full height ordering of critical points from the contour tree.
+Sorting all critical points is computationally unnecessary for constructing the contour tree.
+In \Fig{sorting}, the contour tree consists of two balanced binary trees, one of the joins,
+another of the splits. Again, it is not necessary to know the relative ordering between the mounds
+on the left (or among the depressions on the right) to compute the contour tree. Yet some ordering
+information is necessary: on the left, the little valleys are higher than the big central valley,
+and this is reflected in the contour tree. Leaf paths in the contour tree have points
+in sorted order, but incomparable points in the tree are unconstrained.
+How do we sort exactly what is required, without knowing the contour tree in advance?
+
+\subsection{Breaking $\MM$ into simpler pieces} \label{sec:break}
+Let us begin with the algorithm of Carr, Snoeyink, and Axen~\cite{csa-cctad-03}. The key insight is to build
+two different trees, called the join and split trees, and then merge them together into the contour tree.
+Consider sweeping down the hyperplane $x_{d+1} = h$ and taking the \emph{superlevel} sets. These 
+are the connected components of the portion of $\MM$ above height $h$. For a terrain, the superlevel
+sets are a collection of ``mounds". As we sweep downwards, these mounds keep joining each other,
+until finally, we end up with all of $\MM$. The join tree tracks exactly these events. 
+Formally, let $\MM^+_v$ denote the simplicial complex induced on the subset of 
+vertices which are higher than $v$.
+
+\begin{definition} 
+\label{def:int-criticalJoin}
+The \emph{join tree} $\cJ(\MM)$ is built on the set $\cV$ of all critical points.
+The directed edge $(u,v)$ is present when $u$ is the smallest valued vertex in $\cV$ in a connected component of $\MM^+_v$
+and $v$ is adjacent (in $\MM$) to a vertex in this component. 
+\end{definition}
+
+Refer to \Fig{sorting} for the join tree of a terrain. Note that nothing happens at splits, but these
+are still put as vertices in the join tree. They simply form a long path. The split tree
+is obtained by simply inverting this procedure, sweeping upwards and tracking sublevel sets.
+
+A major insight of~\cite{csa-cctad-03} is an ingeniously simple linear time procedure to construct
+the contour tree from the join and split trees. So the bottleneck is computing these trees. Observe
+in \Fig{sorting} that the split vertices form a long path in the join tree (and vice versa). Therefore, constructing
+these trees forces a global sort of the splits, an unnecessary computation for the contour tree. Unfortunately, in general (i.e.\ unlike \Fig{sorting}) 
+the heights of joins and splits may be interleaved in a complex manner, and hence the final merging of~\cite{csa-cctad-03}
+to get the contour tree requires having the split vertices in the join tree. Without
+this, it is not clear how to get a consistent view of both joins and splits, required for the contour tree.
+
+Our aim is to break $\MM$ into smaller pieces, where this unnecessary computation can be avoided.
+
+\myparagraph{Contour surgery:} We first need a divide-and-conquer lemma. Any contour $\phi$ can be associated
+with an edge $e$ of the contour tree. Suppose we ``cut" $\MM$ along this contour. We prove that $\MM$
+is split into two disconnected pieces, such the contour trees of these pieces is obtained
+by simply cutting $e$ in $\reeb(\MM)$. Alternatively, the contour trees of these pieces can
+be glued together to get $\reeb(\MM)$. This is not particularly surprising, and is fairly easy to prove
+with the right definitions. The idea of loop surgery has been used to reduce Reeb graphs to contour trees~\cite{TiGySi09,DoNa13}.
+Nonetheless, our theorem appears to be new and works for all dimensions.
+
+\begin{figure}[h]\centering
+    \includegraphics[width=.172\linewidth]{figs/tent2}%
+    \hspace{1.2in}
+    \includegraphics[width=.172\linewidth]{figs/tent1}
+    \caption{On left, downward rain spilling only (each shade of gray represents a piece created by each different spilling), producing a grid.  
+    Note we are assuming raining was done in sorted order of the maxima (i.e.\ lowest to highest).  
+    On right, flipping the direction of rain spilling.}
+    \label{fig:order}
+\end{figure}
+
+\myparagraph{Cutting $\MM$ into extremum dominant pieces:} We define a simplicial complex
+endowed with a height to be \emph{minimum dominant} if there exists only a single minimum.
+(Our real definition is more complicated, and involves simplicial complexes
+that allow additional ``trivial'' minima.)
+In such a complex, there exists a non-ascending path from any point to this unique minimum.
+Analogously, we can define \emph{maximum dominant} complexes,
+and both are collectively called \emph{extremum dominant}.
+
+We will cut $\MM$ into disjoint extremum dominant pieces, in linear time.
+One way to think of our procedure is a meteorological analogy. Take an arbitrary maximum $x$,
+and imagine torrential rain at the maximum. The water flows down, wetting any point that has a non-ascending path 
+from $x$. We end up with two portions, the wet part of $\MM$ and the dry part. This is similar
+to \emph{watershed} algorithms used for image segmentation~\cite{RoMe00}. The wet part is obviously connected,
+while there may be numerous disconnected dry parts. The interface between the dry and wet parts
+%is a set of contours\footnote{Technically, they are not contours, but rather the limits of sequences of contours.}, 
+is a set of limits of sequences of contours, which informally we call just a set of contours, which are
+given by the ``water line". The wet part is clearly maximum dominant, since all
+wet points have a non-descending path to $x$. So we can simply cut along the interface
+contours to get the wet maximum dominant piece $\MM'$. By our contour surgery theorem,
+we are left with a set of disconnected dry parts, and we can recur this procedure on them.
+
+But here's the catch. Every time we cut $\MM$ along a contour, we potentially increase
+the complexity of $\MM$. Water flows in the interior of faces, and the interface will naturally
+cut some faces. Each cut introduces new vertices, and a poor choice of repeated raining leads
+to a large increase in complexity. Consider the left of \Fig{order}. Each raining produces a single
+wet and dry piece, and each cut introduces many new vertices. If we wanted to partition this terrain into 
+maximum dominant simplicial complexes, the final complexity would be forbiddingly large.
+
+A simple trick saves the day. Unlike reality, we can choose rain to flow solely downwards or solely upwards.
+Apply the procedure above to get a single wet maximum dominant $\MM'$ and a set of dry pieces.
+Observe that a single dry piece $\NN$ is boundary critical with the newly introduced boundary $\phi$ (the wet-dry interface) behaving as a minimum.
+So we can rain upwards from this minimum, and get a \emph{minimum dominant} portion $\NN'$.
+This ensures that the new interface (after applying the procedure on $\NN$) does not
+cut any face previously cut by $\phi$.
+For each of the new dry pieces, the newly introduced boundary is now a maximum. So we rain downwards from there.
+More formally, we alternate between raining upwards and downwards as we go down the recursion tree.
+We can prove that an original face of $\MM$ is cut at most once, so the final complexity can be bounded.
+In \Fig{order}, regardless of the choice of the starting maximum,  this procedure would yield (at most) two pieces, 
+one maximum dominant, and one minimum dominant.
+
+Using the contour surgery theorem previously discussed, we can build the contour tree of $\MM$
+from the contour trees of the various pieces created. All in all, we prove the following theorem.
+
+\begin{theorem} \label{thm:int-rain} There is an $O(N)$ time procedure that cuts $\MM$ into extremum dominant simplicial complexes
+$\MM_1, \MM_2, \ldots$. Furthermore, given the set of contour trees $\{\cC(\MM_i)\}$, $\cC(\MM)$ can be constructed in $O(N)$ time.
+\end{theorem}
+
+\myparagraph{Extremum dominance simplifies contour trees:} We will focus on minimum dominant simplicial complexes $\MM$. By \Thm{int-rain}, it suffices
+to design an algorithm for contour trees on such inputs. 
+For the $d=2$ case, it helps to visualize such an input as a terrain with no minima,
+except at a unique boundary face (think of a large boundary triangle that is the boundary).
+All the saddles in such a terrain are necessarily joins, and there can be no splits.
+In \Fig{sorting}, the portion on the left is minimum dominant in exactly this fashion, albeit in one dimension lower.
+%Water poured at any critical point must flow to the boundary, and the entire wet portion is connected.
+More formally, $\MM^-_v$ is connected for all $v$, so there are no splits.
+
+We can prove that the split tree is just a path, and the contour tree is exactly
+the join tree. The formal argument is a little involved, and we employ the merging procedure
+of~\cite{csa-cctad-03} to get a proof. We demonstrate that the merging procedure will
+actually just output the join tree, so we do not need to actually compute the split tree.
+(The real definition of minimum dominant is a little more complicated,
+so the contour tree is more than just the join tree. But computationally, it suffices
+to construct the join tree.)
+
+We stress the importance of this step for our approach. Given the algorithm of~\cite{csa-cctad-03},
+one may think that it suffices to design faster algorithms for join trees. But this
+cannot give the sort of optimality we hope for. Again, consider \Fig{sorting}. Any algorithm
+to construct the true join tree must construct the path of splits, which implies
+sorting all of them. Thus it is necessary to cut $\MM$ into pieces where the cost
+of building the join tree can be related to that of building $\cC(\MM)$.
+
+\subsection{Join trees from painted mountaintops} \label{sec:join-paint}
+
+Arguably, everything up to this point is a preamble for the main result: a faster
+algorithm for join trees. Our algorithm does not require the initial input to be extremum
+dominant. This is only required to relate the join trees, of the resulting subcomplexes of Theorem~\ref{thm:int-rain}, to the contour tree of the initial input $\MM$.
+For clarity, we use $\NN$ to denote the input here. Note that in \Def{int-criticalJoin},
+the join tree is defined purely combinatorially in terms of the 1-skeleton (the underlying graph) of $\NN$.
+
+The join tree $\cJ(\NN)$ is a rooted tree with the dominant minimum at the root, and we direct edges downwards (towards the root). So it makes sense to talk
+of comparable vs incomparable vertices.
+We arrive at the main challenge: how to sort only the comparable critical points, without
+constructing the join tree? The join tree algorithm of~\cite{csa-cctad-03} is a typical event-based
+computational geometry algorithm. We have to step away from this viewpoint to avoid the global sort.
+
+The key idea is \emph{paint spilling}. Start with each maximum having a large can of paint, with distinct
+colors for each maximum. In arbitrary order, we spill paint from each maximum, wait till it flows down,
+then spill from the next, etc. Paint is viscous, and only flows down edges. \emph{It does not
+paint the interior of faces with dimension strictly larger than 1.} That is, this process is restricted to the 1-skeleton of $\NN$.
+Furthermore, our paints do not mix, so each edge receives a unique color, decided by the first
+paint to reach it.  In the following, $[n]$ denotes the set $\{1, \ldots, n\}$, for any natural number $n$.
+
+\begin{definition} \label{def:paint1} Let the 1-skeleton of $\NN$ have edge set $E$ and maxima $X$.
+A  \emph{painting} of $\NN$ is a map $\chi:X \cup E \to [|X|]$ with the following property. 
+ Consider an edge $e$. There exists a descending path from some maximum $x$ to $e$
+	consisting of edges in $E$, such that all edges along this path have the same color as $x$.
+
+An \emph{initial} painting has the additional property that the restriction $\chi:X \to [|X|]$ is a bijection.
+\end{definition}
+
+Note that a painting colors edges, and not vertices (except for maxima). Our definition also does not require the timing aspect
+of iterating over colors, though that is one way of painting $\NN$. We begin with an initial painting, since all maximum
+colors are distinct. A few comments on paint vs water. The interface between two regions of different color
+is \emph{not} a contour, and so we cannot apply the divide-and-conquer
+approach of contour surgery. On the other hand, painting does not cut $\NN$, so there is no increase in complexity.
+Clearly, an initial painting can be constructed in $O(N)$ time. This is the tradeoff between water and paint.
+Water allows for an easy divide-and-conquer, at the cost of more complexity in the input. For an extremum
+dominant input, using water to divide the input $\NN$ raises the complexity too much.
+
+Our algorithm incrementally builds $\cJ(\MM)$ from the leaves (maxima) to the root (dominant minimum). 
+We say that vertex $v$ is \emph{touched} by color $c$, if there is a $c$-colored edge with lower endpoint $v$.
+Let us focus on an initial painting, where the colors have 1-1 correspondence with the maxima. Refer to the left part of \Fig{colors}.
+Consider two sibling leaves $\ell_1, \ell_2$ and their common parent $v$. The leaves are maxima,
+and $v$ is a join that ``merges" $\ell_1, \ell_2$. 
+In that case, there are ``mounds" corresponding to $\ell_1$ and $\ell_2$ that merge
+at a valley $v$. Suppose this was the entire input, and $\ell_1$ was colored blue and $\ell_2$ was colored red. 
+Both mounds are colored completely blue or red, while $v$ is touched by both colors.
+So this indicates that $v$ joins the blue maximum and red maximum in $\cJ(\MM)$.
+
+This is precisely how we hope to exploit the information in the painting. We prove later that
+when some join $v$ has exactly two colors among all incident edges with $v$ as the lower endpoint, the corresponding maxima (of those colors) are exactly
+the children of $v$ in $\cJ(\MM)$. To proceed further, we ``merge" the colors red and blue into a new color, purple. In other words,
+we replace all red and blue edges by purple edges. This indicates that the red and blue maxima have been
+handled. Imagine flattening the red and blue mounds until reaching $v$, so that the former join $v$
+is now a new maximum, from which purple paint is poured. In terms of $\cJ(\MM)$, this
+is equivalent to removing leaves $\ell_1$ and $\ell_2$, and making $v$ a new leaf.
+Alternatively, $\cJ(\MM)$ has been constructed up to $v$, and it remains to determine $v$'s parent.
+The merging of the colors is not explicitly performed as that would be too expensive; instead we maintain a union-find
+data structure for that. So the red and blue mounds are not actually recolored purple, 
+and instead say red now points to blue in the union find data structure.
+
+Of course, things are more complicated when there are other mounds. There may be a yellow mound,
+corresponding to $\ell_3$ that joins with the blue mound higher up at some vertex $u$ (see the right part of \Fig{colors}). In $\cJ(\MM)$, $\ell_1$ and $\ell_3$
+are sibling leaves, and $\ell_2$ is a sibling of some ancestor of these leaves. So we cannot
+merge red and blue, until yellow and blue merge. Naturally, we use priority queues to handle this issue.
+We know that $u$ must also be touched by blue. So all critical vertices touched by blue
+are put into a priority queue keyed by height, and vertices are handled in that order.
+
+\begin{figure}[h]\centering
+    \includegraphics[width=.25\linewidth]{figs/redBlue}%
+    \hspace{.07in}
+    \includegraphics[width=.25\linewidth]{figs/redBlue2}%
+    \hspace{.07in}
+    \includegraphics[width=.03\linewidth]{figs/redBlueTree}
+    \hspace{.09in}
+    \includegraphics[width=.3\linewidth]{figs/mound}%
+    \hspace{.07in}
+    \includegraphics[width=.065\linewidth]{figs/moundTree}
+    \caption{On the left, red and blue merge to make purple, followed by the contour tree with initial colors.  On the right, additional maxima and the resulting contour tree.}
+    \label{fig:colors}
+\end{figure}
 
@@
 
-% Relating the running time to a path decomposition is the most technical part of the paper. 
-% All the non-heap operations can be easily bounded by $O(t\alpha(t) + N)$ (the $t\alpha(t)$ is from the union-find data structure
-% for colors). It is not hard to argue that at all times, any heap always contains a subset of a leaf
-% to root path. Unfortunately, this subset is \emph{not} contiguous, as the right part of \Fig{colors} shows. 
-% Specifically, in this figure the far left saddle (labeled $i$) is hit by blue paint.  However, there is another 
-% saddle on the far right (labeled $j$) which is not hit by blue paint.  Since this far right saddle is slightly 
-% higher than the far left one, it will merge into the component containing the blue mound (and also the yellow and red mounds)
-% before the far left one.  Hence, the vertices initially touched by blue are not contiguous in the contour tree.
-% 
-% Nonetheless, we can get a non-trivial (but non-optimal) bound, which is the analogue of \Thm{main-corr} for join trees. 
-% Let $d_v$ denote the distance to the root for vertex $v$ in the join tree.
-% The total cost (of the heap operations) is at most $\sum_v \log d_v$. This immediately proves a bound of $O(t\log D)$, where
-% $D$ is the maximum distance to the root, an improvement over previous work. But this bound is non-optimal.
-% For a balanced binary tree, this bound is $O(t\log\log t)$, whereas the cost of any path decomposition is $O(t)$.
-% \Thm{main-alg}
-%
-% The cost of heap operations depends on heap sizes, which keep changing because
-% of the repeated merging. This causes the analysis to be technically challenging. There are situations where
-% the initial heap sizes are small, but they eventually merge to create larger heaps. 
-% We employ a variant
-% of \emph{heavy path decompositions}, first used by Sleator and Tarjan for analyzing link/cut trees~\cite{st-dsdt-83}.
-% The final analysis charges expensive heap operations to long paths in the decomposition.
-
 \subsection{The lower bound}
@@
 comparisons on all these instances.
-%In general, any algorithm requires $\Omega(C)$ time
-%to solve the set of instances that our algorithm solves with $C$ comparisons.
 
@@
 
-\InSoCGVer{
-
-\newintuitionSection
-
-% \subsection{Outline}
-% The first step is to partition $\MM$ into nicely behaved pieces, such that the contour tree of
-% each piece is a rooted tree.  This is done by cutting $\MM$ at certain contours. Of course, any such ``cut"
-% increases the complexity of the input, so care is required to keep the complexity down.
-% \Sec{surgery} describes our cutting procedure, and \Sec{rain} describes how to use this cutting procedure.
-% It turns out that a simpler construct, the \emph{join tree}~\cite{csa-cctad-00} of each piece is exactly
-% the contour tree. Thus, it suffices to 
-% 
-% Roughly speaking the resulting pieces are ``nice'' because the contour tree of each piece 
-% looks like a rooted tree, i.e.\ the join tree of \cite{csa-cctad-00}, and the precise relation to join trees is the subject of \Sec{extreme}.
-% \Sec{paint} then describes how to optimally compute the join tree (and hence the contour tree) 
-% on each piece (see Appendix~\ref{sec:approach} for discussion on why the join tree should not be computed directly on the input complex). 
-% Finally, \Sec{runTimeSummary} summarizes all sections relating to the running time, specifically \Sec{runTime} which upper bounds the 
-% running time of the algorithm of \Sec{paint}, \Sec{pathDecomp} which uses a sophisticated charging argument to express this bound 
-% in terms of path decompositions, and \Sec{lb} which provides matching running time lower bounds.
-% 
-% The final algorithm is quite technical and has numerous moving parts. However, for the $d=2$ case, 
-% where the input is just a triangulated terrain, the main ideas of the parts of the algorithm 
-% can be explained clearly. Therefore, in \Sec{approach} we provide a high level view of the entire result.
-% We strongly encourage the reader to overview this section, as then the remainder of the paper can be understood 
-% as a simple matter of implementing these ideas.
-% 
-% Alternatively, the reader can view the full version of the paper \CiteFullVer{} which contains this high level overview and all other appendices 
-% in the main body of the text.
-% 
-}
-
-\InNotSoCGVer{
-\intuitionSection
-}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\section{Divide and conquer through contour surgery} \label{sec:surgery}
-
-{\bf The cutting operation:} We define a ``cut" operation on $f:\MM \rightarrow \RR$ that cuts along a regular contour to create
-a new simplicial complex with an added boundary. Given a contour $\phi$, roughly speaking, this constructs the simplicial complex $\MM \setminus \phi$. 
-We will always enforce the condition that $\phi$ never passes through a vertex of $\MM$.
-Again, we use $\eps$ for an infinitesimally small value. We denote $\phi^+$ (resp. $\phi^-$) to be
-the contour at value $f(\phi) + \eps$ (resp. $f(\phi) - \eps$), which is at distance $\eps$
-from $\phi$.
-
-An $h$-contour is achieved by intersecting $\MM$ with the hyperplane $x_{d+1} = h$ and taking a connected component. (Think of the $d+1$-dimension
-as height.) Given some point $x$ on an $h$-contour $\phi$, we can walk along $\MM$ from $x$ to determine $\phi$.
-We can ``cut" along $\phi$ to get a new (possibly) disconnected simplicial complex $\MM'$. This is achieved
-by splitting every face $F$ that $\phi$ intersects into an ``upper" face and ``lower" face. Algorithmically,
-we cut $F$ with $\phi^+$ and take everything above $\phi^+$ in $F$ to make the upper face. Analogously, we cut with $\phi^-$ to get the lower face.
-The faces are then triangulated to ensure that they are all simplices. 
-\Ben{Reviewer wanted clarification on how triangulation is done. Maybe can use bottom-vertex triangulation of Clarkson 88, discussed in Matousek book?}
-This creates the two new boundaries $\phi^+$ and $\phi^-$, and we maintain the property of constant $f$-value at a boundary.
-
-Note that by assumption $\phi$ cannot cut a boundary face, and moreover all non-boundary faces
-have constant size. Therefore, this process takes time linear in $|\phi|$, the number of faces $\phi$ intersects.
-This new simplicial complex is denoted by $\cut(\phi,\MM)$.
-%
-We now describe a high-level approach to construct $\reeb(\MM)$ using this cutting procedure.
-
-\Ben{Surgery should be rewritten to take instead take as input the output of cut, i.e. two complexes and a contour.  
-This is how it is eventually used in Claim\ref{clm:rain-reeb}. Currently it does nothing really.}
-
-\medskip
-\fbox{
-\begin{minipage}{0.9\textwidth}
-{\bf $\surgery(\MM,\phi)$}
-
-\smallskip
-\begin{asparaenum}
-	\item Let $\MM' = \cut(\MM,\phi)$.
-	\item Construct $\reeb(\MM')$ and let $A, B$ be the nodes corresponding to the new boundaries created
-	in $\MM'$. (One is a minimum and the other is maximum.)
-	\item Since $A, B$ are leaves, they each have unique neighbors $A'$ and $B'$, respectively. Insert
-	edge $(A',B')$ and delete $A, B$ to obtain $\reeb(\MM)$.
-\end{asparaenum}
-\end{minipage}}
-
-\medskip
-
-\InSoCGVer{The following theorems are intuitively obvious, and are proven in the full version \CiteFullVer.}
-
-\begin{theorem} \label{thm:surgery} For any regular contour $\phi$, the output of $\surgery(\MM,\phi)$ is $\reeb(\MM)$.
-\end{theorem}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\newcommand{\surgeryBody}{
-%To prove the above theorem, we require some theorems from \cite{c-tmi-04} (Theorems 6.6 and 6.7) that map paths in $\cC(\MM)$ to $\MM$.
-To prove Theorem~\ref{thm:surgery}, we require a theorem from \cite{c-tmi-04} (Theorems 6.6) that map paths in $\cC(\MM)$ to $\MM$.
-
-\begin{theorem} \label{thm:carr-path} For every path $P$ in $\MM$, there exists a path $Q$ in the contour tree corresponding
-to the contours passing through points in $P$.
-For every path $Q$ in the contour tree, there exists at least one path $P$ in $\MM$ through points present
-in contours involving $Q$.
-
-In particular, for every monotone path $P$ in $\MM$, there exists a monotone path $Q$ in the contour tree to which $P$
-maps, and vice versa.
-\end{theorem}
-
-% \begin{theorem} \label{thm:carr-mono} For every monotone path $P$ in $\MM$, there exists a monotone path $Q$ in the contour tree to which $P$
-% maps (as in the previous theorem), and vice versa.
-% \end{theorem}
-
-Theorem~\ref{thm:surgery} is a direct consequence of the following lemma.
-
-\begin{lemma} \label{lem:cut} Consider a regular contour $\phi$ contained in a contour class (of an edge of $\reeb(\MM))$
-$(u,v)$ and let $\MM' = \cut(\MM,\phi)$. Then $\cV(\reeb(\MM')) = \{\phi^+,\phi^-\} \cup \cV(\MM)$
-and $\cE(\reeb(\MM')) = \{(u,\phi^+),(\phi^-,v)\} \cup (\cE(\MM) \setminus (u,v))$.
-\end{lemma}
-
-\begin{proof} First observe that since $\phi$ is a regular contour, the vertex set in the complex $\MM'$ is the same 
-as the vertex set in $\MM$, except with the addition of the newly created vertices on $\phi^+$ and $\phi^-$.  
-Moreover, $\cut(\MM,\phi)$ does not affect the local neighborhood of any vertex in $\MM$. 
-Therefore since a vertex being critical is a local condition, with the exception of new boundary vertices, the 
-critical vertices in $\MM$ and $\MM'$ are the same.  Finally, the new vertices on $\phi^+$ and $\phi^-$ 
-collectively behave as a minimum and maximum, respectively, and so $\cV(\reeb(\MM')) = \{\phi^+,\phi^-\} \cup \cV(\MM)$.
-
-Now consider the edge sets of the contour trees.  Any contour class in $\MM'$ (i.e.\ edge in $\cC(\MM')$) that does not involve $\phi^+$ or $\phi^-$ is also
-a contour class in $\MM$. Furthermore, a maximal contour class satisfying these properties is also
-maximal in $\MM$. So all edges of $\cC(\MM')$ that do not involve $\phi^+$ or $\phi^-$ are edges of $\cC(\MM)$.
-Analogously, every edge of $\cC(\MM)$ not involving $\phi$ is an edge of $\cC(\MM')$.
-
-Consider the contour class corresponding to edge $(u,v)$ of $\cC(\MM)$. There is a natural ordering
-of the contours by function value, ranging from $f(u)$ to $f(v)$. All contours in this class ``above" $\phi$
-form a maximal contour class in $\MM'$, represented by edge $(u,\phi^+)$. Analogously, there is another
-contour class represented by edge $(\phi^-,v)$. We have now accounted for all contours in $\cC(\MM')$,
-completing the proof.
-\end{proof}
-
-A useful corollary of this lemma shows that a contour actually splits the simplicial complex into
-two disconnected complexes.
-}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\InNotSoCGVer{
-\surgeryBody
-}
-
-\begin{theorem} \label{thm:jordan} $\cut(\MM,\phi)$ consists of two
-disconnected simplicial complexes.
-\end{theorem}
-
-\newcommand{\proofofDisconnected}{
-\begin{proof} Denote (as in \Lem{cut}) the edge containing $\phi$ to be $(u,v)$. Suppose for contradiction that there is a path between vertices $u$ and $v$
-in $\MM' = \cut(\MM,\phi)$. By \Thm{carr-path}, there is a path in $\cC(\MM')$ between $u$ and $v$. Since $\phi^+$ and $\phi^-$
-are leaves in $\cC(\MM')$, this path cannot use their incident edges. Therefore by \Lem{cut},
-all the edges of this path are in $\cE(\cC(\MM)) \setminus (u,v)$. So we get a cycle in $\cC(\MM)$, a contradiction.
-To show that there are exactly two connected components in $\cut(\MM,\phi)$, it suffices
-to see that $\cC(\MM')$ has two connected components (by \Lem{cut}) and then applying \Thm{carr-path}.
-\end{proof}
-}
-\InNotSoCGVer{\proofofDisconnected}
-
-% 
-% \begin{lemma} \label{lem:blahblah}
-%  Testing 123 \RefProofInAppendix{blahblah}
-% \end{lemma}
-% 
-% \begin{proof:in:appendix:e}{\ref{lem:blahblah}}{blahblah}
-%  blah balh
-% \end{proof:in:appendix:e}
-
-\section{Raining to partition $\MM$} \label{sec:rain}
-
-In this section, we describe a linear time procedure that partitions $\MM$ into special
-\emph{extremum dominant} simplicial complexes.
-
-\begin{definition} \label{def:dom} A simplicial complex is \emph{minimum dominant} if there exists
-a minimum $x$ such that every non-minimal \emph{vertex} in the complex has a non-ascending path to $x$.
-Analogously define \emph{maximum dominant}. 
-\end{definition}
-
-The first aspect of the partitioning is ``raining''. Start at some point $x \in \MM$ and imagine rain at $x$.
-The water will flow downwards along non-ascending paths and ``wet'' all the points encountered. Note that this procedure considers all points
-of the complex, not just vertices.
-
-\begin{definition} \label{def:wet} Fix $x \in \MM$. The set of points $y \in \MM$ such that there is a non-ascending path from $x$ to $y$
-is denoted by $\wet(x,\MM)$ (which in turn is represented as a simplicial complex). A point $z$ is at the \emph{interface} of $\wet(x,\MM)$ if every neighborhood of $z$
-has non-trivial intersection with $\wet(x,\MM)$ (i.e.\ the intersection is neither empty nor the entire neighborhood).
-\end{definition}
-
-The following claim gives a description of the interface.
-
-\begin{claim} \label{clm:inter} For any $x$, each component of the interface of $\wet(x,\MM)$ contains a join vertex.
-\end{claim}
-
-\begin{proof} If $p \in \wet(x,\MM)$, all the points in any contour containing $p$ are also in $\wet(x,\MM)$.
- (Follow the non-ascending path from $x$ to $p$ and then walk along the contour.) The converse is also true,
- so $\wet(x,\MM)$ contains entire contours.
-
-Let $\eps, \delta$ be sufficiently small as usual. Fix some $y$ at the interface.
-Note that $y \in \wet(x,\MM)$. (Otherwise, $B_\eps(y)\cap \MM$ is dry.)
-The points in $B_\eps(y)\cap \MM$ that lie below $y$ have a descending path from $y$ and hence must be wet.
-There must also be a dry point in $B_\eps(y)\cap \MM$ that is above $y$, and hence,
-there exists a dry, regular $(f(y)+\delta)$-contour $\phi$ intersecting $B_\eps(y)$.
-
-Let $\Gamma_y$ be the contour containing $y$.  Suppose for contradiction
-that $\forall p \in \Gamma_y$, $p$ has up-degree $1$ (see \Def{deg}). Consider the non-ascending path from $x$ to $y$ and let $z$
-be the first point of $\Gamma_y$ encountered. There exists a wet, regular $(f(y) + \delta)$-contour $\psi$ 
-intersecting $B_\eps(z)$. Now, walk from $z$ to $y$ along $\Gamma_y$. If all points $w$ in this walk
-have up-degree $1$, then $\psi$ is the unique $(f(y)+\delta)$-contour
-intersecting $B_\eps(w)$. This would imply that $\phi = \psi$, contradicting the fact that $\psi$ is wet
-and $\phi$ is dry.
-%
-% Therefore, there must exist a join vertex $w$ in $\Gamma_y$, such that $B_\eps(w)$ intersects
-% both $\phi$ and $\psi$. As $\delta, \eps \rightarrow 0^+$, the limit of $\phi$ (which is dry)
-% lies in $\Gamma_y$ (which is wet). 
-% Hence this limit gives a contour of the interface containing the join $w$.
-\end{proof}
-
-Note that $\wet(x,\MM)$ (and its interface) can be computed in time linear in the size of the wet simplicial complex.
-We perform a non-ascending search from $x$. Any face $F$ of $\MM$ encountered is partially (if not entirely) in $\wet(x,\MM)$.
-The wet portion is determined by cutting $F$ along the interface. Since each component of the interface is a contour, this is equivalent
-to locally cutting $F$ by a hyperplane. All these operations can be performed to output $\wet(x,\MM)$ in time linear in $|\wet(x,\MM)|$.
-
-We define a simple \lift{} operation on the interface components. Consider such a component $\phi$ containing
-a join vertex $y$. Take any dry increasing edge incident to $y$, and pick the point $z$ on this edge at height
-$f(y) + \delta$ (where $\delta$ is an infinitesimal, but larger than the value $\eps$ used in the definition of $\cut$). 
-Let $\lift(\phi)$ be the unique contour through the regular point $z$. Note that $\lift(\phi)$ is dry.
-The following claim follows directly from \Thm{jordan}.
-
-\begin{claim} \label{clm:cut-int} Let $\phi$ be a connected component of the interface. Then $\cut(\MM,\lift(\phi))$
-results in two disjoint simplicial complexes, one consisting entirely of dry points.
-\end{claim}
-
-\InNotSoCGVer{
-\begin{proof} By \Thm{jordan}, $\cut(\MM,\lift(\phi))$ results in two disjoint simplicial complexes. Let $\NN$ be the complex containing
-the point $x$ (the argument in $\wet(x,\MM)$), and let $\NN'$ be the other complex. 
-Any path from $x$ to $\NN'$ must intersect $\lift(\phi)$, which is dry. Hence $\NN'$ is dry.
-%
-%
-%If there exists a wet point $z$ in $\NN'$, then there exists a descending path from $x$ to $z$. 
-%As $x$ is in $\NN$ and $z$ is in $\NN'$, again by the Jordan Curve Theorem this path must intersect $\phi$, 
-%contradicting the fact that $\phi$ is dry.
-\end{proof}
-}
-
-We describe the main partitioning procedure that cuts a simplicial complex $\NN$ into extremum
-dominant complexes. It takes an additional input of a maximum $x$. To initialize,
-we begin with $\NN$ set to $\MM$ and $x$ as an arbitrary maximum. 
-% One of the critical aspects
-% of the procedure is that the rain alternately flows downwards and upwards, 
-% since otherwise faces may be cut a super constant number of times (see \Fig{order}\InSoCGVer{ in \Sec{approach}}).
-When
-we start, rain flows downwards. In each recursive call, the direction of rain is \emph{switched} to the 
-opposite direction. This is crucial to ensure a linear running time. 
-The switching is easily implemented by inverting a complex $\NN'$, achieved by negating the height values.
-% While one can implement $\rain$ this way, it is conceptually easier
-% to think of \emph{inverting} a complex $\NN'$ when a recursive call is made. Inversion
-% is easily achieved by just negating the height values. (From a running time standpoint, it suffices
-% to maintain a single bit associated with $\NN'$ that determines whether heights are inverted or not.)
-We can now let rain flow downwards, as it usually does in our world.
-
-%
-%We critically perform
-%an \emph{inversion} step, which is simply achieved by negating
-%
-\medskip
-\fbox{
-\begin{minipage}{0.9\textwidth}
-{\bf $\rain(x,\NN)$}
-
-\smallskip
-\begin{compactenum}
-	\item Determine interface of $\wet(x,\NN)$. 
-	\item If the interface is empty, simply output $\NN$. Otherwise, denote the connected components by $\phi_1, \phi_2, \ldots, \phi_k$ and set $\phi'_i = \lift(\phi_i)$.
-	\item Initialize $\NN_1 = \NN$.
-	\item For $i$ from $1$ to $k$: 
-	\begin{compactenum}
-		\item Construct $\cut(\NN_i,\phi'_i)$, consisting of dry complex $\LL_i$ and remainder $\NN_{i+1}$.
-		\item Let the newly created boundary of $\LL_i$ be $B_i$. Invert $\LL_i$ so that $B_i$ is a maximum. Recursively
-		call $\rain(B_i,\LL_i)$.
-	\end{compactenum}
-	\item Output $\NN_{k+1}$ together with any complexes output by recursive calls.
-\end{compactenum}
-\end{minipage}}
-
-\medskip
-For convenience, denote the total output of $\rain(x,\MM)$ by $\MM_1, \MM_2, \ldots, \MM_r$.
-% We first argue the correctness of $\rain$.
-
-\begin{lemma} \label{lem:rain-1} Each output $\MM_i$ is extremum dominant.
-\end{lemma}
-
-\begin{proof} Consider a call to $\rain(x,\NN)$. If the interface is empty, then all of $\NN$
-is in $\wet(x,\NN)$, so $\NN$ is trivially extremum dominant. So suppose the interface
-is non-empty and consists of $\phi_1, \ldots, \phi_k$ (as denoted in the procedure).
-By repeated applications of \Clm{cut-int}, $\NN_{k+1}$ contains $\wet(x,\MM)$. 
-Consider $\wet(x,\NN_{k+1})$. The interface is exactly $\phi_1, \ldots, \phi_k$.
-So the only dry vertices of $\NN_{k+1}$ are those on the boundaries created by calling $\cut$ for each $\phi_i$, 
-but such boundaries are all maxima.
-%$B_1, \ldots, B_k$. But these boundaries are maxima.
-\end{proof}
-
-As $\rain(x,\MM)$ proceeds, new faces/simplices are created because of repeated cutting. 
-%Indeed, any facet $F$ in any simplicial complex $\NN$ that is recursively invoked, is achieved by cutting facet $F'$ in $\MM$.
-The key to the running time of $\rain(x,\MM)$ is bounding the number of newly created faces, for which we have the following lemma.
-
-\begin{lemma}\label{lem:new-verts}
-A face $F \in \MM$ is cut\footnote{What we are calling a single cut is actually done with two hyperplanes.} at most once during $\rain(x,\MM)$.
-\end{lemma}
-\begin{proof} Notation here follows the pseudocode of $\rain$.
-First, by \Thm{jordan}, all the pieces on which $\rain$ is invoked are disjoint.
-Second, all recursive calls are made on dry complexes.
-
-Consider the first time that $F$ is cut, say, during the call to $\rain(x,\NN)$.
-Specifically, say this happens when $\cut(\NN_i,\phi'_i)$ is constructed. 
-$\cut(\NN_i,\phi'_i)$ will cut $F$ with two horizontal cutting planes, one $\eps$ above $\phi'_i$
-and one $\eps$ below $\phi'_i$.  This breaks $F$ into lower and upper portions which are then triangulated
-(there is also a discarded middle portion).  The lower portion, which is adjacent to $\phi_i$, gets included 
-in $\NN_{k+1}$, the complex containing the wet points, and hence does not participate in any later recursive calls.  
-The upper portion (call it $U$) is in $\LL_i$. Note that the lower
-boundary of $U$ is in the boundary $B_i$. Since a recursive call is made to $\rain(B_i,\LL_i)$
-(and $\LL_i$ is inverted), $U$ becomes wet. Hence $U$, and correspondingly $F$, will not be subsequently cut.
-\end{proof}
-
-The following are direct consequences of \Lem{new-verts} and the $\surgery$ procedure.
-
-\begin{theorem} \label{thm:rain-time} The total running time of $\rain(x,\MM)$ is $O(|\MM|)$.
-\end{theorem}
-
-\InNotSoCGVer{
-\begin{proof} The only non-trivial operations performed are $\wet$ and $\cut$.  
-Since $\cut$ is a linear time procedure, Lemma~\ref{lem:new-verts} implies the total time for all calls to $\cut$ is $O(|\MM|)$.  
-As for the $\wet$ procedure, observe that Lemma~\ref{lem:new-verts} additionally implies there are only $O(|\MM|)$ new faces created by $\rain$.
-Therefore, since $\wet$ is also a linear time procedure, and no face is ever wet twice, the total time for all calls to $\wet$ is $O(|\MM|)$.
-\end{proof}
-}
-
-\begin{claim} \label{clm:rain-reeb} Given $\reeb(\MM_1), \reeb(\MM_2), \ldots, \reeb(\MM_r)$, 
-%be the output of $\rain(x,\MM)$. Suppose that for each reeb graph we know its corresponding vertex in the Manifold Ordering Tree. 
-$\reeb(\MM)$ can be constructed in $O(|\MM|)$ time.
-\end{claim}
-
-\InNotSoCGVer{
-\begin{proof} Consider the tree of recursive calls in $\rain(x,\MM)$, with each node
-labeled with some $\MM_i$.
-Walk through this tree in a leaf first ordering.  Each time we visit a node we connect its contour tree to the 
-contour tree of its children in the tree using the $\surgery$ procedure. 
-Each $\surgery$ call takes constant time, and the total time is the size of the recursion tree.
-\end{proof}
-}
-
-\section{Contour trees of extremum dominant complexes} \label{sec:extreme}
-
-The previous section allows us to restrict attention to extremum dominant complexes.
-We will orient so that the extremum in question is always a \emph{minimum}.
-We will fix such a simplicial complex $\MM$, with the dominant minimum $m^*$. 
-%The set of non-dominant minima is denoted by $M$.
-For vertex $v$, we use $\MM^+_v$ to denote the simplicial complex obtained by only
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+%\InSoCGVer{
+%\newintuitionSection
+%}
+
+% \InNotSoCGVer{
+% \intuitionSection
+% }
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\section{Painting to compute join trees}
+\label{sec:paint}
+
+We now present the main algorithmic contribution, which is a new algorithm for computing join trees of any triangulated simplicial
+complex $\MM$.  In subsequent sections we then show how to reduce computing the contour to computing join by using contour surgery.
+
+We now formally define the \emph{join} and \emph{split} trees as given by~\cite{csa-cctad-03}.
+Conventionally, all edges are directed from higher to lower function value. 
+In the following, for vertex $v$, we use $\MM^+_v$ to denote the simplicial complex obtained by only
 keeping vertices $u$ such that $f(u) > f(v)$. Analogously, define $\MM^-_v$. Note that $\MM^+_v$ may contain
@@
 
-\InNotSoCGVer{
-The main theorem of this section asserts that contour trees of minimum dominant complexes have a simple description.
-The exact statement will require some definitions and notation.
-We require the notions of \emph{join} and \emph{split} trees, as given by~\cite{csa-cctad-03}.
-Conventionally, all edges are directed from higher to lower function value. 
-}
-\InSoCGVer{
-The main theorem of this section asserts that contour trees of minimum dominant complexes have a simple description. 
-Intuitively, the cutting procedure of \Sec{rain} introduced ``stumps'' where we made cuts, which 
-is why we allowed for non-dominant minima and maxima in the definition of extremum dominant complexes.  
-The punchline is that, ignoring these stumps, the contour tree of an extremum dominant complex is equivalent to its \emph{join} (or \emph{split}) tree, defined in~\cite{csa-cctad-03}.
-Hence the critical join tree below is just the join tree without these stumps.  
-Additional definitions and proofs are given in the full version \CiteFullVer.
-}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\newcommand{\joinTreeSplitTreeMerge}{
 \begin{definition} \label{def:join} The \emph{join tree} $\cJ(\MM)$ of $\MM$ is built on vertex set $\cV(\MM)$.
@@
 Some basic facts about these trees. 
-%Note that unlike the contour tree, all vertices in $\MM$ are present in $\cJ(\MM)$ and $\cS(\MM)$. 
 All outdegrees in $\cJ(\MM)$ are at most $1$, all indegree $2$ vertices are joins, all leaves are maxima,
@@
 vertices are splits, all leaves are minima, and the global maximum is the root.
-%
 As these trees are rooted, we can use ancestor-descendant terminology.  
@@
 
-The key observation is that $\cS(\MM)$ is trivial for a minimum dominant $\MM$.
-
-\begin{lemma} \label{lem:split} For minimum dominant $\MM$, $\cS(\MM)$ consists of:
-\begin{asparaitem}
-	\item A single path (in sorted order) with all vertices except non-dominant minima.
-	\item Each non-dominant minimum is attached to a unique split (which is adjacent to it).
-\end{asparaitem}
-\end{lemma}
-
-\begin{proof} It suffices to prove that each split $v$ has one child that is just a leaf, which
-is a non-dominant minimum.  Specifically, any minimum is a leaf in $\cS(\MM)$ and thereby attached to a split, 
-which implies that if we removed all non-dominant minima, we must end up with a path, as asserted above.
-
-Consider a split $v$. For sufficiently small $\eps, \delta$, there are exactly two $(f(v) - \delta)$-contours
-$\phi$ and $\psi$ intersecting $B_\eps(v)$. Both of these are regular contours. There must be a non-ascending
-path from $v$ to the dominant minimum $m^*$. Consider the first edge (necessarily decreasing from $v$)
-on this path. It must intersect one of the $(f(v) - \delta)$-contours, say $\phi$. By \Thm{jordan}, $\cut(\MM,\phi)$ has
-two connected components, with one (call it $\LL$) having $\phi^-$ as a boundary maximum. This complex
-contains $m^*$ as the non-ascending path intersects $\phi$ only once. Let the other
-component be called $\MM'$.
-
-Consider $\cut(\MM',\psi)$ with connected component $\NN$ having $\psi^-$ as a boundary. $\NN$
-does not contain $m^*$, so any path from the interior of $\NN$ to $m^*$ must intersect the boundary $\psi^-$.
-But the latter is a maximum in $\NN$, so there can be no non-ascending path from the interior to $m^*$.
-Since $\MM$ is overall minimum dominant, the interior of $\NN$ can only contain a single vertex $w$, a non-dominant
-minimum.
-
-The split $v$ has two children in $\cS(\MM)$, one in $\NN$ and one in $\LL$. The child in $\NN$ can only
-be the non-dominant minimum $w$, which is a leaf. 
-\end{proof}
-
-It is convenient to denote the non-dominant minima as $m_1, m_2, \ldots, m_k$
-and the corresponding splits (as given by the lemma above) as $s_1, s_2, \ldots, s_k$.
-
-Using the above lemma we can now prove that computing the contour tree for a minimum dominant complex 
-amounts to computing its join tree.  Specifically, to prove our main theorem, we rely on the correctness of the
-merging procedure from~\cite{csa-cctad-03} that constructs the contour tree from the join and split trees. 
-It actually first constructs the \emph{augmented contour tree} $\cA(\MM)$, which is 
-obtained by replacing each edge in the contour tree with a path of all regular vertices 
-(sorted by height) whose corresponding contour belongs to the equivalence class of that edge.
-
-Consider a tree $T$ with a vertex $v$ of in and out degree at most $1$.
-\emph{Erasing} $v$ from $T$ is the following operation: if $v$ is a leaf, just delete $v$. Otherwise, 
-delete $v$ and connect its neighbors by an edge (i.e.\ smooth $v$ out). This tree is denoted by $T \ominus v$.
-%
-The merging procedure of \cite{csa-cctad-03} is shown below as $\merge(\cJ(\MM),\cS(\MM))$.  
-This procedure builds $\cA(\MM)$ by iteratively adding edges that are adjacent to leaves of either $\cJ(\MM)$ or $\cS(\MM)$, 
-erasing the corresponding vertex from both $\cJ(\MM)$ and $\cS(\MM)$, and repeating.  $\cC(\MM)$ is then produced by erasing all regular vertices in $\cA(\MM)$.
-We use the fact that this procedure correctly produces the contour tree (regardless of the order leaves are processed), 
-and refer the interested reader to \cite{csa-cctad-03} for the proof of this fact, as it is not needed for our purposes.
-
-\medskip
-\fbox{
-\begin{minipage}{0.9\textwidth}
-{\bf $\merge(\cJ(\MM),\cS(\MM))$}
-
-\smallskip
-\begin{compactenum}
-	\item Set $\cJ = \cJ(\MM)$ and $\cS = \cS(\MM)$.
-	\item Denote $v$ as a \emph{candidate} if the sum of its indegree in $\cJ$ and outdegree in $\cS$ is $1$. 
-	\item Add all candidates to queue.
-	\item While candidate queue is non-empty: 
-	\begin{compactenum}
-		\item Let $v$ be head of queue. If $v$ is leaf in $\cJ$, consider its edge in $\cJ$. Otherwise
-		consider its edge in $\cS$. In either case, denote the edge by $(v,w)$.
-		\item Insert $(v,w)$ in $\cA(\MM)$. 
-		\item Set $\cJ = \cJ \ominus v$ and $\cS = \cS \ominus v$. Enqueue any new candidates.
-	\end{compactenum}
-	\item Erase all regular vertices in $\cA(\MM)$ to get $\cC(\MM)$.
-\end{compactenum}
-\end{minipage}}
-
-\medskip
-}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\InNotSoCGVer{\joinTreeSplitTreeMerge}
-
-\begin{definition} 
-\label{def:criticalJoin}
-The \emph{critical join tree} $\jc(\MM)$ is built on the set $V'$ of all
-critical points other than the non-dominant minima. 
-The directed edge $(u,v)$ is present when $u$ is the smallest valued vertex in $V'$ in a connected component of $\MM^+_v$
-and $v$ is adjacent (in $\MM$) to a vertex in this component. \InSoCGVer{The \emph{join tree}, $\cJ(\MM)$, is defined analogously, but instead on $\cV(\MM)$.}
-\end{definition}
-
-\InSoCGVer{Each non-dominant minimum, $m_i$, connects to the contour tree at a corresponding split $s_i$. We have the following (see the full version \CiteFullVer for more details).}
-
-\begin{theorem} \label{thm:contour-tree} Let $\MM$ have a dominant minimum. 
-The contour tree $\cC(\MM)$ consists of all edges $\{(s_i, m_i)\}$ and $\jc(\MM)$.
-\end{theorem}
-
-\newcommand{\proofofContourJoinEquiv}{
-\begin{proof} %We first show that $\cA(\MM)$ is $\cJ(\MM) \ominus \{m_i\}$ with edges $\{(s_i,m_i)\}$.
-We have flexibility in choosing the order of processing in $\merge$. We first put the non-dominant
-minima $m_1, \ldots, m_k$ into the queue. As these are processed, the edges $\{(s_i,m_i)\}$ are inserted 
-into $\cA(\MM)$. Once all the $m_i$'s are erased, $\cS$ becomes a path, so all outdegrees are at most $1$.
-Moreover the join tree is now $\cJ(\MM) \ominus \{m_i\}$, and so we can now process $\cJ$ leaf by leaf, and all
-remaining edges of $\cJ$ are inserted into $\cA(\MM)$.
-%
-Note that smoothing out regular points from $\cJ(\MM) \ominus \{m_i\}$ yields the edges of $\jc(\MM)$.
-Thus after smoothing out all regular points from $\cA(\MM)$ we recover the edges of $\jc(\MM)$.  
-Hence the resulting contour tree contains these edges and the $\{(s_i,m_i)\}$ as claimed.
-%Note that $\cC(\MM)$ is obtained by smoothing out all regular points from $\cA(\MM)$. Similarly, smoothing out regular
-%points from $\cJ(\MM) \ominus \{m_i\}$ yields the edges of $\jc(\MM)$.
-\end{proof}
-}
-
-\InNotSoCGVer{\proofofContourJoinEquiv}
-
-\begin{Remark}\label{rem:joinAndCriticalJoin}
-The above theorem, combined with the previous sections, implies that in order to get an efficient contour tree algorithm, 
-it suffices to have an efficient algorithm for computing $\jc(\MM)$.
-In the other sections of the paper it is better to phrase things instead in terms of computing $\cJ(\MM)$ efficiently.  
-%Note however that for minimum dominant complexes output by $\rain$, converting between $\jc$ and $\cJ$ is trivial, as $\cJ$ is just $\jc$ 
-%with each non-dominant minimum $m_i$ augmented along the edge leaving $s_i$.
-Note however that for minimum dominant complexes, converting between $\jc$ and $\cJ$ is trivial, as $\cJ$ is just $\jc$ 
-with each non-dominant minimum $m_i$ augmented along the edge leaving $s_i$.  
-(Indeed $\jc(\MM)$ was only defined to make the above theorem statement clearer.)
-\end{Remark}
-
-\section{Painting to compute contour trees}
-\label{sec:paint}
-
-The main algorithmic contribution is a new algorithm for computing join trees of any triangulated simplicial
-complex $\MM$.
-
-{\bf Painting:} The central tool is a notion 
+\subsection{Painting}
+%{\bf Painting:} 
+The central tool is a notion 
 of \emph{painting} $\MM$. Initially associate a color with each maximum. Imagine there being a large
 can of paint of a distinct color at each maximum $x$. We will spill different paint from each maximum and watch it flow down.
-This is analogous to the raining of \Sec{rain}, but paint is a much more viscous liquid.
+%This is analogous to the raining of \Sec{rain}, but paint is a much more viscous liquid.
 \emph{So paint only flows down edges, and it does not color the interior of higher dimensional faces.} Furthermore, paints
@@
 works purely on the 1-skeleton of $\MM$, which is just a graph.
-%
-
-\InNotSoCGVer{We now restate \Def{paint1}.}
 
@@
 \end{definition}
-
-% A painting colors edges, not vertices. We associate certain
-% sets of colors with vertices.
 
@@
 
-% Non-maximum $v$ participate in at least $1$ and at most $2$ up-stars (the latter iff
-% $v$ a join).
-
 \subsection{The data structures} \label{sec:struct}
-
 \noindent
@@
 
-%\medskip
 \ignore{	
@@
 
-%\Ben{Current color is never defined, but it means the the current rep(c).}
-
 \begin{lemma} \label{lem:full} Suppose vertex $v$ is connected to a component $\PP$ of $\MM^+_v$
@@
 
-% The following crucial claim demonstrates the utility of the stack.
-% 
-% \begin{claim} \label{clm:stack} Vertices are always pushed onto $\stack$ in increasing order of height.
-% \end{claim}
-% 
-% \begin{proof} Consider a single push onto the stack and the current painting. We have $c \in \col(h)$, as given in pseudocode of $\update$.
-% If $h \in T(\rep(c))$, then obviously the next head is higher. So assume that $h \notin T(\rep(c))$.
-% There is obviously an edge of color $\rep(c)$ touching $h$ connected to a component $\PP$ of $\MM^+_v$. 
-% From the initialization, $h \in T(\rep(c'))$, for some $c'$ touching $h$ (where $c'$ is the color
-% of some edge connecting $h$ to $\PP$).
-% Applying \Lem{full},  either all edges in $\PP$ are colored $\rep(c)$ or 
-% there exists critical vertex $w \in \PP$ fully touched by $\rep(c)$ and another color.
-% In the former case, all edges connecting $\PP$ to $h$ are also colored $\rep(c)$
-% (since this is a painting). Hence, $\rep(c') = \rep(c)$ and $h \in T(\rep(c))$.
-% As we assumed this is not the case, we conclude the existence of critical vertex $w$.
-% Since $w$ is fully touched by $\rep(c)$, $w \in T(\rep(c))$. Since $w$
-% is touched by another color, $w$ has not been processed. So the highest of $T(\rep(c))$
-% is higher than $h$, as desired.
-% \ignore{
-% If all edges touching $h$ that connect to this component are also colored $\rep(c)$, then
-% In Case (i), the increase in height is obvious.
-% Consider Case (ii). There is a vertex $v$ touched by color $c \notin \fcol(v)$. Consider the highest
-% vertex $w$ that is touched by $c$ and some other color. Clearly such a vertex exists, since the condition
-% is also satisfied by $v$. Take the $(f(w)+\delta)$-contour $\phi$ that intersects $B_\eps(v)$
-% and intersect some $c$-colored edge incident to $w$. Note that $\phi$ is also colored $c$,
-% since $w$ is the highest vertex to be touched by $c$ and some other color. (Take the path of $c$-colored
-% edges from the maximum to $w$. For any point on this path, the contour passing through this point must
-% be colored $c$.) Hence, $c \in \fcol(w)$ and $w \neq v$. Furthermore, $w$ is a critical point,
-% since $w$ is touched by another color. The corresponding edge cannot intersect $\phi$, so $w$
-% must have up-degree $2$.
-% All in all, when a new vertex is pushed onto the stack, it is always higher than the previous head.
-% }
-% \end{proof}
-
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-%\newcommand{\algorithmRunTimeSection}{
-%\InNotSoCGVer{\subsection{Running Time}}
-%\InSoCGVer{\section{Running Time}}
-\InNotSoCGVer{
 \subsection{Running Time}
@@
 \end{lemma}
-}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\newcommand{\proofofRunTime}{
 \begin{proof}
@@
 \end{proof}
-}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\InNotSoCGVer{
-\proofofRunTime
-}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\InSoCGVer{
-\subsection{Upper Bounds for Running Time}\label{sec:runTime}
-
-The algorithm $\build(\MM)$ processes vertices in $\cJ(\MM)$ one at a time.  The main processing cost  
-comes from priority queue operations.  The cost of these operations is a function of the size of the queue which is in turn a function of the choices made by the subroutine $\init(\MM)$.
-
-\begin{definition}\label{def:initialColoring}
-A \emph{leaf assignment} $\chi$ of a binary tree $T$ assigns \emph{two} distinct leaves to each internal vertex $v$,
-one from the left and one from the right subtree of $v$ (or only one leaf if $v$ has only one child). % naturally if $v$ has only one child it is assigned only one color). 
-The subroutine $\init(\MM)$ naturally defines a leaf assignment to $\cJ(\MM)$ (which 
-is a rooted binary tree) according to the priority queue for each up-star we put a given vertex in.  
-Call this the \emph{initial coloring} of the vertices in $\cJ(\MM)$, and denote it by $\chi$.
-
-For a vertex $v$ in $\cJ(\MM)$, let $L(v)$ denote the set of leaves of the subtree rooted at $v$, and let $A(v)$ denote the set of 
-ancestors of $v$, i.e.\ the vertices on the $v$ to root path. For a vertex $v\in \cJ(\MM)$, and an initial coloring $\chi$, 
-we use $H_v$ to denote the \emph{heap} at $v$. Formally, $H_v = \{u | u \in A(v), \chi(u) \cap L(v) \neq \emptyset\}$, i.e.\
-the set of ancestors colored by some leaf in $L(v)$.
+
+% Note that the analog of \Thm{main-corr} for join trees is an easy corollary of the above lemma.
+% Specifically, as in the \Thm{main-corr} statement, for any critical point $v$ let $\ell_v$ 
+% denote the length of the longest directed path passing through $v$ in the join tree. 
+% As $|H_v|$ only counts vertices in a $v$ to root path, trivially $|H_v|\leq \ell_v$, implying the analog of \Thm{main-corr}.
+
+For any critical point $v$ let $\ell_v$ denote the length of the longest directed path passing through $v$ in the join tree. 
+As $|H_v|$ only counts vertices in a $v$ to root path, trivially $|H_v|\leq \ell_v$, immediately implying the analog of \Thm{main-corr} for join trees.
+Note however that there is fair amount of slack in this argument as $|H_v|$ may be significantly smaller than $\ell_v$.  
+This slack allows for the more refined upper and lower bounds mentioned in \Sec{more-refined}.  Quantifying this slack however is quite challenging, 
+and requires a significantly more sophisticated analysis involving path decompositions, which is the subject of \Sec{lb}.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\section{Divide and conquer through contour surgery} \label{sec:surgery}
+Now that we have an efficient join tree algorithm, we now describe how to 
+reduce computing the contour tree to computing the join tree, as discussed in \Sec{approach}.
+Specifically, in this section we first describe our basic cutting tool, which we call contour surgery.
+In \Sec{rain} we then describe how we use this tool to cut the surface into so called extremum dominant pieces 
+where computing the contour tree reduces to computing the join tree, and this equivalence is proven in \Sec{extreme}.
+
+{\bf The cutting operation:} We define a ``cut" operation on $f:\MM \rightarrow \RR$ that cuts along a regular contour to create
+a new simplicial complex with an added boundary. Given a contour $\phi$, roughly speaking, this constructs the simplicial complex $\MM \setminus \phi$. 
+We will always enforce the condition that $\phi$ never passes through a vertex of $\MM$.
+Again, we use $\eps$ for an infinitesimally small value. We denote $\phi^+$ (resp. $\phi^-$) to be
+the contour at value $f(\phi) + \eps$ (resp. $f(\phi) - \eps$), which is at distance $\eps$
+from $\phi$.
+
+An $h$-contour is achieved by intersecting $\MM$ with the hyperplane $x_{d+1} = h$ and taking a connected component. (Think of the $d+1$-dimension
+as height.) Given some point $x$ on an $h$-contour $\phi$, we can walk along $\MM$ from $x$ to determine $\phi$.
+We can ``cut" along $\phi$ to get a new (possibly) disconnected simplicial complex $\MM'$. This is achieved
+by splitting every face $F$ that $\phi$ intersects into an ``upper" face and ``lower" face. Algorithmically,
+we cut $F$ with $\phi^+$ and take everything above $\phi^+$ in $F$ to make the upper face. Analogously, we cut with $\phi^-$ to get the lower face.
+The faces are then triangulated to ensure that they are all simplices. 
+\Ben{Reviewer wanted clarification on how triangulation is done. Maybe can use bottom-vertex triangulation of Clarkson 88, discussed in Matousek book?}
+This creates the two new boundaries $\phi^+$ and $\phi^-$, and we maintain the property of constant $f$-value at a boundary.
+
+Note that by assumption $\phi$ cannot cut a boundary face, and moreover all non-boundary faces
+have constant size. Therefore, this process takes time linear in $|\phi|$, the number of faces $\phi$ intersects.
+This new simplicial complex is denoted by $\cut(\phi,\MM)$.
+%
+We now describe a high-level approach to construct $\reeb(\MM)$ using this cutting procedure.
+
+\Ben{Surgery should be rewritten to take instead take as input the output of cut, i.e. two complexes and a contour.  
+This is how it is eventually used in Claim\ref{clm:rain-reeb}. Currently it does nothing really.}
+
+\medskip
+\fbox{
+\begin{minipage}{0.9\textwidth}
+{\bf $\surgery(\MM,\phi)$}
+
+\smallskip
+\begin{asparaenum}
+	\item Let $\MM' = \cut(\MM,\phi)$.
+	\item Construct $\reeb(\MM')$ and let $A, B$ be the nodes corresponding to the new boundaries created
+	in $\MM'$. (One is a minimum and the other is maximum.)
+	\item Since $A, B$ are leaves, they each have unique neighbors $A'$ and $B'$, respectively. Insert
+	edge $(A',B')$ and delete $A, B$ to obtain $\reeb(\MM)$.
+\end{asparaenum}
+\end{minipage}}
+
+\medskip
+
+\begin{theorem} \label{thm:surgery} For any regular contour $\phi$, the output of $\surgery(\MM,\phi)$ is $\reeb(\MM)$.
+\end{theorem}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+To prove Theorem~\ref{thm:surgery}, we require a theorem from \cite{c-tmi-04} (Theorems 6.6) that map paths in $\cC(\MM)$ to $\MM$.
+
+\begin{theorem} \label{thm:carr-path} For every path $P$ in $\MM$, there exists a path $Q$ in the contour tree corresponding
+to the contours passing through points in $P$.
+For every path $Q$ in the contour tree, there exists at least one path $P$ in $\MM$ through points present
+in contours involving $Q$.
+
+In particular, for every monotone path $P$ in $\MM$, there exists a monotone path $Q$ in the contour tree to which $P$
+maps, and vice versa.
+\end{theorem}
+
+Theorem~\ref{thm:surgery} is a direct consequence of the following lemma.
+
+\begin{lemma} \label{lem:cut} Consider a regular contour $\phi$ contained in a contour class (of an edge of $\reeb(\MM))$
+$(u,v)$ and let $\MM' = \cut(\MM,\phi)$. Then $\cV(\reeb(\MM')) = \{\phi^+,\phi^-\} \cup \cV(\MM)$
+and $\cE(\reeb(\MM')) = \{(u,\phi^+),(\phi^-,v)\} \cup (\cE(\MM) \setminus (u,v))$.
+\end{lemma}
+
+\begin{proof} First observe that since $\phi$ is a regular contour, the vertex set in the complex $\MM'$ is the same 
+as the vertex set in $\MM$, except with the addition of the newly created vertices on $\phi^+$ and $\phi^-$.  
+Moreover, $\cut(\MM,\phi)$ does not affect the local neighborhood of any vertex in $\MM$. 
+Therefore since a vertex being critical is a local condition, with the exception of new boundary vertices, the 
+critical vertices in $\MM$ and $\MM'$ are the same.  Finally, the new vertices on $\phi^+$ and $\phi^-$ 
+collectively behave as a minimum and maximum, respectively, and so $\cV(\reeb(\MM')) = \{\phi^+,\phi^-\} \cup \cV(\MM)$.
+
+Now consider the edge sets of the contour trees.  Any contour class in $\MM'$ (i.e.\ edge in $\cC(\MM')$) that does not involve $\phi^+$ or $\phi^-$ is also
+a contour class in $\MM$. Furthermore, a maximal contour class satisfying these properties is also
+maximal in $\MM$. So all edges of $\cC(\MM')$ that do not involve $\phi^+$ or $\phi^-$ are edges of $\cC(\MM)$.
+Analogously, every edge of $\cC(\MM)$ not involving $\phi$ is an edge of $\cC(\MM')$.
+
+Consider the contour class corresponding to edge $(u,v)$ of $\cC(\MM)$. There is a natural ordering
+of the contours by function value, ranging from $f(u)$ to $f(v)$. All contours in this class ``above" $\phi$
+form a maximal contour class in $\MM'$, represented by edge $(u,\phi^+)$. Analogously, there is another
+contour class represented by edge $(\phi^-,v)$. We have now accounted for all contours in $\cC(\MM')$,
+completing the proof.
+\end{proof}
+
+A useful corollary of this lemma shows that a contour actually splits the simplicial complex into
+two disconnected complexes.
+
+\begin{theorem} \label{thm:jordan} $\cut(\MM,\phi)$ consists of two
+disconnected simplicial complexes.
+\end{theorem}
+
+\begin{proof} Denote (as in \Lem{cut}) the edge containing $\phi$ to be $(u,v)$. Suppose for contradiction that there is a path between vertices $u$ and $v$
+in $\MM' = \cut(\MM,\phi)$. By \Thm{carr-path}, there is a path in $\cC(\MM')$ between $u$ and $v$. Since $\phi^+$ and $\phi^-$
+are leaves in $\cC(\MM')$, this path cannot use their incident edges. Therefore by \Lem{cut},
+all the edges of this path are in $\cE(\cC(\MM)) \setminus (u,v)$. So we get a cycle in $\cC(\MM)$, a contradiction.
+To show that there are exactly two connected components in $\cut(\MM,\phi)$, it suffices
+to see that $\cC(\MM')$ has two connected components (by \Lem{cut}) and then applying \Thm{carr-path}.
+\end{proof}
+
+% 
+% \begin{lemma} \label{lem:blahblah}
+%  Testing 123 \RefProofInAppendix{blahblah}
+% \end{lemma}
+% 
+% \begin{proof:in:appendix:e}{\ref{lem:blahblah}}{blahblah}
+%  blah balh
+% \end{proof:in:appendix:e}
+
+\section{Raining to partition $\MM$} \label{sec:rain}
+
+In this section, we describe a linear time procedure that partitions $\MM$ into special
+\emph{extremum dominant} simplicial complexes.
+
+\begin{definition} \label{def:dom} A simplicial complex with a height function is \emph{minimum dominant} if there exists
+a minimum $x$ such that every non-minimal \emph{vertex} in the complex has a non-ascending path to $x$.
+Analogously define \emph{maximum dominant}. 
 \end{definition}
 
-Given the above technical definition, the proof of the following lemma is straightforward, though long and so has been moved to the full version \CiteFullVer.
-
-\begin{lemma}\label{lem:runTimeUpper}
-Let $\MM$ be a simplicial complex with $t$ critical points.  
-The running time of $\build(\MM)$ is $O(N+t\alpha(t) + \sum_{v \in \cJ(\MM)} \log |H_v|)$, where $H_v$ is defined by an initial coloring.
+The first aspect of the partitioning is ``raining''. Start at some point $x \in \MM$ and imagine rain at $x$.
+The water will flow downwards along non-ascending paths and ``wet'' all the points encountered. Note that this procedure considers all points
+of the complex, not just vertices.
+
+\begin{definition} \label{def:wet} Fix $x \in \MM$. The set of points $y \in \MM$ such that there is a non-ascending path from $x$ to $y$
+is denoted by $\wet(x,\MM)$ (which in turn is represented as a simplicial complex). A point $z$ is at the \emph{interface} of $\wet(x,\MM)$ if every neighborhood of $z$
+has non-trivial intersection with $\wet(x,\MM)$ (i.e.\ the intersection is neither empty nor the entire neighborhood).
+\end{definition}
+
+The following claim gives a description of the interface.
+
+\begin{claim} \label{clm:inter} For any $x$, each non-empty component of the interface of $\wet(x,\MM)$ contains a join vertex.
+\end{claim}
+
+\begin{proof} If $p \in \wet(x,\MM)$, all the points in any contour containing $p$ are also in $\wet(x,\MM)$.
+ (Follow the non-ascending path from $x$ to $p$ and then walk along the contour.) The converse is also true,
+ so $\wet(x,\MM)$ contains entire contours.
+
+Let $\eps, \delta$ be sufficiently small as usual. Fix some $y$ at the interface.
+Note that $y \in \wet(x,\MM)$. (Otherwise, $B_\eps(y)\cap \MM$ is dry.)
+The points in $B_\eps(y)\cap \MM$ that lie below $y$ have a descending path from $y$ and hence must be wet.
+There must also be a dry point in $B_\eps(y)\cap \MM$ that is above $y$, and hence,
+there exists a dry, regular $(f(y)+\delta)$-contour $\phi$ intersecting $B_\eps(y)$.
+
+Let $\Gamma_y$ be the contour containing $y$.  Suppose for contradiction
+that $\forall p \in \Gamma_y$, $p$ has up-degree $1$ (see \Def{deg}). Consider the non-ascending path from $x$ to $y$ and let $z$
+be the first point of $\Gamma_y$ encountered. There exists a wet, regular $(f(y) + \delta)$-contour $\psi$ 
+intersecting $B_\eps(z)$. Now, walk from $z$ to $y$ along $\Gamma_y$. If all points $w$ in this walk
+have up-degree $1$, then $\psi$ is the unique $(f(y)+\delta)$-contour
+intersecting $B_\eps(w)$. This would imply that $\phi = \psi$, contradicting the fact that $\psi$ is wet
+and $\phi$ is dry.
+%
+% Therefore, there must exist a join vertex $w$ in $\Gamma_y$, such that $B_\eps(w)$ intersects
+% both $\phi$ and $\psi$. As $\delta, \eps \rightarrow 0^+$, the limit of $\phi$ (which is dry)
+% lies in $\Gamma_y$ (which is wet). 
+% Hence this limit gives a contour of the interface containing the join $w$.
+\end{proof}
+
+Note that $\wet(x,\MM)$ (and its interface) can be computed in time linear in the size of the wet simplicial complex.
+We perform a non-ascending search from $x$. Any face $F$ of $\MM$ encountered is partially (if not entirely) in $\wet(x,\MM)$.
+The wet portion is determined by cutting $F$ along the interface. Since each component of the interface is a contour, this is equivalent
+to locally cutting $F$ by a hyperplane. All these operations can be performed to output $\wet(x,\MM)$ in time linear in $|\wet(x,\MM)|$.
+
+We define a simple \lift{} operation on the interface components. Consider such a component $\phi$ containing
+a join vertex $y$. Take any dry increasing edge incident to $y$, and pick the point $z$ on this edge at height
+$f(y) + \delta$ (where $\delta$ is an infinitesimal, but larger than the value $\eps$ used in the definition of $\cut$). 
+Let $\lift(\phi)$ be the unique contour through the regular point $z$. Note that $\lift(\phi)$ is dry.
+The following claim follows directly from \Thm{jordan}.
+
+\begin{claim} \label{clm:cut-int} Let $\phi$ be a connected component of the interface. Then $\cut(\MM,\lift(\phi))$
+results in two disjoint simplicial complexes, one consisting entirely of dry points.
+\end{claim}
+
+\begin{proof} By \Thm{jordan}, $\cut(\MM,\lift(\phi))$ results in two disjoint simplicial complexes. Let $\NN$ be the complex containing
+the point $x$ (the argument in $\wet(x,\MM)$), and let $\NN'$ be the other complex. 
+Any path from $x$ to $\NN'$ must intersect $\lift(\phi)$, which is dry. Hence $\NN'$ is dry.
+%
+%
+%If there exists a wet point $z$ in $\NN'$, then there exists a descending path from $x$ to $z$. 
+%As $x$ is in $\NN$ and $z$ is in $\NN'$, again by the Jordan Curve Theorem this path must intersect $\phi$, 
+%contradicting the fact that $\phi$ is dry.
+\end{proof}
+
+We describe the main partitioning procedure that cuts a simplicial complex $\NN$ into extremum
+dominant complexes. It takes an additional input of a maximum $x$. To initialize,
+we begin with $\NN$ set to $\MM$ and $x$ as an arbitrary maximum. 
+% One of the critical aspects
+% of the procedure is that the rain alternately flows downwards and upwards, 
+% since otherwise faces may be cut a super constant number of times (see \Fig{order}\InSoCGVer{ in \Sec{approach}}).
+When we start, rain flows downwards. In each recursive call, the direction of rain is \emph{switched} to the 
+opposite direction. This is crucial to ensure a linear running time, and \Fig{order} shows how failing to switch 
+directions can lead to a blow up in complexity. 
+The switching is easily implemented by inverting a complex $\NN'$, achieved by negating the height values.
+% While one can implement $\rain$ this way, it is conceptually easier
+% to think of \emph{inverting} a complex $\NN'$ when a recursive call is made. Inversion
+% is easily achieved by just negating the height values. (From a running time standpoint, it suffices
+% to maintain a single bit associated with $\NN'$ that determines whether heights are inverted or not.)
+We can now let rain flow downwards, as it usually does in our world.
+
+\begin{figure}[h]\centering
+    \includegraphics[width=.172\linewidth]{figs/tent2}%
+    \hspace{1.2in}
+    \includegraphics[width=.172\linewidth]{figs/tent1}
+    \caption{On left, downward rain spilling only (each shade of gray represents a piece created by each different spilling), producing a grid.  
+    Note we are assuming raining was done in sorted order of the maxima (i.e.\ lowest to highest).  
+    On right, flipping the direction of rain spilling.}
+    \label{fig:order}
+\end{figure}
+
+\medskip
+\fbox{
+\begin{minipage}{0.9\textwidth}
+{\bf $\rain(x,\NN)$}
+
+\smallskip
+\begin{compactenum}
+	\item Determine interface of $\wet(x,\NN)$. 
+	\item If the interface is empty, simply output $\NN$. Otherwise, denote the connected components by $\phi_1, \phi_2, \ldots, \phi_k$ and set $\phi'_i = \lift(\phi_i)$.
+	\item Initialize $\NN_1 = \NN$.
+	\item For $i$ from $1$ to $k$: 
+	\begin{compactenum}
+		\item Construct $\cut(\NN_i,\phi'_i)$, consisting of dry complex $\LL_i$ and remainder $\NN_{i+1}$.
+		\item Let the newly created boundary of $\LL_i$ be $B_i$. Invert $\LL_i$ so that $B_i$ is a maximum. Recursively
+		call $\rain(B_i,\LL_i)$.
+	\end{compactenum}
+	\item Output $\NN_{k+1}$ together with any complexes output by recursive calls.
+\end{compactenum}
+\end{minipage}}
+
+\medskip
+For convenience, denote the total output of $\rain(x,\MM)$ by $\MM_1, \MM_2, \ldots, \MM_r$.
+% We first argue the correctness of $\rain$.
+
+\begin{lemma} \label{lem:rain-1} Each output $\MM_i$ is extremum dominant.
 \end{lemma}
-}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\InSoCGVer{Our main result, }\Thm{main-corr}\InSoCGVer{,} is an easy corollary of the above lemma.
+
+\begin{proof} Consider a call to $\rain(x,\NN)$. If the interface is empty, then all of $\NN$
+is in $\wet(x,\NN)$, so $\NN$ is trivially extremum dominant. So suppose the interface
+is non-empty and consists of $\phi_1, \ldots, \phi_k$ (as denoted in the procedure).
+By repeated applications of \Clm{cut-int}, $\NN_{k+1}$ contains $\wet(x,\MM)$. 
+Consider $\wet(x,\NN_{k+1})$. The interface is exactly $\phi_1, \ldots, \phi_k$.
+So the only dry vertices of $\NN_{k+1}$ are those on the boundaries created by calling $\cut$ for each $\phi_i$, 
+but such boundaries are all maxima.
+%$B_1, \ldots, B_k$. But these boundaries are maxima.
+\end{proof}
+
+As $\rain(x,\MM)$ proceeds, new faces/simplices are created because of repeated cutting. 
+%Indeed, any facet $F$ in any simplicial complex $\NN$ that is recursively invoked, is achieved by cutting facet $F'$ in $\MM$.
+The key to the running time of $\rain(x,\MM)$ is bounding the number of newly created faces, for which we have the following lemma.
+
+\begin{lemma}\label{lem:new-verts}
+A face $F \in \MM$ is cut\footnote{What we are calling a single cut is actually done with two hyperplanes.} at most once during $\rain(x,\MM)$.
+\end{lemma}
+\begin{proof} Notation here follows the pseudocode of $\rain$.
+First, by \Thm{jordan}, all the pieces on which $\rain$ is invoked are disjoint.
+Second, all recursive calls are made on dry complexes.
+
+Consider the first time that $F$ is cut, say, during the call to $\rain(x,\NN)$.
+Specifically, say this happens when $\cut(\NN_i,\phi'_i)$ is constructed. 
+$\cut(\NN_i,\phi'_i)$ will cut $F$ with two horizontal cutting planes, one $\eps$ above $\phi'_i$
+and one $\eps$ below $\phi'_i$.  This breaks $F$ into lower and upper portions which are then triangulated
+(there is also a discarded middle portion).  The lower portion, which is adjacent to $\phi_i$, gets included 
+in $\NN_{k+1}$, the complex containing the wet points, and hence does not participate in any later recursive calls.  
+The upper portion (call it $U$) is in $\LL_i$. Note that the lower
+boundary of $U$ is in the boundary $B_i$. Since a recursive call is made to $\rain(B_i,\LL_i)$
+(and $\LL_i$ is inverted), $U$ becomes wet. Hence $U$, and correspondingly $F$, will not be subsequently cut.
+\end{proof}
+
+The following are direct consequences of \Lem{new-verts} and the $\surgery$ procedure.
+
+\begin{theorem} \label{thm:rain-time} The total running time of $\rain(x,\MM)$ is $O(|\MM|)$.
+\end{theorem}
+\begin{proof} The only non-trivial operations performed are $\wet$ and $\cut$.  
+Since $\cut$ is a linear time procedure, Lemma~\ref{lem:new-verts} implies the total time for all calls to $\cut$ is $O(|\MM|)$.  
+As for the $\wet$ procedure, observe that Lemma~\ref{lem:new-verts} additionally implies there are only $O(|\MM|)$ new faces created by $\rain$.
+Therefore, since $\wet$ is also a linear time procedure, and no face is ever wet twice, the total time for all calls to $\wet$ is $O(|\MM|)$.
+\end{proof}
+
+\begin{claim} \label{clm:rain-reeb} Given $\reeb(\MM_1), \reeb(\MM_2), \ldots, \reeb(\MM_r)$, 
+%be the output of $\rain(x,\MM)$. Suppose that for each reeb graph we know its corresponding vertex in the Manifold Ordering Tree. 
+$\reeb(\MM)$ can be constructed in $O(|\MM|)$ time.
+\end{claim}
+\begin{proof} Consider the tree of recursive calls in $\rain(x,\MM)$, with each node
+labeled with some $\MM_i$.
+Walk through this tree in a leaf first ordering.  Each time we visit a node we connect its contour tree to the 
+contour tree of its children in the tree using the $\surgery$ procedure. 
+Each $\surgery$ call takes constant time, and the total time is the size of the recursion tree.
+\end{proof}
+
+\section{Contour trees of extremum dominant complexes} \label{sec:extreme}
+
+The previous section allows us to restrict attention to extremum dominant complexes.
+We will orient so that the extremum in question is always a \emph{minimum}.
+We will fix such a simplicial complex $\MM$, with the dominant minimum $m^*$.
+
+% For vertex $v$, we use $\MM^+_v$ to denote the simplicial complex obtained by only
+% keeping vertices $u$ such that $f(u) > f(v)$. Analogously, define $\MM^-_v$. Note that $\MM^+_v$ may contain
+% numerous connected components. 
+%
+%The main theorem of this section asserts that contour trees of minimum dominant complexes have a simple description.
+%The exact statement will require some definitions and notation.
+% We require the notions of \emph{join} and \emph{split} trees, as given by~\cite{csa-cctad-03}.
+% Conventionally, all edges are directed from higher to lower function value. 
+% 
+% \begin{definition} \label{def:join} The \emph{join tree} $\cJ(\MM)$ of $\MM$ is built on vertex set $\cV(\MM)$.
+% The directed edge $(u,v)$ is present when $u$ is the smallest valued vertex in a connected component of $\MM^+_v$
+% \emph{and} $v$ is adjacent to a vertex in this component (in $\MM$). The \emph{split tree} $\cS(\MM)$ is obtained
+% by looking at $\MM^-_v$ (or alternatively, by taking the join tree of the inversion of $\MM$). 
+% \end{definition}
+% 
+% Some basic facts about these trees. 
+% All outdegrees in $\cJ(\MM)$ are at most $1$, all indegree $2$ vertices are joins, all leaves are maxima,
+% and the global minimum is the root. All indegrees in $\cS(\MM)$ are at most $1$, all outdegree $2$
+% vertices are splits, all leaves are minima, and the global maximum is the root.
+% As these trees are rooted, we can use ancestor-descendant terminology.  
+% Specifically, for two adjacent vertices $u$ and $v$, $u$ is the parent of $v$ if $u$ is closer to the root 
+% (i.e.\ each node can have at most one parent, but can have two children).
+
+Recall the definitions of join and split trees from \Sec{paint}.
+The main theorem of this section asserts that contour trees of minimum dominant complexes have a simple description in terms of $\cJ(\MM)$.  
+First we make the key observation that $\cS(\MM)$ is trivial for a minimum dominant $\MM$.
+
+\begin{lemma} \label{lem:split} For minimum dominant $\MM$, $\cS(\MM)$ consists of:
+\begin{asparaitem}
+	\item A single path (in sorted order) with all vertices except non-dominant minima.
+	\item Each non-dominant minimum is attached to a unique split (which is adjacent to it).
+\end{asparaitem}
+\end{lemma}
+
+\begin{proof} It suffices to prove that each split $v$ has one child that is just a leaf, which
+is a non-dominant minimum.  Specifically, any minimum is a leaf in $\cS(\MM)$ and thereby attached to a split, 
+which implies that if we removed all non-dominant minima, we must end up with a path, as asserted above.
+
+Consider a split $v$. For sufficiently small $\eps, \delta$, there are exactly two $(f(v) - \delta)$-contours
+$\phi$ and $\psi$ intersecting $B_\eps(v)$. Both of these are regular contours. There must be a non-ascending
+path from $v$ to the dominant minimum $m^*$. Consider the first edge (necessarily decreasing from $v$)
+on this path. It must intersect one of the $(f(v) - \delta)$-contours, say $\phi$. By \Thm{jordan}, $\cut(\MM,\phi)$ has
+two connected components, with one (call it $\LL$) having $\phi^-$ as a boundary maximum. This complex
+contains $m^*$ as the non-ascending path intersects $\phi$ only once. Let the other
+component be called $\MM'$.
+
+Consider $\cut(\MM',\psi)$ with connected component $\NN$ having $\psi^-$ as a boundary. $\NN$
+does not contain $m^*$, so any path from the interior of $\NN$ to $m^*$ must intersect the boundary $\psi^-$.
+But the latter is a maximum in $\NN$, so there can be no non-ascending path from the interior to $m^*$.
+Since $\MM$ is overall minimum dominant, the interior of $\NN$ can only contain a single vertex $w$, a non-dominant
+minimum.
+
+The split $v$ has two children in $\cS(\MM)$, one in $\NN$ and one in $\LL$. The child in $\NN$ can only
+be the non-dominant minimum $w$, which is a leaf. 
+\end{proof}
+
+It is convenient to denote the non-dominant minima as $m_1, m_2, \ldots, m_k$
+and the corresponding splits (as given by the lemma above) as $s_1, s_2, \ldots, s_k$.
+
+Using the above lemma we can now prove that computing the contour tree for a minimum dominant complex 
+amounts to computing its join tree.  Specifically, to prove our main theorem, we rely on the correctness of the
+merging procedure from~\cite{csa-cctad-03} that constructs the contour tree from the join and split trees. 
+It actually first constructs the \emph{augmented contour tree} $\cA(\MM)$, which is 
+obtained by replacing each edge in the contour tree with a path of all regular vertices 
+(sorted by height) whose corresponding contour belongs to the equivalence class of that edge.
+
+Consider a tree $T$ with a vertex $v$ of in and out degree at most $1$.
+\emph{Erasing} $v$ from $T$ is the following operation: if $v$ is a leaf, just delete $v$. Otherwise, 
+delete $v$ and connect its neighbors by an edge (i.e.\ smooth $v$ out). This tree is denoted by $T \ominus v$.
+%
+The merging procedure of \cite{csa-cctad-03} is shown below as $\merge(\cJ(\MM),\cS(\MM))$.  
+This procedure builds $\cA(\MM)$ by iteratively adding edges that are adjacent to leaves of either $\cJ(\MM)$ or $\cS(\MM)$, 
+erasing the corresponding vertex from both $\cJ(\MM)$ and $\cS(\MM)$, and repeating.  $\cC(\MM)$ is then produced by erasing all regular vertices in $\cA(\MM)$.
+We use the fact that this procedure correctly produces the contour tree (regardless of the order leaves are processed), 
+and refer the interested reader to \cite{csa-cctad-03} for the proof of this fact, as it is not needed for our purposes.
+
+\medskip
+\fbox{
+\begin{minipage}{0.9\textwidth}
+{\bf $\merge(\cJ(\MM),\cS(\MM))$}
+
+\smallskip
+\begin{compactenum}
+	\item Set $\cJ = \cJ(\MM)$ and $\cS = \cS(\MM)$.
+	\item Denote $v$ as a \emph{candidate} if the sum of its indegree in $\cJ$ and outdegree in $\cS$ is $1$. 
+	\item Add all candidates to queue.
+	\item While candidate queue is non-empty: 
+	\begin{compactenum}
+		\item Let $v$ be head of queue. If $v$ is leaf in $\cJ$, consider its edge in $\cJ$. Otherwise
+		consider its edge in $\cS$. In either case, denote the edge by $(v,w)$.
+		\item Insert $(v,w)$ in $\cA(\MM)$. 
+		\item Set $\cJ = \cJ \ominus v$ and $\cS = \cS \ominus v$. Enqueue any new candidates.
+	\end{compactenum}
+	\item Erase all regular vertices in $\cA(\MM)$ to get $\cC(\MM)$.
+\end{compactenum}
+\end{minipage}}
+
+\medskip
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\begin{definition} 
+\label{def:criticalJoin}
+The \emph{critical join tree} $\jc(\MM)$ is built on the set $V'$ of all
+critical points other than the non-dominant minima. 
+The directed edge $(u,v)$ is present when $u$ is the smallest valued vertex in $V'$ in a connected component of $\MM^+_v$
+and $v$ is adjacent (in $\MM$) to a vertex in this component. 
+\end{definition}
+
+\begin{theorem} \label{thm:contour-tree} Let $\MM$ have a dominant minimum. 
+The contour tree $\cC(\MM)$ consists of all edges $\{(s_i, m_i)\}$ and $\jc(\MM)$.
+\end{theorem}
+\begin{proof} %We first show that $\cA(\MM)$ is $\cJ(\MM) \ominus \{m_i\}$ with edges $\{(s_i,m_i)\}$.
+We have flexibility in choosing the order of processing in $\merge$. We first put the non-dominant
+minima $m_1, \ldots, m_k$ into the queue. As these are processed, the edges $\{(s_i,m_i)\}$ are inserted 
+into $\cA(\MM)$. Once all the $m_i$'s are erased, $\cS$ becomes a path, so all outdegrees are at most $1$.
+Moreover the join tree is now $\cJ(\MM) \ominus \{m_i\}$, and so we can now process $\cJ$ leaf by leaf, and all
+remaining edges of $\cJ$ are inserted into $\cA(\MM)$.
+%
+Note that smoothing out regular points from $\cJ(\MM) \ominus \{m_i\}$ yields the edges of $\jc(\MM)$.
+Thus after smoothing out all regular points from $\cA(\MM)$ we recover the edges of $\jc(\MM)$.  
+Hence the resulting contour tree contains these edges and the $\{(s_i,m_i)\}$ as claimed.
+%Note that $\cC(\MM)$ is obtained by smoothing out all regular points from $\cA(\MM)$. Similarly, smoothing out regular
+%points from $\cJ(\MM) \ominus \{m_i\}$ yields the edges of $\jc(\MM)$.
+\end{proof}
+
+\begin{Remark}\label{rem:joinAndCriticalJoin}
+% The above theorem, combined with the previous sections, implies that in order to get an efficient contour tree algorithm, 
+% it suffices to have an efficient algorithm for computing $\jc(\MM)$.
+% In the other sections of the paper it is better to phrase things instead in terms of computing $\cJ(\MM)$ efficiently.  
+% %Note however that for minimum dominant complexes output by $\rain$, converting between $\jc$ and $\cJ$ is trivial, as $\cJ$ is just $\jc$ 
+% %with each non-dominant minimum $m_i$ augmented along the edge leaving $s_i$.
+% Note however that for minimum dominant complexes, converting between $\jc$ and $\cJ$ is trivial, as $\cJ$ is just $\jc$ 
+% with each non-dominant minimum $m_i$ augmented along the edge leaving $s_i$.  
+% (Indeed $\jc(\MM)$ was only defined to make the above theorem statement clearer.)
+The above theorem, combined with the previous sections, implies that in order to get an efficient contour tree algorithm, 
+it suffices to have an efficient algorithm for computing $\jc(\MM)$.
+Note however that for minimum dominant complexes, converting between $\jc$ and $\cJ$ is trivial, 
+as $\cJ$ is just $\jc$ with each non-dominant minimum $m_i$ augmented along the edge leaving $s_i$. 
+(Indeed $\jc(\MM)$ was only defined to make the above theorem statement clearer.)  
+Thus to get an efficient contour tree algorithm it suffices to have an efficient algorithm for computing $\cJ(\MM)$.
+\end{Remark}
+
+\subsection{Putting it all together}
+We now have all the pieces required to prove \Thm{main-corr}.
 \begin{proof}[Proof of \Thm{main-corr}]
@@
 
-Note that there is fair amount of slack in this argument as $|H_v|$ may be significantly smaller than $\ell_v'$.  
-This slack allows for the more refined upper and lower bounds mentioned in \Sec{more-refined}.  Quantifying this slack however is quite challenging, 
-and requires a significantly more sophisticated analysis involving path decompositions, which is provided in the next section.
+Ultimately our goal is to prove the more refined upper and lower bounds mentioned in \Sec{more-refined}.  
+To achieve this, in the next section we give a careful analysis of how heap sizes evolve over the course of our join tree algorithm, 
+rather than using the blunt bound $|H_v|\leq \ell_v'$, as done in the above proof.
 
@@
 
-\newcommand{\leafAssignPathDecomp}{
 \section{Leaf assignments and path decompositions}
@@
 
-In this section, we set up a framework to analyze the time taken to compute a join tree $\cJ(\MM)$ (see \InSoCGVer{\Def{criticalJoin}}\InNotSoCGVer{\Def{join}}).
+In this section, we set up a framework to analyze the time taken to compute a join tree $\cJ(\MM)$ (see \Def{join}), and consequently for contour trees by the above discussion.
 We adopt all notation already defined in \Sec{runTime}.  From here forward we will often assume binary trees are full binary trees 
@@
 compute the contour tree in time
-\InSoCGVer{
-\begin{align*}
-\hspace{.8in} O\pth{N+ \sum_{i=1}^k \pth{\sum_{p \in \pmax(\cJ(\MM_i))} |p|\log |p|} + t_i\alpha(t_i) + N_i}& \\
-=\break O\pth{\pth{\sum_{p \in \pmax(\cC(\MM))} |p|\log |p|} + t\alpha(t) + N}&
-\end{align*}
-}
-\InNotSoCGVer{
 \[
@@
 \]
-}
 
@@
 \end{proof}
-}
 
@@
 
-\InNotSoCGVer{
-\leafAssignPathDecomp
-}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\newcommand{\lowerBoundbyPathDecomp}{
 \section{Lower Bound by Path Decomposition}
@@
 \end{Remark}
-}
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\InNotSoCGVer{
-\lowerBoundbyPathDecomp
-}
 
@@
 
-\InNotSoCGVer{
 \bibliographystyle{alpha}
-%\bibliographystyle{plain}
-}
-
-\InSoCGVer{
-\bibliographystyle{plain}
-}
-
 \bibliography{contour}
-
-% % 
-% % 
-% % \InSoCGVer{
-% % \appendix
-% % 
-% % %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% % 
-% % \section{Some technical remarks}
-% % \label{sec:techmarks}
-% % The following is a continuation from the end of \Sec{basics}.
-% % 
-% % \technicalRemarks
-% % 
-% % %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% % 
-% % \section{Proof of Surgery}
-% % \label{sec:proofOfSurgery}
-% % \surgeryBody
-% % That is, Theorem~\ref{thm:jordan}, which is proven below.
-% % \proofofDisconnected
-% % 
-% % %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% % 
-% % \section{Join and Split Trees of Extremum Dominant Complexes}
-% % \label{sec:jstedm}
-% % In this section we prove Theorem~\ref{thm:contour-tree} (of \Sec{extreme}), which states the equivalence 
-% % of the contour tree on extremum dominant complexes and the critical join tree of Definition~\ref{def:criticalJoin}.
-% % This proof requires some definitions and notation, specially
-% % the notions of \emph{join} and \emph{split} trees, as given by~\cite{csa-cctad-03}.
-% % Conventionally, all edges are directed from higher to lower function value. 
-% % 
-% % \joinTreeSplitTreeMerge
-% % 
-% % We are now ready to prove Theorem~\ref{thm:contour-tree}.
-% % \proofofContourJoinEquiv
-% % 
-% % %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%reeb graph
-% % 
-% % \section{Proof of \Lem{runTimeUpper}}
-% % \label{sec:runTimeProof}
-% % \proofofRunTime
-% % 
-% % %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% % 
-% % \leafAssignPathDecomp
-% % 
-% % %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% % 
-% % \lowerBoundbyPathDecomp
-% % 
-% % }
-% %
 
@@
 
-\appendix
-
 \end{document}

### 2017-01-03 paint-contour-tree.tex
done with my pass

@@
 %\usepackage[dvipsnames,usenames]{color}
+\usepackage{caption}
 \usepackage{color}
@@
       Richardson, TX, 75080, USA; %
-      \texttt{\si{benjamin.raichel}@\si{utdallas}.\si{edu}} %
-  }
+      \texttt{\si{benjamin.raichel}@\si{utdallas}.\si{edu}}.
+  The work was done when this author was an intern
+  at Sandia National Laboratories, Livermore.}
   \and
@@
 \break A Faster Contour Tree Algorithm%
-\footnote{The full updated version of this paper is available on the arXiv \cite{rs-mvrms-14}}%
+% \footnote{The full updated version of this paper is available on the arXiv \cite{rs-mvrms-14}}%
 }
@@
 proceeds various events occur: new contours are created or destroyed, contours merge into each other or
-split into new components, contours acquire or lose handles. 
-The contour tree is a concise representation of all these events.
+split into new components. 
+The contour tree is a concise representation of these events.
 Throughout we follow the definition of contour trees from \cite{kobps-ctsssit-97} which includes all changes in topology.  
@@
 If $f$ is smooth, all points where the gradient of $f$ is zero are \emph{critical points}.
-These points are the ``events" where the contour topology changes and form the vertices of the contour tree. 
+These points are the ``events" where the contour topology changes and form the vertices of the contour tree. (This is formally incorrect,
+but gives the right intuition.)
 An edge of the contour tree connects two critical points if one event immediately ``follows"
@@
 Consider $f: \MM \to \RR$, where $\MM$ is a triangulated mesh with $n$ vertices, $N$ faces in total, and $t \leq n$ critical points. 
-(We assume that $f:\MM \to \RR$ a linear interpolant over distinct valued vertices, where the contour tree
-$T$ has maximum degree $3$. The degree assumption simplifies the presentation, and is commonly made~\cite{kobps-ctsssit-97}.)
+(We assume that $f:\MM \to \RR$ a linear interpolant over distinct valued vertices, with Morse degree at most $3$. We define this formally 
+later, but it suffices to think of this as a general position assumption.
+The degree assumption is standard in this literature~\cite{kobps-ctsssit-97}.)
 % \Ben{It is standard to assume that $f$ has distinct values on the vertices of $\MM$ and the value at interior points is linearly interpolated.}
@@
 \begin{theorem} \label{thm:main-corr} Consider a simplicial complex $f:\MM \to \RR$, described as above, and denote
-the contour tree by $T$ with vertex set (the critical points) $C(T)$. There exists an algorithm to compute $T$
+the contour tree by $T$ with vertex set (the critical points) $C(T)$. 
+Let the Morse degree of $f$ at any point be at most $3$.
+There exists an algorithm to compute $T$
 in $O(\sum_{v \in C(T)} \log \ell_v + t\alpha(t) + N)$ time. Moreover, this algorithm only
@@
 are standard for results in this area, though there is some variability in notation.
-The input is a continuous piecewise-linear function $f:\MM \to \RR$, where $\MM$ is a simply connected and fully triangulated simplicial complex in $\RR^d$,
+The input is a continuous piecewise-linear function $f:\MM \to \RR$, where $\MM$ is a simply connected and fully triangulated simplicial complex of a ball in $\RR^d$,
 except for specially designated \emph{boundary facets}. We assume that all vertices of $\MM$ have distinct function values, 
@@
 $f|B_\eps(x)$ be the restriction of $f$ to $B_\eps(x)$.
+The following definition of a critical point is not standard,
+but is convenient for our presentation.
 
@@
 
-The critical points are exactly where the topology of level sets change.
+The critical points are where certain topological properties
+ of level sets change.
 By assuming that our simplicial complex is boundary critical, the vertices on a given boundary are either collectively all maxima or all minima.  
@@
 %
+
+We assume that the Morse up (or down) degree of any point
+in $\MM$ is at most $2$, and that the total Morse degree
+is at most $3$. This immediately implies analogous bounds
+on the up and down-degrees in $\reeb(\MM)$. This
+is standard in the literature
+among others, though it can be challenging to enforce in 
+practice. Theoretically, a manifold with points of higher Morse degree
+can be converted to one with the degree condition through
+a process of \emph{vertex unfolding} (refer to
+Section 6.3 in~\cite{HaEd10}). This effectively converts
+a multi-saddle to a set of ordinary saddles.
+A detailed discussion is provided in the end
+of Section 2 and beginning of Section 3 in~\cite{BaKr+98}.
+ For 2D manifolds,
+this conversion does not increase the surface complexity.
+In higher dimensions, the unfolding can increase complexity if
+$\MM$ has vertices of non-constant degree.
+
 Similar to \cite{kobps-ctsssit-97} (among others), multi-saddles are treated as a set of ordinary saddles, which can be realized 
@@
 When the Morse up-degree is $2$ but the up-degree is $1$, the topology of the level set changes but not by the number of connected components changing.
-For example, when $d=3$ this is equivalent to the contour gaining a handle.
 When $d=2$, this distinction is not necessary, since any point with Morse degree strictly greater than $1$ will have degree strictly greater than $1$ in $\reeb(\MM)$.
@@
 The contour tree of \Def{tree} is the typical definition in all results on output-sensitive
-contour trees, and is the smallest tree that contains all the topological changes
-of level sets. \Thm{main-alg} is applicable for any augmentation of $\reeb(\MM)$ with a predefined
+contour trees.
+\Thm{main-alg} is applicable for any augmentation of $\reeb(\MM)$ with a predefined
 set of vertices, though we will not delve into these aspects in this paper.
@@
 \begin{asparaitem}
-	 \item An \emph{up-star} of $v$ is the set of edges that all connected to a fixed component of $\MM^+_v$.
+	 \item For every maximal connected component $Y$
+     of $\MM^+_v$, the set of all edges connecting $v$ to $Y$
+     is called an \emph{up-star}.
 	 \item A vertex $v$ is \emph{touched by color $c$} if $v$ is incident to a $c$-colored
@@
 
-\subsection{The data structures} \label{sec:struct}
+\subsection{The algorithm} \label{sec:algo}
+
+We first give an informal description of the algorithm.
+
+\medskip
+
+\begin{compactenum}
+    \item Construct a painting of $\MM$ by a descending BFS 
+    from maxima. This partitions the edges (into various colored
+    sets), and each vertex is incident to at most $2$ colors.
+    \item For each color, construct a (binomial) max-heap
+    indexed by height of all critical points touched by that color.
+    \item Repeat until all heaps are empty:
+    \begin{compactenum}
+        \item Find a critical point $w$ that is at the top
+        of all heaps containing $w$.
+        \item Add $w$ to the join tree. 
+        \item Delete $w$ from all these heaps, and merge them.
+    \end{compactenum}
+\end{compactenum}
+
+\medskip
+
+\noindent
+This motivates the definition of \emph{ripe} vertices.
+
+\begin{definition} \label{def:ripe} A vertex $v$ is \emph{ripe} if: for all $c \in \col(v)$, $v$
+is present in $T(\rep(c))$ and is also the highest vertex in this heap. 
+%Also, for all $c \in \col(v) \setminus \fcol(v)$, $T(\rep(c))$ is empty.
+\end{definition}
+
+For a fast running time, we need to find ripe vertices
+quickly. Thus, the formal algorithm requires a 
+number of data structures, described below.
+
+\subsubsection{The data structures} \label{sec:struct}
 \noindent
@@
 
-\subsection{The algorithm} \label{sec:algo}
-
-We formally describe the algorithm below. 
-We require the following definition of \emph{ripe} vertices.
-
-\begin{definition} \label{def:ripe} A vertex $v$ is \emph{ripe} if: for all $c \in \col(v)$, $v$
-is present in $T(\rep(c))$ and is also the highest vertex in this heap. 
-%Also, for all $c \in \col(v) \setminus \fcol(v)$, $T(\rep(c))$ is empty.
-\end{definition}
+\subsubsection{The formal algorithm} \label{sec:algorithm}
+
+We now give a detailed description of the main algorithm.
+The primary challenge to use the data structures to rapidly find
+ripe vertices.
 
@@
 \end{compactenum}
-\end{minipage}}
+\end{minipage}
+}
 
@@
 	\item At all times, the colors form a valid painting.
-	\item Each vertex is present in at most $2$ heaps. After processing, it is removed from all heaps.
+	\item Each vertex is present in at most $2$ heaps. This is
+    because the Morse up-degree of every point is at most $2$.
+    After processing, it is removed from all heaps.
 	\item After $v$ is processed, all edges incident to $v$ from above have the same (representative) color. %(more technically, the same representative). 
@@
 where computing the contour tree reduces to computing the join tree, and this equivalence is proven in \Sec{extreme}. 
+The pieces were create are technically not simplicial complices,
+in that each piece may have facets that not simplices. We may
+create facets of larger size than the dimension. These can
+easily be completeed into simplicial complices, though we do not
+explicitly need this. For ease of exposition, we do not explicitly
+perform this step in our algorithm.
 
@@
 Let $\chi$ be some fixed leaf assignment to a rooted binary tree $T$, which in turn fixes all the heaps $H_v$. 
-We choose a special path decomposition that is best defined as a subset of edges in $T$ such that each internal
-vertex has degree at most $2$. This naturally gives a path decomposition.
+We partition the edges of $T$ into a \emph{path decomposition}, where each set of the partition (a path) 
+is a subset of edges in $T$ with each internal
+vertex of degree at most $2$. This naturally gives a path decomposition.
 For each internal vertex $v\in T$, add the edge from $v$ to $\arg \max_{v_l, v_r} \{|H_{v_l}|, |H_{v_r}|\}$ 
@@
 %Consider a PL-Morse $f:\MM \to \RR$, where the join tree $\cJ(\MM)$ has maximum degree $3$.
-Let $f:\MM \to \RR$ be the linear interpolant over distinct valued vertices, where the join tree $\cJ(\MM)$ has maximum degree $3$.
+Let $f:\MM \to \RR$ be the linear interpolant over distinct valued vertices, where the Morse degree of any point is at most $3$.
 There is an algorithm to compute the join tree whose running time is $O(\sum_{p \in \pmax(\cJ)} |p|\log |p| + t\alpha(t) + N)$.
@@
 %Consider a PL-Morse $f:\MM \to \RR$, where the contour tree $\cC=\cC(\MM)$ has maximum degree $3$. 
-Let $f:\MM \to \RR$ be the linear interpolant over distinct valued vertices, where the contour tree $\cC=\cC(\MM)$ has maximum degree $3$. 
+Let $f:\MM \to \RR$ be the linear interpolant over distinct valued vertices, where the Morse degree of any point is at most $3$. 
 There is an algorithm to compute $\cC$ whose running time is $O(\sum_{p \in P(\cC)} |p|\log |p| + t\alpha(t) + N)$,

### 2017-01-03 paint-contour-tree.tex
small changes

@@
 If $f$ is smooth, all points where the gradient of $f$ is zero are \emph{critical points}.
-These points are the ``events" where the contour topology changes and form the vertices of the contour tree. (This is formally incorrect,
-but gives the right intuition.)
+These points are the ``events" where the contour topology changes and form the vertices of the contour tree. 
+(This is formally incorrect, but gives the right intuition.)
 An edge of the contour tree connects two critical points if one event immediately ``follows"
@@
 \begin{asparaitem}
-	 \item For every maximal connected component $Y$
+	 \item For each maximal connected component $Y$
      of $\MM^+_v$, the set of all edges connecting $v$ to $Y$
-     is called an \emph{up-star}.
+     is called an \emph{up-star}. (Note $v$ has at most two up-stars, as Morse up-degree is at most 2 by assumption.)
 	 \item A vertex $v$ is \emph{touched by color $c$} if $v$ is incident to a $c$-colored
@@
 where computing the contour tree reduces to computing the join tree, and this equivalence is proven in \Sec{extreme}. 
-The pieces were create are technically not simplicial complices,
-in that each piece may have facets that not simplices. We may
+The pieces we create are technically not simplicial complexes,
+in that each piece may have facets that are not simplices. We may
 create facets of larger size than the dimension. These can
-easily be completeed into simplicial complices, though we do not
-explicitly need this. For ease of exposition, we do not explicitly
+easily be completed into simplicial complexes, though as we do not
+need this, for ease of exposition, we do not explicitly
 perform this step in our algorithm.

### 2017-01-04 paint-contour-tree.tex
small changes

@@
 split into new components. 
-The contour tree is a concise representation of these events.
-Throughout we follow the definition of contour trees from \cite{kobps-ctsssit-97} which includes all changes in topology.  
-For $d>2$, some subsequent works, such as \cite{csa-cctad-03}, only include changes in the number of components.
+The contour tree is a concise representation of these events \cite{kobps-ctsssit-97,csa-cctad-03}.
+%Throughout we follow the definition of contour trees from \cite{kobps-ctsssit-97} which includes all changes in topology.  
+%For $d>2$, some subsequent works, such as \cite{csa-cctad-03}, only include changes in the number of components.
 
@@
 are standard for results in this area, though there is some variability in notation.
-The input is a continuous piecewise-linear function $f:\MM \to \RR$, where $\MM$ is a simply connected and fully triangulated simplicial complex of a ball in $\RR^d$,
+The input is a continuous piecewise-linear function $f:\MM \to \RR$, where $\MM$ is a simply connected and fully triangulated simplicial complex of a topological ball in $\RR^d$,
 except for specially designated \emph{boundary facets}. We assume that all vertices of $\MM$ have distinct function values, 
@@
 
-This is convenient, as we can now assume that $f$ is defined on $\RR^d$. Any point inside a boundary facet
-has a well-defined height, including the infinite facet, which is required to be a boundary facet.  However, 
-we allow for other boundary facets, to capture the resulting surface pieces after our algorithm makes a horizontal cut.
+% This is convenient, as we can now assume that $f$ is defined on $\RR^d$. Any point inside a boundary facet
+% has a well-defined height, including the infinite facet, which is required to be a boundary facet.  However, 
+% we allow for other boundary facets, to capture the resulting surface pieces after our algorithm makes a horizontal cut.
+
+Boundary facets allow us to capture the resulting surface pieces after our algorithm makes a horizontal cut. 
+The above definition is convenient, as any point within a given facet has a well-defined height, including the boundary facets.
 
@@
 a multi-saddle to a set of ordinary saddles.
-A detailed discussion is provided in the end
-of Section 2 and beginning of Section 3 in~\cite{BaKr+98}.
+%A detailed discussion is provided in the end
+%of Section 2 and beginning of Section 3 in~\cite{BaKr+98}.
 For 2D manifolds,
@@
 
-Similar to \cite{kobps-ctsssit-97} (among others), multi-saddles are treated as a set of ordinary saddles, which can be realized 
-via vertex unfolding (which can increase surface complexity if multi-saddle degrees are allowed to be super-constant).
-Therefore, to simplify the presentation, for the remainder of the paper up and down-degrees are at most $2$, and total degree is at most $3$.
+% Similar to \cite{kobps-ctsssit-97} (among others), multi-saddles are treated as a set of ordinary saddles, which can be realized 
+% via vertex unfolding (which can increase surface complexity if multi-saddle degrees are allowed to be super-constant).
+% Therefore, to simplify the presentation, for the remainder of the paper up and down-degrees are at most $2$, and total degree is at most $3$.
 %
@@
 
-We first give an informal description of the algorithm.
+We first give an informal, idealized description of the algorithm. 
+That is, in the actual algorithm the details differ slightly
+and additional data structures are used to improve the running time.
 
@@
 \begin{compactenum}
-    \item Construct a painting of $\MM$ by a descending BFS 
-    from maxima. This partitions the edges (into various colored
-    sets), and each vertex is incident to at most $2$ colors.
-    \item For each color, construct a (binomial) max-heap
-    indexed by height of all critical points touched by that color.
+%     \item Construct a painting of $\MM$ by a descending BFS 
+%     from maxima. This partitions the edges (into various colored
+%     sets), and each vertex is incident to at most $2$ colors.
+%     \item For each color, construct a (binomial) max-heap
+%     indexed by height of all critical points touched by that color.
+    \item Construct a painting of $\MM$ by a descending BFS from each maximum. 
+    This partitions the edges into colored sets, 
+    and assigns each vertex to the color classes of the edges that touch it.
+    \item For each color, construct a (binomial) max-heap of the critical points touched by that color, 
+          indexed by the heights of the critical points.
+    \item Initialize the join tree to consists of a set of isolated vertices, one for each color class.\\
+          (We now iteratively fill in the rest of the join tree from the leaves towards the root.)
     \item Repeat until all heaps are empty:
     \begin{compactenum}
-        \item Find a critical point $w$ that is at the top
-        of all heaps containing $w$.
-        \item Add $w$ to the join tree. 
+        \item Find a critical point $w$ that is at the top of all heaps containing $w$.
+        \item Add $w$ to the join tree by connecting it to the lowest vertex 
+              of each component of the join tree that contains a maximum whose color matches one of the heaps containing $w$.
         \item Delete $w$ from all these heaps, and merge them.
@@
 
-\begin{definition} \label{def:ripe} A vertex $v$ is \emph{ripe} if: for all $c \in \col(v)$, $v$
-is present in $T(\rep(c))$ and is also the highest vertex in this heap. 
-%Also, for all $c \in \col(v) \setminus \fcol(v)$, $T(\rep(c))$ is empty.
+\begin{definition} \label{def:ripe} A vertex $v$ is \emph{ripe} if: for all $c \in \col(v)$, 
+%$v$ is present in $T(\rep(c))$ and is also the highest vertex in this heap. 
+if $v$ is in heap of $c$ then it is the highest vertex in that heap (using notation from \Sec{struct}, if $v\in T(\rep(c))$, then it is the highest in $T(\rep(c))$).
+%%Also, for all $c \in \col(v) \setminus \fcol(v)$, $T(\rep(c))$ is empty.
 \end{definition}
 
-For a fast running time, we need to find ripe vertices
-quickly. Thus, the formal algorithm requires a 
-number of data structures, described below.
+To efficiently implement the above, we need a number of data structures.  
+Namely, binomial heaps for efficient merges, union-find to avoid explicit recoloring, 
+and a stack to hold unripe (though ripening) vertices, which is used to effectively amortize the cost of 
+the otherwise expensive operation of finding ripe vertices. 
+Note we also later show it suffices to assign each vertex to only a single color from each adjacent up-star, 
+rather than the colors of all the edges which touch it.
+
+% For a fast running time, we need to find ripe vertices
+% quickly. Thus, the formal algorithm requires a 
+% number of data structures, described below.
 
@@
 This slack allows for the more refined upper and lower bounds mentioned in \Sec{more-refined}.  Quantifying this slack however is quite challenging, 
-and requires a significantly more sophisticated analysis involving path decompositions, which is the subject of \Sec{lb}.
+and requires a significantly more sophisticated analysis involving path decompositions, which is the subject of \Sec{pathDecomp}.
 
@@
 We describe the main partitioning procedure that cuts a simplicial complex $\NN$ into extremum
-dominant complexes. It takes an additional input of a maximum $x$. To initialize,
+dominant complexes, shown below as $\rain(x,\NN)$. It takes an additional input of a maximum $x$. To initialize,
 we begin with $\NN$ set to $\MM$ and $x$ as an arbitrary maximum. 
@@
 When we start, rain flows downwards. In each recursive call, the direction of rain is \emph{switched} to the 
-opposite direction. This is crucial to ensure a linear running time, and \Fig{order} shows how failing to switch 
+opposite direction. This is crucial to ensure linear running time, and \Fig{order} shows how failing to switch 
 directions can lead to a blow up in complexity. 

### 2017-01-05 paint-contour-tree.tex
typo corrected

@@
 %$v$ is present in $T(\rep(c))$ and is also the highest vertex in this heap. 
-if $v$ is in heap of $c$ then it is the highest vertex in that heap (using notation from \Sec{struct}, if $v\in T(\rep(c))$, then it is the highest in $T(\rep(c))$).
+if $v$ is in the heap of $c$ then it is the highest vertex in that heap (using notation from \Sec{struct}, if $v\in T(\rep(c))$, then it is highest in $T(\rep(c))$).
 %%Also, for all $c \in \col(v) \setminus \fcol(v)$, $T(\rep(c))$ is empty.
@@
 
+\begin{figure}[h]\centering
+    \includegraphics[width=.172\linewidth]{figs/tent2}%
+    \hspace{1.2in}
+    \includegraphics[width=.172\linewidth]{figs/tent1}
+    \caption{On left, downward rain spilling only (each shade of gray represents a piece created by each different spilling), producing a grid.  
+    Note we are assuming raining was done in sorted order of the maxima (i.e.\ lowest to highest).  
+    On right, flipping the direction of rain spilling.}
+    \label{fig:order}
+\end{figure}
+
 We describe the main partitioning procedure that cuts a simplicial complex $\NN$ into extremum
@@
 We can now let rain flow downwards, as it usually does in our world.
-
-\begin{figure}[h]\centering
-    \includegraphics[width=.172\linewidth]{figs/tent2}%
-    \hspace{1.2in}
-    \includegraphics[width=.172\linewidth]{figs/tent1}
-    \caption{On left, downward rain spilling only (each shade of gray represents a piece created by each different spilling), producing a grid.  
-    Note we are assuming raining was done in sorted order of the maxima (i.e.\ lowest to highest).  
-    On right, flipping the direction of rain spilling.}
-    \label{fig:order}
-\end{figure}
 

### 2017-04-15 paint-contour-tree.tex
various small  changes

@@
 All existing algorithms begin with a global sort of at least all critical values of $f$,
-which can require (roughly) $\Omega(n\log n)$ time.
+which can require (roughly) $\Omega(n\log n)$ time, where $n$ is the number of vertices of the mesh.
 Existing lower bounds show that there are pathological instances where this sort is required.
@@
 Consider $f: \MM \to \RR$, where $\MM$ is a triangulated mesh with $n$ vertices, $N$ faces in total, and $t \leq n$ critical points. 
-(We assume that $f:\MM \to \RR$ a linear interpolant over distinct valued vertices, with Morse degree at most $3$. We define this formally 
-later, but it suffices to think of this as a general position assumption.
-The degree assumption is standard in this literature~\cite{kobps-ctsssit-97}.)
+(We assume that $f:\MM \to \RR$ is a linear interpolant over distinct valued vertices, with Morse degree at most $3$. 
+The degree assumption can be achieved via vertex unfolding, as discussed later,
+%We define this formally later, but it suffices to think of this as a general position assumption.
+and is standard in this literature~\cite{kobps-ctsssit-97}.)
 % \Ben{It is standard to assume that $f$ has distinct values on the vertices of $\MM$ and the value at interior points is linearly interpolated.}
@@
 
-\subsection{Previous Work}
+\subsection{Previous work}
 
@@
 meshes and an $O(N^2)$ algorithm for higher dimensions, by van Kreveld \etal \cite{kobps-ctsssit-97}. 
-Tarasov and Vyalya \cite{tv-cct-98} improved the running time to $O(N\log N)$ for the 3D case.
+Tarasov and Vyalyi \cite{tv-cct-98} improved the running time to $O(N\log N)$ for the 3D case.
 The influential paper of Carr \etal \cite{csa-cctad-03} improved the running time for all dimensions to $O(n\log n + N\alpha(N))$.
@@
 The input is a continuous piecewise-linear function $f:\MM \to \RR$, where $\MM$ is a simply connected and fully triangulated simplicial complex of a topological ball in $\RR^d$,
-except for specially designated \emph{boundary facets}. We assume that all vertices of $\MM$ have distinct function values, 
+except for specially designated \emph{boundary facets}. We assume that all pairs of vertices from $\MM$ have distinct function values, except for pairs from the same boundary facet,
 and that $f$ is only explicitly defined on the vertices, and all other values are obtained by linear interpolation.
@@
 \subsection{Some remarks}
+{\bf 1)}
 Note that if one intersects $\MM$ with a given ball $B$, then a single contour in $\MM$ might be split into more than one contour in the intersection. 
@@
 When the Morse up-degree is $2$ but the up-degree is $1$, the topology of the level set changes but not by the number of connected components changing.
-When $d=2$, this distinction is not necessary, since any point with Morse degree strictly greater than $1$ will have degree strictly greater than $1$ in $\reeb(\MM)$.
-
+When $d=2$, this distinction is not necessary, since any point with Morse degree strictly greater than $1$ will have degree strictly greater than $1$ in $\reeb(\MM)$.\\
+
+\noindent{\bf 2)}
 As Carr points out in Chapter 6 of his thesis, the term contour tree can be used for a family of related structures.
@@
 
-The intuition is that the paint spilling from the maxmima in the simplicial complex, corresponds to paint spilling from the leaves in 
+The intuition is that the paint spilling from the maxima in the simplicial complex, corresponds to paint spilling from the leaves in 
 the join tree, which decomposes the join tree into a set of colored paths.  Unfortunately, the situation is more complex since 
@@
 We now present the main algorithmic contribution, which is a new algorithm for computing join trees of any triangulated simplicial
-complex $\MM$.  In subsequent sections we then show how to reduce computing the contour to computing join by using contour surgery.
+complex $\MM$.  In subsequent sections we then show how to reduce computing the contour tree to computing the join tree by using contour surgery.
 
@@
           indexed by the heights of the critical points.
-    \item Initialize the join tree to consists of a set of isolated vertices, one for each color class.\\
+    \item Initialize the join tree to consist of a set of isolated vertices, one for each color class.\\
           (We now iteratively fill in the rest of the join tree from the leaves towards the root.)
@@
 
-\subsection{Running Time}
+\subsection{Running time}
 \label{sec:runTime}
@@
 \newcommand{\pruned}{\cU'}
-\subsection{Paying for Pruned Shrubs: the root is everything} \label{sec:weight}
+\subsection{Paying for pruned shrubs: the root is everything} \label{sec:weight}
 
@@
 
-\subsection{Our Main Result}
+\subsection{Our main result}
 \label{sec:implications}
@@
 
-\section{Lower Bound by Path Decomposition}
+\section{Lower bound by path decomposition}
 \label{sec:lb}
@@
 
-\subsection{Join Trees}
+\subsection{Join trees}
 
@@
 
-\subsection{Contour Trees}
+\subsection{Contour trees}
 We first generalize previous terms to the case of contour trees.
@@
 Nuclear Security Administration under contract DE-AC04-94AL85000.
+Work on this paper by Benjamin Raichel was also partially supported
+by NSF CRII Award 1566137.
 

### 2017-05-03 paint-contour-tree.tex
figure added

@@
 
+\usepackage{subcaption}
 \usepackage{amsfonts,amsmath,amssymb,amsthm,mathtools}
@@
 
+\pagebreak
+\appendix
+
+\section{ \Sec{algo} Algorithm Example }
+
+\textcolor{red}{
+In the example below, I assumed that the maxima are already considered ``processed'', so that there are not additional trivial steps.  
+This actually makes more sense since the $att()$ should really only ever point to already processed vertex.  Therefore, I suggest 
+changing line 3 of init to ``Mark all maxima as processed.  Mark all other critical $v$ as unprocessed.''
+Perhaps line 4 needs to be changed also?
+}
+
+ \begin{figure}[h]\centering
+    \includegraphics[width=.3\linewidth,page=1]{figs/mesh1}
+    \caption{A piece of a potentially larger mesh, with maxima $m_1,m_2,m_3$, saddles $s_1,s_2$, and regular vertices $r_1,r_2,r_3,r_4$.
+             An initial painting is shown where green `g' is spilled from $m_1$, then red `r' from $m_2$, and finally blue `b' from $m_3$.
+             Arrows are directed from higher to lower function value.}
+    \label{fig:mesh1}
+\end{figure} 
+%
+\begin{figure}[h]
+   \begin{subfigure}{.33\textwidth}
+     \centering
+     \includegraphics[width=.7\linewidth,page=1]{figs/mesh2}
+     \caption{}
+     \label{fig:tree1}
+   \end{subfigure}%
+   \begin{subfigure}{.33\textwidth}
+     \centering
+     \includegraphics[width=.7\linewidth,page=2]{figs/mesh2}
+     \caption{}
+     \label{fig:tree2}
+   \end{subfigure}
+   \begin{subfigure}{.33\textwidth}
+     \centering
+     \includegraphics[width=.7\linewidth,page=3]{figs/mesh2}
+     \caption{}
+     \label{fig:tree3}
+   \end{subfigure}
+ \caption{Snapshots of the join tree.}
+ \label{fig:mesh2}
+\end{figure}
+%
+\begin{figure*}[h]%
+    \centerline{%
+       \begin{tabular}{|l|c|c|c|c|c|c|c|c|}%
+         \hline
+         Stage%
+         &%
+           $att(r)$
+         &%
+           $att(g)$
+         &%
+           $att(b)$
+         &%
+           $T(r)$
+         &%
+           $T(g)$
+         &%
+           $T(b)$
+         &%
+           $K$
+         &%
+          Current tree\\
+         \hline%
+         \hline%
+         Initial
+         &%
+           $m_2$
+         &%
+           $m_1$
+         &%
+           $m_3$
+         &%
+           $\{s_2\}$
+         &%
+           $\{s_1\}$
+         &%
+           $\{s_1,s_2\}$
+         &%
+           $()$
+         &%
+           \Fig{tree1}
+         \\%
+         \hline%
+         % 
+         1%
+         &%
+           $m_2$          
+         &%
+           $m_1$
+         &%
+           $m_3$
+         &%
+           $\{s_2\}$
+         &%
+           $\{s_1\}$
+         &%
+           $\{s_1,s_2\}$
+         &%
+           $(s_1)$
+         &%
+           \Fig{tree1}
+         \\%
+         \hline%
+         %
+         2
+         &%
+           $m_2$
+         &%
+           $m_1$
+         &%
+           $m_3$
+         &%
+           $\{s_2\}$
+         &%
+           $\{s_1\}$
+         &%
+           $\{s_1,s_2\}$
+         &%
+           $(s_1,s_2)$
+         &%
+           \Fig{tree1}
+         \\%
+         \hline%
+         %
+         3
+         &%
+           NA
+         &%
+           $m_1$
+         &%
+           $s_2$
+         &%
+           NA
+         &%
+           $\{s_1\}$
+         &%
+           $\{s_1\}$
+         &%
+           $(s_1)$
+         &%
+           \Fig{tree2}
+         \\%
+         \hline%
+         %
+         4
+         &%
+           NA
+         &%
+           NA
+         &%
+           $s_1$
+         &%
+           NA
+         &%
+           NA
+         &%
+           $\{\}$
+         &%
+           $()$
+         &%
+           \Fig{tree3}
+         \\%
+         \hline
+       \end{tabular}%
+    }
+    \caption{Stages of the algorithm run on the initial painting shown in \Fig{mesh1}.
+    The above table assumes that in the stack update, $s_1$ is the first arbitrary unprocessed critical point selected.}
+    \label{fig:exec}
+\end{figure*}
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2017-05-03 response.txt
done with my minor changes

Review #1:

The authors have updated the paper and fixed many issues. They have addressed my concerns. However, these additions have caused some new minor issues, and, in my opinion, still the the paper is not mature enough to be published as it is, as will be seen in the comments below. I will mention several more important points in the following, but I was unable to list all the typos.

Global issues:
1. Since the paper uses the results of CSA03 for merging a split tree with a merge tree to form a contour tree, it is crucial to have the same type of domains. So I suggest that the authors define their initial domain to be the triangulation of a ball, and again crucially, they should mention that the pieces that they break the domain into, i.e. extremum dominant complexes, are also triangulations of topological balls, so that they can always use the results of CSA03. This is the case since the pieces are formed by raining. Please verify this condition to use the theorems from CSA03. In doing so, the authors can update Theorem 5.4 to say that the outputs of cut are triangulated balls. Again, this is necessary since at each recursive step that the results of CSA03 are used the domains must be triangulated topological balls.
The current wording of definition of M is not good. When defined as in CSA03 to be a triangulated ball, then the simply-connectedness assumption is not needed and they can remove it.
--->We have fixed the wording in Theorem 1.1, 1.3, and Theorem 5.4. We have verified that this is consistent with all our previous wording and with the results of CSA03.
(OLD--->We explicitly did not want to triangulate the boundary faces, since we were worried about situations like a double walled hollow upside bowl (i.e. there is another maximum poking through the boundary facet).  However, I am not entirely sure that is possible given we are lifting the input complex one dimension higher?  At this point I am fine just doing what the reviewer says though.  Can you figure this out?
Also small side note, I added "except for pairs from the same boundary facet," to section 2, first paragraph, last sentence:
)

2. Make sure that all headings have the same uppercase lowercase convention, see for example 1.2, 4.4 etc.
--->All section and subsection heading now have the same lowercase convention, except for the leading word being capitalized.

3. I suggest that the authors go through the updated sections more carefully and correct any typos.
---> We went through the updated sections again as suggested to check for and correct typos.

Smaller issues, typos:
1. In the abstract please define n.
---> Defined.

2. In the previous work it seems that Vyalya should be Vyalyi.
--->Corrected

3. In the second par. of the Introduction, the sentence in the parenthesis is very strange indeed. They should remove it and correct the sentence before it. Not all critical points appear in the contour tree. They can simply say some of these critical points affect the connected components of the contour, and those correspond to vertices of the contour tree. Indeed the contour tree is a simple construct and explaining it does not need more complex definitions.
--->We fixed this paragraph.

4. In the third paragraph of the Introduction, the assumption on Morse degree 3 cannot be thought of as general position assumption. Please remove this. You might want to mention that it is possible to do "vertex unfolding" in the domain.
--->We removed this sentence.
(OLD--->I commented out the general position sentence and added "The degree assumption can be achieved via vertex unfolding, as discussed later,". Please take a look.)

5. In 2.1 it is perhaps good to number the remarks 1) and 2).
---> The remarks are now numbered.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

Second review:
The changes and the reorganization of the paper by the authors are all accept-
able. There is a typo and a grammatical error (listed below) which need to be
fixed. I have two other comments, one a required inclusion and one a strong
suggestion.

a. The authors added to Theorem 1.1, the statement ”Let the Morse degree
of f at any point be at most 3.” They should also state this condition in
Theorem 1.3 and in the Abstract. While they note that assuming Morse
degree three ”is a standard assumption in the literature”, this is standard
only for algorithms constructing Morse complexes. I don’t believe that the
algorithms in either CSA03 or CLLR05 make or require this assumption.
This is a real difference so the assumption about Morse degree should be
in the abstract and Theorem 1.3. (The authors also should also add ”See
Definition 2.3 for a definition of Morse Degree” after using the term Morse
Degree in Theorem 1.1.)
--->We updated the abstract. We put a sentence in Theorem 1.3 referring to the conditions in Theorem 1.1.

b. The authors followed my suggestion of moving the join tree algorithm
earlier and adding a higher level description. However, I repeat my rec-
ommendation that they provide a simplicial complex illustration of the
algorithm in Section 4.2.2 The illustration in Figure 3 is good, but it does
not illustrate the details of the algorithm in 4.2.2. If an illustration is
added, the authors should explain in the text how the algorithm operates
on that illustration.
An illustration of a specific simplicial complex and the application of the
algorithm to this complex would greatly help readers of this paper. The
algorithm in Section 4.2 is understandable without an illustration, so the
paper is publishable without this change, but such an addition would
strengthen the paper and increase the likelihood of it being read.
--->We have provided a detailed example in the Appendix.

As long as the authors add the restriction on Morse degree to Theorem 1.3 and
the Abstract (noted in a. above), and correct the typo and grammatical errors
below, I believe the paper is ready for publication.

Typo & grammatical errors:
p. 8, line 4: ”maxmima” - Typo. Should be ”maxima”.
--->Fixed.
p. 9, Algorithm setp 3: ”Initialize the join tree to consists...”
Grammatical error. Should be ”to consist”.
--->Fixed.

### 2017-05-03 paint-contour-tree.tex
done with my minor changes

@@
 We revisit the classical problem of computing the \emph{contour tree}
-of a scalar field $f:\MM \to \RR$, where $\MM$ is a triangulated simplicial mesh in $\RR^d$. 
+of a scalar field $f:\MM \to \RR$, where $\MM$ 
+is a triangulations of a ball in $\RR^d$.
 The contour tree is a fundamental topological structure that tracks
@@
 We present the first algorithm whose time complexity depends
-on the contour tree structure, and avoids the global sort for non-pathological inputs.
+on the contour tree structure, and avoids the global sort for non-pathological inputs. (We assume
+that the Morse degree of the function $f$ at any
+point is 
+at most $3$.)
 If $C$ denotes the set of critical points in $\MM$, the running time is
@@
 If $f$ is smooth, all points where the gradient of $f$ is zero are \emph{critical points}.
-These points are the ``events" where the contour topology changes and form the vertices of the contour tree. 
-(This is formally incorrect, but gives the right intuition.)
-An edge of the contour tree connects two critical points if one event immediately ``follows"
-the other as the sweep plane makes its pass. We provide formal definitions later. 
+These points are the ``events" where the contour topology changes. The contour tree tracks a specific
+subset of events, called ``joins" and ``splits".
+We provide formal definitions later.
+An edge of the contour tree connects two such critical points if one event immediately ``follows"
+the other as the sweep plane makes its pass.  
 \Fig{relative} and \Fig{sorting} 
@@
 
-Consider $f: \MM \to \RR$, where $\MM$ is a triangulated mesh with $n$ vertices, $N$ faces in total, and $t \leq n$ critical points. 
+Consider $f: \MM \to \RR$, where $\MM$ is the triangulation of a $d$-dimensional ball with $n$ vertices, $N$ faces in total, and $t \leq n$ critical points. 
 (We assume that $f:\MM \to \RR$ is a linear interpolant over distinct valued vertices, with Morse degree at most $3$. 
@@
 
-\begin{theorem} \label{thm:main-corr} Consider a simplicial complex $f:\MM \to \RR$, described as above, and denote
+\begin{theorem} \label{thm:main-corr} Consider a simplicial complex $f:\MM \to \RR$, where $\MM$
+is a triangulation of a $d$-dimensional ball.
+Let $N$ be the number of facets in $\MM$.
+Let the Morse degree of $f$ at any point be at most $3$.
+Denote
 the contour tree by $T$ with vertex set (the critical points) $C(T)$. 
-Let the Morse degree of $f$ at any point be at most $3$.
 There exists an algorithm to compute $T$
@@
 %Our algorithm does not need this, but it simplifies the proof and presentation.} 
+(Consider the input setting of \Thm{main-corr}.)
 There is a deterministic algorithm to compute the contour tree, $T$, whose running time is $O(\sum_{p \in P(T)} |p|\log |p| + t\alpha(t) + N)$,
@@
 \begin{theorem} \label{thm:jordan} $\cut(\MM,\phi)$ consists of two
-disconnected simplicial complexes.
+disconnected simplicial complexes, each of which
+is the triangulation of $d$-dimensional ball.
 \end{theorem}

### 2017-05-04 response.txt
figure update

@@
 --->We have fixed the wording in Theorem 1.1, 1.3, and Theorem 5.4. We have verified that this is consistent with all our previous wording and with the results of CSA03.
-(OLD--->We explicitly did not want to triangulate the boundary faces, since we were worried about situations like a double walled hollow upside bowl (i.e.
-there is another maximum poking through the boundary facet).
-However, I am not entirely sure that is possible given we are lifting the input complex one dimension higher?
-At this point I am fine just doing what the reviewer says though.
-Can you figure this out?
-Also small side note, I added "except for pairs from the same boundary facet," to section 2, first paragraph, last sentence:
-)
 
@@
 --->We removed this sentence.
-(OLD--->I commented out the general position sentence and added "The degree assumption can be achieved via vertex unfolding, as discussed later,". Please take a look.)
 

### 2017-05-04 paint-contour-tree.tex
figure update

@@
 Consider $f: \MM \to \RR$, where $\MM$ is the triangulation of a $d$-dimensional ball with $n$ vertices, $N$ faces in total, and $t \leq n$ critical points. 
-(We assume that $f:\MM \to \RR$ is a linear interpolant over distinct valued vertices, with Morse degree at most $3$. 
+(We assume that $f:\MM \to \RR$ is a linear interpolant over distinct valued vertices, with Morse degree at most $3$, see \Def{deg}. 
 The degree assumption can be achieved via vertex unfolding, as discussed later,
@@
 We now give a detailed description of the main algorithm.
-The primary challenge to use the data structures to rapidly find
+The primary challenge is to use the data structures to rapidly find
 ripe vertices.
+An simple example of the execution of the algorithm is given in \Sec{example}.
 
@@
 
-\section{ \Sec{algo} Algorithm Example }
-
-\textcolor{red}{
-In the example below, I assumed that the maxima are already considered ``processed'', so that there are not additional trivial steps.  
-This actually makes more sense since the $att()$ should really only ever point to already processed vertex.  Therefore, I suggest 
-changing line 3 of init to ``Mark all maxima as processed.  Mark all other critical $v$ as unprocessed.''
-Perhaps line 4 needs to be changed also?
-}
+\section{ \Sec{algo} Algorithm Example }\label{sec:example}
+
+% \textcolor{red}{
+% In the example below, I assumed that the maxima are already considered ``processed'', so that there are not additional trivial steps.  
+% This actually makes more sense since the $att()$ should really only ever point to already processed vertex.  Therefore, I suggest 
+% changing line 3 of init to ``Mark all maxima as processed.  Mark all other critical $v$ as unprocessed.''
+% Perhaps line 4 needs to be changed also?
+% }
+
+Here we provide a simple example of the execution of the main algorithm from \Sec{algo}.  
+An example input mesh is shown in \Fig{mesh1}, followed by the different steps of the join construction in \Fig{mesh2}, 
+and finally a table showing the states of the data structures and how they are modified in \Fig{exec}.
+
+Note that in \Fig{exec} a ``stage'' refers to any time at which one of the data structures is modified.  
+For simplicity, the trivial stages where the maxima are removed from their corresponding heaps are not shown. 
+(Alternatively the algorithm in \Sec{algo} could be modified to exclude the maxima from the heaps, 
+but we wanted to keep the algorithm description simple.)
 

### 2017-05-06 paint-contour-tree.tex
typos fixed

@@
 are standard for results in this area, though there is some variability in notation.
-The input is a continuous piecewise-linear function $f:\MM \to \RR$, where $\MM$ is a simply connected and fully triangulated simplicial complex of a topological ball in $\RR^d$,
+The input is a continuous piecewise-linear function $f:\MM \to \RR$, where $\MM$ is a fully triangulated simplicial complex of a topological ball in $\RR^d$,
 except for specially designated \emph{boundary facets}. We assume that all pairs of vertices from $\MM$ have distinct function values, except for pairs from the same boundary facet,
@@
 ripe vertices.
-An simple example of the execution of the algorithm is given in \Sec{example}.
+A simple example of the algorithm execution is given in \Sec{example}.
 
@@
 The pieces we create are technically not simplicial complexes,
-in that each piece may have facets that are not simplices. We may
-create facets of larger size than the dimension. These can
-easily be completed into simplicial complexes, though as we do not
+in that each piece may have facets that are not simplices, that is facets of larger size than the dimension.
+These can easily be completed into simplicial complexes, though as we do not
 need this, for ease of exposition, we do not explicitly

### 2017-05-21 paint-contour-tree.tex
typo corrected

@@
       \texttt{\si{benjamin.raichel}@\si{utdallas}.\si{edu}}.
-  The work was done when this author was an intern
-  at Sandia National Laboratories, Livermore.}
+}
   \and
@@
 \begin{abstract}
-We revisit the classical problem of computing the \emph{contour tree}
-of a scalar field $f:\MM \to \RR$, where $\MM$ 
-is a triangulations of a ball in $\RR^d$.
-The contour tree is a fundamental topological structure that tracks
-the evolution of level sets of $f$ and 
-has numerous applications in data analysis and visualization.
-
-All existing algorithms begin with a global sort of at least all critical values of $f$,
-which can require (roughly) $\Omega(n\log n)$ time, where $n$ is the number of vertices of the mesh.
+We revisit the classical problem of computing the \emph{contour tree} of a scalar field $f:\MM \to \RR$, where $\MM$ is a triangulations of a ball in $\RR^d$.
+The contour tree is a fundamental topological structure that tracks the evolution of level sets of $f$ and has numerous applications in data analysis and visualization.
+
+All existing algorithms begin with a global sort of at least all critical values of $f$, which can require (roughly) $\Omega(n\log n)$ time, where $n$ is the number of vertices of the mesh.
 Existing lower bounds show that there are pathological instances where this sort is required.
-We present the first algorithm whose time complexity depends
-on the contour tree structure, and avoids the global sort for non-pathological inputs. (We assume
-that the Morse degree of the function $f$ at any
-point is 
-at most $3$.)
-If $C$ denotes the set of critical points in $\MM$, the running time is
-roughly $O(N+\sum_{v \in C} \log \ell_v)$, where $\ell_v$ is the depth of $v$ in the contour tree and $N$ is the total complexity of $\MM$.
+We present the first algorithm whose time complexity depends on the contour tree structure, and avoids the global sort for non-pathological inputs.
+(We assume that the Morse degree of the function $f$ at any point is at most $3$.) If $C$ denotes the set of critical points in $\MM$, the running time is roughly $O(N+\sum_{v \in C} \log \ell_v)$, where $\ell_v$ is the depth of $v$ in the contour tree and $N$ is the total complexity of $\MM$.
 This matches all existing upper bounds, but is a significant asymptotic improvement when the contour tree is short and fat.
-Specifically, our approach ensures that any comparison made is between nodes that are either adjacent in $\MM$ 
-or in the same descending path in the contour tree,
-allowing us to argue strong optimality properties of our algorithm.
-
-Our algorithm requires several novel ideas: partitioning $\MM$ in well-behaved portions, 
-a local growing procedure to iteratively build contour trees, and the use of heavy path 
-decompositions for the time complexity analysis. 
+Specifically, our approach ensures that any comparison made is between nodes that are either adjacent in $\MM$ or in the same descending path in the contour tree, allowing us to argue strong optimality properties of our algorithm.
+
+Our algorithm requires several novel ideas: partitioning $\MM$ in well-behaved portions, a local growing procedure to iteratively build contour trees, and the use of heavy path decompositions for the time complexity analysis.
 \end{abstract}
@@
 The degree assumption can be achieved via vertex unfolding, as discussed later,
-%We define this formally later, but it suffices to think of this as a general position assumption.
 and is standard in this literature~\cite{kobps-ctsssit-97}.)
-% \Ben{It is standard to assume that $f$ has distinct values on the vertices of $\MM$ and the value at interior points is linearly interpolated.}
-%It is standard to assume that $f$ is PL-Morse (a well-behaved input assumption) and that the maximum degree in the contour tree is $3$. 
 A fundamental result in this area is the algorithm of Carr, Snoeyink, and Axen to compute contour trees,
@@
 
-% \Sesh{Almost everything below is new/reordered. Please read through.}
 All previous algorithms begin by sorting (at least) the critical points. Can we beat this sorting bound for certain instances,
@@
 to construct the contour tree.
-
-% \begin{figure}[h]\centering
-%     \includegraphics[width=.26\linewidth]{figs/spikey1}%
-%     \hspace{.35in}
-%     \includegraphics[width=.26\linewidth]{figs/spikey2}%
-%     \hspace{.4in}
-%     \includegraphics[width=.215\linewidth]{figs/spikeyTree}
-%     \caption{Two surfaces with different orderings of the maxima, but the same contour tree.}
-%     \label{fig:relative}
-% \end{figure}
 
@@
 
-% First, we require some technical definitions.
-
-% In a contour tree $T$, a \emph{leaf path} is simply a path in $T$
-% containing a leaf, which is also monotonic in the height values of its vertices.
-
 \begin{definition}
@@
 \label{thm:main-alg} 
-%Consider a PL-Morse\footnote{This is a standard ``well-behaved" input assumption that implies that all critical points have distinct values. 
-%Our algorithm does not need this, but it simplifies the proof and presentation.} 
 (Consider the input setting of \Thm{main-corr}.)
@@
 $3$-dimensional structured meshes. Chiang \etal \cite{cllr-sooscctmp-05} provide an unconditional $O(N+t\log t)$ algorithm.
-% Carr's thesis shows relationships between monotone paths
-% in $\MM$ and its contour tree $\cC(\MM)$.
 
@@
 
-% This is convenient, as we can now assume that $f$ is defined on $\RR^d$. Any point inside a boundary facet
-% has a well-defined height, including the infinite facet, which is required to be a boundary facet.  However, 
-% we allow for other boundary facets, to capture the resulting surface pieces after our algorithm makes a horizontal cut.
-
 Boundary facets allow us to capture the resulting surface pieces after our algorithm makes a horizontal cut. 
@@
 
-% We categorize points depending on the local changes in topology.
-% 
-% \begin{definition} \label{def:points} A point $x$ is categorized as follows.
-% \begin{asparaitem}
-% 	\item \emph{Regular:} Both Morse up-degree and down-degrees are $1$.
-% 	\item \emph{Maximum:} Morse up-degree is $0$.
-% 	\item \emph{Minimum:} Morse down-degree is $0$.
-% 	\item \emph{Morse Join:} Morse up-degree is strictly greater than $1$.
-% 	\item \emph{Morse Split:} Morse down-degree is strictly greater than $1$.
-% \end{asparaitem}
-% Non-regular points are \emph{critical}. Morse joins and splits are \emph{saddles}.
-% Maxima and minima are \emph{extrema}.
-% \end{definition}
-%
-
 The set of critical points is denoted by $\cV(f)$.
@@
 This equivalence relation gives a set of \emph{contour classes}. 
-%By continuity, a regular point is present on a single contour,
-%and we can partition the regular contours into these classes. 
 Every such class maps to intervals
@@
 Since this tree has height values associated with the vertices, we can talk about up-degrees and down-degrees in $\reeb(\MM)$.
-%
 
@@
 
-%\pagebreak
 \section{A tour of the new contour tree algorithm} \label{sec:approach}
@@
 of the pieces is linear (see \Sec{rain} for details).
-% 
-% 
-% Each additional cut increases the complexity of the surface,
-% so care is required to keep the total complexity linear. This is achieved by alternately ``raining"
-% upwards and downwards.  We finally get a collection of extremum dominant pieces.
-% We prove a simple contour surgery theorem that builds the contour tree of $\MM$
-% from the contour trees of the various pieces created.
 
@@
 The final analysis charges expensive heap operations to long paths in the decomposition, resulting in the bound stated in \Thm{main-alg}.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
 
-% For a fast running time, we need to find ripe vertices
-% quickly. Thus, the formal algorithm requires a 
-% number of data structures, described below.
-
 \subsubsection{The data structures} \label{sec:struct}
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
 \section{Divide and conquer through contour surgery} \label{sec:surgery}
@@
 disconnected simplicial complexes, each of which
-is the triangulation of $d$-dimensional ball.
+is the triangulation of a $d$-dimensional ball.
 \end{theorem}
@@
 \end{proof}
-
-% 
-% \begin{lemma} \label{lem:blahblah}
-%  Testing 123 \RefProofInAppendix{blahblah}
-% \end{lemma}
-% 
-% \begin{proof:in:appendix:e}{\ref{lem:blahblah}}{blahblah}
-%  blah balh
-% \end{proof:in:appendix:e}
 
@@
 Specifically, $\sum_{p\in \buff} \sum_{w \in p} \log |H_w| = O(\sum_{p\in \buff} V_p) = O(\sum_{p\in \buff}\sum_{v\in \loss(p)} \log |H_v| )$
-
-% \textcolor{red}{\huge Proof missing. Potentially useful observations below}
-%  
-%  Let $r$ be the root of $\buff$.
-First observe that $\loss(r) \geq |r|/100$.
-Suppose otherwise, then in order for the first condition of \Def{buff} to be satisfied for $\buff$, the rooted subtree of some child of $r$, call it $\buff'$, must satisfy $\sum_{p\in \buff'} \loss(p) \geq \frac{1}{100} \sum_{p\in \buff'} |p|$.
-However, this would contradict the second condition of \Def{buff}.
-%  
-%  Consider any node $q\in \buff$.
-Then for any vertex $v\in \loss(q)$, we have $|H_v| = \Omega(|H_q|)$, since $q$ is unsupported and hence by definition less than $|H_q|/100$ vertices of $H_q$ appear in $\loss(q)$.
-%  
-%  \begin{align*}
-%   \sum_{p\in \buff}\sum_{v\in p} \log |H_v| &=  O\left( \sum_{p\in \buff}\sum_{v\in p} \log |H_p| \right)   &\mbox{definition of short} \\
-%   &= O\left( \sum_{p\in \buff} |p| \log |H_p| \right)   &\mbox{} \\
-%   &= ????   &\mbox{buffered is not sufficient to get to next line} \\
-%   &= O\left( \sum_{p\in \buff} |\loss(p)| \log |H_p| \right)   &\mbox{} \\
-%   &= O\left( \sum_{p\in \buff}\sum_{v\in \loss(p)} \log |H_v| \right)
-%  \end{align*}
-%  
-%  \Cor{geometric} should be used somewhere.
 \end{proof}
@@
 The goal of this section is to prove the following lemma.
-% , which ultimately allows us to pay for pruned shrubs.
 
@@
 We combine these bounds to show that $\sum_{p \in \pruned} |p| = O(|H_r|)$.
-% 
-% 
-% 
-% \begin{proof}[Proof of \Lem{cu-root}] We first pro 
-% We use $r$ for $r(\cU)$. %Our aim is to simply apply \Clm{charge} from $r(\cU)$. 
-% Suppose $W_q \geq (4/3)W_{q'}$ (for any choice in $\cU\setminus\{r\}$ of $q$ parent of $q'$), 
-% then by \Cor{chargeCor}, $\sum_{p \in \cU} |p| \leq W_{r}/20+|r|$.
-% By \Clm{loss}, $|\redH_r| \geq W_r - 2\sum_{p \in \cU} |p|$, and so combining these inequalities gives,
-% \[
-% % \sum_{p \in \cU} |p| \leq \pth{|\redH_r|+2\sum_{p \in \cU} |p|}/20+|r| \Rightarrow 
-%  \sum_{p \in \cU} |p|\leq \frac{10}{9} \pth{|\redH_r|/20+|r|} \leq 
-%  \frac{10}{9}\pth{|H_r|/20+\sqrt{|H_r|}/100} \leq |H_r|/10.
-% \]
-% 
-% We now prove that for any $q$ parent of $q'$ (other than $r$), $W_q \geq (4/3)W_{q'}$.
-% Suppose not. Let $p, p'$ be the counterexample furthest from the root,
-% where $p$ is the parent of $p'$. 
-% Note that for $q$ and child $q'$ in $\cU_{p'}$, $W_q \geq (4/3)W_{q'}$.
-% We will apply \Clm{charge} for $\cU_{p'}$ to deduce that $\sum_{q \in \cU_{p'}} |q| \leq W_{p'}/20$.
-% Combining this with \Clm{loss} gives, $|\redH_{p'}| \geq 19W_{p'}/20$. By \Lem{geometric}, $|\redH_p| \geq (3/2)|\redH_{p'}|$.
-% Noting that $W_p \geq |\redH_p|$, we deduce that $W_p \geq (4/3)W_{p'}$.
-% Hence, $p, p'$ is not a counterexample, and more generally, there is no counterexample.
-% That completes the whole proof.
-% \end{proof}
-% 
-% 
-% \textcolor{red}{\huge The rest of this subsection needs updating}
-% 
-% We define a \emph{residue} $R_p$ for each $p \in \cU$. Suppose $p$ has children
-% $q_1, q_2, \ldots, q_k$ in $\cU$. Then $R_p = |\redH_p| - \sum_i |\redH_{q_i}|$. 
-% By definition, $|\redH_p| = \sum_{q \in \cU_p} R_q$. 
-% Note that $R_p$ can be negative. 
-% Now, define $R^+_p = \max(R_p,0)$, and set $W_p = \sum_{q \in \cU_p} R^+_q$, 
-% where $\cU_p$ is the subtree of $\cU$ rooted at $p$.
-% Observe that $W_p \geq |\redH_p|$. We also get an approximate converse.
-% 
-% 
-% \begin{claim} \label{clm:loss} For any path $p\in \cU$, $|\redH_p| \geq W_p - 2\sum_{q \in \cU_p} |q|$.
-% \end{claim}
-% 
-% \begin{proof} We write $W_p - |\redH_p| = \sum_{q \in \cU_p} (R^+_q - R_q)$ $= -\sum_{q \in \cU_p: R_q < 0} R_q$.
-% Consider $q \in \cU_p$ and denote the children in $\cU_p$ by $q'_1, q'_2, \ldots$.
-% Note that $R_q$ is negative exactly when $|\redH_q| < \sum_i |\redH_{q'_i}|$.
-% Traverse $\pmax(T)$ from $q'_i$ to $q$.
-% Other than $q$, all other nodes encountered are in the tall support chain of $q'_i$ and hence
-% do not affect its reduced heap.
-% The vertices of $\redH_{q'_i}$ that are deleted
-% are exactly those present in the path $q$. 
-% Any vertex in $q$ can be deleted from at most two of the reduced heaps (of the children of $q$ in $\cU_p$), since 
-% theses reduced heaps do not have an ancestor-descendant relationship.
-% Therefore when $R_q$ is negative, it is at most by $2|q|$.
-% We sum over all $q$ to complete the proof.
-% \end{proof}
-% 
-% \ignore{Let $D_q = H_q \setminus \bigcup_i H_{q'_i}$. In other words, as we traverse $\cU$,
-% $D_q$ is the set of vertices ``added" at $q$. Suppose we traverse $\pmax(T)$ from $q'_i$
-% to $q$. All vertices of $H_{q'_i}$ that lie on these paths are removed.
-% 
-% 
-%  and traverse $\pmax(T)$ up to $p$. Suppose we encounter
-% the paths $q_0 = q, q_1, q_2, \ldots, q_k = p$. The vertices in the residue $R_q$ that 
-% are removed from $H_p$ are exactly the vertices that lie in these paths. 
-% We split this contribution into tall and short paths, observing that the latter
-% are all unsupported. Hence, we can (crudely) upper bound the number of vertices
-% removed by $\sum_{q \in \cU_p} |q| + \sum_{q \in \cU_p} \sum_{h \in \cH} |h \cap q|$.
-% Each vertex can be removed from at most two distinct $R_q$'s, completing the proof.
-% }
-% 
-% 
-% The main challenge of the entire proof is bounding the sum of path lengths, which is done next.
-% We stress that the all the previous work is mostly the setup for this claim.
-% 
-% 
-% \begin{corollary} \label{cor:chargeCor}
-%  Let $r$ be the root of $\cU$, and suppose that for any paths $q,q' \in \cU\setminus \{r\}$, where $q$ is a parent of $q'$ in $\cU$, $W_q \geq (4/3)W_{q'}$.
-%  Then $\sum_{p \in \cU} |p| \leq W_r/20+|r|$.
-% \end{corollary}
-% \begin{proof}
-%  Let $c_1, \dots, c_m$ be the children of $r$ in $\cU$.  By definition, $W_r = \sum_i W_{c_i}+R^+_r \geq \sum_i W_{c_i}$.
-%  By \Clm{charge}, for each $c_i$ we have $W_{c_i}/20\geq \sum_{p \in \cU_{c_i}} |p|$.  Combining these to facts yields the claim.
-% \end{proof}
-% 
-%
-
-% \begin{claim} \label{clm:charge} Fix $p \in \cU$. Suppose for any $q,q' \in \cU_p$
-% where $q$ is a parent of $q'$ in $\cU_p$, $W_q \geq (4/3)W_{q'}$. Then $\sum_{q \in \cU_p} |q| \leq W_p/20$.
-% \end{claim}
-% 
-% \begin{proof} Since $q$ is an unsupported short path, $|q| < \sqrt{|H_q|}/100 \leq \sqrt{|\redH_q|}/99 \leq \sqrt{|W_q|}/99$.
-% We prove that $\sum_{q \in \cU_p} \sqrt{|W_q|}/99 \leq W_p/20$
-% by a charge redistribution scheme. Assume that each $q \in \cU_p$
-% starts with $\sqrt{W_p}/99$ units of charge. We redistribute this charge over all
-% nodes in $\cU_p$, and then calculate the total charge. For $q \in \cU_p$, spread
-% its charge to all nodes in $\cU_q$ proportional to $R^+$ values. In other words,
-% give $(\sqrt{W_q}/99)\cdot(R^+_{q'}/W_q)$ units of charge to each $q' \in \cU_q$.
-% 
-% After the redistribution, let us compute the charge deposited at $q$. Every ancestor in $\cU_p$ 
-% $q = a_0, a_1, a_2, \ldots, a_k$ contributes to the charge at $q$. The charge is expressed in the following
-% equation. We use the assumption that $W_{a_i} \geq (4/3) W_{a_{i-1}}$ and hence $W_{a_i} \geq (4/3)^i W_{a_0} \geq (4/3)^i$, 
-% as $a_0$ is an unsupported short path and hence $W_{a_0}\geq 1$. 
-% $$ ({R^+_q}/99)\sum_{a_i} 1/\sqrt{W_{a_i}} \leq ({R^+_q}/99)\sum_{a_i} (3/4)^{i/2} \leq R^+_q/20 $$
-% The total charge is $\sum_{q \in \cU_p} R^+_p/20 = W_p/20$. 
-% \end{proof}
-
-% We wrap it all up by proving part (ii) of \Lem{cu-root}.
-%
-% \begin{claim} $\sum_{p \in \cU} |p| \leq |H_{r(\cU)}|/10$.
-% \end{claim}
-% 
-% % \begin{proof} We use $r$ for $r(\cU)$. Our aim is to simply apply \Clm{charge} from $r(\cU)$. 
-% % Suppose $W_q \geq (4/3)W_{q'}$ (for any choice in $\cU$ of $q$ parent of $q'$). By \Clm{charge}, $\sum_{p \in \cU} |p| \leq W_{r}/20$.
-% % By \Clm{loss}, $|\redH_r| \geq W_r - 2\sum_{p \in \cU} |p|$. By simple algebra, we deduce, $\sum_{p \in \cU} |p| \leq |\redH_r|/10 = |H_r|/10$.
-% % 
-% % We now prove that for any $q$ parent of $q'$, $W_q \geq (4/3)W_{q'}$.
-% % Suppose not. Let $p, p'$ be the counterexample furthest from the root,
-% % where $p$ is the parent of $p'$. 
-% % Note that for $q$ and child $q'$ in $\cU_{p'}$, $W_q \geq (4/3)W_{q'}$.
-% % We will apply \Clm{charge} for $\cU_{p'}$ to deduce that $\sum_{q \in \cU_{p'}} |q| \leq W_{p'}/20$.
-% % By \Clm{loss}, $|\redH_{p'}| \geq 19W_{p'}/20$. By \Lem{geometric}, $|\redH_p| \geq (3/2)|\redH_{p'}|$.
-% % Noting that $W_p \geq |\redH_p|$, we deduce that $W_p \geq (4/3)W_{p'}$.
-% % Hence, $p, p'$ is not a counterexample, and more generally, this is no counterexample.
-% % That completes the whole proof.
-% % \end{proof}
 
@@
 thus completing the proof of \Thm{runtime}.
-
-% \begin{claim} \label{clm:above} $\sum_{p \in \cL'} |H_p| \log |H_p| = O(\sum_{q \in \cH} \sum_{v \in q} \log |H_v|)$.
-% \end{claim}
-% 
-% \begin{proof} 
-% Pick $p \in \cL'$. 
-% As we traverse the tall support chain of $p$, there
-% are at least $|H_p|/100$ vertices of $H_p$ that lie in these paths. These are encountered in
-% a fixed order. Let $H'_p$ be the first $|H_p|/200$ of these vertices. When $v \in H'_p$
-% is encountered, there are $|H_p|/200$ vertices of $H_p$ not yet encountered. Hence,
-% $|H_v| \geq |H_p|/200$. Hence, $|H_p|\log |H_p| = O(\sum_{v \in H'_p} \log |H_v|)$.
-% Since all the vertices lie in tall paths,
-% we can write this as $O(\sum_{q \in \cH} \sum_{v \in H'_p \cap q} \log |H_v|)$.
-% Summing over all $p$, the expression is $\sum_{q \in \cH} \sum_{p \in \cL'} \sum_{v \in H'_p \cap q} \log |H_v|$.
-% %and $\sum_{p \in \cL'} |H_p|\log |H_p| = O(\sum_{p \in \cL'} \sum_{v \in H'_p} \log |H_v|)$.
-% 
-% Consider any $v \in H'_p$.  Let $S$ be the set of paths $\widetilde{p}\in \cL'$ such that 
-% $v\in H'_{\widetilde{p}}$.  We now show $|S|\leq 2$ (i.e.\ it contains at most one path other than $p$).
-% First observe that any two paths in $S$ must be unrelated (i.e.\ $S$ is an anti-chain), 
-% since paths which have an ancestor-descendant relationship have disjoint tall support chains.  
-% However, any vertex $v$ receives exactly one color from each of its two subtrees (in $T$), and therefore $|S|\leq 2$ since any two 
-% paths which share descendant leaves in $T$ (i.e.\ their heaps are tracking the same color) must have an ancestor-descendant relationship.
-% 
-% In other words, any $\log |H_v|$ appears at most twice in the above triple summation.
-% Hence, we can bound it by $O(\sum_{q \in \cH} \sum_{v \in q} \log |H_v|)$.
-% \end{proof}
-
-% The first term (unsupported short paths) can be charged to the second term (supported short paths).
-% This is where the critical \Lem{cu-root} plays a role.
 
@@
 \]
-
-%\Thm{rain-time} proves that $\rain$ takes $O(|\MM|)$ time
-
-%\Lem{rain-1} proves that output of $\rain$ are extemum dominant.
-
-%\Clm{rain-reeb} shows that given the contour trees of the output manifolds of $\rain$, the full contour tree can be constructed in $O(|\MM|)$ time:
-
-%\Thm{contour-tree} implies that given an $\jc(\MM)$ for an etremum dominant $\MM$, we can obtain $\cC(\MM)$ in by simply sticking on the non-dominant minima at their respective splits (which can easily be done in linear time).
 \end{proof}
@@
 \section{ \Sec{algo} Algorithm Example }\label{sec:example}
-
-% \textcolor{red}{
-% In the example below, I assumed that the maxima are already considered ``processed'', so that there are not additional trivial steps.  
-% This actually makes more sense since the $att()$ should really only ever point to already processed vertex.  Therefore, I suggest 
-% changing line 3 of init to ``Mark all maxima as processed.  Mark all other critical $v$ as unprocessed.''
-% Perhaps line 4 needs to be changed also?
-% }
 

### 2021-09-23 erratum/updated/erratum.tex
updated erratum

--- 2017-05-21 paint-contour-tree.tex
@@
       Santa Cruz, CA, 95064, USA; %
-      \texttt{\si{scomandu}@\si{ucsc}.\si{edu}} %
+      \texttt{\si{sesh}@\si{ucsc}.\si{edu}} %
   }
@@
 Avoiding the Global Sort: 
-\break A Faster Contour Tree Algorithm%
+\break A Faster Contour Tree Algorithm: Erratum}
 % \footnote{The full updated version of this paper is available on the arXiv \cite{rs-mvrms-14}}%
-}
+
 \date{}
@@
 \begin{abstract}
-We revisit the classical problem of computing the \emph{contour tree} of a scalar field $f:\MM \to \RR$, where $\MM$ is a triangulations of a ball in $\RR^d$.
-The contour tree is a fundamental topological structure that tracks the evolution of level sets of $f$ and has numerous applications in data analysis and visualization.
-
-All existing algorithms begin with a global sort of at least all critical values of $f$, which can require (roughly) $\Omega(n\log n)$ time, where $n$ is the number of vertices of the mesh.
-Existing lower bounds show that there are pathological instances where this sort is required.
-We present the first algorithm whose time complexity depends on the contour tree structure, and avoids the global sort for non-pathological inputs.
-(We assume that the Morse degree of the function $f$ at any point is at most $3$.) If $C$ denotes the set of critical points in $\MM$, the running time is roughly $O(N+\sum_{v \in C} \log \ell_v)$, where $\ell_v$ is the depth of $v$ in the contour tree and $N$ is the total complexity of $\MM$.
-This matches all existing upper bounds, but is a significant asymptotic improvement when the contour tree is short and fat.
-Specifically, our approach ensures that any comparison made is between nodes that are either adjacent in $\MM$ or in the same descending path in the contour tree, allowing us to argue strong optimality properties of our algorithm.
-
-Our algorithm requires several novel ideas: partitioning $\MM$ in well-behaved portions, a local growing procedure to iteratively build contour trees, and the use of heavy path decompositions for the time complexity analysis.
+This erratum points out a serious bug in our paper, ``Avoiding the Global Sort: A Faster Contour Tree Algorithm"~\cite{rs-17},
+as pointed out by Rote~\cite{rote}.
+
 \end{abstract}
 
-\newpage
 \pagenumbering{arabic}
 
-\section{Introduction}
-
-Geometric data is often represented as a function $f: \RR^d \to \RR$. Typically, a finite representation is given
-by considering $f$ to be piecewise linear over some triangulated mesh (i.e.\ simplicial complex) $\MM$ in $\RR^d$.
-\emph{Contour trees} are a topological structure used to represent and visualize
-the function $f$. 
-It is convenient to think of $f$ as a simplicial complex sitting in $\RR^{d+1}$,
-with the last coordinate (i.e.\ height) given by $f$. Imagine sweeping the hyperplane $x_{d+1} = h$
-with $h$ going from $+\infty$ to $-\infty$. At every instance, the intersection of this plane 
-with $f$ gives a set of connected components, the \emph{contours} at height $h$. As the sweeping 
-proceeds various events occur: new contours are created or destroyed, contours merge into each other or
-split into new components. 
-The contour tree is a concise representation of these events \cite{kobps-ctsssit-97,csa-cctad-03}.
-%Throughout we follow the definition of contour trees from \cite{kobps-ctsssit-97} which includes all changes in topology.  
-%For $d>2$, some subsequent works, such as \cite{csa-cctad-03}, only include changes in the number of components.
-
-If $f$ is smooth, all points where the gradient of $f$ is zero are \emph{critical points}.
-These points are the ``events" where the contour topology changes. The contour tree tracks a specific
-subset of events, called ``joins" and ``splits".
-We provide formal definitions later.
-An edge of the contour tree connects two such critical points if one event immediately ``follows"
-the other as the sweep plane makes its pass.  
-\Fig{relative} and \Fig{sorting} 
-show examples of simplicial complexes, with heights and their contour trees.
-Think of the contour tree edges as pointing downwards. Leaves are either maxima or minima,
-and internal nodes are either ``joins" or ``splits".
-
-% so we can talk of the ``up-degree" and ``down-degree" of vertices.
-% Leaves of down-degree $1$ are maxima, because a new contour is created at such points. Leaves of up-degree
-% $1$ are minima, because a contour is destroyed there. Up-degree $2$ vertices are ``joins", where contours
-% merge, and down-degree $2$ vertices are ``splits".  When $d=2$ (as in \Fig{relative}), contours are closed loops.
-% In dimensions $d \geq 3$, we may have internal vertices of both up and down-degree $1$. 
-% For example, these could be events where contours gain or lose handles.
-
-Consider $f: \MM \to \RR$, where $\MM$ is the triangulation of a $d$-dimensional ball with $n$ vertices, $N$ faces in total, and $t \leq n$ critical points. 
-(We assume that $f:\MM \to \RR$ is a linear interpolant over distinct valued vertices, with Morse degree at most $3$, see \Def{deg}. 
-The degree assumption can be achieved via vertex unfolding, as discussed later,
-and is standard in this literature~\cite{kobps-ctsssit-97}.)
-A fundamental result in this area is the algorithm of Carr, Snoeyink, and Axen to compute contour trees,
-which runs in $O(n\log n + N\alpha(N))$ time~\cite{csa-cctad-03} (where $\alpha(\cdot)$ denotes the inverse
-Ackermann function). In practical applications, $N$ is typically $\Theta(n)$
-(certainly true for $d=2$). The most expensive operation is an initial sort of all the vertex
-heights. Chiang \etal build on this approach to get a faster algorithm that only sorts the critical vertices,
-yielding a running time of $O(t\log t + N)$~\cite{cllr-sooscctmp-05}.
-Common applications for contour trees involve turbulent combustion or noisy data, where the number of critical points is likely to be $\Omega(n)$.
-There is a worst-case
-lower bound of $\Omega(t\log t)$ by Chiang \etal \cite{cllr-sooscctmp-05}, based on a construction of Bajaj \etal \cite{BaKr+98}.
-
-All previous algorithms begin by sorting (at least) the critical points. Can we beat this sorting bound for certain instances,
-and can we characterize which inputs are hard? Intuitively, points that are incomparable in the contour tree
-do not need to be compared. Look at \Fig{relative} to see such an example. All previous algorithms waste time
-sorting all the maxima. Also consider the surface of \Fig{sorting}. The final contour tree is basically two binary
-trees joined at their roots, and we do not need the entire sorted order of critical points
-to construct the contour tree.
-
-\begin{figure}[h]\centering
-    \includegraphics[width=.23\linewidth]{figs/spikey1}%
-    \hspace{.37in}
-    \includegraphics[width=.23\linewidth]{figs/spikey2}%
-    \hspace{.41in}
-    \includegraphics[width=.21\linewidth]{figs/spikeyTree}
-    \caption{Two surfaces with different orderings of the maxima, but the same contour tree.}
-    \label{fig:relative}
-\end{figure}
-
-\begin{figure}[h]\centering
-    \includegraphics[width=.35\linewidth,height=.15\linewidth]{figs/balanced4}
-    \hspace{.3in}
-    \includegraphics[width=.35\linewidth,height=.15\linewidth]{figs/trees}
-    \caption{On left, a surface with a balanced contour tree, but whose join and split trees have long tails.  
-    On right (from left to right), the contour, join and split trees.}
-    \label{fig:sorting}
-\end{figure}
-
-Our main result gives an affirmative answer. Remember that we can consider
-the contour tree as directed from top to bottom. For any node $v$ in the tree, let $\ell_v$ denote the length of the 
-longest directed path passing through $v$.
-
-\begin{theorem} \label{thm:main-corr} Consider a simplicial complex $f:\MM \to \RR$, where $\MM$
-is a triangulation of a $d$-dimensional ball.
-Let $N$ be the number of facets in $\MM$.
-Let the Morse degree of $f$ at any point be at most $3$.
-Denote
-the contour tree by $T$ with vertex set (the critical points) $C(T)$. 
-There exists an algorithm to compute $T$
-in $O(\sum_{v \in C(T)} \log \ell_v + t\alpha(t) + N)$ time. Moreover, this algorithm only
-compares function values at pairs of vertices that are ancestor-descendant in $T$ or adjacent in $\MM$.
-\end{theorem}
-
-Essentially, the ``run time per critical point" is determined by the height/depth of the point in the contour tree.
-This bound immediately yields a run time of $O(t\log D + t\alpha(t) + N)$,
-where $D$ is the directed diameter of the contour tree. 
-This is a significant asymptotic improvement for short and fat contour trees. For example, if the tree is balanced,
-then we get a bound of $O(t\log\log t)$.
-Even if $T$ contains a long path of length $O(t/\log t)$, but is otherwise short, we get the improved bound of $O(t\log\log t)$.
-
-\subsection{A refined bound with optimality properties}\label{sec:more-refined}
-
-\Thm{main-corr} is a direct corollary of a stronger but more cumbersome theorem.
-
-\begin{definition}
-\label{def:path} 
-For a contour tree $T$, a \emph{leaf path} is any path in $T$ containing a leaf, 
-which is also monotonic in the height values of its vertices. 
-Then a \emph{path decomposition}, $P(T)$, is a partition of the vertices of $T$ into a set of vertex disjoint leaf paths.  
-\end{definition}
-
-\begin{theorem} 
-\label{thm:main-alg} 
-(Consider the input setting of \Thm{main-corr}.)
-There is a deterministic algorithm to compute the contour tree, $T$, whose running time is $O(\sum_{p \in P(T)} |p|\log |p| + t\alpha(t) + N)$,
-where $P(T)$ is a specific path decomposition (constructed implicitly by the algorithm).
-The number of comparisons made is $O(\sum_{p \in P(T)} |p|\log |p| + N)$.
-In particular, comparisons are only made between pairs of vertices that are ancestor-descendant in $T$ or adjacent in $\MM$.
-\end{theorem}
-
-Note that \Thm{main-corr} is a direct corollary of this statement. For any $v$, $\ell_v$ is at least
-the length of the path in $P(T)$ that contains $v$.
-This bound is strictly stronger,
-since for any balanced contour tree, the run time bound of \Thm{main-alg} is $O(t\alpha(t) + N)$, and $O(N)$
-comparisons are made. (Since for a balanced tree, one can show $\sum_{p \in P(T)} |p|\log |p|= O(t)$.)
-
-The bound of \Thm{main-alg} may seem artificial, since it actually depends on the $P(T)$ that
-is implicitly constructed by the algorithm. Nonetheless, we prove that the algorithm of \Thm{main-alg}
-has strong optimality properties. For convenience, fix some value $t=\Omega(N)$, and consider the set
-of terrains ($d=2$) with $t$ critical points. The bound of \Thm{main-alg} takes values ranging
-from $t$ to $c t\log t$, for some constant $c$. Consider some $\thmConst \in [t,t\log t]$, and consider the set of terrains
-where the algorithm makes $\alpha \thmConst$ comparisons, for any constant $\alpha$. Then \emph{any algorithm} must make roughly $\thmConst$
-comparisons in the worst-case over this set. (Further details are in \Sec{lb}.)
-
-\begin{theorem} 
-\label{thm:main-lb}
-There exists some absolute constant $\alpha$ such that the following holds.
-For sufficiently large $t$ and any $\thmConst\in [t, t\log t]$, consider the set $\bF_\thmConst$ of terrains with $t$ critical points such that
-the number of comparisons made by the algorithm of \Thm{main-alg} on these terrains is in $[\thmConst,\alpha \thmConst]$.
-Any algebraic decision tree that correctly computes
-the contour tree on all of $\bF_\thmConst$ has a worst case running time of $\Omega(\thmConst)$.
-\end{theorem}
-% 
-% \myparagraph{Beyond worst-case analysis.} Our result provides understanding of the complexity
-% of contour tree construction beyond the standard worst-case analysis. Our theorem and lower bound 
-% characterize hard instances.
-% % explain \emph{when}
-% % computing contour trees is expensive, and provide insight into the hardness of particular instances. 
-% %It is intriguing
-% %that the cost (as defined earlier) roughly captures the hardness of an instance.
-% Trees that are short and fat are easy to compute, whereas trees with long paths are hard.
-% This is tangentially related to the concept of instance-optimal algorithms, as defined by Afshani \etal \cite{abc-ioga-09}.
-% Their notion of optimality is far more refined that what we prove, but it shares the flavor of understanding the spectrum
-% of easy to hard instances. In general, our analysis forces one to look closely
-% at the relationship between sorting and the contour tree problem. 
-%
-
-\subsection{Previous work}
-
-Contour trees were first used to study terrain maps by Boyell and Ruston, and Freeman and Morse~\cite{BoRu63,FrMo67}.
-Contour trees have been applied in analysis of fluid mixing, combustion simulations,
-and studying chemical systems~\cite{LaBe+06,BrWe+10,BeWe+11,BrWe+11,MaGr+11}. Carr's thesis~\cite{c-tmi-04} gives various
-applications of contour trees for data visualization and is an excellent reference for contour tree definitions and algorithms.
-
-The first formal result was an $O(N\log N)$ time algorithm for functions over 2D 
-meshes and an $O(N^2)$ algorithm for higher dimensions, by van Kreveld \etal \cite{kobps-ctsssit-97}. 
-Tarasov and Vyalyi \cite{tv-cct-98} improved the running time to $O(N\log N)$ for the 3D case.
-The influential paper of Carr \etal \cite{csa-cctad-03} improved the running time for all dimensions to $O(n\log n + N\alpha(N))$.
-Pascucci and Cole-McLaughlin \cite{pc-ectls-02} provided an $O(n+t\log n)$ time algorithm for 
-$3$-dimensional structured meshes. Chiang \etal \cite{cllr-sooscctmp-05} provide an unconditional $O(N+t\log t)$ algorithm.
-
-Contour trees are a special case of Reeb graphs, a general topological representation for real-valued functions
-on any manifold. Algorithms for computing Reeb graphs
-is an active topic of research~\cite{sk-crgacs-91,cehnp-lrbm-03,PaScBr07,DoNa09,HaWaWe10,Pa12}, where
-two results explicitly reduce to computing contour trees~\cite{TiGySi09,DoNa13}.
-
-% \paragraph{Organization.}  
-% \Sec{basics} contains basic definition for contour trees.  In \Sec{approach} we give a high level overview and intuition for our approach 
-% for computing contour trees.  \Sec{surgery} and \Sec{rain} describe how we break the intial input into nicely behaved (extemum dominant) pieces. 
-% \Sec{extreme} shows that computing the contour tree for these nicely behaved pieces is roughly equivalent to computing the join tree. 
-% \Sec{paint} contains our algorithm for computing join trees, which along with previous sections gives our algorithm for contour trees.  
-% Finally, in \Sec{pathDecomp} and \Sec{lb} we relate the running time of our algorithm (providing upper and lower bounds) to a particular decomposition of the contour tree.
-
-\section{Contour tree basics} \label{sec:basics}
-
-We detail the basic definitions about contour trees, following the terminology of Chapter 6 of Carr's thesis \cite{c-tmi-04}.
-All our assumptions and definitions
-are standard for results in this area, though there is some variability in notation.
-The input is a continuous piecewise-linear function $f:\MM \to \RR$, where $\MM$ is a fully triangulated simplicial complex of a topological ball in $\RR^d$,
-except for specially designated \emph{boundary facets}. We assume that all pairs of vertices from $\MM$ have distinct function values, except for pairs from the same boundary facet,
-and that $f$ is only explicitly defined on the vertices, and all other values are obtained by linear interpolation.
-
-We assume that the boundary values satisfy a special property. This is mainly for convenience
-in presentation.
-
-\begin{definition} \label{def:bound} The function $f$ is \emph{boundary critical} if the following holds.
-Consider a boundary facet $F$. All vertices of $F$ have the same function value. Furthermore, all
-neighbors of vertices in $F$, which are not also in $F$ itself, 
-either have all function values strictly greater than or all function values strictly less than the function value at $F$.
-\end{definition}
-
-Boundary facets allow us to capture the resulting surface pieces after our algorithm makes a horizontal cut. 
-The above definition is convenient, as any point within a given facet has a well-defined height, including the boundary facets.
-
-We think of the dimension $d$, as constant, and assume that $\MM$ is represented in a data structure that allows constant-time access to neighboring simplices
-in $\MM$ (e.g. incidence graphs~\cite{Ed87}). (This is analogous to a doubly connected edge list, but for higher dimensions.)
-Observe that $f:\MM \rightarrow \RR$ can be thought of as a $d$-dimensional simplicial complex living in $\RR^{d+1}$, 
-where $f(x)$ is the ``height" of a point $x \in \MM$, which is encoded in the representation of $\MM$. 
-Specifically, rather than writing our input as $(\MM,f)$, we abuse notation and typically just write $\MM$ to denote the lifted complex.
-
-\begin{definition} \label{def:level} The \emph{level set} at value $h$ is the set $\{x\in \MM| f(x) = h\}$.
-A \emph{contour} is a connected component of a level set. An \emph{$h$-contour} is a contour where $f$-values are $h$.
-\end{definition}
-
-Note that a contour that does not contain a boundary is itself a simplicial complex of one dimension lower, and is represented (in our algorithms) as such.
-We let $\delta$ and $\eps$ denote infinitesimals. Let $B_\eps(x)$ denote a ball of radius $\eps$ around $x$, and let 
-$f|B_\eps(x)$ be the restriction of $f$ to $B_\eps(x)$.
-The following definition of a critical point is not standard,
-but is convenient for our presentation.
-
-\begin{definition} \label{def:deg} The \emph{Morse up-degree} of $x$ is the number of $(f(x) + \delta)$-contours of $f|B_\eps(x)$
-as $\delta, \eps \rightarrow 0^+$. The \emph{Morse down-degree} is the number of $(f(x) - \delta)$-contours of $f|B_\eps(x)$
-as $\delta, \eps \rightarrow 0^+$.
-
-A \emph{regular} point has both Morse up-degree and down-degree $1$. A \emph{maximum} has Morse up-degree $0$, while
-a \emph{minimum} has Morse down-degree $0$. A \emph{Morse Join} has Morse up-degree strictly greater than $1$,
-while a \emph{Morse Split} has Morse down-degree strictly greater than $1$. Non-regular points are called \emph{critical}.
-\end{definition}
-
-The set of critical points is denoted by $\cV(f)$.
-Because $f$ is piecewise-linear, all critical points are vertices in $\MM$. 
-A value $h$ is called \emph{critical}, if $f(v) = h$, for some $v \in \cV(f)$. 
-A contour is called \emph{critical}, if it contains a critical point, and it is called \emph{regular} otherwise.
-
-The critical points are where certain topological properties
- of level sets change.
-By assuming that our simplicial complex is boundary critical, the vertices on a given boundary are either collectively all maxima or all minima.  
-We abuse notation and refer to this entire set of vertices as a maximum or minimum.
-
-\begin{definition} \label{def:equiv} Two regular contours $\psi$ and $\psi'$ are \emph{equivalent} 
-if there exists an $f$-monotone path $p$ connecting a point in $\psi$ to $\psi'$,
-such that no $x \in p$ belongs to a critical contour.
-\end{definition}
-
-This equivalence relation gives a set of \emph{contour classes}. 
-Every such class maps to intervals
-of the form $(f(x_i),f(x_j))$, where $x_i, x_j$ are critical points. Such a class is said
-to be created at $x_i$ and destroyed at $x_j$.
-
-\begin{definition} \label{def:tree} The \emph{contour tree} is the graph on vertex set $\cV= \cV(f)$, where
-edges are formed as follows. For every contour class that is created at $v_i$ and destroyed $v_j$,
-there is an edge $(v_i,v_j)$. (Conventionally, edges are directed from higher to lower function value.)
-\end{definition}
-
-We denote the contour tree of $\MM$
-by $\reeb(\MM)$. The corresponding node and edge sets are denoted as $\cV(\cdot)$ and $\cE(\cdot)$.
-It is not immediately obvious that this graph is a tree, but alternative definitions of the contour tree
-in~\cite{csa-cctad-03} imply this is a tree. 
-Since this tree has height values associated with the vertices, we can talk about up-degrees and down-degrees in $\reeb(\MM)$.
-
-We assume that the Morse up (or down) degree of any point
-in $\MM$ is at most $2$, and that the total Morse degree
-is at most $3$. This immediately implies analogous bounds
-on the up and down-degrees in $\reeb(\MM)$. This
-is standard in the literature
-among others, though it can be challenging to enforce in 
-practice. Theoretically, a manifold with points of higher Morse degree
-can be converted to one with the degree condition through
-a process of \emph{vertex unfolding} (refer to
-Section 6.3 in~\cite{HaEd10}). This effectively converts
-a multi-saddle to a set of ordinary saddles.
-%A detailed discussion is provided in the end
-%of Section 2 and beginning of Section 3 in~\cite{BaKr+98}.
-For 2D manifolds,
-this conversion does not increase the surface complexity.
-In higher dimensions, the unfolding can increase complexity if
-$\MM$ has vertices of non-constant degree.
-
-% Similar to \cite{kobps-ctsssit-97} (among others), multi-saddles are treated as a set of ordinary saddles, which can be realized 
-% via vertex unfolding (which can increase surface complexity if multi-saddle degrees are allowed to be super-constant).
-% Therefore, to simplify the presentation, for the remainder of the paper up and down-degrees are at most $2$, and total degree is at most $3$.
-%
-%We assume there are no ``multi-saddles", so up and down-degrees are at most $2$, and the total degree is at most $3$. 
-%This is again a standard assumption in topological algorithms and can be achieved by vertex unfolding (Section 6.3 in~\cite{HaEd10}).
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\subsection{Some remarks}
-{\bf 1)}
-Note that if one intersects $\MM$ with a given ball $B$, then a single contour in $\MM$ might be split into more than one contour in the intersection. 
-In particular, two $(f(x)+\delta)$-contours of $f|_{B_\eps(x)}$, given by \Def{deg}, might actually be the same contour in $\MM$. 
-Alternatively, one can define the up-degree (as opposed to \emph{Morse} up-degree) as the number of $(f(x)+\delta)$-contours (in the full $\MM$)
-that intersect $B_\eps(x)$, a potentially smaller number. This up-degree
-is exactly the up-degree of $x$ in $\reeb(\MM)$. (Analogously, for down-degree.)
-When the Morse up-degree is $2$ but the up-degree is $1$, the topology of the level set changes but not by the number of connected components changing.
-When $d=2$, this distinction is not necessary, since any point with Morse degree strictly greater than $1$ will have degree strictly greater than $1$ in $\reeb(\MM)$.\\
-
-\noindent{\bf 2)}
-As Carr points out in Chapter 6 of his thesis, the term contour tree can be used for a family of related structures.
-Every regular vertex in $\MM$ is associated with an edge in $\reeb(\MM)$, and sometimes the
-vertex is explicitly placed in $\reeb(\MM)$ (by subdividing the respective edge). This is referred
-to as augmenting the contour tree, and it is common to augment $\reeb(\MM)$ with all vertices.
-Alternatively, one can smooth out all vertices of up-degree and down-degree $1$ to get the
-unaugmented contour tree. (For $d=2$, there are no such vertices in $\reeb(\MM)$.) 
-The contour tree of \Def{tree} is the typical definition in all results on output-sensitive
-contour trees.
-\Thm{main-alg} is applicable for any augmentation of $\reeb(\MM)$ with a predefined
-set of vertices, though we will not delve into these aspects in this paper.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\section{A tour of the new contour tree algorithm} \label{sec:approach}
-We provide a high-level description of the main ideas of subsequent sections.
-
-\myparagraph{Do not globally sort:} The starting point for this work is \Fig{relative}. We have two terrains
-with exactly the same contour tree, but different orderings of (heights of) the critical points.
-Turning it around, we cannot deduce the full height ordering of critical points from the contour tree, and so 
-sorting all critical points is computationally unnecessary.
-In \Fig{sorting}, the contour tree consists of two balanced binary trees, one of the joins,
-another of the splits. Again, it is not necessary to know the relative ordering between the mounds
-on the left (or among the depressions on the right) to compute the contour tree. Yet some ordering
-information is necessary: on the left, the little valleys are higher than the big central valley,
-and this is reflected in the contour tree. Leaf paths in the contour tree have points
-in sorted order, but incomparable points in the tree are unconstrained.
-How do we sort exactly what is required, without knowing the contour tree in advance?
-
-\myparagraph{Join and split trees:}
-The standard contour tree algorithm of Carr, Snoeyink, and Axen~\cite{csa-cctad-03}, builds
-two different trees, called the join and split trees, and then merges them together to get the contour tree.
-Consider sweeping down the hyperplane $x_{d+1} = h$ and taking the \emph{superlevel} sets. These 
-are the connected components of the portion of $\MM$ above height $h$. For a terrain, the superlevel
-sets are a collection of ``mounds". As we sweep downwards, these mounds keep joining each other,
-until finally, we end up with all of $\MM$. The join tree, $\cJ(\MM)$, tracks exactly these events. 
-Formally, let $\MM^+_v$ denote the simplicial complex induced on the subset of 
-vertices higher than $v$. 
-Refer to \Fig{sorting} for the join tree of a terrain. Note that nothing happens at splits, but these
-are still put as vertices in the join tree. The split tree
-is obtained by simply inverting this procedure, sweeping upwards and tracking sublevel sets.
-
-A major insight of~\cite{csa-cctad-03} is an ingeniously simple linear time procedure to construct
-the contour tree from the join and split trees. So the bottleneck is computing these trees. Observe
-in \Fig{sorting} that the split vertices form a long path in the join tree (and vice versa). Therefore, constructing
-these trees forces a global sort of the splits, an unnecessary computation for the contour tree. Unfortunately, in general (i.e.\ unlike \Fig{sorting}) 
-the heights of joins and splits may be interleaved in a complex manner, and hence the final merging of~\cite{csa-cctad-03}
-to get the contour tree requires having the split vertices in the join tree. Without
-this, it is not clear how to get a consistent view of both joins and splits, required for the contour tree.
-
-Our aim is to break $\MM$ into smaller pieces, where this unnecessary computation can be avoided.
-
-\myparagraph{Cutting $\MM$ into extremum dominant pieces:} We define a simplicial complex
-endowed with a height to be \emph{extremum dominant} if there exists only a single minimum, or only a single maximum.
-(When formally defined in \Sec{rain}, extremum dominant complexes will allow additional trivial minima or maxima which are a small complication resulting from the following cutting procedure.)
-We first cut $\MM$ into disjoint extremum dominant pieces in linear time.
-Take an arbitrary maximum $x$ and imagine torrential rain at the maximum. The water flows down, wetting any point that has a non-ascending path 
-from $x$. We end up with two portions, the wet part of $\MM$ and the dry part. This is similar
-to \emph{watershed} algorithms used for image segmentation~\cite{RoMe00}. The wet part is obviously connected and
-can be shown to be extremum dominant. We can cut along the interface of the wet and dry, remove
-the wet part, and recurse on the dry parts. This is carefully done to ensure that the total complexity
-of the pieces is linear (see \Sec{rain} for details).
-
-We now have carved $\MM$ up into a collection of extremum dominant pieces. 
-In \Sec{surgery} we prove a simple contour surgery theorem allowing us to build the contour tree of $\MM$
-from the contour trees of these various pieces.  Specifically, we argue our cutting in $\MM$ was done in such a way that 
-each cut corresponds to cutting only a single edge in the contour tree.  Informally this means gluing $\MM$ back together along these 
-cuts corresponds to gluing the contour tree together.
-
-Ultimately, the reason for cutting $\MM$ into extremum dominant pieces is that such pieces have simpler contour tree structure. 
-Specifically, using ideas from~\cite{csa-cctad-03}, in \Sec{extreme} we prove that the contour tree of an extremum dominant complex is (basically) just the join tree. 
-Thus there is no longer a disconnect between the amount of sorting required for the contour tree versus the join tree, 
-and so the problem of computing contour trees efficiently reduces to that for join trees.
-
-\myparagraph{Join trees from painted mountaintops:} Our main result is a faster algorithm for join trees.
-The key idea is \emph{paint spilling}. Start with each maximum having a large can of paint, with distinct
-colors for each maximum. In arbitrary order, spill paint from each maximum, wait till it flows down,
-then spill from the next, etc. Paint is viscous, and only flows down edges. \emph{It does not
-paint the interior of faces with dimension strictly larger than 1.} 
-Furthermore, paints do not mix, so each edge receives a unique color, decided by the first paint to reach it.  
-%
-Note that the raining procedure, used above to carve the input into extremum dominant pieces, 
-cannot be used here as it wets interiors of higher dimensional faces and so raining from each maximum may significantly increase the input complexity (see \Fig{order}).
-On the other hand, the interface between two different colors is \emph{not} a contour, and so the join trees of each color class cannot just be simply glued together.
-
-% In the following, $[n]$ denotes $\{1, \ldots, n\}$, for any natural number $n$.
-% \begin{definition} \label{def:paint1} \InSoCGVer{(Restatement of \Def{paint2}.)} Let the 1-skeleton of $\NN$ have edge set $E$ and maxima $X$.
-% A  \emph{painting} of $\NN$ is a map $\chi:X \cup E \to [|X|]$ with the following property. 
-%  Consider an edge $e$. There exists a descending path from some maximum $x$ to $e$
-% 	consisting of edges in $E$, such that all edges along this path have the same color as $x$. 
-% 
-% An \emph{initial} painting has the additional property that the restriction $\chi:X \to [|X|]$ is a bijection.
-% \end{definition}
-
-% Note that a painting colors edges, and not vertices (except for maxima). Our definition also does not require the timing aspect
-% of iterating over colors, though that is one way of painting $\NN$. We begin with an initial painting, since all maximum
-% colors are distinct. A few comments on paint vs water. The interface between two regions of different color
-% is \emph{not} a contour, and so we cannot apply the divide-and-conquer
-% approach of contour surgery. On the other hand, painting does not cut $\NN$, so there is no increase in complexity.
-% Clearly, an initial painting can be constructed in $O(N)$ time. This is the tradeoff between water and paint.
-% Water allows for an easy divide-and-conquer, at the cost of more complexity in the input. For an extremum
-% dominant input, using water to divide the input $\NN$ raises the complexity too much.
-
-Our algorithm incrementally builds $\cJ(\MM)$ from the leaves (maxima) to the root (dominant minimum), by repeatedly merging colors and components together.  
-We say that vertex $v$ is \emph{touched} by color $c$, if there is a $c$-colored edge with lower endpoint $v$.
-Let us focus on the initial painting, where the colors have 1-1 correspondence with the maxima. Refer to the left part of \Fig{colors}.
-Consider two sibling leaves $\ell_1, \ell_2$ and their common parent $v$. The leaves are maxima,
-and $v$ is a join that ``merges" $\ell_1, \ell_2$. 
-In that case, there are ``mounds" corresponding to $\ell_1$ and $\ell_2$ that merge
-at a valley $v$. Suppose this was the entire input, and $\ell_1$ was colored blue and $\ell_2$ was colored red. 
-Both mounds are colored completely blue or red, while $v$ is touched by both colors.
-So this indicates that $v$ joins the blue maximum and red maximum in $\cJ(\MM)$.
-
-This is precisely how we hope to exploit the information in the painting. We prove later that
-when some join $v$ has exactly two colors among all incident edges with $v$ as the lower endpoint, the corresponding maxima (of those colors) are exactly
-the children of $v$ in $\cJ(\MM)$. To proceed further, we ``merge" the colors red and blue into a new color, purple. In other words,
-we replace all red and blue edges by purple edges. This indicates that the red and blue maxima have been
-handled. Imagine flattening the red and blue mounds until reaching $v$, so that the former join $v$
-is now a new maximum, from which purple paint is poured. In terms of $\cJ(\MM)$, this
-is equivalent to removing leaves $\ell_1$ and $\ell_2$, and making $v$ a new leaf.
-Alternatively, $\cJ(\MM)$ has been constructed up to $v$, and it remains to determine $v$'s parent.
-The merging of the colors is not explicitly performed as that would be too expensive; instead we maintain a union-find
-data structure for that. So the red and blue mounds are not actually recolored purple, 
-and instead say red now points to blue in the union find data structure.
-
-Of course, things are more complicated when there are other mounds. There may be a yellow mound,
-corresponding to $\ell_3$ that joins with the blue mound higher up at some vertex $u$ (see the right part of \Fig{colors}). In $\cJ(\MM)$, $\ell_1$ and $\ell_3$
-are sibling leaves, and $\ell_2$ is a sibling of some ancestor of these leaves. So we cannot
-merge red and blue, until yellow and blue merge. Naturally, we use priority queues to handle this.
-We know $u$ must also be touched by blue, so all critical vertices touched by blue
-are put in a priority queue keyed by height, and vertices are handled in that order.
-
-\begin{figure}[h]\centering
-    \includegraphics[width=.25\linewidth]{figs/redBlue}%
-    \hspace{.07in}
-    \includegraphics[width=.25\linewidth]{figs/redBlue2}%
-    \hspace{.07in}
-    \includegraphics[width=.03\linewidth]{figs/redBlueTree}
-    \hspace{.09in}
-    \includegraphics[width=.3\linewidth]{figs/mound}%
-    \hspace{.07in}
-    \includegraphics[width=.065\linewidth]{figs/moundTree}
-    \caption{On the left, red and blue merge to make purple, followed by the contour tree with initial colors.  On the right, additional maxima and the resulting contour tree.}
-    \label{fig:colors}
-\end{figure}
-
-What happens when finally blue and red join at $v$? We merge the two colors, but now have
-blue and red queues of critical vertices, which also need to be merged to get a consistent painting. 
-This necessitates using a priority queue with efficient merges.
-Specifically, we use \emph{binomial heaps}~\cite{Vu78}, as they provide logarithmic time merges and deletes (though other similar heaps work). 
-We stress that the feasibility of the entire approach hinges on the use of such an efficient heap structure.
-
-In this discussion, we ignored an annoying problem. Vertices
-may actually be touched by numerous colors, not just one or two as assumed above. A simple solution
-would be to insert vertices into heaps corresponding to all colors touching it. But there could
-be super-constant numbers of copies of a vertex, and handling all these copies would lead
-to extra overhead. We show that it suffices to simply put each vertex $v$ into at most two heaps, one for each
-``side" of a possible join. We are guaranteed that when $v$ needs to be processed, all edges incident from above have
-at most $2$ colors, because of all the color merges that previously occurred.
-
-\myparagraph{Running time intuition:} 
- All non-heap operations can be easily bounded by $O(t\alpha(t) + N)$. 
- One can argue any heap always contains a subset of a single leaf to root path.  
- Bounding each heap by the size of this path suffices to prove the bound of \Thm{main-corr}, stated for join trees. 
- However, this may grossly overestimate heap sizes, as each vertex on a path can have at most two colors (the ones getting merged).
- Thus as we get closer to the root the competition for which two colors a vertex gets assigned to grows, since the number of descendant leaf colors grows. 
- This means for some vertices the size of the heap will be significantly smaller during an associated heap operation.
-
-The intuition is that the paint spilling from the maxima in the simplicial complex, corresponds to paint spilling from the leaves in 
-the join tree, which decomposes the join tree into a set of colored paths.  Unfortunately, the situation is more complex since 
-while a given color class is confined to a single leaf to root path, it may \emph{not} appear contiguously on this path, as the right part of \Fig{colors} shows. 
-Specifically, in this figure the far left saddle (labeled $i$) is hit by blue paint.  However, there is another 
-saddle on the far right (labeled $j$) which is not hit by blue paint.  Since this far right saddle is slightly 
-higher than the far left one, the merge event involving the component containing the blue mound (and also the yellow and red mounds) and corresponding to the far right saddle
-will occur before the one corresponding to the far left saddle.  Hence, the vertices initially touched by blue are not contiguous in the join tree.
-%
-This non-contiguous complication along with the fact that heap sizes keep changing as color classes merge, causes the analysis to be technically challenging. 
-We employ a variant of \emph{heavy path decompositions}, first used by Sleator and Tarjan for analyzing link/cut trees~\cite{st-dsdt-83}.
-The final analysis charges expensive heap operations to long paths in the decomposition, resulting in the bound stated in \Thm{main-alg}.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\newcommand{\newintuitionSection}{
-\section{A tour of the new contour tree algorithm} \label{sec:approach}
-We provide a short, high-level description of the main ideas of subsequent sections. 
-A longer more detailed version of this high-level description can be found in the full version \CiteFullVer.
-
-\myparagraph{Cutting $\MM$ into extremum dominant pieces:} We define a simplicial complex
-endowed with a height to be \emph{extremum dominant} if there exists only a single minimum, or only a single maximum.
-(When formally defined in \Sec{rain}, extremum dominant complexes will allow additional trivial minima or maxima which are a small complication resulting from the following cutting procedure.)
-We first cut $\MM$ into disjoint extremum dominant pieces in linear time.
-Take an arbitrary maximum $x$ and imagine torrential rain at the maximum. The water flows down, wetting any point that has a non-ascending path 
-from $x$. We end up with two portions, the wet part of $\MM$ and the dry part. This is similar
-to \emph{watershed} algorithms used for image segmentation~\cite{RoMe00}. The wet part is obviously connected and
-can be shown to be extremum dominant. We can cut along the interface of the wet and dry, remove
-the wet part, and recurse on the dry parts. This is carefully done to ensure that the total complexity
-of the pieces is linear.
-We prove a simple contour surgery theorem that builds the contour tree of $\MM$
-from the contour trees of the various pieces created.
-
-Using ideas from~\cite{csa-cctad-03}, we prove that the contour tree of an extremum dominant complex is (basically) just the \emph{join tree}, 
-which unlike the contour tree tracks superlevel sets rather than level sets.
-Consider sweeping down the hyperplane $x_{d+1} = h$ and taking the connected components of the portion of $\MM$ above height $h$. For a terrain, these superlevel
-sets are a collection of ``mounds". As we sweep downwards, these mounds keep joining each other,
-until finally, we end up with all of $\MM$. The join tree tracks exactly these events.
-
-\myparagraph{Join trees from painted mountaintops:} Our main result is a faster algorithm for join trees.
-The key idea is \emph{paint spilling}. Start with each maximum having a large can of paint, with distinct
-colors for each maximum. In arbitrary order, we spill paint from each maximum, wait till it flows down,
-then spill from the next, etc. Paint is viscous, and only flows down edges. Furthermore, our paints do not mix, so each edge receives a unique color, decided by the first
-paint to reach it.
-
-Our algorithm incrementally builds the join tree from the leaves (maxima) to the root. 
-Refer to the left part of \Fig{colors}.
-Consider two sibling leaves $\ell_1, \ell_2$ and their common parent $v$. The leaves are maxima,
-and $v$ is a join that ``merges" $\ell_1, \ell_2$. 
-In that case, there are ``mounds" corresponding to $\ell_1$ and $\ell_2$ that merge
-at a valley $v$. Suppose this was the entire input, and $\ell_1$ was colored blue and $\ell_2$ was colored red. 
-Both mounds are colored completely blue or red, while $v$ is touched by both colors.
-So this indicates that $v$ joins the blue maximum and red maximum in the join tree and is a critical point.
-To process $v$, we ``merge" the colors red and blue into a new color, purple. 
-In terms of the join tree, this
-is equivalent to removing leaves $\ell_1$ and $\ell_2$, and making $v$ a new leaf.
-
-Of course, things are more complicated when there are other mounds. There may be a yellow mound,
-corresponding to $\ell_3$ that joins with the blue mound higher up at some vertex $u$ (see the right part of \Fig{colors}). In the join tree, $\ell_1$ and $\ell_3$
-are sibling leaves, and $\ell_2$ is a sibling of some ancestor of these leaves. So we cannot
-merge red and blue, until yellow and blue merge. 
-A critical insight is the use of \emph{binomial heaps}~\cite{Vu78}, or any other efficiently mergeable heaps, to handle all the priorities and the merging.
-This ensures that all comparisons are only made between points that are ancestor-descendant in the join tree or adjacent in $\MM$.
-There are numerous details to handle to prove the final run time bound, but this should provide the reader
-some intuition behind our algorithm.
-
-\begin{figure}[h]\centering
-    \includegraphics[width=.25\linewidth]{figs/redBlue}%
-    \hspace{.07in}
-    \includegraphics[width=.25\linewidth]{figs/redBlue2}%
-    \hspace{.07in}
-    \includegraphics[width=.03\linewidth]{figs/redBlueTree}
-    \hspace{.09in}
-    \includegraphics[width=.3\linewidth]{figs/mound}%
-    \hspace{.07in}
-    \includegraphics[width=.065\linewidth]{figs/moundTree}
-    \caption{On the left, red and blue merge to make purple, followed by the contour tree with initial colors.  On the right, additional maxima and the resulting contour tree.}
-    \label{fig:colors}
-\end{figure}
-}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\newcommand{\intuitionSection}{
-\section{A tour of the new contour tree algorithm} \label{sec:approach}
-Our final algorithm is quite technical and has numerous moving parts. However, for the $d=2$ case, 
-where the input is just a triangulated terrain, the main ideas of the parts of the algorithm 
-can be explained clearly.  Therefore, here we first provide a high level view of the entire result.
-
-In the interest of presentation, the definitions and theorem statements in this section will
-slightly differ from those in the main body. They may also differ from the original definitions proposed
-in earlier work.
-
-\myparagraph{Do not globally sort:} The starting point for this work is \Fig{relative}. We have two terrains
-with exactly the same contour tree, but different orderings of (heights of) the critical points.
-Turning it around, we cannot deduce the full height ordering of critical points from the contour tree.
-Sorting all critical points is computationally unnecessary for constructing the contour tree.
-In \Fig{sorting}, the contour tree consists of two balanced binary trees, one of the joins,
-another of the splits. Again, it is not necessary to know the relative ordering between the mounds
-on the left (or among the depressions on the right) to compute the contour tree. Yet some ordering
-information is necessary: on the left, the little valleys are higher than the big central valley,
-and this is reflected in the contour tree. Leaf paths in the contour tree have points
-in sorted order, but incomparable points in the tree are unconstrained.
-How do we sort exactly what is required, without knowing the contour tree in advance?
-
-\subsection{Breaking $\MM$ into simpler pieces} \label{sec:break}
-Let us begin with the algorithm of Carr, Snoeyink, and Axen~\cite{csa-cctad-03}. The key insight is to build
-two different trees, called the join and split trees, and then merge them together into the contour tree.
-Consider sweeping down the hyperplane $x_{d+1} = h$ and taking the \emph{superlevel} sets. These 
-are the connected components of the portion of $\MM$ above height $h$. For a terrain, the superlevel
-sets are a collection of ``mounds". As we sweep downwards, these mounds keep joining each other,
-until finally, we end up with all of $\MM$. The join tree tracks exactly these events. 
-Formally, let $\MM^+_v$ denote the simplicial complex induced on the subset of 
-vertices which are higher than $v$.
-
-\begin{definition} 
-\label{def:int-criticalJoin}
-The \emph{join tree} $\cJ(\MM)$ is built on the set $\cV$ of all critical points.
-The directed edge $(u,v)$ is present when $u$ is the smallest valued vertex in $\cV$ in a connected component of $\MM^+_v$
-and $v$ is adjacent (in $\MM$) to a vertex in this component. 
-\end{definition}
-
-Refer to \Fig{sorting} for the join tree of a terrain. Note that nothing happens at splits, but these
-are still put as vertices in the join tree. They simply form a long path. The split tree
-is obtained by simply inverting this procedure, sweeping upwards and tracking sublevel sets.
-
-A major insight of~\cite{csa-cctad-03} is an ingeniously simple linear time procedure to construct
-the contour tree from the join and split trees. So the bottleneck is computing these trees. Observe
-in \Fig{sorting} that the split vertices form a long path in the join tree (and vice versa). Therefore, constructing
-these trees forces a global sort of the splits, an unnecessary computation for the contour tree. Unfortunately, in general (i.e.\ unlike \Fig{sorting}) 
-the heights of joins and splits may be interleaved in a complex manner, and hence the final merging of~\cite{csa-cctad-03}
-to get the contour tree requires having the split vertices in the join tree. Without
-this, it is not clear how to get a consistent view of both joins and splits, required for the contour tree.
-
-Our aim is to break $\MM$ into smaller pieces, where this unnecessary computation can be avoided.
-
-\myparagraph{Contour surgery:} We first need a divide-and-conquer lemma. Any contour $\phi$ can be associated
-with an edge $e$ of the contour tree. Suppose we ``cut" $\MM$ along this contour. We prove that $\MM$
-is split into two disconnected pieces, such the contour trees of these pieces is obtained
-by simply cutting $e$ in $\reeb(\MM)$. Alternatively, the contour trees of these pieces can
-be glued together to get $\reeb(\MM)$. This is not particularly surprising, and is fairly easy to prove
-with the right definitions. The idea of loop surgery has been used to reduce Reeb graphs to contour trees~\cite{TiGySi09,DoNa13}.
-Nonetheless, our theorem appears to be new and works for all dimensions.
-
-\begin{figure}[h]\centering
-    \includegraphics[width=.172\linewidth]{figs/tent2}%
-    \hspace{1.2in}
-    \includegraphics[width=.172\linewidth]{figs/tent1}
-    \caption{On left, downward rain spilling only (each shade of gray represents a piece created by each different spilling), producing a grid.  
-    Note we are assuming raining was done in sorted order of the maxima (i.e.\ lowest to highest).  
-    On right, flipping the direction of rain spilling.}
-    \label{fig:order}
-\end{figure}
-
-\myparagraph{Cutting $\MM$ into extremum dominant pieces:} We define a simplicial complex
-endowed with a height to be \emph{minimum dominant} if there exists only a single minimum.
-(Our real definition is more complicated, and involves simplicial complexes
-that allow additional ``trivial'' minima.)
-In such a complex, there exists a non-ascending path from any point to this unique minimum.
-Analogously, we can define \emph{maximum dominant} complexes,
-and both are collectively called \emph{extremum dominant}.
-
-We will cut $\MM$ into disjoint extremum dominant pieces, in linear time.
-One way to think of our procedure is a meteorological analogy. Take an arbitrary maximum $x$,
-and imagine torrential rain at the maximum. The water flows down, wetting any point that has a non-ascending path 
-from $x$. We end up with two portions, the wet part of $\MM$ and the dry part. This is similar
-to \emph{watershed} algorithms used for image segmentation~\cite{RoMe00}. The wet part is obviously connected,
-while there may be numerous disconnected dry parts. The interface between the dry and wet parts
-%is a set of contours\footnote{Technically, they are not contours, but rather the limits of sequences of contours.}, 
-is a set of limits of sequences of contours, which informally we call just a set of contours, which are
-given by the ``water line". The wet part is clearly maximum dominant, since all
-wet points have a non-descending path to $x$. So we can simply cut along the interface
-contours to get the wet maximum dominant piece $\MM'$. By our contour surgery theorem,
-we are left with a set of disconnected dry parts, and we can recur this procedure on them.
-
-But here's the catch. Every time we cut $\MM$ along a contour, we potentially increase
-the complexity of $\MM$. Water flows in the interior of faces, and the interface will naturally
-cut some faces. Each cut introduces new vertices, and a poor choice of repeated raining leads
-to a large increase in complexity. Consider the left of \Fig{order}. Each raining produces a single
-wet and dry piece, and each cut introduces many new vertices. If we wanted to partition this terrain into 
-maximum dominant simplicial complexes, the final complexity would be forbiddingly large.
-
-A simple trick saves the day. Unlike reality, we can choose rain to flow solely downwards or solely upwards.
-Apply the procedure above to get a single wet maximum dominant $\MM'$ and a set of dry pieces.
-Observe that a single dry piece $\NN$ is boundary critical with the newly introduced boundary $\phi$ (the wet-dry interface) behaving as a minimum.
-So we can rain upwards from this minimum, and get a \emph{minimum dominant} portion $\NN'$.
-This ensures that the new interface (after applying the procedure on $\NN$) does not
-cut any face previously cut by $\phi$.
-For each of the new dry pieces, the newly introduced boundary is now a maximum. So we rain downwards from there.
-More formally, we alternate between raining upwards and downwards as we go down the recursion tree.
-We can prove that an original face of $\MM$ is cut at most once, so the final complexity can be bounded.
-In \Fig{order}, regardless of the choice of the starting maximum,  this procedure would yield (at most) two pieces, 
-one maximum dominant, and one minimum dominant.
-
-Using the contour surgery theorem previously discussed, we can build the contour tree of $\MM$
-from the contour trees of the various pieces created. All in all, we prove the following theorem.
-
-\begin{theorem} \label{thm:int-rain} There is an $O(N)$ time procedure that cuts $\MM$ into extremum dominant simplicial complexes
-$\MM_1, \MM_2, \ldots$. Furthermore, given the set of contour trees $\{\cC(\MM_i)\}$, $\cC(\MM)$ can be constructed in $O(N)$ time.
-\end{theorem}
-
-\myparagraph{Extremum dominance simplifies contour trees:} We will focus on minimum dominant simplicial complexes $\MM$. By \Thm{int-rain}, it suffices
-to design an algorithm for contour trees on such inputs. 
-For the $d=2$ case, it helps to visualize such an input as a terrain with no minima,
-except at a unique boundary face (think of a large boundary triangle that is the boundary).
-All the saddles in such a terrain are necessarily joins, and there can be no splits.
-In \Fig{sorting}, the portion on the left is minimum dominant in exactly this fashion, albeit in one dimension lower.
-%Water poured at any critical point must flow to the boundary, and the entire wet portion is connected.
-More formally, $\MM^-_v$ is connected for all $v$, so there are no splits.
-
-We can prove that the split tree is just a path, and the contour tree is exactly
-the join tree. The formal argument is a little involved, and we employ the merging procedure
-of~\cite{csa-cctad-03} to get a proof. We demonstrate that the merging procedure will
-actually just output the join tree, so we do not need to actually compute the split tree.
-(The real definition of minimum dominant is a little more complicated,
-so the contour tree is more than just the join tree. But computationally, it suffices
-to construct the join tree.)
-
-We stress the importance of this step for our approach. Given the algorithm of~\cite{csa-cctad-03},
-one may think that it suffices to design faster algorithms for join trees. But this
-cannot give the sort of optimality we hope for. Again, consider \Fig{sorting}. Any algorithm
-to construct the true join tree must construct the path of splits, which implies
-sorting all of them. Thus it is necessary to cut $\MM$ into pieces where the cost
-of building the join tree can be related to that of building $\cC(\MM)$.
-
-\subsection{Join trees from painted mountaintops} \label{sec:join-paint}
-
-Arguably, everything up to this point is a preamble for the main result: a faster
-algorithm for join trees. Our algorithm does not require the initial input to be extremum
-dominant. This is only required to relate the join trees, of the resulting subcomplexes of Theorem~\ref{thm:int-rain}, to the contour tree of the initial input $\MM$.
-For clarity, we use $\NN$ to denote the input here. Note that in \Def{int-criticalJoin},
-the join tree is defined purely combinatorially in terms of the 1-skeleton (the underlying graph) of $\NN$.
-
-The join tree $\cJ(\NN)$ is a rooted tree with the dominant minimum at the root, and we direct edges downwards (towards the root). So it makes sense to talk
-of comparable vs incomparable vertices.
-We arrive at the main challenge: how to sort only the comparable critical points, without
-constructing the join tree? The join tree algorithm of~\cite{csa-cctad-03} is a typical event-based
-computational geometry algorithm. We have to step away from this viewpoint to avoid the global sort.
-
-The key idea is \emph{paint spilling}. Start with each maximum having a large can of paint, with distinct
-colors for each maximum. In arbitrary order, we spill paint from each maximum, wait till it flows down,
-then spill from the next, etc. Paint is viscous, and only flows down edges. \emph{It does not
-paint the interior of faces with dimension strictly larger than 1.} That is, this process is restricted to the 1-skeleton of $\NN$.
-Furthermore, our paints do not mix, so each edge receives a unique color, decided by the first
-paint to reach it.  In the following, $[n]$ denotes the set $\{1, \ldots, n\}$, for any natural number $n$.
-
-\begin{definition} \label{def:paint1} Let the 1-skeleton of $\NN$ have edge set $E$ and maxima $X$.
-A  \emph{painting} of $\NN$ is a map $\chi:X \cup E \to [|X|]$ with the following property. 
- Consider an edge $e$. There exists a descending path from some maximum $x$ to $e$
-	consisting of edges in $E$, such that all edges along this path have the same color as $x$.
-
-An \emph{initial} painting has the additional property that the restriction $\chi:X \to [|X|]$ is a bijection.
-\end{definition}
-
-Note that a painting colors edges, and not vertices (except for maxima). Our definition also does not require the timing aspect
-of iterating over colors, though that is one way of painting $\NN$. We begin with an initial painting, since all maximum
-colors are distinct. A few comments on paint vs water. The interface between two regions of different color
-is \emph{not} a contour, and so we cannot apply the divide-and-conquer
-approach of contour surgery. On the other hand, painting does not cut $\NN$, so there is no increase in complexity.
-Clearly, an initial painting can be constructed in $O(N)$ time. This is the tradeoff between water and paint.
-Water allows for an easy divide-and-conquer, at the cost of more complexity in the input. For an extremum
-dominant input, using water to divide the input $\NN$ raises the complexity too much.
-
-Our algorithm incrementally builds $\cJ(\MM)$ from the leaves (maxima) to the root (dominant minimum). 
-We say that vertex $v$ is \emph{touched} by color $c$, if there is a $c$-colored edge with lower endpoint $v$.
-Let us focus on an initial painting, where the colors have 1-1 correspondence with the maxima. Refer to the left part of \Fig{colors}.
-Consider two sibling leaves $\ell_1, \ell_2$ and their common parent $v$. The leaves are maxima,
-and $v$ is a join that ``merges" $\ell_1, \ell_2$. 
-In that case, there are ``mounds" corresponding to $\ell_1$ and $\ell_2$ that merge
-at a valley $v$. Suppose this was the entire input, and $\ell_1$ was colored blue and $\ell_2$ was colored red. 
-Both mounds are colored completely blue or red, while $v$ is touched by both colors.
-So this indicates that $v$ joins the blue maximum and red maximum in $\cJ(\MM)$.
-
-This is precisely how we hope to exploit the information in the painting. We prove later that
-when some join $v$ has exactly two colors among all incident edges with $v$ as the lower endpoint, the corresponding maxima (of those colors) are exactly
-the children of $v$ in $\cJ(\MM)$. To proceed further, we ``merge" the colors red and blue into a new color, purple. In other words,
-we replace all red and blue edges by purple edges. This indicates that the red and blue maxima have been
-handled. Imagine flattening the red and blue mounds until reaching $v$, so that the former join $v$
-is now a new maximum, from which purple paint is poured. In terms of $\cJ(\MM)$, this
-is equivalent to removing leaves $\ell_1$ and $\ell_2$, and making $v$ a new leaf.
-Alternatively, $\cJ(\MM)$ has been constructed up to $v$, and it remains to determine $v$'s parent.
-The merging of the colors is not explicitly performed as that would be too expensive; instead we maintain a union-find
-data structure for that. So the red and blue mounds are not actually recolored purple, 
-and instead say red now points to blue in the union find data structure.
-
-Of course, things are more complicated when there are other mounds. There may be a yellow mound,
-corresponding to $\ell_3$ that joins with the blue mound higher up at some vertex $u$ (see the right part of \Fig{colors}). In $\cJ(\MM)$, $\ell_1$ and $\ell_3$
-are sibling leaves, and $\ell_2$ is a sibling of some ancestor of these leaves. So we cannot
-merge red and blue, until yellow and blue merge. Naturally, we use priority queues to handle this issue.
-We know that $u$ must also be touched by blue. So all critical vertices touched by blue
-are put into a priority queue keyed by height, and vertices are handled in that order.
-
-\begin{figure}[h]\centering
-    \includegraphics[width=.25\linewidth]{figs/redBlue}%
-    \hspace{.07in}
-    \includegraphics[width=.25\linewidth]{figs/redBlue2}%
-    \hspace{.07in}
-    \includegraphics[width=.03\linewidth]{figs/redBlueTree}
-    \hspace{.09in}
-    \includegraphics[width=.3\linewidth]{figs/mound}%
-    \hspace{.07in}
-    \includegraphics[width=.065\linewidth]{figs/moundTree}
-    \caption{On the left, red and blue merge to make purple, followed by the contour tree with initial colors.  On the right, additional maxima and the resulting contour tree.}
-    \label{fig:colors}
-\end{figure}
-
-What happens when finally blue and red join at $v$? We merge the two colors, but now have
-blue and red queues of critical vertices, which also need to be merged to get a consistent painting. 
-This necessitates using a priority queue with efficient merges.
-Specifically, we use \emph{binomial heaps}~\cite{Vu78}, as they provide logarithmic time merges and deletes (though other similar heaps work). 
-We stress that the feasibility of the entire approach hinges on the use of such an efficient heap structure.
-
-In this discussion, we ignored an annoying problem. Vertices
-may actually be touched by numerous colors, not just one or two as assumed above. A simple solution
-would be to insert vertices into heaps corresponding to all colors touching it. But there could
-be super-constant numbers of copies of a vertex, and handling all these copies would lead
-to extra overhead. We show that it suffices to simply put each vertex $v$ into at most two heaps, one for each
-``side" of a possible join. We are guaranteed that when $v$ needs to be processed, all edges incident from above have
-at most $2$ colors, because of all the color merges that previously occurred.
-
-\myparagraph{The running time analysis:} 
- All the non-heap operations can be easily bounded by $O(t\alpha(t) + N)$ (the $t\alpha(t)$ is from the union-find data structure
- for colors). It is not hard to argue that at all times, any heap always contains a subset of a leaf to root path.
- This observation suffices to get a running time bound which is the analogue of \Thm{main-corr}, but for join trees. 
- Each heap deletion and merge can be charged to a vertex in the join tree (where each vertex gets charged only a constant number of times).
- Let $d_v$ denote the distance to the root for a vertex $v$ in the join tree, then since a heap's elements are on a single leaf to root 
- path, the size of $v$'s heap at the time an associated heap operation is made is at most $d_v$.
- Therefore, the total cost (of the heap operations) is at most $O(\sum_v \log d_v)$. This immediately proves a bound of $O(t\log D)$, where
- $D$ is the maximum distance to the root, an improvement over previous work.
-
- However, this bound is non-optimal.  For example, for a balanced binary tree, this gives a bound of $O(t\log\log t)$, however, by using 
- an analysis involving path decompositions we can get an $O(t)$ bound.  Imagine walking from some leaf 
- towards the root. Each vertex on this path can have at most two colors (the ones getting merged), however, as we get closer to the root 
- the competition for which two colors a vertex gets assigned to grows, as the number of descendant leaf colors grows. This means that for some vertices, 
- $v$, the size of the heap for an associated heap operation will be significantly smaller than $d_v$.
-
-The intuition is that the paint spilling from the maxmima in the simplicial complex, corresponds to paint spilling from the leaves in 
-the join tree, which decomposes the join tree into a set of colored paths.  Unfortunately, the situation is more complex since 
-while a given color class is confined to a single leaf to root path, it may \emph{not} appear contiguously on this path, as the right part of \Fig{colors} shows. 
-Specifically, in this figure the far left saddle (labeled $i$) is hit by blue paint.  However, there is another 
-saddle on the far right (labeled $j$) which is not hit by blue paint.  Since this far right saddle is slightly 
-higher than the far left one, the merge event involving the component containing the blue mound (and also the yellow and red mounds) and corresponding to the far right saddle
-will occur before the one corresponding to the far left saddle.  Hence, the vertices initially touched by blue are not contiguous in the join tree.
-
-This non-contiguous complication along with the fact that heap size keep changing as color classes merge, causes the analysis to be technically challenging. 
-%There are situations where the initial heap sizes are small, but they eventually merge to create larger heaps. 
-We employ a variant of \emph{heavy path decompositions}, first used by Sleator and Tarjan for analyzing link/cut trees~\cite{st-dsdt-83}.
-The final analysis charges expensive heap operations to long paths in the decomposition, resulting in the bound stated in \Thm{main-alg}.
-
-\subsection{The lower bound}
-
-Consider a contour tree $T$ and the path decomposition $P(T)$ used to bound the running time. 
-Denoting $\cost(P(T)) = \sum_{p \in P(T)} |p|\log |p|$, we construct a set of $\prod_{p \in P(T)} |p|!$
-functions on a fixed domain such that each function has a distinct (labeled) contour tree. By a simple
-entropy argument, any algebraic decision tree that correctly computes the contour tree on all instances
-requires worst case $\Omega(\cost(P(T)))$ time. We prove that our algorithm makes $\Theta(\cost(P(T)))$ 
-comparisons on all these instances.
-
-We have a fairly simple construction that works for terrains. In $P(T)$, consider the path
-$p$ that involves the root. The base of the construction is a conical ``tent", and there
-will be $|p|$ triangular faces that will each have a saddle. The heights of these saddles
-can be varied arbitrarily, and that will give $|p|!$ different choices. Each of these saddles
-will be connected to a recursive construction involving other paths in $P(T)$. Effectively,
-one can think of tiny tents that are sticking out of each face of the main tent. The contour
-trees of these tiny tents attach to a main branch of length $|p|$. Working out the details,
-we get $\prod_{p \in P(T)} |p|!$ terrains each with a distinct contour tree. 
-%Our algorithm requires $\Theta(\cost(P(T)))$ comparisons on all these instances.
-}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-%\InSoCGVer{
-%\newintuitionSection
-%}
-
-% \InNotSoCGVer{
-% \intuitionSection
-% }
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\section{Painting to compute join trees}
-\label{sec:paint}
-
-We now present the main algorithmic contribution, which is a new algorithm for computing join trees of any triangulated simplicial
-complex $\MM$.  In subsequent sections we then show how to reduce computing the contour tree to computing the join tree by using contour surgery.
-
-We now formally define the \emph{join} and \emph{split} trees as given by~\cite{csa-cctad-03}.
-Conventionally, all edges are directed from higher to lower function value. 
-In the following, for vertex $v$, we use $\MM^+_v$ to denote the simplicial complex obtained by only
-keeping vertices $u$ such that $f(u) > f(v)$. Analogously, define $\MM^-_v$. Note that $\MM^+_v$ may contain
-numerous connected components.
-
-\begin{definition} \label{def:join} The \emph{join tree} $\cJ(\MM)$ of $\MM$ is built on vertex set $\cV(\MM)$.
-The directed edge $(u,v)$ is present when $u$ is the smallest valued vertex in a connected component of $\MM^+_v$
-\emph{and} $v$ is adjacent to a vertex in this component (in $\MM$). The \emph{split tree} $\cS(\MM)$ is obtained
-by looking at $\MM^-_v$ (or alternatively, by taking the join tree of the inversion of $\MM$). 
-\end{definition}
-
-Some basic facts about these trees. 
-All outdegrees in $\cJ(\MM)$ are at most $1$, all indegree $2$ vertices are joins, all leaves are maxima,
-and the global minimum is the root. All indegrees in $\cS(\MM)$ are at most $1$, all outdegree $2$
-vertices are splits, all leaves are minima, and the global maximum is the root.
-As these trees are rooted, we can use ancestor-descendant terminology.  
-Specifically, for two adjacent vertices $u$ and $v$, $u$ is the parent of $v$ if $u$ is closer to the root 
-(i.e.\ each node can have at most one parent, but can have two children).
-
-\subsection{Painting}
-%{\bf Painting:} 
-The central tool is a notion 
-of \emph{painting} $\MM$. Initially associate a color with each maximum. Imagine there being a large
-can of paint of a distinct color at each maximum $x$. We will spill different paint from each maximum and watch it flow down.
-%This is analogous to the raining of \Sec{rain}, but paint is a much more viscous liquid.
-\emph{So paint only flows down edges, and it does not color the interior of higher dimensional faces.} Furthermore, paints
-do not mix, so every edge of $\MM$ gets a unique color. This process (and indeed the entire algorithm)
-works purely on the 1-skeleton of $\MM$, which is just a graph.
-
-\begin{definition} \label{def:paint2} Let the 1-skeleton of $\MM$ have edge set $E$ and maxima $X$.
-A  \emph{painting} of $\MM$ is a map $\chi:X \cup E \to [|X|]$, where $\chi(z)$ is referred to as the \emph{color} of $z$,
-with the following property. 
- Consider an edge $e$. There exists a descending path from some maximum $x$ to $e$
-	consisting of edges in $E$, such that all edges along this path have the same color as $x$.
-
-An \emph{initial} painting also requires that the restriction $\chi:X \to [|X|]$ is a bijection.
-\end{definition}
-
-\begin{definition} \label{def:color-set} Fix a painting $\chi$ and vertex $v$.
-\begin{asparaitem}
-	 \item For each maximal connected component $Y$
-     of $\MM^+_v$, the set of all edges connecting $v$ to $Y$
-     is called an \emph{up-star}. (Note $v$ has at most two up-stars, as Morse up-degree is at most 2 by assumption.)
-	 \item A vertex $v$ is \emph{touched by color $c$} if $v$ is incident to a $c$-colored
-	 edge with $v$ at the lower endpoint. For $v$, $\col(v)$ is the set of colors that touch $v$.
-	 \item A color $c \in \col(v)$ \emph{fully touches} $v$ if all edges in an up-star are colored $c$.
-	 \item For any maximum $x\in X$, we say that $x$ is both touched and fully touched by $\chi(x)$.
-\end{asparaitem}
-\end{definition}
-
-\subsection{The algorithm} \label{sec:algo}
-
-We first give an informal, idealized description of the algorithm. 
-That is, in the actual algorithm the details differ slightly
-and additional data structures are used to improve the running time.
-
-\medskip
-
-\begin{compactenum}
-%     \item Construct a painting of $\MM$ by a descending BFS 
-%     from maxima. This partitions the edges (into various colored
-%     sets), and each vertex is incident to at most $2$ colors.
-%     \item For each color, construct a (binomial) max-heap
-%     indexed by height of all critical points touched by that color.
-    \item Construct a painting of $\MM$ by a descending BFS from each maximum. 
-    This partitions the edges into colored sets, 
-    and assigns each vertex to the color classes of the edges that touch it.
-    \item For each color, construct a (binomial) max-heap of the critical points touched by that color, 
-          indexed by the heights of the critical points.
-    \item Initialize the join tree to consist of a set of isolated vertices, one for each color class.\\
-          (We now iteratively fill in the rest of the join tree from the leaves towards the root.)
-    \item Repeat until all heaps are empty:
-    \begin{compactenum}
-        \item Find a critical point $w$ that is at the top of all heaps containing $w$.
-        \item Add $w$ to the join tree by connecting it to the lowest vertex 
-              of each component of the join tree that contains a maximum whose color matches one of the heaps containing $w$.
-        \item Delete $w$ from all these heaps, and merge them.
-    \end{compactenum}
-\end{compactenum}
-
-\medskip
-
-\noindent
-This motivates the definition of \emph{ripe} vertices.
-
-\begin{definition} \label{def:ripe} A vertex $v$ is \emph{ripe} if: for all $c \in \col(v)$, 
-%$v$ is present in $T(\rep(c))$ and is also the highest vertex in this heap. 
-if $v$ is in the heap of $c$ then it is the highest vertex in that heap (using notation from \Sec{struct}, if $v\in T(\rep(c))$, then it is highest in $T(\rep(c))$).
-%%Also, for all $c \in \col(v) \setminus \fcol(v)$, $T(\rep(c))$ is empty.
-\end{definition}
-
-To efficiently implement the above, we need a number of data structures.  
-Namely, binomial heaps for efficient merges, union-find to avoid explicit recoloring, 
-and a stack to hold unripe (though ripening) vertices, which is used to effectively amortize the cost of 
-the otherwise expensive operation of finding ripe vertices. 
-Note we also later show it suffices to assign each vertex to only a single color from each adjacent up-star, 
-rather than the colors of all the edges which touch it.
-
-\subsubsection{The data structures} \label{sec:struct}
-\noindent
-{\bf The binomial heaps $\touch(c)$:} For each color $c$, $\touch(c)$ is a subset of vertices touched by $c$,
-This is stored as a \emph{binomial max-heap} keyed by vertex heights. Abusing notation, $\touch(c)$ refers
-both to the set and the data structure used to store it.
-
-\medskip
-\noindent
-{\bf The union-find data structure on colors:} We will repeatedly perform unions
-of classes of colors, and this will be maintained as a standard union-find data structure.
-For any color $c$, $\rep(c)$ denotes the representative of its class.
-
-\medskip
-\noindent
-{\bf Color list $\col(v)$:} For each vertex $v$, we maintain $\col(v)$ as a simple list.
-In addition, we will maintain another (sub)list of colors $L$ such that $\forall c \in L$,
-$v$ is \emph{not} the highest vertex in $T(\rep(c))$. Note that given $c \in \col(v)$
-and $\rep(c)$, this property can be checked in constant time. If $c$ is ever removed
-from this sublist, it can never enter it again.
-
-For notational convenience, we will not explicitly
-maintain this sublist. We simply assume that, in constant
-time, one can determine (if it exists) an arbitrary color $c \in \col(v)$ such that $v$ is not the highest vertex in $T(\rep(c))$.
-
-\medskip
-\noindent
-{\bf The stack $\stack$:} This consists of non-extremal critical points, with monotonically increasing
-heights as we go from the base to the head.
-\ignore{
-Each point $x \in \stack$ has an associated subset of $\col(x)$, denoted $\mcol(x)$.
-Both $\mcol(x)$ and its complement are stored as hash table. So lookups, inserts, and deletes
-are in these sets are all constant time operations. The stack is guaranteed to satisfy 
-the following invariants.
-\begin{asparaitem}
-	\item For every $x \in \stack$: For every $c \in \mcol(x)$, $x$ is the highest element
-	in $T(c)$. Furthermore, $c = \rep(c)$.
-	\item Consider $x, y \in \stack$ such that $y$ was pushed on $x$. There exists $c \in \col(x) \setminus
-	\mcol(x)$ such that $x$ is not highest in $T(c)$ but $y$ is highest in $T(c)$.
-\end{asparaitem}
-}
-
-\medskip
-\noindent
-{\bf Attachment vertex $\h(c)$:} For each color $c$, we maintain a critical point $\h(c)$ of this color.
-We will maintain the guarantee that the portion of the join tree above (and including) $\h(c)$ has already been constructed.
-%This is initialized to the maximum corresponding to $c$.
-
-\subsubsection{The formal algorithm} \label{sec:algorithm}
-
-We now give a detailed description of the main algorithm.
-The primary challenge is to use the data structures to rapidly find
-ripe vertices.
-A simple example of the algorithm execution is given in \Sec{example}.
-
-\smallskip
-\fbox{
-\begin{minipage}{0.9\textwidth}
-{\bf $\init(\MM)$}
-
-\smallskip
-\begin{compactenum}
-	\item Construct an initial painting of $\MM$ using a descending BFS from maxima that does not explore previously colored edges.
-	\item Determine all critical points in $\MM$. For each $v$, look at $(f(v) \pm \delta)$-contours in $f|_{B_\eps(v)}$ 
-	to determine the up and down degrees. 
-	\item Mark each critical $v$ as unprocessed.
-	\item For each critical $v$ and each up-star, pick an arbitrary color $c$ touching $v$. Insert $v$ into $T(c)$.
-	\item Initialize $\rep(c) = c$ and set $\h(c)$ to be the unique maximum colored $c$.
-	\item Initialize $K$ to be an empty stack.
-\end{compactenum}
-\end{minipage}}
-
-\smallskip
-\fbox{
-\begin{minipage}{0.9\textwidth}
-{\bf $\build(\MM)$}
-
-\smallskip
-\begin{compactenum}
-	\item Run $\init(\MM)$.
-	\item While there are unprocessed critical points:
-	\begin{compactenum}
-		\item Run $\update(\stack)$. Pop $\stack$ to get $h$.
-		\item Let $\cur(h) = \{\rep(c) | c \in \col(h)\}$.
-		\item For all $c' \in \cur(h)$:
-		\begin{compactenum}
-			\item Add edge $(\h(c'),h)$ to $\cJ(\MM)$.
-			\item Delete $h$ from $T(c')$.
-		\end{compactenum}
-	\item Merge heaps $\{T(c') | c' \in \cur(h)\}$.
-	\item Take union of $\cur(h)$ and denote resulting representative color by $\widehat{c}$.
-	\item Set $\h(\widehat{c}) = h$ and mark $h$ as processed.
-	\end{compactenum}
-\end{compactenum}
-\end{minipage}}
-
-\smallskip
-\fbox{
-\begin{minipage}{0.9\textwidth}
-{\bf $\update(\stack)$}
-
-\smallskip
-\begin{compactenum}
-	\item If $K$ is empty, push arbitrary unprocessed critical point $v$.
-	\item Let $h$ be the head of $\stack$.
-	\item While $h$ is not ripe:
-	\begin{compactenum}
-		\item Find $c \in \col(h)$ such that $h$ is not the highest in $T(\rep(c))$.
-		\item Push the highest of $T(\rep(c))$ onto $\stack$, and update head $h$.
-	\end{compactenum}
-\end{compactenum}
-\end{minipage}
-}
-
-\bigskip
-
-A few simple facts:
-\begin{compactitem}
-	\item At all times, the colors form a valid painting.
-	\item Each vertex is present in at most $2$ heaps. This is
-    because the Morse up-degree of every point is at most $2$.
-    After processing, it is removed from all heaps.
-	\item After $v$ is processed, all edges incident to $v$ from above have the same (representative) color. %(more technically, the same representative). 
-%	\item When $\update(\stack)$ terminates, the head $h$ is ripe. 
-	\item Vertices on the stack are in increasing height order. 
-\end{compactitem}
-
-\begin{observation}
-\label{obs:twoQueues}
- Each unprocessed vertex is always in exactly one queue of the colors in each of its up-stars.  Specifically, 
- for a given up-star of a vertex $v$, $\init(\MM)$ puts $v$ into the queue of exactly one of the colors of the up-star, say $c$.  As time goes on this queue 
- may merge with other queues, but while $v$ remains unprocessed, it is only ever (and always) in the queue of $\rep(c)$,  
- since $v$ is never added to a new queue and is not removed until it is processed.  
- In particular, finding the queues of a vertex in $\update(K)$ requires at most two union find operations (assuming each vertex records its two colors from $\init(\MM)$).
-\end{observation}
-
-\ignore{	
-	Suppose $\mcol(h) = \{c\}$ (so $h$ is split). 
-	\begin{compactenum}
-		\item Connect $h$ (in $\reeb(\MM)$) to $\h(c)$ and the unique minimum corresponding to split $h$. 
-		\item Delete $h$ from $T(c)$, set $\h(c) = h$.
-		\item End procedure.
-	\end{compactenum}
-	\item Let $\mcol(h) = \{c_1, c_2\}$ (so $h$ is merge).
-	\item Connect $h$ in $\reeb(\MM)$ to $\h(c_1)$ and $\h(c_2)$.
-	\item Delete all copies of $h$ from $T(c_1)$ and $T(c_2)$.
-	\item Perform union of colors $c_1$ and $c_2$ (denote merged color as $c$). Merge heaps
-	to get $T(c) = T(c_1) \cup T(c_2)$.
-	\item Set $\h(c) = h$.
-
-We state the primary invariant below.
-First, some notation. For any $v$, let $\psi^-_v$ denote the $(f(v)-\delta)$-contour intersecting
-$B_\eps(v)$. If there are two such contours (so $v$ is a split), choose the one that contains
-the dominant minimum. The \emph{palette} $\pal$ is the set of colors currently used,
-which is $\{T(\rep(c)) | c \in |X|\}$.
-
-Observe that $\build(\MM)$ loops over all unprocessed critical point.
-The invariant is true at the starting point of each such iteration.
-
-\medskip
-\textbf{Invariant:} 
-\begin{compactenum}
-	\item For every $c \in \pal$, the subtree of $\cJ(\MM)$ rooted at $\h(c)$ has been constructed.
-	\item Fix $c \in \pal$. Consider the set $S$ of maxima of $\MM$ in the subtree of $\cJ(\MM)$ rooted at $\h(c)$.
-	Then $\rep(c) = \bigcup_{s \in S} \chi(s)$.
-	\item Let $\Psi = \{\psi^-_{\h(c)} | c \in \pal\}$. The coloring given by $\rep(\cdot)$
-	is a valid painting of $\cut(\MM,\Psi)$, where $\psi^-_{\h(c)}$ has color $c$.
-\end{compactenum}
-
-\medskip
-It is easy to see that the invariant is true at the very beginning of the algorithm.
-Each $\h(c)$ is simply the maximum colored with $c$, and we have a valid painting of $\MM$.
-}
-
-\subsection{Proving correctness} \label{sec:correct}
-
-\ignore{
-\begin{claim} \label{clm:process} Assume the invariant. Any vertex with a non-increasing path to some $\h(c)$
-for $c \in \pal$ has been processed.
-\end{claim}
-
-\begin{proof} This is a direct consequence of \Thm{carr-mono}, which relates monotone paths in $\MM$ to $\cC(\MM)$.
-Since the subtree of $\cJ(\MM)$ (which is basically the subtree of $\cC(\MM)$) rooted at $\h(c)$ 
-has been found, all vertices in this subtree must be processed. These are all the vertices with non-increasing paths
-to $\h(c)$ in $\cC(\MM)$, which by \Thm{carr-mono} is the same as those in $\cC(\MM)$.
-\end{proof}
-}
-
-Our main workhorse is the following lemma. The \emph{current color} of an edge, $e$, refers to the value of $\rep(\chi(e))$, 
-where $\chi(e)$ is the color of $e$ from the initial painting.
-
-\begin{lemma} \label{lem:full} Suppose vertex $v$ is connected to a component $\PP$ of $\MM^+_v$
-by an edge $e$ which is currently colored $c$. At all times: either all edges in $\PP$ are currently colored $c$, or there
-exists a critical vertex $w \in \PP$ currently fully touched by $c$ and currently touched by another color.
-\end{lemma}
-
-\begin{proof} Since $e$ has current color $c$,
-there must exist vertices in $\PP$ currently touched by $c$. Consider the highest
-vertex $w$ in $\PP$ that is currently touched by $c$ and some other color. If no such vertex exists,
-this means all edges incident to a vertex currently touched by $c$ are currently colored $c$. By walking through
-$\PP$, we deduce that all edges are currently colored $c$.
-
-So assume $w$ exists. Take the $(f(w)+\delta)$-contour $\phi$ that intersects $B_\eps(w)$
-and intersects some currently $c$-colored edge incident to $w$. Note that all edges intersecting $\phi$ are also currently colored $c$,
-since $w$ is the highest vertex to be currently touched by $c$ and some other color. (Take the path of currently $c$-colored
-edges from the maximum to $w$. For any point on this path, the contour passing through this point must
-be currently colored $c$.) Hence, $c$ currently fully touches $w$. 
-But $w$ is currently touched by another color, and the corresponding edge cannot intersect $\phi$. So $w$
-must have up-degree $2$ and is critical.
-\end{proof}
-
-\begin{corollary}
-\label{cor:terminate}
- Each time $\update(K)$ is called, it terminates with a ripe vertex on top of the stack.
-\end{corollary}
-\begin{proof}
- $\update(K)$ is only called if there are unprocessed vertices remaining, and so by the time we reach step 3 in 
- $\update(K)$, the stack has some unprocessed vertex $h$ on it.  
- If $h$ is ripe, then we are done, so suppose otherwise.
-
- Let $\PP$ be one of the components of $\MM^+_h$.  By construction, $h$ was put in the heap of some initially  
- adjacent color $c$.  Therefore, $h$ must be in the current heap of $\rep(c)$ (see \Obs{twoQueues}).  
- Now by \Lem{full}, either all edges in $\PP$ are colored $\rep(c)$ or  
- there is some vertex $w$ fully touched by $\rep(c)$ and touched by some other color.  
- The former case implies that if there are any unprocessed vertices in $\PP$ then they are all in $T(\rep(c))$, 
- implying that $h$ is not the highest vertex and a new higher up unprocessed vertex will be put on the stack for the next iteration of the while loop.  
- Otherwise, all the vertices in $\PP$ have been processed.  
- However, it cannot be the case that all vertices in all components of $\MM^+_h$ have already been processed, 
- since this would imply that $h$ was ripe, and so one can apply the same argument to the other non-fully processed component.
-
- Now consider the latter case, where we have a non-monochromatic vertex $w$.
- In this case $w$ cannot have been processed (since after being processed it is touched only by one color), 
- and so it must be in $T(\rep(c))$ since it must be in some heap of a color in each up-star (and one up-star is entirely colored $\rep(c)$).
- As $w$ lies above $h$ in $\MM$, this implies $h$ is not on the top of this heap.
-\end{proof}
-
-% We prove a series of claims that will lead to the correctness proof.
-
-\begin{claim} \label{clm:upstar} Consider a ripe vertex $v$ and take the up-star connecting
-to some component of $\MM^+_v$. All edges in this component and the up-star have the same current color.
-\end{claim}
-
-\begin{proof} Let $c$ be the color of some edge in this up-star.
-By ripeness, $v$ is the highest in $T(\rep(c))$.
-Denote the component of $\MM^+_v$ by $\PP$.
-By \Lem{full}, either all edges in $\PP$ are colored $\rep(c)$ or there exists critical vertex $w \in \PP$
-fully touched by $\rep(c)$ and touched by another color. In the latter case, $w$ has not been processed,
-so $w \in T(\rep(c))$ (contradiction to ripeness). Therefore, all edges in $\PP$ are colored $\rep(c)$.
-\end{proof}
-
-\begin{claim} \label{clm:process} The partial output on the processed vertices is exactly
-the restriction of $\cJ(\MM)$ to these vertices.
-\end{claim}
-
-\begin{proof} More generally, we prove the following: all outputs on processed vertices
-are edges of $\cJ(\MM)$ and for any representative color $c$, $\h(c)$ is the lowest processed vertex
-of that representative color. We prove this by induction on the processing order. The base case is trivially
-true, as initially the processed vertices and attachments of the color classes are the set of maxima.
-For the induction step, consider the situation
-when $v$ is being processed.
-
-Since $v$ is being processed, we know by \Cor{terminate} that it is ripe. Take any up-star of $v$, and the corresponding component $\PP$
-of $\MM^+_v$ that it connects to. By \Clm{upstar}, all edges in $\PP$ and the up-star
-have the same current color, say $c$. If some critical vertex in $\PP$ is not processed,
-it must be in $T(c)$, which violates the ripeness of $v$.
-Thus, all critical vertices in $\PP$ have been processed, and so by the induction hypothesis, the restriction of $\cJ(\MM)$ to $\PP$ has been correctly computed.
-Additionally, since all critical vertices in $\PP$ have been processed, they all have the same representative color $c$ of the lowest critical vertex in $\PP$.
-Thus by the strengthened induction hypothesis, this lowest critical vertex is $\h(c)$.
-
-If there is another component of $\MM^+_v$, the same argument implies 
-the lowest critical vertex in this component is $\h(c')$ (where $c'$ is the current color of edges in the respective component).
-Now by the definition of $\cJ(\MM)$, the critical vertex $v$ connects to the lowest critical vertex in each component of $\MM^+_v$, 
-and so by the above $v$ should connect to $\h(c)$ and $\h(c')$, which is precisely what $v$ is connected to by $\build(\MM)$.
-Moreover, $\build$ merges the representative colors $c$ and $c'$ and correctly sets $v$ to be the attachment, 
-as $v$ is the lowest processed vertex of this merged color (as by induction $\h(c)$ and $\h(c')$ were the lowest vertices before merging colors).
-%
-%
-% Consider $w$, the lowest critical vertex in $\PP$. Note that $w$ is a parent of $v$. When $w$ was processed, it was set to $\h(c')$, for
-% some color $c'$. At this stage, all edges in $\PP$ above $w$ were also colored $c'$ (just apply the argument
-% above for $w$). Since there are no critical points below $w$, all edges in $\PP$ below $w$
-% are also colored $c'$ (the painting property). This means that the edges connecting $\PP$
-% are also colored $c'$. So $c' = \rep(c)$, and $\h(\rep(c)) = w$ is a parent of $v$. 
-%
-%This argument can be repeated for all such components $\PP$ to complete the proof.
-\end{proof}
-
-\begin{theorem}
-\label{thm:correct}
- Given an input complex $\MM$, $\build(\MM)$ terminates and outputs $\cJ(\MM)$.
-\end{theorem}
-\begin{proof}
- First observe that each vertex can be processed at most once by $\build(\MM)$.  By \Cor{terminate}, we know that as long as there 
- is an unprocessed vertex, $\update(K)$ will be called and will terminate with a ripe vertex which is ready to be processed. 
- Therefore, eventually all vertices will be processed, and so by \Clm{process} the algorithm will terminate having 
- computed $\cJ(\MM)$.
-\end{proof}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\subsection{Running time}
-\label{sec:runTime}
-
-We now bound the running time of the algorithm of \Sec{algo}.  
-In subsequent sections, through a sophisticated charging argument, this bound is then related to matching upper 
-and lower bounds in terms of path decompositions. 
-%In the subsequent sections we will then provide a matching lower bound for the running time.  
-Therefore, it will be useful to set up some terminology that can 
-be used consistently in both places.  Specifically, the path decomposition bounds will be purely combinatorial 
-statements on colored rooted trees, and so the terminology is of this form.
-
-Any tree $T$ considered in the following will be a rooted binary tree\footnote{Note 
-that technically the trees considered should have a leaf vertex hanging below the root  
-in order to represent the global minimum of the complex.  This vertex is 
-(safely) ignored to simplify presentation.} where the height of a vertex is its distance 
-from the root $r$ (i.e.\ conceptually $T$ will be a join tree with $r$ at the bottom).  
-As such, the children of a vertex $v\in T$ are the adjacent vertices of larger height
-(and $v$ is the parent of such vertices).  Then the subtree rooted at $v$, denoted $T_v$ consists of the graph induced on all 
-vertices which are descendants of $v$ (including $v$ itself).  For two vertices $v$ and $w$ in $T$ let $d(v,w)$ denote the 
-length of the path between $v$ and $w$.
-We use $A(v)$ to denote the set of ancestors of $v$.
-For a set of nodes $U$, $A(U) = \bigcup_{u \in U} A(u)$.
-
-\begin{definition}
-\label{def:leafAssign}
- A \emph{leaf assignment} $\chi$ of a tree $T$ assigns \emph{two} distinct leaves to each internal vertex $v$,
- one from the left child and one from the right child subtree of $v$ (naturally if $v$ has only one child it is assigned 
- only one leaf).
-\end{definition}
-
-For a vertex $v\in T$, we use $H_v$ to denote the \emph{heap} at $v$.
-Formally, $H_v = \{u | u \in A(v), \chi(u) \cap L(T_v) \neq \emptyset\}$, where $L(T_v)$ is the set of leaves of $T_v$.
-In words, $H_v$ is the set of ancestors of $v$ which are colored by some leaf in $T_v$.
-
-% We will use the $H_v$ values for a given $\chi$ to define a specific path decomposition of $T$ \Def{path}.
-% Specifically, this path decomposition is naturally defined by specifying a subset of edges in $T$ such that each internal
-% vertex has degree at most $2$.
-% For each internal vertex $v\in T$, add the edge from $v$ to $\arg \max_{v_l, v_r} \{|H_{v_l}|, |H_{v_r}|\}$ 
-% where $v_l$ and $v_r$ are the children of $v$ (if $|H_{v_l}|=|H_{v_r}|$ then pick one arbitrarily).
-% This is called the \emph{maximum} path decomposition, denoted by $\pmax(T)$ (which is the decomposition referenced in \Thm{main-alg}).
-
-\begin{definition}
-\label{def:initialColoring}
-The subroutine $\init(\MM)$ from \Sec{algo} naturally defines a leaf assignment to $\cJ(\MM)$ 
-according to the priority queue for each up-star we put a given vertex in.  Call this the \emph{initial coloring}
-of the vertices in $\cJ(\MM)$.  This initial coloring also defines the $H_v$ values for all $v\in \cJ(\MM)$.
-\end{definition}
-
-The following lemma should justify these definitions.
-
-\begin{lemma}
-\label{lem:runTimeUpper}
-Let $\MM$ be a simplicial complex with $t$ critical points.  For every vertex in $\cJ(\MM)$, 
-let $H_v$ be defined by the initial coloring of $\MM$.
-The total running time of $\build(\MM)$, including the subroutine call to $\init(\MM)$ and all calls to $\update$, is $O(N+t\alpha(t) + \sum_{v \in \cJ(\MM)} \log |H_v|)$.
-\end{lemma}
-\begin{proof}
-First we look at the initialization procedure $\init(\MM)$.  This procedure runs in $O(N)$ time.
-Indeed, the painting procedure consists of several BFS's but as each vertex is only explored by one of the BFS's, it is linear time overall.
-Determining the critical points is a local computation on the neighborhood of each vertex and so is linear (i.e.\ each edge is viewed at most twice).
-Finally, each vertex is inserted into at most two heaps and so initializing the heaps takes linear time in the number of vertices.
-
-Now consider the union-find operations performed by $\build$ and $\update$.  
-Initially the union find data structure has a singleton component for each leaf (and no new components are ever created), 
-and so each union-find operation takes $O(\alpha(t))$ amortized time.
-For $\update$, by \Obs{twoQueues}, each iteration of the while loop requires a constant number of finds (and no unions).
-Specifically, if a vertex is found to be ripe (and hence processed next) then these can be charged to that vertex.
-If a vertex is not ripe, then these can be charged to the vertex put on the stack.
-As each vertex is put on the stack or processed at most once, $\update$ performs $O(t)$ finds overall.
-Finally, $\build(\MM)$ performs one union and at most two finds for each vertex.  Therefore the total number of union find 
-operations is $O(t)$.
-
-For the remaining operations, observe that for every iteration of the loop in $\update$, a vertex is pushed onto the stack and each
-vertex can only be pushed onto the stack once (since the only way it leaves the stack is by being processed). 
-Therefore the total running time due to $\update$ is linear (ignoring the find operations).
-
-What remains is the time it takes to process a vertex $v$ in $\build(\MM)$.  
-In order to process a vertex there are a few constant time operations, union-find operations, and queue operations.
-Therefore the only thing left to bound are the queue operations.
-Let $v$ be a vertex in $\cJ(\MM)$, and let $c_1$ and $c_2$ be its children (the same argument holds if $v$ has only one child).
-At the time $v$ is processed, the colors and queues of all vertices in a given component of $\MM^+_v$ have merged together.
-In particular, when $v$ is processed we know it is ripe and so all vertices above $v$ in each component of $\MM^+_v$ have been processed, implying 
-these merged queues are the queues of the current colors of $c_1$ and $c_2$.  Again since $v$ is ripe, it must be on the top of these queues and so 
-the only vertices left in these queues are those in $H_{c_1}$ and $H_{c_2}$.
-
-Now when $v$ is handled, three queue operations are performed.  Specifically, $v$ is removed from the queues of $c_1$ and $c_2$, and then the queues are 
-merged together.  By the above arguments the sizes of the queues for each of these operations are $|H_{c_1}|$, $|H_{c_2}|$, and $|H_v|$, respectively.  
-As merging and deleting takes logarithmic time in the heap size for binomial heaps, the claim now follows.
-\end{proof}
-
-% Note that the analog of \Thm{main-corr} for join trees is an easy corollary of the above lemma.
-% Specifically, as in the \Thm{main-corr} statement, for any critical point $v$ let $\ell_v$ 
-% denote the length of the longest directed path passing through $v$ in the join tree. 
-% As $|H_v|$ only counts vertices in a $v$ to root path, trivially $|H_v|\leq \ell_v$, implying the analog of \Thm{main-corr}.
-
-For any critical point $v$ let $\ell_v$ denote the length of the longest directed path passing through $v$ in the join tree. 
-As $|H_v|$ only counts vertices in a $v$ to root path, trivially $|H_v|\leq \ell_v$, immediately implying the analog of \Thm{main-corr} for join trees.
-Note however that there is fair amount of slack in this argument as $|H_v|$ may be significantly smaller than $\ell_v$.  
-This slack allows for the more refined upper and lower bounds mentioned in \Sec{more-refined}.  Quantifying this slack however is quite challenging, 
-and requires a significantly more sophisticated analysis involving path decompositions, which is the subject of \Sec{pathDecomp}.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\section{Divide and conquer through contour surgery} \label{sec:surgery}
-Now that we have an efficient join tree algorithm, we now describe how to 
-reduce computing the contour tree to computing the join tree, as discussed in \Sec{approach}.
-Specifically, in this section we first describe our basic cutting tool, which we call contour surgery.
-In \Sec{rain} we then describe how we use this tool to cut the surface into so called extremum dominant pieces 
-where computing the contour tree reduces to computing the join tree, and this equivalence is proven in \Sec{extreme}. 
-The pieces we create are technically not simplicial complexes,
-in that each piece may have facets that are not simplices, that is facets of larger size than the dimension.
-These can easily be completed into simplicial complexes, though as we do not
-need this, for ease of exposition, we do not explicitly
-perform this step in our algorithm.
-
-{\bf The cutting operation:} We define a ``cut" operation on $f:\MM \rightarrow \RR$ that cuts along a regular contour to create
-a new simplicial complex with an added boundary. Given a contour $\phi$, roughly speaking, this constructs the simplicial complex $\MM \setminus \phi$. 
-We will always enforce the condition that $\phi$ never passes through a vertex of $\MM$.
-Again, we use $\eps$ for an infinitesimally small value. We denote $\phi^+$ (resp. $\phi^-$) to be
-the contour at value $f(\phi) + \eps$ (resp. $f(\phi) - \eps$), which is at distance $\eps$
-from $\phi$.
-
-An $h$-contour is achieved by intersecting $\MM$ with the hyperplane $x_{d+1} = h$ and taking a connected component. (Think of the $d+1$-dimension
-as height.) Given some point $x$ on an $h$-contour $\phi$, we can walk along $\MM$ from $x$ to determine $\phi$.
-We can ``cut" along $\phi$ to get a new (possibly) disconnected simplicial complex $\MM'$. This is achieved
-by splitting every face $F$ that $\phi$ intersects into an ``upper" face and ``lower" face. Algorithmically,
-we cut $F$ with $\phi^+$ and take everything above $\phi^+$ in $F$ to make the upper face. Analogously, we cut with $\phi^-$ to get the lower face.
-The faces are then triangulated to ensure that they are all simplices. 
-\Ben{Reviewer wanted clarification on how triangulation is done. Maybe can use bottom-vertex triangulation of Clarkson 88, discussed in Matousek book?}
-This creates the two new boundaries $\phi^+$ and $\phi^-$, and we maintain the property of constant $f$-value at a boundary.
-
-Note that by assumption $\phi$ cannot cut a boundary face, and moreover all non-boundary faces
-have constant size. Therefore, this process takes time linear in $|\phi|$, the number of faces $\phi$ intersects.
-This new simplicial complex is denoted by $\cut(\phi,\MM)$.
-%
-We now describe a high-level approach to construct $\reeb(\MM)$ using this cutting procedure.
-
-\Ben{Surgery should be rewritten to take instead take as input the output of cut, i.e. two complexes and a contour.  
-This is how it is eventually used in Claim\ref{clm:rain-reeb}. Currently it does nothing really.}
-
-\medskip
-\fbox{
-\begin{minipage}{0.9\textwidth}
-{\bf $\surgery(\MM,\phi)$}
-
-\smallskip
-\begin{asparaenum}
-	\item Let $\MM' = \cut(\MM,\phi)$.
-	\item Construct $\reeb(\MM')$ and let $A, B$ be the nodes corresponding to the new boundaries created
-	in $\MM'$. (One is a minimum and the other is maximum.)
-	\item Since $A, B$ are leaves, they each have unique neighbors $A'$ and $B'$, respectively. Insert
-	edge $(A',B')$ and delete $A, B$ to obtain $\reeb(\MM)$.
-\end{asparaenum}
-\end{minipage}}
-
-\medskip
-
-\begin{theorem} \label{thm:surgery} For any regular contour $\phi$, the output of $\surgery(\MM,\phi)$ is $\reeb(\MM)$.
-\end{theorem}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-To prove Theorem~\ref{thm:surgery}, we require a theorem from \cite{c-tmi-04} (Theorems 6.6) that map paths in $\cC(\MM)$ to $\MM$.
-
-\begin{theorem} \label{thm:carr-path} For every path $P$ in $\MM$, there exists a path $Q$ in the contour tree corresponding
-to the contours passing through points in $P$.
-For every path $Q$ in the contour tree, there exists at least one path $P$ in $\MM$ through points present
-in contours involving $Q$.
-
-In particular, for every monotone path $P$ in $\MM$, there exists a monotone path $Q$ in the contour tree to which $P$
-maps, and vice versa.
-\end{theorem}
-
-Theorem~\ref{thm:surgery} is a direct consequence of the following lemma.
-
-\begin{lemma} \label{lem:cut} Consider a regular contour $\phi$ contained in a contour class (of an edge of $\reeb(\MM))$
-$(u,v)$ and let $\MM' = \cut(\MM,\phi)$. Then $\cV(\reeb(\MM')) = \{\phi^+,\phi^-\} \cup \cV(\MM)$
-and $\cE(\reeb(\MM')) = \{(u,\phi^+),(\phi^-,v)\} \cup (\cE(\MM) \setminus (u,v))$.
-\end{lemma}
-
-\begin{proof} First observe that since $\phi$ is a regular contour, the vertex set in the complex $\MM'$ is the same 
-as the vertex set in $\MM$, except with the addition of the newly created vertices on $\phi^+$ and $\phi^-$.  
-Moreover, $\cut(\MM,\phi)$ does not affect the local neighborhood of any vertex in $\MM$. 
-Therefore since a vertex being critical is a local condition, with the exception of new boundary vertices, the 
-critical vertices in $\MM$ and $\MM'$ are the same.  Finally, the new vertices on $\phi^+$ and $\phi^-$ 
-collectively behave as a minimum and maximum, respectively, and so $\cV(\reeb(\MM')) = \{\phi^+,\phi^-\} \cup \cV(\MM)$.
-
-Now consider the edge sets of the contour trees.  Any contour class in $\MM'$ (i.e.\ edge in $\cC(\MM')$) that does not involve $\phi^+$ or $\phi^-$ is also
-a contour class in $\MM$. Furthermore, a maximal contour class satisfying these properties is also
-maximal in $\MM$. So all edges of $\cC(\MM')$ that do not involve $\phi^+$ or $\phi^-$ are edges of $\cC(\MM)$.
-Analogously, every edge of $\cC(\MM)$ not involving $\phi$ is an edge of $\cC(\MM')$.
-
-Consider the contour class corresponding to edge $(u,v)$ of $\cC(\MM)$. There is a natural ordering
-of the contours by function value, ranging from $f(u)$ to $f(v)$. All contours in this class ``above" $\phi$
-form a maximal contour class in $\MM'$, represented by edge $(u,\phi^+)$. Analogously, there is another
-contour class represented by edge $(\phi^-,v)$. We have now accounted for all contours in $\cC(\MM')$,
-completing the proof.
-\end{proof}
-
-A useful corollary of this lemma shows that a contour actually splits the simplicial complex into
-two disconnected complexes.
-
-\begin{theorem} \label{thm:jordan} $\cut(\MM,\phi)$ consists of two
-disconnected simplicial complexes, each of which
-is the triangulation of a $d$-dimensional ball.
-\end{theorem}
-
-\begin{proof} Denote (as in \Lem{cut}) the edge containing $\phi$ to be $(u,v)$. Suppose for contradiction that there is a path between vertices $u$ and $v$
-in $\MM' = \cut(\MM,\phi)$. By \Thm{carr-path}, there is a path in $\cC(\MM')$ between $u$ and $v$. Since $\phi^+$ and $\phi^-$
-are leaves in $\cC(\MM')$, this path cannot use their incident edges. Therefore by \Lem{cut},
-all the edges of this path are in $\cE(\cC(\MM)) \setminus (u,v)$. So we get a cycle in $\cC(\MM)$, a contradiction.
-To show that there are exactly two connected components in $\cut(\MM,\phi)$, it suffices
-to see that $\cC(\MM')$ has two connected components (by \Lem{cut}) and then applying \Thm{carr-path}.
-\end{proof}
-
-\section{Raining to partition $\MM$} \label{sec:rain}
-
-In this section, we describe a linear time procedure that partitions $\MM$ into special
-\emph{extremum dominant} simplicial complexes.
-
-\begin{definition} \label{def:dom} A simplicial complex with a height function is \emph{minimum dominant} if there exists
-a minimum $x$ such that every non-minimal \emph{vertex} in the complex has a non-ascending path to $x$.
-Analogously define \emph{maximum dominant}. 
-\end{definition}
-
-The first aspect of the partitioning is ``raining''. Start at some point $x \in \MM$ and imagine rain at $x$.
-The water will flow downwards along non-ascending paths and ``wet'' all the points encountered. Note that this procedure considers all points
-of the complex, not just vertices.
-
-\begin{definition} \label{def:wet} Fix $x \in \MM$. The set of points $y \in \MM$ such that there is a non-ascending path from $x$ to $y$
-is denoted by $\wet(x,\MM)$ (which in turn is represented as a simplicial complex). A point $z$ is at the \emph{interface} of $\wet(x,\MM)$ if every neighborhood of $z$
-has non-trivial intersection with $\wet(x,\MM)$ (i.e.\ the intersection is neither empty nor the entire neighborhood).
-\end{definition}
-
-The following claim gives a description of the interface.
-
-\begin{claim} \label{clm:inter} For any $x$, each non-empty component of the interface of $\wet(x,\MM)$ contains a join vertex.
-\end{claim}
-
-\begin{proof} If $p \in \wet(x,\MM)$, all the points in any contour containing $p$ are also in $\wet(x,\MM)$.
- (Follow the non-ascending path from $x$ to $p$ and then walk along the contour.) The converse is also true,
- so $\wet(x,\MM)$ contains entire contours.
-
-Let $\eps, \delta$ be sufficiently small as usual. Fix some $y$ at the interface.
-Note that $y \in \wet(x,\MM)$. (Otherwise, $B_\eps(y)\cap \MM$ is dry.)
-The points in $B_\eps(y)\cap \MM$ that lie below $y$ have a descending path from $y$ and hence must be wet.
-There must also be a dry point in $B_\eps(y)\cap \MM$ that is above $y$, and hence,
-there exists a dry, regular $(f(y)+\delta)$-contour $\phi$ intersecting $B_\eps(y)$.
-
-Let $\Gamma_y$ be the contour containing $y$.  Suppose for contradiction
-that $\forall p \in \Gamma_y$, $p$ has up-degree $1$ (see \Def{deg}). Consider the non-ascending path from $x$ to $y$ and let $z$
-be the first point of $\Gamma_y$ encountered. There exists a wet, regular $(f(y) + \delta)$-contour $\psi$ 
-intersecting $B_\eps(z)$. Now, walk from $z$ to $y$ along $\Gamma_y$. If all points $w$ in this walk
-have up-degree $1$, then $\psi$ is the unique $(f(y)+\delta)$-contour
-intersecting $B_\eps(w)$. This would imply that $\phi = \psi$, contradicting the fact that $\psi$ is wet
-and $\phi$ is dry.
-%
-% Therefore, there must exist a join vertex $w$ in $\Gamma_y$, such that $B_\eps(w)$ intersects
-% both $\phi$ and $\psi$. As $\delta, \eps \rightarrow 0^+$, the limit of $\phi$ (which is dry)
-% lies in $\Gamma_y$ (which is wet). 
-% Hence this limit gives a contour of the interface containing the join $w$.
-\end{proof}
-
-Note that $\wet(x,\MM)$ (and its interface) can be computed in time linear in the size of the wet simplicial complex.
-We perform a non-ascending search from $x$. Any face $F$ of $\MM$ encountered is partially (if not entirely) in $\wet(x,\MM)$.
-The wet portion is determined by cutting $F$ along the interface. Since each component of the interface is a contour, this is equivalent
-to locally cutting $F$ by a hyperplane. All these operations can be performed to output $\wet(x,\MM)$ in time linear in $|\wet(x,\MM)|$.
-
-We define a simple \lift{} operation on the interface components. Consider such a component $\phi$ containing
-a join vertex $y$. Take any dry increasing edge incident to $y$, and pick the point $z$ on this edge at height
-$f(y) + \delta$ (where $\delta$ is an infinitesimal, but larger than the value $\eps$ used in the definition of $\cut$). 
-Let $\lift(\phi)$ be the unique contour through the regular point $z$. Note that $\lift(\phi)$ is dry.
-The following claim follows directly from \Thm{jordan}.
-
-\begin{claim} \label{clm:cut-int} Let $\phi$ be a connected component of the interface. Then $\cut(\MM,\lift(\phi))$
-results in two disjoint simplicial complexes, one consisting entirely of dry points.
-\end{claim}
-
-\begin{proof} By \Thm{jordan}, $\cut(\MM,\lift(\phi))$ results in two disjoint simplicial complexes. Let $\NN$ be the complex containing
-the point $x$ (the argument in $\wet(x,\MM)$), and let $\NN'$ be the other complex. 
-Any path from $x$ to $\NN'$ must intersect $\lift(\phi)$, which is dry. Hence $\NN'$ is dry.
-%
-%
-%If there exists a wet point $z$ in $\NN'$, then there exists a descending path from $x$ to $z$. 
-%As $x$ is in $\NN$ and $z$ is in $\NN'$, again by the Jordan Curve Theorem this path must intersect $\phi$, 
-%contradicting the fact that $\phi$ is dry.
-\end{proof}
-
-\begin{figure}[h]\centering
-    \includegraphics[width=.172\linewidth]{figs/tent2}%
-    \hspace{1.2in}
-    \includegraphics[width=.172\linewidth]{figs/tent1}
-    \caption{On left, downward rain spilling only (each shade of gray represents a piece created by each different spilling), producing a grid.  
-    Note we are assuming raining was done in sorted order of the maxima (i.e.\ lowest to highest).  
-    On right, flipping the direction of rain spilling.}
-    \label{fig:order}
-\end{figure}
-
-We describe the main partitioning procedure that cuts a simplicial complex $\NN$ into extremum
-dominant complexes, shown below as $\rain(x,\NN)$. It takes an additional input of a maximum $x$. To initialize,
-we begin with $\NN$ set to $\MM$ and $x$ as an arbitrary maximum. 
-% One of the critical aspects
-% of the procedure is that the rain alternately flows downwards and upwards, 
-% since otherwise faces may be cut a super constant number of times (see \Fig{order}\InSoCGVer{ in \Sec{approach}}).
-When we start, rain flows downwards. In each recursive call, the direction of rain is \emph{switched} to the 
-opposite direction. This is crucial to ensure linear running time, and \Fig{order} shows how failing to switch 
-directions can lead to a blow up in complexity. 
-The switching is easily implemented by inverting a complex $\NN'$, achieved by negating the height values.
-% While one can implement $\rain$ this way, it is conceptually easier
-% to think of \emph{inverting} a complex $\NN'$ when a recursive call is made. Inversion
-% is easily achieved by just negating the height values. (From a running time standpoint, it suffices
-% to maintain a single bit associated with $\NN'$ that determines whether heights are inverted or not.)
-We can now let rain flow downwards, as it usually does in our world.
-
-\medskip
-\fbox{
-\begin{minipage}{0.9\textwidth}
-{\bf $\rain(x,\NN)$}
-
-\smallskip
-\begin{compactenum}
-	\item Determine interface of $\wet(x,\NN)$. 
-	\item If the interface is empty, simply output $\NN$. Otherwise, denote the connected components by $\phi_1, \phi_2, \ldots, \phi_k$ and set $\phi'_i = \lift(\phi_i)$.
-	\item Initialize $\NN_1 = \NN$.
-	\item For $i$ from $1$ to $k$: 
-	\begin{compactenum}
-		\item Construct $\cut(\NN_i,\phi'_i)$, consisting of dry complex $\LL_i$ and remainder $\NN_{i+1}$.
-		\item Let the newly created boundary of $\LL_i$ be $B_i$. Invert $\LL_i$ so that $B_i$ is a maximum. Recursively
-		call $\rain(B_i,\LL_i)$.
-	\end{compactenum}
-	\item Output $\NN_{k+1}$ together with any complexes output by recursive calls.
-\end{compactenum}
-\end{minipage}}
-
-\medskip
-For convenience, denote the total output of $\rain(x,\MM)$ by $\MM_1, \MM_2, \ldots, \MM_r$.
-% We first argue the correctness of $\rain$.
-
-\begin{lemma} \label{lem:rain-1} Each output $\MM_i$ is extremum dominant.
-\end{lemma}
-
-\begin{proof} Consider a call to $\rain(x,\NN)$. If the interface is empty, then all of $\NN$
-is in $\wet(x,\NN)$, so $\NN$ is trivially extremum dominant. So suppose the interface
-is non-empty and consists of $\phi_1, \ldots, \phi_k$ (as denoted in the procedure).
-By repeated applications of \Clm{cut-int}, $\NN_{k+1}$ contains $\wet(x,\MM)$. 
-Consider $\wet(x,\NN_{k+1})$. The interface is exactly $\phi_1, \ldots, \phi_k$.
-So the only dry vertices of $\NN_{k+1}$ are those on the boundaries created by calling $\cut$ for each $\phi_i$, 
-but such boundaries are all maxima.
-%$B_1, \ldots, B_k$. But these boundaries are maxima.
-\end{proof}
-
-As $\rain(x,\MM)$ proceeds, new faces/simplices are created because of repeated cutting. 
-%Indeed, any facet $F$ in any simplicial complex $\NN$ that is recursively invoked, is achieved by cutting facet $F'$ in $\MM$.
-The key to the running time of $\rain(x,\MM)$ is bounding the number of newly created faces, for which we have the following lemma.
-
-\begin{lemma}\label{lem:new-verts}
-A face $F \in \MM$ is cut\footnote{What we are calling a single cut is actually done with two hyperplanes.} at most once during $\rain(x,\MM)$.
-\end{lemma}
-\begin{proof} Notation here follows the pseudocode of $\rain$.
-First, by \Thm{jordan}, all the pieces on which $\rain$ is invoked are disjoint.
-Second, all recursive calls are made on dry complexes.
-
-Consider the first time that $F$ is cut, say, during the call to $\rain(x,\NN)$.
-Specifically, say this happens when $\cut(\NN_i,\phi'_i)$ is constructed. 
-$\cut(\NN_i,\phi'_i)$ will cut $F$ with two horizontal cutting planes, one $\eps$ above $\phi'_i$
-and one $\eps$ below $\phi'_i$.  This breaks $F$ into lower and upper portions which are then triangulated
-(there is also a discarded middle portion).  The lower portion, which is adjacent to $\phi_i$, gets included 
-in $\NN_{k+1}$, the complex containing the wet points, and hence does not participate in any later recursive calls.  
-The upper portion (call it $U$) is in $\LL_i$. Note that the lower
-boundary of $U$ is in the boundary $B_i$. Since a recursive call is made to $\rain(B_i,\LL_i)$
-(and $\LL_i$ is inverted), $U$ becomes wet. Hence $U$, and correspondingly $F$, will not be subsequently cut.
-\end{proof}
-
-The following are direct consequences of \Lem{new-verts} and the $\surgery$ procedure.
-
-\begin{theorem} \label{thm:rain-time} The total running time of $\rain(x,\MM)$ is $O(|\MM|)$.
-\end{theorem}
-\begin{proof} The only non-trivial operations performed are $\wet$ and $\cut$.  
-Since $\cut$ is a linear time procedure, Lemma~\ref{lem:new-verts} implies the total time for all calls to $\cut$ is $O(|\MM|)$.  
-As for the $\wet$ procedure, observe that Lemma~\ref{lem:new-verts} additionally implies there are only $O(|\MM|)$ new faces created by $\rain$.
-Therefore, since $\wet$ is also a linear time procedure, and no face is ever wet twice, the total time for all calls to $\wet$ is $O(|\MM|)$.
-\end{proof}
-
-\begin{claim} \label{clm:rain-reeb} Given $\reeb(\MM_1), \reeb(\MM_2), \ldots, \reeb(\MM_r)$, 
-%be the output of $\rain(x,\MM)$. Suppose that for each reeb graph we know its corresponding vertex in the Manifold Ordering Tree. 
-$\reeb(\MM)$ can be constructed in $O(|\MM|)$ time.
-\end{claim}
-\begin{proof} Consider the tree of recursive calls in $\rain(x,\MM)$, with each node
-labeled with some $\MM_i$.
-Walk through this tree in a leaf first ordering.  Each time we visit a node we connect its contour tree to the 
-contour tree of its children in the tree using the $\surgery$ procedure. 
-Each $\surgery$ call takes constant time, and the total time is the size of the recursion tree.
-\end{proof}
-
-\section{Contour trees of extremum dominant complexes} \label{sec:extreme}
-
-The previous section allows us to restrict attention to extremum dominant complexes.
-We will orient so that the extremum in question is always a \emph{minimum}.
-We will fix such a simplicial complex $\MM$, with the dominant minimum $m^*$.
-
-% For vertex $v$, we use $\MM^+_v$ to denote the simplicial complex obtained by only
-% keeping vertices $u$ such that $f(u) > f(v)$. Analogously, define $\MM^-_v$. Note that $\MM^+_v$ may contain
-% numerous connected components. 
-%
-%The main theorem of this section asserts that contour trees of minimum dominant complexes have a simple description.
-%The exact statement will require some definitions and notation.
-% We require the notions of \emph{join} and \emph{split} trees, as given by~\cite{csa-cctad-03}.
-% Conventionally, all edges are directed from higher to lower function value. 
-% 
-% \begin{definition} \label{def:join} The \emph{join tree} $\cJ(\MM)$ of $\MM$ is built on vertex set $\cV(\MM)$.
-% The directed edge $(u,v)$ is present when $u$ is the smallest valued vertex in a connected component of $\MM^+_v$
-% \emph{and} $v$ is adjacent to a vertex in this component (in $\MM$). The \emph{split tree} $\cS(\MM)$ is obtained
-% by looking at $\MM^-_v$ (or alternatively, by taking the join tree of the inversion of $\MM$). 
-% \end{definition}
-% 
-% Some basic facts about these trees. 
-% All outdegrees in $\cJ(\MM)$ are at most $1$, all indegree $2$ vertices are joins, all leaves are maxima,
-% and the global minimum is the root. All indegrees in $\cS(\MM)$ are at most $1$, all outdegree $2$
-% vertices are splits, all leaves are minima, and the global maximum is the root.
-% As these trees are rooted, we can use ancestor-descendant terminology.  
-% Specifically, for two adjacent vertices $u$ and $v$, $u$ is the parent of $v$ if $u$ is closer to the root 
-% (i.e.\ each node can have at most one parent, but can have two children).
-
-Recall the definitions of join and split trees from \Sec{paint}.
-The main theorem of this section asserts that contour trees of minimum dominant complexes have a simple description in terms of $\cJ(\MM)$.  
-First we make the key observation that $\cS(\MM)$ is trivial for a minimum dominant $\MM$.
-
-\begin{lemma} \label{lem:split} For minimum dominant $\MM$, $\cS(\MM)$ consists of:
-\begin{asparaitem}
-	\item A single path (in sorted order) with all vertices except non-dominant minima.
-	\item Each non-dominant minimum is attached to a unique split (which is adjacent to it).
-\end{asparaitem}
-\end{lemma}
-
-\begin{proof} It suffices to prove that each split $v$ has one child that is just a leaf, which
-is a non-dominant minimum.  Specifically, any minimum is a leaf in $\cS(\MM)$ and thereby attached to a split, 
-which implies that if we removed all non-dominant minima, we must end up with a path, as asserted above.
-
-Consider a split $v$. For sufficiently small $\eps, \delta$, there are exactly two $(f(v) - \delta)$-contours
-$\phi$ and $\psi$ intersecting $B_\eps(v)$. Both of these are regular contours. There must be a non-ascending
-path from $v$ to the dominant minimum $m^*$. Consider the first edge (necessarily decreasing from $v$)
-on this path. It must intersect one of the $(f(v) - \delta)$-contours, say $\phi$. By \Thm{jordan}, $\cut(\MM,\phi)$ has
-two connected components, with one (call it $\LL$) having $\phi^-$ as a boundary maximum. This complex
-contains $m^*$ as the non-ascending path intersects $\phi$ only once. Let the other
-component be called $\MM'$.
-
-Consider $\cut(\MM',\psi)$ with connected component $\NN$ having $\psi^-$ as a boundary. $\NN$
-does not contain $m^*$, so any path from the interior of $\NN$ to $m^*$ must intersect the boundary $\psi^-$.
-But the latter is a maximum in $\NN$, so there can be no non-ascending path from the interior to $m^*$.
-Since $\MM$ is overall minimum dominant, the interior of $\NN$ can only contain a single vertex $w$, a non-dominant
-minimum.
-
-The split $v$ has two children in $\cS(\MM)$, one in $\NN$ and one in $\LL$. The child in $\NN$ can only
-be the non-dominant minimum $w$, which is a leaf. 
-\end{proof}
-
-It is convenient to denote the non-dominant minima as $m_1, m_2, \ldots, m_k$
-and the corresponding splits (as given by the lemma above) as $s_1, s_2, \ldots, s_k$.
-
-Using the above lemma we can now prove that computing the contour tree for a minimum dominant complex 
-amounts to computing its join tree.  Specifically, to prove our main theorem, we rely on the correctness of the
-merging procedure from~\cite{csa-cctad-03} that constructs the contour tree from the join and split trees. 
-It actually first constructs the \emph{augmented contour tree} $\cA(\MM)$, which is 
-obtained by replacing each edge in the contour tree with a path of all regular vertices 
-(sorted by height) whose corresponding contour belongs to the equivalence class of that edge.
-
-Consider a tree $T$ with a vertex $v$ of in and out degree at most $1$.
-\emph{Erasing} $v$ from $T$ is the following operation: if $v$ is a leaf, just delete $v$. Otherwise, 
-delete $v$ and connect its neighbors by an edge (i.e.\ smooth $v$ out). This tree is denoted by $T \ominus v$.
-%
-The merging procedure of \cite{csa-cctad-03} is shown below as $\merge(\cJ(\MM),\cS(\MM))$.  
-This procedure builds $\cA(\MM)$ by iteratively adding edges that are adjacent to leaves of either $\cJ(\MM)$ or $\cS(\MM)$, 
-erasing the corresponding vertex from both $\cJ(\MM)$ and $\cS(\MM)$, and repeating.  $\cC(\MM)$ is then produced by erasing all regular vertices in $\cA(\MM)$.
-We use the fact that this procedure correctly produces the contour tree (regardless of the order leaves are processed), 
-and refer the interested reader to \cite{csa-cctad-03} for the proof of this fact, as it is not needed for our purposes.
-
-\medskip
-\fbox{
-\begin{minipage}{0.9\textwidth}
-{\bf $\merge(\cJ(\MM),\cS(\MM))$}
-
-\smallskip
-\begin{compactenum}
-	\item Set $\cJ = \cJ(\MM)$ and $\cS = \cS(\MM)$.
-	\item Denote $v$ as a \emph{candidate} if the sum of its indegree in $\cJ$ and outdegree in $\cS$ is $1$. 
-	\item Add all candidates to queue.
-	\item While candidate queue is non-empty: 
-	\begin{compactenum}
-		\item Let $v$ be head of queue. If $v$ is leaf in $\cJ$, consider its edge in $\cJ$. Otherwise
-		consider its edge in $\cS$. In either case, denote the edge by $(v,w)$.
-		\item Insert $(v,w)$ in $\cA(\MM)$. 
-		\item Set $\cJ = \cJ \ominus v$ and $\cS = \cS \ominus v$. Enqueue any new candidates.
-	\end{compactenum}
-	\item Erase all regular vertices in $\cA(\MM)$ to get $\cC(\MM)$.
-\end{compactenum}
-\end{minipage}}
-
-\medskip
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\begin{definition} 
-\label{def:criticalJoin}
-The \emph{critical join tree} $\jc(\MM)$ is built on the set $V'$ of all
-critical points other than the non-dominant minima. 
-The directed edge $(u,v)$ is present when $u$ is the smallest valued vertex in $V'$ in a connected component of $\MM^+_v$
-and $v$ is adjacent (in $\MM$) to a vertex in this component. 
-\end{definition}
-
-\begin{theorem} \label{thm:contour-tree} Let $\MM$ have a dominant minimum. 
-The contour tree $\cC(\MM)$ consists of all edges $\{(s_i, m_i)\}$ and $\jc(\MM)$.
-\end{theorem}
-\begin{proof} %We first show that $\cA(\MM)$ is $\cJ(\MM) \ominus \{m_i\}$ with edges $\{(s_i,m_i)\}$.
-We have flexibility in choosing the order of processing in $\merge$. We first put the non-dominant
-minima $m_1, \ldots, m_k$ into the queue. As these are processed, the edges $\{(s_i,m_i)\}$ are inserted 
-into $\cA(\MM)$. Once all the $m_i$'s are erased, $\cS$ becomes a path, so all outdegrees are at most $1$.
-Moreover the join tree is now $\cJ(\MM) \ominus \{m_i\}$, and so we can now process $\cJ$ leaf by leaf, and all
-remaining edges of $\cJ$ are inserted into $\cA(\MM)$.
-%
-Note that smoothing out regular points from $\cJ(\MM) \ominus \{m_i\}$ yields the edges of $\jc(\MM)$.
-Thus after smoothing out all regular points from $\cA(\MM)$ we recover the edges of $\jc(\MM)$.  
-Hence the resulting contour tree contains these edges and the $\{(s_i,m_i)\}$ as claimed.
-%Note that $\cC(\MM)$ is obtained by smoothing out all regular points from $\cA(\MM)$. Similarly, smoothing out regular
-%points from $\cJ(\MM) \ominus \{m_i\}$ yields the edges of $\jc(\MM)$.
-\end{proof}
-
-\begin{Remark}\label{rem:joinAndCriticalJoin}
-% The above theorem, combined with the previous sections, implies that in order to get an efficient contour tree algorithm, 
-% it suffices to have an efficient algorithm for computing $\jc(\MM)$.
-% In the other sections of the paper it is better to phrase things instead in terms of computing $\cJ(\MM)$ efficiently.  
-% %Note however that for minimum dominant complexes output by $\rain$, converting between $\jc$ and $\cJ$ is trivial, as $\cJ$ is just $\jc$ 
-% %with each non-dominant minimum $m_i$ augmented along the edge leaving $s_i$.
-% Note however that for minimum dominant complexes, converting between $\jc$ and $\cJ$ is trivial, as $\cJ$ is just $\jc$ 
-% with each non-dominant minimum $m_i$ augmented along the edge leaving $s_i$.  
-% (Indeed $\jc(\MM)$ was only defined to make the above theorem statement clearer.)
-The above theorem, combined with the previous sections, implies that in order to get an efficient contour tree algorithm, 
-it suffices to have an efficient algorithm for computing $\jc(\MM)$.
-Note however that for minimum dominant complexes, converting between $\jc$ and $\cJ$ is trivial, 
-as $\cJ$ is just $\jc$ with each non-dominant minimum $m_i$ augmented along the edge leaving $s_i$. 
-(Indeed $\jc(\MM)$ was only defined to make the above theorem statement clearer.)  
-Thus to get an efficient contour tree algorithm it suffices to have an efficient algorithm for computing $\cJ(\MM)$.
-\end{Remark}
-
-\subsection{Putting it all together}
-We now have all the pieces required to prove \Thm{main-corr}.
-\begin{proof}[Proof of \Thm{main-corr}]
-Consider a critical point $v$ of the initial input 
-complex.  By \Thm{jordan} this vertex appears in exactly one of the pieces output by $\rain$.  As in the \Thm{main-corr} statement, let $\ell_v$ 
-denote the length of the longest directed path passing through $v$ in the contour tree of the input complex, and let $\ell_v'$ 
-denote the longest directed path passing through $v$ in the join tree of the piece containing $v$.
-By \Thm{surgery}, ignoring non-dominant extrema introduced from cutting (whose cost can be charged to a corresponding saddle), the join tree on each piece output by $\rain$ is isomorphic to some 
-connected subgraph of the contour tree of the input complex, and hence $\ell_v'\leq \ell_v$.  
-Moreover, $|H_v|$ only counts vertices in a $v$ to root path and so trivially $|H_v|\leq \ell_v'$, implying \Thm{main-corr}.
-\end{proof}
-
-Ultimately our goal is to prove the more refined upper and lower bounds mentioned in \Sec{more-refined}.  
-To achieve this, in the next section we give a careful analysis of how heap sizes evolve over the course of our join tree algorithm, 
-rather than using the blunt bound $|H_v|\leq \ell_v'$, as done in the above proof.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\section{Leaf assignments and path decompositions}
-\label{sec:pathDecomp}
-
-In this section, we set up a framework to analyze the time taken to compute a join tree $\cJ(\MM)$ (see \Def{join}), and consequently for contour trees by the above discussion.
-We adopt all notation already defined in \Sec{runTime}.  From here forward we will often assume binary trees are full binary trees 
-(this assumption simplifies the presentation but is not necessary).
-
-Let $\chi$ be some fixed leaf assignment to a rooted binary tree $T$, which in turn fixes all the heaps $H_v$. 
-We partition the edges of $T$ into a \emph{path decomposition}, where each set of the partition (a path) 
-is a subset of edges in $T$ with each internal
-vertex of degree at most $2$. This naturally gives a path decomposition.
-For each internal vertex $v\in T$, add the edge from $v$ to $\arg \max_{v_l, v_r} \{|H_{v_l}|, |H_{v_r}|\}$ 
- where $v_l$ and $v_r$ are the children of $v$ (if $|H_{v_l}|=|H_{v_r}|$ then pick one arbitrarily).
-This is called the \emph{maximum} path decomposition, denoted by $\pmax(T)$.
-
-Our main goal in this section is to prove the following theorem.  We use $|p|$ to denote the number of vertices in a path $p$.
-
-\begin{theorem} \label{thm:runtime} $\sum_{v \in T} \log |H_v| = O(\sum_{p\in \pmax(T)} |p| \log |p|)$.
-\end{theorem}
-
-We conclude this section in \Sec{implications} by showing that proving this theorem implies our main result \Thm{main-alg}.
-
-\subsection{Tall and short paths}
-\label{sec:shrubs}
-
-The paths in $\pmax(T)$ naturally define a tree\footnote{
-%Please excuse the overloading of the term 'tree', it is the most natural term to use here.
-The term 'tree' is the most natural term to use here, but the reader should be careful not to confuse this tree on paths from $T$ with the tree $T$ itself.
-} of their own.  Specifically, in the original 
-tree $T$ contract each path down to its root.  Call the resulting tree the \emph{shrub} of $T$ corresponding to 
-the path decomposition $\pmax(T)$. Abusing notation, we simply use $\pmax(T)$ to denote the shrub.
-As a result, we use terms like `parent', `child', `sibling', etc.\ for paths as well.
-The shrub gives a handle on the heaps of a path.
-We use $b(p)$ to denote the \emph{base} of the path, which is the vertex in $p$ closest
-to root of $T$. We use $\ell(p)$ to denote the leaf in $p$.
-We use $H_p$ to denote $H_{b(p)}$.
-
-We wish to prove $\sum_{v\in T} \log |H_v| = O(\sum_{p\in \pmax} |p| \log |p|)$. 
-The simplest approach is to prove $\forall p\in \pmax$, $\sum_{v\in p} \log |H_v|= O(|p|\log|p|)$.
-This is unfortunately not true, which is why we divide paths into two categories.
-
-\begin{definition}
- For $p\in \pmax(T)$, $p$ is \emph{short} if $|p| < \sqrt{|H_{p}|}/100$, and \emph{tall} otherwise.
-\end{definition}
-
-The following lemma demonstrates that tall paths can ``pay'' for themselves.
-
-\begin{lemma}
-\label{lem:pathBounds} If $p$ is tall, $\sum_{v\in p} \log |H_v| = O(|p| \log |p|)$.
-If $p$ is short, $\sum_{v\in p} \log |H_v| = O(|H_p| \log |H_p|)$.
-\end{lemma}
-\begin{proof}
-For $v \in p$, $|H_v|\leq |H_{p}| + |p|$ (as $v$ is a descendant of $b(p)$ along $p$).
-Hence, $\sum_{v\in p} \log |H_v| \leq \sum_{v\in p} \log(|H_{p}| + |p|)  = |p| \log (|H_{p}| + |p|)$.
- If $p$ is a tall path, then $|p| \log (|H_{p}| + |p|) = O(|p| \log |p|)$. If $p$ is short, then 
- $|p| \log (|H_{p}| + |p|) = O(|p| \log |H_{p}| )$. For short paths, $|p| = O(|H_p|)$.
-\end{proof}
-
-There are some short paths that we can also ``pay" for.  Consider any short path $p$ in the shrub.  We will refer to the 
-\emph{tall support chain} of $p$ as the tall ancestors of $p$ in the shrub which have a path to $p$ which does not use any 
-short path (i.e.\ it is a chain of tall paths adjacent to $p$).
-
-\begin{definition} \label{def:support} A short path $p$ is \emph{supported} if
-at least $|H_p|/100$ vertices $v$ in $H_p$ lie in paths in the tall support chain of $p$.
-\end{definition}
-
-Let $\cL$ be the set of short paths, $\cL'$ be the set of supported short paths, and $\cH$ be the set of tall paths
-given by $\pmax(T)$ (notation which will be used for the remainder of this section).
-
-\begin{claim} \label{clm:above} 
-$\sum_{p \in \cL'} |H_p| \log |H_p| = O(\sum_{q \in \cH} \sum_{v \in q} \log |H_v|)$.
-\end{claim}
-\begin{proof} 
-Pick $p \in \cL'$. 
-As we traverse the tall support chain of $p$, there
-are at least $|H_p|/100$ vertices of $H_p$ that lie in these paths. These are encountered in
-a fixed order. Let $H'_p$ be the first $|H_p|/200$ of these vertices. When $v \in H'_p$
-is encountered, there are $|H_p|/200$ vertices of $H_p$ not yet encountered. Hence,
-$|H_v| \geq |H_p|/200$. Hence, $|H_p|\log |H_p| = O(\sum_{v \in H'_p} \log |H_v|)$.
-Since all the vertices lie in tall paths,
-we can write this as $O(\sum_{q \in \cH} \sum_{v \in H'_p \cap q} \log |H_v|)$.
-Summing over all $p$, the expression is $\sum_{q \in \cH} \sum_{p \in \cL'} \sum_{v \in H'_p \cap q} \log |H_v|$.
-%and $\sum_{p \in \cL'} |H_p|\log |H_p| = O(\sum_{p \in \cL'} \sum_{v \in H'_p} \log |H_v|)$.
-
-Consider any $v \in H'_p$.  Let $S$ be the set of paths $\widetilde{p}\in \cL'$ such that 
-$v\in H'_{\widetilde{p}}$.  We now show $|S|\leq 2$ (i.e.\ it contains at most one path other than $p$).
-First observe that any two paths in $S$ must be unrelated (i.e.\ $S$ is an anti-chain), 
-since paths which have an ancestor-descendant relationship have disjoint tall support chains.  
-However, any vertex $v$ receives exactly one color from each of its two subtrees (in $T$), and therefore $|S|\leq 2$ since any two 
-paths which share descendant leaves in $T$ (i.e.\ their heaps are tracking the same color) must have an ancestor-descendant relationship.
-
-In other words, any $\log |H_v|$ appears at most twice in the above triple summation.
-Hence, we can bound it by $O(\sum_{q \in \cH} \sum_{v \in q} \log |H_v|)$.
-\end{proof}
-
-\begin{corollary}\label{cor:supported}
-  $\sum_{p \in \cL'} \sum_{v \in p} \log |H_v|= O(\sum_{p\in \pmax} |p|\log |p|)$
-\end{corollary}
-\begin{proof}
- By applying \Lem{pathBounds}, \Clm{above}, and then \Lem{pathBounds} again, we have the following.
- \begin{align*}
-  \sum_{p \in \cL'} \sum_{v \in p} \log |H_v| 
-  =  O\left(\sum_{p \in \cL'} |H_p| \log |H_p|\right)
-  = O\left(\sum_{p \in \cH} \sum_{v \in p} \log |H_v|\right)
-  = O\left(\sum_{p\in \pmax} |p|\log |p|\right)
- \end{align*}
-\end{proof}
-
-\newcommand{\loss}{\lambda}
-
-\subsection{Unsupported shrubs}
-
-So far we know that tall paths and supported short paths can be ``paid for.''  We will also be able to pay for 
-unsupported paths, though doing so is much trickier.
-
-Recall that $\cL$ denotes the set of short paths, $\cL'$ the set of supported short paths, and $\cH$ the set of tall paths
-given by $\pmax(T)$. We now describe a partitioning of the unsupported paths into a forest of shrubs. Consider $p \in \cL \setminus \cL'$,
-and traverse the chain of ancestors from $p$. Eventually, we must reach another short path $q$.
-(If not, we have reached the root $r$ of $\pmax(T)$. Hence, $p$ is supported.)
-Insert an edge from $p$ to $q$, so $q$ is the parent of $p$. This construction leads
-to the \emph{shrub forest} of $\cL \setminus \cL'$, denoted $\cF$, where all the roots are supported short paths, 
-and the remaining nodes are the unsupported short paths.
-
-Let $\cU$ be a shrub in $\cF$.  For any node $q$ in the shrub $\cU$, let $TSC(q)$ denote the set of all vertices of $T$ from all paths in the tall support chain of $q$.
-For $q$ in $\cU$ we then define the \emph{loss} to be $\loss(q) = TSC(q)\cap H_q$. 
-Intuitively, if $c$ is a child node of a parent node $p$ in $\cU$, then $\loss(c)$ are the vertices ``lost'' from the heap of $c$ to its tall support chain, by the time one reaches $p$.
-
-The following is a key property of unsupported paths which 
-we will use multiple times. It is worth noting that the proof of this lemma is where the construction of $\pmax(T)$ enters the picture.
-
-\begin{lemma}
-\label{lem:geometric}
- Let $\cU$ denote a connected component of $\cF$, and let $q$ be the child of some node $p$ in $\cU$. 
- Then $|H_p|\geq \frac{3}{2}|H_q|$. 
-\end{lemma}
-\begin{proof} Let $h(q)$ denote the tall path that is a child of $p$ in $\pmax(T)$,
-and an ancestor of $q$. If no such tall path exists, then by construction $p$ is the parent 
-of $q$ in $\pmax(T)$, and the following argument will go through by setting $h(q)=q$.
-
-The chain of ancestors from $q$ to $h(q)$ consists only of tall paths.
-Since $q$ is unsupported, these paths contain at most $|H_q|/100$ vertices of $H_q$. 
-Thus, $|H_{h(q)}| \geq 99|H_q|/100$.
-
-Consider the base
-of $h(q)$, which is a node $w$ in $T$. Let $v$ denote the sibling of $w$ in $T$.
-Their parent is called $u$. Note that both $u$ and $v$ are nodes in the path $p$.
-Now, the decomposition $\pmax(T)$ put $u$ and $v$ in the same path $p$. This
-implies $|H_v| \geq |H_w|$. Since $|H_u| \geq |H_v| + |H_w| - 2$,
-$|H_u| \geq 2|H_w| - 2$. Let $b$ be the base of $p$.
-We have $|H_p| = |H_b| \geq |H_u| - |p| \geq 2|H_w| - |p| - 2$. Since $p$ is a short path, $|p| < \sqrt{|H_p|}/100$.
-Applying this bound, we get $|H_p| \geq (2-\delta)|H_w|$ (for a small constant $\delta > 0$).
-Since $w$ is the base of $h(q)$, $H_w = H_{h(q)}$. We apply the bound $|H_{h(q)}| \geq 99|H_q|/100$
-to get $|H_p| \geq 197|H_q|/100$, implying the lemma.  
-%For the second part, observe that if $p\neq r(\cU)$, then $p$ is unsupported and so 
-%$|\redH_p| \geq 99|H_p|/100$, and therefore the second part follows since $|H_q| \geq |\redH_q|$.
-\end{proof}
-
-\begin{corollary}\label{cor:geometric}
- Let $\cU$ denote a connected component of $\cF$. Let $p$ and $d$ be any two nodes in $\cU$ such that $d$ is a descendant of $p$, and $p$ 
- is not the root of $\cU$.  Then for any $w \in d$ and $v \in \lambda(p)$, $|H_w| = O(|H_v|)$.
-\end{corollary}
-
-\begin{proof} By the previous lemma, $|H_w| = O(|H_p|)$. Note that there are at most
-$|H_p|/100$ nodes in $\lambda(p)$, since these are all in the tall support chain of $p$
-(which is unsupported). Thus, for any $v \in \lambda(p)$, $|H_v| \geq 99|H_p|/100$,
-completing the proof.
-\end{proof}
-
-To handle the unsupported paths, for every shrub $\cU$ in $\cF$ we first prune away what we call buffered sub-shrubs.  In the following subsection, 
-we show how to charge away the cost of these sub-shrubs to their tall support chains.  In \Sec{weight} we then show how to pay for the remaining pruned shrubs, 
-which is the most difficult part of the argument.  Afterwards in \Sec{thm-runtime}, it is then a straightforward task to put everything together to prove \Thm{runtime}.
-
-%We split the remaining argument into two subsections. We first prove \Thm{runtime} from \Lem{cu-root},
-%which involves routine calculations. Then we prove \Lem{cu-root}, where the interesting work happens.
-
-\newcommand{\buff}{\mathcal{B}}
-\subsection{Pruning away buffered sub-shrubs}\label{sec:buffSub}
-Ultimately our goal in this subsection is to prune away subtrees from $\cU$ which can be paid for by charging to their tall support chains (similarly to supported short paths).
-The remaining portion of the shrub will then be handled in a later subsection.
-We now formally define this pruning process, which is quite intuitive if one thinks of $\cU$ as an actual physical shrub.
-
-\begin{definition}\label{def:prune}
-  Let $\cU$ denote a shrub in $\cF$.
-A \emph{pruning} $\cU'$ of $\cU$ is any subgraph defined by some number of rounds of the following iterative process.
-Initially set $\cU' = \cU$.
-In each iteration select some subtree $\cU'_r$ rooted at a node $r\in \cU'$ and set $\cU' = \cU'\setminus \{\cU'_r\}$.
-\end{definition}
-
-The above definition does not specify which subtrees are pruned, which we now formally define.
-
-\begin{definition}\label{def:buff}
- Let $\cU'$ be a pruning of a shrub of $\cU$ of $\cF$.
-Then a rooted subtree $\buff$ of $\cU'$ is called a \emph{buffered sub-shrub}, if it does not contain the root of $\cU'$ (i.e.\ it is a proper subtree), and the following conditions hold:
-  \begin{enumerate}
-  \item $\sum_{p\in \buff} |\loss(p)| > \frac{1}{100} \sum_{p\in \buff} |p|$
-  \item No rooted subtree of $\buff$ satisfies the above inequality (other than $\buff$ itself). 
- \end{enumerate}
-\end{definition}
-
-The following is quite straightforward. By pruning from the leaves towards
-the root, we can remove all buffered sub-shrubs from a given $\cU$.
-
-\begin{lemma} \label{lem:split-buff} Any shrub $\cU$ can be partitioned
-into $\{\cU', \cB_1, \cB_2, \ldots \cB_r\}$ where each $\cB_i$ is buffered.
-Furthermore, no subtree of $\cU'$ is buffered.
-\end{lemma}
-
-\begin{proof} Take any reverse topological order of the nodes in $\cU$
-(so descendants occur before ancestors). Initialize $\cU'$ to $\cU$.
-We process the nodes in order. For node $p$, let $\cU'_p$ denote the rooted
-subtree of $p$ in $\cU'$. We simply check if $\sum_{q \in \cU'_p} |\loss(q)| \geq \sum_{q \in \cU'_p} |q|/100$. If so, we prune off $\cU'_p$, unless $p$ is the root. 
-(Note that $\cU'$ is redefined
-to be the remainder.)
-
-At the end of this process, we have the desired decomposition. Since we process
-in reverse topological order, the minimality condition of \Def{buff}
-automatically holds. Thus, each pruned sub-shrub is buffered and the remaining $\cU'$
-has the desired property.
-\end{proof}
-
-We prove an important lemma. The cost of heap operations corresponding 
-to a buffered sub-shrub can be charged to tall support chains.
-
-\begin{lemma} \label{lem:buff}
- Let $\buff$ be a buffered sub-shrub.  Then $\sum_{p\in \buff}\sum_{w\in p} \log |H_w| = O(\sum_{p\in \buff}\sum_{v\in \loss(p)} \log |H_v| )$
-\end{lemma}
-\begin{proof} The proof is done by a charging argument. For every $p \in \buff$
-and $v \in \loss(p)$, we create 100 tokens. Each of these tokens is labeled $\log |H_v|$,
-and this is called the ``value" of the token. Observe that the sum of token values is, up to a constant
-factor, the bound we wish to show. The set of tokens created for all $v \in \loss(p)$ is denoted
-$T_p$.
-
-There is a grand total of $100 \sum_{p \in \buff} |\loss(p)|$ tokens, which
-(by the buffered property of $\buff$) is strictly more than $\sum_{p \in \buff} |p|$.
-We will place all these tokens on the nodes in $\buff$, using the following process.
-We reverse topologically sort (descendants before ancestors) the nodes in $\buff$.
-For each $p$ in order, we will place tokens in $T_p$ on the nodes of the subtree $\buff_p$.
-Take an arbitrary token in $T_p$, and find the smallest (in the ordering) descendant $q$
-of $p$ that currently has strictly less than $|q|$ tokens. Place the token on $q$,
-and repeat. This is repeated until all descendants $q$ have $|q|$ tokens, or
-all tokens in $T_p$ are used up. When a node $q$ has $|q|$ tokens, we use
-``$q$ is full".
-
-Now for a useful claim. \emph{For non-root $p$, all tokens of $T_p$ are used up.}
-Suppose not. Consider the first such $p$ in the ordering, and look at the situation
-when we finish processing $T_p$. Every descendant $q$ of $p$ must be full. Furthermore,
-all the tokens on $q \in \buff_p$ must have come from some $T_{p'}$, where $p' \in \buff_p$.
-Thus, $\sum_{q \in \buff_p} |q| < |\bigcup_{p' \in \buff_p} T_{p'}|$ $=100\sum_{p' \in \buff_p} |\loss(p')|$. This violates the second condition of buffered shrubs.
-
-Let $r$ be the root of $\buff$. We argue that $T_r$ is \emph{not used up}.
-Suppose the contrary, so $T_r$ is used up. By the previous paragraph,
-all tokens are used up. But this implies that $\sum_{p \in \buff} |p| \geq |\bigcup_{p \in \buff} T_p|$
-$ = 100\sum_{p \in \buff} |\loss(p)|$, violating the buffered condition.
-
-Since $T_r$ is not used up, it must mean that all $p$'s are full, as $r$ is the root and so $T_r$ tokens can be placed on any path which is not full. 
-Consider any $p$, and let $K_p$ be the set of tokens placed at $p$. Each token is labeled $\log |H_v|$,
-for some $v \in \loss(\hat{p})$, where $\hat{p}$ is an ancestor of $p$ or $p$ itself (as descendants do not place tokens on ancestors). By \Cor{geometric},
-for every $w \in p$, $\log |H_w| = O(\log |H_v|)$. Because $p$ is full,
-there are exactly $|p|$ tokens on $p$. Thus, $\sum_{w \in p} \log |H_w|$
-is $O(V_p)$, where $V_p$ is the sum of token values in $K_p$. 
-Thus summing over all $p$ in $\buff$ completes the proof. 
-Specifically, $\sum_{p\in \buff} \sum_{w \in p} \log |H_w| = O(\sum_{p\in \buff} V_p) = O(\sum_{p\in \buff}\sum_{v\in \loss(p)} \log |H_v| )$
-\end{proof}
-
-For any fixed vertex $v$ on a tall path, there are at most two unsupported short paths such that $v$ is contained in $\loss(\cdot)$. Specifically, 
-$v$ is assigned exactly two colors, one from each of its child subtrees in $T$.  Pick one of these two colors, which corresponds to some leaf in $T$.  
-Any two unsupported short paths which are tracking the color of this leaf, must be ancestor-descendant in $\pmax$, however such paths have disjoint tall support chains, 
-and hence only one such path can count $v$ in $\loss(\cdot)$.
-(Note this is the same argument as in \Clm{above}.)  Thus by summing over all tall support chains we have the following.
-
-%Let $p$ and $q$ be two distinct paths in $\cF$.  Note that while it is possible that $TSC(p)\cap TSC(q)\neq \emptyset$, 
-%it holds that $\loss(p)\cap \loss(q) = \emptyset$.  Thus the above lemma implies that the total heap costs due to vertices in buffered sub-shrubs can be charged to those in tall paths.
-
-\begin{corollary}\label{cor:disjointBuff}
- Let $B= \{\buff_1, \ldots, \buff_k\}$ be any collection of disjoint buffered sub-shrubs from $\cF$. Then, 
- $\sum_{i=1}^k \sum_{p\in \buff_i}\sum_{v\in p} \log |H_v| = O(\sum_{p\in \cH}\sum_{v\in p} \log |H_v| )$.
-\end{corollary}
-
-\newcommand{\pruned}{\cU'}
-\subsection{Paying for pruned shrubs: the root is everything} \label{sec:weight}
-
-We focus attention on the remaining shrub left by \Lem{split-buff}: this is 
-a shrub $\cU$ such that no subtree of $\cU$ is buffered. Refer to such
-a shrub as \emph{buffer-pruned}.
-
-% 
-% \begin{definition}\label{def:pruningAShrub}
-%  Let $\cU$ be a shrub in $\cF$. Consider a reverse topological ordering (i.e.\ leaves first) of the nodes in $\cU$, and use it to define a total ordering.  
-%  We compute a pruning of $\cU$ (see \Def{prune}), by iteratively removing all buffered sub-shrubs (see \Def{buff}), where buffered sub-shrubs whose roots are earlier in the total ordering are pruned first.
-We use $\pruned$ to denote the resulting sub-shrub, which we refer to as the \emph{buffer-pruned} sub-shrub of $\cU$.
-% \end{definition}
-% 
-The goal of this section is to prove the following lemma.
-
-\begin{lemma} \label{lem:cu-root} Let $\pruned$ denote a buffer-pruned sub-shrub of a shrub in $\cF$ and let $r$
-be the root of $\pruned$. (i) For any $v \in p$ such that $p \in \pruned$,
-$|H_v| = O(|H_r|)$. (ii) $\sum_{p \in \pruned} |p| = O(|H_r|)$.
-\end{lemma}
-
-\Lem{cu-root} asserts the root $r$ in $\pruned$ pretty much encompasses all sizes and heaps in $\pruned$.
-Observe that part (i) of \Lem{cu-root} is immediate, as it follows directly from \Lem{geometric}. Proving part (ii) requires much more work.
-% It is helpful to spell out how the various heaps evolve. Consider $p \in \cU$,
-% which has children $q_1, q_2, \ldots, q_k$ in $\pmax(T)$, and a leaf vertex $\ell$ in $T$. 
-% The reduced heap $\redH_p$ is the
-% union of $\redH_\ell$ and $\redH_{q_1}, \redH_{q_2}, \ldots, \redH_{q_k}$, with the vertices in $p$
-% removed. Note that some of the $q_i$'s are the corresponding $h(q')$ vertices,
-% for a child $q'$ in $\cU$.
-
-We need the following technical claim. Let $N(\pruned)$ denote the set of nodes in $\pruned$.
-
-\begin{claim} \label{clm:weights} There exists a function $\alpha:N(\pruned) \mapsto \RR^+$
-with the following properties. (We use $\alpha_p$ to denote the function value.)
-Let $A_p = \sum_{q \in \pruned_p} \alpha_q$.
-Then, $A_p - 3\sum_{q \in \pruned_p} |q| \leq |H_p| \leq A_p$.
-\end{claim}
-
-\begin{proof} We define $\alpha$ inductively, from leaves upwards.
-For a leaf $\ell$ in $\pruned$, set $\alpha_\ell = |H_\ell|$. Consider an internal node $p$
-with children $q_1, q_2, \ldots, q_k$. The heap $H_p$ is formed by the union
-of the $H_{q_i}$ heaps, after deleting certain vertices from them, and potentially
-adding some new vertices. 
-The vertices in $p$ and those in the $\lambda(q_i)$ are deleted.
-Define $\alpha_p$ to be the number of new vertices added to $H_p$.
-
-Thus, we have $|H_p| = \sum_{i \leq k} |H_{q_i}| + \alpha_p$
-$- \textrm{\# vertices deleted from $H_{q_i}$'s}$.
-The number of vertices deleted is at most $2(|p|+\sum_{i \leq k} |\lambda(q_i)|)$.
-There is a factor $2$ because each vertex appears in at most $2$ heaps.
-
-It is evident that $|H_p| \leq \sum_{q \in \pruned_p} \alpha_q = A_p$.
-Note that $|H_p| \geq A_p - 2\sum_{q \in \pruned_p, q \neq p} |\lambda(q)|$ $-2\sum_{q \in \pruned_p} |q|$.
-Crucially, since $\pruned$ is buffer-pruned, $\sum_{q \in \pruned_p, q \neq p} |\lambda(q)|
-\leq \sum_{q \in \pruned_p, q \neq p} |q|/100$ $\leq \sum_{q \in \pruned_p} |q|/100$. This completes the proof.
-\end{proof}
-
-We have the main charging claim. This assumes that $A_p$ is increasing exponentially
-as we go up the tree. That property is proven later by induction, using
-the following claim for sub-shrubs.
-
-\begin{claim} \label{clm:charge} Fix any path $p \in \pruned\setminus r$. Suppose for any $q,q' \in \pruned_p$
-where $q$ is a parent of $q'$ in $\pruned_p$, $A_q \geq (5/4)A_{q'}$. Then $\sum_{q \in \pruned_p} |q| \leq A_p/20$.
-\end{claim}
-
-\begin{proof} Since $q$ is an unsupported short path, $|q| < \sqrt{|H_q|}/100 \leq \sqrt{|A_q|}/99$.
-We prove that $\sum_{q \in \pruned_p} \sqrt{|A_q|}/99 \leq A_p/20$
-by a charge redistribution scheme. Assume that each $q \in \pruned_p$
-starts with $\sqrt{A_q}/99$ units of charge. We redistribute this charge over all
-nodes in $\pruned_q$, and then calculate the total charge. For $q \in \pruned_p$, spread
-its charge to all nodes in $\pruned_q$ proportional to $\alpha$ values. In other words,
-give $(\sqrt{A_q}/99)\cdot(\alpha_{q'}/A_q)$ units of charge to each $q' \in \pruned_q$.
-
-After the redistribution, let us compute the charge deposited at $q$. Every ancestor in $\pruned_p$,
-denoted $a_0, a_1, a_2, \ldots, a_k$, contributes to the charge at $q$. The charge is expressed in the following
-equation. We use the assumption that $A_{a_i} \geq (5/4) A_{a_{i-1}}$ and hence $A_{a_i} \geq (5/4)^i A_{a_0} \geq (5/4)^i$, 
-as $a_0$ is an unsupported short path and hence $A_{a_0}\geq 1$. 
-$$ ({\alpha_q}/99)\sum_{a_i} 1/\sqrt{A_{a_i}} \leq ({\alpha_q}/99)\sum_{a_i} (4/5)^{i/2} \leq \alpha_q/20 $$
-The total charge is $\sum_{q \in \pruned_p} \alpha_qt/20 = A_p/20$. 
-\end{proof}
-
-\begin{claim} \label{clm:exp-A} Fix any path $p \in \pruned \setminus r$.
-%For any $q, q' \in \pruned_p$ where $q$ is a parent of $q'$, $A_q \geq (5/4)A_{q'}$.
-Let $q$ be any path in $\pruned_p$, then for any child path $q'$ of $q$ it holds that $A_q \geq (5/4)A_{q'}$
-\end{claim}
-
-\begin{proof} We prove this by induction over the depth of $q$.
-When $q$ is a leaf, this is vacuously true. Consider some internal node $q$,
-with children $q'_1, q'_2, \ldots, q'_k$. Applying the induction
-hypothesis to the subshrubs rooted at $q'_i$, we can invoke \Clm{charge}
-on each subshrub $\pruned_{q'_i}$. Thus, for all $i$,
-$\sum_{\hat{q} \in \pruned_{q'_i}} |\hat{q}| \leq A_{q'_i}/20$. 
-By \Clm{weights}, 
-$|H_{q'_i}| \geq 17A_{q'_i}/20$ and $A_q \geq |H_q|$.
-By \Lem{geometric}, $|H_q| \geq (3/2)|H_{q'_i}|$. We combine these bounds
-to complete the proof.
-\end{proof}
-
-The proof of \Lem{cu-root} just combines all these claims, and essentially
-repeats the argument in the previous claim. We apply \Clm{exp-A} and \Clm{charge}
-to every child of the root $r$. Denoting the children as $q_1, q_2, \ldots$,
-we get that $\sum_{p \in \pruned_{q_i}} |p| \leq A_{q_i}/20$.
-Summing this over the children, $\sum_{p \in \pruned, p \neq r} |p| \leq A_r/20$.
-By \Clm{weights}, $|H_r| \geq 17A_r/20 - 3|r|$. Since $r$ represented a short path, 
-we have the bound $|r| \leq \sqrt{|H_r|}/100$. Thus, $|H_r| = \Omega(A_r)$.
-We combine these bounds to show that $\sum_{p \in \pruned} |p| = O(|H_r|)$.
-
-\subsection{Proving \Thm{runtime}} \label{sec:thm-runtime}
-
-We split the summation into tall, supported short, and unsupported short paths.
-\begin{eqnarray*} 
-\sum_{p \in \pmax(T)} \sum_{v \in p} \log |H_v| & = & \sum_{p \in \cL \setminus \cL'} \sum_{v \in p} \log |H_v| + \sum_{p \in \cL'} \sum_{v \in p} \log |H_v| + \sum_{p \in \cH} \sum_{v \in p} \log |H_v|
-\end{eqnarray*}
-By \Cor{supported} and \Lem{pathBounds}, the second and last term can be bounded by $O(\sum_{p \in \pmax(T)} |p|\log |p|)$.
-The following claim applies the results of the previous two subsections to give the same bound on the first term, 
-thus completing the proof of \Thm{runtime}.
-
-\begin{claim} \label{clm:first} 
-$\sum_{p \in \cL \setminus \cL'} \sum_{v \in p} \log |H_v| = O(\sum_{p\in \pmax} |p|\log |p|)$.%= O(\sum_{p \in \cL'} |H_p| \log |H_p|)$.
-\end{claim}
-
-\begin{proof} Apply \Lem{split-buff} to $\cU$ to get the family of buffered
-sub-shrubs $B(\cU)$ and buffer-pruned $\pruned$.
-% For any component $\cU$ of the shrub forest $\cF$ of $\cL\setminus \cL'$, let $B(\cU)$ denote the collection of buffered sub-shrubs that we 
-% remo that produced the buffer-pruned sub-shrub $\pruned$
-% (i.e.\ these are disjoint sub-shrubs whose union cover $\cU\setminus (\pruned)$). 
-% Thus we can decompose the sum as follows, 
-\begin{align*}
- \sum_{p \in \cL \setminus \cL'} \sum_{v \in p} \log |H_v| 
- &\leq \sum_{\cU\in \cF} \sum_{p \in \cU} \sum_{v \in p} \log |H_v|
- = \sum_{\cU\in \cF}\sum_{p \in \pruned} \sum_{v \in p} \log |H_v|  +  \sum_{\cU\in \cF}\sum_{\buff\in B(\cU)} \sum_{p \in \buff} \sum_{v \in p} \log |H_v|\\
- &= \sum_{\cU\in \cF}\sum_{p \in \pruned} \sum_{v \in p} \log |H_v|  +  O\left(\sum_{p\in \cH}\sum_{v\in p} \log |H_v| \right)\\
- &= \sum_{\cU\in \cF}\sum_{p \in \pruned} \sum_{v \in p} \log |H_v|  +  O\left(\sum_{p\in \pmax} |p|\log |p|\right),
-\end{align*}
-where the last two lines follow from \Cor{disjointBuff} and \Lem{pathBounds}, respectively.
-
-So what remains is to bound the first term.  
-%We have $\sum_{p \in \cL \setminus \cL'} \sum_{v \in p} \log |H_v| \allowbreak \leq \sum_{\cU} \sum_{p \in \cU} \sum_{v \in p} \log |H_v|$.
-By \Lem{cu-root}, $|H_v| = O(|H_r|)$, where $r$ is the root of $\pruned$, and furthermore, $\sum_{p \in \pruned} |p| = O(|H_r|)$. 
-We have $\sum_{p \in \pruned} \sum_{v \in p} \log |H_v| = O((\log |H_r|) \sum_{p \in \pruned} |p|) = O(|H_r|\log |H_r|)$.
-Note that the roots in the shrub forest are supported short paths, and hence when we sum this over all $\cU$ in the shrub forest we get, 
-\[
-\sum_{\cU\in \cF} O\left(|H_{r(\pruned)}|\log |H_{r(\pruned)}|\right)
-=\sum_{p\in \cL'} O(|H_p|\log |H_p|) = O\left(\sum_{p\in \pmax} |p|\log |p|\right),
-\]
-where the last equality follows by \Clm{above} (via the same argument as in \Cor{supported}).
-\end{proof}
-
-\subsection{Our main result}
-\label{sec:implications}
-
-We now show that \Thm{runtime} allows us to upper bound the running time for our join tree and contour 
-tree algorithms in terms of path decompositions.
-
-\begin{theorem}
-%Consider a PL-Morse $f:\MM \to \RR$, where the join tree $\cJ(\MM)$ has maximum degree $3$.
-Let $f:\MM \to \RR$ be the linear interpolant over distinct valued vertices, where the Morse degree of any point is at most $3$.
-There is an algorithm to compute the join tree whose running time is $O(\sum_{p \in \pmax(\cJ)} |p|\log |p| + t\alpha(t) + N)$.
-\end{theorem}
-\begin{proof}
-By \Thm{correct} we know that $\build(\MM)$ correctly outputs $\cJ(\MM)$, and by \Lem{runTimeUpper}
-we know this takes $O(\sum_{v \in \cJ(\MM)} \log |H_v| + t\alpha(t) + N)$ time, where the $H_v$ values are determined as in \Def{initialColoring}.
-Therefore by \Thm{runtime}, $\build(\MM)$ takes $O(\sum_{p \in \pmax(\cJ)} |p|\log |p| + t\alpha(t) + N)$ 
-time to correctly compute $\cJ(\MM)$.
-\end{proof}
-
-This result for join trees easily implies our main result, \Thm{main-alg}, which we now restate and prove.
-
-\begin{theorem}
-%Consider a PL-Morse $f:\MM \to \RR$, where the contour tree $\cC=\cC(\MM)$ has maximum degree $3$. 
-Let $f:\MM \to \RR$ be the linear interpolant over distinct valued vertices, where the Morse degree of any point is at most $3$. 
-There is an algorithm to compute $\cC$ whose running time is $O(\sum_{p \in P(\cC)} |p|\log |p| + t\alpha(t) + N)$,
-where $P(\cC)$ is a specific path decomposition (constructed implicitly by the algorithm).
-\end{theorem}
-\begin{proof}
-First, lets review the various pieces of our algorithm. 
-On a given input simplicial complex, we first break it into extremum dominant pieces using $\rain(\MM)$ (and in $O(|\MM|)$ time by \Thm{rain-time}).  
-Specifically, \Lem{rain-1} proves that the output of $\rain(\MM)$ is a set of extremum dominant pieces, 
-$\MM_1, \dots, \MM_k$, and \Clm{rain-reeb} shows that given the contour trees, $\cC(\MM_1), \dots, \cC(\MM_k)$, 
-the full contour tree, $\cC(\MM)$, can be constructed (in $O(|\MM|)$ time).
-
-Now one of the key observations was that for extremum dominant complexes, computing the contour tree is roughly the same as 
-computing the join tree.  Specifically, \Thm{contour-tree} implies that given $\jc(\MM_i)$ , 
-we can obtain $\cC(\MM_i)$ by simply sticking on the non-dominant minima at their respective splits (which can easily be done in linear time).
-Remark~\ref{rem:joinAndCriticalJoin} implies that $\jc(\MM_i)$ is trivially obtained from the $\cJ(\MM_i)$, and 
- by the above theorem we know $\cJ(\MM_i)$ can be computed in $O(\sum_{p \in \pmax(\cJ(\MM_i))} |p|\log |p| + t_i\alpha(t_i) + N_i)$
-(where $t_i$ and $N_i$ are the number of critical points and faces when restricted to $\MM_i$).
-
-At this point we can now see what the path decomposition referenced in theorem statement should be.  It is just the union of all the maximum 
-path decompositions across the extremum dominant pieces, $\pmax(\cC(\MM)) = \cup_{i=1}^k \pmax(\cJ(\MM_i))$.
-Since all procedures besides computing the join trees take linear time in the size of the input complex, we can therefore 
-compute the contour tree in time
-\[
-O\pth{N+ \sum_{i=1}^k \pth{\sum_{p \in \pmax(\cJ(\MM_i))} |p|\log |p|} + t_i\alpha(t_i) + N_i}
-=\break O\pth{\pth{\sum_{p \in \pmax(\cC(\MM))} |p|\log |p|} + t\alpha(t) + N}
-\]
-\end{proof}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\section{Lower bound by path decomposition}
-\label{sec:lb}
-We first prove a lower bound for join trees, and then generalize to contour trees.
-Note that the form of the theorem statements in this section differ from \Thm{main-lb}, 
-as they are stated directly in terms of path decompositions.  \Thm{main-lb} 
-is an immediate corollary of the final theorem of this section, \Thm{path-main-lb}.
-
-\subsection{Join trees}
-
-We focus on terrains, so $d=2$.
-Consider any path decomposition $P$ of a valid join tree (i.e.\ any rooted binary tree).
-When we say ``compute the join tree", we require the join tree to
-be labeled with the corresponding vertices of the terrain.
-
-\begin{figure}[h]\centering
-    \includegraphics[width=.18\linewidth]{figs/diamond}\hspace{2cm}
-    \includegraphics[width=.3\linewidth]{figs/cones}
-%    \includegraphics[width=.125\linewidth]{figs/topView}\hspace{2cm}
-%    \includegraphics[width=.095\linewidth]{figs/sideView}
-    \caption{Left: angled view of a tent / Right: a parent and child tent put together}
-    \label{fig:diamond}
-\end{figure}
-
-\begin{lemma}
-\label{lem:pathDecomp}
-Fix any path decomposition $P$. There is a family of terrains, $\bF_P$, all with the same triangulation, 
-such that $|\bF_P| = \Pi_{p_i\in P} (|p_i|-1)!$, and no two terrains in $\bF_P$ define the same labeled join tree.
-\end{lemma}
-
-\begin{proof}
- We describe the basic building block of these terrains, which corresponds to a fixed path $p\in P$.  
- Informally, a \emph{tent} is an upside down cone with $m$ triangular faces (see \Fig{diamond}). 
- Construct a slightly tilted cycle 
- of length $m$ with the two antipodal points at heights $1$ and $0$. These are called the anchor and trap of the tent, respectively.
- The remaining $m-2$ vertices are evenly spread around the cycle and heights decrease monotonically when going from the anchor to the trap.
- Next, create an apex vertex at some appropriately large height, and add an edge to each vertex in the cycle. 
- Note that in order to obtain a terrain, a cycle here means the vertices lie on a circle 
- but are connected with straight line segments rather than by the curved edges of the circle.
-
- Now we describe how to attach two different tents.  In this process, we glue the base of a scaled down ``child'' tent on to a triangular cone face of the larger  ``parent'' tent (see \Fig{diamond}).
- Specifically, the anchor of the child tent is attached directly to a face of the parent tent at some height $h$. The remainder of the base of the child cone is then 
- extended down (at a slight angle) until it hits the face of the parent.
- Note that this gluing process introduces some non-triangular faces, and so these must be triangulated in order to maintain a proper triangulated terrain.
-
- The full terrain is obtained by repeatedly gluing tents. For each path $p_i\in P$, we create a tent of size $|p_i|+1$.
- The two faces adjacent to the anchor are always empty, and the remaining faces are for gluing on other tents.
- (Note that tents have size $|p_i|+1$ since $|p_i|-1$ faces represent the joins of $p_i$, the apex represents the leaf, 
- and we need two empty faces next to the anchor.)
- Now we glue together tents of different paths in the same way the paths are connected in the shrub $\pathTree$ (see \Sec{shrubs}).  
- Specially, for two paths $p,q\in P$ where 
- $p$ is the parent of $q$ in $\pathTree$, we glue $q$ onto a face of the tent for $p$ as described above.
- (Naturally for this construction to work, tents for a given path will be scaled down relative to the size of the tent of their parent).
- By varying the heights of the gluing, we get the family of terrains.
-
- Observe now that the only saddle points in this construction are the anchor points.  Moreover, the only maxima are the apexes of the 
- tents.  We create a global boundary minimum by setting the vertices at the base of the tent representing the root of $\pathTree$ all to the same height (and there are no other minima).  
- Therefore, the saddles on a given tent will appear contiguously on a root to leaf path in the join tree of the terrain, where the leaf corresponds to the maximum of the tent 
- (since all these saddles have a direct line of sight to this apex).  In particular, 
- this implies that, regardless of the heights assigned to the anchors, the join tree has a path decomposition whose corresponding shrub is equivalent to $\pathTree$.
-
- There is a valid instance of this described construction for any relative ordering of the heights of the saddles on a given tent.  
- In particular, there are $(|p_i|-1)!$ possible orderings of the heights of the saddles on the tent for $p_i$, and hence $\Pi_{p_i\in P} (|p_i|-1)!$ possible 
- terrains we can build.  Each one of these functions will result in a different (labeled) join tree. All saddles on a given tent will appear in sorted order in the join tree.
- So, any permutation of the heights on a given tent corresponds to a permutation of the vertices along a path in $P$.
-\end{proof}
-
-% Two path decompositions $P_1$ and $P_2$ (of potentially different complexes and/or height functions) are equivalent 
-% if: there is a 1-1 correspondence between the sizes of the constituent paths, and the shrubs are isomorphic.
-
-\begin{lemma}
-\label{lem:cost}
-For all $\MM \in \bF_P$, the total number of heap operations performed by $\build(\MM)$ is $O(\sum_{p\in P} |p|\log|p|)$.
-\end{lemma}
-\begin{proof}
-%For any choice of $\MM \in \bF$, there is a path decomposition $\overline{P}(T)$ of the join tree equivalent to $\mathcal{P}$.
-%Specifically, this is the decomposition where each path $p\in \overline{P}(T)$ consists exactly of the set of critical points (except for its anchor) from a given tent. 
-The primary ``non-determinism" of the algorithm is the initial painting constructed by $\init(\MM)$. 
-We show that regardless of how paint spilling is done, the number of heap operations is bounded as above.
-
- Consider an arbitrary order of the initial paint spilling over the surface.
- Consider any join on a face of some tent, which is the anchor point of some connecting child tent.  
- The join has two up-stars, each of which has exactly one edge. Each edge connects to a maximum
- and must be colored by that maximum.
- Hence, the two colors touching this join (according to \Def{initialColoring})
- are the colors of the apexes of the child and parent tent.  
-% 
-% Specifically, one of these edges connects directly to the apex of the parent tent and so the join must get the color of this maxima.
-% The other edge connects directly to the apex of the child tent (this is why we left the faces adjacent to an anchor empty) and 
-% so the second color is that of the maxima of the child tent.
-
- Take any join $v$, with two children $w_1$ and $w_2$. Suppose $w_1$ and $v$ belong to the same path
- in the decomposition. The key is that any color from a maximum in the subtree at $w_2$ cannot touch
- any ancestor of $v$. This subtree is exactly the join tree of the child tent attached at $v$.
- The base of this tent is completely contained in a face of the parent tent. So all colors
- from the child ``drain off" to the base of the parent, and do not touch any joins on the parent tent.
-
- Hence, $|H_v|$ is at most the size of the path in $P$ containing $v$. By \Lem{runTimeUpper}, the total number of heap operations
- is at most $\sum_v \log |H_v|$, completing the proof.
-
-% 
-% So let $c$ be the color of some maxima, $m$.  The above proven property of the coloring implies that in the join tree, the vertices for 
-% which $c$ appears are exactly the vertices of the path in  $\overline{P}(T)$ corresponding $m$, as well as 
-% the vertex to which the root of this path connects to (i.e.\ the anchor of the tent of $m$).
-% Now we show this implies $\overline{P}(T)$ is a maximum path decomposition.
-% To see this, consider a vertex $v\in T$, and let $p$ be the path in $\overline{P}(T)$ which contain $v$. 
-% Now $v$ has two children (if it is a join).  Let $c_1$ be the child which is in $p$, and $c_2$ the child which is the root of some adjacent path $p'$.  
-% In order for $\overline{P}(T)$ to be the maximum path decomposition, we must show $|H_{c_1}|\geq |H_{c_2}|$.
-% %(Note, to ensure a unique maximum path decomposition, one can break ties consistently by assigning some arbitrary but fixed ordering to the vertices.)
-% %Now $v$ gets assigned two colors, one 
-% %of which we know is the color assigned to all vertices in $p$ and the other is the one assigned to all the vertices of $p'$. 
-% Observe that none of the colors in the subtree of $c_2$ appear below $v$.  Hence the heap at $c_2$ consists only of $v$.  
-% On the other hand the heap of $c_1$ contains $v$, any other vertex of $p$ below $v$, and the vertex to which the root of $p$ is adjacent to, 
-% and hence $|H_{c_1}|\geq |H_{c_2}|$.
-\end{proof}
-
-The following is the equivalent of \Thm{main-lb} for join trees, and immediately follows from
-the previous lemmas.  Note that the phrasing of the two theorems differ, but can be connected by observing that for any value of $\thmConst$ from \Thm{main-lb}, 
-there is a path decomposition $P$ of some tree $T$ such that $\thmConst= \Theta(\sum_{p \in P} |p|\log |p|)$.
-
-%Call a cost $C$ join tree valid if $C = cost_{\jc}(f)$ for some $f$ defined over a given complex $\MM$.
-
-\begin{theorem}\label{thm:joinLB}
-Consider a rooted tree $T$ and an arbitrary path decomposition $P$ of $T$.
-There is a family $\bF_P$ of terrains such that any algebraic decision
-tree computing the join tree\footnote{Note that for the referenced family of terrains, the join tree and contour tree are equivalent} 
-(on $\bF_P$) requires $\Omega(\sum_{p \in P} |p|\log |p|)$ time.
-Furthermore, our algorithm makes $O(\sum_{p \in P} |p|\log |p|)$ comparisons on all these instances.
-%
-%Consider the set of functions $\bF_C$ that all have the same join tree valid cost $C$.  
-%Any height function decision tree that correctly computes
-%the join tree on all of $\bF_C$, has an average case (uniform over $\bF_C$) running time of $\Omega(C)$.
-\end{theorem}
-
-\begin{proof}
-The proof is a basic entropy argument. Any algebraic decision tree that is correct on all of $\bF_P$
-must distinguish all inputs in this family. By Stirling's approximation, the depth of this tree is $\Omega(\sum_{p_i\in P} |p_i|\log|p_i|)$.
-\Lem{cost} completes the proof.
-%
-%
-% If $C$ is a join tree valid cost then there some $f$ such that $\cost_{\jc}(f)=\sum_{p\in P(f)} |p|\log|p|$.
-% By the above lemmas, there is then a set $\bF$ of $\Pi_{p_i\in P} |p_i|!$ functions, such that for any $g\in \bF$, $C= \cost(g)$.  
-% (Specifically, each sorted ordering of the relative height values of the saddles on each tent yields a function in this class.)
-% Hence $\bF\subseteq \bF_C$, and so any height function decision tree for the set $\bF_C$ must distinguish between these possibilities, and therefore must 
-% have $\Pi_{p_i\in P} |p_i|!$ leaves.  Hence by Stirling's approximation, the average depth in this tree is $\Omega(\sum_{p_i\in P} |p_i|\log|p_i|)$.
-\end{proof}
-
-\subsection{Contour trees}
-We first generalize previous terms to the case of contour trees.
-In this section $T$ will denote an arbitrary contour tree with every internal vertex of degree $3$.
-%and the height function defined over the vertex set is such that every internal vertex in either a split or join 
-%(i.e.\ at most $2$ neighbors lie above or below it).
-
-%We need to generalize the definition of path decompositions, which previously were only defined for rooted trees.
-%Again we define a path decomposition as a partition of the vertices into 
-%disjoint leaf paths (i.e.\ each path is monotonic in height and contains a leaf). 
-For simplicity we now restrict our attention to path decompositions 
-consistent with the raining procedure described in \Sec{rain} 
-(more general decompositions can work, but it is not needed for our purposes).
-
-\begin{definition}
-\label{def:path2} A path decomposition, $P(T)$, is called \emph{rain consistent} if its paths can be obtained as follows.
-Perform an downward BFS from an arbitrary maximum $v$ in $T$, and mark all vertices encountered.  
-Now recursively run a directional BFS from all vertices adjacent to the current marked set.  
-Specifically, for each BFS run, make it an downward BFS if it is at an odd height in the recursion tree and upward otherwise.
-
-This procedure partitions the vertex set into disjoint rooted subtrees of $T$, based on which BFS marked a vertex.  
-For each such subtree, now take any partition of the vertices into leaf paths.\footnote{Note that the subtree  
-of the initial vertex is rooted at a maximum.  For simplicity we require that the path this vertex belongs to 
-also contains a minimum.}
-\end{definition}
-
-The following is analogous to \Lem{pathDecomp}, and in particular uses it as a subroutine.
-
-\begin{lemma}
-\label{lem:pathDecomp2}
-Let $P$ be any rain consistent path decomposition of some contour tree. 
-There is a family of terrains, $\bF_P$, all with the same triangulation, 
-such that the size of $\bF_P$ is $\Pi_{p_i\in P} (|p_i|-1)!$, and no two terrains in $\bF_P$ define the same contour tree.
-\end{lemma}
-\begin{proof}
- As $P$ is rain consistent, the paths can be partitioned into sets $P_1, \dots, P_k$, where $P_i$ is the set of all paths with 
- vertices from a given BFS, as described in \Def{path2}.  Specifically, let $T_i$ be the subtree of $T$ corresponding to $P_i$ and 
- let $r_i$ be the root vertex of this subtree.  
- Note that the $P_i$ sets naturally define a tree where $P_i$ is the parent of $P_j$ if $r_i$ (i.e.\ the root of $T_i$) is adjacent to a vertex in $P_j$.
-
- As the set $P_i$ is a path decomposition of a rooted binary tree $T_i$, the terrain construction of \Lem{pathDecomp} for $P_i$ is well defined.  
- Actually the only difference is that here the rooted tree is not a full binary tree, and so some of the (non-achor adjacent) faces of the constructed tents will be blank.  
- Specifically, these blank faces correspond to the adjacent children of $P_i$, and they tell us how to connect the terrains of the different $P_i$'s.
-
- So for each $P_i$ construct a terrain as described in \Lem{pathDecomp}.  Now each $T_i$ is (roughly speaking) a join or a split tree, depending on 
- whether the BFS which produced it was an upward or downward BFS, respectively.  As the construction in \Lem{pathDecomp} was for join trees, each terrain 
- we constructed for a $P_i$ which came from a split tree, must be flipped upside down.
- Now we must described how to glue the terrains together.
-
- \begin{figure}[h]\centering
-    \includegraphics[width=.28\linewidth]{figs/cones2}
-    \caption{A child tent attached to a parent tent with opposite orientation.}
-    \label{fig:diamond2}
-\end{figure}
-
- By construction, the tents corresponding to the paths in $P_i$ are connected 
- into a tree structure (i.e.\ corresponding to the shrub of $P_i$).  Therefore the bottoms of all these tents are covered except for the one corresponding 
- to the path containing the root $r_i$.  
- If $r_i$ corresponds to the initial maximum that the rain consistent path decomposition was defined from, then this will be flat and corresponds to the global outer face.
- Otherwise, $P_i$ has some parent $P_j$ in which case we connect the bottom of the tent for $r_i$ to a free face of a tent in the construction for $P_j$, specifically, the face corresponding to the 
- vertex in $T$ which $r_i$ is adjacent to.
-This gluing is done in the same manner as in \Lem{pathDecomp}, attaching the anchor for the root of $P_i$ directly the corresponding face of $P_j$, except that now $P_i$ and $P_j$ have opposite orientations.
-See \Fig{diamond2}.
-
- Just as in \Lem{pathDecomp} we now have one fixed terrain structure, such that each different relative ordering of the heights of the join and split vertices 
- on each tent produces a surface with a distinct contour tree.  The specific bound on the size of $\bF_P$, defining these distinct contour trees, follows by applying the 
- bound from \Lem{pathDecomp} to each $P_i$.
-\end{proof}
-
-\begin{lemma}
-For all $\MM \in \bF_P$, the number of heap operations is $\Theta(\sum_{p\in P} |p|\log |p|)$
-\end{lemma}
-\begin{proof}
- This lemma follows immediately from \Lem{cost}.  The heap operations can be partitioned into the
- operations performed in each $P_i$.
- Apply \Lem{cost} to each of the $P_i$ separately and take the sum.
-\end{proof}
-
-We now restate \Thm{main-lb}, which follows immediately from an entropy argument, analogous to \Thm{joinLB}.
-Again, as discussed above for \Thm{joinLB}, the phrasing differs from \Thm{main-lb}.
-
-\begin{theorem}\label{thm:path-main-lb}
-Consider any rain consistent path decomposition $P$. There exists a family $\bF_P$ of terrains ($d=2$) with the following properties.
-Any contour tree algorithm makes $\Omega(\sum_{p \in P} |p|\log |p|)$ comparisons in the worst case over $\bF_P$.
-Furthermore, for any terrain in $\bF_P$, our algorithm makes $O(\sum_{p \in P} |p|\log |p|)$ comparisons.
-\end{theorem}
-
-\begin{Remark}
- Note that for the terrains described in this section, the number of critical points is within a constant factor of 
- the total number of vertices.  In particular, for this family of terrains, all previous algorithms required $\Omega(n\log n)$ time.
-\end{Remark}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\myparagraph{Acknowledgements.}
-We thank Hsien-Chih Chang, Jeff Erickson, and Yusu Wang for numerous useful discussions. This work is
-supported by the Laboratory Directed Research and Development
-(LDRD) program of Sandia National Laboratories. Sandia National
-Laboratories is a multi-program laboratory managed and operated
-by Sandia Corporation, a wholly owned subsidiary of Lockheed
-Martin Corporation, for the U.S. Department of Energy's National
-Nuclear Security Administration under contract DE-AC04-94AL85000.
-Work on this paper by Benjamin Raichel was also partially supported
-by NSF CRII Award 1566137.
+The paper~\cite{rs-17} presents a new algorithm to compute contour trees, whose running time depends on the shape of the tree, and in particular avoids the initial global sort of all the critical points.
+The algorithm has two parts.
+The first part
+is a new algorithm to compute join trees, again whose running time depends on the shape of the tree.
+The second part reduces computing the contour tree to join tree computations, by cutting the input simplicial complex into disjoint pieces.
+For each of these pieces, the contour tree and join tree are the same.
+The paper claims this decomposition can be performed in linear time.
+As pointed out by Rote, the proof of this claim is incorrect~\cite{rote}. As a result, the main theorem (Theorem 1.1) does not hold.
+
+The problems appear in Section 6 of~\cite{rs-17}. 
+For a maximum $x\in \MM$, as defined in Section 6, $\wet(x,\MM)$ is the set of points $y \in \MM$ such that there is a non-ascending path from $x$ to $y$.
+A point $z$ is at the \emph{interface} of $\wet(x,\MM)$ if every neighborhood of $z$ has non-trivial intersection with $\wet(x,\MM)$ (i.e.\ the intersection is neither empty nor the entire neighborhood).
+The bug is that the paper does not describe precisely how to compute $\wet(x,\MM)$.
+(This appears in the first step of the procedure {\tt rain}$(x,\NN)$.) The natural approach is to perform a descending BFS or DFS from $x$.
+Suppose we reach a critical vertex $v$ which is a join.
+We wet an edge in one of its up-stars when we reached it, and its other up-star is entirely dry since it is a join.
+As our procedure wets the interior of faces, at this point we naturally wish to walk around the interface contour adjacent to this join, and then rain down (i.e.
+continue the DFS) from the contour.
+%
+For two distinct join vertices, with a wet edge in one up-star, the edge sets crossed by their corresponding dry contours will be disjoint.  
+%
+Thus the cost of walking along these contours for all joins can be charged to the size of the interface which is linear in the size of $\wet(x,\MM)$. 
+%
+The issue, however, is that while we know locally whether a given vertex $v$ is a critical vertex, we do not know whether it is a join.
+%
+So suppose $v$ is critical with one dry up-star, but that it is not a join.
+Then one can still walk around the contour from the dry up-star and rain down.
+However, if we do this for all such $v$ it may no longer be true that the edge sets crossed by the contours are disjoint.
+Thus depending on the order in which vertices are handled, this could lead to edges being traversed a super-constant number of times.
+
+In summary, while the actual cutting can be performed in linear time, it is not clear how long it takes to determine where to make the cuts.
+Specifically, it is not clear how long it takes to determine the interface of the wet complex, namely the first step in the {\tt rain} procedure of section Section 6, and thus it is unclear whether Theorem 6.7 holds.
+At the most basic level, the issue boils down to efficiently determining which critical points are joins. There does not exist an efficient procedure
+for finding all joins.
+
+We note that that the algorithm and analysis of computing the join tree still hold (Sections 4 and 8 of~\cite{rs-17}) .
 
@@
 
-\pagebreak
-\appendix
-
-\section{ \Sec{algo} Algorithm Example }\label{sec:example}
-
-Here we provide a simple example of the execution of the main algorithm from \Sec{algo}.  
-An example input mesh is shown in \Fig{mesh1}, followed by the different steps of the join construction in \Fig{mesh2}, 
-and finally a table showing the states of the data structures and how they are modified in \Fig{exec}.
-
-Note that in \Fig{exec} a ``stage'' refers to any time at which one of the data structures is modified.  
-For simplicity, the trivial stages where the maxima are removed from their corresponding heaps are not shown. 
-(Alternatively the algorithm in \Sec{algo} could be modified to exclude the maxima from the heaps, 
-but we wanted to keep the algorithm description simple.)
-
- \begin{figure}[h]\centering
-    \includegraphics[width=.3\linewidth,page=1]{figs/mesh1}
-    \caption{A piece of a potentially larger mesh, with maxima $m_1,m_2,m_3$, saddles $s_1,s_2$, and regular vertices $r_1,r_2,r_3,r_4$.
-             An initial painting is shown where green `g' is spilled from $m_1$, then red `r' from $m_2$, and finally blue `b' from $m_3$.
-             Arrows are directed from higher to lower function value.}
-    \label{fig:mesh1}
-\end{figure} 
-%
-\begin{figure}[h]
-   \begin{subfigure}{.33\textwidth}
-     \centering
-     \includegraphics[width=.7\linewidth,page=1]{figs/mesh2}
-     \caption{}
-     \label{fig:tree1}
-   \end{subfigure}%
-   \begin{subfigure}{.33\textwidth}
-     \centering
-     \includegraphics[width=.7\linewidth,page=2]{figs/mesh2}
-     \caption{}
-     \label{fig:tree2}
-   \end{subfigure}
-   \begin{subfigure}{.33\textwidth}
-     \centering
-     \includegraphics[width=.7\linewidth,page=3]{figs/mesh2}
-     \caption{}
-     \label{fig:tree3}
-   \end{subfigure}
- \caption{Snapshots of the join tree.}
- \label{fig:mesh2}
-\end{figure}
-%
-\begin{figure*}[h]%
-    \centerline{%
-       \begin{tabular}{|l|c|c|c|c|c|c|c|c|}%
-         \hline
-         Stage%
-         &%
-           $att(r)$
-         &%
-           $att(g)$
-         &%
-           $att(b)$
-         &%
-           $T(r)$
-         &%
-           $T(g)$
-         &%
-           $T(b)$
-         &%
-           $K$
-         &%
-          Current tree\\
-         \hline%
-         \hline%
-         Initial
-         &%
-           $m_2$
-         &%
-           $m_1$
-         &%
-           $m_3$
-         &%
-           $\{s_2\}$
-         &%
-           $\{s_1\}$
-         &%
-           $\{s_1,s_2\}$
-         &%
-           $()$
-         &%
-           \Fig{tree1}
-         \\%
-         \hline%
-         % 
-         1%
-         &%
-           $m_2$          
-         &%
-           $m_1$
-         &%
-           $m_3$
-         &%
-           $\{s_2\}$
-         &%
-           $\{s_1\}$
-         &%
-           $\{s_1,s_2\}$
-         &%
-           $(s_1)$
-         &%
-           \Fig{tree1}
-         \\%
-         \hline%
-         %
-         2
-         &%
-           $m_2$
-         &%
-           $m_1$
-         &%
-           $m_3$
-         &%
-           $\{s_2\}$
-         &%
-           $\{s_1\}$
-         &%
-           $\{s_1,s_2\}$
-         &%
-           $(s_1,s_2)$
-         &%
-           \Fig{tree1}
-         \\%
-         \hline%
-         %
-         3
-         &%
-           NA
-         &%
-           $m_1$
-         &%
-           $s_2$
-         &%
-           NA
-         &%
-           $\{s_1\}$
-         &%
-           $\{s_1\}$
-         &%
-           $(s_1)$
-         &%
-           \Fig{tree2}
-         \\%
-         \hline%
-         %
-         4
-         &%
-           NA
-         &%
-           NA
-         &%
-           $s_1$
-         &%
-           NA
-         &%
-           NA
-         &%
-           $\{\}$
-         &%
-           $()$
-         &%
-           \Fig{tree3}
-         \\%
-         \hline
-       \end{tabular}%
-    }
-    \caption{Stages of the algorithm run on the initial painting shown in \Fig{mesh1}.
-    The above table assumes that in the stack update, $s_1$ is the first arbitrary unprocessed critical point selected.}
-    \label{fig:exec}
-\end{figure*}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%
-%
-%end of writup
-%
-%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
 \end{document}

### 2021-10-19 erratum/updated/erratum.tex
updates to erratum

@@
 
-In summary, while the actual cutting can be performed in linear time, it is not clear how long it takes to determine where to make the cuts.
+\begin{figure}[h]\centering
+    \includegraphics[width=0.45\linewidth]{../figs/volcano}
+    \hspace{0.5in}
+    \includegraphics[width=0.45\linewidth]{../figs/volcano-top}
+    \caption{The ``volcano" example: the left figure gives a side view and the right figure
+    gives a top view. The dark lines gives the outline of the mountain and the crater rim. The light
+    lines on the right show the triangles on the surface.}
+    \label{fig:volcano}
+\end{figure}
+
+Let us give a concrete example. Consider the surface showing in \Fig{volcano}. It is a 2D-manifold
+in 3D, which we think of as a ``volcano" with a crater. There are two maxima $x, x'$ on the rim of the
+crater, and a critical point $y$ that is the minimum of the rim. Observer that $y$ 
+has Morse up-degree $2$, but has an up-degree of one. Meaning, there is only one contour just above $y$, but locally, it looks like
+there are two contours. So $y$ is critical point, but we cannot locally determine it is \emph{not} a join. (The point $y$
+will appear in the contour tree as a vertex.) One can think of $y$ as a ``false saddle". Consider the top view given in the right of \Fig{volcano}. The rim 
+also a true saddle point, $z$.
+
+Consider raining from the maximum $x$. Note that the contour above $y$ is not part of the interface, since all sides/neighbors of $y$ are reachable
+from $y$ by downward monotone paths from $x$. On the other hand, the only way to \emph{locally} determine this seems to actually construct
+the contour (which would cut all the triangles show in the figure). More precisely, consider the raining procedure from $x$. The rain
+will reach both critical points $y$ and $z$. Yet, to determine that $y$ is not on interface, we would need to first cut the contour
+at $z$ (that encloses the maximum $x'$) and send rain from that contour. This process would ensure that rain reaches $y$ from ``both" sides,
+confirming that $y$ is not at the interface.
+
+This example suggests that we need to process critical points that are reached from $x$ in a specific order. Unfortunately,
+our current algorithm processes critical points in an arbitrary order, to achieve linear running time. In the above example,
+if $y$ is processed before $z$, then we would cut a contour only to realize that it is not in the interface. It appears
+that our current procedure would only work if Morse up-degrees and up-degrees were identical.
+
+At a higher level, this example suggests that it might not be possible to perform contour surgery in linear time,
+since we need to process critical points in order. Such a processing will likely require a priority queue, which
+will introduce further overhead. It is possible that this overhead will be subsumed by the subsequent join/merge
+tree construction. Moreover, it is quite believable that any process that constructs contour trees
+would need to sort (the heights of) $x, y, z$. We think that an optimal algorithm is still possible
+with the current ideas. But as it stands, our current algorithm and proof are incorrect.
+
+% 
+% 
+% 
+% In summary, while the actual cutting can be performed in linear time, it is not clear how long it takes to determine where to make the cuts.
 Specifically, it is not clear how long it takes to determine the interface of the wet complex, namely the first step in the {\tt rain} procedure of section Section 6, and thus it is unclear whether Theorem 6.7 holds.
-At the most basic level, the issue boils down to efficiently determining which critical points are joins. There does not exist an efficient procedure
-for finding all joins.
+% At the most basic level, the issue boils down to efficiently determining which critical points are joins. There does not exist an efficient procedure
+% for finding all joins.
 

### 2021-10-22 erratum/updated/erratum.tex
minor updates

@@
 
-We note that that the algorithm and analysis of computing the join tree still hold (Sections 4 and 8 of~\cite{rs-17}) .
+We note that that the algorithm and analysis of computing the join tree still hold (Sections 4 and 8 of~\cite{rs-17}). 
+Unfortunately, there is no explicit theorem that states the running time bound of computing the join tree. The current
+proof merges various arguments directly into an argument for the final theorem. Theorem 8.1 bounds the running time 
+of the join tree construction, and is still correct. Section 7 (which is not affected by the errors above)
+proves that computing the contour tree of a minimum dominant complex is equivalent to computing
+the join tree. Hence, the main theorem (Theorem 1.1)
+is true if $f$ is a minimum (or maximum) dominant complex.
 
